### 2023-04-06 main.tex

\documentclass{article}

\usepackage[english]{babel}
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

\usepackage{authblk}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{makecell}

\title{Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2}
\author[1]{Shing H. Zhan}
\author[2,5]{Katherine Eaton}
\author[3,5]{Anastasia Ignatieva}
\author[1,5]{Benjamin Jeffery}
\author[1,5]{Duncan S. Palmer}
\author[1,5]{Yan Wong}
\author[2]{Carmen L. Murall}
\author[4]{Sarah P. Otto}
\author[1,6]{Jerome Kelleher}
\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, Oxford, United Kingdom}
\affil[2]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
\affil[3]{Department of Statistics, University of Oxford, Oxford, United Kingdom}
\affil[4]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver, British Columbia, Canada}
\affil[5]{Preliminary alphabetic ordering}
\affil[6]{Correspondence. E-mail: jerome.kelleher@bdi.ox.ac.uk}

\begin{document}
\maketitle

\begin{abstract}
XX
\end{abstract}

\section{Introduction}

Recombination via template switching is a common feature in the evolution of coronaviruses (CoV) \cite{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2 \cite{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing together mutations carried by different lineages, recombination plays an important role in generating genomic diversity, with recombinant lineages associated with CoV adaptation to new host species and with the production of more immune evasive variants \cite{Graham2010-xe,De_Klerk2022-tt}.

Recombination was challenging to detect in SARS-CoV-2 early in the pandemic due to gradual accumulation of genetic variation. In late 2020, the appearance and spread of variants of concern (VoC) harbouring multiple characteristic mutations increased the detectability of recombination. By searching for genomes containing segments from different VoC, multiple recombinant lineages have been detected, for example, between Alpha and pre-VoC \cite{Jackson2021-ik}, Alpha and Epsilon \cite{Wertheim2022-hj}, and Alpha and Delta \cite{Sekizuka2022-xz}). More recently, the high prevalence of Omicron, with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has accelerated the rate of coinfection and the potential for recombination \cite{Bal2022-hq}. Recombinant lineages that have sufficient evidence of onward transmission (more than five descendants) and monophyly (> 70\% bootstrap support) are eligible for a Pango lineage designation \cite{Rambaut2020-dw}, starting with the letter X (“Pango Lineage Nomenclature: Provisional Rules for Naming Recombinant Lineages” 2021).

Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that combine specific subsets of mutations from different lineages (e.g., \cite{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}). This approach only identifies those inter-lineage recombination events resulting in the patterns investigated. Others have inferred recombinant genealogies to disentangle recombination from recurrent mutation, without having to pre-specify characteristic mutations for each lineage \cite{Ignatieva2021-rg}. More recently, Turakhia et al. \cite{Turakhia2022-it} has presented a phylogeny-based approach ("RIPPLES") to identify putative recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find candidate recombinants, RIPPLES first builds a phylogeny without assuming recombination and then scans for long branches (i.e., carrying many mutations) and then determines if these candidates would be better explained by recombination. Specifically, RIPPLES exhaustively breaks each candidate sequence into segments and attempts to find more parsimonious placements for each segment on the phylogeny. If such placements are found, the sequence is identified as a putative recombinant. Although it enables rapid searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that accounts for only mutations, treats recombinants \textit{post hoc}, and is an incomplete representation of the reticulate evolutionary history of SARS-CoV-2. As noted by the authors, a \textit{post hoc} treatment of recombination is possible when recombinant lineages are rare and leave few descendants, but this is becoming increasingly untenable with the proliferation of recombinant lineages (e.g., more than half of the sequences sampled in February, 2023 are from XBB and its descendants; \cite{Chen2022-pz}.

Inferring evolutionary trees while accounting for both mutation and recombination is a long-standing challenge. This challenge is particularly acute for SARS-CoV-2, given the large amount of data that has been shared, with over 15 million sequences in the GISAID database \cite{Shu2017-hp}. It is well known that recombination distorts phylogenies \cite{Schierup2000-fg} and affects the results of downstream analysis (e.g., inference of selection; \cite{Anisimova2003-vr}). However, standard phylogenetic methods do not account for recombination (e.g., \cite{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}). Alternatively, genetic ancestries of recombining lineages can be represented by ancestral recombination graphs (ARG; \cite{Griffiths1981-lw,Hudson1983-if}, which  capture network-like phylogenetic relationships \cite{Gusfield2014-qw}.

Approaches to infer ARGs now exist that can scale to tens of thousands of human genomes and beyond \cite{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}. Recently, Wohns et al. \cite{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands of modern and ancient human genomes, based on over 90 million variable sites. To achieve this, Wohns et al. \cite{Wohns2022-th} utilised several computational advances in ARG inference: (1) the "succinct tree sequence" data structure that can compactly store large ARGs  \cite{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM) implementation of the Li and Stephens (LS) model (Li and Stephens 2003) to infer copying paths between genomes \cite{Kelleher2019-ba}; and (3) efficient, open-source software tools \cite{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

Here, we present a preliminary version of a novel scalable methodology (“sc2ts”, pronounced “scoots”) that builds on the advances above and additionally takes advantage of known collection dates to infer ARGs from millions of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences, we demonstrate that sc2ts can be used to simultaneously (1) detect putative SARS-CoV-2 recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages; (3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record of evolutionary information about SARS-CoV-2 that can be efficiently accessed using "tskit" (tree sequence toolkit) software library \cite{Kelleher2018-xc}. Sc2ts is a promising method to automate detection of SARS-CoV-2 recombinants and to jointly model recombinational and mutational processes shaping the genomic diversity of SARS-CoV-2 at large scale.

\section{Methods}

\subsection{Data preprocessing}

\subsubsection{Sequence-level preprocessing}

The findings of this study are based on sequences and metadata available on GISAID (\href{https://gisaid.org/}) up to Aug 22, 2022 and accessible at doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed if they (1) had ambiguous collection dates; (2) were collected before January 01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host. Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank: MN908947.3) using Nextclade v2.3.0 \cite{Aksamentov2021-hj} (dataset tag 2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG inference, sequences were excluded if they had a “bad” quality control status in any of the four Nextclade columns ("qc.missingData.status”, “qc.mixedSites.status”, “qc.frameShifts.status”, and “qc.stopCodons.status”). These excluded genomes are enriched for low-coverage, contaminated, or misassembled sequences.

\subsubsection{Site-level preprocessing}

Ambiguous nucleotide letters (i.e., not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N. Problematic bases in the alignments, which had two or more ambiguous letters or gaps within a distance of seven bases, were masked as missing data following the protocol in \href{https://github.com/roblanf/sarscov2phylo/}. Masked bases were ignored when attaching a sample genome to the ARG, but they were subsequently imputed. Further, 481 problematic sites flagged as prone to sequencing errors or highly homoplasic were excluded entirely (\href{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473}; \href{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).

\subsubsection{Exclusion of “time travellers”}

Including samples with incorrectly reported collection dates (e.g., an Alpha sample purportedly collected in 2020 from the United States, before Alpha appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce artefacts that distorted the topology of ARGs inferred early on in the study. Hence, to exclude such potential “time travellers”, we employed two filters. The first was to remove any sequence with a collection date that pre-dated the time to the most recent common ancestor (tMRCA) of its corresponding clade. The tMRCA for each clade was obtained from a Nextstrain GISAID reference tree (downloaded on August 22, 2022) where we used the lower bound of the 95\% confidence interval of each clade as the minimum date cut-off. The second filter was a simple threshold on the time delay between the collection date and submission date (maximum submission delay, or MSD). Indeed, we found that ARGs inferred from the MSD-filtered genome set had noticeably fewer branches with a high number of reversions. After some preliminary analysis, a MSD threshold value of 30 days was used when building the ARGs described herein. Future work could estimate collection dates by adding these potential time travellers back on the tree, allowing an automated assessment of collection date discrepancies.

\subsection{Inference}

\subsubsection{Overview of sc2ts}

We introduce an ARG inference method (schematic in Figure 1) that uses (1) an efficient HMM implementation of the Li and Stephens (LS) model \cite{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported collection dates of the SARS-CoV-2 samples to determine the order of genome copying; and (3) several parsimony-based heuristics to locally refine the topology. In brief, this combined approach uses the likelihood model of LS to attach sample genomes to existing nodes and to detect recombinants, but because many of the tip samples attached daily harbour few, if any, mutational changes, we use parsimony to resolve the relationships near the tip samples.

Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by taking advantage of the reported collection dates to incrementally grow the ARG. The inference process occurs forward in time, attaching new genomes aggregated by their collection date (daily batches), from the earliest to the last collection date. Prior to attaching sample genomes, the ARG is initialised such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence of recombination, the ARG is a single tree with no reticulation (Figure 1A); however, in the presence of recombination, a reticulate ARG is generated, with variation in tree topology across the genome (Figure 1B). Unlike standard phylogenetic trees in which internal nodes represent inferred ancestors, internal nodes in the ARGs inferred using sc2ts represent ancestors which can be samples or inferred ancestors.

Each daily batch of genomes is first attached to the growing ARG without any phylogenetic resolution. In cases where the newly added sequence is inferred to be a recombinant, a recombination node is inserted; the “date” of this node is arbitrarily set to the average of the collection dates of the genome and the most recent of its parent nodes. Otherwise, the genomes are attached to the most closely related ancestral node (an attachment node). This creates polytomies that are then locally resolved among the descendants of each attachment node using the Unweighted Pair Group Method with Arithmetic Mean (UPGMA) algorithm \cite{Michener1957-tr}, with average linkage (Figure 1C). This hierarchical clustering algorithm produces a dendrogram relating a set of sequences, which is treated as an ultrametric binary tree here. Subsequently, mutations are placed onto the daily tree by maximum parsimony. An issue with the UPGMA algorithm is that it may create internal nodes supported by no informative site (i.e., having no mutation immediately ancestral to them). In an adjustment step, such unsupported internal nodes are removed to “un-resolve” the daily tree; the remaining internal nodes, which are supported by at least one shared mutation, are retained. The “dates” of the non-sample internal nodes are set to the same date as the samples in the daily batch. The resulting daily tree is grafted onto the attachment node in the global ARG.

During ARG inference, two types of non-sample nodes are inserted to improve parsimonious placement of mutations: mutation-coalescing nodes and pushed-reversion nodes. We call these operations tree adjustments. A mutation-collapsing node is inserted when a shared mutation between sibling nodes is found (Figure 1D). An attempt to create mutation-collapsing nodes is made after grafting a new daily tree onto the global ARG. If shared mutations exist between two or more sibling nodes, then a parsimonious explanation is to create a new node that is the parent of those siblings, on which these mutations occurred. A pushed-reversion node is inserted when an immediate reversion occurs (Figure 1E), i.e. a mutation that reverts another mutation on its parent node. Instead of retaining both the mutation and immediate reversion, only the mutation is kept after a pushed-reversion node is inserted, thereby reducing the number of mutations on the branches. Note that these non-sample nodes are used as reconstructed genomes during HMM copying.

Furthermore, during inference, the ARG is automatically encoded, stored, and updated as a tree sequence using tskit \cite{Kelleher2018-xc}.

\subsubsection{Brief description of the Li and Stephens model}

The LS model is a convenient probabilistic approximation to the coalescent with recombination \cite{Hudson1983-if}, itself a generative model of a genealogical process with recombination \cite{Li2003-ib}. Importantly, it can be implemented as an HMM in which a focal genome is a sequence of nucleotides (observed states) that are probabilistically emitted as an imperfect mosaic of a set of reference genomes (hidden states). As such, an efficient HMM machinery can be employed to make statements about parameters of interest and highlight potential sources of error in genotyping and sequencing data. The LS model is used in a wide variety of applications in genomics. For example, it is the engine that drives statistical genotype phasing and imputation \cite{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it can be used to estimate mutation rates (e.g., REF), recombination rates (e.g., \cite{Hinch2011-tz}), and selection within and across hosts in viral sequence data (e.g., \cite{Palmer2019-wa}).

The generative process of the LS model is summarised in Figure 2. Here, a transition matrix, Q, governs the process of switching (recombining) between members of the reference panel (the hidden states). An emission matrix, \textit{E}, allows for differences between the focal sequence and the hidden state from which it is copied (due to mutation as well as genotyping or sequencing error). Then, we use the Viterbi algorithm \cite{Viterbi1967-ol} to find the most likely copying path through the sequences, given a set of reference sequences, \textit{Q}, and \textit{E}.

Because the HMM is memoryless, the potential parental lineages are independently assessed at each recombinantion breakpoint. This can result in the inference of a recombinant lineage descending immediately from three (or even more) parental lineages. We find that often two of the three parents are identical or similar in sequence in the segments contributed to the recombinant lineage. A more parsimonious explanation would be to switch back to the original parent sequence. In the future, a pragmatic solution would be to detect such cases during the ARG inference stage and to adjust the copying paths accordingly. Meanwhile, however, we urge caution when interpreting the parentage of lineages with multiple breakpoints.

\subsubsection{Parameterization of the HMM implementation of Li and Stephens model}

Recombination is inferred using an efficient HMM implementation of the LS model. Copying paths are estimated from SARS-CoV-2 genomes sampled over time, with younger genomes copying (i.e. inheriting segments and mutations) from older genomes. Here, we were able to apply the LS model on millions of SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of the LS model that is a core component of the tsinfer software. For this application, we have parameterised the HMM using a mismatch ratio (MMR), which tunes how strongly mutations (i.e., no switching of template sequences) are preferred over recombination (i.e., switching between sequences) during copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a focal genome more likely spans at least two ancestral genomes (i.e,. the focal genome is a putative recombinant) than a single ancestral genome. Here, the MMR was set to three in all the HMM runs (see Discussion for future potential development of the HMM).

\subsection{Analysis}

\subsubsection{Imputation of Pango lineage for non-sample nodes}

Non-sample nodes inserted during ARG inference had no Pango lineage assignment from Nextclade. We imputed the Pango lineage of non-sample nodes using a two-step procedure (Figure 3) to help assess how consistently breakpoint intervals are estimated by the HMM runs (see below). If the Pango lineage of the parental node was available and no mutation was immediately ancestral to the non-sample node, then the parental node’s status was directly propagated to the non-sample node. Otherwise, the Pango lineage of the non-sample node was inferred by matching its full set of mutations against the Pango lineage-defining mutations (based on 90\% consensus of the sequences analysed) from the COVID-CG website \cite{Chen2021-zc}; \href{https://covidcg.org/}; accessed on November 04, 2022).

\subsubsection{Breakpoint intervals and HMM consistency}

When recombinant ancestry for a given genome is inferred by the LS HMM, a breakpoint is identified as the last possible position where the copying path must switch. Thus, the breakpoints used in the ARG represent the rightmost end of the interval, within which the recombination event could have occurred. To obtain the leftmost end of this interval, we rerun the HMM using mirrored coordinates (i.e., from the 3’ end to the 5’ end).

A recombinant sequence was deemed to be HMM-consistent if (1) the same number of parent nodes and (2) the same set of mutations immediately ancestral to it were obtained from both the unmirrored and mirrored HMM runs. In the ideal case, the copying paths of both the HMM runs switch between parent sequences at the same position(s); in less ideal cases, however, the copying paths switch at different positions, typically because of a lack of genetic variation between the positions. The switch positions in the copying paths of both the HMM runs were used to define the breakpoint interval(s) of an HMM-consistent recombinant.

\subsubsection{Code availability}

The source code implementing the sc2ts pipeline is available via GitHub \href{https://github.com/jeromekelleher/sc2ts/}. The sc2ts pipeline uses various functions and algorithms implemented in tskit and tsinfer \href{https://github.com/tskit-dev/}. The pipeline uses the SciPy implementation of the UPGMA algorithm to build daily trees \href{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}. The Jupyter notebooks used to produce the results described herein are also available on GitHub \href{https://github.com/jeromekelleher/sc2ts-paper/}.

\section{Results}

\subsection{Inferred ARGs}

To illustrate the use of sc2ts, we employed two different sampling strategies. First, we inferred a densely sampled but time-limited “wide” ARG by attaching all the sample genomes with an MSD of 30 days, one daily batch at a time. The inference procedure ran until the genomes sampled on June 30, 2021 were attached to the ARG, resulting in a total of 1.27 million samples. Second, we inferred a “long” ARG over a more extended timespan, by attaching at most 1,000 genomes, which were randomly subsampled from each daily batch with an MSD within 30 days. For this ARG, the inference procedure ran until the genomes sampled on June 30, 2022 were attached to the ARG. The long ARG covers an extra year of samples but is more sparsely sampled. June 30, 2022 is the collection date of the final batch of sequences which were preprocessed for the development of sc2ts. Once sc2ts is further developed and tested, we will attach samples collected after that date to infer a fully sampled pandemic-scale ARG.

Most of the nodes in the ARGs represent sample genomes (Table 1). Only a small fraction of the nodes represent recombination events (0.28\% and 0.27\% in the wide ARG and long ARG, respectively), over half of which have no descendant samples (besides the recombinant sequence that triggered insertion of the recombination node) (Table 1). There are over one million mutations on the ARGs, the majority of which occur on a terminal edge (Table 1). Notably, the massive amount of genomic variation data in the wide ARG (1.27 million genomes) and long ARG (657 thousand genomes) are compactly stored in ~57 Mb and ~36 Mb, respectively (less than ten bytes per sample). These ARGs can be loaded for analysis in a Jupyter notebook on a typical commercial laptop in merely a fraction of a second using the Python package tskit (Table 1).

\subsection{Computational resources}

The wide ARG (with over one million sample genomes) was built using 128 threads on a machine with 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). Inference ran for 17 days to attach genomes sampled up to June 30, 2021. The long ARG (with ~700 thousand sample genomes) was built using 40 threads on a machine with 256 GB RAM (2x Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz). Inference ran for 23 days to attach genomes sampled up to June 30, 2022. Most of the runtime was spent on running the HMM.

\subsection{Backbone phylogeny}

For validation purposes, we compare the genetic relationships in the sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain project \cite{Hadfield2018-ef}. The NextStrain trees consist of subsamples of SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely restricted to non-recombinant lineages. To allow for comparison, the ARGs were "simplified" \cite{Kelleher2018-xc} to retain only those tips of our genealogy that match the samples present in a specific NextStrain tree. These simplified "backbone" ARGs exhibit very little recombination (6 and 5 trees across the genome for the wide and long ARGs, respectively).

Figure 4 shows a co-phylogeny of the 88 samples shared between the wide ARG and a GISAID global "all-time" tree from NextStrain (downloaded on January 21, 2023). To reduce the ARG to a single tree, we show the very first tree in the wide ARG; however, the five other trees in this simplified ARG show near-identical patterns. Supplementary Figure 3 shows the same comparison for the long ARG. It is clear that the backbone topology of the sc2ts tree shows very close agreement with the NextStrain tree. The sample genomes cluster by their assigned Pango lineage status, and many variants and their descendents form identical monophyletic clades in both the trees (e.g., the Alpha and Delta clades).

There are, however, some notable differences between the trees. Firstly, the sc2ts tree is generally less well resolved (noticeably, February to March, 2020) (Figure 4). Secondly, there are non-identical sample partitions near the tips (e.g., in the Alpha clade, Figure 4). These might have arisen because the samples containing the mutations needed to resolve local topologies were collected later than the cut-off date for the sc2ts ARG, but were for the NextStrain tree. Finally, the branch lengths differ. For example, the branch leading to the ancestral node of the Delta variant samples is much shorter in the sc2ts tree than the NextStrain tree (Figure 4). It is possible that the length of some of the branches was poorly calibrated, probably caused by overly aggressive insertion of mutation-collapsing nodes and reversion-pushed nodes or suboptimal sample attachment. Side-by-side comparison of sc2ts trees with trees inferred using another method (e.g., UShER) should help to understand whether branch length differences are due to poor calibration or methodological differences.

Although here we focus on a single local tree at an arbitrarily chosen position, note that in the original “un-simplified” ARGs, there are 1,496 and 958 trees, respectively (Table 1). Hence, while for a relatively small sample the genealogy may be well described by a single tree, the full dataset shows substantial variation in phylogenetic relationships (or phylogenetic incongruence) across the SARS-CoV-2 genome, much of which may be attributable to recombination. Such complex reticulate relationships are appropriately captured by a genealogy in the form of an ARG.

\subsection{Putative recombination events}

Both the wide and long ARGs contain a large number of recombination nodes (4,123 and 2,078, respectively; Supplementary Tables 1 and 2), with each of these nodes inferred as a result of matching a specific sample (a recombinant) into the ARG. However, ~65\% of recombination nodes are ancestors of only a single descendant sample in each local tree; these could have arisen from sequencing of mixed samples (e.g., from co-infection or contamination), or were simply not transmitted onwards and so were excluded from subsequent analyses (leaving X and Y recombinants in the wide and long ARGs, respectively). In their analysis using RIPPLES, Turakhia et al. (2022) detected 589 recombinants, also excluding singletons.

For simplicity, we focus primarily on “non-complex” recombination nodes. To do this, we apply a filtering step that removes those with more than one breakpoint (and often more than two parent lineages), or where matching is not HMM-consistent or identifies parents with different Pango lineage assignments in the original (i.e., 5’ to 3’ end) and mirrored (i.e., 3’ to 5’ end) HMM runs. Filtering reduces both the number of recombination nodes and breakpoints from  A \& B to C in the wide ARG and from 726 & 806 to 459 in the long ARG (Supplementary Tables 1 and 2). Note that in the unfiltered case, a single recombination node can correspond to multiple points, one for each recombination breakpoint. The effect of this filtering step in the long ARG, which includes time periods of increased rates of coinfection and recombination compared to the wide ARG, can be seen in Figure 3A (unfiltered) and Figure 3B (filtered).

Importantly, the long ARG captured all except one of the Pango designated recombinant lineages (see the section below), but we also detected a large number of additional recombination events, with parents lineages that diverged as far back as two years to as recent as five or six days prior to the recombination event (Figure 3). This provides a degree of resolution of viral recombinants not available with other recombination detection methods. The majority of newly identified recombinants arise from parents that diverged 10 to 30 weeks previously. However, there is also a peak of recombinants that diverged ~95 weeks prior to the event: these correspond to recombinants whose parental common ancestor traces to early 2020, including both Delta-Delta and Delta-Omicron recombinants (Supplementary Figure 4).

The date of common ancestry for the two parents of recombinant lineages is concentrated in several banded rows in Figure 5. These are largely due to a small handful of most recent common ancestors (MRCAs), which are shared between a large number of recombination events, the top four of which are indicated by dotted lines in the figure. These MRCAs are near the root of large expansions of certain lineages (B.1: original variant;  B.1.617.2: Delta, BA.1: first Omicron wave; and BA.2: second Omicron wave), as evidenced by the number of samples with that Pango designation that trace back to these ancestral nodes (listed as percentages on the plot). The specific MRCA nodes to which many recombinants trace frequently represent very large polytomies in the tree (at the extreme, 2,872 immediate children). This is likely to indicate a rapid and under-sampled expansion of a clade in the SARS-CoV-2 genealogy.

Finally, it is clear from Supplementary Figure 4 that of the recombinant parents that can be classified into Alpha, Delta, and Omicron variants, the majority are within-Delta and within-Omicron recombinants, and that Omicron followed by Delta, are the variants associated with the most recombination. This may reflect either sampling intensity, the prevalence of cases (which increases the chance of coinfection and recombination), or possibly heterogeneity in recombination rates among lineages.

\subsection{Early recombinants reported by Jackson et al. (2021)}

By searching for samples combining genomic segments from Alpha (B.1.1.7) and from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations, Jackson et al. (Jackson et al. 2021) reported 16 recombinant sequences among genomes sampled in the United Kingdom up to March 7, 2021. The authors proposed that the recombinants likely arose from eight independent recombination events (categorised into groups A to D and four singletons).

In the wide ARG (sampled to mid-2021), all these sequences are descendants of recombination nodes, except MILK-103C712, which was removed during preprocessing (Table 2); additionally, all are HMM-consistent, except QEUH-1067DEF. Overall, we find excellent agreement between Jackson et al.’s motif-based method and sc2ts in terms of (1) the Pango lineage of the parents and (2) the breakpoint intervals (Table 2). In all cases, the two methods proposed identical or very closely related parent lineages. The general mosaic genome structure is concordant in all the cases, even CAMC-CB7AB3 where there are two switches in the HMM copying paths (indicative of two breakpoints). The breakpoint intervals estimated using the two methods overlap substantially in all the cases (Table 2). Within each of the groups A to D, the constituent sequences have identical breakpoint intervals, because they trace back to a single recombination node in the wide ARG. This result is consistent with the hypothesis that the sequences of the groups A to D emerged from four separate recombination events (Jackson et al. 2021).

\subsection{Pango recombinants}

The wide ARG only contains samples from three Pango-designated recombinant lineages (XA, XB, and XC), due to the fact that it only includes samples up to mid-2021. By contrast, the long ARG (subsampled to mid-2022) contains samples from 33 Pango-designated recombinant lineages (XA to XAK; Supplementary Table 3). For Pango recombinant analysis, we therefore focus on the long ARG.

In the wide ARG, every sequence of the three Pango recombinants trace back to some recombination node. XA, XB, and XC.

In the unfiltered long ARG, the sequences of 28 Pango recombinants trace back to some recombinant node(s); however, the sequences of the five Pango recombinants XP, XU, XAA, XAG, and XAK do not. A likely explanation for the XP recombinant sequences not having a recombinant origin in the long ARG is that its characteristic multibase deletion (https://github.com/cov-lineages/pango-designation/issues/481) was masked during preprocessing and therefore was not incorporated into the HMM.
Upon tracing the origin of XAG through the ARG, XAG might have been caused by incorrect imputation of its parent nodes(?). XAK likely does not appear to have a recombinant origin because of random subsampling, but it may with a more complete tree.

The sequences of 14 Pango recombinants trace back to a single recombination node. The sequences of the remaining 18 Pango recombinants trace back to multiple recombination nodes, however. This may be a result of undersampling in the long ARG, which was built by randomly subsampling at most 1,000 sequences from each daily batch. Having extra recombination nodes in the long ARG appeared to be a more parsimonious explanation for the 19 Pango recombinants than having a single recombination node with many immediately ancestral mutations (including recurrent mutations and reversions), despite that the recombinants may indeed trace to a single origin.

The mosaic genome structure of all the 32 Pango recombinants is concordant with the structure determined by the Pango Lineage Designation Committee (Supplementary Table 3). For example, the genome of XD (also known as “Deltacron”) is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17 (Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.

\subsection{Novel recombinants}

Next, we examine recombination nodes in the filtered long ARG (sampled to mid-2022) that represent previously unnamed putative recombinant sequences, which do not have a Pango recombinant designation. For this proof-of-concept study, we have arbitrarily picked 12 recombination nodes that seem plausible (Supplementary Table 4). Coincidentally, all these nodes involve Omicron subvariants (BA.1, BA.2, BA.4, and BA.5). These recombination nodes (1) were inserted into the long ARG on or after January 8, 2022; (2) have at least 10 descendent samples; and (3) have no mutational differences (including immediate reversions) from their parent nodes.

TODO: Node  740761 (USA/NC-CDC-LC0668306/2022) is a sister clade proposed1006 in the UShER public phylogeny.
Node 628656 (Scotland/QEUH-37794BB/2022) is a sister clade of XAC, and annotated as “miscBA2BA1PostSpike”.

\subsection{Recombinant origins of the Pango lineages XA and XAG}

XX

\subsection{Distribution of recombination breakpoint intervals}

We explore recombination signals across the SARS-CoV-2 genome by inspecting the breakpoint intervals of the filtered recombination nodes from the wide ARG (sampled to mid-2021) (Figure 4). The length of a breakpoint interval reflects uncertainty in the estimated location of a breakpoint. On average, the length of the breakpoint intervals is 2,895 bases (median, 1,670 bases; range, 1 to 23,666 bases). This broad distribution of interval length indicates that the location of a breakpoint often could not be precisely determined given the observed genetic variation.

There is a preponderance of breakpoint intervals spanning the 3’ end of the genome (Figure 4A), as noted previously by Turakhia et al. \cite{Turakhia2022-it}. Inspecting more closely, we find an enrichment in the number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure 4B). Markedly, there is an abrupt drop in the number of intersecting intervals in the N gene following the peak in ORF8 towards the 3’ end.

Next, we investigate the potential recombination signals alongside the number of mutations across the genome (Figure 4C). Regions with a higher density of mutations contain more genetic variation, increasing the power to detect recombination events. The position 28,271 (located in the ORF8-N intergenic region) has a particularly high number of mutations in the samples (Figure 5). At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions (n = 1,605) appear to occur frequently. One possible explanation is that this homopolymeric region is prone to sequencing errors, giving rise to an artifactual recombination signal. At the upstream position 27,972, numerous mutations also occur. This position has been reported to be hypermutable \cite{Jungreis2021-dh}.

We further consider the breakpoint intervals within the context of recombination breakpoint sequence motifs. It is hypothesised that certain palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote template switching during replication in SARS-CoV-2 via formation of base-paired stem loops in the genome structure \cite{Gallaher2020-lb}. There are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019 reference sequence (Supplementary Table 5). We observe that XX (XX\%) of the breakpoint intervals of the HMM-consistent recombinants span at least one of the breakpoint sequences.

\subsection{Mutational spectrum}

In addition to recombination events, the ARGs record mutation events along the SARS-CoV-2 phylogeny. Yi et al. \cite{Yi2021-sc} reconstructed a SARS-CoV-2 phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to January 12, 2021, and systematically catalogued all the mutations occurring along the phylogeny. Notably, in the mutational spectrum described by Yi et al., C-to-U and G-to-U mutations occur more frequently than U-to-C and U-to-G mutations, respectively.

For validation purposes, we compare the mutational spectrum from the wide ARG with the mutational spectrum of Yi et al. We categorised all of the single nucleotide mutations by type (defined by the ancestral state and derived state), excluded mutations inherited by only a single sample (which are enriched for sequencing errors and artefacts), and tallied them up by type. Similarly, we took the data for single nucleotide mutations from Yi et al. \ref{https://github.com/ju-lab/SC2_evol_signature}, excluded mutations occurring along terminal branches, and tallied them up by type. We find that the mutational spectrum from the wide ARG (448,825 mutations) matches the mutational spectrum reported by Yi et al. (92,344 mutations) (Figure X). In both spectra, C-to-U mutations and G-to-U mutations occur more frequently than U-to-C and U-to-G, respectively (Figure X). Similar results are obtained when including the mutations inherited by only a single sample or occurring on terminal branches (data not shown).

\subsection{Mutations under selection}

XX

\section{Discussion}

The COVID-19 pandemic continues to be a global emergency, with persistent high levels of infection. This high prevalence has allowed the proliferation of many variants, with more than 600 Pango-designated lineages circulating globally in the last three months (January to March, 2023; https://gisaid.org/; access on March 27, 2023). High prevalence also brings a higher risk of coinfection, increasing opportunities for new phenotypically distinct recombinants to emerge and spread.

In this proof-of-concept study, we demonstrate the feasibility of reconstructing a large ARG from over one million SARS-CoV-2 genomes, a feat that is simply not possible using other available ARG inference methods (Rasmussen et al. 2014; Ignatieva et al. 2021; Speidel et al. 2019). We have accomplished this feat by building on top of recent advances in storing and processing large ARGs (tskit) and inferring genealogical relationships efficiently at scale (tsinfer). The highly compact representation of the ARGs as succinct tree sequences, coupled with vectorised computing technologies, enables the large SARS-CoV-2 ARGs to be loaded in less than one second in a Jupyter notebook and allows for each SARS-CoV-2 genome to be stored in less than eleven bytes of data. We believe that further development of the sc2ts methodology and software can create a powerful tool (1) to efficiently detect emerging or previously overlooked recombinants in SARS-CoV-2; (2) to jointly model the mutational and recombinational processes shaping global SARS-CoV-2 genomic diversity; and (3) to perform a broad variety of evolutionary analyses of SARS-CoV-2 genomes with unprecedented global sequencing coverage.

ARGs are rich genealogical structures that contain information about the genetic relationships among sampled genomes in the presence of recombination. Here, we have conducted several analyses of the preliminary ARGs to show that the ARGs built using sc2ts can (1) recapitulate the phylogenetic relationships among major SARS-CoV-2 lineages; (2) recover biologically plausible recombination signals in SARS-CoV-2 genomes; and (3) model mutation events in SARS-CoV-2 evolutionary history.

ARGs reconstructed using sc2ts capture the phylogenetic relationships of major SARS-CoV-2 lineages. Our results show that the backbone topology of a densely sampled wide ARG inferred using sc2ts agrees well with the backbone topology of a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference package. However, we have noted several areas of differences (e.g., poorly calibrated branch lengths), which are the focus of future development of sc2ts. We believe that some of these differences can be reduced by improving and refining our approach and strategy to handle samples with erroneously reported collection dates (i.e., time travellers that have passed the maximum submission delay of 30 days) and to modify the topology of daily trees.

Our results show that sc2ts identifies the early recombinants (pre-2022) previously identified by Jackson et al. (2021). Sc2ts also identifies many Pango recombinants, which are designated after community perusal of supporting genomic evidence. We have found several previously unreported and overlooked recombinant sequences in the wide ARG. After further improvements and enhancements, we will use sc2ts to build a comprehensive ARG to search for recombination signals among all the 15 million and counting SARS-CoV-2 genomes available.

“The rate of putative recombination breakpoints is about three times higher towards the 3' of the change point than the 5' interval.“ Consistent with this, the breakpoint intervals of the putative recombinants identified here tend to occur on the 3’ end of the SARS-CoV-2 genome.

\section{Improvements}

Currently, sc2ts is in development. Below, we propose several ways to potentially improve or enhance sc2ts. We welcome suggestions and ideas from experts.

\begin{itemize}
\item A potential source of error in sc2ts ARGs comes from incorrectly reported collection dates of the samples, which can cause sample genomes to be attached to the wrong parts of the ARGs. Thus far, we have attempted to correct the most offending errors in the ARGs by filtering out "time travellers", but the sampling time can still be wrong by days or tens of days. ***describe possible approaches to correct this?***
\item To avoid time travellers, we used a maximum submission delay, but this will systematically bias against submissions from countries with greater submission delays (Kalia et al.). An area of future work is to add back the excluded sequences to a built ARG to assess which ones are likely true time travellers.
\item To avoid time travellers, we used a maximum submission delay, but this will systematically bias against submissions from countries with greater submission delays (Kalia et al.). An area of future work is to add back the excluded sequences to a built ARG to assess which ones are likely true time travellers.
\item We have used a single mismatch value in all the HMM runs, set so that having three mutations is as likely as having a single recombination. This parameter choice is arbitrary, and higher values may reduce the number of origins for those Pango lineages that were inferred to descend from more than one recombinant ancestor. In addition, because both the mutation rate and recombination rate varies across the SARS-CoV-2 genome, it may be helpful to allow the mismatch value to vary along the genome.
\item By treating mutations equally in the current HMM implementation, we implicitly assume that the rates of different types of mutations are identical. However, as reported previously (Yi et al. 2021), there is a high degree of mutational asymmetry in SARS-CoV-2. To account for mutational biases, more sophisticated nucleotide substitution models (e.g., allowing for different rates of transition and transversion) can be incorporated into HMM likelihood calculations.
\item Our current masking strategy handles indels as missing data. Doing this can remove information helpful to detect recombinants. For example, sequences of the Pango recombinant XP were not identified as recombinants, because XP’s characteristic multibase deletion was masked.
\item This "lack of memory" in the HMM can create recombinants with three or more parents when it would be equally parsimonious to have just two. The copying process in the HMM currently does not limit the number of parents to two, but doing so would be a valuable addition.
\item Some recombinants may reasonably have three or more parents, indicating multiple "stacked" recombination events. We have largely ignored these in the current analysis, but with further sc2ts development, there is potential to analyse these putative recombinants and trace their evolutionary origins through the ARG.
\item In the present implementation, HMM traceback is performed once, yielding a single maximum-likelihood genome copying path. A stochastic traceback algorithm (e.g., (Rasmussen et al. 2014)) can be implemented in order to quantify uncertainty about putative parent node(s) and switch positions. These estimates of uncertainty may be used as support values for branches analogous to bootstrap values produced by standard phylogenetic methods.
\item Daily trees are built using the UPGMA algorithm. This entails an extra step to “un-resolve” daily trees in parts of the trees with no phylogenetic resolution. To avoid this step, a better approach is to use a maximum-parsimony algorithm to build daily trees instead.
\end{itemize}

\section{Conclusions}

The global understanding of viral evolution during the COVID-19 pandemic is unprecedented, with over 15 million sequences shared to date. We envisage that once fully developed and tested, sc2ts methodology and software may provide a unified computational framework to study the evolution and epidemiology of SARS-CoV-2 at pandemic scale. The evolutionary analyses presented herein exemplify the numerous applications possible using the ancestral recombination graph inferred by sc2ts. We invite bioinformaticians, evolutionary biologists, and virologists to join us and the broader tskit community to build the most complete representation of the evolutionary history of SARS-CoV-2. Moreover, while we have focussed on the particular application to SARS-CoV-2, the methods we describe are readily applicable to the analysis of many other viral pathogens. Solving the challenges in building and analysing large viral genealogies would therefore contribute to the important goal of improving preparedness for potential future pandemics.

\section{Acknowledgements}

We gratefully acknowledge all data contributors, i.e., the Authors and their Originating laboratories responsible for obtaining the specimens, and their Submitting laboratories for generating the genetic sequence and metadata and sharing via the GISAID Initiative, on which this research is based. Also, we thank Dr. Morag Graham (the National Microbiology Laboratory, Winnipeg, Canada) for kindly providing the genomic coordinates of the breakpoint sequence motifs. SHZ is supported by the Janssen-Oxford Translational Genomics Fellowship. JK, YW, and BJ are supported by the Robertson Foundation.

\begin{table}
\centering
\begin{tabular}{l|l|l}
\hline
\textbf{Property} & \textbf{Wide ARG} & \textbf{Long ARG} \\
\hline
Last sampling date & 2021-06-30 & 2022-06-30 \\
Local trees & 1,496 & 958 \\
\hline
\multicolumn{3}{l}{\textit{Nodes}} \\
All nodes (sample and non-sample) & 1,453,347 & 783,231 \\
Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\
Tree adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107 (10.97\%) \\
Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\
\hline
\multicolumn{3}{l}{\textit{Types of recombination nodes}} \\
Singletons (i.e. no descendant besides the sequence causing insertion of the node) & 2,601 (63\%) & 1,315 (63\%) \\
Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%) \\
HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\
\hline
\multicolumn{3}{l}{\textit{Mutations}} \\
Total mutations & 1,213,193 & 1,062,072 \\
Mutations on a terminal edge & 758,903 (62.55\%) & 767,111 (72.23\%) \\
Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\
Immediate reversions (one intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\
\hline
\multicolumn{3}{l}{\textit{Storage and loading time}} \\
tszipped file (including sample metadata, imputed Pango lineage assignments, and results of HMM runs for recombination nodes) & ~57 Mb & ~36 Mb \\
Bytes to encode one sample & 8.29 & 10.83 \\
Load time (seconds) & 0.9 & 0.5 \\
\hline
\end{tabular}
\end{table}

\begin{table}
\centering
\begin{tabular}{l|c|c|c|c|c}
\hline
\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} & \multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\
\hline
\textbf{Sample} & \textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\ interval(s)} & \textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\
\hline
ALDP-11CF93B & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
ALDP-130BB95 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 & \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
QEUH-CCCB30 & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 & \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 & \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
MILK-1166F52 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
MILK-126FE1F & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
RAND-128FA33 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 & \thead{B.1.177 \\ Alpha} & 16,177–21,255 \\
CAMC-CB7AB3 & n/a & \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} & \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\
MILK-103C712 & n/a & \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} & n/a & n/a \\
QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 & \thead{Alpha \\ B.1.177} & 7,729–10,870 \\
\hline
\end{tabular}
\caption{\label{tab:jackson}Recombinant sequences involving the Alpha (B.1.1.7) variant reported by Jackson et al. \cite{Jackson2021-ik} have recombinant ancestry in the wide ARG. The breakpoint intervals and Pango lineage assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et al., except the Pango lineage assignment of the parents of group B recombinants, which were taken from Table 1 (motif-based results).}
\end{table}

\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
\caption{\label{fig:overview_sc2ts} A schematic of the sc2ts method. Sc2ts reconstructs the genetic relationships among SARS-CoV-2 genomes by copying samples to all possible ancestors collected at earlier time points (curved arrows). An ARG is built by attaching new sample genomes grouped by collection date to an existing ARG rooted at the Wuhan-Hu-1 reference genome, day by day. Each iteration involves three stages: (1) attachment of new samples to the growing ARG (A, B); (2) reconstruction of trees relating the samples under each attachment node (C); and parsimony-based tree topology adjustments (D, E). In the absence of  recombination, sc2ts infers an ARG that is a single tree relating the samples (A). However, in the presence of recombination, sc2ts infers an ARG that is encoded as a sequence of local trees relating segments of the sample genomes (B), inserting a recombination node to represent a putative recombination event. Additionally, mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to make more parsimonious placements of mutations that should be shared or should not be immediately reverted, respectively.}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/ls_diagram_covid.pdf}
\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS) model. The LS model in its implementation as an HMM is governed by the transition and emission matrices, \textit{Q} and \textit{E}, respectively. Each may be a function of the reference panel members, but transitions are generally assumed to be independent of the hidden states themselves (shown in the pink panel). This dramatically increases performance as the state space drops to two states (i.e., switching or not switching). The recombination probability at site \textit{l} is \textit{rl}, while n denotes the number of members in the reference panel. Emissions may also be a function of the nucleotide states, but in our case, we assume that all observed alleles (in both the reference panel and focal sequence, \textit{al}) at the site under consideration are equally likely and occur with mutation probability \textit{l} (shown in the blue panel). These two processes, transition and emission (which encode recombination and mutation, respectively), define the generative process of the HMM, by which an imperfect mosaic is emitted from a reference panel of samples and nodes from the ARG inserted at earlier collection dates (shown in the lower panel). Here, the unseen states in the reference panel are shown as coloured lines enclosed by the grey box, and the true path through the data which leads to the emitted focal sequence is displayed by the black arrow. Examples of transition and emission probabilities along this trajectory are shown by the red and blue arrows, respectively. Black crosses along the focal sequence show genotyping errors or mutations. Note that in practice, we do not know the pattern of colours along the focal sequence and must infer it using the Viterbi algorithm.}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
\caption{\label{fig:mutational_spectra} Mutational spectrum of SARS-CoV-2 in the wide ARG (sampled to mid-2021) compared to that reported by Yi et al. \cite{Yi2021-sc}. Mutations are categorised by type (i.e., ancestral state > derived state). The percentages of each mutation type from the wide ARG are represented by blue bars and the percentages from Yi et al. by orange bars, with the darker colours representing one direction (e.g., C > U) and the lighter colours the reverse (e.g., U > C).}
\end{figure}

\bibliographystyle{alpha}
\bibliography{references}

\end{document}

### 2023-04-06 main.tex
Wrap lines

@@
 
-Recombination via template switching is a common feature in the evolution of coronaviruses (CoV) \cite{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2 \cite{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}.
-By bringing together mutations carried by different lineages, recombination plays an important role in generating genomic diversity, with recombinant lineages associated with CoV adaptation to new host species and with the production of more immune evasive variants \cite{Graham2010-xe,De_Klerk2022-tt}.
+Recombination via template switching is a common feature in the evolution of
+coronaviruses (CoV) \cite{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2
+\cite{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
+together mutations carried by different lineages, recombination plays an
+important role in generating genomic diversity, with recombinant lineages
+associated with CoV adaptation to new host species and with the production of
+more immune evasive variants \cite{Graham2010-xe,De_Klerk2022-tt}.
 
-Recombination was challenging to detect in SARS-CoV-2 early in the pandemic due to gradual accumulation of genetic variation.
-In late 2020, the appearance and spread of variants of concern (VoC) harbouring multiple characteristic mutations increased the detectability of recombination.
-By searching for genomes containing segments from different VoC, multiple recombinant lineages have been detected, for example, between Alpha and pre-VoC \cite{Jackson2021-ik}, Alpha and Epsilon \cite{Wertheim2022-hj}, and Alpha and Delta \cite{Sekizuka2022-xz}).
-More recently, the high prevalence of Omicron, with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has accelerated the rate of coinfection and the potential for recombination \cite{Bal2022-hq}.
-Recombinant lineages that have sufficient evidence of onward transmission (more than five descendants) and monophyly (> 70\% bootstrap support) are eligible for a Pango lineage designation \cite{Rambaut2020-dw}, starting with the letter X (“Pango Lineage Nomenclature: Provisional Rules for Naming Recombinant Lineages” 2021).
+Recombination was challenging to detect in SARS-CoV-2 early in the pandemic due
+to gradual accumulation of genetic variation. In late 2020, the appearance and
+spread of variants of concern (VoC) harbouring multiple characteristic
+mutations increased the detectability of recombination. By searching for
+genomes containing segments from different VoC, multiple recombinant lineages
+have been detected, for example, between Alpha and pre-VoC
+\cite{Jackson2021-ik}, Alpha and Epsilon \cite{Wertheim2022-hj}, and Alpha and
+Delta \cite{Sekizuka2022-xz}). More recently, the high prevalence of Omicron,
+with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
+accelerated the rate of coinfection and the potential for recombination
+\cite{Bal2022-hq}. Recombinant lineages that have sufficient evidence of onward
+transmission (more than five descendants) and monophyly (> 70\% bootstrap
+support) are eligible for a Pango lineage designation \cite{Rambaut2020-dw},
+starting with the letter X (“Pango Lineage Nomenclature: Provisional Rules for
+Naming Recombinant Lineages” 2021).
 
-Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that combine specific subsets of mutations from different lineages (e.g., \cite{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}).
-This approach only identifies those inter-lineage recombination events resulting in the patterns investigated.
-Others have inferred recombinant genealogies to disentangle recombination from recurrent mutation, without having to pre-specify characteristic mutations for each lineage \cite{Ignatieva2021-rg}.
-More recently, Turakhia et al.
-\cite{Turakhia2022-it} has presented a phylogeny-based approach ("RIPPLES") to identify putative recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying motifs.
-To find candidate recombinants, RIPPLES first builds a phylogeny without assuming recombination and then scans for long branches (i.e., carrying many mutations) and then determines if these candidates would be better explained by recombination.
-Specifically, RIPPLES exhaustively breaks each candidate sequence into segments and attempts to find more parsimonious placements for each segment on the phylogeny.
-If such placements are found, the sequence is identified as a putative recombinant.
-Although it enables rapid searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that accounts for only mutations, treats recombinants \textit{post hoc}, and is an incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
-As noted by the authors, a \textit{post hoc} treatment of recombination is possible when recombinant lineages are rare and leave few descendants, but this is becoming increasingly untenable with the proliferation of recombinant lineages (e.g., more than half of the sequences sampled in February, 2023 are from XBB and its descendants; \cite{Chen2022-pz}.
+Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
+combine specific subsets of mutations from different lineages (e.g.,
+\cite{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}).
+This approach only identifies those inter-lineage recombination events
+resulting in the patterns investigated. Others have inferred recombinant
+genealogies to disentangle recombination from recurrent mutation, without
+having to pre-specify characteristic mutations for each lineage
+\cite{Ignatieva2021-rg}. More recently, Turakhia et al. \cite{Turakhia2022-it}
+has presented a phylogeny-based approach ("RIPPLES") to identify putative
+recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying
+motifs. To find candidate recombinants, RIPPLES first builds a phylogeny
+without assuming recombination and then scans for long branches (i.e., carrying
+many mutations) and then determines if these candidates would be better
+explained by recombination. Specifically, RIPPLES exhaustively breaks each
+candidate sequence into segments and attempts to find more parsimonious
+placements for each segment on the phylogeny. If such placements are found, the
+sequence is identified as a putative recombinant. Although it enables rapid
+searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2
+phylogeny that accounts for only mutations, treats recombinants \textit{post
+hoc}, and is an incomplete representation of the reticulate evolutionary
+history of SARS-CoV-2. As noted by the authors, a \textit{post hoc} treatment
+of recombination is possible when recombinant lineages are rare and leave few
+descendants, but this is becoming increasingly untenable with the proliferation
+of recombinant lineages (e.g., more than half of the sequences sampled in
+February, 2023 are from XBB and its descendants; \cite{Chen2022-pz}.
 
-Inferring evolutionary trees while accounting for both mutation and recombination is a long-standing challenge.
-This challenge is particularly acute for SARS-CoV-2, given the large amount of data that has been shared, with over 15 million sequences in the GISAID database \cite{Shu2017-hp}.
-It is well known that recombination distorts phylogenies \cite{Schierup2000-fg} and affects the results of downstream analysis (e.g., inference of selection; \cite{Anisimova2003-vr}).
-However, standard phylogenetic methods do not account for recombination (e.g., \cite{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}).
-Alternatively, genetic ancestries of recombining lineages can be represented by ancestral recombination graphs (ARG; \cite{Griffiths1981-lw,Hudson1983-if}, which  capture network-like phylogenetic relationships \cite{Gusfield2014-qw}.
+Inferring evolutionary trees while accounting for both mutation and
+recombination is a long-standing challenge. This challenge is particularly
+acute for SARS-CoV-2, given the large amount of data that has been shared, with
+over 15 million sequences in the GISAID database \cite{Shu2017-hp}. It is well
+known that recombination distorts phylogenies \cite{Schierup2000-fg} and
+affects the results of downstream analysis (e.g., inference of selection;
+\cite{Anisimova2003-vr}). However, standard phylogenetic methods do not account
+for recombination (e.g., \cite{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}).
+Alternatively, genetic ancestries of recombining lineages can be represented by
+ancestral recombination graphs (ARG; \cite{Griffiths1981-lw,Hudson1983-if},
+which  capture network-like phylogenetic relationships \cite{Gusfield2014-qw}.
 
-Approaches to infer ARGs now exist that can scale to tens of thousands of human genomes and beyond \cite{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}.
-Recently, Wohns et al.
-\cite{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands of modern and ancient human genomes, based on over 90 million variable sites.
-To achieve this, Wohns et al.
-\cite{Wohns2022-th} utilised several computational advances in ARG inference: (1) the "succinct tree sequence" data structure that can compactly store large ARGs  \cite{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM) implementation of the Li and Stephens (LS) model (Li and Stephens 2003) to infer copying paths between genomes \cite{Kelleher2019-ba}; and (3) efficient, open-source software tools \cite{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.
+Approaches to infer ARGs now exist that can scale to tens of thousands of human
+genomes and beyond
+\cite{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}. Recently,
+Wohns et al. \cite{Wohns2022-th} reconstructed a unified whole-genome genealogy
+of thousands of modern and ancient human genomes, based on over 90 million
+variable sites. To achieve this, Wohns et al. \cite{Wohns2022-th} utilised
+several computational advances in ARG inference: (1) the "succinct tree
+sequence" data structure that can compactly store large ARGs
+\cite{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM)
+implementation of the Li and Stephens (LS) model (Li and Stephens 2003) to
+infer copying paths between genomes \cite{Kelleher2019-ba}; and (3) efficient,
+open-source software tools
+\cite{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.
 
-Here, we present a preliminary version of a novel scalable methodology (“sc2ts”, pronounced “scoots”) that builds on the advances above and additionally takes advantage of known collection dates to infer ARGs from millions of full-length SARS-CoV-2 genomes.
-Using publicly available SARS-CoV-2 genome sequences, we demonstrate that sc2ts can be used to simultaneously (1) detect putative SARS-CoV-2 recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages; (3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record of evolutionary information about SARS-CoV-2 that can be efficiently accessed using "tskit" (tree sequence toolkit) software library \cite{Kelleher2018-xc}.
-Sc2ts is a promising method to automate detection of SARS-CoV-2 recombinants and to jointly model recombinational and mutational processes shaping the genomic diversity of SARS-CoV-2 at large scale.
+Here, we present a preliminary version of a novel scalable methodology
+(“sc2ts”, pronounced “scoots”) that builds on the advances above and
+additionally takes advantage of known collection dates to infer ARGs from
+millions of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2
+genome sequences, we demonstrate that sc2ts can be used to simultaneously (1)
+detect putative SARS-CoV-2 recombinants; (2) infer the phylogenetic
+relationships of the major SARS-CoV-2 lineages; (3) reconstruct the mutational
+spectrum of SARS-CoV-2; and (4) provide a compact record of evolutionary
+information about SARS-CoV-2 that can be efficiently accessed using "tskit"
+(tree sequence toolkit) software library \cite{Kelleher2018-xc}. Sc2ts is a
+promising method to automate detection of SARS-CoV-2 recombinants and to
+jointly model recombinational and mutational processes shaping the genomic
+diversity of SARS-CoV-2 at large scale.
 
@@
 
-The findings of this study are based on sequences and metadata available on GISAID (\href{https://gisaid.org/}) up to Aug 22, 2022 and accessible at doi.org/10.55876/gis8.230329cd (Supplementary Table 1).
-Sequences were removed if they (1) had ambiguous collection dates; (2) were collected before January 01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
-Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank: MN908947.3) using Nextclade v2.3.0 \cite{Aksamentov2021-hj} (dataset tag 2022-07-26T12:00:00Z).
-To reduce the impact of poor-quality sequences on ARG inference, sequences were excluded if they had a “bad” quality control status in any of the four Nextclade columns ("qc.missingData.status”, “qc.mixedSites.status”, “qc.frameShifts.status”, and “qc.stopCodons.status”).
-These excluded genomes are enriched for low-coverage, contaminated, or misassembled sequences.
+The findings of this study are based on sequences and metadata available on
+GISAID (\href{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
+doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
+if they (1) had ambiguous collection dates; (2) were collected before January
+01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
+Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
+MN908947.3) using Nextclade v2.3.0 \cite{Aksamentov2021-hj} (dataset tag
+2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
+inference, sequences were excluded if they had a “bad” quality control status
+in any of the four Nextclade columns ("qc.missingData.status”,
+“qc.mixedSites.status”, “qc.frameShifts.status”, and “qc.stopCodons.status”).
+These excluded genomes are enriched for low-coverage, contaminated, or
+misassembled sequences.
 
@@
 
-Ambiguous nucleotide letters (i.e., not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
-Problematic bases in the alignments, which had two or more ambiguous letters or gaps within a distance of seven bases, were masked as missing data following the protocol in \href{https://github.com/roblanf/sarscov2phylo/}.
-Masked bases were ignored when attaching a sample genome to the ARG, but they were subsequently imputed.
-Further, 481 problematic sites flagged as prone to sequencing errors or highly homoplasic were excluded entirely (\href{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473}; \href{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
+Ambiguous nucleotide letters (i.e., not A, C, G, T, or a gap) in the pairwise
+genome alignments were encoded as N. Problematic bases in the alignments, which
+had two or more ambiguous letters or gaps within a distance of seven bases,
+were masked as missing data following the protocol in
+\href{https://github.com/roblanf/sarscov2phylo/}. Masked bases were ignored
+when attaching a sample genome to the ARG, but they were subsequently imputed.
+Further, 481 problematic sites flagged as prone to sequencing errors or highly
+homoplasic were excluded entirely
+(\href{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
+\href{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
 
@@
 
-Including samples with incorrectly reported collection dates (e.g., an Alpha sample purportedly collected in 2020 from the United States, before Alpha appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce artefacts that distorted the topology of ARGs inferred early on in the study.
+Including samples with incorrectly reported collection dates (e.g., an Alpha
+sample purportedly collected in 2020 from the United States, before Alpha
+appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
+artefacts that distorted the topology of ARGs inferred early on in the study.
 Hence, to exclude such potential “time travellers”, we employed two filters.
-The first was to remove any sequence with a collection date that pre-dated the time to the most recent common ancestor (tMRCA) of its corresponding clade.
-The tMRCA for each clade was obtained from a Nextstrain GISAID reference tree (downloaded on August 22, 2022) where we used the lower bound of the 95\% confidence interval of each clade as the minimum date cut-off.
-The second filter was a simple threshold on the time delay between the collection date and submission date (maximum submission delay, or MSD).
-Indeed, we found that ARGs inferred from the MSD-filtered genome set had noticeably fewer branches with a high number of reversions.
-After some preliminary analysis, a MSD threshold value of 30 days was used when building the ARGs described herein.
-Future work could estimate collection dates by adding these potential time travellers back on the tree, allowing an automated assessment of collection date discrepancies.
+The first was to remove any sequence with a collection date that pre-dated the
+time to the most recent common ancestor (tMRCA) of its corresponding clade. The
+tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
+(downloaded on August 22, 2022) where we used the lower bound of the 95\%
+confidence interval of each clade as the minimum date cut-off. The second
+filter was a simple threshold on the time delay between the collection date and
+submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
+inferred from the MSD-filtered genome set had noticeably fewer branches with a
+high number of reversions. After some preliminary analysis, a MSD threshold
+value of 30 days was used when building the ARGs described herein. Future work
+could estimate collection dates by adding these potential time travellers back
+on the tree, allowing an automated assessment of collection date discrepancies.
 
@@
 
-We introduce an ARG inference method (schematic in Figure 1) that uses (1) an efficient HMM implementation of the Li and Stephens (LS) model \cite{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported collection dates of the SARS-CoV-2 samples to determine the order of genome copying; and (3) several parsimony-based heuristics to locally refine the topology.
-In brief, this combined approach uses the likelihood model of LS to attach sample genomes to existing nodes and to detect recombinants, but because many of the tip samples attached daily harbour few, if any, mutational changes, we use parsimony to resolve the relationships near the tip samples.
+We introduce an ARG inference method (schematic in Figure 1) that uses (1) an
+efficient HMM implementation of the Li and Stephens (LS) model \cite{Li2003-ib}
+to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported
+collection dates of the SARS-CoV-2 samples to determine the order of genome
+copying; and (3) several parsimony-based heuristics to locally refine the
+topology. In brief, this combined approach uses the likelihood model of LS to
+attach sample genomes to existing nodes and to detect recombinants, but because
+many of the tip samples attached daily harbour few, if any, mutational changes,
+we use parsimony to resolve the relationships near the tip samples.
 
-Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by taking advantage of the reported collection dates to incrementally grow the ARG.
-The inference process occurs forward in time, attaching new genomes aggregated by their collection date (daily batches), from the earliest to the last collection date.
-Prior to attaching sample genomes, the ARG is initialised such that the Wuhan-Hu-1/2019 reference genome is the root node.
-In the absence of recombination, the ARG is a single tree with no reticulation (Figure 1A); however, in the presence of recombination, a reticulate ARG is generated, with variation in tree topology across the genome (Figure 1B).
-Unlike standard phylogenetic trees in which internal nodes represent inferred ancestors, internal nodes in the ARGs inferred using sc2ts represent ancestors which can be samples or inferred ancestors.
+Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by
+taking advantage of the reported collection dates to incrementally grow the
+ARG. The inference process occurs forward in time, attaching new genomes
+aggregated by their collection date (daily batches), from the earliest to the
+last collection date. Prior to attaching sample genomes, the ARG is initialised
+such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence
+of recombination, the ARG is a single tree with no reticulation (Figure 1A);
+however, in the presence of recombination, a reticulate ARG is generated, with
+variation in tree topology across the genome (Figure 1B). Unlike standard
+phylogenetic trees in which internal nodes represent inferred ancestors,
+internal nodes in the ARGs inferred using sc2ts represent ancestors which can
+be samples or inferred ancestors.
 
-Each daily batch of genomes is first attached to the growing ARG without any phylogenetic resolution.
-In cases where the newly added sequence is inferred to be a recombinant, a recombination node is inserted; the “date” of this node is arbitrarily set to the average of the collection dates of the genome and the most recent of its parent nodes.
-Otherwise, the genomes are attached to the most closely related ancestral node (an attachment node).
-This creates polytomies that are then locally resolved among the descendants of each attachment node using the Unweighted Pair Group Method with Arithmetic Mean (UPGMA) algorithm \cite{Michener1957-tr}, with average linkage (Figure 1C).
-This hierarchical clustering algorithm produces a dendrogram relating a set of sequences, which is treated as an ultrametric binary tree here.
-Subsequently, mutations are placed onto the daily tree by maximum parsimony.
-An issue with the UPGMA algorithm is that it may create internal nodes supported by no informative site (i.e., having no mutation immediately ancestral to them).
-In an adjustment step, such unsupported internal nodes are removed to “un-resolve” the daily tree; the remaining internal nodes, which are supported by at least one shared mutation, are retained.
-The “dates” of the non-sample internal nodes are set to the same date as the samples in the daily batch.
-The resulting daily tree is grafted onto the attachment node in the global ARG.
+Each daily batch of genomes is first attached to the growing ARG without any
+phylogenetic resolution. In cases where the newly added sequence is inferred to
+be a recombinant, a recombination node is inserted; the “date” of this node is
+arbitrarily set to the average of the collection dates of the genome and the
+most recent of its parent nodes. Otherwise, the genomes are attached to the
+most closely related ancestral node (an attachment node). This creates
+polytomies that are then locally resolved among the descendants of each
+attachment node using the Unweighted Pair Group Method with Arithmetic Mean
+(UPGMA) algorithm \cite{Michener1957-tr}, with average linkage (Figure 1C).
+This hierarchical clustering algorithm produces a dendrogram relating a set of
+sequences, which is treated as an ultrametric binary tree here. Subsequently,
+mutations are placed onto the daily tree by maximum parsimony. An issue with
+the UPGMA algorithm is that it may create internal nodes supported by no
+informative site (i.e., having no mutation immediately ancestral to them). In
+an adjustment step, such unsupported internal nodes are removed to “un-resolve”
+the daily tree; the remaining internal nodes, which are supported by at least
+one shared mutation, are retained. The “dates” of the non-sample internal nodes
+are set to the same date as the samples in the daily batch. The resulting daily
+tree is grafted onto the attachment node in the global ARG.
 
-During ARG inference, two types of non-sample nodes are inserted to improve parsimonious placement of mutations: mutation-coalescing nodes and pushed-reversion nodes.
-We call these operations tree adjustments.
-A mutation-collapsing node is inserted when a shared mutation between sibling nodes is found (Figure 1D).
-An attempt to create mutation-collapsing nodes is made after grafting a new daily tree onto the global ARG.
-If shared mutations exist between two or more sibling nodes, then a parsimonious explanation is to create a new node that is the parent of those siblings, on which these mutations occurred.
-A pushed-reversion node is inserted when an immediate reversion occurs (Figure 1E), i.e.
-a mutation that reverts another mutation on its parent node.
-Instead of retaining both the mutation and immediate reversion, only the mutation is kept after a pushed-reversion node is inserted, thereby reducing the number of mutations on the branches.
-Note that these non-sample nodes are used as reconstructed genomes during HMM copying.
+During ARG inference, two types of non-sample nodes are inserted to improve
+parsimonious placement of mutations: mutation-coalescing nodes and
+pushed-reversion nodes. We call these operations tree adjustments. A
+mutation-collapsing node is inserted when a shared mutation between sibling
+nodes is found (Figure 1D). An attempt to create mutation-collapsing nodes is
+made after grafting a new daily tree onto the global ARG. If shared mutations
+exist between two or more sibling nodes, then a parsimonious explanation is to
+create a new node that is the parent of those siblings, on which these
+mutations occurred. A pushed-reversion node is inserted when an immediate
+reversion occurs (Figure 1E), i.e. a mutation that reverts another mutation on
+its parent node. Instead of retaining both the mutation and immediate
+reversion, only the mutation is kept after a pushed-reversion node is inserted,
+thereby reducing the number of mutations on the branches. Note that these
+non-sample nodes are used as reconstructed genomes during HMM copying.
 
-Furthermore, during inference, the ARG is automatically encoded, stored, and updated as a tree sequence using tskit \cite{Kelleher2018-xc}.
+Furthermore, during inference, the ARG is automatically encoded, stored, and
+updated as a tree sequence using tskit \cite{Kelleher2018-xc}.
 
@@
 
-The LS model is a convenient probabilistic approximation to the coalescent with recombination \cite{Hudson1983-if}, itself a generative model of a genealogical process with recombination \cite{Li2003-ib}.
-Importantly, it can be implemented as an HMM in which a focal genome is a sequence of nucleotides (observed states) that are probabilistically emitted as an imperfect mosaic of a set of reference genomes (hidden states).
-As such, an efficient HMM machinery can be employed to make statements about parameters of interest and highlight potential sources of error in genotyping and sequencing data.
-The LS model is used in a wide variety of applications in genomics.
-For example, it is the engine that drives statistical genotype phasing and imputation \cite{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it can be used to estimate mutation rates (e.g., REF), recombination rates (e.g., \cite{Hinch2011-tz}), and selection within and across hosts in viral sequence data (e.g., \cite{Palmer2019-wa}).
+The LS model is a convenient probabilistic approximation to the coalescent with
+recombination \cite{Hudson1983-if}, itself a generative model of a genealogical
+process with recombination \cite{Li2003-ib}. Importantly, it can be implemented
+as an HMM in which a focal genome is a sequence of nucleotides (observed
+states) that are probabilistically emitted as an imperfect mosaic of a set of
+reference genomes (hidden states). As such, an efficient HMM machinery can be
+employed to make statements about parameters of interest and highlight
+potential sources of error in genotyping and sequencing data. The LS model is
+used in a wide variety of applications in genomics. For example, it is the
+engine that drives statistical genotype phasing and imputation
+\cite{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it
+can be used to estimate mutation rates (e.g., REF), recombination rates (e.g.,
+\cite{Hinch2011-tz}), and selection within and across hosts in viral sequence
+data (e.g., \cite{Palmer2019-wa}).
 
-The generative process of the LS model is summarised in Figure 2.
-Here, a transition matrix, Q, governs the process of switching (recombining) between members of the reference panel (the hidden states).
-An emission matrix, \textit{E}, allows for differences between the focal sequence and the hidden state from which it is copied (due to mutation as well as genotyping or sequencing error).
-Then, we use the Viterbi algorithm \cite{Viterbi1967-ol} to find the most likely copying path through the sequences, given a set of reference sequences, \textit{Q}, and \textit{E}.
+The generative process of the LS model is summarised in Figure 2. Here, a
+transition matrix, Q, governs the process of switching (recombining) between
+members of the reference panel (the hidden states). An emission matrix,
+\textit{E}, allows for differences between the focal sequence and the hidden
+state from which it is copied (due to mutation as well as genotyping or
+sequencing error). Then, we use the Viterbi algorithm \cite{Viterbi1967-ol} to
+find the most likely copying path through the sequences, given a set of
+reference sequences, \textit{Q}, and \textit{E}.
 
-Because the HMM is memoryless, the potential parental lineages are independently assessed at each recombinantion breakpoint.
-This can result in the inference of a recombinant lineage descending immediately from three (or even more) parental lineages.
-We find that often two of the three parents are identical or similar in sequence in the segments contributed to the recombinant lineage.
-A more parsimonious explanation would be to switch back to the original parent sequence.
-In the future, a pragmatic solution would be to detect such cases during the ARG inference stage and to adjust the copying paths accordingly.
-Meanwhile, however, we urge caution when interpreting the parentage of lineages with multiple breakpoints.
+Because the HMM is memoryless, the potential parental lineages are
+independently assessed at each recombinantion breakpoint. This can result in
+the inference of a recombinant lineage descending immediately from three (or
+even more) parental lineages. We find that often two of the three parents are
+identical or similar in sequence in the segments contributed to the recombinant
+lineage. A more parsimonious explanation would be to switch back to the
+original parent sequence. In the future, a pragmatic solution would be to
+detect such cases during the ARG inference stage and to adjust the copying
+paths accordingly. Meanwhile, however, we urge caution when interpreting the
+parentage of lineages with multiple breakpoints.
 
-\subsubsection{Parameterization of the HMM implementation of Li and Stephens model}
+\subsubsection{Parameterization of the HMM implementation of Li and Stephens
+model}
 
-Recombination is inferred using an efficient HMM implementation of the LS model.
-Copying paths are estimated from SARS-CoV-2 genomes sampled over time, with younger genomes copying (i.e.
-inheriting segments and mutations) from older genomes.
-Here, we were able to apply the LS model on millions of SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of the LS model that is a core component of the tsinfer software.
-For this application, we have parameterised the HMM using a mismatch ratio (MMR), which tunes how strongly mutations (i.e., no switching of template sequences) are preferred over recombination (i.e., switching between sequences) during copying.
-Under a lower MMR value, the maximum-likelihood HMM copying path of a focal genome more likely spans at least two ancestral genomes (i.e,.
-the focal genome is a putative recombinant) than a single ancestral genome.
-Here, the MMR was set to three in all the HMM runs (see Discussion for future potential development of the HMM).
+Recombination is inferred using an efficient HMM implementation of the LS
+model. Copying paths are estimated from SARS-CoV-2 genomes sampled over time,
+with younger genomes copying (i.e. inheriting segments and mutations) from
+older genomes. Here, we were able to apply the LS model on millions of
+SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of
+the LS model that is a core component of the tsinfer software. For this
+application, we have parameterised the HMM using a mismatch ratio (MMR), which
+tunes how strongly mutations (i.e., no switching of template sequences) are
+preferred over recombination (i.e., switching between sequences) during
+copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
+focal genome more likely spans at least two ancestral genomes (i.e,. the focal
+genome is a putative recombinant) than a single ancestral genome. Here, the MMR
+was set to three in all the HMM runs (see Discussion for future potential
+development of the HMM).
 
@@
 
-Non-sample nodes inserted during ARG inference had no Pango lineage assignment from Nextclade.
-We imputed the Pango lineage of non-sample nodes using a two-step procedure (Figure 3) to help assess how consistently breakpoint intervals are estimated by the HMM runs (see below).
-If the Pango lineage of the parental node was available and no mutation was immediately ancestral to the non-sample node, then the parental node’s status was directly propagated to the non-sample node.
-Otherwise, the Pango lineage of the non-sample node was inferred by matching its full set of mutations against the Pango lineage-defining mutations (based on 90\% consensus of the sequences analysed) from the COVID-CG website \cite{Chen2021-zc}; \href{https://covidcg.org/}; accessed on November 04, 2022).
+Non-sample nodes inserted during ARG inference had no Pango lineage assignment
+from Nextclade. We imputed the Pango lineage of non-sample nodes using a
+two-step procedure (Figure 3) to help assess how consistently breakpoint
+intervals are estimated by the HMM runs (see below). If the Pango lineage of
+the parental node was available and no mutation was immediately ancestral to
+the non-sample node, then the parental node’s status was directly propagated to
+the non-sample node. Otherwise, the Pango lineage of the non-sample node was
+inferred by matching its full set of mutations against the Pango
+lineage-defining mutations (based on 90\% consensus of the sequences analysed)
+from the COVID-CG website \cite{Chen2021-zc}; \href{https://covidcg.org/};
+accessed on November 04, 2022).
 
@@
 
-When recombinant ancestry for a given genome is inferred by the LS HMM, a breakpoint is identified as the last possible position where the copying path must switch.
-Thus, the breakpoints used in the ARG represent the rightmost end of the interval, within which the recombination event could have occurred.
-To obtain the leftmost end of this interval, we rerun the HMM using mirrored coordinates (i.e., from the 3’ end to the 5’ end).
+When recombinant ancestry for a given genome is inferred by the LS HMM, a
+breakpoint is identified as the last possible position where the copying path
+must switch. Thus, the breakpoints used in the ARG represent the rightmost end
+of the interval, within which the recombination event could have occurred. To
+obtain the leftmost end of this interval, we rerun the HMM using mirrored
+coordinates (i.e., from the 3’ end to the 5’ end).
 
-A recombinant sequence was deemed to be HMM-consistent if (1) the same number of parent nodes and (2) the same set of mutations immediately ancestral to it were obtained from both the unmirrored and mirrored HMM runs.
-In the ideal case, the copying paths of both the HMM runs switch between parent sequences at the same position(s); in less ideal cases, however, the copying paths switch at different positions, typically because of a lack of genetic variation between the positions.
-The switch positions in the copying paths of both the HMM runs were used to define the breakpoint interval(s) of an HMM-consistent recombinant.
+A recombinant sequence was deemed to be HMM-consistent if (1) the same number
+of parent nodes and (2) the same set of mutations immediately ancestral to it
+were obtained from both the unmirrored and mirrored HMM runs. In the ideal
+case, the copying paths of both the HMM runs switch between parent sequences at
+the same position(s); in less ideal cases, however, the copying paths switch at
+different positions, typically because of a lack of genetic variation between
+the positions. The switch positions in the copying paths of both the HMM runs
+were used to define the breakpoint interval(s) of an HMM-consistent
+recombinant.
 
@@
 
-The source code implementing the sc2ts pipeline is available via GitHub \href{https://github.com/jeromekelleher/sc2ts/}.
-The sc2ts pipeline uses various functions and algorithms implemented in tskit and tsinfer \href{https://github.com/tskit-dev/}.
-The pipeline uses the SciPy implementation of the UPGMA algorithm to build daily trees \href{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}.
-The Jupyter notebooks used to produce the results described herein are also available on GitHub \href{https://github.com/jeromekelleher/sc2ts-paper/}.
+The source code implementing the sc2ts pipeline is available via GitHub
+\href{https://github.com/jeromekelleher/sc2ts/}. The sc2ts pipeline uses
+various functions and algorithms implemented in tskit and tsinfer
+\href{https://github.com/tskit-dev/}. The pipeline uses the SciPy
+implementation of the UPGMA algorithm to build daily trees
+\href{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}.
+The Jupyter notebooks used to produce the results described herein are also
+available on GitHub \href{https://github.com/jeromekelleher/sc2ts-paper/}.
 
@@
 To illustrate the use of sc2ts, we employed two different sampling strategies.
-First, we inferred a densely sampled but time-limited “wide” ARG by attaching all the sample genomes with an MSD of 30 days, one daily batch at a time.
-The inference procedure ran until the genomes sampled on June 30, 2021 were attached to the ARG, resulting in a total of 1.27 million samples.
-Second, we inferred a “long” ARG over a more extended timespan, by attaching at most 1,000 genomes, which were randomly subsampled from each daily batch with an MSD within 30 days.
-For this ARG, the inference procedure ran until the genomes sampled on June 30, 2022 were attached to the ARG.
-The long ARG covers an extra year of samples but is more sparsely sampled.
-June 30, 2022 is the collection date of the final batch of sequences which were preprocessed for the development of sc2ts.
-Once sc2ts is further developed and tested, we will attach samples collected after that date to infer a fully sampled pandemic-scale ARG.
+First, we inferred a densely sampled but time-limited “wide” ARG by attaching
+all the sample genomes with an MSD of 30 days, one daily batch at a time. The
+inference procedure ran until the genomes sampled on June 30, 2021 were
+attached to the ARG, resulting in a total of 1.27 million samples. Second, we
+inferred a “long” ARG over a more extended timespan, by attaching at most 1,000
+genomes, which were randomly subsampled from each daily batch with an MSD
+within 30 days. For this ARG, the inference procedure ran until the genomes
+sampled on June 30, 2022 were attached to the ARG. The long ARG covers an extra
+year of samples but is more sparsely sampled. June 30, 2022 is the collection
+date of the final batch of sequences which were preprocessed for the
+development of sc2ts. Once sc2ts is further developed and tested, we will
+attach samples collected after that date to infer a fully sampled
+pandemic-scale ARG.
 
-Most of the nodes in the ARGs represent sample genomes (Table 1).
-Only a small fraction of the nodes represent recombination events (0.28\% and 0.27\% in the wide ARG and long ARG, respectively), over half of which have no descendant samples (besides the recombinant sequence that triggered insertion of the recombination node) (Table 1).
-There are over one million mutations on the ARGs, the majority of which occur on a terminal edge (Table 1).
-Notably, the massive amount of genomic variation data in the wide ARG (1.27 million genomes) and long ARG (657 thousand genomes) are compactly stored in ~57 Mb and ~36 Mb, respectively (less than ten bytes per sample).
-These ARGs can be loaded for analysis in a Jupyter notebook on a typical commercial laptop in merely a fraction of a second using the Python package tskit (Table 1).
+Most of the nodes in the ARGs represent sample genomes (Table 1). Only a small
+fraction of the nodes represent recombination events (0.28\% and 0.27\% in the
+wide ARG and long ARG, respectively), over half of which have no descendant
+samples (besides the recombinant sequence that triggered insertion of the
+recombination node) (Table 1). There are over one million mutations on the
+ARGs, the majority of which occur on a terminal edge (Table 1). Notably, the
+massive amount of genomic variation data in the wide ARG (1.27 million genomes)
+and long ARG (657 thousand genomes) are compactly stored in ~57 Mb and ~36 Mb,
+respectively (less than ten bytes per sample). These ARGs can be loaded for
+analysis in a Jupyter notebook on a typical commercial laptop in merely a
+fraction of a second using the Python package tskit (Table 1).
 
@@
 
-The wide ARG (with over one million sample genomes) was built using 128 threads on a machine with 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz).
-Inference ran for 17 days to attach genomes sampled up to June 30, 2021.
-The long ARG (with ~700 thousand sample genomes) was built using 40 threads on a machine with 256 GB RAM (2x Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz).
-Inference ran for 23 days to attach genomes sampled up to June 30, 2022.
-Most of the runtime was spent on running the HMM.
+The wide ARG (with over one million sample genomes) was built using 128 threads
+on a machine with 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). Inference ran for 17
+days to attach genomes sampled up to June 30, 2021. The long ARG (with ~700
+thousand sample genomes) was built using 40 threads on a machine with 256 GB
+RAM (2x Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz). Inference ran for 23 days
+to attach genomes sampled up to June 30, 2022. Most of the runtime was spent on
+running the HMM.
 
@@
 
-For validation purposes, we compare the genetic relationships in the sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain project \cite{Hadfield2018-ef}.
-The NextStrain trees consist of subsamples of SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely restricted to non-recombinant lineages.
-To allow for comparison, the ARGs were "simplified" \cite{Kelleher2018-xc} to retain only those tips of our genealogy that match the samples present in a specific NextStrain tree.
-These simplified "backbone" ARGs exhibit very little recombination (6 and 5 trees across the genome for the wide and long ARGs, respectively).
+For validation purposes, we compare the genetic relationships in the
+sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain
+project \cite{Hadfield2018-ef}. The NextStrain trees consist of subsamples of
+SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely
+restricted to non-recombinant lineages. To allow for comparison, the ARGs were
+"simplified" \cite{Kelleher2018-xc} to retain only those tips of our genealogy
+that match the samples present in a specific NextStrain tree. These simplified
+"backbone" ARGs exhibit very little recombination (6 and 5 trees across the
+genome for the wide and long ARGs, respectively).
 
-Figure 4 shows a co-phylogeny of the 88 samples shared between the wide ARG and a GISAID global "all-time" tree from NextStrain (downloaded on January 21, 2023).
-To reduce the ARG to a single tree, we show the very first tree in the wide ARG; however, the five other trees in this simplified ARG show near-identical patterns.
-Supplementary Figure 3 shows the same comparison for the long ARG.
-It is clear that the backbone topology of the sc2ts tree shows very close agreement with the NextStrain tree.
-The sample genomes cluster by their assigned Pango lineage status, and many variants and their descendents form identical monophyletic clades in both the trees (e.g., the Alpha and Delta clades).
+Figure 4 shows a co-phylogeny of the 88 samples shared between the wide ARG and
+a GISAID global "all-time" tree from NextStrain (downloaded on January 21,
+2023). To reduce the ARG to a single tree, we show the very first tree in the
+wide ARG; however, the five other trees in this simplified ARG show
+near-identical patterns. Supplementary Figure 3 shows the same comparison for
+the long ARG. It is clear that the backbone topology of the sc2ts tree shows
+very close agreement with the NextStrain tree. The sample genomes cluster by
+their assigned Pango lineage status, and many variants and their descendents
+form identical monophyletic clades in both the trees (e.g., the Alpha and Delta
+clades).
 
-There are, however, some notable differences between the trees.
-Firstly, the sc2ts tree is generally less well resolved (noticeably, February to March, 2020) (Figure 4).
-Secondly, there are non-identical sample partitions near the tips (e.g., in the Alpha clade, Figure 4).
-These might have arisen because the samples containing the mutations needed to resolve local topologies were collected later than the cut-off date for the sc2ts ARG, but were for the NextStrain tree.
-Finally, the branch lengths differ.
-For example, the branch leading to the ancestral node of the Delta variant samples is much shorter in the sc2ts tree than the NextStrain tree (Figure 4).
-It is possible that the length of some of the branches was poorly calibrated, probably caused by overly aggressive insertion of mutation-collapsing nodes and reversion-pushed nodes or suboptimal sample attachment.
-Side-by-side comparison of sc2ts trees with trees inferred using another method (e.g., UShER) should help to understand whether branch length differences are due to poor calibration or methodological differences.
+There are, however, some notable differences between the trees. Firstly, the
+sc2ts tree is generally less well resolved (noticeably, February to March,
+2020) (Figure 4). Secondly, there are non-identical sample partitions near the
+tips (e.g., in the Alpha clade, Figure 4). These might have arisen because the
+samples containing the mutations needed to resolve local topologies were
+collected later than the cut-off date for the sc2ts ARG, but were for the
+NextStrain tree. Finally, the branch lengths differ. For example, the branch
+leading to the ancestral node of the Delta variant samples is much shorter in
+the sc2ts tree than the NextStrain tree (Figure 4). It is possible that the
+length of some of the branches was poorly calibrated, probably caused by overly
+aggressive insertion of mutation-collapsing nodes and reversion-pushed nodes or
+suboptimal sample attachment. Side-by-side comparison of sc2ts trees with trees
+inferred using another method (e.g., UShER) should help to understand whether
+branch length differences are due to poor calibration or methodological
+differences.
 
-Although here we focus on a single local tree at an arbitrarily chosen position, note that in the original “un-simplified” ARGs, there are 1,496 and 958 trees, respectively (Table 1).
-Hence, while for a relatively small sample the genealogy may be well described by a single tree, the full dataset shows substantial variation in phylogenetic relationships (or phylogenetic incongruence) across the SARS-CoV-2 genome, much of which may be attributable to recombination.
-Such complex reticulate relationships are appropriately captured by a genealogy in the form of an ARG.
+Although here we focus on a single local tree at an arbitrarily chosen
+position, note that in the original “un-simplified” ARGs, there are 1,496 and
+958 trees, respectively (Table 1). Hence, while for a relatively small sample
+the genealogy may be well described by a single tree, the full dataset shows
+substantial variation in phylogenetic relationships (or phylogenetic
+incongruence) across the SARS-CoV-2 genome, much of which may be attributable
+to recombination. Such complex reticulate relationships are appropriately
+captured by a genealogy in the form of an ARG.
 
@@
 
-Both the wide and long ARGs contain a large number of recombination nodes (4,123 and 2,078, respectively; Supplementary Tables 1 and 2), with each of these nodes inferred as a result of matching a specific sample (a recombinant) into the ARG.
-However, ~65\% of recombination nodes are ancestors of only a single descendant sample in each local tree; these could have arisen from sequencing of mixed samples (e.g., from co-infection or contamination), or were simply not transmitted onwards and so were excluded from subsequent analyses (leaving X and Y recombinants in the wide and long ARGs, respectively).
-In their analysis using RIPPLES, Turakhia et al.
-(2022) detected 589 recombinants, also excluding singletons.
+Both the wide and long ARGs contain a large number of recombination nodes
+(4,123 and 2,078, respectively; Supplementary Tables 1 and 2), with each of
+these nodes inferred as a result of matching a specific sample (a recombinant)
+into the ARG. However, ~65\% of recombination nodes are ancestors of only a
+single descendant sample in each local tree; these could have arisen from
+sequencing of mixed samples (e.g., from co-infection or contamination), or were
+simply not transmitted onwards and so were excluded from subsequent analyses
+(leaving X and Y recombinants in the wide and long ARGs, respectively). In
+their analysis using RIPPLES, Turakhia et al. (2022) detected 589 recombinants,
+also excluding singletons.
 
-For simplicity, we focus primarily on “non-complex” recombination nodes.
-To do this, we apply a filtering step that removes those with more than one breakpoint (and often more than two parent lineages), or where matching is not HMM-consistent or identifies parents with different Pango lineage assignments in the original (i.e., 5’ to 3’ end) and mirrored (i.e., 3’ to 5’ end) HMM runs.
-Filtering reduces both the number of recombination nodes and breakpoints from  A \& B to C in the wide ARG and from 726 & 806 to 459 in the long ARG (Supplementary Tables 1 and 2).
-Note that in the unfiltered case, a single recombination node can correspond to multiple points, one for each recombination breakpoint.
-The effect of this filtering step in the long ARG, which includes time periods of increased rates of coinfection and recombination compared to the wide ARG, can be seen in Figure 3A (unfiltered) and Figure 3B (filtered).
+For simplicity, we focus primarily on “non-complex” recombination nodes. To do
+this, we apply a filtering step that removes those with more than one
+breakpoint (and often more than two parent lineages), or where matching is not
+HMM-consistent or identifies parents with different Pango lineage assignments
+in the original (i.e., 5’ to 3’ end) and mirrored (i.e., 3’ to 5’ end) HMM
+runs. Filtering reduces both the number of recombination nodes and breakpoints
+from  A \& B to C in the wide ARG and from 726 & 806 to 459 in the long ARG
+(Supplementary Tables 1 and 2). Note that in the unfiltered case, a single
+recombination node can correspond to multiple points, one for each
+recombination breakpoint. The effect of this filtering step in the long ARG,
+which includes time periods of increased rates of coinfection and recombination
+compared to the wide ARG, can be seen in Figure 3A (unfiltered) and Figure 3B
+(filtered).
 
-Importantly, the long ARG captured all except one of the Pango designated recombinant lineages (see the section below), but we also detected a large number of additional recombination events, with parents lineages that diverged as far back as two years to as recent as five or six days prior to the recombination event (Figure 3).
-This provides a degree of resolution of viral recombinants not available with other recombination detection methods.
-The majority of newly identified recombinants arise from parents that diverged 10 to 30 weeks previously.
-However, there is also a peak of recombinants that diverged ~95 weeks prior to the event: these correspond to recombinants whose parental common ancestor traces to early 2020, including both Delta-Delta and Delta-Omicron recombinants (Supplementary Figure 4).
+Importantly, the long ARG captured all except one of the Pango designated
+recombinant lineages (see the section below), but we also detected a large
+number of additional recombination events, with parents lineages that diverged
+as far back as two years to as recent as five or six days prior to the
+recombination event (Figure 3). This provides a degree of resolution of viral
+recombinants not available with other recombination detection methods. The
+majority of newly identified recombinants arise from parents that diverged 10
+to 30 weeks previously. However, there is also a peak of recombinants that
+diverged ~95 weeks prior to the event: these correspond to recombinants whose
+parental common ancestor traces to early 2020, including both Delta-Delta and
+Delta-Omicron recombinants (Supplementary Figure 4).
 
-The date of common ancestry for the two parents of recombinant lineages is concentrated in several banded rows in Figure 5.
-These are largely due to a small handful of most recent common ancestors (MRCAs), which are shared between a large number of recombination events, the top four of which are indicated by dotted lines in the figure.
-These MRCAs are near the root of large expansions of certain lineages (B.1: original variant;  B.1.617.2: Delta, BA.1: first Omicron wave; and BA.2: second Omicron wave), as evidenced by the number of samples with that Pango designation that trace back to these ancestral nodes (listed as percentages on the plot).
-The specific MRCA nodes to which many recombinants trace frequently represent very large polytomies in the tree (at the extreme, 2,872 immediate children).
-This is likely to indicate a rapid and under-sampled expansion of a clade in the SARS-CoV-2 genealogy.
+The date of common ancestry for the two parents of recombinant lineages is
+concentrated in several banded rows in Figure 5. These are largely due to a
+small handful of most recent common ancestors (MRCAs), which are shared between
+a large number of recombination events, the top four of which are indicated by
+dotted lines in the figure. These MRCAs are near the root of large expansions
+of certain lineages (B.1: original variant;  B.1.617.2: Delta, BA.1: first
+Omicron wave; and BA.2: second Omicron wave), as evidenced by the number of
+samples with that Pango designation that trace back to these ancestral nodes
+(listed as percentages on the plot). The specific MRCA nodes to which many
+recombinants trace frequently represent very large polytomies in the tree (at
+the extreme, 2,872 immediate children). This is likely to indicate a rapid and
+under-sampled expansion of a clade in the SARS-CoV-2 genealogy.
 
-Finally, it is clear from Supplementary Figure 4 that of the recombinant parents that can be classified into Alpha, Delta, and Omicron variants, the majority are within-Delta and within-Omicron recombinants, and that Omicron followed by Delta, are the variants associated with the most recombination.
-This may reflect either sampling intensity, the prevalence of cases (which increases the chance of coinfection and recombination), or possibly heterogeneity in recombination rates among lineages.
+Finally, it is clear from Supplementary Figure 4 that of the recombinant
+parents that can be classified into Alpha, Delta, and Omicron variants, the
+majority are within-Delta and within-Omicron recombinants, and that Omicron
+followed by Delta, are the variants associated with the most recombination.
+This may reflect either sampling intensity, the prevalence of cases (which
+increases the chance of coinfection and recombination), or possibly
+heterogeneity in recombination rates among lineages.
 
@@
 
-By searching for samples combining genomic segments from Alpha (B.1.1.7) and from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations, Jackson et al.
-(Jackson et al.
-2021) reported 16 recombinant sequences among genomes sampled in the United Kingdom up to March 7, 2021.
-The authors proposed that the recombinants likely arose from eight independent recombination events (categorised into groups A to D and four singletons).
+By searching for samples combining genomic segments from Alpha (B.1.1.7) and
+from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations,
+Jackson et al. (Jackson et al. 2021) reported 16 recombinant sequences among
+genomes sampled in the United Kingdom up to March 7, 2021. The authors proposed
+that the recombinants likely arose from eight independent recombination events
+(categorised into groups A to D and four singletons).
 
-In the wide ARG (sampled to mid-2021), all these sequences are descendants of recombination nodes, except MILK-103C712, which was removed during preprocessing (Table 2); additionally, all are HMM-consistent, except QEUH-1067DEF.
-Overall, we find excellent agreement between Jackson et al.’s motif-based method and sc2ts in terms of (1) the Pango lineage of the parents and (2) the breakpoint intervals (Table 2).
-In all cases, the two methods proposed identical or very closely related parent lineages.
-The general mosaic genome structure is concordant in all the cases, even CAMC-CB7AB3 where there are two switches in the HMM copying paths (indicative of two breakpoints).
-The breakpoint intervals estimated using the two methods overlap substantially in all the cases (Table 2).
-Within each of the groups A to D, the constituent sequences have identical breakpoint intervals, because they trace back to a single recombination node in the wide ARG.
-This result is consistent with the hypothesis that the sequences of the groups A to D emerged from four separate recombination events (Jackson et al.
-2021).
+In the wide ARG (sampled to mid-2021), all these sequences are descendants of
+recombination nodes, except MILK-103C712, which was removed during
+preprocessing (Table 2); additionally, all are HMM-consistent, except
+QEUH-1067DEF. Overall, we find excellent agreement between Jackson et al.’s
+motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
+and (2) the breakpoint intervals (Table 2). In all cases, the two methods
+proposed identical or very closely related parent lineages. The general mosaic
+genome structure is concordant in all the cases, even CAMC-CB7AB3 where there
+are two switches in the HMM copying paths (indicative of two breakpoints). The
+breakpoint intervals estimated using the two methods overlap substantially in
+all the cases (Table 2). Within each of the groups A to D, the constituent
+sequences have identical breakpoint intervals, because they trace back to a
+single recombination node in the wide ARG. This result is consistent with the
+hypothesis that the sequences of the groups A to D emerged from four separate
+recombination events (Jackson et al. 2021).
 
@@
 
-The wide ARG only contains samples from three Pango-designated recombinant lineages (XA, XB, and XC), due to the fact that it only includes samples up to mid-2021.
-By contrast, the long ARG (subsampled to mid-2022) contains samples from 33 Pango-designated recombinant lineages (XA to XAK; Supplementary Table 3).
-For Pango recombinant analysis, we therefore focus on the long ARG.
+The wide ARG only contains samples from three Pango-designated recombinant
+lineages (XA, XB, and XC), due to the fact that it only includes samples up to
+mid-2021. By contrast, the long ARG (subsampled to mid-2022) contains samples
+from 33 Pango-designated recombinant lineages (XA to XAK; Supplementary Table
+3). For Pango recombinant analysis, we therefore focus on the long ARG.
 
-In the wide ARG, every sequence of the three Pango recombinants trace back to some recombination node. XA, XB, and XC.
+In the wide ARG, every sequence of the three Pango recombinants trace back to
+some recombination node. XA, XB, and XC.
 
-In the unfiltered long ARG, the sequences of 28 Pango recombinants trace back to some recombinant node(s); however, the sequences of the five Pango recombinants XP, XU, XAA, XAG, and XAK do not.
-A likely explanation for the XP recombinant sequences not having a recombinant origin in the long ARG is that its characteristic multibase deletion (https://github.com/cov-lineages/pango-designation/issues/481) was masked during preprocessing and therefore was not incorporated into the HMM.
-Upon tracing the origin of XAG through the ARG, XAG might have been caused by incorrect imputation of its parent nodes(?).
-XAK likely does not appear to have a recombinant origin because of random subsampling, but it may with a more complete tree.
+In the unfiltered long ARG, the sequences of 28 Pango recombinants trace back
+to some recombinant node(s); however, the sequences of the five Pango
+recombinants XP, XU, XAA, XAG, and XAK do not. A likely explanation for the XP
+recombinant sequences not having a recombinant origin in the long ARG is that
+its characteristic multibase deletion
+(https://github.com/cov-lineages/pango-designation/issues/481) was masked
+during preprocessing and therefore was not incorporated into the HMM. Upon
+tracing the origin of XAG through the ARG, XAG might have been caused by
+incorrect imputation of its parent nodes(?). XAK likely does not appear to have
+a recombinant origin because of random subsampling, but it may with a more
+complete tree.
 
-The sequences of 14 Pango recombinants trace back to a single recombination node.
-The sequences of the remaining 18 Pango recombinants trace back to multiple recombination nodes, however.
-This may be a result of undersampling in the long ARG, which was built by randomly subsampling at most 1,000 sequences from each daily batch.
-Having extra recombination nodes in the long ARG appeared to be a more parsimonious explanation for the 19 Pango recombinants than having a single recombination node with many immediately ancestral mutations (including recurrent mutations and reversions), despite that the recombinants may indeed trace to a single origin.
+The sequences of 14 Pango recombinants trace back to a single recombination
+node. The sequences of the remaining 18 Pango recombinants trace back to
+multiple recombination nodes, however. This may be a result of undersampling in
+the long ARG, which was built by randomly subsampling at most 1,000 sequences
+from each daily batch. Having extra recombination nodes in the long ARG
+appeared to be a more parsimonious explanation for the 19 Pango recombinants
+than having a single recombination node with many immediately ancestral
+mutations (including recurrent mutations and reversions), despite that the
+recombinants may indeed trace to a single origin.
 
-The mosaic genome structure of all the 32 Pango recombinants is concordant with the structure determined by the Pango Lineage Designation Committee (Supplementary Table 3).
-For example, the genome of XD (also known as “Deltacron”) is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17 (Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.
+The mosaic genome structure of all the 32 Pango recombinants is concordant with
+the structure determined by the Pango Lineage Designation Committee
+(Supplementary Table 3). For example, the genome of XD (also known as
+“Deltacron”) is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
+(Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.
 
@@
 
-Next, we examine recombination nodes in the filtered long ARG (sampled to mid-2022) that represent previously unnamed putative recombinant sequences, which do not have a Pango recombinant designation.
-For this proof-of-concept study, we have arbitrarily picked 12 recombination nodes that seem plausible (Supplementary Table 4).
-Coincidentally, all these nodes involve Omicron subvariants (BA.1, BA.2, BA.4, and BA.5).
-These recombination nodes (1) were inserted into the long ARG on or after January 8, 2022; (2) have at least 10 descendent samples; and (3) have no mutational differences (including immediate reversions) from their parent nodes.
+Next, we examine recombination nodes in the filtered long ARG (sampled to
+mid-2022) that represent previously unnamed putative recombinant sequences,
+which do not have a Pango recombinant designation. For this proof-of-concept
+study, we have arbitrarily picked 12 recombination nodes that seem plausible
+(Supplementary Table 4). Coincidentally, all these nodes involve Omicron
+subvariants (BA.1, BA.2, BA.4, and BA.5). These recombination nodes (1) were
+inserted into the long ARG on or after January 8, 2022; (2) have at least 10
+descendent samples; and (3) have no mutational differences (including immediate
+reversions) from their parent nodes.
 
-TODO: Node  740761 (USA/NC-CDC-LC0668306/2022) is a sister clade proposed1006 in the UShER public phylogeny.
-Node 628656 (Scotland/QEUH-37794BB/2022) is a sister clade of XAC, and annotated as “miscBA2BA1PostSpike”.
+TODO: Node  740761 (USA/NC-CDC-LC0668306/2022) is a sister clade proposed1006
+in the UShER public phylogeny. Node 628656 (Scotland/QEUH-37794BB/2022) is a
+sister clade of XAC, and annotated as “miscBA2BA1PostSpike”.
 
@@
 
-We explore recombination signals across the SARS-CoV-2 genome by inspecting the breakpoint intervals of the filtered recombination nodes from the wide ARG (sampled to mid-2021) (Figure 4).
-The length of a breakpoint interval reflects uncertainty in the estimated location of a breakpoint.
-On average, the length of the breakpoint intervals is 2,895 bases (median, 1,670 bases; range, 1 to 23,666 bases).
-This broad distribution of interval length indicates that the location of a breakpoint often could not be precisely determined given the observed genetic variation.
+We explore recombination signals across the SARS-CoV-2 genome by inspecting the
+breakpoint intervals of the filtered recombination nodes from the wide ARG
+(sampled to mid-2021) (Figure 4). The length of a breakpoint interval reflects
+uncertainty in the estimated location of a breakpoint. On average, the length
+of the breakpoint intervals is 2,895 bases (median, 1,670 bases; range, 1 to
+23,666 bases). This broad distribution of interval length indicates that the
+location of a breakpoint often could not be precisely determined given the
+observed genetic variation.
 
-There is a preponderance of breakpoint intervals spanning the 3’ end of the genome (Figure 4A), as noted previously by Turakhia et al.
-\cite{Turakhia2022-it}.
-Inspecting more closely, we find an enrichment in the number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure 4B).
-Markedly, there is an abrupt drop in the number of intersecting intervals in the N gene following the peak in ORF8 towards the 3’ end.
+There is a preponderance of breakpoint intervals spanning the 3’ end of the
+genome (Figure 4A), as noted previously by Turakhia et al.
+\cite{Turakhia2022-it}. Inspecting more closely, we find an enrichment in the
+number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region,
+and 3’ untranslated region (Figure 4B). Markedly, there is an abrupt drop in
+the number of intersecting intervals in the N gene following the peak in ORF8
+towards the 3’ end.
 
-Next, we investigate the potential recombination signals alongside the number of mutations across the genome (Figure 4C).
-Regions with a higher density of mutations contain more genetic variation, increasing the power to detect recombination events.
-The position 28,271 (located in the ORF8-N intergenic region) has a particularly high number of mutations in the samples (Figure 5).
-At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions (n = 1,605) appear to occur frequently.
-One possible explanation is that this homopolymeric region is prone to sequencing errors, giving rise to an artifactual recombination signal.
-At the upstream position 27,972, numerous mutations also occur.
-This position has been reported to be hypermutable \cite{Jungreis2021-dh}.
+Next, we investigate the potential recombination signals alongside the number
+of mutations across the genome (Figure 4C). Regions with a higher density of
+mutations contain more genetic variation, increasing the power to detect
+recombination events. The position 28,271 (located in the ORF8-N intergenic
+region) has a particularly high number of mutations in the samples (Figure 5).
+At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions
+(n = 1,605) appear to occur frequently. One possible explanation is that this
+homopolymeric region is prone to sequencing errors, giving rise to an
+artifactual recombination signal. At the upstream position 27,972, numerous
+mutations also occur. This position has been reported to be hypermutable
+\cite{Jungreis2021-dh}.
 
-We further consider the breakpoint intervals within the context of recombination breakpoint sequence motifs.
-It is hypothesised that certain palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote template switching during replication in SARS-CoV-2 via formation of base-paired stem loops in the genome structure \cite{Gallaher2020-lb}.
-There are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019 reference sequence (Supplementary Table 5).
-We observe that XX (XX\%) of the breakpoint intervals of the HMM-consistent recombinants span at least one of the breakpoint sequences.
+We further consider the breakpoint intervals within the context of
+recombination breakpoint sequence motifs. It is hypothesised that certain
+palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote
+template switching during replication in SARS-CoV-2 via formation of
+base-paired stem loops in the genome structure \cite{Gallaher2020-lb}. There
+are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019
+reference sequence (Supplementary Table 5). We observe that XX (XX\%) of the
+breakpoint intervals of the HMM-consistent recombinants span at least one of
+the breakpoint sequences.
 
@@
 
-In addition to recombination events, the ARGs record mutation events along the SARS-CoV-2 phylogeny.
-Yi et al.
-\cite{Yi2021-sc} reconstructed a SARS-CoV-2 phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to January 12, 2021, and systematically catalogued all the mutations occurring along the phylogeny.
-Notably, in the mutational spectrum described by Yi et al., C-to-U and G-to-U mutations occur more frequently than U-to-C and U-to-G mutations, respectively.
+In addition to recombination events, the ARGs record mutation events along the
+SARS-CoV-2 phylogeny. Yi et al. \cite{Yi2021-sc} reconstructed a SARS-CoV-2
+phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to
+January 12, 2021, and systematically catalogued all the mutations occurring
+along the phylogeny. Notably, in the mutational spectrum described by Yi et
+al., C-to-U and G-to-U mutations occur more frequently than U-to-C and U-to-G
+mutations, respectively.
 
-For validation purposes, we compare the mutational spectrum from the wide ARG with the mutational spectrum of Yi et al.
-We categorised all of the single nucleotide mutations by type (defined by the ancestral state and derived state), excluded mutations inherited by only a single sample (which are enriched for sequencing errors and artefacts), and tallied them up by type.
+For validation purposes, we compare the mutational spectrum from the wide ARG
+with the mutational spectrum of Yi et al. We categorised all of the single
+nucleotide mutations by type (defined by the ancestral state and derived
+state), excluded mutations inherited by only a single sample (which are
+enriched for sequencing errors and artefacts), and tallied them up by type.
 Similarly, we took the data for single nucleotide mutations from Yi et al.
-\ref{https://github.com/ju-lab/SC2_evol_signature}, excluded mutations occurring along terminal branches, and tallied them up by type.
-We find that the mutational spectrum from the wide ARG (448,825 mutations) matches the mutational spectrum reported by Yi et al.
-(92,344 mutations) (Figure X).
-In both spectra, C-to-U mutations and G-to-U mutations occur more frequently than U-to-C and U-to-G, respectively (Figure X).
-Similar results are obtained when including the mutations inherited by only a single sample or occurring on terminal branches (data not shown).
+\ref{https://github.com/ju-lab/SC2_evol_signature}, excluded mutations
+occurring along terminal branches, and tallied them up by type. We find that
+the mutational spectrum from the wide ARG (448,825 mutations) matches the
+mutational spectrum reported by Yi et al. (92,344 mutations) (Figure X). In
+both spectra, C-to-U mutations and G-to-U mutations occur more frequently than
+U-to-C and U-to-G, respectively (Figure X). Similar results are obtained when
+including the mutations inherited by only a single sample or occurring on
+terminal branches (data not shown).
 
@@
 
-The COVID-19 pandemic continues to be a global emergency, with persistent high levels of infection.
-This high prevalence has allowed the proliferation of many variants, with more than 600 Pango-designated lineages circulating globally in the last three months (January to March, 2023; https://gisaid.org/; access on March 27, 2023).
-High prevalence also brings a higher risk of coinfection, increasing opportunities for new phenotypically distinct recombinants to emerge and spread.
+The COVID-19 pandemic continues to be a global emergency, with persistent high
+levels of infection. This high prevalence has allowed the proliferation of many
+variants, with more than 600 Pango-designated lineages circulating globally in
+the last three months (January to March, 2023; https://gisaid.org/; access on
+March 27, 2023). High prevalence also brings a higher risk of coinfection,
+increasing opportunities for new phenotypically distinct recombinants to emerge
+and spread.
 
-In this proof-of-concept study, we demonstrate the feasibility of reconstructing a large ARG from over one million SARS-CoV-2 genomes, a feat that is simply not possible using other available ARG inference methods (Rasmussen et al.
-2014; Ignatieva et al.
-2021; Speidel et al.
-2019).
-We have accomplished this feat by building on top of recent advances in storing and processing large ARGs (tskit) and inferring genealogical relationships efficiently at scale (tsinfer).
-The highly compact representation of the ARGs as succinct tree sequences, coupled with vectorised computing technologies, enables the large SARS-CoV-2 ARGs to be loaded in less than one second in a Jupyter notebook and allows for each SARS-CoV-2 genome to be stored in less than eleven bytes of data.
-We believe that further development of the sc2ts methodology and software can create a powerful tool (1) to efficiently detect emerging or previously overlooked recombinants in SARS-CoV-2; (2) to jointly model the mutational and recombinational processes shaping global SARS-CoV-2 genomic diversity; and (3) to perform a broad variety of evolutionary analyses of SARS-CoV-2 genomes with unprecedented global sequencing coverage.
+In this proof-of-concept study, we demonstrate the feasibility of
+reconstructing a large ARG from over one million SARS-CoV-2 genomes, a feat
+that is simply not possible using other available ARG inference methods
+(Rasmussen et al. 2014; Ignatieva et al. 2021; Speidel et al. 2019). We have
+accomplished this feat by building on top of recent advances in storing and
+processing large ARGs (tskit) and inferring genealogical relationships
+efficiently at scale (tsinfer). The highly compact representation of the ARGs
+as succinct tree sequences, coupled with vectorised computing technologies,
+enables the large SARS-CoV-2 ARGs to be loaded in less than one second in a
+Jupyter notebook and allows for each SARS-CoV-2 genome to be stored in less
+than eleven bytes of data. We believe that further development of the sc2ts
+methodology and software can create a powerful tool (1) to efficiently detect
+emerging or previously overlooked recombinants in SARS-CoV-2; (2) to jointly
+model the mutational and recombinational processes shaping global SARS-CoV-2
+genomic diversity; and (3) to perform a broad variety of evolutionary analyses
+of SARS-CoV-2 genomes with unprecedented global sequencing coverage.
 
-ARGs are rich genealogical structures that contain information about the genetic relationships among sampled genomes in the presence of recombination.
-Here, we have conducted several analyses of the preliminary ARGs to show that the ARGs built using sc2ts can (1) recapitulate the phylogenetic relationships among major SARS-CoV-2 lineages; (2) recover biologically plausible recombination signals in SARS-CoV-2 genomes; and (3) model mutation events in SARS-CoV-2 evolutionary history.
+ARGs are rich genealogical structures that contain information about the
+genetic relationships among sampled genomes in the presence of recombination.
+Here, we have conducted several analyses of the preliminary ARGs to show that
+the ARGs built using sc2ts can (1) recapitulate the phylogenetic relationships
+among major SARS-CoV-2 lineages; (2) recover biologically plausible
+recombination signals in SARS-CoV-2 genomes; and (3) model mutation events in
+SARS-CoV-2 evolutionary history.
 
-ARGs reconstructed using sc2ts capture the phylogenetic relationships of major SARS-CoV-2 lineages.
-Our results show that the backbone topology of a densely sampled wide ARG inferred using sc2ts agrees well with the backbone topology of a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference package.
-However, we have noted several areas of differences (e.g., poorly calibrated branch lengths), which are the focus of future development of sc2ts.
-We believe that some of these differences can be reduced by improving and refining our approach and strategy to handle samples with erroneously reported collection dates (i.e., time travellers that have passed the maximum submission delay of 30 days) and to modify the topology of daily trees.
+ARGs reconstructed using sc2ts capture the phylogenetic relationships of major
+SARS-CoV-2 lineages. Our results show that the backbone topology of a densely
+sampled wide ARG inferred using sc2ts agrees well with the backbone topology of
+a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference
+package. However, we have noted several areas of differences (e.g., poorly
+calibrated branch lengths), which are the focus of future development of sc2ts.
+We believe that some of these differences can be reduced by improving and
+refining our approach and strategy to handle samples with erroneously reported
+collection dates (i.e., time travellers that have passed the maximum submission
+delay of 30 days) and to modify the topology of daily trees.
 
-Our results show that sc2ts identifies the early recombinants (pre-2022) previously identified by Jackson et al.
-(2021).
-Sc2ts also identifies many Pango recombinants, which are designated after community perusal of supporting genomic evidence.
-We have found several previously unreported and overlooked recombinant sequences in the wide ARG.
-After further improvements and enhancements, we will use sc2ts to build a comprehensive ARG to search for recombination signals among all the 15 million and counting SARS-CoV-2 genomes available.
+Our results show that sc2ts identifies the early recombinants (pre-2022)
+previously identified by Jackson et al. (2021). Sc2ts also identifies many
+Pango recombinants, which are designated after community perusal of supporting
+genomic evidence. We have found several previously unreported and overlooked
+recombinant sequences in the wide ARG. After further improvements and
+enhancements, we will use sc2ts to build a comprehensive ARG to search for
+recombination signals among all the 15 million and counting SARS-CoV-2 genomes
+available.
 
-“The rate of putative recombination breakpoints is about three times higher towards the 3' of the change point than the 5' interval.“ Consistent with this, the breakpoint intervals of the putative recombinants identified here tend to occur on the 3’ end of the SARS-CoV-2 genome.
+“The rate of putative recombination breakpoints is about three times higher
+towards the 3' of the change point than the 5' interval.“ Consistent with this,
+the breakpoint intervals of the putative recombinants identified here tend to
+occur on the 3’ end of the SARS-CoV-2 genome.
 
@@
 
-Currently, sc2ts is in development. Below, we propose several ways to potentially improve or enhance sc2ts. We welcome suggestions and ideas from experts.
+Currently, sc2ts is in development. Below, we propose several ways to
+potentially improve or enhance sc2ts. We welcome suggestions and ideas from
+experts.
 
-\begin{itemize}
-\item A potential source of error in sc2ts ARGs comes from incorrectly reported collection dates of the samples, which can cause sample genomes to be attached to the wrong parts of the ARGs.
-Thus far, we have attempted to correct the most offending errors in the ARGs by filtering out "time travellers", but the sampling time can still be wrong by days or tens of days.
-***describe possible approaches to correct this?***
-\item To avoid time travellers, we used a maximum submission delay, but this will systematically bias against submissions from countries with greater submission delays (Kalia et al.).
-An area of future work is to add back the excluded sequences to a built ARG to assess which ones are likely true time travellers.
-\item To avoid time travellers, we used a maximum submission delay, but this will systematically bias against submissions from countries with greater submission delays (Kalia et al.).
-An area of future work is to add back the excluded sequences to a built ARG to assess which ones are likely true time travellers.
-\item We have used a single mismatch value in all the HMM runs, set so that having three mutations is as likely as having a single recombination.
-This parameter choice is arbitrary, and higher values may reduce the number of origins for those Pango lineages that were inferred to descend from more than one recombinant ancestor.
-In addition, because both the mutation rate and recombination rate varies across the SARS-CoV-2 genome, it may be helpful to allow the mismatch value to vary along the genome.
-\item By treating mutations equally in the current HMM implementation, we implicitly assume that the rates of different types of mutations are identical.
-However, as reported previously (Yi et al.
-2021), there is a high degree of mutational asymmetry in SARS-CoV-2.
-To account for mutational biases, more sophisticated nucleotide substitution models (e.g., allowing for different rates of transition and transversion) can be incorporated into HMM likelihood calculations.
-\item Our current masking strategy handles indels as missing data.
-Doing this can remove information helpful to detect recombinants.
-For example, sequences of the Pango recombinant XP were not identified as recombinants, because XP’s characteristic multibase deletion was masked.
-\item This "lack of memory" in the HMM can create recombinants with three or more parents when it would be equally parsimonious to have just two.
-The copying process in the HMM currently does not limit the number of parents to two, but doing so would be a valuable addition.
-\item Some recombinants may reasonably have three or more parents, indicating multiple "stacked" recombination events.
-We have largely ignored these in the current analysis, but with further sc2ts development, there is potential to analyse these putative recombinants and trace their evolutionary origins through the ARG.
-\item In the present implementation, HMM traceback is performed once, yielding a single maximum-likelihood genome copying path.
-A stochastic traceback algorithm (e.g., (Rasmussen et al.
-2014)) can be implemented in order to quantify uncertainty about putative parent node(s) and switch positions.
-These estimates of uncertainty may be used as support values for branches analogous to bootstrap values produced by standard phylogenetic methods.
-\item Daily trees are built using the UPGMA algorithm.
-This entails an extra step to “un-resolve” daily trees in parts of the trees with no phylogenetic resolution.
-To avoid this step, a better approach is to use a maximum-parsimony algorithm to build daily trees instead.
-\end{itemize}
+\begin{itemize} \item A potential source of error in sc2ts ARGs comes from
+incorrectly reported collection dates of the samples, which can cause sample
+genomes to be attached to the wrong parts of the ARGs. Thus far, we have
+attempted to correct the most offending errors in the ARGs by filtering out
+"time travellers", but the sampling time can still be wrong by days or tens of
+days. ***describe possible approaches to correct this?*** \item To avoid time
+travellers, we used a maximum submission delay, but this will systematically
+bias against submissions from countries with greater submission delays (Kalia
+et al.). An area of future work is to add back the excluded sequences to a
+built ARG to assess which ones are likely true time travellers. \item To avoid
+time travellers, we used a maximum submission delay, but this will
+systematically bias against submissions from countries with greater submission
+delays (Kalia et al.). An area of future work is to add back the excluded
+sequences to a built ARG to assess which ones are likely true time travellers.
+\item We have used a single mismatch value in all the HMM runs, set so that
+having three mutations is as likely as having a single recombination. This
+parameter choice is arbitrary, and higher values may reduce the number of
+origins for those Pango lineages that were inferred to descend from more than
+one recombinant ancestor. In addition, because both the mutation rate and
+recombination rate varies across the SARS-CoV-2 genome, it may be helpful to
+allow the mismatch value to vary along the genome. \item By treating mutations
+equally in the current HMM implementation, we implicitly assume that the rates
+of different types of mutations are identical. However, as reported previously
+(Yi et al. 2021), there is a high degree of mutational asymmetry in SARS-CoV-2.
+To account for mutational biases, more sophisticated nucleotide substitution
+models (e.g., allowing for different rates of transition and transversion) can
+be incorporated into HMM likelihood calculations. \item Our current masking
+strategy handles indels as missing data. Doing this can remove information
+helpful to detect recombinants. For example, sequences of the Pango recombinant
+XP were not identified as recombinants, because XP’s characteristic multibase
+deletion was masked. \item This "lack of memory" in the HMM can create
+recombinants with three or more parents when it would be equally parsimonious
+to have just two. The copying process in the HMM currently does not limit the
+number of parents to two, but doing so would be a valuable addition. \item Some
+recombinants may reasonably have three or more parents, indicating multiple
+"stacked" recombination events. We have largely ignored these in the current
+analysis, but with further sc2ts development, there is potential to analyse
+these putative recombinants and trace their evolutionary origins through the
+ARG. \item In the present implementation, HMM traceback is performed once,
+yielding a single maximum-likelihood genome copying path. A stochastic
+traceback algorithm (e.g., (Rasmussen et al. 2014)) can be implemented in order
+to quantify uncertainty about putative parent node(s) and switch positions.
+These estimates of uncertainty may be used as support values for branches
+analogous to bootstrap values produced by standard phylogenetic methods. \item
+Daily trees are built using the UPGMA algorithm. This entails an extra step to
+“un-resolve” daily trees in parts of the trees with no phylogenetic resolution.
+To avoid this step, a better approach is to use a maximum-parsimony algorithm
+to build daily trees instead. \end{itemize}
 
@@
 
-The global understanding of viral evolution during the COVID-19 pandemic is unprecedented, with over 15 million sequences shared to date.
-We envisage that once fully developed and tested, sc2ts methodology and software may provide a unified computational framework to study the evolution and epidemiology of SARS-CoV-2 at pandemic scale.
-The evolutionary analyses presented herein exemplify the numerous applications possible using the ancestral recombination graph inferred by sc2ts.
-We invite bioinformaticians, evolutionary biologists, and virologists to join us and the broader tskit community to build the most complete representation of the evolutionary history of SARS-CoV-2.
-Moreover, while we have focussed on the particular application to SARS-CoV-2, the methods we describe are readily applicable to the analysis of many other viral pathogens.
-Solving the challenges in building and analysing large viral genealogies would therefore contribute to the important goal of improving preparedness for potential future pandemics.
+The global understanding of viral evolution during the COVID-19 pandemic is
+unprecedented, with over 15 million sequences shared to date. We envisage that
+once fully developed and tested, sc2ts methodology and software may provide a
+unified computational framework to study the evolution and epidemiology of
+SARS-CoV-2 at pandemic scale. The evolutionary analyses presented herein
+exemplify the numerous applications possible using the ancestral recombination
+graph inferred by sc2ts. We invite bioinformaticians, evolutionary biologists,
+and virologists to join us and the broader tskit community to build the most
+complete representation of the evolutionary history of SARS-CoV-2. Moreover,
+while we have focussed on the particular application to SARS-CoV-2, the methods
+we describe are readily applicable to the analysis of many other viral
+pathogens. Solving the challenges in building and analysing large viral
+genealogies would therefore contribute to the important goal of improving
+preparedness for potential future pandemics.
 
@@
 
-We gratefully acknowledge all data contributors, i.e., the Authors and their Originating laboratories responsible for obtaining the specimens, and their Submitting laboratories for generating the genetic sequence and metadata and sharing via the GISAID Initiative, on which this research is based.
-Also, we thank Dr.
-Morag Graham (the National Microbiology Laboratory, Winnipeg, Canada) for kindly providing the genomic coordinates of the breakpoint sequence motifs.
-SHZ is supported by the Janssen-Oxford Translational Genomics Fellowship.
-JK, YW, and BJ are supported by the Robertson Foundation.
+We gratefully acknowledge all data contributors, i.e., the Authors and their
+Originating laboratories responsible for obtaining the specimens, and their
+Submitting laboratories for generating the genetic sequence and metadata and
+sharing via the GISAID Initiative, on which this research is based. Also, we
+thank Dr. Morag Graham (the National Microbiology Laboratory, Winnipeg, Canada)
+for kindly providing the genomic coordinates of the breakpoint sequence motifs.
+SHZ is supported by the Janssen-Oxford Translational Genomics Fellowship. JK,
+YW, and BJ are supported by the Robertson Foundation.
 
-\begin{table}
-\centering
-\begin{tabular}{l|l|l}
-\hline
-\textbf{Property} & \textbf{Wide ARG} & \textbf{Long ARG} \\
-\hline
-Last sampling date & 2021-06-30 & 2022-06-30 \\
-Local trees & 1,496 & 958 \\
-\hline
-\multicolumn{3}{l}{\textit{Nodes}} \\
-All nodes (sample and non-sample) & 1,453,347 & 783,231 \\
-Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\
-Tree adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107 (10.97\%) \\
-Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\
-\hline
-\multicolumn{3}{l}{\textit{Types of recombination nodes}} \\
-Singletons (i.e. no descendant besides the sequence causing insertion of the node) & 2,601 (63\%) & 1,315 (63\%) \\
-Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%) \\
-HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\
-\hline
-\multicolumn{3}{l}{\textit{Mutations}} \\
-Total mutations & 1,213,193 & 1,062,072 \\
-Mutations on a terminal edge & 758,903 (62.55\%) & 767,111 (72.23\%) \\
-Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
-Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\
-Immediate reversions (one intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\
-\hline
-\multicolumn{3}{l}{\textit{Storage and loading time}} \\
-tszipped file (including sample metadata, imputed Pango lineage assignments, and results of HMM runs for recombination nodes) & ~57 Mb & ~36 Mb \\
-Bytes to encode one sample & 8.29 & 10.83 \\
-Load time (seconds) & 0.9 & 0.5 \\
-\hline
-\end{tabular}
-\end{table}
+\begin{table} \centering \begin{tabular}{l|l|l} \hline \textbf{Property} &
+\textbf{Wide ARG} & \textbf{Long ARG} \\ \hline Last sampling date & 2021-06-30
+& 2022-06-30 \\ Local trees & 1,496 & 958 \\ \hline
+\multicolumn{3}{l}{\textit{Nodes}} \\ All nodes (sample and non-sample) &
+1,453,347 & 783,231 \\ Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\ Tree
+adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107
+(10.97\%) \\ Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\ \hline
+\multicolumn{3}{l}{\textit{Types of recombination nodes}} \\ Singletons (i.e.
+no descendant besides the sequence causing insertion of the node) & 2,601
+(63\%) & 1,315 (63\%) \\ Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%)
+\\ HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\ \hline
+\multicolumn{3}{l}{\textit{Mutations}} \\ Total mutations & 1,213,193 &
+1,062,072 \\ Mutations on a terminal edge & 758,903 (62.55\%) & 767,111
+(72.23\%) \\ Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
+Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\ Immediate reversions (one
+intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\ \hline
+\multicolumn{3}{l}{\textit{Storage and loading time}} \\ tszipped file
+(including sample metadata, imputed Pango lineage assignments, and results of
+HMM runs for recombination nodes) & ~57 Mb & ~36 Mb \\ Bytes to encode one
+sample & 8.29 & 10.83 \\ Load time (seconds) & 0.9 & 0.5 \\ \hline
+\end{tabular} \end{table}
 
-\begin{table}
-\centering
-\begin{tabular}{l|c|c|c|c|c}
-\hline
-\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} & \multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\
-\hline
-\textbf{Sample} & \textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\ interval(s)} & \textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\
-\hline
-ALDP-11CF93B & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
-ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
-ALDP-130BB95 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
-LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 & \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
-QEUH-CCCB30 & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 & \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
-QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 & \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
-MILK-1166F52 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
-MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
-QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
-MILK-126FE1F & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
-RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
-RAND-128FA33 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
-CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 & \thead{B.1.177 \\ Alpha} & 16,177–21,255 \\
-CAMC-CB7AB3 & n/a & \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} & \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\
-MILK-103C712 & n/a & \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} & n/a & n/a \\
-QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 & \thead{Alpha \\ B.1.177} & 7,729–10,870 \\
-\hline
-\end{tabular}
-\caption{\label{tab:jackson}Recombinant sequences involving the Alpha (B.1.1.7) variant reported by Jackson et al.
-\cite{Jackson2021-ik} have recombinant ancestry in the wide ARG.
-The breakpoint intervals and Pango lineage assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et al., except the Pango lineage assignment of the parents of group B recombinants, which were taken from Table 1 (motif-based results).}
-\end{table}
+\begin{table} \centering \begin{tabular}{l|c|c|c|c|c} \hline
+\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
+\multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\ \hline \textbf{Sample} &
+\textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\ interval(s)} &
+\textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\ \hline ALDP-11CF93B & A
+& \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
+21,256–22,227 \\ ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
+\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ ALDP-130BB95 & A &
+\thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
+21,256–22,227 \\ LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
+\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ QEUH-CCCB30 & B &
+\thead{B.1.36.28 \\ Alpha} & 6,528–6,953 & \thead{B.1.36 \\ Alpha} &
+6,529–6,954 \\ QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
+\thead{B.1.36 \\ Alpha} & 6,529–6,954 \\ MILK-1166F52 & C & \thead{Alpha \\
+B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
+MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\
+B.1.221} & 25,997–27,972 \\ QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} &
+25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\ MILK-126FE1F & D &
+\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
+22,445–23,063 \\ RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062
+& \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\ RAND-128FA33 & D &
+\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
+22,445–23,063 \\ CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
+\thead{B.1.177 \\ Alpha} & 16,177–21,255 \\ CAMC-CB7AB3 & n/a & \thead{Alpha \\
+B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} & \thead{Alpha \\
+B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\ MILK-103C712 & n/a
+& \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} & n/a & n/a \\
+QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 & \thead{Alpha
+\\ B.1.177} & 7,729–10,870 \\ \hline \end{tabular}
+\caption{\label{tab:jackson}Recombinant sequences involving the Alpha (B.1.1.7)
+variant reported by Jackson et al. \cite{Jackson2021-ik} have recombinant
+ancestry in the wide ARG. The breakpoint intervals and Pango lineage
+assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et
+al., except the Pango lineage assignment of the parents of group B
+recombinants, which were taken from Table 1 (motif-based results).} \end{table}
 
-\begin{figure}
-\centering
+\begin{figure} \centering
 \includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
-\caption{\label{fig:overview_sc2ts} A schematic of the sc2ts method.
-Sc2ts reconstructs the genetic relationships among SARS-CoV-2 genomes by copying samples to all possible ancestors collected at earlier time points (curved arrows).
-An ARG is built by attaching new sample genomes grouped by collection date to an existing ARG rooted at the Wuhan-Hu-1 reference genome, day by day.
-Each iteration involves three stages: (1) attachment of new samples to the growing ARG (A, B); (2) reconstruction of trees relating the samples under each attachment node (C); and parsimony-based tree topology adjustments (D, E).
-In the absence of  recombination, sc2ts infers an ARG that is a single tree relating the samples (A).
-However, in the presence of recombination, sc2ts infers an ARG that is encoded as a sequence of local trees relating segments of the sample genomes (B), inserting a recombination node to represent a putative recombination event.
-Additionally, mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to make more parsimonious placements of mutations that should be shared or should not be immediately reverted, respectively.}
-\end{figure}
+\caption{\label{fig:overview_sc2ts} A schematic of the sc2ts method. Sc2ts
+reconstructs the genetic relationships among SARS-CoV-2 genomes by copying
+samples to all possible ancestors collected at earlier time points (curved
+arrows). An ARG is built by attaching new sample genomes grouped by collection
+date to an existing ARG rooted at the Wuhan-Hu-1 reference genome, day by day.
+Each iteration involves three stages: (1) attachment of new samples to the
+growing ARG (A, B); (2) reconstruction of trees relating the samples under each
+attachment node (C); and parsimony-based tree topology adjustments (D, E). In
+the absence of  recombination, sc2ts infers an ARG that is a single tree
+relating the samples (A). However, in the presence of recombination, sc2ts
+infers an ARG that is encoded as a sequence of local trees relating segments of
+the sample genomes (B), inserting a recombination node to represent a putative
+recombination event. Additionally, mutation-collapsing nodes (D) and
+pushed-reversion nodes (E) are inserted to make more parsimonious placements of
+mutations that should be shared or should not be immediately reverted,
+respectively.} \end{figure}
 
-\begin{figure}
-\centering
+\begin{figure} \centering
 \includegraphics[width=0.8\textwidth]{figures/ls_diagram_covid.pdf}
 \caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS) model.
-The LS model in its implementation as an HMM is governed by the transition and emission matrices, \textit{Q} and \textit{E}, respectively.
-Each may be a function of the reference panel members, but transitions are generally assumed to be independent of the hidden states themselves (shown in the pink panel).
-This dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
-The recombination probability at site \textit{l} is \textit{rl}, while n denotes the number of members in the reference panel.
-Emissions may also be a function of the nucleotide states, but in our case, we assume that all observed alleles (in both the reference panel and focal sequence, \textit{al}) at the site under consideration are equally likely and occur with mutation probability \textit{l} (shown in the blue panel).
-These two processes, transition and emission (which encode recombination and mutation, respectively), define the generative process of the HMM, by which an imperfect mosaic is emitted from a reference panel of samples and nodes from the ARG inserted at earlier collection dates (shown in the lower panel).
-Here, the unseen states in the reference panel are shown as coloured lines enclosed by the grey box, and the true path through the data which leads to the emitted focal sequence is displayed by the black arrow.
-Examples of transition and emission probabilities along this trajectory are shown by the red and blue arrows, respectively.
-Black crosses along the focal sequence show genotyping errors or mutations.
-Note that in practice, we do not know the pattern of colours along the focal sequence and must infer it using the Viterbi algorithm.}
-\end{figure}
+The LS model in its implementation as an HMM is governed by the transition and
+emission matrices, \textit{Q} and \textit{E}, respectively. Each may be a
+function of the reference panel members, but transitions are generally assumed
+to be independent of the hidden states themselves (shown in the pink panel).
+This dramatically increases performance as the state space drops to two states
+(i.e., switching or not switching). The recombination probability at site
+\textit{l} is \textit{rl}, while n denotes the number of members in the
+reference panel. Emissions may also be a function of the nucleotide states, but
+in our case, we assume that all observed alleles (in both the reference panel
+and focal sequence, \textit{al}) at the site under consideration are equally
+likely and occur with mutation probability \textit{l} (shown in the blue
+panel). These two processes, transition and emission (which encode
+recombination and mutation, respectively), define the generative process of the
+HMM, by which an imperfect mosaic is emitted from a reference panel of samples
+and nodes from the ARG inserted at earlier collection dates (shown in the lower
+panel). Here, the unseen states in the reference panel are shown as coloured
+lines enclosed by the grey box, and the true path through the data which leads
+to the emitted focal sequence is displayed by the black arrow. Examples of
+transition and emission probabilities along this trajectory are shown by the
+red and blue arrows, respectively. Black crosses along the focal sequence show
+genotyping errors or mutations. Note that in practice, we do not know the
+pattern of colours along the focal sequence and must infer it using the Viterbi
+algorithm.} \end{figure}
 
-\begin{figure}
-\centering
+\begin{figure} \centering
 \includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
-\caption{\label{fig:mutational_spectra} Mutational spectrum of SARS-CoV-2 in the wide ARG (sampled to mid-2021) compared to that reported by Yi et al.
-\cite{Yi2021-sc}.
-Mutations are categorised by type (i.e., ancestral state > derived state).
-The percentages of each mutation type from the wide ARG are represented by blue bars and the percentages from Yi et al.
-by orange bars, with the darker colours representing one direction (e.g., C > U) and the lighter colours the reverse (e.g., U > C).}
-\end{figure}
+\caption{\label{fig:mutational_spectra} Mutational spectrum of SARS-CoV-2 in
+the wide ARG (sampled to mid-2021) compared to that reported by Yi et al.
+\cite{Yi2021-sc}. Mutations are categorised by type (i.e., ancestral state >
+derived state). The percentages of each mutation type from the wide ARG are
+represented by blue bars and the percentages from Yi et al. by orange bars,
+with the darker colours representing one direction (e.g., C > U) and the
+lighter colours the reverse (e.g., U > C).} \end{figure}
 

### 2023-04-06 paper.tex
Add makefile, use file names from other projects for simplicity

--- 2023-04-06 main.tex
@@
 runs. Filtering reduces both the number of recombination nodes and breakpoints
-from  A \& B to C in the wide ARG and from 726 & 806 to 459 in the long ARG
+from  A \& B to C in the wide ARG and from 726 \& 806 to 459 in the long ARG
 (Supplementary Tables 1 and 2). Note that in the unfiltered case, a single
@@
 \bibliographystyle{alpha}
-\bibliography{references}
+\bibliography{paper}
 

### 2023-04-06 paper.tex
Fixup refs

@@
 \usepackage{graphicx}
-\usepackage[colorlinks=true, allcolors=blue]{hyperref}
+\usepackage{natbib}
+\usepackage{url}
+% JK: turning this off for the moment as I keep clicking through on links
+% to the bibliography while reading the text and it's intensely annoying.
+% Can reinstate when we're ready to preprint
+% \usepackage[hidelinks]{hyperref}
+
 \usepackage{makecell}
@@
 Recombination via template switching is a common feature in the evolution of
-coronaviruses (CoV) \cite{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2
-\cite{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
+coronaviruses (CoV) \citep{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2
+\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
 together mutations carried by different lineages, recombination plays an
@@
 associated with CoV adaptation to new host species and with the production of
-more immune evasive variants \cite{Graham2010-xe,De_Klerk2022-tt}.
+more immune evasive variants \citep{Graham2010-xe,De_Klerk2022-tt}.
 
@@
 have been detected, for example, between Alpha and pre-VoC
-\cite{Jackson2021-ik}, Alpha and Epsilon \cite{Wertheim2022-hj}, and Alpha and
-Delta \cite{Sekizuka2022-xz}). More recently, the high prevalence of Omicron,
+\citep{Jackson2021-ik}, Alpha and Epsilon \citep{Wertheim2022-hj}, and Alpha and
+Delta \citep{Sekizuka2022-xz}). More recently, the high prevalence of Omicron,
 with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
 accelerated the rate of coinfection and the potential for recombination
-\cite{Bal2022-hq}. Recombinant lineages that have sufficient evidence of onward
+\citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of onward
 transmission (more than five descendants) and monophyly (> 70\% bootstrap
-support) are eligible for a Pango lineage designation \cite{Rambaut2020-dw},
+support) are eligible for a Pango lineage designation \citep{Rambaut2020-dw},
 starting with the letter X (“Pango Lineage Nomenclature: Provisional Rules for
@@
 combine specific subsets of mutations from different lineages (e.g.,
-\cite{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}).
+\citep{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}).
 This approach only identifies those inter-lineage recombination events
@@
 having to pre-specify characteristic mutations for each lineage
-\cite{Ignatieva2021-rg}. More recently, Turakhia et al. \cite{Turakhia2022-it}
+\citep{Ignatieva2021-rg}. More recently, Turakhia et al. \citep{Turakhia2022-it}
 has presented a phylogeny-based approach ("RIPPLES") to identify putative
@@
 of recombinant lineages (e.g., more than half of the sequences sampled in
-February, 2023 are from XBB and its descendants; \cite{Chen2022-pz}.
+February, 2023 are from XBB and its descendants; \citep{Chen2022-pz}.
 
@@
 acute for SARS-CoV-2, given the large amount of data that has been shared, with
-over 15 million sequences in the GISAID database \cite{Shu2017-hp}. It is well
-known that recombination distorts phylogenies \cite{Schierup2000-fg} and
+over 15 million sequences in the GISAID database \citep{Shu2017-hp}. It is well
+known that recombination distorts phylogenies \citep{Schierup2000-fg} and
 affects the results of downstream analysis (e.g., inference of selection;
-\cite{Anisimova2003-vr}). However, standard phylogenetic methods do not account
-for recombination (e.g., \cite{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}).
+\citep{Anisimova2003-vr}). However, standard phylogenetic methods do not account
+for recombination (e.g., \citep{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}).
 Alternatively, genetic ancestries of recombining lineages can be represented by
-ancestral recombination graphs (ARG; \cite{Griffiths1981-lw,Hudson1983-if},
-which  capture network-like phylogenetic relationships \cite{Gusfield2014-qw}.
+ancestral recombination graphs (ARG; \citep{Griffiths1981-lw,Hudson1983-if},
+which  capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.
 
@@
 genomes and beyond
-\cite{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}. Recently,
-Wohns et al. \cite{Wohns2022-th} reconstructed a unified whole-genome genealogy
+\citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}. Recently,
+Wohns et al. \citep{Wohns2022-th} reconstructed a unified whole-genome genealogy
 of thousands of modern and ancient human genomes, based on over 90 million
-variable sites. To achieve this, Wohns et al. \cite{Wohns2022-th} utilised
+variable sites. To achieve this, Wohns et al. \citep{Wohns2022-th} utilised
 several computational advances in ARG inference: (1) the "succinct tree
 sequence" data structure that can compactly store large ARGs
-\cite{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM)
+\citep{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM)
 implementation of the Li and Stephens (LS) model (Li and Stephens 2003) to
-infer copying paths between genomes \cite{Kelleher2019-ba}; and (3) efficient,
+infer copying paths between genomes \citep{Kelleher2019-ba}; and (3) efficient,
 open-source software tools
-\cite{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.
+\citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.
 
@@
 information about SARS-CoV-2 that can be efficiently accessed using "tskit"
-(tree sequence toolkit) software library \cite{Kelleher2018-xc}. Sc2ts is a
+(tree sequence toolkit) software library \citep{Kelleher2018-xc}. Sc2ts is a
 promising method to automate detection of SARS-CoV-2 recombinants and to
@@
 The findings of this study are based on sequences and metadata available on
-GISAID (\href{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
+GISAID (\url{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
 doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
@@
 Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
-MN908947.3) using Nextclade v2.3.0 \cite{Aksamentov2021-hj} (dataset tag
+MN908947.3) using Nextclade v2.3.0 \citep{Aksamentov2021-hj} (dataset tag
 2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
@@
 were masked as missing data following the protocol in
-\href{https://github.com/roblanf/sarscov2phylo/}. Masked bases were ignored
+\url{https://github.com/roblanf/sarscov2phylo/}. Masked bases were ignored
 when attaching a sample genome to the ARG, but they were subsequently imputed.
@@
 homoplasic were excluded entirely
-(\href{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
-\href{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
+(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
+\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
 
@@
 We introduce an ARG inference method (schematic in Figure 1) that uses (1) an
-efficient HMM implementation of the Li and Stephens (LS) model \cite{Li2003-ib}
+efficient HMM implementation of the Li and Stephens (LS) model \citep{Li2003-ib}
 to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported
@@
 attachment node using the Unweighted Pair Group Method with Arithmetic Mean
-(UPGMA) algorithm \cite{Michener1957-tr}, with average linkage (Figure 1C).
+(UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
 This hierarchical clustering algorithm produces a dendrogram relating a set of
@@
 Furthermore, during inference, the ARG is automatically encoded, stored, and
-updated as a tree sequence using tskit \cite{Kelleher2018-xc}.
+updated as a tree sequence using tskit \citep{Kelleher2018-xc}.
 
@@
 The LS model is a convenient probabilistic approximation to the coalescent with
-recombination \cite{Hudson1983-if}, itself a generative model of a genealogical
-process with recombination \cite{Li2003-ib}. Importantly, it can be implemented
+recombination \citep{Hudson1983-if}, itself a generative model of a genealogical
+process with recombination \citep{Li2003-ib}. Importantly, it can be implemented
 as an HMM in which a focal genome is a sequence of nucleotides (observed
@@
 engine that drives statistical genotype phasing and imputation
-\cite{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it
+\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it
 can be used to estimate mutation rates (e.g., REF), recombination rates (e.g.,
-\cite{Hinch2011-tz}), and selection within and across hosts in viral sequence
-data (e.g., \cite{Palmer2019-wa}).
+\citep{Hinch2011-tz}), and selection within and across hosts in viral sequence
+data (e.g., \citep{Palmer2019-wa}).
 
@@
 state from which it is copied (due to mutation as well as genotyping or
-sequencing error). Then, we use the Viterbi algorithm \cite{Viterbi1967-ol} to
+sequencing error). Then, we use the Viterbi algorithm \citep{Viterbi1967-ol} to
 find the most likely copying path through the sequences, given a set of
@@
 lineage-defining mutations (based on 90\% consensus of the sequences analysed)
-from the COVID-CG website \cite{Chen2021-zc}; \href{https://covidcg.org/};
+from the COVID-CG website \citep{Chen2021-zc}; \url{https://covidcg.org/};
 accessed on November 04, 2022).
@@
 The source code implementing the sc2ts pipeline is available via GitHub
-\href{https://github.com/jeromekelleher/sc2ts/}. The sc2ts pipeline uses
+\url{https://github.com/jeromekelleher/sc2ts/}. The sc2ts pipeline uses
 various functions and algorithms implemented in tskit and tsinfer
-\href{https://github.com/tskit-dev/}. The pipeline uses the SciPy
+\url{https://github.com/tskit-dev/}. The pipeline uses the SciPy
 implementation of the UPGMA algorithm to build daily trees
-\href{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}.
+\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}.
 The Jupyter notebooks used to produce the results described herein are also
-available on GitHub \href{https://github.com/jeromekelleher/sc2ts-paper/}.
+available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.
 
@@
 sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain
-project \cite{Hadfield2018-ef}. The NextStrain trees consist of subsamples of
+project \citep{Hadfield2018-ef}. The NextStrain trees consist of subsamples of
 SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely
 restricted to non-recombinant lineages. To allow for comparison, the ARGs were
-"simplified" \cite{Kelleher2018-xc} to retain only those tips of our genealogy
+"simplified" \citep{Kelleher2018-xc} to retain only those tips of our genealogy
 that match the samples present in a specific NextStrain tree. These simplified
@@
 genome (Figure 4A), as noted previously by Turakhia et al.
-\cite{Turakhia2022-it}. Inspecting more closely, we find an enrichment in the
+\citep{Turakhia2022-it}. Inspecting more closely, we find an enrichment in the
 number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region,
@@
 mutations also occur. This position has been reported to be hypermutable
-\cite{Jungreis2021-dh}.
+\citep{Jungreis2021-dh}.
 
@@
 template switching during replication in SARS-CoV-2 via formation of
-base-paired stem loops in the genome structure \cite{Gallaher2020-lb}. There
+base-paired stem loops in the genome structure \citep{Gallaher2020-lb}. There
 are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019
@@
 In addition to recombination events, the ARGs record mutation events along the
-SARS-CoV-2 phylogeny. Yi et al. \cite{Yi2021-sc} reconstructed a SARS-CoV-2
+SARS-CoV-2 phylogeny. Yi et al. \citep{Yi2021-sc} reconstructed a SARS-CoV-2
 phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to
@@
 Similarly, we took the data for single nucleotide mutations from Yi et al.
-\ref{https://github.com/ju-lab/SC2_evol_signature}, excluded mutations
+\url{https://github.com/ju-lab/SC2_evol_signature}, excluded mutations
 occurring along terminal branches, and tallied them up by type. We find that
@@
 \caption{\label{tab:jackson}Recombinant sequences involving the Alpha (B.1.1.7)
-variant reported by Jackson et al. \cite{Jackson2021-ik} have recombinant
+variant reported by Jackson et al. \citep{Jackson2021-ik} have recombinant
 ancestry in the wide ARG. The breakpoint intervals and Pango lineage
@@
 the wide ARG (sampled to mid-2021) compared to that reported by Yi et al.
-\cite{Yi2021-sc}. Mutations are categorised by type (i.e., ancestral state >
+\citep{Yi2021-sc}. Mutations are categorised by type (i.e., ancestral state >
 derived state). The percentages of each mutation type from the wide ARG are
@@
 
-\bibliographystyle{alpha}
+\bibliographystyle{plainnat}
 \bibliography{paper}

### 2023-04-06 paper.tex
Light edits to intro

@@
 
+\usepackage[round]{natbib}
 \usepackage[english]{babel}
@@
 \usepackage{graphicx}
-\usepackage{natbib}
 \usepackage{url}
@@
 Recombination via template switching is a common feature in the evolution of
-coronaviruses (CoV) \citep{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2
+coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2
 \citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
@@
 important role in generating genomic diversity, with recombinant lineages
-associated with CoV adaptation to new host species and with the production of
-more immune evasive variants \citep{Graham2010-xe,De_Klerk2022-tt}.
+associated with adaptation to new host species and with the production of
+more immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
 
@@
 \citep{Jackson2021-ik}, Alpha and Epsilon \citep{Wertheim2022-hj}, and Alpha and
-Delta \citep{Sekizuka2022-xz}). More recently, the high prevalence of Omicron,
+Delta~\citep{Sekizuka2022-xz}. More recently, the high prevalence of Omicron,
 with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
@@
 \citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of onward
-transmission (more than five descendants) and monophyly (> 70\% bootstrap
-support) are eligible for a Pango lineage designation \citep{Rambaut2020-dw},
-starting with the letter X (“Pango Lineage Nomenclature: Provisional Rules for
-Naming Recombinant Lineages” 2021).
+transmission (more than five descendants) and monophyly
+($>70\%$ bootstrap support) are eligible for a Pango lineage
+designation~\citep{Rambaut2020-dw},
+starting with the letter X
+% FIXME: should this be a citation?
+(``Pango Lineage Nomenclature: Provisional Rules for Naming Recombinant Lineages'' 2021).
 
 Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
-combine specific subsets of mutations from different lineages (e.g.,
-\citep{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}).
+combine specific subsets of mutations from different lineages
+\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
 This approach only identifies those inter-lineage recombination events
@@
 having to pre-specify characteristic mutations for each lineage
-\citep{Ignatieva2021-rg}. More recently, Turakhia et al. \citep{Turakhia2022-it}
-has presented a phylogeny-based approach ("RIPPLES") to identify putative
-recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying
-motifs. To find candidate recombinants, RIPPLES first builds a phylogeny
+\citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it}
+presented a phylogeny-based approach (``RIPPLES'') to identify putative
+recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying motifs.
+To find candidate recombinants, RIPPLES first builds a phylogeny
 without assuming recombination and then scans for long branches (i.e., carrying
@@
 searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2
-phylogeny that accounts for only mutations, treats recombinants \textit{post
-hoc}, and is an incomplete representation of the reticulate evolutionary
+phylogeny that accounts for only mutations, treats recombinants
+\textit{post hoc}, and is an incomplete representation of the reticulate evolutionary
 history of SARS-CoV-2. As noted by the authors, a \textit{post hoc} treatment
 of recombination is possible when recombinant lineages are rare and leave few
-descendants, but this is becoming increasingly untenable with the proliferation
-of recombinant lineages (e.g., more than half of the sequences sampled in
-February, 2023 are from XBB and its descendants; \citep{Chen2022-pz}.
+descendants. However, this is becoming increasingly untenable with the proliferation
+of recombinant lineages; for example, more than half of the sequences sampled in
+February 2023 are from XBB and its descendants~\citep{Chen2022-pz}.
 
@@
 known that recombination distorts phylogenies \citep{Schierup2000-fg} and
-affects the results of downstream analysis (e.g., inference of selection;
-\citep{Anisimova2003-vr}). However, standard phylogenetic methods do not account
-for recombination (e.g., \citep{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}).
+affects the results of downstream analysis, e.g., inference of
+selection~\citep{Anisimova2003-vr}.
+However, standard phylogenetic methods do not account
+for recombination \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}.
 Alternatively, genetic ancestries of recombining lineages can be represented by
-ancestral recombination graphs (ARG; \citep{Griffiths1981-lw,Hudson1983-if},
+ancestral recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-if},
 which  capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.
@@
 \citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}. Recently,
-Wohns et al. \citep{Wohns2022-th} reconstructed a unified whole-genome genealogy
+\cite{Wohns2022-th} reconstructed a unified whole-genome genealogy
 of thousands of modern and ancient human genomes, based on over 90 million
-variable sites. To achieve this, Wohns et al. \citep{Wohns2022-th} utilised
+variable sites. To achieve this, \cite{Wohns2022-th} utilised
 several computational advances in ARG inference: (1) the "succinct tree
@@
 Here, we present a preliminary version of a novel scalable methodology
-(“sc2ts”, pronounced “scoots”) that builds on the advances above and
+(``sc2ts'', optionally pronounced ``scoots'') that builds on the advances above and
 additionally takes advantage of known collection dates to infer ARGs from

### 2023-04-06 paper.tex
Replace quotes

@@
 GISAID (\url{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
-doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
-if they (1) had ambiguous collection dates; (2) were collected before January
-01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
-Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
-MN908947.3) using Nextclade v2.3.0 \citep{Aksamentov2021-hj} (dataset tag
-2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
-inference, sequences were excluded if they had a “bad” quality control status
-in any of the four Nextclade columns ("qc.missingData.status”,
-“qc.mixedSites.status”, “qc.frameShifts.status”, and “qc.stopCodons.status”).
+% FIXME
+doi.org/10.55876/gis8.230329cd (Supplementary Table 1).
+Sequences were removed
+if they (1) had ambiguous collection dates;
+(2) were collected before January 01, 2020 or after June 30, 2022; or
+(3) were isolated from a non-human host.
+Sequences were aligned to the
+Wuhan-Hu-1/2019 reference sequence (GenBank: MN908947.3)
+using Nextclade v2.3.0~\citep{Aksamentov2021-hj}
+(dataset tag 2022-07-26T12:00:00Z).
+To reduce the impact of poor-quality sequences on ARG inference,
+sequences were excluded if they had a ``bad'' quality control status
+in any of the four Nextclade columns (``qc.missingData.status'',
+``qc.mixedSites.status'', ``qc.frameShifts.status'' and
+``qc.stopCodons.status'').
 These excluded genomes are enriched for low-coverage, contaminated, or
@@
 were masked as missing data following the protocol in
-\url{https://github.com/roblanf/sarscov2phylo/}. Masked bases were ignored
-when attaching a sample genome to the ARG, but they were subsequently imputed.
+\url{https://github.com/roblanf/sarscov2phylo/}.
+Masked bases were ignored when attaching a sample genome to the ARG,
+but they were subsequently imputed.
 Further, 481 problematic sites flagged as prone to sequencing errors or highly
 homoplasic were excluded entirely
+% Do we need this first URL?
 (\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
@@
 
-\subsubsection{Exclusion of “time travellers”}
+\subsubsection{Exclusion of ``time travellers''}
 
@@
 artefacts that distorted the topology of ARGs inferred early on in the study.
-Hence, to exclude such potential “time travellers”, we employed two filters.
+Hence, to exclude such potential ``time travellers'' we employed two filters.
 The first was to remove any sequence with a collection date that pre-dated the
@@
 phylogenetic resolution. In cases where the newly added sequence is inferred to
-be a recombinant, a recombination node is inserted; the “date” of this node is
+be a recombinant, a recombination node is inserted; the ``date'' of this node is
 arbitrarily set to the average of the collection dates of the genome and the
@@
 informative site (i.e., having no mutation immediately ancestral to them). In
-an adjustment step, such unsupported internal nodes are removed to “un-resolve”
+an adjustment step, such unsupported internal nodes are removed to ``un-resolve''
 the daily tree; the remaining internal nodes, which are supported by at least
-one shared mutation, are retained. The “dates” of the non-sample internal nodes
+one shared mutation, are retained. The ``dates'' of the non-sample internal nodes
 are set to the same date as the samples in the daily batch. The resulting daily
@@
 To illustrate the use of sc2ts, we employed two different sampling strategies.
-First, we inferred a densely sampled but time-limited “wide” ARG by attaching
+First, we inferred a densely sampled but time-limited ``wide'' ARG by attaching
 all the sample genomes with an MSD of 30 days, one daily batch at a time. The
@@
 attached to the ARG, resulting in a total of 1.27 million samples. Second, we
-inferred a “long” ARG over a more extended timespan, by attaching at most 1,000
+inferred a ``long'' ARG over a more extended timespan, by attaching at most 1,000
 genomes, which were randomly subsampled from each daily batch with an MSD
@@
 Although here we focus on a single local tree at an arbitrarily chosen
-position, note that in the original “un-simplified” ARGs, there are 1,496 and
+position, note that in the original ``un-simplified'' ARGs, there are 1,496 and
 958 trees, respectively (Table 1). Hence, while for a relatively small sample
@@
 
-For simplicity, we focus primarily on “non-complex” recombination nodes. To do
+For simplicity, we focus primarily on ``non-complex'' recombination nodes. To do
 this, we apply a filtering step that removes those with more than one
@@
 (Supplementary Table 3). For example, the genome of XD (also known as
-“Deltacron”) is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
+``Deltacron'') is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
 (Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.
@@
 in the UShER public phylogeny. Node 628656 (Scotland/QEUH-37794BB/2022) is a
-sister clade of XAC, and annotated as “miscBA2BA1PostSpike”.
+sister clade of XAC, and annotated as ``miscBA2BA1PostSpike''.
 
@@
 
-“The rate of putative recombination breakpoints is about three times higher
-towards the 3' of the change point than the 5' interval.“ Consistent with this,
+``The rate of putative recombination breakpoints is about three times higher
+towards the 3' of the change point than the 5' interval.`` Consistent with this,
 the breakpoint intervals of the putative recombinants identified here tend to
@@
 Daily trees are built using the UPGMA algorithm. This entails an extra step to
-“un-resolve” daily trees in parts of the trees with no phylogenetic resolution.
+``un-resolve'' daily trees in parts of the trees with no phylogenetic resolution.
 To avoid this step, a better approach is to use a maximum-parsimony algorithm

### 2023-04-06 paper.tex

@@
 \usepackage{graphicx}
+\usepackage{makecell}
 \usepackage{url}
@@
 % \usepackage[hidelinks]{hyperref}
-
-\usepackage{makecell}
 
@@
 \citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it
-can be used to estimate mutation rates (e.g., REF), recombination rates (e.g.,
-\citep{Hinch2011-tz}), and selection within and across hosts in viral sequence
-data (e.g., \citep{Palmer2019-wa}).
+can be used to estimate mutation rates (e.g., REF), recombination rates
+\citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral sequence
+data \citep[e.g.][]{Palmer2019-wa}.
 
@@
 (leaving X and Y recombinants in the wide and long ARGs, respectively). In
-their analysis using RIPPLES, Turakhia et al. (2022) detected 589 recombinants,
+their analysis using RIPPLES, \cite{Turakhia2022-it} detected 589 recombinants,
 also excluding singletons.
@@
 There is a preponderance of breakpoint intervals spanning the 3’ end of the
-genome (Figure 4A), as noted previously by Turakhia et al.
-\citep{Turakhia2022-it}. Inspecting more closely, we find an enrichment in the
+genome (Figure 4A), as noted previously by
+\cite{Turakhia2022-it}.
+Inspecting more closely, we find an enrichment in the
 number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region,
@@
 In addition to recombination events, the ARGs record mutation events along the
-SARS-CoV-2 phylogeny. Yi et al. \citep{Yi2021-sc} reconstructed a SARS-CoV-2
+SARS-CoV-2 phylogeny. \cite{Yi2021-sc} reconstructed a SARS-CoV-2
 phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to
@@
 Our results show that sc2ts identifies the early recombinants (pre-2022)
-previously identified by Jackson et al. (2021). Sc2ts also identifies many
+previously identified by \cite{Jackson2021-ik}. Sc2ts also identifies many
 Pango recombinants, which are designated after community perusal of supporting
@@
 
-\begin{itemize} \item A potential source of error in sc2ts ARGs comes from
+\begin{itemize}
+\item A potential source of error in sc2ts ARGs comes from
 incorrectly reported collection dates of the samples, which can cause sample
@@
 "time travellers", but the sampling time can still be wrong by days or tens of
-days. ***describe possible approaches to correct this?*** \item To avoid time
+days. ***describe possible approaches to correct this?***
+\item To avoid time
 travellers, we used a maximum submission delay, but this will systematically
@@
 et al.). An area of future work is to add back the excluded sequences to a
-built ARG to assess which ones are likely true time travellers. \item To avoid
+built ARG to assess which ones are likely true time travellers.
+\item To avoid
 time travellers, we used a maximum submission delay, but this will
@@
 recombination rate varies across the SARS-CoV-2 genome, it may be helpful to
-allow the mismatch value to vary along the genome. \item By treating mutations
+allow the mismatch value to vary along the genome.
+\item By treating mutations
 equally in the current HMM implementation, we implicitly assume that the rates
@@
 models (e.g., allowing for different rates of transition and transversion) can
-be incorporated into HMM likelihood calculations. \item Our current masking
+be incorporated into HMM likelihood calculations.
+\item Our current masking
 strategy handles indels as missing data. Doing this can remove information
@@
 XP were not identified as recombinants, because XP’s characteristic multibase
-deletion was masked. \item This "lack of memory" in the HMM can create
+deletion was masked.
+\item This "lack of memory" in the HMM can create
 recombinants with three or more parents when it would be equally parsimonious
 to have just two. The copying process in the HMM currently does not limit the
-number of parents to two, but doing so would be a valuable addition. \item Some
+number of parents to two, but doing so would be a valuable addition.
+\item Some
 recombinants may reasonably have three or more parents, indicating multiple
@@
 these putative recombinants and trace their evolutionary origins through the
-ARG. \item In the present implementation, HMM traceback is performed once,
+ARG.
+\item In the present implementation, HMM traceback is performed once,
 yielding a single maximum-likelihood genome copying path. A stochastic
@@
 These estimates of uncertainty may be used as support values for branches
-analogous to bootstrap values produced by standard phylogenetic methods. \item
+analogous to bootstrap values produced by standard phylogenetic methods.
+\item
 Daily trees are built using the UPGMA algorithm. This entails an extra step to
@@
 To avoid this step, a better approach is to use a maximum-parsimony algorithm
-to build daily trees instead. \end{itemize}
+to build daily trees instead.
+\end{itemize}
 

### 2023-04-06 paper.tex

@@
 
-We introduce an ARG inference method (schematic in Figure 1) that uses (1) an
+\begin{figure} \centering
+\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
+\caption{\label{fig:overview_sc2ts} A schematic of the sc2ts method. Sc2ts
+reconstructs the genetic relationships among SARS-CoV-2 genomes by copying
+samples to all possible ancestors collected at earlier time points (curved
+arrows). An ARG is built by attaching new sample genomes grouped by collection
+date to an existing ARG rooted at the Wuhan-Hu-1 reference genome, day by day.
+Each iteration involves three stages: (1) attachment of new samples to the
+growing ARG (A, B); (2) reconstruction of trees relating the samples under each
+attachment node (C); and parsimony-based tree topology adjustments (D, E). In
+the absence of  recombination, sc2ts infers an ARG that is a single tree
+relating the samples (A). However, in the presence of recombination, sc2ts
+infers an ARG that is encoded as a sequence of local trees relating segments of
+the sample genomes (B), inserting a recombination node to represent a putative
+recombination event. Additionally, mutation-collapsing nodes (D) and
+pushed-reversion nodes (E) are inserted to make more parsimonious placements of
+mutations that should be shared or should not be immediately reverted,
+respectively.}
+\end{figure}
+
+We introduce an ARG inference method, summarised in
+Figure~\ref{fig:overview_sc2ts}.
+% Note going to rewrite this anyway, so not trying to make it flow.
+It uses (1) an
 efficient HMM implementation of the Li and Stephens (LS) model \citep{Li2003-ib}
@@
 
-The generative process of the LS model is summarised in Figure 2. Here, a
-transition matrix, Q, governs the process of switching (recombining) between
+\begin{figure} \centering
+\includegraphics[width=0.6\textwidth]{figures/ls_diagram_covid.pdf}
+% FIXME this caption is too long
+\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS) model.
+The LS model in its implementation as an HMM is governed by the transition and
+emission matrices, $Q$ and $E$, respectively. Each may be a
+function of the reference panel members, but transitions are generally assumed
+to be independent of the hidden states themselves (shown in the pink panel).
+This dramatically increases performance as the state space drops to two states
+(i.e., switching or not switching). The recombination probability at site
+$\ell$ is $r\ell$, while n denotes the number of members in the
+reference panel. Emissions may also be a function of the nucleotide states, but
+in our case, we assume that all observed alleles (in both the reference panel
+and focal sequence, $a\ell$) at the site under consideration are equally
+likely and occur with mutation probability $\ell$ (shown in the blue
+panel). These two processes, transition and emission (which encode
+recombination and mutation, respectively), define the generative process of the
+HMM, by which an imperfect mosaic is emitted from a reference panel of samples
+and nodes from the ARG inserted at earlier collection dates (shown in the lower
+panel). Here, the unseen states in the reference panel are shown as coloured
+lines enclosed by the grey box, and the true path through the data which leads
+to the emitted focal sequence is displayed by the black arrow. Examples of
+transition and emission probabilities along this trajectory are shown by the
+red and blue arrows, respectively. Black crosses along the focal sequence show
+genotyping errors or mutations. Note that in practice, we do not know the
+pattern of colours along the focal sequence and must infer it using the Viterbi
+algorithm.} \end{figure}
+
+The generative process of the LS model is summarised in
+Figure~\ref{fig:ls_diagram}.
+Here, a
+transition matrix, $Q$, governs the process of switching (recombining) between
 members of the reference panel (the hidden states). An emission matrix,
-\textit{E}, allows for differences between the focal sequence and the hidden
+$E$, allows for differences between the focal sequence and the hidden
 state from which it is copied (due to mutation as well as genotyping or
@@
 find the most likely copying path through the sequences, given a set of
-reference sequences, \textit{Q}, and \textit{E}.
+reference sequences, $Q$, and $E$.
 
@@
 pandemic-scale ARG.
+
+\begin{table}
+\centering
+\begin{tabular}{l|l|l} \hline \textbf{Property} &
+\textbf{Wide ARG} & \textbf{Long ARG} \\ \hline Last sampling date & 2021-06-30
+& 2022-06-30 \\ Local trees & 1,496 & 958 \\ \hline
+\multicolumn{3}{l}{\textit{Nodes}} \\ All nodes (sample and non-sample) &
+1,453,347 & 783,231 \\ Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\ Tree
+adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107
+(10.97\%) \\ Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\ \hline
+\multicolumn{3}{l}{\textit{Types of recombination nodes}} \\ Singletons (i.e.
+no descendant besides the sequence causing insertion of the node) & 2,601
+(63\%) & 1,315 (63\%) \\ Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%)
+\\ HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\ \hline
+\multicolumn{3}{l}{\textit{Mutations}} \\ Total mutations & 1,213,193 &
+1,062,072 \\ Mutations on a terminal edge & 758,903 (62.55\%) & 767,111
+(72.23\%) \\ Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
+Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\ Immediate reversions (one
+intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\ \hline
+\multicolumn{3}{l}{\textit{Storage and loading time}} \\ tszipped file
+(including sample metadata, imputed Pango lineage assignments, and results of
+HMM runs for recombination nodes) & ~57 Mb & ~36 Mb \\ Bytes to encode one
+sample & 8.29 & 10.83 \\ Load time (seconds) & 0.9 & 0.5 \\ \hline
+\end{tabular}
+\end{table}
 
@@
 
+\begin{table} \centering
+\begin{tabular}{l|c|c|c|c|c} \hline
+\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
+\multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\ \hline
+\textbf{Sample} &
+\textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\
+interval(s)} &
+\textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\
+\hline ALDP-11CF93B & A
+& \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
+21,256–22,227 \\ ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
+\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ ALDP-130BB95 & A &
+\thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
+21,256–22,227 \\ LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
+\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ QEUH-CCCB30 & B &
+\thead{B.1.36.28 \\ Alpha} & 6,528–6,953 & \thead{B.1.36 \\ Alpha} &
+6,529–6,954 \\ QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
+\thead{B.1.36 \\ Alpha} & 6,529–6,954 \\ MILK-1166F52 & C & \thead{Alpha \\
+B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
+MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\
+B.1.221} & 25,997–27,972 \\ QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} &
+25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\ MILK-126FE1F & D &
+\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
+22,445–23,063 \\ RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062
+& \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\ RAND-128FA33 & D &
+\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
+22,445–23,063 \\ CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
+\thead{B.1.177 \\ Alpha} & 16,177–21,255 \\ CAMC-CB7AB3 & n/a & \thead{Alpha \\
+B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} & \thead{Alpha \\
+B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\ MILK-103C712 & n/a
+& \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} & n/a & n/a \\
+QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 & \thead{Alpha
+\\ B.1.177} & 7,729–10,870 \\ \hline
+\end{tabular}
+\caption{\label{tab:jackson}Recombinant sequences involving the Alpha (B.1.1.7)
+variant reported by \cite{Jackson2021-ik} have recombinant
+ancestry in the wide ARG. The breakpoint intervals and Pango lineage
+assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et
+al., except the Pango lineage assignment of the parents of group B
+recombinants, which were taken from Table 1 (motif-based results).} \end{table}
+
 In the wide ARG (sampled to mid-2021), all these sequences are descendants of
@@
 hypothesis that the sequences of the groups A to D emerged from four separate
-recombination events (Jackson et al. 2021).
+recombination events~\cite{Jackson2021-ik}.
 
@@
 mutations, respectively.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
+\caption{\label{fig:mutational_spectra}
+Mutational spectrum of SARS-CoV-2 in
+the wide ARG (sampled to mid-2021) compared to \cite{Yi2021-sc}.
+Mutations are categorised by type (i.e., ancestral state >
+derived state). The percentages of each mutation type from the wide ARG are
+represented by blue bars and the percentages from Yi et al. by orange bars,
+with the darker colours representing one direction (e.g., C > U) and the
+lighter colours the reverse (e.g., U > C).}
+\end{figure}
 
@@
 
-\begin{table} \centering \begin{tabular}{l|l|l} \hline \textbf{Property} &
-\textbf{Wide ARG} & \textbf{Long ARG} \\ \hline Last sampling date & 2021-06-30
-& 2022-06-30 \\ Local trees & 1,496 & 958 \\ \hline
-\multicolumn{3}{l}{\textit{Nodes}} \\ All nodes (sample and non-sample) &
-1,453,347 & 783,231 \\ Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\ Tree
-adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107
-(10.97\%) \\ Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\ \hline
-\multicolumn{3}{l}{\textit{Types of recombination nodes}} \\ Singletons (i.e.
-no descendant besides the sequence causing insertion of the node) & 2,601
-(63\%) & 1,315 (63\%) \\ Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%)
-\\ HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\ \hline
-\multicolumn{3}{l}{\textit{Mutations}} \\ Total mutations & 1,213,193 &
-1,062,072 \\ Mutations on a terminal edge & 758,903 (62.55\%) & 767,111
-(72.23\%) \\ Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
-Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\ Immediate reversions (one
-intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\ \hline
-\multicolumn{3}{l}{\textit{Storage and loading time}} \\ tszipped file
-(including sample metadata, imputed Pango lineage assignments, and results of
-HMM runs for recombination nodes) & ~57 Mb & ~36 Mb \\ Bytes to encode one
-sample & 8.29 & 10.83 \\ Load time (seconds) & 0.9 & 0.5 \\ \hline
-\end{tabular} \end{table}
-
-\begin{table} \centering \begin{tabular}{l|c|c|c|c|c} \hline
-\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
-\multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\ \hline \textbf{Sample} &
-\textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\ interval(s)} &
-\textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\ \hline ALDP-11CF93B & A
-& \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
-21,256–22,227 \\ ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
-\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ ALDP-130BB95 & A &
-\thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
-21,256–22,227 \\ LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
-\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ QEUH-CCCB30 & B &
-\thead{B.1.36.28 \\ Alpha} & 6,528–6,953 & \thead{B.1.36 \\ Alpha} &
-6,529–6,954 \\ QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
-\thead{B.1.36 \\ Alpha} & 6,529–6,954 \\ MILK-1166F52 & C & \thead{Alpha \\
-B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
-MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\
-B.1.221} & 25,997–27,972 \\ QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} &
-25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\ MILK-126FE1F & D &
-\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
-22,445–23,063 \\ RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062
-& \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\ RAND-128FA33 & D &
-\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
-22,445–23,063 \\ CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
-\thead{B.1.177 \\ Alpha} & 16,177–21,255 \\ CAMC-CB7AB3 & n/a & \thead{Alpha \\
-B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} & \thead{Alpha \\
-B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\ MILK-103C712 & n/a
-& \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} & n/a & n/a \\
-QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 & \thead{Alpha
-\\ B.1.177} & 7,729–10,870 \\ \hline \end{tabular}
-\caption{\label{tab:jackson}Recombinant sequences involving the Alpha (B.1.1.7)
-variant reported by Jackson et al. \citep{Jackson2021-ik} have recombinant
-ancestry in the wide ARG. The breakpoint intervals and Pango lineage
-assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et
-al., except the Pango lineage assignment of the parents of group B
-recombinants, which were taken from Table 1 (motif-based results).} \end{table}
-
-\begin{figure} \centering
-\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
-\caption{\label{fig:overview_sc2ts} A schematic of the sc2ts method. Sc2ts
-reconstructs the genetic relationships among SARS-CoV-2 genomes by copying
-samples to all possible ancestors collected at earlier time points (curved
-arrows). An ARG is built by attaching new sample genomes grouped by collection
-date to an existing ARG rooted at the Wuhan-Hu-1 reference genome, day by day.
-Each iteration involves three stages: (1) attachment of new samples to the
-growing ARG (A, B); (2) reconstruction of trees relating the samples under each
-attachment node (C); and parsimony-based tree topology adjustments (D, E). In
-the absence of  recombination, sc2ts infers an ARG that is a single tree
-relating the samples (A). However, in the presence of recombination, sc2ts
-infers an ARG that is encoded as a sequence of local trees relating segments of
-the sample genomes (B), inserting a recombination node to represent a putative
-recombination event. Additionally, mutation-collapsing nodes (D) and
-pushed-reversion nodes (E) are inserted to make more parsimonious placements of
-mutations that should be shared or should not be immediately reverted,
-respectively.} \end{figure}
-
-\begin{figure} \centering
-\includegraphics[width=0.8\textwidth]{figures/ls_diagram_covid.pdf}
-\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS) model.
-The LS model in its implementation as an HMM is governed by the transition and
-emission matrices, \textit{Q} and \textit{E}, respectively. Each may be a
-function of the reference panel members, but transitions are generally assumed
-to be independent of the hidden states themselves (shown in the pink panel).
-This dramatically increases performance as the state space drops to two states
-(i.e., switching or not switching). The recombination probability at site
-\textit{l} is \textit{rl}, while n denotes the number of members in the
-reference panel. Emissions may also be a function of the nucleotide states, but
-in our case, we assume that all observed alleles (in both the reference panel
-and focal sequence, \textit{al}) at the site under consideration are equally
-likely and occur with mutation probability \textit{l} (shown in the blue
-panel). These two processes, transition and emission (which encode
-recombination and mutation, respectively), define the generative process of the
-HMM, by which an imperfect mosaic is emitted from a reference panel of samples
-and nodes from the ARG inserted at earlier collection dates (shown in the lower
-panel). Here, the unseen states in the reference panel are shown as coloured
-lines enclosed by the grey box, and the true path through the data which leads
-to the emitted focal sequence is displayed by the black arrow. Examples of
-transition and emission probabilities along this trajectory are shown by the
-red and blue arrows, respectively. Black crosses along the focal sequence show
-genotyping errors or mutations. Note that in practice, we do not know the
-pattern of colours along the focal sequence and must infer it using the Viterbi
-algorithm.} \end{figure}
-
-\begin{figure} \centering
-\includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
-\caption{\label{fig:mutational_spectra} Mutational spectrum of SARS-CoV-2 in
-the wide ARG (sampled to mid-2021) compared to that reported by Yi et al.
-\citep{Yi2021-sc}. Mutations are categorised by type (i.e., ancestral state >
-derived state). The percentages of each mutation type from the wide ARG are
-represented by blue bars and the percentages from Yi et al. by orange bars,
-with the darker colours representing one direction (e.g., C > U) and the
-lighter colours the reverse (e.g., U > C).} \end{figure}
-
 \bibliographystyle{plainnat}

### 2023-04-06 paper.tex

@@
 \end{tabular}
+\caption{\label{tab:args}Summary of the inferred ARGs.
+}
 \end{table}
@@
 fraction of a second using the Python package tskit (Table 1).
-
-\subsection{Computational resources}
 
@@
 restricted to non-recombinant lineages. To allow for comparison, the ARGs were
-"simplified" \citep{Kelleher2018-xc} to retain only those tips of our genealogy
-that match the samples present in a specific NextStrain tree. These simplified
-"backbone" ARGs exhibit very little recombination (6 and 5 trees across the
+``simplified'' \citep{Kelleher2018-xc} to retain only those samples of our genealogy
+that match those present in a specific NextStrain tree. These simplified
+``backbone'' ARGs exhibit very little recombination (6 and 5 trees across the
 genome for the wide and long ARGs, respectively).
@@
 
-\subsection{Putative recombination events}
+\subsection{Mutational spectrum}
+
+In addition to recombination events, the ARGs record mutation events along the
+SARS-CoV-2 phylogeny. \cite{Yi2021-sc} reconstructed a SARS-CoV-2
+phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to
+January 12, 2021, and systematically catalogued all the mutations occurring
+along the phylogeny. Notably, in the mutational spectrum described by Yi et
+al., C-to-U and G-to-U mutations occur more frequently than U-to-C and U-to-G
+mutations, respectively.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
+\caption{\label{fig:mutational_spectra}
+Mutational spectrum of SARS-CoV-2 in
+the wide ARG (sampled to mid-2021) compared to \cite{Yi2021-sc}.
+Mutations are categorised by type (i.e., ancestral state $>$
+derived state). The percentages of each mutation type from the wide ARG are
+represented by blue bars and the percentages from Yi et al. by orange bars,
+with the darker colours representing one direction (e.g., C$>$U) and the
+lighter colours the reverse (e.g., U$>$C).}
+\end{figure}
+
+For validation purposes, we compare the mutational spectrum from the wide ARG
+with the mutational spectrum of Yi et al. We categorised all of the single
+nucleotide mutations by type (defined by the ancestral state and derived
+state), excluded mutations inherited by only a single sample (which are
+enriched for sequencing errors and artefacts), and tallied them up by type.
+Similarly, we took the data for single nucleotide mutations from Yi et al.
+\url{https://github.com/ju-lab/SC2_evol_signature}, excluded mutations
+occurring along terminal branches, and tallied them up by type. We find that
+the mutational spectrum from the wide ARG (448,825 mutations) matches the
+mutational spectrum reported by Yi et al. (92,344 mutations) (Figure X). In
+both spectra, C-to-U mutations and G-to-U mutations occur more frequently than
+U-to-C and U-to-G, respectively (Figure X). Similar results are obtained when
+including the mutations inherited by only a single sample or occurring on
+terminal branches (data not shown).
+
+% If we do this, goes here
+% \subsection{Mutations under selection}
+% XX
+
+\subsection{Early recombinants reported by Jackson et al. (2021)}
+% - Early Recombinants from Jackson et al
+%     Discuss results from Wide ARG in detail. Go into the details of the
+% forward-backward HMM runs for each example. Be much more explicity about the
+% single origin for the different groups (maybe give one of Ana's graphs for
+% Group A (XA I think?)
+
+By searching for samples combining genomic segments from Alpha (B.1.1.7) and
+from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations,
+Jackson et al. (Jackson et al. 2021) reported 16 recombinant sequences among
+genomes sampled in the United Kingdom up to March 7, 2021. The authors proposed
+that the recombinants likely arose from eight independent recombination events
+(categorised into groups A to D and four singletons).
+
+\begin{table} \centering
+\begin{tabular}{l|c|c|c|c|c} \hline
+\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
+\multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\ \hline
+\textbf{Sample} &
+\textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\
+interval(s)} &
+\textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\
+\hline ALDP-11CF93B & A
+& \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
+21,256–22,227 \\ ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
+\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ ALDP-130BB95 & A &
+\thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
+21,256–22,227 \\ LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
+\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ QEUH-CCCB30 & B &
+\thead{B.1.36.28 \\ Alpha} & 6,528–6,953 & \thead{B.1.36 \\ Alpha} &
+6,529–6,954 \\ QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
+\thead{B.1.36 \\ Alpha} & 6,529–6,954 \\ MILK-1166F52 & C & \thead{Alpha \\
+B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
+MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\
+B.1.221} & 25,997–27,972 \\ QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} &
+25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\ MILK-126FE1F & D &
+\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
+22,445–23,063 \\ RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062
+& \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\ RAND-128FA33 & D &
+\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
+22,445–23,063 \\ CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
+\thead{B.1.177 \\ Alpha} & 16,177–21,255 \\ CAMC-CB7AB3 & n/a & \thead{Alpha \\
+B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} & \thead{Alpha \\
+B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\ MILK-103C712 & n/a
+& \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} & n/a & n/a \\
+QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 & \thead{Alpha
+\\ B.1.177} & 7,729–10,870 \\ \hline
+\end{tabular}
+\caption{\label{tab:jackson}Recombinant sequences involving the Alpha (B.1.1.7)
+variant reported by \cite{Jackson2021-ik} have recombinant
+ancestry in the wide ARG. The breakpoint intervals and Pango lineage
+assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et
+al., except the Pango lineage assignment of the parents of group B
+recombinants, which were taken from Table 1 (motif-based results).}
+\end{table}
+
+In the wide ARG (sampled to mid-2021), all these sequences are descendants of
+recombination nodes, except MILK-103C712, which was removed during
+preprocessing (Table 2); additionally, all are HMM-consistent, except
+QEUH-1067DEF. Overall, we find excellent agreement between Jackson et al.’s
+motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
+and (2) the breakpoint intervals (Table 2). In all cases, the two methods
+proposed identical or very closely related parent lineages. The general mosaic
+genome structure is concordant in all the cases, even CAMC-CB7AB3 where there
+are two switches in the HMM copying paths (indicative of two breakpoints). The
+breakpoint intervals estimated using the two methods overlap substantially in
+all the cases (Table 2). Within each of the groups A to D, the constituent
+sequences have identical breakpoint intervals, because they trace back to a
+single recombination node in the wide ARG. This result is consistent with the
+hypothesis that the sequences of the groups A to D emerged from four separate
+recombination events~\cite{Jackson2021-ik}.
+
+\subsection{Distribution of recombination breakpoint intervals}
+
+% - Distribution of breakpoint intervals
+%     Study the breakpoint intervals in the Wide ARG. Because we are interested
+% in the intervals, we want to be sure that these are robustly reported by the
+% HMM. Hence, we focus on those that have two parents in the mirrored and
+% unmirrored HMM runs, the same set of mutations, and the left and right parents
+% are consistently imputed for Pango lineage status.
+%     We focus on the Wide ARG here because it covers (roughly?) the same time
+% period as the RIPPLES analysis.
+
+We explore recombination signals across the SARS-CoV-2 genome by inspecting the
+breakpoint intervals of the filtered recombination nodes from the wide ARG
+(sampled to mid-2021) in Figure~\ref{fig:breakpoint-distribution}.
+The length of a breakpoint interval reflects
+uncertainty in the estimated location of a breakpoint. On average, the length
+of the breakpoint intervals is 2,895 bases (median, 1,670 bases; range, 1 to
+23,666 bases). This broad distribution of interval length indicates that the
+location of a breakpoint often could not be precisely determined given the
+observed genetic variation.
+
+\begin{figure}
+\centering
+% \includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
+\caption{\label{fig:breakpoint-distribution}
+Distribution of recombination breakpoints and mutations along the genome.}
+\end{figure}
+
+There is a preponderance of breakpoint intervals spanning the 3’ end of the
+genome (Figure 4A), as noted previously by
+\cite{Turakhia2022-it}.
+Inspecting more closely, we find an enrichment in the
+number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region,
+and 3’ untranslated region (Figure 4B). Markedly, there is an abrupt drop in
+the number of intersecting intervals in the N gene following the peak in ORF8
+towards the 3’ end.
+
+Next, we investigate the potential recombination signals alongside the number
+of mutations across the genome (Figure 4C). Regions with a higher density of
+mutations contain more genetic variation, increasing the power to detect
+recombination events. The position 28,271 (located in the ORF8-N intergenic
+region) has a particularly high number of mutations in the samples (Figure 5).
+At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions
+(n = 1,605) appear to occur frequently. One possible explanation is that this
+homopolymeric region is prone to sequencing errors, giving rise to an
+artifactual recombination signal. At the upstream position 27,972, numerous
+mutations also occur. This position has been reported to be hypermutable
+\citep{Jungreis2021-dh}.
+
+We further consider the breakpoint intervals within the context of
+recombination breakpoint sequence motifs. It is hypothesised that certain
+palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote
+template switching during replication in SARS-CoV-2 via formation of
+base-paired stem loops in the genome structure \citep{Gallaher2020-lb}. There
+are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019
+reference sequence (Supplementary Table 5). We observe that XX (XX\%) of the
+breakpoint intervals of the HMM-consistent recombinants span at least one of
+the breakpoint sequences.
+
+\subsection{Divergence between recombinant parents}
+% - Divergence between recombinant parents
+%     We study the recombination events in the Long ARG in terms. A recombinant
+% will not always have two parents, for reasons X, Y and Z. As a first
+% approximation, we can assume that each breakpoint corresponds to a
+% recombination event (although there are some exceptions). Here we have to get
+% into all the hairy details about which sets of breakpoints we filter and don't
+% filter.
 
@@
 
-\subsection{Early recombinants reported by Jackson et al. (2021)}
+\subsection{Recombinant Pango lineages}
 
-By searching for samples combining genomic segments from Alpha (B.1.1.7) and
-from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations,
-Jackson et al. (Jackson et al. 2021) reported 16 recombinant sequences among
-genomes sampled in the United Kingdom up to March 7, 2021. The authors proposed
-that the recombinants likely arose from eight independent recombination events
-(categorised into groups A to D and four singletons).
+In this section we examine the Long ARG to determine how well it reflects
+the structure of previously identified recombinant lineages.
+There are [high level summary of the pango recombinant lineages and samples
+in the Long ARG].
+long ARG (subsampled to mid-2022) contains samples
+from 33 Pango-designated recombinant lineages (XA to XAK; Supplementary Table)
+[Probably could have a table here to summarise?]
 
-\begin{table} \centering
-\begin{tabular}{l|c|c|c|c|c} \hline
-\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
-\multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\ \hline
-\textbf{Sample} &
-\textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\
-interval(s)} &
-\textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\
-\hline ALDP-11CF93B & A
-& \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
-21,256–22,227 \\ ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
-\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ ALDP-130BB95 & A &
-\thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
-21,256–22,227 \\ LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
-\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ QEUH-CCCB30 & B &
-\thead{B.1.36.28 \\ Alpha} & 6,528–6,953 & \thead{B.1.36 \\ Alpha} &
-6,529–6,954 \\ QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
-\thead{B.1.36 \\ Alpha} & 6,529–6,954 \\ MILK-1166F52 & C & \thead{Alpha \\
-B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
-MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\
-B.1.221} & 25,997–27,972 \\ QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} &
-25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\ MILK-126FE1F & D &
-\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
-22,445–23,063 \\ RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062
-& \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\ RAND-128FA33 & D &
-\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
-22,445–23,063 \\ CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
-\thead{B.1.177 \\ Alpha} & 16,177–21,255 \\ CAMC-CB7AB3 & n/a & \thead{Alpha \\
-B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} & \thead{Alpha \\
-B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\ MILK-103C712 & n/a
-& \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} & n/a & n/a \\
-QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 & \thead{Alpha
-\\ B.1.177} & 7,729–10,870 \\ \hline
-\end{tabular}
-\caption{\label{tab:jackson}Recombinant sequences involving the Alpha (B.1.1.7)
-variant reported by \cite{Jackson2021-ik} have recombinant
-ancestry in the wide ARG. The breakpoint intervals and Pango lineage
-assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et
-al., except the Pango lineage assignment of the parents of group B
-recombinants, which were taken from Table 1 (motif-based results).} \end{table}
-
-In the wide ARG (sampled to mid-2021), all these sequences are descendants of
-recombination nodes, except MILK-103C712, which was removed during
-preprocessing (Table 2); additionally, all are HMM-consistent, except
-QEUH-1067DEF. Overall, we find excellent agreement between Jackson et al.’s
-motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
-and (2) the breakpoint intervals (Table 2). In all cases, the two methods
-proposed identical or very closely related parent lineages. The general mosaic
-genome structure is concordant in all the cases, even CAMC-CB7AB3 where there
-are two switches in the HMM copying paths (indicative of two breakpoints). The
-breakpoint intervals estimated using the two methods overlap substantially in
-all the cases (Table 2). Within each of the groups A to D, the constituent
-sequences have identical breakpoint intervals, because they trace back to a
-single recombination node in the wide ARG. This result is consistent with the
-hypothesis that the sequences of the groups A to D emerged from four separate
-recombination events~\cite{Jackson2021-ik}.
-
-\subsection{Pango recombinants}
-
-The wide ARG only contains samples from three Pango-designated recombinant
-lineages (XA, XB, and XC), due to the fact that it only includes samples up to
-mid-2021. By contrast, the long ARG (subsampled to mid-2022) contains samples
-from 33 Pango-designated recombinant lineages (XA to XAK; Supplementary Table
-3). For Pango recombinant analysis, we therefore focus on the long ARG.
-
-In the wide ARG, every sequence of the three Pango recombinants trace back to
-some recombination node. XA, XB, and XC.
-
-In the unfiltered long ARG, the sequences of 28 Pango recombinants trace back
+The sequences of 28 Pango recombinants trace back
 to some recombinant node(s); however, the sequences of the five Pango
-recombinants XP, XU, XAA, XAG, and XAK do not. A likely explanation for the XP
+% jk: really? I thought it was just XP that wasn't inferred to be a
+% recombinant.
+recombinants XP, XU, XAA, XAG, and XAK do not.
+A likely explanation for the XP
 recombinant sequences not having a recombinant origin in the long ARG is that
 its characteristic multibase deletion
-(https://github.com/cov-lineages/pango-designation/issues/481) was masked
-during preprocessing and therefore was not incorporated into the HMM. Upon
-tracing the origin of XAG through the ARG, XAG might have been caused by
+(\url{https://github.com/cov-lineages/pango-designation/issues/481}) was masked
+during preprocessing and therefore was not incorporated into the HMM.
+% Eh?
+Upon tracing the origin of XAG through the ARG, XAG might have been caused by
 incorrect imputation of its parent nodes(?). XAK likely does not appear to have
@@
 
-The sequences of 14 Pango recombinants trace back to a single recombination
-node. The sequences of the remaining 18 Pango recombinants trace back to
-multiple recombination nodes, however. This may be a result of undersampling in
-the long ARG, which was built by randomly subsampling at most 1,000 sequences
-from each daily batch. Having extra recombination nodes in the long ARG
-appeared to be a more parsimonious explanation for the 19 Pango recombinants
-than having a single recombination node with many immediately ancestral
-mutations (including recurrent mutations and reversions), despite that the
-recombinants may indeed trace to a single origin.
-
-The mosaic genome structure of all the 32 Pango recombinants is concordant with
+[FIXME this isn't quite right - not all the sequences for a given pango
+lineage trace back to the same recombination event, so won't have the same
+intervals.]
+The mosaic genome structure of all the 32 Pango recombinant lineages is concordant with
 the structure determined by the Pango Lineage Designation Committee
@@
 
+The wide ARG contains samples from three Pango-designated recombinant
+lineages (XA, XB, and XC) and N samples, since it only includes data
+up to mid-2021.
+In the wide ARG, every sequence of the three Pango recombinants trace back to
+some recombination node. XA, XB, and XC.
+
+For the long ARG, we examine the lineages in increasing complexity of the
+inferred evolutionary relationships between samples assigned to Pango
+X lineages.
+
+% - Recombinant Pango lineages
+%     How well do we recover recombinant Pango lineages? Work through the
+% examples in increasing compexity, using headings to split things up if
+% necessary. (i.e., Single origin, Multiple origin, Interellated lineages). We'd
+% need to be clear about what these terms mean, of course. We could include
+% lineage graphs for all of these examples?
+
+% OLD TEXT
+
+% The sequences of 14 Pango recombinants trace back to a single recombination
+% node. The sequences of the remaining 18 Pango recombinants trace back to
+% multiple recombination nodes, however. This may be a result of undersampling in
+% the long ARG, which was built by randomly subsampling at most 1,000 sequences
+% from each daily batch. Having extra recombination nodes in the long ARG
+% appeared to be a more parsimonious explanation for the 19 Pango recombinants
+% than having a single recombination node with many immediately ancestral
+% mutations (including recurrent mutations and reversions), despite that the
+% recombinants may indeed trace to a single origin.
+
+\subsubsection{Single origin}
+
+We would usually assume that all sequences assigned to a given Pango X
+lineage are descendants of a single recombinant sequence, arising as a
+result of a mixed infection followed by onward transmission. Thus, we
+would also expect our ARG to reflect this evolutionary history, where
+we have a all the samples assigned to this recombinant lineage tracing
+back to a single recombination node, representing this original
+recombination event.
+
+The sequences of 14 Pango recombinants trace back to a single recombination
+node. The details are given in Figure X, where we [give the details of
+the numbers of samples, etc for each of these 14 lineages] and
+show the lineage subgraph for X?, which makes this simple structure
+clear. [Explain the lineage subgraph format a bit]
+
+\subsubsection{Multiple origin}
+The ARG inference has no pre-defined knowledge of Pango X lineage assignments,
+and there is therefore no particular requirement that all the samples
+assigned to a given lineage must trace back to a single recombination
+event. In this section we examine the Pango X lineage samples that
+are inferred to have multiple recombinant origins (i.e., trace
+back to different recombination nodes in the ARG), but are still
+form monophyletic clades % is this the right term?
+descending from these nodes .
+
+In Figure XXX we give the details of these multiple pango lineages,
+along with an example lineage subgraph.
+
+Interestingly, XC is inferred to be single origin in the Wide ARG but
+not in the long ARG. See lineage subgraphs in supplemental figure Y.
+This suggests that insufficient sampling may be the explanation for
+many of these multiple origin events.
+
+\subsubsection{Complex relationships between X lineages}
+
+Some lineages are inferred to have overlapping ancestral histories.
+Discuss XAG example in Figure XXX.
+
 \subsection{Novel recombinants}
 
-Next, we examine recombination nodes in the filtered long ARG (sampled to
+Next, we examine recombination nodes in the long ARG (sampled to
 mid-2022) that represent previously unnamed putative recombinant sequences,
@@
 study, we have arbitrarily picked 12 recombination nodes that seem plausible
-(Supplementary Table 4). Coincidentally, all these nodes involve Omicron
+(Supplementary Table 4). All these nodes involve Omicron
 subvariants (BA.1, BA.2, BA.4, and BA.5). These recombination nodes (1) were
@@
 sister clade of XAC, and annotated as ``miscBA2BA1PostSpike''.
-
-\subsection{Recombinant origins of the Pango lineages XA and XAG}
-
-XX
-
-\subsection{Distribution of recombination breakpoint intervals}
-
-We explore recombination signals across the SARS-CoV-2 genome by inspecting the
-breakpoint intervals of the filtered recombination nodes from the wide ARG
-(sampled to mid-2021) (Figure 4). The length of a breakpoint interval reflects
-uncertainty in the estimated location of a breakpoint. On average, the length
-of the breakpoint intervals is 2,895 bases (median, 1,670 bases; range, 1 to
-23,666 bases). This broad distribution of interval length indicates that the
-location of a breakpoint often could not be precisely determined given the
-observed genetic variation.
-
-There is a preponderance of breakpoint intervals spanning the 3’ end of the
-genome (Figure 4A), as noted previously by
-\cite{Turakhia2022-it}.
-Inspecting more closely, we find an enrichment in the
-number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region,
-and 3’ untranslated region (Figure 4B). Markedly, there is an abrupt drop in
-the number of intersecting intervals in the N gene following the peak in ORF8
-towards the 3’ end.
-
-Next, we investigate the potential recombination signals alongside the number
-of mutations across the genome (Figure 4C). Regions with a higher density of
-mutations contain more genetic variation, increasing the power to detect
-recombination events. The position 28,271 (located in the ORF8-N intergenic
-region) has a particularly high number of mutations in the samples (Figure 5).
-At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions
-(n = 1,605) appear to occur frequently. One possible explanation is that this
-homopolymeric region is prone to sequencing errors, giving rise to an
-artifactual recombination signal. At the upstream position 27,972, numerous
-mutations also occur. This position has been reported to be hypermutable
-\citep{Jungreis2021-dh}.
-
-We further consider the breakpoint intervals within the context of
-recombination breakpoint sequence motifs. It is hypothesised that certain
-palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote
-template switching during replication in SARS-CoV-2 via formation of
-base-paired stem loops in the genome structure \citep{Gallaher2020-lb}. There
-are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019
-reference sequence (Supplementary Table 5). We observe that XX (XX\%) of the
-breakpoint intervals of the HMM-consistent recombinants span at least one of
-the breakpoint sequences.
-
-\subsection{Mutational spectrum}
-
-In addition to recombination events, the ARGs record mutation events along the
-SARS-CoV-2 phylogeny. \cite{Yi2021-sc} reconstructed a SARS-CoV-2
-phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to
-January 12, 2021, and systematically catalogued all the mutations occurring
-along the phylogeny. Notably, in the mutational spectrum described by Yi et
-al., C-to-U and G-to-U mutations occur more frequently than U-to-C and U-to-G
-mutations, respectively.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
-\caption{\label{fig:mutational_spectra}
-Mutational spectrum of SARS-CoV-2 in
-the wide ARG (sampled to mid-2021) compared to \cite{Yi2021-sc}.
-Mutations are categorised by type (i.e., ancestral state >
-derived state). The percentages of each mutation type from the wide ARG are
-represented by blue bars and the percentages from Yi et al. by orange bars,
-with the darker colours representing one direction (e.g., C > U) and the
-lighter colours the reverse (e.g., U > C).}
-\end{figure}
-
-For validation purposes, we compare the mutational spectrum from the wide ARG
-with the mutational spectrum of Yi et al. We categorised all of the single
-nucleotide mutations by type (defined by the ancestral state and derived
-state), excluded mutations inherited by only a single sample (which are
-enriched for sequencing errors and artefacts), and tallied them up by type.
-Similarly, we took the data for single nucleotide mutations from Yi et al.
-\url{https://github.com/ju-lab/SC2_evol_signature}, excluded mutations
-occurring along terminal branches, and tallied them up by type. We find that
-the mutational spectrum from the wide ARG (448,825 mutations) matches the
-mutational spectrum reported by Yi et al. (92,344 mutations) (Figure X). In
-both spectra, C-to-U mutations and G-to-U mutations occur more frequently than
-U-to-C and U-to-G, respectively (Figure X). Similar results are obtained when
-including the mutations inherited by only a single sample or occurring on
-terminal branches (data not shown).
-
-\subsection{Mutations under selection}
-
-XX
 

### 2023-04-06 paper.tex
Some text formatting for Table 2

@@
 
+% FIXME there's too much vspace between the rows here. Probably best to redo
+% the formatting.
 \begin{table} \centering
@@
     & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
-21,256–22,227 \\ ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
-\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ ALDP-130BB95 & A &
-\thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
-21,256–22,227 \\ LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
-\thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\ QEUH-CCCB30 & B &
-\thead{B.1.36.28 \\ Alpha} & 6,528–6,953 & \thead{B.1.36 \\ Alpha} &
-6,529–6,954 \\ QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
-\thead{B.1.36 \\ Alpha} & 6,529–6,954 \\ MILK-1166F52 & C & \thead{Alpha \\
-B.1.221.1} & 25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
-MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 & \thead{Alpha \\
-B.1.221} & 25,997–27,972 \\ QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} &
-25,996–27,441 & \thead{Alpha \\ B.1.221} & 25,997–27,972 \\ MILK-126FE1F & D &
-\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
-22,445–23,063 \\ RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062
-& \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\ RAND-128FA33 & D &
-\thead{B.1.36.39 \\ Alpha} & 20,703–23,062 & \thead{B.1.36.39 \\ Alpha} &
-22,445–23,063 \\ CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
-\thead{B.1.177 \\ Alpha} & 16,177–21,255 \\ CAMC-CB7AB3 & n/a & \thead{Alpha \\
-B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} & \thead{Alpha \\
-B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\ MILK-103C712 & n/a
-& \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} & n/a & n/a \\
-QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 & \thead{Alpha
-\\ B.1.177} & 7,729–10,870 \\ \hline
+21,256–22,227 \\
+    ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
+ALDP-130BB95 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
+LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
+QEUH-CCCB30 & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
+    \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
+QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
+    \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
+MILK-1166F52 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
+    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
+MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
+    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
+QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
+    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
+MILK-126FE1F & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
+    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
+RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
+    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
+RAND-128FA33 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
+    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
+CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
+    \thead{B.1.177 \\ Alpha} & 16,177–21,255 \\
+CAMC-CB7AB3 & n/a & \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} &
+     \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\
+MILK-103C712 & n/a & \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} &
+    n/a & n/a \\
+QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 &
+    \thead{Alpha \\ B.1.177} & 7,729–10,870 \\ \hline
 \end{tabular}

### 2023-04-06 paper.tex
Update table 1

@@
 
+\usepackage{booktabs}
 \usepackage{authblk}
@@
 \begin{table}
-\centering
-\begin{tabular}{l|l|l} \hline \textbf{Property} &
-\textbf{Wide ARG} & \textbf{Long ARG} \\ \hline Last sampling date & 2021-06-30
-& 2022-06-30 \\ Local trees & 1,496 & 958 \\ \hline
-\multicolumn{3}{l}{\textit{Nodes}} \\ All nodes (sample and non-sample) &
-1,453,347 & 783,231 \\ Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\ Tree
-adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107
-(10.97\%) \\ Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\ \hline
-\multicolumn{3}{l}{\textit{Types of recombination nodes}} \\ Singletons (i.e.
-no descendant besides the sequence causing insertion of the node) & 2,601
-(63\%) & 1,315 (63\%) \\ Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%)
-\\ HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\ \hline
-\multicolumn{3}{l}{\textit{Mutations}} \\ Total mutations & 1,213,193 &
-1,062,072 \\ Mutations on a terminal edge & 758,903 (62.55\%) & 767,111
-(72.23\%) \\ Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
-Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\ Immediate reversions (one
-intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\ \hline
-\multicolumn{3}{l}{\textit{Storage and loading time}} \\ tszipped file
-(including sample metadata, imputed Pango lineage assignments, and results of
-HMM runs for recombination nodes) & ~57 Mb & ~36 Mb \\ Bytes to encode one
-sample & 8.29 & 10.83 \\ Load time (seconds) & 0.9 & 0.5 \\ \hline
+\begin{center}
+    \begin{tabular}{llrlrl}\toprule
+        &  & \multicolumn{2}{l}{Wide ARG} & \multicolumn{2}{l}{Long ARG} \\
+    \cmidrule{3-6}
+    \multicolumn{2}{l}{Sample filtering}
+             & \multicolumn{2}{l}{collection $\leq$ 2021-06-30}
+                 & \multicolumn{2}{l}{collection $\leq$ 2022-06-30} \\
+           & & \multicolumn{2}{l}{max-delay=30} & \multicolumn{2}{l}{max-delay=30} \\
+        &  & &                               & \multicolumn{2}{l}{max-daily=1000} \\
+    \midrule
+
+    Nodes & & 1,453,347 & & 783,231  & \\
+    & Samples & 1,265,685 & (87\%) & 657,239 & (84\%) \\
+    & Local (UPGMA) & 103K & (7.1\%) & 52K & (6.6\%)  \\
+    & Tree adjustment & 80,830 & (5.6\%) & 72,107 & (9.2\%) \\
+    & Recombination & 4,123 & (0.33\%) & 2,078 & (0.27\%) \\
+    \cmidrule{2-6}
+    % TODO add more information on mutations
+    Mutations & & 1.21M &  & 1.06M & \\
+    & On one sample & 759K & (63\%) & 767K & (72\%) \\
+    & Reversions & 72K & (5.98\%) & 48K & (4.54\%) \\
+    & per node & 0.83  &   & 1.36 \\
+    \cmidrule{2-6}
+    Recombination events & \\
+    & Total & 4,123 & & 2,078 & \\
+    & Singletons & 2,601  & (63\%) & 1,315 & (63\%) \\
+    % clearer for many than "local trees?"
+    & Unique breakpoints & 1,496 &  & 958 & \\
+    \cmidrule{2-6}
+    \multicolumn{2}{l}{Size (inc metadata)} & 57MB & & 36MB&  \\
+    \multicolumn{2}{l}{Bytes/sample (exc metadata)}  & 8.29  & & 10.83 \\
+    Load time & & 0.9s & & 0.5s & \\
+    \bottomrule
 \end{tabular}
+\end{center}
+
 \caption{\label{tab:args}Summary of the inferred ARGs.
@@
 \end{table}
+
+% \centering
+% \begin{tabular}{l|l|l} \hline \textbf{Property} &
+% \textbf{Wide ARG} & \textbf{Long ARG} \\ \hline Last sampling date & 2021-06-30
+% & 2022-06-30 \\ Local trees & 1,496 & 958 \\ \hline
+% \multicolumn{3}{l}{\textit{Nodes}} \\ All nodes (sample and non-sample) &
+% 1,453,347 & 783,231 \\ Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\ Tree
+% adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107
+% (10.97\%) \\ Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\ \hline
+% \multicolumn{3}{l}{\textit{Types of recombination nodes}} \\ Singletons  & 2,601
+% (63\%) & 1,315 (63\%) \\ Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%)
+% \\ HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\ \hline
+% \multicolumn{3}{l}{\textit{Mutations}} \\ Total mutations & 1,213,193 &
+% 1,062,072 \\ Mutations on a terminal edge & 758,903 (62.55\%) & 767,111
+% (72.23\%) \\ Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
+% Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\ Immediate reversions (one
+% intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\ \hline
+% \multicolumn{3}{l}{\textit{Storage and loading time}} \\ tszipped file
+%  & ~57 Mb & ~36 Mb \\ Bytes to encode one
+% sample & 8.29 & 10.83 \\ Load time (seconds) & 0.9 & 0.5 \\ \hline
+% \end{tabular}
 
@@
 
-The wide ARG (with over one million sample genomes) was built using 128 threads
-on a machine with 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). Inference ran for 17
-days to attach genomes sampled up to June 30, 2021. The long ARG (with ~700
-thousand sample genomes) was built using 40 threads on a machine with 256 GB
-RAM (2x Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz). Inference ran for 23 days
-to attach genomes sampled up to June 30, 2022. Most of the runtime was spent on
-running the HMM.
+Inferring the wide ARG required 17 days of computation time on a machine
+with 128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz).
+The long ARG required 23 days on a (much older) machine
+with 40 threads and 256 GB RAM (2x Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz).
+The majority of the time is spent on running the HMM to find the copying
+path for each sequence, and the process is therefore highly amenable to
+distributing across multiple machines.
 

### 2023-04-06 paper.tex
Add proposed table from Jackson et al, commented out
Add in some rough notes for pango lineage analyses

@@
 
+% jk: this is the proposal from issue #33
+
+% \begin{tabular}{ll|clll}
+% \toprule
+% A Group (XA) & Jackson        & 21,255–21,613 & B.1.177/B.1.1.7 \\
+%              & \texttt{sc2ts} &  21,256--22,227 & B.1.177.18/B.1.1.7 \\
+% \midrule
+% B Group & Jackson        &  6,528--6,954 & B.1.36/B.1.1.7  \\
+%         & \texttt{sc2ts} &  6,529--6,954 & B.1.36/B.1.1.7  \\
+% \midrule
+% C Group & Jackson        &  25,997--27,972 &  B.1.1.7/B.1.221 \\
+%         & \texttt{sc2ts} &  25,996--27,441 &  B.1.1.7/B.1.221 \\
+% \midrule
+% D Group & Jackson        &  21,575--23,063 &  B.1.36.17/B.1.1.7 \\
+%         & \texttt{sc2ts} &  22,445--23,063 &  B.1.36.39/B.1.1.7 \\
+% \midrule
+% CAMC-CBA018 & Jackson        &  20,389--21,254 & B.1.177/B.1.1.7 \\
+%             & \texttt{sc2ts} & 16,177--21,255 & B.1.177/B.1.1.7 \\
+% \midrule
+% \end{tabular}
+
 In the wide ARG (sampled to mid-2021), all these sequences are descendants of
@@
 in the Long ARG].
-long ARG (subsampled to mid-2022) contains samples
+long ARG (subsampled to mid-2022) contains 749 samples
 from 33 Pango-designated recombinant lineages (XA to XAK; Supplementary Table)
 [Probably could have a table here to summarise?]
+
+% \item Total of 2078 recombinants.
+% \item 749 samples from 33 Pango X lineage (recombinant) designations
+% \item \alert{All but} 2 XP samples inferred to have recombinant ancestry
+% (Almost certainly missing XP because of our alignment masking approach)
+
+All but 2 XP samples inferred to have recombinant ancestry.
 
@@
 
+[TODO need to be clear if we're looking at single monophyletic origins
+here, or whether we include complicated things like XAG etc or not.
+Probably not?]
+
+% \item 14 out of 33 of the Pango X lineage inferred to have a single
+% recombinant origin. (XA, XAA, XAC, XAE, XAG, XAK, XF, XG, XH, XL,
+%  XR, XT, XW, XY)
+% \item Total of 148 samples (max 32, average of 10 per group)
+% \item XAA and XAG share the same origin, all others are distinct
+%     \begin{itemize}
+%         \item Ours: BA.1 \& BA.2.9 (6472 -- 8393)
+%         \item XAA: BA.1* \& BA.2* (8393 -- 9343)
+%         \item XAG: BA.2* \& BA.1* (6516 -- 8392)
+%         \item At least one XAB lineage (50 samples, with 10 origins)
+%         also descends from this recombinant---unclear what's happening
+%     \end{itemize}
+% \end{itemize}
+% \end{block}
+% \end{frame}
+
 \subsubsection{Multiple origin}
@@
 descending from these nodes .
+
+% \begin{frame}
+% \begin{block}{Multiple origin Pango X lineages}
+% \begin{itemize}
+% \item For the remaining 18, the story is more complicated.
+% \item XJ, XU and XV each have 3 samples and 3 recombinants
+% \item One recombinant shared among 5 lineages (XAA, XAB, XAG, XQ, XU)
+% \item XE has 170 samples and 12 recombinants; some are very similar
+% (Official: BA.1* \& BA.2, 10448--11287):
+% \\
+% \vspace{1em}
+% \begin{tabular}{llll}
+% \toprule
+% Left parent & Right parent & \multicolumn{2}{l}{Breakpoint interval}\\
+% \midrule
+% BA.1.15 & BA.2  &9867  &10198\\
+% BA.1.15 & BA.2  &9867  &10198\\
+% BA.1.1  &BA.2  &10451  &11537\\
+% BA.1.1.5 & BA.2 &10448 & 11537\\
+% BA.1.15 & BA.2  &10448  &12880\\
+% BA.1.17 & BA.2  &10448 & 11537\\
+% BA.1  &BA.2.9  &10448 & 11537\\
+% BA.1  &BA.2  &12764  &13195\\
+% \bottomrule
+% \end{tabular}
+% % BA.3  &BA.1.17  &22674  &22674\\
+% % BA.1.1.15 Recombinant 10448 11537
+% % BA.1.1 Recombinant 10030 10447
+% % Recombinant BA.2 17411 19955
+% \end{itemize}
+% \end{block}
+% \end{frame}
+
+\begin{figure}
+\begin{tabular}{cc}
+
+Example lineage subgraph
+ &
+
+\begin{tabular}{llll}
+\toprule
+Left parent & Right parent & \multicolumn{2}{l}{Breakpoint interval}\\
+\midrule
+BA.1.15 & BA.2  &9867  &10198\\
+BA.1.15 & BA.2  &9867  &10198\\
+BA.1.1  &BA.2  &10451  &11537\\
+BA.1.1.5 & BA.2 &10448 & 11537\\
+BA.1.15 & BA.2  &10448  &12880\\
+BA.1.17 & BA.2  &10448 & 11537\\
+BA.1  &BA.2.9  &10448 & 11537\\
+BA.1  &BA.2  &12764  &13195\\
+\bottomrule
+\end{tabular}
+% BA.3  &BA.1.17  &22674  &22674\\
+% BA.1.1.15 Recombinant 10448 11537
+% BA.1.1 Recombinant 10030 10447
+% Recombinant BA.2 17411 19955
+\end{tabular}
+\caption{A. Example subgraph for a multi origin example. B. details
+of inferred breakpoints for different recombination events inferred
+for XE samples (170 samples and 12 inferred recombination nodes; some are very
+similar (Official: BA.1* \& BA.2, 10448–11287)}
+\end{figure}
 

### 2023-04-10 plot.py
Add graph code and pdfs

@@ -4,16 +4,20 @@ from datetime import datetime, timedelta
 import hashlib
 import re
 import sys
+import json
 import os
 import tempfile
 import subprocess
 import logging
+import gzip
+import shutil
 
 import matplotlib as mpl
 
 mpl.use("Agg")
 from matplotlib.colors import rgb2hex
 from matplotlib.cm import get_cmap
+from matplotlib.lines import Line2D
 from matplotlib import pyplot as plt
 import numpy as np
 import tqdm
@@ -148,7 +152,7 @@ class Figure:
 
 class Cophylogeny(Figure):
     name = None
-    pos = 22000  # Position along tree seq to plot trees
+    pos = 0  # Position along tree seq to plot trees
     sc2ts_filename = None
     nextstrain_ts_fn = "nextstrain_ncov_gisaid_global_all-time_timetree-2023-01-21.nex"
 
@@ -193,7 +197,17 @@ class Cophylogeny(Figure):
         stored in self.sc2ts and self.nxstr
         """
         sc2ts_arg, basetime = utils.load_tsz(self.ts_dir, self.sc2ts_filename)
-        nextstrain = Nextstrain(self.nextstrain_ts_fn, span=sc2ts_arg.sequence_length)
+        fn = os.path.join(self.ts_dir, self.nextstrain_ts_fn)
+        try:
+            nextstrain = Nextstrain(fn, span=sc2ts_arg.sequence_length)
+        except FileNotFoundError:
+            logging.info("Attempting to extract gz compressed file")
+            with tempfile.TemporaryDirectory() as tmpdir:
+                with gzip.open(fn + ".gz") as f_in:
+                    fn = os.path.join(tmpdir, self.nextstrain_ts_fn)
+                    with open(fn, "wb") as f_out:
+                        shutil.copyfileobj(f_in, f_out)
+                    nextstrain = Nextstrain(fn, span=sc2ts_arg.sequence_length)
 
         # Slow step: find the samples in sc2ts_arg.ts also in nextstrain.ts, and subset
         sc2ts_its, nxstr_its = sc2ts.subset_to_intersection(
@@ -816,150 +830,439 @@ class RecombinationNodeMrcas_filtered_subset(RecombinationNodeMrcas_subset):
         self.df = self.filter(self.df)
 
 
-class XAG_graph(Figure):
+def change_pos(pos, key, dx=None, dy=None, x=None, y=None, xy_from=None):
+    """
[…]

### 2023-04-11 paper.tex
Rough outline for single origin Pango

@@
 recombination event.
+
+\begin{figure}
+\begin{tabular}{cc}
+\includegraphics[width=0.4\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
+&
+\begin{tabular}{llr}
+Lineage & number of samples & other details... \\
+XA & & \\
+XF & & \\
+XG & & \\
+XH & & \\
+XL & & \\
+XR & & \\
+XT & & \\
+XW & & \\
+XY & & \\
+XAC & & \\
+XAE & & \\
+XAK & & \\
+\end{tabular}
+\end{tabular}
+\caption{A. Subgraph for XA (wide ARG). B. Details for the the single-origin Pango
+X lineages in the long ARG.}
+\end{figure}
 

### 2023-04-11 paper.tex
Add shared origin example to fig.

--- 2023-04-06 paper.tex
@@
 \usepackage{booktabs}
+\usepackage{tabularx}
 \usepackage{authblk}
@@
 genome for the wide and long ARGs, respectively).
+
+\begin{figure} \centering
+\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
+\caption{\label{fig:overview_sc2ts} A schematic of the sc2ts method. Sc2ts
+reconstructs the genetic relationships among SARS-CoV-2 genomes by copying
+samples to all possible ancestors collected at earlier time points (curved
+arrows). An ARG is built by attaching new sample genomes grouped by collection
+date to an existing ARG rooted at the Wuhan-Hu-1 reference genome, day by day.
+Each iteration involves three stages: (1) attachment of new samples to the
+growing ARG (A, B); (2) reconstruction of trees relating the samples under each
+attachment node (C); and parsimony-based tree topology adjustments (D, E). In
+the absence of  recombination, sc2ts infers an ARG that is a single tree
+relating the samples (A). However, in the presence of recombination, sc2ts
+infers an ARG that is encoded as a sequence of local trees relating segments of
+the sample genomes (B), inserting a recombination node to represent a putative
+recombination event. Additionally, mutation-collapsing nodes (D) and
+pushed-reversion nodes (E) are inserted to make more parsimonious placements of
+mutations that should be shared or should not be immediately reverted,
+respectively.}
+\end{figure}
 
@@
 
+\begin{figure}
+\begin{tabularx}{\textwidth}{cc}
+\includegraphics[width=0.4\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
+&
+\begin{tabular}[b]{llr}
+Lineage & number of samples & other details... \\
+XA & & \\
+XF & & \\
+XG & & \\
+XH & & \\
+XL & & \\
+XR & & \\
+XT & & \\
+XW & & \\
+XY & & \\
+XAC & & \\
+XAE & & \\
+XAK & & \\
+XAA & & \\
+XAB & & \\
+XAG & & \\
+\end{tabular}\\
+\multicolumn{2}{c}{
+\includegraphics[width=\textwidth]{figures/Pango_XAG_nxcld_tight_graph.pdf}}
+\end{tabularx}
+\caption{Single origin Pango X lineages. (A) Subgraph for XA (Wide ARG).
+(B) Details for the the single-origin Pango
+X lineages in the long ARG. (C) XAA, XAB and XAG share a single recombinant
+origin in the Long ARG. }
+\end{figure}
+
 The sequences of 14 Pango recombinants trace back to a single recombination
@@
 show the lineage subgraph for X?, which makes this simple structure
-clear. [Explain the lineage subgraph format a bit]
+clear. [Explan the lineage subgraph format a bit]
 
@@
 % \end{frame}
+
+[Discuss panel C where XAA, XAB,XAG and share a common recombinant origin]
 

### 2023-04-11 paper.tex
Add cophylogeny fig

@@
 \begin{figure} \centering
-\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
-\caption{\label{fig:overview_sc2ts} A schematic of the sc2ts method. Sc2ts
-reconstructs the genetic relationships among SARS-CoV-2 genomes by copying
-samples to all possible ancestors collected at earlier time points (curved
-arrows). An ARG is built by attaching new sample genomes grouped by collection
-date to an existing ARG rooted at the Wuhan-Hu-1 reference genome, day by day.
-Each iteration involves three stages: (1) attachment of new samples to the
-growing ARG (A, B); (2) reconstruction of trees relating the samples under each
-attachment node (C); and parsimony-based tree topology adjustments (D, E). In
-the absence of  recombination, sc2ts infers an ARG that is a single tree
-relating the samples (A). However, in the presence of recombination, sc2ts
-infers an ARG that is encoded as a sequence of local trees relating segments of
-the sample genomes (B), inserting a recombination node to represent a putative
-recombination event. Additionally, mutation-collapsing nodes (D) and
-pushed-reversion nodes (E) are inserted to make more parsimonious placements of
-mutations that should be shared or should not be immediately reverted,
-respectively.}
+\includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
+\caption{\label{fig:cophlogeny} DESCRIBE ME.
+}
 \end{figure}
 
-Figure 4 shows a co-phylogeny of the 88 samples shared between the wide ARG and
+Figure~\ref{fig:cophlogeny} shows a co-phylogeny of the 88 samples shared between the wide ARG and
 a GISAID global "all-time" tree from NextStrain (downloaded on January 21,

### 2023-04-11 paper.tex
Updates to figures and display items

@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
+\includegraphics[width=0.4\textwidth]{figures/mutational_spectra.pdf}
 \caption{\label{fig:mutational_spectra}
@@
 
-% FIXME there's too much vspace between the rows here. Probably best to redo
-% the formatting.
+% jk: this is the proposal from issue #33
+
 \begin{table} \centering
-\begin{tabular}{l|c|c|c|c|c} \hline
-\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
-\multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\ \hline
-\textbf{Sample} &
-\textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\
-interval(s)} &
-\textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\
-\hline ALDP-11CF93B & A
-    & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
-21,256–22,227 \\
-    ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
-    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
-ALDP-130BB95 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
-    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
-LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
-    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
-QEUH-CCCB30 & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
-    \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
-QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
-    \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
-MILK-1166F52 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
-    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
-MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
-    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
-QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
-    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
-MILK-126FE1F & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
-    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
-RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
-    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
-RAND-128FA33 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
-    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
-CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
-    \thead{B.1.177 \\ Alpha} & 16,177–21,255 \\
-CAMC-CB7AB3 & n/a & \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} &
-     \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\
-MILK-103C712 & n/a & \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} &
-    n/a & n/a \\
-QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 &
-    \thead{Alpha \\ B.1.177} & 7,729–10,870 \\ \hline
+\begin{tabular}{ll|clll}
+\toprule
+A Group (XA) & Jackson        & 21,255–21,613 & B.1.177/B.1.1.7 \\
+             & \texttt{sc2ts} &  21,256--22,227 & B.1.177.18/B.1.1.7 \\
+\midrule
+B Group & Jackson        &  6,528--6,954 & B.1.36/B.1.1.7  \\
+        & \texttt{sc2ts} &  6,529--6,954 & B.1.36/B.1.1.7  \\
+\midrule
+C Group & Jackson        &  25,997--27,972 &  B.1.1.7/B.1.221 \\
+        & \texttt{sc2ts} &  25,996--27,441 &  B.1.1.7/B.1.221 \\
+\midrule
+D Group & Jackson        &  21,575--23,063 &  B.1.36.17/B.1.1.7 \\
+        & \texttt{sc2ts} &  22,445--23,063 &  B.1.36.39/B.1.1.7 \\
+\midrule
+CAMC-CBA018 & Jackson        &  20,389--21,254 & B.1.177/B.1.1.7 \\
+            & \texttt{sc2ts} & 16,177--21,255 & B.1.177/B.1.1.7 \\
+\midrule
 \end{tabular}
-\caption{\label{tab:jackson}Recombinant sequences involving the Alpha (B.1.1.7)
-variant reported by \cite{Jackson2021-ik} have recombinant
-ancestry in the wide ARG. The breakpoint intervals and Pango lineage
-assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et
-al., except the Pango lineage assignment of the parents of group B
-recombinants, which were taken from Table 1 (motif-based results).}
+\caption{\label{tab:jackson}Comparison of recombination breakpoint intervals
+and parent lineages for a subset of the data
+reported by \cite{Jackson2021-ik}. [FIXME ADD MORE DETAIL]
+More detailed results are given in
+Table~\ref{tab:jackson_supplement}.}
 \end{table}
-
-% jk: this is the proposal from issue #33
-
-% \begin{tabular}{ll|clll}
-% \toprule
-% A Group (XA) & Jackson        & 21,255–21,613 & B.1.177/B.1.1.7 \\
-%              & \texttt{sc2ts} &  21,256--22,227 & B.1.177.18/B.1.1.7 \\
-% \midrule
-% B Group & Jackson        &  6,528--6,954 & B.1.36/B.1.1.7  \\
-%         & \texttt{sc2ts} &  6,529--6,954 & B.1.36/B.1.1.7  \\
-% \midrule
-% C Group & Jackson        &  25,997--27,972 &  B.1.1.7/B.1.221 \\
-%         & \texttt{sc2ts} &  25,996--27,441 &  B.1.1.7/B.1.221 \\
-% \midrule
-% D Group & Jackson        &  21,575--23,063 &  B.1.36.17/B.1.1.7 \\
-%         & \texttt{sc2ts} &  22,445--23,063 &  B.1.36.39/B.1.1.7 \\
-% \midrule
-% CAMC-CBA018 & Jackson        &  20,389--21,254 & B.1.177/B.1.1.7 \\
-%             & \texttt{sc2ts} & 16,177--21,255 & B.1.177/B.1.1.7 \\
-% \midrule
-% \end{tabular}
 
@@
 
+\section{Supplementary}
+
+% FIXME there's too much vspace between the rows here. Probably best to redo
+% the formatting.
+\begin{table} \centering
+\begin{tabular}{l|c|c|c|c|c} \hline
+\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
+\multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\ \hline
+\textbf{Sample} &
+\textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\
+interval(s)} &
+\textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\
+\hline ALDP-11CF93B & A
+    & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
+21,256–22,227 \\
+    ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
+ALDP-130BB95 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
+LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
+QEUH-CCCB30 & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
+    \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
+QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
+    \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
+MILK-1166F52 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
+    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
+MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
+    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
+QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
+    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
+MILK-126FE1F & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
+    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
+RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
+    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
+RAND-128FA33 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
+    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
+CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
+    \thead{B.1.177 \\ Alpha} & 16,177–21,255 \\
+CAMC-CB7AB3 & n/a & \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} &
+     \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\
+MILK-103C712 & n/a & \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} &
+    n/a & n/a \\
+QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 &
+    \thead{Alpha \\ B.1.177} & 7,729–10,870 \\ \hline
+\end{tabular}
+\caption{\label{tab:jackson_supplement}Recombinant sequences involving the Alpha (B.1.1.7)
+variant reported by \cite{Jackson2021-ik} have recombinant
+ancestry in the wide ARG. The breakpoint intervals and Pango lineage
+assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et
+al., except the Pango lineage assignment of the parents of group B
+recombinants, which were taken from Table 1 (motif-based results).}
+\end{table}
+
 \end{document}

### 2023-04-12 paper.tex

@@
 \includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
-\caption{\label{fig:cophlogeny} DESCRIBE ME.
+\caption{\label{fig:cophylogeny} DESCRIBE ME. See
+Figure~\ref{fig:cophlogeny_long} for the equivalent cophylogeny
+for the Long ARG.
 }
@@
 
-Figure~\ref{fig:cophlogeny} shows a co-phylogeny of the 88 samples shared between the wide ARG and
+Figure~\ref{fig:cophylogeny} shows a co-phylogeny of the 88 samples shared between the wide ARG and
 a GISAID global "all-time" tree from NextStrain (downloaded on January 21,
@@
 % filter.
+
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
+\caption{\label{fig:recomb_mrcas} Divergance between parents at
+recombination breakpoints. (A) All recombination nodes that are ancestral
+to least two samples. (B) Further filtered as described in the text.
+MORE DETAIL. Horizontal lines represent WHAT,
+Figure~\ref{fig:recomb_mrcas_lineage_breakdown} shows the same data
+broken down by parental lineage.}
+\end{figure}
 
@@
 
-\section{Supplementary}
+\renewcommand\thefigure{S\arabic{figure}}
+\setcounter{figure}{0}
+\renewcommand\thetable{S\arabic{table}}
+\setcounter{table}{0}
+\section*{Supplementary}
+
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/supp_cophylogeny_long.pdf}
+\caption{\label{fig:cophlogeny_long}Copyhlogeny for the Long ARG. MORE DETAIL.}
+\end{figure}
 
@@
 
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/supp_recombination_node_mrcas.pdf}
+\caption{\label{fig:recomb_mrcas_lineage_breakdown}  DESCRIBE ME}
+\end{figure}
+
 \end{document}

### 2023-04-13 paper.tex

@@
 \includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
-\caption{\label{fig:cophylogeny} DESCRIBE ME. See
-Figure~\ref{fig:cophlogeny_long} for the equivalent cophylogeny
-for the Long ARG.
-}
+\caption{\label{fig:cophylogeny} Tanglegram comparing the first local tree in the
+wide ARG (sampled to mid-2021) and an “all-time” global NextStrain tree (sampled
+to January 21, 2023). The trees were pruned down to the samples present in both
+the trees. Light grey lines match the corresponding samples between the two
+trees; black circles indicate identical sample partitions between the two trees.
+Terminal branches are colour-coded according to the Pango lineage status
+assigned to the tip samples; no Pango recombinants are shown, as none are shared
+between the trees. The tanglegram was generated with the aid of the Neighbor-Net
+algorithm \citep{Scornavacca2011-mg} implemented in Dendroscope
+version 3.8.5 \citep{Huson2012-ys}. See Figure~\ref{fig:cophlogeny_long} for the equivalent cophylogeny
+for the Long ARG.}
 \end{figure}
@@
 \includegraphics[width=\textwidth]{figures/supp_cophylogeny_long.pdf}
-\caption{\label{fig:cophlogeny_long}Copyhlogeny for the Long ARG. MORE DETAIL.}
+\caption{\label{fig:cophlogeny_long}Tanglegram equivalent to that in
+Figure~{fig:cophlogeny}, but for the long ARG (i.e. subsampled to mid-2022). }
 \end{figure}

### 2023-04-13 plot.py
Change colourscheme

@@ -853,10 +853,10 @@ class Pango_X_graph(Figure):
         # Pango X-lineages in red/brown/pink
         'XA': '#fe812e',
         'XB': '#fc31fb',
-        'XQ': '#e780a9',
-        'XU': '#c35aab',
-        'XAA': '#fdacfe',
-        'XAB': '#f2b6b9',
+        'XQ': '#db5884',
+        'XU': '#894a4e',
+        'XAA': '#ffc9a5',
+        'XAB': '#fdacfe',
         'XAG': '#fc3131',
 
         # B lineages in blue/green (AY.100 is B.1.617.2.100)

### 2023-04-13 plot.py
Don't distinguish 2-parent re nodes

@@ -672,15 +672,7 @@ class RecombinationNodeMrcas(Figure):
             if (self.basetime - datetime(y, m, 1)).days > -2
         ]
 
-        for num_parents_2, colour in zip([True, False], ["green", "blue"]):
-            df_ = df[df.num_parents == 2] if num_parents_2 else df[df.num_parents > 2]
-            main_ax.scatter(
-                df_.tmrca_delta / 7,
-                df_.tmrca,
-                alpha=0.1,
-                c=colour,
-                label="num_parents=2" if num_parents_2 else "num_parents>2",
-            )
+        main_ax.scatter(df.tmrca_delta / 7, df.tmrca, alpha=0.1)
         if xlab:
             hist_ax.set_xlabel("Estimated divergence between lineage pairs (weeks)")
         if ylab:
@@ -833,7 +825,7 @@ class RecombinationNodeMrcas_filtered_subset(RecombinationNodeMrcas_subset):
 
     def __init__(self, args):
         super().__init__(args)
-        self.df = self.filter(self.df)
+        self.df = self._filter(self.df)
 
 
 def change_pos(pos, key, dx=None, dy=None, x=None, y=None, xy_from=None):

### 2023-04-13 paper.tex
Some analysis of the pango recombinants

@@
 algorithm \citep{Scornavacca2011-mg} implemented in Dendroscope
-version 3.8.5 \citep{Huson2012-ys}. See Figure~\ref{fig:cophlogeny_long} for the equivalent cophylogeny
+version 3.8.5 \citep{Huson2012-ys}. See
+Figure~\ref{fig:cophylogeny_long} for the equivalent cophylogeny
 for the Long ARG.}
@@
 
+\begin{table}
+% \begin{footnotesize}
+\begin{tabular}{lllrl}
+\toprule
+{} &         parent\_lineages &          pango\_parents &   n &
+descendant\_lineages \\
+lineage &                         &                        &     &
+\\
+\midrule
+XA      &   [B.1.177.18, B.1.1.7] &     [B.1.1.7, B.1.177] &   5 & \{'XA': 5\} \\
+XAA     &          [BA.1, BA.2.9] &         [BA.1*, BA.2*] &   8 &  \{'XAB':
+39, 'XAG': 17, 'XAA': 8, 'XQ': 2, 'XU'... \\
+XAC     &     [BA.2.3, BA.1.17.2] &  [BA.2*, BA.1*, BA.2*] &  27 &
+\{'XAC': 27\} \\
+XAE     &            [BA.2, BA.1] &         [BA.2*, BA.1*] &   9 &
+\{'XAE': 9\} \\
+XAG     &          [BA.1, BA.2.9] &         [BA.1*, BA.2*] &  17 &  \{'XAB':
+39, 'XAG': 17, 'XAA': 8, 'XQ': 2, 'XU'... \\
+XF      &            [AY.4, BA.1] &    [B.1.617.2*, BA.1*] &   2 &
+\{'XF': 2\} \\
+XG      &         [BA.1.17, BA.2] &         [BA.1*, BA.2*] &  32 &
+\{'XG': 32, 'XAB': 1\} \\
+XH      &       [BA.1.20, BA.2.9] &         [BA.1*, BA.2*] &  11 &  \{'XAF':
+34, 'XH': 11, 'B.1.1.529': 3, 'XE': 3,... \\
+XL      &       [BA.1.17.2, BA.2] &         [BA.1*, BA.2*] &  10 &
+\{'XL': 10, 'XAB': 1, 'XU': 1, 'XQ': 1\} \\
+XR      &          [BA.1.1, BA.2] &       [BA.1.1*, BA.2*] &   8 &
+\{'XQ': 9, 'XR': 8, 'XAB': 2\} \\
+XT      &  [BA.2.23, Unknown (R)] &         [BA.2*, BA.1*] &   1 &  \{'BA.2':
+1, 'BA.1': 1, 'BA.1.9': 1, 'BA.2.23':... \\
+XW      &       [BA.1.1.15, BA.2] &         [BA.1*, BA.2*] &  11 &
+\{'XW': 11, 'XN': 2\} \\
+XY      &          [BA.1.1, BA.2] &         [BA.1*, BA.2*] &   6 &
+\{'XY': 6, 'XAF': 1\} \\
+\bottomrule
+\end{tabular}
+\caption{\label{tab:pango-single-origin}
+Table summarising the status of the Pango X-lineages in the
+Long ARG that are inferred to have a single recombinant origin
+based on the NextClade Pango lineage assignments. Imputed parent lineages
+are shown from the ARG and also to parent lineages as defined by
+the official Pango designation alias key.
+FIXME: if we like this table then we can reformat it so that lists
+at the end are spread over multiple lines rather than using Python
+dict notation }
+\end{table}
+
 \begin{figure}
-\begin{tabularx}{\textwidth}{cc}
+\begin{tabularx}{\textwidth}{c}
+
+% \end{footnotesize}
+% \\
+
 \includegraphics[width=0.4\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
-&
-\begin{tabular}[b]{llr}
-Lineage & number of samples & other details... \\
-XA & & \\
-XF & & \\
-XG & & \\
-XH & & \\
-XL & & \\
-XR & & \\
-XT & & \\
-XW & & \\
-XY & & \\
-XAC & & \\
-XAE & & \\
-XAK & & \\
-XAA & & \\
-XAB & & \\
-XAG & & \\
-\end{tabular}\\
-\multicolumn{2}{c}{
-\includegraphics[width=\textwidth]{figures/Pango_XAG_nxcld_tight_graph.pdf}}
+\\ % &
+% \multicolumn{2}{c}{
+\includegraphics[width=\textwidth]{figures/Pango_XAG_nxcld_tight_graph.pdf}
+% }
 \end{tabularx}
-\caption{Single origin Pango X lineages. (A) Subgraph for XA (Wide ARG).
+\caption{\label{fig:pango-single-origin-graph}
+Single origin Pango X lineages. (A) Subgraph for XA (Wide ARG).
 (B) Details for the the single-origin Pango
@@
 
-The sequences of 14 Pango recombinants trace back to a single recombination
-node. The details are given in Figure X, where we [give the details of
+The sequences of 13 Pango recombinant lineages trace back to a single recombination
+node, as summarised in Table~\ref{tab:pango-single-origin}. We can
+see that XA, XAC, XAE, and XF form monophyletic clades, in which all of the
+samples given that Pango designation descendand from a single recombination
+node.
+
+The details are given in Figure X, where we [give the details of
 the numbers of samples, etc for each of these 14 lineages] and
@@
 clear. [Explan the lineage subgraph format a bit]
+
+We exclude XAK it only has a single sample in this dataset, and
+that sample traces to an older recombination node that
+that is likely to be spurious since it has a very large number of descendants.
 
@@
 \includegraphics[width=\textwidth]{figures/supp_cophylogeny_long.pdf}
-\caption{\label{fig:cophlogeny_long}Tanglegram equivalent to that in
-Figure~{fig:cophlogeny}, but for the long ARG (i.e. subsampled to mid-2022). }
+\caption{\label{fig:cophylogeny_long}Tanglegram equivalent to that in
+Figure~\ref{fig:cophylogeny}, but for the long ARG (i.e. subsampled to mid-2022). }
 \end{figure}

### 2023-04-13 plot.py
Add captions for the MRCA plots
And change captions. HMM consistent / inconsistent are plotted in different colours in the supp plot.

@@ -618,22 +618,17 @@ class RecombinationNodeMrcas(Figure):
         df["tmrca_delta"] = df.tmrca - self.ts.nodes_time[df.node.values]
         logging.info(f"{len(df)} breakpoints | {len(np.unique(df.node))} re nodes read")
         # Remove potential contaminents
-        df = df[df.max_descendant_samples > 1]
-        # For all plots, omit the breakpoints which are not in the tree seq but listed
-        # in the HMM metadata, see https://github.com/jeromekelleher/sc2ts/issues/121
-        self.df = df[df.is_arg_hmm_path_length_consistent == True]
+        self.df = df[df.max_descendant_samples > 1]
         logging.info(
-            f"{len(self.df)} breakpoints | {len(np.unique(self.df.node))} "
-            "re nodes initially retained"
+            f"{len(self.df)} breakpoints in the ARG | " +
+            f"{len(np.unique(self.df.node))} re nodes initially retained"
         )
 
     @staticmethod
-    def _filter(df):
-        return df[
-            np.logical_and(
-                df.fwd_bck_parents_max_mut_dist == 0, df.is_hmm_mutation_consistent
-            )
-        ]
+    def is_strict_HMM_consistent(df):
+        return np.logical_and(
+            df.fwd_bck_parents_max_mut_dist == 0, df.is_hmm_mutation_consistent
+        )
 
     @staticmethod
     def add_common_lines(ax, num, ts, common_proportions):
@@ -659,20 +654,59 @@ class RecombinationNodeMrcas(Figure):
                 bbox=dict(facecolor="white", edgecolor="none", pad=0),
             )
 
-    def do_plot(self, main_ax, hist_ax, df, title, label_tweak, xlab=True, ylab=True):
+    def do_plot(
+            self,
+            main_ax,
+            hist_ax,
+            df,
+            title,
+            label_tweak,
+            xlab=True,
+            ylab=True,
+            HMM_consistent_col="green",  # or "" to not plot
+            HMM_inconsistent_col="blue",
+    ):
         logging.info(f"Plotting {title}")
-        logging.info(f" {len(df)} points, {len(np.unique(df.node))} rec nodes")
-        logging.info(
-            f" time diff: min={df.tmrca_delta.min()}, max={df.tmrca_delta.max()}"
-        )
         dates = [
             datetime(y, m, 1)
             for y in (2020, 2021, 2022)
             for m in range(1, 13, 3)
[…]

### 2023-04-14 paper.tex
Update analysis on exclusion criteria

@@
 \subsection{Recombinant Pango lineages}
-
+[Intro bit saying what we're doing]
+
+[Quick overview of the wide ARG.]
+The wide ARG contains samples from three Pango-designated recombinant
+lineages (XA, XB, and XC) and N samples, since it only includes data
+up to mid-2021.
+In the wide ARG, every sequence of the three Pango recombinants trace back to
+some recombination node
+
+[Details of the long ARG]
 In this section we examine the Long ARG to determine how well it reflects
@@
 from 33 Pango-designated recombinant lineages (XA to XAK; Supplementary Table)
-[Probably could have a table here to summarise?]
-
-% \item Total of 2078 recombinants.
-% \item 749 samples from 33 Pango X lineage (recombinant) designations
-% \item \alert{All but} 2 XP samples inferred to have recombinant ancestry
-% (Almost certainly missing XP because of our alignment masking approach)
-
-All but 2 XP samples inferred to have recombinant ancestry.
-
-The sequences of 28 Pango recombinants trace back
-to some recombinant node(s); however, the sequences of the five Pango
-% jk: really? I thought it was just XP that wasn't inferred to be a
-% recombinant.
-recombinants XP, XU, XAA, XAG, and XAK do not.
-A likely explanation for the XP
-recombinant sequences not having a recombinant origin in the long ARG is that
-its characteristic multibase deletion
-(\url{https://github.com/cov-lineages/pango-designation/issues/481}) was masked
-during preprocessing and therefore was not incorporated into the HMM.
-% Eh?
-Upon tracing the origin of XAG through the ARG, XAG might have been caused by
-incorrect imputation of its parent nodes(?). XAK likely does not appear to have
-a recombinant origin because of random subsampling, but it may with a more
-complete tree.
-
-[FIXME this isn't quite right - not all the sequences for a given pango
-lineage trace back to the same recombination event, so won't have the same
-intervals.]
-The mosaic genome structure of all the 32 Pango recombinant lineages is concordant with
-the structure determined by the Pango Lineage Designation Committee
-(Supplementary Table 3). For example, the genome of XD (also known as
-``Deltacron'') is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
-(Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.
-
-The wide ARG contains samples from three Pango-designated recombinant
-lineages (XA, XB, and XC) and N samples, since it only includes data
-up to mid-2021.
-In the wide ARG, every sequence of the three Pango recombinants trace back to
-some recombination node. XA, XB, and XC.
-
-For the long ARG, we examine the lineages in increasing complexity of the
-inferred evolutionary relationships between samples assigned to Pango
-X lineages.
+
+Of these 749 samples, all but 2 are inferred to descend from a recombination
+node. These two samples are designated XP, and a likely explanation
+for their not have an inferred recombinant origin in the long ARG is that
+the characteristic multibase deletion for XP
+(\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
+during preprocessing. An improved masking pipeline is an important avenue
+for future development (see the Discussion).
+
+38 of the remaining X designated samples descend from recombination
+nodes that are ancestral only to that sample
+(10 are XZ; 6 are XE;  3 each from
+XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1 sample from each
+of XB, XM, XJ, XAF, XAH and XAJ).
+Such samples are likely to be enriched for sequencing errors and
+lineage designation artefacts, and so we exclude them from further
+analysis in this section.
+
+The most recent recombination node for 79 samples
+(XN: 53, XZ: 16, XAJ: 6, XE: 1, XAD: 1, XAH: 1, XAK: 1)
+is the same. This node is ancestral to more than 127K samples,
+and is likely to be a spuriously inferred recombination event.
+Understanding the origin of this and other similar nodes, is
+an important direction for future work (see Discussion).
+For simplicity, we exclude these samples from further
+analyses in this section.
+
+We analyse the inferred evolutionary patterns of the
+remaining 630 Pango X designated sequences in the following
+subsections.
+
+% The sequences of 28 Pango recombinants trace back
+% to some recombinant node(s); however, the sequences of the five Pango
+% % jk: really? I thought it was just XP that wasn't inferred to be a
+% % recombinant.
+% recombinants XP, XU, XAA, XAG, and XAK do not.
+% A likely explanation for the XP
+% recombinant sequences not having a recombinant origin in the long ARG is that
+% its characteristic multibase deletion
+% (\url{https://github.com/cov-lineages/pango-designation/issues/481}) was masked
+% during preprocessing and therefore was not incorporated into the HMM.
+% % Eh?
+% Upon tracing the origin of XAG through the ARG, XAG might have been caused by
+% incorrect imputation of its parent nodes(?). XAK likely does not appear to have
+% a recombinant origin because of random subsampling, but it may with a more
+% complete tree.
+
+% [FIXME this isn't quite right - not all the sequences for a given pango
+% lineage trace back to the same recombination event, so won't have the same
+% intervals.]
+% The mosaic genome structure of all the 32 Pango recombinant lineages is concordant with
+% the structure determined by the Pango Lineage Designation Committee
+% (Supplementary Table 3). For example, the genome of XD (also known as
+% ``Deltacron'') is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
+% (Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.
+
+% For the long ARG, we examine the lineages in increasing complexity of the
+% inferred evolutionary relationships between samples assigned to Pango
+% X lineages.
 

### 2023-04-14 paper.tex

@@
 
-We analyse the inferred evolutionary patterns of the
-remaining 630 Pango X designated sequences in the following
-subsections.
+The remaining remaining 630 Pango X designated sequences trace
+back to 50 different recombination nodes. We examine the
+inferred inheritance structure in the following subsections.
 
@@
 \midrule
-XA      &   [B.1.177.18, B.1.1.7] &     [B.1.1.7, B.1.177] &   5 & \{'XA': 5\} \\
-XAA     &          [BA.1, BA.2.9] &         [BA.1*, BA.2*] &   8 &  \{'XAB':
-39, 'XAG': 17, 'XAA': 8, 'XQ': 2, 'XU'... \\
+XA      &        [B.1.177.18, B.1.1.7] &     [B.1.1.7, B.1.177] &   5 &
+\{'XA': 5\} \\
+XAA     &               [BA.1, BA.2.9] &         [BA.1*, BA.2*] &   8 &
+\{'XAB': 39, 'XAG': 17, 'XAA': 8, 'XQ': 2, 'XU'... \\
 XAC     &          [BA.2.3, BA.1.17.2] &  [BA.2*, BA.1*, BA.2*] &  27 &
@@
 \{'XAE': 9\} \\
-XAG     &          [BA.1, BA.2.9] &         [BA.1*, BA.2*] &  17 &  \{'XAB':
-39, 'XAG': 17, 'XAA': 8, 'XQ': 2, 'XU'... \\
+XAG     &               [BA.1, BA.2.9] &         [BA.1*, BA.2*] &  17 &
+\{'XAB': 39, 'XAG': 17, 'XAA': 8, 'XQ': 2, 'XU'... \\
 XF      &                 [AY.4, BA.1] &    [B.1.617.2*, BA.1*] &   2 &
@@
 \{'XG': 32, 'XAB': 1\} \\
-XH      &       [BA.1.20, BA.2.9] &         [BA.1*, BA.2*] &  11 &  \{'XAF':
-34, 'XH': 11, 'B.1.1.529': 3, 'XE': 3,... \\
+XH      &            [BA.1.20, BA.2.9] &         [BA.1*, BA.2*] &  11 &
+\{'XAF': 34, 'XH': 11, 'B.1.1.529': 3, 'XE': 3,... \\
+XK      &             [BA.1.1.1, BA.2] &         [BA.1*, BA.2*] &   6 &
+\{'XK': 3\} \\
 XL      &            [BA.1.17.2, BA.2] &         [BA.1*, BA.2*] &  10 &
@@
 \{'XQ': 9, 'XR': 8, 'XAB': 2\} \\
-XT      &  [BA.2.23, Unknown (R)] &         [BA.2*, BA.1*] &   1 &  \{'BA.2':
-1, 'BA.1': 1, 'BA.1.9': 1, 'BA.2.23':... \\
+XS      &              [AY.36, BA.1.1] &  [B.1.617.2*, BA.1.1*] &   6 &
+\{'XS': 4\} \\
+XT      &       [BA.2.23, Unknown (R)] &         [BA.2*, BA.1*] &   1 &
+\{'BA.2': 1, 'BA.1': 1, 'BA.1.9': 1, 'BA.2.23':... \\
+XV      &  [BA.1.1, BA.2, Unknown (R)] &         [BA.1*, BA.2*] &   3 &
+\{'BA.2': 8, 'XV': 1\} \\
 XW      &            [BA.1.1.15, BA.2] &         [BA.1*, BA.2*] &  11 &
@@
 \end{tabular}
+
 \caption{\label{tab:pango-single-origin}
@@
 
-The sequences of 13 Pango recombinant lineages trace back to a single recombination
-node, as summarised in Table~\ref{tab:pango-single-origin}. We can
+The sequences of 16 Pango recombinant lineages trace back to a single recombination
+node, as summarised in Table~\ref{tab:pango-single-origin}.
+[FIXME UPDATE THESE]
+We can
 see that XA, XAC, XAE, and XF form monophyletic clades, in which all of the
@@
 
-The details are given in Figure X, where we [give the details of
-the numbers of samples, etc for each of these 14 lineages] and
-show the lineage subgraph for X?, which makes this simple structure
-clear. [Explan the lineage subgraph format a bit]
-
-We exclude XAK it only has a single sample in this dataset, and
-that sample traces to an older recombination node that
-that is likely to be spurious since it has a very large number of descendants.
-
-[TODO need to be clear if we're looking at single monophyletic origins
-here, or whether we include complicated things like XAG etc or not.
-Probably not?]
-
-% \item 14 out of 33 of the Pango X lineage inferred to have a single
-% recombinant origin. (XA, XAA, XAC, XAE, XAG, XAK, XF, XG, XH, XL,
-%  XR, XT, XW, XY)
-% \item Total of 148 samples (max 32, average of 10 per group)
-% \item XAA and XAG share the same origin, all others are distinct
-%     \begin{itemize}
-%         \item Ours: BA.1 \& BA.2.9 (6472 -- 8393)
-%         \item XAA: BA.1* \& BA.2* (8393 -- 9343)
-%         \item XAG: BA.2* \& BA.1* (6516 -- 8392)
-%         \item At least one XAB lineage (50 samples, with 10 origins)
-%         also descends from this recombinant---unclear what's happening
-%     \end{itemize}
-% \end{itemize}
-% \end{block}
-% \end{frame}
+[Discuss remaining]
+
+There is generally good agreement on the imputed parental lineages from
+the ARG and the official Pango designations. [Discuss more]
 

### 2023-04-18 plot.py
Add XD plot
And remove singleton recombinants

@@ -848,15 +848,6 @@ class RecombinationNodeMrcas_subset(RecombinationNodeMrcas):
         plt.savefig(prefix + f".{args.outtype}", bbox_inches="tight")
 
 
-class RecombinationNodeMrcas_filtered_subset(RecombinationNodeMrcas_subset):
-    name = "supp_recombination_node_mrcas_filtered"
-    # Only used to get information out really.
-
-    def __init__(self, args):
-        super().__init__(args)
-        self.df = self._filter(self.df)
-
-
 def change_pos(pos, key, dx=None, dy=None, x=None, y=None, xy_from=None):
     """
     Helper function to change the position of a node in a graph layout. `pos`
@@ -872,27 +863,36 @@ class Pango_X_graph(Figure):
     sample_metadata_labels = ""  # Don't show the strain name
     node_colours = {
         # Pango X-lineages in red/brown/pink
-        "XA": "#fe812e",
-        "XB": "#fc31fb",
-        "XQ": "#db5884",
-        "XU": "#894a4e",
-        "XAA": "#ffc9a5",
-        "XAB": "#fdacfe",
-        "XAG": "#fc3131",
+        'XA': '#bb0000',
+        'XB': '#fe812e',
+        'XD': '#fc31fb',
+        'XQ': '#db5884',
+        'XU': '#894a4e',
+        'XAA': '#ffc9a5',
+        'XAB': '#fdacfe',
+        'XAG': '#ff0055',
+
         # B lineages in blue/green (AY.100 is B.1.617.2.100)
         # Basic in blue
-        "B.1": "#212fe1",
-        "B.1.177.18": "#93e8ff",
-        "B.1.631": "#add9e5",
-        "B.1.634": "#2092e1",
-        "B.1.627": "#7b96cc",
-        "B.1.1": "#aea4ff",
-        "B.1.1.7": "#8666e3",  # Alpha
-        "AY.100": "#c6c076",  # Delta
+        'B.1': '#212fe1',
+        'B.1.177.18': '#93e8ff',
+        'B.1.631': '#add9e5',
+        'B.1.634': '#2092e1',
+        'B.1.627': '#7b96cc',
+        'B.1.1': '#aea4ff',
+
+        'B.1.1.7': '#8666e3', # Alpha
+
+        'B.1.617.2': '#e2da91',  # Delta
+        'AY.4': '#b78c19',  # Delta
+
[…]

### 2023-04-19 plot.py
Partially account for removed singleton recombinants

@@ -1106,7 +1106,7 @@ class Pango_XD_nxcld_tight_graph(Pango_X_tight_graph):
     name = "Pango_XD_nxcld_tight_graph"
     imputed_lineage = "Nextclade_pango"
 
-    figsize=(4, 4)
+    figsize=(4, 3.5)
 
     @staticmethod
     def define_nodes(ts, imputed_lineage):
@@ -1147,7 +1147,7 @@ class Pango_XB_nxcld_tight_graph(Pango_X_tight_graph):
             334261,
             337869,
             345330,
-            394058,  # most recent nodes, product of the most recent recombination
+            # 394058, # most recent nodes, product of the most recent recombination 
         ]
         for n in basic_nodes:
             pango = ts.node(n).metadata.get("Imputed_" + imputed_lineage)
@@ -1173,12 +1173,12 @@ class Pango_XB_nxcld_tight_graph(Pango_X_tight_graph):
 
         # All adjustments below found by tedious trial and error
 
-        change_pos(pos, 351074, xy_from=73385)
-        change_pos(pos, 73385, xy_from=322395, dx=5 * dx)
-        pos[322395], pos[345330] = pos[345330], pos[322395]
-        change_pos(pos, 345330, dx=dx)
-        change_pos(pos, 394059, x=pos[345330][0] / 2 + pos[73385][0] / 2)
-        change_pos(pos, 394058, xy_from=394059, y=pos[394058][1])
+        #change_pos(pos, 351074, xy_from=73385)
+        #change_pos(pos, 73385, xy_from=322395, dx=5 * dx)
+        #pos[322395], pos[345330] = pos[345330], pos[322395]
+        #change_pos(pos, 345330, dx=dx)
+        #change_pos(pos, 394059, x=pos[345330][0] / 2 + pos[73385][0] / 2)
+        #change_pos(pos, 394058, xy_from=394059, y=pos[394058][1])
         pos[329950], pos[280287] = pos[280287], pos[329950]
         pos[359092], pos[339516] = pos[339516], pos[359092]
         change_pos(pos, 359092, dx=dx)
@@ -1189,10 +1189,10 @@ class Pango_XB_nxcld_tight_graph(Pango_X_tight_graph):
         for p in [380107, 335144, 338964]:
             change_pos(pos, p, dx=-dx, dy=-dy)
 
-        change_pos(pos, 423732, dx=-5 * dx)
[…]

### 2023-04-21 paper.tex
Syncing

@@
 \section{Introduction}
-
-Recombination via template switching is a common feature in the evolution of
-coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2
-\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
-together mutations carried by different lineages, recombination plays an
-important role in generating genomic diversity, with recombinant lineages
-associated with adaptation to new host species and with the production of
-more immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
-
-Recombination was challenging to detect in SARS-CoV-2 early in the pandemic due
-to gradual accumulation of genetic variation. In late 2020, the appearance and
-spread of variants of concern (VoC) harbouring multiple characteristic
-mutations increased the detectability of recombination. By searching for
-genomes containing segments from different VoC, multiple recombinant lineages
-have been detected, for example, between Alpha and pre-VoC
-\citep{Jackson2021-ik}, Alpha and Epsilon \citep{Wertheim2022-hj}, and Alpha and
-Delta~\citep{Sekizuka2022-xz}. More recently, the high prevalence of Omicron,
-with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
-accelerated the rate of coinfection and the potential for recombination
-\citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of onward
-transmission (more than five descendants) and monophyly
-($>70\%$ bootstrap support) are eligible for a Pango lineage
-designation~\citep{Rambaut2020-dw},
-starting with the letter X
+Recombination via template switching is a common feature in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2 \citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}.
+By bringing together mutations carried by different lineages, recombination plays an important role in generating genomic diversity, with recombinant lineages associated with adaptation to new host species and with the production of more immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
+
+Recombination was challenging to detect early in the SARS-CoV-2 pandemic due to gradual accumulation of genetic variation.
+In late 2020, the appearance and spread of variants of concern (VoC), designated into classes such as Alpha and Delta which harboured multiple characteristic mutations, increased the detectability of recombination.
+By searching for genomes containing segments from different VoC designations, multiple recombinant lineages have been detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha and Epsilon \citep{Wertheim2022-hj}, and Alpha and Delta~\citep{Sekizuka2022-xz}.
+More recently, the high prevalence of Omicron, with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has accelerated the rate of coinfection and the potential for recombination \citep{Bal2022-hq}.
+Recombinant lineages that have sufficient evidence of onward transmission (more than five descendants) and monophyly
+($>70\%$ bootstrap support) are eligible for a Pango lineage designation~\citep{Rambaut2020-dw}, starting with the letter X
 % FIXME: should this be a citation?
@@
 
-Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
-combine specific subsets of mutations from different lineages
-\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
-This approach only identifies those inter-lineage recombination events
-resulting in the patterns investigated. Others have inferred recombinant
-genealogies to disentangle recombination from recurrent mutation, without
-having to pre-specify characteristic mutations for each lineage
-\citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it}
-presented a phylogeny-based approach (``RIPPLES'') to identify putative
-recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying motifs.
-To find candidate recombinants, RIPPLES first builds a phylogeny
-without assuming recombination and then scans for long branches (i.e., carrying
-many mutations) and then determines if these candidates would be better
-explained by recombination. Specifically, RIPPLES exhaustively breaks each
-candidate sequence into segments and attempts to find more parsimonious
-placements for each segment on the phylogeny. If such placements are found, the
-sequence is identified as a putative recombinant. Although it enables rapid
-searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2
-phylogeny that accounts for only mutations, treats recombinants
-\textit{post hoc}, and is an incomplete representation of the reticulate evolutionary
-history of SARS-CoV-2. As noted by the authors, a \textit{post hoc} treatment
-of recombination is possible when recombinant lineages are rare and leave few
-descendants. However, this is becoming increasingly untenable with the proliferation
-of recombinant lineages; for example, more than half of the sequences sampled in
-February 2023 are from XBB and its descendants~\citep{Chen2022-pz}.
-
-Inferring evolutionary trees while accounting for both mutation and
-recombination is a long-standing challenge. This challenge is particularly
-acute for SARS-CoV-2, given the large amount of data that has been shared, with
-over 15 million sequences in the GISAID database \citep{Shu2017-hp}. It is well
-known that recombination distorts phylogenies \citep{Schierup2000-fg} and
-affects the results of downstream analysis, e.g., inference of
-selection~\citep{Anisimova2003-vr}.
-However, standard phylogenetic methods do not account
-for recombination \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}.
-Alternatively, genetic ancestries of recombining lineages can be represented by
-ancestral recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-if},
-which  capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.
-
-Approaches to infer ARGs now exist that can scale to tens of thousands of human
-genomes and beyond
-\citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}. Recently,
-\cite{Wohns2022-th} reconstructed a unified whole-genome genealogy
-of thousands of modern and ancient human genomes, based on over 90 million
-variable sites. To achieve this, \cite{Wohns2022-th} utilised
-several computational advances in ARG inference: (1) the "succinct tree
-sequence" data structure that can compactly store large ARGs
-\citep{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM)
-implementation of the Li and Stephens (LS) model (Li and Stephens 2003) to
-infer copying paths between genomes \citep{Kelleher2019-ba}; and (3) efficient,
-open-source software tools
-\citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.
+Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that combine specific subsets of mutations from different lineages \citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
+This approach only identifies those inter-lineage recombination events resulting in the patterns investigated.
+Others have inferred recombinant genealogies to disentangle recombination from recurrent mutation, without having to pre-specify characteristic mutations for each lineage \citep{Ignatieva2021-rg}.
+More recently, \cite{Turakhia2022-it} presented a phylogeny-based approach (``RIPPLES'') to identify putative recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying motifs.
+To find candidate recombinants, RIPPLES first builds a phylogeny without assuming recombination and then scans for long branches (i.e., carrying many mutations) and then determines if these candidates would be better explained by recombination.
+Specifically, RIPPLES exhaustively breaks each candidate sequence into segments and attempts to find more parsimonious placements for each segment on the phylogeny.
+If such placements are found, the sequence is identified as a putative recombinant.
+Although it enables rapid searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that accounts for only mutations, treats recombinants \textit{post hoc}, and is an incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
+As noted by the authors, a \textit{post hoc} treatment of recombination is possible when recombinant lineages are rare and leave few descendants.
+However, this is becoming increasingly untenable with the proliferation of recombinant lineages; for example, more than half of the sequences sampled in February 2023 are from XBB and its descendants~\citep{Chen2022-pz}.
+This also means that future evolution of SARS-CoV-2 is likely to involve multiple sequential recombination events on top of existing recombinant lineages, creating a highly reticulated genealogy.
+
+Inferring evolutionary trees while accounting for both mutation and recombination is a long-standing computational challenge.
+This challenge is particularly acute for SARS-CoV-2, given the large amount of data that has been shared, with over 15 million sequences in the GISAID database \citep{Shu2017-hp}.
+It is well known that recombination distorts phylogenies \citep{Schierup2000-fg} and affects the results of downstream analysis, e.g., inference of selection~\citep{Anisimova2003-vr}.
+However, standard phylogenetic methods do not account for recombination \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}.
+Alternatively, genetic ancestries of recombining lineages can be represented by ancestral recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-if}, which capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.
+
+Approaches to infer ARGs now exist that can scale to tens of thousands of human genomes and beyond  \citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}, and which deal with levels of sequential recombination far in excess of those seen in viral phylogenies.
+For example, using over 90 million variable sites \citet{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands of modern and ancient human genomes, accounting for millions of recombination events (over 5 million "local trees").
+To achieve this, \cite{Wohns2022-th} took advantage of several computational advances in ARG inference: (1) the "succinct tree sequence" data structure that can compactly store large ARGs \citep{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM) implementation of the Li and Stephens (LS) model \citep{Li2003-ib} to infer copying paths between genomes \citep{Kelleher2019-ba}; and (3) efficient, open-source software tools \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.
 
 Here, we present a preliminary version of a novel scalable methodology
-(``sc2ts'', optionally pronounced ``scoots'') that builds on the advances above and
-additionally takes advantage of known collection dates to infer ARGs from
-millions of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2
-genome sequences, we demonstrate that sc2ts can be used to simultaneously (1)
-detect putative SARS-CoV-2 recombinants; (2) infer the phylogenetic
-relationships of the major SARS-CoV-2 lineages; (3) reconstruct the mutational
-spectrum of SARS-CoV-2; and (4) provide a compact record of evolutionary
-information about SARS-CoV-2 that can be efficiently accessed using "tskit"
-(tree sequence toolkit) software library \citep{Kelleher2018-xc}. Sc2ts is a
-promising method to automate detection of SARS-CoV-2 recombinants and to
-jointly model recombinational and mutational processes shaping the genomic
-diversity of SARS-CoV-2 at large scale.
+(``sc2ts'', optionally pronounced ``scoots'') that builds on the advances above and additionally takes advantage of known collection dates to infer ARGs from millions of full-length SARS-CoV-2 genomes.
+Using publicly available SARS-CoV-2 genome sequences, we demonstrate that sc2ts can be used to simultaneously (1) detect putative SARS-CoV-2 recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages; (3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record of evolutionary information about SARS-CoV-2 that can be efficiently accessed using "tskit" (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
+Sc2ts is a promising method to automate detection of SARS-CoV-2 recombinants and to jointly model recombinational and mutational processes shaping the genomic diversity of SARS-CoV-2 at scale.
 
@@
 \subsection{Data preprocessing}
-
 \subsubsection{Sequence-level preprocessing}
-
-The findings of this study are based on sequences and metadata available on
-GISAID (\url{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
+The findings of this study are based on sequences and metadata available on GISAID (\url{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
 % FIXME
 doi.org/10.55876/gis8.230329cd (Supplementary Table 1).
-Sequences were removed
-if they (1) had ambiguous collection dates;
-(2) were collected before January 01, 2020 or after June 30, 2022; or
-(3) were isolated from a non-human host.
-Sequences were aligned to the
-Wuhan-Hu-1/2019 reference sequence (GenBank: MN908947.3)
-using Nextclade v2.3.0~\citep{Aksamentov2021-hj}
-(dataset tag 2022-07-26T12:00:00Z).
-To reduce the impact of poor-quality sequences on ARG inference,
-sequences were excluded if they had a ``bad'' quality control status
-in any of the four Nextclade columns (``qc.missingData.status'',
-``qc.mixedSites.status'', ``qc.frameShifts.status'' and
-``qc.stopCodons.status'').
-These excluded genomes are enriched for low-coverage, contaminated, or
-misassembled sequences.
+Sequences were removed if they (1) had ambiguous collection dates; (2) were collected before January 01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
+Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank: MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag 2022-07-26T12:00:00Z).
+To reduce the impact of poor-quality sequences on ARG inference, sequences were excluded if they had a ``bad'' quality control status in any of the four Nextclade columns (``qc.missingData.status'', ``qc.mixedSites.status'', ``qc.frameShifts.status'' and `qc.stopCodons.status'').
+These excluded genomes are enriched for low-coverage, contaminated, or misassembled sequences.
 
 \subsubsection{Site-level preprocessing}
-
-Ambiguous nucleotide letters (i.e., not A, C, G, T, or a gap) in the pairwise
-genome alignments were encoded as N. Problematic bases in the alignments, which
-had two or more ambiguous letters or gaps within a distance of seven bases,
-were masked as missing data following the protocol in
-\url{https://github.com/roblanf/sarscov2phylo/}.
-Masked bases were ignored when attaching a sample genome to the ARG,
-but they were subsequently imputed.
-Further, 481 problematic sites flagged as prone to sequencing errors or highly
-homoplasic were excluded entirely
+Ambiguous nucleotide letters (i.e., not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
+Problematic bases in the alignments, which had two or more ambiguous letters or gaps within a distance of seven bases, were masked as missing data following the protocol in \url{https://github.com/roblanf/sarscov2phylo/}.
+Masked bases were ignored when attaching a sample genome to the ARG, but they were subsequently imputed.
+Further, 481 problematic sites flagged as prone to sequencing errors or highly homoplasic were excluded entirely
 % Do we need this first URL?
-(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
-\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
+(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473}; \url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
 
 \subsubsection{Exclusion of ``time travellers''}
-
-Including samples with incorrectly reported collection dates (e.g., an Alpha
-sample purportedly collected in 2020 from the United States, before Alpha
-appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
-artefacts that distorted the topology of ARGs inferred early on in the study.
+Including samples with incorrectly reported collection dates (e.g., an Alpha sample purportedly collected in 2020 from the United States, before Alpha appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce artefacts that distorted the topology of ARGs inferred early on in the study.
 Hence, to exclude such potential ``time travellers'' we employed two filters.
-The first was to remove any sequence with a collection date that pre-dated the
-time to the most recent common ancestor (tMRCA) of its corresponding clade. The
-tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
-(downloaded on August 22, 2022) where we used the lower bound of the 95\%
-confidence interval of each clade as the minimum date cut-off. The second
-filter was a simple threshold on the time delay between the collection date and
-submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
-inferred from the MSD-filtered genome set had noticeably fewer branches with a
-high number of reversions. After some preliminary analysis, a MSD threshold
-value of 30 days was used when building the ARGs described herein. Future work
-could estimate collection dates by adding these potential time travellers back
-on the tree, allowing an automated assessment of collection date discrepancies.
+The first was to remove any sequence with a collection date that pre-dated the time to the most recent common ancestor (tMRCA) of its corresponding clade.
+The tMRCA for each clade was obtained from a Nextstrain GISAID reference tree (downloaded on August 22, 2022) where we used the lower bound of the 95\% confidence interval of each clade as the minimum date cut-off.
+The second filter was a simple threshold on the time delay between the collection date and submission date (maximum submission delay, or MSD).
+Indeed, we found that ARGs inferred from the MSD-filtered genome set had noticeably fewer branches with a high number of reversions.
+After some preliminary analysis, a MSD threshold value of 30 days was used when building the ARGs described herein.
+Future work could estimate collection dates by adding these potential time travellers back on the tree, allowing an automated assessment of collection date discrepancies.
 
 \subsection{Inference}
-
 \subsubsection{Overview of sc2ts}
-
 \begin{figure} \centering
 \includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
-\caption{\label{fig:overview_sc2ts} A schematic of the sc2ts method. Sc2ts
-reconstructs the genetic relationships among SARS-CoV-2 genomes by copying
-samples to all possible ancestors collected at earlier time points (curved
-arrows). An ARG is built by attaching new sample genomes grouped by collection
-date to an existing ARG rooted at the Wuhan-Hu-1 reference genome, day by day.
-Each iteration involves three stages: (1) attachment of new samples to the
-growing ARG (A, B); (2) reconstruction of trees relating the samples under each
-attachment node (C); and parsimony-based tree topology adjustments (D, E). In
-the absence of  recombination, sc2ts infers an ARG that is a single tree
-relating the samples (A). However, in the presence of recombination, sc2ts
-infers an ARG that is encoded as a sequence of local trees relating segments of
-the sample genomes (B), inserting a recombination node to represent a putative
-recombination event. Additionally, mutation-collapsing nodes (D) and
-pushed-reversion nodes (E) are inserted to make more parsimonious placements of
-mutations that should be shared or should not be immediately reverted,
-respectively.}
+\caption{\label{fig:overview_sc2ts} A schematic of the sc2ts method.
+Sc2ts reconstructs the genetic relationships among SARS-CoV-2 genomes by copying samples to all possible ancestors collected at earlier time points (curved arrows).
+An ARG is built by attaching new sample genomes grouped by collection date to an existing ARG rooted at the Wuhan-Hu-1 reference genome, day by day.
+Each iteration involves three stages: (1) attachment of new samples to the growing ARG (A, B); (2) reconstruction of trees relating the samples under each attachment node (C); and parsimony-based tree topology adjustments (D, E).
+In the absence of  recombination, sc2ts infers an ARG that is a single tree relating the samples (A).
+However, in the presence of recombination, sc2ts infers an ARG that is encoded as a sequence of local trees relating segments of the sample genomes (B), inserting a recombination node to represent a putative recombination event.
+Additionally, mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to make more parsimonious placements of mutations that should be shared or should not be immediately reverted, respectively.}
 \end{figure}
@@
 % Note going to rewrite this anyway, so not trying to make it flow.
-It uses (1) an
-efficient HMM implementation of the Li and Stephens (LS) model \citep{Li2003-ib}
-to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported
-collection dates of the SARS-CoV-2 samples to determine the order of genome
-copying; and (3) several parsimony-based heuristics to locally refine the
-topology. In brief, this combined approach uses the likelihood model of LS to
-attach sample genomes to existing nodes and to detect recombinants, but because
-many of the tip samples attached daily harbour few, if any, mutational changes,
-we use parsimony to resolve the relationships near the tip samples.
-
-Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by
-taking advantage of the reported collection dates to incrementally grow the
-ARG. The inference process occurs forward in time, attaching new genomes
-aggregated by their collection date (daily batches), from the earliest to the
-last collection date. Prior to attaching sample genomes, the ARG is initialised
-such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence
-of recombination, the ARG is a single tree with no reticulation (Figure 1A);
-however, in the presence of recombination, a reticulate ARG is generated, with
-variation in tree topology across the genome (Figure 1B). Unlike standard
-phylogenetic trees in which internal nodes represent inferred ancestors,
-internal nodes in the ARGs inferred using sc2ts represent ancestors which can
-be samples or inferred ancestors.
-
-Each daily batch of genomes is first attached to the growing ARG without any
-phylogenetic resolution. In cases where the newly added sequence is inferred to
-be a recombinant, a recombination node is inserted; the ``date'' of this node is
-arbitrarily set to the average of the collection dates of the genome and the
-most recent of its parent nodes. Otherwise, the genomes are attached to the
-most closely related ancestral node (an attachment node). This creates
-polytomies that are then locally resolved among the descendants of each
-attachment node using the Unweighted Pair Group Method with Arithmetic Mean
-(UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
-This hierarchical clustering algorithm produces a dendrogram relating a set of
-sequences, which is treated as an ultrametric binary tree here. Subsequently,
-mutations are placed onto the daily tree by maximum parsimony. An issue with
-the UPGMA algorithm is that it may create internal nodes supported by no
-informative site (i.e., having no mutation immediately ancestral to them). In
-an adjustment step, such unsupported internal nodes are removed to ``un-resolve''
-the daily tree; the remaining internal nodes, which are supported by at least
-one shared mutation, are retained. The ``dates'' of the non-sample internal nodes
-are set to the same date as the samples in the daily batch. The resulting daily
-tree is grafted onto the attachment node in the global ARG.
-
-During ARG inference, two types of non-sample nodes are inserted to improve
-parsimonious placement of mutations: mutation-coalescing nodes and
-pushed-reversion nodes. We call these operations tree adjustments. A
-mutation-collapsing node is inserted when a shared mutation between sibling
-nodes is found (Figure 1D). An attempt to create mutation-collapsing nodes is
-made after grafting a new daily tree onto the global ARG. If shared mutations
-exist between two or more sibling nodes, then a parsimonious explanation is to
-create a new node that is the parent of those siblings, on which these
-mutations occurred. A pushed-reversion node is inserted when an immediate
-reversion occurs (Figure 1E), i.e. a mutation that reverts another mutation on
-its parent node. Instead of retaining both the mutation and immediate
-reversion, only the mutation is kept after a pushed-reversion node is inserted,
-thereby reducing the number of mutations on the branches. Note that these
-non-sample nodes are used as reconstructed genomes during HMM copying.
-
-Furthermore, during inference, the ARG is automatically encoded, stored, and
-updated as a tree sequence using tskit \citep{Kelleher2018-xc}.
+It uses (1) an efficient HMM implementation of the Li and Stephens (LS) model \citep{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported collection dates of the SARS-CoV-2 samples to determine the order of genome copying; and (3) several parsimony-based heuristics to locally refine the topology.
+In brief, this combined approach uses the likelihood model of LS to attach sample genomes to existing nodes and to detect recombinants, but because many of the daily tip samples harbour few, if any, mutational changes, we use parsimony to resolve their close relationships.
+
+Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by taking advantage of the reported collection dates to incrementally grow the ARG.
+The inference process occurs forward in time, attaching new genomes aggregated by their collection date (daily batches), from the earliest to the last collection date.
+Prior to attaching sample genomes, the ARG is initialised such that the Wuhan-Hu-1/2019 reference genome is the root node.
+In the absence of recombination, the ARG is a single tree with no reticulation (Figure 1A); however, in the presence of recombination, a reticulate ARG is generated, with variation in tree topology across the genome (Figure 1B).
+Unlike standard phylogenetic trees in which internal nodes always represent inferred ancestors, internal nodes in the sc2ts ARGs represent sequences which can either be inferred ancestors or known samples.
+
+Each daily batch of genomes is first attached to the growing ARG without any phylogenetic resolution.
+In cases where the newly added sequence is inferred to be a recombinant, a recombination node is inserted; the ``date'' of this node is arbitrarily set to the average of the collection dates of the genome and the most recent of its parent nodes.
+Otherwise, the genomes are attached to the most closely related ancestral node (an attachment node).
+This creates polytomies that are then locally resolved among the descendants of each attachment node using the Unweighted Pair Group Method with Arithmetic Mean (UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
+This hierarchical clustering algorithm produces a dendrogram relating a set of sequences, which is treated as an ultrametric binary tree here.
+Subsequently, mutations are placed onto the daily tree by maximum parsimony.
+An issue with the UPGMA algorithm is that it may create internal nodes supported by no informative site (i.e., having no mutation immediately ancestral to them).
+In an adjustment step, such unsupported internal nodes are removed to ``un-resolve'' the daily tree; the remaining internal nodes, which are supported by at least one shared mutation, are retained.
+The ``dates'' of the non-sample internal nodes are set to the same date as the samples in the daily batch.
+The resulting daily tree is grafted onto the attachment node in the global ARG.
+
+During ARG inference, two types of non-sample nodes are inserted to improve parsimonious placement of mutations: mutation-coalescing nodes and pushed-reversion nodes.
+We call these operations tree adjustments.
+A mutation-collapsing node is inserted when a shared mutation between sibling nodes is found (Figure 1D).
+An attempt to create mutation-collapsing nodes is made after grafting a new daily tree onto the global ARG.
+If shared mutations exist between two or more sibling nodes, then a parsimonious explanation is to create a new node that is the parent of those siblings, on which these mutations occurred.
+A pushed-reversion node is inserted when an immediate reversion occurs (Figure 1E), i.e.
+a mutation that reverts another mutation on its parent node.
+Instead of retaining both the mutation and immediate reversion, only the mutation is kept after a pushed-reversion node is inserted, thereby reducing the number of mutations on the branches.
+Note that these non-sample nodes are used as reconstructed genomes during HMM copying.
+
+Furthermore, during inference, the ARG is automatically encoded, stored, and updated as a tree sequence using tskit \citep{Kelleher2018-xc}.
 
 \subsubsection{Brief description of the Li and Stephens model} 
-
-The LS model is a convenient probabilistic approximation to the coalescent with
-recombination \citep{Hudson1983-if}, itself a generative model of a genealogical
-process with recombination \citep{Li2003-ib}. Importantly, it can be implemented
-as an HMM in which a focal genome is a sequence of nucleotides (observed
-states) that are probabilistically emitted as an imperfect mosaic of a set of
-reference genomes (hidden states). As such, an efficient HMM machinery can be
-employed to make statements about parameters of interest and highlight
-potential sources of error in genotyping and sequencing data. The LS model is
-used in a wide variety of applications in genomics. For example, it is the
-engine that drives statistical genotype phasing and imputation
-\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it
-can be used to estimate mutation rates (e.g., REF), recombination rates
-\citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral sequence
-data \citep[e.g.][]{Palmer2019-wa}.
+The LS model is a convenient probabilistic approximation to the coalescent with recombination \citep{Hudson1983-if}, itself a generative model of a genealogical process with recombination \citep{Li2003-ib}.
+Importantly, it can be implemented as an HMM in which a focal genome is a sequence of nucleotides (observed states) that are probabilistically emitted as an imperfect mosaic of a set of reference genomes (hidden states).
+As such, an efficient HMM machinery can be employed to make statements about parameters of interest and highlight potential sources of error in genotyping and sequencing data.
+The LS model is used in a wide variety of applications in genomics.
+For example, it is the engine that drives statistical genotype phasing and imputation \citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it can be used to estimate mutation rates (e.g., REF), recombination rates \citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral sequence data \citep[e.g.][]{Palmer2019-wa}.
 
@@
 \caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS) model.
-The LS model in its implementation as an HMM is governed by the transition and
-emission matrices, $Q$ and $E$, respectively. Each may be a
-function of the reference panel members, but transitions are generally assumed
-to be independent of the hidden states themselves (shown in the pink panel).
-This dramatically increases performance as the state space drops to two states
-(i.e., switching or not switching). The recombination probability at site
-$\ell$ is $r\ell$, while n denotes the number of members in the
-reference panel. Emissions may also be a function of the nucleotide states, but
-in our case, we assume that all observed alleles (in both the reference panel
-and focal sequence, $a\ell$) at the site under consideration are equally
-likely and occur with mutation probability $\ell$ (shown in the blue
-panel). These two processes, transition and emission (which encode
-recombination and mutation, respectively), define the generative process of the
-HMM, by which an imperfect mosaic is emitted from a reference panel of samples
-and nodes from the ARG inserted at earlier collection dates (shown in the lower
-panel). Here, the unseen states in the reference panel are shown as coloured
-lines enclosed by the grey box, and the true path through the data which leads
-to the emitted focal sequence is displayed by the black arrow. Examples of
-transition and emission probabilities along this trajectory are shown by the
-red and blue arrows, respectively. Black crosses along the focal sequence show
-genotyping errors or mutations. Note that in practice, we do not know the
-pattern of colours along the focal sequence and must infer it using the Viterbi
-algorithm.} \end{figure}
-
-The generative process of the LS model is summarised in
-Figure~\ref{fig:ls_diagram}.
-Here, a
-transition matrix, $Q$, governs the process of switching (recombining) between
-members of the reference panel (the hidden states). An emission matrix,
-$E$, allows for differences between the focal sequence and the hidden
-state from which it is copied (due to mutation as well as genotyping or
-sequencing error). Then, we use the Viterbi algorithm \citep{Viterbi1967-ol} to
-find the most likely copying path through the sequences, given a set of
-reference sequences, $Q$, and $E$.
-
-Because the HMM is memoryless, the potential parental lineages are
-independently assessed at each recombinantion breakpoint. This can result in
-the inference of a recombinant lineage descending immediately from three (or
-even more) parental lineages. We find that often two of the three parents are
-identical or similar in sequence in the segments contributed to the recombinant
-lineage. A more parsimonious explanation would be to switch back to the
-original parent sequence. In the future, a pragmatic solution would be to
-detect such cases during the ARG inference stage and to adjust the copying
-paths accordingly. Meanwhile, however, we urge caution when interpreting the
-parentage of lineages with multiple breakpoints.
-
-\subsubsection{Parameterization of the HMM implementation of Li and Stephens
-model}
-
-Recombination is inferred using an efficient HMM implementation of the LS
-model. Copying paths are estimated from SARS-CoV-2 genomes sampled over time,
-with younger genomes copying (i.e. inheriting segments and mutations) from
-older genomes. Here, we were able to apply the LS model on millions of
-SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of
-the LS model that is a core component of the tsinfer software. For this
-application, we have parameterised the HMM using a mismatch ratio (MMR), which
-tunes how strongly mutations (i.e., no switching of template sequences) are
-preferred over recombination (i.e., switching between sequences) during
-copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
-focal genome more likely spans at least two ancestral genomes (i.e,. the focal
-genome is a putative recombinant) than a single ancestral genome. Here, the MMR
-was set to three in all the HMM runs (see Discussion for future potential
-development of the HMM).
+The LS model in its implementation as an HMM is governed by the transition and emission matrices, $Q$ and $E$, respectively.
+Each may be a function of the reference panel members, but transitions are generally assumed to be independent of the hidden states themselves (shown in the pink panel).
+This dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
+The recombination probability at site $\ell$ is $r\ell$, while n denotes the number of members in the reference panel.
+Emissions may also be a function of the nucleotide states, but in our case, we assume that all observed alleles (in both the reference panel and focal sequence, $a\ell$) at the site under consideration are equally likely and occur with mutation probability $\ell$ (shown in the blue panel).
+These two processes, transition and emission (which encode recombination and mutation, respectively), define the generative process of the HMM, by which an imperfect mosaic is emitted from a reference panel of samples and nodes from the ARG inserted at earlier collection dates (shown in the lower panel).
+Here, the unseen states in the reference panel are shown as coloured lines enclosed by the grey box, and the true path through the data which leads to the emitted focal sequence is displayed by the black arrow.
+Examples of transition and emission probabilities along this trajectory are shown by the red and blue arrows, respectively.
+Black crosses along the focal sequence show genotyping errors or mutations.
+Note that in practice, we do not know the pattern of colours along the focal sequence and must infer it using the Viterbi algorithm.} \end{figure}
+
+The generative process of the LS model is summarised in Figure~\ref{fig:ls_diagram}.
+Here, a transition matrix, $Q$, governs the process of switching (recombining) between members of the reference panel (the hidden states).
+An emission matrix, $E$, allows for differences between the focal sequence and the hidden state from which it is copied (due to mutation as well as genotyping or sequencing error).
+Then, we use the Viterbi algorithm \citep{Viterbi1967-ol} to find the most likely copying path through the sequences, given a set of reference sequences, $Q$, and $E$.
+
+Because the HMM is memoryless, the potential parental lineages are independently assessed at each recombination breakpoint.
+This can result in the inference of a recombinant lineage descending immediately from three (or even more) parental lineages.
+We find that often two of the three parents are identical or similar in sequence in the segments contributed to the recombinant lineage.
+A more parsimonious explanation would be to switch back to the original parent sequence.
+In the future, a pragmatic solution would be to detect such cases during the ARG inference stage and to adjust the copying paths accordingly (see ``Improvements'', below).
+Meanwhile, however, we urge caution when interpreting the parentage of lineages with multiple breakpoints.
+
+\subsubsection{Parameterization of the HMM implementation of Li and Stephens model} 
+Recombination is inferred using an efficient HMM implementation of the LS model.
+Copying paths are estimated from SARS-CoV-2 genomes sampled over time, with younger genomes copying (i.e.
+inheriting segments and mutations) from older genomes.
+Here, we were able to apply the LS model on millions of SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of the LS model that is a core component of the tsinfer software.
+For this application, we have parameterised the HMM using a mismatch ratio (MMR), which tunes how strongly mutations (i.e., no switching of template sequences) are preferred over recombination (i.e., switching between sequences) during copying.
+Under a lower MMR value, the maximum-likelihood HMM copying path of a focal genome more likely spans at least two ancestral genomes (i.e,.
+the focal genome is a putative recombinant) than a single ancestral genome.
+Here, the MMR was set to three in all the HMM runs (see ``Improvements'' for future potential development of the HMM).
 
 \subsection{Analysis} 
-
 \subsubsection{Imputation of Pango lineage for non-sample nodes} 
-
-Non-sample nodes inserted during ARG inference had no Pango lineage assignment
-from Nextclade. We imputed the Pango lineage of non-sample nodes using a
-two-step procedure (Figure 3) to help assess how consistently breakpoint
-intervals are estimated by the HMM runs (see below). If the Pango lineage of
-the parental node was available and no mutation was immediately ancestral to
-the non-sample node, then the parental node’s status was directly propagated to
-the non-sample node. Otherwise, the Pango lineage of the non-sample node was
-inferred by matching its full set of mutations against the Pango
-lineage-defining mutations (based on 90\% consensus of the sequences analysed)
-from the COVID-CG website \citep{Chen2021-zc}; \url{https://covidcg.org/};
-accessed on November 04, 2022).
+Non-sample nodes inserted during ARG inference had no Pango lineage assignment from Nextclade.
+We imputed the Pango lineage of non-sample nodes using a two-step procedure (Figure 3) to help assess how consistently breakpoint intervals are estimated by the HMM runs (see below).
+If the Pango lineage of the parental node was available and no mutation was immediately ancestral to the non-sample node, then the parental node’s status was directly propagated to the non-sample node.
+Otherwise, the Pango lineage of the non-sample node was inferred by matching its full set of mutations against the Pango lineage-defining mutations (based on 90\% consensus of the sequences analysed) from the COVID-CG website \citep{Chen2021-zc}; \url{https://covidcg.org/}; accessed on November 04, 2022).
 
 \subsubsection{Breakpoint intervals and HMM consistency}
-
-When recombinant ancestry for a given genome is inferred by the LS HMM, a
-breakpoint is identified as the last possible position where the copying path
-must switch. Thus, the breakpoints used in the ARG represent the rightmost end
-of the interval, within which the recombination event could have occurred. To
-obtain the leftmost end of this interval, we rerun the HMM using mirrored
-coordinates (i.e., from the 3’ end to the 5’ end).
-
-A recombinant sequence was deemed to be HMM-consistent if (1) the same number
-of parent nodes and (2) the same set of mutations immediately ancestral to it
-were obtained from both the unmirrored and mirrored HMM runs. In the ideal
-case, the copying paths of both the HMM runs switch between parent sequences at
-the same position(s); in less ideal cases, however, the copying paths switch at
-different positions, typically because of a lack of genetic variation between
-the positions. The switch positions in the copying paths of both the HMM runs
-were used to define the breakpoint interval(s) of an HMM-consistent
-recombinant.
+ When recombinant ancestry for a given genome is inferred by the LS HMM, a breakpoint is identified as the last possible position where the copying path must switch.
+Thus, the breakpoints used in the ARG represent the rightmost end of the interval, within which the recombination event could have occurred.
+To obtain the leftmost end of this interval, we rerun the HMM using mirrored coordinates (i.e., from the 3’ end to the 5’ end).
+
+A recombinant sequence was deemed to be HMM-consistent if (1) the same number of parent nodes and (2) the same set of mutations immediately ancestral to it were obtained from both the unmirrored and mirrored HMM runs.
+In the ideal case, the copying paths of both the HMM runs switch between parent sequences at the same position(s); in less ideal cases, however, the copying paths switch at different positions, typically because of a lack of genetic variation between the positions.
+The switch positions in the copying paths of both the HMM runs were used to define the breakpoint interval(s) of an HMM-consistent recombinant.
 
 \subsubsection{Code availability}
-
-The source code implementing the sc2ts pipeline is available via GitHub
-\url{https://github.com/jeromekelleher/sc2ts/}. The sc2ts pipeline uses
-various functions and algorithms implemented in tskit and tsinfer
-\url{https://github.com/tskit-dev/}. The pipeline uses the SciPy
-implementation of the UPGMA algorithm to build daily trees
-\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}.
-The Jupyter notebooks used to produce the results described herein are also
-available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.
+ The source code implementing the sc2ts pipeline is available via GitHub \url{https://github.com/jeromekelleher/sc2ts/}.
+The sc2ts pipeline uses various functions and algorithms implemented in tskit and tsinfer \url{https://github.com/tskit-dev/}.
+The pipeline uses the SciPy implementation of the UPGMA algorithm to build daily trees \url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}.
+The Jupyter notebooks used to produce the results described herein are also available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.
 
 \section{Results}
-
 \subsection{Inferred ARGs} 
-
 To illustrate the use of sc2ts, we employed two different sampling strategies.
-First, we inferred a densely sampled but time-limited ``wide'' ARG by attaching
-all the sample genomes with an MSD of 30 days, one daily batch at a time. The
-inference procedure ran until the genomes sampled on June 30, 2021 were
-attached to the ARG, resulting in a total of 1.27 million samples. Second, we
-inferred a ``long'' ARG over a more extended timespan, by attaching at most 1,000
-genomes, which were randomly subsampled from each daily batch with an MSD
-within 30 days. For this ARG, the inference procedure ran until the genomes
-sampled on June 30, 2022 were attached to the ARG. The long ARG covers an extra
-year of samples but is more sparsely sampled. June 30, 2022 is the collection
-date of the final batch of sequences which were preprocessed for the
-development of sc2ts. Once sc2ts is further developed and tested, we will
-attach samples collected after that date to infer a fully sampled
-pandemic-scale ARG.
+First, we inferred a densely sampled but time-limited ``wide'' ARG by attaching all the sample genomes with an MSD of 30 days, one daily batch at a time.
+The inference procedure ran until the genomes sampled on June 30, 2021 were attached to the ARG, resulting in a total of 1.27 million samples.
+Second, we inferred a ``long'' ARG over a more extended timespan, by attaching at most 1,000 genomes, which were randomly subsampled from each daily batch with an MSD within 30 days.
+For this ARG, the inference procedure ran until the genomes sampled on June 30, 2022 were attached to the ARG.
+The long ARG covers an extra year of samples but is more sparsely sampled.
+June 30, 2022 is the collection date of the final batch of sequences which were preprocessed for the development of sc2ts.
+Once sc2ts is further developed and tested, we will attach samples collected after that date to infer a fully sampled pandemic-scale ARG.
 
@@
 
-Most of the nodes in the ARGs represent sample genomes (Table 1). Only a small
-fraction of the nodes represent recombination events (0.28\% and 0.27\% in the
-wide ARG and long ARG, respectively), over half of which have no descendant
-samples (besides the recombinant sequence that triggered insertion of the
-recombination node) (Table 1). There are over one million mutations on the
-ARGs, the majority of which occur on a terminal edge (Table 1). Notably, the
-massive amount of genomic variation data in the wide ARG (1.27 million genomes)
-and long ARG (657 thousand genomes) are compactly stored in ~57 Mb and ~36 Mb,
-respectively (less than ten bytes per sample). These ARGs can be loaded for
-analysis in a Jupyter notebook on a typical commercial laptop in merely a
-fraction of a second using the Python package tskit (Table 1).
-
-Inferring the wide ARG required 17 days of computation time on a machine
-with 128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz).
-The long ARG required 23 days on a (much older) machine
-with 40 threads and 256 GB RAM (2x Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz).
-The majority of the time is spent on running the HMM to find the copying
-path for each sequence, and the process is therefore highly amenable to
-distributing across multiple machines.
+Most of the nodes in the ARGs represent sample genomes (Table 1).
+Only a small fraction of the nodes represent recombination events (0.28\% and 0.27\% in the wide ARG and long ARG, respectively), over half of which have no descendant samples (besides the recombinant sequence that triggered insertion of the recombination node) (Table 1).
+There are over one million mutations on the ARGs, the majority of which occur on a terminal edge (Table 1).
+Notably, the massive amount of genomic variation data in the wide ARG (1.27 million genomes) and long ARG (657 thousand genomes) are compactly stored in ~57 Mb and ~36 Mb, respectively (less than ten bytes per sample).
+These ARGs can be loaded for analysis in a Jupyter notebook on a typical commercial laptop in merely a fraction of a second using the Python package tskit (Table 1).
+
+Inferring the wide ARG required 17 days of computation time on a machine with 128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz).
+Inferring the long ARG required 23 days on a (much older) machine with 40 threads and 256 GB RAM (2x Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz).
+The majority of the time is spent on running the HMM to find the copying path for each sequence, and the process is therefore highly amenable to distributing across multiple machines.
 
 \subsection{Backbone phylogeny} 
-
-For validation purposes, we compare the genetic relationships in the
-sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain
-project \citep{Hadfield2018-ef}. The NextStrain trees consist of subsamples of
-SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely
-restricted to non-recombinant lineages. To allow for comparison, the ARGs were
-``simplified'' \citep{Kelleher2018-xc} to retain only those samples of our genealogy
-that match those present in a specific NextStrain tree. These simplified
-``backbone'' ARGs exhibit very little recombination (6 and 5 trees across the
-genome for the wide and long ARGs, respectively).
+For validation purposes, we compare the genetic relationships in the sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain project \citep{Hadfield2018-ef}.
+The NextStrain trees consist of subsamples of SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely restricted to non-recombinant lineages.
+To allow for comparison, the ARGs were ``simplified'' \citep{Kelleher2018-xc} to retain only those samples of our genealogy that match those present in a specific NextStrain tree.
+These simplified ``backbone'' ARGs exhibit very little recombination (6 and 5 trees across the genome for the wide and long ARGs, respectively).
 
@@
 \includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
-\caption{\label{fig:cophylogeny} Tanglegram comparing the first local tree in the
-wide ARG (sampled to mid-2021) and an “all-time” global NextStrain tree (sampled
-to January 21, 2023). The trees were pruned down to the samples present in both
-the trees. Light grey lines match the corresponding samples between the two
-trees; black circles indicate identical sample partitions between the two trees.
-Terminal branches are colour-coded according to the Pango lineage status
-assigned to the tip samples; no Pango recombinants are shown, as none are shared
-between the trees. The tanglegram was generated with the aid of the Neighbor-Net
-algorithm \citep{Scornavacca2011-mg} implemented in Dendroscope
-version 3.8.5 \citep{Huson2012-ys}. See
-Figure~\ref{fig:cophylogeny_long} for the equivalent cophylogeny
-for the Long ARG.}
-\end{figure}
-
-Figure~\ref{fig:cophylogeny} shows a co-phylogeny of the 88 samples shared between the wide ARG and
-a GISAID global "all-time" tree from NextStrain (downloaded on January 21,
-2023). To reduce the ARG to a single tree, we show the very first tree in the
-wide ARG; however, the five other trees in this simplified ARG show
-near-identical patterns. Supplementary Figure 3 shows the same comparison for
-the long ARG. It is clear that the backbone topology of the sc2ts tree shows
-very close agreement with the NextStrain tree. The sample genomes cluster by
-their assigned Pango lineage status, and many variants and their descendents
-form identical monophyletic clades in both the trees (e.g., the Alpha and Delta
-clades).
-
-There are, however, some notable differences between the trees. Firstly, the
-sc2ts tree is generally less well resolved (noticeably, February to March,
-2020) (Figure 4). Secondly, there are non-identical sample partitions near the
-tips (e.g., in the Alpha clade, Figure 4). These might have arisen because the
-samples containing the mutations needed to resolve local topologies were
-collected later than the cut-off date for the sc2ts ARG, but were for the
-NextStrain tree. Finally, the branch lengths differ. For example, the branch
-leading to the ancestral node of the Delta variant samples is much shorter in
-the sc2ts tree than the NextStrain tree (Figure 4). It is possible that the
-length of some of the branches was poorly calibrated, probably caused by overly
-aggressive insertion of mutation-collapsing nodes and reversion-pushed nodes or
-suboptimal sample attachment. Side-by-side comparison of sc2ts trees with trees
-inferred using another method (e.g., UShER) should help to understand whether
-branch length differences are due to poor calibration or methodological
-differences.
-
-Although here we focus on a single local tree at an arbitrarily chosen
-position, note that in the original ``un-simplified'' ARGs, there are 1,496 and
-958 trees, respectively (Table 1). Hence, while for a relatively small sample
-the genealogy may be well described by a single tree, the full dataset shows
-substantial variation in phylogenetic relationships (or phylogenetic
-incongruence) across the SARS-CoV-2 genome, much of which may be attributable
-to recombination. Such complex reticulate relationships are appropriately
-captured by a genealogy in the form of an ARG.
+\caption{\label{fig:cophylogeny} Tanglegram comparing the first local tree in the wide ARG (sampled to mid-2021) and an “all-time” global NextStrain tree (sampled to January 21, 2023).
+The trees were pruned down to the samples present in both the trees.
+Light grey lines match the corresponding samples between the two trees; black circles indicate identical sample partitions between the two trees.
+Terminal branches are colour-coded according to the Pango lineage status assigned to the tip samples; no Pango recombinants are shown, as none are shared between the trees.
+The tanglegram was generated with the aid of the Neighbor-Net algorithm \citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5 \citep{Huson2012-ys}.
+See Figure~\ref{fig:cophylogeny_long} for the equivalent cophylogeny for the Long ARG.}
+\end{figure}
+
+Figure~\ref{fig:cophylogeny} shows a co-phylogeny of the 88 samples shared between the wide ARG and a GISAID global "all-time" tree from NextStrain (downloaded on January 21, 2023).
+To reduce the ARG to a single tree, we show the very first tree in the wide ARG; however, the five other trees in this simplified ARG show near-identical patterns.
+Supplementary Figure~\ref{fig:cophylogeny_long} shows the same comparison for the long ARG.
+It is clear that the backbone topology of the sc2ts tree shows very close agreement with the NextStrain tree.
+The sample genomes cluster by their assigned Pango lineage status, and many variants and their descendents form identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC clades, labelled).
+
+Figure~\ref{fig:cophylogeny} reveals some notable differences between the trees.
+Firstly, the sc2ts tree is generally less well resolved (noticeably, February to March, 2020).
+Secondly, there are non-identical sample partitions near the tips (e.g., in the Alpha clade).
+These might have arisen because the samples containing the mutations needed to resolve local topologies were collected later than the cut-off date for the sc2ts ARG, but were for the NextStrain tree.
+Finally, the branch lengths differ.
+For example, the branch leading to the ancestral node of the Delta variant samples is much shorter in the sc2ts tree than the equivalent NextStrain tree.
+It is possible that the length of some of the branches was poorly calibrated, probably caused by overly aggressive insertion of mutation-collapsing nodes and reversion-pushed nodes or suboptimal sample attachment.
+Side-by-side comparison of sc2ts trees with trees inferred using another method (e.g., UShER) should help to understand whether branch length differences are due to poor calibration or methodological differences.
+
+Although the simplified backbone ARGs have few local trees, all of which show essentially the same topologies, note that in the original ``un-simplified'' wide and long ARGs, there are 1,496 and 958 local trees respectively (Table 1).
+Hence, while for a relatively small subset of the genealogy may be well described by a single tree, the full dataset shows substantial variation in phylogenetic relationships (i.e.
+``phylogenetic incongruence'') across the SARS-CoV-2 genome, much of which may be attributable to recombination.
+An ARG representation is therefore essential to fully capture these reticulate genealogical relationships.
 
 \subsection{Mutational spectrum}
-
-In addition to recombination events, the ARGs record mutation events along the
-SARS-CoV-2 phylogeny. \cite{Yi2021-sc} reconstructed a SARS-CoV-2
-phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to
-January 12, 2021, and systematically catalogued all the mutations occurring
-along the phylogeny. Notably, in the mutational spectrum described by Yi et
-al., C-to-U and G-to-U mutations occur more frequently than U-to-C and U-to-G
-mutations, respectively.
+ In addition to recombination events, the ARGs record mutation events along the SARS-CoV-2 phylogeny.
+\cite{Yi2021-sc} reconstructed a SARS-CoV-2 phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to January 12, 2021, and systematically catalogued all the mutations occurring along the phylogeny.
+Notably, they described a mutational spectrum in which C-to-U and G-to-U mutations occur more frequently than U-to-C and U-to-G mutations, respectively.
 
@@
 \caption{\label{fig:mutational_spectra}
-Mutational spectrum of SARS-CoV-2 in
-the wide ARG (sampled to mid-2021) compared to \cite{Yi2021-sc}.
-Mutations are categorised by type (i.e., ancestral state $>$
-derived state). The percentages of each mutation type from the wide ARG are
-represented by blue bars and the percentages from Yi et al. by orange bars,
-with the darker colours representing one direction (e.g., C$>$U) and the
-lighter colours the reverse (e.g., U$>$C).}
-\end{figure}
-
-For validation purposes, we compare the mutational spectrum from the wide ARG
-with the mutational spectrum of Yi et al. We categorised all of the single
-nucleotide mutations by type (defined by the ancestral state and derived
-state), excluded mutations inherited by only a single sample (which are
-enriched for sequencing errors and artefacts), and tallied them up by type.
-Similarly, we took the data for single nucleotide mutations from Yi et al.
-\url{https://github.com/ju-lab/SC2_evol_signature}, excluded mutations
-occurring along terminal branches, and tallied them up by type. We find that
-the mutational spectrum from the wide ARG (448,825 mutations) matches the
-mutational spectrum reported by Yi et al. (92,344 mutations) (Figure X). In
-both spectra, C-to-U mutations and G-to-U mutations occur more frequently than
-U-to-C and U-to-G, respectively (Figure X). Similar results are obtained when
-including the mutations inherited by only a single sample or occurring on
-terminal branches (data not shown).
+Mutational spectrum of SARS-CoV-2 in the wide ARG (sampled to mid-2021) compared to \cite{Yi2021-sc}.
+Mutations are categorised by type (i.e., ancestral state $>$ derived state).
+The percentages of each mutation type from the wide ARG are represented by blue bars and the percentages from Yi et al.
+by orange bars, with the darker colours representing one direction (e.g., C$>$U) and the lighter colours the reverse (e.g., U$>$C).}
+\end{figure}
+
+For validation purposes, we compare the mutational spectrum from the wide ARG with the mutational spectrum of \citet{Yi2021-sc}.
+We categorised all of the single nucleotide mutations by type (defined by the ancestral state and derived state), excluded mutations inherited by only a single sample (which are enriched for sequencing errors and artefacts), and tallied them up by type.
+Similarly, we took the data for single nucleotide mutations from \url{https://github.com/ju-lab/SC2_evol_signature} \citep{Yi2021-sc}, excluded mutations occurring along terminal branches, and tallied them up by type.
+Figure~\ref{fig:mutational_spectra} shows that the mutational spectrum from the wide ARG (based on 448,825 mutations) matches that reported by \citet[based on 92,344 mutations]{Yi2021-sc}.
+In both spectra, C-to-U mutations and G-to-U mutations occur more frequently than U-to-C and U-to-G, respectively.
+Similar results are obtained when including the mutations inherited by only a single sample or those occurring on terminal branches (data not shown).
 
@@
 
-By searching for samples combining genomic segments from Alpha (B.1.1.7) and
-from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations,
-Jackson et al. (Jackson et al. 2021) reported 16 recombinant sequences among
-genomes sampled in the United Kingdom up to March 7, 2021. The authors proposed
-that the recombinants likely arose from eight independent recombination events
-(categorised into groups A to D and four singletons).
+By searching for samples combining genomic segments from Alpha (B.1.1.7) and from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations, \citet{Jackson2021-ik} reported 16 recombinant sequences among genomes sampled in the United Kingdom up to March 7, 2021.
+The authors proposed that the recombinants likely arose from eight independent recombination events (categorised into groups A to D and four singletons).
 
@@
 \end{tabular}
-\caption{\label{tab:jackson}Comparison of recombination breakpoint intervals
-and parent lineages for a subset of the data
-reported by \cite{Jackson2021-ik}. [FIXME ADD MORE DETAIL]
-More detailed results are given in
-Table~\ref{tab:jackson_supplement}.}
+\caption{\label{tab:jackson}Comparison of recombination breakpoint intervals and parent lineages for a subset of the data reported by \cite{Jackson2021-ik}.
+[FIXME ADD MORE DETAIL] More detailed results are given in Table~\ref{tab:jackson_supplement}.}
 \end{table}
 
-In the wide ARG (sampled to mid-2021), all these sequences are descendants of
-recombination nodes, except MILK-103C712, which was removed during
-preprocessing (Table 2); additionally, all are HMM-consistent, except
-QEUH-1067DEF. Overall, we find excellent agreement between Jackson et al.’s
-motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
-and (2) the breakpoint intervals (Table 2). In all cases, the two methods
-proposed identical or very closely related parent lineages. The general mosaic
-genome structure is concordant in all the cases, even CAMC-CB7AB3 where there
-are two switches in the HMM copying paths (indicative of two breakpoints). The
-breakpoint intervals estimated using the two methods overlap substantially in
-all the cases (Table 2). Within each of the groups A to D, the constituent
-sequences have identical breakpoint intervals, because they trace back to a
-single recombination node in the wide ARG. This result is consistent with the
-hypothesis that the sequences of the groups A to D emerged from four separate
-recombination events~\cite{Jackson2021-ik}.
+In the wide ARG (sampled to mid-2021), all these sequences are descendants of recombination nodes, except MILK-103C712, which was removed during preprocessing (Table 2); additionally, all are HMM-consistent, except QEUH-1067DEF.
+Overall, we find excellent agreement between Jackson et al.’s motif-based method and sc2ts in terms of (1) the Pango lineage of the parents and (2) the breakpoint intervals (Table 2).
+In all cases, the two methods proposed identical or very closely related parent lineages.
+The general mosaic genome structure is concordant in all the cases, even CAMC-CB7AB3 where there are two switches in the HMM copying paths (indicative of two breakpoints).
+The breakpoint intervals estimated using the two methods overlap substantially in all the cases (Table 2).
+Within each of the groups A to D, the constituent sequences have identical breakpoint intervals, because they trace back to a single recombination node in the wide ARG.
+This result is consistent with the hypothesis that the sequences of the groups A to D emerged from four separate recombination events~\cite{Jackson2021-ik}.
 
@@
 
-We explore recombination signals across the SARS-CoV-2 genome by inspecting the
-breakpoint intervals of the filtered recombination nodes from the wide ARG
-(sampled to mid-2021) in Figure~\ref{fig:breakpoint-distribution}.
-The length of a breakpoint interval reflects
-uncertainty in the estimated location of a breakpoint. On average, the length
-of the breakpoint intervals is 2,895 bases (median, 1,670 bases; range, 1 to
-23,666 bases). This broad distribution of interval length indicates that the
-location of a breakpoint often could not be precisely determined given the
-observed genetic variation.
+We explore recombination signals across the SARS-CoV-2 genome by inspecting the breakpoint intervals of the filtered recombination nodes from the wide ARG (sampled to mid-2021) in Figure~\ref{fig:breakpoint-distribution}.
+The length of a breakpoint interval reflects uncertainty in the estimated location of a breakpoint.
+On average, the length of the breakpoint intervals is 2,895 bases (median, 1,670 bases; range, 1 to 23,666 bases).
+This broad distribution of interval length indicates that the location of a breakpoint often could not be precisely determined given the observed genetic variation.
 
@@
 
-There is a preponderance of breakpoint intervals spanning the 3’ end of the
-genome (Figure 4A), as noted previously by
-\cite{Turakhia2022-it}.
-Inspecting more closely, we find an enrichment in the
-number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region,
-and 3’ untranslated region (Figure 4B). Markedly, there is an abrupt drop in
-the number of intersecting intervals in the N gene following the peak in ORF8
-towards the 3’ end.
-
-Next, we investigate the potential recombination signals alongside the number
-of mutations across the genome (Figure 4C). Regions with a higher density of
-mutations contain more genetic variation, increasing the power to detect
-recombination events. The position 28,271 (located in the ORF8-N intergenic
-region) has a particularly high number of mutations in the samples (Figure 5).
-At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions
-(n = 1,605) appear to occur frequently. One possible explanation is that this
-homopolymeric region is prone to sequencing errors, giving rise to an
-artifactual recombination signal. At the upstream position 27,972, numerous
-mutations also occur. This position has been reported to be hypermutable
+There is a preponderance of breakpoint intervals spanning the 3’ end of the genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}.
+Inspecting more closely, we find an enrichment in the number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure 4B).
+Markedly, there is an abrupt drop in the number of intersecting intervals in the N gene following the peak in ORF8 towards the 3’ end.
+
+Next, we investigate the potential recombination signals alongside the number of mutations across the genome (Figure 4C).
+Regions with a higher density of mutations contain more genetic variation, increasing the power to detect recombination events.
+The position 28,271 (located in the ORF8-N intergenic region) has a particularly high number of mutations in the samples (Figure 5).
+At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions (n = 1,605) appear to occur frequently.
+One possible explanation is that this homopolymeric region is prone to sequencing errors, giving rise to an artifactual recombination signal.
+At the upstream position 27,972, numerous mutations also occur.
+This position has been reported to be hypermutable
 \citep{Jungreis2021-dh}.
 
-We further consider the breakpoint intervals within the context of
-recombination breakpoint sequence motifs. It is hypothesised that certain
-palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote
-template switching during replication in SARS-CoV-2 via formation of
-base-paired stem loops in the genome structure \citep{Gallaher2020-lb}. There
-are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019
-reference sequence (Supplementary Table 5). We observe that XX (XX\%) of the
-breakpoint intervals of the HMM-consistent recombinants span at least one of
-the breakpoint sequences.
+We further consider the breakpoint intervals within the context of recombination breakpoint sequence motifs.
+It is hypothesised that certain palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote template switching during replication in SARS-CoV-2 via formation of base-paired stem loops in the genome structure \citep{Gallaher2020-lb}.
+There are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019 reference sequence (Supplementary Table 5).
+We observe that XX (XX\%) of the breakpoint intervals of the HMM-consistent recombinants span at least one of the breakpoint sequences.
 
@@
 \includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
-\caption{\label{fig:recomb_mrcas} Divergance between parents at
-recombination breakpoints. (A) All recombination nodes that are ancestral
-to least two samples. (B) Further filtered as described in the text.
+\caption{\label{fig:recomb_mrcas} Divergance between parents at recombination breakpoints.
+(A) All recombination nodes that are ancestral to least two samples.
+(B) Further filtered as described in the text.
 MORE DETAIL. Horizontal lines represent WHAT,
-Figure~\ref{fig:recomb_mrcas_lineage_breakdown} shows the same data
-broken down by parental lineage.}
-\end{figure}
-
-Both the wide and long ARGs contain a large number of recombination nodes
-(4,123 and 2,078, respectively; Supplementary Tables 1 and 2), with each of
-these nodes inferred as a result of matching a specific sample (a recombinant)
-into the ARG. However, ~65\% of recombination nodes are ancestors of only a
-single descendant sample in each local tree; these could have arisen from
-sequencing of mixed samples (e.g., from co-infection or contamination), or were
-simply not transmitted onwards and so were excluded from subsequent analyses
-(leaving X and Y recombinants in the wide and long ARGs, respectively). In
-their analysis using RIPPLES, \cite{Turakhia2022-it} detected 589 recombinants,
-also excluding singletons.
-
-For simplicity, we focus primarily on ``non-complex'' recombination nodes. To do
-this, we apply a filtering step that removes those with more than one
-breakpoint (and often more than two parent lineages), or where matching is not
-HMM-consistent or identifies parents with different Pango lineage assignments
-in the original (i.e., 5’ to 3’ end) and mirrored (i.e., 3’ to 5’ end) HMM
-runs. Filtering reduces both the number of recombination nodes and breakpoints
-from  A \& B to C in the wide ARG and from 726 \& 806 to 459 in the long ARG
-(Supplementary Tables 1 and 2). Note that in the unfiltered case, a single
-recombination node can correspond to multiple points, one for each
-recombination breakpoint. The effect of this filtering step in the long ARG,
-which includes time periods of increased rates of coinfection and recombination
-compared to the wide ARG, can be seen in Figure 3A (unfiltered) and Figure 3B
-(filtered).
-
-Importantly, the long ARG captured all except one of the Pango designated
-recombinant lineages (see the section below), but we also detected a large
-number of additional recombination events, with parents lineages that diverged
-as far back as two years to as recent as five or six days prior to the
-recombination event (Figure 3). This provides a degree of resolution of viral
-recombinants not available with other recombination detection methods. The
-majority of newly identified recombinants arise from parents that diverged 10
-to 30 weeks previously. However, there is also a peak of recombinants that
-diverged ~95 weeks prior to the event: these correspond to recombinants whose
-parental common ancestor traces to early 2020, including both Delta-Delta and
-Delta-Omicron recombinants (Supplementary Figure 4).
-
-The date of common ancestry for the two parents of recombinant lineages is
-concentrated in several banded rows in Figure 5. These are largely due to a
-small handful of most recent common ancestors (MRCAs), which are shared between
-a large number of recombination events, the top four of which are indicated by
-dotted lines in the figure. These MRCAs are near the root of large expansions
-of certain lineages (B.1: original variant;  B.1.617.2: Delta, BA.1: first
-Omicron wave; and BA.2: second Omicron wave), as evidenced by the number of
-samples with that Pango designation that trace back to these ancestral nodes
-(listed as percentages on the plot). The specific MRCA nodes to which many
-recombinants trace frequently represent very large polytomies in the tree (at
-the extreme, 2,872 immediate children). This is likely to indicate a rapid and
-under-sampled expansion of a clade in the SARS-CoV-2 genealogy.
-
-Finally, it is clear from Supplementary Figure 4 that of the recombinant
-parents that can be classified into Alpha, Delta, and Omicron variants, the
-majority are within-Delta and within-Omicron recombinants, and that Omicron
-followed by Delta, are the variants associated with the most recombination.
-This may reflect either sampling intensity, the prevalence of cases (which
-increases the chance of coinfection and recombination), or possibly
-heterogeneity in recombination rates among lineages.
+Figure~\ref{fig:recomb_mrcas_voc_breakdown} shows the same data broken down by parental ``VoC'' classification.}
+\end{figure}
+
+Both the wide and long ARGs contain a large number of recombination nodes (4123 and 2078 respectively; Supplementary Tables 1 and 2).
+Each of these nodes were inferred as a result of matching a specific recombinant sample into the ARG.
+However, ~63\% of recombination nodes are ancestral to only a single recombinant sample; we exclude these "singleton recombinants" from subsequent analyses as they may have arisen from sequencing of mixed samples (e.g., from co-infection or contamination), and in any case, never result in onward transmission.
+This leaves 1522 and 763 recombination nodes in the wide and long ARGs, respectively.
+In their analysis using RIPPLES, \cite{Turakhia2022-it} detected 589 non-singleton recombinants.
+
+We focus on the long ARG because compared to the wide ARG,  it includes time periods of increased rates of coinfection and recombination, and contains many samples previously identified as the product of recombination (i.e.
+with a Pango lineage designation starting with an "X").
+Not only does the long ARG capture all but one of the Pango designated recombinant lineages during this time period (see the section below), but it also includes a large number of non-Pango-identified recombinants.
+Many of these involve recombination between relatively closely related lineages, as measured by the divergence time between the recombination events and the most recent common ancestor (MRCA) shared by  the left and right parent at each breakpoint.
+As can be seen from Figure~\ref{fig:recomb_mrcas}, this divergence time ranges from as little as five or six days, to as much as several years in the case of recent recombinants.
+This demonstrates that sc2ts can identify viral recombination between very closely related lineages, providing a degree of resolution not available with other recombination detection methods.
+
+The histograms in Figure~\ref{fig:recomb_mrcas} show that
+majority of newly identified recombinants arise from parents that diverged 10 to 30 weeks previously.
+However, there is also a peak of recombinants that diverged ~95 weeks prior to the event: these correspond to recombinants whose parental common ancestor traces to early 2020, including both Delta-Delta and Delta-Omicron recombinants (Supplementary Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
+Note that each point in these plots represents a breakpoint, so that recombination events with more than one breakpoint (e.g.
+with 3 or more parents, which comprise X \% of the recombination events) are represented by several plotted points.
+
+As a check on the reliability of these identified recombination events, Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are "HMM-consistent"; that is, where the mirrored HMM runs (i.e., 5’ to 3’ end versus 3’ to 5’ end) identify the same set of originating mutations and the same number of breakpoints, and the parents in the mirrored runs are sequence-identical.
+This reduces both the number of recombination nodes / breakpoints from 1522 / B to C / D in the wide ARG and from 763 / 806 to 459 in the long ARG (Supplementary Tables 1 and 2).
+
+The date of common ancestry for the two parents of recombinant lineages is concentrated in several banded rows in Figure~\ref{fig:recomb_mrcas}.
+These are largely due to a small handful of MRCAs which are shared between a large number of recombination events; the top four are indicated by dotted lines in the figure.
+These shared MRCAs are near the root of large expansions of certain lineages (B.1: original variant;  B.1.617.2: Delta, BA.1: first Omicron wave; and BA.2: second Omicron wave), as evidenced by the number of samples with that Pango designation that trace back to these ancestral nodes (listed as percentages on the plot).
+The specific MRCA nodes to which many recombinants trace frequently represent very large polytomies in the tree (at the extreme, 2,872 immediate children).
+This is likely to indicate a rapid and under-sampled expansion of a clade in the SARS-CoV-2 genealogy.
+
+Finally, it is clear from Supplementary Figure 4 that of the recombinant parents that fall into the Alpha, Delta, and Omicron VoC classes, the majority are within-Delta and within-Omicron recombinants, and that Omicron followed by Delta, are the variants associated with the most recombination.
+This may reflect either sampling intensity, the prevalence of cases (which increases the chance of coinfection and recombination), or possibly heterogeneity in recombination probabilities among lineages.
 
 \subsection{Recombinant Pango lineages}
-[Intro bit saying what we're doing]
-
+
+In this section we focus on the detailed genealogy of samples that have been previously identified as recombinants, i.e.
+given a Pango designation that starts with an ``X''.
+Since the wide ARG is restricted to data collected prior to mid-2021, it contains samples from only three Pango-designated recombinant lineages: XA, XB, and XC.
+To investigate the diversity of genealogies associated with recombinants, we therefore devote more analysis to patterns in the long ARG.
+As in previous sections, we exclude singleton recombinants where possible.
+
+\subsubsection{Wide ARG}
 [Quick overview of the wide ARG.]
-The wide ARG contains samples from three Pango-designated recombinant
-lineages (XA, XB, and XC) and N samples, since it only includes data
-up to mid-2021.
-In the wide ARG, every sequence of the three Pango recombinants trace back to
-some recombination node
-
-[Details of the long ARG]
-In this section we examine the Long ARG to determine how well it reflects
-the structure of previously identified recombinant lineages.
-There are [high level summary of the pango recombinant lineages and samples
-in the Long ARG].
-long ARG (subsampled to mid-2022) contains 749 samples
-from 33 Pango-designated recombinant lineages (XA to XAK; Supplementary Table)
-
-Of these 749 samples, all but 2 are inferred to descend from a recombination
-node. These two samples are designated XP, and a likely explanation
-for their not have an inferred recombinant origin in the long ARG is that
-the characteristic multibase deletion for XP
-(\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
-during preprocessing. An improved masking pipeline is an important avenue
-for future development (see the Discussion).
-
-38 of the remaining X designated samples descend from recombination
-nodes that are ancestral only to that sample
-(10 are XZ; 6 are XE;  3 each from
-XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1 sample from each
-of XB, XM, XJ, XAF, XAH and XAJ).
-Such samples are likely to be enriched for sequencing errors and
-lineage designation artefacts, and so we exclude them from further
-analysis in this section.
-
-The most recent recombination node for 79 samples
-(XN: 53, XZ: 16, XAJ: 6, XE: 1, XAD: 1, XAH: 1, XAK: 1)
-is the same. This node is ancestral to more than 127K samples,
-and is likely to be a spuriously inferred recombination event.
-Understanding the origin of this and other similar nodes, is
-an important direction for future work (see Discussion).
-For simplicity, we exclude these samples from further
-analyses in this section.
-
-The remaining remaining 630 Pango X designated sequences trace
-back to 50 different recombination nodes. We examine the
-inferred inheritance structure in the following subsections.
+In the wide ARG, 44 samples are designated as XA by both Nextclade and GISAID, 237 are designated as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
+After removing singleton recombinants, XA numbers remain unchanged, but XB is reduced to 235 (229) and XC to 4 (0).
+We correctly identify that all samples designated as XB by any method have one or more recombination nodes in their ancestry.
+
+The simplest Pango recombinant genealogy is XA: all XA samples trace back to a unique originating recombination node.
+This is the product of a recombination between a B.1.177.18 sample (specifically the strain Wales/ALDP-115BF41/2021, EPI\_ISL: 1012804) which contributed the majority of the genome from the start to a maximum position of 22227bp, and an unknown (inserted) node with imputed Pango lineage B.1.1.7, which contributed the remaining right hand portion.
+The recombination node has five immediate children: four sample leaves (EPI\_ISLs: 989697, 1019487, 1104468 and 1122630) and a UPGMA-inserted node which is the ancestor of all other XA samples in the dataset.
+
+For XB, all samples trace back to a originating node which is the product of a recombination between a B.1 sample (specifically the strain England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of the genome from the start up to a maximum position of 23604bp), and two other UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of 27389bp) and B.1.36.8 (the remaining fragment of the genome).
+As described below, in the long ARG the equivalent recombination node has only 2 parents, with no involvement of  B.1.36.8, and it is possible that the third parent in the wide ARG is artifactual.
+An added complication is that in both ARGs, additional non-X-designated samples such as descend from the not only a small number of are six further recombination nodes nested within the XB ``clade''.
+However, five are ancestral to only negligible fractions of the designated XB samples.
+The sixth accounts for about 17\% of the designated XB nodes, and involves a recombination between descendants of the originating XB recombination.
+More specifically strain USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination between an XB grandparent and its XB grandchild.
+This could well be artifactual.
+
+For XC, the wide ARG identifies more than one originating recombination node (see "multiple origins" below).
+However, as none of the samples designated as XC by Nextclade are designated as XC by GISAID, these patterns could be due to mistaken labelling, hence we do not examine XC further.
+
+\subsubsection{Long ARG}
+[high level summary of the pango recombinant lineages and samples in the Long ARG]
+The long ARG (subsampled to mid-2022) contains 33 Nextclade-designated Pango recombinant lineages (XA to XAK; Supplementary Table XXX) comprising 749 samples (711 when singleton recombinants are removed).
+If the GISAID designations are used, 38 Pango lineages exist in the long ARG, comprising only 515 samples.
+In the analyses below, we mainly focus on the larger number of Nextclade designated samples.
+
+Of the 749 Nextclade recombinant samples, all but 2 are inferred to descend from a recombination node.
+These two samples are designated XP, and a likely explanation for their not have an inferred recombinant origin in the long ARG is that the characteristic multibase deletion for XP (\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked during preprocessing.
+An improved masking pipeline is an important avenue for future development (see the Discussion).
+
+38 of the remaining X designated samples are singleton recombinants, descending from recombination nodes that are ancestral only to that sample (10 are XZ; 6 are XE;  3 each from XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1 sample from each of XB, XM, XJ, XAF, XAH and XAJ).
+Such samples are likely to be enriched for sequencing errors and lineage designation artefacts, and so we exclude them from further analysis in this section.
+
+The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE: 1, XAD: 1, XAH: 1, XAK: 1) is the same.
+This node is ancestral to more than 127K samples, and is likely to be a spuriously inferred recombination event.
+Understanding the origin of this and other similar nodes, is an important direction for future work (see Discussion).
+For simplicity, we exclude these samples from further analyses in this section.
+
+The remaining remaining 630 Pango X designated sequences trace back to 50 different recombination nodes. We examine the inferred inheritance structure in the following subsections.
 
@@
 \subsubsection{Single origin}
-
-We would usually assume that all sequences assigned to a given Pango X
-lineage are descendants of a single recombinant sequence, arising as a
-result of a mixed infection followed by onward transmission. Thus, we
-would also expect our ARG to reflect this evolutionary history, where
-we have a all the samples assigned to this recombinant lineage tracing
-back to a single recombination node, representing this original
-recombination event.
+We would usually assume that all sequences assigned to a given Pango X lineage are descendants of a single recombinant sequence, arising as a result of a mixed infection followed by onward transmission.
+Thus, we would also expect our ARG to reflect this evolutionary history, where we have a all the samples assigned to this recombinant lineage tracing back to a single recombination node, representing this original recombination event.
 
@@
 \end{tabular}
-
 \caption{\label{tab:pango-single-origin}
-Table summarising the status of the Pango X-lineages in the
-Long ARG that are inferred to have a single recombinant origin
-based on the NextClade Pango lineage assignments. Imputed parent lineages
-are shown from the ARG and also to parent lineages as defined by
-the official Pango designation alias key.
-FIXME: if we like this table then we can reformat it so that lists
-at the end are spread over multiple lines rather than using Python
-dict notation }
+Table summarising the status of the Pango X-lineages in the Long ARG that are inferred to have a single recombinant origin based on the NextClade Pango lineage assignments.
+Imputed parent lineages are shown from the ARG and also to parent lineages as defined by the official Pango designation alias key.
+FIXME: if we like this table then we can reformat it so that lists at the end are spread over multiple lines rather than using Python dict notation }
 \end{table}
@@
 \caption{\label{fig:pango-single-origin-graph}
-Single origin Pango X lineages. (A) Subgraph for XA (Wide ARG).
-(B) Details for the the single-origin Pango
-X lineages in the long ARG. (C) XAA, XAB and XAG share a single recombinant
-origin in the Long ARG. }
-\end{figure}
-
-The sequences of 16 Pango recombinant lineages trace back to a single recombination
-node, as summarised in Table~\ref{tab:pango-single-origin}.
+Single origin Nextclade Pango X lineages.
+(A) Subgraph for XA in the long ARG: all XA samples and their ancestral lineages are shown back to the nearest sampled viral genome, with dotted lines showing continuation of the ARG.
+Nodes are coloured by Nextclade Pango designation; smaller nodes are not samples and their Pango designation was imputed where possible.
+(B) Details for the single-origin Pango X lineages in the long ARG.
+(C) XAG origins in the long ARG.}
+\end{figure}
+
+In the long ARG, the sequences of 16 Pango recombinant lineages trace back to a single recombination node, as summarised in Table~\ref{tab:pango-single-origin}.
 [FIXME UPDATE THESE]
-We can
-see that XA, XAC, XAE, and XF form monophyletic clades, in which all of the
-samples given that Pango designation descendand from a single recombination
-node.
+We can see that XA, XAC, XAE, and XF form monophyletic clades, in which all of the samples given that Pango designation descendand from a single recombination node.
 
@@
 
-There is generally good agreement on the imputed parental lineages from
-the ARG and the official Pango designations. [Discuss more]
-
-[Discuss panel C where XAA, XAB,XAG and share a common recombinant origin]
+There is generally good agreement on the imputed parental lineages from the ARG and the official Pango designations. [Discuss more]
+
+[Discuss panel C where XAG shares a common recombinant origin with samples from several other X lineages]
+All XAG samples share a single origin; however this is also shared with all XAA samples, and includes some, but not all, of the samples identified as XAB, XQ, and XU by Nextclade.
+Note that if the GISAID designations are used, many of the samples marked here as XAB are reclassified as BA.2, and XAG becomes fully monophyletic (supplementary figure \ref{fig:pango_XAG_gisaid_graph}; this also gives identities of the samples in the graph)
 
 \subsubsection{Multiple origin}
-The ARG inference has no pre-defined knowledge of Pango X lineage assignments,
-and there is therefore no particular requirement that all the samples
-assigned to a given lineage must trace back to a single recombination
-event. In this section we examine the Pango X lineage samples that
-are inferred to have multiple recombinant origins (i.e., trace
-back to different recombination nodes in the ARG), but are still
-form monophyletic clades % is this the right term?
+The ARG inference has no pre-defined knowledge of Pango X lineage assignments, and there is therefore no particular requirement that all the samples assigned to a given lineage must trace back to a single recombination event.
+In this section we examine the Pango X lineage samples that are inferred to have multiple recombinant origins (i.e., trace back to different recombination nodes in the ARG), but still form monophyletic clades % is this the right term?
 descending from these nodes .
@@
 Example lineage subgraph
+% \includegraphics[width=0.4\textwidth]{figures/Pango_XD_nxcld_tight_graph.pdf}
+
  &
@@
 \end{tabular}
-\caption{A. Example subgraph for a multi origin example. B. details
-of inferred breakpoints for different recombination events inferred
-for XE samples (170 samples and 12 inferred recombination nodes; some are very
-similar (Official: BA.1* \& BA.2, 10448–11287)}
-\end{figure}
-
-In Figure XXX we give the details of these multiple pango lineages,
-along with an example lineage subgraph.
-
-Interestingly, XC is inferred to be single origin in the Wide ARG but
-not in the long ARG. See lineage subgraphs in supplemental figure Y.
-This suggests that insufficient sampling may be the explanation for
-many of these multiple origin events.
-
-\subsubsection{Complex relationships between X lineages}
-
-Some lineages are inferred to have overlapping ancestral histories.
-Discuss XAG example in Figure XXX.
+\caption{A.
+Example subgraph for XD in the long ARG, which shows .
+B.
+details of inferred breakpoints for different recombination events inferred for XE samples (170 samples and 12 inferred recombination nodes; some are very similar (Official: BA.1* \& BA.2, 10448–11287)}
+\end{figure}
+
+In Figure XXX we give the details of these multiple pango lineages, along with an example lineage subgraph.
+
+% Interestingly, XC is inferred to be single origin in the Wide ARG but not in the long ARG.
+See lineage subgraphs in supplemental figure Y.
+This suggests that insufficient sampling may be the explanation for many of these multiple origin events.
+
+\subsubsection{Complex origins}
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
+\caption{\label{fig:complex_origins_graph}  DESCRIBE ME}
+\end{figure}
+
+As well as cases where a single recombination event explains several X lineages (such as XAG above), many of the Pango X lineages involve more complex origins, in which samples which have not been previously identified as recombinants also descend from originating recombination events.
+In addition these complex topologies often include extra recombination nodes within the topology that descends from the main originating event.
+An example is XB in which recombination between a B.1 sample and B.1.627 sample leads both to the XB samples and to a "hairball" of further recombination nodes whose descendants have not been identified as recombinants by Nextclade (blue nodes, Figure~\ref{fig:complex_origins_graph})  ...
 
 \subsection{Novel recombinants}
-
-Next, we examine recombination nodes in the long ARG (sampled to
-mid-2022) that represent previously unnamed putative recombinant sequences,
-which do not have a Pango recombinant designation. For this proof-of-concept
-study, we have arbitrarily picked 12 recombination nodes that seem plausible
-(Supplementary Table 4). All these nodes involve Omicron
-subvariants (BA.1, BA.2, BA.4, and BA.5). These recombination nodes (1) were
-inserted into the long ARG on or after January 8, 2022; (2) have at least 10
-descendent samples; and (3) have no mutational differences (including immediate
-reversions) from their parent nodes.
-
-TODO: Node  740761 (USA/NC-CDC-LC0668306/2022) is a sister clade proposed1006
-in the UShER public phylogeny. Node 628656 (Scotland/QEUH-37794BB/2022) is a
-sister clade of XAC, and annotated as ``miscBA2BA1PostSpike''.
+Next, we examine recombination nodes in the long ARG (sampled to mid-2022) that represent previously unnamed putative recombinant sequences, which do not have a Pango recombinant designation.
+For this proof-of-concept study, we have arbitrarily picked 12 recombination nodes that seem plausible (Supplementary Table 4).
+All these nodes involve Omicron subvariants (BA.1, BA.2, BA.4, and BA.5).
+These recombination nodes (1) were inserted into the long ARG on or after January 8, 2022; (2) have at least 10 descendent samples; and (3) have no mutational differences (including immediate reversions) from their parent nodes.
+
+TODO: Node  740761 (USA/NC-CDC-LC0668306/2022) is a sister clade proposed1006 in the UShER public phylogeny.
+Node 628656 (Scotland/QEUH-37794BB/2022) is a sister clade of XAC, and annotated as ``miscBA2BA1PostSpike''.
 
 \section{Discussion}
-
-The COVID-19 pandemic continues to be a global emergency, with persistent high
-levels of infection. This high prevalence has allowed the proliferation of many
-variants, with more than 600 Pango-designated lineages circulating globally in
-the last three months (January to March, 2023; https://gisaid.org/; access on
-March 27, 2023). High prevalence also brings a higher risk of coinfection,
-increasing opportunities for new phenotypically distinct recombinants to emerge
-and spread.
-
-In this proof-of-concept study, we demonstrate the feasibility of
-reconstructing a large ARG from over one million SARS-CoV-2 genomes, a feat
-that is simply not possible using other available ARG inference methods
-(Rasmussen et al. 2014; Ignatieva et al. 2021; Speidel et al. 2019). We have
-accomplished this feat by building on top of recent advances in storing and
-processing large ARGs (tskit) and inferring genealogical relationships
-efficiently at scale (tsinfer). The highly compact representation of the ARGs
-as succinct tree sequences, coupled with vectorised computing technologies,
-enables the large SARS-CoV-2 ARGs to be loaded in less than one second in a
-Jupyter notebook and allows for each SARS-CoV-2 genome to be stored in less
-than eleven bytes of data. We believe that further development of the sc2ts
-methodology and software can create a powerful tool (1) to efficiently detect
-emerging or previously overlooked recombinants in SARS-CoV-2; (2) to jointly
-model the mutational and recombinational processes shaping global SARS-CoV-2
-genomic diversity; and (3) to perform a broad variety of evolutionary analyses
-of SARS-CoV-2 genomes with unprecedented global sequencing coverage.
-
-ARGs are rich genealogical structures that contain information about the
-genetic relationships among sampled genomes in the presence of recombination.
-Here, we have conducted several analyses of the preliminary ARGs to show that
-the ARGs built using sc2ts can (1) recapitulate the phylogenetic relationships
-among major SARS-CoV-2 lineages; (2) recover biologically plausible
-recombination signals in SARS-CoV-2 genomes; and (3) model mutation events in
-SARS-CoV-2 evolutionary history.
-
-ARGs reconstructed using sc2ts capture the phylogenetic relationships of major
-SARS-CoV-2 lineages. Our results show that the backbone topology of a densely
-sampled wide ARG inferred using sc2ts agrees well with the backbone topology of
-a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference
-package. However, we have noted several areas of differences (e.g., poorly
-calibrated branch lengths), which are the focus of future development of sc2ts.
-We believe that some of these differences can be reduced by improving and
-refining our approach and strategy to handle samples with erroneously reported
-collection dates (i.e., time travellers that have passed the maximum submission
-delay of 30 days) and to modify the topology of daily trees.
-
-Our results show that sc2ts identifies the early recombinants (pre-2022)
-previously identified by \cite{Jackson2021-ik}. Sc2ts also identifies many
-Pango recombinants, which are designated after community perusal of supporting
-genomic evidence. We have found several previously unreported and overlooked
-recombinant sequences in the wide ARG. After further improvements and
-enhancements, we will use sc2ts to build a comprehensive ARG to search for
-recombination signals among all the 15 million and counting SARS-CoV-2 genomes
-available.
-
-``The rate of putative recombination breakpoints is about three times higher
-towards the 3' of the change point than the 5' interval.`` Consistent with this,
-the breakpoint intervals of the putative recombinants identified here tend to
-occur on the 3’ end of the SARS-CoV-2 genome.
+The COVID-19 pandemic continues to be a global emergency, with persistent high levels of infection.
+This high prevalence has allowed the proliferation of many variants, with more than 600 Pango-designated lineages circulating globally in the last three months (January to March, 2023; https://gisaid.org/; access on March 27, 2023).
+High prevalence also brings a higher risk of coinfection, increasing opportunities for new phenotypically distinct recombinants to emerge and spread.
+
+In this proof-of-concept study, we demonstrate the feasibility of reconstructing a large ARG from over one million SARS-CoV-2 genomes, a feat that is simply not possible using other available ARG inference methods \citep{Rasmussen2014-el,Ignatieva2021-rg, Speidel2019-yh}.
+We have accomplished this feat by building on top of recent advances in storing and processing large ARGs (tskit) and inferring genealogical relationships efficiently at scale (tsinfer).
+The highly compact representation of the ARGs as succinct tree sequences, coupled with vectorised computing technologies, enables the large SARS-CoV-2 ARGs to be loaded in less than one second in a Jupyter notebook and allows for each SARS-CoV-2 genome to be stored in less than eleven bytes of data.
+We believe that further development of the sc2ts methodology and software can create a powerful tool (1) to efficiently detect emerging or previously overlooked recombinants in SARS-CoV-2; (2) to jointly model the mutational and recombinational processes shaping global SARS-CoV-2 genomic diversity; and (3) to perform a broad variety of evolutionary analyses of SARS-CoV-2 genomes with unprecedented global sequencing coverage.
+
+ARGs are rich genealogical structures that contain information about the genetic relationships among sampled genomes in the presence of recombination.
+Here, we have conducted several analyses of the preliminary ARGs to show that the ARGs built using sc2ts can (1) recapitulate the phylogenetic relationships among major SARS-CoV-2 lineages; (2) recover biologically plausible recombination signals in SARS-CoV-2 genomes; and (3) model mutation events in SARS-CoV-2 evolutionary history.
+
+ARGs reconstructed using sc2ts capture the phylogenetic relationships of major SARS-CoV-2 lineages.
+Our results show that the backbone topology of a densely sampled wide ARG inferred using sc2ts agrees well with the backbone topology of a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference package.
+However, we have noted several areas of differences (e.g., poorly calibrated branch lengths), which are the focus of future development of sc2ts.
+We believe that some of these differences can be reduced by improving and refining our approach and strategy to handle samples with erroneously reported collection dates (i.e., time travellers that have passed the maximum submission delay of 30 days) and to modify the topology of daily trees.
+
+Our results show that sc2ts identifies the early recombinants (pre-2022) previously identified by \cite{Jackson2021-ik}.
+Sc2ts also identifies many Pango recombinants, which are designated after community perusal of supporting genomic evidence.
+We have found several previously unreported and overlooked recombinant sequences in the wide ARG.
+After further improvements and enhancements, we will use sc2ts to build a comprehensive ARG to search for recombination signals among all the 15 million and counting SARS-CoV-2 genomes available.
+
+``The rate of putative recombination breakpoints is about three times higher towards the 3' of the change point than the 5' interval.`` Consistent with this, the breakpoint intervals of the putative recombinants identified here tend to occur on the 3’ end of the SARS-CoV-2 genome.
 
 \section{Improvements}
-
-Currently, sc2ts is in development. Below, we propose several ways to
-potentially improve or enhance sc2ts. We welcome suggestions and ideas from
-experts.
+Currently, sc2ts is in development. Below, we propose several ways to potentially improve or enhance sc2ts. We welcome suggestions and ideas from experts.
 
 \begin{itemize}
-\item A potential source of error in sc2ts ARGs comes from
-incorrectly reported collection dates of the samples, which can cause sample
-genomes to be attached to the wrong parts of the ARGs. Thus far, we have
-attempted to correct the most offending errors in the ARGs by filtering out
-"time travellers", but the sampling time can still be wrong by days or tens of
-days. ***describe possible approaches to correct this?***
-\item To avoid time
-travellers, we used a maximum submission delay, but this will systematically
-bias against submissions from countries with greater submission delays (Kalia
-et al.). An area of future work is to add back the excluded sequences to a
-built ARG to assess which ones are likely true time travellers.
-\item To avoid
-time travellers, we used a maximum submission delay, but this will
-systematically bias against submissions from countries with greater submission
-delays (Kalia et al.). An area of future work is to add back the excluded
-sequences to a built ARG to assess which ones are likely true time travellers.
-\item We have used a single mismatch value in all the HMM runs, set so that
-having three mutations is as likely as having a single recombination. This
-parameter choice is arbitrary, and higher values may reduce the number of
-origins for those Pango lineages that were inferred to descend from more than
-one recombinant ancestor. In addition, because both the mutation rate and
-recombination rate varies across the SARS-CoV-2 genome, it may be helpful to
-allow the mismatch value to vary along the genome.
-\item By treating mutations
-equally in the current HMM implementation, we implicitly assume that the rates
-of different types of mutations are identical. However, as reported previously
-(Yi et al. 2021), there is a high degree of mutational asymmetry in SARS-CoV-2.
-To account for mutational biases, more sophisticated nucleotide substitution
-models (e.g., allowing for different rates of transition and transversion) can
-be incorporated into HMM likelihood calculations.
-\item Our current masking
-strategy handles indels as missing data. Doing this can remove information
-helpful to detect recombinants. For example, sequences of the Pango recombinant
-XP were not identified as recombinants, because XP’s characteristic multibase
-deletion was masked.
-\item This "lack of memory" in the HMM can create
-recombinants with three or more parents when it would be equally parsimonious
-to have just two. The copying process in the HMM currently does not limit the
-number of parents to two, but doing so would be a valuable addition.
-\item Some
-recombinants may reasonably have three or more parents, indicating multiple
-"stacked" recombination events. We have largely ignored these in the current
-analysis, but with further sc2ts development, there is potential to analyse
-these putative recombinants and trace their evolutionary origins through the
-ARG.
-\item In the present implementation, HMM traceback is performed once,
-yielding a single maximum-likelihood genome copying path. A stochastic
-traceback algorithm (e.g., (Rasmussen et al. 2014)) can be implemented in order
-to quantify uncertainty about putative parent node(s) and switch positions.
-These estimates of uncertainty may be used as support values for branches
-analogous to bootstrap values produced by standard phylogenetic methods.
-\item
-Daily trees are built using the UPGMA algorithm. This entails an extra step to
-``un-resolve'' daily trees in parts of the trees with no phylogenetic resolution.
-To avoid this step, a better approach is to use a maximum-parsimony algorithm
-to build daily trees instead.
+\item A potential source of error in sc2ts ARGs comes from incorrectly reported collection dates of the samples, which can cause sample genomes to be attached to the wrong parts of the ARGs.
+Thus far, we have attempted to correct the most offending errors in the ARGs by filtering out "time travellers", but the sampling time can still be wrong by days or tens of days.
+***describe possible approaches to correct this?***
+\item To avoid time travellers, we used a maximum submission delay, but this will systematically bias against submissions from countries with greater submission delays (Kalia et al.).
+An area of future work is to add back the excluded sequences to a built ARG to assess which ones are likely true time travellers.
+\item We have used a single mismatch value in all the HMM runs, set so that having three mutations is as likely as having a single recombination.
+This parameter choice is arbitrary, and higher values may reduce the number of origins for those Pango lineages that were inferred to descend from more than one recombinant ancestor.
+In addition, because both the mutation rate and recombination rate varies across the SARS-CoV-2 genome, it may be helpful to allow the mismatch value to vary along the genome.
+\item By treating mutations equally in the current HMM implementation, we implicitly assume that the rates of different types of mutations are identical.
+However, as reported previously \citep{Yi2021-sc}, there is a high degree of mutational asymmetry in SARS-CoV-2.
+To account for mutational biases, more sophisticated nucleotide substitution models (e.g., allowing for different rates of transition and transversion) can be incorporated into HMM likelihood calculations.
+\item Our current masking strategy handles indels as missing data.
+Doing this can remove information helpful to detect recombinants.
+For example, sequences of the Pango recombinant XP were not identified as recombinants, because XP’s characteristic multibase deletion was masked.
+\item The "lack of memory" in the HMM can result in  recombinants with two or more breakpoints to have different parents for each recombined portion of the genome, when it is sometimes equally parsimonious to match against the same parent in separate regions.
+The copying process in the HMM currently does not preferentially "switch back" to a previoulsy matched parent, but incorporating this would be a valuable addition.
+\item Some recombinants may reasonably have three or more parents, indicating multiple "stacked" recombination events.
+We have not investigated this in the current analysis, but with further sc2ts development, there is potential to analyse these putative recombinants and trace their evolutionary origins through the ARG.
+\item In the present implementation, HMM traceback is performed once, yielding a single maximum-likelihood genome copying path.
+A stochastic traceback algorithm (e.g., (Rasmussen et al.
+2014)) can be implemented in order to quantify uncertainty about putative parent node(s) and switch positions.
+These estimates of uncertainty may be used as support values for branches analogous to bootstrap values produced by standard phylogenetic methods.
+\item Daily trees are built using the UPGMA algorithm.
+This entails an extra step to ``un-resolve'' daily trees in parts of the trees with no phylogenetic resolution.
+Using a maximum-parsimony algorithm to build daily trees would leave these "soft polytomies" in, removing the need for the un-resolving step.
 \end{itemize}
@@
 \section{Conclusions}
-
-The global understanding of viral evolution during the COVID-19 pandemic is
-unprecedented, with over 15 million sequences shared to date. We envisage that
-once fully developed and tested, sc2ts methodology and software may provide a
-unified computational framework to study the evolution and epidemiology of
-SARS-CoV-2 at pandemic scale. The evolutionary analyses presented herein
-exemplify the numerous applications possible using the ancestral recombination
-graph inferred by sc2ts. We invite bioinformaticians, evolutionary biologists,
-and virologists to join us and the broader tskit community to build the most
-complete representation of the evolutionary history of SARS-CoV-2. Moreover,
-while we have focussed on the particular application to SARS-CoV-2, the methods
-we describe are readily applicable to the analysis of many other viral
-pathogens. Solving the challenges in building and analysing large viral
-genealogies would therefore contribute to the important goal of improving
-preparedness for potential future pandemics.
+The global understanding of viral evolution during the COVID-19 pandemic is unprecedented, with over 15 million sequences shared to date.
+We envisage that once fully developed and tested, sc2ts methodology and software may provide a unified computational framework to study the evolution and epidemiology of SARS-CoV-2 at pandemic scale.
+The evolutionary analyses presented herein exemplify the numerous applications possible using the ancestral recombination graph inferred by sc2ts.
+We invite bioinformaticians, evolutionary biologists, and virologists to join us and the broader tskit community to build the most complete representation of the evolutionary history of SARS-CoV-2.
+Moreover, while we have focussed on the particular application to SARS-CoV-2, the methods we describe are readily applicable to the analysis of many other viral pathogens.
+Solving the challenges in building and analysing large viral genealogies would therefore contribute to the important goal of improving preparedness for potential future pandemics.
 
 \section{Acknowledgements}
-
-We gratefully acknowledge all data contributors, i.e., the Authors and their
-Originating laboratories responsible for obtaining the specimens, and their
-Submitting laboratories for generating the genetic sequence and metadata and
-sharing via the GISAID Initiative, on which this research is based. Also, we
-thank Dr. Morag Graham (the National Microbiology Laboratory, Winnipeg, Canada)
-for kindly providing the genomic coordinates of the breakpoint sequence motifs.
-SHZ is supported by the Janssen-Oxford Translational Genomics Fellowship. JK,
-YW, and BJ are supported by the Robertson Foundation.
+We gratefully acknowledge all data contributors, i.e., the Authors and their Originating laboratories responsible for obtaining the specimens, and their Submitting laboratories for generating the genetic sequence and metadata and sharing via the GISAID Initiative, on which this research is based.
+Also, we thank Dr.
+Morag Graham (the National Microbiology Laboratory, Winnipeg, Canada) for kindly providing the genomic coordinates of the breakpoint sequence motifs.
+SHZ is supported by the Janssen-Oxford Translational Genomics Fellowship.
+JK, YW, and BJ are supported by the Robertson Foundation.
 
@@
 \includegraphics[width=\textwidth]{figures/supp_cophylogeny_long.pdf}
-\caption{\label{fig:cophylogeny_long}Tanglegram equivalent to that in
-Figure~\ref{fig:cophylogeny}, but for the long ARG (i.e. subsampled to mid-2022). }
+\caption{\label{fig:cophylogeny_long}Tanglegram equivalent to that in Figure~\ref{fig:cophylogeny}, but for the long ARG (i.e. subsampled to mid-2022). }
 \end{figure}
@@
 \end{tabular}
-\caption{\label{tab:jackson_supplement}Recombinant sequences involving the Alpha (B.1.1.7)
-variant reported by \cite{Jackson2021-ik} have recombinant
-ancestry in the wide ARG. The breakpoint intervals and Pango lineage
-assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et
-al., except the Pango lineage assignment of the parents of group B
-recombinants, which were taken from Table 1 (motif-based results).}
+\caption{\label{tab:jackson_supplement}Recombinant sequences involving the Alpha (B.1.1.7) variant reported by \cite{Jackson2021-ik} have recombinant ancestry in the wide ARG.
+The breakpoint intervals and Pango lineage assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et al., except the Pango lineage assignment of the parents of group B recombinants, which were taken from Table 1 (motif-based results).}
 \end{table}
@@
 \includegraphics[width=\textwidth]{figures/supp_recombination_node_mrcas.pdf}
-\caption{\label{fig:recomb_mrcas_lineage_breakdown}  DESCRIBE ME}
+\caption{\label{fig:recomb_mrcas_voc_breakdown}  DESCRIBE ME}
+\end{figure}
+
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/Pango_XAG_gisaid_large_graph.pdf}
+\caption{\label{fig:pango_XAG_gisaid_graph}  DESCRIBE ME}
 \end{figure}

### 2023-04-21 paper.tex
Update initial Results section

@@
 \subsection{Inferred ARGs}
-To illustrate the use of sc2ts, we employed two different sampling strategies.
-First, we inferred a densely sampled but time-limited ``wide'' ARG by attaching all the sample genomes with an MSD of 30 days, one daily batch at a time.
-The inference procedure ran until the genomes sampled on June 30, 2021 were attached to the ARG, resulting in a total of 1.27 million samples.
-Second, we inferred a ``long'' ARG over a more extended timespan, by attaching at most 1,000 genomes, which were randomly subsampled from each daily batch with an MSD within 30 days.
-For this ARG, the inference procedure ran until the genomes sampled on June 30, 2022 were attached to the ARG.
-The long ARG covers an extra year of samples but is more sparsely sampled.
-June 30, 2022 is the collection date of the final batch of sequences which were preprocessed for the development of sc2ts.
-Once sc2ts is further developed and tested, we will attach samples collected after that date to infer a fully sampled pandemic-scale ARG.
+For this prelimary study our goals are to illustrate the utility of
+\texttt{sc2ts}, but to also closely study the properties of the
+inferred ARGs to inform subsequent development. Thus, rather
+than attempting to be comprehensive we chose to work with a
+representative subset of the available data, and limited to
+inferences that could be performed on a single server in
+a few weeks (see below for further details on timings and
+computer hardware used). The cut-off dates for sampling were chosen
+arbitrarily. We anticipate that further development of the
+inference methods and scaling out across multiple servers will
+enable inferences at the full pandemic scale using \texttt{sc2ts}
+(see Discussion).
+
+To investigate the properties of the inferred ARGs under different
+sampling conditions, we inferred two ARGs, which we refer to as the
+``Wide'' and ``Long'' ARGs throughout.
+The Wide ARG is densely sampled but time-limited and
+includes all 1.27 million sample genomes  with a
+maximum sampling delay of 30 days (see Methods) up to June 30, 2021.
+The Long ARG is more sparsely sampled but
+includes data for an additional year (June 30, 2022).
+For the long ARG, we use at most 1,000 randomly subsampled
+genomes per day from those meeting the maximum sampling delay criterion.
 
@@
 
-Most of the nodes in the ARGs represent sample genomes (Table 1).
-Only a small fraction of the nodes represent recombination events (0.28\% and 0.27\% in the wide ARG and long ARG, respectively), over half of which have no descendant samples (besides the recombinant sequence that triggered insertion of the recombination node) (Table 1).
-There are over one million mutations on the ARGs, the majority of which occur on a terminal edge (Table 1).
-Notably, the massive amount of genomic variation data in the wide ARG (1.27 million genomes) and long ARG (657 thousand genomes) are compactly stored in ~57 Mb and ~36 Mb, respectively (less than ten bytes per sample).
-These ARGs can be loaded for analysis in a Jupyter notebook on a typical commercial laptop in merely a fraction of a second using the Python package tskit (Table 1).
+The properties of the inferred ARGs are summarised in Table~\ref{tab:args}.
+% TODO more discussion here, making sure we explain anything that might not be
+% obvious.
+Most of the nodes in the ARGs represent sample genomes
+Only a small fraction of the nodes represent recombination events (0.28\% and
+0.27\% in the wide ARG and long ARG, respectively), over half of which have no
+descendant samples (besides the recombinant sequence that triggered insertion
+of the recombination node). There are over one million mutations on
+the ARGs, the majority of which occur on a terminal edge. Notably,
+the massive amount of genomic variation data in the wide ARG (1.27 million
+genomes) and long ARG (657 thousand genomes) are compactly stored in ~57 Mb and
+~36 Mb, respectively (less than ten bytes per sample). These ARGs can be loaded
+for analysis in a Jupyter notebook on a typical laptop in a
+fraction of a second using the Python package tskit.
 
-Inferring the wide ARG required 17 days of computation time on a machine with 128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz).
-Inferring the long ARG required 23 days on a (much older) machine with 40 threads and 256 GB RAM (2x Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz).
-The majority of the time is spent on running the HMM to find the copying path for each sequence, and the process is therefore highly amenable to distributing across multiple machines.
+Inferring the wide ARG required 17 days of computation time on a machine with
+128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). Inferring the long ARG
+required 23 days on a (much older) machine with 40 threads and 256 GB RAM (2x
+Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz). The majority of the time is spent
+on running the HMM to find the copying path for each sequence, and the process
+is therefore highly amenable to distributing across multiple machines.
 

### 2023-04-21 paper.tex
changes from Yan
Fix line break (again)

@@
 More recently, the high prevalence of Omicron, with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has accelerated the rate of coinfection and the potential for recombination \citep{Bal2022-hq}.
-Recombinant lineages that have sufficient evidence of onward transmission (more than five descendants) and monophyly
-($>70\%$ bootstrap support) are eligible for a Pango lineage designation~\citep{Rambaut2020-dw}, starting with the letter X
+Recombinant lineages that have sufficient evidence of onward transmission (more than five descendants) and monophyly ($>70\%$ bootstrap support) are eligible for a Pango lineage designation~\citep{Rambaut2020-dw}, starting with the letter X
 % FIXME: should this be a citation?
@@
 \subsection{Data preprocessing}
+
 \subsubsection{Sequence-level preprocessing}
-The findings of this study are based on sequences and metadata available on GISAID (\url{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
+The findings of this study are
+based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
+up to Aug 22, 2022 and accessible at
 % FIXME
-doi.org/10.55876/gis8.230329cd (Supplementary Table 1).
-Sequences were removed if they (1) had ambiguous collection dates; (2) were collected before January 01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
-Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank: MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag 2022-07-26T12:00:00Z).
-To reduce the impact of poor-quality sequences on ARG inference, sequences were excluded if they had a ``bad'' quality control status in any of the four Nextclade columns (``qc.missingData.status'', ``qc.mixedSites.status'', ``qc.frameShifts.status'' and `qc.stopCodons.status'').
-These excluded genomes are enriched for low-coverage, contaminated, or misassembled sequences.
-
-\subsubsection{Site-level preprocessing}
-Ambiguous nucleotide letters (i.e., not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
-Problematic bases in the alignments, which had two or more ambiguous letters or gaps within a distance of seven bases, were masked as missing data following the protocol in \url{https://github.com/roblanf/sarscov2phylo/}.
-Masked bases were ignored when attaching a sample genome to the ARG, but they were subsequently imputed.
-Further, 481 problematic sites flagged as prone to sequencing errors or highly homoplasic were excluded entirely
+doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
+if they (1) had ambiguous collection dates; (2) were collected before January
+01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
+Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
+MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
+2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
+inference, sequences were excluded if they had a ``bad'' quality control status
+in any of the four Nextclade columns (``qc.missingData.status'',
+``qc.mixedSites.status'', ``qc.frameShifts.status'' and
+`qc.stopCodons.status''). These excluded genomes are enriched for low-coverage,
+contaminated, or misassembled sequences.
+
+\subsubsection{Site-level preprocessing} Ambiguous nucleotide letters (i.e.,
+not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
+Problematic bases in the alignments, which had two or more ambiguous letters or
+gaps within a distance of seven bases, were masked as missing data following
+the protocol in \url{https://github.com/roblanf/sarscov2phylo/}. Masked bases
+were ignored when attaching a sample genome to the ARG, but they were
+subsequently imputed. Further, 481 problematic sites flagged as prone to
+sequencing errors or highly homoplasic were excluded entirely
 % Do we need this first URL?
-(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473}; \url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
+(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
+\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
 
 \subsubsection{Exclusion of ``time travellers''}
-Including samples with incorrectly reported collection dates (e.g., an Alpha sample purportedly collected in 2020 from the United States, before Alpha appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce artefacts that distorted the topology of ARGs inferred early on in the study.
+
+Including samples with incorrectly reported collection dates (e.g., an Alpha
+sample purportedly collected in 2020 from the United States, before Alpha
+appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
+artefacts that distorted the topology of ARGs inferred early on in the study.
 Hence, to exclude such potential ``time travellers'' we employed two filters.
-The first was to remove any sequence with a collection date that pre-dated the time to the most recent common ancestor (tMRCA) of its corresponding clade.
-The tMRCA for each clade was obtained from a Nextstrain GISAID reference tree (downloaded on August 22, 2022) where we used the lower bound of the 95\% confidence interval of each clade as the minimum date cut-off.
-The second filter was a simple threshold on the time delay between the collection date and submission date (maximum submission delay, or MSD).
-Indeed, we found that ARGs inferred from the MSD-filtered genome set had noticeably fewer branches with a high number of reversions.
-After some preliminary analysis, a MSD threshold value of 30 days was used when building the ARGs described herein.
-Future work could estimate collection dates by adding these potential time travellers back on the tree, allowing an automated assessment of collection date discrepancies.
+The first was to remove any sequence with a collection date that pre-dated the
+time to the most recent common ancestor (tMRCA) of its corresponding clade. The
+tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
+(downloaded on August 22, 2022) where we used the lower bound of the 95\%
+confidence interval of each clade as the minimum date cut-off. The second
+filter was a simple threshold on the time delay between the collection date and
+submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
+inferred from the MSD-filtered genome set had noticeably fewer branches with a
+high number of reversions. After some preliminary analysis, a MSD threshold
+value of 30 days was used when building the ARGs described herein. Future work
+could estimate collection dates by adding these potential time travellers back
+on the tree, allowing an automated assessment of collection date discrepancies.
 
@@
 \includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
-\caption{\label{fig:overview_sc2ts} A schematic of the sc2ts method.
-Sc2ts reconstructs the genetic relationships among SARS-CoV-2 genomes by copying samples to all possible ancestors collected at earlier time points (curved arrows).
-An ARG is built by attaching new sample genomes grouped by collection date to an existing ARG rooted at the Wuhan-Hu-1 reference genome, day by day.
-Each iteration involves three stages: (1) attachment of new samples to the growing ARG (A, B); (2) reconstruction of trees relating the samples under each attachment node (C); and parsimony-based tree topology adjustments (D, E).
-In the absence of  recombination, sc2ts infers an ARG that is a single tree relating the samples (A).
-However, in the presence of recombination, sc2ts infers an ARG that is encoded as a sequence of local trees relating segments of the sample genomes (B), inserting a recombination node to represent a putative recombination event.
-Additionally, mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to make more parsimonious placements of mutations that should be shared or should not be immediately reverted, respectively.}
+\caption{\label{fig:overview_sc2ts}
+A schematic of the sc2ts method. Sc2ts reconstructs the genetic relationships
+among SARS-CoV-2 genomes by copying samples to all possible ancestors collected
+at earlier time points (curved arrows). An ARG is built by attaching new sample
+genomes grouped by collection date to an existing ARG rooted at the Wuhan-Hu-1
+reference genome, day by day. Each iteration involves three stages: (1)
+attachment of new samples to the growing ARG (A, B); (2) reconstruction of
+trees relating the samples under each attachment node (C); and parsimony-based
+tree topology adjustments (D, E). In the absence of  recombination, sc2ts
+infers an ARG that is a single tree relating the samples (A). However, in the
+presence of recombination, sc2ts infers an ARG that is encoded as a sequence of
+local trees relating segments of the sample genomes (B), inserting a
+recombination node to represent a putative recombination event. Additionally,
+mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to
+make more parsimonious placements of mutations that should be shared or should
+not be immediately reverted, respectively.}
 \end{figure}
@@
 % Note going to rewrite this anyway, so not trying to make it flow.
-It uses (1) an efficient HMM implementation of the Li and Stephens (LS) model \citep{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported collection dates of the SARS-CoV-2 samples to determine the order of genome copying; and (3) several parsimony-based heuristics to locally refine the topology.
-In brief, this combined approach uses the likelihood model of LS to attach sample genomes to existing nodes and to detect recombinants, but because many of the daily tip samples harbour few, if any, mutational changes, we use parsimony to resolve their close relationships.
-
-Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by taking advantage of the reported collection dates to incrementally grow the ARG.
-The inference process occurs forward in time, attaching new genomes aggregated by their collection date (daily batches), from the earliest to the last collection date.
-Prior to attaching sample genomes, the ARG is initialised such that the Wuhan-Hu-1/2019 reference genome is the root node.
-In the absence of recombination, the ARG is a single tree with no reticulation (Figure 1A); however, in the presence of recombination, a reticulate ARG is generated, with variation in tree topology across the genome (Figure 1B).
-Unlike standard phylogenetic trees in which internal nodes always represent inferred ancestors, internal nodes in the sc2ts ARGs represent sequences which can either be inferred ancestors or known samples.
-
-Each daily batch of genomes is first attached to the growing ARG without any phylogenetic resolution.
-In cases where the newly added sequence is inferred to be a recombinant, a recombination node is inserted; the ``date'' of this node is arbitrarily set to the average of the collection dates of the genome and the most recent of its parent nodes.
-Otherwise, the genomes are attached to the most closely related ancestral node (an attachment node).
-This creates polytomies that are then locally resolved among the descendants of each attachment node using the Unweighted Pair Group Method with Arithmetic Mean (UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
-This hierarchical clustering algorithm produces a dendrogram relating a set of sequences, which is treated as an ultrametric binary tree here.
-Subsequently, mutations are placed onto the daily tree by maximum parsimony.
-An issue with the UPGMA algorithm is that it may create internal nodes supported by no informative site (i.e., having no mutation immediately ancestral to them).
-In an adjustment step, such unsupported internal nodes are removed to ``un-resolve'' the daily tree; the remaining internal nodes, which are supported by at least one shared mutation, are retained.
-The ``dates'' of the non-sample internal nodes are set to the same date as the samples in the daily batch.
-The resulting daily tree is grafted onto the attachment node in the global ARG.
-
-During ARG inference, two types of non-sample nodes are inserted to improve parsimonious placement of mutations: mutation-coalescing nodes and pushed-reversion nodes.
-We call these operations tree adjustments.
-A mutation-collapsing node is inserted when a shared mutation between sibling nodes is found (Figure 1D).
-An attempt to create mutation-collapsing nodes is made after grafting a new daily tree onto the global ARG.
-If shared mutations exist between two or more sibling nodes, then a parsimonious explanation is to create a new node that is the parent of those siblings, on which these mutations occurred.
-A pushed-reversion node is inserted when an immediate reversion occurs (Figure 1E), i.e.
-a mutation that reverts another mutation on its parent node.
-Instead of retaining both the mutation and immediate reversion, only the mutation is kept after a pushed-reversion node is inserted, thereby reducing the number of mutations on the branches.
-Note that these non-sample nodes are used as reconstructed genomes during HMM copying.
-
-Furthermore, during inference, the ARG is automatically encoded, stored, and updated as a tree sequence using tskit \citep{Kelleher2018-xc}.
-
-\subsubsection{Brief description of the Li and Stephens model}
-The LS model is a convenient probabilistic approximation to the coalescent with recombination \citep{Hudson1983-if}, itself a generative model of a genealogical process with recombination \citep{Li2003-ib}.
-Importantly, it can be implemented as an HMM in which a focal genome is a sequence of nucleotides (observed states) that are probabilistically emitted as an imperfect mosaic of a set of reference genomes (hidden states).
-As such, an efficient HMM machinery can be employed to make statements about parameters of interest and highlight potential sources of error in genotyping and sequencing data.
-The LS model is used in a wide variety of applications in genomics.
-For example, it is the engine that drives statistical genotype phasing and imputation \citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it can be used to estimate mutation rates (e.g., REF), recombination rates \citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral sequence data \citep[e.g.][]{Palmer2019-wa}.
+It uses (1) an efficient HMM implementation of the Li and Stephens (LS) model
+\citep{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2)
+the reported collection dates of the SARS-CoV-2 samples to determine the order
+of genome copying; and (3) several parsimony-based heuristics to locally refine
+the topology. In brief, this combined approach uses the likelihood model of LS
+to attach sample genomes to existing nodes and to detect recombinants, but
+because many of the daily tip samples harbour few, if any, mutational changes,
+we use parsimony to resolve their close relationships.
+
+Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by
+taking advantage of the reported collection dates to incrementally grow the
+ARG. The inference process occurs forward in time, attaching new genomes
+aggregated by their collection date (daily batches), from the earliest to the
+last collection date. Prior to attaching sample genomes, the ARG is initialised
+such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence
+of recombination, the ARG is a single tree with no reticulation (Figure 1A);
+however, in the presence of recombination, a reticulate ARG is generated, with
+variation in tree topology across the genome (Figure 1B). Unlike standard
+phylogenetic trees in which internal nodes always represent inferred ancestors,
+internal nodes in the sc2ts ARGs represent sequences which can either be
+inferred ancestors or known samples.
+
+Each daily batch of genomes is first attached to the growing ARG without any
+phylogenetic resolution. In cases where the newly added sequence is inferred to
+be a recombinant, a recombination node is inserted; the ``date'' of this node
+is arbitrarily set to the average of the collection dates of the genome and the
+most recent of its parent nodes. Otherwise, the genomes are attached to the
+most closely related ancestral node (an attachment node). This creates
+polytomies that are then locally resolved among the descendants of each
+attachment node using the Unweighted Pair Group Method with Arithmetic Mean
+(UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
+This hierarchical clustering algorithm produces a dendrogram relating a set of
+sequences, which is treated as an ultrametric binary tree here. Subsequently,
+mutations are placed onto the daily tree by maximum parsimony. An issue with
+the UPGMA algorithm is that it may create internal nodes supported by no
+informative site (i.e., having no mutation immediately ancestral to them). In
+an adjustment step, such unsupported internal nodes are removed to
+``un-resolve'' the daily tree; the remaining internal nodes, which are
+supported by at least one shared mutation, are retained. The ``dates'' of the
+non-sample internal nodes are set to the same date as the samples in the daily
+batch. The resulting daily tree is grafted onto the attachment node in the
+global ARG.
+
+During ARG inference, two types of non-sample nodes are inserted to improve
+parsimonious placement of mutations: mutation-coalescing nodes and
+pushed-reversion nodes. We call these operations tree adjustments. A
+mutation-collapsing node is inserted when a shared mutation between sibling
+nodes is found (Figure 1D). An attempt to create mutation-collapsing nodes is
+made after grafting a new daily tree onto the global ARG. If shared mutations
+exist between two or more sibling nodes, then a parsimonious explanation is to
+create a new node that is the parent of those siblings, on which these
+mutations occurred. A pushed-reversion node is inserted when an immediate
+reversion occurs (Figure 1E), i.e. a mutation that reverts another mutation on
+its parent node. Instead of retaining both the mutation and immediate
+reversion, only the mutation is kept after a pushed-reversion node is inserted,
+thereby reducing the number of mutations on the branches. Note that these
+non-sample nodes are used as reconstructed genomes during HMM copying.
+
+Furthermore, during inference, the ARG is automatically encoded, stored, and
+updated as a tree sequence using tskit \citep{Kelleher2018-xc}.
+
+\subsubsection{Brief description of the Li and Stephens model} The LS model is
+a convenient probabilistic approximation to the coalescent with recombination
+\citep{Hudson1983-if}, itself a generative model of a genealogical process with
+recombination \citep{Li2003-ib}. Importantly, it can be implemented as an HMM
+in which a focal genome is a sequence of nucleotides (observed states) that are
+probabilistically emitted as an imperfect mosaic of a set of reference genomes
+(hidden states). As such, an efficient HMM machinery can be employed to make
+statements about parameters of interest and highlight potential sources of
+error in genotyping and sequencing data. The LS model is used in a wide variety
+of applications in genomics. For example, it is the engine that drives
+statistical genotype phasing and imputation
+\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and
+it can be used to estimate mutation rates (e.g., REF), recombination rates
+\citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral
+sequence data \citep[e.g.][]{Palmer2019-wa}.
 
@@
 \caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS) model.
-The LS model in its implementation as an HMM is governed by the transition and emission matrices, $Q$ and $E$, respectively.
-Each may be a function of the reference panel members, but transitions are generally assumed to be independent of the hidden states themselves (shown in the pink panel).
-This dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
-The recombination probability at site $\ell$ is $r\ell$, while n denotes the number of members in the reference panel.
-Emissions may also be a function of the nucleotide states, but in our case, we assume that all observed alleles (in both the reference panel and focal sequence, $a\ell$) at the site under consideration are equally likely and occur with mutation probability $\ell$ (shown in the blue panel).
-These two processes, transition and emission (which encode recombination and mutation, respectively), define the generative process of the HMM, by which an imperfect mosaic is emitted from a reference panel of samples and nodes from the ARG inserted at earlier collection dates (shown in the lower panel).
-Here, the unseen states in the reference panel are shown as coloured lines enclosed by the grey box, and the true path through the data which leads to the emitted focal sequence is displayed by the black arrow.
-Examples of transition and emission probabilities along this trajectory are shown by the red and blue arrows, respectively.
-Black crosses along the focal sequence show genotyping errors or mutations.
-Note that in practice, we do not know the pattern of colours along the focal sequence and must infer it using the Viterbi algorithm.} \end{figure}
-
-The generative process of the LS model is summarised in Figure~\ref{fig:ls_diagram}.
-Here, a transition matrix, $Q$, governs the process of switching (recombining) between members of the reference panel (the hidden states).
-An emission matrix, $E$, allows for differences between the focal sequence and the hidden state from which it is copied (due to mutation as well as genotyping or sequencing error).
-Then, we use the Viterbi algorithm \citep{Viterbi1967-ol} to find the most likely copying path through the sequences, given a set of reference sequences, $Q$, and $E$.
-
-Because the HMM is memoryless, the potential parental lineages are independently assessed at each recombination breakpoint.
-This can result in the inference of a recombinant lineage descending immediately from three (or even more) parental lineages.
-We find that often two of the three parents are identical or similar in sequence in the segments contributed to the recombinant lineage.
-A more parsimonious explanation would be to switch back to the original parent sequence.
-In the future, a pragmatic solution would be to detect such cases during the ARG inference stage and to adjust the copying paths accordingly (see ``Improvements'', below).
-Meanwhile, however, we urge caution when interpreting the parentage of lineages with multiple breakpoints.
-
-\subsubsection{Parameterization of the HMM implementation of Li and Stephens model}
-Recombination is inferred using an efficient HMM implementation of the LS model.
-Copying paths are estimated from SARS-CoV-2 genomes sampled over time, with younger genomes copying (i.e.
-inheriting segments and mutations) from older genomes.
-Here, we were able to apply the LS model on millions of SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of the LS model that is a core component of the tsinfer software.
-For this application, we have parameterised the HMM using a mismatch ratio (MMR), which tunes how strongly mutations (i.e., no switching of template sequences) are preferred over recombination (i.e., switching between sequences) during copying.
-Under a lower MMR value, the maximum-likelihood HMM copying path of a focal genome more likely spans at least two ancestral genomes (i.e,.
-the focal genome is a putative recombinant) than a single ancestral genome.
-Here, the MMR was set to three in all the HMM runs (see ``Improvements'' for future potential development of the HMM).
-
-\subsection{Analysis}
-\subsubsection{Imputation of Pango lineage for non-sample nodes}
-Non-sample nodes inserted during ARG inference had no Pango lineage assignment from Nextclade.
-We imputed the Pango lineage of non-sample nodes using a two-step procedure (Figure 3) to help assess how consistently breakpoint intervals are estimated by the HMM runs (see below).
-If the Pango lineage of the parental node was available and no mutation was immediately ancestral to the non-sample node, then the parental node’s status was directly propagated to the non-sample node.
-Otherwise, the Pango lineage of the non-sample node was inferred by matching its full set of mutations against the Pango lineage-defining mutations (based on 90\% consensus of the sequences analysed) from the COVID-CG website \citep{Chen2021-zc}; \url{https://covidcg.org/}; accessed on November 04, 2022).
-
-\subsubsection{Breakpoint intervals and HMM consistency}
- When recombinant ancestry for a given genome is inferred by the LS HMM, a breakpoint is identified as the last possible position where the copying path must switch.
-Thus, the breakpoints used in the ARG represent the rightmost end of the interval, within which the recombination event could have occurred.
-To obtain the leftmost end of this interval, we rerun the HMM using mirrored coordinates (i.e., from the 3’ end to the 5’ end).
-
-A recombinant sequence was deemed to be HMM-consistent if (1) the same number of parent nodes and (2) the same set of mutations immediately ancestral to it were obtained from both the unmirrored and mirrored HMM runs.
-In the ideal case, the copying paths of both the HMM runs switch between parent sequences at the same position(s); in less ideal cases, however, the copying paths switch at different positions, typically because of a lack of genetic variation between the positions.
-The switch positions in the copying paths of both the HMM runs were used to define the breakpoint interval(s) of an HMM-consistent recombinant.
-
-\subsubsection{Code availability}
- The source code implementing the sc2ts pipeline is available via GitHub \url{https://github.com/jeromekelleher/sc2ts/}.
-The sc2ts pipeline uses various functions and algorithms implemented in tskit and tsinfer \url{https://github.com/tskit-dev/}.
-The pipeline uses the SciPy implementation of the UPGMA algorithm to build daily trees \url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}.
-The Jupyter notebooks used to produce the results described herein are also available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.
+The LS model in its implementation as an HMM is governed by the transition and
+emission matrices, $Q$ and $E$, respectively. Each may be a function of the
+reference panel members, but transitions are generally assumed to be
+independent of the hidden states themselves (shown in the pink panel). This
+dramatically increases performance as the state space drops to two states
+(i.e., switching or not switching). The recombination probability at site
+$\ell$ is $r\ell$, while n denotes the number of members in the reference
+panel. Emissions may also be a function of the nucleotide states, but in our
+case, we assume that all observed alleles (in both the reference panel and
+focal sequence, $a\ell$) at the site under consideration are equally likely and
+occur with mutation probability $\ell$ (shown in the blue panel). These two
+processes, transition and emission (which encode recombination and mutation,
+respectively), define the generative process of the HMM, by which an imperfect
+mosaic is emitted from a reference panel of samples and nodes from the ARG
+inserted at earlier collection dates (shown in the lower panel). Here, the
+unseen states in the reference panel are shown as coloured lines enclosed by
+the grey box, and the true path through the data which leads to the emitted
+focal sequence is displayed by the black arrow. Examples of transition and
+emission probabilities along this trajectory are shown by the red and blue
+arrows, respectively. Black crosses along the focal sequence show genotyping
+errors or mutations. Note that in practice, we do not know the pattern of
+colours along the focal sequence and must infer it using the Viterbi
+algorithm.} \end{figure}
+
+The generative process of the LS model is summarised in
+Figure~\ref{fig:ls_diagram}. Here, a transition matrix, $Q$, governs the
+process of switching (recombining) between members of the reference panel (the
+hidden states). An emission matrix, $E$, allows for differences between the
+focal sequence and the hidden state from which it is copied (due to mutation as
+well as genotyping or sequencing error). Then, we use the Viterbi algorithm
+\citep{Viterbi1967-ol} to find the most likely copying path through the
+sequences, given a set of reference sequences, $Q$, and $E$.
+
+Because the HMM is memoryless, the potential parental lineages are
+independently assessed at each recombination breakpoint. This can result in the
+inference of a recombinant lineage descending immediately from three (or even
+more) parental lineages. We find that often two of the three parents are
+identical or similar in sequence in the segments contributed to the recombinant
+lineage. A more parsimonious explanation would be to switch back to the
+original parent sequence. In the future, a pragmatic solution would be to
+detect such cases during the ARG inference stage and to adjust the copying
+paths accordingly (see ``Improvements'', below). Meanwhile, however, we urge
+caution when interpreting the parentage of lineages with multiple breakpoints.
+
+\subsubsection{Parameterization of the HMM implementation of Li and Stephens
+model} Recombination is inferred using an efficient HMM implementation of the
+LS model. Copying paths are estimated from SARS-CoV-2 genomes sampled over
+time, with younger genomes copying (i.e. inheriting segments and mutations)
+from older genomes. Here, we were able to apply the LS model on millions of
+SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of
+the LS model that is a core component of the tsinfer software. For this
+application, we have parameterised the HMM using a mismatch ratio (MMR), which
+tunes how strongly mutations (i.e., no switching of template sequences) are
+preferred over recombination (i.e., switching between sequences) during
+copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
+focal genome more likely spans at least two ancestral genomes (i.e,. the focal
+genome is a putative recombinant) than a single ancestral genome. Here, the MMR
+was set to three in all the HMM runs (see ``Improvements'' for future potential
+development of the HMM).
+
+\subsection{Analysis} \subsubsection{Imputation of Pango lineage for non-sample
+nodes} Non-sample nodes inserted during ARG inference had no Pango lineage
+assignment from Nextclade. We imputed the Pango lineage of non-sample nodes
+using a two-step procedure (Figure 3) to help assess how consistently
+breakpoint intervals are estimated by the HMM runs (see below). If the Pango
+lineage of the parental node was available and no mutation was immediately
+ancestral to the non-sample node, then the parental node’s status was directly
+propagated to the non-sample node. Otherwise, the Pango lineage of the
+non-sample node was inferred by matching its full set of mutations against the
+Pango lineage-defining mutations (based on 90\% consensus of the sequences
+analysed) from the COVID-CG website \citep{Chen2021-zc};
+\url{https://covidcg.org/}; accessed on November 04, 2022).
+
+\subsubsection{Breakpoint intervals and HMM consistency} When recombinant
+ancestry for a given genome is inferred by the LS HMM, a breakpoint is
+identified as the last possible position where the copying path must switch.
+Thus, the breakpoints used in the ARG represent the rightmost end of the
+interval, within which the recombination event could have occurred. To obtain
+the leftmost end of this interval, we rerun the HMM using mirrored coordinates
+(i.e., from the 3’ end to the 5’ end).
+
+A recombinant sequence was deemed to be HMM-consistent if (1) the same number
+of parent nodes and (2) the same set of mutations immediately ancestral to it
+were obtained from both the unmirrored and mirrored HMM runs. In the ideal
+case, the copying paths of both the HMM runs switch between parent sequences at
+the same position(s); in less ideal cases, however, the copying paths switch at
+different positions, typically because of a lack of genetic variation between
+the positions. The switch positions in the copying paths of both the HMM runs
+were used to define the breakpoint interval(s) of an HMM-consistent
+recombinant.
+
+\subsubsection{Code availability} The source code implementing the sc2ts
+pipeline is available via GitHub
+\url{https://github.com/jeromekelleher/sc2ts/}. The sc2ts pipeline uses various
+functions and algorithms implemented in tskit and tsinfer
+\url{https://github.com/tskit-dev/}. The pipeline uses the SciPy implementation
+of the UPGMA algorithm to build daily trees
+\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}.
+The Jupyter notebooks used to produce the results described herein are also
+available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.
 

### 2023-04-21 paper.tex
Move code availability down and add data availability statement

@@
 recombinant.
-
-\subsubsection{Code availability} The source code implementing the sc2ts
-pipeline is available via GitHub
-\url{https://github.com/jeromekelleher/sc2ts/}. The sc2ts pipeline uses various
-functions and algorithms implemented in tskit and tsinfer
-\url{https://github.com/tskit-dev/}. The pipeline uses the SciPy implementation
-of the UPGMA algorithm to build daily trees
-\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}.
-The Jupyter notebooks used to produce the results described herein are also
-available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.
 
@@
 \section{Acknowledgements}
-We gratefully acknowledge all data contributors, i.e., the Authors and their Originating laboratories responsible for obtaining the specimens, and their Submitting laboratories for generating the genetic sequence and metadata and sharing via the GISAID Initiative, on which this research is based.
-Also, we thank Dr.
-Morag Graham (the National Microbiology Laboratory, Winnipeg, Canada) for kindly providing the genomic coordinates of the breakpoint sequence motifs.
-SHZ is supported by the Janssen-Oxford Translational Genomics Fellowship.
-JK, YW, and BJ are supported by the Robertson Foundation.
+We gratefully acknowledge all data contributors, i.e., the Authors and their
+Originating laboratories responsible for obtaining the specimens, and their
+Submitting laboratories for generating the genetic sequence and metadata and
+sharing via the GISAID Initiative, on which this research is based. Also, we
+thank Dr. Morag Graham (the National Microbiology Laboratory, Winnipeg, Canada)
+for kindly providing the genomic coordinates of the breakpoint sequence motifs.
+SHZ is supported by the Janssen-Oxford Translational Genomics Fellowship. JK,
+YW, and BJ are supported by the Robertson Foundation.
+
+\section{Data Availability}
+The source code for \texttt{sc2ts} is
+is available on GitHub
+\url{https://github.com/jeromekelleher/sc2ts/}.
+The Jupyter notebooks and code used to produce the results described here are also
+available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.
+
+% TODO what's the right way to say this?
+The inferred ARGs described here are available on request to those with
+the appropriate GISAID data access.
 

### 2023-04-21 paper.tex
More reshuffling

@@
 
-\subsection{Data preprocessing}
-
-\subsubsection{Sequence-level preprocessing}
-The findings of this study are
-based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
-up to Aug 22, 2022 and accessible at
-% FIXME
-doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
-if they (1) had ambiguous collection dates; (2) were collected before January
-01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
-Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
-MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
-2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
-inference, sequences were excluded if they had a ``bad'' quality control status
-in any of the four Nextclade columns (``qc.missingData.status'',
-``qc.mixedSites.status'', ``qc.frameShifts.status'' and
-`qc.stopCodons.status''). These excluded genomes are enriched for low-coverage,
-contaminated, or misassembled sequences.
-
-\subsubsection{Site-level preprocessing} Ambiguous nucleotide letters (i.e.,
-not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
-Problematic bases in the alignments, which had two or more ambiguous letters or
-gaps within a distance of seven bases, were masked as missing data following
-the protocol in \url{https://github.com/roblanf/sarscov2phylo/}. Masked bases
-were ignored when attaching a sample genome to the ARG, but they were
-subsequently imputed. Further, 481 problematic sites flagged as prone to
-sequencing errors or highly homoplasic were excluded entirely
-% Do we need this first URL?
-(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
-\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
-
-\subsubsection{Exclusion of ``time travellers''}
-
-Including samples with incorrectly reported collection dates (e.g., an Alpha
-sample purportedly collected in 2020 from the United States, before Alpha
-appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
-artefacts that distorted the topology of ARGs inferred early on in the study.
-Hence, to exclude such potential ``time travellers'' we employed two filters.
-The first was to remove any sequence with a collection date that pre-dated the
-time to the most recent common ancestor (tMRCA) of its corresponding clade. The
-tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
-(downloaded on August 22, 2022) where we used the lower bound of the 95\%
-confidence interval of each clade as the minimum date cut-off. The second
-filter was a simple threshold on the time delay between the collection date and
-submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
-inferred from the MSD-filtered genome set had noticeably fewer branches with a
-high number of reversions. After some preliminary analysis, a MSD threshold
-value of 30 days was used when building the ARGs described herein. Future work
-could estimate collection dates by adding these potential time travellers back
-on the tree, allowing an automated assessment of collection date discrepancies.
-
-\subsection{Inference}
-\subsubsection{Overview of sc2ts}
+\subsection{Overview}
+Sc2ts is a method for inferring ARGs from densely sampled pandemic-scale data
+in real time, in which recombination occurs at a low but significant rate.
+The essential idea is to incrementally update the ARG each day
+with the sequences collected on that day. These updates are performed by
+first finding likely ``copying paths'' for each sequence in the daily
+batch to the current ARG, using the ARG-based implementation of
+the Li and Stephens model~\citep{Li2003-ib} introduced as
+part of \texttt{tsinfer}~\citep{Kelleher2019-ba}. These copying paths
+will mostly consist of a new sequence copying from a node in the ARG
+with a small number of mutations, and many sequences
+from a daily batch often copy from the same ARG node. The second
+step is then to ``resolve'' these implied polytomies by using
+standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
+Such local (greedy) ARG updates inevitably lead to unparsimonious
+topologies, and the third step is then to attempt to increase the
+overall parsimony of the inference by making some simple topological
+updates (Figure~\ref{fig:overview_sc2ts}C and D).
+
 \begin{figure} \centering
@@
 
-We introduce an ARG inference method, summarised in
-Figure~\ref{fig:overview_sc2ts}.
-% Note going to rewrite this anyway, so not trying to make it flow.
-It uses (1) an efficient HMM implementation of the Li and Stephens (LS) model
-\citep{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2)
-the reported collection dates of the SARS-CoV-2 samples to determine the order
-of genome copying; and (3) several parsimony-based heuristics to locally refine
-the topology. In brief, this combined approach uses the likelihood model of LS
-to attach sample genomes to existing nodes and to detect recombinants, but
-because many of the daily tip samples harbour few, if any, mutational changes,
-we use parsimony to resolve their close relationships.
-
-Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by
-taking advantage of the reported collection dates to incrementally grow the
-ARG. The inference process occurs forward in time, attaching new genomes
-aggregated by their collection date (daily batches), from the earliest to the
-last collection date. Prior to attaching sample genomes, the ARG is initialised
-such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence
-of recombination, the ARG is a single tree with no reticulation (Figure 1A);
-however, in the presence of recombination, a reticulate ARG is generated, with
-variation in tree topology across the genome (Figure 1B). Unlike standard
-phylogenetic trees in which internal nodes always represent inferred ancestors,
-internal nodes in the sc2ts ARGs represent sequences which can either be
-inferred ancestors or known samples.
-
-Each daily batch of genomes is first attached to the growing ARG without any
-phylogenetic resolution. In cases where the newly added sequence is inferred to
-be a recombinant, a recombination node is inserted; the ``date'' of this node
-is arbitrarily set to the average of the collection dates of the genome and the
-most recent of its parent nodes. Otherwise, the genomes are attached to the
-most closely related ancestral node (an attachment node). This creates
-polytomies that are then locally resolved among the descendants of each
-attachment node using the Unweighted Pair Group Method with Arithmetic Mean
-(UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
-This hierarchical clustering algorithm produces a dendrogram relating a set of
-sequences, which is treated as an ultrametric binary tree here. Subsequently,
-mutations are placed onto the daily tree by maximum parsimony. An issue with
-the UPGMA algorithm is that it may create internal nodes supported by no
-informative site (i.e., having no mutation immediately ancestral to them). In
-an adjustment step, such unsupported internal nodes are removed to
-``un-resolve'' the daily tree; the remaining internal nodes, which are
-supported by at least one shared mutation, are retained. The ``dates'' of the
-non-sample internal nodes are set to the same date as the samples in the daily
-batch. The resulting daily tree is grafted onto the attachment node in the
-global ARG.
-
-During ARG inference, two types of non-sample nodes are inserted to improve
-parsimonious placement of mutations: mutation-coalescing nodes and
-pushed-reversion nodes. We call these operations tree adjustments. A
-mutation-collapsing node is inserted when a shared mutation between sibling
-nodes is found (Figure 1D). An attempt to create mutation-collapsing nodes is
-made after grafting a new daily tree onto the global ARG. If shared mutations
-exist between two or more sibling nodes, then a parsimonious explanation is to
-create a new node that is the parent of those siblings, on which these
-mutations occurred. A pushed-reversion node is inserted when an immediate
-reversion occurs (Figure 1E), i.e. a mutation that reverts another mutation on
-its parent node. Instead of retaining both the mutation and immediate
-reversion, only the mutation is kept after a pushed-reversion node is inserted,
-thereby reducing the number of mutations on the branches. Note that these
-non-sample nodes are used as reconstructed genomes during HMM copying.
-
-Furthermore, during inference, the ARG is automatically encoded, stored, and
-updated as a tree sequence using tskit \citep{Kelleher2018-xc}.
-
-\subsubsection{Brief description of the Li and Stephens model} The LS model is
+[Quick overview of LS and ARGs/tskit computation here]
+As used here, the Li and Stephens model essentially searches among
+all sequences in the ARG (both sampled and inferred ancestral nodes)
+
+The details of each of these steps are discussed in subsequent
+sections.
+
+% We introduce an ARG inference method, summarised in
+% Figure~\ref{fig:overview_sc2ts}.
+% % Note going to rewrite this anyway, so not trying to make it flow.
+% It uses (1) an efficient HMM implementation of the Li and Stephens (LS) model
+% \citep{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2)
+% the reported collection dates of the SARS-CoV-2 samples to determine the order
+% of genome copying; and (3) several parsimony-based heuristics to locally refine
+% the topology. In brief, this combined approach uses the likelihood model of LS
+% to attach sample genomes to existing nodes and to detect recombinants, but
+% because many of the daily tip samples harbour few, if any, mutational changes,
+% we use parsimony to resolve their close relationships.
+
+% Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by
+% taking advantage of the reported collection dates to incrementally grow the
+% ARG. The inference process occurs forward in time, attaching new genomes
+% aggregated by their collection date (daily batches), from the earliest to the
+% last collection date. Prior to attaching sample genomes, the ARG is initialised
+% such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence
+% of recombination, the ARG is a single tree with no reticulation (Figure 1A);
+% however, in the presence of recombination, a reticulate ARG is generated, with
+% variation in tree topology across the genome (Figure 1B). Unlike standard
+% phylogenetic trees in which internal nodes always represent inferred ancestors,
+% internal nodes in the sc2ts ARGs represent sequences which can either be
+% inferred ancestors or known samples.
+
+% Each daily batch of genomes is first attached to the growing ARG without any
+% phylogenetic resolution. In cases where the newly added sequence is inferred to
+% be a recombinant, a recombination node is inserted; the ``date'' of this node
+% is arbitrarily set to the average of the collection dates of the genome and the
+% most recent of its parent nodes. Otherwise, the genomes are attached to the
+% most closely related ancestral node (an attachment node). This creates
+% polytomies that are then locally resolved among the descendants of each
+% attachment node using the Unweighted Pair Group Method with Arithmetic Mean
+% (UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
+% This hierarchical clustering algorithm produces a dendrogram relating a set of
+% sequences, which is treated as an ultrametric binary tree here. Subsequently,
+% mutations are placed onto the daily tree by maximum parsimony. An issue with
+% the UPGMA algorithm is that it may create internal nodes supported by no
+% informative site (i.e., having no mutation immediately ancestral to them). In
+% an adjustment step, such unsupported internal nodes are removed to
+% ``un-resolve'' the daily tree; the remaining internal nodes, which are
+% supported by at least one shared mutation, are retained. The ``dates'' of the
+% non-sample internal nodes are set to the same date as the samples in the daily
+% batch. The resulting daily tree is grafted onto the attachment node in the
+% global ARG.
+
+% During ARG inference, two types of non-sample nodes are inserted to improve
+% parsimonious placement of mutations: mutation-coalescing nodes and
+% pushed-reversion nodes. We call these operations tree adjustments. A
+% mutation-collapsing node is inserted when a shared mutation between sibling
+% nodes is found (Figure 1D). An attempt to create mutation-collapsing nodes is
+% made after grafting a new daily tree onto the global ARG. If shared mutations
+% exist between two or more sibling nodes, then a parsimonious explanation is to
+% create a new node that is the parent of those siblings, on which these
+% mutations occurred. A pushed-reversion node is inserted when an immediate
+% reversion occurs (Figure 1E), i.e. a mutation that reverts another mutation on
+% its parent node. Instead of retaining both the mutation and immediate
+% reversion, only the mutation is kept after a pushed-reversion node is inserted,
+% thereby reducing the number of mutations on the branches. Note that these
+% non-sample nodes are used as reconstructed genomes during HMM copying.
+
+% Furthermore, during inference, the ARG is automatically encoded, stored, and
+% updated as a tree sequence using tskit \citep{Kelleher2018-xc}.
+
+\subsection{Ancestral Recombination Graphs}
+[Quick overview of ARGs and Tree Sequences/tskit. Refer to the ARG paper for
+more discussion and details.]
+
+\subsection{The Li and Stephens model}
+[Edit down, cutting the text in the fig caption significantly]
+
+The Li and Stephens (LS) model is
 a convenient probabilistic approximation to the coalescent with recombination
@@
 
-\subsubsection{Parameterization of the HMM implementation of Li and Stephens
-model} Recombination is inferred using an efficient HMM implementation of the
+Recombination is inferred using an efficient HMM implementation of the
 LS model. Copying paths are estimated from SARS-CoV-2 genomes sampled over
@@
 
-\subsection{Analysis} \subsubsection{Imputation of Pango lineage for non-sample
-nodes} Non-sample nodes inserted during ARG inference had no Pango lineage
+\subsection{Tree inference from clusters of samples}
+[Discuss the "local" tree inference, but be careful not to use the term]
+
+\subsection{Parsimony updates}
+[Discuss the parsimony tweaks we make to the trees. Get a better name
+for the subsection title]
+
+\subsection{Treatment of recombinants}
+[Discuss the special treatment we give to recombinants.]
+Include the discussion about mirrored coords from here :]
+
+% \subsubsection{Breakpoint intervals and HMM consistency} When recombinant
+% ancestry for a given genome is inferred by the LS HMM, a breakpoint is
+% identified as the last possible position where the copying path must switch.
+% Thus, the breakpoints used in the ARG represent the rightmost end of the
+% interval, within which the recombination event could have occurred. To obtain
+% the leftmost end of this interval, we rerun the HMM using mirrored coordinates
+% (i.e., from the 3’ end to the 5’ end).
+
+% A recombinant sequence was deemed to be HMM-consistent if (1) the same number
+% of parent nodes and (2) the same set of mutations immediately ancestral to it
+% were obtained from both the unmirrored and mirrored HMM runs. In the ideal
+% case, the copying paths of both the HMM runs switch between parent sequences at
+% the same position(s); in less ideal cases, however, the copying paths switch at
+% different positions, typically because of a lack of genetic variation between
+% the positions. The switch positions in the copying paths of both the HMM runs
+% were used to define the breakpoint interval(s) of an HMM-consistent
+% recombinant.
+
+\subsection{Data preprocessing}
+
+The findings of this study are
+based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
+up to Aug 22, 2022 and accessible at
+% FIXME
+doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
+if they (1) had ambiguous collection dates; (2) were collected before January
+01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
+Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
+MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
+2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
+inference, sequences were excluded if they had a ``bad'' quality control status
+in any of the four Nextclade columns (``qc.missingData.status'',
+``qc.mixedSites.status'', ``qc.frameShifts.status'' and
+`qc.stopCodons.status''). These excluded genomes are enriched for low-coverage,
+contaminated, or misassembled sequences.
+
+Ambiguous nucleotide letters (i.e.,
+not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
+Problematic bases in the alignments, which had two or more ambiguous letters or
+gaps within a distance of seven bases, were masked as missing data following
+the protocol in \url{https://github.com/roblanf/sarscov2phylo/}. Masked bases
+were ignored when attaching a sample genome to the ARG, but they were
+subsequently imputed. Further, 481 problematic sites flagged as prone to
+sequencing errors or highly homoplasic were excluded entirely
+% Do we need this first URL?
+(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
+\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
+
+\subsection{Exclusion of ``time travellers''}
+[FIXME I'm pretty sure we don't have two filters here, we just use
+maximum sampling delay. Also not sure the MSD acronym is worth it,
+we don't use it much.]
+
+Including samples with incorrectly reported collection dates (e.g., an Alpha
+sample purportedly collected in 2020 from the United States, before Alpha
+appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
+artefacts that distorted the topology of ARGs inferred early on in the study.
+Hence, to exclude such potential ``time travellers'' we employed two filters.
+The first was to remove any sequence with a collection date that pre-dated the
+time to the most recent common ancestor (tMRCA) of its corresponding clade. The
+tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
+(downloaded on August 22, 2022) where we used the lower bound of the 95\%
+confidence interval of each clade as the minimum date cut-off. The second
+filter was a simple threshold on the time delay between the collection date and
+submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
+inferred from the MSD-filtered genome set had noticeably fewer branches with a
+high number of reversions. After some preliminary analysis, a MSD threshold
+value of 30 days was used when building the ARGs described herein. Future work
+could estimate collection dates by adding these potential time travellers back
+on the tree, allowing an automated assessment of collection date discrepancies.
+
+\subsection{Imputation of Pango lineage for non-sample nodes}
+Non-sample nodes inserted during ARG inference had no Pango lineage
 assignment from Nextclade. We imputed the Pango lineage of non-sample nodes
@@
 \url{https://covidcg.org/}; accessed on November 04, 2022).
-
-\subsubsection{Breakpoint intervals and HMM consistency} When recombinant
-ancestry for a given genome is inferred by the LS HMM, a breakpoint is
-identified as the last possible position where the copying path must switch.
-Thus, the breakpoints used in the ARG represent the rightmost end of the
-interval, within which the recombination event could have occurred. To obtain
-the leftmost end of this interval, we rerun the HMM using mirrored coordinates
-(i.e., from the 3’ end to the 5’ end).
-
-A recombinant sequence was deemed to be HMM-consistent if (1) the same number
-of parent nodes and (2) the same set of mutations immediately ancestral to it
-were obtained from both the unmirrored and mirrored HMM runs. In the ideal
-case, the copying paths of both the HMM runs switch between parent sequences at
-the same position(s); in less ideal cases, however, the copying paths switch at
-different positions, typically because of a lack of genetic variation between
-the positions. The switch positions in the copying paths of both the HMM runs
-were used to define the breakpoint interval(s) of an HMM-consistent
-recombinant.
 

### 2023-04-24 paper.tex
Updates from Yan

@@
 \subsection{Imputation of Pango lineage for non-sample nodes}
-Non-sample nodes inserted during ARG inference had no Pango lineage
-assignment from Nextclade. We imputed the Pango lineage of non-sample nodes
-using a two-step procedure (Figure 3) to help assess how consistently
-breakpoint intervals are estimated by the HMM runs (see below). If the Pango
+Initially, only sample nodes have Pango lineage assignments.
+For analysis of recombinant parents, it is also helpful to assign
+Pango lineages to non-sample nodes. We do this
+using a two-step procedure (Figure 3). If the Pango
 lineage of the parental node was available and no mutation was immediately
@@
 \includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
-\caption{\label{fig:recomb_mrcas} Divergance between parents at recombination breakpoints.
-(A) All recombination nodes that are ancestral to least two samples.
-(B) Further filtered as described in the text.
-MORE DETAIL. Horizontal lines represent WHAT,
-Figure~\ref{fig:recomb_mrcas_voc_breakdown} shows the same data broken down by parental ``VoC'' classification.}
+\caption{\label{fig:recomb_mrcas} Date of common ancestry between the parents either side of recombination breakpoints, as a function of degree of parental divergence.
+MRCAs of parents inferred to have produced a Pango designated recombinant (XA, XB, etc.) are identified in orange (A) Data after removing singleton recombinants (XXX parent pairs).
+Many recombinations involve parental strains that descend from the same MRCA: dates for the four most frequently identified MRCA nodes are indicated by horizontal dotted lines.
+These frequent MRCA nodes tend to be associated with major waves and commonly have many immediate children (i.e.
+are large polytomies, likely to indicate undersampled outbreaks) (B) Further filtered by removing HMM inconsistent breakpoints (YYY parent pairs).
+Note that Figure~\ref{fig:recomb_mrcas_voc_breakdown} shows the same data broken down by parental VoC classification.}
 \end{figure}
@@
 \includegraphics[width=\textwidth]{figures/supp_recombination_node_mrcas.pdf}
-\caption{\label{fig:recomb_mrcas_voc_breakdown}  DESCRIBE ME}
+\caption{\label{fig:recomb_mrcas_voc_breakdown}  Divergence between parent lineages for recombination events within and among different VoC categories.
+HMM-consistent recombinants are shown in green, HMM-inconsistent in blue (including many of the named Delta-Omicron recombinants).
+We find 69 HMM consistent (77 inconsistent) Alpha-Alpha recombination breakpoints = 13\% (10\%) of the total; 84 HMM consistent (143 inconsistent) Delta-Delta recombination breakpoints = 18\% (18\%) of the total; 99 HMM consistent (155 inconsistent) Omicron-Omicron recombination breakpoints = 22\% (19\%) of the total; and 2 HMM consistent (20 inconsistent) Delta-Omicron recombination breakpoints = 2\% (< 1\%) of the total.
+These plots do not include recombination breakpoints involved lineages that were not classified into these VoC categories}
 \end{figure}

### 2023-04-24 paper.tex
Updates from Overleaf

@@
 Pango lineages to non-sample nodes. We do this
-using a two-step procedure (Figure 3). If the Pango
+using a two-step procedure (Figure~\ref{fig:imputation}). If the Pango
 lineage of the parental node was available and no mutation was immediately
@@
 \url{https://covidcg.org/}; accessed on November 04, 2022).
+
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/imputation.pdf}
+\caption{\label{fig:imputation}A schematic of the two-step procedure to impute Pango lineage for inserted, non-sample nodes.
+Here, three nodes (question marks) have unknown Pango lineage (A).
+Because the parent of node 1 is available (i.e., L2) and there is no mutation occurring on the edge connecting the two nodes, the Pango lineage of the parent is directly propagated to node 1 (B).
+However, if there is a mutation (red X’s), then the Pango lineage is inferred by comparison to the set of lineage-defining mutations from COVID-CG\citep{Chen2021-zc}, as in case (C).
+Finally, the inferred status is propagated to any non-mutant child nodes, as in case (D).
+}
+\end{figure}
 
@@
 
-\begin{figure}
-\centering
+\begin{figure} \centering
 \includegraphics[width=0.4\textwidth]{figures/mutational_spectra.pdf}
 \caption{\label{fig:mutational_spectra}
-Mutational spectrum of SARS-CoV-2 in the wide ARG (sampled to mid-2021) compared to \cite{Yi2021-sc}.
-Mutations are categorised by type (i.e., ancestral state $>$ derived state).
-The percentages of each mutation type from the wide ARG are represented by blue bars and the percentages from Yi et al.
-by orange bars, with the darker colours representing one direction (e.g., C$>$U) and the lighter colours the reverse (e.g., U$>$C).}
+Mutational spectrum of SARS-CoV-2 in
+the wide ARG (sampled to mid-2021) compared to \cite{Yi2021-sc}. Mutations are
+categorised by type (i.e., ancestral state $>$ derived state). The percentages
+of each mutation type from the wide ARG are represented by blue bars and the
+percentages from Yi et al. by orange bars, with the darker colours representing
+one direction (e.g., C$>$U) and the lighter colours the reverse (e.g., U$>$C).}
 \end{figure}

### 2023-04-25 paper.tex

@@
 
-Inferring evolutionary trees while accounting for both mutation and recombination is a long-standing computational challenge.
-This challenge is particularly acute for SARS-CoV-2, given the large amount of data that has been shared, with over 15 million sequences in the GISAID database \citep{Shu2017-hp}.
-It is well known that recombination distorts phylogenies \citep{Schierup2000-fg} and affects the results of downstream analysis, e.g., inference of selection~\citep{Anisimova2003-vr}.
-However, standard phylogenetic methods do not account for recombination \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}.
-Alternatively, genetic ancestries of recombining lineages can be represented by ancestral recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-if}, which capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.
+Inferring evolutionary trees while accounting for both mutation and
+recombination is a long-standing computational challenge. This challenge is
+particularly acute for SARS-CoV-2, given the large amount of data that has been
+shared, with over 15 million sequences in the GISAID database
+\citep{Shu2017-hp}. It is well known that recombination distorts phylogenies
+\citep{Schierup2000-fg} and affects the results of downstream analysis, e.g.,
+inference of selection~\citep{Anisimova2003-vr}. However, standard phylogenetic
+methods do not account for recombination
+\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Alternatively,
+genetic ancestries of recombining lineages can be represented by ancestral
+recombination graphs (ARGs)~\citep{Griffiths1998-ancestral,Wong2023-efficient},
+which
+capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.
 
@@
 batch to the current ARG, using the ARG-based implementation of
-the Li and Stephens model~\citep{Li2003-ib} introduced as
-part of \texttt{tsinfer}~\citep{Kelleher2019-ba}. These copying paths
-will mostly consist of a new sequence copying from a node in the ARG
-with a small number of mutations, and many sequences
-from a daily batch often copy from the same ARG node. The second
+the Li and Stephens model~\citep{Li2003-ib} introduced
+in \texttt{tsinfer}~\citep{Kelleher2019-ba}.
+These copying paths
+will mostly consist of a new sample sequence copying from a node in the ARG
+with a small number of mutations, and often many samples
+from a daily batch will copy from the same ARG node. The second
 step is then to ``resolve'' these implied polytomies by using
 standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
-Such local (greedy) ARG updates inevitably lead to unparsimonious
-topologies, and the third step is then to attempt to increase the
-overall parsimony of the inference by making some simple topological
+This greedy update strategy inevitably lead to unparsimonious
+topologies, and the third step is to then increase the
+overall parsimony of the inferred ARG by making some simple topological
 updates (Figure~\ref{fig:overview_sc2ts}C and D).
+The operations are carried out efficiently using the well-established
+\texttt{tskit} library~\citep{Kelleher2018-xc,Ralph2020-efficiently},
+and the resulting structure can be conveniently and interactively
+analysed using the many features provided by \texttt{tskit}.
 
@@
 \end{figure}
-
-[Quick overview of LS and ARGs/tskit computation here]
-As used here, the Li and Stephens model essentially searches among
-all sequences in the ARG (both sampled and inferred ancestral nodes)
-
-The details of each of these steps are discussed in subsequent
-sections.
 
@@
 \subsection{Ancestral Recombination Graphs}
-[Quick overview of ARGs and Tree Sequences/tskit. Refer to the ARG paper for
-more discussion and details.]
+The term ``Ancestral Recombination Graph'' was introduced by
+Griffiths and colleagues~\citep{Griffiths1991-two,Griffiths1998-ancestral}
+and originally defined an alternative formulation of the coalescent
+with recombination stochastic process~\citep{Hudson1983-properties}.
+Subsequently, the term ARG came to be used in a more general way to
+describe not just realisations of this model, but to any
+recombinant genetic ancestry~\citep{Minichiello2006-mapping}.
+While there is some subtlety in the details~\citep{Wong2023-efficient},
+we can think of an ARG as being any graph that encodes the
+reticulate genetic ancestry of a sample of sequences under
+the influence of recombination.
+
+The ``succinct tree sequence'' is an ARG data structure
+that is both general (in terms of the types of ancestry that can
+be described) and computationally efficient~\citep{Wong2023-efficient}.
+Originally developed to facilitate large-scale coalescent
+simulations~\citep{Kelleher2016-wk}, the methods have been
+extended and applied to forwards-time
+simulations~\citep{Kelleher2018-xc,Haller2018-tree},
+calculation of population genetics statistics~\citep{Ralph2020-efficiently}
+and ARG inference~\citep{Kelleher2019-ba,Wohns2022-th}.
+The succinct tree sequence is based on a simple tabular representation,
+which defines a set of nodes, edges, sites and mutations. A \emph{node}
+represents a particular genome, which may be an observed sample
+or an inferred genetic ancestor. The genetic inheritance between
+a pair of nodes along a segment of genome is defined by
+the \emph{edge} $(\ell, r, p, c)$, which states that
+child node $c$ inherited its genome from parent node $p$
+from left coordinate $\ell$ to right coordinate $r$. A \emph{site}
+defines the genomic position at which some variation occurs,
+and a \emph{mutation} defines the site, node and derived state
+that generates variation.
+Given these tables, we can then efficiently construct the local
+genealogical trees along the genome (arising from recombination),
+and also perform a range of calculations efficiently by
+reasoning about the differences between these local
+trees~\citep{Kelleher2016-wk,Ralph2020-efficiently}. These
+algorithms have led to peformance increases of several orders
+of magnitude over the state-of-the-art in a range of
+applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,
+Ralph2020-efficiently,Baumdicker2022-ep}.
+The succinct tree sequence encoding
+is also very concise, allowing, for example, for millions of
+complete human genomes to be stored in a few gigabytes of
+space~\citep{Kelleher2019-ba}.
+
+The \texttt{tskit} library~\citep{Tskit2023-tskit} is a liberally
+licensed open source toolkit that provides a comprehensive suite
+of tools for working with ARGs. Based on core functionality written
+in C, it provides interfaces in C, Python and Rust. The Python interface
+is based on NumPy~\citep{Harris2020-array}, and provides a convenient
+platform for interactive analysis of large-scale data using, for
+example, Jupyter notebooks~\citep{Kluyver2016-jupyter} and taking
+advantage of the analysis tools in the burgeoning PyData ecosystem.
+Tskit is mature software, widely used in population genetics, and
+has been incorporated into several downstream
+% TODO Add more - geonomics?
+applications~\citep[e.g.][]{Speidel2019-yh,Haller2019-slim}
 
@@
 a convenient probabilistic approximation to the coalescent with recombination
-\citep{Hudson1983-if}, itself a generative model of a genealogical process with
+\citep{Hudson1983-properties}, itself a generative model of a genealogical process with
 recombination \citep{Li2003-ib}. Importantly, it can be implemented as an HMM
@@
 \begin{figure} \centering
-\includegraphics[width=\textwidth]{figures/imputation.pdf}
-\caption{\label{fig:imputation}A schematic of the two-step procedure to impute Pango lineage for inserted, non-sample nodes.
-Here, three nodes (question marks) have unknown Pango lineage (A).
-Because the parent of node 1 is available (i.e., L2) and there is no mutation occurring on the edge connecting the two nodes, the Pango lineage of the parent is directly propagated to node 1 (B).
-However, if there is a mutation (red X’s), then the Pango lineage is inferred by comparison to the set of lineage-defining mutations from COVID-CG\citep{Chen2021-zc}, as in case (C).
-Finally, the inferred status is propagated to any non-mutant child nodes, as in case (D).
+\includegraphics[width=.7\textwidth]{figures/imputation.pdf}
+\caption{\label{fig:imputation}A schematic of the two-step procedure to impute
+Pango lineage for inserted, non-sample nodes. Here, three nodes (question
+marks) have unknown Pango lineage (A). Because the parent of node 1 is
+available (i.e., L2) and there is no mutation occurring on the edge connecting
+the two nodes, the Pango lineage of the parent is directly propagated to node 1
+(B). However, if there is a mutation (red X’s), then the Pango lineage is
+inferred by comparison to the set of lineage-defining mutations from
+COVID-CG\citep{Chen2021-zc}, as in case (C). Finally, the inferred status is
+propagated to any non-mutant child nodes, as in case (D).
 }

### 2023-04-25 paper.tex
Updates from Overleaf

@@
 
-Inferring evolutionary trees while accounting for both mutation and
-recombination is a long-standing computational challenge. This challenge is
-particularly acute for SARS-CoV-2, given the large amount of data that has been
-shared, with over 15 million sequences in the GISAID database
-\citep{Shu2017-hp}. It is well known that recombination distorts phylogenies
-\citep{Schierup2000-fg} and affects the results of downstream analysis, e.g.,
-inference of selection~\citep{Anisimova2003-vr}. However, standard phylogenetic
-methods do not account for recombination
-\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Alternatively,
+%% ARGs are good for *representing* viral genealogies
+
+It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
+and affects the results of downstream analysis, e.g., inference of
+selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
+not account for recombination
+\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
 genetic ancestries of recombining lineages can be represented by ancestral
-recombination graphs (ARGs)~\citep{Griffiths1998-ancestral,Wong2023-efficient},
-which
-capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.
-
-Approaches to infer ARGs now exist that can scale to tens of thousands of human genomes and beyond  \citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}, and which deal with levels of sequential recombination far in excess of those seen in viral phylogenies.
-For example, using over 90 million variable sites \citet{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands of modern and ancient human genomes, accounting for millions of recombination events (over 5 million "local trees").
-To achieve this, \cite{Wohns2022-th} took advantage of several computational advances in ARG inference: (1) the "succinct tree sequence" data structure that can compactly store large ARGs \citep{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM) implementation of the Li and Stephens (LS) model \citep{Li2003-ib} to infer copying paths between genomes \citep{Kelleher2019-ba}; and (3) efficient, open-source software tools \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.
+recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-if}, which
+capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently, 
+development of the ``succinct tree sequence'' data
+structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
+encoding of large ARGs, and this representation should be well-suited to storing
+recombinant viral genealogies, even those with millions of sampled sequence,
+such as SARS-CoV-2 (with currently over 15 million sequences in the GISAID database
+\citep{Shu2017-hp}). Moreover the ``tskit'' software library, which implements this
+data structure, enables efficient manipulation of the stored genealogies and is
+integrated with a suite of open-source software tools to simulate, analyse, and
+infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.
+
+%% ARG inference is hard
+
+Inferring recombinant phylogenies has been a long-standing computational challenge.
+However, approaches to infer ARGs now exist that can scale to tens of
+thousands of human genomes and beyond
+\citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}, and which
+deal with levels of repeated recombination far in excess of those seen in viral
+phylogenies. For example, using over 90 million variable sites
+\citet{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands
+of modern and ancient human genomes, accounting for millions of recombination
+events (over 5 million "local trees"). To achieve this, \cite{Wohns2022-th}
+took advantage of a highly efficient Hidden Markov Model (HMM) implementation
+of the Li and Stephens (LS) model \citep{Li2003-ib} to infer copying paths between
+genomes \citep{Kelleher2019-ba}.
+
+% What is new here?
 
 Here, we present a preliminary version of a novel scalable methodology
-(``sc2ts'', optionally pronounced ``scoots'') that builds on the advances above and additionally takes advantage of known collection dates to infer ARGs from millions of full-length SARS-CoV-2 genomes.
-Using publicly available SARS-CoV-2 genome sequences, we demonstrate that sc2ts can be used to simultaneously (1) detect putative SARS-CoV-2 recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages; (3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record of evolutionary information about SARS-CoV-2 that can be efficiently accessed using "tskit" (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
-Sc2ts is a promising method to automate detection of SARS-CoV-2 recombinants and to jointly model recombinational and mutational processes shaping the genomic diversity of SARS-CoV-2 at scale.
+(``sc2ts'', optionally pronounced ``scoots'') that builds on the advances above
+and additionally takes advantage of known collection dates to infer ARGs from millions
+of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
+we demonstrate that sc2ts can be used to simultaneously (1) detect putative SARS-CoV-2
+recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
+(3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record
+of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
+"tskit" (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
+Sc2ts is a promising method to automate detection of SARS-CoV-2 recombinants
+and to jointly model recombinational and mutational processes shaping the genomic
+diversity of SARS-CoV-2 at scale.
 

### 2023-04-25 paper.tex

@@
 \midrule
-XA      &        [B.1.177.18, B.1.1.7] &     [B.1.1.7, B.1.177] &   5 &
-\{'XA': 5\} \\
+XA      &        [B.1.177.18, B.1.1.7] &     [B.1.1.7, B.1.177] &   5 & XA: 5 \\
 XAA     &               [BA.1, BA.2.9] &         [BA.1*, BA.2*] &   8 &
-\{'XAB': 39, 'XAG': 17, 'XAA': 8, 'XQ': 2, 'XU'... \\
-XAC     &          [BA.2.3, BA.1.17.2] &  [BA.2*, BA.1*, BA.2*] &  27 &
-\{'XAC': 27\} \\
-XAE     &                 [BA.2, BA.1] &         [BA.2*, BA.1*] &   9 &
-\{'XAE': 9\} \\
+    XAB: 39, XAG: 17, XAA: 8 \\
+    &&&&   XAA: 8, XQ: 2, XU: ... \\
+XAC     &          [BA.2.3, BA.1.17.2] &  [BA.2*, BA.1*, BA.2*] &  27 & XAC: 27 \\
+XAE     &                 [BA.2, BA.1] &         [BA.2*, BA.1*] &   9 & XAE: 9 \\
 XAG     &               [BA.1, BA.2.9] &         [BA.1*, BA.2*] &  17 &
-\{'XAB': 39, 'XAG': 17, 'XAA': 8, 'XQ': 2, 'XU'... \\
-XF      &                 [AY.4, BA.1] &    [B.1.617.2*, BA.1*] &   2 &
-\{'XF': 2\} \\
-XG      &              [BA.1.17, BA.2] &         [BA.1*, BA.2*] &  32 &
-\{'XG': 32, 'XAB': 1\} \\
+    XAB: 39, XAG: 17, XAA: 8 \\
+    &&&&   XAA: 8, XQ: 2, XU: ... \\
+XF      &                 [AY.4, BA.1] &    [B.1.617.2*, BA.1*] &   2 & XF: 2 \\
+XG      &              [BA.1.17, BA.2] &         [BA.1*, BA.2*] &  32 & XG: 32, XAB: 1 \\
 XH      &            [BA.1.20, BA.2.9] &         [BA.1*, BA.2*] &  11 &
-\{'XAF': 34, 'XH': 11, 'B.1.1.529': 3, 'XE': 3,... \\
-XK      &             [BA.1.1.1, BA.2] &         [BA.1*, BA.2*] &   6 &
-\{'XK': 3\} \\
+    XAF: 34, XH: 11 \\
+&&&& B.1.1.529: 3, XE: 3,... \\
+XK      &             [BA.1.1.1, BA.2] &         [BA.1*, BA.2*] &   6 & XK: 3 \\
 XL      &            [BA.1.17.2, BA.2] &         [BA.1*, BA.2*] &  10 &
-\{'XL': 10, 'XAB': 1, 'XU': 1, 'XQ': 1\} \\
+    XL: 10, XAB: 1, XU: 1, \\
+    &&&& XQ: 1 \\
 XR      &               [BA.1.1, BA.2] &       [BA.1.1*, BA.2*] &   8 &
-\{'XQ': 9, 'XR': 8, 'XAB': 2\} \\
-XS      &              [AY.36, BA.1.1] &  [B.1.617.2*, BA.1.1*] &   6 &
-\{'XS': 4\} \\
+    XQ: 9, XR: 8, XAB: 2\\
+XS      &              [AY.36, BA.1.1] &  [B.1.617.2*, BA.1.1*] &   6 & XS: 4\\
 XT      &       [BA.2.23, Unknown (R)] &         [BA.2*, BA.1*] &   1 &
-\{'BA.2': 1, 'BA.1': 1, 'BA.1.9': 1, 'BA.2.23':... \\
-XV      &  [BA.1.1, BA.2, Unknown (R)] &         [BA.1*, BA.2*] &   3 &
-\{'BA.2': 8, 'XV': 1\} \\
-XW      &            [BA.1.1.15, BA.2] &         [BA.1*, BA.2*] &  11 &
-\{'XW': 11, 'XN': 2\} \\
-XY      &               [BA.1.1, BA.2] &         [BA.1*, BA.2*] &   6 &
-\{'XY': 6, 'XAF': 1\} \\
+    BA.2: 1, BA.1: 1, BA.1.9: 1, \\
+    &&&& BA.2.23:... \\
+XV      &  [BA.1.1, BA.2, Unknown (R)] &         [BA.1*, BA.2*] &   3 & BA.2: 8, XV: 1 \\
+XW      &            [BA.1.1.15, BA.2] &         [BA.1*, BA.2*] &  11 & XW: 11, XN: 2 \\
+XY      &               [BA.1.1, BA.2] &         [BA.1*, BA.2*] &   6 & XY: 6, XAF: 1 \\
 \bottomrule
@@
 \caption{\label{tab:pango-single-origin}
-Table summarising the status of the Pango X-lineages in the Long ARG that are inferred to have a single recombinant origin based on the NextClade Pango lineage assignments.
+Table summarising the status of the Pango X-lineages in the Long ARG thare inferred to have a single recombinant origin based on the NextClade Pango lineage assignments.
 Imputed parent lineages are shown from the ARG and also to parent lineages as defined by the official Pango designation alias key.

### 2023-04-25 paper.tex
Partial update of the LS section

@@
 genetic ancestries of recombining lineages can be represented by ancestral
-recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-if}, which
+recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
 capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
@@
 \subsection{The Li and Stephens model}
-[Edit down, cutting the text in the fig caption significantly]
-
-The Li and Stephens (LS) model is
-a convenient probabilistic approximation to the coalescent with recombination
-\citep{Hudson1983-properties}, itself a generative model of a genealogical process with
-recombination \citep{Li2003-ib}. Importantly, it can be implemented as an HMM
-in which a focal genome is a sequence of nucleotides (observed states) that are
-probabilistically emitted as an imperfect mosaic of a set of reference genomes
-(hidden states). As such, an efficient HMM machinery can be employed to make
-statements about parameters of interest and highlight potential sources of
-error in genotyping and sequencing data. The LS model is used in a wide variety
-of applications in genomics. For example, it is the engine that drives
+The Li and Stephens (LS) model\citep{Li2003-ib}  is an approximation of the
+coalescent with recombination~\citep{Hudson1983-properties}, which captures
+many of the important features of the joint processes of mutation and
+recombination. It is a Hidden Markov Model (HMM) in which a focal genome is a
+sequence of nucleotides that are probabilistically emitted as
+an imperfect mosaic of a set of reference genomes
+(Figure~\ref{fig:ls_diagram}).
+The LS model is used in a wide variety
+of applications in genomics, including modern approaches to
 statistical genotype phasing and imputation
-\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and
-it can be used to estimate mutation rates (e.g., REF), recombination rates
-\citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral
+\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa},
+and estimation of parameters such as
+recombination rates \citep[e.g.][]{Hinch2011-tz},
+and selection within and across hosts in viral
 sequence data \citep[e.g.][]{Palmer2019-wa}.
+See~\cite{Mcvean2019-linkage} for further review and discussion
+of the LS model.
 
 \begin{figure} \centering
-\includegraphics[width=0.6\textwidth]{figures/ls_diagram_covid.pdf}
+\includegraphics[width=0.5\textwidth]{figures/ls_diagram_covid.pdf}
 % FIXME this caption is too long
-\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS) model.
-The LS model in its implementation as an HMM is governed by the transition and
-emission matrices, $Q$ and $E$, respectively. Each may be a function of the
-reference panel members, but transitions are generally assumed to be
-independent of the hidden states themselves (shown in the pink panel). This
-dramatically increases performance as the state space drops to two states
-(i.e., switching or not switching). The recombination probability at site
-$\ell$ is $r\ell$, while n denotes the number of members in the reference
-panel. Emissions may also be a function of the nucleotide states, but in our
-case, we assume that all observed alleles (in both the reference panel and
-focal sequence, $a\ell$) at the site under consideration are equally likely and
-occur with mutation probability $\ell$ (shown in the blue panel). These two
-processes, transition and emission (which encode recombination and mutation,
-respectively), define the generative process of the HMM, by which an imperfect
-mosaic is emitted from a reference panel of samples and nodes from the ARG
-inserted at earlier collection dates (shown in the lower panel). Here, the
-unseen states in the reference panel are shown as coloured lines enclosed by
-the grey box, and the true path through the data which leads to the emitted
-focal sequence is displayed by the black arrow. Examples of transition and
-emission probabilities along this trajectory are shown by the red and blue
-arrows, respectively. Black crosses along the focal sequence show genotyping
-errors or mutations. Note that in practice, we do not know the pattern of
-colours along the focal sequence and must infer it using the Viterbi
-algorithm.} \end{figure}
+\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
+model, in which a query sequence (bottom) is described as an
+imperfect mosaic of the sequences in a reference panel.
+Black crosses along the focal sequence show genotyping
+errors or mutations.
+In the standard formulation, at site $\ell$, the recombination probability is $r_\ell$,
+the mutation probability is $\mu_\ell$ and $n$
+denotes the size of the reference panel.
+The Viterbi algorithm can be used to find a
+``copying path'' through the reference panel for a given query sequence that
+maximises the likelihood under these parameters.
+% The LS HMM is governed by the transition and
+% emission matrices, $Q$ and $E$, respectively.
+% Each may be a function of the
+% reference panel members, but transitions are generally assumed to be
+% independent of the hidden states themselves (shown in the pink panel). This
+% dramatically increases performance as the state space drops to two states
+% (i.e., switching or not switching).
+% Emissions may also be a function of the nucleotide states, but in our
+% case, we assume that all observed alleles (in both the reference panel and
+% focal sequence, $a\ell$) at the site under consideration are equally likely and
+% occur with mutation probability $\ell$ (shown in the blue panel).
+% These two
+% processes, transition and emission (which encode recombination and mutation,
+% respectively), define the generative process of the HMM, by which an imperfect
+% mosaic is emitted from a reference panel.
+ % of samples and nodes from the ARG
+% inserted at earlier collection dates (shown in the lower panel). Here, the
+% unseen states in the reference panel are shown as coloured lines enclosed by
+% the grey box, and the true path through the data which leads to the emitted
+% focal sequence is displayed by the black arrow. Examples of transition and
+% emission probabilities along this trajectory are shown by the red and blue
+% arrows, respectively.
+% Note that in practice, we do not know the pattern of
+% colours along the focal sequence and must infer it using the Viterbi algorithm.
+}
+\end{figure}
 
@@
 sequences, given a set of reference sequences, $Q$, and $E$.
-
-Because the HMM is memoryless, the potential parental lineages are
-independently assessed at each recombination breakpoint. This can result in the
-inference of a recombinant lineage descending immediately from three (or even
-more) parental lineages. We find that often two of the three parents are
-identical or similar in sequence in the segments contributed to the recombinant
-lineage. A more parsimonious explanation would be to switch back to the
-original parent sequence. In the future, a pragmatic solution would be to
-detect such cases during the ARG inference stage and to adjust the copying
-paths accordingly (see ``Improvements'', below). Meanwhile, however, we urge
-caution when interpreting the parentage of lineages with multiple breakpoints.
+The relative rates of mutation and recombination are usually controlled by
+the site-specific parameters $\mu_\ell$ and $r_\ell$, respectively.
+For this work we used a simpler formulation, which uses a single parameter
+(the mismatch ratio, MMR) to control the relative importance of mutation
+and recombination in the HMM. Specifically, an MMR value of $k$ will
+prefer $k$ mismatches (mutations) to a recombination.[DUNCAN MORE DETAILS
+HERE PLEASE]
+
+[JK: TODO, expand this next paragraph to two, explaining a bit more that
+we're searching among every node in the ARG, using the efficient tree
+implementation in tsinfer, and a few words about how we can use the
+HMM more, to do things like stochastic traceback, etc.]
 
@@
 \toprule
-A Group (XA) & Jackson        & 21,255–21,613 & B.1.177/B.1.1.7 \\
+A Group (XA) & Jackson        &  21,256–21,613 & B.1.177/B.1.1.7 \\
              & \texttt{sc2ts} &  21,256--22,227 & B.1.177.18/B.1.1.7 \\
 \midrule
-B Group & Jackson        &  6,528--6,954 & B.1.36/B.1.1.7  \\
+B Group & Jackson        &  6,529--6,953 & B.1.36/B.1.1.7  \\
         & \texttt{sc2ts} &  6,529--6,954 & B.1.36/B.1.1.7  \\
 \midrule
-C Group & Jackson        &  25,997--27,972 &  B.1.1.7/B.1.221 \\
-        & \texttt{sc2ts} &  25,996--27,441 &  B.1.1.7/B.1.221 \\
+C Group & Jackson        &  25,997--27,441 &  B.1.1.7/B.1.221 \\
+        & \texttt{sc2ts} &  25,997--27,972 &  B.1.1.7/B.1.221 \\
 \midrule
-D Group & Jackson        &  21,575--23,063 &  B.1.36.17/B.1.1.7 \\
+D Group & Jackson        &  21,576--23,063 &  B.1.36.17/B.1.1.7 \\
         & \texttt{sc2ts} &  22,445--23,063 &  B.1.36.39/B.1.1.7 \\
 \midrule
-CAMC-CBA018 & Jackson        &  20,389--21,254 & B.1.177/B.1.1.7 \\
+CAMC-CBA018 & Jackson        &  20,390--21,254 & B.1.177/B.1.1.7 \\
             & \texttt{sc2ts} &  16,177--21,255 & B.1.177/B.1.1.7 \\

### 2023-04-25 plot.py
Add plots with legends

@@ -2,6 +2,7 @@ import argparse
 import collections
 from datetime import datetime, timedelta
 import hashlib
+import itertools
 import re
 import sys
 import json
@@ -18,6 +19,7 @@ mpl.use("Agg")
 from matplotlib.colors import rgb2hex
 from matplotlib.cm import get_cmap
 from matplotlib.lines import Line2D
+from matplotlib.patches import Patch
 from matplotlib import pyplot as plt
 import numpy as np
 import tqdm
@@ -862,15 +864,6 @@ def change_pos(pos, key, dx=None, dy=None, x=None, y=None, xy_from=None):
 class Pango_X_graph(Figure):
     sample_metadata_labels = ""  # Don't show the strain name
     node_colours = {
-        # Pango X-lineages in red/brown/pink
-        'XA': '#bb0000',
-        'XB': '#fe812e',
-        'XD': '#fc31fb',
-        'XQ': '#db5884',
-        'XU': '#894a4e',
-        'XAA': '#ffc9a5',
-        'XAB': '#fdacfe',
-        'XAG': '#ff0055',
 
         # B lineages in blue/green (AY.100 is B.1.617.2.100)
         # Basic in blue
@@ -881,9 +874,12 @@ class Pango_X_graph(Figure):
         'B.1.627': '#7b96cc',
         'B.1.1': '#aea4ff',
 
-        'B.1.1.7': '#8666e3', # Alpha
+        # Alpha: purple / lilac
+        'B.1.1.7': '#ae9cff',
 
+        # Delta: yellow/brown
         'B.1.617.2': '#e2da91',  # Delta
+        'AY.1': '#ffe44a',  # Delta
         'AY.4': '#b78c19',  # Delta
 
         # Omicron (B.1.1.529...) in green-ish
@@ -892,6 +888,15 @@ class Pango_X_graph(Figure):
         'BA.1.15': '#248f0f',  # Omicron
         'BA.2.9': '#21e1ab', # Omicron
 
+        # Pango X-lineages in red/orange/pink
+        'XA': '#ff0055',
+        'XB': '#fc31fb',
+        'XD': '#bb0000',
+        'XQ': '#db5884',
+        'XU': '#894a4e',
+        'XAA': '#ffc9a5',
+        'XAB': '#fdacfe',
+        'XAG': '#fe812e',
 
         None: "lightgray",  # Default
         "Unknown (R)": "k",
[…]

### 2023-04-26 plot.py
Add the detailed supplementary subgraphs

@@ -193,7 +193,7 @@ class Cophylogeny(Figure):
                     order.append([int(n[1:]) for n in re.findall(r"n\d+", line)])
         return order
 
-    def __init__(self, args):
+    def __init__(self):
         """
         Defines two simplified tree sequences, focussed on a specific tree. These are
         stored in self.sc2ts and self.nxstr
@@ -611,7 +611,7 @@ class RecombinationNodeMrcas(Figure):
     csv_fn = "breakpoints_{}.csv"
     data_dir = "data"
 
-    def __init__(self, args):
+    def __init__(self):
         self.ts, self.basetime = utils.load_tsz(self.data_dir, self.sc2ts_filename)
 
         prefix = utils.snip_tsz_suffix(self.sc2ts_filename)
@@ -925,7 +925,7 @@ class Pango_X_graph(Figure):
         _, idx = np.unique(nodes, return_index=True)
         return nodes[np.sort(idx)]
 
-    def __init__(self, args):
+    def __init__(self):
         ts, self.basetime = utils.load_tsz(self.ts_dir, self.long_fn)
         self.ts = sc2ts.detach_singleton_recombinants(ts)
         logging.info(
@@ -958,8 +958,8 @@ class Pango_X_tight_graph(Pango_X_graph):
     legend=None
     node_def_lineage = "Nextclade_pango"  # Always use this when defining which nodes to pick
 
-    def __init__(self, args):
-        super().__init__(args)
+    def __init__(self):
+        super().__init__()
         if not hasattr(self, "label_replace"):
             self.label_replace = {
                 "Unknown": "",
@@ -1079,11 +1079,12 @@ class Pango_XA_nxcld_tight_graph(Pango_X_tight_graph):
     sample_metadata_labels = "Imputed_" + imputed_lineage  # For the first plot, use labels
     figsize=(7, 4)
 
-    def __init__(self, args):
-        super().__init__(args)
-        # For the first plot, leave the "mutation" labels as they are
-        del self.label_replace[" mutation"]
-        del self.label_replace[" mutations"]
+    def __init__(self):
+        super().__init__()
+        # If this is the base class, don't abbreviate the "mutation" labels
[…]

### 2023-04-26 plot.py
Print out tsk-ISL mappings

@@ -1029,7 +1029,7 @@ class Pango_X_tight_graph(Pango_X_graph):
         self.nodes = self.define_nodes(self.ts)
         G, pos = self.plot_subgraph(
             self.nodes, ax=ax, node_positions=self.node_positions)
-        self.plot_legend(ax)
+        self.legend_key(ax)
         self.post_process(ax)
 
         if self.node_positions is None:
@@ -1039,7 +1039,8 @@ class Pango_X_tight_graph(Pango_X_graph):
             fn = self.fn_prefix
         plt.savefig(fn + f".{args.outtype}", bbox_inches="tight")
 
-    def plot_legend(self, ax):
+    def legend_key(self, ax):
+        # Plot or otherwise output a legend or key. Define in the subcless
         pass
 
     @classmethod
@@ -1110,7 +1111,7 @@ class Pango_XA_nxcld_tight_graph(Pango_X_tight_graph):
         ax.set_xlim(x_min - (x_max - x_min) * 0.25, x_max)
         # ax.set_ylim(y_min, y_max + (y_max - y_min) * 0.3)
 
-    def plot_legend(self, ax):
+    def legend_key(self, ax):
         legend_pts = plt.legend(
             title=f"Node types",
             handles=self.make_legend_elements(
@@ -1147,7 +1148,7 @@ class Pango_XAG_nxcld_tight_graph(Pango_X_tight_graph):
     legend = "upper left"
     figsize=(11, 4.5)
 
-    def plot_legend(self, ax):
+    def legend_key(self, ax):
         used_colours = self.used_pango_colours(self.nodes)
         del used_colours["Unknown (R)"]
         ax.legend(
@@ -1239,7 +1240,7 @@ class Pango_XD_nxcld_tight_graph(Pango_X_tight_graph):
         change_pos(pos, 575079, dx=shift)
 
 
-    def plot_legend(self, ax):
+    def legend_key(self, ax):
         used_colours = self.used_pango_colours(self.nodes)
         del used_colours["Unknown (R)"]
         ax.legend(
@@ -1320,7 +1321,7 @@ class Pango_XB_nxcld_tight_graph(Pango_X_tight_graph):
 
         change_pos(pos, 339516, dx=dx)
 
-    def plot_legend(self, ax):
+    def legend_key(self, ax):
         used_colours = self.used_pango_colours(self.nodes)
         del used_colours["Unknown (R)"]
         ax.legend(
[…]

### 2023-04-26 paper.tex
Updates from Overleaf

@@
 For XB, all samples trace back to a originating node which is the product of a recombination between a B.1 sample (specifically the strain England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of the genome from the start up to a maximum position of 23604bp), and two other UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of 27389bp) and B.1.36.8 (the remaining fragment of the genome).
-As described below, in the long ARG the equivalent recombination node has only 2 parents, with no involvement of  B.1.36.8, and it is possible that the third parent in the wide ARG is artifactual.
-An added complication is that in both ARGs, additional non-X-designated samples such as descend from the not only a small number of are six further recombination nodes nested within the XB ``clade''.
-However, five are ancestral to only negligible fractions of the designated XB samples.
-The sixth accounts for about 17\% of the designated XB nodes, and involves a recombination between descendants of the originating XB recombination.
+As described later, in the long ARG the equivalent recombination node has only 2 parents, with no involvement of  B.1.36.8, and it is possible that the third parent in the wide ARG is artifactual.
+Additional complications arise because additional non-X-designated samples such as B.1.634 also descend from this recombination (also true in the long ARG, see below), and there are also a small number of further recombination nodes nested within the XB ``clade''.
+However, all but one of these nested recombinations are ancestral to only a negligible fractions of the designated XB samples.
+The exception accounts for about 17\% of the designated XB nodes, and involves a recombination between descendants of the originating XB recombination.
 More specifically strain USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination between an XB grandparent and its XB grandchild.
@@
 Single origin Nextclade Pango X lineages.
-(A) Subgraph for XA in the long ARG: all XA samples and their ancestral lineages are shown back to the nearest sampled viral genome, with dotted lines showing continuation of the ARG.
-Nodes are coloured by Nextclade Pango designation; smaller nodes are not samples and their Pango designation was imputed where possible.
+(A) Subgraph for XA in the long ARG: all XA samples and their ancestral lineages are shown outwards to the nearest sampled viral genome; dotted lines show where the ARG continues.
+Nodes are coloured by Nextclade Pango designation; smaller symbols are nodes inserted by  sc2ts (i.e.
+not samples); their Pango designation has been imputed where possible.
 (B) Details for the single-origin Pango X lineages in the long ARG.
@@
 All XAG samples share a single origin; however this is also shared with all XAA samples, and includes some, but not all, of the samples identified as XAB, XQ, and XU by Nextclade.
-Note that if the GISAID designations are used, many of the samples marked here as XAB are reclassified as BA.2, and XAG becomes fully monophyletic (supplementary figure \ref{fig:pango_XAG_gisaid_graph}; this also gives identities of the samples in the graph)
+Note that if the GISAID designations are used, many of the samples marked here as XAB are reclassified as BA.2, and XAG becomes fully monophyletic (supplementary figure \ref{fig:pango_XAG_gisaid_graph}; this also gives identifying labels to the sample nodes, which can be converted to EPI\_ISL indentifiers using Supplemeentary table XXX )
 
@@
 In Figure XXX we give the details of these multiple pango lineages, along with an example lineage subgraph.
+As with the subgraphs in Figure~\ref{tab:pango-single-origin}, the equivalent XD plot with GISAID Pango designations, along with identifying labels, is shown for comparison in Supplementary Figure XXX
 
@@
 In addition these complex topologies often include extra recombination nodes within the topology that descends from the main originating event.
-An example is XB in which recombination between a B.1 sample and B.1.627 sample leads both to the XB samples and to a "hairball" of further recombination nodes whose descendants have not been identified as recombinants by Nextclade (blue nodes, Figure~\ref{fig:complex_origins_graph})  ...
+An example is XB in the long ARG, plotted in Figure~\ref{fig:complex_origins_graph}.
+Here, a recombination between a B.1 sample and B.1.627 sample leads both to the XB samples and to a "hairball" of further recombination nodes whose descendants have not been identified as recombinants by Nextclade (blue nodes, "pre-Alpha" VoC).
+However, in this subgraph, there is only a single sample, identified as XB by Nextclade, which is descended from these
+nested recombination events; moreover, this sample is not identified as XB by GISAID (Supplementary Figure~\ref{}).
+We suggest that investigating cases of complex origins, and identifing which (if any) of the nested, further recombination events may be artifactual, is an important area of future research.
 
@@
 \begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/Pango_XA_gisaid_large_graph.pdf}
+\caption{\label{fig:pango_XA_gisaid_graph}  DESCRIBE ME}
+\end{figure}
+
+\begin{figure} \centering
 \includegraphics[width=\textwidth]{figures/Pango_XAG_gisaid_large_graph.pdf}
@@
 
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/Pango_XD_gisaid_large_graph.pdf}
+\caption{\label{fig:pango_XD_gisaid_graph}  DESCRIBE ME}
+\end{figure}
+
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/Pango_XB_gisaid_large_graph.pdf}
+\caption{\label{fig:pango_XB_gisaid_graph}  DESCRIBE ME}
+\end{figure}
+
 \end{document}

### 2023-04-27 paper.tex
Updates from Overleaf

@@
 
-\includegraphics[width=0.4\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
+\includegraphics[width=0.6\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
 \\ % &
@@
 
-In the long ARG, the sequences of 16 Pango recombinant lineages trace back to a single recombination node, as summarised in Table~\ref{tab:pango-single-origin}.
+In Figure~\ref{fig:pango-single-origin-graph}A we illustrate the case of XA.
+We construct a small subgraph of the long ARG, tracing routes from Nextclade-identified XA samples to the closest other sample strains in our dataset.
+Where a route passes upwards through a recombination node, we only show the continuing parental lineages to the most recent sample ancestors, rather than descending the ARG downward to cousins of the recombinant parents.
+Dotted lines show where this subgraph links to other nodes within the ARG.
+It is clear from the subgraph that all the XA samples trace to a single originating recombination node.
+
+Table~\ref{tab:pango-single-origin} summarises the sequences of all 16 of the Pango recombinant lineages that trace back to a single recombination node in the long ARG.
 [FIXME UPDATE THESE]
@@
 All XAG samples share a single origin; however this is also shared with all XAA samples, and includes some, but not all, of the samples identified as XAB, XQ, and XU by Nextclade.
-Note that if the GISAID designations are used, many of the samples marked here as XAB are reclassified as BA.2, and XAG becomes fully monophyletic (supplementary figure \ref{fig:pango_XAG_gisaid_graph}; this also gives identifying labels to the sample nodes, which can be converted to EPI\_ISL indentifiers using Supplemeentary table XXX )
+Note that if the GISAID designations are used, many of the samples marked here as XAB are reclassified as BA.2, and XAG becomes fully monophyletic (supplementary figure \ref{fig:pango_XAG_gisaid_graph}; this also gives identifying labels to the sample nodes, which can be converted to EPI\_ISL indentifiers using Supplementary table XXX )
 
@@
 
-Example lineage subgraph
-% \includegraphics[width=0.4\textwidth]{figures/Pango_XD_nxcld_tight_graph.pdf}
+\includegraphics[width=0.45\textwidth]{figures/Pango_XD_nxcld_tight_graph.pdf}
 
@@
 \end{tabular}
-\caption{A.
-Example subgraph for XD in the long ARG, which shows .
+\caption{\label{fig:multiple_origins_graph} A.
+Example subgraph for XD in the long ARG, which shows two origins for the four Nextclade-labelled XD samples in the long ARG.
 B.
@@
 
-In Figure XXX we give the details of these multiple pango lineages, along with an example lineage subgraph.
-As with the subgraphs in Figure~\ref{tab:pango-single-origin}, the equivalent XD plot with GISAID Pango designations, along with identifying labels, is shown for comparison in Supplementary Figure XXX
+In Figure~\ref{fig:multiple_origins_graph} we give the details of these multiple pango lineages, along with an example lineage subgraph.
+The  earliest descendant sample of each of these putative independent XD origination nodes has EPI\_ISL: 11222324 (dated 2022-02-26) for the left hand plot, and EPI\_ISL 8514045 (dated 2021-12-30) for the right hand one.
+These two samples differ by  Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the equivalent subgraph with sample identifiers and GISAID Pango labels, none of which are XD (there are no samples designated by GISAID as XD in the long ARG).
 
@@
 However, in this subgraph, there is only a single sample, identified as XB by Nextclade, which is descended from these
-nested recombination events; moreover, this sample is not identified as XB by GISAID (Supplementary Figure~\ref{}).
+nested recombination events; moreover, this sample is not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph}).
 We suggest that investigating cases of complex origins, and identifing which (if any) of the nested, further recombination events may be artifactual, is an important area of future research.

### 2023-04-27 paper.tex

@@
 has been incorporated into several downstream
-% TODO Add more - geonomics?
-applications~\citep[e.g.][]{Speidel2019-yh,Haller2019-slim}
+applications~\citep[e.g.][]{Haller2019-slim,Speidel2019-yh,
+Terasaki2021-geonomics,
+Fan2022-genealogical,Korfmann2022-weak,
+Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}.
 

### 2023-04-27 paper.tex
Updates from Overleaf

@@
 Single origin Nextclade Pango X lineages.
-(A) Subgraph for XA in the long ARG: all XA samples and their ancestral lineages are shown outwards to the nearest sampled viral genome; dotted lines show where the ARG continues.
-Nodes are coloured by Nextclade Pango designation; smaller symbols are nodes inserted by  sc2ts (i.e.
-not samples); their Pango designation has been imputed where possible.
+(A) Subgraph for XA in the long ARG: all XA samples and their ancestral lineages are shown outwards to the nearest sampled viral genome; dotted lines show ARG continuation.
+Nodes are coloured by Nextclade Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with imputed Pango status.
 (B) Details for the single-origin Pango X lineages in the long ARG.
@@
 In Figure~\ref{fig:multiple_origins_graph} we give the details of these multiple pango lineages, along with an example lineage subgraph.
-The  earliest descendant sample of each of these putative independent XD origination nodes has EPI\_ISL: 11222324 (dated 2022-02-26) for the left hand plot, and EPI\_ISL 8514045 (dated 2021-12-30) for the right hand one.
-These two samples differ by  Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the equivalent subgraph with sample identifiers and GISAID Pango labels, none of which are XD (there are no samples designated by GISAID as XD in the long ARG).
+In this XD example the left hand subgraph (the origin of three XD samples) has an earliest sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand subgraph has a single sample with EPI\_ISL 8514045 (dated 2021-12-30).
+These two samples differ by 35 mutations; considering the time between the samples, it does seem possible that they represent true independent origins of XD.
+Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the same subgraphs with sample identifiers and coloured by GISAID Pango labels (however, no samples are designated by GISAID as XD in the long ARG).
 

### 2023-04-27 paper.tex
Update LS section

@@
 is also very concise, allowing, for example, for millions of
-complete human genomes to be stored in a few gigabytes of
+complete human genomes to potentially be stored in a few gigabytes of
 space~\citep{Kelleher2019-ba}.
@@
 
-[JK: TODO, expand this next paragraph to two, explaining a bit more that
-we're searching among every node in the ARG, using the efficient tree
-implementation in tsinfer, and a few words about how we can use the
-HMM more, to do things like stochastic traceback, etc.]
-
-Recombination is inferred using an efficient HMM implementation of the
-LS model. Copying paths are estimated from SARS-CoV-2 genomes sampled over
-time, with younger genomes copying (i.e. inheriting segments and mutations)
-from older genomes. Here, we were able to apply the LS model on millions of
-SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of
-the LS model that is a core component of the tsinfer software. For this
-application, we have parameterised the HMM using a mismatch ratio (MMR), which
-tunes how strongly mutations (i.e., no switching of template sequences) are
-preferred over recombination (i.e., switching between sequences) during
-copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
-focal genome more likely spans at least two ancestral genomes (i.e,. the focal
-genome is a putative recombinant) than a single ancestral genome. Here, the MMR
-was set to three in all the HMM runs (see ``Improvements'' for future potential
-development of the HMM).
+% OLD text, ready to be adapted (and deleted once done)
+% For this
+% application, we have parameterised the HMM using a mismatch ratio (MMR), which
+% tunes how strongly mutations (i.e., no switching of template sequences) are
+% preferred over recombination (i.e., switching between sequences) during
+% copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
+% focal genome more likely spans at least two ancestral genomes (i.e,. the focal
+% genome is a putative recombinant) than a single ancestral genome. Here, the MMR
+% was set to three in all the HMM runs .
+
+In \texttt{sc2ts} we use the efficient ARG-based implementation of the
+LS Viterbi algorithm from \texttt{tsinfer}~\citep{Kelleher2019-ba} to find
+the most likely copying path for each sample sequence
+among all sequences (sampled and inferred) in the current ARG.
+In the majority of cases, with non-recombinant sample sequences,
+the most likely solution is to copy from one
+of the nodes in the ARG that minimised the number of mutations required
+on the query sequence. Importantly, because the reference panel here consists of
+\emph{every} node in the ARG, we can match to both older sample
+sequences or internal nodes representing an inferred ancestral sequence
+(see subsequent subsections for details about how these are added).
+Thus, when no recombination is present, the LS Viterbi algorithm is
+implementing a version of parsimony, in which we are guaranteed to
+find a sequence that minimises the number of additional mutations required
+to incorporate a newly-added sample into the ARG.
+Recombination is then inferred when the most likely solution to the LS
+HMM is to copy from more than one ARG node along the genome, for a
+given sample sequece. [TODO MORE]
+
+We can solve this massive optimisation problem exactly because the ARG-based
+implementation of the LS HMM used in \texttt{tsinfer} scales approximately
+logarithmically with reference panel size (as opposed to linearly,
+for standard matrix-based approaches)~\citep{Kelleher2019-ba}.
+This efficient HMM algorithm is the main reason for \texttt{tsinfer}'s
+scalability, and here allows us to find closely matching
+sequences and recombination paths among millions of SARS-CoV-2
+genomes exactly under a well-defined statistical model.
+
+It is important to note that the Viterbi algorithm only returns
+\emph{one of} the copying paths that maximise the likelihood under
+ the given mutation and recombination parameters. There may be many
+such paths, from which we choose one arbitrarily. Also, the
+present choice of using a single mismatch-ratio parameter to control
+the likelihood of recombination vs mutation may lead to relatively
+flat likelihood spaces where many different paths have equal likelihood.
+There are many possibilities in using established HMM methodology
+to reason about and explore the space of possible matches, which may
+be a fruitful avenue for future work. See the Discussion for further
+potential improvements that may be made to the HMM.
 

### 2023-04-27 paper.tex

@@
 \subsection{Tree inference from clusters of samples}
-[Discuss the "local" tree inference, but be careful not to use the term]
+For a given daily batch of sample sequences that are to be attached to the
+ARG, the LS HMM provides the most likely copying path and required mutations
+for each of those sequences. We could directly attach these samples to
+the ARG according to the copying paths and sets of mutations, but
+such an approach would be highly unparsimonious because there are
+often many samples that share the same copying path with
+complex patterns of shared mutations.
+In this section we discuss the strategy used to infer likely tree
+toplogies among the samples that share the same copying path.
+In the majority of cases this copying path will consist of a single
+node (no recombination), and so for simplicity we will assume in the
+following discussion that all of the samples in a HMM-defined cluster
+attach to a single node. The differences required to deal with
+newly-detected putative recombinants are discussed below
+in Section XXX.
+
+With tens of thousands of samples being added to the ARG per day,
+such clusters of sequences attaching to the same node
+can be large, with complex patterns of shared mutations.
+To avoid the implied large polytomy and unparsimonious
+repetition of identical mutations among sibling nodes, we
+infer a tree among these samples (the LS HMM has already
+
+A natural
+way to
+
+We can think of the sequences attaching to a given node in the ARG
+from a daily batch as a polytomy,
+but with potentially shared mutations among the sibling nodes
+(Figure~\ref{fig:overview_sc2ts}C).
 

### 2023-04-28 paper.tex
Move methods to end
Re-break long lines

@@
 \section{Introduction}
-Recombination via template switching is a common feature in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2 \citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}.
-By bringing together mutations carried by different lineages, recombination plays an important role in generating genomic diversity, with recombinant lineages associated with adaptation to new host species and with the production of more immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
-
-Recombination was challenging to detect early in the SARS-CoV-2 pandemic due to gradual accumulation of genetic variation.
-In late 2020, the appearance and spread of variants of concern (VoC), designated into classes such as Alpha and Delta which harboured multiple characteristic mutations, increased the detectability of recombination.
-By searching for genomes containing segments from different VoC designations, multiple recombinant lineages have been detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha and Epsilon \citep{Wertheim2022-hj}, and Alpha and Delta~\citep{Sekizuka2022-xz}.
-More recently, the high prevalence of Omicron, with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has accelerated the rate of coinfection and the potential for recombination \citep{Bal2022-hq}.
-Recombinant lineages that have sufficient evidence of onward transmission (more than five descendants) and monophyly ($>70\%$ bootstrap support) are eligible for a Pango lineage designation~\citep{Rambaut2020-dw}, starting with the letter X
+Recombination via template switching is a common feature
+in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt}),
+including SARS-CoV-2
+\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
+together mutations carried by different lineages, recombination plays an
+important role in generating genomic diversity, with recombinant lineages
+associated with adaptation to new host species and with the production of more
+immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
+
+Recombination was challenging to detect early in the SARS-CoV-2 pandemic due to
+gradual accumulation of genetic variation. In late 2020, the appearance and
+spread of variants of concern (VoC), designated into classes such as Alpha and
+Delta which harboured multiple characteristic mutations, increased the
+detectability of recombination. By searching for genomes containing segments
+from different VoC designations, multiple recombinant lineages have been
+detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha
+and Epsilon \citep{Wertheim2022-hj}, and Alpha and
+Delta~\citep{Sekizuka2022-xz}. More recently, the high prevalence of Omicron,
+with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
+accelerated the rate of coinfection and the potential for recombination
+\citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of
+onward transmission (more than five descendants) and monophyly ($>70\%$
+bootstrap support) are eligible for a Pango lineage
+designation~\citep{Rambaut2020-dw}, starting with the letter X
 % FIXME: should this be a citation?
-(``Pango Lineage Nomenclature: Provisional Rules for Naming Recombinant Lineages'' 2021).
-
-Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that combine specific subsets of mutations from different lineages \citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
-This approach only identifies those inter-lineage recombination events resulting in the patterns investigated.
-Others have inferred recombinant genealogies to disentangle recombination from recurrent mutation, without having to pre-specify characteristic mutations for each lineage \citep{Ignatieva2021-rg}.
-More recently, \cite{Turakhia2022-it} presented a phylogeny-based approach (``RIPPLES'') to identify putative recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying motifs.
-To find candidate recombinants, RIPPLES first builds a phylogeny without assuming recombination and then scans for long branches (i.e., carrying many mutations) and then determines if these candidates would be better explained by recombination.
-Specifically, RIPPLES exhaustively breaks each candidate sequence into segments and attempts to find more parsimonious placements for each segment on the phylogeny.
-If such placements are found, the sequence is identified as a putative recombinant.
-Although it enables rapid searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that accounts for only mutations, treats recombinants \textit{post hoc}, and is an incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
-As noted by the authors, a \textit{post hoc} treatment of recombination is possible when recombinant lineages are rare and leave few descendants.
-However, this is becoming increasingly untenable with the proliferation of recombinant lineages; for example, more than half of the sequences sampled in February 2023 are from XBB and its descendants~\citep{Chen2022-pz}.
-This also means that future evolution of SARS-CoV-2 is likely to involve multiple sequential recombination events on top of existing recombinant lineages, creating a highly reticulated genealogy.
+(``Pango Lineage Nomenclature: Provisional Rules for Naming Recombinant
+Lineages'' 2021).
+
+Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
+combine specific subsets of mutations from different lineages
+\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
+This approach only identifies those inter-lineage recombination events
+resulting in the patterns investigated. Others have inferred recombinant
+genealogies to disentangle recombination from recurrent mutation, without
+having to pre-specify characteristic mutations for each lineage
+\citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it} presented a
+phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
+over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find
+candidate recombinants, RIPPLES first builds a phylogeny without assuming
+recombination and then scans for long branches (i.e., carrying many mutations)
+and then determines if these candidates would be better explained by
+recombination. Specifically, RIPPLES exhaustively breaks each candidate
+sequence into segments and attempts to find more parsimonious placements for
+each segment on the phylogeny. If such placements are found, the sequence is
+identified as a putative recombinant. Although it enables rapid searches for
+genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
+accounts for only mutations, treats recombinants \textit{post hoc}, and is an
+incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
+As noted by the authors, a \textit{post hoc} treatment of recombination is
+possible when recombinant lineages are rare and leave few descendants. However,
+this is becoming increasingly untenable with the proliferation of recombinant
+lineages; for example, more than half of the sequences sampled in February 2023
+are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
+future evolution of SARS-CoV-2 is likely to involve multiple sequential
+recombination events on top of existing recombinant lineages, creating a highly
+reticulated genealogy.
 
@@
 diversity of SARS-CoV-2 at scale.
-
-\section{Methods}
-
-\subsection{Overview}
-Sc2ts is a method for inferring ARGs from densely sampled pandemic-scale data
-in real time, in which recombination occurs at a low but significant rate.
-The essential idea is to incrementally update the ARG each day
-with the sequences collected on that day. These updates are performed by
-first finding likely ``copying paths'' for each sequence in the daily
-batch to the current ARG, using the ARG-based implementation of
-the Li and Stephens model~\citep{Li2003-ib} introduced
-in \texttt{tsinfer}~\citep{Kelleher2019-ba}.
-These copying paths
-will mostly consist of a new sample sequence copying from a node in the ARG
-with a small number of mutations, and often many samples
-from a daily batch will copy from the same ARG node. The second
-step is then to ``resolve'' these implied polytomies by using
-standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
-This greedy update strategy inevitably lead to unparsimonious
-topologies, and the third step is to then increase the
-overall parsimony of the inferred ARG by making some simple topological
-updates (Figure~\ref{fig:overview_sc2ts}C and D).
-The operations are carried out efficiently using the well-established
-\texttt{tskit} library~\citep{Kelleher2018-xc,Ralph2020-efficiently},
-and the resulting structure can be conveniently and interactively
-analysed using the many features provided by \texttt{tskit}.
-
-\begin{figure} \centering
-\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
-\caption{\label{fig:overview_sc2ts}
-A schematic of the sc2ts method. Sc2ts reconstructs the genetic relationships
-among SARS-CoV-2 genomes by copying samples to all possible ancestors collected
-at earlier time points (curved arrows). An ARG is built by attaching new sample
-genomes grouped by collection date to an existing ARG rooted at the Wuhan-Hu-1
-reference genome, day by day. Each iteration involves three stages: (1)
-attachment of new samples to the growing ARG (A, B); (2) reconstruction of
-trees relating the samples under each attachment node (C); and parsimony-based
-tree topology adjustments (D, E). In the absence of  recombination, sc2ts
-infers an ARG that is a single tree relating the samples (A). However, in the
-presence of recombination, sc2ts infers an ARG that is encoded as a sequence of
-local trees relating segments of the sample genomes (B), inserting a
-recombination node to represent a putative recombination event. Additionally,
-mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to
-make more parsimonious placements of mutations that should be shared or should
-not be immediately reverted, respectively.}
-\end{figure}
-
-% We introduce an ARG inference method, summarised in
-% Figure~\ref{fig:overview_sc2ts}.
-% % Note going to rewrite this anyway, so not trying to make it flow.
-% It uses (1) an efficient HMM implementation of the Li and Stephens (LS) model
-% \citep{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2)
-% the reported collection dates of the SARS-CoV-2 samples to determine the order
-% of genome copying; and (3) several parsimony-based heuristics to locally refine
-% the topology. In brief, this combined approach uses the likelihood model of LS
-% to attach sample genomes to existing nodes and to detect recombinants, but
-% because many of the daily tip samples harbour few, if any, mutational changes,
-% we use parsimony to resolve their close relationships.
-
-% Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by
-% taking advantage of the reported collection dates to incrementally grow the
-% ARG. The inference process occurs forward in time, attaching new genomes
-% aggregated by their collection date (daily batches), from the earliest to the
-% last collection date. Prior to attaching sample genomes, the ARG is initialised
-% such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence
-% of recombination, the ARG is a single tree with no reticulation (Figure 1A);
-% however, in the presence of recombination, a reticulate ARG is generated, with
-% variation in tree topology across the genome (Figure 1B). Unlike standard
-% phylogenetic trees in which internal nodes always represent inferred ancestors,
-% internal nodes in the sc2ts ARGs represent sequences which can either be
-% inferred ancestors or known samples.
-
-% Each daily batch of genomes is first attached to the growing ARG without any
-% phylogenetic resolution. In cases where the newly added sequence is inferred to
-% be a recombinant, a recombination node is inserted; the ``date'' of this node
-% is arbitrarily set to the average of the collection dates of the genome and the
-% most recent of its parent nodes. Otherwise, the genomes are attached to the
-% most closely related ancestral node (an attachment node). This creates
-% polytomies that are then locally resolved among the descendants of each
-% attachment node using the Unweighted Pair Group Method with Arithmetic Mean
-% (UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
-% This hierarchical clustering algorithm produces a dendrogram relating a set of
-% sequences, which is treated as an ultrametric binary tree here. Subsequently,
-% mutations are placed onto the daily tree by maximum parsimony. An issue with
-% the UPGMA algorithm is that it may create internal nodes supported by no
-% informative site (i.e., having no mutation immediately ancestral to them). In
-% an adjustment step, such unsupported internal nodes are removed to
-% ``un-resolve'' the daily tree; the remaining internal nodes, which are
-% supported by at least one shared mutation, are retained. The ``dates'' of the
-% non-sample internal nodes are set to the same date as the samples in the daily
-% batch. The resulting daily tree is grafted onto the attachment node in the
-% global ARG.
-
-% During ARG inference, two types of non-sample nodes are inserted to improve
-% parsimonious placement of mutations: mutation-coalescing nodes and
-% pushed-reversion nodes. We call these operations tree adjustments. A
-% mutation-collapsing node is inserted when a shared mutation between sibling
-% nodes is found (Figure 1D). An attempt to create mutation-collapsing nodes is
-% made after grafting a new daily tree onto the global ARG. If shared mutations
-% exist between two or more sibling nodes, then a parsimonious explanation is to
-% create a new node that is the parent of those siblings, on which these
-% mutations occurred. A pushed-reversion node is inserted when an immediate
-% reversion occurs (Figure 1E), i.e. a mutation that reverts another mutation on
-% its parent node. Instead of retaining both the mutation and immediate
-% reversion, only the mutation is kept after a pushed-reversion node is inserted,
-% thereby reducing the number of mutations on the branches. Note that these
-% non-sample nodes are used as reconstructed genomes during HMM copying.
-
-% Furthermore, during inference, the ARG is automatically encoded, stored, and
-% updated as a tree sequence using tskit \citep{Kelleher2018-xc}.
-
-\subsection{Ancestral Recombination Graphs}
-The term ``Ancestral Recombination Graph'' was introduced by
-Griffiths and colleagues~\citep{Griffiths1991-two,Griffiths1998-ancestral}
-and originally defined an alternative formulation of the coalescent
-with recombination stochastic process~\citep{Hudson1983-properties}.
-Subsequently, the term ARG came to be used in a more general way to
-describe not just realisations of this model, but to any
-recombinant genetic ancestry~\citep{Minichiello2006-mapping}.
-While there is some subtlety in the details~\citep{Wong2023-efficient},
-we can think of an ARG as being any graph that encodes the
-reticulate genetic ancestry of a sample of sequences under
-the influence of recombination.
-
-The ``succinct tree sequence'' is an ARG data structure
-that is both general (in terms of the types of ancestry that can
-be described) and computationally efficient~\citep{Wong2023-efficient}.
-Originally developed to facilitate large-scale coalescent
-simulations~\citep{Kelleher2016-wk}, the methods have been
-extended and applied to forwards-time
-simulations~\citep{Kelleher2018-xc,Haller2018-tree},
-calculation of population genetics statistics~\citep{Ralph2020-efficiently}
-and ARG inference~\citep{Kelleher2019-ba,Wohns2022-th}.
-The succinct tree sequence is based on a simple tabular representation,
-which defines a set of nodes, edges, sites and mutations. A \emph{node}
-represents a particular genome, which may be an observed sample
-or an inferred genetic ancestor. The genetic inheritance between
-a pair of nodes along a segment of genome is defined by
-the \emph{edge} $(\ell, r, p, c)$, which states that
-child node $c$ inherited its genome from parent node $p$
-from left coordinate $\ell$ to right coordinate $r$. A \emph{site}
-defines the genomic position at which some variation occurs,
-and a \emph{mutation} defines the site, node and derived state
-that generates variation.
-Given these tables, we can then efficiently construct the local
-genealogical trees along the genome (arising from recombination),
-and also perform a range of calculations efficiently by
-reasoning about the differences between these local
-trees~\citep{Kelleher2016-wk,Ralph2020-efficiently}. These
-algorithms have led to peformance increases of several orders
-of magnitude over the state-of-the-art in a range of
-applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,
-Ralph2020-efficiently,Baumdicker2022-ep}.
-The succinct tree sequence encoding
-is also very concise, allowing, for example, for millions of
-complete human genomes to potentially be stored in a few gigabytes of
-space~\citep{Kelleher2019-ba}.
-
-The \texttt{tskit} library~\citep{Tskit2023-tskit} is a liberally
-licensed open source toolkit that provides a comprehensive suite
-of tools for working with ARGs. Based on core functionality written
-in C, it provides interfaces in C, Python and Rust. The Python interface
-is based on NumPy~\citep{Harris2020-array}, and provides a convenient
-platform for interactive analysis of large-scale data using, for
-example, Jupyter notebooks~\citep{Kluyver2016-jupyter} and taking
-advantage of the analysis tools in the burgeoning PyData ecosystem.
-Tskit is mature software, widely used in population genetics, and
-has been incorporated into several downstream
-applications~\citep[e.g.][]{Haller2019-slim,Speidel2019-yh,
-Terasaki2021-geonomics,
-Fan2022-genealogical,Korfmann2022-weak,
-Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}.
-
-\subsection{The Li and Stephens model}
-The Li and Stephens (LS) model\citep{Li2003-ib}  is an approximation of the
-coalescent with recombination~\citep{Hudson1983-properties}, which captures
-many of the important features of the joint processes of mutation and
-recombination. It is a Hidden Markov Model (HMM) in which a focal genome is a
-sequence of nucleotides that are probabilistically emitted as
-an imperfect mosaic of a set of reference genomes
-(Figure~\ref{fig:ls_diagram}).
-The LS model is used in a wide variety
-of applications in genomics, including modern approaches to
-statistical genotype phasing and imputation
-\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa},
-and estimation of parameters such as
-recombination rates \citep[e.g.][]{Hinch2011-tz},
-and selection within and across hosts in viral
-sequence data \citep[e.g.][]{Palmer2019-wa}.
-See~\cite{Mcvean2019-linkage} for further review and discussion
-of the LS model.
-
-\begin{figure} \centering
-\includegraphics[width=0.5\textwidth]{figures/ls_diagram_covid.pdf}
-% FIXME this caption is too long
-\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
-model, in which a query sequence (bottom) is described as an
-imperfect mosaic of the sequences in a reference panel.
-Black crosses along the focal sequence show genotyping
-errors or mutations.
-In the standard formulation, at site $\ell$, the recombination probability is $r_\ell$,
-the mutation probability is $\mu_\ell$ and $n$
-denotes the size of the reference panel.
-The Viterbi algorithm can be used to find a
-``copying path'' through the reference panel for a given query sequence that
-maximises the likelihood under these parameters.
-% The LS HMM is governed by the transition and
-% emission matrices, $Q$ and $E$, respectively.
-% Each may be a function of the
-% reference panel members, but transitions are generally assumed to be
-% independent of the hidden states themselves (shown in the pink panel). This
-% dramatically increases performance as the state space drops to two states
-% (i.e., switching or not switching).
-% Emissions may also be a function of the nucleotide states, but in our
-% case, we assume that all observed alleles (in both the reference panel and
-% focal sequence, $a\ell$) at the site under consideration are equally likely and
-% occur with mutation probability $\ell$ (shown in the blue panel).
-% These two
-% processes, transition and emission (which encode recombination and mutation,
-% respectively), define the generative process of the HMM, by which an imperfect
-% mosaic is emitted from a reference panel.
- % of samples and nodes from the ARG
-% inserted at earlier collection dates (shown in the lower panel). Here, the
-% unseen states in the reference panel are shown as coloured lines enclosed by
-% the grey box, and the true path through the data which leads to the emitted
-% focal sequence is displayed by the black arrow. Examples of transition and
-% emission probabilities along this trajectory are shown by the red and blue
-% arrows, respectively.
-% Note that in practice, we do not know the pattern of
-% colours along the focal sequence and must infer it using the Viterbi algorithm.
-}
-\end{figure}
-
-The generative process of the LS model is summarised in
-Figure~\ref{fig:ls_diagram}. Here, a transition matrix, $Q$, governs the
-process of switching (recombining) between members of the reference panel (the
-hidden states). An emission matrix, $E$, allows for differences between the
-focal sequence and the hidden state from which it is copied (due to mutation as
-well as genotyping or sequencing error). Then, we use the Viterbi algorithm
-\citep{Viterbi1967-ol} to find the most likely copying path through the
-sequences, given a set of reference sequences, $Q$, and $E$.
-The relative rates of mutation and recombination are usually controlled by
-the site-specific parameters $\mu_\ell$ and $r_\ell$, respectively.
-For this work we used a simpler formulation, which uses a single parameter
-(the mismatch ratio, MMR) to control the relative importance of mutation
-and recombination in the HMM. Specifically, an MMR value of $k$ will
-prefer $k$ mismatches (mutations) to a recombination.[DUNCAN MORE DETAILS
-HERE PLEASE]
-
-% OLD text, ready to be adapted (and deleted once done)
-% For this
-% application, we have parameterised the HMM using a mismatch ratio (MMR), which
-% tunes how strongly mutations (i.e., no switching of template sequences) are
-% preferred over recombination (i.e., switching between sequences) during
-% copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
-% focal genome more likely spans at least two ancestral genomes (i.e,. the focal
-% genome is a putative recombinant) than a single ancestral genome. Here, the MMR
-% was set to three in all the HMM runs .
-
-In \texttt{sc2ts} we use the efficient ARG-based implementation of the
-LS Viterbi algorithm from \texttt{tsinfer}~\citep{Kelleher2019-ba} to find
-the most likely copying path for each sample sequence
-among all sequences (sampled and inferred) in the current ARG.
-In the majority of cases, with non-recombinant sample sequences,
-the most likely solution is to copy from one
-of the nodes in the ARG that minimised the number of mutations required
-on the query sequence. Importantly, because the reference panel here consists of
-\emph{every} node in the ARG, we can match to both older sample
-sequences or internal nodes representing an inferred ancestral sequence
-(see subsequent subsections for details about how these are added).
-Thus, when no recombination is present, the LS Viterbi algorithm is
-implementing a version of parsimony, in which we are guaranteed to
-find a sequence that minimises the number of additional mutations required
-to incorporate a newly-added sample into the ARG.
-Recombination is then inferred when the most likely solution to the LS
-HMM is to copy from more than one ARG node along the genome, for a
-given sample sequece. [TODO MORE]
-
-We can solve this massive optimisation problem exactly because the ARG-based
-implementation of the LS HMM used in \texttt{tsinfer} scales approximately
-logarithmically with reference panel size (as opposed to linearly,
-for standard matrix-based approaches)~\citep{Kelleher2019-ba}.
-This efficient HMM algorithm is the main reason for \texttt{tsinfer}'s
-scalability, and here allows us to find closely matching
-sequences and recombination paths among millions of SARS-CoV-2
-genomes exactly under a well-defined statistical model.
-
-It is important to note that the Viterbi algorithm only returns
-\emph{one of} the copying paths that maximise the likelihood under
- the given mutation and recombination parameters. There may be many
-such paths, from which we choose one arbitrarily. Also, the
-present choice of using a single mismatch-ratio parameter to control
-the likelihood of recombination vs mutation may lead to relatively
-flat likelihood spaces where many different paths have equal likelihood.
-There are many possibilities in using established HMM methodology
-to reason about and explore the space of possible matches, which may
-be a fruitful avenue for future work. See the Discussion for further
-potential improvements that may be made to the HMM.
-
-\subsection{Tree inference from clusters of samples}
-For a given daily batch of sample sequences that are to be attached to the
-ARG, the LS HMM provides the most likely copying path and required mutations
-for each of those sequences. We could directly attach these samples to
-the ARG according to the copying paths and sets of mutations, but
-such an approach would be highly unparsimonious because there are
-often many samples that share the same copying path with
-complex patterns of shared mutations.
-In this section we discuss the strategy used to infer likely tree
-toplogies among the samples that share the same copying path.
-In the majority of cases this copying path will consist of a single
-node (no recombination), and so for simplicity we will assume in the
-following discussion that all of the samples in a HMM-defined cluster
-attach to a single node. The differences required to deal with
-newly-detected putative recombinants are discussed below
-in Section XXX.
-
-With tens of thousands of samples being added to the ARG per day,
-such clusters of sequences attaching to the same node
-can be large, with complex patterns of shared mutations.
-To avoid the implied large polytomy and unparsimonious
-repetition of identical mutations among sibling nodes, we
-infer a tree among these samples (the LS HMM has already
-
-A natural
-way to
-
-We can think of the sequences attaching to a given node in the ARG
-from a daily batch as a polytomy,
-but with potentially shared mutations among the sibling nodes
-(Figure~\ref{fig:overview_sc2ts}C).
-
-\subsection{Parsimony updates}
-[Discuss the parsimony tweaks we make to the trees. Get a better name
-for the subsection title]
-
-\subsection{Treatment of recombinants}
-[Discuss the special treatment we give to recombinants.]
-Include the discussion about mirrored coords from here :]
-
-% \subsubsection{Breakpoint intervals and HMM consistency} When recombinant
-% ancestry for a given genome is inferred by the LS HMM, a breakpoint is
-% identified as the last possible position where the copying path must switch.
-% Thus, the breakpoints used in the ARG represent the rightmost end of the
-% interval, within which the recombination event could have occurred. To obtain
-% the leftmost end of this interval, we rerun the HMM using mirrored coordinates
-% (i.e., from the 3’ end to the 5’ end).
-
-% A recombinant sequence was deemed to be HMM-consistent if (1) the same number
-% of parent nodes and (2) the same set of mutations immediately ancestral to it
-% were obtained from both the unmirrored and mirrored HMM runs. In the ideal
-% case, the copying paths of both the HMM runs switch between parent sequences at
-% the same position(s); in less ideal cases, however, the copying paths switch at
-% different positions, typically because of a lack of genetic variation between
-% the positions. The switch positions in the copying paths of both the HMM runs
-% were used to define the breakpoint interval(s) of an HMM-consistent
-% recombinant.
-
-\subsection{Data preprocessing}
-
-The findings of this study are
-based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
-up to Aug 22, 2022 and accessible at
-% FIXME
-doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
-if they (1) had ambiguous collection dates; (2) were collected before January
-01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
-Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
-MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
-2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
-inference, sequences were excluded if they had a ``bad'' quality control status
-in any of the four Nextclade columns (``qc.missingData.status'',
-``qc.mixedSites.status'', ``qc.frameShifts.status'' and
-`qc.stopCodons.status''). These excluded genomes are enriched for low-coverage,
-contaminated, or misassembled sequences.
-
-Ambiguous nucleotide letters (i.e.,
-not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
-Problematic bases in the alignments, which had two or more ambiguous letters or
-gaps within a distance of seven bases, were masked as missing data following
-the protocol in \url{https://github.com/roblanf/sarscov2phylo/}. Masked bases
-were ignored when attaching a sample genome to the ARG, but they were
-subsequently imputed. Further, 481 problematic sites flagged as prone to
-sequencing errors or highly homoplasic were excluded entirely
-% Do we need this first URL?
-(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
-\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
-
-\subsection{Exclusion of ``time travellers''}
-[FIXME I'm pretty sure we don't have two filters here, we just use
-maximum sampling delay. Also not sure the MSD acronym is worth it,
-we don't use it much.]
-
-Including samples with incorrectly reported collection dates (e.g., an Alpha
-sample purportedly collected in 2020 from the United States, before Alpha
-appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
-artefacts that distorted the topology of ARGs inferred early on in the study.
-Hence, to exclude such potential ``time travellers'' we employed two filters.
-The first was to remove any sequence with a collection date that pre-dated the
-time to the most recent common ancestor (tMRCA) of its corresponding clade. The
-tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
-(downloaded on August 22, 2022) where we used the lower bound of the 95\%
-confidence interval of each clade as the minimum date cut-off. The second
-filter was a simple threshold on the time delay between the collection date and
-submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
-inferred from the MSD-filtered genome set had noticeably fewer branches with a
-high number of reversions. After some preliminary analysis, a MSD threshold
-value of 30 days was used when building the ARGs described herein. Future work
-could estimate collection dates by adding these potential time travellers back
-on the tree, allowing an automated assessment of collection date discrepancies.
-
-\subsection{Imputation of Pango lineage for non-sample nodes}
-Initially, only sample nodes have Pango lineage assignments.
-For analysis of recombinant parents, it is also helpful to assign
-Pango lineages to non-sample nodes. We do this
-using a two-step procedure (Figure~\ref{fig:imputation}). If the Pango
-lineage of the parental node was available and no mutation was immediately
-ancestral to the non-sample node, then the parental node’s status was directly
-propagated to the non-sample node. Otherwise, the Pango lineage of the
-non-sample node was inferred by matching its full set of mutations against the
-Pango lineage-defining mutations (based on 90\% consensus of the sequences
-analysed) from the COVID-CG website \citep{Chen2021-zc};
-\url{https://covidcg.org/}; accessed on November 04, 2022).
-
-\begin{figure} \centering
-\includegraphics[width=.7\textwidth]{figures/imputation.pdf}
-\caption{\label{fig:imputation}A schematic of the two-step procedure to impute
-Pango lineage for inserted, non-sample nodes. Here, three nodes (question
-marks) have unknown Pango lineage (A). Because the parent of node 1 is
-available (i.e., L2) and there is no mutation occurring on the edge connecting
-the two nodes, the Pango lineage of the parent is directly propagated to node 1
-(B). However, if there is a mutation (red X’s), then the Pango lineage is
-inferred by comparison to the set of lineage-defining mutations from
-COVID-CG\citep{Chen2021-zc}, as in case (C). Finally, the inferred status is
-propagated to any non-mutant child nodes, as in case (D).
-}
-\end{figure}
 
@@
 \subsection{Backbone phylogeny}
-For validation purposes, we compare the genetic relationships in the sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain project \citep{Hadfield2018-ef}.
-The NextStrain trees consist of subsamples of SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely restricted to non-recombinant lineages.
-To allow for comparison, the ARGs were ``simplified'' \citep{Kelleher2018-xc} to retain only those samples of our genealogy that match those present in a specific NextStrain tree.
-These simplified ``backbone'' ARGs exhibit very little recombination (6 and 5 trees across the genome for the wide and long ARGs, respectively).
+For validation purposes, we compare the genetic
+relationships in the sc2ts-inferred ARGs with phylogenetic relationships
+inferred by the NextStrain project \citep{Hadfield2018-ef}. The NextStrain
+trees consist of subsamples of SARS-CoV-2 genomes, which over the time period
+of our ARGs are almost entirely restricted to non-recombinant lineages. To
+allow for comparison, the ARGs were ``simplified'' \citep{Kelleher2018-xc} to
+retain only those samples of our genealogy that match those present in a
+specific NextStrain tree. These simplified ``backbone'' ARGs exhibit very
+little recombination (6 and 5 trees across the genome for the wide and long
+ARGs, respectively).
 
@@
 \includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
-\caption{\label{fig:cophylogeny} Tanglegram comparing the first local tree in the wide ARG (sampled to mid-2021) and an “all-time” global NextStrain tree (sampled to January 21, 2023).
-The trees were pruned down to the samples present in both the trees.
-Light grey lines match the corresponding samples between the two trees; black circles indicate identical sample partitions between the two trees.
-Terminal branches are colour-coded according to the Pango lineage status assigned to the tip samples; no Pango recombinants are shown, as none are shared between the trees.
-The tanglegram was generated with the aid of the Neighbor-Net algorithm \citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5 \citep{Huson2012-ys}.
-See Figure~\ref{fig:cophylogeny_long} for the equivalent cophylogeny for the Long ARG.}
+\caption{\label{fig:cophylogeny}
+Tanglegram comparing the first local tree in the wide ARG (sampled to mid-2021)
+and an ``all-time'' global NextStrain tree (sampled to January 21, 2023). The
+trees were pruned down to the samples present in both the trees. Light grey
+lines match the corresponding samples between the two trees; black circles
+indicate identical sample partitions between the two trees. Terminal branches
+are colour-coded according to the Pango lineage status assigned to the tip
+samples; no Pango recombinants are shown, as none are shared between the trees.
+The tanglegram was generated with the aid of the Neighbor-Net algorithm
+\citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5
+\citep{Huson2012-ys}. See Figure~\ref{fig:cophylogeny_long} for the equivalent
+cophylogeny for the Long ARG.}
 \end{figure}
 
-Figure~\ref{fig:cophylogeny} shows a co-phylogeny of the 88 samples shared between the wide ARG and a GISAID global "all-time" tree from NextStrain (downloaded on January 21, 2023).
-To reduce the ARG to a single tree, we show the very first tree in the wide ARG; however, the five other trees in this simplified ARG show near-identical patterns.
-Supplementary Figure~\ref{fig:cophylogeny_long} shows the same comparison for the long ARG.
-It is clear that the backbone topology of the sc2ts tree shows very close agreement with the NextStrain tree.
-The sample genomes cluster by their assigned Pango lineage status, and many variants and their descendents form identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC clades, labelled).
-
-Figure~\ref{fig:cophylogeny} reveals some notable differences between the trees.
-Firstly, the sc2ts tree is generally less well resolved (noticeably, February to March, 2020).
-Secondly, there are non-identical sample partitions near the tips (e.g., in the Alpha clade).
-These might have arisen because the samples containing the mutations needed to resolve local topologies were collected later than the cut-off date for the sc2ts ARG, but were for the NextStrain tree.
-Finally, the branch lengths differ.
-For example, the branch leading to the ancestral node of the Delta variant samples is much shorter in the sc2ts tree than the equivalent NextStrain tree.
-It is possible that the length of some of the branches was poorly calibrated, probably caused by overly aggressive insertion of mutation-collapsing nodes and reversion-pushed nodes or suboptimal sample attachment.
-Side-by-side comparison of sc2ts trees with trees inferred using another method (e.g., UShER) should help to understand whether branch length differences are due to poor calibration or methodological differences.
-
-Although the simplified backbone ARGs have few local trees, all of which show essentially the same topologies, note that in the original ``un-simplified'' wide and long ARGs, there are 1,496 and 958 local trees respectively (Table 1).
-Hence, while for a relatively small subset of the genealogy may be well described by a single tree, the full dataset shows substantial variation in phylogenetic relationships (i.e.
-``phylogenetic incongruence'') across the SARS-CoV-2 genome, much of which may be attributable to recombination.
-An ARG representation is therefore essential to fully capture these reticulate genealogical relationships.
+Figure~\ref{fig:cophylogeny} shows a co-phylogeny of the 88 samples shared
+between the wide ARG and a GISAID global "all-time" tree from NextStrain
+(downloaded on January 21, 2023). To reduce the ARG to a single tree, we show
+the very first tree in the wide ARG; however, the five other trees in this
+simplified ARG show near-identical patterns. Supplementary
+Figure~\ref{fig:cophylogeny_long} shows the same comparison for the long ARG.
+It is clear that the backbone topology of the sc2ts tree shows very close
+agreement with the NextStrain tree. The sample genomes cluster by their
+assigned Pango lineage status, and many variants and their descendents form
+identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
+clades, labelled).
+
+Figure~\ref{fig:cophylogeny} reveals some notable differences between the
+trees. Firstly, the sc2ts tree is generally less well resolved (noticeably,
+February to March, 2020). Secondly, there are non-identical sample partitions
+near the tips (e.g., in the Alpha clade). These might have arisen because the
+samples containing the mutations needed to resolve local topologies were
+collected later than the cut-off date for the sc2ts ARG, but were for the
+NextStrain tree. Finally, the branch lengths differ. For example, the branch
+leading to the ancestral node of the Delta variant samples is much shorter in
+the sc2ts tree than the equivalent NextStrain tree. It is possible that the
+length of some of the branches was poorly calibrated, probably caused by overly
+aggressive insertion of mutation-collapsing nodes and reversion-pushed nodes or
+suboptimal sample attachment. Side-by-side comparison of sc2ts trees with trees
+inferred using another method (e.g., UShER) should help to understand whether
+branch length differences are due to poor calibration or methodological
+differences.
+
+Although the simplified backbone ARGs have few local trees, all of which show
+essentially the same topologies, note that in the original ``un-simplified''
+wide and long ARGs, there are 1,496 and 958 local trees respectively (Table 1).
+Hence, while for a relatively small subset of the genealogy may be well
+described by a single tree, the full dataset shows substantial variation in
+phylogenetic relationships (i.e. ``phylogenetic incongruence'') across the
+SARS-CoV-2 genome, much of which may be attributable to recombination. An ARG
+representation is therefore essential to fully capture these reticulate
+genealogical relationships.
 
 \subsection{Mutational spectrum}
- In addition to recombination events, the ARGs record mutation events along the SARS-CoV-2 phylogeny.
-\cite{Yi2021-sc} reconstructed a SARS-CoV-2 phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to January 12, 2021, and systematically catalogued all the mutations occurring along the phylogeny.
-Notably, they described a mutational spectrum in which C-to-U and G-to-U mutations occur more frequently than U-to-C and U-to-G mutations, respectively.
+ In addition to recombination events, the ARGs record mutation events along the
+SARS-CoV-2 phylogeny. \cite{Yi2021-sc} reconstructed a SARS-CoV-2 phylogeny of
+over 350,000 genomes sampled globally from December 24, 2019, to January 12,
+2021, and systematically catalogued all the mutations occurring along the
+phylogeny. Notably, they described a mutational spectrum in which C-to-U and
+G-to-U mutations occur more frequently than U-to-C and U-to-G mutations,
+respectively.
 
@@
 
-For validation purposes, we compare the mutational spectrum from the wide ARG with the mutational spectrum of \citet{Yi2021-sc}.
-We categorised all of the single nucleotide mutations by type (defined by the ancestral state and derived state), excluded mutations inherited by only a single sample (which are enriched for sequencing errors and artefacts), and tallied them up by type.
-Similarly, we took the data for single nucleotide mutations from \url{https://github.com/ju-lab/SC2_evol_signature} \citep{Yi2021-sc}, excluded mutations occurring along terminal branches, and tallied them up by type.
-Figure~\ref{fig:mutational_spectra} shows that the mutational spectrum from the wide ARG (based on 448,825 mutations) matches that reported by \citet[based on 92,344 mutations]{Yi2021-sc}.
-In both spectra, C-to-U mutations and G-to-U mutations occur more frequently than U-to-C and U-to-G, respectively.
-Similar results are obtained when including the mutations inherited by only a single sample or those occurring on terminal branches (data not shown).
+For validation purposes, we compare the mutational spectrum from the wide ARG
+with the mutational spectrum of \citet{Yi2021-sc}. We categorised all of the
+single nucleotide mutations by type (defined by the ancestral state and derived
+state), excluded mutations inherited by only a single sample (which are
+enriched for sequencing errors and artefacts), and tallied them up by type.
+Similarly, we took the data for single nucleotide mutations from
+\url{https://github.com/ju-lab/SC2_evol_signature} \citep{Yi2021-sc}, excluded
+mutations occurring along terminal branches, and tallied them up by type.
+Figure~\ref{fig:mutational_spectra} shows that the mutational spectrum from the
+wide ARG (based on 448,825 mutations) matches that reported by \citet[based on
+92,344 mutations]{Yi2021-sc}. In both spectra, C-to-U mutations and G-to-U
+mutations occur more frequently than U-to-C and U-to-G, respectively. Similar
+results are obtained when including the mutations inherited by only a single
+sample or those occurring on terminal branches (data not shown).
 
@@
 
-By searching for samples combining genomic segments from Alpha (B.1.1.7) and from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations, \citet{Jackson2021-ik} reported 16 recombinant sequences among genomes sampled in the United Kingdom up to March 7, 2021.
-The authors proposed that the recombinants likely arose from eight independent recombination events (categorised into groups A to D and four singletons).
+By searching for samples combining genomic segments from Alpha (B.1.1.7) and
+from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations,
+\citet{Jackson2021-ik} reported 16 recombinant sequences among genomes sampled
+in the United Kingdom up to March 7, 2021. The authors proposed that the
+recombinants likely arose from eight independent recombination events
+(categorised into groups A to D and four singletons).
 
@@
 
-In the wide ARG (sampled to mid-2021), all these sequences are descendants of recombination nodes, except MILK-103C712, which was removed during preprocessing (Table 2); additionally, all are HMM-consistent, except QEUH-1067DEF.
-Overall, we find excellent agreement between Jackson et al.’s motif-based method and sc2ts in terms of (1) the Pango lineage of the parents and (2) the breakpoint intervals (Table 2).
-In all cases, the two methods proposed identical or very closely related parent lineages.
-The general mosaic genome structure is concordant in all the cases, even CAMC-CB7AB3 where there are two switches in the HMM copying paths (indicative of two breakpoints).
-The breakpoint intervals estimated using the two methods overlap substantially in all the cases (Table 2).
-Within each of the groups A to D, the constituent sequences have identical breakpoint intervals, because they trace back to a single recombination node in the wide ARG.
-This result is consistent with the hypothesis that the sequences of the groups A to D emerged from four separate recombination events~\cite{Jackson2021-ik}.
+In the wide ARG (sampled to mid-2021), all these sequences are descendants of
+recombination nodes, except MILK-103C712, which was removed during
+preprocessing (Table 2); additionally, all are HMM-consistent, except
+QEUH-1067DEF. Overall, we find excellent agreement between Jackson et al.’s
+motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
+and (2) the breakpoint intervals (Table 2). In all cases, the two methods
+proposed identical or very closely related parent lineages. The general mosaic
+genome structure is concordant in all the cases, even CAMC-CB7AB3 where there
+are two switches in the HMM copying paths (indicative of two breakpoints). The
+breakpoint intervals estimated using the two methods overlap substantially in
+all the cases (Table 2). Within each of the groups A to D, the constituent
+sequences have identical breakpoint intervals, because they trace back to a
+single recombination node in the wide ARG. This result is consistent with the
+hypothesis that the sequences of the groups A to D emerged from four separate
+recombination events~\cite{Jackson2021-ik}.
 
@@
 
-We explore recombination signals across the SARS-CoV-2 genome by inspecting the breakpoint intervals of the filtered recombination nodes from the wide ARG (sampled to mid-2021) in Figure~\ref{fig:breakpoint-distribution}.
-The length of a breakpoint interval reflects uncertainty in the estimated location of a breakpoint.
-On average, the length of the breakpoint intervals is 2,895 bases (median, 1,670 bases; range, 1 to 23,666 bases).
-This broad distribution of interval length indicates that the location of a breakpoint often could not be precisely determined given the observed genetic variation.
+We explore recombination signals across the SARS-CoV-2 genome by inspecting the
+breakpoint intervals of the filtered recombination nodes from the wide ARG
+(sampled to mid-2021) in Figure~\ref{fig:breakpoint-distribution}. The length
+of a breakpoint interval reflects uncertainty in the estimated location of a
+breakpoint. On average, the length of the breakpoint intervals is 2,895 bases
+(median, 1,670 bases; range, 1 to 23,666 bases). This broad distribution of
+interval length indicates that the location of a breakpoint often could not be
+precisely determined given the observed genetic variation.
 
@@
 
-There is a preponderance of breakpoint intervals spanning the 3’ end of the genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}.
-Inspecting more closely, we find an enrichment in the number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure 4B).
-Markedly, there is an abrupt drop in the number of intersecting intervals in the N gene following the peak in ORF8 towards the 3’ end.
-
-Next, we investigate the potential recombination signals alongside the number of mutations across the genome (Figure 4C).
-Regions with a higher density of mutations contain more genetic variation, increasing the power to detect recombination events.
-The position 28,271 (located in the ORF8-N intergenic region) has a particularly high number of mutations in the samples (Figure 5).
-At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions (n = 1,605) appear to occur frequently.
-One possible explanation is that this homopolymeric region is prone to sequencing errors, giving rise to an artifactual recombination signal.
-At the upstream position 27,972, numerous mutations also occur.
-This position has been reported to be hypermutable
+There is a preponderance of breakpoint intervals spanning the 3’ end of the
+genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}. Inspecting
+more closely, we find an enrichment in the number of intersecting intervals in
+the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure
+4B). Markedly, there is an abrupt drop in the number of intersecting intervals
+in the N gene following the peak in ORF8 towards the 3’ end.
+
+Next, we investigate the potential recombination signals alongside the number
+of mutations across the genome (Figure 4C). Regions with a higher density of
+mutations contain more genetic variation, increasing the power to detect
+recombination events. The position 28,271 (located in the ORF8-N intergenic
+region) has a particularly high number of mutations in the samples (Figure 5).
+At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions
+(n = 1,605) appear to occur frequently. One possible explanation is that this
+homopolymeric region is prone to sequencing errors, giving rise to an
+artifactual recombination signal. At the upstream position 27,972, numerous
+mutations also occur. This position has been reported to be hypermutable
 \citep{Jungreis2021-dh}.
 
-We further consider the breakpoint intervals within the context of recombination breakpoint sequence motifs.
-It is hypothesised that certain palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote template switching during replication in SARS-CoV-2 via formation of base-paired stem loops in the genome structure \citep{Gallaher2020-lb}.
-There are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019 reference sequence (Supplementary Table 5).
-We observe that XX (XX\%) of the breakpoint intervals of the HMM-consistent recombinants span at least one of the breakpoint sequences.
+We further consider the breakpoint intervals within the context of
+recombination breakpoint sequence motifs. It is hypothesised that certain
+palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote
+template switching during replication in SARS-CoV-2 via formation of
+base-paired stem loops in the genome structure \citep{Gallaher2020-lb}. There
+are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019
+reference sequence (Supplementary Table 5). We observe that XX (XX\%) of the
+breakpoint intervals of the HMM-consistent recombinants span at least one of
+the breakpoint sequences.
 
@@
 \includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
-\caption{\label{fig:recomb_mrcas} Date of common ancestry between the parents either side of recombination breakpoints, as a function of degree of parental divergence.
-MRCAs of parents inferred to have produced a Pango designated recombinant (XA, XB, etc.) are identified in orange (A) Data after removing singleton recombinants (XXX parent pairs).
-Many recombinations involve parental strains that descend from the same MRCA: dates for the four most frequently identified MRCA nodes are indicated by horizontal dotted lines.
-These frequent MRCA nodes tend to be associated with major waves and commonly have many immediate children (i.e.
-are large polytomies, likely to indicate undersampled outbreaks) (B) Further filtered by removing HMM inconsistent breakpoints (YYY parent pairs).
-Note that Figure~\ref{fig:recomb_mrcas_voc_breakdown} shows the same data broken down by parental VoC classification.}
+\caption{\label{fig:recomb_mrcas}
+Date of common ancestry between the parents either side of recombination
+breakpoints, as a function of degree of parental divergence. MRCAs of parents
+inferred to have produced a Pango designated recombinant (XA, XB, etc.) are
+identified in orange (A) Data after removing singleton recombinants (XXX parent
+pairs). Many recombinations involve parental strains that descend from the same
+MRCA: dates for the four most frequently identified MRCA nodes are indicated by
+horizontal dotted lines. These frequent MRCA nodes tend to be associated with
+major waves and commonly have many immediate children (i.e. are large
+polytomies, likely to indicate undersampled outbreaks) (B) Further filtered by
+removing HMM inconsistent breakpoints (YYY parent pairs). Note that
+Figure~\ref{fig:recomb_mrcas_voc_breakdown} shows the same data broken down by
+parental VoC classification.}
 \end{figure}
 
-Both the wide and long ARGs contain a large number of recombination nodes (4123 and 2078 respectively; Supplementary Tables 1 and 2).
-Each of these nodes were inferred as a result of matching a specific recombinant sample into the ARG.
-However, ~63\% of recombination nodes are ancestral to only a single recombinant sample; we exclude these "singleton recombinants" from subsequent analyses as they may have arisen from sequencing of mixed samples (e.g., from co-infection or contamination), and in any case, never result in onward transmission.
-This leaves 1522 and 763 recombination nodes in the wide and long ARGs, respectively.
-In their analysis using RIPPLES, \cite{Turakhia2022-it} detected 589 non-singleton recombinants.
-
-We focus on the long ARG because compared to the wide ARG, it includes time periods of increased rates of coinfection and recombination, and contains many samples previously identified as the product of recombination (i.e.
-with a Pango lineage designation starting with an "X").
-Not only does the long ARG capture all but one of the Pango designated recombinant lineages during this time period (see the section below), but it also includes a large number of non-Pango-identified recombinants.
-Many of these involve recombination between relatively closely related lineages, as measured by the divergence time between the recombination events and the most recent common ancestor (MRCA) shared by  the left and right parent at each breakpoint.
-As can be seen from Figure~\ref{fig:recomb_mrcas}, this divergence time ranges from as little as five or six days, to as much as several years in the case of recent recombinants.
-This demonstrates that sc2ts can identify viral recombination between very closely related lineages, providing a degree of resolution not available with other recombination detection methods.
+Both the wide and long ARGs contain a large number of recombination nodes (4123
+and 2078 respectively; Supplementary Tables 1 and 2). Each of these nodes were
+inferred as a result of matching a specific recombinant sample into the ARG.
+However, ~63\% of recombination nodes are ancestral to only a single
+recombinant sample; we exclude these "singleton recombinants" from subsequent
+analyses as they may have arisen from sequencing of mixed samples (e.g., from
+co-infection or contamination), and in any case, never result in onward
+transmission.  This leaves 1522 and 763 recombination nodes in the wide and
+long ARGs, respectively. In their analysis using RIPPLES,
+\cite{Turakhia2022-it} detected 589 non-singleton recombinants.
+
+We focus on the long ARG because compared to the wide ARG, it includes time
+periods of increased rates of coinfection and recombination, and contains many
+samples previously identified as the product of recombination (i.e. with a
+Pango lineage designation starting with an "X"). Not only does the long ARG
+capture all but one of the Pango designated recombinant lineages during this
+time period (see the section below), but it also includes a large number of
+non-Pango-identified recombinants. Many of these involve recombination between
+relatively closely related lineages, as measured by the divergence time between
+the recombination events and the most recent common ancestor (MRCA) shared by
+the left and right parent at each breakpoint. As can be seen from
+Figure~\ref{fig:recomb_mrcas}, this divergence time ranges from as little as
+five or six days, to as much as several years in the case of recent
+recombinants. This demonstrates that sc2ts can identify viral recombination
+between very closely related lineages, providing a degree of resolution not
+available with other recombination detection methods.
 
@@
 
-As a check on the reliability of these identified recombination events, Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are "HMM-consistent"; that is, where the mirrored HMM runs (i.e., 5’ to 3’ end versus 3’ to 5’ end) identify the same set of originating mutations and the same number of breakpoints, and the parents in the mirrored runs are sequence-identical.
-This reduces both the number of recombination nodes / breakpoints from 1522 / B to C / D in the wide ARG and from 763 / 806 to 459 in the long ARG (Supplementary Tables 1 and 2).
-
-The date of common ancestry for the two parents of recombinant lineages is concentrated in several banded rows in Figure~\ref{fig:recomb_mrcas}.
-These are largely due to a small handful of MRCAs which are shared between a large number of recombination events; the top four are indicated by dotted lines in the figure.
-These shared MRCAs are near the root of large expansions of certain lineages (B.1: original variant;  B.1.617.2: Delta, BA.1: first Omicron wave; and BA.2: second Omicron wave), as evidenced by the number of samples with that Pango designation that trace back to these ancestral nodes (listed as percentages on the plot).
-The specific MRCA nodes to which many recombinants trace frequently represent very large polytomies in the tree (at the extreme, 2,872 immediate children).
-This is likely to indicate a rapid and under-sampled expansion of a clade in the SARS-CoV-2 genealogy.
-
-Finally, it is clear from Supplementary Figure 4 that of the recombinant parents that fall into the Alpha, Delta, and Omicron VoC classes, the majority are within-Delta and within-Omicron recombinants, and that Omicron followed by Delta, are the variants associated with the most recombination.
-This may reflect either sampling intensity, the prevalence of cases (which increases the chance of coinfection and recombination), or possibly heterogeneity in recombination probabilities among lineages.
+As a check on the reliability of these identified recombination events,
+Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are
+"HMM-consistent"; that is, where the mirrored HMM runs (i.e., 5’ to 3’ end
+versus 3’ to 5’ end) identify the same set of originating mutations and the
+same number of breakpoints, and the parents in the mirrored runs are
+sequence-identical. This reduces both the number of recombination nodes /
+breakpoints from 1522 / B to C / D in the wide ARG and from 763 / 806 to 459 in
+the long ARG (Supplementary Tables 1 and 2).
+
+The date of common ancestry for the two parents of recombinant lineages is
+concentrated in several banded rows in Figure~\ref{fig:recomb_mrcas}. These are
+largely due to a small handful of MRCAs which are shared between a large number
+of recombination events; the top four are indicated by dotted lines in the
+figure. These shared MRCAs are near the root of large expansions of certain
+lineages (B.1: original variant;  B.1.617.2: Delta, BA.1: first Omicron wave;
+and BA.2: second Omicron wave), as evidenced by the number of samples with that
+Pango designation that trace back to these ancestral nodes (listed as
+percentages on the plot). The specific MRCA nodes to which many recombinants
+trace frequently represent very large polytomies in the tree (at the extreme,
+2,872 immediate children). This is likely to indicate a rapid and under-sampled
+expansion of a clade in the SARS-CoV-2 genealogy.
+
+Finally, it is clear from Supplementary Figure 4 that of the recombinant
+parents that fall into the Alpha, Delta, and Omicron VoC classes, the majority
+are within-Delta and within-Omicron recombinants, and that Omicron followed by
+Delta, are the variants associated with the most recombination. This may
+reflect either sampling intensity, the prevalence of cases (which increases the
+chance of coinfection and recombination), or possibly heterogeneity in
+recombination probabilities among lineages.
 
@@
 
-In this section we focus on the detailed genealogy of samples that have been previously identified as recombinants, i.e.
-given a Pango designation that starts with an ``X''.
-Since the wide ARG is restricted to data collected prior to mid-2021, it contains samples from only three Pango-designated recombinant lineages: XA, XB, and XC.
-To investigate the diversity of genealogies associated with recombinants, we therefore devote more analysis to patterns in the long ARG.
-As in previous sections, we exclude singleton recombinants where possible.
+In this section we focus on the detailed genealogy of samples that have been
+previously identified as recombinants, i.e. given a Pango designation that
+starts with an ``X''. Since the wide ARG is restricted to data collected prior
+to mid-2021, it contains samples from only three Pango-designated recombinant
+lineages: XA, XB, and XC. To investigate the diversity of genealogies
+associated with recombinants, we therefore devote more analysis to patterns in
+the long ARG. As in previous sections, we exclude singleton recombinants where
+possible.
 
@@
 [Quick overview of the wide ARG.]
-In the wide ARG, 44 samples are designated as XA by both Nextclade and GISAID, 237 are designated as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
-After removing singleton recombinants, XA numbers remain unchanged, but XB is reduced to 235 (229) and XC to 4 (0).
-We correctly identify that all samples designated as XB by any method have one or more recombination nodes in their ancestry.
-
-The simplest Pango recombinant genealogy is XA: all XA samples trace back to a unique originating recombination node.
-This is the product of a recombination between a B.1.177.18 sample (specifically the strain Wales/ALDP-115BF41/2021, EPI\_ISL: 1012804) which contributed the majority of the genome from the start to a maximum position of 22227bp, and an unknown (inserted) node with imputed Pango lineage B.1.1.7, which contributed the remaining right hand portion.
-The recombination node has five immediate children: four sample leaves (EPI\_ISLs: 989697, 1019487, 1104468 and 1122630) and a UPGMA-inserted node which is the ancestor of all other XA samples in the dataset.
-
-For XB, all samples trace back to a originating node which is the product of a recombination between a B.1 sample (specifically the strain England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of the genome from the start up to a maximum position of 23604bp), and two other UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of 27389bp) and B.1.36.8 (the remaining fragment of the genome).
-As described later, in the long ARG the equivalent recombination node has only 2 parents, with no involvement of  B.1.36.8, and it is possible that the third parent in the wide ARG is artifactual.
-Additional complications arise because additional non-X-designated samples such as B.1.634 also descend from this recombination (also true in the long ARG, see below), and there are also a small number of further recombination nodes nested within the XB ``clade''.
-However, all but one of these nested recombinations are ancestral to only a negligible fractions of the designated XB samples.
-The exception accounts for about 17\% of the designated XB nodes, and involves a recombination between descendants of the originating XB recombination.
-More specifically strain USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination between an XB grandparent and its XB grandchild.
-This could well be artifactual.
-
-For XC, the wide ARG identifies more than one originating recombination node (see "multiple origins" below).
-However, as none of the samples designated as XC by Nextclade are designated as XC by GISAID, these patterns could be due to mistaken labelling, hence we do not examine XC further.
+In the wide ARG, 44
+samples are designated as XA by both Nextclade and GISAID, 237 are designated
+as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
+After removing singleton recombinants, XA numbers remain unchanged, but XB is
+reduced to 235 (229) and XC to 4 (0). We correctly identify that all samples
+designated as XB by any method have one or more recombination nodes in their
+ancestry.
+
+The simplest Pango recombinant genealogy is XA: all XA samples trace back to a
+unique originating recombination node. This is the product of a recombination
+between a B.1.177.18 sample (specifically the strain Wales/ALDP-115BF41/2021,
+EPI\_ISL: 1012804) which contributed the majority of the genome from the start
+to a maximum position of 22227bp, and an unknown (inserted) node with imputed
+Pango lineage B.1.1.7, which contributed the remaining right hand portion. The
+recombination node has five immediate children: four sample leaves (EPI\_ISLs:
+989697, 1019487, 1104468 and 1122630) and a UPGMA-inserted node which is the
+ancestor of all other XA samples in the dataset.
+
+For XB, all samples trace back to a originating node which is the product of a
+recombination between a B.1 sample (specifically the strain
+England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of
+the genome from the start up to a maximum position of 23604bp), and two other
+UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of
+27389bp) and B.1.36.8 (the remaining fragment of the genome). As described
+later, in the long ARG the equivalent recombination node has only 2 parents,
+with no involvement of  B.1.36.8, and it is possible that the third parent in
+the wide ARG is artifactual. Additional complications arise because additional
+non-X-designated samples such as B.1.634 also descend from this recombination
+(also true in the long ARG, see below), and there are also a small number of
+further recombination nodes nested within the XB ``clade''. However, all but
+one of these nested recombinations are ancestral to only a negligible fractions
+of the designated XB samples. The exception accounts for about 17\% of the
+designated XB nodes, and involves a recombination between descendants of the
+originating XB recombination. More specifically strain
+USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination
+between an XB grandparent and its XB grandchild. This could well be
+artifactual.
+
+For XC, the wide ARG identifies more than one originating recombination node
+(see "multiple origins" below). However, as none of the samples designated as
+XC by Nextclade are designated as XC by GISAID, these patterns could be due to
+mistaken labelling, hence we do not examine XC further.
 
 \subsubsection{Long ARG}
-[high level summary of the pango recombinant lineages and samples in the Long ARG]
-The long ARG (subsampled to mid-2022) contains 33 Nextclade-designated Pango recombinant lineages (XA to XAK; Supplementary Table XXX) comprising 749 samples (711 when singleton recombinants are removed).
-If the GISAID designations are used, 38 Pango lineages exist in the long ARG, comprising only 515 samples.
-In the analyses below, we mainly focus on the larger number of Nextclade designated samples.
-
-Of the 749 Nextclade recombinant samples, all but 2 are inferred to descend from a recombination node.
-These two samples are designated XP, and a likely explanation for their not have an inferred recombinant origin in the long ARG is that the characteristic multibase deletion for XP (\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked during preprocessing.
-An improved masking pipeline is an important avenue for future development (see the Discussion).
-
-38 of the remaining X designated samples are singleton recombinants, descending from recombination nodes that are ancestral only to that sample (10 are XZ; 6 are XE;  3 each from XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1 sample from each of XB, XM, XJ, XAF, XAH and XAJ).
-Such samples are likely to be enriched for sequencing errors and lineage designation artefacts, and so we exclude them from further analysis in this section.
-
-The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE: 1, XAD: 1, XAH: 1, XAK: 1) is the same.
-This node is ancestral to more than 127K samples, and is likely to be a spuriously inferred recombination event.
-Understanding the origin of this and other similar nodes, is an important direction for future work (see Discussion).
-For simplicity, we exclude these samples from further analyses in this section.
-
-The remaining remaining 630 Pango X designated sequences trace back to 50 different recombination nodes. We examine the inferred inheritance structure in the following subsections.
+[high level summary of the pango recombinant lineages
+and samples in the Long ARG]
+The long ARG (subsampled to mid-2022) contains 33
+Nextclade-designated Pango recombinant lineages (XA to XAK; Supplementary Table
+XXX) comprising 749 samples (711 when singleton recombinants are removed). If
+the GISAID designations are used, 38 Pango lineages exist in the long ARG,
+comprising only 515 samples. In the analyses below, we mainly focus on the
+larger number of Nextclade designated samples.
+
+Of the 749 Nextclade recombinant samples, all but 2 are inferred to descend
+from a recombination node. These two samples are designated XP, and a likely
+explanation for their not have an inferred recombinant origin in the long ARG
+is that the characteristic multibase deletion for XP
+(\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
+during preprocessing. An improved masking pipeline is an important avenue for
+future development (see the Discussion).
+
+38 of the remaining X designated samples are singleton recombinants, descending
+from recombination nodes that are ancestral only to that sample (10 are XZ; 6
+are XE;  3 each from XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1
+sample from each of XB, XM, XJ, XAF, XAH and XAJ). Such samples are likely to
+be enriched for sequencing errors and lineage designation artefacts, and so we
+exclude them from further analysis in this section.
+
+The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
+1, XAD: 1, XAH: 1, XAK: 1) is the same. This node is ancestral to more than
+127K samples, and is likely to be a spuriously inferred recombination event.
+Understanding the origin of this and other similar nodes, is an important
+direction for future work (see Discussion). For simplicity, we exclude these
+samples from further analyses in this section.
+
+The remaining remaining 630 Pango X designated sequences trace back to 50
+different recombination nodes. We examine the inferred inheritance structure in
+the following subsections.
 
@@
 \subsubsection{Single origin}
-We would usually assume that all sequences assigned to a given Pango X lineage are descendants of a single recombinant sequence, arising as a result of a mixed infection followed by onward transmission.
-Thus, we would also expect our ARG to reflect this evolutionary history, where we have a all the samples assigned to this recombinant lineage tracing back to a single recombination node, representing this original recombination event.
+We would usually assume that all sequences
+assigned to a given Pango X lineage are descendants of a single recombinant
+sequence, arising as a result of a mixed infection followed by onward
+transmission. Thus, we would also expect our ARG to reflect this evolutionary
+history, where we have a all the samples assigned to this recombinant lineage
+tracing back to a single recombination node, representing this original
+recombination event.
 
@@
 % }
-\end{tabularx}
-\caption{\label{fig:pango-single-origin-graph}
-Single origin Nextclade Pango X lineages.
-(A) Subgraph for XA in the long ARG: all XA samples and their ancestral lineages are shown outwards to the nearest sampled viral genome; dotted lines show ARG continuation.
-Nodes are coloured by Nextclade Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with imputed Pango status.
-(B) Details for the single-origin Pango X lineages in the long ARG.
-(C) XAG origins in the long ARG.}
-\end{figure}
+\end{tabularx} \caption{\label{fig:pango-single-origin-graph} Single origin
+Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all XA samples
+and their ancestral lineages are shown outwards to the nearest sampled viral
+genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
+Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
+imputed Pango status. (B) Details for the single-origin Pango X lineages in the
+long ARG. (C) XAG origins in the long ARG.
+} \end{figure}
 
@@
 
-Table~\ref{tab:pango-single-origin} summarises the sequences of all 16 of the Pango recombinant lineages that trace back to a single recombination node in the long ARG.
-[FIXME UPDATE THESE]
-We can see that XA, XAC, XAE, and XF form monophyletic clades, in which all of the samples given that Pango designation descendand from a single recombination node.
+Table~\ref{tab:pango-single-origin} summarises the sequences of all 16 of the
+Pango recombinant lineages that trace back to a single recombination node in
+the long ARG. [FIXME UPDATE THESE] We can see that XA, XAC, XAE, and XF form
+monophyletic clades, in which all of the samples given that Pango designation
+descendand from a single recombination node.
 
@@
 
-There is generally good agreement on the imputed parental lineages from the ARG and the official Pango designations. [Discuss more]
-
-[Discuss panel C where XAG shares a common recombinant origin with samples from several other X lineages]
-All XAG samples share a single origin; however this is also shared with all XAA samples, and includes some, but not all, of the samples identified as XAB, XQ, and XU by Nextclade.
-Note that if the GISAID designations are used, many of the samples marked here as XAB are reclassified as BA.2, and XAG becomes fully monophyletic (supplementary figure \ref{fig:pango_XAG_gisaid_graph}; this also gives identifying labels to the sample nodes, which can be converted to EPI\_ISL indentifiers using Supplementary table XXX )
-
-\subsubsection{Multiple origin}
-The ARG inference has no pre-defined knowledge of Pango X lineage assignments, and there is therefore no particular requirement that all the samples assigned to a given lineage must trace back to a single recombination event.
-In this section we examine the Pango X lineage samples that are inferred to have multiple recombinant origins (i.e., trace back to different recombination nodes in the ARG), but still form monophyletic clades % is this the right term?
-descending from these nodes .
+There is generally good agreement on the imputed parental lineages from the ARG
+and the official Pango designations. [Discuss more]
+
+[Discuss panel C where XAG shares a common recombinant origin with samples from
+several other X lineages] All XAG samples share a single origin; however this
+is also shared with all XAA samples, and includes some, but not all, of the
+samples identified as XAB, XQ, and XU by Nextclade. Note that if the GISAID
+designations are used, many of the samples marked here as XAB are reclassified
+as BA.2, and XAG becomes fully monophyletic (supplementary figure
+\ref{fig:pango_XAG_gisaid_graph}; this also gives identifying labels to the
+sample nodes, which can be converted to EPI\_ISL indentifiers using
+Supplementary table XXX )
+
+\subsubsection{Multiple origin} The ARG inference has no pre-defined knowledge
+of Pango X lineage assignments, and there is therefore no particular
+requirement that all the samples assigned to a given lineage must trace back to
+a single recombination event. In this section we examine the Pango X lineage
+samples that are inferred to have multiple recombinant origins (i.e., trace
+back to different recombination nodes in the ARG), but still form monophyletic
+clades % is this the right term? descending from these nodes .
 
@@
 \end{tabular}
-\caption{\label{fig:multiple_origins_graph} A.
-Example subgraph for XD in the long ARG, which shows two origins for the four Nextclade-labelled XD samples in the long ARG.
-B.
-details of inferred breakpoints for different recombination events inferred for XE samples (170 samples and 12 inferred recombination nodes; some are very similar (Official: BA.1* \& BA.2, 10448–11287)}
+\caption{\label{fig:multiple_origins_graph}
+A. Example subgraph for XD in the long ARG, which shows two origins for the
+four Nextclade-labelled XD samples in the long ARG. B. details of inferred
+breakpoints for different recombination events inferred for XE samples (170
+samples and 12 inferred recombination nodes; some are very similar (Official:
+BA.1* \& BA.2, 10448–11287)}
 \end{figure}
 
-In Figure~\ref{fig:multiple_origins_graph} we give the details of these multiple pango lineages, along with an example lineage subgraph.
-In this XD example the left hand subgraph (the origin of three XD samples) has an earliest sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand subgraph has a single sample with EPI\_ISL 8514045 (dated 2021-12-30).
-These two samples differ by 35 mutations; considering the time between the samples, it does seem possible that they represent true independent origins of XD.
-Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the same subgraphs with sample identifiers and coloured by GISAID Pango labels (however, no samples are designated by GISAID as XD in the long ARG).
-
-% Interestingly, XC is inferred to be single origin in the Wide ARG but not in the long ARG.
-See lineage subgraphs in supplemental figure Y.
-This suggests that insufficient sampling may be the explanation for many of these multiple origin events.
+In Figure~\ref{fig:multiple_origins_graph} we give the details of these
+multiple pango lineages, along with an example lineage subgraph. In this XD
+example the left hand subgraph (the origin of three XD samples) has an earliest
+sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
+subgraph has a single sample with EPI\_ISL 8514045 (dated 2021-12-30). These
+two samples differ by 35 mutations; considering the time between the samples,
+it does seem possible that they represent true independent origins of XD.
+Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the same subgraphs
+with sample identifiers and coloured by GISAID Pango labels (however, no
+samples are designated by GISAID as XD in the long ARG).
 
@@
 
-As well as cases where a single recombination event explains several X lineages (such as XAG above), many of the Pango X lineages involve more complex origins, in which samples which have not been previously identified as recombinants also descend from originating recombination events.
-In addition these complex topologies often include extra recombination nodes within the topology that descends from the main originating event.
-An example is XB in the long ARG, plotted in Figure~\ref{fig:complex_origins_graph}.
-Here, a recombination between a B.1 sample and B.1.627 sample leads both to the XB samples and to a "hairball" of further recombination nodes whose descendants have not been identified as recombinants by Nextclade (blue nodes, "pre-Alpha" VoC).
-However, in this subgraph, there is only a single sample, identified as XB by Nextclade, which is descended from these
-nested recombination events; moreover, this sample is not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph}).
-We suggest that investigating cases of complex origins, and identifing which (if any) of the nested, further recombination events may be artifactual, is an important area of future research.
-
-\subsection{Novel recombinants}
-Next, we examine recombination nodes in the long ARG (sampled to mid-2022) that represent previously unnamed putative recombinant sequences, which do not have a Pango recombinant designation.
-For this proof-of-concept study, we have arbitrarily picked 12 recombination nodes that seem plausible (Supplementary Table 4).
-All these nodes involve Omicron subvariants (BA.1, BA.2, BA.4, and BA.5).
-These recombination nodes (1) were inserted into the long ARG on or after January 8, 2022; (2) have at least 10 descendent samples; and (3) have no mutational differences (including immediate reversions) from their parent nodes.
-
-TODO: Node  740761 (USA/NC-CDC-LC0668306/2022) is a sister clade proposed1006 in the UShER public phylogeny.
-Node 628656 (Scotland/QEUH-37794BB/2022) is a sister clade of XAC, and annotated as ``miscBA2BA1PostSpike''.
+As well as cases where a single recombination event explains several X lineages
+(such as XAG above), many of the Pango X lineages involve more complex origins,
+in which samples which have not been previously identified as recombinants also
+descend from originating recombination events. In addition these complex
+topologies often include extra recombination nodes within the topology that
+descends from the main originating event. An example is XB in the long ARG,
+plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
+between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
+"hairball" of further recombination nodes whose descendants have not been
+identified as recombinants by Nextclade (blue nodes, "pre-Alpha" VoC). However,
+in this subgraph, there is only a single sample, identified as XB by Nextclade,
+which is descended from these nested recombination events; moreover, this
+sample is not identified as XB by GISAID (Supplementary
+Figure~\ref{fig:pango_XB_gisaid_graph}). We suggest that investigating cases of
+complex origins, and identifing which (if any) of the nested, further
+recombination events may be artifactual, is an important area of future
+research.
+
+% JK: removing this for now
+
+% \subsection{Novel recombinants}
+% Next, we examine recombination nodes in the
+% long ARG (sampled to mid-2022) that represent previously unnamed putative
+% recombinant sequences, which do not have a Pango recombinant designation. For
+% this proof-of-concept study, we have arbitrarily picked 12 recombination nodes
+% that seem plausible (Supplementary Table 4). All these nodes involve Omicron
+% subvariants (BA.1, BA.2, BA.4, and BA.5). These recombination nodes (1) were
+% inserted into the long ARG on or after January 8, 2022; (2) have at least 10
+% descendent samples; and (3) have no mutational differences (including immediate
+% reversions) from their parent nodes.
+
+% TODO: Node  740761 (USA/NC-CDC-LC0668306/2022) is a sister clade proposed1006
+% in the UShER public phylogeny. Node 628656 (Scotland/QEUH-37794BB/2022) is a
+% sister clade of XAC, and annotated as ``miscBA2BA1PostSpike''.
 
 \section{Discussion}
-The COVID-19 pandemic continues to be a global emergency, with persistent high levels of infection.
-This high prevalence has allowed the proliferation of many variants, with more than 600 Pango-designated lineages circulating globally in the last three months (January to March, 2023; https://gisaid.org/; access on March 27, 2023).
-High prevalence also brings a higher risk of coinfection, increasing opportunities for new phenotypically distinct recombinants to emerge and spread.
-
-In this proof-of-concept study, we demonstrate the feasibility of reconstructing a large ARG from over one million SARS-CoV-2 genomes, a feat that is simply not possible using other available ARG inference methods \citep{Rasmussen2014-el,Ignatieva2021-rg, Speidel2019-yh}.
-We have accomplished this feat by building on top of recent advances in storing and processing large ARGs (tskit) and inferring genealogical relationships efficiently at scale (tsinfer).
-The highly compact representation of the ARGs as succinct tree sequences, coupled with vectorised computing technologies, enables the large SARS-CoV-2 ARGs to be loaded in less than one second in a Jupyter notebook and allows for each SARS-CoV-2 genome to be stored in less than eleven bytes of data.
-We believe that further development of the sc2ts methodology and software can create a powerful tool (1) to efficiently detect emerging or previously overlooked recombinants in SARS-CoV-2; (2) to jointly model the mutational and recombinational processes shaping global SARS-CoV-2 genomic diversity; and (3) to perform a broad variety of evolutionary analyses of SARS-CoV-2 genomes with unprecedented global sequencing coverage.
-
-ARGs are rich genealogical structures that contain information about the genetic relationships among sampled genomes in the presence of recombination.
-Here, we have conducted several analyses of the preliminary ARGs to show that the ARGs built using sc2ts can (1) recapitulate the phylogenetic relationships among major SARS-CoV-2 lineages; (2) recover biologically plausible recombination signals in SARS-CoV-2 genomes; and (3) model mutation events in SARS-CoV-2 evolutionary history.
-
-ARGs reconstructed using sc2ts capture the phylogenetic relationships of major SARS-CoV-2 lineages.
-Our results show that the backbone topology of a densely sampled wide ARG inferred using sc2ts agrees well with the backbone topology of a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference package.
-However, we have noted several areas of differences (e.g., poorly calibrated branch lengths), which are the focus of future development of sc2ts.
-We believe that some of these differences can be reduced by improving and refining our approach and strategy to handle samples with erroneously reported collection dates (i.e., time travellers that have passed the maximum submission delay of 30 days) and to modify the topology of daily trees.
-
-Our results show that sc2ts identifies the early recombinants (pre-2022) previously identified by \cite{Jackson2021-ik}.
-Sc2ts also identifies many Pango recombinants, which are designated after community perusal of supporting genomic evidence.
-We have found several previously unreported and overlooked recombinant sequences in the wide ARG.
-After further improvements and enhancements, we will use sc2ts to build a comprehensive ARG to search for recombination signals among all the 15 million and counting SARS-CoV-2 genomes available.
+The COVID-19 pandemic continues to be a global emergency,
+with persistent high levels of infection. This high prevalence has allowed the
+proliferation of many variants, with more than 600 Pango-designated lineages
+circulating globally in the last three months (January to March, 2023;
+https://gisaid.org/; access on March 27, 2023). High prevalence also brings a
+higher risk of coinfection, increasing opportunities for new phenotypically
+distinct recombinants to emerge and spread.
+
+In this proof-of-concept study, we demonstrate the feasibility of
+reconstructing a large ARG from over one million SARS-CoV-2 genomes, a feat
+that is simply not possible using other available ARG inference methods
+\citep{Rasmussen2014-el,Ignatieva2021-rg, Speidel2019-yh}. We have accomplished
+this feat by building on top of recent advances in storing and processing large
+ARGs (tskit) and inferring genealogical relationships efficiently at scale
+(tsinfer). The highly compact representation of the ARGs as succinct tree
+sequences, coupled with vectorised computing technologies, enables the large
+SARS-CoV-2 ARGs to be loaded in less than one second in a Jupyter notebook and
+allows for each SARS-CoV-2 genome to be stored in less than eleven bytes of
+data. We believe that further development of the sc2ts methodology and software
+can create a powerful tool (1) to efficiently detect emerging or previously
+overlooked recombinants in SARS-CoV-2; (2) to jointly model the mutational and
+recombinational processes shaping global SARS-CoV-2 genomic diversity; and (3)
+to perform a broad variety of evolutionary analyses of SARS-CoV-2 genomes with
+unprecedented global sequencing coverage.
+
+ARGs are rich genealogical structures that contain information about the
+genetic relationships among sampled genomes in the presence of recombination.
+Here, we have conducted several analyses of the preliminary ARGs to show that
+the ARGs built using sc2ts can (1) recapitulate the phylogenetic relationships
+among major SARS-CoV-2 lineages; (2) recover biologically plausible
+recombination signals in SARS-CoV-2 genomes; and (3) model mutation events in
+SARS-CoV-2 evolutionary history.
+
+ARGs reconstructed using sc2ts capture the phylogenetic relationships of major
+SARS-CoV-2 lineages. Our results show that the backbone topology of a densely
+sampled wide ARG inferred using sc2ts agrees well with the backbone topology of
+a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference
+package. However, we have noted several areas of differences (e.g., poorly
+calibrated branch lengths), which are the focus of future development of sc2ts.
+We believe that some of these differences can be reduced by improving and
+refining our approach and strategy to handle samples with erroneously reported
+collection dates (i.e., time travellers that have passed the maximum submission
+delay of 30 days) and to modify the topology of daily trees.
+
+Our results show that sc2ts identifies the early recombinants (pre-2022)
+previously identified by \cite{Jackson2021-ik}. Sc2ts also identifies many
+Pango recombinants, which are designated after community perusal of supporting
+genomic evidence. We have found several previously unreported and overlooked
+recombinant sequences in the wide ARG. After further improvements and
+enhancements, we will use sc2ts to build a comprehensive ARG to search for
+recombination signals among all the 15 million and counting SARS-CoV-2 genomes
+available.
 
@@
 \section{Conclusions}
-The global understanding of viral evolution during the COVID-19 pandemic is unprecedented, with over 15 million sequences shared to date.
-We envisage that once fully developed and tested, sc2ts methodology and software may provide a unified computational framework to study the evolution and epidemiology of SARS-CoV-2 at pandemic scale.
-The evolutionary analyses presented herein exemplify the numerous applications possible using the ancestral recombination graph inferred by sc2ts.
-We invite bioinformaticians, evolutionary biologists, and virologists to join us and the broader tskit community to build the most complete representation of the evolutionary history of SARS-CoV-2.
-Moreover, while we have focussed on the particular application to SARS-CoV-2, the methods we describe are readily applicable to the analysis of many other viral pathogens.
-Solving the challenges in building and analysing large viral genealogies would therefore contribute to the important goal of improving preparedness for potential future pandemics.
+The global understanding of viral evolution during the COVID-19 pandemic is
+unprecedented, with over 15 million sequences shared to date. We envisage that
+once fully developed and tested, sc2ts methodology and software may provide a
+unified computational framework to study the evolution and epidemiology of
+SARS-CoV-2 at pandemic scale. The evolutionary analyses presented herein
+exemplify the numerous applications possible using the ancestral recombination
+graph inferred by sc2ts. We invite bioinformaticians, evolutionary biologists,
+and virologists to join us and the broader tskit community to build the most
+complete representation of the evolutionary history of SARS-CoV-2. Moreover,
+while we have focussed on the particular application to SARS-CoV-2, the methods
+we describe are readily applicable to the analysis of many other viral
+pathogens. Solving the challenges in building and analysing large viral
+genealogies would therefore contribute to the important goal of improving
+preparedness for potential future pandemics.
+
+\section{Methods}
+
+[TODO update this summary para to avoid duplications with
+later sections and to add the appropriate signposts
+when they are done(ish)]
+
+Sc2ts is a method for inferring ARGs from densely sampled pandemic-scale data
+in real time, in which recombination occurs at a low but significant rate.
+The essential idea is to incrementally update the ARG each day
+with the sequences collected on that day. These updates are performed by
+first finding likely ``copying paths'' for each sequence in the daily
+batch to the current ARG, using the ARG-based implementation of
+the Li and Stephens model~\citep{Li2003-ib} introduced
+in \texttt{tsinfer}~\citep{Kelleher2019-ba}.
+These copying paths
+will mostly consist of a new sample sequence copying from a node in the ARG
+with a small number of mutations, and often many samples
+from a daily batch will copy from the same ARG node. The second
+step is then to ``resolve'' these implied polytomies by using
+standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
+This greedy update strategy inevitably lead to unparsimonious
+topologies, and the third step is to then increase the
+overall parsimony of the inferred ARG by making some simple topological
+updates (Figure~\ref{fig:overview_sc2ts}C and D).
+The operations are carried out efficiently using the well-established
+\texttt{tskit}
+library~\citep{Kelleher2018-xc,Ralph2020-efficiently,Tskit2023-tskit},
+and the resulting structure can be conveniently and interactively
+analysed using the many features provided by \texttt{tskit}.
+
+\begin{figure} \centering
+\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
+\caption{\label{fig:overview_sc2ts}
+A schematic of the sc2ts method. Sc2ts reconstructs the genetic relationships
+among SARS-CoV-2 genomes by copying samples to all possible ancestors collected
+at earlier time points (curved arrows). An ARG is built by attaching new sample
+genomes grouped by collection date to an existing ARG rooted at the Wuhan-Hu-1
+reference genome, day by day. Each iteration involves three stages: (1)
+attachment of new samples to the growing ARG (A, B); (2) reconstruction of
+trees relating the samples under each attachment node (C); and parsimony-based
+tree topology adjustments (D, E). In the absence of  recombination, sc2ts
+infers an ARG that is a single tree relating the samples (A). However, in the
+presence of recombination, sc2ts infers an ARG that is encoded as a sequence of
+local trees relating segments of the sample genomes (B), inserting a
+recombination node to represent a putative recombination event. Additionally,
+mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to
+make more parsimonious placements of mutations that should be shared or should
+not be immediately reverted, respectively.}
+\end{figure}
+
+% We introduce an ARG inference method, summarised in
+% Figure~\ref{fig:overview_sc2ts}.
+% % Note going to rewrite this anyway, so not trying to make it flow.
+% It uses (1) an efficient HMM implementation of the Li and Stephens (LS) model
+% \citep{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2)
+% the reported collection dates of the SARS-CoV-2 samples to determine the order
+% of genome copying; and (3) several parsimony-based heuristics to locally refine
+% the topology. In brief, this combined approach uses the likelihood model of LS
+% to attach sample genomes to existing nodes and to detect recombinants, but
+% because many of the daily tip samples harbour few, if any, mutational changes,
+% we use parsimony to resolve their close relationships.
+
+% Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by
+% taking advantage of the reported collection dates to incrementally grow the
+% ARG. The inference process occurs forward in time, attaching new genomes
+% aggregated by their collection date (daily batches), from the earliest to the
+% last collection date. Prior to attaching sample genomes, the ARG is initialised
+% such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence
+% of recombination, the ARG is a single tree with no reticulation (Figure 1A);
+% however, in the presence of recombination, a reticulate ARG is generated, with
+% variation in tree topology across the genome (Figure 1B). Unlike standard
+% phylogenetic trees in which internal nodes always represent inferred ancestors,
+% internal nodes in the sc2ts ARGs represent sequences which can either be
+% inferred ancestors or known samples.
+
+% Each daily batch of genomes is first attached to the growing ARG without any
+% phylogenetic resolution. In cases where the newly added sequence is inferred to
+% be a recombinant, a recombination node is inserted; the ``date'' of this node
+% is arbitrarily set to the average of the collection dates of the genome and the
+% most recent of its parent nodes. Otherwise, the genomes are attached to the
+% most closely related ancestral node (an attachment node). This creates
+% polytomies that are then locally resolved among the descendants of each
+% attachment node using the Unweighted Pair Group Method with Arithmetic Mean
+% (UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
+% This hierarchical clustering algorithm produces a dendrogram relating a set of
+% sequences, which is treated as an ultrametric binary tree here. Subsequently,
+% mutations are placed onto the daily tree by maximum parsimony. An issue with
+% the UPGMA algorithm is that it may create internal nodes supported by no
+% informative site (i.e., having no mutation immediately ancestral to them). In
+% an adjustment step, such unsupported internal nodes are removed to
+% ``un-resolve'' the daily tree; the remaining internal nodes, which are
+% supported by at least one shared mutation, are retained. The ``dates'' of the
+% non-sample internal nodes are set to the same date as the samples in the daily
+% batch. The resulting daily tree is grafted onto the attachment node in the
+% global ARG.
+
+% During ARG inference, two types of non-sample nodes are inserted to improve
+% parsimonious placement of mutations: mutation-coalescing nodes and
+% pushed-reversion nodes. We call these operations tree adjustments. A
+% mutation-collapsing node is inserted when a shared mutation between sibling
+% nodes is found (Figure 1D). An attempt to create mutation-collapsing nodes is
+% made after grafting a new daily tree onto the global ARG. If shared mutations
+% exist between two or more sibling nodes, then a parsimonious explanation is to
+% create a new node that is the parent of those siblings, on which these
+% mutations occurred. A pushed-reversion node is inserted when an immediate
+% reversion occurs (Figure 1E), i.e. a mutation that reverts another mutation on
+% its parent node. Instead of retaining both the mutation and immediate
+% reversion, only the mutation is kept after a pushed-reversion node is inserted,
+% thereby reducing the number of mutations on the branches. Note that these
+% non-sample nodes are used as reconstructed genomes during HMM copying.
+
+% Furthermore, during inference, the ARG is automatically encoded, stored, and
+% updated as a tree sequence using tskit \citep{Kelleher2018-xc}.
+
+\subsection{Ancestral Recombination Graphs}
+The term ``Ancestral Recombination Graph'' was introduced by
+Griffiths and colleagues~\citep{Griffiths1991-two,Griffiths1998-ancestral}
+and originally defined an alternative formulation of the coalescent
+with recombination stochastic process~\citep{Hudson1983-properties}.
+Subsequently, the term ARG came to be used in a more general way to
+describe not just realisations of this model, but to any
+recombinant genetic ancestry~\citep{Minichiello2006-mapping}.
+While there is some subtlety in the details~\citep{Wong2023-efficient},
+we can think of an ARG as being any graph that encodes the
+reticulate genetic ancestry of a sample of sequences under
+the influence of recombination.
+
+The ``succinct tree sequence'' is an ARG data structure
+that is both general (in terms of the types of ancestry that can
+be described) and computationally efficient~\citep{Wong2023-efficient}.
+Originally developed to facilitate large-scale coalescent
+simulations~\citep{Kelleher2016-wk}, the methods have been
+extended and applied to forwards-time
+simulations~\citep{Kelleher2018-xc,Haller2018-tree},
+calculation of population genetics statistics~\citep{Ralph2020-efficiently}
+and ARG inference~\citep{Kelleher2019-ba,Wohns2022-th}.
+The succinct tree sequence is based on a simple tabular representation,
+which defines a set of nodes, edges, sites and mutations. A \emph{node}
+represents a particular genome, which may be an observed sample
+or an inferred genetic ancestor. The genetic inheritance between
+a pair of nodes along a segment of genome is defined by
+the \emph{edge} $(\ell, r, p, c)$, which states that
+child node $c$ inherited its genome from parent node $p$
+from left coordinate $\ell$ to right coordinate $r$. A \emph{site}
+defines the genomic position at which some variation occurs,
+and a \emph{mutation} defines the site, node and derived state
+that generates variation.
+Given these tables, we can then efficiently construct the local
+genealogical trees along the genome (arising from recombination),
+and also perform a range of calculations efficiently by
+reasoning about the differences between these local
+trees~\citep{Kelleher2016-wk,Ralph2020-efficiently}. These
+algorithms have led to peformance increases of several orders
+of magnitude over the state-of-the-art in a range of
+applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,
+Ralph2020-efficiently,Baumdicker2022-ep}.
+The succinct tree sequence encoding
+is also very concise, allowing, for example, for millions of
+complete human genomes to potentially be stored in a few gigabytes of
+space~\citep{Kelleher2019-ba}.
+
+The \texttt{tskit} library~\citep{Tskit2023-tskit} is a liberally
+licensed open source toolkit that provides a comprehensive suite
+of tools for working with ARGs. Based on core functionality written
+in C, it provides interfaces in C, Python and Rust. The Python interface
+is based on NumPy~\citep{Harris2020-array}, and provides a convenient
+platform for interactive analysis of large-scale data using, for
+example, Jupyter notebooks~\citep{Kluyver2016-jupyter} and taking
+advantage of the analysis tools in the burgeoning PyData ecosystem.
+Tskit is mature software, widely used in population genetics, and
+has been incorporated into several downstream
+applications~\citep[e.g.][]{Haller2019-slim,Speidel2019-yh,
+Terasaki2021-geonomics,
+Fan2022-genealogical,Korfmann2022-weak,
+Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}.
+
+\subsection{The Li and Stephens model}
+The Li and Stephens (LS) model\citep{Li2003-ib}  is an approximation of the
+coalescent with recombination~\citep{Hudson1983-properties}, which captures
+many of the important features of the joint processes of mutation and
+recombination. It is a Hidden Markov Model (HMM) in which a focal genome is a
+sequence of nucleotides that are probabilistically emitted as
+an imperfect mosaic of a set of reference genomes
+(Figure~\ref{fig:ls_diagram}).
+The LS model is used in a wide variety
+of applications in genomics, including modern approaches to
+statistical genotype phasing and imputation
+\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa},
+and estimation of parameters such as
+recombination rates \citep[e.g.][]{Hinch2011-tz},
+and selection within and across hosts in viral
+sequence data \citep[e.g.][]{Palmer2019-wa}.
+See~\cite{Mcvean2019-linkage} for further review and discussion
+of the LS model.
+
+\begin{figure} \centering
+\includegraphics[width=0.5\textwidth]{figures/ls_diagram_covid.pdf}
+% FIXME this caption is too long
+\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
+model, in which a query sequence (bottom) is described as an
+imperfect mosaic of the sequences in a reference panel.
+Black crosses along the focal sequence show genotyping
+errors or mutations.
+In the standard formulation, at site $\ell$, the recombination probability is $r_\ell$,
+the mutation probability is $\mu_\ell$ and $n$
+denotes the size of the reference panel.
+The Viterbi algorithm can be used to find a
+``copying path'' through the reference panel for a given query sequence that
+maximises the likelihood under these parameters.
+% The LS HMM is governed by the transition and
+% emission matrices, $Q$ and $E$, respectively.
+% Each may be a function of the
+% reference panel members, but transitions are generally assumed to be
+% independent of the hidden states themselves (shown in the pink panel). This
+% dramatically increases performance as the state space drops to two states
+% (i.e., switching or not switching).
+% Emissions may also be a function of the nucleotide states, but in our
+% case, we assume that all observed alleles (in both the reference panel and
+% focal sequence, $a\ell$) at the site under consideration are equally likely and
+% occur with mutation probability $\ell$ (shown in the blue panel).
+% These two
+% processes, transition and emission (which encode recombination and mutation,
+% respectively), define the generative process of the HMM, by which an imperfect
+% mosaic is emitted from a reference panel.
+ % of samples and nodes from the ARG
+% inserted at earlier collection dates (shown in the lower panel). Here, the
+% unseen states in the reference panel are shown as coloured lines enclosed by
+% the grey box, and the true path through the data which leads to the emitted
+% focal sequence is displayed by the black arrow. Examples of transition and
+% emission probabilities along this trajectory are shown by the red and blue
+% arrows, respectively.
+% Note that in practice, we do not know the pattern of
+% colours along the focal sequence and must infer it using the Viterbi algorithm.
+}
+\end{figure}
+
+The generative process of the LS model is summarised in
+Figure~\ref{fig:ls_diagram}. Here, a transition matrix, $Q$, governs the
+process of switching (recombining) between members of the reference panel (the
+hidden states). An emission matrix, $E$, allows for differences between the
+focal sequence and the hidden state from which it is copied (due to mutation as
+well as genotyping or sequencing error). Then, we use the Viterbi algorithm
+\citep{Viterbi1967-ol} to find the most likely copying path through the
+sequences, given a set of reference sequences, $Q$, and $E$.
+The relative rates of mutation and recombination are usually controlled by
+the site-specific parameters $\mu_\ell$ and $r_\ell$, respectively.
+For this work we used a simpler formulation, which uses a single parameter
+(the mismatch ratio, MMR) to control the relative importance of mutation
+and recombination in the HMM. Specifically, an MMR value of $k$ will
+prefer $k$ mismatches (mutations) to a recombination.[DUNCAN MORE DETAILS
+HERE PLEASE]
+
+% OLD text, ready to be adapted (and deleted once done)
+% For this
+% application, we have parameterised the HMM using a mismatch ratio (MMR), which
+% tunes how strongly mutations (i.e., no switching of template sequences) are
+% preferred over recombination (i.e., switching between sequences) during
+% copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
+% focal genome more likely spans at least two ancestral genomes (i.e,. the focal
+% genome is a putative recombinant) than a single ancestral genome. Here, the MMR
+% was set to three in all the HMM runs .
+
+In \texttt{sc2ts} we use the efficient ARG-based implementation of the
+LS Viterbi algorithm from \texttt{tsinfer}~\citep{Kelleher2019-ba} to find
+the most likely copying path for each sample sequence
+among all sequences (sampled and inferred) in the current ARG.
+In the majority of cases, with non-recombinant sample sequences,
+the most likely solution is to copy from one
+of the nodes in the ARG that minimised the number of mutations required
+on the query sequence. Importantly, because the reference panel here consists of
+\emph{every} node in the ARG, we can match to both older sample
+sequences or internal nodes representing an inferred ancestral sequence
+(see subsequent subsections for details about how these are added).
+Thus, when no recombination is present, the LS Viterbi algorithm is
+implementing a version of parsimony, in which we are guaranteed to
+find a sequence that minimises the number of additional mutations required
+to incorporate a newly-added sample into the ARG.
+Recombination is then inferred when the most likely solution to the LS
+HMM is to copy from more than one ARG node along the genome, for a
+given sample sequece. [TODO MORE]
+
+We can solve this massive optimisation problem exactly because the ARG-based
+implementation of the LS HMM used in \texttt{tsinfer} scales approximately
+logarithmically with reference panel size (as opposed to linearly,
+for standard matrix-based approaches)~\citep{Kelleher2019-ba}.
+This efficient HMM algorithm is the main reason for \texttt{tsinfer}'s
+scalability, and here allows us to find closely matching
+sequences and recombination paths among millions of SARS-CoV-2
+genomes exactly under a well-defined statistical model.
+
+It is important to note that the Viterbi algorithm only returns
+\emph{one of} the copying paths that maximise the likelihood under
+ the given mutation and recombination parameters. There may be many
+such paths, from which we choose one arbitrarily. Also, the
+present choice of using a single mismatch-ratio parameter to control
+the likelihood of recombination vs mutation may lead to relatively
+flat likelihood spaces where many different paths have equal likelihood.
+There are many possibilities in using established HMM methodology
+to reason about and explore the space of possible matches, which may
+be a fruitful avenue for future work. See the Discussion for further
+potential improvements that may be made to the HMM.
+
+\subsection{Tree inference from clusters of samples}
+For a given daily batch of sample sequences that are to be attached to the
+ARG, the LS HMM provides the most likely copying path and required mutations
+for each of those sequences. We could directly attach these samples to
+the ARG according to the copying paths and sets of mutations, but
+such an approach would be highly unparsimonious because there are
+often many samples that share the same copying path with
+complex patterns of shared mutations.
+In this section we discuss the strategy used to infer likely tree
+toplogies among the samples that share the same copying path.
+In the majority of cases this copying path will consist of a single
+node (no recombination), and so for simplicity we will assume in the
+following discussion that all of the samples in a HMM-defined cluster
+attach to a single node. The differences required to deal with
+newly-detected putative recombinants are discussed below
+in Section XXX.
+
+With tens of thousands of samples being added to the ARG per day,
+such clusters of sequences attaching to the same node
+can be large, with complex patterns of shared mutations.
+To avoid the implied large polytomy and unparsimonious
+repetition of identical mutations among sibling nodes, we
+infer a tree among these samples (the LS HMM has already
+
+A natural
+way to
+
+We can think of the sequences attaching to a given node in the ARG
+from a daily batch as a polytomy,
+but with potentially shared mutations among the sibling nodes
+(Figure~\ref{fig:overview_sc2ts}C).
+
+\subsection{Parsimony updates}
+[Discuss the parsimony tweaks we make to the trees. Get a better name
+for the subsection title]
+
+\subsection{Treatment of recombinants}
+[Discuss the special treatment we give to recombinants.]
+Include the discussion about mirrored coords from here :]
+
+% \subsubsection{Breakpoint intervals and HMM consistency} When recombinant
+% ancestry for a given genome is inferred by the LS HMM, a breakpoint is
+% identified as the last possible position where the copying path must switch.
+% Thus, the breakpoints used in the ARG represent the rightmost end of the
+% interval, within which the recombination event could have occurred. To obtain
+% the leftmost end of this interval, we rerun the HMM using mirrored coordinates
+% (i.e., from the 3’ end to the 5’ end).
+
+% A recombinant sequence was deemed to be HMM-consistent if (1) the same number
+% of parent nodes and (2) the same set of mutations immediately ancestral to it
+% were obtained from both the unmirrored and mirrored HMM runs. In the ideal
+% case, the copying paths of both the HMM runs switch between parent sequences at
+% the same position(s); in less ideal cases, however, the copying paths switch at
+% different positions, typically because of a lack of genetic variation between
+% the positions. The switch positions in the copying paths of both the HMM runs
+% were used to define the breakpoint interval(s) of an HMM-consistent
+% recombinant.
+
+\subsection{Data preprocessing}
+
+The findings of this study are
+based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
+up to Aug 22, 2022 and accessible at
+% FIXME
+doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
+if they (1) had ambiguous collection dates; (2) were collected before January
+01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
+Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
+MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
+2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
+inference, sequences were excluded if they had a ``bad'' quality control status
+in any of the four Nextclade columns (``qc.missingData.status'',
+``qc.mixedSites.status'', ``qc.frameShifts.status'' and
+`qc.stopCodons.status''). These excluded genomes are enriched for low-coverage,
+contaminated, or misassembled sequences.
+
+Ambiguous nucleotide letters (i.e.,
+not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
+Problematic bases in the alignments, which had two or more ambiguous letters or
+gaps within a distance of seven bases, were masked as missing data following
+the protocol in \url{https://github.com/roblanf/sarscov2phylo/}. Masked bases
+were ignored when attaching a sample genome to the ARG, but they were
+subsequently imputed. Further, 481 problematic sites flagged as prone to
+sequencing errors or highly homoplasic were excluded entirely
+% Do we need this first URL?
+(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
+\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
+
+\subsection{Exclusion of ``time travellers''}
+[FIXME I'm pretty sure we don't have two filters here, we just use
+maximum sampling delay. Also not sure the MSD acronym is worth it,
+we don't use it much.]
+
+Including samples with incorrectly reported collection dates (e.g., an Alpha
+sample purportedly collected in 2020 from the United States, before Alpha
+appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
+artefacts that distorted the topology of ARGs inferred early on in the study.
+Hence, to exclude such potential ``time travellers'' we employed two filters.
+The first was to remove any sequence with a collection date that pre-dated the
+time to the most recent common ancestor (tMRCA) of its corresponding clade. The
+tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
+(downloaded on August 22, 2022) where we used the lower bound of the 95\%
+confidence interval of each clade as the minimum date cut-off. The second
+filter was a simple threshold on the time delay between the collection date and
+submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
+inferred from the MSD-filtered genome set had noticeably fewer branches with a
+high number of reversions. After some preliminary analysis, a MSD threshold
+value of 30 days was used when building the ARGs described herein. Future work
+could estimate collection dates by adding these potential time travellers back
+on the tree, allowing an automated assessment of collection date discrepancies.
+
+\subsection{Imputation of Pango lineage for non-sample nodes}
+Initially, only sample nodes have Pango lineage assignments.
+For analysis of recombinant parents, it is also helpful to assign
+Pango lineages to non-sample nodes. We do this
+using a two-step procedure (Figure~\ref{fig:imputation}). If the Pango
+lineage of the parental node was available and no mutation was immediately
+ancestral to the non-sample node, then the parental node’s status was directly
+propagated to the non-sample node. Otherwise, the Pango lineage of the
+non-sample node was inferred by matching its full set of mutations against the
+Pango lineage-defining mutations (based on 90\% consensus of the sequences
+analysed) from the COVID-CG website \citep{Chen2021-zc};
+\url{https://covidcg.org/}; accessed on November 04, 2022).
+
+\begin{figure} \centering
+\includegraphics[width=.7\textwidth]{figures/imputation.pdf}
+\caption{\label{fig:imputation}A schematic of the two-step procedure to impute
+Pango lineage for inserted, non-sample nodes. Here, three nodes (question
+marks) have unknown Pango lineage (A). Because the parent of node 1 is
+available (i.e., L2) and there is no mutation occurring on the edge connecting
+the two nodes, the Pango lineage of the parent is directly propagated to node 1
+(B). However, if there is a mutation (red X’s), then the Pango lineage is
+inferred by comparison to the set of lineage-defining mutations from
+COVID-CG\citep{Chen2021-zc}, as in case (C). Finally, the inferred status is
+propagated to any non-mutant child nodes, as in case (D).
+}
+\end{figure}
 

### 2023-04-28 paper.tex
Finish first pass on sample cluster trees

@@
 such an approach would be highly unparsimonious because there are
-often many samples that share the same copying path with
-complex patterns of shared mutations.
+often many samples that share the same copying path, with
+complex patterns of shared mutations among those clusters of samples.
 In this section we discuss the strategy used to infer likely tree
-toplogies among the samples that share the same copying path.
-In the majority of cases this copying path will consist of a single
-node (no recombination), and so for simplicity we will assume in the
+topologies within these clusters of samples sharing the same copying path.
+In the majority of cases the copying path will consist of a single
+node (i.e., no recombination), and so for simplicity we will assume in the
 following discussion that all of the samples in a HMM-defined cluster
-attach to a single node. The differences required to deal with
+attach to a single node. The special treatment afforded to
 newly-detected putative recombinants are discussed below
@@
 With tens of thousands of samples being added to the ARG per day,
-such clusters of sequences attaching to the same node
-can be large, with complex patterns of shared mutations.
-To avoid the implied large polytomy and unparsimonious
-repetition of identical mutations among sibling nodes, we
-infer a tree among these samples (the LS HMM has already
-
-A natural
-way to
-
-We can think of the sequences attaching to a given node in the ARG
-from a daily batch as a polytomy,
-but with potentially shared mutations among the sibling nodes
-(Figure~\ref{fig:overview_sc2ts}C).
+we can have clusters of hundreds of sequences attaching to the same node.
+While some of these samples will require no extra mutations
+(i.e, they are identical to the attachment node), in general there
+will be complex patterns of shared mutations among the samples
+reflecting their evolutionary relationships. A natural way to infer
+these within-cluster evolutionary relationships is to use standard
+tree-building algorithms.
+We can infer a likely tree relating the
+samples in a cluster independently of the other samples in a
+daily batch, and then attach the tree (and mutations)
+to the ARG at the node identified by the HMM.
+
+We currently use the implementation of the UPGMA algorithm~\citep{Michener1957-tr}
+from SciPy~\citep{Pauli2020-scipy} to build trees from sample
+clusters, and map mutations back to this topology using maximum parsimony.
+This approach was chosen mainly for simplicity and because of the
+speed and reliability of the implementation.
+An issue with the UPGMA algorithm is that it generates a strictly
+binary tree, creating internal nodes
+supported by no informative site (i.e., having no mutation immediately
+ancestral to them). To avoid such false precision we post-process
+to remove unsupported internal nodes, representing the relationship
+between $k$ identical descendants of the node as a polytomy of size $k$.
+
+There are well-known issues with using such a simple algorithm for inferring
+evolutionary relationships~\citep{Felsenstein2004-inferring}.
+Since this within-cluster tree building has a significant influence on the
+overall ARG topology (see Results), implementing more sophisticated
+tree building methods which keep track of the location of the required mutations
+(rather then inferring post-hoc by parsimony), is a likely avenue
+for improvements in overall inference quality.
 
@@
 Include the discussion about mirrored coords from here :]
+
+\subsection{Node dating}
+[Discuss how we date nodes, and suggest caution on interpretation]
 

### 2023-04-28 paper.tex
Updates on parsimony algs

@@
 
-\subsection{Tree inference from clusters of samples}
+\subsection{Tree inference from HMM sample clusters}
 For a given daily batch of sample sequences that are to be attached to the
@@
 
-\subsection{Parsimony updates}
-[Discuss the parsimony tweaks we make to the trees. Get a better name
-for the subsection title]
+\subsection{Parsimony increasing hueristics}
+Attaching trees built from the clusters of samples that copy from
+a particular node (or path of nodes for recombinants, see the
+next section) under the HMM is an inherently greedy strategy,
+and can produce inferences that are clearly unparsimonious.
+The final step in adding a daily batch of samples to the ARG
+is then to perform some local updates that target specific
+types of parsimony violation in the just-updated regions of the
+ARG. There are currently two parsimony-increasing operations
+applied, which we refer to as ``mutation collapsing'' and ``reversion
+pushing''.
+
+Mutation collapsing inspects the siblings of a newly attached node
+to check if any of them share (a subset) of the mutations that
+it carries. If so, we increase the overall parsimony of the inference
+by creating a new node representing the ancestor that carried
+those shared mutations, and make that new node the parent of the
+siblings carrying those shared mutations. The patterns of
+shared mutations between siblings can be complex, and the current
+implementation uses a simple greedy strategy for choosing
+the particular mutations to collapse. [TODO back ref to the
+Results section to discuss how important these nodes are]
+
+The reversion push operation inspects newly added nodes to see
+if any of its mutations are ``immediate reversions''; that is,
+are reversions of a mutation that occured on the new node's
+immediate parent. We increase the overall parsimony of the
+inference by ``pushing in'' a new node in which descends from the
+original parent which carries all its mutations except those
+causing the reversions on the newly added node.
+[TODO back ref to the
+Results section to discuss how important these nodes are]
+
+Together these operations substantially increase the overall parsimony
+of the inferred ARGs, most noticably in terms of the number of
+reversion mutations inferred. However, they are both simple
+greedy operations, just examining the local parts of the
+ARG topology affected by newly added samples. Since the inferred
+ARGs still contain a large number of reversion mutations which
+are likely to be mostly artefactual, it is clear that there
+is room for improvement, and that further parsimony increasing
+hueristics would likely be of benefit.
 

### 2023-04-28 paper.tex

@@
 \centering
-% \includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
+\includegraphics[width=\textwidth]{figures/recombination_intervals.pdf}
 \caption{\label{fig:breakpoint-distribution}
@@
 \subsection{Tree inference from HMM sample clusters}
+\label{sec-sample-cluster-tree-inference}
 For a given daily batch of sample sequences that are to be attached to the
@@
 \subsection{Treatment of recombinants}
-[Discuss the special treatment we give to recombinants.]
-Include the discussion about mirrored coords from here :]
+A sample sequence is designated as a recombinant if the most likely
+path inferred by the LS HMM for that sample contains at least
+one switch between parents. Recombinant sequences are mostly treated
+identically to non-recombinants, as we simply need to reason about
+a path of parent nodes along genome intervals rather than a
+single parent over the whole genome, which is naturally handled by
+the succinct tree sequence data structure and \texttt{tskit} library
+(see the Ancestral Recombination Graphs section above). To facilitate
+analysis and to help understand the robustness of recombinants
+we perform some additional steps in \texttt{sc3ts}.
+
+The LS HMM may infer identical paths and patterns of mutations
+for multiple samples in a daily batch, and so we
+create a ``recombination`` node (marked with a specific ``flags'' value)
+for each distinct recombinant. (This node is not strictly necessary
+but makes it convenient to find recombinants for subsequent analysis.)
+Variation within a cluster
+of recombinant sequences is handled in the same way as non-recombinants
+(see Section~\ref{sec-sample-cluster-tree-inference} above for details).
+
+[TODO ADD IN HMM RERUN STUFF]
 

### 2023-05-02 paper.tex
Fisrt pass at recombinants

@@
 For example, sequences of the Pango recombinant XP were not identified as recombinants, because XP’s characteristic multibase deletion was masked.
-\item The "lack of memory" in the HMM can result in  recombinants with two or more breakpoints to have different parents for each recombined portion of the genome, when it is sometimes equally parsimonious to match against the same parent in separate regions.
-The copying process in the HMM currently does not preferentially "switch back" to a previoulsy matched parent, but incorporating this would be a valuable addition.
-\item Some recombinants may reasonably have three or more parents, indicating multiple "stacked" recombination events.
-We have not investigated this in the current analysis, but with further sc2ts development, there is potential to analyse these putative recombinants and trace their evolutionary origins through the ARG.
-\item In the present implementation, HMM traceback is performed once, yielding a single maximum-likelihood genome copying path.
-A stochastic traceback algorithm (e.g., (Rasmussen et al.
-2014)) can be implemented in order to quantify uncertainty about putative parent node(s) and switch positions.
-These estimates of uncertainty may be used as support values for branches analogous to bootstrap values produced by standard phylogenetic methods.
 \item Daily trees are built using the UPGMA algorithm.
@@
 
-[TODO ADD IN HMM RERUN STUFF]
+When the Viterbi algorithm implementation used by the LS HMM infers
+recombinant ancestry for a given genome, the point at which inheritance
+switches from one parent node to another is the last possible
+position.
+Thus, the breakpoints used in the ARG represent the rightmost end of the
+interval, within which the recombination event could have occurred. To obtain
+the leftmost end of this interval, we rerun the HMM using mirrored coordinates
+(i.e., from the 3’ end to the 5’ end) and store the information as
+metadata on the recombination node.
+These left- and rightmost positions
+are the recombination intervals reported throughout.
+Ultimately all the recombination intervals mean is that the parent sequences
+are identical within that interval, and there is therefore no information
+about where the recombination occured. There are many different processes
+that can contribute to the size of these intervals, and therefore
+understanding how these processes interact to produce wider or narrower
+intervals is an important avenue for future work.
+
+The LS HMM machinery, and the interpretation of inferred recombinant paths and
+breakpoint intervals is a central part of \texttt{sc2ts}, and there are many
+ways to extend and improve. For example, the current parameterisation of using
+a single ``mismatch ratio'' is very simplistic (see section XXX), and likely
+results in a very flat likelihood space where many sequences have equal
+likelihood of being chosen. More sophisticated parameterisations where we
+condition on known per-site mutation rates (and perhaps per-base subsitution
+rates) may lead to better results, although perhaps at the cost of some
+reduction in interpretability. Post-processing the match results to produce
+more parsimonious breakpoints may also be a worthwhile avenue for development.
+In particular, we may choose to insert breakpoints for the ARG that are choosed
+from within the possible interval, rather than the current approach of taking
+the right-most value. Cases where we have more than two parents may either be
+the ``stacking'' of multiple recombination events, or instances where
+the HMM has chosen to switch to a third sequence rather than back to
+the original parent (where this is equally parsimonious). Many putative
+recombination events, however, will represent poor quality data, where
+a recombinant copying path happens to be a more likely explanation
+of a highly divergent sequence.
+A thorough analysis of the behaviour of the LS model in the context
+of a pandemic-scale ARG may lead to significant improvements in our
+ability identify recombinants and to filter poor quality data.
 
@@
 [Discuss how we date nodes, and suggest caution on interpretation]
-
-% \subsubsection{Breakpoint intervals and HMM consistency} When recombinant
-% ancestry for a given genome is inferred by the LS HMM, a breakpoint is
-% identified as the last possible position where the copying path must switch.
-% Thus, the breakpoints used in the ARG represent the rightmost end of the
-% interval, within which the recombination event could have occurred. To obtain
-% the leftmost end of this interval, we rerun the HMM using mirrored coordinates
-% (i.e., from the 3’ end to the 5’ end).
 

### 2023-05-02 paper.tex

@@
 \subsection{Tree inference from HMM sample clusters}
-\label{sec-sample-cluster-tree-inference}
+\label{sec:sample-cluster-tree-inference}
 For a given daily batch of sample sequences that are to be attached to the
@@
 \subsection{Parsimony increasing hueristics}
+\label{sec:parsimony-heuristics}
 Attaching trees built from the clusters of samples that copy from
@@
 of recombinant sequences is handled in the same way as non-recombinants
-(see Section~\ref{sec-sample-cluster-tree-inference} above for details).
+(see Section~\ref{sec:sample-cluster-tree-inference} above for details).
 
@@
 \subsection{Node dating}
-[Discuss how we date nodes, and suggest caution on interpretation]
-
-% A recombinant sequence was deemed to be HMM-consistent if (1) the same number
-% of parent nodes and (2) the same set of mutations immediately ancestral to it
-% were obtained from both the unmirrored and mirrored HMM runs. In the ideal
-% case, the copying paths of both the HMM runs switch between parent sequences at
-% the same position(s); in less ideal cases, however, the copying paths switch at
-% different positions, typically because of a lack of genetic variation between
-% the positions. The switch positions in the copying paths of both the HMM runs
-% were used to define the breakpoint interval(s) of an HMM-consistent
-% recombinant.
+The approaches to assigning a date to nodes in \texttt{sc2ts} is currently
+ad-hoc, and the inferred timing of events from the ARGs reported
+here should be treated with caution (e.g., Figure~\ref{fig:recomb_mrcas}).
+Sample nodes (those corresponding to observed sample sequences) are the
+most accurately dated, as we use the reported collection date for
+these nodes. These are not entirely accurate, but our data filtering
+criteria should remove the most egregious errors (see
+Section~\ref{sec:filtering_time_travellers}). Other nodes in the ARG
+are dated by splitting the time between the attached samples
+and the chosen parent nodes equally (in the case of
+trees inferred from daily sample clusters,
+Section~\ref{sec:sample-cluster-tree-inference}) or by adding arbitrary
+small values when creating new nodes using parsimony rules
+(Section~\ref{sec:parsimony-heuristics}). Substantial improvements could
+likely be made using simple node dating methods.
 
 \subsection{Data preprocessing}
-
 The findings of this study are
@@
 up to Aug 22, 2022 and accessible at
-% FIXME
-doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
-if they (1) had ambiguous collection dates; (2) were collected before January
-01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
-Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
-MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
+\url{https://doi.org/10.55876/gis8.230329cd} (Supplementary Table 1).
+We removed sequences
+if they (1) had ambiguous collection dates;
+(2) were collected before January 01, 2020 or after June 30, 2022; or
+(3) were isolated from a non-human host.
+We aligned sequences to the Wuhan-Hu-1/2019 reference sequence
+(GenBank: MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
 2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
-inference, sequences were excluded if they had a ``bad'' quality control status
+inference, we also excluded sequences if they had a ``bad'' quality control status
 in any of the four Nextclade columns (``qc.missingData.status'',
-``qc.mixedSites.status'', ``qc.frameShifts.status'' and
-`qc.stopCodons.status''). These excluded genomes are enriched for low-coverage,
-contaminated, or misassembled sequences.
-
-Ambiguous nucleotide letters (i.e.,
+``qc.mixedSites.status'', ``qc.frameShifts.status'' and `qc.stopCodons.status'').
+
+% FIXME this isn't quite right
+We encode ambiguous nucleotide letters (i.e.,
 not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
@@
 
-\subsection{Exclusion of ``time travellers''}
+\subsection{Filtering ``time travellers''}
+\label{sec:filtering_time_travellers}
 [FIXME I'm pretty sure we don't have two filters here, we just use

### 2023-05-02 paper.tex

@@
 \subsection{The Li and Stephens model}
+\label{sec:ls}
 The Li and Stephens (LS) model\citep{Li2003-ib}  is an approximation of the
@@
 
+[ALSO MENTION MISSING DATA, where we deal with Ns in the alignment]
+
 % OLD text, ready to be adapted (and deleted once done)
@@
 (GenBank: MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
-2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
-inference, we also excluded sequences if they had a ``bad'' quality control status
+2022-07-26T12:00:00Z). We also excluded sequences if they had a
+``bad'' quality control status
 in any of the four Nextclade columns (``qc.missingData.status'',
@@
 
-% FIXME this isn't quite right
 We encode ambiguous nucleotide letters (i.e.,
-not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
-Problematic bases in the alignments, which had two or more ambiguous letters or
-gaps within a distance of seven bases, were masked as missing data following
-the protocol in \url{https://github.com/roblanf/sarscov2phylo/}. Masked bases
-were ignored when attaching a sample genome to the ARG, but they were
-subsequently imputed. Further, 481 problematic sites flagged as prone to
-sequencing errors or highly homoplasic were excluded entirely
-% Do we need this first URL?
-(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
-\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).
+not A, C, G, T, or a gap) in the pairwise genome alignments as missing data
+(N). Problematic bases in the alignments, which had two or more Ns or
+gaps within a distance of seven bases, are masked as missing data following
+the approach used in the ``faToVCF`` tool used by
+UShER~\citep{Turakhia2022-it}.
+Sites that are masked by this process are
+treated as missing data by the LS HMM (Section~\ref{sec:ls}).
+In addition, we exclude 481 problematic sites flagged as prone to
+sequencing errors or as highly homoplasic
+(\url{https://github.com/W-L/ProblematicSites_SARS-CoV2/},
+accessed 2022-09-22) entirely.
+
+Although the masking strategy used currently is simple and robust,
+there are significant disadvantages, excluding, for example, any
+deletions of length greater than one base. Exploring more sophisticated
+masking strategies is an important route for future improvements.
 
@@
 \label{sec:filtering_time_travellers}
-[FIXME I'm pretty sure we don't have two filters here, we just use
-maximum sampling delay. Also not sure the MSD acronym is worth it,
-we don't use it much.]
-
-Including samples with incorrectly reported collection dates (e.g., an Alpha
-sample purportedly collected in 2020 from the United States, before Alpha
-appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
-artefacts that distorted the topology of ARGs inferred early on in the study.
-Hence, to exclude such potential ``time travellers'' we employed two filters.
-The first was to remove any sequence with a collection date that pre-dated the
-time to the most recent common ancestor (tMRCA) of its corresponding clade. The
-tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
-(downloaded on August 22, 2022) where we used the lower bound of the 95\%
-confidence interval of each clade as the minimum date cut-off. The second
-filter was a simple threshold on the time delay between the collection date and
-submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
-inferred from the MSD-filtered genome set had noticeably fewer branches with a
-high number of reversions. After some preliminary analysis, a MSD threshold
-value of 30 days was used when building the ARGs described herein. Future work
-could estimate collection dates by adding these potential time travellers back
-on the tree, allowing an automated assessment of collection date discrepancies.
+A major source of error in early versions of \texttt{sc2ts} was
+the existance of ``time traveller'' sequences: those with erroneously
+early collection dates.
+For example, an Alpha sample purportedly collected in 2020 from the United States,
+before Alpha appeared in the United Kingdom (USA/MN-Mayo-1563/2020)
+produced significant topological distortions in
+ARGs inferred early on in the study.
+Hence, to exclude such potential ``time travellers'' we employ two filters.
+The first is to remove any sequence with a collection date that pre-dates the
+time to the most recent common ancestor (tMRCA) of its corresponding clade.
+We obtained the tMRCA for each clade from a Nextstrain GISAID reference tree
+(downloaded on August 22, 2022), and we used the lower bound of the 95\%
+confidence interval of each clade as the minimum date cut-off.
+This removed NN [Katherine can you fill in please?] sequences.
+
+This filter removes the most egregious time travellers, but many still
+remain. We thus employed a second filter which is a simple threshold on
+the time delay between the collection date and submission date.
+After some preliminary analysis we settled on a maximum submission delay
+of 30 days when building the ARGs described here as a reasonable compromise.
+
+As the results of \texttt{sc2ts} are sensitive to the existance of
+time-travellers, an imporant aspect of future work is to find better ways
+to identify them. One possibility is to use the LS HMM itself to flag
+overly divergent sequences, and to exclude them from attachment to the ARG.
+We might also estimate collection dates by adding these potential time travellers
+back to the ARG, allowing an automated assessment of collection date discrepancies.
 
 \subsection{Imputation of Pango lineage for non-sample nodes}
-Initially, only sample nodes have Pango lineage assignments.
-For analysis of recombinant parents, it is also helpful to assign
-Pango lineages to non-sample nodes. We do this
-using a two-step procedure (Figure~\ref{fig:imputation}). If the Pango
-lineage of the parental node was available and no mutation was immediately
-ancestral to the non-sample node, then the parental node’s status was directly
-propagated to the non-sample node. Otherwise, the Pango lineage of the
-non-sample node was inferred by matching its full set of mutations against the
-Pango lineage-defining mutations (based on 90\% consensus of the sequences
-analysed) from the COVID-CG website \citep{Chen2021-zc};
-\url{https://covidcg.org/}; accessed on November 04, 2022).
-
+% FIXME the caption and the text are basically saying the same thing here. Can
+% we streamline, and maybe give a bit more detail in the text about iteration,
+% and the concrete implementation?
 \begin{figure} \centering
@@
 \end{figure}
+For many applications it is helpful to infer the Pango lineage status
+of the internal ARG nodes estimated by \texttt{sc2ts}.
+% Initially, only sample nodes have Pango lineage assignments.
+% For analysis of recombinant parents, it is also helpful to assign
+% Pango lineages to non-sample nodes.
+We do this using a two-step procedure (Figure~\ref{fig:imputation}).
+Initially, only sample nodes have Pango lineage assignments.
+For a given non-sample node $u$, if the Pango lineage of its parent is
+available and there are no intervening mutations,
+then we impute that $u$'s Pango lineage is the same as its parent.
+Otherwise, the Pango lineage of the
+non-sample node is inferred by matching its full set of mutations against the
+Pango lineage-defining mutations (based on 90\% consensus of the sequences
+analysed) from the COVID-CG website \citep{Chen2021-zc};
+\url{https://covidcg.org/}; accessed on November 04, 2022).
+
+[TODO how accurate is this approach? We should probably discuss it somewhere
+in the Results, and refer back from here? ]
 

### 2023-05-02 plot.py
Put XA, XAG, XD into a single plot

@@ -881,12 +881,14 @@ class Pango_X_graph(Figure):
     node_colours = {
         # B lineages in blue/green (AY.100 is B.1.617.2.100)
         # Basic in blue
-        "B.1": "#212fe1",
-        "B.1.177.18": "#93e8ff",
-        "B.1.631": "#add9e5",
-        "B.1.634": "#2092e1",
-        "B.1.627": "#7b96cc",
-        "B.1.1": "#aea4ff",
+        'B.1': '#212fe1',
+        'B.1.384': '#4ac7ff',
+        'B.1.177.18': '#93e8ff',
+        'B.1.631': '#add9e5',
+        'B.1.634': '#2092e1',
+        'B.1.627': '#7b96cc',
+        'B.1.1': '#aea4ff',
+
         # Alpha: purple / lilac
         "B.1.1.7": "#ae9cff",
         # Delta: yellow/brown
@@ -894,10 +896,12 @@ class Pango_X_graph(Figure):
         "AY.1": "#ffe44a",  # Delta
         "AY.4": "#b78c19",  # Delta
         # Omicron (B.1.1.529...) in green-ish
-        "BA.1": "#74b058",  # Omicron
-        "BA.1.17": "#57de3c",  # Omicron
-        "BA.1.15": "#248f0f",  # Omicron
-        "BA.2.9": "#21e1ab",  # Omicron
+        'BA.1': '#74b058',  # Omicron
+        'BA.1.17': '#57de3c',  # Omicron
+        'BA.1.15': '#248f0f',  # Omicron
+        'BA.2': '#68ffa5', # Omicron
+        'BA.2.9': '#21e1ab', # Omicron
+
         # Pango X-lineages in red/orange/pink
         "XA": "#ff0055",
         "XB": "#fc31fb",
@@ -989,14 +993,15 @@ class Pango_X_tight_graph(Pango_X_graph):
     def define_nodes(cls, ts):
         raise NotImplementedError()
 
-    def used_pango_colours(self, used_nodes):
+    @classmethod
+    def used_pango_colours(cls, used_nodes, ts):
         used_pango = set(
             [
-                self.ts.node(n).metadata.get("Imputed_" + self.imputed_lineage, None)
+                ts.node(n).metadata.get("Imputed_" + cls.imputed_lineage, None)
                 for n in used_nodes
             ]
         )
-        return {k: v for k, v in self.node_colours.items() if k in used_pango}
[…]

### 2023-05-02 paper.tex
`Modify text to match new subgraph plots

@@
 
-\includegraphics[width=0.6\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
-\\ % &
-% \multicolumn{2}{c}{
-\includegraphics[width=\textwidth]{figures/Pango_XAG_nxcld_tight_graph.pdf}
+\includegraphics[width=\textwidth]{figures/Pango_XA_XAG_XD_nxcld_tight_graph.pdf}
 % }
-\end{tabularx} \caption{\label{fig:pango-single-origin-graph} Single origin
-Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all XA samples
-and their ancestral lineages are shown outwards to the nearest sampled viral
-genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
+\end{tabularx} \caption{\label{fig:pango-simple-origin-graph} Simple origin
+Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
+as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
+sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
 Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
-imputed Pango status. (B) Details for the single-origin Pango X lineages in the
-long ARG. (C) XAG origins in the long ARG.
+imputed Pango status. (B) Equivalent plot for the 17 XAG samples in the long ARG
+(C) Equivalent plot for both origination events for the four XD samples in the long ARG.
 } \end{figure}
 
-In Figure~\ref{fig:pango-single-origin-graph}A we illustrate the case of XA.
+In Figure~\ref{fig:pango-simple-origin-graph}A we illustrate the case of XA.
 We construct a small subgraph of the long ARG, tracing routes from Nextclade-identified XA samples to the closest other sample strains in our dataset.
@@
 monophyletic clades, in which all of the samples given that Pango designation
-descendand from a single recombination node.
+descend from a single recombination node.
 
@@
 
-[Discuss panel C where XAG shares a common recombinant origin with samples from
-several other X lineages] All XAG samples share a single origin; however this
-is also shared with all XAA samples, and includes some, but not all, of the
-samples identified as XAB, XQ, and XU by Nextclade. Note that if the GISAID
+[Discuss panel B where XAG shares a common recombinant origin with samples from
+several other X lineages]
+
+Figure~\ref{fig:pango-simple-origin-graph}B shows a case of another
+Pango lineage, XAG, whose Nextclade-designated samples also trace to a
+single originating recombination. This is a more intricate example, as the same origin
+is also shared with all samples designated XAA, and some, but not all, of the
+samples identified as XAB, XQ, and XU by Nextclade. Note, however, that if the GISAID
 designations are used, many of the samples marked here as XAB are reclassified
 as BA.2, and XAG becomes fully monophyletic (supplementary figure
-\ref{fig:pango_XAG_gisaid_graph}; this also gives identifying labels to the
-sample nodes, which can be converted to EPI\_ISL indentifiers using
+\ref{fig:pango_XAG_gisaid_graph}, which also gives identifying labels to the
+sample nodes, convertable to EPI\_ISL identifiers using
 Supplementary table XXX )
 
-\subsubsection{Multiple origin} The ARG inference has no pre-defined knowledge
+\subsubsection{Multiple origins} The ARG inference has no pre-defined knowledge
 of Pango X lineage assignments, and there is therefore no particular
@@
 \begin{figure}
-\begin{tabular}{cc}
-
-\includegraphics[width=0.45\textwidth]{figures/Pango_XD_nxcld_tight_graph.pdf}
-
- &
 
@@
 % Recombinant BA.2 17411 19955
-\end{tabular}
-\caption{\label{fig:multiple_origins_graph}
-A. Example subgraph for XD in the long ARG, which shows two origins for the
-four Nextclade-labelled XD samples in the long ARG. B. details of inferred
+\caption{\label{tab:multiple_origins_table}
+details of inferred
 breakpoints for different recombination events inferred for XE samples (170
@@
 
-In Figure~\ref{fig:multiple_origins_graph} we give the details of these
-multiple pango lineages, along with an example lineage subgraph. In this XD
-example the left hand subgraph (the origin of three XD samples) has an earliest
+Figure~\ref{fig:pango-simple-origin-graph}C shows a simple example, consisting of
+the 4 samples labelled XD by Nextclade in the long ARG. The left hand subgraph
+(the origin of three XD samples) has an earliest
 sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
 subgraph has a single sample with EPI\_ISL 8514045 (dated 2021-12-30). These
-two samples differ by 35 mutations; considering the time between the samples,
+two samples differ by 35 mutations; considering this, and the time between the samples,
 it does seem possible that they represent true independent origins of XD.
@@
 
+Table \ref{tab:multiple_origins_table} summarises other Pango recombinants which
+we infer to have multiple origins. [TODO: needs more detail]
+
 \subsubsection{Complex origins}
@@
 \includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
-\caption{\label{fig:complex_origins_graph}  DESCRIBE ME}
+\caption{\label{fig:complex_origins_graph}  A Pango lineage subgraph showing
+complex origins of the recombinant lineage XB. All XB samples in the long ARG
+trace to a single recombination node (top centre), but three further recombinations
+also descend from this node, with many samples descending from these nested
+recombinations not being identified as XB, but instead assigned by Nextclade to
+various pre-Alpha lineages.}
 \end{figure}
 
-As well as cases where a single recombination event explains several X lineages
-(such as XAG above), many of the Pango X lineages involve more complex origins,
-in which samples which have not been previously identified as recombinants also
-descend from originating recombination events. In addition these complex
-topologies often include extra recombination nodes within the topology that
-descends from the main originating event. An example is XB in the long ARG,
+As well as cases where single recombination events explain an X lineage,
+such as the examples in Figure~\ref{fig:pango-simple-origin-graph}, many of the
+Pango X lineages involve more complex origins,
+in which additional recombination events are nested inside the earliest shared
+recombination. Frequently, the descendants of the original recombination
+also include samples which have not been previously identified as recombinants.
+An example is XB in the long ARG,
 plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
 between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
-"hairball" of further recombination nodes whose descendants have not been
-identified as recombinants by Nextclade (blue nodes, "pre-Alpha" VoC). However,
+``hairball'' of further recombination nodes whose descendants have not been
+identified as recombinants by Nextclade (blue nodes, ``pre-Alpha'' VoC). However,
 in this subgraph, there is only a single sample, identified as XB by Nextclade,
@@
 Figure~\ref{fig:pango_XB_gisaid_graph}). We suggest that investigating cases of
-complex origins, and identifing which (if any) of the nested, further
+complex origins, and identifying which (if any) of the nested, further
 recombination events may be artifactual, is an important area of future
@@
 
-\subsection{Parsimony increasing hueristics}
+\subsection{Parsimony increasing heuristics}
 \label{sec:parsimony-heuristics}
@@
 is room for improvement, and that further parsimony increasing
-hueristics would likely be of benefit.
+heuristics would likely be of benefit.
 

### 2023-05-02 plot.py

@@ -1328,6 +1328,7 @@ class Pango_XB_nxcld_tight_graph(Pango_X_tight_graph):
     def post_process(cls, ax):
         x_min, x_max = ax.get_xlim()
         ax.set_xlim(x_min + (x_max - x_min) * 0.06, x_max - (x_max - x_min) * 0.06)
+        ax.text(0.98, 0.05,  "XB subgraph", fontsize=18, transform=ax.transAxes, ha='right')
 
     @staticmethod
     def adjust_positions(pos):

### 2023-05-02 paper.tex
Briefly define HMM consistency when we first say it

--- 2023-05-02 paper.tex
@@
 recombination nodes, except MILK-103C712, which was removed during
-preprocessing (Table 2); additionally, all are HMM-consistent, except
-QEUH-1067DEF. Overall, we find excellent agreement between Jackson et al.’s
+preprocessing (Table 2). Additionally, all except QEUH-1067DEF are HMM-consistent:
+that is, the mirrored HMM runs (i.e. forward from 5’ to 3’ end versus backwards from 3’ to 5’ end)
+identify the same set of originating mutations, have the same number of breakpoints, and have
+equivalent (sequence-identical) parents from forward and backward runs.
+Overall, we find excellent agreement between Jackson et al.’s
 motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
@@
 Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are
-"HMM-consistent"; that is, where the mirrored HMM runs (i.e., 5’ to 3’ end
-versus 3’ to 5’ end) identify the same set of originating mutations and the
-same number of breakpoints, and the parents in the mirrored runs are
-sequence-identical. This reduces both the number of recombination nodes /
+HMM-consistent. This reduces both the number of recombination nodes /
 breakpoints from 1522 / B to C / D in the wide ARG and from 763 / 806 to 459 in

### 2023-05-03 paper.tex
Yans' updates

--- 2023-05-02 paper.tex
@@
 of our ARGs are almost entirely restricted to non-recombinant lineages. To
-allow for comparison, the ARGs were ``simplified'' \citep{Kelleher2018-xc} to
+enable this comparison, the ARGs were ``simplified'' \citep{Kelleher2018-xc} to
 retain only those samples of our genealogy that match those present in a
@@
 
-\subsection{Tree inference from HMM sample clusters}
+\subsection{Tree inference from HMM daily sample clusters}
 \label{sec:sample-cluster-tree-inference}

### 2023-05-04 paper.tex
Yan's updates

@@
 \subsection{Backbone phylogeny}
-For validation purposes, we compare the genetic
-relationships in the sc2ts-inferred ARGs with phylogenetic relationships
-inferred by the NextStrain project \citep{Hadfield2018-ef}. The NextStrain
-trees consist of subsamples of SARS-CoV-2 genomes, which over the time period
-of our ARGs are almost entirely restricted to non-recombinant lineages. To
-enable this comparison, the ARGs were ``simplified'' \citep{Kelleher2018-xc} to
-retain only those samples of our genealogy that match those present in a
-specific NextStrain tree. These simplified ``backbone'' ARGs exhibit very
-little recombination (6 and 5 trees across the genome for the wide and long
-ARGs, respectively).
 
@@
 \end{figure}
+
+For validation purposes, we compare the genetic
+relationships in the sc2ts-inferred ARGs with phylogenetic relationships
+inferred by the NextStrain project \citep{Hadfield2018-ef}. The NextStrain
+trees consist of subsamples of SARS-CoV-2 genomes, which over the time period
+of our ARGs are almost entirely restricted to non-recombinant lineages. To
+enable this comparison, the ARGs were ``simplified'' \citep{Kelleher2018-xc} to
+retain only those samples of our genealogy that match those present in a
+specific NextStrain tree. These simplified ``backbone'' ARGs exhibit very
+little recombination (6 and 5 trees across the genome for the wide and long
+ARGs, respectively).
 
@@
 Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
-imputed Pango status. (B) Equivalent plot for the 17 XAG samples in the long ARG
-(C) Equivalent plot for both origination events for the four XD samples in the long ARG.
+imputed Pango status. (B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
+(C) Equivalent plot for both origination events involving the four XD samples in the long ARG.
+Details of the mutations and sample node identities for all three plots are provided in supplementary figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
 } \end{figure}
@@
 trace to a single recombination node (top centre), but three further recombinations
-also descend from this node, with many samples descending from these nested
-recombinations not being identified as XB, but instead assigned by Nextclade to
-various pre-Alpha lineages.}
+also descend from this node. The samples descending from these nested
+recombinations include 9 that are assigned by Nextclade to
+various non-recombinant pre-Alpha lineages (blue).}
 \end{figure}
@@
 between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
-``hairball'' of further recombination nodes whose descendants have not been
-identified as recombinants by Nextclade (blue nodes, ``pre-Alpha'' VoC). However,
-in this subgraph, there is only a single sample, identified as XB by Nextclade,
-which is descended from these nested recombination events; moreover, this
-sample is not identified as XB by GISAID (Supplementary
-Figure~\ref{fig:pango_XB_gisaid_graph}). We suggest that investigating cases of
-complex origins, and identifying which (if any) of the nested, further
-recombination events may be artifactual, is an important area of future
-research.
+``hairball'' of further recombination nodes whose descendants are mostly not
+identified as recombinants by Nextclade, and plotted in blue. A similar
+pattern is seen when examining XB in the wide ARG (discussion above, plot not shown).
+Note that the descendants of nested recombination events include only one
+XB sample (pink upper left, with 7 mutations above it); moreover, this sample is
+not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph})
+indicating some uncertainty in lineage assignment in this part of the ARG.
+Investigating examples of complex recombinant origins, and identifying which
+(if any) of the nested recombination events may be artifactual, is an important
+area of future research.
 

### 2023-05-04 analyse.py
First pass at a script to output stats used in the paper

@@ -1,0 +1,139 @@
+import argparse
+
+import numpy as np
+import tszip
+
+import sc2ts
+
+class CompressedSc2tsARG:
+    # Keep details of the compressed ARG so we can extract e.g. compressed sizes
+    def __init__(self, fn):
+        if not fn.endswith(".tsz"):
+            raise ValueError("Requires a tszip compressed file ending in .tsz")
+        self.fn = fn
+        self.ts = tszip.decompress(fn)
+        self.treeinfo = None
+        
+    @property
+    def ti(self):
+        # Slow to calculate, so cache it
+        if self.treeinfo is None:
+            self.treeinfo = sc2ts.TreeInfo(self.ts)
+        return self.treeinfo
+
+    def compressed_bytes(self):
+        # return compressed size in bytes
+        raise NotImplementedError  # TODO
+        #return tszip.print_summary()
+
+
+def perc(x, tot):
+    # return percentage, formatted nicely
+    return f"{x/tot * 100:.2g}%"
+
+class Total:
+    # Keep track of a total, and print values as a percentage of the total
+    def __init__(self, total):
+        self._total = total
+    def suffix(self, x, suffix=None):
+        # print with "M", "k", etc
+        if suffix is None:
+            return str(x)
+        elif suffix == "k":
+            return f"{x/1000:.3g}k"
+        elif suffix == "M":
+            return f"{x/1000_000:.3g}M"
+    def total(self, suffix=None):
+        return self.suffix(self._total, suffix)
+    def x_perc(self, x, suffix=None):
+        return f"{self.suffix(x)} {perc(x, self._total)}"
+
+def tot(df, *rownames):
+    # return sum of named rows in the df, or just a single row value
+    return df.loc[list(rownames)].sum().value
+
+def prt(*args):
+    # shortcut for printing separated by a tab
+    print(*args, sep="\t")
+
+def print_stats(wide, long, use_treeinfo=True):
+    if use_treeinfo:
+        summaryW = wide.ti.summary()
+        summaryL = long.ti.summary()
+    prt("* Results")
+    prt("** Inferred ARGs")
+    prt("", "", "Wide ARG", "Long ARG")
[…]

### 2023-05-04 analyse.py
Add sizes
And correct long_ts naming bug

@@ -1,4 +1,5 @@
 import argparse
+import os.path
 
 import numpy as np
 import tszip
@@ -22,9 +23,7 @@ class CompressedSc2tsARG:
         return self.treeinfo
 
     def compressed_bytes(self):
-        # return compressed size in bytes
-        raise NotImplementedError  # TODO
-        #return tszip.print_summary()
+        return os.path.getsize(self.fn)
 
 
 def perc(x, tot):
@@ -118,7 +117,7 @@ def print_stats(wide, long, use_treeinfo=True):
     #prt("", "3+ breaks",
     
     
-    # SIZES
+    # SIZES - currently quoted in MB not MiB - I think this is more useful to the layman
     prt("Size (inc metadata)", Total(wide.compressed_bytes()).total("M")+"B", Total(long.compressed_bytes()).total("M")+"B", )
 
 
@@ -129,7 +128,7 @@ def main():
 
     wide = CompressedSc2tsARG("data/upgma-full-md-30-mm-3-2021-06-30-recinfo-gisaid-il.ts.tsz")
     print("Read wide ARG")
-    long_ts = CompressedSc2tsARG("data/upgma-mds-1000-md-30-mm-3-2022-06-30-recinfo-gisaid-il.ts.tsz")
+    long = CompressedSc2tsARG("data/upgma-mds-1000-md-30-mm-3-2022-06-30-recinfo-gisaid-il.ts.tsz")
     print("Read long ARG")
     print("-------------")
 

### 2023-05-04 paper.tex
Updates from Duncan and Yan

@@
 \usepackage{url}
+% For rotating figures, tables, etc.
+%  including their captions - only for supplementary figures
+\usepackage{rotating}
+
 % JK: turning this off for the moment as I keep clicking through on links
@@
 such as SARS-CoV-2 (with currently over 15 million sequences in the GISAID database
-\citep{Shu2017-hp}). Moreover the ``tskit'' software library, which implements this
+\citep{Shu2017-hp}). Moreover the \emph{tskit} software library, which implements this
 data structure, enables efficient manipulation of the stored genealogies and is
@@
 of modern and ancient human genomes, accounting for millions of recombination
-events (over 5 million "local trees"). To achieve this, \cite{Wohns2022-th}
+events (over 5 million ``local trees''). To achieve this, \cite{Wohns2022-th}
 took advantage of a highly efficient Hidden Markov Model (HMM) implementation
@@
 Here, we present a preliminary version of a novel scalable methodology
-(``sc2ts'', optionally pronounced ``scoots'') that builds on the advances above
+(\emph{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
 and additionally takes advantage of known collection dates to infer ARGs from millions
@@
 of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
-"tskit" (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
+\emph{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
 Sc2ts is a promising method to automate detection of SARS-CoV-2 recombinants
@@
 \subsection{Inferred ARGs}
-For this prelimary study our goals are to illustrate the utility of
+For this preliminary study our goals are to illustrate the utility of
 \texttt{sc2ts}, but to also closely study the properties of the
@@
 
-To investigate the properties of the inferred ARGs under different
+To investigate the properties of the method under different
 sampling conditions, we inferred two ARGs, which we refer to as the
@@
 includes all 1.27 million sample genomes  with a
-maximum sampling delay of 30 days (see Methods) up to June 30, 2021.
+maximum sampling delay of 30 days (see Methods Section~\ref{sec:filtering_time_travellers}) 
+up to June 30, 2021.
 The Long ARG is more sparsely sampled but
@@
 equivalent (sequence-identical) parents from forward and backward runs.
-Overall, we find excellent agreement between Jackson et al.’s
+Overall, we find excellent agreement between Jackson et al.'s
 motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
@@
 recombination event.
-
-\begin{table}
-% \begin{footnotesize}
-\begin{tabular}{lllrl}
-\toprule
-{} &              parent\_lineages &          pango\_parents &   n &
-descendant\_lineages \\
-lineage &                              &                        &     &
-\\
-\midrule
-XA      &        [B.1.177.18, B.1.1.7] &     [B.1.1.7, B.1.177] &   5 & XA: 5 \\
-XAA     &               [BA.1, BA.2.9] &         [BA.1*, BA.2*] &   8 &
-    XAB: 39, XAG: 17, XAA: 8 \\
-    &&&&   XAA: 8, XQ: 2, XU: ... \\
-XAC     &          [BA.2.3, BA.1.17.2] &  [BA.2*, BA.1*, BA.2*] &  27 & XAC: 27 \\
-XAE     &                 [BA.2, BA.1] &         [BA.2*, BA.1*] &   9 & XAE: 9 \\
-XAG     &               [BA.1, BA.2.9] &         [BA.1*, BA.2*] &  17 &
-    XAB: 39, XAG: 17, XAA: 8 \\
-    &&&&   XAA: 8, XQ: 2, XU: ... \\
-XF      &                 [AY.4, BA.1] &    [B.1.617.2*, BA.1*] &   2 & XF: 2 \\
-XG      &              [BA.1.17, BA.2] &         [BA.1*, BA.2*] &  32 & XG: 32, XAB: 1 \\
-XH      &            [BA.1.20, BA.2.9] &         [BA.1*, BA.2*] &  11 &
-    XAF: 34, XH: 11 \\
-&&&& B.1.1.529: 3, XE: 3,... \\
-XK      &             [BA.1.1.1, BA.2] &         [BA.1*, BA.2*] &   6 & XK: 3 \\
-XL      &            [BA.1.17.2, BA.2] &         [BA.1*, BA.2*] &  10 &
-    XL: 10, XAB: 1, XU: 1, \\
-    &&&& XQ: 1 \\
-XR      &               [BA.1.1, BA.2] &       [BA.1.1*, BA.2*] &   8 &
-    XQ: 9, XR: 8, XAB: 2\\
-XS      &              [AY.36, BA.1.1] &  [B.1.617.2*, BA.1.1*] &   6 & XS: 4\\
-XT      &       [BA.2.23, Unknown (R)] &         [BA.2*, BA.1*] &   1 &
-    BA.2: 1, BA.1: 1, BA.1.9: 1, \\
-    &&&& BA.2.23:... \\
-XV      &  [BA.1.1, BA.2, Unknown (R)] &         [BA.1*, BA.2*] &   3 & BA.2: 8, XV: 1 \\
-XW      &            [BA.1.1.15, BA.2] &         [BA.1*, BA.2*] &  11 & XW: 11, XN: 2 \\
-XY      &               [BA.1.1, BA.2] &         [BA.1*, BA.2*] &   6 & XY: 6, XAF: 1 \\
-\bottomrule
-\end{tabular}
-\caption{\label{tab:pango-single-origin}
-Table summarising the status of the Pango X-lineages in the Long ARG thare inferred to have a single recombinant origin based on the NextClade Pango lineage assignments.
-Imputed parent lineages are shown from the ARG and also to parent lineages as defined by the official Pango designation alias key.
-FIXME: if we like this table then we can reformat it so that lists at the end are spread over multiple lines rather than using Python dict notation }
-\end{table}
 
@@
 \begin{figure}
-
+\centering
 \begin{tabular}{llll}
@@
 \midrule
-BA.1.15 & BA.2  &9867  &10198\\
-BA.1.15 & BA.2  &9867  &10198\\
-BA.1.1  &BA.2  &10451  &11537\\
-BA.1.1.5 & BA.2 &10448 & 11537\\
-BA.1.15 & BA.2  &10448  &12880\\
-BA.1.17 & BA.2  &10448 & 11537\\
-BA.1  &BA.2.9  &10448 & 11537\\
-BA.1  &BA.2  &12764  &13195\\
+BA.1.15  & BA.2   & 9,867  & 10,198 \\
+BA.1.15  & BA.2   & 9,867  & 10,198 \\
+BA.1.1   & BA.2   & 10,451 & 11,537 \\
+BA.1.1.5 & BA.2   & 10,448 & 11,537 \\
+BA.1.15  & BA.2   & 10,448 & 12,880 \\
+BA.1.17  & BA.2   & 10,448 & 11,537 \\
+BA.1     & BA.2.9 & 10,448 & 11,537 \\
+BA.1     & BA.2   & 12,764 & 13,195 \\
 \bottomrule
@@
 \caption{\label{tab:multiple_origins_table}
-details of inferred
+Details of inferred
 breakpoints for different recombination events inferred for XE samples (170
@@
 
-``The rate of putative recombination breakpoints is about three times higher towards the 3' of the change point than the 5' interval.`` Consistent with this, the breakpoint intervals of the putative recombinants identified here tend to occur on the 3’ end of the SARS-CoV-2 genome.
+``The rate of putative recombination breakpoints is about three times higher towards the 3' of the change point than the 5' interval.'' Consistent with this, the breakpoint intervals of the putative recombinants identified here tend to occur on the 3’ end of the SARS-CoV-2 genome.
 
@@
 \item A potential source of error in sc2ts ARGs comes from incorrectly reported collection dates of the samples, which can cause sample genomes to be attached to the wrong parts of the ARGs.
-Thus far, we have attempted to correct the most offending errors in the ARGs by filtering out "time travellers", but the sampling time can still be wrong by days or tens of days.
+Thus far, we have attempted to correct the most offending errors in the ARGs by filtering out ``time travellers'', but the sampling time can still be wrong by days or tens of days.
 ***describe possible approaches to correct this?***
@@
 ``copying path'' through the reference panel for a given query sequence that
-maximises the likelihood under these parameters.
+maximises the likelihood under these parameters. Unseen states in the reference panel are shown as coloured lines enclosed by
+the grey box. The black arrow describes the true path through the data which leads to the emitted
+focal sequence below. Examples of transition and
+emission probabilities along this trajectory are shown by the red and blue
+arrows, respectively.
+}
+\end{figure}
+
 % The LS HMM is governed by the transition and
 % emission matrices, $Q$ and $E$, respectively.
-% Each may be a function of the
-% reference panel members, but transitions are generally assumed to be
-% independent of the hidden states themselves (shown in the pink panel). This
-% dramatically increases performance as the state space drops to two states
-% (i.e., switching or not switching).
-% Emissions may also be a function of the nucleotide states, but in our
-% case, we assume that all observed alleles (in both the reference panel and
-% focal sequence, $a\ell$) at the site under consideration are equally likely and
-% occur with mutation probability $\ell$ (shown in the blue panel).
+% 
 % These two
@@
 % inserted at earlier collection dates (shown in the lower panel). Here, the
-% unseen states in the reference panel are shown as coloured lines enclosed by
-% the grey box, and the true path through the data which leads to the emitted
-% focal sequence is displayed by the black arrow. Examples of transition and
-% emission probabilities along this trajectory are shown by the red and blue
-% arrows, respectively.
+% 
 % Note that in practice, we do not know the pattern of
 % colours along the focal sequence and must infer it using the Viterbi algorithm.
-}
-\end{figure}
 
@@
 focal sequence and the hidden state from which it is copied (due to mutation as
-well as genotyping or sequencing error). Then, we use the Viterbi algorithm
+well as genotyping or sequencing error). Both $E$ and $Q$ may be a function of the reference panel members, 
+but transitions are generally assumed to be independent of the hidden states (Figure~\ref{fig:ls_diagram}, pink panel). 
+This assumption dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
+Emissions may also be a function of the nucleotide states, but in our
+case, we assume that all observed alleles (in both the reference panel and
+focal sequence, $a\ell$) at the site under consideration are equally likely and
+occur with mutation probability $\frac{\mu_\ell}{a_\ell-1}$ (Figure~\ref{fig:ls_diagram}, blue panel).
+We use the Viterbi algorithm
 \citep{Viterbi1967-ol} to find the most likely copying path through the
@@
 the site-specific parameters $\mu_\ell$ and $r_\ell$, respectively.
-For this work we used a simpler formulation, which uses a single parameter
-(the mismatch ratio, MMR) to control the relative importance of mutation
+For this work we used a slightly different formulation, which uses a parameter, the mismatch ratio (MMR), to control the relative importance of mutation
 and recombination in the HMM. Specifically, an MMR value of $k$ will
-prefer $k$ mismatches (mutations) to a recombination.[DUNCAN MORE DETAILS
-HERE PLEASE]
-
-[ALSO MENTION MISSING DATA, where we deal with Ns in the alignment]
-
+prefer $k$ mismatches (mutations) to a single recombination which results in copying from a template with no mismatches.
+To map between recombination probability and mutation probability for a particular mismatch ratio, we simply consider the two paths which we wish to be equally likely, and rearrange for the mutation or recombination probability.
+In the simple case when we assume $\mu_l=\mu$ and $r_l=r$, we obtain (supplementary methods):
+\begin{align*}
+r = \frac{n\mu^k}{(1-\mu)^k + (n-1)\mu^k}, \quad \quad
+\mu = \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}.
+\end{align*}
+Under a lower MMR value (for which recombinations are favoured over mutations), the maximum-likelihood HMM copying path of a
+focal genome more likely spans at least two ancestral genomes (i.e,. the focal
+genome is a putative recombinant) than a single ancestral genome. Here, the MMR 
+was set to three in all the HMM runs.
+
+A varying degree of missingness is present in the genetic data, that is, for a given sample, a subset of nucleotides coded as `unknown' at a set of positions.
+As a result, we had to consider missingness in our implementation of LS Viterbi.
+One approach would be to `impute' missing data, and hard-call or define some distribution over the possible missing nucleotide, and use that in the determination of path probabilities.
+Another is to not attempt to estimate the true nucleotide, but rather to consider such sites as uninformative to the path probability.
+We chose the latter.
+To do this, we simply encode the emission probability where the underlying nucleotide transition is $\mathit{unknown}\rightarrow \{A,C,G,T\}$, $\{A,C,G,T\}\rightarrow\mathit{unknown}$, or $\mathit{unknown}\rightarrow\mathit{unknown}$ at 1.
+
+% [ALSO MENTION MISSING DATA, where we deal with Ns in the alignment]
 % OLD text, ready to be adapted (and deleted once done)
@@
 % preferred over recombination (i.e., switching between sequences) during
-% copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
-% focal genome more likely spans at least two ancestral genomes (i.e,. the focal
-% genome is a putative recombinant) than a single ancestral genome. Here, the MMR
-% was set to three in all the HMM runs .
+% copying.
 
 In \texttt{sc2ts} we use the efficient ARG-based implementation of the
-LS Viterbi algorithm from \texttt{tsinfer}~\citep{Kelleher2019-ba} to find
+LS Viterbi algorithm from \texttt{tsinfer} \citep{Kelleher2019-ba} to find
 the most likely copying path for each sample sequence
@@
 HMM is to copy from more than one ARG node along the genome, for a
-given sample sequece. [TODO MORE]
+given sample sequence. [TODO MORE]
 
@@
 gaps within a distance of seven bases, are masked as missing data following
-the approach used in the ``faToVCF`` tool used by
+the approach used in the ``faToVCF'' tool used by
 UShER~\citep{Turakhia2022-it}.
@@
 \section*{Supplementary}
+\subsection*{Supplementary methods}
+\subsubsection*{Derivation of relation between recombination probability and mismatch probability conditional on mismatch ratio}
+Without loss of generality, consider a region of length $m$. Up until this region, two paths are equally likely, so we can re-scale by the likelihood up to the site before the region starts.
+
+For a MMR of $k$, we only need to consider two of the $n$ members of the reference panel.
+One; $\rho_1$, for which there are no mismatches, but we need to recombine in, and a second; $\rho_2$, for which there are $k$ (randomly chosen) mismatches in the region.
+
+We can simply follow the probability of tracking along each of those paths:
+\begin{align*}
+P(\rho_1) &= c \frac{r}{n} \left(\left(1-r\right)+\frac{r}{n}\right)^m\left(1-\mu\right)^m &\text{recombine to a template with no mismatches,}\\
+P(\rho_2) &= c \left(\left(1-r\right) + \frac{r}{n}\right)\left(\left(1-r\right)+\frac{r}{n}\right)^m \left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
+\end{align*}
+where $r$ is the recombination probability between adjacent variants, $\mu$ is the mismatch probability between a target sequence at a variant and a particular member of the reference panel, and $c$ is the likelihood of the path up to this point, which we assume to be equal.
+We then set these path probabilities equal, and rearrange to obtain:
+\begin{align*}
+r &= \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k},\\
+\mu &= \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}
+\end{align*}
 
@@
 
+\begin{table}
+% \begin{footnotesize}
+\begin{tabular}{lllrl}
+\toprule
+lineage & ARG parents & Official Pango parents & n & Other descendants \\
+\midrule
+XA      &       B.1.1.7 + B.1.177.18 &     B.1.1.7 + B.1.177 &   5 & -- \\
+XAA, XAG     &              BA.1 + BA.2.9 &         BA.1* + BA.2* &   8 &
+    XAB: 39, XAG: 17, XAA: 8 \\
+    &&&&   XAA: 8, XQ: 2, XU: ... \\
+XAC     &          [BA.2.3, BA.1.17.2] &  [BA.2*, BA.1*, BA.2*] &  27 & -- \\
+XAE     &                 [BA.2, BA.1] &         [BA.2*, BA.1*] &   9 & XAE: 9 \\
+XAG     &               [BA.1, BA.2.9] &         [BA.1*, BA.2*] &  17 &
+    XAB: 39, XAG: 17, XAA: 8 \\
+    &&&&   XAA: 8, XQ: 2, XU: ... \\
+XF      &                 [AY.4, BA.1] &    [B.1.617.2*, BA.1*] &   2 & XF: 2 \\
+XG      &              [BA.1.17, BA.2] &         [BA.1*, BA.2*] &  32 & XG: 32, XAB: 1 \\
+XH      &            [BA.1.20, BA.2.9] &         [BA.1*, BA.2*] &  11 &
+    XAF: 34, XH: 11 \\
+&&&& B.1.1.529: 3, XE: 3,... \\
+XK      &             [BA.1.1.1, BA.2] &         [BA.1*, BA.2*] &   6 & XK: 3 \\
+XL      &            [BA.1.17.2, BA.2] &         [BA.1*, BA.2*] &  10 &
+    XL: 10, XAB: 1, XU: 1, \\
+    &&&& XQ: 1 \\
+XR      &               [BA.1.1, BA.2] &       [BA.1.1*, BA.2*] &   8 &
+    XQ: 9, XR: 8, XAB: 2\\
+XS      &              [AY.36, BA.1.1] &  [B.1.617.2*, BA.1.1*] &   6 & XS: 4\\
+XT      &       [BA.2.23, Unknown (R)] &         [BA.2*, BA.1*] &   1 &
+    BA.2: 1, BA.1: 1, BA.1.9: 1, \\
+    &&&& BA.2.23:... \\
+XV      &  [BA.1.1, BA.2, Unknown (R)] &         [BA.1*, BA.2*] &   3 & BA.2: 8, XV: 1 \\
+XW      &            [BA.1.1.15, BA.2] &         [BA.1*, BA.2*] &  11 & XW: 11, XN: 2 \\
+XY      &               [BA.1.1, BA.2] &         [BA.1*, BA.2*] &   6 & XY: 6, XAF: 1 \\
+\bottomrule
+\end{tabular}
+\caption{\label{tab:pango-single-origin}
+Summary of the Pango X-lineages in the long ARG that are inferred to have a single recombinant origin based on the NextClade Pango lineage assignments.
+Parent lineages from the ARG (as designated by Nextclade and, where relevant and also to parent lineages as defined by the official Pango designation alias key.
+(TODO: explain the asterisks) FIXME: if we like this table then we can reformat it so that lists at the end are spread over multiple lines rather than using Python dict notation }
+\end{table}
+
 \begin{figure} \centering
-\includegraphics[width=\textwidth]{figures/Pango_XA_gisaid_large_graph.pdf}
-\caption{\label{fig:pango_XA_gisaid_graph}  DESCRIBE ME}
+\includegraphics[width=0.5\textwidth]{figures/Pango_XA_gisaid_large_graph.pdf}
+\caption{\label{fig:pango_XA_gisaid_graph}
+Detailed version of Figure~\ref{fig:pango-simple-origin-graph}A. All single nucleotide mutations are listed
+with the initial nucleotide state, followed by the reference genome position, followed by the derived
+nucleotide state (after mutation). Sample nodes are shown with \emph{tskit} IDs, which can be mapped to
+GISAID EPI ISL identifiers and strain names using supplementary file XXX. In contrast to
+Figure~\ref{fig:pango-simple-origin-graph}A, Pango lineages shown here are those assigned by GISAID
+rather than Nextclade. In the case of XA, Nextclade and GISAID agree completely on the lineage designations.
+}
 \end{figure}
 
-\begin{figure} \centering
+\begin{sidewaysfigure} \centering
 \includegraphics[width=\textwidth]{figures/Pango_XAG_gisaid_large_graph.pdf}
-\caption{\label{fig:pango_XAG_gisaid_graph}  DESCRIBE ME}
-\end{figure}
+\caption{\label{fig:pango_XAG_gisaid_graph}
+Detailed version of Figure~\ref{fig:pango-simple-origin-graph}B, with node and mutation labels as in
+Figure~\ref{fig:pango_XA_gisaid_graph}. 
+}
+\end{sidewaysfigure}
 
@@
 \includegraphics[width=\textwidth]{figures/Pango_XD_gisaid_large_graph.pdf}
-\caption{\label{fig:pango_XD_gisaid_graph}  DESCRIBE ME}
+\caption{\label{fig:pango_XD_gisaid_graph}
+Detailed version of Figure~\ref{fig:pango-simple-origin-graph}C , with node and mutation labels as in
+Figure~\ref{fig:pango_XA_gisaid_graph}
+}
 \end{figure}
 
-\begin{figure} \centering
+\begin{sidewaysfigure} \centering
 \includegraphics[width=\textwidth]{figures/Pango_XB_gisaid_large_graph.pdf}
-\caption{\label{fig:pango_XB_gisaid_graph}  DESCRIBE ME}
-\end{figure}
+\caption{\label{fig:pango_XB_gisaid_graph}
+Detailed version of Figure~\ref{fig:complex_origins_graph}, with node and mutation labels as in
+Figure~\ref{fig:pango_XA_gisaid_graph}
+}
+\end{sidewaysfigure}
 

### 2023-05-04 paper.tex
Updates from Overleaf

@@
 Detailed version of Figure~\ref{fig:pango-simple-origin-graph}B, with node and mutation labels as in
-Figure~\ref{fig:pango_XA_gisaid_graph}. 
+Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file XXX to map tskID to EPI_ISL and strain name).
+Note that GISAID Pango designations render XAG is monophyletic a
 }
@@
 \begin{figure} \centering
-\includegraphics[width=\textwidth]{figures/Pango_XD_gisaid_large_graph.pdf}
+\includegraphics[width=0.6\textwidth]{figures/Pango_XD_gisaid_large_graph.pdf}
 \caption{\label{fig:pango_XD_gisaid_graph}
 Detailed version of Figure~\ref{fig:pango-simple-origin-graph}C , with node and mutation labels as in
-Figure~\ref{fig:pango_XA_gisaid_graph}
+Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file XXX to map tskID to EPI_ISL and strain name).
 }
@@
 Detailed version of Figure~\ref{fig:complex_origins_graph}, with node and mutation labels as in
-Figure~\ref{fig:pango_XA_gisaid_graph}
+Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file XXX to map tskID to EPI_ISL and strain name).
 }

### 2023-05-04 paper.tex
fixes

--- 2023-05-04 paper.tex
@@
 Detailed version of Figure~\ref{fig:pango-simple-origin-graph}B, with node and mutation labels as in
-Figure~\ref{fig:pango_XA_gisaid_graph}. 
+Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file XXX to map tskID to EPI\_ISL and strain name).
+Note that GISAID Pango designations assign an extra 2 samples to XAG, which are labelled as XAB
+by Nextclade. This renders XAG fully monophyletic. However, unlike Nextclade, the GISAID designations
+also mark many of the samples in this subgraph as non-recombinants (designating them BA.2), and in general
+we find that Nextclade assignments agree more with our ARG structure than GISAID assignments.
 }
@@
 \begin{figure} \centering
-\includegraphics[width=\textwidth]{figures/Pango_XD_gisaid_large_graph.pdf}
+\includegraphics[width=0.6\textwidth]{figures/Pango_XD_gisaid_large_graph.pdf}
 \caption{\label{fig:pango_XD_gisaid_graph}
 Detailed version of Figure~\ref{fig:pango-simple-origin-graph}C , with node and mutation labels as in
-Figure~\ref{fig:pango_XA_gisaid_graph}
+Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file XXX to map tskID to EPI\_ISL and strain name).
+Note that GISAID does not designate any nodes as XD in the long ARG, hence no recmbninant Pango designations
+are marked in this plot. From inspection of the samples, we believe the GISAID designations to be erroneous
+in this case.
 }
@@
 Detailed version of Figure~\ref{fig:complex_origins_graph}, with node and mutation labels as in
-Figure~\ref{fig:pango_XA_gisaid_graph}
+Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file XXX to map tskID to EPI\_ISL and strain name).
 }

### 2023-05-04 paper.tex
Update values for nodes in Table 1

@@
     Nodes & & 1,453,347 & & 783,231  & \\
-    & Samples & 1,265,685 & (87\%) & 657,239 & (84\%) \\
-    & Local (UPGMA) & 103K & (7.1\%) & 52K & (6.6\%)  \\
-    & Tree adjustment & 80,830 & (5.6\%) & 72,107 & (9.2\%) \\
-    & Recombination & 4,123 & (0.33\%) & 2,078 & (0.27\%) \\
+    & Samples & 1,265,685 & (87.09\%) & 657,239 & (83.91\%) \\
+    & Local (UPGMA) & 102,709 & (7.07\%) & 51,807 & (6.61\%)  \\
+    & Reversion push & 40,538 & (2.79\%) & 34,358 & (4.39\%)  \\
+    & Mutation collapse & 40,292 & (2.77\%) & 37,749 & (4.82\%)  \\
+    & Recombination & 4,123 & (0.28\%) & 2,078 & (0.27\%) \\
     \cmidrule{2-6}
@@
 samples that are inferred to have multiple recombinant origins (i.e., trace
-back to different recombination nodes in the ARG), but still form monophyletic
-clades % is this the right term? descending from these nodes .
+back to independent recombination nodes in the ARG).
 
@@
 the 4 samples labelled XD by Nextclade in the long ARG. The left hand subgraph
-(the origin of three XD samples) has an earliest
+(the origin of three XD samples, all sampled in France) has an earliest
 sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
-subgraph has a single sample with EPI\_ISL 8514045 (dated 2021-12-30). These
-two samples differ by 35 mutations; considering this, and the time between the samples,
+subgraph has a single sample with EPI\_ISL 8514045 (sampled in Turkey and dated 2021-12-30).
+These two samples differ by 35 mutations; considering this, plus the geographical
+locations and time between the samples,
 it does seem possible that they represent true independent origins of XD.
@@
 
-Table \ref{tab:multiple_origins_table} summarises other Pango recombinants which
+Another X lineage with multiple originating events is XE. In this case, however,
+there are a large number of independent events. These
+
+\ref{tab:multiple_origins_table} summarises other Pango recombinants which
 we infer to have multiple origins. [TODO: needs more detail]
+As with XD, none of the samples in the long ARG are identified as XE by GISAID.
 
@@
 Pango X lineages involve more complex origins,
-in which additional recombination events are nested inside the earliest shared
-recombination. Frequently, the descendants of the original recombination
-also include samples which have not been previously identified as recombinants.
+in which multiple non-independent (``nested'') recombination events are involved.
+In these cases, the descendants of the original recombination
+frequently include samples which have not been previously identified as recombinants.
 An example is XB in the long ARG,
@@
 Investigating examples of complex recombinant origins, and identifying which
-(if any) of the nested recombination events may be artifactual, is an important
+(if any) of the nested recombination events may be artefactual, is an important
 area of future research.
@@
 \label{sec:ls}
-The Li and Stephens (LS) model\citep{Li2003-ib}  is an approximation of the
+The Li and Stephens (LS) model \citep{Li2003-ib}  is an approximation of the
 coalescent with recombination~\citep{Hudson1983-properties}, which captures
@@
 are identical within that interval, and there is therefore no information
-about where the recombination occured. There are many different processes
+about where the recombination occurred. There are many different processes
 that can contribute to the size of these intervals, and therefore
@@
 \subsubsection*{Derivation of relation between recombination probability and mismatch probability conditional on mismatch ratio}
-Without loss of generality, consider a region of length $m$. Up until this region, two paths are equally likely, so we can re-scale by the likelihood up to the site before the region starts.
+Without loss of generality, consider a region of length $m$. Up until this region, two paths are equally likely, so we can re-scale by the likelihood up to the site before the region starts, $c$.
 
@@
 % \begin{footnotesize}
+%% FIXME - commas and braces are confusing - I suggest separating by "+". The code to
+%% produce this table is in Pango-recombinant-analysis.ipynb
 \begin{tabular}{lllrl}
@@
 with the initial nucleotide state, followed by the reference genome position, followed by the derived
-nucleotide state (after mutation). Sample nodes are shown with \emph{tskit} IDs, which can be mapped to
-GISAID EPI ISL identifiers and strain names using supplementary file XXX. In contrast to
-Figure~\ref{fig:pango-simple-origin-graph}A, Pango lineages shown here are those assigned by GISAID
-rather than Nextclade. In the case of XA, Nextclade and GISAID agree completely on the lineage designations.
+nucleotide state (after mutation). If an ancestral (``parent'') mutation exists at the same site in the path
+to the root, it is possible for a mutation to be a ``reversion'' to the initial state of the parent mutation:
+such reversions are indicated by lowercase nucleotide letters. Sample nodes are shown with \emph{tskit} IDs,
+which can be mapped to GISAID EPI ISL identifiers and strain names using supplementary file \protect\path{Pango_X_graph_sample_maps.txt}. In contrast to
+Figure~\ref{fig:pango-simple-origin-graph}A, Pango lineages shown here are those assigned by GISAID rather
+than Nextclade; however, in the specific case of XA, Nextclade and GISAID exactly agree on the lineage designations.
 }
@@
 Detailed version of Figure~\ref{fig:pango-simple-origin-graph}B, with node and mutation labels as in
-Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file XXX to map tskID to EPI\_ISL and strain name).
+Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file \protect\path{Pango_X_graph_sample_maps.txt}
+to map tskID to EPI\_ISL and strain name).
 Note that GISAID Pango designations assign an extra 2 samples to XAG, which are labelled as XAB
-by Nextclade. This renders XAG fully monophyletic. However, unlike Nextclade, the GISAID designations
-also mark many of the samples in this subgraph as non-recombinants (designating them BA.2), and in general
-we find that Nextclade assignments agree more with our ARG structure than GISAID assignments.
+by Nextclade in Figure~\ref{fig:pango-simple-origin-graph}B. This renders XAG fully monophyletic.
+However, unlike Nextclade, the GISAID designations also mark many of the samples in this subgraph
+as non-recombinants (designating them BA.2), and in general we find that Nextclade assignments
+agree more with our ARG structure than GISAID assignments.
 }
@@
 \caption{\label{fig:pango_XD_gisaid_graph}
-Detailed version of Figure~\ref{fig:pango-simple-origin-graph}C , with node and mutation labels as in
-Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file XXX to map tskID to EPI\_ISL and strain name).
+Detailed version of Figure~\ref{fig:pango-simple-origin-graph}C, with node and mutation labels as in
+Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file \protect\path{Pango_X_graph_sample_maps.txt}
+to map tskID to EPI\_ISL and strain name).
 Note that GISAID does not designate any nodes as XD in the long ARG, hence no recmbninant Pango designations
@@
 Detailed version of Figure~\ref{fig:complex_origins_graph}, with node and mutation labels as in
-Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file XXX to map tskID to EPI\_ISL and strain name).
+Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file \protect\path{Pango_X_graph_sample_maps.txt}
+to map tskID to EPI\_ISL and strain name).
+The GISAID XB designations agree with the Nextclade ones except in two cases: an unplotted
+singleton recombinant nested within the main grouping, and the tsk285180 node marked BA.1 in this plot,
+but which is labelled XB by Nextstrain.
 }
@@
 
+%% Could insert the Pango_X_graph_sample_maps.txt with something like this
+%% but maybe best to leave it as a separate file?
+%% \lstinputlisting[caption={XXX}]{data/Pango_X_graph_sample_maps.txt}
+
 \end{document}

### 2023-05-05 paper.tex
Updates from Overleaf

--- 2023-05-04 paper.tex
@@
 Note that GISAID Pango designations assign an extra 2 samples to XAG, which are labelled as XAB
-by Nextclade. This renders XAG fully monophyletic. However, unlike Nextclade, the GISAID designations
-also mark many of the samples in this subgraph as non-recombinants (designating them BA.2), and in general
-we find that Nextclade assignments agree more with our ARG structure than GISAID assignments.
+by Nextclade in Figure~\ref{fig:pango-simple-origin-graph}B. This renders XAG fully monophyletic.
+However, unlike Nextclade, the GISAID designations also mark many of the samples in this subgraph
+as non-recombinants (designating them BA.2), and in general we find that Nextclade assignments
+agree more with our ARG structure than GISAID assignments.
 }
@@
 \caption{\label{fig:pango_XD_gisaid_graph}
-Detailed version of Figure~\ref{fig:pango-simple-origin-graph}C , with node and mutation labels as in
+Detailed version of Figure~\ref{fig:pango-simple-origin-graph}C, with node and mutation labels as in
 Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file XXX to map tskID to EPI\_ISL and strain name).
@@
 Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file XXX to map tskID to EPI\_ISL and strain name).
+The GISAID XB designations agree with the Nextclade ones except in two cases: an unplotted
+singleton recombinant nested within the main grouping, and the tsk285180 node marked BA.1 in this plot,
+but which is labelled XB by Nextstrain.
 }

### 2023-05-05 paper.tex
First pass through "Inferred ARGs" section done

--- 2023-05-04 paper.tex
@@
 For this preliminary study our goals are to illustrate the utility of
-\texttt{sc2ts}, but to also closely study the properties of the
-inferred ARGs to inform subsequent development. Thus, rather
+\texttt{sc2ts}, and to closely study the properties of the
+inferred ARGs to inform subsequent development. Rather
 than attempting to be comprehensive we chose to work with a
@@
 computer hardware used). The cut-off dates for sampling were chosen
-arbitrarily. We anticipate that further development of the
-inference methods and scaling out across multiple servers will
-enable inferences at the full pandemic scale using \texttt{sc2ts}
-(see Discussion).
-
-To investigate the properties of the method under different
-sampling conditions, we inferred two ARGs, which we refer to as the
+arbitrarily.
+We inferred two ARGs, which we refer to as the
 ``Wide'' and ``Long'' ARGs throughout.
@@
 includes all 1.27 million sample genomes  with a
-maximum sampling delay of 30 days (see Methods Section~\ref{sec:filtering_time_travellers})
+maximum sampling delay of 30 days
+(see Section~\ref{sec:filtering_time_travellers})
 up to June 30, 2021.
@@
     Nodes & & 1,453,347 & & 783,231  & \\
+     & \emph{Node type} & \\
     & Samples & 1,265,685 & (87.09\%) & 657,239 & (83.91\%) \\
-    & Local (UPGMA) & 102,709 & (7.07\%) & 51,807 & (6.61\%)  \\
+    & UPGMA   & 102,709 & (7.07\%) & 51,807 & (6.61\%)  \\
     & Reversion push & 40,538 & (2.79\%) & 34,358 & (4.39\%)  \\
@@
     \cmidrule{2-6}
-    % TODO add more information on mutations
-    Mutations & & 1.21M &  & 1.06M & \\
-    & On one sample & 759K & (63\%) & 767K & (72\%) \\
-    & Reversions & 72K & (5.98\%) & 48K & (4.54\%) \\
-    & per node & 0.83  &   & 1.36 \\
+    Edges & & 1,458,146 &  & 785,539 & \\
+     & \emph{Parent node type} & \\
+    & Samples & 610,729 & (41.88\%) & 319,626 & (40.69\%) \\
+    & UPGMA   & 470,545 & (32.27\%) & 156,881 & (19.97\%)  \\
+    & Reversion push & 184,608 & (12.66\%) & 144,991 & (18.46\%)  \\
+    & Mutation collapse &186,218  & (12.77\%) & 160,833 & (20.47\%)  \\
+    & Recombination & 6,046 & (0.41\%) & 3,208 & (0.41\%) \\
     \cmidrule{2-6}
-    Recombination events & \\
-    & Total & 4,123 & & 2,078 & \\
-    & Singletons & 2,601  & (63\%) & 1,315 & (63\%) \\
-    % clearer for many than "local trees?"
-    & Unique breakpoints & 1,496 &  & 958 & \\
+    Mutations & & 1,231,193 &  & 1,062,072 & \\
+    & Per node & 0.83  & $\pm$1.40 & 1.36 & $\pm$1.72 \\
+    & Per sample & 0.77  & $\pm$1.39 & 1.38 & $\pm$1.77 \\
+    & Per site & 41.23  & $\pm$108.16    & 36.10 & $\pm$80.03\\
     \cmidrule{2-6}
-    \multicolumn{2}{l}{Size (inc metadata)} & 57MB & & 36MB&  \\
+    % size as reported by ls -lh
+    \multicolumn{2}{l}{Compressed size (inc metadata)} & 58MB & & 37MB&  \\
     \multicolumn{2}{l}{Bytes/sample (exc metadata)}  & 8.29  & & 10.83 \\
@@
 \end{center}
-
-\caption{\label{tab:args}Summary of the inferred ARGs.
+\caption{\label{tab:args}Summary of the inferred ARGs. Nodes are classified
+as either samples or by the inference process which produced them to aid
+understanding of the relative importances of these processes to the overall
+topology. Similarly, edges are classified by the type of the parent node,
+giving some insights into how often these nodes are subsequently
+copied from after insertion (although the relationship is more complicated
+for UPGMA inserted nodes). The mean and standard deviation ($\pm$)
+are reported for the number of mutations per node, sample and site.
 }
 \end{table}
-
-% \centering
-% \begin{tabular}{l|l|l} \hline \textbf{Property} &
-% \textbf{Wide ARG} & \textbf{Long ARG} \\ \hline Last sampling date & 2021-06-30
-% & 2022-06-30 \\ Local trees & 1,496 & 958 \\ \hline
-% \multicolumn{3}{l}{\textit{Nodes}} \\ All nodes (sample and non-sample) &
-% 1,453,347 & 783,231 \\ Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\ Tree
-% adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107
-% (10.97\%) \\ Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\ \hline
-% \multicolumn{3}{l}{\textit{Types of recombination nodes}} \\ Singletons  & 2,601
-% (63\%) & 1,315 (63\%) \\ Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%)
-% \\ HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\ \hline
-% \multicolumn{3}{l}{\textit{Mutations}} \\ Total mutations & 1,213,193 &
-% 1,062,072 \\ Mutations on a terminal edge & 758,903 (62.55\%) & 767,111
-% (72.23\%) \\ Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
-% Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\ Immediate reversions (one
-% intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\ \hline
-% \multicolumn{3}{l}{\textit{Storage and loading time}} \\ tszipped file
-%  & ~57 Mb & ~36 Mb \\ Bytes to encode one
-% sample & 8.29 & 10.83 \\ Load time (seconds) & 0.9 & 0.5 \\ \hline
-% \end{tabular}
 
@@
 Most of the nodes in the ARGs represent sample genomes
-Only a small fraction of the nodes represent recombination events (0.28\% and
-0.27\% in the wide ARG and long ARG, respectively), over half of which have no
-descendant samples (besides the recombinant sequence that triggered insertion
-of the recombination node). There are over one million mutations on
-the ARGs, the majority of which occur on a terminal edge. Notably,
-the massive amount of genomic variation data in the wide ARG (1.27 million
-genomes) and long ARG (657 thousand genomes) are compactly stored in ~57 Mb and
-~36 Mb, respectively (less than ten bytes per sample). These ARGs can be loaded
-for analysis in a Jupyter notebook on a typical laptop in a
-fraction of a second using the Python package tskit.
-
-Inferring the wide ARG required 17 days of computation time on a machine with
-128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). Inferring the long ARG
+(Wide ARG: 87\%, Long ARG: 84\%), but the inferred ancestral sequences
+represented by the nodes inferred from daily sample clusters
+by UPGMA (Section~\ref{sec:sample-cluster-tree-inference})
+and parsimony heuristics (Section~\ref{sec:parsimony-heuristics})
+contribute significantly.
+Table~\ref{tab:args} shows the counts of edges (see
+Section~\ref{sec:args} for a technical definition) in the ARGs
+categorised by parent node type, and demonstrates that each
+of these contributes significantly to the overall topology.
+In particular, we can see that approximately 40\% of the edges
+in both ARGs involve inheriting directly from sampled genomes.
+Both ARGs contain 29,422 sites (\ref{sec:data_preprocessing}),
+and a large number of mutations.
+The average number of mutations per-site is high, although much
+of this may be explained by outlier sites with artefactually high
+mutation counts (see Figure~\ref{fig:breakpoint-distribution}).
+Despite this, however, the number of mutations per node
+is small. In the Wide ARG, for example, we have a mean of 0.77
+mutations per sample, demonstrating that new samples are generally
+well represented by the it. See Section~\ref{sec:mutation_spectrum}
+for more analyis on the patterns of mutations inferred.
+Recombination plays a relatively minor role, with $<0.3\%$
+of nodes in the ARGs representing inferred recombination events.
+We analyse these events in Sections~\ref{sec:jackson_recombs},
+\ref{sec:breakpoint_intervals}, \ref{sec:parent_divergence},
+and \ref{sec:pango_x_lineages}.
+
+Table~\ref{tab:args} also summarises some of the computational
+properties of the inferred ARGs.
+The ARGs are encoded as a ``succinct tree sequence'' using
+the \texttt{tskit} library, which provides an extensive
+suite of operations for constructing and analysing ARGs
+(Section~\ref{sec:args}). For exampl, the Wide ARG
+which contains complete genomes (with imputed missing data)
+for around 1.2 million samples, along with extensive sample
+and debugging metadata, requires only 58MB of space (compressed
+using the \texttt{tszip} utility). The majority of this
+space is used by metadata, which when discarded results in
+an encoding that requires an average of only 8.29 bytes per
+SARS-CoV-2 genome stored.
+Loading these ARGs takes less than a second, and they can be interactively
+analysed using Jupyter notebooks on a standard laptop. The majority
+of the analyses in this preprint can be carried out in seconds
+with the \texttt{tskit} Python API, using a few gigabytes of RAM.
+
+Inferring these ARGs does require substantial computation.
+The wide ARG required 17 days of computation time on a machine with
+128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). The long ARG
 required 23 days on a (much older) machine with 40 threads and 256 GB RAM (2x
 Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz). The majority of the time is spent
-on running the HMM to find the copying path for each sequence, and the process
+on running the LS HMM (Section~\ref{sec:ls}) to find the copying
+path for each sequence, and the process
 is therefore highly amenable to distributing across multiple machines.
+We therefore anticipate that further development of the
+inference methods and scaling out across multiple servers will
+enable inferences at the full pandemic scale
+(see Discussion).
 
@@
 \subsection{Mutational spectrum}
+\label{sec:mutation_spectrum}
  In addition to recombination events, the ARGs record mutation events along the
@@
 \subsection{Early recombinants reported by Jackson et al. (2021)}
+\label{sec:jackson_recombs}
 % - Early Recombinants from Jackson et al
@@
 \subsection{Distribution of recombination breakpoint intervals}
+\label{sec:breakpoint_intervals}
 
@@
 \subsection{Divergence between recombinant parents}
+\label{sec:parent_divergence}
 % - Divergence between recombinant parents
@@
 \subsection{Recombinant Pango lineages}
+\label{sec:pango_x_lineages}
 
@@
 \subsection{Ancestral Recombination Graphs}
+\label{sec:args}
 The term ``Ancestral Recombination Graph'' was introduced by
@@
 \subsection{Node dating}
+\label{sec:node_dating}
 The approaches to assigning a date to nodes in \texttt{sc2ts} is currently
@@
 \subsection{Data preprocessing}
+\label{sec:data_preprocessing}
 The findings of this study are

### 2023-05-05 paper.tex
Yans updates

--- 2023-05-04 paper.tex
@@
     Nodes & & 1,453,347 & & 783,231  & \\
-    & Samples & 1,265,685 & (87.09\%) & 657,239 & (83.91\%) \\
-    & Local (UPGMA) & 102,709 & (7.07\%) & 51,807 & (6.61\%)  \\
-    & Reversion push & 40,538 & (2.79\%) & 34,358 & (4.39\%)  \\
-    & Mutation collapse & 40,292 & (2.77\%) & 37,749 & (4.82\%)  \\
-    & Recombination & 4,123 & (0.28\%) & 2,078 & (0.27\%) \\
+    & Samples & 1,265,685 & (87\%) & 657,239 & (84\%) \\
+    & Local (UPGMA) & 103K & (7.1\%) & 52K & (6.6\%)  \\
+    & Tree adjustment & 80,830 & (5.6\%) & 72,107 & (9.2\%) \\
+    & Recombination & 4,123 & (0.33\%) & 2,078 & (0.27\%) \\
     \cmidrule{2-6}

### 2023-05-05 paper.tex
Jerome's pass through Backbone phylo section

--- 2023-05-05 paper.tex
@@
 \subsection{Backbone phylogeny}
+\label{sec:backbone_phylogeny}
+For validation purposes, we compare the genetic
+relationships in our ARGs with phylogenetic relationships
+inferred by the NextStrain project \citep{Hadfield2018-ef}. 
+The NextStrain
+trees consist of a relatively small set of SARS-CoV-2 genomes, which over the time period
+of our ARGs are almost entirely restricted to non-recombinant lineages. 
+To enable this comparison, we ``simplified''~\citep{Kelleher2018-xc} 
+the ARGs to  retain only samples that are in a 
+specific NextStrain tree. These simplified ``backbone'' ARGs exhibit very
+little recombination (6 and 5 trees across the genome for the Wide and Long
+ARGs, respectively). 
+% JK: This leaves a big question hanging - why do we ahave recombination
+% in here at all? It's probably just artefactual, right?
 
@@
 \caption{\label{fig:cophylogeny}
-Tanglegram comparing the first local tree in the wide ARG (sampled to mid-2021)
+Tanglegram comparing the leftmost local tree in the Wide ARG 
+(sampled to mid-2021)
 and an ``all-time'' global NextStrain tree (sampled to January 21, 2023). The
-trees were pruned down to the samples present in both the trees. Light grey
+trees are pruned down to the samples present in both the trees. Light grey
 lines match the corresponding samples between the two trees; black circles
@@
 samples; no Pango recombinants are shown, as none are shared between the trees.
-The tanglegram was generated with the aid of the Neighbor-Net algorithm
+The tanglegram was generated using the Neighbor-Net algorithm
 \citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5
@@
 
-For validation purposes, we compare the genetic
-relationships in the sc2ts-inferred ARGs with phylogenetic relationships
-inferred by the NextStrain project \citep{Hadfield2018-ef}. The NextStrain
-trees consist of subsamples of SARS-CoV-2 genomes, which over the time period
-of our ARGs are almost entirely restricted to non-recombinant lineages. To
-enable this comparison, the ARGs were ``simplified'' \citep{Kelleher2018-xc} to
-retain only those samples of our genealogy that match those present in a
-specific NextStrain tree. These simplified ``backbone'' ARGs exhibit very
-little recombination (6 and 5 trees across the genome for the wide and long
-ARGs, respectively).
-
 Figure~\ref{fig:cophylogeny} shows a co-phylogeny of the 88 samples shared
-between the wide ARG and a GISAID global "all-time" tree from NextStrain
-(downloaded on January 21, 2023). To reduce the ARG to a single tree, we show
-the very first tree in the wide ARG; however, the five other trees in this
+between the wide ARG and a GISAID global ``all-time'' tree from NextStrain
+(downloaded on January 21, 2023). 
+To reduce the ARG to a single tree, we show
+the left-most tree (covering the genome interval 0--64) in the Wide ARG; 
+however, the five other trees in this
 simplified ARG show near-identical patterns. Supplementary
-Figure~\ref{fig:cophylogeny_long} shows the same comparison for the long ARG.
+Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG.
 It is clear that the backbone topology of the sc2ts tree shows very close
@@
 
-Figure~\ref{fig:cophylogeny} reveals some notable differences between the
-trees. Firstly, the sc2ts tree is generally less well resolved (noticeably,
-February to March, 2020). Secondly, there are non-identical sample partitions
-near the tips (e.g., in the Alpha clade). These might have arisen because the
-samples containing the mutations needed to resolve local topologies were
-collected later than the cut-off date for the sc2ts ARG, but were for the
-NextStrain tree. Finally, the branch lengths differ. For example, the branch
-leading to the ancestral node of the Delta variant samples is much shorter in
-the sc2ts tree than the equivalent NextStrain tree. It is possible that the
-length of some of the branches was poorly calibrated, probably caused by overly
-aggressive insertion of mutation-collapsing nodes and reversion-pushed nodes or
-suboptimal sample attachment. Side-by-side comparison of sc2ts trees with trees
-inferred using another method (e.g., UShER) should help to understand whether
-branch length differences are due to poor calibration or methodological
-differences.
-
-Although the simplified backbone ARGs have few local trees, all of which show
-essentially the same topologies, note that in the original ``un-simplified''
-wide and long ARGs, there are 1,496 and 958 local trees respectively (Table 1).
-Hence, while for a relatively small subset of the genealogy may be well
-described by a single tree, the full dataset shows substantial variation in
-phylogenetic relationships (i.e. ``phylogenetic incongruence'') across the
-SARS-CoV-2 genome, much of which may be attributable to recombination. An ARG
-representation is therefore essential to fully capture these reticulate
-genealogical relationships.
+Figure~\ref{fig:cophylogeny} also reveals some notable differences between the
+trees. Firstly, the \texttt{sc2ts} tree is generally less well resolved,
+particularly in early 2020 when sampling density was much lower than
+later in the pandemic.
+Secondly, there are non-identical sample partitions
+near the tips (e.g., in the Alpha clade).
+Finally, and most noticably, the branch lengths differ. 
+As discussed in Section~\ref{sec:node_dating}, the dating of nodes 
+other than samples in \texttt{sc2ts} is currently quite crude, and 
+it is likely that branches not ``anchored'' by samples will be 
+poorly calibrated. See [SOMEWHERE ELSE] for a discussion of how
+these issues may be alleviated in future versions of \texttt{sc2ts}.
+
+% JK: I don't think this is the place to get into discussing the benefits 
+% of an ARG, and I don't want to get bogged down in explaining what it 
+% means to have 1500 trees along the genome.
+
+% Although the simplified backbone ARGs have few local trees, all of which show
+% essentially the same topologies, note that in the original ``un-simplified''
+% Wide and Long ARGs, there are 1,496 and 958 local trees respectively (Table 1).
+% Hence, while for a relatively small subset of the genealogy may be well
+% described by a single tree, the full dataset shows substantial variation in
+% phylogenetic relationships (i.e. ``phylogenetic incongruence'') across the
+% SARS-CoV-2 genome, much of which may be attributable to recombination. An ARG
+% representation is therefore essential to fully capture these reticulate
+% genealogical relationships.
 

### 2023-05-09 plot.py
Move cophylo plot to middle of genome

@@ -169,7 +169,7 @@ class Figure:
 
 class Cophylogeny(Figure):
     name = None
-    pos = 0  # Position along tree seq to plot trees
+    pos = 14952  # Position along tree seq to plot trees: 14952 is halfway along
     sc2ts_filename = None
     nextstrain_ts_fn = "nextstrain_ncov_gisaid_global_all-time_timetree-2023-01-21.nex"
 

### 2023-05-09 paper.tex

@@
 \begin{figure} \centering
-\includegraphics[width=0.4\textwidth]{figures/mutational_spectra.pdf}
+\newcolumntype{Y}{>{\centering\arraybackslash}X}
+\begin{tabularx}{\textwidth}{Yc}
+
+\begin{tabular}[b]{llrlr}\toprule
+            & \multicolumn{2}{l}{Wide ARG} & \multicolumn{2}{l}{Long ARG} \\
+% \midrule
+    \cmidrule{2-5}
+Total      & \multicolumn{2}{l}{1,213,193} & \multicolumn{2}{l}{1,062,072} \\
+Transitions & 873,487   & (72.00\%) & 783,773  & (73.80\%) \\
+Transversions & 326,053 & (26.88\%) & 270,333  & (25.45\%) \\
+Insertions  & 6191  & (0.51\%) & 2814   & (0.26\%) \\
+Deletions   & 7462  & (0.62\%) & 5152   & (0.49\%) \\
+Recurrent   & 74,719 & (6.16\%) & 50,099 & (4.72\%) \\
+Reversions  & 72,617 & (5.99\%) & 48,226 & (4.54\%) \\
+Private     & 758,903 & (62.55\%) &  767,111 & (72.23\%)\\
+\bottomrule
+\end{tabular}
+&
+\includegraphics[width=0.35\textwidth]{figures/mutational_spectra.pdf}\\
+(A) & (B)\\
+\end{tabularx}
 \caption{\label{fig:mutational_spectra}
-Mutational spectrum of SARS-CoV-2 in
-the wide ARG (sampled to mid-2021) compared to \cite{Yi2021-sc}. Mutations are
+(A) Summary of mutations in the Long and Wide ARGs.
+(B) Mutational spectrum of SARS-CoV-2 in
+the wide ARG compared to \cite{Yi2021-sc}. Mutations are
 categorised by type (i.e., ancestral state $>$ derived state). The percentages
@@
 
-\subsubsection{Multiple origins} The ARG inference has no pre-defined knowledge
+\subsubsection{Multiple origins}
+The ARG inference has no pre-defined knowledge
 of Pango X lineage assignments, and there is therefore no particular
@@
 \subsubsection{Complex origins}
-\begin{figure} \centering
-\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
-\caption{\label{fig:complex_origins_graph}  A Pango lineage subgraph showing
-complex origins of the recombinant lineage XB. All XB samples in the long ARG
-trace to a single recombination node (top centre), but three further recombinations
-also descend from this node. The samples descending from these nested
-recombinations include 9 that are assigned by Nextclade to
-various non-recombinant pre-Alpha lineages (blue).}
-\end{figure}
-
 As well as cases where single recombination events explain an X lineage,
@@
 area of future research.
+
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
+\caption{\label{fig:complex_origins_graph}  A Pango lineage subgraph showing
+complex origins of the recombinant lineage XB. All XB samples in the long ARG
+trace to a single recombination node (top centre), but three further recombinations
+also descend from this node. The samples descending from these nested
+recombinations include 9 that are assigned by Nextclade to
+various non-recombinant pre-Alpha lineages (blue).}
+\end{figure}
 

### 2023-05-10 plot.py
Add some extra logging output
To indicate recombinants

@@ -301,7 +301,8 @@ class Cophylogeny(Figure):
         )
 
         logging.info(
-            f"{self.sc2ts.ts.num_trees} trees in the simplified 'backbone' ARG"
+            f"{self.sc2ts.ts.num_trees} trees in the simplified 'backbone' ARG. Using the one " +
+            f"between pos {self.sc2ts.tree.interval.left} and {self.sc2ts.tree.interval.right}."
         )
 
     def plot(self, args):
@@ -1495,8 +1496,17 @@ def print_sample_map(ts, nodes):
                 md["gisaid_epi_isl"] if "gisaid_epi_isl" in md else "EPI_ISL_unknown",
                 md["strain"] if "strain" in md else "Strain_unknown",
             )
+    # Find the nodes that descend from a recombinant node
+    ts, node_map = ts.simplify(nodes, map_nodes=True)
+    new_nodes = node_map[nodes]
+    re_nodes = np.flatnonzero(ts.nodes_flags & sc2ts.NODE_IS_RECOMBINANT)
+    recombinant_samples = set()
+    for tree in ts.trees():
+        for u in re_nodes:
+            recombinant_samples.update(tree.samples(u))
     for k in sorted(isl_map.keys()):
-        print(f"tsk{k}: {isl_map[k][0]}, {isl_map[k][1]}")
+        prefix = "*tsk" if node_map[k] in recombinant_samples else "tsk"
+        print(f"{prefix}{k}: {isl_map[k][0]}, {isl_map[k][1]}")
 
 
 class Pango_XA_gisaid_large_graph(Pango_XA_nxcld_tight_graph):

### 2023-05-10 paper.tex
Updates from Overleaf

@@
 than attempting to be comprehensive we chose to work with a
-representative subset of the available data, and limited to
+representative subset of the available data, limited to
 inferences that could be performed on a single server in
@@
 well represented by the it. See Section~\ref{sec:mutation_spectrum}
-for more analyis on the patterns of mutations inferred.
+for more analysis of the patterns of inferred mutations.
 Recombination plays a relatively minor role, with $<0.3\%$
@@
 suite of operations for constructing and analysing ARGs
-(Section~\ref{sec:args}). For exampl, the Wide ARG
+(Section~\ref{sec:args}). For example, the Wide ARG
 which contains complete genomes (with imputed missing data)
@@
 inferred by the NextStrain project \citep{Hadfield2018-ef}.
-The NextStrain
-trees consist of a relatively small set of SARS-CoV-2 genomes, which over the time period
-of our ARGs are almost entirely restricted to non-recombinant lineages.
 To enable this comparison, we ``simplified''~\citep{Kelleher2018-xc}
-the ARGs to  retain only samples that are in a
-specific NextStrain tree. These simplified ``backbone'' ARGs exhibit very
+the sc2ts ARGs to a ``backbone'' containing only samples present
+in a specific NextStrain tree. For this purpose, we picked an
+``all-data''  NextStrain tree restricted to the time period covered
+by the appropriate ARG. This results in a relatively small
+set of shared samples, almost entirely restricted to non-recombinant lineages.
+These simplified backbone ARGs exhibit very
 little recombination (6 and 5 trees across the genome for the Wide and Long
@@
 To reduce the ARG to a single tree, we show
-the left-most tree (covering the genome interval 0--64) in the Wide ARG;
+the tree in the middle of the viral genome (covering the genome interval 8393 to 21618)
+in the Wide ARG;
 however, the five other trees in this
@@
 Griffiths and colleagues~\citep{Griffiths1991-two,Griffiths1998-ancestral}
-and originally defined an alternative formulation of the coalescent
+and originally defined as an alternative formulation of the coalescent
 with recombination stochastic process~\citep{Hudson1983-properties}.
@@
 reticulate genetic ancestry of a sample of sequences under
-the influence of recombination.
+the influence of recombination. This definition encompasses various types of graphs traditionally described using the broader term of phylogenetic networks.
 
@@
 and a \emph{mutation} defines the site, node and derived state
-that generates variation.
+that generates variation (note that this required that the sequences can be aligned and phased).
 Given these tables, we can then efficiently construct the local
@@
 trees~\citep{Kelleher2016-wk,Ralph2020-efficiently}. These
-algorithms have led to peformance increases of several orders
+algorithms have led to performance increases of several orders
 of magnitude over the state-of-the-art in a range of
@@
 Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}.
+We emphasise that this ecosystem for storing and manipulating ARGs can generally be used to efficiently record and analyse SARS-CoV-2 genealogies reconstructed using other methods, not only the new approach that we describe here.
 

### 2023-05-10 paper.tex
Editing imputation section and figure

@@
 \includegraphics[width=.7\textwidth]{figures/imputation.pdf}
-\caption{\label{fig:imputation}A schematic of the two-step procedure to impute
+\caption{\label{fig:imputation}A schematic of the iterative procedure to impute
 Pango lineage for inserted, non-sample nodes. Here, three nodes (question
-marks) have unknown Pango lineage (A). Because the parent of node 1 is
-available (i.e., L2) and there is no mutation occurring on the edge connecting
-the two nodes, the Pango lineage of the parent is directly propagated to node 1
-(B). However, if there is a mutation (red X’s), then the Pango lineage is
-inferred by comparison to the set of lineage-defining mutations from
-COVID-CG\citep{Chen2021-zc}, as in case (C). Finally, the inferred status is
-propagated to any non-mutant child nodes, as in case (D).
-}
+marks) have unknown Pango lineage (A). The lineage for node N1 can be directly 
+copied from its parent (L2). The lineage for node N2 must be inferred from 
+that of the parent (L2) plus the lineage-defining mutation (red X) 
+on the connecting edge. The lineage for node N3 can then be copied from 
+that of node N2.}
 \end{figure}
-For many applications it is helpful to infer the Pango lineage status
-of the internal ARG nodes estimated by \texttt{sc2ts}.
-% Initially, only sample nodes have Pango lineage assignments.
-% For analysis of recombinant parents, it is also helpful to assign
-% Pango lineages to non-sample nodes.
-We do this using a two-step procedure (Figure~\ref{fig:imputation}).
-Initially, only sample nodes have Pango lineage assignments.
-For a given non-sample node $u$, if the Pango lineage of its parent is
-available and there are no intervening mutations,
-then we impute that $u$'s Pango lineage is the same as its parent.
-Otherwise, the Pango lineage of the
-non-sample node is inferred by matching its full set of mutations against the
-Pango lineage-defining mutations (based on 90\% consensus of the sequences
-analysed) from the COVID-CG website \citep{Chen2021-zc};
-\url{https://covidcg.org/}; accessed on November 04, 2022).
-
-[TODO how accurate is this approach? We should probably discuss it somewhere
-in the Results, and refer back from here? ]
+While the Pango lineages of sample nodes are imported directly from 
+GISAID metadata, the lineage status of internal nodes inserted into 
+the \texttt{sc2ts} ARG must be imputed. We do this using 
+the list of lineage-defining mutations (based on 90\% consensus 
+of the sequences analysed) from the COVID-CG website \citep[][\url{https://covidcg.org/}; accessed on 4 November 2022]{Chen2021-zc}.
+
+For a given non-sample node $u$, if the Pango lineage of its parent or 
+one of its children is already known, and there are no lineage-defining mutations
+on the connecting edge, then $u$ copies this Pango lineage exactly.
+Otherwise, the lineage for $u$ is inferred by 
+matching its full set of mutations against the COVID-CG list. 
+We apply these two steps to the internal nodes of the ARG iteratively,
+as illustrated in Figure~\ref{fig:imputation}, until all internal nodes 
+are assigned a lineage (where possible---note that sometimes a lineage cannot
+be assigned to the children of a recombination node).
+
+This method is fast, as the lineages for most of the internal nodes can be 
+imputed by copying from the surrounding nodes (66\% of nodes for the Long ARG, 
+80\% of nodes for the Wide ARG). The accuracy depends on
+the quality of the list of lineage-defining mutations; we find that when
+masking the lineages of sample nodes one-by-one and re-imputing these using
+the remaining information results in 98\% (resp.~99\%) of lineages being
+imputed correctly.
 

### 2023-05-11 paper.tex
Add table for XE and move to appendix

@@
 
-% \begin{frame}
-% \begin{block}{Multiple origin Pango X lineages}
-% \begin{itemize}
-% \item For the remaining 18, the story is more complicated.
-% \item XJ, XU and XV each have 3 samples and 3 recombinants
-% \item One recombinant shared among 5 lineages (XAA, XAB, XAG, XQ, XU)
-% \item XE has 170 samples and 12 recombinants; some are very similar
-% (Official: BA.1* \& BA.2, 10448--11287):
-% \\
-% \vspace{1em}
-% \begin{tabular}{llll}
-% \toprule
-% Left parent & Right parent & \multicolumn{2}{l}{Breakpoint interval}\\
-% \midrule
-% BA.1.15 & BA.2  &9867  &10198\\
-% BA.1.15 & BA.2  &9867  &10198\\
-% BA.1.1  &BA.2  &10451  &11537\\
-% BA.1.1.5 & BA.2 &10448 & 11537\\
-% BA.1.15 & BA.2  &10448  &12880\\
-% BA.1.17 & BA.2  &10448 & 11537\\
-% BA.1  &BA.2.9  &10448 & 11537\\
-% BA.1  &BA.2  &12764  &13195\\
-% \bottomrule
-% \end{tabular}
-% % BA.3  &BA.1.17  &22674  &22674\\
-% % BA.1.1.15 Recombinant 10448 11537
-% % BA.1.1 Recombinant 10030 10447
-% % Recombinant BA.2 17411 19955
-% \end{itemize}
-% \end{block}
-% \end{frame}
-
-\begin{figure}
-\centering
-\begin{tabular}{llll}
-\toprule
-Left parent & Right parent & \multicolumn{2}{l}{Breakpoint interval}\\
-\midrule
-BA.1.15  & BA.2   & 9,867  & 10,198 \\
-BA.1.15  & BA.2   & 9,867  & 10,198 \\
-BA.1.1   & BA.2   & 10,451 & 11,537 \\
-BA.1.1.5 & BA.2   & 10,448 & 11,537 \\
-BA.1.15  & BA.2   & 10,448 & 12,880 \\
-BA.1.17  & BA.2   & 10,448 & 11,537 \\
-BA.1     & BA.2.9 & 10,448 & 11,537 \\
-BA.1     & BA.2   & 12,764 & 13,195 \\
-\bottomrule
-\end{tabular}
-% BA.3  &BA.1.17  &22674  &22674\\
-% BA.1.1.15 Recombinant 10448 11537
-% BA.1.1 Recombinant 10030 10447
-% Recombinant BA.2 17411 19955
-\caption{\label{tab:multiple_origins_table}
-Details of inferred
-breakpoints for different recombination events inferred for XE samples (170
-samples and 12 inferred recombination nodes; some are very similar (Official:
-BA.1* \& BA.2, 10448–11287)}
-\end{figure}
-
 Figure~\ref{fig:pango-simple-origin-graph}C shows a simple example, consisting of
@@
 
-As the results of \texttt{sc2ts} are sensitive to the existance of
-time-travellers, an imporant aspect of future work is to find better ways
+As the results of \texttt{sc2ts} are sensitive to the existence of
+time-travellers, an important aspect of future work is to find better ways
 to identify them. One possibility is to use the LS HMM itself to flag
@@
 \subsection{Imputation of Pango lineage for non-sample nodes}
-% FIXME the caption and the text are basically saying the same thing here. Can
-% we streamline, and maybe give a bit more detail in the text about iteration,
-% and the concrete implementation?
+
 \begin{figure} \centering
@@
 This method is fast, as the lineages for most of the internal nodes can be
-imputed by copying from the surrounding nodes (66\% of nodes for the Long ARG, 
-80\% of nodes for the Wide ARG). The accuracy depends on
-the quality of the list of lineage-defining mutations; we find that when
-masking the lineages of sample nodes one-by-one and re-imputing these using
-the remaining information results in 98\% (resp.~99\%) of lineages being
-imputed correctly.
+imputed by copying from the surrounding nodes (66\% of nodes for the long ARG,
+80\% of nodes for the wide ARG), and significantly more efficient than extracting
+the haplotypes for each internal node and using existing Pangolin assignment tools
+\citep{OToole2021-assignment}. The
+accuracy depends on the quality of the list of lineage-defining mutations,
+as well as the source of lineage designation for the sample nodes: we obtain
+slightly different results when using those recorded on GISAID (which uses
+pangoLEARN), and those assigned by Nextclade. To gauge the accuracy of
+imputation, we have used our method to re-impute the lineage designations of each
+sample node using the surrounding information; this results in 98\% (resp.~99\%) of
+nodes being assigned the same lineage as per the source metadata in the long
+(resp.~wide) ARG.
 
@@
 \caption{\label{tab:pango-single-origin}
-Summary of the Pango X-lineages in the long ARG that are inferred to have a single recombinant origin based on the NextClade Pango lineage assignments.
-Parent lineages from the ARG (as designated by Nextclade and, where relevant and also to parent lineages as defined by the official Pango designation alias key.
-(TODO: explain the asterisks) FIXME: if we like this table then we can reformat it so that lists at the end are spread over multiple lines rather than using Python dict notation }
+Summary of the Pango X-lineages in the long ARG that are inferred to have a
+single recombinant origin based on the NextClade Pango lineage assignments.
+Parent lineages from the ARG (as designated by Nextclade and, where relevant
+and also to parent lineages as defined by the official Pango designation alias
+key. (TODO: explain the asterisks) FIXME: if we like this table then we can
+reformat it so that lists at the end are spread over multiple lines rather than
+using Python dict notation }
+\end{table}
+
+\begin{table}
+\begin{tabular}{llll}
+\toprule
+ interval &                 causal & parent &                                descendant\_lineages \\
+\midrule
+ (9867, 10198) &             XE  & [BA.1, BA.2] &                                          \{'XE': 3\} \\
+(10448, 11537) &             XH  & [BA.1.20, BA.2.9] &  \{'XAF': 34, 'XH': 11, 'B.1.1.529': 3, 'XE': 3,... \\
+(10448, 11537) &             XE  & [BA.1.1.15, Unknown (R)] & \{'XE': 2\} \\
+(10448, 11537) &             XE  & [BA.1.17.2, BA.2] &                                        \{'XE': 155\} \\
+\bottomrule
+\end{tabular}
+\caption{\label{tab:multiple_origins_table}
+Details of inferred
+breakpoints for different recombination events inferred for XE samples
+(Official: BA.1* \& BA.2, 10448–11287). TODO ELABORATE}
 \end{table}

### 2023-05-11 paper.tex
Ana updates

--- 2023-05-10 paper.tex
@@
 the list of lineage-defining mutations (based on 90\% consensus 
-of the sequences analysed) from the COVID-CG website \citep[][\url{https://covidcg.org/}; accessed on 4 November 2022]{Chen2021-zc}.
+of the sequences analysed) from the COVID-CG website 
+\citep[][\url{https://covidcg.org/}; accessed on 4 November 2022]{Chen2021-zc}.
 
@@
 This method is fast, as the lineages for most of the internal nodes can be 
-imputed by copying from the surrounding nodes (66\% of nodes for the Long ARG, 
-80\% of nodes for the Wide ARG). The accuracy depends on
-the quality of the list of lineage-defining mutations; we find that when
-masking the lineages of sample nodes one-by-one and re-imputing these using
-the remaining information results in 98\% (resp.~99\%) of lineages being
-imputed correctly.
+imputed by copying from the surrounding nodes (66\% of nodes for the long ARG, 
+80\% of nodes for the wide ARG), and significantly more efficient than extracting
+the haplotypes for each internal node and using existing Pangolin assignment tools 
+\citep{OToole2021-assignment}. The 
+accuracy depends on the quality of the list of lineage-defining mutations, 
+as well as the source of lineage designation for the sample nodes: we obtain 
+slightly different results when using those recorded on GISAID (which uses 
+pangoLEARN), and those assigned by Nextclade. To gauge the accuracy of
+imputation, we have used our method to re-impute the lineage designations of each 
+sample node using the surrounding information; this results in 98\% (resp.~99\%) of 
+nodes being assigned the same lineage as per the source metadata in the long 
+(resp.~wide) ARG.
 

### 2023-05-11 plot.py
Updates to intervals figure

@@ -4,7 +4,6 @@ from datetime import datetime, timedelta
 import hashlib
 import itertools
 import re
-import sys
 import json
 import os
 import tempfile
@@ -15,17 +14,16 @@ import shutil
 
 import matplotlib as mpl
 
-mpl.use("Agg")
+mpl.use("Agg")  # NOQA
 from matplotlib.colors import rgb2hex
-from matplotlib.cm import get_cmap
+from matplotlib.cm import get_cmap, ScalarMappable
 from matplotlib.lines import Line2D
 from matplotlib.patches import Patch
 from matplotlib import collections as mc
 from matplotlib import pyplot as plt
+from mpl_toolkits.axes_grid1.inset_locator import inset_axes
 import numpy as np
 import tqdm
-import tszip
-import tskit
 import pandas as pd
 
 import tsconvert  # Not on pip. Install with python -m pip install git+http://github.com/tskit-dev/tsconvert
@@ -301,8 +299,8 @@ class Cophylogeny(Figure):
         )
 
         logging.info(
-            f"{self.sc2ts.ts.num_trees} trees in the simplified 'backbone' ARG. Using the one " +
-            f"between pos {self.sc2ts.tree.interval.left} and {self.sc2ts.tree.interval.right}."
+            f"{self.sc2ts.ts.num_trees} trees in the simplified 'backbone' ARG. Using the one "
+            + f"between pos {self.sc2ts.tree.interval.left} and {self.sc2ts.tree.interval.right}."
         )
 
     def plot(self, args):
@@ -882,14 +880,13 @@ class Pango_X_graph(Figure):
     node_colours = {
         # B lineages in blue/green (AY.100 is B.1.617.2.100)
         # Basic in blue
-        'B.1': '#212fe1',
-        'B.1.384': '#4ac7ff',
-        'B.1.177.18': '#93e8ff',
-        'B.1.631': '#add9e5',
-        'B.1.634': '#2092e1',
-        'B.1.627': '#7b96cc',
-        'B.1.1': '#aea4ff',
-
+        "B.1": "#212fe1",
+        "B.1.384": "#4ac7ff",
+        "B.1.177.18": "#93e8ff",
+        "B.1.631": "#add9e5",
+        "B.1.634": "#2092e1",
+        "B.1.627": "#7b96cc",
+        "B.1.1": "#aea4ff",
         # Alpha: purple / lilac
         "B.1.1.7": "#ae9cff",
         # Delta: yellow/brown
[…]

### 2023-05-11 plot.py
Inset length distribution

@@ -1797,6 +1797,14 @@ class RecombinationIntervals(Figure):
             ScalarMappable(norm=norm, cmap=cmap), cax=axins1, orientation="vertical"
         )
 
+        axins2 = inset_axes(
+            ax1,
+            width="20%",
+            height="50%",
+            loc="upper center",
+        )
+        axins2.hist(length)
+
         covers = np.zeros(df_sites.position.max())
         for left, right in intervals:
             covers[left:right] += 1

### 2023-05-12 paper.tex
Yan's updates

--- 2023-05-11 paper.tex
@@
 \label{sec:backbone_phylogeny}
+%% The ARG is mostly tree-like
+Compared to organisms like humans which recombine every generation,
+SARS-CoV2 shows little recombination, with recombination nodes accounting
+for <0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
+relationships between strains can often be adequately represented by a
+single tree, particularly early in the pandemic, or when looking at
+a subset of sampled genomes e.g. restricted by time or limited to close
+relatives.
+
 For validation purposes, we compare the genetic

### 2023-05-12 plot.py
Allow png output of figures

@@ -1877,7 +1877,7 @@ def main():
     parser.add_argument(
         "-o",
         "--outtype",
-        choices=["pdf", "svg"],
+        choices=["pdf", "png", "svg"],
         default="pdf",
         help="The format of the output file",
     )

### 2023-05-12 paper.tex
Summarise mutations

--- 2023-05-11 paper.tex
@@
 \label{sec:backbone_phylogeny}
-For validation purposes, we compare the genetic
-relationships in our ARGs with phylogenetic relationships
-inferred by the NextStrain project \citep{Hadfield2018-ef}.
-To enable this comparison, we ``simplified''~\citep{Kelleher2018-xc}
+%% The ARG is mostly tree-like
+Compared to organisms like humans which recombine every generation,
+in SARS-CoV2 recombination is rare, with recombination nodes accounting
+for <0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
+relationships between strains can often be represented by a
+single phylogenetic tree, particularly when looking at
+a subset of sampled genomes e.g. restricted by time or limited to close
+relatives. We expect the ARG to be particularly tree-like
+early in the pandemic, when co-infection was less likely.
+
+A classic tree-based summary of SARS-CoV2 ancestry is provided
+by the NextStrain project \citep{Hadfield2018-ef}. The trees
+available from Nextstrain are based on small subsamples of the
+dataset, and early in the pandemic tend to be restricted to
+a sample of strains which are not thought to be recombinants.
+For validation purposes, we our compare our ARGs with
+a downloaded ``all-data'' NextStrain tree, restricted to the time period
+covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
 the sc2ts ARGs to a ``backbone'' containing only samples present
-in a specific NextStrain tree. For this purpose, we picked an
-``all-data''  NextStrain tree restricted to the time period covered
-by the appropriate ARG. This results in a relatively small
-set of shared samples, almost entirely restricted to non-recombinant lineages.
-These simplified backbone ARGs exhibit very
-little recombination (6 and 5 trees across the genome for the Wide and Long
-ARGs, respectively).
-% JK: This leaves a big question hanging - why do we ahave recombination
-% in here at all? It's probably just artefactual, right?
+the NextStrain tree. This results in a small set of shared
+samples (180 for the wide ARG, 88 for the long ARG), none of which are
+designated as Pango recombinant lineages by Nextclade (see subsection
+\ref{sec:pango_x_lineages}).
+
+These sc2ts backbone phylogenies do contain small amounts of recombination
 
@@
 \label{sec:mutation_spectrum}
- In addition to recombination events, the ARGs record mutation events along the
-SARS-CoV-2 phylogeny. \cite{Yi2021-sc} reconstructed a SARS-CoV-2 phylogeny of
-over 350,000 genomes sampled globally from December 24, 2019, to January 12,
-2021, and systematically catalogued all the mutations occurring along the
-phylogeny. Notably, they described a mutational spectrum in which C-to-U and
-G-to-U mutations occur more frequently than U-to-C and U-to-G mutations,
-respectively.
+The ARGs inferred by \texttt{sc2ts} and represented using the \texttt{tskit}
+library (Section~\ref{sec:args}) are a joint estimate of the processes
+of inheritance, recombination and mutation. Unlike most approaches to
+phylogenetic analysis, mutations are included in the \texttt{tskit}
+data model alongside
+the topological representation of genetic inheritance.
+This has many
+advantages, for example allowing us to compute statistics of the observed
+sequences efficiently~\citep{Kelleher2016-wk,Ralph2020-efficiently} and
+to provide high-levels of data compression~\citep{Kelleher2019-ba}.
+The same idea has recently been used to represent
+SARS-CoV-2 data in UShER's ``mutation annotated tree'' format~\citep{Turakhia2022-it}.
 
@@
 Total      & \multicolumn{2}{l}{1,213,193} & \multicolumn{2}{l}{1,062,072} \\
+Private     & 758,903 & (62.55\%) &  767,111 & (72.23\%)\\
+    \cmidrule{2-5}
 Transitions & 873,487   & (72.00\%) & 783,773  & (73.80\%) \\
 Transversions & 326,053 & (26.88\%) & 270,333  & (25.45\%) \\
+    \cmidrule{2-5}
 Insertions  & 6191  & (0.51\%) & 2814   & (0.26\%) \\
 Deletions   & 7462  & (0.62\%) & 5152   & (0.49\%) \\
+    \cmidrule{2-5}
 Recurrent   & 74,719 & (6.16\%) & 50,099 & (4.72\%) \\
 Reversions  & 72,617 & (5.99\%) & 48,226 & (4.54\%) \\
-Private     & 758,903 & (62.55\%) &  767,111 & (72.23\%)\\
 \bottomrule
@@
 \caption{\label{fig:mutational_spectra}
-(A) Summary of mutations in the Long and Wide ARGs.
-(B) Mutational spectrum of SARS-CoV-2 in
-the wide ARG compared to \cite{Yi2021-sc}. Mutations are
-categorised by type (i.e., ancestral state $>$ derived state). The percentages
+(A) Summary of mutations in the Long and Wide ARGs. Private mutations
+occur on terminal branches. Insertions are mutations in which the
+inherited state is the gap state ``-'' and the derived state is a
+nucleotide, and vice versa for deletions. Recurrent mutations are
+those that inherit from at least one other mutation (at that site),
+and a reversion is a recurrent mutation that reverses the state change
+of the mutation immediately ancestral to it.
+(B) Mutational spectrum in
+the Wide ARG compared to \cite{Yi2021-sc}. Mutations are
+categorised by type (i.e., inherited state $>$ derived state). The percentages
 of each mutation type from the wide ARG are represented by blue bars and the
@@
 
-For validation purposes, we compare the mutational spectrum from the wide ARG
-with the mutational spectrum of \citet{Yi2021-sc}. We categorised all of the
-single nucleotide mutations by type (defined by the ancestral state and derived
-state), excluded mutations inherited by only a single sample (which are
-enriched for sequencing errors and artefacts), and tallied them up by type.
+The properties of the mutations inferred in the Wide and Long
+ARGs are summarised in Figure~\ref{fig:mutational_spectra}A. In both
+cases we have a large number of mutations, and a majority of these
+(Wide ARG: 62.55\%, Long ARG: 72.23) are private to a single
+sample, i.e.\ on terminal branches.
+Although the average number
+of mutations per sample is small
+(Wide ARG: 0.77, Long ARG: 1.38; c.f.\ Table~\ref{tab:args}), the average number per site
+is large (Wide ARG: 41.23, Long ARG: 36.10; c.f.\ Table~\ref{tab:args}). This is
+partly driven by some highly mutagenic sites (e.g., site 28271 has over 7,000
+mutations in the Wide ARG, c.f.\ Figure~\ref{fig:breakpoint-distribution}), which
+may be artefactual.
+The main weakness of the current version of \texttt{sc2ts} is an excess
+of inferred reversion mutations, with 5.99\% of all mutations in the Wide
+ARG (Long ARG: 4.54\%) reverting the state change of the immediately
+ancestral mutation. The majority of these mutations are likely to be
+artefactual, and a key goal of future development is to improve
+our parsimony heuristics (Section~\ref{sec:parsimony-heuristics}) to
+address this issue.
+
+Despite the presence of some artefactual mutations,
+the properties of the mutations inferred by \texttt{sc2ts} largely follow
+established results.
+In Figure~\ref{fig:mutational_spectra}B we compare the mutational spectrum
+in the Wide ARG to the results of \cite{Yi2021-sc},
+who reconstructed a SARS-CoV-2 phylogeny of
+over 350,000 genomes sampled globally from December 24, 2019, to January 12,
+2021, and classified the mutations occurring along the phylogeny.
+We categorised all of single nucleotide
+mutations in the Wide ARG by type (defined by the inherited and derived states),
+excluded mutations inherited by only a single sample (which are
+enriched for sequencing errors and artefacts).
 Similarly, we took the data for single nucleotide mutations from
-\url{https://github.com/ju-lab/SC2_evol_signature} \citep{Yi2021-sc}, excluded
+\citep[][\url{https://github.com/ju-lab/SC2_evol_signature}]{Yi2021-sc}, excluded
 mutations occurring along terminal branches, and tallied them up by type.
-Figure~\ref{fig:mutational_spectra} shows that the mutational spectrum from the
+Figure~\ref{fig:mutational_spectra}B shows that the mutational spectrum from the
 wide ARG (based on 448,825 mutations) matches that reported by \citet[based on
@@
 sample or those occurring on terminal branches (data not shown).
-
-% If we do this, goes here
-% \subsection{Mutations under selection}
-% XX
 

### 2023-05-12 paper.tex
Edites by yan

--- 2023-05-12 paper.tex
@@
 Compared to organisms like humans which recombine every generation,
-SARS-CoV2 shows little recombination, with recombination nodes accounting
+in SARS-CoV2 recombination is rare, with recombination nodes accounting
 for <0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
-relationships between strains can often be adequately represented by a
-single tree, particularly early in the pandemic, or when looking at
+relationships between strains can often be represented by a
+single phylogenetic tree, particularly when looking at
 a subset of sampled genomes e.g. restricted by time or limited to close
-relatives.
-
-For validation purposes, we compare the genetic
-relationships in our ARGs with phylogenetic relationships
-inferred by the NextStrain project \citep{Hadfield2018-ef}.
-To enable this comparison, we ``simplified''~\citep{Kelleher2018-xc}
+relatives. We expect the ARG to be particularly tree-like
+early in the pandemic, when co-infection was less likely.
+
+A classic tree-based summary of SARS-CoV2 ancestry is provided
+by the NextStrain project \citep{Hadfield2018-ef}. The trees
+available from Nextstrain are based on small subsamples of the
+dataset, and early in the pandemic tend to be restricted to
+a sample of strains which are not thought to be recombinants.
+For validation purposes, we our compare our ARGs with
+a downloaded ``all-data'' NextStrain tree, restricted to the time period
+covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
 the sc2ts ARGs to a ``backbone'' containing only samples present
-in a specific NextStrain tree. For this purpose, we picked an
-``all-data''  NextStrain tree restricted to the time period covered
-by the appropriate ARG. This results in a relatively small
-set of shared samples, almost entirely restricted to non-recombinant lineages.
-These simplified backbone ARGs exhibit very
-little recombination (6 and 5 trees across the genome for the Wide and Long
-ARGs, respectively).
-% JK: This leaves a big question hanging - why do we ahave recombination
-% in here at all? It's probably just artefactual, right?
+the NextStrain tree. This results in a small set of shared
+samples (180 for the wide ARG, 88 for the long ARG), none of which are
+designated as Pango recombinant lineages by Nextclade (see subsection
+\ref{sec:pango_x_lineages}).
+
+These sc2ts backbone phylogenies do contain small amounts of recombination
 

### 2023-05-12 paper.tex
Change title
Updates

--- 2023-05-12 paper.tex
@@
 
-\title{Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2}
+\title{Toward Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2}
 \author[1]{Shing H. Zhan}
@@
 early in the pandemic, when co-infection was less likely.
-
-A classic tree-based summary of SARS-CoV2 ancestry is provided
-by the NextStrain project \citep{Hadfield2018-ef}. The trees
-available from Nextstrain are based on small subsamples of the
-dataset, and early in the pandemic tend to be restricted to
-a sample of strains which are not thought to be recombinants.
-For validation purposes, we our compare our ARGs with
-a downloaded ``all-data'' NextStrain tree, restricted to the time period
-covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
-the sc2ts ARGs to a ``backbone'' containing only samples present
-the NextStrain tree. This results in a small set of shared
-samples (180 for the wide ARG, 88 for the long ARG), none of which are
-designated as Pango recombinant lineages by Nextclade (see subsection
-\ref{sec:pango_x_lineages}).
-
-These sc2ts backbone phylogenies do contain small amounts of recombination
-
 \begin{figure} \centering
@@
 \end{figure}
+
+A classic tree-based summary of SARS-CoV2 ancestry is provided
+by the NextStrain project \citep{Hadfield2018-ef}. The trees
+available from Nextstrain are based on small subsamples of the
+dataset, and early in the pandemic tend to be restricted to
+a sample of strains which are not thought to be recombinants.
+For validation purposes, we our compare our ARGs with
+a downloaded ``all-data'' NextStrain tree, restricted to the time period
+covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
+the sc2ts ARGs to a ``backbone'' containing only samples present
+the NextStrain tree. This results in a small set of shared
+samples (180 for the wide ARG, 88 for the long ARG), none of which are
+designated as Pango recombinant lineages by Nextclade (see subsection
+\ref{sec:pango_x_lineages}).
+
+Our sc2ts backbone phylogenies do contain small amounts of recombination:
+7 recombination nodes in the backbone of the wide ARG and 8 in the backbone
+of the long ARG. However, the backbones remain extremely tree-like, as these
+recombination events only involve minor and local topological rearrangements:
+for instance recombination between two sibling lineages, or even recombination
+within a single lineage (in which the parent lineages above a recombination event
+coalesce with each other before coalescing with any other lineages in
+the phylogeny
 

### 2023-05-12 paper.tex
Tweaks to backbone phylo section

--- 2023-05-12 paper.tex
@@
 in SARS-CoV2 recombination is rare, with recombination nodes accounting
-for <0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
+for $<$0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
 relationships between strains can often be represented by a
@@
 
-Figure~\ref{fig:cophylogeny} shows a co-phylogeny of the 88 samples shared
-between the wide ARG and a GISAID global ``all-time'' tree from NextStrain
-(downloaded on January 21, 2023).
-To reduce the ARG to a single tree, we show
-the tree in the middle of the viral genome (covering the genome interval 8393 to 21618)
-in the Wide ARG;
-however, the five other trees in this
-simplified ARG show near-identical patterns. Supplementary
-Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG.
-It is clear that the backbone topology of the sc2ts tree shows very close
+A classic tree-based summary of SARS-CoV2 ancestry is provided
+by the NextStrain project \citep{Hadfield2018-ef}. The trees
+available from Nextstrain are based on small subsamples of the
+dataset, and early in the pandemic tend to be restricted to
+a sample of strains which are not thought to be recombinants.
+For validation purposes, we our compare our ARGs with
+a downloaded NextStrain tree, restricted to the time period
+covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
+the \texttt{sc2ts} ARGs to a ``backbone'' containing only samples present
+the NextStrain tree. This results in a small set of shared
+samples (180 for the Wide ARG, 88 for the Long ARG), none of which are
+assigned to Pango recombinant lineages by Nextclade (see Section
+\ref{sec:pango_x_lineages}).
+
+Our \texttt{sc2ts} backbone phylogenies do contain small amounts of recombination:
+there are 7 recombination nodes in the backbone of the Wide ARG and 8 in the backbone
+of the Long ARG. However, the backbones remain extremely tree-like, as these
+recombination events involve minor, local topological rearrangements.
+All involve recombination between closely related lineages (e.g.\ sibling lineages)
+and a number even represent recombination within a single lineage, in which the parent
+lineages above a recombination event coalesce with each other before coalescing
+with any other lineages in the phylogeny. These could either represent true
+recombination events between closely related strains, which have previously gone unnoticed,
+or could be artefacts of the current \texttt{sc2ts} method and its parameterization.
+
+Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
+a GISAID global ``all-time'' tree from NextStrain (downloaded on January 21, 2023).
+We illustrate the backbone phylogeny by visualising
+a single tree in the middle of the viral genome
+(covering the interval 8393--21618), since other regions of the 
+genome show almost identical topologies.
+It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
 agreement with the NextStrain tree. The sample genomes cluster by their
@@
 clades, labelled).
+Supplementary Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.
 
@@
 near the tips (e.g., in the Alpha clade).
-Finally, and most noticably, the branch lengths differ.
+Finally, and most noticeably, the branch lengths differ.
 As discussed in Section~\ref{sec:node_dating}, the dating of nodes

### 2023-05-12 paper.tex
Yans updates

@@
 early in the pandemic, when co-infection was less likely.
+
 \begin{figure} \centering
@@
 \caption{\label{fig:cophylogeny}
-Tanglegram comparing the leftmost local tree in the Wide ARG
-(sampled to mid-2021)
-and an ``all-time'' global NextStrain tree (sampled to January 21, 2023). The
-trees are pruned down to the samples present in both the trees. Light grey
+Tanglegram comparing a local tree from the Wide ARG
+(sampled to mid-2021) and an ``all-time'' global NextStrain tree
+(downloaded on January 21, 2023).
+Phylogenies are pruned down to those samples present as tips in both datasets. Light grey
 lines match the corresponding samples between the two trees; black circles
@@
 of the Long ARG. However, the backbones remain extremely tree-like, as these
-recombination events involve minor, local topological rearrangements.
-All involve recombination between closely related lineages (e.g.\ sibling lineages)
+recombination events involve minor, local topological rearrangements. In particular,
+all involve recombination between closely related lineages (e.g.\ sibling lineages)
 and a number even represent recombination within a single lineage, in which the parent
@@
 recombination events between closely related strains, which have previously gone unnoticed,
-or could be artefacts of the current \texttt{sc2ts} method and its parameterization.
+or could be artifacts of the current \texttt{sc2ts} method and its parameterization.
 
 Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
-a GISAID global ``all-time'' tree from NextStrain (downloaded on January 21, 2023).
-We illustrate the backbone phylogeny by visualising
+a GISAID global ``all-time'' tree from NextStrain
+% We give the date below in the caption - we could probably remove if either from there
+% or from here (I suggest from here)
+(downloaded on January 21, 2023).
+We illustrate the backbone phylogeny by visualizing
 a single tree in the middle of the viral genome
-(covering the interval 8393--21618), since other regions of the 
+%% (covering the interval 8393--21618),  % removed as interval now in the figure title
+since other regions of the 
 genome show almost identical topologies.
@@
 agreement with the NextStrain tree. The sample genomes cluster by their
-assigned Pango lineage status, and many variants and their descendents form
+assigned Pango lineage status, and many variants and their descendants form
 identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
@@
 \caption{\label{fig:recomb_mrcas}
-Date of common ancestry between the parents either side of recombination
-breakpoints, as a function of degree of parental divergence. MRCAs of parents
-inferred to have produced a Pango designated recombinant (XA, XB, etc.) are
-identified in orange (A) Data after removing singleton recombinants (XXX parent
-pairs). Many recombinations involve parental strains that descend from the same
-MRCA: dates for the four most frequently identified MRCA nodes are indicated by
-horizontal dotted lines. These frequent MRCA nodes tend to be associated with
-major waves and commonly have many immediate children (i.e. are large
-polytomies, likely to indicate undersampled outbreaks) (B) Further filtered by
-removing HMM inconsistent breakpoints (YYY parent pairs). Note that
-Figure~\ref{fig:recomb_mrcas_voc_breakdown} shows the same data broken down by
-parental VoC classification.}
+Date of common ancestry between the parents on either side of recombination
+breakpoints, as a function of parental divergence. MRCAs of parents
+inferred to have caused a Pango designated recombinant (XA, XB, etc.)
+identified in orange (A) 851 parent pairs (singleton recombinants removed).
+Parental strains often descend from the same MRCA; each horizontal dotted line
+indicates a frequently found MRCA: these tend to be associated with major waves
+and often have many immediate children (i.e. are large
+polytomies, likely to indicate undersampled outbreaks) (B) Further filtered
+to 358 parent pairs by removing HMM inconsistent breakpoints. For equivalent
+plots broken down by parental VoC classification, see
+Figure~\ref{fig:recomb_mrcas_voc_breakdown}
+}
 \end{figure}
@@
 majority of newly identified recombinants arise from parents that diverged 10 to 30 weeks previously. 
-However, there is also a peak of recombinants that diverged ~95 weeks prior to the event: these correspond to recombinants whose parental common ancestor traces to early 2020, including both Delta-Delta and Delta-Omicron recombinants (Supplementary Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
-Note that each point in these plots represents a breakpoint, so that recombination events with more than one breakpoint (e.g.
-with 3 or more parents, which comprise X \% of the recombination events) are represented by several plotted points.
-
-As a check on the reliability of these identified recombination events,
+However, there is also a peak of recombinants that diverged ${\sim}95$ weeks prior to the event:
+these correspond to recombinants whose parental common ancestor traces to early 2020,
+including both Delta-Delta and Delta-Omicron recombinants
+(Supplementary Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
+Note that each plotted point represents a breakpoint, so that a recombination
+event with more than one breakpoint (e.g. with 3 or more parents, comprising
+${\sim}10 \%$ of the recombination events in Figure ~\ref{fig:recomb_mrcas}A and
+${\sim}3 \%$ in Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
+
+As a check on the reliability of these identified recombinations,
 Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are
-HMM-consistent. This reduces both the number of recombination nodes /
-breakpoints from 1522 / B to C / D in the wide ARG and from 763 / 806 to 459 in
-the long ARG (Supplementary Tables 1 and 2).
+HMM-consistent. This reduces both the number of recombination nodes \&
+breakpoints in the long ARG from 763 \& 851 to 347 \& 358 (
+and from from 1522 \&  to C / D in the wide ARG, see Supplementary Tables 1 and 2).
 
@@
 \subsubsection{Wide ARG}
-[Quick overview of the wide ARG.]
+%% Quick overview of the wide ARG.
 In the wide ARG, 44
@@
 
+%% Textual description of the 3 recombinant lineages present in the wide ARG
 The simplest Pango recombinant genealogy is XA: all XA samples trace back to a
@@
 with no involvement of  B.1.36.8, and it is possible that the third parent in
-the wide ARG is artifactual. Additional complications arise because additional
+the wide ARG is artefactual. Additional complications arise because additional
 non-X-designated samples such as B.1.634 also descend from this recombination
@@
 further recombination nodes nested within the XB ``clade''. However, all but
-one of these nested recombinations are ancestral to only a negligible fractions
+one of these nested recombinations are ancestral to negligible fractions
 of the designated XB samples. The exception accounts for about 17\% of the
@@
 between an XB grandparent and its XB grandchild. This could well be
-artifactual.
+artefactual.
 

### 2023-05-12 paper.tex

@@
 
-\title{Toward Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2}
+\title{Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2}
 \author[1]{Shing H. Zhan}
@@
 a single tree in the middle of the viral genome
-%% (covering the interval 8393--21618),  % removed as interval now in the figure title
 since other regions of the
@@
 
-\subsection{Early recombinants reported by Jackson et al. (2021)}
+\subsection{Early recombinants}
 \label{sec:jackson_recombs}
-% - Early Recombinants from Jackson et al
-%     Discuss results from Wide ARG in detail. Go into the details of the
-% forward-backward HMM runs for each example. Be much more explicity about the
-% single origin for the different groups (maybe give one of Ana's graphs for
-% Group A (XA I think?)
-
+% FIXME: this is rubbish, can someone do better?
+Although coronaviruses are generally known to recombine [CITATION?]
+it was not clear that SARS-CoV-2 recombinants with ongoing transmission
+existed early in the pandemic.
+\citet{Jackson2021-ik} performed a careful analysis of sequences
+circulating in the UK in late 2020 to early 2021, and found
+evidence of multiple independent recombination events and onward
+transmission.
 By searching for samples combining genomic segments from Alpha (B.1.1.7) and
 from the parental lineage B.1.1 based on a list of 22 Alpha-defining mutations,
-\citet{Jackson2021-ik} reported 16 recombinant sequences among genomes sampled
-in the United Kingdom up to March 7, 2021. The authors proposed that the
-recombinants likely arose from eight independent recombination events
-(categorised into groups A to D and four singletons).
-
-% jk: this is the proposal from issue #33
+they found 16 recombinant sequences from 8 putative
+origins (groups A to D and four singletons).
+These findings are closely replicated in both the Wide and Long ARGs.
 
 \begin{table} \centering
-\begin{tabular}{ll|clll}
+\begin{tabular}{lllll}
 \toprule
-A Group (XA) & Jackson        &  21,256–21,613 & B.1.177/B.1.1.7 \\
-             & \texttt{sc2ts} &  21,256--22,227 & B.1.177.18/B.1.1.7 \\
+Group        & Sequences & Method & Interval & Parents \\
 \midrule
-B Group & Jackson        &  6,529--6,953 & B.1.36/B.1.1.7  \\
-        & \texttt{sc2ts} &  6,529--6,954 & B.1.36/B.1.1.7  \\
-\midrule
-C Group & Jackson        &  25,997--27,441 &  B.1.1.7/B.1.221 \\
-        & \texttt{sc2ts} &  25,997--27,972 &  B.1.1.7/B.1.221 \\
-\midrule
-D Group & Jackson        &  21,576--23,063 &  B.1.36.17/B.1.1.7 \\
-        & \texttt{sc2ts} &  22,445--23,063 &  B.1.36.39/B.1.1.7 \\
-\midrule
-CAMC-CBA018 & Jackson        &  20,390--21,254 & B.1.177/B.1.1.7 \\
-            & \texttt{sc2ts} &  16,177--21,255 & B.1.177/B.1.1.7 \\
-\midrule
+A (XA)       & 4   & Jackson        &  21,256--21,613 & B.1.177/B.1.1.7 \\
+             &     &\texttt{sc2ts} &  21,256--22,227 & B.1.177.18/B.1.1.7 \\
+\cmidrule{3-5}
+B            & 2   & Jackson        &  6,529--6,953 & B.1.36/B.1.1.7  \\
+             &     &\texttt{sc2ts} &  6,529--6,954 & B.1.36/B.1.1.7  \\
+\cmidrule{3-5}
+C            & 3   &Jackson        &  25,997--27,441 &  B.1.1.7/B.1.221 \\
+             &     & \texttt{sc2ts} &  25,997--27,972 &  B.1.1.7/B.1.221 \\
+\cmidrule{3-5}
+D            & 3   & Jackson        &  21,576--23,063 &  B.1.36.17/B.1.1.7 \\
+             &     & \texttt{sc2ts} &  22,445--23,063 &  B.1.36.39/B.1.1.7 \\
+\bottomrule
 \end{tabular}
-\caption{\label{tab:jackson}Comparison of recombination breakpoint intervals and parent lineages for a subset of the data reported by \cite{Jackson2021-ik}.
-[FIXME ADD MORE DETAIL] More detailed results are given in Table~\ref{tab:jackson_supplement}.}
+\caption{\label{tab:jackson}Comparison of recombination breakpoint
+intervals and parent lineages for Groups A-D
+reported by \cite{Jackson2021-ik} with the corresponding
+recombination events in the Wide ARG.
+The second column gives the number of sequences in the group.
+Intervals are given in left-inclusive, right-exclusive form.
+Details for all 16 sequences are
+given in Table~\ref{tab:jackson_supplement}.}
 \end{table}
 
-In the wide ARG (sampled to mid-2021), all these sequences are descendants of
-recombination nodes, except MILK-103C712, which was removed during
-preprocessing (Table 2). Additionally, all except QEUH-1067DEF are HMM-consistent:
-that is, the mirrored HMM runs (i.e. forward from 5’ to 3’ end versus backwards from 3’ to 5’ end)
-identify the same set of originating mutations, have the same number of breakpoints, and have
-equivalent (sequence-identical) parents from forward and backward runs.
-Overall, we find excellent agreement between Jackson et al.'s
-motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
-and (2) the breakpoint intervals (Table 2). In all cases, the two methods
-proposed identical or very closely related parent lineages. The general mosaic
-genome structure is concordant in all the cases, even CAMC-CB7AB3 where there
-are two switches in the HMM copying paths (indicative of two breakpoints). The
-breakpoint intervals estimated using the two methods overlap substantially in
-all the cases (Table 2). Within each of the groups A to D, the constituent
-sequences have identical breakpoint intervals, because they trace back to a
-single recombination node in the wide ARG. This result is consistent with the
-hypothesis that the sequences of the groups A to D emerged from four separate
-recombination events~\cite{Jackson2021-ik}.
+The Wide ARG contains 15 of these 16 recombinant sequences
+(sample MILK-103C712 was removed during preprocessing; see
+Section~\ref{sec:data_preprocessing}), and the results are summarised
+in Tables~\ref{tab:jackson} and \ref{tab:jackson_supplement}.
+Table~\ref{tab:jackson} shows the groups of sequences identified by Jackson
+et al.\ as likely independent recombination events with onward transmission.
+In each case we have a corresponding recombination node in the Wide ARG,
+from which all the sequences in the group descend. The parent lineages
+and breakpoint intervals agree closely (see
+Section~\ref{sec:breakpoint_intervals} for more details on breakpoint
+intervals).
+For groups B, C and D,
+these recombination nodes form clades consisting only of the identified
+sequences.
+Group A sequences were subsequently given the Pango XA designation
+following onward transmission,
+and there are 44 XA designated samples in the Wide ARG (including the
+4 sequences analysed by Jackson et al.). The Group A recombination
+node forms a monophyletic clade of these 44 samples.
+Table~\ref{tab:jackson_supplement} shows the details for each of the
+16 sequences individually, and showing generally a strong concordance
+in mosaic structure and parent lineages
+(including CAMC-CB7AB3 which is inferred to have two breakpoints under both
+methods).
+
+The Long ARG contains 5 of the sequences: two each from groups A and B
+and QEUH-1067DEF. These cluster under three recombination nodes, as expected,
+and have identical breakpoint intervals and parental lineage assignments
+to those of Wide ARG.
+The recombination nodes for Group B and QEUH-1067DEF are ancestral only
+to the sequences involved.
+The recombination node for group A forms a monophyletic clade of all
+5 XA samples present in the Long ARG
+(Figure~\ref{fig:pango-simple-origin-graph}A)
 

### 2023-05-12 paper.tex
Some text on breakpoints

@@
 \label{sec:breakpoint_intervals}
+It is rarely possible to be precise about the position on the genome at which
+a recombinant sequence switches from inheriting from one parent to another.
+Even if we observe the recombinant sequence before signficant
+divergence occurs (during onward transmission), if the two parent sequences
+are similar there is no way to distinguish them. Here we define the
+interval for a particular breakpoint as the interval over which the
+sequences for the left and right parent nodes are identical. The right-hand
+extreme of the interval is chosen by the LS HMM Viterbi algorithm
+(Section~\ref{sec:ls}, and the left bound can then be derived by
+sequence comparison of the parents.
 
@@
 HMM-consistent. This reduces both the number of recombination nodes \&
-breakpoints in the long ARG from 763 \& 851 to 347 \& 358 (
-and from from 1522 \&  to C / D in the wide ARG, see Supplementary Tables 1 and 2).
+breakpoints in the long ARG from 763 \& 851 to 347 \& 358
+(and in the wide ARG from 1522 \& 1769 to XXX \& YYY , see
+Supplementary Tables 1 and 2).
 
@@
 
-Finally, it is clear from Supplementary Figure 4 that of the recombinant
+Finally, it is clear from Supplementary Figure\ref{fig:recomb_mrcas_voc_breakdown} that of the recombinant
 parents that fall into the Alpha, Delta, and Omicron VoC classes, the majority
@@
 \subsubsection{Long ARG}
-[high level summary of the pango recombinant lineages
-and samples in the Long ARG]
+%% high level summary of the pango recombinant lineages and samples in the Long ARG
 The long ARG (subsampled to mid-2022) contains 33
 Nextclade-designated Pango recombinant lineages (XA to XAK; Supplementary Table
-XXX) comprising 749 samples (711 when singleton recombinants are removed). If
-the GISAID designations are used, 38 Pango lineages exist in the long ARG,
+XXX) comprising 749 samples (711 when singleton recombinants are removed). In
+contrast, GISAID designates 38 Pango recombinant lineages in the long ARG,
 comprising only 515 samples. In the analyses below, we mainly focus on the
@@
 
-Of the 749 Nextclade recombinant samples, all but 2 are inferred to descend
+Of the 749 Nextclade-designated recombinants, we infer that all but two descend
 from a recombination node. These two samples are designated XP, and a likely
-explanation for their not have an inferred recombinant origin in the long ARG
+explanation for the absence of a corresponding recombination event in the long ARG
 is that the characteristic multibase deletion for XP
@@
 
-The remaining remaining 630 Pango X designated sequences trace back to 50
+The remaining remaining 630 Pango recombinant samples trace back to 50
 different recombination nodes. We examine the inferred inheritance structure in
@@
 \subsubsection{Single origin}
-We would usually assume that all sequences
+In the absence of genealogical information, a reasonable initial assumption is
+that all sequences
 assigned to a given Pango X lineage are descendants of a single recombinant
 sequence, arising as a result of a mixed infection followed by onward
-transmission. Thus, we would also expect our ARG to reflect this evolutionary
-history, where we have a all the samples assigned to this recombinant lineage
-tracing back to a single recombination node, representing this original
+transmission. We would expect our ARG to reveal evolutionary histories of this
+nature, where all the samples assigned to a given recombinant lineage
+trace back to one recombination node, representing a single originating
 recombination event.
@@
 \begin{figure}
-\begin{tabularx}{\textwidth}{c}
-
-% \end{footnotesize}
-% \\
-
+\centering
 \includegraphics[width=\textwidth]{figures/Pango_XA_XAG_XD_nxcld_tight_graph.pdf}
-% }
-\end{tabularx} \caption{\label{fig:pango-simple-origin-graph} Simple origin
+
+\caption{\label{fig:pango-simple-origin-graph} Simple origin
 Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
@@
 
-In Figure~\ref{fig:pango-simple-origin-graph}A we illustrate the case of XA.
-We construct a small subgraph of the long ARG, tracing routes from Nextclade-identified XA samples to the closest other sample strains in our dataset.
-Where a route passes upwards through a recombination node, we only show the continuing parental lineages to the most recent sample ancestors, rather than descending the ARG downward to cousins of the recombinant parents.
-Dotted lines show where this subgraph links to other nodes within the ARG.
-It is clear from the subgraph that all the XA samples trace to a single originating recombination node.
-
-Table~\ref{tab:pango-single-origin} summarises the sequences of all 16 of the
-Pango recombinant lineages that trace back to a single recombination node in
-the long ARG. [FIXME UPDATE THESE] We can see that XA, XAC, XAE, and XF form
-monophyletic clades, in which all of the samples given that Pango designation
-descend from a single recombination node.
-
-[Discuss remaining]
-
-There is generally good agreement on the imputed parental lineages from the ARG
-and the official Pango designations. [Discuss more]
-
-[Discuss panel B where XAG shares a common recombinant origin with samples from
-several other X lineages]
+A simple example is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
+This shows a subgraph of the long ARG, in which paths have been traced above all
+Nextclade-identified XA samples to the closest other sample strains in our dataset.
+Dotted lines show where this subgraph links to other nodes within the ARG, and above
+recombination nodes only
+ancestral nodes are shown (i.e. cousins of the recombinant parents are omitted).
+It is clear from the subgraph that all the XA samples trace to a single originating
+recombination node (plotted in black), with a breakpoint position around 22227bp.
+
+%% Discuss panel B where XAG shares a common recombinant origin with samples from several other X lineages
 
@@
 Supplementary table XXX )
+
+Supplementary Table~\ref{tab:pango-single-origin} summarises the sequences of all 16 of the
+Pango recombinant lineages that trace back to a single recombination node in
+the long ARG. [FIXME UPDATE THESE] We can see that XA, XAC, XAE, and XF form
+monophyletic clades, in which all of the samples given that Pango designation
+descend from a single recombination node.
+
+[Discuss remaining]
+
+There is generally good agreement on the imputed parental lineages from the ARG
+and the official Pango designations. [Discuss more]
 

### 2023-05-12 paper.tex
Yans updates

@@
 \label{sec:breakpoint_intervals}
-It is rarely possible to be precise about the position on the genome at which
-a recombinant sequence switches from inheriting from one parent to another.
-Even if we observe the recombinant sequence before signficant
-divergence occurs (during onward transmission), if the two parent sequences
-are similar there is no way to distinguish them. Here we define the
-interval for a particular breakpoint as the interval over which the
-sequences for the left and right parent nodes are identical. The right-hand
-extreme of the interval is chosen by the LS HMM Viterbi algorithm
-(Section~\ref{sec:ls}, and the left bound can then be derived by
-sequence comparison of the parents.
 

### 2023-05-12 paper.tex
Updates from Overleaf

@@
 A simple example is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
-This shows a subgraph of the long ARG, in which paths have been traced above all
+This shows a subgraph of the long ARG, in which paths have been traced from all
 Nextclade-identified XA samples to the closest other sample strains in our dataset.
 Dotted lines show where this subgraph links to other nodes within the ARG, and above
-recombination nodes only
-ancestral nodes are shown (i.e. cousins of the recombinant parents are omitted).
+recombination nodes only ancestral nodes are shown (i.e. we do not descent to include
+cousins of recombinant parents).
 It is clear from the subgraph that all the XA samples trace to a single originating
@@
 
-Figure~\ref{fig:pango-simple-origin-graph}B shows a case of another
+A more complex single-origin case is Figure~\ref{fig:pango-simple-origin-graph}B shows a case of another
 Pango lineage, XAG, whose Nextclade-designated samples also trace to a

### 2023-05-15 paper.tex
Updates to early recomb from Ana

@@
 Recombination via template switching is a common feature
-in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt}),
+in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
 including SARS-CoV-2
@@
 
-Recombination was challenging to detect early in the SARS-CoV-2 pandemic due to
-gradual accumulation of genetic variation. In late 2020, the appearance and
+The lack of genetic variation made recombination challenging to detect early in the SARS-CoV-2 pandemic. In late 2020, the appearance and
 spread of variants of concern (VoC), designated into classes such as Alpha and
@@
 structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
-encoding of large ARGs, and this representation should be well-suited to storing
-recombinant viral genealogies, even those with millions of sampled sequence,
-such as SARS-CoV-2 (with currently over 15 million sequences in the GISAID database
+encoding of large ARGs. This representation should be well-suited to storing
+large recombinant viral genealogies such as SARS-CoV-2,
+which currently encompasses over 15 million sequences in the GISAID database
 \citep{Shu2017-hp}). Moreover the \emph{tskit} software library, which implements this
@@
 of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
-we demonstrate that sc2ts can be used to simultaneously (1) detect putative SARS-CoV-2
+we demonstrate that \emph{sc2ts} can be used to simultaneously (1) detect putative SARS-CoV-2
 recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
@@
 \emph{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
-Sc2ts is a promising method to automate detection of SARS-CoV-2 recombinants
+\emph{Sc2ts} is a promising method to automate detection of SARS-CoV-2 recombinants
 and to jointly model recombinational and mutational processes shaping the genomic
@@
 mutations per sample, demonstrating that new samples are generally
-well represented by the it. See Section~\ref{sec:mutation_spectrum}
+well represented by it. See Section~\ref{sec:mutation_spectrum}
 for more analysis of the patterns of inferred mutations.
@@
 Tanglegram comparing a local tree from the Wide ARG
-(sampled to mid-2021) and an ``all-time'' global NextStrain tree
+(sampled to mid-2021) and an ``all-time'' global Nextstrain tree
 (downloaded on January 21, 2023).
@@
 A classic tree-based summary of SARS-CoV2 ancestry is provided
-by the NextStrain project \citep{Hadfield2018-ef}. The trees
+by the Nextstrain project \citep{Hadfield2018-ef}. The trees
 available from Nextstrain are based on small subsamples of the
@@
 For validation purposes, we our compare our ARGs with
-a downloaded NextStrain tree, restricted to the time period
+a downloaded Nextstrain tree, restricted to the time period
 covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
-the \texttt{sc2ts} ARGs to a ``backbone'' containing only samples present
-the NextStrain tree. This results in a small set of shared
+the \texttt{sc2ts} ARGs to a backbone containing only samples present
+the Nextstrain tree. This results in a small set of shared
 samples (180 for the Wide ARG, 88 for the Long ARG), none of which are
@@
 recombination events involve minor, local topological rearrangements. In particular,
-all involve recombination between closely related lineages (e.g.\ sibling lineages)
-and a number even represent recombination within a single lineage, in which the parent
-lineages above a recombination event coalesce with each other before coalescing
-with any other lineages in the phylogeny. These could either represent true
-recombination events between closely related strains, which have previously gone unnoticed,
+recombination in the backbone phylogenies only occurs between close relatives
+(e.g.\ sibling lineages), sometimes even within a single lineage such that the parent
+lineages above a recombination event coalesce with each other before involvement with
+any other part of the phylogeny. These local reticulations could either represent
+real but previously unobserved recombination events between closely related strains,
 or could be artifacts of the current \texttt{sc2ts} method and its parameterization.
@@
 Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
-a GISAID global ``all-time'' tree from NextStrain
+a GISAID global ``all-time'' tree from Nextstrain
 % We give the date below in the caption - we could probably remove if either from there
@@
 It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
-agreement with the NextStrain tree. The sample genomes cluster by their
+agreement with the Nextstrain tree. The sample genomes cluster by their
 assigned Pango lineage status, and many variants and their descendants form
@@
 % FIXME: this is rubbish, can someone do better?
-Although coronaviruses are generally known to recombine [CITATION?]
-it was not clear that SARS-CoV-2 recombinants with ongoing transmission
-existed early in the pandemic.
-\citet{Jackson2021-ik} performed a careful analysis of sequences
+RNA viruses are known to recombine at high rates 
+\citep{Simon2011-rna}, and recombination has been widely 
+documented to be commonplace in animal and human coronaviruses 
+\citep{Su2016-epidemiology}. While recombination in SARS-CoV-2 
+was shown early on to be frequent \emph{in-vitro} 
+\citep{Gribble2021-coronavirus}, the relatively slow accumulation 
+of genetic diversity early on in the pandemic hampered efforts to
+detect recombinant strains. A number of early studies relying on 
+analysing patterns of linkage disequilibrium and searching for 
+mosaic genomes carrying characteristic mutations of different
+lineages either failed to detect recombination or posited 
+that this occurred at low rates \citep[e.g.~][]{Nie2020-phylogenetic,Tang2020-origin,VanInsberghe2021-eu,Varabyou2021-rw}.
+The first clear evidence of recombinant lineages was presented by
+\citet{Jackson2021-ik}, who performed a careful analysis of sequences
 circulating in the UK in late 2020 to early 2021, and found
@@
 
-Both the wide and long ARGs contain a large number of recombination nodes (4123
-and 2078 respectively; Supplementary Tables 1 and 2). Each of these nodes were
+Although recombination nodes are relatively infrequent, the wide and long ARGs still
+contain many of them (4123 and 2078 respectively; Supplementary Tables 1 and 2).
+Each of these nodes were
 inferred as a result of matching a specific recombinant sample into the ARG.
@@
 figure. These shared MRCAs are near the root of large expansions of certain
-lineages (B.1: original variant;  B.1.617.2: Delta, BA.1: first Omicron wave;
-and BA.2: second Omicron wave), as evidenced by the number of samples with that
-Pango designation that trace back to these ancestral nodes (listed as
-percentages on the plot). The specific MRCA nodes to which many recombinants
-trace frequently represent very large polytomies in the tree (at the extreme,
-2,872 immediate children). This is likely to indicate a rapid and under-sampled
-expansion of a clade in the SARS-CoV-2 genealogy.
+lineages (B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
+and BA.2: second Omicron wave), as evidenced by the number of samples with those
+Pango designations which trace back to each ancestral node (listed as
+percentages on the plot). The majority of these frequently-shared MRCAs
+are associated with extensive polytomies in the tree (at the extreme,
+the BA.2-linked node has 2,872 immediate children). This is likely to
+indicate a rapid and under-sampled expansion of a clade in the SARS-CoV-2 genealogy.
 
@@
 reduced to 235 (229) and XC to 4 (0). We correctly identify that all samples
-designated as XB by any method have one or more recombination nodes in their
-ancestry.
+designated as XA, XB, or XC by any method have one or more recombination nodes
+in their ancestry.
 
@@
 
-The remaining remaining 630 Pango recombinant samples trace back to 50
-different recombination nodes. We examine the inferred inheritance structure in
-the following subsections.
+The remaining remaining 630 Pango recombinant samples, representing 31
+Nextclade-designated Pango lineages, trace back to 50 different recombination nodes.
+We examine the inferred inheritance structure in the following subsections.
 
@@
 This shows a subgraph of the long ARG, in which paths have been traced from all
-Nextclade-identified XA samples to the closest other sample strains in our dataset.
-Dotted lines show where this subgraph links to other nodes within the ARG, and above
-recombination nodes only ancestral nodes are shown (i.e. we do not descent to include
-cousins of recombinant parents).
-It is clear from the subgraph that all the XA samples trace to a single originating
+Nextclade-identified XA samples (in red) to the closest other sample nodes in our dataset. The
+subgraph may thus include inserted, non-sample nodes (i.e. UPGMA, recombination, mutation
+collapse, or reversion push nodes, all plotted as smaller circles) on the path to the nearest
+sample nodes (larger circles). Dotted lines show where this subgraph links to the rest of the
+ARG. Above recombination nodes only ancestral nodes are shown, so that e.g. cousins of non-sample
+recombinant parents are not shown. It is clear from the subgraph that all the XA samples trace
+to a single originating
 recombination node (plotted in black), with a breakpoint position around 22227bp.
@@
 
-A more complex single-origin case is Figure~\ref{fig:pango-simple-origin-graph}B shows a case of another
-Pango lineage, XAG, whose Nextclade-designated samples also trace to a
-single originating recombination. This is a more intricate example, as the same origin
-is also shared with all samples designated XAA, and some, but not all, of the
-samples identified as XAB, XQ, and XU by Nextclade. Note, however, that if the GISAID
-designations are used, many of the samples marked here as XAB are reclassified
-as BA.2, and XAG becomes fully monophyletic (supplementary figure
-\ref{fig:pango_XAG_gisaid_graph}, which also gives identifying labels to the
-sample nodes, convertable to EPI\_ISL identifiers using
-Supplementary table XXX )
-
-Supplementary Table~\ref{tab:pango-single-origin} summarises the sequences of all 16 of the
-Pango recombinant lineages that trace back to a single recombination node in
-the long ARG. [FIXME UPDATE THESE] We can see that XA, XAC, XAE, and XF form
-monophyletic clades, in which all of the samples given that Pango designation
-descend from a single recombination node.
-
-[Discuss remaining]
-
-There is generally good agreement on the imputed parental lineages from the ARG
-and the official Pango designations. [Discuss more]
+A more complex single-origin case is XAG, illustrated in
+Figure~\ref{fig:pango-simple-origin-graph}B. Here, the XAG samples
+all trace back to the same most recent common recombination node (black), but
+we infer this recombination event to also be the originating event for
+all the recombinant samples designated XAA,
+and some, but not all, of those identified as XAB, XQ, and XU by Nextclade.
+
+The classification of originating recombination events is dependent on
+accurate designation of Pango lineages to samples. It is therefore important
+to note that if the GISAID Pango designations are used, many of the samples
+marked here as XAB are reclassified
+as BA.2 (and XAG becomes fully monophyletic, see supplementary figure
+\ref{fig:pango_XAG_gisaid_graph}).
+
+In total we identify 16/31 single-origin Pango recombinant lineages in the long ARG.
+Supplementary Table~\ref{tab:pango-single-origin} summarises these origination events
+together with their inferred parents, showing good agreement between \texttt{sc2ts} and
+the official Pango designated parents, with \texttt{sc2ts} providing more a precise
+parent designation. Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
+simple-origin (XA) type, whereas the rest are of the XAG type, and include multiple Pango
+designated lineages as descendants of the originating recombination event. In some cases,
+these may be caused by samples which have been designated as the wrong recombinant lineage
+by Nextclade.
 
@@
 a single recombination event. In this section we examine the Pango X lineage
-samples that are inferred to have multiple recombinant origins (i.e., trace
+samples that are inferred to have multiple recombinant origins (i.e. trace
 back to independent recombination nodes in the ARG).
@@
 subgraph has a single sample with EPI\_ISL 8514045 (sampled in Turkey and dated 2021-12-30).
-These two samples differ by 35 mutations; considering this, plus the geographical
-locations and time between the samples,
-it does seem possible that they represent true independent origins of XD.
+The different geographical locations, the time between the samples, and
+the fact that the two samples differ at 23 nucleotide positions,
+suggests that these may indeed represent true independent origins of XD.
 Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the same subgraphs
@@
 
-Another X lineage with multiple originating events is XE. In this case, however,
-there are a large number of independent events. These
-
-\ref{tab:multiple_origins_table} summarises other Pango recombinants which
-we infer to have multiple origins. [TODO: needs more detail]
-As with XD, none of the samples in the long ARG are identified as XE by GISAID.
-
-\subsubsection{Complex origins}
+Another X lineage we infer to have multiple origins is XE. In this case, 
+we identify at least 4 independent origination events of the Nextclade-designated
+XE samples (Supplementary Table~\ref{tab:multiple_origins_table}).
+
+[TODO: needs more detail]
+
+\subsubsection{Nested recombinant origins}
 As well as cases where single recombination events explain an X lineage,
@@
 ``hairball'' of further recombination nodes whose descendants are mostly not
-identified as recombinants by Nextclade, and plotted in blue. A similar
+identified as recombinants by Nextclade (plotted in blue). A similar
 pattern is seen when examining XB in the wide ARG (discussion above, plot not shown).
@@
 indicating some uncertainty in lineage assignment in this part of the ARG.
-Investigating examples of complex recombinant origins, and identifying which
+Investigating examples of nested recombinant origins, and identifying which
 (if any) of the nested recombination events may be artefactual, is an important
@@
 %% produce this table is in Pango-recombinant-analysis.ipynb
+\small
 \begin{tabular}{lllrl}
 \toprule
-lineage & ARG parents & Official Pango parents & n & Other descendants \\
+ & Sc2ts inferred parents & Official Pango parents & n & Other descendants \\
+Pango &  &  &  &  \\
 \midrule
-XA      &       B.1.1.7 + B.1.177.18 &     B.1.1.7 + B.1.177 &   5 & -- \\
-XAA, XAG     &              BA.1 + BA.2.9 &         BA.1* + BA.2* &   8 &
-    XAB: 39, XAG: 17, XAA: 8 \\
-    &&&&   XAA: 8, XQ: 2, XU: ... \\
-XAC     &          [BA.2.3, BA.1.17.2] &  [BA.2*, BA.1*, BA.2*] &  27 & -- \\
-XAE     &                 [BA.2, BA.1] &         [BA.2*, BA.1*] &   9 & XAE: 9 \\
-XAG     &               [BA.1, BA.2.9] &         [BA.1*, BA.2*] &  17 &
-    XAB: 39, XAG: 17, XAA: 8 \\
-    &&&&   XAA: 8, XQ: 2, XU: ... \\
-XF      &                 [AY.4, BA.1] &    [B.1.617.2*, BA.1*] &   2 & XF: 2 \\
-XG      &              [BA.1.17, BA.2] &         [BA.1*, BA.2*] &  32 & XG: 32, XAB: 1 \\
-XH      &            [BA.1.20, BA.2.9] &         [BA.1*, BA.2*] &  11 &
-    XAF: 34, XH: 11 \\
-&&&& B.1.1.529: 3, XE: 3,... \\
-XK      &             [BA.1.1.1, BA.2] &         [BA.1*, BA.2*] &   6 & XK: 3 \\
-XL      &            [BA.1.17.2, BA.2] &         [BA.1*, BA.2*] &  10 &
-    XL: 10, XAB: 1, XU: 1, \\
-    &&&& XQ: 1 \\
-XR      &               [BA.1.1, BA.2] &       [BA.1.1*, BA.2*] &   8 &
-    XQ: 9, XR: 8, XAB: 2\\
-XS      &              [AY.36, BA.1.1] &  [B.1.617.2*, BA.1.1*] &   6 & XS: 4\\
-XT      &       [BA.2.23, Unknown (R)] &         [BA.2*, BA.1*] &   1 &
-    BA.2: 1, BA.1: 1, BA.1.9: 1, \\
-    &&&& BA.2.23:... \\
-XV      &  [BA.1.1, BA.2, Unknown (R)] &         [BA.1*, BA.2*] &   3 & BA.2: 8, XV: 1 \\
-XW      &            [BA.1.1.15, BA.2] &         [BA.1*, BA.2*] &  11 & XW: 11, XN: 2 \\
-XY      &               [BA.1.1, BA.2] &         [BA.1*, BA.2*] &   6 & XY: 6, XAF: 1 \\
+\bfseries XA & B.1.1.7 + B.1.177.18 & B.1.1.7 + B.1.177 & 5 & – \\
+\bfseries XAA & BA.1 + BA.2.9 & BA.1* + BA.2* & 8 & XAB:~38/48, XAG:~17/17, XU:~1/3, XQ:~2/12 \\
+\bfseries XAC & BA.1.17.2 + BA.2.3 & BA.1* + BA.2* & 27 & – \\
+\bfseries XAE & BA.1 + BA.2 & BA.1* + BA.2* & 9 & – \\
+\bfseries XAG & BA.1 + BA.2.9 & BA.1* + BA.2* & 17 & XAB:~38/48, XAA:~8/8, XU:~1/3, XQ:~2/12 \\
+\bfseries XF & AY.4 + BA.1 & B.1.617.2* + BA.1* & 2 & – \\
+\bfseries XG & BA.1.17 + BA.2 & BA.1* + BA.2* & 32 & XAB:~1/48 \\
+\bfseries XH & BA.1.20 + BA.2.9 & BA.1* + BA.2* & 11 & XAF:~34/35, B.1.1.529:~2, XE:~3/163 \\
+\bfseries XK & BA.1.1.1 + BA.2 & BA.1* + BA.2* & 3 & – \\
+\bfseries XL & BA.1.17.2 + BA.2 & BA.1* + BA.2* & 10 & XAB:~1/48, XU:~1/3 \\
+\bfseries XR & BA.1.1 + BA.2 & BA.1.1* + BA.2* & 8 & XQ:~9/12, XAB:~2/48 \\
+\bfseries XS & AY.36 + BA.1.1 & B.1.617.2* + BA.1.1* & 4 & – \\
+\bfseries XT & BA.2.23 + Unknown & BA.1* + BA.2* & 1 & BA.2.23:~1 \\
+\bfseries XV & BA.1.1 + BA.2 + Unknown & BA.1* + BA.2* & 1 & BA.2:~8 \\
+\bfseries XW & BA.1.1.15 + BA.2 & BA.1* + BA.2* & 11 & XN:~2/13 \\
+\bfseries XY & BA.1.1 + BA.2 & BA.1* + BA.2* & 6 & XAF:~1/35 \\
 \bottomrule
@@
 single recombinant origin based on the NextClade Pango lineage assignments.
-Parent lineages from the ARG (as designated by Nextclade and, where relevant
-and also to parent lineages as defined by the official Pango designation alias
-key. (TODO: explain the asterisks) FIXME: if we like this table then we can
-reformat it so that lists at the end are spread over multiple lines rather than
-using Python dict notation }
+Sc2ts inferred parents are based on imputed NextClade Pango lineages. Official parents
+are taken from Pango designation alias key (TODO: explain the asterisks). The final column
+list other inferred descendants of the originating recombination event, summarised
+by NextClade Pango designation and sample count / total ARG count (total count in the ARG
+is omitted for non-recombinant samples).
+}
 \end{table}

### 2023-05-16 paper.tex
Updates from duncan and Yan

@@
 which currently encompasses over 15 million sequences in the GISAID database
-\citep{Shu2017-hp}). Moreover the \emph{tskit} software library, which implements this
+\citep{Shu2017-hp}). Moreover the \texttt{tskit} software library, which implements this
 data structure, enables efficient manipulation of the stored genealogies and is
@@
 Here, we present a preliminary version of a novel scalable methodology
-(\emph{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
+(\texttt{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
 and additionally takes advantage of known collection dates to infer ARGs from millions
 of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
-we demonstrate that \emph{sc2ts} can be used to simultaneously (1) detect putative SARS-CoV-2
+we demonstrate that \texttt{sc2ts} can be used to simultaneously (1) detect putative SARS-CoV-2
 recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
@@
 of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
-\emph{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
-\emph{Sc2ts} is a promising method to automate detection of SARS-CoV-2 recombinants
+\texttt{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
+\texttt{Sc2ts} is a promising method to automate detection of SARS-CoV-2 recombinants
 and to jointly model recombinational and mutational processes shaping the genomic
@@
 
-Although recombination nodes are relatively infrequent, the wide and long ARGs still
-contain many of them (4123 and 2078 respectively; Supplementary Tables 1 and 2).
-Each of these nodes were
-inferred as a result of matching a specific recombinant sample into the ARG.
-However, ~63\% of recombination nodes are ancestral to only a single
+Although recombination nodes are relatively infrequent, there are still many
+present in the wide and long ARGs (4123 and 2078 respectively; Supplementary
+Tables 1 and 2). Each recombination node was inferred as a
+result of matching a specific (``causal'') recombinant sample into the ARG.
+However, ~63\% of them are ancestral to only a single
 recombinant sample; we exclude these "singleton recombinants" from subsequent
@@
 As a check on the reliability of these identified recombinations,
-Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are
+Figure~\ref{fig:recomb_mrcas}B  plots only breakpoints which are
 HMM-consistent. This reduces both the number of recombination nodes \&
@@
 of recombination events; the top four are indicated by dotted lines in the
-figure. These shared MRCAs are near the root of large expansions of certain
+figure. These shared MRCAs lie near the root of large expansions of certain
 lineages (B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
@@
 The remaining remaining 630 Pango recombinant samples, representing 31
-Nextclade-designated Pango lineages, trace back to 50 different recombination nodes.
+Nextclade-designated Pango lineages, trace back to 50 different most recent
+recombination nodes.
 We examine the inferred inheritance structure in the following subsections.
@@
 Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
-imputed Pango status. (B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
+imputed Pango status. Genomic regions inherited by the recombination node use 
+breakpoint positions in the ARG, corresponding to the maximum inferred breakpoint position
+(B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
 (C) Equivalent plot for both origination events involving the four XD samples in the long ARG.
@@
 This shows a subgraph of the long ARG, in which paths have been traced from all
-Nextclade-identified XA samples (in red) to the closest other sample nodes in our dataset. The
-subgraph may thus include inserted, non-sample nodes (i.e. UPGMA, recombination, mutation
-collapse, or reversion push nodes, all plotted as smaller circles) on the path to the nearest
-sample nodes (larger circles). Dotted lines show where this subgraph links to the rest of the
-ARG. Above recombination nodes only ancestral nodes are shown, so that e.g. cousins of non-sample
-recombinant parents are not shown. It is clear from the subgraph that all the XA samples trace
-to a single originating
-recombination node (plotted in black), with a breakpoint position around 22227bp.
+Nextclade-identified XA samples (in red) to the closest other sample nodes in our dataset.
+Sample nodes are plotted as larger circles, but the subgraph also includes intermediate,
+non-sample nodes: i.e. inserted nodes such as UPGMA, recombination, mutation collapse, and
+reversion push nodes, plotted as smaller circles. Dotted lines show where this subgraph links
+to the rest of the ARG. Above recombination nodes, only ancestral nodes are shown, meaning
+that the subgraph is not extended to show other descendants of recombinant parents. It is clear
+from the subgraph that all the XA
+samples trace to a single originating recombination node,
+whose genome is a composite of a B.1.177.18 lineage on the left of the genome (up to a maximum
+position of ${\sim}22227$ basepairs) and a B.1.1.7 lineage for the remainder of the genome.
 
@@
 Figure~\ref{fig:pango-simple-origin-graph}B. Here, the XAG samples
-all trace back to the same most recent common recombination node (black), but
+all trace back to the same most recent common recombination node 
+(combining BA.1 on the left and BA.2.9 on the right), but
 we infer this recombination event to also be the originating event for
@@
 to note that if the GISAID Pango designations are used, many of the samples
-marked here as XAB are reclassified
-as BA.2 (and XAG becomes fully monophyletic, see supplementary figure
-\ref{fig:pango_XAG_gisaid_graph}).
-
-In total we identify 16/31 single-origin Pango recombinant lineages in the long ARG.
-Supplementary Table~\ref{tab:pango-single-origin} summarises these origination events
+marked here as XAB are reclassified as BA.2 (and XAG becomes fully monophyletic,
+see supplementary figure \ref{fig:pango_XAG_gisaid_graph}). This is an
+independent confirmation of the uncertainty in designation of these XAG-related
+samples.
+
+In total out of the 31 Pango recombinant lineages in the long ARG we identify 16 as
+having a single origin. Supplementary Table~\ref{tab:pango-single-origin} summarises these
 together with their inferred parents, showing good agreement between \texttt{sc2ts} and
@@
 designated lineages as descendants of the originating recombination event. In some cases,
-these may be caused by samples which have been designated as the wrong recombinant lineage
-by Nextclade.
+these may be, however, be a result of erroneous Pango designations.
 
@@
 requirement that all the samples assigned to a given lineage must trace back to
-a single recombination event. In this section we examine the Pango X lineage
+a single recombination event. In this section we examine several Pango X lineage
 samples that are inferred to have multiple recombinant origins (i.e. trace
@@
 sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
-subgraph has a single sample with EPI\_ISL 8514045 (sampled in Turkey and dated 2021-12-30).
-The different geographical locations, the time between the samples, and
-the fact that the two samples differ at 23 nucleotide positions,
-suggests that these may indeed represent true independent origins of XD.
-Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the same subgraphs
-with sample identifiers and coloured by GISAID Pango labels (however, no
-samples are designated by GISAID as XD in the long ARG).
-
-Another X lineage we infer to have multiple origins is XE. In this case, 
-we identify at least 4 independent origination events of the Nextclade-designated
-XE samples (Supplementary Table~\ref{tab:multiple_origins_table}).
-
-[TODO: needs more detail]
+subgraph has a single sample with EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
+Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
+the breakpoints in each case are slightly different: the start of the Omicron insertion
+in the French samples has an estimated maximum position of 21641 basepairs
+(and minimum of 21619) with the insertion end occurring at a maximum position of 25584
+(and a minimum of 25470). In contrast, the Omicron insertion in the Turkish sample
+is inferred to have occurred between positions 22578 (min 21619) and 24130 (min 23605).
+The breakpoint difference, the different geographical locations, the time between the
+samples, and the fact that the two samples differ at 23 nucleotide positions,
+suggests that these may indeed represent true independent Deltacron recombinants.
+The canonical XD definition is based entirely on samples from northern Europe, particularly
+France (\url{https://github.com/cov-lineages/pango-designation/issues/444}) so it seems
+plausible that the earlier Turkish sample has been mislabelled XD by Nextclade.
+Indeed, GISAID does not label any of these samples XD (see Supplementary
+Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).
+
+A similar pattern is seen for several of the origins inferred for the Pango
+XE lineage, which was originally defined on the basis of sequences from the UK and Ireland. The
+Long ARG infers five independent origins for nonsingleton Nextclade-designated XE
+samples (Supplementary Table~\ref{tab:multiple_origins_table}). One of the recombinant origins,
+caused in our ARG by an English sample, accounts for 155 out of 164 XE samples
+(primarily UK and Irish, but with a few northern European and north American strains).
+This certainly represents the canonical XE recombinant lineage. The other recombination nodes
+only give rise for one, two, or three XE samples. Three of these recombination
+nodes are likely to represent Nextclade mislabelling: they either have different
+inferred breakpoint positions or (in the case of the earliest recombination node),
+descendants that are primarily labelled as other Pango X-lineages. Additional support
+for the suggestion that they result from incorrect labelling comes from the fact that
+none of their XE-labelled descendant samples are from the UK or Ireland.
+However, the most recently derived XE cluster, consisting
+of two samples (one from Canada and one from England) may not be due to mislabelling: not
+only does it contain an English sample but it has exactly the same breakpoint interval as
+the canonical XE clade. It is possible that instead, this is a case where the \texttt{sc2ts} algorithm
+has taken a "true" XE sample and mistakenly inferred an additional, spurious origin. [WHY? - this
+is presumably because it has changed too much from the nearest in the main XE cluster - need to discuss the
+weighting of 3 mutations for one recombination here. NB: the number
+of nucleotide differences between the closest XE in the two clusters is 11: see the
+Pango-recombinant-analysis workbook]
 
@@
 \includegraphics[width=0.5\textwidth]{figures/ls_diagram_covid.pdf}
-% FIXME this caption is too long
 \caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
@@
 imperfect mosaic of the sequences in a reference panel.
-Black crosses along the focal sequence show genotyping
+Black crosses along the focal sequence show genotyping/sequencing
 errors or mutations.
@@
 case, we assume that all observed alleles (in both the reference panel and
-focal sequence, $a\ell$) at the site under consideration are equally likely and
+focal sequence, $a_\ell$) at the site under consideration are equally likely and
 occur with mutation probability $\frac{\mu_\ell}{a_\ell-1}$ (Figure~\ref{fig:ls_diagram}, blue panel).
@@
 \citep{Viterbi1967-ol} to find the most likely copying path through the
-sequences, given a set of reference sequences, $Q$, and $E$.
-The relative rates of mutation and recombination are usually controlled by
-the site-specific parameters $\mu_\ell$ and $r_\ell$, respectively.
+sequences, given a set of reference sequences, $Q$, and $E$. Throughout, we refer to the probabilities of mismatching to a member of the reference panel at site $\ell$; $\mu_\ell$,
+and the probability of recombining between two members of the reference panel between site $\ell-1$ and site $\ell$; $r_\ell$). $\mu_\ell$ is often referred to as the 
+`mutation probability' for convenience, but we note that this probability of 
+mismatch also encompasses various other error modes which result in a mismatch
+(such as sequencing and alignment errors). Note also that it is these probabilities,
+not the rates of mutation and recombination, that are required to fully define the HMM.
+That said, we can approximately relate these probabilities to the coalescent with recombination process which LS is approximating. In the case of recombination,
+suppose that there is some constant underlying recombination rate, $\rho$.
+Assuming that recombination events occur as a Poisson process, 
+then the probability of recombining between adjacent sites is:
+\begin{align*}
+    r_\ell = 1 - \exp\left(-\frac{\rho\left(m_{i-1} - m_{i}\right)}{n}\right),
+\end{align*}
+where $m_j$ is the nucleotide position of variant $j$. Similarly, we may consider the probability of mutation to represent to the 
+probability of mutation along the branch joining the focal sequence to a coalescent
+tree containing a reference panel with $n$ members, assuming a Poisson process
+along the branch. The expected length of this branch is $\frac{4}{n}$, and so the mismatch 
+probability represents the probability of observing a mutation or error mode, 
+along this branch. We can then approximate the probability of observing a mismatch as:
+\begin{align}
+    \mu_\ell \approx 1 - \exp{\left(-\frac{4\nu}{n}\right)}
+\end{align}
+where $\nu$ is the mutation rate per $\mathit{ploidy} \times N_e$ generations.
+The probabilities of these competing process of mismatch and recombination are usually controlled by the site-specific parameters $\mu_\ell$ and $r_\ell$, respectively.
 For this work we used a slightly different formulation, which uses a parameter, the mismatch ratio (MMR), to control the relative importance of mutation
@@
 prefer $k$ mismatches (mutations) to a single recombination which results in copying from a template with no mismatches. 
-To map between recombination probability and mutation probability for a particular mismatch ratio, we simply consider the two paths which we wish to be equally likely, and rearrange for the mutation or recombination probability.
-In the simple case when we assume $\mu_l=\mu$ and $r_l=r$, we obtain (supplementary methods):
+To map between recombination probability and mutation probability for a particular mismatch
+ratio, we simply consider the two paths which we wish to be equally likely, and rearrange 
+for the mutation or recombination probability. Consider the simple case when we assume 
+$\mu_l=\mu$ and $r_l=r$. Without loss of generality, consider a region of length $m$. Up 
+until this region, two paths are equally likely, so we can re-scale by the likelihood of 
+observing the focal sequence up to the site before the region starts, $c$. For a MMR of $k$,
+we only need to consider two of the $n$ members of the reference panel. One; 
+$\mathcal{P}_1$, for which there are no mismatches, but we need to recombine in, and a 
+second; $\mathcal{P}_2$, for which there are $k$ (randomly chosen) mismatches in the region.
+Now, follow the probability of tracking along each of those paths:
 \begin{align*}
-r = \frac{n\mu^k}{(1-\mu)^k + (n-1)\mu^k}, \quad \quad
+P(\mathcal{P}_1) &= \frac{cr}{n} \left(\left(1-r\right)+\frac{r}{n}\right)^m\left(1-\mu\right)^m &\text{recombine to a template with no mismatches,}\\
+P(\mathcal{P}_2) &= c \left(\left(1-r\right) + \frac{r}{n}\right)\left(\left(1-r\right)+\frac{r}{n}\right)^m \left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
+\end{align*}
+where $r$ is the recombination probability between adjacent variants, $\mu$ is the mismatch
+probability between a target sequence at a variant and a particular member of the reference
+panel, and $c$ is the likelihood of the path up to this point, which we assume to be equal.
+We then set these path probabilities equal, and rearrange to obtain:
+\begin{align*}
+r = \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k}, \quad\quad
 \mu = \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}.
@@
 
-A varying degree of missingness is present in the genetic data, that is, for a given sample, a subset of nucleotides coded as `unknown' at a set of positions.
-As a result, we had to consider missingness in our implementation of LS Viterbi.
-One approach would be to `impute' missing data, and hard-call or define some distribution over the possible missing nucleotide, and use that in the determination of path probabilities.
-Another is to not attempt to estimate the true nucleotide, but rather to consider such sites as uninformative to the path probability.
-We chose the latter.
-To do this, we simply encode the emission probability where the underlying nucleotide transition is $\mathit{unknown}\rightarrow \{A,C,G,T\}$, $\{A,C,G,T\}\rightarrow\mathit{unknown}$, or $\mathit{unknown}\rightarrow\mathit{unknown}$ at 1.
-
-% [ALSO MENTION MISSING DATA, where we deal with Ns in the alignment]
-% OLD text, ready to be adapted (and deleted once done)
-% For this
-% application, we have parameterised the HMM using a mismatch ratio (MMR), which
-% tunes how strongly mutations (i.e., no switching of template sequences) are
-% preferred over recombination (i.e., switching between sequences) during
-% copying.
+A varying degree of missingness is present in the genetic data, that is, for a given sample, 
+a subset of nucleotides coded as `unknown' at a set of positions. As a result, we had to
+consider missingness in our implementation of LS Viterbi. One approach would be to `impute'
+all missing data beforehand using external software to hard-call or define a 
+distribution over the possible missing nucleotides, and use that in the determination of path
+probabilities. Another is to not attempt to estimate the true nucleotide, but rather to 
+consider such sites as uninformative to the path probability, but to impute missing data 
+within that path on the fly, placing them within the reference panel iteratively. We chose the latter. To do this, we simply encode the emission
+probability where the underlying nucleotide transition is $\mathit{unknown}\rightarrow \{A,C,G,T\}$ as 1.
 
@@
 HMM is to copy from more than one ARG node along the genome, for a
-given sample sequence. [TODO MORE]
+given sample sequence.
+This statement warrants emphasis: the Viterbi algorithm enables us to find a path through the reference panel from among the $n^m$ paths that is provably at the optimum, under the LS model.
 
@@
 to reason about and explore the space of possible matches, which may
-be a fruitful avenue for future work. See the Discussion for further
-potential improvements that may be made to the HMM.
+be a fruitful avenue for future work.
+Examples include a stochastic traceback through the collection of paths at the global optimum to glean further information about our the likelihood surface, and determine whether there are downstream implications for our conclusions.
+Here, we have considered the Viterbi algorithm to make statements about the most likely paths through the data.
+The machinery used here can be modified to run the forwards and backwards algorithms which determine the probability of observing a focal sequence, integrating over all possible paths through the data, under LS model [cite D-Palms].
+This presents an opportunity to estimate parameters of interest while exploiting the tree-structure to gain log based speed-ups in exact their exact implementation.
 
@@
 \section*{Supplementary}
-\subsection*{Supplementary methods}
-\subsubsection*{Derivation of relation between recombination probability and mismatch probability conditional on mismatch ratio}
-Without loss of generality, consider a region of length $m$. Up until this region, two paths are equally likely, so we can re-scale by the likelihood up to the site before the region starts, $c$.
-
-For a MMR of $k$, we only need to consider two of the $n$ members of the reference panel.
-One; $\rho_1$, for which there are no mismatches, but we need to recombine in, and a second; $\rho_2$, for which there are $k$ (randomly chosen) mismatches in the region.
-
-We can simply follow the probability of tracking along each of those paths:
-\begin{align*}
-P(\rho_1) &= c \frac{r}{n} \left(\left(1-r\right)+\frac{r}{n}\right)^m\left(1-\mu\right)^m &\text{recombine to a template with no mismatches,}\\
-P(\rho_2) &= c \left(\left(1-r\right) + \frac{r}{n}\right)\left(\left(1-r\right)+\frac{r}{n}\right)^m \left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
-\end{align*}
-where $r$ is the recombination probability between adjacent variants, $\mu$ is the mismatch probability between a target sequence at a variant and a particular member of the reference panel, and $c$ is the likelihood of the path up to this point, which we assume to be equal.
-We then set these path probabilities equal, and rearrange to obtain:
-\begin{align*}
-r &= \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k},\\
-\mu &= \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}
-\end{align*}
-
 \begin{figure} \centering
@@
 
-\begin{table}
+\begin{sidewaystable}
 % \begin{footnotesize}
-%% FIXME - commas and braces are confusing - I suggest separating by "+". The code to
-%% produce this table is in Pango-recombinant-analysis.ipynb
-\small
+\centering
 \begin{tabular}{lllrl}
 \toprule
- & Sc2ts inferred parents & Official Pango parents & n & Other descendants \\
-Pango &  &  &  &  \\
+Pango & Sc2ts inferred parents & Official Pango parents & $n$ & Other descendants \\
 \midrule
@@
 }
-\end{table}
-
-\begin{table}
-\begin{tabular}{llll}
+\end{sidewaystable}
+
+\begin{sidewaystable}
+\begin{tabular}{llp{1.5cm}llrl}
 \toprule
- interval &                 causal & parent &                                descendant\_lineages \\
+Causal date & Breakpoint & Causal lineage & Causal strain & Sc2ts inferred parents & $n_{XE}$ & Other descendants \\
 \midrule
- (9867, 10198) &             XE  & [BA.1, BA.2] &                                          \{'XE': 3\} \\
-(10448, 11537) &             XH  & [BA.1.20, BA.2.9] &  \{'XAF': 34, 'XH': 11, 'B.1.1.529': 3, 'XE': 3,... \\
-(10448, 11537) &             XE  & [BA.1.1.15, Unknown (R)] & \{'XE': 2\} \\
-(10448, 11537) &             XE  & [BA.1.17.2, BA.2] &                                        \{'XE': 155\} \\
+2022-01-06 & \{10448-11537\} & XH & Denmark/DCGC-323778/2022 & BA.1.20 + BA.2.9 & 3 & XAF: 34/35, XH: 11/11, B.1.1.529: 2 \\
+2022-02-11 & {10448-11537} & XE & England/MILK-3674F1A/2022 & BA.1.17.2 + BA.2 & 155 &  \\
+2022-02-17 & \{17411-19955\} & XE & Turkey/HSGM-G2220/2022 & Unknown + BA.2 & 1 & B.1.1.529: 1 \\
+2022-04-09 & \{9867-10198\} & XE & Belgium/ULG-25782/2022 & BA.1 + BA.2 & 3 &  \\
+2022-04-20 & \{10448-11537\} & XE & Canada/BC-BCCDC-424071/2022 & BA.1.1.15 + Unknown & 2 &  \\
 \bottomrule
@@
 \caption{\label{tab:multiple_origins_table}
-Details of inferred
-breakpoints for different recombination events inferred for XE samples
-(Official: BA.1* \& BA.2, 10448–11287). TODO ELABORATE}
-\end{table}
+Summary of inferred multiple ``origins'' for Nextclade-designated XE samples
+(Official: BA.1* \& BA.2, 10448–11287, defined from UK and Irish strains, earliest
+on 2022/1/16;  see \url{https://github.com/cov-lineages/pango-designation/issues/454}).
+XE samples are grouped by their most recent ancestral recombination node;
+Details are also provided for the ``causal strain'' that led to the recombination
+being proposed by \texttt{sc2ts}.
+}
+\end{sidewaystable}
 
@@
 to the root, it is possible for a mutation to be a ``reversion'' to the initial state of the parent mutation:
-such reversions are indicated by lowercase nucleotide letters. Sample nodes are shown with \emph{tskit} IDs,
+such reversions are indicated by lowercase nucleotide letters. Sample nodes are shown with \texttt{tskit} IDs,
 which can be mapped to GISAID EPI ISL identifiers and strain names using supplementary file \protect\path{Pango_X_graph_sample_maps.txt}. In contrast to
@@
 to map tskID to EPI\_ISL and strain name).
-Note that GISAID does not designate any nodes as XD in the long ARG, hence no recmbninant Pango designations
+Note that GISAID does not designate any nodes as XD in the long ARG, hence no recombinant Pango lineages
 are marked in this plot. From inspection of the samples, we believe the GISAID designations to be erroneous

### 2023-05-17 paper.tex
Yan's updates

@@
 A similar pattern is seen for several of the origins inferred for the Pango
-XE lineage, which was originally defined on the basis of sequences from the UK and Ireland. The
+XE lineage (originally defined on the basis of sequences from the UK and Ireland). The
 Long ARG infers five independent origins for nonsingleton Nextclade-designated XE
@@
 (primarily UK and Irish, but with a few northern European and north American strains).
-This certainly represents the canonical XE recombinant lineage. The other recombination nodes
+This certainly represents the XE recombinant lineage as originally defined. The other recombination nodes
 only give rise for one, two, or three XE samples. Three of these recombination
@@
 none of their XE-labelled descendant samples are from the UK or Ireland.
-However, the most recently derived XE cluster, consisting
-of two samples (one from Canada and one from England) may not be due to mislabelling: not
-only does it contain an English sample but it has exactly the same breakpoint interval as
-the canonical XE clade. It is possible that instead, this is a case where the \texttt{sc2ts} algorithm
+It is less clear if the most recent XE cluster is a mislabelling or a case of \texttt{sc2ts}
+inferring multiple recombinations when there should be just one. This cluster of 2 samples
+(one of which is English) has exactly the same breakpoint estimates as the
+main cluster. If the only samples available are rather genetically distant from an originating
+recombinant, we might expect sc2ts to infer multiple similar recombination events
+(rather than a single event with extra reversions)
+
+a recombinant outbreak
 has taken a "true" XE sample and mistakenly inferred an additional, spurious origin. [WHY? - this
@@
 As well as cases where single recombination events explain an X lineage,
-such as the examples in Figure~\ref{fig:pango-simple-origin-graph}, many of the
-Pango X lineages involve more complex origins,
-in which multiple non-independent (``nested'') recombination events are involved.
-In these cases, the descendants of the original recombination
-frequently include samples which have not been previously identified as recombinants.
-An example is XB in the long ARG,
+such as the examples in Figure~\ref{fig:pango-simple-origin-graph}, four of the
+Pango X lineages (XB, XE, XAB, and XAJ) have Nextclade-designated samples which
+are related in a more complex manner, where multiple non-independent (``nested'')
+recombination events occur. An example is XB in the long ARG,
 plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
 between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
-``hairball'' of further recombination nodes whose descendants are mostly not
+``hairball'' of further recombination nodes whose descendants are often not
 identified as recombinants by Nextclade (plotted in blue). A similar
 pattern is seen when examining XB in the wide ARG (discussion above, plot not shown).
+Similarly, XE and XAJ
 Note that the descendants of nested recombination events include only one
@@
 indicating some uncertainty in lineage assignment in this part of the ARG.
-Investigating examples of nested recombinant origins, and identifying which
-(if any) of the nested recombination events may be artefactual, is an important
-area of future research.
+Also note that the number of mutations on the lineages immediately above and
+below the recombination node (totalling 18+2) is rather large, suggesting
+that the sampled recombinant which induced the recombination node in the Long ARG
+is only distantly related to the true originating recombinant. This could account
+for complex and potentially artefactual relationships around these nodes, and
+is likely to be due to undersampling of the XB outbreak. Investigating examples
+of nested recombinant origins, and identifying which (if any) of the nested
+recombination events may be artefactual, is an important area of future research.
 
@@
 \centering
-\begin{tabular}{lllrl}
+\begin{tabular}{p{1cm}p{1.2cm}p{4.2cm}lll}
 \toprule
-Pango & Sc2ts inferred parents & Official Pango parents & $n$ & Other descendants \\
+Focal Pango  & Num origins & Num focal samples (further split by origin, *=nested) & Official Pango parents & \textbf{Main clade}: sc2ts parents & \textbf{Main clade}: other descendants \\
 \midrule
-\bfseries XA & B.1.1.7 + B.1.177.18 & B.1.1.7 + B.1.177 & 5 & – \\
-\bfseries XAA & BA.1 + BA.2.9 & BA.1* + BA.2* & 8 & XAB:~38/48, XAG:~17/17, XU:~1/3, XQ:~2/12 \\
-\bfseries XAC & BA.1.17.2 + BA.2.3 & BA.1* + BA.2* & 27 & – \\
-\bfseries XAE & BA.1 + BA.2 & BA.1* + BA.2* & 9 & – \\
-\bfseries XAG & BA.1 + BA.2.9 & BA.1* + BA.2* & 17 & XAB:~38/48, XAA:~8/8, XU:~1/3, XQ:~2/12 \\
-\bfseries XF & AY.4 + BA.1 & B.1.617.2* + BA.1* & 2 & – \\
-\bfseries XG & BA.1.17 + BA.2 & BA.1* + BA.2* & 32 & XAB:~1/48 \\
-\bfseries XH & BA.1.20 + BA.2.9 & BA.1* + BA.2* & 11 & XAF:~34/35, B.1.1.529:~2, XE:~3/163 \\
-\bfseries XK & BA.1.1.1 + BA.2 & BA.1* + BA.2* & 3 & – \\
-\bfseries XL & BA.1.17.2 + BA.2 & BA.1* + BA.2* & 10 & XAB:~1/48, XU:~1/3 \\
-\bfseries XR & BA.1.1 + BA.2 & BA.1.1* + BA.2* & 8 & XQ:~9/12, XAB:~2/48 \\
-\bfseries XS & AY.36 + BA.1.1 & B.1.617.2* + BA.1.1* & 4 & – \\
-\bfseries XT & BA.2.23 + Unknown & BA.1* + BA.2* & 1 & BA.2.23:~1 \\
-\bfseries XV & BA.1.1 + BA.2 + Unknown & BA.1* + BA.2* & 1 & BA.2:~8 \\
-\bfseries XW & BA.1.1.15 + BA.2 & BA.1* + BA.2* & 11 & XN:~2/13 \\
-\bfseries XY & BA.1.1 + BA.2 & BA.1* + BA.2* & 6 & XAF:~1/35 \\
+\bfseries XA & 1 & \textbf{5} & B.1.1.7 + B.1.177 & B.1.1.7 + B.1.177.18 &  \\
+\bfseries XF & 1 & \textbf{2} & B.1.617.2* + BA.1* & AY.4 + BA.1 &  \\
+\bfseries XG & 1 & \textbf{32} & BA.1* + BA.2* & BA.1.17 + BA.2 & XAB: 1/48 \\
+\bfseries XH & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XAF: 34/35, B.1.1.529: 2, XE: 3/163 \\
+\bfseries XK & 1 & \textbf{3} & BA.1* + BA.2* & BA.1.1.1 + BA.2 &  \\
+\bfseries XL & 1 & \textbf{10} & BA.1* + BA.2* & BA.1.17.2 + BA.2 & XAB: 1/48, XU: 1/3 \\
+\bfseries XR & 1 & \textbf{8} & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XQ: 9/12, XAB: 2/48 \\
+\bfseries XS & 1 & \textbf{4} & B.1.617.2* + BA.1.1* & AY.36 + BA.1.1 &  \\
+\bfseries XT & 1 & \textbf{1} & BA.1* + BA.2* & BA.2.23 + Unknown & BA.2.23: 1 \\
+\bfseries XV & 1 & \textbf{1} & BA.1* + BA.2* & BA.1.1 + BA.2 + Unknown & BA.2: 8 \\
+\bfseries XW & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.1.15 + BA.2 & XN: 2/13 \\
+\bfseries XY & 1 & \textbf{6} & BA.1* + BA.2* & BA.1.1 + BA.2 & XAF: 1/35 \\
+\bfseries XAA & 1 & \textbf{8} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
+\bfseries XAC & 1 & \textbf{27} & BA.1* + BA.2* & BA.1.17.2 + BA.2.3 &  \\
+\bfseries XAE & 1 & \textbf{9} & BA.1* + BA.2* & BA.1 + BA.2 &  \\
+\bfseries XAG & 1 & \textbf{17} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAA: 8/8, XU: 1/3, XQ: 2/12 \\
+\bfseries XB & 2 & 57 (\textbf{57}, 1)* & B.1.631 + B.1.634 & B.1 + B.1.627 & B.1.634: 3, B.1.631: 7 \\
+\bfseries XC & 2 & 3 (1, \textbf{2}) & AY.29 + B.1.1.7 & AY.103 + B.1.1.7 &  \\
+\bfseries XD & 2 & 4 (1, \textbf{3}) & B.1.617.2* + BA.1* & AY.4 + BA.1.15 &  \\
+\bfseries XJ & 2 & 2 (1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
+\bfseries XAD & 2 & 5 (1, \textbf{4}) & BA.1* + BA.2* & BA.1.1 + BA.2 &  \\
+\bfseries XAF & 2 & 35 (\textbf{34}, 1) & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XH: 11/11, B.1.1.529: 2, XE: 3/163 \\
+\bfseries XAH & 2 & 12 (\textbf{10}, 2) & BA.1* + BA.2* & BA.1 + BA.2.10 & XZ: 61/66, XAD: 1/5 \\
+\bfseries XQ & 3 & 12 (2, \textbf{9}, 1) & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XR: 8/8, XAB: 2/48 \\
+\bfseries XU & 3 & 3 (1, 1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
+\bfseries XM & 4 & 47 (2, 1, \textbf{40}, 4) & BA.1.1* + BA.2* & BA.1.1 + BA.2 &  \\
+\bfseries XN & 4 & 13 (2, 2, \textbf{7}, 2) & BA.1* + BA.2* & BA.1 + BA.2 &  \\
+\bfseries XAJ & 4 & 5 (1, 1, \textbf{3}, 1)* & BA.2.12.1* + BA.4* & Unknown & BA.5: 1 \\
+\bfseries XE & 5 & 163 (3, \textbf{155}, 1, 3, 2)* & BA.1* + BA.2* & BA.1.17.2 + BA.2 &  \\
+\bfseries XZ & 5 & 66 (1, 1, \textbf{61}, 2, 1) & BA.1* + BA.2* & BA.1 + BA.2.10 & XAH: 10/12, XAD: 1/5 \\
+\bfseries XAB & 8 & 48 (1, \textbf{38}, 1, 2, 2, 2, 18, 2)* & BA.1* + BA.2* & BA.1 + BA.2.9 & XAA: 8/8, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
 \bottomrule
@@
 \caption{\label{tab:pango-single-origin}
-Summary of the Pango X-lineages in the long ARG that are inferred to have a
-single recombinant origin based on the NextClade Pango lineage assignments.
-Sc2ts inferred parents are based on imputed NextClade Pango lineages. Official parents
-are taken from Pango designation alias key (TODO: explain the asterisks). The final column
-list other inferred descendants of the originating recombination event, summarised
-by NextClade Pango designation and sample count / total ARG count (total count in the ARG
-is omitted for non-recombinant samples).
+Summary of all Pango X-lineages in the long ARG. In cases of multiple origins, most X-lineages have a single ``main clade'' (in bold). 
+Sc2ts inferred parents for the main clade are based on imputed NextClade Pango lineages. Official parents
+are taken from Pango designation alias key (TODO: explain the asterisks).
+Other inferred samples within the main clade are listed, grouped by NextClade Pango designation and clade count / total ARG count (the latter being omitted for non-recombinant designations).
 }
@@
 \begin{sidewaystable}
-\begin{tabular}{llp{1.5cm}llrl}
+\centering
+\begin{tabular}{llp{1.2cm}llrl}
 \toprule
-Causal date & Breakpoint & Causal lineage & Causal strain & Sc2ts inferred parents & $n_{XE}$ & Other descendants \\
+Causal date & Sc2ts inferred parents & Causal lineage & Causal strain & Breakpoint & $n_{XE}$ & Other descendants \\
 \midrule
-2022-01-06 & \{10448-11537\} & XH & Denmark/DCGC-323778/2022 & BA.1.20 + BA.2.9 & 3 & XAF: 34/35, XH: 11/11, B.1.1.529: 2 \\
-2022-02-11 & {10448-11537} & XE & England/MILK-3674F1A/2022 & BA.1.17.2 + BA.2 & 155 &  \\
-2022-02-17 & \{17411-19955\} & XE & Turkey/HSGM-G2220/2022 & Unknown + BA.2 & 1 & B.1.1.529: 1 \\
-2022-04-09 & \{9867-10198\} & XE & Belgium/ULG-25782/2022 & BA.1 + BA.2 & 3 &  \\
-2022-04-20 & \{10448-11537\} & XE & Canada/BC-BCCDC-424071/2022 & BA.1.1.15 + Unknown & 2 &  \\
+2022-01-06 & BA.1.20 + BA.2.9 & XH & Denmark/DCGC-323778/2022 & (10448,11538) & 3 & XAF: 34/35, XH: 11/11, B.1.1.529: 2 \\
+2022-02-11 & BA.1.17.2 + BA.2 & XE & England/MILK-3674F1A/2022 & (10448,11538) & 155 &  \\
+2022-02-17 & Unknown + BA.2 & XE & Turkey/HSGM-G2220/2022 & (17411,19956) & 1 & B.1.1.529: 1 \\
+2022-04-09 & BA.1 + BA.2 & XE & Belgium/ULG-25782/2022 & (9867,10199) & 3 &  \\
+2022-04-20 & BA.1.1.15 + Unknown & XE & Canada/BC-BCCDC-424071/2022 & (10448,11538) & 2 &  \\
 \bottomrule
-\end{tabular}
-\caption{\label{tab:multiple_origins_table}
+\end{tabular}\caption{\label{tab:multiple_origins_table}
 Summary of inferred multiple ``origins'' for Nextclade-designated XE samples
 (Official: BA.1* \& BA.2, 10448–11287, defined from UK and Irish strains, earliest
-on 2022/1/16;  see \url{https://github.com/cov-lineages/pango-designation/issues/454}).
+on 2022/1/16;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/454}).
 XE samples are grouped by their most recent ancestral recombination node;

### 2023-05-17 paper.tex
Update Jackson et al breakpoints for sc2ts

@@
 A (XA)       & 4   & Jackson        &  21,256--21,613 & B.1.177/B.1.1.7 \\
-             &     &\texttt{sc2ts} &  21,256--22,227 & B.1.177.18/B.1.1.7 \\
+             &     &\texttt{sc2ts} &  21,256--22,228 & B.1.177.18/B.1.1.7 \\
 \cmidrule{3-5}
-B            & 2   & Jackson        &  6,529--6,953 & B.1.36/B.1.1.7  \\
-             &     &\texttt{sc2ts} &  6,529--6,954 & B.1.36/B.1.1.7  \\
+B            & 2   & Jackson        &  6,529--6,954 & B.1.36/B.1.1.7  \\
+             &     &\texttt{sc2ts} &  6,529--6,955 & B.1.36/B.1.1.7  \\
 \cmidrule{3-5}
 C            & 3   &Jackson        &  25,997--27,441 &  B.1.1.7/B.1.221 \\
-             &     & \texttt{sc2ts} &  25,997--27,972 &  B.1.1.7/B.1.221 \\
+             &     & \texttt{sc2ts} &  25,997--27,973 &  B.1.1.7/B.1.221 \\
 \cmidrule{3-5}
 D            & 3   & Jackson        &  21,576--23,063 &  B.1.36.17/B.1.1.7 \\
-             &     & \texttt{sc2ts} &  22,445--23,063 &  B.1.36.39/B.1.1.7 \\
+             &     & \texttt{sc2ts} &  22,445--23,064 &  B.1.36.39/B.1.1.7 \\
 \bottomrule
@@
 The second column gives the number of sequences in the group.
-Intervals are given in left-inclusive, right-exclusive form.
+Intervals are half-closed, so that for an interval $\ell$--$r$
+the breakpoint $x$ can be $\ell \leq x < r$.
 Details for all 16 sequences are
@@
 previously identified as recombinants, i.e. given a Pango designation that
-starts with an ``X''. Since the wide ARG is restricted to data collected prior
+starts with an ``X''. Since the Wide ARG is restricted to data collected prior
 to mid-2021, it contains samples from only three Pango-designated recombinant
@@
 associated with recombinants, we therefore devote more analysis to patterns in
-the long ARG. As in previous sections, we exclude singleton recombinants where
+the Long ARG. As in previous sections, we exclude singleton recombinants where
 possible.
@@
 %% Quick overview of the wide ARG.
-In the wide ARG, 44
+In the Wide ARG, 44
 samples are designated as XA by both Nextclade and GISAID, 237 are designated
@@
 
-For XC, the wide ARG identifies more than one originating recombination node
+For XC, the Wide ARG identifies more than one originating recombination node
 (see "multiple origins" below). However, as none of the samples designated as
@@
 %% high level summary of the pango recombinant lineages and samples in the Long ARG
-The long ARG (subsampled to mid-2022) contains 33
-Nextclade-designated Pango recombinant lineages (XA to XAK; Supplementary Table
-XXX) comprising 749 samples (711 when singleton recombinants are removed). In
+The Long ARG (subsampled to mid-2022) contains 33
+Nextclade-designated Pango recombinant lineages comprising 749 samples (711 when singleton recombinants are removed). In
 contrast, GISAID designates 38 Pango recombinant lineages in the long ARG,
@@
 
-In total out of the 31 Pango recombinant lineages in the long ARG we identify 16 as
+In total out of the 31 Pango recombinant lineages in the Long ARG we identify 16 as
 having a single origin. Supplementary Table~\ref{tab:pango-single-origin} summarises these
@@
     & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
-21,256–22,227 \\
+21,256–22,228 \\
     ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
-    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
+    \thead{B.1.177.18 \\ Alpha} & 21,256–22,228 \\
 ALDP-130BB95 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
-    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
+    \thead{B.1.177.18 \\ Alpha} & 21,256–22,228 \\
 LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
-    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
+    \thead{B.1.177.18 \\ Alpha} & 21,256–22,228 \\
 QEUH-CCCB30 & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
-    \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
+    \thead{B.1.36 \\ Alpha} & 6,529–6,955 \\
 QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
-    \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
+    \thead{B.1.36 \\ Alpha} & 6,529–6,955 \\
 MILK-1166F52 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
-    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
+    \thead{Alpha \\ B.1.221} & 25,997–27,973 \\
 MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
-    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
+    \thead{Alpha \\ B.1.221} & 25,997–27,973 \\
 QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
-    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
+    \thead{Alpha \\ B.1.221} & 25,997–27,973 \\
 MILK-126FE1F & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
-    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
+    \thead{B.1.36.39 \\ Alpha} & 22,445–23,064 \\
 RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
-    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
+    \thead{B.1.36.39 \\ Alpha} & 22,445–23,064 \\
 RAND-128FA33 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
-    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
+    \thead{B.1.36.39 \\ Alpha} & 22,445–23,064 \\
 CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
-    \thead{B.1.177 \\ Alpha} & 16,177–21,255 \\
+    \thead{B.1.177 \\ Alpha} & 17,616--21,256 \\
 CAMC-CB7AB3 & n/a & \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} &
-     \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\
+     \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3268--5389\\17,616--21,256} \\
 MILK-103C712 & n/a & \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} &
@@
 QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 &
-    \thead{Alpha \\ B.1.177} & 7,729–10,870 \\ \hline
+    \thead{Alpha \\ B.1.177} & 7729--10871 \\ \hline
 \end{tabular}
@@
 \caption{\label{tab:pango-single-origin}
-Summary of all Pango X-lineages in the long ARG. In cases of multiple origins, most X-lineages have a single ``main clade'' (in bold). 
+Summary of the Pango X-lineages in the long ARG (excluding XP and XAK whose samples are entirely filtered out, see text).
+In cases of multiple origins, most X-lineages have a single ``main clade'' (in bold).
 Sc2ts inferred parents for the main clade are based on imputed NextClade Pango lineages. Official parents
@@
 
-\begin{sidewaystable}
+\begin{table}
+\small
 \centering
-\begin{tabular}{llp{1.2cm}llrl}
+\begin{tabular}{lp{0.2cm}p{0.8cm}lp{0.9cm}ll}
 \toprule
-Causal date & Sc2ts inferred parents & Causal lineage & Causal strain & Breakpoint & $n_{XE}$ & Other descendants \\
+Date & $n$ & Causal Pango & Causal strain & Causal \#muts & Sc2ts parents & Breakpoint \\
 \midrule
-2022-01-06 & BA.1.20 + BA.2.9 & XH & Denmark/DCGC-323778/2022 & (10448,11538) & 3 & XAF: 34/35, XH: 11/11, B.1.1.529: 2 \\
-2022-02-11 & BA.1.17.2 + BA.2 & XE & England/MILK-3674F1A/2022 & (10448,11538) & 155 &  \\
-2022-02-17 & Unknown + BA.2 & XE & Turkey/HSGM-G2220/2022 & (17411,19956) & 1 & B.1.1.529: 1 \\
-2022-04-09 & BA.1 + BA.2 & XE & Belgium/ULG-25782/2022 & (9867,10199) & 3 &  \\
-2022-04-20 & BA.1.1.15 + Unknown & XE & Canada/BC-BCCDC-424071/2022 & (10448,11538) & 2 &  \\
+2022-03-24 & 2 & XN & Germany/BY-RKI-I-674341/2022 & 3 & Unknown + BA.2 & (2833,9867) \\
+2022-03-29 & 2 & XN & Spain/AB-HGUA-26086757/2022 & 2 & BA.1.1 + BA.2.9 & (2833,4185) \\
+2022-04-23 & 7 & XN & Spain/MD-HRYC-12296321/2022 & 2 & BA.1 + BA.2 & (4185,4322) \\
+2022-05-17 & 2 & XN & USA/TX-HMH-MCoV-97903/2022 & 5 & BA.2 + BA.2 & (22776,22787) \\
 \bottomrule
-\end{tabular}\caption{\label{tab:multiple_origins_table}
-Summary of inferred multiple ``origins'' for Nextclade-designated XE samples
-(Official: BA.1* \& BA.2, 10448–11287, defined from UK and Irish strains, earliest
-on 2022/1/16;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/454}).
-XE samples are grouped by their most recent ancestral recombination node;
+\end{tabular}
+\caption{\label{tab:multiple_origins_table}
+Summary of inferred multiple ``origins'' for Nextclade-designated XN samples after filtering
+(Official: BA.1* \& BA.2*, breakpoint between 2834 \& 4183, defined from UK strains, earliest
+on 2022-02-01;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/480}).
+XN samples are grouped into rows based on their most recent ancestral recombination node;
 Details are also provided for the ``causal strain'' that led to the recombination
 being proposed by \texttt{sc2ts}.
+Note that a large number of XN samples were previously removed by filtering.
+When undersampling causes three or more mutations to lie immediately above the causal strain, \texttt{sc2ts} may mistakenly propose multiple independent origins rather than one.
 }
-\end{sidewaystable}
+\end{table}
 

### 2023-05-17 paper.tex
Add data for Long ARG recomb intervals

--- 2023-05-17 paper.tex
@@
 are similar there is no way to distinguish them. Here we define the
-interval for a particular breakpoint as the interval over which the
+interval for a particular breakpoint as the genome coordinates over which the
 sequences for the left and right parent nodes are identical. The right-hand
-extreme of the interval is chosen by the LS HMM Viterbi algorithm
-(Section~\ref{sec:ls}, and the left bound can then be derived by
+extreme of the breakpoint interval is chosen by the LS HMM Viterbi algorithm
+(Section~\ref{sec:ls}) and the left bound can then be derived by
 sequence comparison of the parents.

### 2023-05-17 paper.tex
Add long arg recomb intervals figure

@@
 extreme of the breakpoint interval is chosen by the LS HMM Viterbi algorithm
-(Section~\ref{sec:ls}) and the left bound can then be derived by
+(Section~\ref{sec:ls}) and the left bound is then derived by
 sequence comparison of the parents.
@@
 \centering
-\includegraphics[width=\textwidth]{figures/recombination_intervals.pdf}
+\includegraphics[width=\textwidth]{figures/wide_arg_recombination_intervals.pdf}
 \caption{\label{fig:breakpoint-distribution}
-Distribution of recombination breakpoints and mutations along the genome.}
+Distribution of recombination breakpoints and mutations along the genome in
+the Wide ARG.
+See Figure~\ref{fig:long_arg_breakpoint_distribution} for the equivalent plot
+for the Long ARG.}
 \end{figure}
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/long_arg_recombination_intervals.pdf}
+\caption{\label{fig:long_arg_breakpoint_distribution}
+Distribution of recombination breakpoints and mutations along the genome in
+the Long ARG.}
+\end{figure}
+
 \begin{figure} \centering

### 2023-05-17 plot.py

@@ -1758,8 +1758,6 @@ class MutationalSpectra(Figure):
 
 
 class RecombinationIntervals(Figure):
-    name = "recombination_intervals"
-
     def plot_breakpoints(
         self, df, df_sites, mutations_threshold=0.9, unique_only=False
     ):
@@ -1850,12 +1848,30 @@ class RecombinationIntervals(Figure):
             axs.set_xticklabels(["" for _ in genes.keys()], rotation="vertical")
         axs.set_xticklabels(list(genes.keys()), rotation="vertical")
 
+
+class WideRecombinationIntervals(RecombinationIntervals):
+    name = "wide_arg_recombination_intervals"
+
     def plot(self, args):
-        df_sites = pd.read_csv("data/site_info.csv")
+        df_sites = pd.read_csv("data/wide_arg_site_info.csv")
         df_recombs = pd.read_csv("data/wide_arg_recombinants.csv")
         df_recombs = df_recombs[df_recombs.max_descendant_samples > 1]
         self.plot_breakpoints(
             df_recombs, df_sites, mutations_threshold=0.25, unique_only=False
+        )
+        prefix = os.path.join("figures", self.name)
+        plt.savefig(prefix + f".{args.outtype}", bbox_inches="tight")
+
+
+class LongRecombinationIntervals(RecombinationIntervals):
+    name = "long_arg_recombination_intervals"
+
+    def plot(self, args):
+        df_sites = pd.read_csv("data/long_arg_site_info.csv")
+        df_recombs = pd.read_csv("data/long_arg_recombinants.csv")
+        df_recombs = df_recombs[df_recombs.max_descendant_samples > 1]
+        self.plot_breakpoints(
+            df_recombs, df_sites, mutations_threshold=0.40, unique_only=False
         )
         prefix = os.path.join("figures", self.name)
         plt.savefig(prefix + f".{args.outtype}", bbox_inches="tight")

### 2023-05-17 paper.tex
Update plots, and add description for main text

@@
 
-\subsection{Distribution of recombination breakpoint intervals}
+\subsection{Recombination breakpoint intervals}
 \label{sec:breakpoint_intervals}
@@
 Distribution of recombination breakpoints and mutations along the genome in
-the Wide ARG.
+the Wide ARG. Top panel shows the intervals for 1769 breakpoints associated
+with 1522 recombination nodes $>1$ descending samples plotted along the genome
+as line segments, coloured by interval width. The inset histogram shows the
+distribution of these interval widths.
+The bottom panel shows the number of intervals that intersect with
+each site along the genome (left axis, orange), and the number of mutations
+per site (right axis, blue).
 See Figure~\ref{fig:long_arg_breakpoint_distribution} for the equivalent plot

### 2023-05-17 plot.py

@@ -1787,8 +1787,10 @@ class RecombinationIntervals(Figure):
         ax1.set_yticks([])
         axins1 = inset_axes(
             ax1,
-            width="1%",
-            height="50%",
+            width=0.1,
+            height=1.5,
+            bbox_to_anchor=(0.01, 1.0),
+            bbox_transform=ax1.transAxes,
             loc="upper left",
         )
         fig.colorbar(
@@ -1797,11 +1799,18 @@ class RecombinationIntervals(Figure):
 
         axins2 = inset_axes(
             ax1,
-            width="20%",
-            height="50%",
-            loc="upper center",
+            width=2.0,
+            height=1.5,
+            bbox_to_anchor=(0.15, 1.0),
+            bbox_transform=ax1.transAxes,
+            loc="upper left",
         )
         axins2.hist(length)
+        axins2.set_xlabel("Width of interval")
+        axins2.set_ylabel("Count")
+
+        print("Unique recombs = ", len(df.node.unique()))
+        print(length.describe())
 
         covers = np.zeros(df_sites.position.max())
         for left, right in intervals:
@@ -1816,7 +1825,7 @@ class RecombinationIntervals(Figure):
         pos = df_sites.position
         color = "tab:blue"
         ax3 = ax2.twinx()
-        ax3.plot(pos, count, color=color)
+        ax3.plot(pos, count, color=color, alpha=0.7)
         ax3.tick_params(axis="y", labelcolor=color)
         ax3.set_ylabel("Mutations per site", color=color)
 

### 2023-05-17 paper.tex
Some updates on intervals code.

@@
 
-% - Distribution of breakpoint intervals
-%     Study the breakpoint intervals in the Wide ARG. Because we are interested
-% in the intervals, we want to be sure that these are robustly reported by the
-% HMM. Hence, we focus on those that have two parents in the mirrored and
-% unmirrored HMM runs, the same set of mutations, and the left and right parents
-% are consistently imputed for Pango lineage status.
-%     We focus on the Wide ARG here because it covers (roughly?) the same time
-% period as the RIPPLES analysis.
-
-We explore recombination signals across the SARS-CoV-2 genome by inspecting the
-breakpoint intervals of the filtered recombination nodes from the wide ARG
-(sampled to mid-2021) in Figure~\ref{fig:breakpoint-distribution}. The length
-of a breakpoint interval reflects uncertainty in the estimated location of a
-breakpoint. On average, the length of the breakpoint intervals is 2,895 bases
-(median, 1,670 bases; range, 1 to 23,666 bases). This broad distribution of
-interval length indicates that the location of a breakpoint often could not be
-precisely determined given the observed genetic variation.
-
 \begin{figure}
@@
 Distribution of recombination breakpoints and mutations along the genome in
-the Wide ARG. Top panel shows the intervals for 1769 breakpoints associated
-with 1522 recombination nodes $>1$ descending samples plotted along the genome
-as line segments, coloured by interval width. The inset histogram shows the
+the Wide ARG. Top panel shows the intervals for 1,769 breakpoints associated
+with 1,522 recombination nodes with at least two descending samples, plotted along the genome
+as line segments (coloured by interval width). The inset histogram shows the
 distribution of these interval widths.
@@
 per site (right axis, blue).
+The top-ten sites by mutation count are annotated.
 See Figure~\ref{fig:long_arg_breakpoint_distribution} for the equivalent plot
@@
 
-There is a preponderance of breakpoint intervals spanning the 3’ end of the
-genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}. Inspecting
-more closely, we find an enrichment in the number of intersecting intervals in
-the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure
-4B). Markedly, there is an abrupt drop in the number of intersecting intervals
-in the N gene following the peak in ORF8 towards the 3’ end.
+Figure~\ref{fig:breakpoint-distribution} shows the distribution of breakpoint
+intervals and patterns of recurrent mutation along the genome in the Wide ARG
+(see Figure~\ref{fig:long_arg_breakpoint_distribution} for the same information
+of the Long ARG). We focus on the Wide ARG here as it covers roughly the same time
+period as the analyses of \cite{Turakhia2022-it}, facilitating comparisons
+of the results. To reduce the effect of artefactual recombinants, we
+consider only the breakpoints associated with the 1,522 recombination nodes
+that are ancestral to more than one sample. The mean length of these intervals
+is 1,685 bases (median 962); the length distribution is summarised in the
+inset histogram in Figure~\ref{fig:breakpoint-distribution}.
+
+Figure~\ref{fig:breakpoint-distribution} [ TODO what can we really say here?
+Other than the boundaries happen to lie on gene breaks, not an awful lot I
+think?]
+
+% This isn't really true any more, is it?
+% There is a preponderance of breakpoint intervals spanning the 3’ end of the
+% genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}. Inspecting
+% more closely, we find an enrichment in the number of intersecting intervals in
+% the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure
+% 4B). Markedly, there is an abrupt drop in the number of intersecting intervals
+% in the N gene following the peak in ORF8 towards the 3’ end.
 
@@
 
-The remaining remaining 630 Pango recombinant samples, representing 31
-Nextclade-designated Pango lineages, trace back to 50 different most recent
-recombination nodes.
-We examine the inferred inheritance structure in the following subsections.
+The remaining remaining 630 Pango recombinant samples trace back to 50 different
+most recent recombination nodes, representing 31
+Nextclade-designated Pango lineages. These are summarised
+in Supplementary Table~\ref{tab:pango-recombinants}, and fall
+into three classes: single origin, multiple origin,
+and multiple nested origins.
 
@@
 trace back to one recombination node, representing a single originating
-recombination event.
+recombination event. In the Long ARG, 16 of the 31 Pango recombinants lineages identified by Nextclade
+ fall into this category (Supplementary Table~\ref{tab:pango-recombinants})
 
@@
 Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
-imputed Pango status. Genomic regions inherited by the recombination node use
-breakpoint positions in the ARG, corresponding to the maximum inferred breakpoint position
+imputed Pango status. Genomic regions inherited by the recombination node are shown
+corresponding to the maximum inferred breakpoint position from the HMM process.
 (B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
@@
 
-A simple example is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
-This shows a subgraph of the long ARG, in which paths have been traced from all
+One of the simplest examples is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
+Here, a subgraph of the long ARG has been plotted, where paths have been traced from all
 Nextclade-identified XA samples (in red) to the closest other sample nodes in our dataset.
@@
 to the rest of the ARG. Above recombination nodes, only ancestral nodes are shown, meaning
-that the subgraph is not extended to show other descendants of recombinant parents. It is clear
-from the subgraph that all the XA
-samples trace to a single originating recombination node,
+that the subgraph is not extended to show additional descendants of recombinant parents.
+
+It is clear from the XA subgraph that all the samples labelled XA by Nextclade trace to a single originating recombination node,
 whose genome is a composite of a B.1.177.18 lineage on the left of the genome (up to a maximum
@@
 Figure~\ref{fig:pango-simple-origin-graph}B. Here, the XAG samples
-all trace back to the same most recent common recombination node
+all trace back to the same most recent recombination node
 (combining BA.1 on the left and BA.2.9 on the right), but
@@
 
-In total out of the 31 Pango recombinant lineages in the Long ARG we identify 16 as
-having a single origin. Supplementary Table~\ref{tab:pango-single-origin} summarises these
-together with their inferred parents, showing good agreement between \texttt{sc2ts} and
-the official Pango designated parents, with \texttt{sc2ts} providing more a precise
-parent designation. Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
-simple-origin (XA) type, whereas the rest are of the XAG type, and include multiple Pango
-designated lineages as descendants of the originating recombination event. In some cases,
-these may be, however, be a result of erroneous Pango designations.
-
-\subsubsection{Multiple origins}
-The ARG inference has no pre-defined knowledge
+Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
+simple-origin (XA) type with no other Pango designations among their descendants. 
+The remaining 10 are of the XAG type with multiple Pango
+designated lineages as additional descendants of the originating recombination event. In some cases,
+these may be, however, be a result of erroneous Pango designations. Supplementary Table~\ref{tab:pango-recombinants} also shows the official Pango designated parent lineages for
+each recombinant lineage. The parent lineages in the Long ARG agree to a great extent with these, although \texttt{sc2ts} provides more a precise
+parent designation.
+
+\subsubsection{Independent multiple origins}
+The \texttt{sc2ts} inference process has no pre-defined knowledge
 of Pango X lineage assignments, and there is therefore no particular
 requirement that all the samples assigned to a given lineage must trace back to
-a single recombination event. In this section we examine several Pango X lineage
-samples that are inferred to have multiple recombinant origins (i.e. trace
-back to independent recombination nodes in the ARG).
+a single recombination event. 15 Pango X lineages are inferred to have multiple
+recombinant origins, such that their samples trace to >1 most recent
+recombination node in the ARG. Of these, 11 are cases where none of the
+recombinants are nested (i.e. there is no overlap between the descendant samples
+of each recombination node). In 9 of these 11, there is a single ``main''
+recombination event from which the majority of the corresponding recombinant samples
+descend (in other two, there are too few Pango X samples to decide on a ``main'' clade:
+see Supplementary Table~\ref{tab:pango-recombinants}).
 
@@
 A similar pattern is seen for several of the origins inferred for the Pango
-XE lineage (originally defined on the basis of sequences from the UK and Ireland). The
-Long ARG infers five independent origins for nonsingleton Nextclade-designated XE
+XN lineage (originally defined on the basis of sequences from the UK). The
+Long ARG infers four independent origins for nonsingleton Nextclade-designated XN
 samples (Supplementary Table~\ref{tab:multiple_origins_table}). One of the recombinant origins,
@@
 \subsubsection{Nested recombinant origins}
-As well as cases where single recombination events explain an X lineage,
-such as the examples in Figure~\ref{fig:pango-simple-origin-graph}, four of the
-Pango X lineages (XB, XE, XAB, and XAJ) have Nextclade-designated samples which
-are related in a more complex manner, where multiple non-independent (``nested'')
-recombination events occur. An example is XB in the long ARG,
-plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
-between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
+
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
+\caption{\label{fig:complex_origins_graph}  A subgraph of the Long ARG showing
+nested recombination events involving Pango lineage XB. All XB samples
+trace to a single recombination node (top centre), but three further recombinations
+also descend from this node. The samples descending from these nested
+recombinations include 9 that are assigned by Nextclade to
+various non-recombinant pre-Alpha lineages (blue).}
+\end{figure}
+
+As well as cases where Pango X lineage origins are attributed to independent
+recombination events, four Pango X lineages in the Long ARG have Nextclade-designated
+samples whose ancestry involves multiple non-independent (``nested'') recombination events
+(marked with asterisks in Supplementary Table~\ref{tab:pango-single-origin}; the most
+extensive is for XAB). The earliest example, present in both the Wide and Long ARGs, is XB
+The Logn ARG subgraph is plotted in Figure~\ref{fig:complex_origins_graph}, showing a recombination
+between a B.1 sample and B.1.627 sample that leads both to the XB samples and to a
 ``hairball'' of further recombination nodes whose descendants are often not
 identified as recombinants by Nextclade (plotted in blue). A similar
-pattern is seen when examining XB in the wide ARG (discussion above, plot not shown).
-Similarly, XE and XAJ
-Note that the descendants of nested recombination events include only one
+pattern is seen when examining XB in the Wide ARG (but as discussed previously,
+in the Wide ARG, XB exhibits multiple nested recombinations, some with many XB
+descendants).
+
+Note that in the Long ARG, the nested recombination events account for only one
 XB sample (pink upper left, with 7 mutations above it); moreover, this sample is
@@
 recombination events may be artefactual, is an important area of future research.
-
-\begin{figure} \centering
-\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
-\caption{\label{fig:complex_origins_graph}  A Pango lineage subgraph showing
-complex origins of the recombinant lineage XB. All XB samples in the long ARG
-trace to a single recombination node (top centre), but three further recombinations
-also descend from this node. The samples descending from these nested
-recombinations include 9 that are assigned by Nextclade to
-various non-recombinant pre-Alpha lineages (blue).}
-\end{figure}
 
@@
 Distribution of recombination breakpoints and mutations along the genome in
-the Long ARG.}
+the Long ARG.
+Top panel shows the intervals for 851 breakpoints associated
+with 763 recombination nodes with at least two descending samples, plotted along the genome
+as line segments (coloured by interval width).
+Other details as described in Figure~\ref{fig:breakpoint-distribution}.}
 \end{figure}
@@
 \includegraphics[width=\textwidth]{figures/supp_recombination_node_mrcas.pdf}
-\caption{\label{fig:recomb_mrcas_voc_breakdown}  Divergence between parent lineages for recombination events within and among different VoC categories.
-HMM-consistent recombinants are shown in green, HMM-inconsistent in blue (including many of the named Delta-Omicron recombinants).
-We find 69 HMM consistent (77 inconsistent) Alpha-Alpha recombination breakpoints = 13\% (10\%) of the total; 84 HMM consistent (143 inconsistent) Delta-Delta recombination breakpoints = 18\% (18\%) of the total; 99 HMM consistent (155 inconsistent) Omicron-Omicron recombination breakpoints = 22\% (19\%) of the total; and 2 HMM consistent (20 inconsistent) Delta-Omicron recombination breakpoints = 2\% (< 1\%) of the total.
-These plots do not include recombination breakpoints involved lineages that were not classified into these VoC categories}
+\caption{\label{fig:recomb_mrcas_voc_breakdown}  Divergence between parent
+lineages for recombination events within and among different VoC categories.
+HMM-consistent recombinants are shown in green, HMM-inconsistent in blue
+(including many of the named Delta-Omicron recombinants). We find 69 HMM
+consistent (77 inconsistent) Alpha-Alpha recombination breakpoints = 13\%
+(10\%) of the total; 84 HMM consistent (143 inconsistent) Delta-Delta
+recombination breakpoints = 18\% (18\%) of the total; 99 HMM consistent (155
+inconsistent) Omicron-Omicron recombination breakpoints = 22\% (19\%) of the
+total; and 2 HMM consistent (20 inconsistent) Delta-Omicron recombination
+breakpoints = 2\% (< 1\%) of the total. These plots do not include
+recombination breakpoints involved lineages that were not classified into these
+VoC categories}
 \end{figure}
@@
 \toprule
-Focal Pango  & Num origins & Num focal samples (further split by origin, *=nested) & Official Pango parents & \textbf{Main clade}: sc2ts parents & \textbf{Main clade}: other descendants \\
+Focal Pango  & Num origins & Num focal samples (further split by origin, *=nested) & Official Pango parents & \textbf{Main clade}: sc2ts parents & \textbf{Main clade}: additional descendants \\
 \midrule
@@
 \end{tabular}
-\caption{\label{tab:pango-single-origin}
+\caption{\label{tab:pango-recombinants}
 Summary of the Pango X-lineages in the long ARG (excluding XP and XAK whose samples are entirely filtered out, see text).

### 2023-05-17 plot.py

@@ -1758,9 +1758,7 @@ class MutationalSpectra(Figure):
 
 
 class RecombinationIntervals(Figure):
-    def plot_breakpoints(
-        self, df, df_sites, mutations_threshold=0.9, unique_only=False
-    ):
+    def plot_breakpoints(self, df, df_sites, unique_only=False):
         dfs = df.sort_values(["breakpoint_interval_left", "breakpoint_interval_right"])
         length = dfs.breakpoint_interval_right - dfs.breakpoint_interval_left
         intervals = [
@@ -1809,7 +1807,9 @@ class RecombinationIntervals(Figure):
         axins2.set_xlabel("Width of interval")
         axins2.set_ylabel("Count")
 
-        print("Unique recombs = ", len(df.node.unique()))
+        print("Recombination nodes = ", len(df.node.unique()))
+        print("Num intervals:", len(intervals))
+        print("Unique intervals:", len(set(intervals)))
         print(length.describe())
 
         covers = np.zeros(df_sites.position.max())
@@ -1831,8 +1831,8 @@ class RecombinationIntervals(Figure):
 
         ax2.set_xlabel("Genome position")
 
-        threshold = np.max(count) * mutations_threshold
-        top_sites = np.where(count > threshold)[0]
+        top_count = 10
+        top_sites = np.argsort(count)[-top_count:]
         for site in top_sites:
             ax3.annotate(
                 f"{int(pos[site])}", xy=(pos[site], count[site]), xycoords="data"
@@ -1865,9 +1865,7 @@ class WideRecombinationIntervals(RecombinationIntervals):
         df_sites = pd.read_csv("data/wide_arg_site_info.csv")
         df_recombs = pd.read_csv("data/wide_arg_recombinants.csv")
         df_recombs = df_recombs[df_recombs.max_descendant_samples > 1]
-        self.plot_breakpoints(
-            df_recombs, df_sites, mutations_threshold=0.25, unique_only=False
-        )
+        self.plot_breakpoints(df_recombs, df_sites, unique_only=False)
         prefix = os.path.join("figures", self.name)
         plt.savefig(prefix + f".{args.outtype}", bbox_inches="tight")
 
@@ -1879,9 +1877,7 @@ class LongRecombinationIntervals(RecombinationIntervals):
         df_sites = pd.read_csv("data/long_arg_site_info.csv")
         df_recombs = pd.read_csv("data/long_arg_recombinants.csv")
[…]

### 2023-05-18 paper.tex
Yans updates

--- 2023-05-17 paper.tex
@@
 per site (right axis, blue).
+TODO describe mutation threshold.
 See Figure~\ref{fig:long_arg_breakpoint_distribution} for the equivalent plot
@@
 
-In total out of the 31 Pango recombinant lineages in the Long ARG we identify 16 as
-having a single origin. Supplementary Table~\ref{tab:pango-single-origin} summarises these
-together with their inferred parents, showing good agreement between \texttt{sc2ts} and
-the official Pango designated parents, with \texttt{sc2ts} providing more a precise
-parent designation. Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
-simple-origin (XA) type, whereas the rest are of the XAG type, and include multiple Pango
-designated lineages as descendants of the originating recombination event. In some cases,
-these may be, however, be a result of erroneous Pango designations.
-
-\subsubsection{Multiple origins}
+Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
+simple-origin (XA) type with no other Pango designations among their descendants. 
+The remaining 10 are of the XAG type with multiple Pango
+designated lineages as additional descendants of the originating recombination event. In some cases,
+these may be, however, be a result of erroneous Pango designations. Supplementary Table~\ref{tab:pango-recombinants} also shows the official Pango designated parent lineages for
+each recombinant lineage. The parent lineages in the Long ARG agree to a great extent with these, although \texttt{sc2ts} provides more a precise
+parent designation.
+
+\subsubsection{Independent multiple origins}
 The ARG inference has no pre-defined knowledge
@@
 requirement that all the samples assigned to a given lineage must trace back to
-a single recombination event. In this section we examine several Pango X lineage
-samples that are inferred to have multiple recombinant origins (i.e. trace
-back to independent recombination nodes in the ARG).
+a single recombination event. 15 Pango X lineages are inferred to have multiple
+recombinant origins, such that their samples trace to more than one most recent
+recombination node in the ARG. Of these, 11 are cases where none of the
+recombinants are nested (i.e. there is no overlap between the descendant samples
+of each recombination node). In 9 of these 11, there is a single ``main''
+recombination event from which the majority of the corresponding recombinant samples
+descend. In other two, there are too few Pango X samples to decide on a ``main'' clade.
 

### 2023-05-18 paper.tex
Yans updates

--- 2023-05-17 paper.tex
@@
 
+% - Distribution of breakpoint intervals
+%     Study the breakpoint intervals in the Wide ARG. Because we are interested
+% in the intervals, we want to be sure that these are robustly reported by the
+% HMM. Hence, we focus on those that have two parents in the mirrored and
+% unmirrored HMM runs, the same set of mutations, and the left and right parents
+% are consistently imputed for Pango lineage status.
+%     We focus on the Wide ARG here because it covers (roughly?) the same time
+% period as the RIPPLES analysis.
+
+We explore recombination signals across the SARS-CoV-2 genome by inspecting the
+breakpoint intervals of the filtered recombination nodes from the wide ARG
+(sampled to mid-2021) in Figure~\ref{fig:breakpoint-distribution}. The length
+of a breakpoint interval reflects uncertainty in the estimated location of a
+breakpoint. On average, the length of the breakpoint intervals is 2,895 bases
+(median, 1,670 bases; range, 1 to 23,666 bases). This broad distribution of
+interval length indicates that the location of a breakpoint often could not be
+precisely determined given the observed genetic variation.
+
 \begin{figure}
@@
 Distribution of recombination breakpoints and mutations along the genome in
-the Wide ARG. Top panel shows the intervals for 1,769 breakpoints associated
-with 1,522 recombination nodes with at least two descending samples, plotted along the genome
-as line segments (coloured by interval width). The inset histogram shows the
+the Wide ARG. Top panel shows the intervals for 1769 breakpoints associated
+with 1522 recombination nodes $>1$ descending samples plotted along the genome
+as line segments, coloured by interval width. The inset histogram shows the
 distribution of these interval widths.
@@
 per site (right axis, blue).
-The top-ten sites by mutation count are annotated.
+TODO describe mutation threshold.
 See Figure~\ref{fig:long_arg_breakpoint_distribution} for the equivalent plot
@@
 
-Figure~\ref{fig:breakpoint-distribution} shows the distribution of breakpoint
-intervals and patterns of recurrent mutation along the genome in the Wide ARG
-(see Figure~\ref{fig:long_arg_breakpoint_distribution} for the same information
-of the Long ARG). We focus on the Wide ARG here as it covers roughly the same time
-period as the analyses of \cite{Turakhia2022-it}, facilitating comparisons
-of the results. To reduce the effect of artefactual recombinants, we
-consider only the breakpoints associated with the 1,522 recombination nodes
-that are ancestral to more than one sample. The mean length of these intervals
-is 1,685 bases (median 962); the length distribution is summarised in the
-inset histogram in Figure~\ref{fig:breakpoint-distribution}.
-
-Figure~\ref{fig:breakpoint-distribution} [ TODO what can we really say here?
-Other than the boundaries happen to lie on gene breaks, not an awful lot I
-think?]
-
-% This isn't really true any more, is it?
-% There is a preponderance of breakpoint intervals spanning the 3’ end of the
-% genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}. Inspecting
-% more closely, we find an enrichment in the number of intersecting intervals in
-% the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure
-% 4B). Markedly, there is an abrupt drop in the number of intersecting intervals
-% in the N gene following the peak in ORF8 towards the 3’ end.
+There is a preponderance of breakpoint intervals spanning the 3’ end of the
+genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}. Inspecting
+more closely, we find an enrichment in the number of intersecting intervals in
+the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure
+4B). Markedly, there is an abrupt drop in the number of intersecting intervals
+in the N gene following the peak in ORF8 towards the 3’ end.
 
@@
 Distribution of recombination breakpoints and mutations along the genome in
-the Long ARG.
-Top panel shows the intervals for 851 breakpoints associated
-with 763 recombination nodes with at least two descending samples, plotted along the genome
-as line segments (coloured by interval width).
-Other details as described in Figure~\ref{fig:breakpoint-distribution}.}
+the Long ARG.}
 \end{figure}
@@
 \includegraphics[width=\textwidth]{figures/supp_recombination_node_mrcas.pdf}
-\caption{\label{fig:recomb_mrcas_voc_breakdown}  Divergence between parent
-lineages for recombination events within and among different VoC categories.
-HMM-consistent recombinants are shown in green, HMM-inconsistent in blue
-(including many of the named Delta-Omicron recombinants). We find 69 HMM
-consistent (77 inconsistent) Alpha-Alpha recombination breakpoints = 13\%
-(10\%) of the total; 84 HMM consistent (143 inconsistent) Delta-Delta
-recombination breakpoints = 18\% (18\%) of the total; 99 HMM consistent (155
-inconsistent) Omicron-Omicron recombination breakpoints = 22\% (19\%) of the
-total; and 2 HMM consistent (20 inconsistent) Delta-Omicron recombination
-breakpoints = 2\% (< 1\%) of the total. These plots do not include
-recombination breakpoints involved lineages that were not classified into these
-VoC categories}
+\caption{\label{fig:recomb_mrcas_voc_breakdown}  Divergence between parent lineages for recombination events within and among different VoC categories.
+HMM-consistent recombinants are shown in green, HMM-inconsistent in blue (including many of the named Delta-Omicron recombinants).
+We find 69 HMM consistent (77 inconsistent) Alpha-Alpha recombination breakpoints = 13\% (10\%) of the total; 84 HMM consistent (143 inconsistent) Delta-Delta recombination breakpoints = 18\% (18\%) of the total; 99 HMM consistent (155 inconsistent) Omicron-Omicron recombination breakpoints = 22\% (19\%) of the total; and 2 HMM consistent (20 inconsistent) Delta-Omicron recombination breakpoints = 2\% (< 1\%) of the total.
+These plots do not include recombination breakpoints involved lineages that were not classified into these VoC categories}
 \end{figure}

### 2023-05-18 paper.tex

--- 2023-05-17 paper.tex
@@
 that are ancestral to more than one sample. The mean length of these intervals
-is 1,685 bases (median 962); the length distribution is summarised in the
+is 1,685 bases (median 962), and the length distribution is summarised in the
 inset histogram in Figure~\ref{fig:breakpoint-distribution}.
 
-Figure~\ref{fig:breakpoint-distribution} [ TODO what can we really say here?
-Other than the boundaries happen to lie on gene breaks, not an awful lot I
-think?]
-
-% This isn't really true any more, is it?
-% There is a preponderance of breakpoint intervals spanning the 3’ end of the
-% genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}. Inspecting
-% more closely, we find an enrichment in the number of intersecting intervals in
-% the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure
-% 4B). Markedly, there is an abrupt drop in the number of intersecting intervals
-% in the N gene following the peak in ORF8 towards the 3’ end.
-
-Next, we investigate the potential recombination signals alongside the number
-of mutations across the genome (Figure 4C). Regions with a higher density of
-mutations contain more genetic variation, increasing the power to detect
-recombination events. The position 28,271 (located in the ORF8-N intergenic
-region) has a particularly high number of mutations in the samples (Figure 5).
-At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions
-(n = 1,605) appear to occur frequently. One possible explanation is that this
-homopolymeric region is prone to sequencing errors, giving rise to an
-artifactual recombination signal. At the upstream position 27,972, numerous
-mutations also occur. This position has been reported to be hypermutable
-\citep{Jungreis2021-dh}.
-
-We further consider the breakpoint intervals within the context of
-recombination breakpoint sequence motifs. It is hypothesised that certain
-palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote
-template switching during replication in SARS-CoV-2 via formation of
-base-paired stem loops in the genome structure \citep{Gallaher2020-lb}. There
-are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019
-reference sequence (Supplementary Table 5). We observe that XX (XX\%) of the
-breakpoint intervals of the HMM-consistent recombinants span at least one of
-the breakpoint sequences.
+While the detailed interpretation of Figure~\ref{fig:breakpoint-distribution}
+requires further study, a few key properties stand out.[TODO, write this
+better. ] Firstly, in several cases there is a clear relationship between
+gene boundaries and the breakpoint intervals. For example, ORF7a and OR7b
+show sharp differences in the number of intersecting intervals within
+these genes. This could be due to a number of factors, such as [BIOLOGY]
+or the presence of sites with artefactually high numbers of mutations
+in the intergenic regions (see the next paragraph for more discussion).
+Secondly, we see substantially fewer intervals intersecting with the N
+gene than other regions, in agreement with the results of
+\cite{Turakhia2022-it}.
+
+It is important to note that the exact endpoints of intervals can be
+somewhat arbitrary, since they are defined by exact comparisons
+of a recombinant's parents. The presence of problematic sites that are enriched
+for errors will therefore tend to artificially truncate breakpoint intervals
+at these sites.
+It is therefore helpful to compare the number of interecting intervals
+with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
+For example, position 28,271 (located in the ORF8-N intergenic
+region) has the largest number of mutatations and corresponds to the
+left endpoint of 9 intervals and the right of 57.
+The 7,572 mutations  (including 5,782 insertions and 1,605 deletions)
+at this site are a likely indicator that this
+homopolymeric region is prone to sequencing errors
+and should be included in the list of ``problematic sites'' that
+are excluded from analysis (Section~\ref{sec:data_preprocessing}).
+% There can be good biological reasons for this too though
+However, hypermutable sites such as 27,972 \citep{Jungreis2021-dh}
+have the same effect of truncating intervals. At this site we have
+770 mutations (including 425 C$>$T, 314 T$>$C) with potential signals
+of polymorphism maintained by selection [FIXME!!], and 58 intervals
+with endpoints within one base of this site.
+
+% JK Dropping this for now:
+% We further consider the breakpoint intervals within the context of
+% recombination breakpoint sequence motifs. It is hypothesised that certain
+% palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote
+% template switching during replication in SARS-CoV-2 via formation of
+% base-paired stem loops in the genome structure \citep{Gallaher2020-lb}. There
+% are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019
+% reference sequence (Supplementary Table 5). We observe that XX (XX\%) of the
+% breakpoint intervals of the HMM-consistent recombinants span at least one of
+% the breakpoint sequences.
 

### 2023-05-18 paper.tex
Pass through recombination parents section

@@
 \label{sec:parent_divergence}
-% - Divergence between recombinant parents
-%     We study the recombination events in the Long ARG in terms. A recombinant
-% will not always have two parents, for reasons X, Y and Z. As a first
-% approximation, we can assume that each breakpoint corresponds to a
-% recombination event (although there are some exceptions). Here we have to get
-% into all the hairy details about which sets of breakpoints we filter and don't
-% filter.
+% Paragraph 1, what does all this mean?
+
+% Purpose of this paragraph - remind the reader that we can only do this
+% sort of things because we have an ARG, and we can do  this at a far
+% greater level of granularity than existing methods.
+[ FIXME this is a rough outline of what an intro para might say]
+An important advantage of inferring a joint model of inheritance,
+mutation and recombination in an ARG (Section~\ref{sec:args})
+is that is allows us to precisely trace the origins of recombinants.
+[Katherine - need some input from you here to say this properly?]
+Existing methods only provide us with approximate estimates of
+the origins of recombination events, and are not able to distinguish
+recombinations between closely related lineages.
+
+% Paragraph 2, what do we plot, specifically?
+In this section we investigate the patterns of divergence between
+the parents of recombination nodes.
+We focus on the Long ARG because  it covers time
+periods where recombination is known to have had a significant
+impact, and includes
+[HOW MANY?]
+samples designated to [HOW MANY] Pango X lineages (i.e., those inferred
+to have recombinant ancestry; see Section~\ref{sec:pango_x_lineages} for
+further analysis of Pango recombinant lineages in the Long and Wide ARGs).
+Figure~\ref{fig:recomb_mrcas} shows the estimated date of the most recent
+common ancestor (MRCA) of the parent nodes for each recombination breakpoint,
+plotted against the divergence between these parents (i.e., the
+total branch length from the parents to their MRCA in the trees to
+the immediate left and right of the breakpoint).
+As in Section~\ref{sec:breakpoint_intervals}, we exclude recombination nodes
+that are ancestral to only one sample from this analysis.
+Figure~\ref{fig:recomb_mrcas}A plots breakpoints for all recombinants that
+are ancestral to more than one sample, and
+Figure~\ref{fig:recomb_mrcas}B plots a subset which pass some further
+reliability filters. Specifically, only recombination nodes that
+[FILL ME IN].
+This reduces the number of recombination nodes
+to 347 and breakpoints to 358.
+Note that each plotted point
+represents a breakpoint, so that a recombination node with more than one
+breakpoint (e.g. with 3 or more parents, comprising ${\sim}10 \%$ of the
+recombination nodes in Figure ~\ref{fig:recomb_mrcas}A and ${\sim}3 \%$ in
+Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
+It is important to note here that time estimates in these figures may be unreliable,
+since nodes other than samples are dated somewhat crudely in the current version
+of \texttt{sc2ts} (Section~\ref{sec:node_dating}).
 
@@
 Date of common ancestry between the parents on either side of recombination
-breakpoints, as a function of parental divergence. MRCAs of parents
-inferred to have caused a Pango designated recombinant (XA, XB, etc.)
-identified in orange (A) 851 parent pairs (singleton recombinants removed).
-Parental strains often descend from the same MRCA; each horizontal dotted line
-indicates a frequently found MRCA: these tend to be associated with major waves
-and often have many immediate children (i.e. are large
-polytomies, likely to indicate undersampled outbreaks) (B) Further filtered
-to 358 parent pairs by removing HMM inconsistent breakpoints. For equivalent
-plots broken down by parental VoC classification, see
-Figure~\ref{fig:recomb_mrcas_voc_breakdown}
+breakpoints, as a function of the divergence time between the parents.
+MRCAs of parents associated with Pango designated recombinants (XA, etc)
+are identified in orange. (A) 851 parent pairs (singleton recombinants removed).
+(B) Further filtered
+to 358 parent pairs by removing HMM inconsistent breakpoints (see text).
+Horizontal dotted lines show the four most common MRCA nodes,
+which tend to be associated with major outbreaks and with many immediate children.
+See Figure~\ref{fig:recomb_mrcas_voc_breakdown}
+for equivalent plots broken down by parental VoC classification.
 }
@@
 
-Although recombination nodes are relatively infrequent, there are still many
-present in the wide and long ARGs (4123 and 2078 respectively; Supplementary
-Tables 1 and 2). Each recombination node was inferred as a
-result of matching a specific (``causal'') recombinant sample into the ARG.
-However, ~63\% of them are ancestral to only a single
-recombinant sample; we exclude these "singleton recombinants" from subsequent
-analyses as they may have arisen from sequencing of mixed samples (e.g., from
-co-infection or contamination), and in any case, never result in onward
-transmission.  This leaves 1522 and 763 recombination nodes in the wide and
-long ARGs, respectively. In their analysis using RIPPLES,
-\cite{Turakhia2022-it} detected 589 non-singleton recombinants.
-
-We focus on the long ARG because compared to the wide ARG, it includes time
-periods of increased rates of coinfection and recombination, and contains many
-samples previously identified as the product of recombination (i.e. with a
-Pango lineage designation starting with an "X"). Not only does the long ARG
-capture all but one of the Pango designated recombinant lineages during this
-time period (see the section below), but it also includes a large number of
-non-Pango-identified recombinants. Many of these involve recombination between
-relatively closely related lineages, as measured by the divergence time between
-the recombination events and the most recent common ancestor (MRCA) shared by
-the left and right parent at each breakpoint. As can be seen from
-Figure~\ref{fig:recomb_mrcas}, this divergence time ranges from as little as
-five or six days, to as much as several years in the case of recent
-recombinants. This demonstrates that sc2ts can identify viral recombination
-between very closely related lineages, providing a degree of resolution not
-available with other recombination detection methods.
-
-The histograms in Figure~\ref{fig:recomb_mrcas} show that
-majority of newly identified recombinants arise from parents that diverged 10 to 30 weeks previously.
-However, there is also a peak of recombinants that diverged ${\sim}95$ weeks prior to the event:
-these correspond to recombinants whose parental common ancestor traces to early 2020,
-including both Delta-Delta and Delta-Omicron recombinants
-(Supplementary Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
-Note that each plotted point represents a breakpoint, so that a recombination
-event with more than one breakpoint (e.g. with 3 or more parents, comprising
-${\sim}10 \%$ of the recombination events in Figure ~\ref{fig:recomb_mrcas}A and
-${\sim}3 \%$ in Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
-
-As a check on the reliability of these identified recombinations,
-Figure~\ref{fig:recomb_mrcas}B  plots only breakpoints which are
-HMM-consistent. This reduces both the number of recombination nodes \&
-breakpoints in the long ARG from 763 \& 851 to 347 \& 358
-(and in the wide ARG from 1522 \& 1769 to XXX \& YYY , see
-Supplementary Tables 1 and 2).
-
-The date of common ancestry for the two parents of recombinant lineages is
+% Paragraph 3, what do we say about the dates/nodes of the MRCAs?
+The date of the MRCA of recombinant parents is
 concentrated in several banded rows in Figure~\ref{fig:recomb_mrcas}. These are
-largely due to a small handful of MRCAs which are shared between a large number
-of recombination events; the top four are indicated by dotted lines in the
-figure. These shared MRCAs lie near the root of large expansions of certain
-lineages (B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
-and BA.2: second Omicron wave), as evidenced by the number of samples with those
+largely due to a few MRCAs shared by many
+recombinants; the top four are indicated by dotted lines in the
+figure. These shared MRCAs lie near the root of large expansions
+(B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
+and BA.2: second Omicron wave), as demonstrated by the number of samples with those
 Pango designations which trace back to each ancestral node (listed as
 percentages on the plot). The majority of these frequently-shared MRCAs
-are associated with extensive polytomies in the tree (at the extreme,
-the BA.2-linked node has 2,872 immediate children). This is likely to
-indicate a rapid and under-sampled expansion of a clade in the SARS-CoV-2 genealogy.
-
-Finally, it is clear from Supplementary Figure\ref{fig:recomb_mrcas_voc_breakdown} that of the recombinant
+are associated with large polytomies (at the extreme,
+the BA.2-linked node has 2,872 immediate children), likely
+indicating a rapid and under-sampled expansion of a clade.
+
+% Paragraph 4, what do we say about the divergence times?
+The histograms in Figure~\ref{fig:recomb_mrcas} show that there is a
+broad range in divergance times among recombinant parents.
+We can also see that the majority of recombinants arise
+from parents that diverged 10 to 30 weeks
+previously. However, there is also a peak of recombinants that diverged
+${\sim}95$ weeks prior to the event: these correspond to recombinants whose
+parental common ancestor traces to early 2020, including both Delta-Delta and
+Delta-Omicron recombinants
+(Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
+Figure~\ref{fig:recomb_mrcas_voc_breakdown} also shows that, of the recombinant
 parents that fall into the Alpha, Delta, and Omicron VoC classes, the majority
-are within-Delta and within-Omicron recombinants, and that Omicron followed by
-Delta, are the variants associated with the most recombination. This may
+are within-Delta and within-Omicron recombinants, and that Omicron and
+Delta are the variants associated with the most recombination. This may
 reflect either sampling intensity, the prevalence of cases (which increases the
-chance of coinfection and recombination), or possibly heterogeneity in
+chance of coinfection and recombination), or possible heterogeneity in
 recombination probabilities among lineages.
@@
 The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
-1, XAD: 1, XAH: 1, XAK: 1) is the same. This node is ancestral to more than
-127K samples, and is likely to be a spuriously inferred recombination event.
-Understanding the origin of this and other similar nodes, is an important
+1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, caused by strain Denmark/DCGC-281594/2021,
+is ancestral to more than 127K samples, most of which are not designated as
+recombinants by any other method. It is highly likely to be a spuriously inferred recombination event.
+Understanding the origin of this and other similar nodes is an important
 direction for future work (see Discussion). For simplicity, we exclude these
@@
 
-\caption{\label{fig:pango-simple-origin-graph} Simple origin
+\caption{\label{fig:pango-simple-origin-graph} Non-nested
 Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
@@
 Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
-simple-origin (XA) type with no other Pango designations among their descendants.
-The remaining 10 are of the XAG type with multiple Pango
-designated lineages as additional descendants of the originating recombination event. In some cases,
-these may be, however, be a result of erroneous Pango designations. Supplementary Table~\ref{tab:pango-recombinants} also shows the official Pango designated parent lineages for
-each recombinant lineage. The parent lineages in the Long ARG agree to a great extent with these, although \texttt{sc2ts} provides more a precise
-parent designation.
+basic (XA) type with no other Pango designations among their descendants.
+The remaining ten are of the XAG type with multiple Pango
+designated lineages as additional descendants of the originating recombination event.
+In some cases,
+these may be, however, be a result of erroneous Pango designations.
+Supplementary Table~\ref{tab:pango-recombinants} also shows the official
+Pango designated parent lineages and \texttt{sc2ts} inferred parent lineages,
+which extensively agree (although \texttt{sc2ts} provides more a precise
+parent designation).
 
@@
 recombinant origins, such that their samples trace to >1 most recent
-recombination node in the ARG. Of these, 11 are cases where none of the
-recombinants are nested (i.e. there is no overlap between the descendant samples
-of each recombination node). In 9 of these 11, there is a single ``main''
+recombination node in the ARG. Of these, 11 are cases where the
+recombinants are independent rather than nested (i.e. there is no overlap in the list
+of descendant samples for each recombination node). Most have a single ``main''
 recombination event from which the majority of the corresponding recombinant samples
-descend (in other two, there are too few Pango X samples to decide on a ``main'' clade:
+descend (but in XJ and XU there are too few Pango X samples to decide on a ``main'' clade:
 see Supplementary Table~\ref{tab:pango-recombinants}).
 
-Figure~\ref{fig:pango-simple-origin-graph}C shows a simple example, consisting of
+Figure~\ref{fig:pango-simple-origin-graph}C shows a simple multiple-origin example, consisting of
 the 4 samples labelled XD by Nextclade in the long ARG. The left hand subgraph
@@
 A similar pattern is seen for several of the origins inferred for the Pango
-XN lineage (originally defined on the basis of sequences from the UK). The
-Long ARG infers four independent origins for nonsingleton Nextclade-designated XN
-samples (Supplementary Table~\ref{tab:multiple_origins_table}). One of the recombinant origins,
-caused in our ARG by an English sample, accounts for 155 out of 164 XE samples
-(primarily UK and Irish, but with a few northern European and north American strains).
-This certainly represents the XE recombinant lineage as originally defined. The other recombination nodes
-only give rise for one, two, or three XE samples. Three of these recombination
-nodes are likely to represent Nextclade mislabelling: they either have different
-inferred breakpoint positions or (in the case of the earliest recombination node),
-descendants that are primarily labelled as other Pango X-lineages. Additional support
-for the suggestion that they result from incorrect labelling comes from the fact that
-none of their XE-labelled descendant samples are from the UK or Ireland.
-It is less clear if the most recent XE cluster is a mislabelling or a case of \texttt{sc2ts}
-inferring multiple recombinations when there should be just one. This cluster of 2 samples
-(one of which is English) has exactly the same breakpoint estimates as the
-main cluster. If the only samples available are rather genetically distant from an originating
+XN lineage (originally defined on the basis of sequences from the UK). After removing
+singleton recombinants and the descendants of the erroneously recombinant
+Denmark/DCGC-281594/2021 strain (which, however, includes 53 XN-designated samples)
+the Long ARG contains four independent origins for Nextclade-designated XN
+samples (Supplementary Table~\ref{tab:multiple_origins_table}). The oldest
+of these four origins has 3 mutations above the sample that caused the
+recombination to be inferred. It therefore seems likely that we have not managed
+to sample an XN strain that is very close to the true recombinant. This may
+lead to `texttt{sc2ts} mistakenly inferring multiple recombination events when there
+
+If the only samples available are rather genetically distant from an originating
 recombinant, we might expect sc2ts to infer multiple similar recombination events
@@
 samples whose ancestry involves multiple non-independent (``nested'') recombination events
-(marked with asterisks in Supplementary Table~\ref{tab:pango-single-origin}; the most
+(marked with asterisks in Supplementary Table~\ref{tab:pango-recombinants}; the most
 extensive is for XAB). The earliest example, present in both the Wide and Long ARGs, is XB
@@
 on 2022-02-01;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/480}).
-XN samples are grouped into rows based on their most recent ancestral recombination node;
+The 13 XN samples are grouped into rows based on their most recent ancestral recombination node;
+all the descendant samples of the four recombination nodes are designated XN by Nextclade.
 Details are also provided for the ``causal strain'' that led to the recombination
-being proposed by \texttt{sc2ts}.
-Note that a large number of XN samples were previously removed by filtering.
-When undersampling causes three or more mutations to lie immediately above the causal strain, \texttt{sc2ts} may mistakenly propose multiple independent origins rather than one.
+being proposed by \texttt{sc2ts}. When undersampling causes three or more mutations to lie
+immediately above the causal strain, \texttt{sc2ts} may mistakenly propose multiple
+independent origins rather than one. Note that a large number of XN samples were previously
+removed by filtering. 
 }

### 2023-05-18 paper.tex
Minor edits

--- 2023-05-18 paper.tex
@@
 Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
-It is important to note here that time estimates in these figures may be unreliable,
+It is important to note here that some of the time estimates in these figures
+come with a substantial degree of uncertainly,
 since nodes other than samples are dated somewhat crudely in the current version
@@
 The histograms in Figure~\ref{fig:recomb_mrcas} show that there is a
-broad range in divergance times among recombinant parents.
+broad range in divergence times among recombinant parents.
 We can also see that the majority of recombinants arise
@@
 (C) Equivalent plot for both origination events involving the four XD samples in the long ARG.
-Details of the mutations and sample node identities for all three plots are provided in supplementary figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
+Details of the mutations and sample node identities for all three plots are provided in supplementary Figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
 } \end{figure}
@@
 marked here as XAB are reclassified as BA.2 (and XAG becomes fully monophyletic,
-see supplementary figure \ref{fig:pango_XAG_gisaid_graph}). This is an
+see supplementary Figure \ref{fig:pango_XAG_gisaid_graph}). This is an
 independent confirmation of the uncertainty in designation of these XAG-related
@@
 a single recombination event. 15 Pango X lineages are inferred to have multiple
-recombinant origins, such that their samples trace to >1 most recent
+recombinant origins, such that their samples trace to $>$1 most recent
 recombination node in the ARG. Of these, 11 are cases where the
@@
 Figure~\ref{fig:pango-simple-origin-graph}C shows a simple multiple-origin example, consisting of
-the 4 samples labelled XD by Nextclade in the long ARG. The left hand subgraph
+the 4 samples labelled XD by Nextclade in the Long ARG. The left hand subgraph
 (the origin of three XD samples, all sampled in France) has an earliest
@@
 to sample an XN strain that is very close to the true recombinant. This may
-lead to `texttt{sc2ts} mistakenly inferring multiple recombination events when there
+lead to \texttt{sc2ts} mistakenly inferring multiple recombination events when there
 
@@
 a recombinant outbreak
-has taken a "true" XE sample and mistakenly inferred an additional, spurious origin. [WHY? - this
+has taken a ``true'' XE sample and mistakenly inferred an additional, spurious origin. [WHY? - this
 is presumably because it has changed too much from the nearest in the main XE cluster - need to discuss the
@@
 \subsubsection{Nested recombinant origins}
-
-\begin{figure} \centering
-\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
-\caption{\label{fig:complex_origins_graph}  A subgraph of the Long ARG showing
-nested recombination events involving Pango lineage XB. All XB samples
-trace to a single recombination node (top centre), but three further recombinations
-also descend from this node. The samples descending from these nested
-recombinations include 9 that are assigned by Nextclade to
-various non-recombinant pre-Alpha lineages (blue).}
-\end{figure}
-
 As well as cases where Pango X lineage origins are attributed to independent
@@
 descendants).
+
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
+\caption{\label{fig:complex_origins_graph}  A subgraph of the Long ARG showing
+nested recombination events involving Pango lineage XB. All XB samples
+trace to a single recombination node (top centre), but three further recombinations
+also descend from this node. The samples descending from these nested
+recombinations include 9 that are assigned by Nextclade to
+various non-recombinant pre-Alpha lineages (blue).}
+\end{figure}
 

### 2023-05-18 paper.tex
First pass at abstract, some initial stuff on intro

@@
 \begin{abstract}
-XX
+Recombination is an important and ongoing feature of SARS-CoV-2 evolution,
+and presents major challenges to [things important to public health].
+Detection of new recombinant strains is challenging, with methods
+[being a bit rubbish].
+In addition, many basic approaches fundamental to the study of viral
+evolution assume that recombination is negligible, in that a single
+phylogenetic tree can represent the genetic ancestry of the
+circulating strains. Here we present an initial version of
+\texttt{sc2ts}, a method to automatically detect recombinants
+in real-time and to fully integrate them into a cohesive joint
+model of mutation, recombination and genetic inheritance known
+as an Ancestral Recombination Graph (ARG). We infer two ARGs under
+different sampling strategies, and study their properties.
+One is densely sampled, and includes 1.27 million sequences
+sampled up to June 30, 2021.
+The second is more sparsely sampled,
+consisting of 784K sequences sampled up to June 30, 2022.
+We find that both ARGs are
+% better words?
+highly consistent with
+known features of SARS-CoV-2 evolution, recovering the basic
+backbone phylogeny, mutational spectra, and recapitulating
+details on the majority of known recombinant lineages.
+Using the well-established and fully-featured \texttt{tskit} library,
+the ARGs can also be stored concisely and processed efficiently
+using standard Python tools. For example, the ARG for 1.27 million
+sequences---encoding the inferred reticulate ancestry,
+full alignments, and extensive metadata---requires
+58MB of storage,
+and loads in less than a second.
+%JK Not sure if we'd actually say it like this, but this is the
+% basic gist I'd like to get accross[
+The ability to fully integrate the effects of recombination into
+downstream analyses, to quickly and automatically detect new recombinants,
+and to provide an efficient and convenient platform for computation
+makes \texttt{sc2ts} a promising approach, worthy of
+further investment and development.
 \end{abstract}
@@
 \section{Introduction}
+% Recombination is an important force in SARS-CoV-2, recombinants
+% have arisen and they have spread to high frequencies
 Recombination via template switching is a common feature
@@
 immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
-
 The lack of genetic variation made recombination challenging to detect early in the SARS-CoV-2 pandemic. In late 2020, the appearance and
@@
 Delta which harboured multiple characteristic mutations, increased the
-detectability of recombination. By searching for genomes containing segments
-from different VoC designations, multiple recombinant lineages have been
-detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha
-and Epsilon \citep{Wertheim2022-hj}, and Alpha and
-Delta~\citep{Sekizuka2022-xz}. More recently, the high prevalence of Omicron,
+detectability of recombination.
+More recently, the high prevalence of Omicron,
 with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
-accelerated the rate of coinfection and the potential for recombination
-\citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of
-onward transmission (more than five descendants) and monophyly ($>70\%$
-bootstrap support) are eligible for a Pango lineage
-designation~\citep{Rambaut2020-dw}, starting with the letter X
-% FIXME: should this be a citation?
-(``Pango Lineage Nomenclature: Provisional Rules for Naming Recombinant
-Lineages'' 2021).
-
-Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
-combine specific subsets of mutations from different lineages
-\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
-This approach only identifies those inter-lineage recombination events
-resulting in the patterns investigated. Others have inferred recombinant
-genealogies to disentangle recombination from recurrent mutation, without
-having to pre-specify characteristic mutations for each lineage
-\citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it} presented a
-phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
-over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find
-candidate recombinants, RIPPLES first builds a phylogeny without assuming
-recombination and then scans for long branches (i.e., carrying many mutations)
-and then determines if these candidates would be better explained by
-recombination. Specifically, RIPPLES exhaustively breaks each candidate
-sequence into segments and attempts to find more parsimonious placements for
-each segment on the phylogeny. If such placements are found, the sequence is
-identified as a putative recombinant. Although it enables rapid searches for
-genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
-accounts for only mutations, treats recombinants \textit{post hoc}, and is an
-incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
-As noted by the authors, a \textit{post hoc} treatment of recombination is
-possible when recombinant lineages are rare and leave few descendants. However,
-this is becoming increasingly untenable with the proliferation of recombinant
-lineages; for example, more than half of the sequences sampled in February 2023
-are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
-future evolution of SARS-CoV-2 is likely to involve multiple sequential
-recombination events on top of existing recombinant lineages, creating a highly
-reticulated genealogy.
-
-%% ARGs are good for *representing* viral genealogies
-
-It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
-and affects the results of downstream analysis, e.g., inference of
-selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
-not account for recombination
-\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
-genetic ancestries of recombining lineages can be represented by ancestral
-recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
-capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
-development of the ``succinct tree sequence'' data
-structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
-encoding of large ARGs. This representation should be well-suited to storing
-large recombinant viral genealogies such as SARS-CoV-2,
-which currently encompasses over 15 million sequences in the GISAID database
-\citep{Shu2017-hp}). Moreover the \texttt{tskit} software library, which implements this
-data structure, enables efficient manipulation of the stored genealogies and is
-integrated with a suite of open-source software tools to simulate, analyse, and
-infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.
-
-%% ARG inference is hard
-
-Inferring recombinant phylogenies has been a long-standing computational challenge.
-However, approaches to infer ARGs now exist that can scale to tens of
-thousands of human genomes and beyond
-\citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}, and which
-deal with levels of repeated recombination far in excess of those seen in viral
-phylogenies. For example, using over 90 million variable sites
-\citet{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands
-of modern and ancient human genomes, accounting for millions of recombination
-events (over 5 million ``local trees''). To achieve this, \cite{Wohns2022-th}
-took advantage of a highly efficient Hidden Markov Model (HMM) implementation
-of the Li and Stephens (LS) model \citep{Li2003-ib} to infer copying paths between
-genomes \citep{Kelleher2019-ba}.
-
-% What is new here?
-
-Here, we present a preliminary version of a novel scalable methodology
-(\texttt{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
-and additionally takes advantage of known collection dates to infer ARGs from millions
-of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
-we demonstrate that \texttt{sc2ts} can be used to simultaneously (1) detect putative SARS-CoV-2
-recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
-(3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record
-of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
-\texttt{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
-\texttt{Sc2ts} is a promising method to automate detection of SARS-CoV-2 recombinants
-and to jointly model recombinational and mutational processes shaping the genomic
-diversity of SARS-CoV-2 at scale.
+accelerated the rate of coinfection and the potential for
+recombination~\citep{Bal2022-hq}.
+[FIXME say this better and add citation.]
+Multiple recombinant lineages have successfully
+established and spread to high frequency.
+
+[PARA 2: recombination is hard to detect. Quick overview of state-of-the-art.
+Finish with phylogeny-based method, RIPPLES.]
+
+[PARA 3: but you can't use a phylogeny when there's recomb. Need a joint
+model of mutation and recomb. ARGs, tskit, forward refs to ARG section for
+details.]
+
+[PARA 4: we present sc2ts based on these recent advances. Summarise
+results, with forward refs to Results sections.]
+
+% By searching for genomes containing segments
+% from different VoC designations, multiple recombinant lineages have been
+% detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha
+% and Epsilon \citep{Wertheim2022-hj}, and Alpha and
+% Delta~\citep{Sekizuka2022-xz}.
+
+% Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
+% combine specific subsets of mutations from different lineages
+% \citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
+% This approach only identifies those inter-lineage recombination events
+% resulting in the patterns investigated.
+% Others have inferred recombinant
+% genealogies to disentangle recombination from recurrent mutation, without
+% having to pre-specify characteristic mutations for each lineage
+% \citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it} presented a
+% phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
+% over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find
+% candidate recombinants, RIPPLES first builds a phylogeny without assuming
+% recombination and then scans for long branches (i.e., carrying many mutations)
+% and then determines if these candidates would be better explained by
+% recombination. Specifically, RIPPLES exhaustively breaks each candidate
+% sequence into segments and attempts to find more parsimonious placements for
+% each segment on the phylogeny. If such placements are found, the sequence is
+% identified as a putative recombinant. Although it enables rapid searches for
+% genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
+% accounts for only mutations, treats recombinants \textit{post hoc}, and is an
+% incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
+% As noted by the authors, a \textit{post hoc} treatment of recombination is
+% possible when recombinant lineages are rare and leave few descendants. However,
+% this is becoming increasingly untenable with the proliferation of recombinant
+% lineages; for example, more than half of the sequences sampled in February 2023
+% are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
+% future evolution of SARS-CoV-2 is likely to involve multiple sequential
+% recombination events on top of existing recombinant lineages, creating a highly
+% reticulated genealogy.
+
+% %% ARGs are good for *representing* viral genealogies
+
+% It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
+% and affects the results of downstream analysis, e.g., inference of
+% selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
+% not account for recombination
+% \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
+% genetic ancestries of recombining lineages can be represented by ancestral
+% recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
+% capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
+% development of the ``succinct tree sequence'' data
+% structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
+% encoding of large ARGs. This representation should be well-suited to storing
+% large recombinant viral genealogies such as SARS-CoV-2,
+% which currently encompasses over 15 million sequences in the GISAID database
+% \citep{Shu2017-hp}). Moreover the \texttt{tskit} software library, which implements this
+% data structure, enables efficient manipulation of the stored genealogies and is
+% integrated with a suite of open-source software tools to simulate, analyse, and
+% infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.
+
+% %% ARG inference is hard
+
+% Inferring recombinant phylogenies has been a long-standing computational challenge.
+% However, approaches to infer ARGs now exist that can scale to tens of
+% thousands of human genomes and beyond
+% \citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}, and which
+% deal with levels of repeated recombination far in excess of those seen in viral
+% phylogenies. For example, using over 90 million variable sites
+% \citet{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands
+% of modern and ancient human genomes, accounting for millions of recombination
+% events (over 5 million ``local trees''). To achieve this, \cite{Wohns2022-th}
+% took advantage of a highly efficient Hidden Markov Model (HMM) implementation
+% of the Li and Stephens (LS) model \citep{Li2003-ib} to infer copying paths between
+% genomes \citep{Kelleher2019-ba}.
+
+% % What is new here?
+
+% Here, we present a preliminary version of a novel scalable methodology
+% (\texttt{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
+% and additionally takes advantage of known collection dates to infer ARGs from millions
+% of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
+% we demonstrate that \texttt{sc2ts} can be used to simultaneously (1) detect putative SARS-CoV-2
+% recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
+% (3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record
+% of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
+% \texttt{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
+% \texttt{Sc2ts} is a promising method to automate detection of SARS-CoV-2 recombinants
+% and to jointly model recombinational and mutational processes shaping the genomic
+% diversity of SARS-CoV-2 at scale.
 

### 2023-05-18 paper.tex
Minor edits to Results

@@
 in both ARGs involve inheriting directly from sampled genomes.
-Both ARGs contain 29,422 sites (\ref{sec:data_preprocessing}),
+Both ARGs contain 29,422 sites (Section \ref{sec:data_preprocessing}),
 and a large number of mutations.
@@
 in SARS-CoV2 recombination is rare, with recombination nodes accounting
-for $<$0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
+for $<$0.3\% of the inferred ancestry (Table~\ref{tab:args}). As a result,
 relationships between strains can often be represented by a
 single phylogenetic tree, particularly when looking at
-a subset of sampled genomes e.g. restricted by time or limited to close
-relatives. We expect the ARG to be particularly tree-like
+a subset of sampled genomes, e.g.\ restricted by time or limited to close
+relatives. We expect ARGs to be particularly tree-like
 early in the pandemic, when co-infection was less likely.
@@
 
-Our \texttt{sc2ts} backbone phylogenies do contain small amounts of recombination:
-there are 7 recombination nodes in the backbone of the Wide ARG and 8 in the backbone
-of the Long ARG. However, the backbones remain extremely tree-like, as these
-recombination events involve minor, local topological rearrangements. In particular,
-recombination in the backbone phylogenies only occurs between close relatives
-(e.g.\ sibling lineages), sometimes even within a single lineage such that the parent
-lineages above a recombination event coalesce with each other before involvement with
-any other part of the phylogeny. These local reticulations could either represent
+The \texttt{sc2ts} backbone phylogenies contain small amounts of recombination,
+with 7 recombination nodes in the Wide ARG backbone (8 for the Long ARG
+backbone). However, the backbones are highly tree-like, as these
+recombination events involve minor, local topological rearrangements
+where recombination only occurs between close relatives.
+% sometimes even within a single lineage such that the parent
+% lineages above a recombination event coalesce with each other before involvement with
+% any other part of the phylogeny.
+These local reticulations could either represent
 real but previously unobserved recombination events between closely related strains,
-or could be artifacts of the current \texttt{sc2ts} method and its parameterization.
+or could be artefacts of the current \texttt{sc2ts} method and its parameterization.
 
@@
 library (Section~\ref{sec:args}) are a joint estimate of the processes
-of inheritance, recombination and mutation. Unlike most approaches to
+of genetic inheritance, recombination and mutation. Unlike most approaches to
 phylogenetic analysis, mutations are included in the \texttt{tskit}
@@
 The same idea has recently been used to represent
-SARS-CoV-2 data in UShER's ``mutation annotated tree'' format~\citep{Turakhia2022-it}.
+SARS-CoV-2 data in UShER's ``mutation annotated tree''
+format~\citep{Turakhia2021-ur}.
 
@@
 cases we have a large number of mutations, and a majority of these
-(Wide ARG: 62.55\%, Long ARG: 72.23) are private to a single
+(Wide ARG: 62.55\%, Long ARG: 72.23\%) are private to a single
 sample, i.e.\ on terminal branches.
@@
 may be artefactual.
-The main weakness of the current version of \texttt{sc2ts} is an excess
+A significant issue with the current version of \texttt{sc2ts} is an excess
 of inferred reversion mutations, with 5.99\% of all mutations in the Wide
@@
 2021, and classified the mutations occurring along the phylogeny.
-We categorised all of single nucleotide
+We categorised all single nucleotide
 mutations in the Wide ARG by type (defined by the inherited and derived states),
@@
 Figure~\ref{fig:mutational_spectra}B shows that the mutational spectrum from the
-wide ARG (based on 448,825 mutations) matches that reported by \citet[based on
+Wide ARG (based on 448,825 mutations) matches that reported by \citet[based on
 92,344 mutations]{Yi2021-sc}. In both spectra, C-to-U mutations and G-to-U

### 2023-05-19 paper.tex

@@
 details on the majority of known recombinant lineages.
-Using the well-established and fully-featured \texttt{tskit} library,
+Using the well-established and featured-rich \texttt{tskit} library,
 the ARGs can also be stored concisely and processed efficiently

### 2023-05-19 paper.tex
Goodish versions of first three paras

@@
 different sampling strategies, and study their properties.
-One is densely sampled, and includes 1.27 million sequences
-sampled up to June 30, 2021.
-The second is more sparsely sampled,
+One contains 1.27 million sequences
+sampled up to June 30, 2021, and the second is more sparsely sampled,
 consisting of 784K sequences sampled up to June 30, 2022.
@@
 details on the majority of known recombinant lineages.
-Using the well-established and featured-rich \texttt{tskit} library,
+Using the well-established and feature-rich \texttt{tskit} library,
 the ARGs can also be stored concisely and processed efficiently
@@
 sequences---encoding the inferred reticulate ancestry,
-full alignments, and extensive metadata---requires
+genetic variation, and extensive metadata---requires
 58MB of storage,
@@
 immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
-The lack of genetic variation made recombination challenging to detect early in the SARS-CoV-2 pandemic. In late 2020, the appearance and
+Early in the SARS-CoV-2 pandemic, the lack of genetic variation
+meant that recombination had little effect.
+In late 2020, the appearance and
 spread of variants of concern (VoC), designated into classes such as Alpha and
-Delta which harboured multiple characteristic mutations, increased the
-detectability of recombination.
+Delta which harboured multiple characteristic mutations,
+created the conditions required for recombination to occur
+and recombinant strains with onward transmission were detected at
+low frequency~\citep{Jackson2021-ik}.
 More recently, the high prevalence of Omicron,
@@
 recombination~\citep{Bal2022-hq}.
-[FIXME say this better and add citation.]
-Multiple recombinant lineages have successfully
-established and spread to high frequency.
-
-[PARA 2: recombination is hard to detect. Quick overview of state-of-the-art.
-Finish with phylogeny-based method, RIPPLES.]
-
-[PARA 3: but you can't use a phylogeny when there's recomb. Need a joint
-model of mutation and recomb. ARGs, tskit, forward refs to ARG section for
-details.]
-
-[PARA 4: we present sc2ts based on these recent advances. Summarise
-results, with forward refs to Results sections.]
-
-% By searching for genomes containing segments
-% from different VoC designations, multiple recombinant lineages have been
-% detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha
-% and Epsilon \citep{Wertheim2022-hj}, and Alpha and
-% Delta~\citep{Sekizuka2022-xz}.
-
-% Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
-% combine specific subsets of mutations from different lineages
-% \citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
-% This approach only identifies those inter-lineage recombination events
-% resulting in the patterns investigated.
+In early 2023, multiple recombinant lineages have successfully
+established and spread to high frequency, and accounting for recombinant
+ancestry is now essential in understanding the evolution of the SARS-CoV-2
+virus.
+
+% Recombination is hard to detect
+Detecting recombination in SARS-CoV-2 is difficult, and identifying new
+recombinant strains is a time-consuming, often manual process [Citation?].
+Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
+combine specific subsets of mutations from different lineages
+\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz},
+which can only identify inter-lineage recombination events specific to those motifs.
+% % JK: I've left this out because I felt it broke the narrative up a bit, but
+% % happy to put back in. (We're still citing the paper in a few places).
 % Others have inferred recombinant
@@
 % having to pre-specify characteristic mutations for each lineage
-% \citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it} presented a
-% phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
-% over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find
-% candidate recombinants, RIPPLES first builds a phylogeny without assuming
-% recombination and then scans for long branches (i.e., carrying many mutations)
-% and then determines if these candidates would be better explained by
-% recombination. Specifically, RIPPLES exhaustively breaks each candidate
-% sequence into segments and attempts to find more parsimonious placements for
-% each segment on the phylogeny. If such placements are found, the sequence is
-% identified as a putative recombinant. Although it enables rapid searches for
-% genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
-% accounts for only mutations, treats recombinants \textit{post hoc}, and is an
-% incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
-% As noted by the authors, a \textit{post hoc} treatment of recombination is
-% possible when recombinant lineages are rare and leave few descendants. However,
-% this is becoming increasingly untenable with the proliferation of recombinant
-% lineages; for example, more than half of the sequences sampled in February 2023
-% are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
-% future evolution of SARS-CoV-2 is likely to involve multiple sequential
-% recombination events on top of existing recombinant lineages, creating a highly
-% reticulated genealogy.
+% \citep{Ignatieva2021-rg}.
+More recently, \cite{Turakhia2022-it} presented a
+phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
+over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find
+candidate recombinants, RIPPLES first builds a phylogeny without assuming
+recombination and then scans for long branches (i.e., carrying many mutations)
+and then determines if these candidates would be better explained by
+recombination. Specifically, RIPPLES exhaustively breaks each candidate
+sequence into segments and attempts to find more parsimonious placements for
+each segment on the phylogeny. If such placements are found, the sequence is
+identified as a putative recombinant. Although it enables rapid searches for
+genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
+accounts for only mutations, treats recombinants \textit{post hoc}, and is an
+incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
+As noted by the authors, a \textit{post hoc} treatment of recombination is
+possible when recombinant lineages are rare and leave few descendants. However,
+the proliferation of recombinant lineages is making this increasingly
+untenable; for example, more than half of the sequences sampled in February 2023
+are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
+future evolution of SARS-CoV-2 is likely to involve multiple sequential
+recombination events on top of existing recombinant lineages, creating a highly
+reticulated ancestry.
+
+% PARA 3: but you can't use a phylogeny when there's recomb. Need a joint
+% model of mutation and recomb. ARGs, tskit, forward refs to ARG section for
+% details.
+
+It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
+and affects the results of downstream analysis, e.g., inference of
+selection~\citep{Anisimova2003-vr}. Standard phylogenetic methods do
+not account for recombination
+\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}, and there is
+no standard method for incorporating the effects of recombination into
+phylogenetic analysis.
+Ancestral Recombination Graphs (ARGs) are a means of describing such
+network-like ancestry~\citep{Griffiths1981-lw,Gusfield2014-qw}, but
+until recently lacked software support and sufficiently scalable
+inference methods to be of practical use.
+However, approaches to infer ARGs now exist that can scale to tens of
+thousands of human genomes and beyond
+\citep{Speidel2019-yh,Kelleher2019-ba,Schaefer2021-yg,Zhang2023-lf}, and which
+deal with levels of repeated recombination far in excess of those seen in viral
+phylogenies. The ``succinct tree sequence'' is an ARG data structure
+which has led to significant computational advances across a range
+of
+applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,Ralph2020-efficiently,
+Wohns2022-th}, and the supporting \texttt{tskit} software library is now widely
+is now widely used in population genetic applications.
+The methods in \texttt{tskit} have been developed to support millions of
+whole human genomes~\citep{Kelleher2019-ba}, and so it is particularly well suited
+to representing large SARS-CoV-2 genealogies,
+which currently encompasses over 15 million sequences in the GISAID database
+\citep{Shu2017-hp}.
+See Section~\ref{sec:args} for more details on ARGs and the succinct
+tree sequence data structure.
+
+[PARA 4: we present sc2ts based on these recent advances. Summarise
+results, with forward refs to Results sections.]
 
 % %% ARGs are good for *representing* viral genealogies
-
-% It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
-% and affects the results of downstream analysis, e.g., inference of
-% selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
-% not account for recombination
-% \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
-% genetic ancestries of recombining lineages can be represented by ancestral
-% recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
-% capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
-% development of the ``succinct tree sequence'' data
-% structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
-% encoding of large ARGs. This representation should be well-suited to storing
-% large recombinant viral genealogies such as SARS-CoV-2,
-% which currently encompasses over 15 million sequences in the GISAID database
-% \citep{Shu2017-hp}). Moreover the \texttt{tskit} software library, which implements this
-% data structure, enables efficient manipulation of the stored genealogies and is
-% integrated with a suite of open-source software tools to simulate, analyse, and
-% infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.
 
@@
 Loading these ARGs takes less than a second, and they can be interactively
-analysed using Jupyter notebooks on a standard laptop. The majority
+analysed using Jupyter notebooks~\citep{Kluyver2016-jupyter}
+on a standard laptop. The majority
 of the analyses in this preprint can be carried out in seconds
@@
 describe not just realisations of this model, but to any
-recombinant genetic ancestry~\citep{Minichiello2006-mapping}.
+recombinant genetic ancestry~\citep{Minichiello2006-mapping,Zhang2023-lf}.
 While there is some subtlety in the details~\citep{Wong2023-efficient},
@@
 reticulate genetic ancestry of a sample of sequences under
-the influence of recombination. This definition encompasses various types of graphs traditionally described using the broader term of phylogenetic networks.
+the influence of recombination.
+This definition encompasses various types of graphs traditionally
+described using the broader term of phylogenetic networks.
 
@@
 Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}.
-We emphasise that this ecosystem for storing and manipulating ARGs can generally be used to efficiently record and analyse SARS-CoV-2 genealogies reconstructed using other methods, not only the new approach that we describe here.
+This ecosystem for storing and manipulating ARGs can
+generally be used to efficiently record and analyse SARS-CoV-2 genealogies reconstructed
+using other methods, not only the new approach that we describe here.
 

### 2023-05-19 paper.tex
Reasonable version of the intro

@@
 
-[PARA 4: we present sc2ts based on these recent advances. Summarise
-results, with forward refs to Results sections.]
-
-% %% ARGs are good for *representing* viral genealogies
-
-% %% ARG inference is hard
-
-% Inferring recombinant phylogenies has been a long-standing computational challenge.
-% However, approaches to infer ARGs now exist that can scale to tens of
-% thousands of human genomes and beyond
-% \citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}, and which
-% deal with levels of repeated recombination far in excess of those seen in viral
-% phylogenies. For example, using over 90 million variable sites
-% \citet{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands
-% of modern and ancient human genomes, accounting for millions of recombination
-% events (over 5 million ``local trees''). To achieve this, \cite{Wohns2022-th}
-% took advantage of a highly efficient Hidden Markov Model (HMM) implementation
-% of the Li and Stephens (LS) model \citep{Li2003-ib} to infer copying paths between
-% genomes \citep{Kelleher2019-ba}.
-
-% % What is new here?
-
-% Here, we present a preliminary version of a novel scalable methodology
-% (\texttt{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
-% and additionally takes advantage of known collection dates to infer ARGs from millions
-% of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
-% we demonstrate that \texttt{sc2ts} can be used to simultaneously (1) detect putative SARS-CoV-2
-% recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
-% (3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record
-% of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
-% \texttt{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
-% \texttt{Sc2ts} is a promising method to automate detection of SARS-CoV-2 recombinants
-% and to jointly model recombinational and mutational processes shaping the genomic
-% diversity of SARS-CoV-2 at scale.
+% [PARA 4: we present sc2ts based on these recent advances. Summarise
+% results, with forward refs to Results sections.]
+Here we present a preliminary version of \texttt{sc2ts}, a novel method
+for inferring ARGs for SARS-CoV-2 at pandemic scale in real-time.
+Building on the open-source \texttt{tskit} library, the method infers
+a joint model of mutation, genetic inheritance and recombination, which
+can be conveniently and efficiently analysed using standard data science
+tools. Inference is based on incrementally adding batches of sequences
+based on their collection dates, and proceeds in three phases. Firstly,
+likely attachment nodes in the current ARG
+(or paths, if inferred to be a recombinant) are
+chosen for each of the samples in the daily batch using the
+Li and Stephen's model (Section~\ref{sec:ls}). Then, since many samples
+in a daily batch can share an attachment path,
+we infer trees for each of the clusters separately using standard
+methods (Section~\ref{sec:sample-cluster-tree-inference}).
+Finally, we attach the trees for these sample-clusters to the current
+ARG, and apply some parsimony heuristics to address issues
+introduced by the inherent greediness of the strategy
+(Section~\ref{sec:parsimony-heuristics}).
+Using the current preliminary version of \texttt{sc2ts} we infer two
+large ARGs (1,265,685 and 657,239 samples, respectively), and study
+the properties of these ARGs to illustrate the power of the method
+and to inform subsequent development. We find that these
+ARGs accurately reflect known patterns of phylogenetic relationships
+between samples (Section~\ref{sec:backbone_phylogeny}),
+and mutational spectra (Section~\ref{sec:mutation_spectrum})
+while simultaneously identifying the majority of known recombinant
+lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),
+with an unprecedented degree of precision in the
+genomic location of recombination breakpoints
+(Section~\ref{sec:breakpoint_intervals})
+and the relationship between parental sequences
+(Section~\ref{sec:parent_divergence}).
+We hope that these benefits of an accurate joint model of
+mutation, genetic inheritance and recombination, will generate community
+interest and development of the \texttt{sc2ts} method, and
+more generally in applying the efficient and mature software
+of the \texttt{tskit} ecosystem to pandemic-scale SARS-CoV-2 data.
 
@@
 \subsection{Inferred ARGs}
-For this preliminary study our goals are to illustrate the utility of
+The goals of this preliminary study are to illustrate the utility of
 \texttt{sc2ts}, and to closely study the properties of the
@@
 For example, position 28,271 (located in the ORF8-N intergenic
-region) has the largest number of mutatations and corresponds to the
+region) has the largest number of mutations and corresponds to the
 left endpoint of 9 intervals and the right of 57.
@@
 
-Sc2ts is a method for inferring ARGs from densely sampled pandemic-scale data
+Sc2ts (pronounced ``scoots'', optionally) is a method for inferring
+ARGs from densely sampled pandemic-scale data
 in real time, in which recombination occurs at a low but significant rate.
@@
 advantage of the analysis tools in the burgeoning PyData ecosystem.
+(It is possible to access the toolkit from R via the \texttt{reticulate}
+package, and the \texttt{slendr} library~\citep{Petr2022-slendr}
+also provides some native R support. A full R interface would be a
+valuable addition to the ecosystem.)
 Tskit is mature software, widely used in population genetics, and
@@
 Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}.
-This ecosystem for storing and manipulating ARGs can
+It is important to note that
+this ecosystem for storing and manipulating ARGs can
 generally be used to efficiently record and analyse SARS-CoV-2 genealogies reconstructed

### 2023-05-19 paper.tex
Reasonable version of the intro

@@
 For example, position 28,271 (located in the ORF8-N intergenic
-region) has the largest number of mutations and corresponds to the
+region) has the largest number of mutatations and corresponds to the
 left endpoint of 9 intervals and the right of 57.

### 2023-05-19 paper.tex
Updates from Yan

--- 2023-05-19 paper.tex
@@
 different sampling strategies, and study their properties.
-One contains 1.27 million sequences
-sampled up to June 30, 2021, and the second is more sparsely sampled,
+One is densely sampled, and includes 1.27 million sequences
+sampled up to June 30, 2021.
+The second is more sparsely sampled,
 consisting of 784K sequences sampled up to June 30, 2022.
@@
 details on the majority of known recombinant lineages.
-Using the well-established and feature-rich \texttt{tskit} library,
+Using the well-established and featured-rich \texttt{tskit} library,
 the ARGs can also be stored concisely and processed efficiently
@@
 sequences---encoding the inferred reticulate ancestry,
-genetic variation, and extensive metadata---requires
+full alignments, and extensive metadata---requires
 58MB of storage,
@@
 immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
-Early in the SARS-CoV-2 pandemic, the lack of genetic variation
-meant that recombination had little effect.
-In late 2020, the appearance and
+The lack of genetic variation made recombination challenging to detect early in the SARS-CoV-2 pandemic. In late 2020, the appearance and
 spread of variants of concern (VoC), designated into classes such as Alpha and
-Delta which harboured multiple characteristic mutations,
-created the conditions required for recombination to occur
-and recombinant strains with onward transmission were detected at
-low frequency~\citep{Jackson2021-ik}.
+Delta which harboured multiple characteristic mutations, increased the
+detectability of recombination.
 More recently, the high prevalence of Omicron,
@@
 recombination~\citep{Bal2022-hq}.
-In early 2023, multiple recombinant lineages have successfully
-established and spread to high frequency, and accounting for recombinant
-ancestry is now essential in understanding the evolution of the SARS-CoV-2
-virus.
-
-% Recombination is hard to detect
-Detecting recombination in SARS-CoV-2 is difficult, and identifying new
-recombinant strains is a time-consuming, often manual process [Citation?].
-Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
-combine specific subsets of mutations from different lineages
-\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz},
-which can only identify inter-lineage recombination events specific to those motifs.
-% % JK: I've left this out because I felt it broke the narrative up a bit, but
-% % happy to put back in. (We're still citing the paper in a few places).
+[FIXME say this better and add citation.]
+Multiple recombinant lineages have successfully
+established and spread to high frequency.
+
+[PARA 2: recombination is hard to detect. Quick overview of state-of-the-art.
+Finish with phylogeny-based method, RIPPLES.]
+
+[PARA 3: but you can't use a phylogeny when there's recomb. Need a joint
+model of mutation and recomb. ARGs, tskit, forward refs to ARG section for
+details.]
+
+[PARA 4: we present sc2ts based on these recent advances. Summarise
+results, with forward refs to Results sections.]
+
+% By searching for genomes containing segments
+% from different VoC designations, multiple recombinant lineages have been
+% detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha
+% and Epsilon \citep{Wertheim2022-hj}, and Alpha and
+% Delta~\citep{Sekizuka2022-xz}.
+
+% Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
+% combine specific subsets of mutations from different lineages
+% \citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
+% This approach only identifies those inter-lineage recombination events
+% resulting in the patterns investigated.
 % Others have inferred recombinant
@@
 % having to pre-specify characteristic mutations for each lineage
-% \citep{Ignatieva2021-rg}.
-More recently, \cite{Turakhia2022-it} presented a
-phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
-over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find
-candidate recombinants, RIPPLES first builds a phylogeny without assuming
-recombination and then scans for long branches (i.e., carrying many mutations)
-and then determines if these candidates would be better explained by
-recombination. Specifically, RIPPLES exhaustively breaks each candidate
-sequence into segments and attempts to find more parsimonious placements for
-each segment on the phylogeny. If such placements are found, the sequence is
-identified as a putative recombinant. Although it enables rapid searches for
-genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
-accounts for only mutations, treats recombinants \textit{post hoc}, and is an
-incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
-As noted by the authors, a \textit{post hoc} treatment of recombination is
-possible when recombinant lineages are rare and leave few descendants. However,
-the proliferation of recombinant lineages is making this increasingly
-untenable; for example, more than half of the sequences sampled in February 2023
-are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
-future evolution of SARS-CoV-2 is likely to involve multiple sequential
-recombination events on top of existing recombinant lineages, creating a highly
-reticulated ancestry.
-
-% PARA 3: but you can't use a phylogeny when there's recomb. Need a joint
-% model of mutation and recomb. ARGs, tskit, forward refs to ARG section for
-% details.
-
-It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
-and affects the results of downstream analysis, e.g., inference of
-selection~\citep{Anisimova2003-vr}. Standard phylogenetic methods do
-not account for recombination
-\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}, and there is
-no standard method for incorporating the effects of recombination into
-phylogenetic analysis.
-Ancestral Recombination Graphs (ARGs) are a means of describing such
-network-like ancestry~\citep{Griffiths1981-lw,Gusfield2014-qw}, but
-until recently lacked software support and sufficiently scalable
-inference methods to be of practical use.
-However, approaches to infer ARGs now exist that can scale to tens of
-thousands of human genomes and beyond
-\citep{Speidel2019-yh,Kelleher2019-ba,Schaefer2021-yg,Zhang2023-lf}, and which
-deal with levels of repeated recombination far in excess of those seen in viral
-phylogenies. The ``succinct tree sequence'' is an ARG data structure
-which has led to significant computational advances across a range
-of
-applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,Ralph2020-efficiently,
-Wohns2022-th}, and the supporting \texttt{tskit} software library is now widely
-is now widely used in population genetic applications.
-The methods in \texttt{tskit} have been developed to support millions of
-whole human genomes~\citep{Kelleher2019-ba}, and so it is particularly well suited
-to representing large SARS-CoV-2 genealogies,
-which currently encompasses over 15 million sequences in the GISAID database
-\citep{Shu2017-hp}.
-See Section~\ref{sec:args} for more details on ARGs and the succinct
-tree sequence data structure.
-
-[PARA 4: we present sc2ts based on these recent advances. Summarise
-results, with forward refs to Results sections.]
+% \citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it} presented a
+% phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
+% over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find
+% candidate recombinants, RIPPLES first builds a phylogeny without assuming
+% recombination and then scans for long branches (i.e., carrying many mutations)
+% and then determines if these candidates would be better explained by
+% recombination. Specifically, RIPPLES exhaustively breaks each candidate
+% sequence into segments and attempts to find more parsimonious placements for
+% each segment on the phylogeny. If such placements are found, the sequence is
+% identified as a putative recombinant. Although it enables rapid searches for
+% genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
+% accounts for only mutations, treats recombinants \textit{post hoc}, and is an
+% incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
+% As noted by the authors, a \textit{post hoc} treatment of recombination is
+% possible when recombinant lineages are rare and leave few descendants. However,
+% this is becoming increasingly untenable with the proliferation of recombinant
+% lineages; for example, more than half of the sequences sampled in February 2023
+% are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
+% future evolution of SARS-CoV-2 is likely to involve multiple sequential
+% recombination events on top of existing recombinant lineages, creating a highly
+% reticulated genealogy.
 
 % %% ARGs are good for *representing* viral genealogies
+
+% It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
+% and affects the results of downstream analysis, e.g., inference of
+% selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
+% not account for recombination
+% \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
+% genetic ancestries of recombining lineages can be represented by ancestral
+% recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
+% capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
+% development of the ``succinct tree sequence'' data
+% structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
+% encoding of large ARGs. This representation should be well-suited to storing
+% large recombinant viral genealogies such as SARS-CoV-2,
+% which currently encompasses over 15 million sequences in the GISAID database
+% \citep{Shu2017-hp}). Moreover the \texttt{tskit} software library, which implements this
+% data structure, enables efficient manipulation of the stored genealogies and is
+% integrated with a suite of open-source software tools to simulate, analyse, and
+% infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.
 
@@
 Loading these ARGs takes less than a second, and they can be interactively
-analysed using Jupyter notebooks~\citep{Kluyver2016-jupyter}
-on a standard laptop. The majority
+analysed using Jupyter notebooks on a standard laptop. The majority
 of the analyses in this preprint can be carried out in seconds
@@
 describe not just realisations of this model, but to any
-recombinant genetic ancestry~\citep{Minichiello2006-mapping,Zhang2023-lf}.
+recombinant genetic ancestry~\citep{Minichiello2006-mapping}.
 While there is some subtlety in the details~\citep{Wong2023-efficient},
@@
 reticulate genetic ancestry of a sample of sequences under
-the influence of recombination.
-This definition encompasses various types of graphs traditionally
-described using the broader term of phylogenetic networks.
+the influence of recombination. This definition encompasses various types of graphs traditionally described using the broader term of phylogenetic networks.
 
@@
 Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}.
-This ecosystem for storing and manipulating ARGs can
-generally be used to efficiently record and analyse SARS-CoV-2 genealogies reconstructed
-using other methods, not only the new approach that we describe here.
+We emphasise that this ecosystem for storing and manipulating ARGs can generally be used to efficiently record and analyse SARS-CoV-2 genealogies reconstructed using other methods, not only the new approach that we describe here.
 

### 2023-05-19 paper.tex
First pass at the discussion

--- 2023-05-19 paper.tex
@@
 \section{Discussion}
-The COVID-19 pandemic continues to be a global emergency,
+% PARA 1. SARS-CoV-2 is still important, and the vast hordes of data provide an
+% unprecedented opportunity to study the evolution of a virus. Even if the
+% public health emergency is officially over, it doesn't mean it's not coming
+% back and we should be prepared. Recombination is now a vital factor, and
+% present a way in which it can be incorporated. Our results are accurate.
+
+The COVID-19 pandemic continues to be a global emergency [DOES IT? Didn't the
+WHO downgrade recently?]
 with persistent high levels of infection. This high prevalence has allowed the
@@
 distinct recombinants to emerge and spread.
-
-In this proof-of-concept study, we demonstrate the feasibility of
-reconstructing a large ARG from over one million SARS-CoV-2 genomes, a feat
-that is simply not possible using other available ARG inference methods
-\citep{Rasmussen2014-el,Ignatieva2021-rg, Speidel2019-yh}. We have accomplished
-this feat by building on top of recent advances in storing and processing large
-ARGs (tskit) and inferring genealogical relationships efficiently at scale
-(tsinfer). The highly compact representation of the ARGs as succinct tree
-sequences, coupled with vectorised computing technologies, enables the large
-SARS-CoV-2 ARGs to be loaded in less than one second in a Jupyter notebook and
-allows for each SARS-CoV-2 genome to be stored in less than eleven bytes of
-data. We believe that further development of the sc2ts methodology and software
-can create a powerful tool (1) to efficiently detect emerging or previously
-overlooked recombinants in SARS-CoV-2; (2) to jointly model the mutational and
-recombinational processes shaping global SARS-CoV-2 genomic diversity; and (3)
-to perform a broad variety of evolutionary analyses of SARS-CoV-2 genomes with
-unprecedented global sequencing coverage.
-
-ARGs are rich genealogical structures that contain information about the
-genetic relationships among sampled genomes in the presence of recombination.
-Here, we have conducted several analyses of the preliminary ARGs to show that
-the ARGs built using sc2ts can (1) recapitulate the phylogenetic relationships
-among major SARS-CoV-2 lineages; (2) recover biologically plausible
-recombination signals in SARS-CoV-2 genomes; and (3) model mutation events in
-SARS-CoV-2 evolutionary history.
-
-ARGs reconstructed using sc2ts capture the phylogenetic relationships of major
-SARS-CoV-2 lineages. Our results show that the backbone topology of a densely
-sampled wide ARG inferred using sc2ts agrees well with the backbone topology of
-a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference
-package. However, we have noted several areas of differences (e.g., poorly
-calibrated branch lengths), which are the focus of future development of sc2ts.
-We believe that some of these differences can be reduced by improving and
-refining our approach and strategy to handle samples with erroneously reported
-collection dates (i.e., time travellers that have passed the maximum submission
-delay of 30 days) and to modify the topology of daily trees.
-
-Our results show that sc2ts identifies the early recombinants (pre-2022)
-previously identified by \cite{Jackson2021-ik}. Sc2ts also identifies many
-Pango recombinants, which are designated after community perusal of supporting
-genomic evidence. We have found several previously unreported and overlooked
-recombinant sequences in the wide ARG. After further improvements and
-enhancements, we will use sc2ts to build a comprehensive ARG to search for
-recombination signals among all the 15 million and counting SARS-CoV-2 genomes
-available.
-
-``The rate of putative recombination breakpoints is about three times higher towards the 3' of the change point than the 5' interval.'' Consistent with this, the breakpoint intervals of the putative recombinants identified here tend to occur on the 3’ end of the SARS-CoV-2 genome.
-
-\section{Improvements}
-Currently, sc2ts is in development. Below, we propose several ways to potentially improve or enhance sc2ts. We welcome suggestions and ideas from experts.
-
-\begin{itemize}
-\item A potential source of error in sc2ts ARGs comes from incorrectly reported collection dates of the samples, which can cause sample genomes to be attached to the wrong parts of the ARGs.
-Thus far, we have attempted to correct the most offending errors in the ARGs by filtering out ``time travellers'', but the sampling time can still be wrong by days or tens of days.
-***describe possible approaches to correct this?***
-\item To avoid time travellers, we used a maximum submission delay, but this will systematically bias against submissions from countries with greater submission delays (Kalia et al.).
-An area of future work is to add back the excluded sequences to a built ARG to assess which ones are likely true time travellers.
-\item We have used a single mismatch value in all the HMM runs, set so that having three mutations is as likely as having a single recombination.
-This parameter choice is arbitrary, and higher values may reduce the number of origins for those Pango lineages that were inferred to descend from more than one recombinant ancestor.
-In addition, because both the mutation rate and recombination rate varies across the SARS-CoV-2 genome, it may be helpful to allow the mismatch value to vary along the genome.
-\item By treating mutations equally in the current HMM implementation, we implicitly assume that the rates of different types of mutations are identical.
-However, as reported previously \citep{Yi2021-sc}, there is a high degree of mutational asymmetry in SARS-CoV-2.
-To account for mutational biases, more sophisticated nucleotide substitution models (e.g., allowing for different rates of transition and transversion) can be incorporated into HMM likelihood calculations.
-\item Our current masking strategy handles indels as missing data.
-Doing this can remove information helpful to detect recombinants.
-For example, sequences of the Pango recombinant XP were not identified as recombinants, because XP’s characteristic multibase deletion was masked.
-\item Daily trees are built using the UPGMA algorithm.
-This entails an extra step to ``un-resolve'' daily trees in parts of the trees with no phylogenetic resolution.
-Using a maximum-parsimony algorithm to build daily trees would leave these "soft polytomies" in, removing the need for the un-resolving step.
-\end{itemize}
-
-\section{Conclusions}
-The global understanding of viral evolution during the COVID-19 pandemic is
-unprecedented, with over 15 million sequences shared to date. We envisage that
-once fully developed and tested, sc2ts methodology and software may provide a
-unified computational framework to study the evolution and epidemiology of
-SARS-CoV-2 at pandemic scale. The evolutionary analyses presented herein
-exemplify the numerous applications possible using the ancestral recombination
-graph inferred by sc2ts. We invite bioinformaticians, evolutionary biologists,
-and virologists to join us and the broader tskit community to build the most
-complete representation of the evolutionary history of SARS-CoV-2. Moreover,
-while we have focussed on the particular application to SARS-CoV-2, the methods
-we describe are readily applicable to the analysis of many other viral
-pathogens. Solving the challenges in building and analysing large viral
-genealogies would therefore contribute to the important goal of improving
-preparedness for potential future pandemics.
+[Handling recombination properly as an ongoing process is vital. Our results
+here show that recombination can be incorporated into a cohesive joint
+model of viral evolution. We basically recapitulate what is known, plus
+there's a whole bunch of other stuff that we haven't explored because
+we're in the validation stage.]
+
+% PARA 2. Here are things that could be improved in the short term. Here's a
+% few more, longer term ideas.
+Nonetheless, \texttt{sc2ts} is currently ``alpha'' quality software, and
+there are some relatively straightforward changes that could be made to
+improve inference quality. Firstly, the pipeline used to identify and
+mask erroneous sites in the input alignments is simplistic, and, among
+other issues, results in multi-base indels being marked as missing data
+(Section~\ref{sec:data_preprocessing}). A more sophisticated
+approach~\citep[e.g.][]{Aksamentov2021-hj} would
+likely yield significant improvements, and reduce the effect sites with
+artefactually high levels of mutation (e.g., site 28271;
+Section~\ref{sec:breakpoint_intervals}).
+Using a pre-existing tree built using state-of-the-art phylogenetics methods
+for the early stages of the pandemic (Section~\ref{sec:backbone_phylogeny}),
+and minor adaptations to standard node-dating methods
+(Section~\ref{sec:node_dating}) should help resolve the most notable issues with
+the inferred backbone phylogeny (Figure~\ref{fig:cophylogeny}).
+Trees constructed from daily sample clusters have a surprisingly large
+influence on the overall ARG topology
+(Section~\ref{sec:sample-cluster-tree-inference}), and so using a more
+sophisticated tree building approach should yield clear improvements.
+The unrealistically large number of reversion mutations
+(Section~\ref{sec:mutation_spectrum}) may be reduced by
+improvements to the current parsimony
+hueristics (Section~\ref{sec:parsimony-heuristics}).
+A major source of errors are the ``time-traveller'' samples, whose
+provided collection dates are months (or years) too early
+(Section~\ref{sec:filtering_time_travellers}).
+% Hmm, not sure about this. Wouldn't this also rule out new major lineages
+% also (which tend to show up with a bunch of new mutations?) I guess it
+% depends on the threshold, and we could set it according to previous
+% experience with new lineages?
+While it is unclear how
+we might solve this problem in general, some simple expedients such
+as filtering out sequences that exceed a given cost in the LS HMM
+(i.e., number of mutations and recombination switches) may work well
+in practise. Such an approach would also reduce the impact of sequences
+with high levels of sequencing error (which contribute a large number of
+mutations, currently).
+Taken together, these and other relatively minor improvements should
+enable inference over much larger subsets of the dataset, and
+give a clearer picture of the combined processes of recombination and
+mutation over the pandemic so far.
+
+% What are the more blue-sky type things? Basically these are things we don't
+% really want to do for the first "real" version of sc2ts.
+An attractive feature of \texttt{sc2ts} is that the most difficult part of
+the inference problem---finding likely recombinant paths through the existing ARG
+for new samples---is solved exactly under a well-defined statistical model,
+using established HMM methodology (Section~\ref{sec:ls}).
+The implementation currently uses
+a single, arbitrarily chosen, maximum likelihood path via the
+Viterbi algorithm, but there are numerous ways in which the HMM machinery
+could be extended in order to explore the set of possible paths, or to
+quantify the uncertainty around it. Similarly, the current parameterisation
+of the HMM with a single mismatch ratio is simplistic, and it would
+be straightforward to condition on per-site mutation rates. (Per nucleotide
+state transition, is also possible, with some additional development.)
+Recombination breakpoints for the ARG are currently inserted at the
+right-most extent of the possible interval
+(Section~\ref{sec:breakpoint_intervals}). More likely locations for the
+breakpoint could be chosen within the interval, for example based
+on the sequence motifs.
+
+% PARA 3. Handling the data is hard, so methods for making this easier
+% are important. Classical phylo approaches don't scale to millions.
+% The ability to quickly develop new analyses, at scale, will be crucial.
+% Sc2ts is largely composed of other parts being reused, with mainly being
+% concerned with masking input alignments. We wrote no C at all. This kind
+% of code reuse and ease of analysis is a major bonus.
+The vast volumes of whole genome sequence data generated during the pandemic
+has presented classical phylogenetics methods and software with major difficulties.
+Standard interchange formats such as FASTA, Newick and VCF were simply not designed
+to deal with millions of samples, and their limitations have come sharply
+into focus~\citep{Turakhia2021-ur,de2023maximum}.
+[Finish this up with a discussion of how easy it is to do stuff with our
+ARGS, how awesome tskit is, and how little code there is in sc2ts.
+Being able to quickly develop new methods by putting together high
+performance, high quality components is an essential element of
+preparedness for futture pandemics.]
+
+% The COVID-19 pandemic continues to be a global emergency,
+% with persistent high levels of infection. This high prevalence has allowed the
+% proliferation of many variants, with more than 600 Pango-designated lineages
+% circulating globally in the last three months (January to March, 2023;
+% https://gisaid.org/; access on March 27, 2023). High prevalence also brings a
+% higher risk of coinfection, increasing opportunities for new phenotypically
+% distinct recombinants to emerge and spread.
+
+% In this proof-of-concept study, we demonstrate the feasibility of
+% reconstructing a large ARG from over one million SARS-CoV-2 genomes, a feat
+% that is simply not possible using other available ARG inference methods
+% \citep{Rasmussen2014-el,Ignatieva2021-rg, Speidel2019-yh}. We have accomplished
+% this feat by building on top of recent advances in storing and processing large
+% ARGs (tskit) and inferring genealogical relationships efficiently at scale
+% (tsinfer). The highly compact representation of the ARGs as succinct tree
+% sequences, coupled with vectorised computing technologies, enables the large
+% SARS-CoV-2 ARGs to be loaded in less than one second in a Jupyter notebook and
+% allows for each SARS-CoV-2 genome to be stored in less than eleven bytes of
+% data. We believe that further development of the sc2ts methodology and software
+% can create a powerful tool (1) to efficiently detect emerging or previously
+% overlooked recombinants in SARS-CoV-2; (2) to jointly model the mutational and
+% recombinational processes shaping global SARS-CoV-2 genomic diversity; and (3)
+% to perform a broad variety of evolutionary analyses of SARS-CoV-2 genomes with
+% unprecedented global sequencing coverage.
+
+% ARGs are rich genealogical structures that contain information about the
+% genetic relationships among sampled genomes in the presence of recombination.
+% Here, we have conducted several analyses of the preliminary ARGs to show that
+% the ARGs built using sc2ts can (1) recapitulate the phylogenetic relationships
+% among major SARS-CoV-2 lineages; (2) recover biologically plausible
+% recombination signals in SARS-CoV-2 genomes; and (3) model mutation events in
+% SARS-CoV-2 evolutionary history.
+
+% ARGs reconstructed using sc2ts capture the phylogenetic relationships of major
+% SARS-CoV-2 lineages. Our results show that the backbone topology of a densely
+% sampled wide ARG inferred using sc2ts agrees well with the backbone topology of
+% a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference
+% package. However, we have noted several areas of differences (e.g., poorly
+% calibrated branch lengths), which are the focus of future development of sc2ts.
+% We believe that some of these differences can be reduced by improving and
+% refining our approach and strategy to handle samples with erroneously reported
+% collection dates (i.e., time travellers that have passed the maximum submission
+% delay of 30 days) and to modify the topology of daily trees.
+
+% Our results show that sc2ts identifies the early recombinants (pre-2022)
+% previously identified by \cite{Jackson2021-ik}. Sc2ts also identifies many
+% Pango recombinants, which are designated after community perusal of supporting
+% genomic evidence. We have found several previously unreported and overlooked
+% recombinant sequences in the wide ARG. After further improvements and
+% enhancements, we will use sc2ts to build a comprehensive ARG to search for
+% recombination signals among all the 15 million and counting SARS-CoV-2 genomes
+% available.
+
+% ``The rate of putative recombination breakpoints is about three times higher towards the 3' of the change point than the 5' interval.'' Consistent with this, the breakpoint intervals of the putative recombinants identified here tend to occur on the 3’ end of the SARS-CoV-2 genome.
+
+% \section{Improvements}
+% Currently, sc2ts is in development. Below, we propose several ways to potentially improve or enhance sc2ts. We welcome suggestions and ideas from experts.
+
+% \begin{itemize} \item A potential source of error in sc2ts ARGs comes from
+% incorrectly reported collection dates of the samples, which can cause sample
+% genomes to be attached to the wrong parts of the ARGs. Thus far, we have
+% attempted to correct the most offending errors in the ARGs by filtering out
+% ``time travellers'', but the sampling time can still be wrong by days or tens
+% of days. ***describe possible approaches to correct this?*** \item To avoid
+% time travellers, we used a maximum submission delay, but this will
+% systematically bias against submissions from countries with greater submission
+% delays (Kalia et al.). An area of future work is to add back the excluded
+% sequences to a built ARG to assess which ones are likely true time travellers.
+% \item We have used a single mismatch value in all the HMM runs, set so that
+% having three mutations is as likely as having a single recombination. This
+% parameter choice is arbitrary, and higher values may reduce the number of
+% origins for those Pango lineages that were inferred to descend from more than
+% one recombinant ancestor. In addition, because both the mutation rate and
+% recombination rate varies across the SARS-CoV-2 genome, it may be helpful to
+% allow the mismatch value to vary along the genome. \item By treating mutations
+% equally in the current HMM implementation, we implicitly assume that the rates
+% of different types of mutations are identical. However, as reported previously
+% \citep{Yi2021-sc}, there is a high degree of mutational asymmetry in
+% SARS-CoV-2. To account for mutational biases, more sophisticated nucleotide
+% substitution models (e.g., allowing for different rates of transition and
+% transversion) can be incorporated into HMM likelihood calculations. \item Our
+% current masking strategy handles indels as missing data. Doing this can remove
+% information helpful to detect recombinants. For example, sequences of the Pango
+% recombinant XP were not identified as recombinants, because XP’s characteristic
+% multibase deletion was masked. \item Daily trees are built using the UPGMA
+% algorithm. This entails an extra step to ``un-resolve'' daily trees in parts of
+% the trees with no phylogenetic resolution. Using a maximum-parsimony algorithm
+% to build daily trees would leave these "soft polytomies" in, removing the need
+% for the un-resolving step. \end{itemize}
+
+% \section{Conclusions}
+% The global understanding of viral evolution during the COVID-19 pandemic is
+% unprecedented, with over 15 million sequences shared to date. We envisage that
+% once fully developed and tested, sc2ts methodology and software may provide a
+% unified computational framework to study the evolution and epidemiology of
+% SARS-CoV-2 at pandemic scale. The evolutionary analyses presented herein
+% exemplify the numerous applications possible using the ancestral recombination
+% graph inferred by sc2ts. We invite bioinformaticians, evolutionary biologists,
+% and virologists to join us and the broader tskit community to build the most
+% complete representation of the evolutionary history of SARS-CoV-2. Moreover,
+% while we have focussed on the particular application to SARS-CoV-2, the methods
+% we describe are readily applicable to the analysis of many other viral
+% pathogens. Solving the challenges in building and analysing large viral
+% genealogies would therefore contribute to the important goal of improving
+% preparedness for potential future pandemics.
 
@@
 
-It is important to note that the Viterbi algorithm only returns
-\emph{one of} the copying paths that maximise the likelihood under
- the given mutation and recombination parameters. There may be many
-such paths, from which we choose one arbitrarily. Also, the
-present choice of using a single mismatch-ratio parameter to control
-the likelihood of recombination vs mutation may lead to relatively
-flat likelihood spaces where many different paths have equal likelihood.
-There are many possibilities in using established HMM methodology
-to reason about and explore the space of possible matches, which may
-be a fruitful avenue for future work.
-Examples include a stochastic traceback through the collection of paths at the global optimum to glean further information about our the likelihood surface, and determine whether there are downstream implications for our conclusions.
-Here, we have considered the Viterbi algorithm to make statements about the most likely paths through the data.
-The machinery used here can be modified to run the forwards and backwards algorithms which determine the probability of observing a focal sequence, integrating over all possible paths through the data, under LS model [cite D-Palms].
-This presents an opportunity to estimate parameters of interest while exploiting the tree-structure to gain log based speed-ups in exact their exact implementation.
+It is important to note that the Viterbi algorithm only returns \emph{one of}
+the copying paths that maximise the likelihood under the given mutation and
+recombination parameters. There may be many such paths, from which we choose
+one arbitrarily. Also, the present choice of using a single mismatch-ratio
+parameter to control the likelihood of recombination vs mutation may lead to
+relatively flat likelihood spaces where many different paths have equal
+likelihood. There are many possibilities in using established HMM methodology
+to reason about and explore the space of possible matches, which may be a
+fruitful avenue for future work. Examples include a stochastic traceback
+through the collection of paths at the global optimum to glean further
+information about our the likelihood surface, and determine whether there are
+downstream implications for our conclusions. Here, we have considered the
+Viterbi algorithm to make statements about the most likely paths through the
+data. The machinery used here can be modified to run the forwards and backwards
+algorithms which determine the probability of observing a focal sequence,
+integrating over all possible paths through the data, under LS model [cite
+D-Palms]. This presents an opportunity to estimate parameters of interest while
+exploiting the tree-structure to gain log based speed-ups in exact their exact
+implementation.
 

### 2023-05-22 paper.tex
Ana's updates

@@
 divergence occurs (during onward transmission), if the two parent sequences
-are similar there is no way to distinguish them. Here we define the
-interval for a particular breakpoint as the genome coordinates over which the
+are similar there is no way to 
+% distinguish them. 
+accurately specify the breakpoint.
+Here we define the
+interval around a particular breakpoint as the genome coordinates over which the
 sequences for the left and right parent nodes are identical. The right-hand
 extreme of the breakpoint interval is chosen by the LS HMM Viterbi algorithm
-(Section~\ref{sec:ls}) and the left bound is then derived by
-sequence comparison of the parents.
+(Section~\ref{sec:ls}), and the left endpoint is then derived by directly
+comparing the parent sequences.
 
@@
 
-While the detailed interpretation of Figure~\ref{fig:breakpoint-distribution}
-requires further study, a few key properties stand out.[TODO, write this
-better. ] Firstly, in several cases there is a clear relationship between
+Emphasising that further work is required to filter out spuriously inferred
+recombination events and improve the accuracy of the inferred breakpoint
+intervals, we can draw two preliminary conclusions from 
+Figure~\ref{fig:breakpoint-distribution}.
+% While the detailed interpretation of Figure~\ref{fig:breakpoint-distribution}
+% requires further study, a few key properties stand out.[TODO, write this
+% better. ] 
+Firstly, in several cases there is a clear relationship between
 gene boundaries and the breakpoint intervals. For example, ORF7a and OR7b
@@
 
-It is important to note that the exact endpoints of intervals can be
+It is important to note that the precise endpoints of intervals can be
 somewhat arbitrary, since they are defined by exact comparisons
-of a recombinant's parents. The presence of problematic sites that are enriched
-for errors will therefore tend to artificially truncate breakpoint intervals
-at these sites.
-It is therefore helpful to compare the number of interecting intervals
+of a recombinant's parents. 
+% The presence of problematic sites that are enriched
+% for errors will therefore tend to artificially truncate breakpoint intervals
+% at these sites.
+% It is therefore helpful to compare the number of interecting intervals
+% with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
+% For example, position 28,271 (located in the ORF8-N intergenic
+% region) has the largest number of mutations and corresponds to the
+% left endpoint of 9 intervals and the right of 57.
+% The 7,572 mutations  (including 5,782 insertions and 1,605 deletions)
+% at this site are a likely indicator that this
+% homopolymeric region is prone to sequencing errors
+% and should be included in the list of ``problematic sites'' that
+% are excluded from analysis (Section~\ref{sec:data_preprocessing}).
+% There can be good biological reasons for this too though
+% However, hypermutable sites such as 27,972 \citep{Jungreis2021-dh}
+% have the same effect of truncating intervals. At this site we have
+% 770 mutations (including 425 C$>$T, 314 T$>$C) with potential signals
+% of polymorphism maintained by selection [FIXME!!], and 58 intervals
+% with endpoints within one base of this site.
+Thus, breakpoint intervals will tend to be truncated at sites which are
+hypermutable, either due to being prone to sequencing errors or
+for biological reasons such as selection pressure.
+It is therefore helpful to compare the number of intersecting intervals
 with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
-For example, position 28,271 (located in the ORF8-N intergenic
-region) has the largest number of mutations and corresponds to the
-left endpoint of 9 intervals and the right of 57.
+For instance, position 28,271 (located in the ORF8-N intergenic
+region) has the largest number of mutations and appears as an endpoint
+of 66 intervals.
 The 7,572 mutations (including 5,782 insertions and 1,605 deletions)
-at this site are a likely indicator that this
-homopolymeric region is prone to sequencing errors
+at this site are an indicator that this
+homopolymeric region may be prone to sequencing errors
 and should be included in the list of ``problematic sites'' that
 are excluded from analysis (Section~\ref{sec:data_preprocessing}).
-% There can be good biological reasons for this too though
-However, hypermutable sites such as 27,972 \citep{Jungreis2021-dh}
-have the same effect of truncating intervals. At this site we have
-770 mutations (including 425 C$>$T, 314 T$>$C) with potential signals
-of polymorphism maintained by selection [FIXME!!], and 58 intervals
-with endpoints within one base of this site.
+On the other hand, 58 of the breakpoint intervals have endpoints
+within one base of site 27,972, which has undergone 770 mutation events 
+(including 425 C$>$T, 314 T$>$C). The C$>$T mutation has the effect of
+truncating ORF8, and it has been posited that the truncation is neutral
+or advantageous for transmission, and disadvantageous within-host 
+\citep{Jungreis2021-dh}, suggesting that the high rate of recurrent mutation 
+at this site is due to selection.
 

### 2023-05-22 paper.tex
Minor updates to intro

@@
 immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
-Early in the SARS-CoV-2 pandemic, the lack of genetic variation
-meant that recombination had little effect.
+Early in the SARS-CoV-2 pandemic there was insufficient 
+genetic variation for meaningful recombination.
 In late 2020, the appearance and
@@
 Detecting recombination in SARS-CoV-2 is difficult, and identifying new
-recombinant strains is a time-consuming, often manual process [Citation?].
+recombinant strains is a time-consuming, manual process, subject 
+to biases [FIXME: have we a citation for how tedious this is? Also what 
+are the biases?]
 Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
@@
 % \citep{Ignatieva2021-rg}.
-More recently, \cite{Turakhia2022-it} presented a
+\cite{Turakhia2022-it} presented a
 phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
-over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find
-candidate recombinants, RIPPLES first builds a phylogeny without assuming
-recombination and then scans for long branches (i.e., carrying many mutations)
-and then determines if these candidates would be better explained by
-recombination. Specifically, RIPPLES exhaustively breaks each candidate
-sequence into segments and attempts to find more parsimonious placements for
+over ten million SARS-CoV-2 genomes, without pre-specifying motifs. 
+RIPPLES finds candidate recombinants by using an existing phylogeny
+(built assuming no recombination), and finding potential recombinants
+by  scanning for branches containing many mutations.
+It then determines if these candidates would be better explained by
+recombination by exhaustively breaking each 
+sequence into segments and attempting to find more parsimonious placements for
 each segment on the phylogeny. If such placements are found, the sequence is
@@
 applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,Ralph2020-efficiently,
-Wohns2022-th}, and the supporting \texttt{tskit} software library is now widely
+Wohns2022-th}, and the supporting \texttt{tskit} software library 
 is now widely used in population genetic applications.
@@
 Here we present a preliminary version of \texttt{sc2ts}, a novel method
-for inferring ARGs for SARS-CoV-2 at pandemic scale in real-time.
+for inferring ARGs for SARS-CoV-2 at pandemic scale, in real-time.
 Building on the open-source \texttt{tskit} library, the method infers
@@
 tools. Inference is based on incrementally adding batches of sequences
-based on their collection dates, and proceeds in three phases. Firstly,
-likely attachment nodes in the current ARG
+based on their collection dates, and proceeds in three phases. 
+First, likely attachment nodes in the current ARG
 (or paths, if inferred to be a recombinant) are
 chosen for each of the samples in the daily batch using the
-Li and Stephen's model (Section~\ref{sec:ls}). Then, since many samples
+Li and Stephens model (Section~\ref{sec:ls}). Then, since many samples
 in a daily batch can share an attachment path,
-we infer trees for each of the clusters separately using standard
+we infer phylogenetic trees for each of the clusters separately using standard
 methods (Section~\ref{sec:sample-cluster-tree-inference}).
@@
 between samples (Section~\ref{sec:backbone_phylogeny}),
-and mutational spectra (Section~\ref{sec:mutation_spectrum})
-while simultaneously identifying the majority of known recombinant
+mutational spectra (Section~\ref{sec:mutation_spectrum}),
+and automatically identify the majority of known recombinant
 lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),

### 2023-05-22 paper.tex
Tweaks to inferred ARGs section

@@
 \texttt{sc2ts}, and to closely study the properties of the
-inferred ARGs to inform subsequent development. Rather
-than attempting to be comprehensive we chose to work with a
+inferred ARGs to inform subsequent development. 
+We work with a 
 representative subset of the available data, limited to
-inferences that could be performed on a single server in
+inferences that can be performed on a single server in
 a few weeks (see below for further details on timings and
@@
 of nodes in the ARGs representing inferred recombination events.
-We analyse these events in Sections~\ref{sec:jackson_recombs},
+Of these recombination nodes, the majority are ancestral 
+to only one sample (Wide ARG: FIXME\%, Long ARG: FIXME \%).
+We analyse signals of recombination in Sections~\ref{sec:jackson_recombs},
 \ref{sec:breakpoint_intervals}, \ref{sec:parent_divergence},

### 2023-05-23 paper.tex
Yan's updates

@@
 We focus on the Long ARG because it covers time
-periods where recombination is known to have had a significant
-impact, and includes
-[HOW MANY?]
-samples designated to [HOW MANY] Pango X lineages (i.e., those inferred
+periods where recombination is known to have occurred
+a large number of times, for example it contains samples from over
+30 Pango X lineages (i.e., those inferred
 to have recombinant ancestry; see Section~\ref{sec:pango_x_lineages} for
@@
 common ancestor (MRCA) of the parent nodes for each recombination breakpoint,
-plotted against the divergence between these parents (i.e., the
+plotted against the divergence between these parents (i.e. the
 total branch length from the parents to their MRCA in the trees to
 the immediate left and right of the breakpoint).
-As in Section~\ref{sec:breakpoint_intervals}, we exclude recombination nodes
-that are ancestral to only one sample from this analysis.
-Figure~\ref{fig:recomb_mrcas}A plots breakpoints for all recombinants that
-are ancestral to more than one sample, and
-Figure~\ref{fig:recomb_mrcas}B plots a subset which pass some further
-reliability filters. Specifically, only recombination nodes that
-[FILL ME IN].
+As in Section~\ref{sec:breakpoint_intervals}, in these plots we
+exclude ``singleton'' recombination nodes (those ancestral to only one sample).
+Figure~\ref{fig:recomb_mrcas}A plots breakpoints for all nonsingleton recombinants,
+and Figure~\ref{fig:recomb_mrcas}B plots a subset of potentially more
+reliable breakpoints that occur in nodes
+ancestral to 10 or more samples (note that this is one of the official criteria
+used when proposing Pango recombinant lineages [CITE]).
+% that we deem to be more reliable by the measure of ``HMM consistency''. 
+% We define an  in that when we compare the LS HMM results when run from
+% 5' to 3' versus 3' to 5', it gives the
+% same mutations patterns and sequence-identical 5' parents as when run 3' to 5'
 This reduces the number of recombination nodes
-to 347 and breakpoints to 358.
+to XXX and breakpoints to YYY.
 Note that each plotted point
@@
 since nodes other than samples are dated somewhat crudely in the current version
-of \texttt{sc2ts} (Section~\ref{sec:node_dating}).
+of \texttt{sc2ts} (Section~\ref{sec:node_dating}). Nevertheless, the histograms
+clearly show that \texttt{sc2ts} can identify recombination between lineages
+that are only a matter of weeks apart.
 
@@
 previously identified as recombinants, i.e. given a Pango designation that
-starts with an ``X''. Since the Wide ARG is restricted to data collected prior
-to mid-2021, it contains samples from only three Pango-designated recombinant
-lineages: XA, XB, and XC. To investigate the diversity of genealogies
-associated with recombinants, we therefore devote more analysis to patterns in
-the Long ARG. As in previous sections, we exclude singleton recombinants where
-possible.
-
-\subsubsection{Wide ARG}
-%% Quick overview of the wide ARG.
-In the Wide ARG, 44
-samples are designated as XA by both Nextclade and GISAID, 237 are designated
-as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
-After removing singleton recombinants, XA numbers remain unchanged, but XB is
-reduced to 235 (229) and XC to 4 (0). We correctly identify that all samples
-designated as XA, XB, or XC by any method have one or more recombination nodes
-in their ancestry.
-
-%% Textual description of the 3 recombinant lineages present in the wide ARG
-The simplest Pango recombinant genealogy is XA: all XA samples trace back to a
-unique originating recombination node. This is the product of a recombination
-between a B.1.177.18 sample (specifically the strain Wales/ALDP-115BF41/2021,
-EPI\_ISL: 1012804) which contributed the majority of the genome from the start
-to a maximum position of 22227bp, and an unknown (inserted) node with imputed
-Pango lineage B.1.1.7, which contributed the remaining right hand portion. The
-recombination node has five immediate children: four sample leaves (EPI\_ISLs:
-989697, 1019487, 1104468 and 1122630) and a UPGMA-inserted node which is the
-ancestor of all other XA samples in the dataset.
-
-For XB, all samples trace back to a originating node which is the product of a
-recombination between a B.1 sample (specifically the strain
-England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of
-the genome from the start up to a maximum position of 23604bp), and two other
-UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of
-27389bp) and B.1.36.8 (the remaining fragment of the genome). As described
-later, in the long ARG the equivalent recombination node has only 2 parents,
-with no involvement of  B.1.36.8, and it is possible that the third parent in
-the wide ARG is artefactual. Additional complications arise because additional
-non-X-designated samples such as B.1.634 also descend from this recombination
-(also true in the long ARG, see below), and there are also a small number of
-further recombination nodes nested within the XB ``clade''. However, all but
-one of these nested recombinations are ancestral to negligible fractions
-of the designated XB samples. The exception accounts for about 17\% of the
-designated XB nodes, and involves a recombination between descendants of the
-originating XB recombination. More specifically strain
-USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination
-between an XB grandparent and its XB grandchild. This could well be
-artefactual.
-
-For XC, the Wide ARG identifies more than one originating recombination node
-(see "multiple origins" below). However, as none of the samples designated as
-XC by Nextclade are designated as XC by GISAID, these patterns could be due to
-mistaken labelling, hence we do not examine XC further.
+starts with an ``X''. As the Wide ARG was only sampled until mid 2021,
+the majority of these Pango X lineages are found only in the Long ARG.
 
@@
 The Long ARG (subsampled to mid-2022) contains 33
-Nextclade-designated Pango recombinant lineages comprising 749 samples (711 when singleton recombinants are removed). In
+Nextclade-designated Pango recombinant lineages comprising 749 samples
+(711 when singleton recombinants are removed). In
 contrast, GISAID designates 38 Pango recombinant lineages in the long ARG,
@@
 
-One of the simplest examples is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
-Here, a subgraph of the long ARG has been plotted, where paths have been traced from all
-Nextclade-identified XA samples (in red) to the closest other sample nodes in our dataset.
+One of the simplest examples is the case of XA, corresponding to group A
+of \citet{Jackson2021-ik} as discussed above. Figure~\ref{fig:pango-simple-origin-graph}A
+shows the exact relationships inferred by \texttt{sc2ts} as a subgraph of the Long ARG. 
+Here, paths have been traced from all
+Nextclade-identified XA samples (in red) to the closest other sample nodes in the ARG.
 Sample nodes are plotted as larger circles, but the subgraph also includes intermediate,
@@
 
-It is clear from the XA subgraph that all the samples labelled XA by Nextclade trace to a single originating recombination node,
-whose genome is a composite of a B.1.177.18 lineage on the left of the genome (up to a maximum
-position of ${\sim}22227$ basepairs) and a B.1.1.7 lineage for the remainder of the genome.
+It is clear from the XA subgraph that all the samples labelled XA by Nextclade trace to a
+single originating recombination node, whose genome is a composite of a B.1.177.18 lineage
+on the left of the genome and a B.1.1.7 lineage on the right. In the subgraph we show
+the maximum genomic position for the recombination breakpoint, here at 22227 basepairs
+(corresponding to a breakpoint strictly less than 22228, see Table~\ref{tab:jackson})
 
@@
 recombination event from which the majority of the corresponding recombinant samples
-descend (but in XJ and XU there are too few Pango X samples to decide on a ``main'' clade:
-see Supplementary Table~\ref{tab:pango-recombinants}).
+descend, and which agrees with the offical Pango designated parent lineages
+(see Supplementary Table~\ref{tab:pango-recombinants}, but note that
+in XJ and XU there are too few Pango X samples to decide on a ``main'' clade).
 
@@
 in the French samples has an estimated maximum position of 21641 basepairs
-(and minimum of 21619) with the insertion end occurring at a maximum position of 25584
+(and minimum of 21619, not shown) with the insertion end occurring at a maximum position of 25584
 (and a minimum of 25470). In contrast, the Omicron insertion in the Turkish sample
@@
 Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).
-
-A similar pattern is seen for several of the origins inferred for the Pango
-XN lineage (originally defined on the basis of sequences from the UK). After removing
-singleton recombinants and the descendants of the erroneously recombinant
-Denmark/DCGC-281594/2021 strain (which, however, includes 53 XN-designated samples)
-the Long ARG contains four independent origins for Nextclade-designated XN
-samples (Supplementary Table~\ref{tab:multiple_origins_table}). The oldest
-of these four origins has 3 mutations above the sample that caused the
-recombination to be inferred. It therefore seems likely that we have not managed
-to sample an XN strain that is very close to the true recombinant. This may
-lead to \texttt{sc2ts} mistakenly inferring multiple recombination events when there
-
-If the only samples available are rather genetically distant from an originating
-recombinant, we might expect sc2ts to infer multiple similar recombination events
-(rather than a single event with extra reversions)
-
-a recombinant outbreak
-has taken a ``true'' XE sample and mistakenly inferred an additional, spurious origin. [WHY? - this
-is presumably because it has changed too much from the nearest in the main XE cluster - need to discuss the
-weighting of 3 mutations for one recombination here. NB: the number
-of nucleotide differences between the closest XE in the two clusters is 11: see the
-Pango-recombinant-analysis workbook]
+Investigation of other multiple-origin examples reveals somewhat similar patterns,
+suggesting that most of the simple multiple origin examples are due to Nextclade mislabelling.
 
@@
 (marked with asterisks in Supplementary Table~\ref{tab:pango-recombinants}; the most
-extensive is for XAB). The earliest example, present in both the Wide and Long ARGs, is XB
-The Logn ARG subgraph is plotted in Figure~\ref{fig:complex_origins_graph}, showing a recombination
+extensive is for XAB). The earliest example (present in both the Wide and Long ARGs) is XB
+The Long ARG subgraph is plotted in Figure~\ref{fig:complex_origins_graph}, showing a recombination
 between a B.1 sample and B.1.627 sample that leads both to the XB samples and to a
@@
 identified as recombinants by Nextclade (plotted in blue). A similar
-pattern is seen when examining XB in the Wide ARG (but as discussed previously,
-in the Wide ARG, XB exhibits multiple nested recombinations, some with many XB
-descendants).
+pattern is seen when examining XB in the Wide ARG (see below).
 
@@
 recombination events may be artefactual, is an important area of future research.
+
+\subsubsection{Wide ARG}
+%% Quick overview of the wide ARG.
+Since the Wide ARG is restricted to data collected prior
+to mid-2021, it contains samples from only three Pango-designated recombinant
+lineages: XA, XB, and XC. Both Nextclade and GISAID designate 44
+samples as XA, while 237 samples are designated
+as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
+After removing singleton recombinants, XA numbers remain unchanged, but XB is
+reduced to 235 Nextclade-designated samples (229 GISAID) and XC is reduced to 4
+(none in the GISAID designations). We correctly identify that all samples
+designated as XA, XB, or XC by any method have one or more recombination nodes
+in their ancestry.
+
+%% Textual description of the 3 recombinant lineages present in the wide ARG
+As in the Long ARG, all XA samples in the wide ARG trace back to a
+unique originating recombination node. This is the product of a recombination
+between a B.1.177.18 sample (specifically the strain Wales/ALDP-115BF41/2021,
+EPI\_ISL: 1012804) which contributed the majority of the genome from the start
+to a maximum position of 22227, and an unknown (inserted) node with imputed
+Pango lineage B.1.1.7, which contributed the remaining right hand portion. The
+recombination node has five immediate children: four sample leaves (EPI\_ISLs:
+989697, 1019487, 1104468 and 1122630) and a UPGMA-inserted node which is the
+ancestor of all other XA samples in the dataset.
+
+For XB, all samples trace back to an originating node which is the product of a
+recombination between a B.1 sample (specifically the strain
+England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of
+the genome from the start up to a maximum position of 23604bp), and two other
+UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of
+27389bp) and B.1.36.8 (the remaining fragment of the genome). Above, we noted that
+in the long ARG the equivalent recombination node has only 2 parents,
+with no involvement of  B.1.36.8; it is possible that the third parent in
+the wide ARG is artefactual. As in the Long ARG, additional
+non-X-designated samples such as B.1.634 also descend from this recombination,
+and there are also a small number of nested recombination nodes. However, all but
+one of these nested recombinations are unimportant, being ancestral to negligible fractions
+of the Nextclade-designated XB samples. The one exception accounts for about 17\% of the
+designated XB nodes, and involves a recombination between descendants of the
+originating XB recombination. More specifically strain
+USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination
+between an XB grandparent and its XB grandchild. This could well be
+artefactual.
+
+For the few XC-labelled samples, the Wide ARG identifies more than one
+originating recombination node. However, as none of the samples designated as
+XC by Nextclade are designated as XC by GISAID, these patterns could be due to
+mislabelling, and greater number of XC samples would be needed to draw reasonable
+conclusions.
 
@@
 The source code for \texttt{sc2ts} is 
-is available on GitHub
+available on GitHub
 \url{https://github.com/jeromekelleher/sc2ts/}.
@@
 \toprule
-Focal Pango  & Num origins & Num focal samples (further split by origin, *=nested) & Official Pango parents & \textbf{Main clade}: sc2ts parents & \textbf{Main clade}: additional descendants \\
+Focal Pango  & Num origins & Num focal samples (further split by origin, \textdagger=nested) & Official Pango parents & \textbf{Main clade}: sc2ts parents & \textbf{Main clade}: additional descendants \\
 \midrule
@@
 \bfseries XAG & 1 & \textbf{17} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAA: 8/8, XU: 1/3, XQ: 2/12 \\
-\bfseries XB & 2 & 57 (\textbf{57}, 1)* & B.1.631 + B.1.634 & B.1 + B.1.627 & B.1.634: 3, B.1.631: 7 \\
+\bfseries XB & 2 & 57 (\textbf{57}, 1)\textdagger & B.1.631 + B.1.634 & B.1 + B.1.627 & B.1.634: 3, B.1.631: 7 \\
 \bfseries XC & 2 & 3 (1, \textbf{2}) & AY.29 + B.1.1.7 & AY.103 + B.1.1.7 &  \\
@@
 \bfseries XN & 4 & 13 (2, 2, \textbf{7}, 2) & BA.1* + BA.2* & BA.1 + BA.2 &  \\
-\bfseries XAJ & 4 & 5 (1, 1, \textbf{3}, 1)* & BA.2.12.1* + BA.4* & Unknown & BA.5: 1 \\
-\bfseries XE & 5 & 163 (3, \textbf{155}, 1, 3, 2)* & BA.1* + BA.2* & BA.1.17.2 + BA.2 &  \\
+\bfseries XAJ & 4 & 5 (1, 1, \textbf{3}, 1)\textdagger & BA.2.12.1* + BA.4* & Unknown & BA.5: 1 \\
+\bfseries XE & 5 & 163 (3, \textbf{155}, 1, 3, 2)\textdagger & BA.1* + BA.2* & BA.1.17.2 + BA.2 &  \\
 \bfseries XZ & 5 & 66 (1, 1, \textbf{61}, 2, 1) & BA.1* + BA.2* & BA.1 + BA.2.10 & XAH: 10/12, XAD: 1/5 \\
-\bfseries XAB & 8 & 48 (1, \textbf{38}, 1, 2, 2, 2, 18, 2)* & BA.1* + BA.2* & BA.1 + BA.2.9 & XAA: 8/8, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
+\bfseries XAB & 8 & 48 (1, \textbf{38}, 1, 2, 2, 2, 18, 2)\textdagger & BA.1* + BA.2* & BA.1 + BA.2.9 & XAA: 8/8, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
 \bottomrule
@@
 In cases of multiple origins, most X-lineages have a single ``main clade'' (in bold).
-Sc2ts inferred parents for the main clade are based on imputed NextClade Pango lineages. Official parents
-are taken from Pango designation alias key (TODO: explain the asterisks).
-Other inferred samples within the main clade are listed, grouped by NextClade Pango designation and clade count / total ARG count (the latter being omitted for non-recombinant designations).
+Sc2ts inferred parents for the main clade are based on Nextclade designations, imputed where necessary.
+Official parents are taken from Pango designation alias key (TODO: explain the asterisks).
+Note that B.1.617.2 is the origin of the Delta VoC, which includes all AY.* classes.
+Additional descendants within the main clade are summarised giving their NextClade Pango designation and clade count / total ARG count (the latter being omitted for non-recombinant designations).
 }
@@
 
-\begin{table}
-\small
-\centering
-\begin{tabular}{lp{0.2cm}p{0.8cm}lp{0.9cm}ll}
-\toprule
-Date & $n$ & Causal Pango & Causal strain & Causal \#muts & Sc2ts parents & Breakpoint \\
-\midrule
-2022-03-24 & 2 & XN & Germany/BY-RKI-I-674341/2022 & 3 & Unknown + BA.2 & (2833,9867) \\
-2022-03-29 & 2 & XN & Spain/AB-HGUA-26086757/2022 & 2 & BA.1.1 + BA.2.9 & (2833,4185) \\
-2022-04-23 & 7 & XN & Spain/MD-HRYC-12296321/2022 & 2 & BA.1 + BA.2 & (4185,4322) \\
-2022-05-17 & 2 & XN & USA/TX-HMH-MCoV-97903/2022 & 5 & BA.2 + BA.2 & (22776,22787) \\
-\bottomrule
-\end{tabular}
-\caption{\label{tab:multiple_origins_table}
-Summary of inferred multiple ``origins'' for Nextclade-designated XN samples after filtering
-(Official: BA.1* \& BA.2*, breakpoint between 2834 \& 4183, defined from UK strains, earliest
-on 2022-02-01;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/480}).
-The 13 XN samples are grouped into rows based on their most recent ancestral recombination node;
-all the descendant samples of the four recombination nodes are designated XN by Nextclade.
-Details are also provided for the ``causal strain'' that led to the recombination
-being proposed by \texttt{sc2ts}. When undersampling causes three or more mutations to lie
-immediately above the causal strain, \texttt{sc2ts} may mistakenly propose multiple
-independent origins rather than one. Note that a large number of XN samples were previously
-removed by filtering.
-}
-\end{table}
+% \begin{table}
+% \small
+% \centering
+% \begin{tabular}{lp{0.2cm}p{0.8cm}lp{0.9cm}ll}
+% \toprule
+% Date & $n$ & Causal Pango & Causal strain & Causal \#muts & Sc2ts parents & Breakpoint \\
+% \midrule
+% 2022-03-24 & 2 & XN & Germany/BY-RKI-I-674341/2022 & 3 & Unknown + BA.2 & (2833,9867) \\
+% 2022-03-29 & 2 & XN & Spain/AB-HGUA-26086757/2022 & 2 & BA.1.1 + BA.2.9 & (2833,4185) \\
+% 2022-04-23 & 7 & XN & Spain/MD-HRYC-12296321/2022 & 2 & BA.1 + BA.2 & (4185,4322) \\
+% 2022-05-17 & 2 & XN & USA/TX-HMH-MCoV-97903/2022 & 5 & BA.2 + BA.2 & (22776,22787) \\
+% \bottomrule
+% \end{tabular}
+% \caption{\label{tab:multiple_origins_table}
+% Summary of inferred multiple ``origins'' for Nextclade-designated XN samples after filtering
+% (Official: BA.1* \& BA.2*, breakpoint between 2834 \& 4183, defined from UK strains, earliest
+% on 2022-02-01;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/480}).
+% The 13 XN samples are grouped into rows based on their most recent ancestral recombination node;
+% all the descendant samples of the four recombination nodes are designated XN by Nextclade.
+% Details are also provided for the ``causal strain'' that led to the recombination
+% being proposed by \texttt{sc2ts}. When undersampling causes three or more mutations to lie
+% immediately above the causal strain, \texttt{sc2ts} may mistakenly propose multiple
+% independent origins rather than one. Note that a large number of XN samples were previously
+% removed by filtering.
+% }
+% \end{table}
 

### 2023-05-23 paper.tex
Fix line ends

@@
 immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
-Early in the SARS-CoV-2 pandemic there was insufficient 
-genetic variation for meaningful recombination.
-In late 2020, the appearance and
+% Early in the SARS-CoV-2 pandemic there was insufficient 
+% genetic variation for meaningful recombination.
+Early on in the COVID-19 pandemic, the levels of genetic diversity
+were too low for the emergence of distinctive recombinant strains.
+By late 2020, however, the appearance and
 spread of variants of concern (VoC), designated into classes such as Alpha and
@@
 Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
-combine specific subsets of mutations from different lineages
+combine specific subsets of characteristic mutations from different lineages
 \citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz},
-which can only identify inter-lineage recombination events specific to those motifs.
+and as a result can only identify inter-lineage recombination events.
+% specific to those motifs.
 % % JK: I've left this out because I felt it broke the narrative up a bit, but
@@
 phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
-over ten million SARS-CoV-2 genomes, without pre-specifying motifs. 
+over ten million SARS-CoV-2 genomes, without pre-specifying sets of characteristic
+mutations.
 RIPPLES finds candidate recombinants by using an existing phylogeny
@@
 It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
-and affects the results of downstream analysis, e.g., inference of
+and affects the results of downstream analyses, e.g., inference of
 selection~\citep{Anisimova2003-vr}. Standard phylogenetic methods do
@@
 (Section~\ref{sec:parsimony-heuristics}).
-Using the current preliminary version of \texttt{sc2ts} we infer two
+Using the current preliminary version of \texttt{sc2ts}, we infer two
 large ARGs (1,265,685 and 657,239 samples, respectively), and study
@@
 This reduces the number of recombination nodes
-to XXX and breakpoints to YYY.
+to 161 and breakpoints to 174.
 Note that each plotted point
@@
 Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
-It is important to note here that some of the time estimates in these figures
+We caution that some of the time estimates in these figures
 come with a substantial degree of uncertainly,
-since nodes other than samples are dated somewhat crudely in the current version
+since non-sample nodes are dated somewhat crudely in the current version
 of \texttt{sc2ts} (Section~\ref{sec:node_dating}). Nevertheless, the histograms

### 2023-05-23 make_csv_files.py
Update to new tree sequence versions (with interval metadata)
And also switch to using <=10 descendant samples for MRCA filtering

@@ -7,11 +7,15 @@ import utils
 
 if __name__ == "__main__":
 
-    # Make a breakpoints file for the long ARG
-    fn = "upgma-mds-1000-md-30-mm-3-2022-06-30-recinfo-gisaid-il.ts.tsz"
-    prefix = utils.snip_tsz_suffix(fn)
-    long, basetime = utils.load_tsz("data", fn)
-    logging.info("Collecting tree information, may take a while")
-    long_treeinfo = sc2ts.TreeInfo(long)
-    df = long_treeinfo.export_recombinant_breakpoints()
-    df.to_csv(f"data/breakpoints_{prefix}.csv")
+    # Make breakpoints files
+    for fn in (
+        "upgma-mds-1000-md-30-mm-3-2022-06-30-recinfo2-gisaid-il.ts.tsz",
+        "upgma-full-md-30-mm-3-2021-06-30-recinfo2-gisaid-il.ts.tsz",
+    ):
+        prefix = utils.snip_tsz_suffix(fn)
+        ts, basetime = utils.load_tsz("data", fn)
+        logging.info("Collecting tree information, may take a while")
+        treeinfo = sc2ts.TreeInfo(ts)
+        df = treeinfo.export_recombinant_breakpoints()
+        df[f"parents_dist_{ts.time_units}"] = ts.nodes_time[df["mrca"]] - ts.nodes_time[df["node"]]
+        df.to_csv(f"data/breakpoints_{prefix}.csv")

### 2023-05-23 plot.py

@@ -158,8 +158,8 @@ class Figure:
 
     name = None
     ts_dir = "data"
-    wide_fn = "upgma-full-md-30-mm-3-2021-06-30-recinfo-gisaid-il.ts.tsz"
-    long_fn = "upgma-mds-1000-md-30-mm-3-2022-06-30-recinfo-gisaid-il.ts.tsz"
+    wide_fn = "upgma-full-md-30-mm-3-2021-06-30-recinfo2-gisaid-il.ts.tsz"
+    long_fn = "upgma-mds-1000-md-30-mm-3-2022-06-30-recinfo2-gisaid-il.ts.tsz"
 
     def plot(self, args):
         raise NotImplementedError()
@@ -608,19 +608,19 @@ class Cophylogeny(Figure):
 
 class CophylogenyWide(Cophylogeny):
     name = "cophylogeny_wide"
-    sc2ts_filename = "upgma-full-md-30-mm-3-2021-06-30-recinfo-gisaid-il.ts.tsz"
+    sc2ts_filename = "upgma-full-md-30-mm-3-2021-06-30-recinfo2-gisaid-il.ts.tsz"
     use_colour = "Pango"
 
 
 class CophylogenyLong(Cophylogeny):
     name = "supp_cophylogeny_long"
-    sc2ts_filename = "upgma-mds-1000-md-30-mm-3-2022-06-30-recinfo-gisaid-il.ts.tsz"
+    sc2ts_filename = "upgma-mds-1000-md-30-mm-3-2022-06-30-recinfo2-gisaid-il.ts.tsz"
     use_colour = "Pango"
 
 
 class RecombinationNodeMrcas(Figure):
     name = None
-    sc2ts_filename = "upgma-mds-1000-md-30-mm-3-2022-06-30-recinfo-gisaid-il.ts.tsz"
+    sc2ts_filename = "upgma-mds-1000-md-30-mm-3-2022-06-30-recinfo2-gisaid-il.ts.tsz"
     csv_fn = "breakpoints_{}.csv"
     data_dir = "data"
 
@@ -628,9 +628,9 @@ class RecombinationNodeMrcas(Figure):
         self.ts, self.basetime = utils.load_tsz(self.data_dir, self.sc2ts_filename)
 
         prefix = utils.snip_tsz_suffix(self.sc2ts_filename)
-        df = pd.read_csv(os.path.join(self.data_dir, self.csv_fn.format(prefix)))
-        df["tmrca"] = self.ts.nodes_time[df.mrca.values]
[…]

### 2023-05-23 paper.tex
stuff

@@
 higher risk of coinfection, increasing opportunities for new phenotypically
-distinct recombinants to emerge and spread.
-[Handling recombination properly as an ongoing process is vital. Our results
-here show that recombination can be incorporated into a cohesive joint
-model of viral evolution. We basically recapitulate what is known, plus
-there's a whole bunch of other stuff that we haven't explored because
-we're in the validation stage.]
+distinct recombinants to emerge and spread. Phylogenetic approaches have
+been central to pandemic responses thus far [TODO: Examples and citations].
+However, as recombinant lineages are now endemic and it is clear that
+recombination is a major feature of ongoing SARS-CoV-2 evolution,
+it is imperative that these methods are updated to incorporate the
+effects of recombination, and future public health interventions are
+not based on incomplete and potentially biased evolutionary models.
+Here we have introduced the first method to infer an evolutionary
+history that jointly models genetic inheritance, mutation and
+recombination at pandemic scale, and illustrated how this single
+structure accurately captures results derived by many different means.
 
@@
 Nonetheless, \texttt{sc2ts} is currently ``alpha'' quality software, and
-there are some relatively straightforward changes that could be made to
-improve inference quality. Firstly, the pipeline used to identify and
+we caution against overinterpreting current results. As we have sought
+to illustrate throughout, there are some clear areas for improvement
+and some relatively straightforward changes that could be made to
+improve inference. The pipeline used to identify and
 mask erroneous sites in the input alignments is simplistic, and, among
@@
 approach~\citep[e.g.][]{Aksamentov2021-hj} would
-likely yield significant improvements, and reduce the effect sites with
+likely yield significant improvements, and reduce the effect of sites with
 artefactually high levels of mutation (e.g., site 28271;
@@
 of the HMM with a single mismatch ratio is simplistic, and it would
-be straightforward to condition on per-site mutation rates. (Per nucleotide
-state transition, is also possible, with some additional development.)
+be straightforward to condition on per-site mutation rates (per nucleotide
+state transition, is also possible, with some additional development).
 Recombination breakpoints for the ARG are currently inserted at the
@@
 breakpoint could be chosen within the interval, for example based
-on the sequence motifs.
+on sequence motifs [CITATION?].
+% Hint: there's probably a bunch of other stuff you could do with this
+It is likely that the basic machinery of finding matches and quantifying
+the undertainty around them under a well-defined statistic model
+in large ARGs would have many applications besides those explored here.
 
@@
 into focus~\citep{Turakhia2021-ur,de2023maximum}.
-[Finish this up with a discussion of how easy it is to do stuff with our
-ARGS, how awesome tskit is, and how little code there is in sc2ts.
-Being able to quickly develop new methods by putting together high
-performance, high quality components is an essential element of
-preparedness for futture pandemics.]
+Replacements that can scale to millions of genomes have had to be developed
+at speed, usually focusing on compiled programming languages to maximise
+performance. Here, however, we have developed a new method based on
+an existing data structure and library infrastructure, designed
+from the beginning to scale to millions of
+samples~\citep{Kelleher2016-wk,Kelleher2019-ba}.
+The \texttt{sc2ts} package is written entirely in Python,
+and by reusing existing high-performance components can infer
+recombinant viral ancestry at unprecedented scale.
+Similarly, all of the analyses shown here are written in Python,
+using the \texttt{tskit} API, mostly running in a few seconds
+on standard laptop computers (see the Code Availability
+section for details of the corresponding Jupyter notebooks).
+The ability to quickly and easily develop new methods and
+analyses, at scale, by using existing high-quality software
+components and data formats is an essential element of future
+pandemic preparedness.
 

### 2023-05-23 paper.tex
Tidy up intro to xlins

@@
 ``Wide'' and ``Long'' ARGs throughout.
-The Wide ARG is densely sampled but time-limited and
-includes all 1.27 million sample genomes  with a
-maximum sampling delay of 30 days
-(see Section~\ref{sec:filtering_time_travellers})
-up to June 30, 2021.
-The Long ARG is more sparsely sampled but
-includes data for an additional year (June 30, 2022).
-For the long ARG, we use at most 1,000 randomly subsampled
-genomes per day from those meeting the maximum sampling delay criterion.
+The Wide ARG is densely sampled but time-limited, and
+includes all 1.27 million sample genomes collected up to June 30, 2021
+with a maximum delay between sampling and submission dates of 30 days
+(see Section~\ref{sec:filtering_time_travellers}).
+% The Long ARG is more sparsely sampled but
+% includes data for an additional year (June 30, 2022).
+% For the Long ARG, we use at most 1,000 randomly subsampled
+% genomes per day from those meeting the maximum sampling delay criterion.
+For the Long ARG, we randomly sub-sample a maximum of 1,000 genomes per
+day (again restricting the delay between sampling and submission to 30 days),
+and include an additional year worth of samples (to June 30, 2022).
 
@@
 previously identified as recombinants, i.e. given a Pango designation that
-starts with an ``X''. As the Wide ARG was only sampled until mid 2021,
-the majority of these Pango X lineages are found only in the Long ARG.
-
-\subsubsection{Long ARG}
-%% high level summary of the pango recombinant lineages and samples in the Long ARG
-The Long ARG (subsampled to mid-2022) contains 33
-Nextclade-designated Pango recombinant lineages comprising 749 samples
-(711 when singleton recombinants are removed). In
-contrast, GISAID designates 38 Pango recombinant lineages in the long ARG,
-comprising only 515 samples. In the analyses below, we mainly focus on the
-larger number of Nextclade designated samples.
-
-Of the 749 Nextclade-designated recombinants, we infer that all but two descend
-from a recombination node. These two samples are designated XP, and a likely
+starts with an ``X''. We primarily focus on the Long ARG in this section
+since it contains many more recombinants, briefly summarising the status
+of Pango X lineages for the Wide ARG in Section~\ref{sec-pango_x_wide_arg}.
+
+When using Nextclade Pango lineage designations, the
+Long ARG contains 749 samples from 33 recombinant lineages; this
+reduces toe 711 and XX when singleton recombinants (i.e., recombination
+nodes subtending only one sample) are removed.
+In contrast, when using GISAID Pango lineage assignments (for the same
+data) we have 515 samples from 38 Pango recombinant lineages,
+indicating that there is substantial uncertainty in the assignment of
+Pango X lineages to samples.
+
+Of the 749 Nextclade-designated recombinant samples, all but two descend
+from a recombination node in the Long ARG.
+These two samples are designated XP, and a likely
 explanation for the absence of a corresponding recombination event in the long ARG
@@
 (\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
-during preprocessing. An improved masking pipeline is an important avenue for
-future development (see the Discussion).
-
-38 of the remaining X designated samples are singleton recombinants, descending
+during preprocessing (See~\ref{sec:data_preprocessing} for details and potential
+improvements). 38 of the remaining X designated samples are singleton recombinants,
+descending
 from recombination nodes that are ancestral only to that sample (10 are XZ; 6
@@
 The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
-1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, caused by strain Denmark/DCGC-281594/2021,
-is ancestral to more than 127K samples, most of which are not designated as
-recombinants by any other method. It is highly likely to be a spuriously inferred recombination event.
-Understanding the origin of this and other similar nodes is an important
-direction for future work (see Discussion). For simplicity, we exclude these
-samples from further analyses in this section.
+1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, caused by strain
+Denmark/DCGC-281594/2021, is ancestral to more than 127K samples,
+and is highly likely to be a spuriously inferred recombination event.
+Understanding the origin of this
+and other similar high-impact, but likely spurious recombination
+events is an important aspect of future work.
+For simplicity, we exclude these samples from further analyses in
+this section.
 
 The remaining remaining 630 Pango recombinant samples trace back to 50 different
-most recent recombination nodes, representing 31
+most recent recombination nodes, corresponding to 31
 Nextclade-designated Pango lineages. These are summarised
@@
 into three classes: single origin, multiple origin,
-and multiple nested origins.
-
-% The sequences of 28 Pango recombinants trace back
-% to some recombinant node(s); however, the sequences of the five Pango
-% % jk: really? I thought it was just XP that wasn't inferred to be a
-% % recombinant.
-% recombinants XP, XU, XAA, XAG, and XAK do not.
-% A likely explanation for the XP
-% recombinant sequences not having a recombinant origin in the long ARG is that
-% its characteristic multibase deletion
-% (\url{https://github.com/cov-lineages/pango-designation/issues/481}) was masked
-% during preprocessing and therefore was not incorporated into the HMM.
-% % Eh?
-% Upon tracing the origin of XAG through the ARG, XAG might have been caused by
-% incorrect imputation of its parent nodes(?). XAK likely does not appear to have
-% a recombinant origin because of random subsampling, but it may with a more
-% complete tree.
-
-% [FIXME this isn't quite right - not all the sequences for a given pango
-% lineage trace back to the same recombination event, so won't have the same
-% intervals.]
-% The mosaic genome structure of all the 32 Pango recombinant lineages is concordant with
-% the structure determined by the Pango Lineage Designation Committee
-% (Supplementary Table 3). For example, the genome of XD (also known as
-% ``Deltacron'') is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
-% (Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.
-
-% For the long ARG, we examine the lineages in increasing complexity of the
-% inferred evolutionary relationships between samples assigned to Pango
-% X lineages.
-
-% - Recombinant Pango lineages
-%     How well do we recover recombinant Pango lineages? Work through the
-% examples in increasing compexity, using headings to split things up if
-% necessary. (i.e., Single origin, Multiple origin, Interellated lineages). We'd
-% need to be clear about what these terms mean, of course. We could include
-% lineage graphs for all of these examples?
-
-% OLD TEXT
-
-% The sequences of 14 Pango recombinants trace back to a single recombination
-% node. The sequences of the remaining 18 Pango recombinants trace back to
-% multiple recombination nodes, however. This may be a result of undersampling in
-% the long ARG, which was built by randomly subsampling at most 1,000 sequences
-% from each daily batch. Having extra recombination nodes in the long ARG
-% appeared to be a more parsimonious explanation for the 19 Pango recombinants
-% than having a single recombination node with many immediately ancestral
-% mutations (including recurrent mutations and reversions), despite that the
-% recombinants may indeed trace to a single origin.
+and multiple nested origins, which we discuss in the following sections.
 
@@
 sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
-Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
+Pango designation; smaller symbols are non-sample nodes inserted by
+\texttt{sc2ts}, with
 imputed Pango status. Genomic regions inherited by the recombination node are shown
@@
 on the left of the genome and a B.1.1.7 lineage on the right. In the subgraph we show
-the maximum genomic position for the recombination breakpoint, here at 22227 basepairs
-(corresponding to a breakpoint strictly less than 22228, see Table~\ref{tab:jackson})
+the maximum genomic position for the recombination breakpoint, here at 22,227 basepairs
+(corresponding to a breakpoint strictly less than 22,228, see Table~\ref{tab:jackson})
 
@@
 \subsubsection{Wide ARG}
+\label{sec-pango_x_wide_arg}.
 %% Quick overview of the wide ARG.

### 2023-05-23 paper.tex
Updates

--- 2023-05-23 paper.tex
@@
 Recombination via template switching is a common feature
-in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
+of the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
 including SARS-CoV-2
@@
 together mutations carried by different lineages, recombination plays an
-important role in generating genomic diversity, with recombinant lineages
+important role in generating genetic diversity, with recombinant lineages
 associated with adaptation to new host species and with the production of more
@@
 Early on in the COVID-19 pandemic, the levels of genetic diversity
-were too low for the emergence of distinctive recombinant strains.
+were too low to precipitate the emergence of distinctive recombinant strains.
 By late 2020, however, the appearance and
@@
 Delta which harboured multiple characteristic mutations,
-created the conditions required for recombination to occur
-and recombinant strains with onward transmission were detected at
-low frequency~\citep{Jackson2021-ik}.
+created the conditions required for recombination to occur,
+and recombinant strains with onward transmission became detectable at
+low frequencies~\citep{Jackson2021-ik}.
 More recently, the high prevalence of Omicron,
@@
 recombination events on top of existing recombinant lineages, creating a highly
-reticulated ancestry.
+reticulated genealogy.
 
@@
 It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
-and affects the results of downstream analyses, e.g., inference of
+and affects the results of downstream analyses, such as inference of
 selection~\citep{Anisimova2003-vr}. Standard phylogenetic methods do
@@
 Wohns2022-th}, and the supporting \texttt{tskit} software library
-is now widely used in population genetic applications.
+is now widely used in population genetics applications.
 The methods in \texttt{tskit} have been developed to support millions of
@@
 based on their collection dates, and proceeds in three phases.
-First, likely attachment nodes in the current ARG
-(or paths, if inferred to be a recombinant) are
-chosen for each of the samples in the daily batch using the
-Li and Stephens model (Section~\ref{sec:ls}). Then, since many samples
-in a daily batch can share an attachment path,
+% First, likely attachment nodes in the current ARG
+% (or paths, if inferred to be a recombinant) are
+First, possible paths connecting each sample to the current ARG
+are inferred (allowing for recombination) using the
+% chosen for each of the samples in the daily batch using the
+Li and Stephens (LS) model (Section~\ref{sec:ls}). Then, since many samples
+in the batch can share an attachment path,
 we infer phylogenetic trees for each of the clusters separately using standard
@@
 Finally, we attach the trees for these sample-clusters to the current
-ARG, and apply some parsimony heuristics to address issues
-introduced by the inherent greediness of the strategy
+ARG, and apply some parsimony-based heuristics to address issues
+introduced by the inherent greediness of this strategy
 (Section~\ref{sec:parsimony-heuristics}).
@@
 The goals of this preliminary study are to illustrate the utility of
-\texttt{sc2ts}, and to closely study the properties of the
+\texttt{sc2ts}, and to investigate the properties of the
 inferred ARGs to inform subsequent development.
@@
 using the \texttt{tszip} utility). The majority of this
-space is used by metadata, which when discarded results in
+space is used by the metadata, which when discarded results in
 an encoding that requires an average of only 8.29 bytes per
@@
 %% The ARG is mostly tree-like
-Compared to organisms like humans which recombine every generation,
-in SARS-CoV2 recombination is rare, with recombination nodes accounting
+Compared to organisms like humans which recombine in every generation,
+in SARS-CoV-2 recombination is relatively rare, with recombination nodes accounting
 for $<$0.3\% of the inferred ancestry (Table~\ref{tab:args}). As a result,
@@
 relatives. We expect ARGs to be particularly tree-like
-early in the pandemic, when co-infection was less likely.
+early in the pandemic, when co-infection was less likely and divergence between
+lineages relatively low.
 
@@
 
-A classic tree-based summary of SARS-CoV2 ancestry is provided
+A classic tree-based summary of SARS-CoV-2 ancestry is provided
 by the Nextstrain project \citep{Hadfield2018-ef}. The trees
@@
 near the tips (e.g., in the Alpha clade).
-Finally, and most noticeably, the branch lengths differ.
+Finally, and most noticeably, some branch lengths differ substantially.
 As discussed in Section~\ref{sec:node_dating}, the dating of nodes
@@
 sequences efficiently~\citep{Kelleher2016-wk,Ralph2020-efficiently} and
-to provide high-levels of data compression~\citep{Kelleher2019-ba}.
+to provide high levels of data compression~\citep{Kelleher2019-ba}.
 The same idea has recently been used to represent
@@
 mutations in the Wide ARG by type (defined by the inherited and derived states),
-excluded mutations inherited by only a single sample (which are
-enriched for sequencing errors and artefacts).
+excluding mutations inherited by only a single sample (which are
+more likely to be sequencing errors).
 Similarly, we took the data for single nucleotide mutations from
-\citep[][\url{https://github.com/ju-lab/SC2_evol_signature}]{Yi2021-sc}, excluded
+\citet[][\url{https://github.com/ju-lab/SC2_evol_signature}]{Yi2021-sc}, excluding
 mutations occurring along terminal branches, and tallied them up by type.
@@
 in mosaic structure and parent lineages
-(including CAMC-CB7AB3 which is inferred to have two breakpoints under both
+(including sample CAMC-CB7AB3, which is inferred to have two breakpoints under both
 methods).
@@
 The Long ARG contains 5 of the sequences: two each from groups A and B
-and QEUH-1067DEF. These cluster under three recombination nodes, as expected,
+and sample QEUH-1067DEF. These cluster under three recombination nodes, as expected,
 and have identical breakpoint intervals and parental lineage assignments
 to those of Wide ARG.
-The recombination nodes for Group B and QEUH-1067DEF are ancestral only
+The recombination nodes for Group B and sample QEUH-1067DEF are ancestral only
 to the sequences involved.
@@
 Long ARG contains 749 samples from 33 recombinant lineages; this
-reduces toe 711 and XX when singleton recombinants (i.e., recombination
+reduces to 711 and FIXMEXX when singleton recombinants (i.e., recombination
 nodes subtending only one sample) are removed.

### 2023-05-23 plot.py
Nicer inset histograms for interval plot

@@ -1786,15 +1786,17 @@ class RecombinationIntervals(Figure):
 
         axins2 = inset_axes(
             ax1,
-            width=2.0,
+            width=2.7,
             height=1.5,
-            bbox_to_anchor=(0.15, 1.0),
+            bbox_to_anchor=(0.13, 1.0),
             bbox_transform=ax1.transAxes,
             loc="upper left",
         )
-        axins2.hist(length)
+        y, x, _ = axins2.hist(length, bins=np.arange(0, 15_000, 500))
         axins2.set_xlabel("Width of interval")
-        axins2.set_ylabel("Count")
+        axins2.set_ylabel("Interval count (100s)")
+        axins2.yaxis.set_ticks(np.arange(0, y.max() + 20, 100))
+        axins2.yaxis.set_major_formatter(mpl.ticker.FuncFormatter(lambda y, pos: int(y/100)))
 
         print("Recombination nodes = ", len(df.node.unique()))
         print("Num intervals:", len(intervals))

### 2023-05-23 plot.py
Make a single MRCA plot
And label histogram axis, etc

@@ -640,11 +640,6 @@ class RecombinationNodeMrcas(Figure):
         )
 
     @staticmethod
-    def filter_breaks(df):
-        # return np.logical_and(df.fwd_bck_parents_max_dist == 0, df.is_hmm_mutation_consistent)
-        return df.max_descendant_samples >= 10
-
-    @staticmethod
     def add_common_lines(ax, num, ts, common_proportions):
         v_pos = {k: v for v, k in enumerate(common_proportions.keys())}
         for i, (u, (pango, prop, date)) in enumerate(common_proportions.items()):
@@ -660,9 +655,9 @@ class RecombinationNodeMrcas(Figure):
             most = "Most" if v_pos[u] == 0 else ordinal(i + 1) + " most"
             ax.text(
                 (ts.node(u).time + 5) / 7,  # Hand tweaked to get nice label positions
-                date,
-                f"{most} freq. MRCA,{sep}{n_children} children,{sep}{prop * 100:.1f} % of {pango}",
-                fontsize=8,
+                date if i !=2 else date - timedelta(days=7),  # Push apart the 2nd and 3rd labels
+                f"{most} frequent{sep}MRCA,{sep}{n_children} children,{sep}{prop * 100:.1f} % of {pango}",
+                fontsize=10,
                 va="center",
                 bbox=dict(facecolor="white", edgecolor="none", pad=0),
             )
@@ -676,60 +671,80 @@ class RecombinationNodeMrcas(Figure):
         label_tweak,
         xlab=True,
         ylab=True,
-        filter_pass_colour="green",  # or "" to not plot
-        filter_fail_colour="blue",
+        filter_pass_size=20,
+        filter_fail_size=5,
+        filter_num_desc = 5
     ):
-        logging.info(f"Plotting {title}")
-        filter_pass = (filter_pass_colour, self.filter_breaks(df), "# descendants >= 10")
-        filter_fail = (filter_fail_colour, ~self.filter_breaks(df), "# descendants < 10")
 
-        points = []
-        pangoX = []
-        tmrca_delta = []
-        for colour, use, label in (filter_pass, filter_fail):
-            if colour != "":
-                plot_df = df[use]
-                points.extend(plot_df.node.values)
-                tmrca_delta.extend(plot_df.parents_dist_days.values)
-                is_pangoX = df.causal_lineage.str.startswith("X")
-                pangoX.extend(np.flatnonzero(np.logical_and(use, is_pangoX)))
[…]

### 2023-05-25 paper.tex
Yans updates

@@
 as either samples or by the inference process which produced them to aid
-understanding of the relative importances of these processes to the overall
+understanding of the relative importance of these processes to the overall
 topology. Similarly, edges are classified by the type of the parent node,
@@
 Inferring these ARGs does require substantial computation.
-The wide ARG required 17 days of computation time on a machine with
-128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). The long ARG
+The Wide ARG required 17 days of computation time on a machine with
+128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). The Long ARG
 required 23 days on a (much older) machine with 40 threads and 256 GB RAM (2x
@@
 a sample of strains which are not thought to be recombinants.
-For validation purposes, we our compare our ARGs with
+For validation purposes, we compare our ARGs with
 a downloaded Nextstrain tree, restricted to the time period
-covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
-the \texttt{sc2ts} ARGs to a backbone containing only samples present
-the Nextstrain tree. This results in a small set of shared
+covered by each ARG. To enable this, we ``simplify''~\citep{Kelleher2018-xc}
+the \texttt{sc2ts} ARGs to a backbone containing only those samples
+present in the Nextstrain tree. This results in a small set of shared
 samples (180 for the Wide ARG, 88 for the Long ARG), none of which are
@@
 Secondly, there are non-identical sample partitions
-near the tips (e.g., in the Alpha clade).
+near the tips (e.g. within the Alpha clade).
 Finally, and most noticeably, some branch lengths differ substantially.
@@
 these issues may be alleviated in future versions of \texttt{sc2ts}.
-
-% JK: I don't think this is the place to get into discussing the benefits
-% of an ARG, and I don't want to get bogged down in explaining what it
-% means to have 1500 trees along the genome.
-
-% Although the simplified backbone ARGs have few local trees, all of which show
-% essentially the same topologies, note that in the original ``un-simplified''
-% Wide and Long ARGs, there are 1,496 and 958 local trees respectively (Table 1).
-% Hence, while for a relatively small subset of the genealogy may be well
-% described by a single tree, the full dataset shows substantial variation in
-% phylogenetic relationships (i.e. ``phylogenetic incongruence'') across the
-% SARS-CoV-2 genome, much of which may be attributable to recombination. An ARG
-% representation is therefore essential to fully capture these reticulate
-% genealogical relationships.
 
@@
 categorised by type (i.e., inherited state $>$ derived state). The percentages
-of each mutation type from the wide ARG are represented by blue bars and the
+of each mutation type from the Wide ARG are represented by blue bars and the
 percentages from Yi et al. by orange bars, with the darker colours representing
@@
 mutation and recombination in an ARG (Section~\ref{sec:args})
-is that is allows us to precisely trace the origins of recombinants.
+is that this allows us to precisely trace the origins of recombinants.
 [Katherine - need some input from you here to say this properly?]
@@
 As in Section~\ref{sec:breakpoint_intervals}, in these plots we
-exclude ``singleton'' recombination nodes (those ancestral to only one sample).
-Figure~\ref{fig:recomb_mrcas}A plots breakpoints for all nonsingleton recombinants,
-and Figure~\ref{fig:recomb_mrcas}B plots a subset of potentially more
-reliable breakpoints that occur in nodes
-ancestral to 10 or more samples (note that this is one of the official criteria
-used when proposing Pango recombinant lineages [CITE]).
+exclude breakpoints associated with ``singleton'' recombination nodes
+(those ancestral to only one sample).
+Larger points distinguish those breakpoints which occur in nodes
+ancestral to 5 or more samples, comprising 316 breakpoints from
+291 recombination nodes. The criterion of 5 descendants matches the
+minimum number required to designate a new Pango recombinant lineage
+\citep{Rambaut2020-dw}, and serves to highlight those recombinants which
+have greater epidemiological significance.
+
 % that we deem to be more reliable by the measure of ``HMM consistency''.
@@
 % same mutations patterns and sequence-identical 5' parents as when run 3' to 5'
-This reduces the number of recombination nodes
-to 161 and breakpoints to 174.
-Note that each plotted point
-represents a breakpoint, so that a recombination node with more than one
-breakpoint (e.g. with 3 or more parents, comprising ${\sim}10 \%$ of the
-recombination nodes in Figure ~\ref{fig:recomb_mrcas}A and ${\sim}3 \%$ in
-Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
+Note that as each plotted point represents a breakpoint, a recombination
+node with more than one breakpoint (e.g. with 3 or more parents,
+comprising ${\sim}10 \%$ of the
+recombination nodes in Figure ~\ref{fig:recomb_mrcas}) will be represented by several points.
 We caution that some of the time estimates in these figures
@@
 clearly show that \texttt{sc2ts} can identify recombination between lineages
-that are only a matter of weeks apart.
+that are only a few weeks apart.
 
@@
 MRCAs of parents associated with Pango designated recombinants (XA, etc)
-are identified in orange. (A) 851 parent pairs (singleton recombinants removed).
-(B) Further filtered
-to 358 parent pairs by removing HMM inconsistent breakpoints (see text).
+are identified in orange. Larger symbols represent breakpoints in recombination
+nodes which are ancestral to five or more samples. 
 Horizontal dotted lines show the four most common MRCA nodes,
 which tend to be associated with major outbreaks and with many immediate children.
-See Figure~\ref{fig:recomb_mrcas_voc_breakdown}
+The stacked histogram shows the distribution of parental divergence times, ranging
+from parents that have diverged only a handful of days previously, to much more
+divergent parent lineages. See Figure~\ref{fig:recomb_mrcas_voc_breakdown}
 for equivalent plots broken down by parental VoC classification.
@@
 ${\sim}95$ weeks prior to the event: these correspond to recombinants whose
-parental common ancestor traces to early 2020, including both Delta-Delta and
-Delta-Omicron recombinants
-(Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
-Figure~\ref{fig:recomb_mrcas_voc_breakdown} also shows that, of the recombinant
-parents that fall into the Alpha, Delta, and Omicron VoC classes, the majority
-are within-Delta and within-Omicron recombinants, and that Omicron and
-Delta are the variants associated with the most recombination. This may
-reflect either sampling intensity, the prevalence of cases (which increases the
+parental common ancestor traces to early 2020. We would expect, for example
+Delta-Omicron recombinants to fall into this group, hence it can be
+useful to classify the breakpoints further by the VoC combinations of their
+parent lineages. Considering only the Alpha, Delta, and Omicron VoC classes,
+such a classification reveals that the majority of breakpoints have two Delta or
+two Omicron parents, and that Omicron and Delta are the variants associated
+with the most recombination (Supplementary Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
+This may reflect either sampling intensity, the
+prevalence of cases (which increases the
 chance of coinfection and recombination), or possible heterogeneity in
@@
 previously identified as recombinants, i.e. given a Pango designation that
-starts with an ``X''. We primarily focus on the Long ARG in this section
-since it contains many more recombinants, briefly summarising the status
-of Pango X lineages for the Wide ARG in Section~\ref{sec-pango_x_wide_arg}.
-
-When using Nextclade Pango lineage designations, the
-Long ARG contains 749 samples from 33 recombinant lineages; this
-reduces to 711 and FIXMEXX when singleton recombinants (i.e., recombination
+starts with an ``X''. We focus primarily on the Long ARG in this section,
+since it contains many more recombinants; the status
+of Pango X lineages in the Wide ARG is briefly summarised in Section~\ref{sec-pango_x_wide_arg}.
+
+The Long ARG contains 749 samples designated by Nextclade [CITE] as Pango
+recombinants, classed into 33 Pango lineages; this
+reduces to 711 samples from the same 33 lineages
+when singleton recombinants (i.e. recombination
 nodes subtending only one sample) are removed.
-In contrast, when using GISAID Pango lineage assignments (for the same
-data) we have 515 samples from 38 Pango recombinant lineages,
-indicating that there is substantial uncertainty in the assignment of
-Pango X lineages to samples.
-
-Of the 749 Nextclade-designated recombinant samples, all but two descend
-from a recombination node in the Long ARG.
+In contrast, using the alternative GISAID designation process [CITE], only
+515 samples in the Long ARG are labelled as Pango recombinants,
+which fall into 38 lineages (when singleton recombinants are excluded this
+drops to 514 samples from 35 lineages). The variation in Nextclade versus
+GISAID classifications highlights the frequent uncertainty that exists when
+assigning Pango X lineages to samples.
+
+Below we focus on the 749 Nextclade-designated recombinant samples, all
+but two of which descend from a recombination node in the Long ARG.
 These two samples are designated XP, and a likely
-explanation for the absence of a corresponding recombination event in the long ARG
+explanation for the absence of a corresponding recombination event in the Long ARG
 is that the characteristic multibase deletion for XP
@@
 
-\caption{\label{fig:pango-simple-origin-graph} Non-nested
-Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
+\caption{\label{fig:pango-simple-origin-graph} Examples of non-nested
+Nextclade Pango X lineages. (A) Subgraph for XA in the Long ARG: all five samples designated
 as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
@@
 Pango designation; smaller symbols are non-sample nodes inserted by
-\texttt{sc2ts}, with
-imputed Pango status. Genomic regions inherited by the recombination node are shown
-corresponding to the maximum inferred breakpoint position from the HMM process.
-(B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
-(C) Equivalent plot for both origination events involving the four XD samples in the long ARG.
-Details of the mutations and sample node identities for all three plots are provided in supplementary Figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
+\texttt{sc2ts}, whose Pango status is imputed. Genomic regions inherited by the recombination node are shown;
+breakpoints correspond to the maximum breakpoint position inferred by \texttt{sc2ts}.
+(B) Equivalent plot for the 17 XAG samples in the Long ARG, with abbreviated labelling.
+(C) Equivalent plot for both origination events involving the four XD samples in the Long ARG.
+Details of the mutations and sample node identities for all three plots are provided in
+supplementary Figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}, which also provide alternative GISAID Pango designations.
 } \end{figure}
@@
 
-Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
+Six of the 16 Nextclade designated lineages (XA, XAC, XAE, XF, XK, and XS) are of the
 basic (XA) type with no other Pango designations among their descendants.
@@
 requirement that all the samples assigned to a given lineage must trace back to
-a single recombination event. 15 Pango X lineages are inferred to have multiple
+a single recombination event. Using Nextclade-designations, 15 Pango X lineages are
+inferred to have multiple
 recombinant origins, such that their samples trace to $>$1 most recent
@@
 Investigation of other multiple-origin examples reveals somewhat similar patterns,
-suggesting that most of the simple multiple origin examples are due to Nextclade mislabelling.
+suggesting that most of the simple multiple origin examples are due to incorrect
+Pango labelling.
 
@@
 samples whose ancestry involves multiple non-independent (``nested'') recombination events
-(marked with asterisks in Supplementary Table~\ref{tab:pango-recombinants}; the most
-extensive is for XAB). The earliest example (present in both the Wide and Long ARGs) is XB
-The Long ARG subgraph is plotted in Figure~\ref{fig:complex_origins_graph}, showing a recombination
-between a B.1 sample and B.1.627 sample that leads both to the XB samples and to a
-``hairball'' of further recombination nodes whose descendants are often not
-identified as recombinants by Nextclade (plotted in blue). A similar
-pattern is seen when examining XB in the Wide ARG (see below).
+(marked by~\textdagger ~in Supplementary Table~\ref{tab:pango-recombinants}; the most
+complex appears to be XAB). Figure~\ref{fig:complex_origins_graph} plots the earliest example,
+XB,  which is present in both the Wide and Long ARGs. The subgraph shows a recombination
+between a B.1 sample and B.1.627 sample that leads not only to all the XB-labelled samples
+but also to a ``hairball'' of further recombination nodes whose descendants are often not
+identified as recombinants by Nextclade (plotted on the left, in blue). A similar
+pattern is seen when examining XB in the Wide ARG (see discussion below).
 
@@
 \subsubsection{Wide ARG}
-\label{sec-pango_x_wide_arg}.
+\label{sec-pango_x_wide_arg}
 %% Quick overview of the wide ARG.
@@
 %% Textual description of the 3 recombinant lineages present in the wide ARG
-As in the Long ARG, all XA samples in the wide ARG trace back to a
+As in the Long ARG, all XA samples in the Wide ARG trace back to a
 unique originating recombination node. This is the product of a recombination
@@
 England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of
-the genome from the start up to a maximum position of 23604bp), and two other
+the genome from the start up to a maximum position of 23604bp), and two
 UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of
-27389bp) and B.1.36.8 (the remaining fragment of the genome). Above, we noted that
-in the long ARG the equivalent recombination node has only 2 parents,
+27389bp) and B.1.36.8 (the remaining fragment of the genome). Figure~\ref{fig:complex_origins_graph}
+shows that in the Long ARG the equivalent recombination node has only 2 parents,
 with no involvement of  B.1.36.8; it is possible that the third parent in
-the wide ARG is artefactual. As in the Long ARG, additional
+the Wide ARG is artefactual. As in the Long ARG, additional
 non-X-designated samples such as B.1.634 also descend from this recombination,
@@
 % Next, we examine recombination nodes in the
-% long ARG (sampled to mid-2022) that represent previously unnamed putative
+% Long ARG (sampled to mid-2022) that represent previously unnamed putative
 % recombinant sequences, which do not have a Pango recombinant designation. For
@@
 % subvariants (BA.1, BA.2, BA.4, and BA.5). These recombination nodes (1) were
-% inserted into the long ARG on or after January 8, 2022; (2) have at least 10
+% inserted into the Long ARG on or after January 8, 2022; (2) have at least 10
 % descendent samples; and (3) have no mutational differences (including immediate
@@
 improvements to the current parsimony
-hueristics (Section~\ref{sec:parsimony-heuristics}).
+heuristics (Section~\ref{sec:parsimony-heuristics}).
 A major source of errors are the ``time-traveller'' samples, whose
@@
 While it is unclear how
-we might solve this problem in general, some simple expedients such
+we might solve this problem in general, some simple solutions such
 as filtering out sequences that exceed a given cost in the LS HMM
 (i.e., number of mutations and recombination switches) may work well
-in practise. Such an approach would also reduce the impact of sequences
+in practice. Such an approach would also reduce the impact of sequences
 with high levels of sequencing error (which contribute a large number of
@@
 of the HMM with a single mismatch ratio is simplistic, and it would
-be straightforward to condition on per-site mutation rates (per nucleotide
-state transition, is also possible, with some additional development).
+be straightforward to condition on per-site mutation rates (and nucleotide-dependent
+state transitions, with some additional development).
 Recombination breakpoints for the ARG are currently inserted at the
@@
 It is likely that the basic machinery of finding matches and quantifying
-the undertainty around them under a well-defined statistic model
+the uncertainty around them under a well-defined statistical model
 in large ARGs would have many applications besides those explored here.
@@
 % SARS-CoV-2 lineages. Our results show that the backbone topology of a densely
-% sampled wide ARG inferred using sc2ts agrees well with the backbone topology of
+% sampled Wide ARG inferred using sc2ts agrees well with the backbone topology of
 % a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference
@@
 % genomic evidence. We have found several previously unreported and overlooked
-% recombinant sequences in the wide ARG. After further improvements and
+% recombinant sequences in the Wide ARG. After further improvements and
 % enhancements, we will use sc2ts to build a comprehensive ARG to search for
@@
 standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
-This greedy update strategy inevitably lead to unparsimonious
+This greedy update strategy inevitably leads to unparsimonious
 topologies, and the third step is to then increase the
@@
 and a \emph{mutation} defines the site, node and derived state
-that generates variation (note that this required that the sequences can be aligned and phased).
+that generates variation (note that this requires that the sequences can be aligned and phased).
 Given these tables, we can then efficiently construct the local
@@
 generally be used to efficiently record and analyse SARS-CoV-2 genealogies reconstructed
-using other methods, not only the new approach that we describe here.
+using other methods, not only the \texttt{sc2ts} approach that we describe here.
 
@@
 \citep{Viterbi1967-ol} to find the most likely copying path through the
-sequences, given a set of reference sequences, $Q$, and $E$. Throughout, we refer to the probabilities of mismatching to a member of the reference panel at site $\ell$; $\mu_\ell$,
-and the probability of recombining between two members of the reference panel between site $\ell-1$ and site $\ell$; $r_\ell$). $\mu_\ell$ is often referred to as the
-`mutation probability' for convenience, but we note that this probability of
+sequences, given a set of reference sequences, $Q$, and $E$. Throughout, we refer to the probabilities of mismatching to a member of the reference panel at site $\ell$ as $\mu_\ell$,
+and the probability of recombining between two members of the reference panel between site $\ell-1$ and site $\ell$ as $r_\ell$.
+For convenience, $\mu_\ell$ is commonly referred to as the `mutation probability', but we note that this probability of
 mismatch also encompasses various other error modes which result in a mismatch
@@
 observing the focal sequence up to the site before the region starts, $c$. For a MMR of $k$,
-we only need to consider two of the $n$ members of the reference panel. One;
+we only need to consider two of the $n$ members of the reference panel. One,
 $\mathcal{P}_1$, for which there are no mismatches, but we need to recombine in, and a
-second; $\mathcal{P}_2$, for which there are $k$ (randomly chosen) mismatches in the region.
+second, $\mathcal{P}_2$, for which there are $k$ (randomly chosen) mismatches in the region.
 Now, follow the probability of tracking along each of those paths:
@@
 panel, and $c$ is the likelihood of the path up to this point, which we assume to be equal.
-We then set these path probabilities equal, and rearrange to obtain:
+We then set these path probabilities to be equal, and rearrange to obtain:
 \begin{align*}
@@
 a subset of nucleotides coded as `unknown' at a set of positions. As a result, we had to
-consider missingness in our implementation of LS Viterbi. One approach would be to `impute'
+consider missingness in our implementation of the LS Viterbi algorithm. One approach would be to `impute'
 all missing data beforehand using external software to hard-call or define a
@@
 
-In \texttt{sc2ts} we use the efficient ARG-based implementation of the
+In \texttt{sc2ts}, we use the efficient ARG-based implementation of the
 LS Viterbi algorithm from \texttt{tsinfer} \citep{Kelleher2019-ba} to find
@@
 the most likely solution is to copy from one
-of the nodes in the ARG that minimised the number of mutations required
-on the query sequence. Importantly, because the reference panel here consists of
+of the nodes in the ARG that minimises the number of mutations required
+to insert the query sequence. Importantly, because the reference panel here consists of
 \emph{every} node in the ARG, we can match to both older sample
@@
 algorithms which determine the probability of observing a focal sequence,
-integrating over all possible paths through the data, under LS model [cite
+integrating over all possible paths through the data, under the LS model [cite
 D-Palms]. This presents an opportunity to estimate parameters of interest while
-exploiting the tree-structure to gain log based speed-ups in exact their exact
+exploiting the tree structure to gain log-based speed-ups in their exact
 implementation.
@@
 While some of these samples will require no extra mutations
-(i.e, they are identical to the attachment node), in general there
+(because they are identical to the attachment node), in general there
 will be complex patterns of shared mutations among the samples
@@
 
-\subsection{Parsimony increasing heuristics}
+\subsection{Parsimony-increasing heuristics}
 \label{sec:parsimony-heuristics}
@@
 Mutation collapsing inspects the siblings of a newly attached node
-to check if any of them share (a subset) of the mutations that
+to check if any of them share (a subset of) the mutations that
 it carries. If so, we increase the overall parsimony of the inference
@@
 
-The reversion push operation inspects newly added nodes to see
+The reversion push operation inspects a newly added node to see
 if any of its mutations are ``immediate reversions''; that is,
@@
 immediate parent. We increase the overall parsimony of the
-inference by ``pushing in'' a new node in which descends from the
+inference by ``pushing in'' a new node which descends from the
 original parent which carries all its mutations except those
@@
 are likely to be mostly artefactual, it is clear that there
-is room for improvement, and that further parsimony increasing
+is room for improvement, and that further parsimony-increasing
 heuristics would likely be of benefit.
@@
 analysis and to help understand the robustness of recombinants
-we perform some additional steps in \texttt{sc3ts}.
+we perform some additional steps in \texttt{sc2ts}.
 
@@
 more parsimonious breakpoints may also be a worthwhile avenue for development.
-In particular, we may choose to insert breakpoints for the ARG that are choosed
+In particular, we may choose to insert breakpoints for the ARG that are chosen
 from within the possible interval, rather than the current approach of taking
-the right-most value. Cases where we have more than two parents may either be
+the rightmost value. Cases where we have more than two parents may either be
 the ``stacking'' of multiple recombination events, or instances where
@@
 if they (1) had ambiguous collection dates;
-(2) were collected before January 01, 2020 or after June 30, 2022; or
+(2) were collected before January 1, 2020 or after June 30, 2022; or
 (3) were isolated from a non-human host.
@@
 A major source of error in early versions of \texttt{sc2ts} was
-the existance of ``time traveller'' sequences: those with erroneously
+the existence of ``time traveller'' sequences: those with erroneously
 early collection dates.
 For example, an Alpha sample purportedly collected in 2020 from the United States,
-before Alpha appeared in the United Kingdom (USA/MN-Mayo-1563/2020)
+before Alpha appeared in the United Kingdom (USA/MN-Mayo-1563/2020),
 produced significant topological distortions in
-ARGs inferred early on in the study.
+inferred ARGs.
 Hence, to exclude such potential ``time travellers'' we employ two filters.
@@
 This method is fast, as the lineages for most of the internal nodes can be
-imputed by copying from the surrounding nodes (66\% of nodes for the long ARG,
-80\% of nodes for the wide ARG), and significantly more efficient than extracting
+imputed by copying from the surrounding nodes (66\% of nodes for the Long ARG,
+80\% of nodes for the Wide ARG), and significantly more efficient than extracting
 the haplotypes for each internal node and using existing Pangolin assignment tools
@@
 \includegraphics[width=\textwidth]{figures/supp_cophylogeny_long.pdf}
-\caption{\label{fig:cophylogeny_long}Tanglegram equivalent to that in Figure~\ref{fig:cophylogeny}, but for the long ARG (i.e. subsampled to mid-2022). }
+\caption{\label{fig:cophylogeny_long}Tanglegram equivalent to that in Figure~\ref{fig:cophylogeny},
+but for the Long ARG (i.e. subsampled to mid-2022).}
 \end{figure}
@@
 \multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
-\multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\ \hline
+\multicolumn{2}{c}{\textbf{sc2ts (Wide ARG)}} \\ \hline
 \textbf{Sample} &
@@
 \end{tabular}
-\caption{\label{tab:jackson_supplement}Recombinant sequences involving the Alpha (B.1.1.7) variant reported by \cite{Jackson2021-ik} have recombinant ancestry in the wide ARG.
+\caption{\label{tab:jackson_supplement}Recombinant sequences involving the Alpha (B.1.1.7) variant reported by \cite{Jackson2021-ik} have recombinant ancestry in the Wide ARG.
 The breakpoint intervals and Pango lineage assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et al., except the Pango lineage assignment of the parents of group B recombinants, which were taken from Table 1 (motif-based results).}
@@
 \caption{\label{tab:pango-recombinants}
-Summary of the Pango X-lineages in the long ARG (excluding XP and XAK whose samples are entirely filtered out, see text).
+Summary of the Pango X-lineages in the Long ARG (excluding XP and XAK whose samples are entirely filtered out, see text).
 In cases of multiple origins, most X-lineages have a single ``main clade'' (in bold).
@@
 to map tskID to EPI\_ISL and strain name).
-Note that GISAID does not designate any nodes as XD in the long ARG, hence no recombinant Pango lineages
+Note that GISAID does not designate any nodes as XD in the Long ARG, hence no recombinant Pango lineages
 are marked in this plot. From inspection of the samples, we believe the GISAID designations to be erroneous

### 2023-05-25 paper.tex
Responses to comments

@@
 % obvious.
-Most of the nodes in the ARGs represent sample genomes
-(Wide ARG: 87\%, Long ARG: 84\%), but the inferred ancestral sequences
-represented by the nodes inferred from daily sample clusters
-by UPGMA (Section~\ref{sec:sample-cluster-tree-inference})
-and parsimony heuristics (Section~\ref{sec:parsimony-heuristics})
-contribute significantly.
+The majority of the nodes in the ARGs represent sample genomes
+(Wide ARG: 87\%, Long ARG: 84\%), with the remainder mostly representing
+the ancestral sequences inferred by UPGMA 
+from daily sample clusters (Section~\ref{sec:sample-cluster-tree-inference})
+and parsimony heuristics (Section~\ref{sec:parsimony-heuristics}).
 Table~\ref{tab:args} shows the counts of edges (see
@@
 as line segments (coloured by interval width). The inset histogram shows the
-distribution of these interval widths.
+distribution of these interval widths (truncated at 10kb).
 The bottom panel shows the number of intervals that intersect with
@@
 % present a way in which it can be incorporated. Our results are accurate.
-
-The COVID-19 pandemic continues to be a global emergency [DOES IT? Didn't the
-WHO downgrade recently?]
+Although the COVID-19 pandemic is no longer considered a global emergency
+by the WHO, the prevalence of SARS-CoV-2 continues to be high worldwide
 with persistent high levels of infection. This high prevalence has allowed the

### 2023-05-25 paper.tex
Tweaks to first methods sections

@@
 sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
-Pango designation; smaller symbols are non-sample nodes inserted by
-\texttt{sc2ts}, whose Pango status is imputed. Genomic regions inherited by the recombination node are shown;
+Pango designation; smaller symbols are non-sample nodes inserted by \texttt{sc2ts}, whose
+Pango status is imputed. Genomic regions inherited by the recombination node are shown;
 breakpoints correspond to the maximum breakpoint position inferred by \texttt{sc2ts}.
-(B) Equivalent plot for the 17 XAG samples in the Long ARG, with abbreviated labelling.
+(B) Equivalent plot for the 17 XAG samples in the Long ARG, with abbreviated labelling. Where
+non-XAG samples are ancestral to other samples, the number of descendant samples is marked as ``+1'',
+``+7'', etc.
 (C) Equivalent plot for both origination events involving the four XD samples in the Long ARG.
@@
 \section{Methods}
-
-[TODO update this summary para to avoid duplications with
-later sections and to add the appropriate signposts
-when they are done(ish)]
-
 Sc2ts (pronounced ``scoots'', optionally) is a method for inferring
-ARGs from densely sampled pandemic-scale data
+Ancestral Recombination Graphs (ARGs; see Section~\ref{sec:args})
+from densely sampled pandemic-scale data
 in real time, in which recombination occurs at a low but significant rate.
-The essential idea is to incrementally update the ARG each day
-with the sequences collected on that day. These updates are performed by
-first finding likely ``copying paths'' for each sequence in the daily
-batch to the current ARG, using the ARG-based implementation of
-the Li and Stephens model~\citep{Li2003-ib} introduced
-in \texttt{tsinfer}~\citep{Kelleher2019-ba}.
+% The overall algorithm is summarised in Figure~\ref{fig:overview_sc2ts},
+% and with further details in subsequent subsections.
+The basic idea is to incrementally update an ARG each day
+with the sequences collected on that day (Figure~\ref{fig:overview_sc2ts}).
+The first step is to find likely ``copying paths'' under the Li and Stephens model
+for each sequence in the daily batch to the current ARG (Section~\ref{sec:ls}).
 These copying paths
@@
 with a small number of mutations, and often many samples
-from a daily batch will copy from the same ARG node. The second
-step is then to ``resolve'' these implied polytomies by using
-standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
+from a daily batch will copy from the same ARG node.
+The second step is then to ``resolve'' these implied polytomies by using
+standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C;
+Section~\ref{sec:sample-cluster-tree-inference}).
 This greedy update strategy inevitably leads to unparsimonious
@@
 overall parsimony of the inferred ARG by making some simple topological
-updates (Figure~\ref{fig:overview_sc2ts}C and D).
-The operations are carried out efficiently using the well-established
+updates (Figure~\ref{fig:overview_sc2ts}C,D;
+Section~\ref{sec:parsimony-heuristics}).
+Recombination is inferred as an integral and ongoing part of this
+process, requiring only a few additional steps to facilitate
+later analysis (Section~\ref{sec:treatment_recombinants}).
+The results of inference for a given day is then
+a joint model of mutation, genetic inheritance and recombination
+for all of the sequences inserted into the ARG up to that day,
+which can be conveniently and efficiently analysed using the
+mature and feature-rich
 \texttt{tskit}
-library~\citep{Kelleher2018-xc,Ralph2020-efficiently,Tskit2023-tskit},
-and the resulting structure can be conveniently and interactively
-analysed using the many features provided by \texttt{tskit}.
+library~\citep{Kelleher2018-xc,Ralph2020-efficiently,Tskit2023-tskit}.
 
@@
 \caption{\label{fig:overview_sc2ts}
-A schematic of the sc2ts method. Sc2ts reconstructs the genetic relationships
+A schematic of the \texttt{sc2ts} method.
+Sc2ts reconstructs the genetic relationships
 among SARS-CoV-2 genomes by copying samples to all possible ancestors collected
-at earlier time points (curved arrows). An ARG is built by attaching new sample
-genomes grouped by collection date to an existing ARG rooted at the Wuhan-Hu-1
-reference genome, day by day. Each iteration involves three stages: (1)
-attachment of new samples to the growing ARG (A, B); (2) reconstruction of
+at earlier time points (curved arrows). Each daily iteration involves three stages:
+(1) attachment of new samples to the growing ARG (A, B); (2) reconstruction of
 trees relating the samples under each attachment node (C); and parsimony-based
@@
 we can think of an ARG as being any graph that encodes the
-reticulate genetic ancestry of a sample of sequences under
+reticulate genetic ancestry of a sample of colinear sequences under
 the influence of recombination.
-This definition encompasses various types of graphs traditionally
+This definition encompasses various types of graphs often
 described using the broader term of phylogenetic networks.
@@
 and a \emph{mutation} defines the site, node and derived state
-that generates variation (note that this requires that the sequences can be aligned and phased).
+that generates variation.
+% JK: Note sure this clarifies much at the representational level?
+% (note that this requires that the sequences can be aligned and phased).
+
 Given these tables, we can then efficiently construct the local
@@
 Fan2022-genealogical,Korfmann2022-weak,
-Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}.
+Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier,
+Zhang2023-lf}.
 It is important to note that
@@
 using other methods, not only the \texttt{sc2ts} approach that we describe here.
+Note also that there is no requirement that recombination be present,
+and the methods are also very efficient when working with a single tree.
 
@@
 \subsection{Treatment of recombinants}
+\label{sec:treatment_recombinants}
 A sample sequence is designated as a recombinant if the most likely

### 2023-05-25 paper.tex
Pass through LS model

@@
 \usepackage{amsmath}
+\usepackage{amsfonts}
 \usepackage{graphicx}
@@
 many of the important features of the joint processes of mutation and
-recombination. It is a Hidden Markov Model (HMM) in which a focal genome is a
+recombination. It is a Hidden Markov Model (HMM) in which a focal genome
+is modelled as a
 sequence of nucleotides that are probabilistically emitted as
@@
 imperfect mosaic of the sequences in a reference panel.
-Black crosses along the focal sequence show genotyping/sequencing
+Black crosses along the focal sequence show sequencing
 errors or mutations.
@@
 focal sequence and the hidden state from which it is copied (due to mutation as
-well as genotyping or sequencing error). Both $E$ and $Q$ may be a function of the reference panel members,
-but transitions are generally assumed to be independent of the hidden states (Figure~\ref{fig:ls_diagram}, pink panel).
+well as sequencing error).
+Both $E$ and $Q$ may be a function of the reference panel members,
+but transitions are generally assumed to be independent of the
+hidden states (Figure~\ref{fig:ls_diagram}, pink panel).
 This assumption dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
@@
 focal sequence, $a_\ell$) at the site under consideration are equally likely and
-occur with mutation probability $\frac{\mu_\ell}{a_\ell-1}$ (Figure~\ref{fig:ls_diagram}, blue panel).
-We use the Viterbi algorithm
-\citep{Viterbi1967-ol} to find the most likely copying path through the
-sequences, given a set of reference sequences, $Q$, and $E$. Throughout, we refer to the probabilities of mismatching to a member of the reference panel at site $\ell$ as $\mu_\ell$,
+occur with mutation probability $\mu_\ell/(a_\ell-1)$
+ (Figure~\ref{fig:ls_diagram}, blue panel).
+We use the Viterbi algorithm~\citep{Viterbi1967-ol}
+to find the most likely copying path, given a set of reference sequences, $Q$, and $E$.
+Throughout, we refer to the probabilities of mismatching to a
+member of the reference panel at site $\ell$ as $\mu_\ell$,
 and the probability of recombining between two members of the reference panel between site $\ell-1$ and site $\ell$ as $r_\ell$.
@@
 Assuming that recombination events occur as a Poisson process,
-then the probability of recombining between adjacent sites is:
+then the probability of recombining between adjacent sites is
 \begin{align*}
@@
 \end{align*}
-where $m_j$ is the nucleotide position of variant $j$. Similarly, we may consider the probability of mutation to represent to the
+where $m_j$ is the nucleotide position of variant $j$.
+Similarly, we may consider the probability of mutation to represent to the
 probability of mutation along the branch joining the focal sequence to a coalescent
 tree containing a reference panel with $n$ members, assuming a Poisson process
-along the branch. The expected length of this branch is $\frac{4}{n}$, and so the mismatch
+along the branch. The expected length of this branch is $4/n$, and so the mismatch
 probability represents the probability of observing a mutation or error mode,
@@
 \end{align}
-where $\nu$ is the mutation rate per $\mathit{ploidy} \times N_e$ generations.
-The probabilities of these competing process of mismatch and recombination are usually controlled by the site-specific parameters $\mu_\ell$ and $r_\ell$, respectively.
-For this work we used a slightly different formulation, which uses a parameter, the mismatch ratio (MMR), to control the relative importance of mutation
-and recombination in the HMM. Specifically, an MMR value of $k$ will
-prefer $k$ mismatches (mutations) to a single recombination which results in copying from a template with no mismatches.
-To map between recombination probability and mutation probability for a particular mismatch
-ratio, we simply consider the two paths which we wish to be equally likely, and rearrange
-for the mutation or recombination probability. Consider the simple case when we assume
-$\mu_l=\mu$ and $r_l=r$. Without loss of generality, consider a region of length $m$. Up
-until this region, two paths are equally likely, so we can re-scale by the likelihood of
-observing the focal sequence up to the site before the region starts, $c$. For a MMR of $k$,
-we only need to consider two of the $n$ members of the reference panel. One,
-$\mathcal{P}_1$, for which there are no mismatches, but we need to recombine in, and a
-second, $\mathcal{P}_2$, for which there are $k$ (randomly chosen) mismatches in the region.
+where $\nu$ is the mutation rate per $\mbox{ploidy} \times N_e$ generations.
+
+The probabilities of these competing process of mismatch and recombination are
+usually controlled by the site-specific parameters $\mu_\ell$ and $r_\ell$,
+respectively. For this work we used a slightly different formulation, which
+uses one parameter, the mismatch ratio (MMR), to control the relative
+importance of mutation and recombination in the HMM. Specifically, an MMR value
+of $k$ will prefer $k$ mismatches (mutations) to a single recombination which
+results in copying from a template with no mismatches. To map between
+recombination probability and mutation probability for a particular mismatch
+ratio, we simply consider the two paths which we wish to be equally likely, and
+rearrange for the mutation or recombination probability. Consider the simple
+case when we assume $\mu_l=\mu$ and $r_l=r$. Without loss of generality,
+consider a region of length $m$. Up until this region, two paths are equally
+likely, so we can re-scale by the likelihood of observing the focal sequence up
+to the site before the region starts, $c$. For an MMR of $k$, we only need to
+consider two of the $n$ members of the reference panel.
+
+One, $\mathcal{P}_1$,
+for which there are no mismatches, but we need to recombine in, and a second,
+$\mathcal{P}_2$, for which there are $k$ (randomly chosen) mismatches in the
+region.
 Now, follow the probability of tracking along each of those paths:
 \begin{align*}
-P(\mathcal{P}_1) &= \frac{cr}{n} \left(\left(1-r\right)+\frac{r}{n}\right)^m\left(1-\mu\right)^m &\text{recombine to a template with no mismatches,}\\
-P(\mathcal{P}_2) &= c \left(\left(1-r\right) + \frac{r}{n}\right)\left(\left(1-r\right)+\frac{r}{n}\right)^m \left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
+\mathbb{P}[\mathcal{P}_1] &= \frac{cr}{n}
+\left(\left(1-r\right)+\frac{r}{n}\right)^m\left(1-\mu\right)^m
+&\text{recombine to a template with no mismatches,}\\
+\mathbb{P}[\mathcal{P}_2] &= c
+\left(\left(1-r\right) +
+\frac{r}{n}\right)\left(\left(1-r\right)+\frac{r}{n}\right)^m
+\left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
 \end{align*}
-where $r$ is the recombination probability between adjacent variants, $\mu$ is the mismatch
-probability between a target sequence at a variant and a particular member of the reference
-panel, and $c$ is the likelihood of the path up to this point, which we assume to be equal.
-We then set these path probabilities to be equal, and rearrange to obtain:
-\begin{align*}
-r = \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k}, \quad\quad
-\mu = \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}.
-\end{align*}
-Under a lower MMR value (for which recombinations are favoured over mutations), the maximum-likelihood HMM copying path of a
-focal genome more likely spans at least two ancestral genomes (i.e,. the focal
-genome is a putative recombinant) than a single ancestral genome. Here, the MMR
-was set to three in all the HMM runs.
-
-A varying degree of missingness is present in the genetic data, that is, for a given sample,
-a subset of nucleotides coded as `unknown' at a set of positions. As a result, we had to
-consider missingness in our implementation of the LS Viterbi algorithm. One approach would be to `impute'
-all missing data beforehand using external software to hard-call or define a
-distribution over the possible missing nucleotides, and use that in the determination of path
-probabilities. Another is to not attempt to estimate the true nucleotide, but rather to
-consider such sites as uninformative to the path probability, but to impute missing data
-within that path on the fly, placing them within the reference panel iteratively. We chose the latter. To do this, we simply encode the emission
-probability where the underlying nucleotide transition is $\mathit{unknown}\rightarrow \{A,C,G,T\}$ as 1.
+% where $r$ is the recombination probability between adjacent
+% variants, $\mu$ is the mismatch probability between a target sequence at a
+% variant and a particular member of the reference panel,
+where $c$ is the
+likelihood of the path up to this point, which are assumed to be
+equal by construction.
+We then
+set these path probabilities to be equal, and rearrange to obtain:
+\begin{align*} r = \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k},
+\quad\quad \mu = \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}. \end{align*}
+Thus, for lower MMR values ($k$ here),
+recombinations are favoured over mutations
+than a single ancestral genome. We use an MMR value of $k=3$ in this work,
+as it provided results that tallied well with empirical observations.
+Exploring different mismatch ratios, and more sophisticated parameterisations
+of the HMM are important avenues for future work.
+
+A varying degree of missingness is present in the genetic data, that is, for a
+given sample, a subset of nucleotides coded as `unknown' at a set of positions.
+As a result, we had to consider missingness in our implementation of the LS
+Viterbi algorithm. One approach would be to `impute' all missing data
+beforehand using external software to hard-call or define a distribution over
+the possible missing nucleotides, and use that in the determination of path
+probabilities. Another is to not attempt to estimate the true nucleotide, but
+rather to consider such sites as uninformative to the path probability, and
+impute missing data within that path on the fly, placing them within the
+the emission probability transition from unknown to
+the emission probability from unknown to
+to any other state as 1.
+% $\mbox{N} \rightarrow \star$ as 1.
 
@@
 given sample sequence.
-This statement warrants emphasis: the Viterbi algorithm enables us to find a path through the reference panel from among the $n^m$ paths that is provably at the optimum, under the LS model.
-
+
+The Viterbi algorithm enables us to find a path through
+the reference panel from among the $n^m$ paths that is provably at the
+optimum, under the LS model.
 We can solve this massive optimisation problem exactly because the ARG-based
@@
 relatively flat likelihood spaces where many different paths have equal
-likelihood. There are many possibilities in using established HMM methodology
+likelihood.
+There are many possibilities in using established HMM methodology
 to reason about and explore the space of possible matches, which may be a
-fruitful avenue for future work. Examples include a stochastic traceback
+fruitful avenue for future work.
+% Could talk about running the HMM in reverse here, but is there much
+% point? It's confusing the main point which is that there lots of
+% ways in which this could be done properly.
+Examples include a stochastic traceback
 through the collection of paths at the global optimum to glean further

### 2023-05-25 paper.tex
Add in-prep LS ARG paper

--- 2023-05-25 paper.tex
@@
 algorithms which determine the probability of observing a focal sequence,
-integrating over all possible paths through the data, under the LS model [cite
-D-Palms]. This presents an opportunity to estimate parameters of interest while
-exploiting the tree structure to gain log-based speed-ups in their exact
-implementation.
+integrating over all possible paths through the data, under the LS
+model~\citep{Palmer2023-efficient}.
+This presents an opportunity to estimate parameters of interest
+under the LS model at pandemic scale.
 

### 2023-05-25 paper.tex
More updates

@@
 % of days. ***describe possible approaches to correct this?*** \item To avoid
+
 % time travellers, we used a maximum submission delay, but this will
@@
 attach to a single node. The special treatment afforded to
-newly-detected putative recombinants are discussed below
-in Section XXX.
+newly-detected putative recombinants are discussed
+in Section~\ref{sec:treatment_recombinants}.
 
 With tens of thousands of samples being added to the ARG per day,
-we can have clusters of hundreds of sequences attaching to the same node.
+there are often clusters of hundreds of sequences attaching to the same node.
 While some of these samples will require no extra mutations
@@
 
-We currently use the implementation of the UPGMA algorithm~\citep{Michener1957-tr}
-from SciPy~\citep{Pauli2020-scipy} to build trees from sample
+We currently use the UPGMA algorithm~\citep{Michener1957-tr}
+as implemented in SciPy~\citep{Pauli2020-scipy} to build trees from sample
 clusters, and map mutations back to this topology using maximum parsimony.
@@
 supported by no informative site (i.e., having no mutation immediately
-ancestral to them). To avoid such false precision we post-process
+ancestral to them). We avoid such false precision by post-processing
 to remove unsupported internal nodes, representing the relationship
-between $k$ identical descendants of the node as a polytomy of size $k$.
+between $k$ identical descendants of a node as a polytomy of size $k$.
 
@@
 evolutionary relationships~\citep{Felsenstein2004-inferring}.
-Since this within-cluster tree building has a significant influence on the
-overall ARG topology (see Results), implementing more sophisticated
-tree building methods which keep track of the location of the required mutations
+Table~\ref{tab:args} show that
+this within-cluster tree building has a significant influence on the
+overall ARG topology (Wide ARG: 7.05\% of nodes and parent of
+32.27\% of edges; Long ARG: 6.61\% and 19.97\%, respectively),
+and therefore applying more sophisticated
+tree building methods which keep track of the required mutations
 (rather then inferring post-hoc by parsimony), is a likely avenue
@@
 Attaching trees built from the clusters of samples that copy from
-a particular node (or path of nodes for recombinants, see the
-next section) under the HMM is an inherently greedy strategy,
+a particular node (or path of nodes for recombinants,
+see Section~\ref{sec:treatment_recombinants}) under the
+HMM is an inherently greedy strategy,
 and can produce inferences that are clearly unparsimonious.
 The final step in adding a daily batch of samples to the ARG
-is then to perform some local updates that target specific
+is therefore to perform some local updates that target specific
 types of parsimony violation in the just-updated regions of the
@@
 applied, which we refer to as ``mutation collapsing'' and ``reversion
-pushing''.
+pushing'' (Figure~\ref{fig:overview_sc2ts}D, E).
 
@@
 implementation uses a simple greedy strategy for choosing
-the particular mutations to collapse. [TODO back ref to the
-Results section to discuss how important these nodes are]
+the particular mutations to collapse.
+Table~\ref{tab:args} shows that
+while these nodes constitute a small fraction of the
+overall total (Wide ARG: 2.77\%, Long ARG: 4.82\%),
+they are parental to a much larger fraction of the
+edges (Wide ARG: 12.77\%, Long ARG: 20.47\%).
 
@@
 causing the reversions on the newly added node.
-[TODO back ref to the
-Results section to discuss how important these nodes are]
+Table~\ref{tab:args} shows that
+these nodes constitute a similar fraction of the
+overall numbers of nodes (Wide ARG: 2.79\%, Long ARG: 4.39\%)
+and edges (Wide ARG: 12.66\%, Long ARG: 18.46\%)
+as mutation collapsing.
 
@@
 of the inferred ARGs, most noticably in terms of the number of
-reversion mutations inferred. However, they are both simple
+reversion mutations inferred. These nodes are also being chosen by the
+LS HMM as likely choices of parent (as indicated by the node and edge
+counts quoted above), also demostrating that the hueristics are successfully
+capturing features of real sequences.
+However, they are both simple
 greedy operations, just examining the local parts of the
@@
 the succinct tree sequence data structure and \texttt{tskit} library
-(see the Ancestral Recombination Graphs section above). To facilitate
+(Section~\ref{sec:args}). To facilitate
 analysis and to help understand the robustness of recombinants
@@
 (see Section~\ref{sec:sample-cluster-tree-inference} above for details).
-
 When the Viterbi algorithm implementation used by the LS HMM infers
@@
 switches from one parent node to another is the last possible
-position.
-Thus, the breakpoints used in the ARG represent the rightmost end of the
-interval, within which the recombination event could have occurred. To obtain
-the leftmost end of this interval, we rerun the HMM using mirrored coordinates
-(i.e., from the 3’ end to the 5’ end) and store the information as
-metadata on the recombination node.
-These left- and rightmost positions
-are the recombination intervals reported throughout.
-Ultimately all the recombination intervals mean is that the parent sequences
-are identical within that interval, and there is therefore no information
-about where the recombination occurred. There are many different processes
-that can contribute to the size of these intervals, and therefore
-understanding how these processes interact to produce wider or narrower
-intervals is an important avenue for future work.
+position. The left-most extent of the breakpoint interval is derived
+by sequence comparison between the parents, as described in
+Section~\ref{sec:breakpoint_intervals}.
+
+% Thus, the breakpoints used in the ARG represent the rightmost end of the
+% interval, within which the recombination event could have occurred. To obtain
+% the leftmost end of this interval, we rerun the HMM using mirrored coordinates
+% (i.e., from the 3’ end to the 5’ end) and store the information as
+% metadata on the recombination node.
+% These left- and rightmost positions
+% are the recombination intervals reported throughout.
+% Ultimately all the recombination intervals mean is that the parent sequences
+% are identical within that interval, and there is therefore no information
+% about where the recombination occurred. There are many different processes
+% that can contribute to the size of these intervals, and therefore
+% understanding how these processes interact to produce wider or narrower
+% intervals is an important avenue for future work.
 
@@
 ways to extend and improve. For example, the current parameterisation of using
-a single ``mismatch ratio'' is very simplistic (see section XXX), and likely
-results in a very flat likelihood space where many sequences have equal
-likelihood of being chosen. More sophisticated parameterisations where we
+a single ``mismatch ratio'' is very simplistic (Section~\ref{sec:ls}), and likely
+results in a flat likelihood space where many recombinant paths have equal
+probability of being chosen.
+More sophisticated parameterisations where we
 condition on known per-site mutation rates (and perhaps per-base subsitution

### 2023-05-25 paper.tex

@@
 logarithmically with reference panel size (as opposed to linearly,
-for standard matrix-based approaches)~\citep{Kelleher2019-ba}.
+for standard matrix-based approaches).
 This efficient HMM algorithm is the main reason for \texttt{tsinfer}'s
@@
 % ways in which this could be done properly.
-Examples include a stochastic traceback
+Examples include stochastic traceback~\citep[e.g.,][]{rasmussen2014genome}
 through the collection of paths at the global optimum to glean further
@@
 small values when creating new nodes using parsimony rules
-(Section~\ref{sec:parsimony-heuristics}). Substantial improvements could
-likely be made using simple node dating methods.
+(Section~\ref{sec:parsimony-heuristics}). Since the ARGs here are
+very treelike, with recombination nodes consituting a tiny fraction
+of the overall topology (Table~\ref{tab:args}), existing methods
+\cite[e.g.,][]{to2016fast} could likely be adapted to accurately
+date the vast majority of the nodes.
 
@@
 based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
-up to Aug 22, 2022 and accessible at
-\url{https://doi.org/10.55876/gis8.230329cd} (Supplementary Table 1).
+up to 2022-08-22 and accessible at
+% Do we need to upload a separate Table or is this DOI sufficient?
+\url{https://doi.org/10.55876/gis8.230329cd}.
 We removed sequences
-if they (1) had ambiguous collection dates;
-(2) were collected before January 1, 2020 or after June 30, 2022; or
-(3) were isolated from a non-human host.
+if they had ambiguous collection dates, were collected before 2020-01-01
+or were isolated from a non-human host.
 We aligned sequences to the Wuhan-Hu-1/2019 reference sequence
@@
 the approach used in the ``faToVCF'' tool used by
-UShER~\citep{Turakhia2022-it}.
+UShER~\citep{Turakhia2021-ur}.
 Sites that are masked by this process are
@@
 Hence, to exclude such potential ``time travellers'' we employ two filters.
-The first is to remove any sequence with a collection date that pre-dates the
+
+The first filter is a simple threshold on
+the time delay between the collection date and submission date.
+After some preliminary analysis we settled on a maximum submission delay
+of 30 days when building the ARGs described here.
+The second filter is to remove any sequence with a collection date that
+pre-dates the
 time to the most recent common ancestor (tMRCA) of its corresponding clade.
 We obtained the tMRCA for each clade from a Nextstrain GISAID reference tree
-(downloaded on August 22, 2022), and we used the lower bound of the 95\%
+(downloaded on 2022-08-22), and we used the lower bound of the 95\%
 confidence interval of each clade as the minimum date cut-off.
-This removed NN [Katherine can you fill in please?] sequences.
-
-This filter removes the most egregious time travellers, but many still
-remain. We thus employed a second filter which is a simple threshold on
-the time delay between the collection date and submission date.
-After some preliminary analysis we settled on a maximum submission delay
-of 30 days when building the ARGs described here as a reasonable compromise.
+This excludes a further 618 samples not covered by the maximum submission
+delay filter.
 
@@
 the list of lineage-defining mutations (based on 90\% consensus
-of the sequences analysed) from the COVID-CG website \citep[][\url{https://covidcg.org/}; accessed on 4 November 2022]{Chen2021-zc}.
+of the sequences analysed) from the
+COVID-CG website \citep[][\url{https://covidcg.org/};
+accessed on 2022-11-04]{Chen2021-zc}.
 
@@
 This method is fast, as the lineages for most of the internal nodes can be
-imputed by copying from the surrounding nodes (66\% of nodes for the Long ARG,
-80\% of nodes for the Wide ARG), and significantly more efficient than extracting
+imputed by copying from the surrounding nodes
+(Wide ARG: 80\% of nodes,
+Long ARG: 66\%),
+and significantly more efficient than extracting
 the haplotypes for each internal node and using existing Pangolin assignment tools
@@
 imputation, we have used our method to re-impute the lineage designations of each
-sample node using the surrounding information; this results in 98\% (resp.~99\%) of
-nodes being assigned the same lineage as per the source metadata in the long
-(resp.~wide) ARG.
+sample node using the surrounding information; this results in
+99\% of nodes being assigned the same lineage as per the source metadata in the
+Wide ARG, and 98\% in the Long ARG.
 
@@
 Submitting laboratories for generating the genetic sequence and metadata and
-sharing via the GISAID Initiative, on which this research is based. Also, we
-thank Dr. Morag Graham (the National Microbiology Laboratory, Winnipeg, Canada)
-for kindly providing the genomic coordinates of the breakpoint sequence motifs.
+sharing via the GISAID Initiative, on which this research is based.
+% Not using these now:
+% Also, we
+% thank Dr. Morag Graham (the National Microbiology Laboratory, Winnipeg, Canada)
+% for kindly providing the genomic coordinates of the breakpoint sequence motifs.
 SHZ is supported by the Janssen-Oxford Translational Genomics Fellowship. JK,

### 2023-05-25 paper.tex
Tidyup pass through bib

@@
 along this branch. We can then approximate the probability of observing a mismatch as:
-\begin{align}
+\begin{align*}
     \mu_\ell \approx 1 - \exp{\left(-\frac{4\nu}{n}\right)}
-\end{align}
+\end{align*}
 where $\nu$ is the mutation rate per $\mbox{ploidy} \times N_e$ generations.
@@
 
+\clearpage
 \renewcommand\thefigure{S\arabic{figure}}
@@
 \section*{Supplementary}
-\begin{figure} \centering
+
+\begin{figure}[h] \centering
 \includegraphics[width=\textwidth]{figures/supp_cophylogeny_long.pdf}
@@
 total; and 2 HMM consistent (20 inconsistent) Delta-Omicron recombination
-breakpoints = 2\% (< 1\%) of the total. These plots do not include
+breakpoints = 2\% ($<1$\%) of the total. These plots do not include
 recombination breakpoints involved lineages that were not classified into these

### 2023-05-26 paper.tex
Yans updates

@@
 as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
-sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
+sampled viral genome; dotted lines show ARG continuation. Vertical position of nodes does not
+show absolute time, but simply relative rank (parents above children). Nodes are coloured by Nextclade
 Pango designation; smaller symbols are non-sample nodes inserted by \texttt{sc2ts}, whose
@@
 breakpoints correspond to the maximum breakpoint position inferred by \texttt{sc2ts}.
-(B) Equivalent plot for the 17 XAG samples in the Long ARG, with abbreviated labelling. Where
+(B) Equivalent subgraph for the 17 XAG samples in the Long ARG, with abbreviated labelling. Where
 non-XAG samples are ancestral to other samples, the number of descendant samples is marked as ``+1'',
 ``+7'', etc.
-(C) Equivalent plot for both origination events involving the four XD samples in the Long ARG.
+(C) Equivalent subgraphs for both origination events involving the four XD samples in the Long ARG.
 Details of the mutations and sample node identities for all three plots are provided in
@@
 lineages for recombination events within and among different VoC categories.
-HMM-consistent recombinants are shown in green, HMM-inconsistent in blue
-(including many of the named Delta-Omicron recombinants). We find 69 HMM
-consistent (77 inconsistent) Alpha-Alpha recombination breakpoints = 13\%
-(10\%) of the total; 84 HMM consistent (143 inconsistent) Delta-Delta
-recombination breakpoints = 18\% (18\%) of the total; 99 HMM consistent (155
-inconsistent) Omicron-Omicron recombination breakpoints = 22\% (19\%) of the
-total; and 2 HMM consistent (20 inconsistent) Delta-Omicron recombination
-breakpoints = 2\% ($<1$\%) of the total. These plots do not include
-recombination breakpoints involved lineages that were not classified into these
-VoC categories}
+There are 78 Alpha|Alpha recombination breakpoints corresponding to 75 recombination nodes 
+(25 breakpoints / 24 nodes with $\geq5$ descendant samples).
+148 breakpoints from 142 nodes are Delta|Delta recombinations (45 breakpoints / 45 nodes with  $\geq5$ descendants),
+and 148 breakpoints from 142 nodes are Omicron|Omicron (71 breakpoints / 68 nodes with $\geq5$ descendants).
+The equivalent figures for Alpha|Delta are 9 / 8 (3 / 2),
+for Alpha|Omicron are 2 / 1 (0 / 0),
+and for Delta|Omicron are 20 / 13 (6 / 3). 
+Note that these plots do not include recombination breakpoints involving lineages that were not classified into these VoC categories.
+}
 \end{figure}

### 2023-05-26 paper.tex
Fix 3seq coords for the supplemental table

@@
 \textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\
-\hline ALDP-11CF93B & A
-    & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
-21,256–22,228 \\
-    ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
-    \thead{B.1.177.18 \\ Alpha} & 21,256–22,228 \\
-ALDP-130BB95 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
-    \thead{B.1.177.18 \\ Alpha} & 21,256–22,228 \\
-LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
-    \thead{B.1.177.18 \\ Alpha} & 21,256–22,228 \\
-QEUH-CCCB30 & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
-    \thead{B.1.36 \\ Alpha} & 6,529–6,955 \\
-QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
-    \thead{B.1.36 \\ Alpha} & 6,529–6,955 \\
-MILK-1166F52 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
-    \thead{Alpha \\ B.1.221} & 25,997–27,973 \\
-MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
-    \thead{Alpha \\ B.1.221} & 25,997–27,973 \\
-QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
-    \thead{Alpha \\ B.1.221} & 25,997–27,973 \\
-MILK-126FE1F & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
-    \thead{B.1.36.39 \\ Alpha} & 22,445–23,064 \\
-RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
-    \thead{B.1.36.39 \\ Alpha} & 22,445–23,064 \\
-RAND-128FA33 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
-    \thead{B.1.36.39 \\ Alpha} & 22,445–23,064 \\
-CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
+\hline ALDP-11CF93B & A &
+    \thead{B.1.177 \\ Alpha} & 21,256--21,615 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256--22,228 \\
+ALDP-125C4D7 & A &
+    \thead{B.1.177 \\ Alpha} & 21,256--21,615 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256--22,228 \\
+ALDP-130BB95 & A &
+    \thead{B.1.177 \\ Alpha} & 21,256--21,615 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256--22,228 \\
+LIVE-DFCFFE & A &
+    \thead{B.1.177 \\ Alpha} & 18,999--20,296 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256--22,228 \\
+QEUH-CCCB30 & B &
+    \thead{B.1.36.28 \\ Alpha} & 6,529--6,955 &
+    \thead{B.1.36 \\ Alpha} & 6,529--6,955 \\
+QEUH-CD0F1F & B &
+    \thead{B.1.36.28 \\ Alpha} & 6,529--6,955 &
+    \thead{B.1.36 \\ Alpha} & 6,529--6,955 \\
+MILK-1166F52 & C &
+    \thead{Alpha \\ B.1.221.1} & 25,997--27,443 &
+    \thead{Alpha \\ B.1.221} & 25,997--27,973 \\
+MILK-11C95A6 & C &
+    \thead{Alpha \\ B.1.221.1} & 25,997--27,443 &
+    \thead{Alpha \\ B.1.221} & 25,997--27,973 \\
+QEUH-109B25C & C &
+    \thead{Alpha \\ B.1.221.1} & 25,997--27,443 &
+    \thead{Alpha \\ B.1.221} & 25,997--27,973 \\
+MILK-126FE1F & D &
+    \thead{B.1.36.39 \\ Alpha} & 20,704--23,064 &
+    \thead{B.1.36.39 \\ Alpha} & 22,445--23,064 \\
+RAND-12671E1 & D &
+    \thead{B.1.36.39 \\ Alpha} & 20,704--23,064 &
+    \thead{B.1.36.39 \\ Alpha} & 22,445--23,064 \\
+RAND-128FA33 & D &
+    \thead{B.1.36.39 \\ Alpha} & 20,704--23,064 &
+    \thead{B.1.36.39 \\ Alpha} & 22,445--23,064 \\
+CAMC-CBA018 & n/a &
+    \thead{B.1.177 \\ Alpha} & 20,390--21,256 &
     \thead{B.1.177 \\ Alpha} & 17,616--21,256 \\
-CAMC-CB7AB3 & n/a & \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} &
-     \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3268--5389\\17,616--21,256} \\
-MILK-103C712 & n/a & \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} &
+CAMC-CB7AB3 & n/a &
+    \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,268--4,476 \\ 20,390--21,256} &
+    \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,268--5,389\\17,616--21,256} \\
+MILK-103C712 & n/a &
+    \thead{B.1.177.17 \\ Alpha} & \thead{409--446 \\ 26,802--27,878} &
     n/a & n/a \\
-QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 &
-    \thead{Alpha \\ B.1.177} & 7729--10871 \\ \hline
+QEUH-1067DEF & n/a &
+    \thead{Alpha \\ B.1.177.9} & 10,524--10,871 &
+    \thead{Alpha \\ B.1.177} & 7,729--10,871 \\ \hline
 \end{tabular}
-\caption{\label{tab:jackson_supplement}Recombinant sequences involving the Alpha (B.1.1.7) variant reported by \cite{Jackson2021-ik} have recombinant ancestry in the Wide ARG.
-The breakpoint intervals and Pango lineage assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et al., except the Pango lineage assignment of the parents of group B recombinants, which were taken from Table 1 (motif-based results).}
+\caption{\label{tab:jackson_supplement}
+Recombinant sequences involving the
+Alpha (B.1.1.7) variant reported by \cite{Jackson2021-ik} have recombinant
+ancestry in the Wide ARG. The breakpoint intervals and Pango lineage
+assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et
+al., except the Pango lineage assignment of the parents of group B
+recombinants, which were taken from Table 1 (motif-based results).}
 \end{table}
@@
 lineages for recombination events within and among different VoC categories.
-There are 78 Alpha|Alpha recombination breakpoints corresponding to 75 recombination nodes 
+There are 78 Alpha+Alpha recombination breakpoints corresponding to 75 recombination nodes
 (25 breakpoints / 24 nodes with $\geq5$ descendant samples).
-148 breakpoints from 142 nodes are Delta|Delta recombinations (45 breakpoints / 45 nodes with  $\geq5$ descendants),
-and 148 breakpoints from 142 nodes are Omicron|Omicron (71 breakpoints / 68 nodes with $\geq5$ descendants).
-The equivalent figures for Alpha|Delta are 9 / 8 (3 / 2),
-for Alpha|Omicron are 2 / 1 (0 / 0),
-and for Delta|Omicron are 20 / 13 (6 / 3). 
+148 breakpoints from 142 nodes are Delta+Delta recombinations (45 breakpoints / 45 nodes with  $\geq5$ descendants),
+and 148 breakpoints from 142 nodes are Omicron+Omicron (71 breakpoints / 68 nodes with $\geq5$ descendants).
+The equivalent figures for Alpha+Delta are 9 / 8 (3 / 2),
+for Alpha+Omicron are 2 / 1 (0 / 0),
+and for Delta+Omicron are 20 / 13 (6 / 3).
 Note that these plots do not include recombination breakpoints involving lineages that were not classified into these VoC categories.

### 2023-05-26 paper.tex
Fix 3SEQ coords in main text table
Closes #32
Define the breakpoint intervals

@@
 \label{sec:jackson_recombs}
-% FIXME: this is rubbish, can someone do better?
 RNA viruses are known to recombine at high rates
@@
 \midrule
-A (XA)       & 4   & Jackson        &  21,256--21,613 & B.1.177/B.1.1.7 \\
+A (XA)       & 4   & Jackson        &  21,256--21,615 & B.1.177/B.1.1.7 \\
              &     &\texttt{sc2ts} &  21,256--22,228 & B.1.177.18/B.1.1.7 \\
 \cmidrule{3-5}
-B            & 2   & Jackson        &  6,529--6,954 & B.1.36/B.1.1.7  \\
+B            & 2   & Jackson        &  6,529--6,955 & B.1.36/B.1.1.7  \\
              &     &\texttt{sc2ts} &  6,529--6,955 & B.1.36/B.1.1.7  \\
 \cmidrule{3-5}
-C            & 3   &Jackson        &  25,997--27,441 &  B.1.1.7/B.1.221 \\
+C            & 3   &Jackson        &  25,997--27,443 &  B.1.1.7/B.1.221 \\
              &     & \texttt{sc2ts} &  25,997--27,973 &  B.1.1.7/B.1.221 \\
 \cmidrule{3-5}
-D            & 3   & Jackson        &  21,576--23,063 &  B.1.36.17/B.1.1.7 \\
+D            & 3   & Jackson        &  21,576--23,064 &  B.1.36.17/B.1.1.7 \\
              &     & \texttt{sc2ts} &  22,445--23,064 &  B.1.36.39/B.1.1.7 \\
@@
 The second column gives the number of sequences in the group.
-Intervals are half-closed, so that for an interval $\ell$--$r$
-the breakpoint $x$ can be $\ell \leq x < r$.
+The 3SEQ~\citep{boni2007exact} coordinates
+reported by Jackson et al.\ have been altered as follows: we
+add one to both left and right coordinates to correspond to the
+\texttt{tskit} definition of inheritance on either side of a breakpoint,
+and add one to the right coordinate to make the intervals
+right-exclusive. See Section~\ref{sec:treatment_recombinants}
+for a precise definition of sequence inheritance at recombination
+events, and the corresponding breakpoint intervals.
 Details for all 16 sequences are
@@
 Section~\ref{sec:breakpoint_intervals}.
+
+For a particular recombination node in an ARG, a  breakpoint
+is defined as the location at which inheritance switches from one parent
+to another.
+In the \texttt{tskit} encoding (see Section~\ref{sec:args}) inheritance between
+nodes is defined by edges $(\ell, r, p, c)$, which state that child node $c$
+inherits from parent node $p$
+over the half-closed genome interval $[\ell, r)$.
+For simplicity, suppose that a recombination node $u$ inherits from two
+parents $p_1$ and $p_2$ with a breakpoint of $x$.
+In the ARG, this is defined by two edges
+$(0, x, p_1, u)$ and $(x, L, p_2, u)$ where $L$ is the length of the genome.
+Since inheritance intervals are half-closed,
+$u$ inherits all positions up to $x$ (exclusive) from parent $p_1$
+and all positions from $x$ (inclusive) to the end of the genome from parent
+$p_2$. We then define the breakpoint \emph{interval} $[b_\ell, b_r)$ as the
+half-closed interval defining the range of possible values for $x$, such
+that $b_\ell \leq x < b_r$.
 
@@
 al., except the Pango lineage assignment of the parents of group B
-recombinants, which were taken from Table 1 (motif-based results).}
+recombinants, which were taken from Table 1 (motif-based results).
+3SEQ interval coordinate modifications are described in the caption for
+Table~\ref{tab:jackson}.}
 \end{table}

### 2023-05-26 paper.tex
Yan's updates

--- 2023-05-26 paper.tex
@@
 lineages for recombination events within and among different VoC categories.
-There are 78 Alpha|Alpha recombination breakpoints corresponding to 75 recombination nodes 
+There are 78 Alpha+Alpha recombination breakpoints corresponding to 75 recombination nodes 
 (25 breakpoints / 24 nodes with $\geq5$ descendant samples).
-148 breakpoints from 142 nodes are Delta|Delta recombinations (45 breakpoints / 45 nodes with  $\geq5$ descendants),
-and 148 breakpoints from 142 nodes are Omicron|Omicron (71 breakpoints / 68 nodes with $\geq5$ descendants).
-The equivalent figures for Alpha|Delta are 9 / 8 (3 / 2),
-for Alpha|Omicron are 2 / 1 (0 / 0),
-and for Delta|Omicron are 20 / 13 (6 / 3). 
+148 breakpoints from 142 nodes are Delta+Delta recombinations (45 breakpoints / 45 nodes with  $\geq5$ descendants),
+and 148 breakpoints from 142 nodes are Omicron+Omicron (71 breakpoints / 68 nodes with $\geq5$ descendants).
+The equivalent figures for Alpha+Delta are 9 / 8 (3 / 2),
+for Alpha+Omicron are 2 / 1 (0 / 0),
+and for Delta+Omicron are 20 / 13 (6 / 3). 
 Note that these plots do not include recombination breakpoints involving lineages that were not classified into these VoC categories.

### 2023-05-26 paper.tex
Better typesetting for jackson parents
Update the Summary tables notebook, fill in values

--- 2023-05-26 paper.tex
@@
 Of these recombination nodes, the majority are ancestral
-to only one sample (Wide ARG: FIXME\%, Long ARG: FIXME \%).
+to only one sample (Wide ARG: 63.1\%, Long ARG: 63.3\%).
 We analyse signals of recombination in Sections~\ref{sec:jackson_recombs},
@@
 \begin{table} \centering
-\begin{tabular}{lllll}
+\begin{tabular}{lllr@{--}lr@{+}l}
 \toprule
-Group        & Sequences & Method & Interval & Parents \\
+Group        & Sequences & Method & \multicolumn{2}{c}{Interval}
+    & \multicolumn{2}{r}{Parent lineages} \\
 \midrule
-A (XA)       & 4   & Jackson        &  21,256--21,615 & B.1.177/B.1.1.7 \\
-             &     &\texttt{sc2ts} &  21,256--22,228 & B.1.177.18/B.1.1.7 \\
-\cmidrule{3-5}
-B            & 2   & Jackson        &  6,529--6,955 & B.1.36/B.1.1.7  \\
-             &     &\texttt{sc2ts} &  6,529--6,955 & B.1.36/B.1.1.7  \\
-\cmidrule{3-5}
-C            & 3   &Jackson        &  25,997--27,443 &  B.1.1.7/B.1.221 \\
-             &     & \texttt{sc2ts} &  25,997--27,973 &  B.1.1.7/B.1.221 \\
-\cmidrule{3-5}
-D            & 3   & Jackson        &  21,576--23,064 &  B.1.36.17/B.1.1.7 \\
-             &     & \texttt{sc2ts} &  22,445--23,064 &  B.1.36.39/B.1.1.7 \\
+A (XA)       & 4   & Jackson        &  21,256&21,615 & B.1.177&B.1.1.7 \\
+             &     &\texttt{sc2ts} &  21,256&22,228 & B.1.177.18&B.1.1.7 \\
+\cmidrule{3-7}
+B            & 2   & Jackson        &  6,529&6,955 & B.1.36&B.1.1.7  \\
+             &     &\texttt{sc2ts} &  6,529&6,955 & B.1.36&B.1.1.7  \\
+\cmidrule{3-7}
+C            & 3   &Jackson        &  25,997&27,443 &  B.1.1.7&B.1.221 \\
+             &     & \texttt{sc2ts} &  25,997&27,973 &  B.1.1.7&B.1.221 \\
+\cmidrule{3-7}
+D            & 3   & Jackson        &  21,576&23,064 &  B.1.36.17&B.1.1.7 \\
+             &     & \texttt{sc2ts} &  22,445&23,064 &  B.1.36.39&B.1.1.7 \\
 \bottomrule

### 2023-05-26 paper.tex
Pass through the backbone phylo section

@@
 understanding of the relative importance of these processes to the overall
-topology. Similarly, edges are classified by the type of the parent node,
-giving some insights into how often these nodes are subsequently
-copied from after insertion (although the relationship is more complicated
-for UPGMA inserted nodes). The mean and standard deviation ($\pm$)
+topology. Similarly, edges are classified by the type of the parent node.
+The mean and standard deviation ($\pm$)
 are reported for the number of mutations per node, sample and site.
@@
 inference methods and scaling out across multiple servers will
-enable inferences at the full pandemic scale
-(see Discussion).
+enable inferences at the full pandemic scale.
 
@@
 present in the Nextstrain tree. This results in a small set of shared
-samples (180 for the Wide ARG, 88 for the Long ARG), none of which are
+samples (Wide ARG: 180, Long ARG: 88), none of which are
 assigned to Pango recombinant lineages by Nextclade (see Section
@@
 particularly in early 2020 when sampling density was much lower than
-later in the pandemic.
-Secondly, there are non-identical sample partitions
-near the tips (e.g. within the Alpha clade).
-Finally, and most noticeably, some branch lengths differ substantially.
+later in the pandemic. Resolution early in the pandemic could be improved
+by using a tree inferred using classical phylogenetic
+approaches for the first few months of the pandemic, before the scale of
+data began to overwhelm these methods. Indeed, this is the approach
+taken by UShER~\citep{Turakhia2021-ur}. Using a pre-existing tree for
+the early stages of the pandemic would be straightforward in \texttt{sc2ts},
+and the main reason we did not do this for the initial
+version under consideration here was to evaluate the algorithm's
+performance with sparse data. Given
+the simplicity of the algorithm, tree inference for the early pandemic
+is surprisingly good.
+The second difference between the \texttt{sc2ts} and Nextstrain trees
+we would like to highlight, are the non-identical sample partitions
+near the tips (e.g.\ within the Alpha clade). It is unclear what
+particular features of the \texttt{sc2ts} algorithm is driving these
+differences, and more study is required to characterise and address them.
+Finally, some branch lengths differ substantially between the \texttt{sc2ts}
+and Nextstrain trees.
 As discussed in Section~\ref{sec:node_dating}, the dating of nodes
-other than samples in \texttt{sc2ts} is currently quite crude, and
-it is likely that branches not ``anchored'' by samples will be
-poorly calibrated. See [SOMEWHERE ELSE] for a discussion of how
-these issues may be alleviated in future versions of \texttt{sc2ts}.
+other than samples in \texttt{sc2ts} is currently quite crude,
+but there are likely straightforward expedients that would yield
+substantial improvements.
 

### 2023-05-26 plot.py
Tweak some labels and positions
For cophylogenies (labels & dates), subplots (add all XB nodes), and supp_recombination nodes ("+" as separator)

@@ -165,6 +165,19 @@ class Figure:
         raise NotImplementedError()
 
 
+def isomonths_iter(start):
+    """
+    start as 2020-01, 2021-07
+    yield each month as a YYYY-MM string
+    """
+    year, month = [int(x) for x in start.split("-")]
+    while True:
+        yield f"{year}-{month:02}"
+        month += 1
+        if month > 12:
+            month = 1
+            year += 1
+
 class Cophylogeny(Figure):
     name = None
     pos = 14952  # Position along tree seq to plot trees: 14952 is halfway along
@@ -432,28 +445,8 @@ class Cophylogeny(Figure):
             symbol_size=1,
             y_axis=True,
             y_ticks={
-                self.sc2ts.timediff(isodate): (isodate[:7] if show else "")
-                for isodate, show in {
-                    "2020-01-01": True,
-                    "2020-02-01": False,
-                    "2020-03-01": False,
-                    "2020-04-01": True,
-                    "2020-05-01": False,
-                    "2020-06-01": False,
-                    "2020-07-01": True,
-                    "2020-08-01": False,
-                    "2020-09-01": False,
-                    "2020-10-01": True,
-                    "2020-11-01": False,
-                    "2020-12-01": False,
-                    "2021-01-01": True,
-                    "2021-02-01": False,
-                    "2021-03-01": False,
-                    "2021-04-01": True,
-                    "2021-05-01": False,
-                    "2021-06-01": False,
-                    "2021-07-01": True,
-                }.items()
+                self.sc2ts.timediff(date+"-01"): (date if show else "")
+                for date, show in self.x_axis_ticks.items()
             },
             y_label=" ",
         )
@@ -468,28 +461,8 @@ class Cophylogeny(Figure):
             symbol_size=1,
             y_axis=True,
             y_ticks={
-                self.nxstr.timediff(isodate): (isodate[:7] if show else "")
-                for isodate, show in {
-                    "2020-01-01": True,
-                    "2020-02-01": False,
-                    "2020-03-01": False,
[…]

### 2023-05-26 plot.py
Remove breakpoints files
Switch to using the straight export_breakpoints csv dump. Fixes #120

@@ -602,7 +602,7 @@ class CophylogenyLong(Cophylogeny):
 class RecombinationNodeMrcas(Figure):
     name = None
     sc2ts_filename = "upgma-mds-1000-md-30-mm-3-2022-06-30-recinfo2-gisaid-il.ts.tsz"
-    csv_fn = "breakpoints_{}.csv"
+    csv_fn = "long_arg_recombinants.csv"
     data_dir = "data"
 
     def __init__(self):
@@ -612,6 +612,9 @@ class RecombinationNodeMrcas(Figure):
         df = pd.read_csv(
             os.path.join(self.data_dir, self.csv_fn.format(prefix)),
             parse_dates=["causal_date", "mrca_date"])
+        df[f"parents_dist_{self.ts.time_units}"] = (
+            self.ts.nodes_time[df["mrca"]] - self.ts.nodes_time[df["node"]]
+        )
         logging.info(f"{len(df)} breakpoints | {len(np.unique(df.node))} re nodes read")
         # Remove potential contaminents
         self.df = df[df.max_descendant_samples > 1]

### 2023-05-26 paper.tex

@@
 (sampled to mid-2021) and an ``all-time'' global Nextstrain tree
-(downloaded on January 21, 2023).
+(downloaded on 2023-01-21).
 Phylogenies are pruned down to those samples present as tips in both datasets. Light grey
@@
 Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
-a GISAID global ``all-time'' tree from Nextstrain
-% We give the date below in the caption - we could probably remove if either from there
-% or from here (I suggest from here)
-(downloaded on January 21, 2023).
+a GISAID global ``all-time'' tree from Nextstrain.
 We illustrate the backbone phylogeny by visualizing
@@
 of mutations per sample is small
-(Wide ARG: 0.77, Long ARG: 1.38; c.f.\ Table~\ref{tab:args}), the average number per site
-is large (Wide ARG: 41.23, Long ARG: 36.10; c.f.\ Table~\ref{tab:args}). This is
-partly driven by some highly mutagenic sites (e.g., site 28271 has over 7,000
-mutations in the Wide ARG, c.f.\ Figure~\ref{fig:breakpoint-distribution}), which
+(Wide ARG: 0.77, Long ARG: 1.38; Table~\ref{tab:args}), the average number per site
+is large (Wide ARG: 41.23, Long ARG: 36.10; Table~\ref{tab:args}). This is
+partly driven by some highly mutagenic sites (e.g., site 28,271 has over 7,000
+mutations in the Wide ARG, Figure~\ref{fig:breakpoint-distribution}), which
 may be artefactual.
@@
 who reconstructed a SARS-CoV-2 phylogeny of
-over 350,000 genomes sampled globally from December 24, 2019, to January 12,
-2021, and classified the mutations occurring along the phylogeny.
+over 350,000 genomes sampled globally from 2019-12-24 to 2021-01-12
+and classified the mutations occurring along the phylogeny.
 We categorised all single nucleotide
@@
 \citep{Gribble2021-coronavirus}, the relatively slow accumulation
-of genetic diversity early on in the pandemic hampered efforts to
+of genetic diversity early in the pandemic hampered efforts to
 detect recombinant strains. A number of early studies relying on
@@
 are similar there is no way to
-% distinguish them.
-accurately specify the breakpoint.
+accurately identify the breakpoint.
 Here we define the
@@
 (Section~\ref{sec:ls}), and the left endpoint is then derived by directly
-comparing the parent sequences.
+comparing the parent sequences. See Section~\ref{sec:treatment_recombinants} for
+a precise definition of breakpoints and their intervals.
 
@@
 (see Figure~\ref{fig:long_arg_breakpoint_distribution} for the same information
-of the Long ARG). We focus on the Wide ARG here as it covers roughly the same time
+for the Long ARG). We focus on the Wide ARG here as it covers roughly the same time
 period as the analyses of \cite{Turakhia2022-it}, facilitating comparisons
@@
 
-Emphasising that further work is required to filter out spuriously inferred
-recombination events and improve the accuracy of the inferred breakpoint
-intervals, we can draw two preliminary conclusions from
+Although we emphasise that further work is required to filter out spuriously inferred
+recombination events, we can draw two preliminary conclusions from
 Figure~\ref{fig:breakpoint-distribution}.
@@
 show sharp differences in the number of intersecting intervals within
-these genes. This could be due to a number of factors, such as [BIOLOGY]
+these genes. This could be due to a number of factors, such as [FIXME: BIOLOGY?]
 or the presence of sites with artefactually high numbers of mutations
@@
 of a recombinant's parents.
-% The presence of problematic sites that are enriched
-% for errors will therefore tend to artificially truncate breakpoint intervals
-% at these sites.
-% It is therefore helpful to compare the number of interecting intervals
-% with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
-% For example, position 28,271 (located in the ORF8-N intergenic
-% region) has the largest number of mutations and corresponds to the
-% left endpoint of 9 intervals and the right of 57.
-% The 7,572 mutations  (including 5,782 insertions and 1,605 deletions)
-% at this site are a likely indicator that this
-% homopolymeric region is prone to sequencing errors
-% and should be included in the list of ``problematic sites'' that
-% are excluded from analysis (Section~\ref{sec:data_preprocessing}).
-% There can be good biological reasons for this too though
-% However, hypermutable sites such as 27,972 \citep{Jungreis2021-dh}
-% have the same effect of truncating intervals. At this site we have
-% 770 mutations (including 425 C$>$T, 314 T$>$C) with potential signals
-% of polymorphism maintained by selection [FIXME!!], and 58 intervals
-% with endpoints within one base of this site.
-Thus, breakpoint intervals will tend to be truncated at sites which are
+Thus, breakpoint intervals will tend to be truncated at sites that are
 hypermutable, either due to being prone to sequencing errors or

### 2023-05-26 paper.tex
pass through divergence section

@@
 % Paragraph 1, what does all this mean?
-
 % Purpose of this paragraph - remind the reader that we can only do this
@@
 % greater level of granularity than existing methods.
-[ FIXME this is a rough outline of what an intro para might say]
 An important advantage of inferring a joint model of inheritance,
@@
 is that this allows us to precisely trace the origins of recombinants.
-[Katherine - need some input from you here to say this properly?]
-Existing methods only provide us with approximate estimates of
-the origins of recombination events, and are not able to distinguish
-recombinations between closely related lineages.
+Methods based on motifs among pre-specified lineages
+provide us with estimates of the origins of recombinants as
+parents belonging to one or other of these groups. Defining the
+groups requires care [FIXME and other fiddly things to mention Katherine?]
+RIPPLES~\citep{Turakhia2022-it} does provide precise estimates
+of the parental origins of a recombinant, but since it depends
+on an existing phylogenetic tree, cannot reason about
+nested recombinant origins.
+
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
+\caption{\label{fig:recomb_mrcas}
+Date of common ancestry between the parents on either side of recombination
+breakpoints, as a function of the divergence time between the parents.
+MRCAs of parents associated with Pango designated recombinants (XA, etc)
+are identified in orange. Larger symbols represent breakpoints in recombination
+nodes ancestral to five or more samples.
+Horizontal dotted lines show the four most common MRCA nodes,
+which tend to be associated with major outbreaks and with many immediate children.
+The stacked histogram shows the distribution of parental divergence times, ranging
+from parents that have diverged only a handful of days previously, to much more
+divergent parent lineages. See Figure~\ref{fig:recomb_mrcas_voc_breakdown}
+for equivalent plots broken down by parental VoC classification.
+}
+\end{figure}
 
 % Paragraph 2, what do we plot, specifically?
-In this section we investigate the patterns of divergence between
-the parents of recombination nodes.
+In this section we explore the detailed ancestral relationships between
+recombinant parents by investigate the patterns of divergence between
+them.
 We focus on the Long ARG because it covers time
-periods where recombination is known to have occurred
-a large number of times, for example it contains samples from over
-30 Pango X lineages (i.e., those inferred
+periods where substantial recombination is known to have occurred,
+and contains samples from 33 Pango X lineages (i.e., those inferred
 to have recombinant ancestry; see Section~\ref{sec:pango_x_lineages} for
-further analysis of Pango recombinant lineages in the Long and Wide ARGs).
+further analysis).
 Figure~\ref{fig:recomb_mrcas} shows the estimated date of the most recent
@@
 minimum number required to designate a new Pango recombinant lineage
-\citep{Rambaut2020-dw}, and serves to highlight those recombinants which
-have greater epidemiological significance.
-
-% that we deem to be more reliable by the measure of ``HMM consistency''.
-% We define an  in that when we compare the LS HMM results when run from
-% 5' to 3' versus 3' to 5', it gives the
-% same mutations patterns and sequence-identical 5' parents as when run 3' to 5'
+\citep{Rambaut2020-dw}.
 Note that as each plotted point represents a breakpoint, a recombination
@@
 comprising ${\sim}10 \%$ of the
-recombination nodes in Figure ~\ref{fig:recomb_mrcas}) will be represented by several points.
-We caution that some of the time estimates in these figures
-come with a substantial degree of uncertainly,
-since non-sample nodes are dated somewhat crudely in the current version
-of \texttt{sc2ts} (Section~\ref{sec:node_dating}). Nevertheless, the histograms
-clearly show that \texttt{sc2ts} can identify recombination between lineages
-that are only a few weeks apart.
-
-\begin{figure} \centering
-\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
-\caption{\label{fig:recomb_mrcas}
-Date of common ancestry between the parents on either side of recombination
-breakpoints, as a function of the divergence time between the parents.
-MRCAs of parents associated with Pango designated recombinants (XA, etc)
-are identified in orange. Larger symbols represent breakpoints in recombination
-nodes which are ancestral to five or more samples.
-Horizontal dotted lines show the four most common MRCA nodes,
-which tend to be associated with major outbreaks and with many immediate children.
-The stacked histogram shows the distribution of parental divergence times, ranging
-from parents that have diverged only a handful of days previously, to much more
-divergent parent lineages. See Figure~\ref{fig:recomb_mrcas_voc_breakdown}
-for equivalent plots broken down by parental VoC classification.
-}
-\end{figure}
+recombination nodes in
+Figure~\ref{fig:recomb_mrcas}) will be represented by several points.
 
@@
 largely due to a few MRCAs shared by many
-recombinants; the top four are indicated by dotted lines in the
-figure. These shared MRCAs lie near the root of large expansions
-(B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
-and BA.2: second Omicron wave), as demonstrated by the number of samples with those
-Pango designations which trace back to each ancestral node (listed as
-percentages on the plot). The majority of these frequently-shared MRCAs
-are associated with large polytomies (at the extreme,
-the BA.2-linked node has 2,872 immediate children), likely
+recombinants (the top four are indicated in the figure).
+These shared MRCAs lie near the root of large expansions
+% Anyone reading this will know what these are
+% (B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
+% and BA.2: second Omicron wave)
+% as demonstrated by the number of samples with those
+% Pango designations traback to each ancestral node.
+% The majority of these frequently-shared MRCAs
+and the majority are associated with large polytomies, likely
 indicating a rapid and under-sampled expansion of a clade.
@@
 % Paragraph 4, what do we say about the divergence times?
-The histograms in Figure~\ref{fig:recomb_mrcas} show that there is a
+Figure~\ref{fig:recomb_mrcas} shows that there is a
 broad range in divergence times among recombinant parents.
-We can also see that the majority of recombinants arise
+We can see that the majority of recombinants arise
 from parents that diverged 10 to 30 weeks
-previously. However, there is also a peak of recombinants that diverged
-${\sim}95$ weeks prior to the event: these correspond to recombinants whose
+previously. There is also a peak of recombinants that diverged
+${\sim}95$ weeks prior to the event, which correspond to recombinants whose
 parental common ancestor traces to early 2020. We would expect, for example
@@
 two Omicron parents, and that Omicron and Delta are the variants associated
-with the most recombination (Supplementary Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
+with the most recombination (Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
 This may reflect either sampling intensity, the
@@
 recombination probabilities among lineages.
+
+The time estimates in these figures should be
+treated with a degree of caution,
+since non-sample nodes are crudely dated in the current version
+of \texttt{sc2ts} (see Section~\ref{sec:node_dating}, in particular
+for discussion of how these dates might be improved using existing
+methods).
+Nevertheless, it is clear
+that \texttt{sc2ts} can identify recombination between lineages
+that are only a few weeks diverged.
 

### 2023-05-26 paper.tex
Updates from Overleaf

--- 2023-05-26 paper.tex
@@
 \caption{\label{fig:recomb_mrcas}
-Date of common ancestry between the parents on either side of recombination
-breakpoints, as a function of the divergence time between the parents.
-MRCAs of parents associated with Pango designated recombinants (XA, etc)
-are identified in orange. Larger symbols represent breakpoints in recombination
-nodes which are ancestral to five or more samples.
-Horizontal dotted lines show the four most common MRCA nodes,
-which tend to be associated with major outbreaks and with many immediate children.
+Most recent common ancestor (MRCA) date for pairs of parents on either side
+of a recombination breakpoint, as a function of the divergence time between parent
+pairs. Breakpoints from recombination events caused by Pango designated recombinants
+(XA, etc) are identified in orange. Larger symbols represent breakpoints from
+recombination nodes which are ancestral to five or more samples.
+Horizontal dotted lines show the four most common MRCA nodes; these
+tend to be associated with major outbreaks and with many immediate children.
 The stacked histogram shows the distribution of parental divergence times, ranging

### 2023-05-26 paper.tex

--- 2023-05-26 paper.tex
@@
 previously identified as recombinants, i.e. given a Pango designation that
-starts with an ``X''. We focus primarily on the Long ARG in this section,
-since it contains many more recombinants; the status
-of Pango X lineages in the Wide ARG is briefly summarised in Section~\ref{sec-pango_x_wide_arg}.
+starts with an ``X''. We focus primarily on the Long ARG
+since it contains many more recombinants. The status
+of Pango X lineages in the Wide ARG is briefly
+summarised in Section~\ref{sec-pango_x_wide_arg}.
 

### 2023-05-26 paper.tex
Yans updates

@@
 
-The Long ARG contains 749 samples designated by Nextclade [CITE] as Pango
+The Long ARG contains 749 samples designated by Nextclade [CITATION] as Pango
 recombinants, classed into 33 Pango lineages; this
@@
 nodes subtending only one sample) are removed.
-In contrast, using the alternative GISAID designation process [CITE], only
+In contrast, using the alternative GISAID designation process [CITATION], only
 515 samples in the Long ARG are labelled as Pango recombinants,
@@
 distinct recombinants to emerge and spread. Phylogenetic approaches have
-been central to pandemic responses thus far [TODO: Examples and citations].
+been central to pandemic responses thus far [FIXME: Examples and citations].
 However, as recombinant lineages are now endemic and it is clear that

### 2023-05-26 paper.tex

@@
 clades, labelled).
-Supplementary Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.
+Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.
 
@@
 \label{sec:pango_x_lineages}
-
-In this section we focus on the detailed genealogy of samples that have been
-previously identified as recombinants, i.e. given a Pango designation that
-starts with an ``X''. We focus primarily on the Long ARG
-since it contains many more recombinants. The status
+In this section we focus on the detailed ancestry of samples that have been
+previously identified as recombinants, i.e., designated as belonging
+to a Pango lineage with a name starting with an ``X''.
+We focus primarily on the Long ARG
+since it contains many more recombinants (the status
 of Pango X lineages in the Wide ARG is briefly
-summarised in Section~\ref{sec-pango_x_wide_arg}.
-
-The Long ARG contains 749 samples designated by Nextclade [CITATION] as Pango
-recombinants, classed into 33 Pango lineages; this
-reduces to 711 samples from the same 33 lineages
-when singleton recombinants (i.e. recombination
-nodes subtending only one sample) are removed.
-In contrast, using the alternative GISAID designation process [CITATION], only
-515 samples in the Long ARG are labelled as Pango recombinants,
-which fall into 38 lineages (when singleton recombinants are excluded this
-drops to 514 samples from 35 lineages). The variation in Nextclade versus
-GISAID classifications highlights the frequent uncertainty that exists when
-assigning Pango X lineages to samples.
-
-Below we focus on the 749 Nextclade-designated recombinant samples, all
+summarised in Section~\ref{sec-pango_x_wide_arg}).
+Designation of samples to Pango lineages is not a straightforward task,
+and there can be significant variation between methods.
+Here, we consider two different assignments of Pango lineages to samples in
+the Long ARG, those provided by Nextclade and by GISAID.
+Using the Nextclade assignments, the Long ARG contains 749 samples from
+33 Pango X lineages (711 samples from 33 lineages when we
+remove singleton recombinants).
+In contrast, using the GISAID designations we
+have 515 samples from 38 X lineages
+[FIXME this doesn't make sense! How did we drop 1 sample and 3 lineages?]
+(514 samples from 35 lineages when we filter singletons).
+This variation in classifications highlights the uncertainty that exists when
+assigning Pango X lineages to samples, and is important to keep in mind
+when interpreting the results here.
+
+We focus here on the 749 Nextclade-designated recombinant samples, all
 but two of which descend from a recombination node in the Long ARG.
@@
 (\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
-during preprocessing (See~\ref{sec:data_preprocessing} for details and potential
-improvements). 38 of the remaining X designated samples are singleton recombinants,
+during preprocessing (See Section~\ref{sec:data_preprocessing} for details and potential
+improvements). Of the remaining X designated samples, 38 are singleton recombinants,
 descending
@@
 The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
-1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, caused by strain
+1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, introduced as a
+result of adding strain
 Denmark/DCGC-281594/2021, is ancestral to more than 127K samples,
@@
 
-The remaining remaining 630 Pango recombinant samples trace back to 50 different
-most recent recombination nodes, corresponding to 31
-Nextclade-designated Pango lineages. These are summarised
-in Supplementary Table~\ref{tab:pango-recombinants}, and fall
-into three classes: single origin, multiple origin,
+The remaining 630 samples (31 Pango lineages) trace back to 50 different
+most recent recombination nodes, summarised
+in Table~\ref{tab:pango-recombinants}.
+These fall into three classes: single origin, multiple origin,
 and multiple nested origins, which we discuss in the following sections.
@@
 recombination event. In the Long ARG, 16 of the 31 Pango recombinants lineages identified by Nextclade
-fall into this category (Supplementary Table~\ref{tab:pango-recombinants})
+fall into this category (Table~\ref{tab:pango-recombinants})
 
@@
 these may be, however, be a result of erroneous Pango designations.
-Supplementary Table~\ref{tab:pango-recombinants} also shows the official
+Table~\ref{tab:pango-recombinants} also shows the official
 Pango designated parent lineages and \texttt{sc2ts} inferred parent lineages,
@@
 descend, and which agrees with the offical Pango designated parent lineages
-(see Supplementary Table~\ref{tab:pango-recombinants}, but note that
+(see Table~\ref{tab:pango-recombinants}, but note that
 in XJ and XU there are too few Pango X samples to decide on a ``main'' clade).
@@
 plausible that the earlier Turkish sample has been mislabelled XD by Nextclade.
-Indeed, GISAID does not label any of these samples XD (see Supplementary
+Indeed, GISAID does not label any of these samples XD (see
 Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).
@@
 samples whose ancestry involves multiple non-independent (``nested'') recombination events
-(marked by~\textdagger ~in Supplementary Table~\ref{tab:pango-recombinants}; the most
+(marked by~\textdagger ~in Table~\ref{tab:pango-recombinants}; the most
 complex appears to be XAB). Figure~\ref{fig:complex_origins_graph} plots the earliest example,
@@
 XB sample (pink upper left, with 7 mutations above it); moreover, this sample is
-not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph})
+not identified as XB by GISAID (Figure~\ref{fig:pango_XB_gisaid_graph})
 indicating some uncertainty in lineage assignment in this part of the ARG.

### 2023-05-30 paper.tex
Edits from Kat, Ana and Sally

@@
 and presents major challenges to [things important to public health].
+% ...
+genomic surveillance programs, predictive models for symptoms and severity, policy recommendations such as vaccine effectiveness and booster frequency, etc.
+(https://doi.org/10.1016/S1473-3099(22)00831-3)
 Detection of new recombinant strains is challenging, with methods
 [being a bit rubbish].
+% ... with few methods that facilitate rapid and systematic (reproducible) classification .https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/
 In addition, many basic approaches fundamental to the study of viral
@@
 \texttt{sc2ts}, a method to automatically detect recombinants
-in real-time and to fully integrate them into a cohesive joint
-model of mutation, recombination and genetic inheritance known
-as an Ancestral Recombination Graph (ARG). We infer two ARGs under
+in real time and to cohesively integrate them into a
+genealogy in the form of an ancestral recombination graph (ARG),
+which jointly records mutation, recombination and genetic
+inheritance. We infer two ARGs under
 different sampling strategies, and study their properties.
@@
 established and spread to high frequency, and accounting for recombinant
-ancestry is now essential in understanding the evolution of the SARS-CoV-2
-virus.
+ancestry is now essential in understanding the evolution of SARS-CoV-2.
 
@@
 are the biases?]
+% This resource shows the typical workflow for manually identifying recombinants that the pango team uses: https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/
+% Regarding bias, not peer-reviewed but in the official pango designation post for XAC, you can see that me (Katherine) and the authors had a multi-month long discussion on how many breakpoints there should be and where.
+Reproducibility was difficult: https://github.com/cov-lineages/pango-designation/issues/590
 Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
 combine specific subsets of characteristic mutations from different lineages
-\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz},
+\citep[e.g.,][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz},
 and as a result can only identify inter-lineage recombination events.
@@
 not account for recombination
-\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}, and there is
+\citep[e.g.,][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}, and there is
 no standard method for incorporating the effects of recombination into
-phylogenetic analysis.
+phylogenetic analyses.
 Ancestral Recombination Graphs (ARGs) are a means of describing such
@@
 Here we present a preliminary version of \texttt{sc2ts}, a novel method
-for inferring ARGs for SARS-CoV-2 at pandemic scale, in real-time.
-Building on the open-source \texttt{tskit} library, the method infers
-a joint model of mutation, genetic inheritance and recombination, which
+for inferring ARGs for SARS-CoV-2 at pandemic scale, in real time.
+Building on the open-source \texttt{tskit} library, the method 
+explicitly reconstructs genealogies with both
+mutation and recombination, which
 can be conveniently and efficiently analysed using standard data science
@@
 Using the current preliminary version of \texttt{sc2ts}, we infer two
-large ARGs (1,265,685 and 657,239 samples, respectively), and study
+large ARGs (with 1,265,685 and 657,239 samples, respectively), and study
 the properties of these ARGs to illustrate the power of the method
@@
 (Section~\ref{sec:parent_divergence}).
-We hope that these benefits of an accurate joint model of
-mutation, genetic inheritance and recombination, will generate community
+We hope that these benefits of accurate joint estimation of
+genetic inheritance with mutation and recombination events will generate community
 interest and development of the \texttt{sc2ts} method, and
@@
 single phylogenetic tree, particularly when looking at
-a subset of sampled genomes, e.g.\ restricted by time or limited to close
+a subset of sampled genomes, e.g., restricted by time or limited to close
 relatives. We expect ARGs to be particularly tree-like
@@
 we would like to highlight, are the non-identical sample partitions
-near the tips (e.g.\ within the Alpha clade). It is unclear what
+near the tips (e.g., within the Alpha clade). It is unclear what
 particular features of the \texttt{sc2ts} algorithm is driving these
@@
 The ARGs inferred by \texttt{sc2ts} and represented using the \texttt{tskit}
-library (Section~\ref{sec:args}) are a joint estimate of the processes
-of genetic inheritance, recombination and mutation. Unlike most approaches to
+library (Section~\ref{sec:args}) are a joint estimate of the genealogy with recombination and mutation. Unlike most approaches to
 phylogenetic analysis, mutations are included in the \texttt{tskit}
@@
 lineages either failed to detect recombination or posited
-that this occurred at low rates \citep[e.g.~][]{Nie2020-phylogenetic,Tang2020-origin,VanInsberghe2021-eu,Varabyou2021-rw}.
+that this occurred at low rates \citep[e.g.,~][]{Nie2020-phylogenetic,Tang2020-origin,VanInsberghe2021-eu,Varabyou2021-rw}.
 The first clear evidence of recombinant lineages was presented by
@@
 these genes. This could be due to a number of factors, such as [FIXME: BIOLOGY?]
+% reduced frequency of breakpoint sequence motifs, or secondary structure.
+Sorry, I don't have a citation for this, just theories.
+The RIPPLES paper vaguely mentions "important mechanistic bias shapes the distribution of recombination events".
+But maybe we just don't know yet?
 or the presence of sites with artefactually high numbers of mutations
@@
 % greater level of granularity than existing methods.
-An important advantage of inferring a joint model of inheritance,
-mutation and recombination in an ARG (Section~\ref{sec:args})
+An important advantage of explicitly inferring an ARG compatible
+with a given sample of sequences (Section~\ref{sec:args}) 
 is that this allows us to precisely trace the origins of recombinants.
@@
 groups requires care [FIXME and other fiddly things to mention Katherine?]
+% ...
+Defining the groups is a highly subjective process, involving numerous manual steps that require subject matter expertise of the organism, that often use web sites as data sources which are difficult to version-control, and thus hinder reproducibility.
+The web paper by the PH4GE group is one of the best sources for the reality (and pain) of how it's done: https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/
 RIPPLES~\citep{Turakhia2022-it} does provide precise estimates
@@
 common ancestor (MRCA) of the parent nodes for each recombination breakpoint,
-plotted against the divergence between these parents (i.e. the
+plotted against the divergence between these parents (i.e., the
 total branch length from the parents to their MRCA in the trees to
@@
 Note that as each plotted point represents a breakpoint, a recombination
-node with more than one breakpoint (e.g. with 3 or more parents,
+node with more than one breakpoint (e.g., with 3 or more parents,
 comprising ${\sim}10 \%$ of the
@@
 the Long ARG, those provided by Nextclade and by GISAID.
+% GISAID lineage calls come from the pangoLEARN algorithm.
 Using the Nextclade assignments, the Long ARG contains 749 samples from
@@
 have 515 samples from 38 X lineages
-[FIXME this doesn't make sense! How did we drop 1 sample and 3 lineages?]
-(514 samples from 35 lineages when we filter singletons).
+(511 samples from 35 lineages when we filter singletons).
 This variation in classifications highlights the uncertainty that exists when
@@
 \includegraphics[width=\textwidth]{figures/Pango_XA_XAG_XD_nxcld_tight_graph.pdf}
+% This figure is gorgeous!!
 
@@
 Sample nodes are plotted as larger circles, but the subgraph also includes intermediate,
-non-sample nodes: i.e. inserted nodes such as UPGMA, recombination, mutation collapse, and
+non-sample nodes: i.e., inserted nodes such as UPGMA, recombination, mutation collapse, and
 reversion push nodes, plotted as smaller circles. Dotted lines show where this subgraph links
@@
 on the left of the genome and a B.1.1.7 lineage on the right. In the subgraph we show
-the maximum genomic position for the recombination breakpoint, here at 22,227 basepairs
+the maximum genomic position for the recombination breakpoint, here at position 22,227
 (corresponding to a breakpoint strictly less than 22,228, see Table~\ref{tab:jackson})
@@
 In some cases,
-these may be, however, be a result of erroneous Pango designations.
+these may, however, be a result of erroneous Pango designations.
 Table~\ref{tab:pango-recombinants} also shows the official
@@
 requirement that all the samples assigned to a given lineage must trace back to
-a single recombination event. Using Nextclade-designations, 15 Pango X lineages are
+a single recombination event. Using Nextclade designations, 15 Pango X lineages are
 inferred to have multiple
-recombinant origins, such that their samples trace to $>$1 most recent
+recombinant origins, such that their samples trace to more than one most recent
 recombination node in the ARG. Of these, 11 are cases where the
-recombinants are independent rather than nested (i.e. there is no overlap in the list
+recombinants are independent rather than nested (i.e., there is no overlap in the list
 of descendant samples for each recombination node). Most have a single ``main''
@@
 (the origin of three XD samples, all sampled in France) has an earliest
-sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
-subgraph has a single sample with EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
+sample EPI\_ISL~11222324 (dated 2022-02-26), while the right hand
+subgraph has a single sample EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
 Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
 the breakpoints in each case are slightly different: the start of the Omicron insertion
-in the French samples has an estimated maximum position of 21641 basepairs
+in the French samples has an estimated maximum position of 21641bp
 (and minimum of 21619, not shown) with the insertion end occurring at a maximum position of 25584
@@
 samples, and the fact that the two samples differ at 23 nucleotide positions,
-suggests that these may indeed represent true independent Deltacron recombinants.
+suggests that these may indeed represent true independent Delta--Omicron recombinants.
 The canonical XD definition is based entirely on samples from northern Europe, particularly
 France (\url{https://github.com/cov-lineages/pango-designation/issues/444}) so it seems
-plausible that the earlier Turkish sample has been mislabelled XD by Nextclade.
+plausible that the earlier Turkish sample has been mislabelled as XD by Nextclade.
 Indeed, GISAID does not label any of these samples XD (see
@@
 between a B.1.177.18 sample (specifically the strain Wales/ALDP-115BF41/2021,
-EPI\_ISL: 1012804) which contributed the majority of the genome from the start
+EPI\_ISL~1012804) which contributed the majority of the genome from the start
 to a maximum position of 22227, and an unknown (inserted) node with imputed
 Pango lineage B.1.1.7, which contributed the remaining right hand portion. The
-recombination node has five immediate children: four sample leaves (EPI\_ISLs:
+recombination node has five immediate children: four sample leaves (EPI\_ISLs
 989697, 1019487, 1104468 and 1122630) and a UPGMA-inserted node which is the
@@
 recombination between a B.1 sample (specifically the strain
-England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of
+England/CAMB-7B47D/2020, EPI\_ISL 433960, which contributed the majority of
 the genome from the start up to a maximum position of 23604bp), and two
@@
 designated XB nodes, and involves a recombination between descendants of the
-originating XB recombination. More specifically strain
-USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination
-between an XB grandparent and its XB grandchild. This could well be
+originating XB recombination. More specifically, strain
+USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombinant
+between an XB grandparent and its XB grandchild, which could be
 artefactual.
@@
 by the WHO, the prevalence of SARS-CoV-2 continues to be high worldwide
-with persistent high levels of infection. This high prevalence has allowed the
+with persistent high levels of infection. This has allowed the
 proliferation of many variants, with more than 600 Pango-designated lineages
@@
 been central to pandemic responses thus far [FIXME: Examples and citations].
+% 1. Form the basis of variant classifications (UShER paper),
+% 2.
+Global context.
+In 2020, Nextstrain provided situation reports that were interactive, phylogenetically-driven visualizations of SARS-CoV-2 characteristics such as transmission https://nextstrain.org/ncov-sit-reps/
+% 3.
+The WHO's recommendations on Omicron asked countries to "report initial cases/clusters associated with VOC infection to WHO".
+Phylogenetics being one way to identify those crucial early clusters.
+https://www.who.int/news/item/26-11-2021-classification-of-omicron-(b.1.1.529)-sars-cov-2-variant-of-concern
 However, as recombinant lineages are now endemic and it is clear that
@@
 Here we have introduced the first method to infer an evolutionary
-history that jointly models genetic inheritance, mutation and
+history that jointly estimates genealogies with both mutation and
 recombination at pandemic scale, and illustrated how this single
@@
 (Section~\ref{sec:data_preprocessing}). A more sophisticated
-approach~\citep[e.g.][]{Aksamentov2021-hj} would
+approach~\citep[e.g.,][]{Aksamentov2021-hj} would
 likely yield significant improvements, and reduce the effect of sites with
@@
 on sequence motifs [CITATION?].
+% Citation: https://doi.org/10.1007/s00705-020-04750-z . This is the data source we used to create the breakpoints motif look up table.
 % Hint: there's probably a bunch of other stuff you could do with this
@@
 The results of inference for a given day is then
-a joint model of mutation, genetic inheritance and recombination
+a genealogy recording genetic inheritance as well as 
+mutation and recombination events
 for all of the sequences inserted into the ARG up to that day,
@@
 % mutations occurred. A pushed-reversion node is inserted when an immediate
-% reversion occurs (Figure 1E), i.e. a mutation that reverts another mutation on
+% reversion occurs (Figure 1E), i.e., a mutation that reverts another mutation on
 % its parent node. Instead of retaining both the mutation and immediate
@@
 has been incorporated into several downstream
-applications~\citep[e.g.][]{Haller2019-slim,Speidel2019-yh,
+applications~\citep[e.g.,][]{Haller2019-slim,Speidel2019-yh,
 Terasaki2021-geonomics,
@@
 and estimation of parameters such as
-recombination rates \citep[e.g.][]{Hinch2011-tz},
+recombination rates \citep[e.g.,][]{Hinch2011-tz},
 and selection within and across hosts in viral
-sequence data \citep[e.g.][]{Palmer2019-wa}.
+sequence data \citep[e.g.,][]{Palmer2019-wa}.
 See~\cite{Mcvean2019-linkage} for further review and discussion
@@
 SHZ is supported by the Janssen-Oxford Translational Genomics Fellowship. JK,
-YW, and BJ are supported by the Robertson Foundation.
+YW, and BJ are supported by the Robertson Foundation. AI is supported by the Wellcome Trust.
 
@@
 \caption{\label{fig:cophylogeny_long}Tanglegram equivalent to that in Figure~\ref{fig:cophylogeny},
-but for the Long ARG (i.e. subsampled to mid-2022).}
+but for the Long ARG (i.e., subsampled to mid-2022).}
 \end{figure}

### 2023-05-31 paper.tex

@@
 have 515 samples from 38 X lineages
-(511 samples from 35 lineages when we filter singletons).
+(511 samples from 35 lineages when we filter singleton recombinants). Of the GISAID designations,
+28 are shared with Nextclade (26 after filtering singletons).
 This variation in classifications highlights the uncertainty that exists when
@@
 (B) Equivalent subgraph for the 17 XAG samples in the Long ARG, with abbreviated labelling. Where
-non-XAG samples are ancestral to other samples, the number of descendant samples is marked as ``+1'',
-``+7'', etc.
+non-XAG samples are ancestral to further unplotted samples, the number of unplotted descendant samples
+is marked as ``+1'', ``+7'', etc.
 (C) Equivalent subgraphs for both origination events involving the four XD samples in the Long ARG.
@@
 Pango lineage B.1.1.7, which contributed the remaining right hand portion. The
-recombination node has five immediate children: four sample leaves (EPI\_ISLs
-989697, 1019487, 1104468 and 1122630) and a UPGMA-inserted node which is the
+recombination node has five immediate children: four sample leaves (strains
+Wales/ALDP-11CF93B/2021, Wales/ALDP-125C4D7/2021, Wales/LIVE-DFCFFE/2021, and
+Wales/ALDP-130BB95/2021) and a reversion-push node which is the
 ancestor of all other XA samples in the dataset.
@@
 Although the COVID-19 pandemic is no longer considered a global emergency
-by the WHO, the prevalence of SARS-CoV-2 continues to be high worldwide
-with persistent high levels of infection. This has allowed the
+by the WHO, the prevalence of SARS-CoV-2 continues to be high worldwide. This fuels 
 proliferation of many variants, with more than 600 Pango-designated lineages
@@
 Section~\ref{sec:breakpoint_intervals}).
-Using a pre-existing tree built using state-of-the-art phylogenetics methods
+Using a pre-existing tree built using state-of-the-art phylogenetic methods
 for the early stages of the pandemic (Section~\ref{sec:backbone_phylogeny}),
@@
 The vast volumes of whole genome sequence data generated during the pandemic
-has presented classical phylogenetics methods and software with major difficulties.
+has presented classical phylogenetic methods and software with major difficulties.
 Standard interchange formats such as FASTA, Newick and VCF were simply not designed

### 2023-05-31 plot.py
Update Jackson et al recombinants

@@ -27,7 +27,7 @@ import tqdm
 import pandas as pd
 
 import tsconvert  # Not on pip. Install with python -m pip install git+http://github.com/tskit-dev/tsconvert
-import sc2ts  # install with python -m pip install git+https://github.com/jeromekelleher/sc2ts
+import sc2ts
 
 import utils
 

### 2023-06-01 paper.tex
Updates from Overleaf

@@
 In this section we explore the detailed ancestral relationships between
-recombinant parents by investigate the patterns of divergence between
+recombinant parents by investigating the patterns of divergence between
 them.
@@
 and for Delta+Omicron are 20 / 13 (6 / 3).
-Note that these plots do not include recombination breakpoints involving lineages that were not classified into these VoC categories.
+Note only recombination breakpoints involving lineages
+classified into Alpha, Delta, and Omicron VoC categories are plotted above:
+all other breakpoints are omitted.
 }

### 2023-06-01 paper.tex
Update author section

@@
 \author[1]{Shing H. Zhan}
-\author[2,5]{Katherine Eaton}
-\author[3,5]{Anastasia Ignatieva}
-\author[1,5]{Benjamin Jeffery}
-\author[1,5]{Duncan S. Palmer}
-\author[1,5]{Yan Wong}
+\author[3$\star$]{Anastasia Ignatieva}
+\author[1$\star$]{Yan Wong}
+\author[2]{Katherine Eaton}
+\author[1]{Benjamin Jeffery}
+\author[1]{Duncan S. Palmer}
 \author[2]{Carmen L. Murall}
 \author[4]{Sarah P. Otto}
-\author[1,6]{Jerome Kelleher}
-\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, Oxford, United Kingdom}
+\author[1$\dagger$]{Jerome Kelleher}
+\affil[1]{Big Data Institute, Li Ka Shing Centre for Health
+Information and Discovery, University of Oxford, Oxford, United Kingdom}
 \affil[2]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
 \affil[3]{Department of Statistics, University of Oxford, Oxford, United Kingdom}
-\affil[4]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver, British Columbia, Canada}
-\affil[5]{Preliminary alphabetic ordering}
-\affil[6]{Correspondence. E-mail: jerome.kelleher@bdi.ox.ac.uk}
+\affil[4]{Department of Zoology and Biodiversity Research Centre,
+University of British Columbia, Vancouver, British Columbia, Canada}
+\affil[$\star$]{Joint second author}
+\affil[$\dagger$]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
 
@@
 Figure~\ref{fig:recomb_mrcas} shows that there is a
-broad range in divergence times among recombinant parents.
-We can see that the majority of recombinants arise
-from parents that diverged 10 to 30 weeks
-previously. There is also a peak of recombinants that diverged
-${\sim}95$ weeks prior to the event, which correspond to recombinants whose
+large range in divergence times among recombinant parents, with a broad
+spread of times from about 10 to 80 weeks prior to the recombination event
+and a minor additional peak of recombinants that diverged
+${\sim}95$ weeks, corresponding to recombinants whose
 parental common ancestor traces to early 2020. We would expect, for example

### 2023-06-01 paper.tex
Update data access section

@@
 up to 2022-08-22 and accessible at
-% Do we need to upload a separate Table or is this DOI sufficient?
 \url{https://doi.org/10.55876/gis8.230329cd}.
@@
 \section{Data Availability}
-The source code for \texttt{sc2ts} is
-available on GitHub
-\url{https://github.com/jeromekelleher/sc2ts/}.
-The Jupyter notebooks and code used to produce the results described here are also
-available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.
-
-% TODO what's the right way to say this?
+The source code for \texttt{sc2ts} and notebooks and code used to produce the
+results described here available on GitHub
+at \url{https://github.com/jeromekelleher/sc2ts/}
+and \url{https://github.com/jeromekelleher/sc2ts-paper/}.
+
+Details of the GISAID data used are available at
+\url{https://doi.org/10.55876/gis8.230329cd} and included in the
+Supplemental Table.
+
 The inferred ARGs described here are available on request to those with

### 2023-06-01 paper.tex
Add BMRC acks
Closes #62

@@
 
+Computation used the Oxford Biomedical Research Computing (BMRC) facility, a
+joint development between the Wellcome Centre for Human Genetics and the Big
+Data Institute supported by Health Data Research UK and the NIHR Oxford
+Biomedical Research Centre. The views expressed are those of the author(s) and
+not necessarily those of the NHS, the NIHR or the Department of Health.
+
 \section{Data Availability}
 The source code for \texttt{sc2ts} and notebooks and code used to produce the
-results described here available on GitHub
-at \url{https://github.com/jeromekelleher/sc2ts/}
-and \url{https://github.com/jeromekelleher/sc2ts-paper/}.
+results described here available on GitHub:
+\begin{itemize}
+\item \url{https://github.com/jeromekelleher/sc2ts/}
+\item\url{https://github.com/jeromekelleher/sc2ts-paper/}
+\end{itemize}
 

### 2023-06-01 paper.tex
Yans updates
Clear up some author comments

@@
 % Can reinstate when we're ready to preprint
-% \usepackage[hidelinks]{hyperref}
+\usepackage[hidelinks]{hyperref}
 
@@
 mutation and recombination, which
-can be conveniently and efficiently analysed using standard data science
+can be conveniently and efficiently analysed using standard Python data science
 tools. Inference is based on incrementally adding batches of sequences
@@
 single phylogenetic tree, particularly when looking at
-a subset of sampled genomes, e.g., restricted by time or limited to close
-relatives. We expect ARGs to be particularly tree-like
+a subset of strains. We expect ARGs to be particularly tree-like
 early in the pandemic, when co-infection was less likely and divergence between
@@
 ${\sim}95$ weeks, corresponding to recombinants whose
-parental common ancestor traces to early 2020. We would expect, for example
-Delta-Omicron recombinants to fall into this group, hence it can be
-useful to classify the breakpoints further by the VoC combinations of their
+parental common ancestor traces to early 2020. This latter group should
+contain, for example Delta-Omicron recombinants. More widely, we can
+further classify the breakpoints by the VoC combinations of their
 parent lineages. Considering only the Alpha, Delta, and Omicron VoC classes,
@@
 circulating globally in the last three months (January to March, 2023;
-https://gisaid.org/; access on March 27, 2023). High prevalence also brings a
-higher risk of coinfection, increasing opportunities for new phenotypically
+https://gisaid.org/; access on March 27, 2023). High prevalence increases the risk of coinfection, providing opportunities for new phenotypically
 distinct recombinants to emerge and spread. Phylogenetic approaches have
@@
 
-% We introduce an ARG inference method, summarised in
-% Figure~\ref{fig:overview_sc2ts}.
-% % Note going to rewrite this anyway, so not trying to make it flow.
-% It uses (1) an efficient HMM implementation of the Li and Stephens (LS) model
-% \citep{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2)
-% the reported collection dates of the SARS-CoV-2 samples to determine the order
-% of genome copying; and (3) several parsimony-based heuristics to locally refine
-% the topology. In brief, this combined approach uses the likelihood model of LS
-% to attach sample genomes to existing nodes and to detect recombinants, but
-% because many of the daily tip samples harbour few, if any, mutational changes,
-% we use parsimony to resolve their close relationships.
-
-% Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by
-% taking advantage of the reported collection dates to incrementally grow the
-% ARG. The inference process occurs forward in time, attaching new genomes
-% aggregated by their collection date (daily batches), from the earliest to the
-% last collection date. Prior to attaching sample genomes, the ARG is initialised
-% such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence
-% of recombination, the ARG is a single tree with no reticulation (Figure 1A);
-% however, in the presence of recombination, a reticulate ARG is generated, with
-% variation in tree topology across the genome (Figure 1B). Unlike standard
-% phylogenetic trees in which internal nodes always represent inferred ancestors,
-% internal nodes in the sc2ts ARGs represent sequences which can either be
-% inferred ancestors or known samples.
-
-% Each daily batch of genomes is first attached to the growing ARG without any
-% phylogenetic resolution. In cases where the newly added sequence is inferred to
-% be a recombinant, a recombination node is inserted; the ``date'' of this node
-% is arbitrarily set to the average of the collection dates of the genome and the
-% most recent of its parent nodes. Otherwise, the genomes are attached to the
-% most closely related ancestral node (an attachment node). This creates
-% polytomies that are then locally resolved among the descendants of each
-% attachment node using the Unweighted Pair Group Method with Arithmetic Mean
-% (UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
-% This hierarchical clustering algorithm produces a dendrogram relating a set of
-% sequences, which is treated as an ultrametric binary tree here. Subsequently,
-% mutations are placed onto the daily tree by maximum parsimony. An issue with
-% the UPGMA algorithm is that it may create internal nodes supported by no
-% informative site (i.e., having no mutation immediately ancestral to them). In
-% an adjustment step, such unsupported internal nodes are removed to
-% ``un-resolve'' the daily tree; the remaining internal nodes, which are
-% supported by at least one shared mutation, are retained. The ``dates'' of the
-% non-sample internal nodes are set to the same date as the samples in the daily
-% batch. The resulting daily tree is grafted onto the attachment node in the
-% global ARG.
-
-% During ARG inference, two types of non-sample nodes are inserted to improve
-% parsimonious placement of mutations: mutation-coalescing nodes and
-% pushed-reversion nodes. We call these operations tree adjustments. A
-% mutation-collapsing node is inserted when a shared mutation between sibling
-% nodes is found (Figure 1D). An attempt to create mutation-collapsing nodes is
-% made after grafting a new daily tree onto the global ARG. If shared mutations
-% exist between two or more sibling nodes, then a parsimonious explanation is to
-% create a new node that is the parent of those siblings, on which these
-% mutations occurred. A pushed-reversion node is inserted when an immediate
-% reversion occurs (Figure 1E), i.e., a mutation that reverts another mutation on
-% its parent node. Instead of retaining both the mutation and immediate
-% reversion, only the mutation is kept after a pushed-reversion node is inserted,
-% thereby reducing the number of mutations on the branches. Note that these
-% non-sample nodes are used as reconstructed genomes during HMM copying.
-
-% Furthermore, during inference, the ARG is automatically encoded, stored, and
-% updated as a tree sequence using tskit \citep{Kelleher2018-xc}.
-
 \subsection{Ancestral Recombination Graphs}
@@
 probability of being chosen.
-More sophisticated parameterisations where we
-condition on known per-site mutation rates (and perhaps per-base subsitution
-rates) may lead to better results, although perhaps at the cost of some
-reduction in interpretability. Post-processing the match results to produce
+Post-processing the match results to produce
 more parsimonious breakpoints may also be a worthwhile avenue for development.
@@
 \item \url{https://github.com/jeromekelleher/sc2ts/}
-\item\url{https://github.com/jeromekelleher/sc2ts-paper/}
+\item \url{https://github.com/jeromekelleher/sc2ts-paper/}
 \end{itemize}
@@
 
-%% Could insert the Pango_X_graph_sample_maps.txt with something like this
-%% but maybe best to leave it as a separate file?
-%% \lstinputlisting[caption={XXX}]{data/Pango_X_graph_sample_maps.txt}
-
 \end{document}

### 2023-06-01 paper.tex
Various changes implementing Sally's comments

@@
 \begin{abstract}
-Recombination is an important and ongoing feature of SARS-CoV-2 evolution,
-and presents major challenges to [things important to public health].
-% ...
-genomic surveillance programs, predictive models for symptoms and severity, policy recommendations such as vaccine effectiveness and booster frequency, etc.
-(https://doi.org/10.1016/S1473-3099(22)00831-3)
-Detection of new recombinant strains is challenging, with methods
-[being a bit rubbish].
-% ... with few methods that facilitate rapid and systematic (reproducible) classification .https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/
+Recombination is an ongoing and increasingly important feature of circulating lineages of SARS-CoV-2, challenging how we represent the evolutionary history of this virus and giving rise to new variants of potential public health concern by combining transmission and immune evasion properties of different lineages.
+Detection of new recombinant strains is challenging, with most methods looking for breaks between sets of mutations that characterize distinct lineages.
 In addition, many basic approaches fundamental to the study of viral
@@
 immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
-% Early in the SARS-CoV-2 pandemic there was insufficient
-% genetic variation for meaningful recombination.
 Early on in the COVID-19 pandemic, the levels of genetic diversity
@@
 
+\begin{figure} \centering
+\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
+\caption{\label{fig:overview_sc2ts}
+A schematic of the \texttt{sc2ts} method.
+Sc2ts reconstructs the genetic relationships
+among SARS-CoV-2 genomes by copying samples to all possible ancestors collected
+at earlier time points (curved arrows). Each daily iteration involves three stages:
+(1) attachment of new samples to the growing ARG (A, B); (2) reconstruction of
+trees relating the samples under each attachment node (C); and parsimony-based
+tree topology adjustments (D, E). In the absence of  recombination, sc2ts
+infers an ARG that is a single tree relating the samples (A). However, in the
+presence of recombination, sc2ts infers an ARG that is encoded as a sequence of
+local trees relating segments of the sample genomes (B), inserting a
+recombination node to represent a putative recombination event. Additionally,
+mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to
+make more parsimonious placements of mutations that should be shared or should
+not be immediately reverted, respectively.}
+\end{figure}
+
 % [PARA 4: we present sc2ts based on these recent advances. Summarise
@@
 can be conveniently and efficiently analysed using standard Python data science
-tools. Inference is based on incrementally adding batches of sequences
+tools. As illustrated in Figure~\ref{fig:overview_sc2ts}, inference is based on incrementally adding batches of sequences
 based on their collection dates, and proceeds in three phases.
@@
 % chosen for each of the samples in the daily batch using the
-Li and Stephens (LS) model (Section~\ref{sec:ls}). Then, since many samples
+Li and Stephens (LS) model (Figure~\ref{fig:overview_sc2ts}A, B; Section~\ref{sec:ls}). Then, since many samples
 in the batch can share an attachment path,
 we infer phylogenetic trees for each of the clusters separately using standard
-methods (Section~\ref{sec:sample-cluster-tree-inference}).
+methods (Figure~\ref{fig:overview_sc2ts}C; Section~\ref{sec:sample-cluster-tree-inference}).
 Finally, we attach the trees for these sample-clusters to the current
@@
 introduced by the inherent greediness of this strategy
-(Section~\ref{sec:parsimony-heuristics}).
+(Figure~\ref{fig:overview_sc2ts}D, E; Section~\ref{sec:parsimony-heuristics}).
 Using the current preliminary version of \texttt{sc2ts}, we infer two
@@
 \label{sec:parent_divergence}
-% Paragraph 1, what does all this mean?
-% Purpose of this paragraph - remind the reader that we can only do this
-% sort of things because we have an ARG, and we can do  this at a far
-% greater level of granularity than existing methods.
-An important advantage of explicitly inferring an ARG compatible
-with a given sample of sequences (Section~\ref{sec:args})
-is that this allows us to precisely trace the origins of recombinants.
-Methods based on motifs among pre-specified lineages
-provide us with estimates of the origins of recombinants as
-parents belonging to one or other of these groups. Defining the
-groups requires care [FIXME and other fiddly things to mention Katherine?]
-% ...
-Defining the groups is a highly subjective process, involving numerous manual steps that require subject matter expertise of the organism, that often use web sites as data sources which are difficult to version-control, and thus hinder reproducibility.
-The web paper by the PH4GE group is one of the best sources for the reality (and pain) of how it's done: https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/
-RIPPLES~\citep{Turakhia2022-it} does provide precise estimates
-of the parental origins of a recombinant, but since it depends
-on an existing phylogenetic tree, cannot reason about
-nested recombinant origins.
-
-\begin{figure} \centering
-\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
-\caption{\label{fig:recomb_mrcas}
-Date of common ancestry between the parents on either side of recombination
-breakpoints, as a function of the divergence time between the parents.
-MRCAs of parents associated with Pango designated recombinants (XA, etc)
-are identified in orange. Larger symbols represent breakpoints in recombination
-nodes ancestral to five or more samples.
-Horizontal dotted lines show the four most common MRCA nodes,
-which tend to be associated with major outbreaks and with many immediate children.
-The stacked histogram shows the distribution of parental divergence times, ranging
-from parents that have diverged only a handful of days previously, to much more
-divergent parent lineages. See Figure~\ref{fig:recomb_mrcas_voc_breakdown}
-for equivalent plots broken down by parental VoC classification.
-}
-\end{figure}
-
 % Paragraph 2, what do we plot, specifically?
@@
 We focus on the Long ARG because it covers time
-periods where substantial recombination is known to have occurred,
+periods where substantial recombination is known to have occurred
 and contains samples from 33 Pango X lineages (i.e., those inferred
@@
 
+\begin{figure} \centering
+\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
+\caption{\label{fig:recomb_mrcas}
+Date of common ancestry between the parents on either side of recombination
+breakpoints, as a function of the divergence time between the parents.
+MRCAs of parents associated with Pango designated recombinants (XA, etc)
+are identified in orange. Larger symbols represent breakpoints in recombination
+nodes ancestral to five or more samples.
+Horizontal dotted lines show the four most common MRCA nodes,
+which tend to be associated with major outbreaks and with many immediate children.
+The stacked histogram shows the distribution of parental divergence times, ranging
+from parents that have diverged only a handful of days ago, to much more
+divergent parent lineages. See Figure~\ref{fig:recomb_mrcas_voc_breakdown}
+for equivalent plots broken down by parental VoC classification.
+}
+\end{figure}
+
 % Paragraph 3, what do we say about the dates/nodes of the MRCAs?
@@
 recombinants (the top four are indicated in the figure).
-These shared MRCAs lie near the root of large expansions
+These shared MRCAs lie near the root of large expansions,
 % Anyone reading this will know what these are
@@
 and the majority are associated with large polytomies, likely
-indicating a rapid and under-sampled expansion of a clade.
+indicating a rapid and under-sampled expansion of a clade (e.g. major
+Delta and Omicron waves).
 
@@
 Figure~\ref{fig:recomb_mrcas} shows that there is a
-large range in divergence times among recombinant parents, with a broad
+large range in divergence times among parents of detected recombinants, with a broad
 spread of times from about 10 to 80 weeks prior to the recombination event
-and a minor additional peak of recombinants that diverged
-${\sim}95$ weeks, corresponding to recombinants whose
-parental common ancestor traces to early 2020. This latter group should
+and a minor additional peak involving recombination between lineages that diverged
+${\sim}95$ weeks ago, corresponding to common ancestors which trace back to early 2020.
+This latter group should
 contain, for example Delta-Omicron recombinants. More widely, we can
@@
 treated with a degree of caution,
-since non-sample nodes are crudely dated in the current version
+because non-sample nodes are crudely dated in the current version
 of \texttt{sc2ts} (see Section~\ref{sec:node_dating}, in particular
@@
 We focus primarily on the Long ARG
-since it contains many more recombinants (the status
+which contains many more recombinants (the status
 of Pango X lineages in the Wide ARG is briefly
@@
 but two of which descend from a recombination node in the Long ARG.
-These two samples are designated XP, and a likely
+These two exceptions are both designated XP, and a likely
 explanation for the absence of a corresponding recombination event in the Long ARG
@@
 (\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
-during preprocessing (See Section~\ref{sec:data_preprocessing} for details and potential
+during our preprocessing (see Section~\ref{sec:data_preprocessing} for details and potential
 improvements). Of the remaining X designated samples, 38 are singleton recombinants,
 descending
-from recombination nodes that are ancestral only to that sample (10 are XZ; 6
+from recombination nodes that are ancestral only to that sample (10 are labelled XZ; 6
 are XE;  3 each from XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1
@@
 Pango status is imputed. Genomic regions inherited by the recombination node are shown;
-breakpoints correspond to the maximum breakpoint position inferred by \texttt{sc2ts}.
+breakpoints correspond to the rightmost breakpoint position inferred by \texttt{sc2ts}.
 (B) Equivalent subgraph for the 17 XAG samples in the Long ARG, with abbreviated labelling. Where
@@
 Sample nodes are plotted as larger circles, but the subgraph also includes intermediate,
-non-sample nodes: i.e., inserted nodes such as UPGMA, recombination, mutation collapse, and
-reversion push nodes, plotted as smaller circles. Dotted lines show where this subgraph links
+non-sample nodes (i.e. inserted by sc2ts, see Sections~\ref{sec:sample-cluster-tree-inference} and \ref{sec:parsimony-heuristics}). Dotted lines show where this subgraph links
 to the rest of the ARG. Above recombination nodes, only ancestral nodes are shown, meaning
@@
 on the left of the genome and a B.1.1.7 lineage on the right. In the subgraph we show
-the maximum genomic position for the recombination breakpoint, here at position 22,227
+the rightmost genomic position for the recombination breakpoint, here at position 22,227
 (corresponding to a breakpoint strictly less than 22,228, see Table~\ref{tab:jackson})
@@
 to note that if the GISAID Pango designations are used, many of the samples
-marked here as XAB are reclassified as BA.2 (and XAG becomes fully monophyletic,
-see supplementary Figure \ref{fig:pango_XAG_gisaid_graph}). This is an
+marked here as XAB are reclassified as BA.2 and XAG becomes fully monophyletic
+(although still not an immediate descendant of the originating recombination node,
+see Supplementary Figure \ref{fig:pango_XAG_gisaid_graph}). This is an
 independent confirmation of the uncertainty in designation of these XAG-related
@@
 Pango designated parent lineages and \texttt{sc2ts} inferred parent lineages,
-which extensively agree (although \texttt{sc2ts} provides more a precise
+which extensively agree (although \texttt{sc2ts} provides a more precise
 parent designation).
@@
 recombination event from which the majority of the corresponding recombinant samples
-descend, and which agrees with the offical Pango designated parent lineages
+descend and which agrees with the offical Pango designated parent lineages
 (see Table~\ref{tab:pango-recombinants}, but note that
@@
 the 4 samples labelled XD by Nextclade in the Long ARG. The left hand subgraph
-(the origin of three XD samples, all sampled in France) has an earliest
-sample EPI\_ISL~11222324 (dated 2022-02-26), while the right hand
-subgraph has a single sample EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
+(containing three XD samples, all sampled in France) has an earliest
+sample (EPI\_ISL~11222324, dated 2022-02-26), while the right hand
+subgraph has a single XD sample (EPI\_ISL~8514045, sampled in Turkey and dated 2021-12-30).
 Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
 the breakpoints in each case are slightly different: the start of the Omicron insertion
-in the French samples has an estimated maximum position of 21641bp
-(and minimum of 21619, not shown) with the insertion end occurring at a maximum position of 25584
-(and a minimum of 25470). In contrast, the Omicron insertion in the Turkish sample
-is inferred to have occurred between positions 22578 (min 21619) and 24130 (min 23605).
+in the French samples has an estimated rightmost position of 21641bp
+(and a leftmost of 21619, not shown) with the insertion end occurring at a rightmost position of 25584
+(and a leftmost of 25470). By contrast, the Omicron insertion in the Turkish sample
+is inferred to have occurred from position 21619-22578 to position 23605-24130.
 The breakpoint difference, the different geographical locations, the time between the
 samples, and the fact that the two samples differ at 23 nucleotide positions,
-suggests that these may indeed represent true independent Delta--Omicron recombinants.
+suggests that these may indeed represent independent Delta--Omicron recombinants.
 The canonical XD definition is based entirely on samples from northern Europe, particularly
@@
 recombination events, four Pango X lineages in the Long ARG have Nextclade-designated
-samples whose ancestry involves multiple non-independent (``nested'') recombination events
+samples whose ancestry involves further recombination events
 (marked by~\textdagger ~in Table~\ref{tab:pango-recombinants}; the most
@@
 is only distantly related to the true originating recombinant. This could account
-for complex and potentially artefactual relationships around these nodes, and
+for complex and potentially artefactual relationships around these nodes and
 is likely to be due to undersampling of the XB outbreak. Investigating examples
@@
 %% Quick overview of the wide ARG.
-Since the Wide ARG is restricted to data collected prior
+Because the Wide ARG is restricted to data collected prior
 to mid-2021, it contains samples from only three Pango-designated recombinant
@@
 reduced to 235 Nextclade-designated samples (229 GISAID) and XC is reduced to 4
-(none in the GISAID designations). We correctly identify that all samples
+(none in the GISAID designations). We confirm that all samples
 designated as XA, XB, or XC by any method have one or more recombination nodes
@@
 As in the Long ARG, all XA samples in the Wide ARG trace back to a
-unique originating recombination node. This is the product of a recombination
-between a B.1.177.18 sample (specifically the strain Wales/ALDP-115BF41/2021,
-EPI\_ISL~1012804) which contributed the majority of the genome from the start
-to a maximum position of 22227, and an unknown (inserted) node with imputed
+unique originating recombination node, consistent with Figure \ref{fig:pango-simple-origin-graph}A.
+This is the product of a recombination
+between a B.1.177.18 sample (specifically the strain Wales/ALDP-115BF41/2021)
+which contributed the majority of the genome from the start
+to a rightmost position of 22227, and an unknown (inserted) node with imputed
 Pango lineage B.1.1.7, which contributed the remaining right hand portion. The
@@
 Wales/ALDP-11CF93B/2021, Wales/ALDP-125C4D7/2021, Wales/LIVE-DFCFFE/2021, and
-Wales/ALDP-130BB95/2021) and a reversion-push node which is the
-ancestor of all other XA samples in the dataset.
+Wales/ALDP-130BB95/2021) and a inserted node which is the
+ancestor of all other XA samples in the dataset. The geographical clustering inferred by \texttt{sc2ts} 
+for these samples matches the findings of \citet{Jackson2021-ik}.
 
@@
 recombination between a B.1 sample (specifically the strain
-England/CAMB-7B47D/2020, EPI\_ISL 433960, which contributed the majority of
-the genome from the start up to a maximum position of 23604bp), and two
-UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of
+England/CAMB-7B47D/2020, which contributed the majority of
+the genome from the start up to a rightmost position of 23604bp), and two
+UPGMA nodes with imputed Pango lineages B.1.627 (up to a rightmost position of
 27389bp) and B.1.36.8 (the remaining fragment of the genome). Figure~\ref{fig:complex_origins_graph}
@@
 of the Nextclade-designated XB samples. The one exception accounts for about 17\% of the
-designated XB nodes, and involves a recombination between descendants of the
+designated XB nodes and involves a recombination between descendants of the
 originating XB recombination. More specifically, strain
-USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombinant
-between an XB grandparent and its XB grandchild, which could be
-artefactual.
+USA/TX-HMH-MCoV-43092/2021 is inferred to be a recombinant
+between an XB grandparent and its XB grandchild, which 
+is likely to be artefactual.
 
@@
 XC by Nextclade are designated as XC by GISAID, these patterns could be due to
-mislabelling, and greater number of XC samples would be needed to draw reasonable
+mislabelling, and a greater number of XC samples would be needed to draw reasonable
 conclusions.
@@
 distinct recombinants to emerge and spread. Phylogenetic approaches have
-been central to pandemic responses thus far [FIXME: Examples and citations].
+been central to pandemic responses thus far \citep{Attwood2022-ab}.
 % 1. Form the basis of variant classifications (UShER paper),
@@
 https://www.who.int/news/item/26-11-2021-classification-of-omicron-(b.1.1.529)-sars-cov-2-variant-of-concern
-However, as recombinant lineages are now endemic and it is clear that
-recombination is a major feature of ongoing SARS-CoV-2 evolution,
+However, with the rise to high frequency of recombinant lineages \citep[e.g. XBB; ][]{Tamura2023-ab},
 it is imperative that these methods are updated to incorporate the
-effects of recombination, and future public health interventions are
+effects of recombination, so that future public health interventions are
 not based on incomplete and potentially biased evolutionary models.
@@
 history that jointly estimates genealogies with both mutation and
-recombination at pandemic scale, and illustrated how this single
+recombination at pandemic scale and illustrated how this single
 structure accurately captures results derived by many different means.
@@
 we caution against overinterpreting current results. As we have sought
-to illustrate throughout, there are some clear areas for improvement
-and some relatively straightforward changes that could be made to
-improve inference. The pipeline used to identify and
+to illustrate throughout, there are some clear areas for improvement.
+The pipeline used to identify and
 mask erroneous sites in the input alignments is simplistic, and, among
@@
 approach~\citep[e.g.,][]{Aksamentov2021-hj} would
-likely yield significant improvements, and reduce the effect of sites with
+likely yield significant improvements and reduce the effect of sites with
 artefactually high levels of mutation (e.g., site 28271;
@@
 A major source of errors are the ``time-traveller'' samples, whose
-provided collection dates are months (or years) too early
+recorded collection dates are months (or years) too early
 (Section~\ref{sec:filtering_time_travellers}).
@@
 Taken together, these and other relatively minor improvements should
-enable inference over much larger subsets of the dataset, and
+enable inference over much larger subsets of the dataset and
 give a clearer picture of the combined processes of recombination and
@@
 breakpoint could be chosen within the interval, for example based
-on sequence motifs [CITATION?].
+on sequence motifs \citep{Gallaher2020-mr, Yang2020-ct}.
 % Citation: https://doi.org/10.1007/s00705-020-04750-z . This is the data source we used to create the breakpoints motif look up table.
@@
 % of code reuse and ease of analysis is a major bonus.
-The vast volumes of whole genome sequence data generated during the pandemic
-has presented classical phylogenetic methods and software with major difficulties.
+The vast volume of whole genome sequence data generated during the pandemic
+has presented classical phylogenetic methods and software with major
+difficulties~\citep{Hodcroft2021-wt}.
 Standard interchange formats such as FASTA, Newick and VCF were simply not designed
@@
 samples~\citep{Kelleher2016-wk,Kelleher2019-ba}.
-The \texttt{sc2ts} package is written entirely in Python,
+The \texttt{sc2ts} package is written entirely in Python
 and by reusing existing high-performance components can infer
@@
 section for details of the corresponding Jupyter notebooks).
-The ability to quickly and easily develop new methods and
-analyses, at scale, by using existing high-quality software
-components and data formats is an essential element of future
-pandemic preparedness.
-
-% The COVID-19 pandemic continues to be a global emergency,
-% with persistent high levels of infection. This high prevalence has allowed the
-% proliferation of many variants, with more than 600 Pango-designated lineages
-% circulating globally in the last three months (January to March, 2023;
-% https://gisaid.org/; access on March 27, 2023). High prevalence also brings a
-% higher risk of coinfection, increasing opportunities for new phenotypically
-% distinct recombinants to emerge and spread.
-
-% In this proof-of-concept study, we demonstrate the feasibility of
-% reconstructing a large ARG from over one million SARS-CoV-2 genomes, a feat
-% that is simply not possible using other available ARG inference methods
-% \citep{Rasmussen2014-el,Ignatieva2021-rg, Speidel2019-yh}. We have accomplished
-% this feat by building on top of recent advances in storing and processing large
-% ARGs (tskit) and inferring genealogical relationships efficiently at scale
-% (tsinfer). The highly compact representation of the ARGs as succinct tree
-% sequences, coupled with vectorised computing technologies, enables the large
-% SARS-CoV-2 ARGs to be loaded in less than one second in a Jupyter notebook and
-% allows for each SARS-CoV-2 genome to be stored in less than eleven bytes of
-% data. We believe that further development of the sc2ts methodology and software
-% can create a powerful tool (1) to efficiently detect emerging or previously
-% overlooked recombinants in SARS-CoV-2; (2) to jointly model the mutational and
-% recombinational processes shaping global SARS-CoV-2 genomic diversity; and (3)
-% to perform a broad variety of evolutionary analyses of SARS-CoV-2 genomes with
-% unprecedented global sequencing coverage.
-
-% ARGs are rich genealogical structures that contain information about the
-% genetic relationships among sampled genomes in the presence of recombination.
-% Here, we have conducted several analyses of the preliminary ARGs to show that
-% the ARGs built using sc2ts can (1) recapitulate the phylogenetic relationships
-% among major SARS-CoV-2 lineages; (2) recover biologically plausible
-% recombination signals in SARS-CoV-2 genomes; and (3) model mutation events in
-% SARS-CoV-2 evolutionary history.
-
-% ARGs reconstructed using sc2ts capture the phylogenetic relationships of major
-% SARS-CoV-2 lineages. Our results show that the backbone topology of a densely
-% sampled Wide ARG inferred using sc2ts agrees well with the backbone topology of
-% a NextStrain tree inferred using IQ-Tree2, a widely used phylogenetic inference
-% package. However, we have noted several areas of differences (e.g., poorly
-% calibrated branch lengths), which are the focus of future development of sc2ts.
-% We believe that some of these differences can be reduced by improving and
-% refining our approach and strategy to handle samples with erroneously reported
-% collection dates (i.e., time travellers that have passed the maximum submission
-% delay of 30 days) and to modify the topology of daily trees.
-
-% Our results show that sc2ts identifies the early recombinants (pre-2022)
-% previously identified by \cite{Jackson2021-ik}. Sc2ts also identifies many
-% Pango recombinants, which are designated after community perusal of supporting
-% genomic evidence. We have found several previously unreported and overlooked
-% recombinant sequences in the Wide ARG. After further improvements and
-% enhancements, we will use sc2ts to build a comprehensive ARG to search for
-% recombination signals among all the 15 million and counting SARS-CoV-2 genomes
-% available.
-
-% ``The rate of putative recombination breakpoints is about three times higher towards the 3' of the change point than the 5' interval.'' Consistent with this, the breakpoint intervals of the putative recombinants identified here tend to occur on the 3’ end of the SARS-CoV-2 genome.
-
-% \section{Improvements}
-% Currently, sc2ts is in development. Below, we propose several ways to potentially improve or enhance sc2ts. We welcome suggestions and ideas from experts.
-
-% \begin{itemize} \item A potential source of error in sc2ts ARGs comes from
-% incorrectly reported collection dates of the samples, which can cause sample
-% genomes to be attached to the wrong parts of the ARGs. Thus far, we have
-% attempted to correct the most offending errors in the ARGs by filtering out
-% ``time travellers'', but the sampling time can still be wrong by days or tens
-% of days. ***describe possible approaches to correct this?*** \item To avoid
-
-% time travellers, we used a maximum submission delay, but this will
-% systematically bias against submissions from countries with greater submission
-% delays (Kalia et al.). An area of future work is to add back the excluded
-% sequences to a built ARG to assess which ones are likely true time travellers.
-% \item We have used a single mismatch value in all the HMM runs, set so that
-% having three mutations is as likely as having a single recombination. This
-% parameter choice is arbitrary, and higher values may reduce the number of
-% origins for those Pango lineages that were inferred to descend from more than
-% one recombinant ancestor. In addition, because both the mutation rate and
-% recombination rate varies across the SARS-CoV-2 genome, it may be helpful to
-% allow the mismatch value to vary along the genome. \item By treating mutations
-% equally in the current HMM implementation, we implicitly assume that the rates
-% of different types of mutations are identical. However, as reported previously
-% \citep{Yi2021-sc}, there is a high degree of mutational asymmetry in
-% SARS-CoV-2. To account for mutational biases, more sophisticated nucleotide
-% substitution models (e.g., allowing for different rates of transition and
-% transversion) can be incorporated into HMM likelihood calculations. \item Our
-% current masking strategy handles indels as missing data. Doing this can remove
-% information helpful to detect recombinants. For example, sequences of the Pango
-% recombinant XP were not identified as recombinants, because XP’s characteristic
-% multibase deletion was masked. \item Daily trees are built using the UPGMA
-% algorithm. This entails an extra step to ``un-resolve'' daily trees in parts of
-% the trees with no phylogenetic resolution. Using a maximum-parsimony algorithm
-% to build daily trees would leave these "soft polytomies" in, removing the need
-% for the un-resolving step. \end{itemize}
-
-% \section{Conclusions}
-% The global understanding of viral evolution during the COVID-19 pandemic is
-% unprecedented, with over 15 million sequences shared to date. We envisage that
-% once fully developed and tested, sc2ts methodology and software may provide a
-% unified computational framework to study the evolution and epidemiology of
-% SARS-CoV-2 at pandemic scale. The evolutionary analyses presented herein
-% exemplify the numerous applications possible using the ancestral recombination
-% graph inferred by sc2ts. We invite bioinformaticians, evolutionary biologists,
-% and virologists to join us and the broader tskit community to build the most
-% complete representation of the evolutionary history of SARS-CoV-2. Moreover,
-% while we have focussed on the particular application to SARS-CoV-2, the methods
-% we describe are readily applicable to the analysis of many other viral
-% pathogens. Solving the challenges in building and analysing large viral
-% genealogies would therefore contribute to the important goal of improving
-% preparedness for potential future pandemics.
+Retooling methods to scale rapidly and to encode recombination
+promises to improve tracking during this and any future pandemic.
 
@@
 will mostly consist of a new sample sequence copying from a node in the ARG
-with a small number of mutations, and often many samples
+with a small number of mutations and often many samples
 from a daily batch will copy from the same ARG node.
@@
 
-\begin{figure} \centering
-\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
-\caption{\label{fig:overview_sc2ts}
-A schematic of the \texttt{sc2ts} method.
-Sc2ts reconstructs the genetic relationships
-among SARS-CoV-2 genomes by copying samples to all possible ancestors collected
-at earlier time points (curved arrows). Each daily iteration involves three stages:
-(1) attachment of new samples to the growing ARG (A, B); (2) reconstruction of
-trees relating the samples under each attachment node (C); and parsimony-based
-tree topology adjustments (D, E). In the absence of  recombination, sc2ts
-infers an ARG that is a single tree relating the samples (A). However, in the
-presence of recombination, sc2ts infers an ARG that is encoded as a sequence of
-local trees relating segments of the sample genomes (B), inserting a
-recombination node to represent a putative recombination event. Additionally,
-mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to
-make more parsimonious placements of mutations that should be shared or should
-not be immediately reverted, respectively.}
-\end{figure}
-
 \subsection{Ancestral Recombination Graphs}
@@
 simulations~\citep{Kelleher2016-wk}, the methods have been
-extended and applied to forwards-time
+extended and applied to forward-time
 simulations~\citep{Kelleher2018-xc,Haller2018-tree},
@@
 Given these tables, we can then efficiently construct the local
-genealogical trees along the genome (arising from recombination),
+genealogical trees along the genome (arising from recombination)
 and also perform a range of calculations efficiently by
@@
 in C, it provides interfaces in C, Python and Rust. The Python interface
-is based on NumPy~\citep{Harris2020-array}, and provides a convenient
+is based on NumPy~\citep{Harris2020-array} and provides a convenient
 platform for interactive analysis of large-scale data using, for
@@
 is therefore to perform some local updates that target specific
-types of parsimony violation in the just-updated regions of the
+types of parsimony violations in the just-updated regions of the
 ARG. There are currently two parsimony-increasing operations
@@
 
-Mutation collapsing inspects the siblings of a newly attached node
-to check if any of them share (a subset of) the mutations that
+Given a newly attached node, mutation collapsing inspects its siblings from
+previous sample days to check if any of them share (a subset of) the mutations that
 it carries. If so, we increase the overall parsimony of the inference
@@
 those shared mutations, and make that new node the parent of the
-siblings carrying those shared mutations. The patterns of
+siblings carrying those shared mutations (Figure \ref{fig:overview_sc2ts}D). The patterns of
 shared mutations between siblings can be complex, and the current
@@
 if any of its mutations are ``immediate reversions''; that is,
-are reversions of a mutation that occured on the new node's
+are reversions of a mutation that occurred on the new node's
 immediate parent. We increase the overall parsimony of the
@@
 original parent which carries all its mutations except those
-causing the reversions on the newly added node.
+causing the reversions on the newly added node (Figure \ref{fig:overview_sc2ts}E).
 Table~\ref{tab:args} shows that
@@
 greedy operations, just examining the local parts of the
-ARG topology affected by newly added samples. Since the inferred
+ARG topology affected by newly added samples. Because the inferred
 ARGs still contain a large number of reversion mutations which
 are likely to be mostly artefactual, it is clear that there
-is room for improvement, and that further parsimony-increasing
+is room for improvement and that further parsimony-increasing
 heuristics would likely be of benefit.

### 2023-06-01 paper.tex
Partial changes from Sally for intervals section

@@
 and as a result can only identify inter-lineage recombination events.
-% specific to those motifs.
-% % JK: I've left this out because I felt it broke the narrative up a bit, but
-% % happy to put back in. (We're still citing the paper in a few places).
-% Others have inferred recombinant
-% genealogies to disentangle recombination from recurrent mutation, without
-% having to pre-specify characteristic mutations for each lineage
-% \citep{Ignatieva2021-rg}.
 \cite{Turakhia2022-it} presented a
@@
 a recombinant sequence switches from inheriting from one parent to another.
-Even if we observe the recombinant sequence before signficant
+Even if we observe the recombinant sequence before subsequent
 divergence occurs (during onward transmission), if the two parent sequences
@@
 distribution of these interval widths (truncated at 10kb).
-The bottom panel shows the number of intervals that intersect with
-each site along the genome (left axis, orange), and the number of mutations
+The bottom panel shows the number of intervals that span 
+each site along the genome (left axis, orange) and the number of mutations
 per site (right axis, blue).
@@
 
-Although we emphasise that further work is required to filter out spuriously inferred
-recombination events, we can draw two preliminary conclusions from
-Figure~\ref{fig:breakpoint-distribution}.
-% While the detailed interpretation of Figure~\ref{fig:breakpoint-distribution}
-% requires further study, a few key properties stand out.[TODO, write this
-% better. ]
-Firstly, in several cases there is a clear relationship between
-gene boundaries and the breakpoint intervals. For example, ORF7a and OR7b
-show sharp differences in the number of intersecting intervals within
-these genes. This could be due to a number of factors, such as [FIXME: BIOLOGY?]
-% reduced frequency of breakpoint sequence motifs, or secondary structure.
-Sorry, I don't have a citation for this, just theories.
-The RIPPLES paper vaguely mentions "important mechanistic bias shapes the distribution of recombination events".
-But maybe we just don't know yet?
-or the presence of sites with artefactually high numbers of mutations
-in the intergenic regions (see the next paragraph for more discussion).
-Secondly, we see substantially fewer intervals intersecting with the N
-gene than other regions, in agreement with the results of
-\cite{Turakhia2022-it}.
+Although we emphasise that further work is required to filter out spuriously inferred recombination events, 
+we can draw some preliminary conclusions from Figure~\ref{fig:breakpoint-distribution}.
+The number of intervals spanning a site (orange curve) is lower at the ends of the genome, 
+as expected due to the lack of information about recombination with few flanking sites.
+In addition, the number of intervals spanning a site often 
+drops near the beginning of a gene.
+This is particularly apparent at the ORF8/N gene interface, 
+with the N gene containing fewer potential recombination breakpoints 
+than other genes, in agreement with the results of \cite{Turakhia2022-it}. 
+Noticeable declines in the number of spanning intervals are also 
+seen near the beginning of S, M, ORF7a, and ORF8.
+These declines are sometimes associated with hypermutated sites (e.g., 27,384 and 28,271 
+near the beginning of ORF7a and N, respectively), as expected because sites that 
+undergo mutation at high rates are more likely to differ between the 
+parents of a recombinant and so provide information about the location of a breakpoint.
+This pattern may, however, also be an artefact of sequencing errors causing sites 
+to appear different between the parents when they are not 
+(indeed, there is evidence that 28271
+
+The 7,572 mutations (including 5,782 insertions and 1,605 deletions)
+at this site are an indicator that this
+homopolymeric region may be prone to sequencing errors
+may…).
+In other cases, however, the drop in intervals does not appear to 
+coincide with hypermutability and may reflect shifts in the actual 
+rate of recombination between genes.  
+Indeed, template switching is known to occur at hotspots, 
+which often involve transcription-regulatory sequences preceding genes 
+in SARS-CoV-2~\citep{Yang2021-characterizing}
+The rate of recombination may also depend on the relative abundance 
+of different subgenomic RNA intermediates that span different genes, 
+affecting the availability of templates and influencing the rate of 
+homologous recombination~\citep{Kim2020-gt,Zou2021-sars}  
+[Sally: KATHERINE, WHAT DO YOU THINK?  These not-full-length subgenomic RNA intermediates may provide a lot of homologous templates for some genes more than other (e.g., fig 3B here)]
 
@@
 (containing three XD samples, all sampled in France) has an earliest
-sample (EPI\_ISL~11222324, dated 2022-02-26), while the right hand
-subgraph has a single XD sample (EPI\_ISL~8514045, sampled in Turkey and dated 2021-12-30).
+sample (strain France/HDF-biopath-7747831001/2022) dated 2022-02-26, while the right hand
+subgraph has a single XD sample (strain Turkey/HSGM-F12594/2021) dated 2021-12-30.
 Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
@@
 Together these operations substantially increase the overall parsimony
-of the inferred ARGs, most noticably in terms of the number of
+of the inferred ARGs, most noticeably in terms of the number of
 reversion mutations inferred. These nodes are also being chosen by the
 LS HMM as likely choices of parent (as indicated by the node and edge
-counts quoted above), also demostrating that the hueristics are successfully
+counts quoted above), also demonstrating that the heuristics are successfully
 capturing features of real sequences.
@@
 ways to extend and improve. For example, the current parameterisation of using
-a single ``mismatch ratio'' is very simplistic (Section~\ref{sec:ls}), and likely
+a single ``mismatch ratio'' is very simplistic (Section~\ref{sec:ls}) and likely
 results in a flat likelihood space where many recombinant paths have equal
@@
 the rightmost value. Cases where we have more than two parents may either be
-the ``stacking'' of multiple recombination events, or instances where
+the ``stacking'' of multiple recombination events or instances where
 the HMM has chosen to switch to a third sequence rather than back to
@@
 of a pandemic-scale ARG may lead to significant improvements in our
-ability identify recombinants and to filter poor quality data.
+ability to identify recombinants and to filter poor quality data.
 
@@
 small values when creating new nodes using parsimony rules
-(Section~\ref{sec:parsimony-heuristics}). Since the ARGs here are
-very treelike, with recombination nodes consituting a tiny fraction
+(Section~\ref{sec:parsimony-heuristics}). Because the ARGs for SARS-CoV-2 are
+very treelike, with recombination nodes constituting a tiny fraction
 of the overall topology (Table~\ref{tab:args}), existing methods
@@
 \url{https://doi.org/10.55876/gis8.230329cd}.
-We removed sequences
+We~removed sequences
 if they had ambiguous collection dates, were collected before 2020-01-01
@@
 to identify them. One possibility is to use the LS HMM itself to flag
-overly divergent sequences, and to exclude them from attachment to the ARG.
+overly divergent sequences and to exclude them from attachment to the ARG.
 We might also estimate collection dates by adding these potential time travellers
@@
 marks) have unknown Pango lineage (A). The lineage for node N1 can be directly
-copied from its parent (L2). The lineage for node N2 must be inferred from
+copied from its parent (L2), which has an identical sequence. The lineage for node N2 must be inferred from
 that of the parent (L2) plus the lineage-defining mutation (red X)
@@
 the appropriate GISAID data access.
+
+The mapping of \texttt{tskit} IDs to strain and EPI\_ISL identifiers for the
+subgraph plots in Supplementary Figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, \ref{fig:pango_XD_gisaid_graph} and \ref{fig:pango_XB_gisaid_graph} are at
+\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/data/Pango_X_graph_sample_maps.txt}.
+
+The
 

### 2023-06-01 paper.tex
Finish up Sally's comments

@@
 to appear different between the parents when they are not
-(indeed, there is evidence that 28271
-
-The 7,572 mutations (including 5,782 insertions and 1,605 deletions)
-at this site are an indicator that this
-homopolymeric region may be prone to sequencing errors
-may…).
+(see the discussion of potential errors at site 28,271 below).
 In other cases, however, the drop in intervals does not appear to
@@
 Thus, breakpoint intervals will tend to be truncated at sites that are
-hypermutable, either due to being prone to sequencing errors or
-for biological reasons such as selection pressure.
+hypermutable, either due to increased information about parentage
+or spurious inferences caused by sequencing errors.
 It is therefore helpful to compare the number of intersecting intervals
@@
 \citep{Jungreis2021-dh}, suggesting that the high rate of recurrent mutation
-at this site is due to selection.
+at this site may be due to selection.
 

### 2023-06-02 paper.tex
typo
Ana's updates

@@
 \author[1]{Shing H. Zhan}
-\author[3$\star$]{Anastasia Ignatieva}
+\author[2$\star$]{Anastasia Ignatieva}
 \author[1$\star$]{Yan Wong}
-\author[2]{Katherine Eaton}
+\author[3]{Katherine Eaton}
 \author[1]{Benjamin Jeffery}
@@
 Information and Discovery, University of Oxford, Oxford, United Kingdom}
-\affil[2]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
-\affil[3]{Department of Statistics, University of Oxford, Oxford, United Kingdom}
+\affil[2]{Department of Statistics, University of Oxford, Oxford, United Kingdom}
+\affil[3]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
 \affil[4]{Department of Zoology and Biodiversity Research Centre,
@@
 associated with adaptation to new host species and with the production of more
-immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
+immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt,Tamura2023-ab}.
 Early on in the COVID-19 pandemic, the levels of genetic diversity
-were too low to precipitate the emergence of distinctive recombinant strains.
+were too low to enable the detection of distinctive recombinant strains.
 By late 2020, however, the appearance and
@@
 Delta which harboured multiple characteristic mutations,
-created the conditions required for recombination to occur,
-and recombinant strains with onward transmission became detectable at
-low frequencies~\citep{Jackson2021-ik}.
+created the conditions required to detect
+recombinant strains and their onward transmission~\citep{Jackson2021-ik}.
 More recently, the high prevalence of Omicron,
@@
 combine specific subsets of characteristic mutations from different lineages
-\citep[e.g.,][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz},
+\citep[e.g.,][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}
 and as a result can only identify inter-lineage recombination events.
@@
 RIPPLES finds candidate recombinants by using an existing phylogeny
-(built assuming no recombination), and finding potential recombinants
+(built assuming no recombination) and finding potential recombinants
 by  scanning for branches containing many mutations.
@@
 untenable; for example, more than half of the sequences sampled in February 2023
-are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
+are from the recombinant strain XBB and its descendants~\citep{Chen2022-pz}. This also means that
 future evolution of SARS-CoV-2 is likely to involve multiple sequential
@@
 
-It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
+It is well known that recombination distorts phylogenies~\citep{Schierup2000-fg}
 and affects the results of downstream analyses, such as inference of
@@
 thousands of human genomes and beyond
-\citep{Speidel2019-yh,Kelleher2019-ba,Schaefer2021-yg,Zhang2023-lf}, and which
-deal with levels of repeated recombination far in excess of those seen in viral
+\citep{Speidel2019-yh,Kelleher2019-ba,Schaefer2021-yg,Zhang2023-lf}, dealing with levels of recombination far in excess of those seen in viral
 phylogenies. The ``succinct tree sequence'' is an ARG data structure
@@
 tools. As illustrated in Figure~\ref{fig:overview_sc2ts}, inference is based on incrementally adding batches of sequences
-based on their collection dates, and proceeds in three phases.
+based on their collection dates and proceeds in three phases.
 % First, likely attachment nodes in the current ARG
@@
 Finally, we attach the trees for these sample-clusters to the current
-ARG, and apply some parsimony-based heuristics to address issues
+ARG and apply some parsimony-based heuristics to address issues
 introduced by the inherent greediness of this strategy
@@
 Using the current preliminary version of \texttt{sc2ts}, we infer two
-large ARGs (with 1,265,685 and 657,239 samples, respectively), and study
+large ARGs (with 1,265,685 and 657,239 samples, respectively) and study
 the properties of these ARGs to illustrate the power of the method
@@
 lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),
-with an unprecedented degree of precision in the
+with a high level of precision in the
 genomic location of recombination breakpoints
@@
 The goals of this preliminary study are to illustrate the utility of
-\texttt{sc2ts}, and to investigate the properties of the
+\texttt{sc2ts} and to investigate the properties of the
 inferred ARGs to inform subsequent development.
@@
 ``Wide'' and ``Long'' ARGs throughout.
-The Wide ARG is densely sampled but time-limited, and
+The Wide ARG is densely sampled but time-limited and
 includes all 1.27 million sample genomes collected up to June 30, 2021
@@
 For the Long ARG, we randomly sub-sample a maximum of 1,000 genomes per
-day (again restricting the delay between sampling and submission to 30 days),
+day (again restricting the delay between sampling and submission to 30 days)
 and include an additional year worth of samples (to June 30, 2022).
@@
     Mutations & & 1,231,193 &  & 1,062,072 & \\
-    & Per node & 0.83  & $\pm$1.40 & 1.36 & $\pm$1.72 \\
-    & Per sample & 0.77  & $\pm$1.39 & 1.38 & $\pm$1.77 \\
-    & Per site & 41.23  & $\pm$108.16    & 36.10 & $\pm$80.03\\
+    & Per node per genome & 0.83  & $\pm$1.40 & 1.36 & $\pm$1.72 \\
+    & Per sample per genome & 0.77  & $\pm$1.39 & 1.38 & $\pm$1.77 \\
+    & Per site per ARG & 41.23  & $\pm$108.16    & 36.10 & $\pm$80.03\\
     \cmidrule{2-6}
@@
 understanding of the relative importance of these processes to the overall
-topology. Similarly, edges are classified by the type of the parent node.
+topology (``UPGMA" indicates a node inferred to be an ancestor connecting a sample to the existing tree). Similarly, edges are classified by the type of the parent node.
 The mean and standard deviation ($\pm$)
@@
 in both ARGs involve inheriting directly from sampled genomes.
-Both ARGs contain 29,422 sites (Section \ref{sec:data_preprocessing}),
+Both ARGs contain 29,422 sites, due to the large number of sequences (Section \ref{sec:data_preprocessing}),
 and a large number of mutations.
-The average number of mutations per-site is high, although much
+The average number of mutations per-site is high, although some
 of this may be explained by outlier sites with artefactually high
@@
 is small. In the Wide ARG, for example, we have a mean of 0.77
-mutations per sample, demonstrating that new samples are generally
-well represented by it. See Section~\ref{sec:mutation_spectrum}
+mutations per sampled genome, demonstrating that most added samples fit into the ARG parsimoniously. See Section~\ref{sec:mutation_spectrum}
 for more analysis of the patterns of inferred mutations.
@@
 %% The ARG is mostly tree-like
-Compared to organisms like humans which recombine in every generation,
-in SARS-CoV-2 recombination is relatively rare, with recombination nodes accounting
+Compared to organisms like humans that recombine in every generation,
+recombination is relatively rare in SARS-CoV-2, with recombination nodes accounting
 for $<$0.3\% of the inferred ancestry (Table~\ref{tab:args}). As a result,
@@
 (downloaded on 2023-01-21).
-Phylogenies are pruned down to those samples present as tips in both datasets. Light grey
+Phylogenies are pruned down to those samples present as tips in both datasets, with the horizontal axis representing time (tips end at the sample collection date). Light grey
 lines match the corresponding samples between the two trees; black circles
@@
 dataset, and early in the pandemic tend to be restricted to
-a sample of strains which are not thought to be recombinants.
+a sample of strains that were not thought to be recombinants.
 For validation purposes, we compare our ARGs with
@@
 
-The \texttt{sc2ts} backbone phylogenies contain small amounts of recombination,
+The \texttt{sc2ts} backbone phylogenies for these Nextstrain subsamples contain small amounts of recombination,
 with 7 recombination nodes in the Wide ARG backbone (8 for the Long ARG
@@
 These local reticulations could either represent
-real but previously unobserved recombination events between closely related strains,
+real but previously unobserved recombination events between closely related strains
 or could be artefacts of the current \texttt{sc2ts} method and its parameterization.
@@
 We illustrate the backbone phylogeny by visualizing
-a single tree in the middle of the viral genome
-since other regions of the
+a single tree in the middle of the viral genome,
+although other regions of the
 genome show almost identical topologies.
@@
 The second difference between the \texttt{sc2ts} and Nextstrain trees
-we would like to highlight, are the non-identical sample partitions
-near the tips (e.g., within the Alpha clade). It is unclear what
-particular features of the \texttt{sc2ts} algorithm is driving these
+that we would like to highlight are a few non-identical sample partitions
+near the tips (e.g.\ cross-crossing assingments within the Alpha clade). It is unclear what
+particular differences between the phylogenetic reconstruction algorithms are driving these
 differences, and more study is required to characterise and address them.
@@
 \label{sec:jackson_recombs}
-RNA viruses are known to recombine at high rates
+RNA viruses are known to recombine at high rates when cells are co-infected
 \citep{Simon2011-rna}, and recombination has been widely
@@
 \citet{Jackson2021-ik}, who performed a careful analysis of sequences
-circulating in the UK in late 2020 to early 2021, and found
+circulating in the UK in late 2020 to early 2021 and found
 evidence of multiple independent recombination events and onward
@@
 origins (groups A to D and four singletons).
-These findings are closely replicated in both the Wide and Long ARGs.
+These findings are closely replicated in both the Wide and Long ARGs (Table \ref{tab:jackson}).
 
@@
 recombination events in the Wide ARG.
-The second column gives the number of sequences in the group.
+The second column gives the number of sequences in the group,
+limited to the samples considered by \citet{Jackson2021-ik}.
 The 3SEQ~\citep{boni2007exact} coordinates
@@
 add one to both left and right coordinates to correspond to the
-\texttt{tskit} definition of inheritance on either side of a breakpoint,
+\texttt{tskit} definition of inheritance on either side of a breakpoint
 and add one to the right coordinate to make the intervals
@@
 for a precise definition of sequence inheritance at recombination
-events, and the corresponding breakpoint intervals.
+events and the corresponding breakpoint intervals.
 Details for all 16 sequences are
@@
 (sample MILK-103C712 was removed during preprocessing; see
-Section~\ref{sec:data_preprocessing}), and the results are summarised
-in Tables~\ref{tab:jackson} and \ref{tab:jackson_supplement}.
+Section~\ref{sec:data_preprocessing}).
 Table~\ref{tab:jackson} shows the groups of sequences identified by Jackson
@@
 Table~\ref{tab:jackson_supplement} shows the details for each of the
-16 sequences individually, and showing generally a strong concordance
+16 sequences individually and showing generally a strong concordance
 in mosaic structure and parent lineages
@@
 using the \texttt{tskit} API, mostly running in a few seconds
-on standard laptop computers (see the Code Availability
+on standard laptop computers (see the Data Availability
 section for details of the corresponding Jupyter notebooks).
@@
 
-The
+[FIXME/TODO - include a reference to the GISAID EPI SET pdf]
 

### 2023-06-02 paper.tex
Batch up updates from Sally

@@
 to representing large SARS-CoV-2 genealogies,
-which currently encompasses over 15 million sequences in the GISAID database
-\citep{Shu2017-hp}.
+which currently encompasses over 15 million sequences in the GISAID 
+database~\citep{Shu2017-hp}.
 See Section~\ref{sec:args} for more details on ARGs and the succinct
@@
 based on their collection dates and proceeds in three phases.
-% First, likely attachment nodes in the current ARG
-% (or paths, if inferred to be a recombinant) are
 First, possible paths connecting each sample to the current ARG
 are inferred (allowing for recombination) using the
-% chosen for each of the samples in the daily batch using the
-Li and Stephens (LS) model (Figure~\ref{fig:overview_sc2ts}A, B; Section~\ref{sec:ls}). Then, since many samples
+Li and Stephens (LS) 
+model (Figure~\ref{fig:overview_sc2ts}A, B; Section~\ref{sec:ls}). 
+Then, since many samples
 in the batch can share an attachment path,
-we infer phylogenetic trees for each of the clusters separately using standard
-methods (Figure~\ref{fig:overview_sc2ts}C; Section~\ref{sec:sample-cluster-tree-inference}).
+we infer phylogenetic trees for each of the clusters separately using standard methods 
+(Figure~\ref{fig:overview_sc2ts}C; Section~\ref{sec:sample-cluster-tree-inference}).
 Finally, we attach the trees for these sample-clusters to the current
@@
 (see Section~\ref{sec:filtering_time_travellers}).
-% The Long ARG is more sparsely sampled but
-% includes data for an additional year (June 30, 2022).
-% For the Long ARG, we use at most 1,000 randomly subsampled
-% genomes per day from those meeting the maximum sampling delay criterion.
 For the Long ARG, we randomly sub-sample a maximum of 1,000 genomes per
@@
 The properties of the inferred ARGs are summarised in Table~\ref{tab:args}.
-% TODO more discussion here, making sure we explain anything that might not be
-% obvious.
 The majority of the nodes in the ARGs represent sample genomes
@@
 is small. In the Wide ARG, for example, we have a mean of 0.77
-mutations per sampled genome, demonstrating that most added samples fit into the ARG parsimoniously. See Section~\ref{sec:mutation_spectrum}
+mutations per sampled genome, demonstrating that most 
+added samples fit into the ARG parsimoniously. 
+See Section~\ref{sec:mutation_spectrum}
 for more analysis of the patterns of inferred mutations.
@@
 that we would like to highlight are a few non-identical sample partitions
-near the tips (e.g.\ cross-crossing assingments within the Alpha clade). It is unclear what
+near the tips (e.g.\ criss-crossing assignments within the Alpha clade). It is unclear what
 particular differences between the phylogenetic reconstruction algorithms are driving these
@@
 origins (groups A to D and four singletons).
-These findings are closely replicated in both the Wide and Long ARGs (Table \ref{tab:jackson}).
+These findings are closely replicated in both the Wide and Long ARGs.
 
@@
 The stacked histogram shows the distribution of parental divergence times, ranging
-from parents that have diverged only a handful of days ago, to much more
+from parents that have diverged only a few days ago, to much more
 divergent parent lineages. See Figure~\ref{fig:recomb_mrcas_voc_breakdown}
@@
 These shared MRCAs lie near the root of large expansions,
-% Anyone reading this will know what these are
-% (B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
-% and BA.2: second Omicron wave)
-% as demonstrated by the number of samples with those
-% Pango designations traback to each ancestral node.
-% The majority of these frequently-shared MRCAs
 and the majority are associated with large polytomies, likely
@@
 \includegraphics[width=\textwidth]{figures/Pango_XA_XAG_XD_nxcld_tight_graph.pdf}
-% This figure is gorgeous!!
-
 \caption{\label{fig:pango-simple-origin-graph} Examples of non-nested
@@
 sampled viral genome; dotted lines show ARG continuation. Vertical position of nodes does not
-show absolute time, but simply relative rank (parents above children). Nodes are coloured by Nextclade
+show absolute time, but relative rank (parents above children). Nodes are coloured by Nextclade
 Pango designation; smaller symbols are non-sample nodes inserted by \texttt{sc2ts}, whose
@@
 Sample nodes are plotted as larger circles, but the subgraph also includes intermediate,
-non-sample nodes (i.e. inserted by sc2ts, see Sections~\ref{sec:sample-cluster-tree-inference} and \ref{sec:parsimony-heuristics}). Dotted lines show where this subgraph links
+non-sample nodes (i.e., inserted by \texttt{sc2ts}, see Sections~\ref{sec:sample-cluster-tree-inference} and \ref{sec:parsimony-heuristics}). Dotted lines show where this subgraph links
 to the rest of the ARG. Above recombination nodes, only ancestral nodes are shown, meaning
@@
 (and a leftmost of 25470). By contrast, the Omicron insertion in the Turkish sample
-is inferred to have occurred from position 21619-22578 to position 23605-24130.
+is inferred to have occurred from position 21619--22578 to position 23605--24130.
 The breakpoint difference, the different geographical locations, the time between the
@@
 in practice. Such an approach would also reduce the impact of sequences
-with high levels of sequencing error (which contribute a large number of
-mutations, currently).
+with high levels of sequencing error (which currently contribute a large number of
+mutations).
 Taken together, these and other relatively minor improvements should
@@
 right-most extent of the possible interval
-(Section~\ref{sec:breakpoint_intervals}). More likely locations for the
+(Section~\ref{sec:breakpoint_intervals}). 
+More likely locations for the
 breakpoint could be chosen within the interval, for example based
-on sequence motifs \citep{Gallaher2020-mr, Yang2020-ct}.
-% Citation: https://doi.org/10.1007/s00705-020-04750-z . This is the data source we used to create the breakpoints motif look up table.
+on sequence motifs~\citep{Gallaher2020-mr, Yang2020-ct}.
 % Hint: there's probably a bunch of other stuff you could do with this
@@
 will mostly consist of a new sample sequence copying from a node in the ARG
-with a small number of mutations and often many samples
+with a small number of mutations, and often many samples
 from a daily batch will copy from the same ARG node.
@@
 as implemented in SciPy~\citep{Pauli2020-scipy} to build trees from sample
-clusters, and map mutations back to this topology using maximum parsimony.
-This approach was chosen mainly for simplicity and because of the
-speed and reliability of the implementation.
+clusters, and then map mutations back to this topology using maximum parsimony.
+We chose this approach mainly for simplicity, and because of the
+speed and reliability of the SciPy implementation.
 An issue with the UPGMA algorithm is that it generates a strictly
@@
 this within-cluster tree building has a significant influence on the
-overall ARG topology (Wide ARG: 7.05\% of nodes and parent of
+overall ARG topology (Wide ARG: 7.07\% of nodes and parent of
 32.27\% of edges; Long ARG: 6.61\% and 19.97\%, respectively),
@@
 is room for improvement and that further parsimony-increasing
-heuristics would likely be of benefit.
+heuristics 
+(e.g., resolving reversions beyond those on immediately adjacent edges)
+would likely be of benefit .
 
@@
 that $b_\ell \leq x < b_r$.
-
-% Thus, the breakpoints used in the ARG represent the rightmost end of the
-% interval, within which the recombination event could have occurred. To obtain
-% the leftmost end of this interval, we rerun the HMM using mirrored coordinates
-% (i.e., from the 3’ end to the 5’ end) and store the information as
-% metadata on the recombination node.
-% These left- and rightmost positions
-% are the recombination intervals reported throughout.
-% Ultimately all the recombination intervals mean is that the parent sequences
-% are identical within that interval, and there is therefore no information
-% about where the recombination occurred. There are many different processes
-% that can contribute to the size of these intervals, and therefore
-% understanding how these processes interact to produce wider or narrower
-% intervals is an important avenue for future work.
 
@@
 
-Although the masking strategy used currently is simple and robust,
-there are significant disadvantages, excluding, for example, any
+Although the current masking strategy is simple and robust,
+there are significant disadvantages because it excludes, for example, any
 deletions of length greater than one base. Exploring more sophisticated
@@
 \section{Data Availability}
+\label{sec-data-availability}
 The source code for \texttt{sc2ts} and notebooks and code used to produce the
@@
 \url{https://doi.org/10.55876/gis8.230329cd} and included in the
-Supplemental Table.
+Supplemental Table (GISAID EPI SET PDF).
 
@@
 \url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/data/Pango_X_graph_sample_maps.txt}.
-
-[FIXME/TODO - include a reference to the GISAID EPI SET pdf]
 
@@
 
-% \begin{table}
-% \small
-% \centering
-% \begin{tabular}{lp{0.2cm}p{0.8cm}lp{0.9cm}ll}
-% \toprule
-% Date & $n$ & Causal Pango & Causal strain & Causal \#muts & Sc2ts parents & Breakpoint \\
-% \midrule
-% 2022-03-24 & 2 & XN & Germany/BY-RKI-I-674341/2022 & 3 & Unknown + BA.2 & (2833,9867) \\
-% 2022-03-29 & 2 & XN & Spain/AB-HGUA-26086757/2022 & 2 & BA.1.1 + BA.2.9 & (2833,4185) \\
-% 2022-04-23 & 7 & XN & Spain/MD-HRYC-12296321/2022 & 2 & BA.1 + BA.2 & (4185,4322) \\
-% 2022-05-17 & 2 & XN & USA/TX-HMH-MCoV-97903/2022 & 5 & BA.2 + BA.2 & (22776,22787) \\
-% \bottomrule
-% \end{tabular}
-% \caption{\label{tab:multiple_origins_table}
-% Summary of inferred multiple ``origins'' for Nextclade-designated XN samples after filtering
-% (Official: BA.1* \& BA.2*, breakpoint between 2834 \& 4183, defined from UK strains, earliest
-% on 2022-02-01;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/480}).
-% The 13 XN samples are grouped into rows based on their most recent ancestral recombination node;
-% all the descendant samples of the four recombination nodes are designated XN by Nextclade.
-% Details are also provided for the ``causal strain'' that led to the recombination
-% being proposed by \texttt{sc2ts}. When undersampling causes three or more mutations to lie
-% immediately above the causal strain, \texttt{sc2ts} may mistakenly propose multiple
-% independent origins rather than one. Note that a large number of XN samples were previously
-% removed by filtering.
-% }
-% \end{table}
-
 \begin{figure} \centering

### 2023-06-02 paper.tex
Remove spaces

@@
 \begin{abstract}
-Recombination is an ongoing and increasingly important feature of circulating lineages of SARS-CoV-2, challenging how we represent the evolutionary history of this virus and giving rise to new variants of potential public health concern by combining transmission and immune evasion properties of different lineages.
-Detection of new recombinant strains is challenging, with most methods looking for breaks between sets of mutations that characterize distinct lineages.
+Recombination is an ongoing and increasingly important feature of circulating
+lineages of SARS-CoV-2, challenging how we represent the evolutionary history
+of this virus and giving rise to new variants of potential public health
+concern by combining transmission and immune evasion properties of different
+lineages. Detection of new recombinant strains is challenging, with most
+methods looking for breaks between sets of mutations that characterize distinct
+lineages.
 In addition, many basic approaches fundamental to the study of viral
@@
 circulating globally in the last three months (January to March, 2023;
-https://gisaid.org/; access on March 27, 2023). High prevalence increases the risk of coinfection, providing opportunities for new phenotypically
+\url{https://gisaid.org/}; accessed on 2023-03-27).
+High prevalence increases the risk of coinfection,
+providing opportunities for new phenotypically
 distinct recombinants to emerge and spread. Phylogenetic approaches have
-been central to pandemic responses thus far \citep{Attwood2022-ab}.
-% 1. Form the basis of variant classifications (UShER paper),
-% 2.
-Global context.
-In 2020, Nextstrain provided situation reports that were interactive, phylogenetically-driven visualizations of SARS-CoV-2 characteristics such as transmission https://nextstrain.org/ncov-sit-reps/
-% 3.
-The WHO's recommendations on Omicron asked countries to "report initial cases/clusters associated with VOC infection to WHO".
-Phylogenetics being one way to identify those crucial early clusters.
-https://www.who.int/news/item/26-11-2021-classification-of-omicron-(b.1.1.529)-sars-cov-2-variant-of-concern
-However, with the rise to high frequency of recombinant lineages \citep[e.g. XBB; ][]{Tamura2023-ab},
+been central to pandemic responses thus
+far~\citep{Attwood2022-ab,Bloom2023-fitness,Abbas2022-reconstruction,
+Mclaughlin2022-genomic}.
+However, with the rise to high frequency of recombinant
+lineages \citep[e.g.\ XBB;][]{Tamura2023-ab},
 it is imperative that these methods are updated to incorporate the

### 2023-06-02 paper.tex
Switch to abbrvnat

@@
 Detecting recombination in SARS-CoV-2 is difficult, and identifying new
-recombinant strains is a time-consuming, manual process, subject
-to biases [FIXME: have we a citation for how tedious this is? Also what
-are the biases?]
+recombinant strains is a time-consuming, manual
+process~\citep{Smith2023-identifying}.
+% (e.g., \url{https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/}).
 % This resource shows the typical workflow for manually identifying recombinants that the pango team uses: https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/
@@
 occur on terminal branches. Insertions are mutations in which the
-inherited state is the gap state ``-'' and the derived state is a
-nucleotide, and vice versa for deletions. Recurrent mutations are
-those that inherit from at least one other mutation (at that site),
+ancestral state is the gap state ``-'' and the derived state is a
+nucleotide, and vice versa for deletions. A recurrent mutation is
+one that occurs at a site which has previously mutated (i.e. where one
+or more mutations at that site occur on the edges ancestral to the mutation),
 and a reversion is a recurrent mutation that reverses the state change
@@
 (Wide ARG: 0.77, Long ARG: 1.38; Table~\ref{tab:args}), the average number per site
-is large (Wide ARG: 41.23, Long ARG: 36.10; Table~\ref{tab:args}). This is
+is large across each ARG (Wide ARG mean: 41.23, median: XXX;  Long ARG mean: 36.10, median: YYY; Table~\ref{tab:args}). This is
 partly driven by some highly mutagenic sites (e.g., site 28,271 has over 7,000
@@
 In addition, we exclude 481 problematic sites flagged as prone to
-sequencing errors or as highly homoplasic
+sequencing errors or as highly homoplasic entirely
 (\url{https://github.com/W-L/ProblematicSites_SARS-CoV2/},
-accessed 2022-09-22) entirely.
+accessed 2022-09-22).
 
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{abbrvnat}
 \bibliography{paper}

### 2023-06-02 paper.tex
Define LS HMM earlier

@@
 immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt,Tamura2023-ab}.
-Early on in the COVID-19 pandemic, the levels of genetic diversity
+Early in the COVID-19 pandemic, the levels of genetic diversity
 were too low to enable the detection of distinctive recombinant strains.
@@
 % Recombination is hard to detect
-Detecting recombination in SARS-CoV-2 is difficult, and identifying new
+Detecting recombination in SARS-CoV-2 is difficult and identifying new
 recombinant strains is a time-consuming, manual
 process~\citep{Smith2023-identifying}.
-% (e.g., \url{https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/}).
-% This resource shows the typical workflow for manually identifying recombinants that the pango team uses: https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/
-% Regarding bias, not peer-reviewed but in the official pango designation post for XAC, you can see that me (Katherine) and the authors had a multi-month long discussion on how many breakpoints there should be and where.
-Reproducibility was difficult: https://github.com/cov-lineages/pango-designation/issues/590
 Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
@@
 A schematic of the \texttt{sc2ts} method.
-Sc2ts reconstructs the genetic relationships
-among SARS-CoV-2 genomes by copying samples to all possible ancestors collected
-at earlier time points (curved arrows). Each daily iteration involves three stages:
-(1) attachment of new samples to the growing ARG (A, B); (2) reconstruction of
-trees relating the samples under each attachment node (C); and parsimony-based
-tree topology adjustments (D, E). In the absence of  recombination, sc2ts
-infers an ARG that is a single tree relating the samples (A). However, in the
-presence of recombination, sc2ts infers an ARG that is encoded as a sequence of
-local trees relating segments of the sample genomes (B), inserting a
-recombination node to represent a putative recombination event. Additionally,
-mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to
+The genetic relationships among SARS-CoV-2 genomes is
+reconstructed by using the Li and Stephens
+model to infer attachment paths for samples to an existing ARG (curved lines).
+Each daily iteration involves three stages:
+attachment of new samples to the growing ARG (A, B);
+reconstruction of trees relating the samples under each attachment node (C);
+and parsimony-based tree topology adjustments (D, E).
+In the absence of  recombination, \texttt{sc2ts}
+infers an ARG that is a single tree relating the samples (A).
+When recombination is detected, \texttt{sc2ts} infers
+an ARG that concisely encodes a sequence of local trees relating segments
+of the sample genomes (B). Additionally,
+mutation-collapsing nodes (D) and reversion-push nodes (E) are inserted to
 make more parsimonious placements of mutations that should be shared or should
@@
 are inferred (allowing for recombination) using the
-Li and Stephens (LS)
-model (Figure~\ref{fig:overview_sc2ts}A, B; Section~\ref{sec:ls}).
+Li and Stephens (LS) model (Figure~\ref{fig:overview_sc2ts}A, B);
+the LS ``copying processs'' is a Hidden Markov Model (HMM)
+approximating the effects of mutation
+and recombination, widely used in large-scale genomics (Section~\ref{sec:ls}).
 Then, since many samples

### 2023-06-02 paper.tex
edits from yan

--- 2023-06-02 paper.tex
@@
 Detecting recombination in SARS-CoV-2 is difficult, and identifying new
-recombinant strains is a time-consuming, manual
-process~\citep{Smith2023-identifying}.
-% (e.g., \url{https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/}).
+recombinant strains is a time-consuming, manual process, subject
+to biases [FIXME: have we a citation for how tedious this is? Also what
+are the biases?]
 % This resource shows the typical workflow for manually identifying recombinants that the pango team uses: https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/
@@
 In addition, we exclude 481 problematic sites flagged as prone to
-sequencing errors or as highly homoplasic entirely
+sequencing errors or as highly homoplasic
 (\url{https://github.com/W-L/ProblematicSites_SARS-CoV2/},
-accessed 2022-09-22).
+accessed 2022-09-22) entirely.
 
@@
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{plainnat}
 \bibliography{paper}

### 2023-06-02 paper.tex
Ana's updates

--- 2023-06-02 paper.tex
@@
 occur on terminal branches. Insertions are mutations in which the
-ancestral state is the gap state ``-'' and the derived state is a
+inherited state is the gap state ``-'' and the derived state is a
 nucleotide, and vice versa for deletions. A recurrent mutation is
@@
 Designation of samples to Pango lineages is not a straightforward task,
-and there can be significant variation between methods.
+and there can be significant variation between methods \citep{deBernardiSchneider2023-sars}.
 Here, we consider two different assignments of Pango lineages to samples in

### 2023-06-02 paper.tex
Define mutation classifications

@@
 Li and Stephens (LS) model (Figure~\ref{fig:overview_sc2ts}A, B);
-the LS ``copying processs'' is a Hidden Markov Model (HMM)
+the LS ``copying process'' is a Hidden Markov Model (HMM)
 approximating the effects of mutation
@@
 (Figure~\ref{fig:overview_sc2ts}C; Section~\ref{sec:sample-cluster-tree-inference}).
-Finally, we attach the trees for these sample-clusters to the current
+Finally, we attach the trees for these sample clusters to the current
 ARG and apply some parsimony-based heuristics to address issues
@@
 day (again restricting the delay between sampling and submission to 30 days)
-and include an additional year worth of samples (to June 30, 2022).
+and include an additional year's worth of samples (to June 30, 2022).
 
@@
 and a large number of mutations.
-The average number of mutations per-site is high, although some
+The average number of mutations per site is high, although some
 of this may be explained by outlier sites with artefactually high
@@
 distinct recombinants to emerge and spread. Phylogenetic approaches have
-been central to pandemic responses thus
+been central to responses to the pandemic thus
 far~\citep{Attwood2022-ab,Bloom2023-fitness,Abbas2022-reconstruction,
@@
 later analysis (Section~\ref{sec:treatment_recombinants}).
-The results of inference for a given day is then
+The result of inference for a given day is then
 a genealogy recording genetic inheritance as well as
@@
 from left coordinate $\ell$ to right coordinate $r$. A \emph{site}
-defines the genomic position at which some variation occurs,
-and a \emph{mutation} defines the site, node and derived state
-that generates variation.
-% JK: Note sure this clarifies much at the representational level?
-% (note that this requires that the sequences can be aligned and phased).
-
-Given these tables, we can then efficiently construct the local
+defines a position on the genome
+and the ancestral state (allele) at that site.
+A \emph{mutation} records the site and node IDs where a mutation
+occurs and the derived state (allele).
+In addition to these basic elements of the data model, the \texttt{tskit} library
+supports additional tables and fields,
+the ability to associate arbitrary metadata with
+table rows, and facilities to record provenance
+information~\citep{Tskit2023-tskit}.
+
+Tskit supports arbitrarily complex patterns of mutation at a particular
+site, and it is useful
+to define some terminology to classify them.
+A mutation's ``parent''
+is the first mutation encountered (at that site) on the path to
+root from the mutation's node in the local tree corresponding to
+the site's position. If no other mutation is encountered on the
+path to root, the mutation's parent is null.
+The ``derived state'' of a mutation is the allelic state inherited by nodes
+in the subtree rooted at the mutation's node (in the local tree), assuming
+there are no subsequent descendant mutations.
+The ``inherited state'' of a mutation is the derived state of the
+parent mutation, if it exists, or the site's ancestral state otherwise.
+A ``recurrent mutation`` is a mutation with a non-null parent,
+and a ``reversion`` is a recurrent mutation that reverses the state
+change of its parent. For example, if we have two mutations
+$a$ and $b$, such that $a$ is the parent of $b$,
+with state transitions (inherited state $\rightarrow$ derived state)
+$a: A\rightarrow T$ and $b: T\rightarrow A$, we define $b$ as a reversion.
+
+Given the node, edge, site and mutation tables we can
+efficiently construct the local
 genealogical trees along the genome (arising from recombination)
-and also perform a range of calculations efficiently by
+and perform a range of calculations efficiently by
 reasoning about the differences between these local
@@
 
-The probabilities of these competing process of mismatch and recombination are
+The probabilities of these competing processes of mismatch and recombination are
 usually controlled by the site-specific parameters $\mu_\ell$ and $r_\ell$,
@@
 rather to consider such sites as uninformative to the path probability, and
-impute missing data within that path on the fly, placing them within the
-the emission probability transition from unknown to
+impute missing data within that path on the fly, placing them within the emission probability transition from unknown to
 the emission probability from unknown to
@@
 through the collection of paths at the global optimum to glean further
-information about our the likelihood surface, and determine whether there are
+information about the likelihood surface, and determine whether there are
 downstream implications for our conclusions. Here, we have considered the
@@
 for multiple samples in a daily batch, and so we
-create a ``recombination`` node (marked with a specific ``flags'' value)
+create a ``recombination'' node (marked with a specific ``flags'' value)
 for each distinct recombinant. (This node is not strictly necessary
@@
 \label{sec:node_dating}
-The approaches to assigning a date to nodes in \texttt{sc2ts} is currently
+The approach to assigning a date to nodes in \texttt{sc2ts} is currently
 ad-hoc, and the inferred timing of events from the ARGs reported
@@
 
-\bibliographystyle{abbrvnat}
+% \bibliographystyle{abbrvnat}
+\bibliographystyle{stylish}
 \bibliography{paper}

### 2023-06-02 paper.tex
Add medians and refer to precise mutation defs

@@
     \cmidrule{2-5}
-Insertions  & 6191  & (0.51\%) & 2814   & (0.26\%) \\
-Deletions   & 7462  & (0.62\%) & 5152   & (0.49\%) \\
+Insertions  & 6,191  & (0.51\%) & 2,814   & (0.26\%) \\
+Deletions   & 7,462  & (0.62\%) & 5,152   & (0.49\%) \\
     \cmidrule{2-5}
@@
 inherited state is the gap state ``-'' and the derived state is a
-nucleotide, and vice versa for deletions. A recurrent mutation is
-one that occurs at a site which has previously mutated (i.e. where one
-or more mutations at that site occur on the edges ancestral to the mutation),
-and a reversion is a recurrent mutation that reverses the state change
-of the mutation immediately ancestral to it.
+nucleotide, and vice versa for deletions. See Section~\ref{sec:args}
+for precise definitions of mutations, and the recurrent and reversion
+classifications.
 (B) Mutational spectrum in
@@
 of each mutation type from the Wide ARG are represented by blue bars and the
-percentages from Yi et al. by orange bars, with the darker colours representing
+percentages from Yi et al.\ by orange bars, with the darker colours representing
 one direction (e.g., C$>$U) and the lighter colours the reverse (e.g., U$>$C).}
@@
 of mutations per sample is small
-(Wide ARG: 0.77, Long ARG: 1.38; Table~\ref{tab:args}), the average number per site
-is large across each ARG (Wide ARG mean: 41.23, median: XXX;  Long ARG mean: 36.10, median: YYY; Table~\ref{tab:args}). This is
-partly driven by some highly mutagenic sites (e.g., site 28,271 has over 7,000
+(Wide ARG: 0.77, Long ARG: 1.38; Table~\ref{tab:args}),
+the average number per site
+is large in both ARGs (Wide ARG: 41.23; Long ARG:36.10; Table~\ref{tab:args}).
+However, a lower median count (Wide ARG: 14; Long ARG:13),
+suggests that these high mutation counts are
+partly driven by some hypermutable sites (e.g., site 28,271 has over 7,000
 mutations in the Wide ARG, Figure~\ref{fig:breakpoint-distribution}), which
@@
 ARG (Long ARG: 4.54\%) reverting the state change of the immediately
-ancestral mutation. The majority of these mutations are likely to be
+ancestral mutation (see Section~\ref{sec:args} for precise definitions).
+The majority of these mutations are likely to be
 artefactual, and a key goal of future development is to improve

### 2023-06-02 paper.tex
Edits from Ben and Ana

--- 2023-06-02 paper.tex
@@
 from left coordinate $\ell$ to right coordinate $r$. A \emph{site}
-defines a position on the genome
-and the ancestral state (allele) at that site.
-A \emph{mutation} records the site and node IDs where a mutation
-occurs and the derived state (allele).
-In addition to these basic elements of the data model, the \texttt{tskit} library
-supports additional tables and fields,
-the ability to associate arbitrary metadata with
-table rows, and facilities to record provenance
-information~\citep{Tskit2023-tskit}.
-
-Tskit supports arbitrarily complex patterns of mutation at a particular
-site, and it is useful
-to define some terminology to classify them.
-A mutation's ``parent''
-is the first mutation encountered (at that site) on the path to
-root from the mutation's node in the local tree corresponding to
-the site's position. If no other mutation is encountered on the
-path to root, the mutation's parent is null.
-The ``derived state'' of a mutation is the allelic state inherited by nodes
-in the subtree rooted at the mutation's node (in the local tree), assuming
-there are no subsequent descendant mutations.
-The ``inherited state'' of a mutation is the derived state of the
-parent mutation, if it exists, or the site's ancestral state otherwise.
-A ``recurrent mutation`` is a mutation with a non-null parent,
-and a ``reversion`` is a recurrent mutation that reverses the state
-change of its parent. For example, if we have two mutations
-$a$ and $b$, such that $a$ is the parent of $b$,
-with state transitions (inherited state $\rightarrow$ derived state)
-$a: A\rightarrow T$ and $b: T\rightarrow A$, we define $b$ as a reversion.
-
-Given the node, edge, site and mutation tables we can
-efficiently construct the local
+defines the genomic position at which some variation occurs,
+and a \emph{mutation} defines the site, node and derived state
+that generates variation.
+% JK: Note sure this clarifies much at the representational level?
+% (note that this requires that the sequences can be aligned and phased).
+
+Given these tables, we can then efficiently construct the local
 genealogical trees along the genome (arising from recombination)
-and perform a range of calculations efficiently by
+and also perform a range of calculations efficiently by
 reasoning about the differences between these local

### 2023-06-02 paper.tex
Tweaks

--- 2023-06-02 paper.tex
@@
 (Wide ARG: 87\%, Long ARG: 84\%), with the remainder mostly representing
-the ancestral sequences inferred by UPGMA
+the ancestral sequences inferred 
 from daily sample clusters (Section~\ref{sec:sample-cluster-tree-inference})
@@
 add one to both left and right coordinates to correspond to the
-\texttt{tskit} definition of inheritance on either side of a breakpoint
+\texttt{tskit} definition of inheritance on either side of a breakpoint,
 and add one to the right coordinate to make the intervals

### 2023-06-02 paper.tex
Remove unhelpful UPGMAs and edge counts
Closes #165

@@
      & \emph{Node type} & \\
-    & Samples & 1,265,685 & (87.09\%) & 657,239 & (83.91\%) \\
-    & UPGMA   & 102,709 & (7.07\%) & 51,807 & (6.61\%)  \\
+    & Sample & 1,265,685 & (87.09\%) & 657,239 & (83.91\%) \\
+    & Daily sample cluster tree  & 102,709 & (7.07\%) & 51,807 & (6.61\%)  \\
     & Reversion push & 40,538 & (2.79\%) & 34,358 & (4.39\%)  \\
@@
     \cmidrule{2-6}
-    Edges & & 1,458,146 &  & 785,539 & \\
-     & \emph{Parent node type} & \\
-    & Samples & 610,729 & (41.88\%) & 319,626 & (40.69\%) \\
-    & UPGMA   & 470,545 & (32.27\%) & 156,881 & (19.97\%)  \\
-    & Reversion push & 184,608 & (12.66\%) & 144,991 & (18.46\%)  \\
-    & Mutation collapse &186,218  & (12.77\%) & 160,833 & (20.47\%)  \\
-    & Recombination & 6,046 & (0.41\%) & 3,208 & (0.41\%) \\
-    \cmidrule{2-6}
+    % Edges & & 1,458,146 &  & 785,539 & \\
+    %  & \emph{Parent node type} & \\
+    % & Samples & 610,729 & (41.88\%) & 319,626 & (40.69\%) \\
+    % & UPGMA   & 470,545 & (32.27\%) & 156,881 & (19.97\%)  \\
+    % & Reversion push & 184,608 & (12.66\%) & 144,991 & (18.46\%)  \\
+    % & Mutation collapse &186,218  & (12.77\%) & 160,833 & (20.47\%)  \\
+    % & Recombination & 6,046 & (0.41\%) & 3,208 & (0.41\%) \\
+    % \cmidrule{2-6}
     Mutations & & 1,231,193 &  & 1,062,072 & \\
@@
 \caption{\label{tab:args}Summary of the inferred ARGs. Nodes are classified
-as either samples or by the inference process which produced them to aid
-understanding of the relative importance of these processes to the overall
-topology (``UPGMA" indicates a node inferred to be an ancestor connecting a sample to the existing tree). Similarly, edges are classified by the type of the parent node.
+as either samples or by the inference process that produced them
+(see Sections~\ref{sec:sample-cluster-tree-inference},
+\ref{sec:parsimony-heuristics} and \ref{sec:treatment_recombinants} for
+details).
 The mean and standard deviation ($\pm$)
@@
 and parsimony heuristics (Section~\ref{sec:parsimony-heuristics}).
-Table~\ref{tab:args} shows the counts of edges (see
-Section~\ref{sec:args} for a technical definition) in the ARGs
-categorised by parent node type, and demonstrates that each
-of these contributes significantly to the overall topology.
-In particular, we can see that approximately 40\% of the edges
-in both ARGs involve inheriting directly from sampled genomes.
-Both ARGs contain 29,422 sites, due to the large number of sequences (Section \ref{sec:data_preprocessing}),
+Both ARGs contain 29,422 sites (Section \ref{sec:data_preprocessing}),
 and a large number of mutations.
@@
 \label{sec:sample-cluster-tree-inference}
-For a given daily batch of sample sequences that are to be attached to the
-ARG, the LS HMM provides the most likely copying path and required mutations
-for each of those sequences. We could directly attach these samples to
-the ARG according to the copying paths and sets of mutations, but
-such an approach would be highly unparsimonious because there are
-often many samples that share the same copying path, with
-complex patterns of shared mutations among those clusters of samples.
-In this section we discuss the strategy used to infer likely tree
-topologies within these clusters of samples sharing the same copying path.
-In the majority of cases the copying path will consist of a single
-node (i.e., no recombination), and so for simplicity we will assume in the
-following discussion that all of the samples in a HMM-defined cluster
-attach to a single node. The special treatment afforded to
-newly-detected putative recombinants are discussed
-in Section~\ref{sec:treatment_recombinants}.
-
 With tens of thousands of samples being added to the ARG per day,
-there are often clusters of hundreds of sequences attaching to the same node.
+there are often clusters of hundreds of sequences attaching to the same node
+(or more generally, recombinant path;
+see Section~\ref{sec:treatment_recombinants}).
 While some of these samples will require no extra mutations
@@
 samples in a cluster independently of the other samples in a
-daily batch, and then attach the tree (and mutations)
+daily batch and then attach the tree (and mutations)
 to the ARG at the node identified by the HMM.
@@
 this within-cluster tree building has a significant influence on the
-overall ARG topology (Wide ARG: 7.07\% of nodes and parent of
-32.27\% of edges; Long ARG: 6.61\% and 19.97\%, respectively),
+overall ARG topology
 and therefore applying more sophisticated
@@
 % \bibliographystyle{abbrvnat}
-\bibliographystyle{stylish}
+\bibliographystyle{refstyle}
 \bibliography{paper}
@@
 \setcounter{table}{0}
-\section*{Supplementary}
+\section*{Supplementary Material}
 

### 2023-06-02 paper.tex
Anas updates
Updates from Overleaf
Tidy up the mutational spectrum plot
Switching A and B so that the Y axis label doesn't get confused with the table text.

This does, however, mean that Fig 3B is mentioned before 3A. Does that matter?

@@
 \newcolumntype{Y}{>{\centering\arraybackslash}X}
-\begin{tabularx}{\textwidth}{Yc}
-
+\begin{tabularx}{\textwidth}{cY}
+\hspace*{-2.5mm}\includegraphics[width=0.32\textwidth]{figures/mutational_spectra.pdf}
+&
 \begin{tabular}[b]{llrlr}\toprule
@@
 Recurrent   & 74,719 & (6.16\%) & 50,099 & (4.72\%) \\
-Reversions  & 72,617 & (5.99\%) & 48,226 & (4.54\%) \\
+Of which reversions  & 72,617 & (5.99\%) & 48,226 & (4.54\%) \\
 \bottomrule
-\end{tabular}
-&
-\includegraphics[width=0.35\textwidth]{figures/mutational_spectra.pdf}\\
+\vspace{-2mm}
+\end{tabular}\\
 (A) & (B)\\
@@
 \caption{\label{fig:mutational_spectra}
-(A) Summary of mutations in the Long and Wide ARGs. Private mutations
+(A) Mutational spectrum in
+the Wide ARG compared to \cite{Yi2021-sc}. Mutations are
+categorised by type (i.e., inherited state $>$ derived state). The percentages
+of each mutation type from the Wide ARG are represented by blue bars and the
+percentages from Yi et al.\ by orange bars, with the darker colours representing
+one direction (e.g., C$>$U) and the lighter colours the reverse (e.g., U$>$C).
+(B) Summary of mutations in the Long and Wide ARGs. Private mutations
 occur on terminal branches. Insertions are mutations in which the
@@
 classifications.
-(B) Mutational spectrum in
-the Wide ARG compared to \cite{Yi2021-sc}. Mutations are
-categorised by type (i.e., inherited state $>$ derived state). The percentages
-of each mutation type from the Wide ARG are represented by blue bars and the
-percentages from Yi et al.\ by orange bars, with the darker colours representing
-one direction (e.g., C$>$U) and the lighter colours the reverse (e.g., U$>$C).}
+}
 \end{figure}
@@
 The properties of the mutations inferred in the Wide and Long
-ARGs are summarised in Figure~\ref{fig:mutational_spectra}A. In both
+ARGs are summarised in Figure~\ref{fig:mutational_spectra}B. In both
 cases we have a large number of mutations, and a majority of these
@@
 established results.
-In Figure~\ref{fig:mutational_spectra}B we compare the mutational spectrum
+In Figure~\ref{fig:mutational_spectra}A we compare the mutational spectrum
 in the Wide ARG to the results of \cite{Yi2021-sc},
@@
 mutations occurring along terminal branches, and tallied them up by type.
-Figure~\ref{fig:mutational_spectra}B shows that the mutational spectrum from the
+Figure~\ref{fig:mutational_spectra}A shows that the mutational spectrum from the
 Wide ARG (based on 448,825 mutations) matches that reported by \citet[based on

### 2023-06-02 paper.tex
Reset hyperlinks
Fixes #170
Updates from Overleaf

--- 2023-06-02 paper.tex
@@
 \author[1]{Shing H. Zhan}
-\author[2$\star$]{Anastasia Ignatieva}
+\author[2,3$\star$]{Anastasia Ignatieva}
 \author[1$\star$]{Yan Wong}
-\author[3]{Katherine Eaton}
+\author[4]{Katherine Eaton}
 \author[1]{Benjamin Jeffery}
 \author[1]{Duncan S. Palmer}
-\author[2]{Carmen L. Murall}
-\author[4]{Sarah P. Otto}
+\author[4]{Carmen L. Murall}
+\author[5]{Sarah P. Otto}
 \author[1$\dagger$]{Jerome Kelleher}
@@
 \affil[2]{Department of Statistics, University of Oxford, Oxford, United Kingdom}
-\affil[3]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
-\affil[4]{Department of Zoology and Biodiversity Research Centre,
+\affil[3]{School of Mathematics and Statistics, University of Glasgow, Glasgow, United Kingdom}
+\affil[4]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
+\affil[5]{Department of Zoology and Biodiversity Research Centre,
 University of British Columbia, Vancouver, British Columbia, Canada}
@@
 including SARS-CoV-2
-\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
+\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2022-st}. By bringing
 together mutations carried by different lineages, recombination plays an
@@
 Recurrent   & 74,719 & (6.16\%) & 50,099 & (4.72\%) \\
-Reversions  & 72,617 & (5.99\%) & 48,226 & (4.54\%) \\
+Of which reversions  & 72,617 & (5.99\%) & 48,226 & (4.54\%) \\
 \bottomrule
@@
 USA/TX-HMH-MCoV-43092/2021 is inferred to be a recombinant
-between an XB grandparent and its XB grandchild, which
-is likely to be artefactual.
+between an XB grandparent and its XB grandchild, involving an intermediate UPGMA node.
+We assess it is a relatively likely that this is an artefactual recombination event,
+caused by undersampling. Alternatively, it could reflect true (but hard to detect)
+recombination between closely related lineages.
 

### 2023-06-02 paper.tex

--- 2023-06-02 paper.tex
@@
 \renewcommand\thefigure{S\arabic{figure}}
+\renewcommand{\theHfigure}{S\arabic{figure}}
 \setcounter{figure}{0}
 \renewcommand\thetable{S\arabic{table}}
+\renewcommand\theHtable{S\arabic{table}}
 \setcounter{table}{0}

### 2023-06-03 paper.tex
Updates from Overleaf

@@
 
-\title{Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2}
+\title{\bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2}
 \author[1]{Shing H. Zhan}
@@
 \author[1$\dagger$]{Jerome Kelleher}
-\affil[1]{Big Data Institute, Li Ka Shing Centre for Health
+\affil[1]{\normalsize Big Data Institute, Li Ka Shing Centre for Health
 Information and Discovery, University of Oxford, Oxford, United Kingdom}
@@
 \affil[$\dagger$]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
+
+% \author{
+% Shing H.~Zhan \thanks{Big Data Institute, Li Ka Shing Centre for Health
+% Information and Discovery, University of Oxford, Oxford, United Kingdom} \and Anastasia Ignatieva
+% \thanks{Department of Statistics, University of Oxford, Oxford, United Kingdom}
+% }
 

### 2023-06-05 paper.tex
edits from Duncan and Ana

--- 2023-06-02 paper.tex
@@
 \usepackage{authblk}
+\usepackage[symbol]{footmisc}
 \usepackage{amsmath}
@@
 
-\title{Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2}
-\author[1]{Shing H. Zhan}
-\author[2,3$\star$]{Anastasia Ignatieva}
-\author[1$\star$]{Yan Wong}
-\author[4]{Katherine Eaton}
-\author[1]{Benjamin Jeffery}
-\author[1]{Duncan S. Palmer}
-\author[4]{Carmen L. Murall}
-\author[5]{Sarah P. Otto}
-\author[1$\dagger$]{Jerome Kelleher}
-\affil[1]{Big Data Institute, Li Ka Shing Centre for Health
-Information and Discovery, University of Oxford, Oxford, United Kingdom}
-\affil[2]{Department of Statistics, University of Oxford, Oxford, United Kingdom}
-\affil[3]{School of Mathematics and Statistics, University of Glasgow, Glasgow, United Kingdom}
-\affil[4]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
-\affil[5]{Department of Zoology and Biodiversity Research Centre,
-University of British Columbia, Vancouver, British Columbia, Canada}
-\affil[$\star$]{Joint second author}
-\affil[$\dagger$]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
+\title{\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2}
+
+\author[1]{Shing~H.~Zhan}
+\author[2,3$\star$]{Anastasia~Ignatieva}
+\author[1$\star$]{Yan~Wong}
+\author[4]{Katherine~Eaton}
+\author[1]{Benjamin~Jeffery}
+\author[1]{Duncan~S.~Palmer}
+\author[4]{Carmen~L.~Murall}
+\author[5]{Sarah~P.~Otto}
+\author[1$\dagger$]{Jerome~Kelleher}
+% \affil[1]{\normalsize Big Data Institute, Li Ka Shing Centre for Health
+% Information and Discovery, University of Oxford, United Kingdom}
+% \affil[2]{Department of Statistics, University of Oxford, United Kingdom}
+% \affil[3]{School of Mathematics and Statistics, University of Glasgow, United Kingdom}
+% \affil[4]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
+% \affil[5]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Canada}
+% \affil[$\star$]{Joint second author}
+% \affil[$\dagger$]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
+\affil[ ]{\mbox{}\vspace{-2.5em}}
 
@@
 
+\setlength{\skip\footins}{1em}
+\setlength{\footnotemargin}{0.5em}
+\renewcommand{\thefootnote}{\arabic{footnote}}
+\footnotetext[1]{Big Data Institute, Li Ka Shing Centre for Health
+Information and Discovery, University of Oxford, United Kingdom}
+\footnotetext[2]{Department of Statistics, University of Oxford, United Kingdom}
+\footnotetext[3]{School of Mathematics and Statistics, University of Glasgow, United Kingdom}
+\footnotetext[4]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
+\footnotetext[5]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Canada}
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+\footnotetext[1]{Joint second author}
+\footnotetext[2]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
+
+\vspace{-1em}
 \begin{abstract}
@@
 \caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
-model, in which a query sequence (bottom) is described as an
+model, in which a focal sequence (bottom) is described as an
 imperfect mosaic of the sequences in a reference panel.
@@
 The Viterbi algorithm can be used to find a
-``copying path'' through the reference panel for a given query sequence that
+``copying path'' through the reference panel for a given focal sequence that
 maximises the likelihood under these parameters. Unseen states in the reference panel are shown as coloured lines enclosed by
@@
 Emissions may also be a function of the nucleotide states, but in our
-case, we assume that all observed alleles (in both the reference panel and
+case, we assume that all observed alleles in both the reference panel and
 focal sequence, $a_\ell$) at the site under consideration are equally likely and
@@
 For convenience, $\mu_\ell$ is commonly referred to as the `mutation probability', but we note that this probability of
-mismatch also encompasses various other error modes which result in a mismatch
+mismatch also encompasses various other error modes that result in a mismatch
 (such as sequencing and alignment errors). Note also that it is these probabilities,
 not the rates of mutation and recombination, that are required to fully define the HMM.
-That said, we can approximately relate these probabilities to the coalescent with recombination process which LS is approximating. In the case of recombination,
+That said, we can approximately relate these probabilities to the coalescent in a population with constant effective population size ($N_e$), with recombination process which LS is approximating to guide intuition.
+We may consider the probability of mutation to represent the
+probability of mutation along the branches joining the focal sequence to the closest member of a coalescent
+tree containing a reference panel with $n$ members, assuming a Poisson process
+along the branch. The expected length of these two branches is $4/n$ in units of $N_e$ generations, and so the mismatch
+probability represents the probability of observing a mutation or error mode,
+along these branches. We can then approximate the probability of observing a mismatch as:
+\begin{align*}
+    \mu_\ell \approx 1 - \exp{\left(-\frac{4\nu}{n}\right)}
+\end{align*}
+where $\nu$ is the mutation rate per $\mbox{ploidy} \times N_e$ generations.
+In the case of recombination,
 suppose that there is some constant underlying recombination rate, $\rho$.
 Assuming that recombination events occur as a Poisson process,
-then the probability of recombining between adjacent sites is
+then the probability of detecting a recombination event before coalescence between the focal sequence and the reference panel, between adjacent variant sites is
 \begin{align*}
@@
 where $m_j$ is the nucleotide position of variant $j$.
-Similarly, we may consider the probability of mutation to represent to the
-probability of mutation along the branch joining the focal sequence to a coalescent
-tree containing a reference panel with $n$ members, assuming a Poisson process
-along the branch. The expected length of this branch is $4/n$, and so the mismatch
-probability represents the probability of observing a mutation or error mode,
-along this branch. We can then approximate the probability of observing a mismatch as:
-\begin{align*}
-    \mu_\ell \approx 1 - \exp{\left(-\frac{4\nu}{n}\right)}
-\end{align*}
-where $\nu$ is the mutation rate per $\mbox{ploidy} \times N_e$ generations.
 
@@
 importance of mutation and recombination in the HMM. Specifically, an MMR value
-of $k$ will prefer $k$ mismatches (mutations) to a single recombination which
+of $k$ will prefer $k$ mismatches (mutations) to a single recombination that
  results in copying from a template with no mismatches. To map between
 recombination probability and mutation probability for a particular mismatch
-ratio, we simply consider the two paths which we wish to be equally likely, and
+ratio, we simply consider the two paths that we wish to be equally likely, and
 rearrange for the mutation or recombination probability. Consider the simple
-case when we assume $\mu_l=\mu$ and $r_l=r$. Without loss of generality,
+case where we assume $\mu_l=\mu$ and $r_l=r$. Without loss of generality,
 consider a region of length $m$. Up until this region, two paths are equally
@@
 We then
-set these path probabilities to be equal, and rearrange to obtain:
+set these path probabilities to be equal, and rearrange to relate $r$ and $\mu$ to one another::
 \begin{align*} r = \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k},
@@
 A varying degree of missingness is present in the genetic data, that is, for a
-given sample, a subset of nucleotides coded as `unknown' at a set of positions.
+given sample, a subset of nucleotides may be coded as `unknown' at a set of positions.
 As a result, we had to consider missingness in our implementation of the LS
@@
 the possible missing nucleotides, and use that in the determination of path
-probabilities. Another is to not attempt to estimate the true nucleotide, but
+probabilities. Another, which we carry out within \texttt{sc2ts}, is to not attempt to estimate the true nucleotide, but
 rather to consider such sites as uninformative to the path probability, and
@@
 of the nodes in the ARG that minimises the number of mutations required
-to insert the query sequence. Importantly, because the reference panel here consists of
+to insert the focal sequence. Importantly, because the reference panel here consists of
 \emph{every} node in the ARG, we can match to both older sample
@@
 data. The machinery used here can be modified to run the forwards and backwards
-algorithms which determine the probability of observing a focal sequence,
+algorithms that determine the probability of observing a focal sequence,
 integrating over all possible paths through the data, under the LS

### 2023-06-05 paper.tex
Minor edits from JK

@@
 sampled up to June 30, 2021, and the second is more sparsely sampled,
-consisting of 784K sequences sampled up to June 30, 2022.
+consisting of 657K sequences sampled up to June 30, 2022.
 We find that both ARGs are
@@
 downstream analyses, to quickly and automatically detect new recombinants,
-and to provide an efficient and convenient platform for computation
-makes \texttt{sc2ts} a promising approach, worthy of
-further investment and development.
+and to utilise an efficient and convenient platform for computation
+based on well-engineered technologies
+makes \texttt{sc2ts} a promising approach.
 \end{abstract}
@@
 established and spread to high frequency, and accounting for recombinant
-ancestry is now essential in understanding the evolution of SARS-CoV-2.
+ancestry is now essential in understanding the ongoing evolution of SARS-CoV-2.
 
@@
 Then, since many samples
-in the batch can share an attachment path,
-we infer phylogenetic trees for each of the clusters separately using standard methods
+in a batch can share an attachment path,
+we infer phylogenetic trees for each of these clusters separately using standard methods
 (Figure~\ref{fig:overview_sc2ts}C; Section~\ref{sec:sample-cluster-tree-inference}).
@@
 the properties of these ARGs to illustrate the power of the method
-and to inform subsequent development. We find that these
-ARGs accurately reflect known patterns of phylogenetic relationships
-between samples (Section~\ref{sec:backbone_phylogeny}),
+and to inform subsequent development. 
+We find that these
+ARGs accurately capture known phylogenetic 
+relationships (Section~\ref{sec:backbone_phylogeny}) and
 mutational spectra (Section~\ref{sec:mutation_spectrum}),
-and automatically identify the majority of known recombinant
-lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),
+as well as automatically identifying the majority of known recombinant
+lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages})
 with a high level of precision in the
@@
 (Section~\ref{sec:breakpoint_intervals})
-and the relationship between parental sequences
+and relationship between parental sequences
 (Section~\ref{sec:parent_divergence}).
@@
 a few weeks (see below for further details on timings and
-computer hardware used). The cut-off dates for sampling were chosen
-arbitrarily.
+computer hardware used). The cut-off dates for sampling 
+are arbitrary.
 We inferred two ARGs, which we refer to as the
@@
 The Wide ARG is densely sampled but time-limited and
-includes all 1.27 million sample genomes collected up to June 30, 2021
-with a maximum delay between sampling and submission dates of 30 days
-(see Section~\ref{sec:filtering_time_travellers}).
+includes 1.27 million strains collected up to June 30, 2021
+which 
+pass some quality-control filters (Section~\ref{sec:data_preprocessing})
+and  have a maximum delay between sampling and submission dates of 30 days
+(Section~\ref{sec:filtering_time_travellers}).
 For the Long ARG, we randomly sub-sample a maximum of 1,000 genomes per
@@
 Inferring these ARGs does require substantial computation.
-The Wide ARG required 17 days of computation time on a machine with
+The Wide ARG required 17 days to infer on a server with
 128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). The Long ARG
@@
 are colour-coded according to the Pango lineage status assigned to the tip
-samples; no Pango recombinants are shown, as none are shared between the trees.
+samples.
 The tanglegram was generated using the Neighbor-Net algorithm
@@
 partly driven by some hypermutable sites (e.g., site 28,271 has over 7,000
-mutations in the Wide ARG, Figure~\ref{fig:breakpoint-distribution}), which
+mutations in the Wide ARG; Figure~\ref{fig:breakpoint-distribution}), which
 may be artefactual.
@@
 291 recombination nodes. The criterion of 5 descendants matches the
-minimum number required to designate a new Pango recombinant lineage
-\citep{Rambaut2020-dw}.
+minimum number required to designate a new Pango 
+lineage~\citep{Rambaut2020-dw}.
 Note that as each plotted point represents a breakpoint, a recombination
@@
 sequence, arising as a result of a mixed infection followed by onward
-transmission. We would expect our ARG to reveal evolutionary histories of this
+transmission. We would expect our ARGs to reveal evolutionary histories of this
 nature, where all the samples assigned to a given recombinant lineage
@@
 Emissions may also be a function of the nucleotide states, but in our
-case, we assume that all observed alleles in both the reference panel and
-focal sequence, $a_\ell$) at the site under consideration are equally likely and
-occur with mutation probability $\mu_\ell/(a_\ell-1)$
+case, we assume that all alleles ($A,C,G,T,-$) can be mutated to with equal probability $\mu_\ell/5$.
+This is reasonable for rapidly evolving pathogens, but we note that setting the number of alleles at site $\ell$ ($a_\ell$) to the set of observed alleles across all analysed samples ($\mu_\ell/(a_\ell-1)$) will often be more appropriate
  (Figure~\ref{fig:ls_diagram}, blue panel).
@@
 \mathbb{P}[\mathcal{P}_1] &= \frac{cr}{n}
-\left(\left(1-r\right)+\frac{r}{n}\right)^m\left(1-\mu\right)^m
+\left(\left(1-r\right)+\frac{r}{n}\right)^{m-1}\left(1-\mu\right)^m
 &\text{recombine to a template with no mismatches,}\\
@@
 \left(\left(1-r\right) +
-\frac{r}{n}\right)\left(\left(1-r\right)+\frac{r}{n}\right)^m
+\frac{r}{n}\right)\left(\left(1-r\right)+\frac{r}{n}\right)^{m-1}
 \left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
@@
 We then
-set these path probabilities to be equal, and rearrange to relate $r$ and $\mu$ to one another::
+set these path probabilities to be equal, and rearrange to relate $r$ and $\mu$ to one another:
 \begin{align*} r = \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k},

### 2023-06-05 paper.tex
Various updates

@@
 \author[1]{Duncan~S.~Palmer}
-\author[4]{Carmen~L.~Murall}
+\author[4]{Carmen~Lia~Murall}
 \author[5]{Sarah~P.~Otto}
@@
 and estimation of parameters such as
-recombination rates \citep[e.g.,][]{Hinch2011-tz},
+recombination rates \citep[e.g.,][]{Hinch2011-tz}
 and selection within and across hosts in viral
@@
 not the rates of mutation and recombination, that are required to fully define the HMM.
-That said, we can approximately relate these probabilities to the coalescent in a population with constant effective population size ($N_e$), with recombination process which LS is approximating to guide intuition.
+That said, we can approximately relate these probabilities to the coalescent process in a population with constant effective population size ($N_e$), which LS is approximating to guide intuition.
 We may consider the probability of mutation to represent the
@@
 recombinations are favoured over mutations
-than a single ancestral genome. We use an MMR value of $k=3$ in this work,
+than a single ancestral genome. 
+We use an MMR value of $k=3$ in this work,
 as it provided results that tallied well with empirical observations. 
-Exploring different mismatch ratios, and more sophisticated parameterisations
+Exploring different mismatch ratios and more sophisticated parameterisations
 of the HMM are important avenues for future work.
@@
 We can solve this massive optimisation problem exactly because the ARG-based
-implementation of the LS HMM used in \texttt{tsinfer} scales approximately
+implementation of the LS HMM used in \texttt{tsinfer}~\citep{Kelleher2019-ba} 
+scales approximately
 logarithmically with reference panel size (as opposed to linearly,

### 2023-06-05 paper.tex
Start on details for false positives table

@@
 % sister clade of XAC, and annotated as ``miscBA2BA1PostSpike''.
+
+\subsection{False positive recombinants}
+The Long and Wide ARGs both contain several recombination nodes that are
+ancestral to large numbers of samples, and are almost certainly
+artefactual. The
+
+\begin{table} \centering
+\begin{tabular}{lllll}
+Strain & Lineage & Descendants & HMM Cost & Lineage rank\\
+Germany/HH-RKI-I-061284/2021 & B.1.617.2 & 178405 & 17 & 2 \\
+India/ILSGS00961/2021   & B.1.617.2 & 177649 & 25  & 7  \\
+Denmark/DCGC-281594/2021 & BA.2.9 & 127227 & 32  & 1\\
+USA/NJ-GBW-EWR000001/2021 & BA.3 & 127230 & 25 & 1\\
+\end{tabular}
+\caption{\label{tab:false_positive}
+False positive recombination events.}
+\end{table}
+
+Further study is required to systematically identify likely false
+positive
 

### 2023-06-06 paper.tex

--- 2023-06-05 paper.tex
@@
 % sister clade of XAC, and annotated as ``miscBA2BA1PostSpike''.
+
+\subsection{False positive recombinants}
+The Long and Wide ARGs both contain several recombination nodes that are
+ancestral to large numbers of samples, and are almost certainly
+artefactual. Table~\ref{tab:false_positive} shows details for four recombinants
+in the Long ARG that are likely false positives, and have come to be
+the ancestors of a large number of samples. These are the top-four recombinants
+from the Long ARG in terms of numbers of descendant samples
+(the fifth-largest has substantially fewer, at 14,869),
+labelled A--D. These false positives are associated with the origins
+of large outbreaks. For each row, we show the sample's Pango lineage
+designation and the rank of that sample in the overall list of
+samples with that designation. Thus, rows C and D are the first
+samples seen in the Long ARG from the BA.2.9 and BA.3 lineages,
+and A and B are the third and eighth samples from the B.1.617.2
+lineage (Delta VoC). Also shown is the overall ``cost'' of the Viterbi
+solution computed by the LS HMM (Section~\ref{sec:ls}), which
+is $3 \times$ number of recombinations + number of additional mutations
+(for a mismatch ratio of 3; Section~\ref{sec:ls}).
+The column shows that each of these samples was a large
+``distance'' from the current ARG when it was added. The mean
+HMM cost over all 2078 recombinants in the Long ARG is 8.39
+(median: 7), and the mean cost for the  763 non-singleton
+recombinants is 8.5 (median: 7). Thus, the recombinants in
+Table~\ref{tab:false_positive} are in the tail of this distribution.
+
+\begin{table} \centering
+\begin{tabular}{llllll}
+\toprule
+& Strain & Descendants & Lineage & Rank in lineage&HMM Cost \\
+\midrule
+A&Germany/HH-RKI-I-061284/2021 & 178,405    & B.1.617.2   & 3& 17  \\
+B&India/ILSGS00961/2021        & 177,649    & B.1.617.2   & 8& 25   \\
+C&Denmark/DCGC-281594/2021     & 127,227    & BA.2.9      & 1& 32 \\
+D&USA/NJ-GBW-EWR000001/2021    & 127,230    & BA.3        & 1& 25 \\
+\bottomrule
+\end{tabular}
+\caption{\label{tab:false_positive}
+False positive recombination events. Details show are for the top
+four recombination nodes in the Long ARG ordered by number of
+descending samples. See the text for details on the remaining columns.
+Rows are labelled A, B, C and D for ease of reference.}
+\end{table}
+
+Occasional evolutionary leaps, in which a large number of mutations
+are acquired in sudden jumps
+is a signature feature of SARS-CoV-2~\citep{Corey2021-sars,Nielsen2023-host}.
+Such ``saltations'' naturally present challenges to \texttt{sc2ts} and the
+current HMM parameterisation of three mutations per recombination (mismatch
+ratio). The first few sequences from these new lineages will match in to
+the existing ARG with very low likelihood, and the HMM will therefore
+search for ways to reduce the number of mutations required by recombining
+in segments. If this is combined with data quality issues and
+``time travellers'' (Section~\ref{sec:filtering_time_travellers}),
+the early stages of the VoC emergence can be quite muddled.
+This appears to be the case for the Delta VoC,
+which corresponded to the sudden emergence of
+[HOW MANY, citation?] mutations,
+as show in Table~\ref{tab:false_positive}A and B.
+Similar patterns around the emergence of Delta are seen in the Wide ARG
+(not shown).
+
+Given these considerations, it  is important to note that the
+number of ultimate descendants is not an entirely reliable
+indicator of reliability for recombinants.
+Further study is required to systematically identify such false
+positives recombinants, and to update the topology with a
+more parsimonious explanation of the data.
 

### 2023-06-06 paper.tex
Minor updates

@@
 from the Long ARG in terms of numbers of descendant samples
-(the fifth-largest has substantially fewer, at 14,869),
+(the fifth-largest has substantially fewer, at 14,869 descendant samples),
 labelled A--D. These false positives are associated with the origins
@@
 and A and B are the third and eighth samples from the B.1.617.2
-lineage (Delta VoC). Also shown is the overall ``cost'' of the Viterbi
+lineage (Delta VoC), respectively. Also shown is the overall ``cost'' of the Viterbi
 solution computed by the LS HMM (Section~\ref{sec:ls}), which
@@
 ``distance'' from the current ARG when it was added. The mean
-HMM cost over all 2078 recombinants in the Long ARG is 8.39
+HMM cost over all 2,078 recombinants in the Long ARG is 8.39
 (median: 7), and the mean cost for the  763 non-singleton
@@
 to any other state as 1.
-% $\mbox{N} \rightarrow \star$ as 1.
 
@@
 the particular mutations to collapse.
-Table~\ref{tab:args} shows that
-while these nodes constitute a small fraction of the
-overall total (Wide ARG: 2.77\%, Long ARG: 4.82\%),
-they are parental to a much larger fraction of the
-edges (Wide ARG: 12.77\%, Long ARG: 20.47\%).
 
@@
 inference by ``pushing in'' a new node which descends from the
-original parent which carries all its mutations except those
+original parent, and carries all its mutations except those
 causing the reversions on the newly added node (Figure \ref{fig:overview_sc2ts}E).
-Table~\ref{tab:args} shows that
-these nodes constitute a similar fraction of the
-overall numbers of nodes (Wide ARG: 2.79\%, Long ARG: 4.39\%)
-and edges (Wide ARG: 12.66\%, Long ARG: 18.46\%)
-as mutation collapsing.
-
-Together these operations substantially increase the overall parsimony
-of the inferred ARGs, most noticeably in terms of the number of
-reversion mutations inferred. These nodes are also being chosen by the
-LS HMM as likely choices of parent (as indicated by the node and edge
-counts quoted above), also demonstrating that the heuristics are successfully
+
+Table~\ref{tab:args} shows that nodes generated by 
+these operations constitute 
+roughly the same fraction of the total in both the Long and Wide ARGs,
+and contribute significantly to the overall topology.
+These nodes are also being chosen by the
+LS HMM as likely choices of parent (data not shown) 
+demonstrating that the heuristics are successfully
 capturing features of real sequences.
@@
 ARGs still contain a large number of reversion mutations which
-are likely to be mostly artefactual, it is clear that there
+are likely to be mostly artefactual (Section~\ref{sec:mutation_spectrum}), 
+it is clear that there
 is room for improvement and that further parsimony-increasing

### 2023-06-06 paper.tex
Simplify some things in the LS section

@@
 (such as sequencing and alignment errors). Note also that it is these probabilities,
-not the rates of mutation and recombination, that are required to fully define the HMM.
-That said, we can approximately relate these probabilities to the coalescent process in a population with constant effective population size ($N_e$), which LS is approximating to guide intuition.
-We may consider the probability of mutation to represent the
-probability of mutation along the branches joining the focal sequence to the closest member of a coalescent
-tree containing a reference panel with $n$ members, assuming a Poisson process
-along the branch. The expected length of these two branches is $4/n$ in units of $N_e$ generations, and so the mismatch
-probability represents the probability of observing a mutation or error mode,
-along these branches. We can then approximate the probability of observing a mismatch as:
-\begin{align*}
-    \mu_\ell \approx 1 - \exp{\left(-\frac{4\nu}{n}\right)}
-\end{align*}
-where $\nu$ is the mutation rate per $\mbox{ploidy} \times N_e$ generations.
-In the case of recombination,
-suppose that there is some constant underlying recombination rate, $\rho$.
-Assuming that recombination events occur as a Poisson process,
-then the probability of detecting a recombination event before coalescence between the focal sequence and the reference panel, between adjacent variant sites is
-\begin{align*}
-    r_\ell = 1 - \exp\left(-\frac{\rho\left(m_{i-1} - m_{i}\right)}{n}\right),
-\end{align*}
-where $m_j$ is the nucleotide position of variant $j$.
+not the rates of mutation and recombination, that are required to fully define
+the HMM; see~\cite{Donnelly2010-coalescent} for a discussion of how these
+parameters relate to the coalescent process.
+
+% That said, we can approximately relate these probabilities to the coalescent
+% process in a population with constant effective population size ($N_e$), which
+% LS is approximating to guide intuition. We may consider the probability of
+% mutation to represent the probability of mutation along the branches joining
+% the focal sequence to the closest member of a coalescent tree containing a
+% reference panel with $n$ members, assuming a Poisson process along the branch.
+% The expected length of these two branches is $4/n$ in units of $N_e$
+% generations, and so the mismatch probability represents the probability of
+% observing a mutation or error mode, along these branches. We can then
+% approximate the probability of observing a mismatch as: \begin{align*} \mu_\ell
+% \approx 1 - \exp{\left(-\frac{4\nu}{n}\right)} \end{align*} where $\nu$ is the
+% mutation rate per $\mbox{ploidy} \times N_e$ generations. In the case of
+% recombination, suppose that there is some constant underlying recombination
+% rate, $\rho$. Assuming that recombination events occur as a Poisson process,
+% then the probability of detecting a recombination event before coalescence
+% between the focal sequence and the reference panel, between adjacent variant
+% sites is \begin{align*} r_\ell = 1 - \exp\left(-\frac{\rho\left(m_{i-1} -
+% m_{i}\right)}{n}\right), \end{align*} where $m_j$ is the nucleotide position of
+% variant $j$.
 
@@
 consider two of the $n$ members of the reference panel.
-
 One, $\mathcal{P}_1$,
@@
 \mathbb{P}[\mathcal{P}_1] &= \frac{cr}{n}
-\left(\left(1-r\right)+\frac{r}{n}\right)^{m-1}\left(1-\mu\right)^m
+\alpha^{m-1}\left(1-\mu\right)^m
 &\text{recombine to a template with no mismatches,}\\
 \mathbb{P}[\mathcal{P}_2] &= c
-\left(\left(1-r\right) +
-\frac{r}{n}\right)\left(\left(1-r\right)+\frac{r}{n}\right)^{m-1}
+\alpha^{m}
+% \alpha \alpha^{m-1}
 \left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
@@
 % variant and a particular member of the reference panel,
-where $c$ is the
+where $\alpha = (1 - r + r / n) $ and $c$ is the
 likelihood of the path up to this point, which are assumed to be

### 2023-06-07 paper.tex
Updates from Overleaf

@@
 
-We focus here on the 749 Nextclade-designated recombinant samples, all
-but two of which descend from a recombination node in the Long ARG.
-These two exceptions are both designated XP, and a likely
+We focus here on the 749 Nextclade-designated recombinant samples.
+There are two samples designated XP which do not descend from a recombination node, 
+and a likely
 explanation for the absence of a corresponding recombination event in the Long ARG
@@
 between an XB grandparent and its XB grandchild, involving an intermediate UPGMA node.
-We assess it is a relatively likely that this is an artefactual recombination event,
-caused by undersampling. Alternatively, it could reflect true (but hard to detect)
+It seems likely to be artefactual recombination event
+caused by undersampling, but could reflect true 
 recombination between closely related lineages.
@@
 \subsection{False positive recombinants}
+\label{sec:false_positives}
 The Long and Wide ARGs both contain several recombination nodes that are
@@
 This appears to be the case for the Delta VoC,
-which corresponded to the sudden emergence of
-[HOW MANY, citation?] mutations,
-as show in Table~\ref{tab:false_positive}A and B.
+corresponding to rows A and B of Table~\ref{tab:false_positive},
+which emerged and quickly rose to high prevalence in early-mid 2021,
+carrying 30 characteristic mutations compared to the reference sequence 
+\citep{McCrone2022-context}, including nine mutations in the S gene 
+not seen in earlier VoCs.
 Similar patterns around the emergence of Delta are seen in the Wide ARG
@@
 number of ultimate descendants is not an entirely reliable
-indicator of reliability for recombinants.
+indicator of the quality of inferred recombinants.
 Further study is required to systematically identify such false
@@
 recombination rates \citep[e.g.,][]{Hinch2011-tz}
-and selection within and across hosts in viral
+and intensity of selection within and across hosts in viral
 sequence data \citep[e.g.,][]{Palmer2019-wa}.
@@
 Note that B.1.617.2 is the origin of the Delta VoC, which includes all AY.* classes.
-Additional descendants within the main clade are summarised giving their NextClade Pango designation and clade count / total ARG count (the latter being omitted for non-recombinant designations).
+Additional descendants within the main clade are summarised giving their Nextclade Pango designation and clade count / total ARG count (the latter being omitted for non-recombinant designations).
 }

### 2023-06-07 plot.py
Change order of labels
And reorganise code to prep moving untangling routine to a notebook

@@ -53,12 +53,12 @@ genes = {
 }
 
 
-class FocalTreeTs:
+class SingleTreeTs:
     """Convenience class to access a single focal tree in a tree sequence"""
 
-    def __init__(self, ts, pos, basetime=None):
-        self.tree = ts.at(pos, sample_lists=True)
-        self.pos = pos
+    def __init__(self, ts, basetime=None):
+        # .tree is the first non-empty tree
+        self.tree = next(t for t in ts.trees() if t.num_edges > 0)
         self.basetime = basetime
 
     @property
@@ -89,7 +89,7 @@ class Nextstrain:
     def __init__(self, filename, span, prefix="data"):
         """
         Load from a nextstrain nexus file.
-        Note that NextClade also produces a tree with  more samples but no branch lengths
+        Note that NextClade also produces a tree with more samples but no branch lengths
         e.g. at
             https://github.com/nextstrain/nextclade_data/tree/
             release/data/datasets/sars-cov-2/references/MN908947/versions
@@ -296,23 +296,30 @@ class Cophylogeny(Figure):
             [sc2ts_tip.at(self.pos), nxstr_tip.first()]
         )
 
-        # Align the time in the nextstrain tree to the sc2ts tree
-        ns_sc2_time_difference = []
-        for s1, s2 in zip(sc2ts_tip.samples(), nxstr_tip.samples()):
-            n1 = sc2ts_tip.node(s1)
-            n2 = nxstr_tip.node(s2)
-            assert n1.metadata["strain"] == n2.metadata["strain"]
-            ns_sc2_time_difference.append(n1.time - n2.time)
-        dt = timedelta(**{nxstr_tip.time_units: np.median(ns_sc2_time_difference)})
-
         nxstr_order = list(
             reversed(nxstr_order)
         )  # RH tree rotated so reverse the order
 
-        self.sc2ts = FocalTreeTs(sc2ts_tip.simplify(sc2ts_order), self.pos, basetime)
-        self.nxstr = FocalTreeTs(
-            nxstr_tip.simplify(nxstr_order), self.pos, basetime - dt
-        )
+        plotted_sc2ts_ts = sc2ts_tip.simplify(sc2ts_order).keep_intervals([sc2ts_tip.at(self.pos).interval])
+        plotted_nxstr_ts = nxstr_tip.simplify(nxstr_order).keep_intervals([nxstr_tip.at(self.pos).interval])
+
[…]

### 2023-06-07 plot.py
Add new supplementary figure PDF

@@ -948,6 +948,7 @@ class Pango_X_tight_graph(Pango_X_graph):
     node_def_lineage = (
         "Nextclade_pango"  # Always use this when defining which nodes to pick
     )
+    show_descendant_samples = "sample_tips"
 
     def __init__(self):
         super().__init__()
@@ -1026,6 +1027,7 @@ class Pango_X_tight_graph(Pango_X_graph):
             edge_font_size=self.edge_font_size,
             colour_metadata_key="Imputed_" + self.imputed_lineage,
             node_positions=node_positions,
+            show_descendant_samples=self.show_descendant_samples
         )
 
     def plot(self, args):
@@ -1623,6 +1625,109 @@ class Pango_XB_gisaid_large_graph(Pango_XB_nxcld_tight_graph):
     def post_process(cls, ax):
         x_min, x_max = ax.get_xlim()
         ax.set_xlim(x_min + (x_max - x_min) * 0.06, x_max - (x_max - x_min) * 0.06)
+
+class LongTopTwoFalsePositiveGraph(Pango_X_tight_graph):
+    name = "false_positive_top2_nxcld_large_graph"
+    focal_nodes = [229998, 235293]
+    main_focal_node_parents = [21605, 204945]
+    main_focal_child = 248215
+    main_focal_grandchild = 243149
+    main_delta_source = 261771
+    node_colours = {
+        **{
+            main_delta_source: "gold",
+            "B.1.617.2": "#e2da91",
+            "AY.100": "#e2da91",
+            "AY.112": "#e2da91",
+            None: "lightgray",
+            "Unknown (R)": "k",
+            "Unknown": "white",
+        },
+        **{u: "darkred" for u in focal_nodes}
+    }
+    imputed_lineage = "Nextclade_pango"
+    label_replace = {"Unknown (R)": "Rec node"}
+    show_ids = True # Show IDs for all nodes
+    figsize = (20, 30)
+    node_size = 5000
+    edge_font_size = 7
+    node_font_size = 7
+    mutations_fn = None
+    show_metadata = True
+    show_descendant_samples = "tips"
+
+    @classmethod
+    def define_nodes(cls, ts):
+        to_plot = set()
+        for u in set(cls.focal_nodes) | {cls.main_focal_child, cls.main_focal_grandchild}:
+            to_plot.add(u)
+            for child in np.unique(ts.edges_child[ts.edges_parent == u]):
+                to_plot.add(child)
+
[…]

### 2023-06-07 plot.py
Minor label change

@@ -1697,7 +1697,7 @@ class LongTopTwoFalsePositiveGraph(Pango_X_tight_graph):
             "Unknown (R)": "Additional recombination node",
             "B.1.617.2": "Nextclade Pango Delta VoC",
             cls.main_delta_source: "Large Delta polytomy (>100 children)",
-            "Unknown": "Unknown Pango lineage",
+            "Unknown": "Unimputed Pango lineage",
         }
         ax.legend(
             title=f"$\\bf{{Highlighted\ nodes}}$",

### 2023-06-08 paper.tex

@@
 \usepackage[hidelinks]{hyperref}
+
+% sets colours for notes to each other in the text
+\usepackage{xcolor}
+\newcommand{\sally}[1]{\textcolor{red}{#1}}
+\newcommand{\red}[1]{\textcolor{red}{#1}}
+\newcommand{\green}[1]{\textcolor{green}{#1}}
+\newcommand{\blue}[1]{\textcolor{blue}{#1}}
 
@@
 the average number per site
-is large in both ARGs (Wide ARG: 41.23; Long ARG:36.10; Table~\ref{tab:args}).
-However, a lower median count (Wide ARG: 14; Long ARG:13),
+is large in both ARGs (Wide ARG: 41.23; Long ARG: 36.10; Table~\ref{tab:args}).
+However, a lower median count (Wide ARG: 14; Long ARG: 13),
 suggests that these high mutation counts are
@@
 ancestral mutation (see Section~\ref{sec:args} for precise definitions).
+\sally{[What confuses me is that I would think that none of these reversions would remain after the reversion push step of Figure 1D.
+Is this because ``immediately ancestral'' is incorrect or because the problem is that some of these immediate reversions are created deeper in the tree than the newly added samples (i.e., uncorrected by the reversion push).
+If the former, maybe say ``reverting the state change of a recent ancestral mutation'', if the latter, maybe say  ``(see Section~\ref{sec:args} for precise definitions, note that the reversion push algorithm of Figure 1D only eliminates reversions in the newly added portion of the tree and not reversions around earlier nodes created by this algorithm)''.]}
 The majority of these mutations are likely to be
@@
 Even if we observe the recombinant sequence before subsequent
-divergence occurs (during onward transmission), if the two parent sequences
-are similar there is no way to
-accurately identify the breakpoint.
+divergence occurs (during onward transmission), there is no way to identify the exact breakpoint if the two parent sequences
+are similar.
 Here we define the
-interval around a particular breakpoint as the genome coordinates over which the
+interval \sally{within which a particular breakpoint may have occurred} as the genome coordinates over which the
 sequences for the left and right parent nodes are identical. The right-hand
@@
 homologous recombination~\citep{Kim2020-gt,Zou2021-sars}
-[Sally: KATHERINE, WHAT DO YOU THINK?  These not-full-length subgenomic RNA intermediates may provide a lot of homologous templates for some genes more than other (e.g., fig 3B here)]
+\sally{[Sally: KATHERINE, WHAT DO YOU THINK?  These not-full-length subgenomic RNA intermediates may provide a lot of homologous templates for some genes more than other (e.g., fig 3B here).]}
 
 It is important to note that the precise endpoints of intervals can be
-somewhat arbitrary, since they are defined by exact comparisons
-of a recombinant's parents.
+somewhat arbitrary, because they are defined by \sally{the sequence differences that happen to be present in the} recombinant's parents.
 Thus, breakpoint intervals will tend to be truncated at sites that are
@@
 It is therefore helpful to compare the number of intersecting intervals
-with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
+with the number of mutations per site (Figure~\ref{fig:breakpoint-distribution}).
 For instance, position 28,271 (located in the ORF8-N intergenic
@@
 homopolymeric region may be prone to sequencing errors
-and should be included in the list of ``problematic sites'' that
+and \sally{potentially} should be included in the list of ``problematic sites'' that
 are excluded from analysis (Section~\ref{sec:data_preprocessing}).
@@
 
-The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
+\sally{[Delete this paragraph now?
+The new section 4.8 gets at the real issue, that the multi-step mutations common in new VoC generate false positive signals of recombination.]} The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
 1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, introduced as a
 result of adding strain
-Denmark/DCGC-281594/2021, is ancestral to more than 127K samples,
+Denmark/DCGC-281594/2021 (BA.2.9), is ancestral to more than 127K samples, 
 and is highly likely to be a spuriously inferred recombination event.
+\sally{[Replace with ``...is ancestral to more than 127K samples and XX immediate descendants.
+While such a hyper-connected node may reflect a large outbreak with little phylogenetic resolution, it may also be spurious, e.g., caused by the attraction of sequences with similar constellations of mutations or another phylogenetic artefact.'']}
 Understanding the origin of this
-and other similar high-impact, but likely spurious recombination
+and other similar high-impact, but potentially spurious, recombination
 events is an important aspect of future work.
@@
 between an XB grandparent and its XB grandchild, involving an intermediate UPGMA node.
-It seems likely to be artefactual recombination event
-caused by undersampling, but could reflect true 
-recombination between closely related lineages.
+It seems likely to be an artefactual recombination event caused by undersampling, but it could also reflect true recombination between closely related lineages.
 
@@
 The Long and Wide ARGs both contain several recombination nodes that are
-ancestral to large numbers of samples, and are almost certainly
-artefactual. Table~\ref{tab:false_positive} shows details for four recombinants
-in the Long ARG that are likely false positives, and have come to be
+ancestral to a large number of samples\sally{.
+These inferences of recombination may well be artefacts caused by the appearance of new variants of concern carrying more than the expected number of mutations \cite{otto2021origins}}.
+Table~\ref{tab:false_positive} shows details for four recombinants
+in the Long ARG that are likely false positives and have come to be
 the ancestors of a large number of samples. These are the top-four recombinants
@@
 (the fifth-largest has substantially fewer, at 14,869 descendant samples),
-labelled A--D. These false positives are associated with the origins
-of large outbreaks. For each row, we show the sample's Pango lineage
+labelled A--D. 
+\sally{[DELETED SENTENCE]}%These false positives are associated with the origins of large outbreaks. 
+For each row, we show the sample's Pango lineage
 designation and the rank of that sample in the overall list of
@@
 Occasional evolutionary leaps, in which a large number of mutations
-are acquired in sudden jumps
-is a signature feature of SARS-CoV-2~\citep{Corey2021-sars,Nielsen2023-host}.
+are acquired in sudden jumps,
+is a signature feature of SARS-CoV-2~\citep{Corey2021-sars,otto2021origins,Nielsen2023-host}.
 Such ``saltations'' naturally present challenges to \texttt{sc2ts} and the
 current HMM parameterisation of three mutations per recombination (mismatch
-ratio). The first few sequences from these new lineages will match in to
-the existing ARG with very low likelihood, and the HMM will therefore
+ratio). The first few sequences from these new lineages will \sally{be a poor match} to
+the existing ARG, and the HMM will therefore
 search for ways to reduce the number of mutations required by recombining
-in segments. If this is combined with data quality issues and
-``time travellers'' (Section~\ref{sec:filtering_time_travellers}),
-the early stages of the VoC emergence can be quite muddled.
+segments. 
+\sally{[DELETED SENTENCE (TRUE WHETHER OR NOT QUALITY ISSUES OCCUR).] }%If this is combined with data quality issues and ``time travellers'' (Section~\ref{sec:filtering_time_travellers}), the early stages of the VoC emergence can be quite muddled.
 This appears to be the case for the Delta VoC,
@@
 section for details of the corresponding Jupyter notebooks).
-Retooling methods to scale rapidly and to encode recombination
+Retooling methods to scale up rapidly with expanding data and to encode recombination
 promises to improve tracking during this and any future pandemic.
@@
 This assumption dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
-Emissions may also be a function of the nucleotide states, but in our
-case, we assume that all alleles ($A,C,G,T,-$) can be mutated to with equal probability $\mu_\ell/5$.
-This is reasonable for rapidly evolving pathogens, but we note that setting the number of alleles at site $\ell$ ($a_\ell$) to the set of observed alleles across all analysed samples ($\mu_\ell/(a_\ell-1)$) will often be more appropriate
+Emissions may also be a function of the nucleotide states, but in our RNA virus case, we assume that \sally{mutations occur to all possible alternate alleles} ($A,C,G,U,-$) with equal probability $\mu_\ell/5$.
+\sally{[Why is this 5?
+I wouldn't have counted mutations to the same allele as a mutation.]} This is reasonable for rapidly evolving pathogens, but we note that setting the number of alleles at site $\ell$ ($a_\ell$) to the set of observed alleles across all analysed samples ($\mu_\ell/(a_\ell-1)$) will often be more appropriate
  (Figure~\ref{fig:ls_diagram}, blue panel).
@@
 recombination probability and mutation probability for a particular mismatch
-ratio, we simply consider the two paths that we wish to be equally likely, and
+ratio, we simply consider the two paths that we wish to be equally likely and
 rearrange for the mutation or recombination probability. Consider the simple
@@
 % variant and a particular member of the reference panel,
-where $\alpha = (1 - r + r / n) $ and $c$ is the
+where \sally{$\alpha = (1 - r(n-1)/n) $ represents the probability of a sequence not recombining at a given position with any of the other $n-1$ members of the reference panel [REVISION OK?]} and $c$ is the
 likelihood of the path up to this point, which are assumed to be
@@
 Thus, for lower MMR values ($k$ here),
-recombinations are favoured over mutations
-than a single ancestral genome.
-We use an MMR value of $k=3$ in this work,
-as it provided results that tallied well with empirical observations.
+\sally{recombination is increasingly favoured over multiple mutations in a specific ancestral genome}.
+%We use an MMR value of $k=3$ in this work, as it provided results that tallied well with empirical observations.
+\sally{We use an MMR value of $k=3$ in this work,
+because of the relatively high rate of recombination relative to mutation typical of coronaviruses \cite{amoutzias2022remarkable}.}
 Exploring different mismatch ratios and more sophisticated parameterisations
@@
 beforehand using external software to hard-call or define a distribution over
-the possible missing nucleotides, and use that in the determination of path
-probabilities. Another, which we carry out within \texttt{sc2ts}, is to not attempt to estimate the true nucleotide, but
+the possible missing nucleotides and use that in the determination of path
+probabilities. Another, which we carry out within \texttt{sc2ts}, is to not attempt to estimate the true nucleotide but
 rather to consider such sites as uninformative to the path probability, and
@@
 to any other state as 1.
+\sally{[Repeated clause in the last sentence?
+Also, I'm a bit unclear if this is enough.
+Let's say that the nodes were G->Unknown->G.
+The latter is given a probability of one, but is the former G->Unknown also allowed with probability one?
+Note, however, that G->G shouldn't occur with probability one but Exp[-mu/t]]}
 
@@
 evolutionary relationships~\citep{Felsenstein2004-inferring}.
-Table~\ref{tab:args} show that
+Table~\ref{tab:args} shows that
 this within-cluster tree building has a significant influence on the
-overall ARG topology
+overall ARG topology,
 and therefore applying more sophisticated
-tree building methods which keep track of the required mutations
-(rather then inferring post-hoc by parsimony), is a likely avenue
+tree building methods that keep track of the required mutations
+(rather then inferring post-hoc by parsimony) is a likely avenue
 for improvements in overall inference quality.
@@
 see Section~\ref{sec:treatment_recombinants}) under the
-HMM is an inherently greedy strategy,
+HMM is an inherently greedy strategy
 and can produce inferences that are clearly unparsimonious.
@@
 by creating a new node representing the ancestor that carried
-those shared mutations, and make that new node the parent of the
+those shared mutations and make that new node the parent of the
 siblings carrying those shared mutations (Figure \ref{fig:overview_sc2ts}D). The patterns of
@@
 (e.g., resolving reversions beyond those on immediately adjacent edges)
-would likely be of benefit .
+would likely be of benefit.
 

### 2023-06-08 paper.tex
Various updates

@@
 There are two samples designated XP which do not descend from a recombination node, 
-and a likely
-explanation for the absence of a corresponding recombination event in the Long ARG
-is that the characteristic multibase deletion for XP
+and a likely explanation for the absence of a corresponding recombination event
+in the Long ARG is that the characteristic multibase deletion for XP
 (\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
@@
 improvements). Of the remaining X designated samples, 38 are singleton recombinants,
-descending
-from recombination nodes that are ancestral only to that sample (10 are labelled XZ; 6
-are XE;  3 each from XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1
-sample from each of XB, XM, XJ, XAF, XAH and XAJ). Such samples are likely to
+descending from recombination nodes that are ancestral only to that sample
+(10 are labelled XZ; 6 are XE;  3 each from XN and XK; 2 each from XC, XS, XV, XQ and XAB;
+and 1 sample from each of XB, XM, XJ, XAF, XAH and XAJ). Such samples are likely to
 be enriched for sequencing errors and lineage designation artefacts, and so we
-exclude them from further analysis in this section.
-
-\sally{[Delete this paragraph now?
-The new section 4.8 gets at the real issue, that the multi-step mutations common in new VoC generate false positive signals of recombination.]} The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
-1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, introduced as a
-result of adding strain
-Denmark/DCGC-281594/2021 (BA.2.9), is ancestral to more than 127K samples, 
-and is highly likely to be a spuriously inferred recombination event.
-\sally{[Replace with ``...is ancestral to more than 127K samples and XX immediate descendants.
-While such a hyper-connected node may reflect a large outbreak with little phylogenetic resolution, it may also be spurious, e.g., caused by the attraction of sequences with similar constellations of mutations or another phylogenetic artefact.'']}
-Understanding the origin of this
-and other similar high-impact, but potentially spurious, recombination
-events is an important aspect of future work.
-For simplicity, we exclude these samples from further analyses in
-this section.
+exclude them from further analysis in this section. A further 79 samples
+(XN: 53, XZ: 16, XAJ: 6, XE:1, XAD: 1, XAH: 1, XAK: 1) trace back
+to a most recent recombination node that is likely to be a false positive
+(row C in Table~\ref{tab:false_positive}, see Section~\ref{sec:false_positives}).
+For simplicity these samples are likewise excluded from further analyses.
 
@@
 The Long and Wide ARGs both contain several recombination nodes that are
-ancestral to a large number of samples\sally{.
-These inferences of recombination may well be artefacts caused by the appearance of new variants of concern carrying more than the expected number of mutations \cite{otto2021origins}}.
-Table~\ref{tab:false_positive} shows details for four recombinants
+ancestral to a large number of samples.
+These inferences of recombination may well be artefacts 
+caused by the appearance of new variants of concern carrying 
+more than the expected number of mutations~\citep{otto2021origins}.
+Table~\ref{tab:false_positive} shows details of four recombinants
 in the Long ARG that are likely false positives and have come to be
@@
 labelled A--D. 
-\sally{[DELETED SENTENCE]}%These false positives are associated with the origins of large outbreaks. 
 For each row, we show the sample's Pango lineage
@@
 is $3 \times$ number of recombinations + number of additional mutations
-(for a mismatch ratio of 3; Section~\ref{sec:ls}).
-The column shows that each of these samples was a large
+(for a mismatch ratio of $k=3$).
+This column shows that each of these samples was a large
 ``distance'' from the current ARG when it was added. The mean
@@
 are acquired in sudden jumps,
-is a signature feature of SARS-CoV-2~\citep{Corey2021-sars,otto2021origins,Nielsen2023-host}.
+is a signature feature of 
+SARS-CoV-2~\citep{Corey2021-sars,otto2021origins,Nielsen2023-host}.
 Such ``saltations'' naturally present challenges to \texttt{sc2ts} and the
 current HMM parameterisation of three mutations per recombination (mismatch
-ratio). The first few sequences from these new lineages will \sally{be a poor match} to
+ratio). The first few sequences from these new lineages will be a poor match to
 the existing ARG, and the HMM will therefore
@@
 segments. 
-\sally{[DELETED SENTENCE (TRUE WHETHER OR NOT QUALITY ISSUES OCCUR).] }%If this is combined with data quality issues and ``time travellers'' (Section~\ref{sec:filtering_time_travellers}), the early stages of the VoC emergence can be quite muddled.
 This appears to be the case for the Delta VoC,
@@
 not seen in earlier VoCs.
+The origins of the  Delta VoC are
+are illustrated in  Figure~\ref{fig:false_positives}, 
+which shows the large numbers of mutations involved,
+and the multiple closely related recombinations inferred
+before tree-like behaviour is resumed at the ancestor of 
+[FIXME what percentage?] Delta samples (node \texttt{tsk261771}).
 Similar patterns around the emergence of Delta are seen in the Wide ARG
@@
 Further study is required to systematically identify such false
-positives recombinants, and to update the topology with a
+positive recombinants, and to update the topology with a
 more parsimonious explanation of the data.
@@
 This assumption dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
-Emissions may also be a function of the nucleotide states, but in our RNA virus case, we assume that \sally{mutations occur to all possible alternate alleles} ($A,C,G,U,-$) with equal probability $\mu_\ell/5$.
-\sally{[Why is this 5?
-I wouldn't have counted mutations to the same allele as a mutation.]} This is reasonable for rapidly evolving pathogens, but we note that setting the number of alleles at site $\ell$ ($a_\ell$) to the set of observed alleles across all analysed samples ($\mu_\ell/(a_\ell-1)$) will often be more appropriate
+Emissions may also be a function of the nucleotide states, but in our RNA virus case, we assume that \sally{mutations occur to all possible alternate alleles} ($A,C,G,U,-$) with equal probability for a mutation to any other allele (including a gap in the alignment; $-$) are equally likely $\mu_\ell/4$.
+This is reasonable for rapidly evolving pathogens, but we note that setting the number of alleles at site $\ell$ ($a_\ell$) to the set of observed alleles across all analysed samples ($\mu_\ell/(a_\ell-1)$) will often be more appropriate
  (Figure~\ref{fig:ls_diagram}, blue panel).
 We use the Viterbi algorithm~\citep{Viterbi1967-ol}
-to find the most likely copying path, given a set of reference sequences, $Q$, and $E$.
+to find the most likely copying path, given $Q$, $E$, and a set of reference sequences.
 Throughout, we refer to the probabilities of mismatching to a
@@
 \end{align*}
-% where $r$ is the recombination probability between adjacent
-% variants, $\mu$ is the mismatch probability between a target sequence at a
-% variant and a particular member of the reference panel,
-where \sally{$\alpha = (1 - r(n-1)/n) $ represents the probability of a sequence not recombining at a given position with any of the other $n-1$ members of the reference panel [REVISION OK?]} and $c$ is the
-likelihood of the path up to this point, which are assumed to be
-equal by construction.
+% Note: brackets here emphasise the two different components 
+% of alpha: (1-r) = don't leave this reference member, 
+% and r/n = move out, but back to yourself.
+where $\alpha = (1 - r) + r/n$ represents the probability of a sequence not recombining at a given position with any of the other $n-1$ members of the reference panel (Figure~\ref{fig:ls_diagram}, red panel) and $c$ is the
+likelihood of the path up to this point (assumed to be
+equal by construction).
 We then
@@
 Thus, for lower MMR values ($k$ here),
-\sally{recombination is increasingly favoured over multiple mutations in a specific ancestral genome}.
-%We use an MMR value of $k=3$ in this work, as it provided results that tallied well with empirical observations.
-\sally{We use an MMR value of $k=3$ in this work,
-because of the relatively high rate of recombination relative to mutation typical of coronaviruses \cite{amoutzias2022remarkable}.}
+recombination is increasingly favoured over multiple mutations in a specific ancestral genome.
+We use an MMR value of $k=3$ in this work,
+because of the relatively high rate of recombination relative 
+to mutation typical of 
+coronaviruses~\citep{amoutzias2022remarkable}.
 Exploring different mismatch ratios and more sophisticated parameterisations
@@
 the possible missing nucleotides and use that in the determination of path
-probabilities. Another, which we carry out within \texttt{sc2ts}, is to not attempt to estimate the true nucleotide but
-rather to consider such sites as uninformative to the path probability, and
-impute missing data within that path on the fly, placing them within the emission probability transition from unknown to
-the emission probability from unknown to
-to any other state as 1.
-\sally{[Repeated clause in the last sentence?
-Also, I'm a bit unclear if this is enough.
-Let's say that the nodes were G->Unknown->G.
-The latter is given a probability of one, but is the former G->Unknown also allowed with probability one?
-Note, however, that G->G shouldn't occur with probability one but Exp[-mu/t]]}
+probabilities.
+Another, which we carry out within \texttt{sc2ts}, is to assume sites with missing nucleotides are uninformative to the path probability, and impute them as each focal sequence is analysed, on the fly.
+We achieve this by setting the emission probability from any state $(A,C,G,U,-)$ to `unknown' as 1.
+Then, after determining a most likely copying path, we can impute these missing data by simply copying nucleotides present on the most likely copying path at the sites of missing data in the focal sequence.
+Note that by iteratively imputing missing data in focal sequences in this way, the reference panel being used for any newly observed focal sequence is guaranteed to be devoid of missingness.
 
@@
 subgraph plots in Supplementary Figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, \ref{fig:pango_XD_gisaid_graph} and \ref{fig:pango_XB_gisaid_graph} are at
-\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/data/Pango_X_graph_sample_maps.txt}.
+\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/data/Subgraph_sample_mapping.txt}.
 
@@
 Detailed version of Figure~\ref{fig:pango-simple-origin-graph}A. All single nucleotide mutations are listed
-with the initial nucleotide state, followed by the reference genome position, followed by the derived
-nucleotide state (after mutation). If an ancestral (``parent'') mutation exists at the same site in the path
-to the root, it is possible for a mutation to be a ``reversion'' to the initial state of the parent mutation:
-such reversions are indicated by lowercase nucleotide letters. Sample nodes are shown with \texttt{tskit} IDs,
-which can be mapped to GISAID EPI ISL identifiers and strain names using supplementary file \protect\path{Pango_X_graph_sample_maps.txt}. In contrast to
+with the inherited nucleotide state, followed by the reference genome position, followed by the derived
+nucleotide state (after mutation). Recurrent mutations (see Section~\ref{sec:args}) are highlighted in bold,
+with reversions indicated by lowercase nucleotide letters. Sample nodes are shown with \texttt{tskit} IDs,
+which can be mapped to GISAID EPI ISL identifiers and strain names using supplementary file \protect\path{Subgraph_sample_mapping.txt}. In contrast to
 Figure~\ref{fig:pango-simple-origin-graph}A, Pango lineages shown here are those assigned by GISAID rather
@@
 Detailed version of Figure~\ref{fig:pango-simple-origin-graph}B, with node and mutation labels as in
-Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file \protect\path{Pango_X_graph_sample_maps.txt}
+Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file \protect\path{Subgraph_sample_mapping.txt}
 to map tskID to EPI\_ISL and strain name).
@@
 Detailed version of Figure~\ref{fig:pango-simple-origin-graph}C, with node and mutation labels as in
-Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file \protect\path{Pango_X_graph_sample_maps.txt}
+Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file \protect\path{Subgraph_sample_mapping.txt}
 to map tskID to EPI\_ISL and strain name).
@@
 Detailed version of Figure~\ref{fig:complex_origins_graph}, with node and mutation labels as in
-Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file \protect\path{Pango_X_graph_sample_maps.txt}
+Figure~\ref{fig:pango_XA_gisaid_graph} (see supplementary file \protect\path{Subgraph_sample_mapping.txt}
 to map tskID to EPI\_ISL and strain name).
@@
 
+\begin{figure} \centering
+\includegraphics[width=0.85\textwidth]{figures/false_positive_top2_nxcld_large_graph.pdf}
+\caption{\label{fig:false_positives}
+Subgraph of the Long ARG, focussing on two likely false positive 
+recombination nodes at the start of the Delta wave 
+(``focal'' nodes, in red, corresponding to rows A and B in 
+Table~\ref{tab:false_positive}).
+The path to node \texttt{tsk261771} has been expanded: 
+this node (in gold) is a large polytomy with
+107 immediate children, and is the ancestor of the majority of Delta samples.
+The suspected-incorrect recombination path to an early B.1 MRCA is also shown.
+Note the large amount of additional recombination (black nodes) among close 
+descendants of the focal nodes.
+% Why say this here and not at for the other subgraph figs?
+%See supplementary file \protect\path{Subgraph_sample_mapping.txt} to  map tskID to EPI\_ISL and strain name.
+}
+\end{figure}
+
 \end{document}

### 2023-06-08 paper.tex
Various updates

@@
 The Li and Stephens (LS) model \citep{Li2003-ib} is an approximation of the
-coalescent with recombination~\citep{Hudson1983-properties}, which captures
+coalescent with recombination~\citep{Hudson1983-properties} which captures
 many of the important features of the joint processes of mutation and
@@
 This assumption dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
-Emissions may also be a function of the nucleotide states, but in our RNA virus case, we assume that \sally{mutations occur to all possible alternate alleles} ($A,C,G,U,-$) with equal probability for a mutation to any other allele (including a gap in the alignment; $-$) are equally likely $\mu_\ell/4$.
-This is reasonable for rapidly evolving pathogens, but we note that setting the number of alleles at site $\ell$ ($a_\ell$) to the set of observed alleles across all analysed samples ($\mu_\ell/(a_\ell-1)$) will often be more appropriate
+Emissions may also be a function of the nucleotide states, 
+but in our RNA virus case
+we assume that mutations occur from all possible alleles 
+($A,C,G,U$ and a gap in the alignment, $-$) to any other with
+equal probability $\mu_\ell/4$.
+This is reasonable for rapidly evolving pathogens, 
+but we note that setting the number of alleles at site $\ell$ ($a_\ell$) 
+to the set of observed alleles across all analysed 
+samples ($\mu_\ell/(a_\ell-1)$) will often be more appropriate 
 (Figure~\ref{fig:ls_diagram}, blue panel).
@@
 region.
-Now, follow the probability of tracking along each of those paths:
+The probability of tracking along each of those paths is given by:
 \begin{align*}
@@
 The path to node \texttt{tsk261771} has been expanded: 
-this node (in gold) is a large polytomy with
-107 immediate children, and is the ancestor of the majority of Delta samples.
+this node (in gold) is the ancestor of ${\sim}90\%$ of Delta samples in the Long ARG,
+and represents a large polytomy with 107 immediate children.
 The suspected-incorrect recombination path to an early B.1 MRCA is also shown.

### 2023-06-08 plot.py
Add summaries of "Majority Delta" node

@@ -1696,7 +1696,7 @@ class LongTopTwoFalsePositiveGraph(Pango_X_tight_graph):
             cls.focal_nodes[0]: "Focal recombination node",
             "Unknown (R)": "Additional recombination node",
             "B.1.617.2": "Nextclade Pango Delta VoC",
-            cls.main_delta_source: "Large Delta polytomy (>100 children)",
+            cls.main_delta_source: "Majority Delta ancestor\n(root of 90% of Delta samples, >100 children)",
             "Unknown": "Unimputed Pango lineage",
         }
         ax.legend(

### 2023-06-08 paper.tex
Fix spaces
Duncan's updates

@@
 % and r/n = move out, but back to yourself.
-where $\alpha = (1 - r) + r/n$ represents the probability of a sequence not recombining at a given position with any of the other $n-1$ members of the reference panel (Figure~\ref{fig:ls_diagram}, red panel) and $c$ is the
+where $\alpha = (1 - r) + r/n$ represents the probability of a sequence not recombining at a
+given position with any of the other $n-1$ members of the reference panel
+(Figure~\ref{fig:ls_diagram}, red panel) and $c$ is the
 likelihood of the path up to this point (assumed to be
@@
 A varying degree of missingness is present in the genetic data, that is, for a
-given sample, a subset of nucleotides may be coded as `unknown' at a set of positions.
-As a result, we had to consider missingness in our implementation of the LS
-Viterbi algorithm. One approach would be to `impute' all missing data
+given sample, a subset of nucleotides may be coded as `unknown' at a set of
+positions. As a result, we had to consider missingness in our implementation of
+the LS Viterbi algorithm.
+
+To deal with this missingness, we could explicitly model the missingness structure and generative process, or impute the missing data and revert back to our model without missingness.
+We choose the latter.
+To do this, we estimate a most likely path through the data assuming that missing nucleotides in the focal sequence are uninformative to the probability of any given path through the reference data.
+We then simply fill in the missing data in the focal sequence by copying it from the relevant positions in the most likely path.
+To ensure that missing data is not considered in determining the most likely copying path, we set the emission probability from any state $(A,C,G,U,-)$ to `unknown' as
+1 (though we could have chosen any constant). As a result, emissions of unknown nucleotides will not contribute to differences in path probabilities. Note that by
+iteratively imputing missing data in focal sequences in this way, the reference
+panel being used for any newly observed focal sequence is guaranteed to be
+devoid of missingness.
+
+One approach would be to `impute' all missing data
 beforehand using external software to hard-call or define a distribution over
 the possible missing nucleotides and use that in the determination of path
-probabilities.
-Another, which we carry out within \texttt{sc2ts}, is to assume sites with missing nucleotides are uninformative to the path probability, and impute them as each focal sequence is analysed, on the fly.
-We achieve this by setting the emission probability from any state $(A,C,G,U,-)$ to `unknown' as 1.
-Then, after determining a most likely copying path, we can impute these missing data by simply copying nucleotides present on the most likely copying path at the sites of missing data in the focal sequence.
-Note that by iteratively imputing missing data in focal sequences in this way, the reference panel being used for any newly observed focal sequence is guaranteed to be devoid of missingness.
+probabilities. Another, which we carry out within \texttt{sc2ts}, is to assume
+sites with missing nucleotides are uninformative to estimating the most likely path through the data which gives rise to the focal sequence.
+If we are able to estimate the most likely copying path among these paths, a reasonable imputation method is to fill in the missing data in the focal sequence by copying it from the relevant positions in the most likely path.
+To ensure that missing data is not considered in determining the most likely copying path, we set the emission probability from any state $(A,C,G,U,-)$ to `unknown' as
+1 (though we could have chosen any constant). As a result, emissions of unknown nucleotides will not contribute to differences in path probabilities. Note that by
+iteratively imputing missing data in focal sequences in this way, the reference
+panel being used for any newly observed focal sequence is guaranteed to be
+devoid of missingness.
 
@@
 The path to node \texttt{tsk261771} has been expanded:
-this node (in gold) is the ancestor of ${\sim}90\%$ of Delta samples in the Long ARG,
+this node (in gold) is the ancestor of ${\sim}89.8\%$ of Delta samples in the Long ARG,
 and represents a large polytomy with 107 immediate children.

### 2023-06-08 paper.tex
Partial update

@@
 devoid of missingness.
+
+Genetic data for SARS-CoV-2 contains substantial amounts of missingness,
+(nucleotides coded as `unknown' or masked out
+as described in Section~\ref{sec:data_preprocessing})
+and it is important to accounting for this missinging in a systematic way.
+In \texttt{sc2ts} we automatically impute missing data as samples are added
+into the ARG, using the LS model.
+To do this, we assume that sites with missing nucleotides are uninformative
+to the path probability, by setting the emission probability from any
+state $(A,C,G,U,-)$ to `unknown' equal to 1.
+Once the most likely copying path is determined, we then attach the
+sample to the ARG (Figure~\ref{fig:overview_sc2ts} and following section).
+For each sample attached to the ARG, we encode its nucleotide sequence by
+recoding mutations where it differs from the nucleotide sequence
+of its attachment node, but (importantly) ignoring sites with missing data.
+Thus, we impute the
+recording mutations
+
+% As a result, we had to consider missingness in our implementation of
+% A varying degree of missingness is present in the genetic data, that is, for a
+% given sample, a subset of nucleotides may be coded as `unknown' at a set of
+% positions. As a result, we had to consider missingness in our implementation of
+% the LS Viterbi algorithm. One approach would be to `impute' all missing data
+% beforehand using external software to hard-call or define a distribution over
+% the possible missing nucleotides and use that in the determination of path
+% probabilities.
+
+% Another, which we carry out within \texttt{sc2ts}, is to assume
+% sites with missing nucleotides are uninformative to the path probability, and
+% impute them as each focal sequence is analysed, on the fly. We achieve this by
+% setting the emission probability from any state $(A,C,G,U,-)$ to `unknown' as
+% 1. Then, after determining a most likely copying path, we can impute these
+% missing data by simply copying nucleotides present on the most likely copying
+% path at the sites of missing data in the focal sequence. Note that by
+% iteratively imputing missing data in focal sequences in this way, the reference
+% panel being used for any newly observed focal sequence is guaranteed to be
+% devoid of missingness.
 

### 2023-06-08 paper.tex
Pass at missingness

--- 2023-06-08 paper.tex
@@
 member of the reference panel at site $\ell$ as $\mu_\ell$,
-and the probability of recombining between two members of the reference panel between site $\ell-1$ and site $\ell$ as $r_\ell$.
-For convenience, $\mu_\ell$ is commonly referred to as the `mutation probability', but we note that this probability of
+and the probability of recombining between two members of the reference panel
+between site $\ell-1$ and site $\ell$ as $r_\ell$. For convenience, $\mu_\ell$ is commonly referred to as the `mutation probability', but we note that this probability of
 mismatch also encompasses various other error modes that result in a mismatch
@@
 
-A varying degree of missingness is present in the genetic data, that is, for a
-given sample, a subset of nucleotides may be coded as `unknown' at a set of positions.
-As a result, we had to consider missingness in our implementation of the LS
-Viterbi algorithm. One approach would be to `impute' all missing data
-beforehand using external software to hard-call or define a distribution over
-the possible missing nucleotides and use that in the determination of path
-probabilities.
-Another, which we carry out within \texttt{sc2ts}, is to assume sites with missing nucleotides are uninformative to the path probability, and impute them as each focal sequence is analysed, on the fly.
-We achieve this by setting the emission probability from any state $(A,C,G,U,-)$ to `unknown' as 1.
-Then, after determining a most likely copying path, we can impute these missing data by simply copying nucleotides present on the most likely copying path at the sites of missing data in the focal sequence.
-Note that by iteratively imputing missing data in focal sequences in this way, the reference panel being used for any newly observed focal sequence is guaranteed to be devoid of missingness.
+% A varying degree of missingness is present in the genetic data, that is, for a
+% given sample, a subset of nucleotides may be coded as `unknown' at a set of
+% positions. As a result, we had to consider missingness in our implementation of
+% the LS Viterbi algorithm.
+
+% To deal with this missingness, we could explicitly model the missingness
+% structure and generative process, or impute the missing data and revert back to
+% our model without missingness. We choose the latter. To do this, we estimate a
+% most likely path through the data assuming that missing nucleotides in the
+% focal sequence are uninformative to the probability of any given path through
+% the reference data. We then simply fill in the missing data in the focal
+% sequence by copying it from the relevant positions in the most likely path. To
+% ensure that missing data is not considered in determining the most likely
+% copying path, we set the emission probability from any state $(A,C,G,U,-)$ to
+% `unknown' as 1 (though we could have chosen any constant). As a result,
+% emissions of unknown nucleotides will not contribute to differences in path
+% probabilities. Note that by iteratively imputing missing data in focal
+% sequences in this way, the reference panel being used for any newly observed
+% focal sequence is guaranteed to be devoid of missingness.
+
+% One approach would be to `impute' all missing data beforehand using external
+% software to hard-call or define a distribution over the possible missing
+% nucleotides and use that in the determination of path probabilities. Another,
+% which we carry out within \texttt{sc2ts}, is to assume sites with missing
+% nucleotides are uninformative to estimating the most likely path through the
+% data which gives rise to the focal sequence. If we are able to estimate the
+% most likely copying path among these paths, a reasonable imputation method is
+% to fill in the missing data in the focal sequence by copying it from the
+% relevant positions in the most likely path. To ensure that missing data is
+% not considered in determining the most likely copying path, we set the
+% emission probability from any state $(A,C,G,U,-)$ to `unknown' as 1 (though
+% we could have chosen any constant). As a result, emissions of unknown
+% nucleotides will not contribute to differences in path probabilities. Note
+% that by iteratively imputing missing data in focal sequences in this way, the
+% reference panel being used for any newly observed focal sequence is
+% guaranteed to be devoid of missingness.
+
+Genetic data for SARS-CoV-2 contains substantial amounts of missingness,
+(nucleotides coded as `unknown' or masked out
+as described in Section~\ref{sec:data_preprocessing})
+and it is important to account for this missinging in a systematic way.
+In \texttt{sc2ts} we automatically impute missing data as samples are added
+into the ARG, using the LS model.
+To do this, we assume that sites with missing nucleotides are uninformative
+to the path probability, by setting the emission probability from any
+state $(A,C,G,U,-)$ to `unknown' equal to 1 (though we could have chosen any constant).
+As a result, emissions of unknown nucleotides will not contribute to
+differences in path probabilities.
+Once the most likely copying path is determined, we then attach the
+sample to the ARG (see Figure~\ref{fig:overview_sc2ts} and following subsections).
+For each newly attached sample we encode its nucleotide sequence by
+recording mutations where it differs from the nucleotide sequence
+of its parental node, but (importantly) ignoring sites with missing data
+in the new sample.
+Thus, once the newly added sequence has been attached to the ARG
+any missing data is imputed from its parental node. There is therefore
+no missing data in the ARG; all missing bases are ``hard called''
+at attachment time.
+Note that this approach is equivalent to
+using state-of-the-art imputation
+methods~\citep[e.g.][]{Browning2018-nk,Delaneau2019-wl} with a reference
+panel consisting of all sequences in the ARG,
+since these methods are also based on the LS model.
+% Could say more here, but let's leave it.
+Evaluating the accuracy of missing data imputation
+using \texttt{sc2ts} is an important facet of future work.
 

### 2023-06-08 paper.tex
Updates from Overleaf
Tried to clarify reversions

@@
 may be artefactual.
-A significant issue with the current version of \texttt{sc2ts} is an excess
-of inferred reversion mutations, with 5.99\% of all mutations in the Wide
+
+The current version of \texttt{sc2ts} infers a
+large number of ``reversion'' mutations, with 5.99\% of all mutations in the Wide
 ARG (Long ARG: 4.54\%) reverting the state change of the immediately
 ancestral mutation (see Section~\ref{sec:args} for precise definitions).
-\sally{[What confuses me is that I would think that none of these reversions would remain after the reversion push step of Figure 1D.
-Is this because ``immediately ancestral'' is incorrect or because the problem is that some of these immediate reversions are created deeper in the tree than the newly added samples (i.e., uncorrected by the reversion push).
-If the former, maybe say ``reverting the state change of a recent ancestral mutation'', if the latter, maybe say  ``(see Section~\ref{sec:args} for precise definitions, note that the reversion push algorithm of Figure 1D only eliminates reversions in the newly added portion of the tree and not reversions around earlier nodes created by this algorithm)''.]}
-The majority of these mutations are likely to be
-artefactual, and a key goal of future development is to improve
-our parsimony heuristics (Section~\ref{sec:parsimony-heuristics}) to
-address this issue.
+These are symptomatic of both data quality issues
+such as ``time travellers'' (Section~\ref{sec:filtering_time_travellers})
+and problematic sites (e.g., 28,271 as discussed in
+Section~\ref{sec:breakpoint_intervals}; see also
+Figure~\ref{fig:pango_XB_gisaid_graph} for an example of multiple
+reversions at this site), as well as indicating opportunities
+for improvement in tree building hueristics
+(Section~\ref{sec:parsimony-heuristics}).
+For example, the current ``reversion push'' operation
+only eliminates reversions in the newly added portion of the tree
+and not reversions around earlier nodes created by this algorithm.
 
@@
 as described in Section~\ref{sec:data_preprocessing})
-and it is important to account for this missinging in a systematic way.
+and it is important to account for this missingness in a systematic way.
 In \texttt{sc2ts} we automatically impute missing data as samples are added

### 2023-06-08 paper.tex

--- 2023-06-08 paper.tex
@@
 Here we define the
-interval \sally{within which a particular breakpoint may have occurred} as the genome coordinates over which the
+interval within which a particular breakpoint may have occurred
+as the genome coordinates over which the
 sequences for the left and right parent nodes are identical. The right-hand
@@
 It is important to note that the precise endpoints of intervals can be
-somewhat arbitrary, because they are defined by \sally{the sequence differences that happen to be present in the} recombinant's parents.
+somewhat arbitrary, because they are defined by the sequence differences
+that happen to be present in the recombinant's parents.
 Thus, breakpoint intervals will tend to be truncated at sites that are
@@
 homopolymeric region may be prone to sequencing errors
-and \sally{potentially} should be included in the list of ``problematic sites'' that
+and potentially should be included in the list of ``problematic sites'' that
 are excluded from analysis (Section~\ref{sec:data_preprocessing}).

### 2023-06-08 paper.tex
Updates from Overleaf

@@
 The Wide ARG is densely sampled but time-limited and
-includes 1.27 million strains collected up to June 30, 2021
+includes 1.27 million sequences collected up to June 30, 2021
 which
@@
 For each row, we show the sample's Pango lineage
-designation and the rank of that sample in the overall list of
-samples with that designation. Thus, rows C and D are the first
+designation and the temporal rank of that sample out of all
+samples with that designation. Thus, rows C and D are the earliest
 samples seen in the Long ARG from the BA.2.9 and BA.3 lineages,
-and A and B are the third and eighth samples from the B.1.617.2
-lineage (Delta VoC), respectively. Also shown is the overall ``cost'' of the Viterbi
+and A and B are respectively the third and eighth earliest samples among 12829 samples
+assigned by Nextclade to the B.1.617.2
+lineage (Delta VoC). Also shown is the overall ``cost'' of the Viterbi
 solution computed by the LS HMM (Section~\ref{sec:ls}), which
@@
 \midrule
-A&Germany/HH-RKI-I-061284/2021 & 178,405    & B.1.617.2   & 3& 17  \\
-B&India/ILSGS00961/2021        & 177,649    & B.1.617.2   & 8& 25   \\
-C&Denmark/DCGC-281594/2021     & 127,227    & BA.2.9      & 1& 32 \\
-D&USA/NJ-GBW-EWR000001/2021    & 127,230    & BA.3        & 1& 25 \\
+A&Germany/HH-RKI-I-061284/2021 & 178,405    & B.1.617.2   & 3 / 12829& 17  \\
+B&India/ILSGS00961/2021        & 177,649    & B.1.617.2   & 8 / 12829& 25   \\
+C&Denmark/DCGC-281594/2021     & 127,227    & BA.2.9      & 1 / 10897& 32 \\
+D&USA/NJ-GBW-EWR000001/2021    & 127,230    & BA.3        & 1 / 15& 25 \\
 \bottomrule
@@
 \begin{figure} \centering
-\includegraphics[width=0.5\textwidth]{figures/ls_diagram_covid.pdf}
+\includegraphics[width=0.5\textwidth]{figures/Li_S.pdf}
 \caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
@@
 this node (in gold) is the ancestor of ${\sim}89.8\%$ of Delta samples in the Long ARG,
-and represents a large polytomy with 107 immediate children.
+and represents a large polytomy with 107 immediate children. Of the remaining Delta samples, 10\% of Delta
+samples are mostly descendants of the node \texttt{tsk232088} on the far right,
+a sibling of the earliest focal recombination node.
 The suspected-incorrect recombination path to an early B.1 MRCA is also shown.

### 2023-06-08 paper.tex

@@
 lineages. Detection of new recombinant strains is challenging, with most
-methods looking for breaks between sets of mutations that characterize distinct
+methods looking for breaks between sets of mutations that characterise distinct
 lineages.
@@
 a GISAID global ``all-time'' tree from Nextstrain.
-We illustrate the backbone phylogeny by visualizing
+We illustrate the backbone phylogeny by visualising
 a single tree in the middle of the viral genome,
@@
 reversions at this site), as well as indicating opportunities
-for improvement in tree building hueristics
+for improvement in tree building heuristics
 (Section~\ref{sec:parsimony-heuristics}).
@@
 recombination event from which the majority of the corresponding recombinant samples
-descend and which agrees with the offical Pango designated parent lineages
+descend and which agrees with the official Pango designated parent lineages
 (see Table~\ref{tab:pango-recombinants}, but note that
@@
 Such ``saltations'' naturally present challenges to \texttt{sc2ts} and the
-current HMM parameterisation of three mutations per recombination (mismatch
+current HMM parameterization of three mutations per recombination (mismatch
 ratio). The first few sequences from these new lineages will be a poor match to
@@
 Nonetheless, \texttt{sc2ts} is currently ``alpha'' quality software, and
-we caution against overinterpreting current results. As we have sought
+we caution against over interpreting current results. As we have sought
 to illustrate throughout, there are some clear areas for improvement.
@@
 could be extended in order to explore the set of possible paths, or to
-quantify the uncertainty around it. Similarly, the current parameterisation
+quantify the uncertainty around it. Similarly, the current parameterization
 of the HMM with a single mismatch ratio is simplistic, and it would
@@
 coronaviruses~\citep{amoutzias2022remarkable}.
-Exploring different mismatch ratios and more sophisticated parameterisations
+Exploring different mismatch ratios and more sophisticated parameterizations
 of the HMM are important avenues for future work.
@@
 breakpoint intervals is a central part of \texttt{sc2ts}, and there are many
-ways to extend and improve. For example, the current parameterisation of using
+ways to extend and improve. For example, the current parameterization of using
 a single ``mismatch ratio'' is very simplistic (Section~\ref{sec:ls}) and likely
@@
 \caption{\label{fig:false_positives}
-Subgraph of the Long ARG, focussing on two likely false positive
+Subgraph of the Long ARG, focusing on two likely false positive
 recombination nodes at the start of the Delta wave

### 2023-06-08 paper.tex
Tweaks

@@
 before tree-like behaviour is resumed at the ancestor of
-[FIXME what percentage?] Delta samples (node \texttt{tsk261771}).
+90\% of Delta samples (node \texttt{tsk261771}).
 Similar patterns around the emergence of Delta are seen in the Wide ARG
@@
 this node (in gold) is the ancestor of ${\sim}89.8\%$ of Delta samples in the Long ARG,
-and represents a large polytomy with 107 immediate children. Of the remaining Delta samples, 10\% of Delta
-samples are mostly descendants of the node \texttt{tsk232088} on the far right,
+and represents a large polytomy with 107 immediate children. Of the remaining Delta samples, most (8.4\%)
+are descendants of the node \texttt{tsk232088} on the far right,
 a sibling of the earliest focal recombination node.

### 2023-06-08 paper.tex
Tweaks

@@
 mutational spectra (Section~\ref{sec:mutation_spectrum}),
-as well as automatically identifying the majority of known recombinant
+and automatically identify  the majority of known recombinant
 lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages})
@@
 We hope that these benefits of accurate joint estimation of
-genetic inheritance with mutation and recombination events will generate community
+genetic inheritance with mutation and recombination will generate community
 interest and development of the \texttt{sc2ts} method, and
@@
 \label{sec:backbone_phylogeny}
-%% The ARG is mostly tree-like
+% FIXME the development of thought is quite muddled here, we should rearrange
+% in later revisions. We should discuss the expected lack of recombination
+% and our likely false +s in one paragraph.
 Compared to organisms like humans that recombine in every generation,
@@
 single phylogenetic tree, particularly when looking at
-a subset of strains. We expect ARGs to be particularly tree-like
+a subset of strains. We expect ARGs to be particularly treelike
 early in the pandemic, when co-infection was less likely and divergence between
@@
 with 7 recombination nodes in the Wide ARG backbone (8 for the Long ARG
-backbone). However, the backbones are highly tree-like, as these
-recombination events involve minor, local topological rearrangements
+backbone). However, the recombination events involve minor, local topological rearrangements
 where recombination only occurs between close relatives.
-% sometimes even within a single lineage such that the parent
-% lineages above a recombination event coalesce with each other before involvement with
-% any other part of the phylogeny.
-These local reticulations could either represent
-real but previously unobserved recombination events between closely related strains
-or could be artefacts of the current \texttt{sc2ts} method and its parameterization.
+(The majority of these recombinations are likely
+false-positives, as discussed in Section~\ref{sec:false_positives}.)
 

### 2023-06-08 paper.tex
Done up to S2.6

@@
 
-Although we emphasise that further work is required to filter out spuriously inferred recombination events,
+Although further work is required to filter
+out spuriously inferred recombination events (see
+Section~\ref{sec:false_positives})
 we can draw some preliminary conclusions from Figure~\ref{fig:breakpoint-distribution}.
@@
 which often involve transcription-regulatory sequences preceding genes
-in SARS-CoV-2~\citep{Yang2021-characterizing}
+in SARS-CoV-2~\citep{Yang2021-characterizing}.
 The rate of recombination may also depend on the relative abundance
@@
 affecting the availability of templates and influencing the rate of
-homologous recombination~\citep{Kim2020-gt,Zou2021-sars}
+homologous recombination~\citep{Kim2020-gt,Zou2021-sars}.
 \sally{[Sally: KATHERINE, WHAT DO YOU THINK?  These not-full-length subgenomic RNA intermediates may provide a lot of homologous templates for some genes more than other (e.g., fig 3B here).]}

### 2023-06-08 paper.tex
Final version for Biorxiv

@@
 homologous recombination~\citep{Kim2020-gt,Zou2021-sars}.
-\sally{[Sally: KATHERINE, WHAT DO YOU THINK?  These not-full-length subgenomic RNA intermediates may provide a lot of homologous templates for some genes more than other (e.g., fig 3B here).]}
+%\sally{[Sally: KATHERINE, WHAT DO YOU THINK?  These not-full-length subgenomic RNA intermediates may provide a lot of homologous templates for some genes more than other (e.g., fig 3B here).]}
 
@@
 
-One of the simplest examples is the case of XA, corresponding to group A
-of \citet{Jackson2021-ik} as discussed above. Figure~\ref{fig:pango-simple-origin-graph}A
+One of the simplest examples is XA, corresponding to group A
+of \citet{Jackson2021-ik} as discussed in Section~\ref{sec:jackson_recombs}. 
+Figure~\ref{fig:pango-simple-origin-graph}A
 shows the exact relationships inferred by \texttt{sc2ts} as a subgraph of the Long ARG.
@@
 \begin{figure} \centering
-\includegraphics[width=0.85\textwidth]{figures/false_positive_top2_nxcld_large_graph.pdf}
+\includegraphics[width=1\textwidth]{figures/false_positive_top2_nxcld_large_graph.pdf}
 \caption{\label{fig:false_positives}

### 2023-06-08 plot.py
Put tanglegram-creating code into notebook
So that plot.py doesn't need to run dendroscope. The intermediate (sorted) files are saved in

@@ -27,15 +27,13 @@ import tqdm
 import pandas as pd
 
 import tsconvert  # Not on pip. Install with python -m pip install git+http://github.com/tskit-dev/tsconvert
+import tszip
 import sc2ts
 import sc2ts.utils
 
 import utils
 
 # Redefine the path to your local dendroscope Java app & chromium app here
-dendroscope_binary = (
-    "/Applications/Dendroscope/Dendroscope.app/Contents/MacOS/JavaApplicationStub"
-)
 chromium_binary = "/usr/local/bin/chromium"
 
 genes = {
@@ -180,10 +178,12 @@ def isomonths_iter(start):
             year += 1
 
 class Cophylogeny(Figure):
+    """
+    Plot a cophylogeny of a sc2ts tree and a nextstrain tree
+    requires .tsz files produced by running s2.2_cophylo_recombinants.ipynb
+    """
     name = None
-    pos = 14952  # Position along tree seq to plot trees: 14952 is halfway along
-    sc2ts_filename = None
-    nextstrain_ts_fn = "nextstrain_ncov_gisaid_global_all-time_timetree-2023-01-21.nex"
+    file_prefix = None
 
     # Utility functions
     @staticmethod
@@ -196,127 +196,24 @@ class Cophylogeny(Figure):
         leaves = [u for u in tree.nodes(order="minlex_postorder") if tree.is_leaf(u)]
         return {focal_tree_ts.strain(v): i for i, v in enumerate(leaves)}
 
-    @staticmethod
-    def run_nnet_untangle(trees):
-        assert len(trees) == 2
-        with tempfile.TemporaryDirectory() as tmpdirname:
-            newick_path = os.path.join(tmpdirname, "cophylo.nwk")
-            command_path = os.path.join(tmpdirname, "commands.txt")
-            with open(newick_path, "wt") as file:
-                for tree in trees:
-                    print(tree.as_newick(), file=file)
-            with open(command_path, "wt") as file:
-                print(f"open file='{newick_path}';", file=file)
-                print("compute tanglegram method=nnet", file=file)
-                print(
-                    f"save format=newick file='{newick_path}'", file=file
-                )  # overwrite
-                print("quit;", file=file)
-            subprocess.run([dendroscope_binary, "-g", "-c", command_path])
-            order = []
[…]

### 2025-08-11 latex/main.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%  Welcome to the Cell Press LaTeX template,     
%%%  version 1.10. This is a minimalist template    
%%%  to help you organize your article for            
%%%  publication at Cell Press. PLEASE NOTE:

%%%  (1) If you submit your final manuscript materials 
%%%  in LaTeX format, our typesetters will prepare 
%%%  a Word file for use in production. This conversion 
%%%  process allows us to copyedit your paper, fix 
%%%  any typos, and add formatting and tagging. The 
%%%  conversion process will add approximately 3 
%%%  business days to the production timeline. 
%%%  Authors using LaTeX should keep this in mind 
%%%  when considering deadlines.

%%%  (2) Keep your LaTeX files as simple as possible. 
%%%  Avoid the use of elaborate local macros and/or 
%%%  customized external style files. If you need 
%%%  additional macros, please keep them simple and 
%%%  include them in the actual .tex document preamble. 
%%%  Source code should be set up so that all .sty 
%%%  and .bst files called by the main .tex file are 
%%%  in the same directory as the main .tex file.

%%%  (3) Cell Press publishes more than 40 journals, 
%%%  some of which may have different or additional 
%%%  formatting requirements not specified in this 
%%%  template. When revising your paper before 
%%%  acceptance, please review the formatting 
%%%  guidelines, including the Final Files Requirements, 
%%%  for the journal you are publishing with.

%%%  Please send feedback on this template to lshipp@cell.com.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[12pt,letterpaper]{article}
\usepackage[a4paper, total={7in, 10in}]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{graphicx}
\usepackage{helvet}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{orcidlink} 
\usepackage[super,comma,sort&compress]  
   {natbib}\bibliographystyle{numbered}
\usepackage[right]{lineno} \linenumbers

\makeatletter
\renewcommand{\maketitle}{\bgroup\setlength{\parindent}{0pt}
\begin{flushleft}
  \textbf{\@title}

  \@author
\end{flushleft}\egroup}
\makeatother

%%%  Insert title below; leave date empty.

\title{Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time}
\date{}

%%%  Author first and last names should be spelled 
%%%  out in their entirety (do not abbreviate "J.H. 
%%%  Watson" unless this is how the author's name 
%%%  always appears). Middle initials are OK. Do 
%%%  not include titles, positions, or degrees.

%%%  Use numbered footnotes to indicate institutional 
%%%  affiliations. Authors may have multiple 
%%%  institutional affiliations, and affiliations 
%%%  may be shared among multiple authors.

%%%  After the institutional affiliations, numbered 
%%%  footnotes may be used to indicate an author's 
%%%  present address, equal contribution status, 
%%%  and/or senior author status.

%%%  The final numbered footnote should indicate 
%%%  which author is the lead contact (required). 
%%%  One author must be designated as the lead contact. 
%%%  There can be no more than one lead contact.

%%%  Corresponding authors should be indicated with 
%%%  asterisks (*). Use 2 asterisks (**) for the 
%%%  second-listed corresponding author, 3 (***) for 
%%%  the third-listed, and so on. The lead contact 
%%%  must be a corresponding author. Additional 
%%%  authors may also serve as corresponding authors.

\author[1,\orcidlink{0000-0000-0000-0000}]{First Last}
\author[1,2]{First Middle Last}
\author[2,3]{First M.M. Last}
\author[3,*]{First Last, Jr.}
\author[3,4,5,**]{F. Middle Last}

%%%  Institutional affiliations should contain the 
%%%  following information at minimum: department(s)/
%%%  subunit(s), institution, city, state/region (if 
%%%  applicable), and country.

\affil[1]{University A, London SW7 2AZ, UK}
\affil[2]{Department B, University B, Toronto, ON M5S 3H6, Canada}
\affil[3]{Division C, Department C, University C, Cambridge, MA, USA}

\affil[4]{Senior author}
\affil[5]{Lead contact}

%%%  List only one email address per corresponding author.

\affil[*]{Correspondence: first.last.jr@university.edu}
\affil[**]{Correspondence: f.middle.last@university.edu}

\begin{document}

\maketitle

\section*{SUMMARY}

The summary (abstract) should consist of a single paragraph of 150 words or fewer.

\section*{KEYWORDS}

%%%  Include up to 10 keywords, separated by commas. 
%%%  Keywords entered in EM are not carried over; only 
%%%  keywords included in the main text will be used 
%%%  in the final article metadata. Please note that 
%%%  for some journals, keywords are chosen by the editors.

cell, press, LaTeX, template

\section*{INTRODUCTION}

No subheadings, please. Use of abbreviations should be kept to a minimum, and nonstandard abbreviations should be defined when first used in the text.

\section*{RESULTS}

The results may be divided with subheadings. In our view, good subheadings convey information about the findings, so we encourage you to be specific. For example, "Factor X requires factor Y to function in process Z" is better than "Analysis of factors X and Y by approach Q." We recommend that you use similar language in your figure titles for clarity and structural harmony. Please do not use numbered subheadings. For some formats, ``results and discussion" may be combined under a single heading.

\subsection*{Figures}

For initial submissions, figures and their titles and legends can be run-in with the text. For final submissions, the image files must be uploaded as separate files, one complete file per figure. The figure titles and legends should be placed toward the end of the main text, after the declaration of interests. Our typesetters will place the figures and their titles and legends appropriately based on your in-text citations.

\subsection*{Citations}

To cite/link to display items (figures and tables) and/or sections of the manuscript, simply write, for example, ``(Figure 1)" or ``see discussion." Our team will do the rest. Please note that we do not number sections or subsections.

To cite references, you may use the cite command, e.g., ``Recent articles in \textit{Matter} and \textit{Cell} \cite{wang2020conformational,king2002,cates1984statics}  have shown ..." or  ``Many interesting discoveries have been reported, \cite{radzihovsky1995new,kapoor2022leakage,hantke2018my,alcock2006} which ..."

\subsection*{Equations}

Simple formulae should appear in line with the text whenever possible. You can write inline math by enclosing it between \verb|\(| and \verb|\)|, as in this example: \(x^2 + y^2 = z^2\). You can also enclose it between dollar signs (\texttt{\$}), as in this example: $E=mc^2$.

Larger, more complex formulae may appear on a new line, either by enclosing them between \verb|\[| and \verb|\]| or by using the \verb|displaymath| environment:

\[ x^n + y^n = z^n \]

\begin{displaymath}
\sqrt{x^2+1}
\end{displaymath}

If any equations or formulae need to be referred to or cited again later in the text, use the \verb|equation| environment to number them. Later, you can cite these as "Equation 1," "Equation 2," etc.

\begin{equation}
f(x) = \sum_{i=0}^{n} \frac{a_i}{1+x}
\end{equation}

\section*{DISCUSSION}

The discussion should explain the significance of the results and place them into a broader context. Subheadings are permitted.

\subsection*{Limitations of the study}

A "limitations" or "limitations of the study" subsection in the discussion is encouraged and may be required for some journals and some article formats.

\newpage

\noindent{The following Cell Press journals require most research articles to include their methods in a \textbf{METHODS} section, which should appear after the discussion: \textit{Cell Biomaterials}, \textit{Chem}, \textit{Chem Catalysis}, \textit{Cell Reports Physical Science}, \textit{Cell Reports Sustainability}, \textit{Device}, \textit{Joule}, \textit{Matter}, \textit{Newton}, \textit{One Earth}, and \textit{Patterns}. For all other Cell Press journals, please \textbf{delete} the contents of this page from the template and refer instead to \textbf{STAR Methods} on page 9.}

\section*{METHODS}

%%%  The methods (also/formerly called "experimental 
%%%  procedures") should appear 
%%%  immediately after the discussion. Subheadings 
%%%  may be customized. Please consult your handling 
%%%  editor if you have questions about what content 
%%%  should appear here.

\subsection*{Custom methods subheading 1}
\subsection*{Custom methods subheading 2}
\subsection*{Custom methods subheading 3}
\subsection*{Custom methods subheading 4}

\newpage

%%%  The following components should appear after the 
%%%  methods. 
%%%  For journals using STAR Methods, these components 
%%%  should appear immediately after the discussion 
%%%  (after any "limitations" or "conclusions" subsection 
%%%  within the discussion).

\section*{RESOURCE AVAILABILITY}

%%%  The resource availability section is required 
%%%  for all research articles. This component 
%%%  has 3 subsections: "lead contact," "materials 
%%%  availability," and "data and code availability." 
%%%  All 3 subsections must be included, even if no 
%%%  unique materials were generated in the study. 
%%%  Do not edit or change the names of the subheadings. 
%%%  No other subheadings or text are allowed in the 
%%%  resource availability section.

\subsection*{Lead contact}

%%%  Authors are required to designate a lead contact, 
%%%  who will be responsible for communication with 
%%%  the journal before and after publication and is 
%%%  the arbiter of disputes, including concerns 
%%%  related to reagents or resource sharing. Only 
%%%  one author can be named the lead contact, and 
%%%  only the lead contact’s information may be 
%%%  provided in this section.

Requests for further information and resources should be directed to and will be fulfilled by the lead contact, Sally White (s.white@university.edu).

\subsection*{Materials availability}

%%%  This subsection must include a statement describing 
%%%  the availability of newly generated materials 
%%%  associated with the paper, including any conditions 
%%%  for access. If there are no newly generated materials 
%%%  associated with the paper, the statement should 
%%%  state this, e.g.: This study did not generate new 
%%%  materials.

Plasmids generated in this study have been deposited to [Addgene, name and catalog number or unique identifier].

\subsection*{Data and code availability}

%%%  All original research papers must include a 
%%%  comprehensive and accurate “data and code 
%%%  availability” statement within the “resource 
%%%  availability” component of the paper before it 
%%%  is accepted for publication. These statements 
%%%  are structured and consist of three bulleted 
%%%  components. Each component must be present.

\begin{itemize}
    \item Small-molecule crystallography data have been deposited at CCDC under the database identifier ABCDEF and are publicly available as of the date of publication. 1H NMR spectra have been deposited at Mendeley under the DOI 10.12345/a1b2c3d4e5f6g7.1 and are publicly available as of the date of publication. All other data reported in this paper will be shared by the lead contact upon request.
    \item All original code has been deposited at Zenodo under the DOI 10.56789/a1b2c3d4e5f6g7.1 and is publicly available as of the date of publication.\cite{georgios_rizos_2023_10253149}
    \item Any additional information required to reanalyze the data reported in this paper is available from the lead contact upon request.    
\end{itemize}

\section*{ACKNOWLEDGMENTS}

%%%  Use this section to acknowledge contributions 
%%%  from non-authors and list funding sources, 
%%%  including grant numbers.

This work was funded by [FUNDER] via grant [GRANT NO.]. The authors thank all members of the lab for their support.

\section*{AUTHOR CONTRIBUTIONS}

%%%  This component is required for most research papers. 
%%%  Mention each individual author with a statement 
%%%  outlining the contribution of each author to the work.

Conceptualization, S.C.P. and S.Y.W.; methodology, A.B., S.C.P., and S.Y.W.; investigation, M.E., A.N.V., N.A.V., S.C.P., and S.Y.W.; writing-–original draft, S.C.P. and S.Y.W.; writing-–review \& editing, S.C.P. and S.Y.W.; funding acquisition, S.C.P. and S.Y.W.; resources, M.E.V and C.K.B.; supervision, A.B., N.L.W., and A.A.D.

\section*{DECLARATION OF INTERESTS}

%%%  This component is required for all articles, even 
%%%  if the authors have no competing interests; if 
%%%  this is the case, insert "The authors declare no 
%%%  competing interests." Please refer to the 
%%%  declaration of interests policy: 
%%%  https://www.cell.com/declaration-of-interests

S.Y.W. is an employee and shareholder of COMPANY. M.E.V. is a founder of COMPANY and a member of its scientific advisory board.

\section*{DECLARATION OF GENERATIVE AI AND AI-ASSISTED TECHNOLOGIES}

%%%  If generative AI and AI-assisted technologies 
%%%  were used in the writing process, this must 
%%%  be disclosed in the paper. This declaration 
%%%  does not apply to the use of basic tools for 
%%%  checking grammar, spelling, references, etc. 
%%%  If you have nothing to disclose, please do not 
%%%  include this component.

During the preparation of this work, the author(s) used [NAME OF TOOL OR SERVICE] in order to [REASON]. After using this tool or service, the author(s) reviewed and edited the content as needed and take(s) full responsibility for the content of the publication.

\section*{SUPPLEMENTAL INFORMATION INDEX}

%%%  Supplemental information must be uploaded as 
%%%  separate files. In the main text, please list the 
%%%  files to be included in a brief index. For details, 
%%%  please review the supplemental information guidelines:

%%%  Journals with a general "methods" section: 
%%%  http://www.cell.com/supplemental-information

%%%  Journals with STAR Methods: 
%%%  https://www.cell.com/STAR-supplemental-information

\begin{description}
  \item Figures S1-S5 and their legends in a PDF
  \item Table S1. A descriptive title for an Excel file that was too large to appear in the PDF
  \item Table S2. Another descriptive title for a different Excel file
  \item Data S1. Raw data on x, y, and z
\end{description}

\newpage

\section*{MAIN FIGURE TITLES AND LEGENDS}

%%%  At final submission, figure files MUST be 
%%%  provided separately as high-resolution image 
%%%  files. All of the panels for a figure should 
%%%  be in the same file. Figures should have 
%%%  clear labels/file names (Figure 1, Figure 2, 
%%%  etc.).

%%%  Figure titles and legends should be placed 
%%%  at the end of the main text. You do not 
%%%  need to place the figures, nor their titles 
%%%  and legends, within the main text. While 
%%%  typesetting your article, our team will 
%%%  place each figure in the best location 
%%%  based on the final layout and on your 
%%%  figure citations, e.g., (Figures 1A and 1B).

%%%  Please review the figure guidelines before 
%%%  submitting your final materials: 
%%%  https://www.cell.com/figureguidelines.

\noindent\includegraphics[width=0.85\linewidth]{Figure1.jpg}

\subsection*{Figure 1. A brief title that describes the entire figure without citing specific panels}

The figure legend can be all one paragraph and describe the images (A), graphs (B), and plots (C), etc., together.
\newline
(A) Or each panel or group of panels can be described separately, as shown here and below.
\newline
(B) Graph of X, Y, and Z.
\newline
(C and D) If panels are grouped like this, please explicitly describe each panel, e.g., “Images showing SEM (C) and TEM (D).”
\newline
Please define all scale and error bars, and please review the Cell Press figure guidelines before submission: \href{https://www.cell.com/figureguidelines}{https://www.cell.com/figureguidelines}. Example figure created by Cassie Comeau, Cell Press.

\newpage

\section*{MAIN TABLES, INCLUDING TITLES AND LEGENDS}

%%%  Whenever possible, we encourage you to submit 
%%%  your main-text tables as Microsoft Word documents, 
%%%  using Word's table function. This will ensure 
%%%  the best results during conversion. Tables 
%%%  should be numbered Table 1, Table 2, etc. and 
%%%  should not include subpanels (do not use Table 1A, 
%%%  1B, etc.). Give each table a brief descriptive 
%%%  title. Table legends are optional but encouraged. 
%%%  Footnotes (superscript lowercase letters) should 
%%%  be used where necessary to indicate some feature 
%%%  of the data; please do not use bold, italic, 
%%%  colored text, or shading for this purpose. Use 
%%%  separate cells, not line breaks or spaces, for 
%%%  all discrete data elements. Small embedded 
%%%  graphics with color are OK.

%%%  Like figures, all tables must be cited within 
%%%  the main text, and our typesetting team will 
%%%  place the tables within the typeset paper at 
%%%  the appropriate locations.

\subsection*{Table 1. A table with clear organization of data}

\begin{tabular}{|l | l | l | l|} 
 \hline
 \textbf{Column 1} & \textbf{Column 2} & \textbf{Column 3} & \textbf{Column 4} \\ [1ex] 
 \hline
 Row A\textsuperscript{a} & 6 & 87,837 & 787 \\ [1ex] 
 \hline
 Row B & 7 & 78 & 5,415 \\ [1ex] 
 \hline
 Row C & 545 & 778 & 7507 \\ [1ex] 
 \hline
 Row D & 545 & 18,744 & 7,560 \\ [1ex] 
 \hline
 Row E & 88 & 788 & 6,344 \\ [1ex] 
 \hline
\end{tabular}

\bigskip

The table legend (optional) follows the table itself. The legend should be used to provide additional info that relates to the table as a whole.

\textsuperscript{a}Footnotes can be used to provide additional info on specific content within the table, such as this footnote to the first row (row A). Do not use footnotes in the table title.

\textsuperscript{b}More footnotes

\newpage

%%%  REFERENCES: As of 2023, all Cell Press journals 
%%%  use Numbered (AMA) style. We recommend placing 
%%%  your references in the included "references.bib" 
%%%  file.

\bibliography{references}

\bigskip

%%%  In your References, please include only articles 
%%%  that are published (online publication and 
%%%  preprint servers are OK). Unpublished data, 
%%%  submitted and/or accepted manuscripts, abstracts, 
%%%  and personal communications should be cited within 
%%%  the text only ("unpublished data," "data not 
%%%  shown," "Alice Smith, personal communication") 
%%%  and not included in the references list. Personal 
%%%  communication should be documented by a letter 
%%%  of permission. Whenever possible, please make 
%%%  sure your .bib file has the complete author lists 
%%%  for each item (at minimum, the first 11 authors 
%%%  listed).

\newpage

All Cell Press \textbf{life and medical science} journals, and the multidisciplinary journal \textbf{iScience}, use the \textbf{\href{https://www.cell.com/star-authors-guide}{STAR Methods}} format for reporting materials and methods. Other Cell Press journals use a non-structured \textbf{methods} format; for those journals, please \textbf{delete} the contents of this page from the template and refer instead to the \textbf{methods} template on page 3.

If you are publishing in a STAR Methods journal, please refer to the appropriate guide for authors:
\begin{itemize}
\item \textit{Cell} authors should download the guide for \textit{Cell} authors \href{https://www.cell.com/pb-assets/journals/research/cell/methods/Methods_Guide_Cell-1678470557763.pdf}{available here}
\item All other journal authors should use \href{https://www.cell.com/pb-assets/journals/research/cell/methods/Methods_Guide_general-1678470557763.pdf}{this version of the guide}
\end{itemize}

\section*{STAR METHODS}

%%%  The STAR Methods should appear in your main 
%%%  manuscript file after the figure legends, main 
%%%  table(s) and table legend(s).

\subsection*{Key resources table}

\textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}

\subsection*{Experimental model and study participant details}

%%%  Please list here under separate headings 
%%%  all the experimental models/study participants 
%%%  (animals, human participants, plants, microbe 
%%%  strains, cell lines, primary cell cultures) 
%%%  used in the study. For each model, provide 
%%%  information related to their species/strain, 
%%%  genotype, age/developmental stage, sex (and 
%%%  gender, ancestry, race, and ethnicity if 
%%%  reported for human studies), maintenance, 
%%%  and care, including institutional permission 
%%%  and oversight information for the studies 
%%%  the experimental animal/human participant 
%%%  study. The influence (or association) of sex, 
%%%  gender, or both on the results of the study 
%%%  must be reported. In cases where it cannot, 
%%%  authors should discuss this as a limitation 
%%%  to their research’s generalizability.

%%%  Please omit this component if your study does 
%%%  not use experimental models typical in the 
%%%  life sciences (e.g., if your study is 
%%%  computational or physical science research).

\subsection*{Method details}

%%%  Please provide precise details of all the 
%%%  procedures in the paper (behavioral task, 
%%%  generation of reagents, biological assays, 
%%%  modeling, etc.) such that it is clear how, 
%%%  when, where, and why procedures were 
%%%  performed. We encourage authors to provide 
%%%  information related to the experimental 
%%%  design as suggested by NIH and ARRIVE 
%%%  guidelines (e.g., information about 
%%%  replicates, randomization, blinding, sample 
%%%  size estimation, and the criteria for 
%%%  inclusion and exclusion of any data or 
%%%  subjects).

\subsection*{Quantification and statistical analysis}

%%%  Please describe here all statistical analysis 
%%%  and software used. We ask authors to indicate 
%%%  in this component where all of the statistical 
%%%  details of experiments can be found (e.g., in 
%%%  the figure legends, figures, results, etc.), 
%%%  including the statistical tests used, exact 
%%%  value of n, what n represents (e.g., number 
%%%  of animals, number of cells, etc.), definition 
%%%  of center, and dispersion and precision 
%%%  measures (e.g., mean, median, SD, SEM, confidence 
%%%  intervals). Also, please summarize in this 
%%%  component how significance was defined, the 
%%%  statistical methods used to determine strategies 
%%%  for randomization and/or stratification, sample 
%%%  size estimation, and inclusion and exclusion 
%%%  of any data or subjects, as well as any methods 
%%%  used to determine whether the data met 
%%%  assumptions of the statistical approach.

\subsection*{Additional resources}

%%%  Please provide links to websites that provide 
%%%  further information relevant to the study 
%%%  (e.g., protocol download, troubleshooting 
%%%  forum, etc.). Clinical trial registry numbers 
%%%  and links should also be placed here. Please 
%%%  briefly describe the resource and its 
%%%  relevance for the paper. Please report this 
%%%  information as: “Description: URL.”

Cell.com homepage: https://www.cell.com
\newline Templates for Cell Press authors: https://www.cell.com/article-templates

%%%  ADDITIONAL MANUSCRIPT COMPONENTS:

%%%  Depending on the journal and the article 
%%%  type, you may be asked to upload the 
%%%  following as separate files: graphical 
%%%  abstract, highlights, eTOC blurb (In 
%%%  Brief), and/or other article components 
%%%  such as a "bigger picture" statement. 
%%%  These items are typically not required for 
%%%  initial submissions. Please refer to the 
%%%  journal's website, your acceptance 
%%%  letter, and/or the Final Files 
%%%  Requirements checklist to see if any of
%%%  these items are required (at any stage).

\end{document}

### 2025-08-14 arg_postprocessing/scripts/extract_recombinants.py
Add script to get basic recombinant info

@@ -1,0 +1,24 @@
+import tszip
+import click
+import sc2ts
+import numpy as np
+import pandas as pd
+
+
+@click.command()
+@click.argument("ts")
+@click.argument("output")
+def run(ts, output):
+    ts = tszip.load(ts)
+
+    recomb_nodes = np.where(ts.nodes_flags & sc2ts.NODE_IS_RECOMBINANT > 0)[0]
+    node_mutations = np.bincount(ts.mutations_node)
+    df = pd.DataFrame(
+        {"recombinant": recomb_nodes, "mutations": node_mutations[recomb_nodes]}
+    )
+
+    df.to_csv(output, index=False)
+
+
+if __name__ == "__main__":
+    run()

### 2025-08-14 arg_postprocessing/scripts/add_pangonet_distance_to_csv.py
Update formatting

@@ -3,6 +3,7 @@ import pandas as pd
 import numpy as np
 from pangonet.pangonet import PangoNet
 
+
 def get_pangonet_distance(pango, label_1, label_2):
     label_1 = pango.compress(label_1)
     label_2 = pango.compress(label_2)
@@ -26,7 +27,6 @@ def get_pangonet_distance(pango, label_1, label_2):
     right_path = parent_paths["right"]
     assert left_path[-1] == right_path[-1]
     return len(left_path) + len(right_path) - 2
-
 
 
 @click.command()

### 2025-08-14 arg_postprocessing/scripts/get_all_nodes_fasta.py

@@ -1,4 +1,3 @@
-
 import click
 import tszip
 import tskit

### 2025-08-14 arg_postprocessing/scripts/rematch_recombinants.py
Remove old rematch script

@@ -1,89 +1,0 @@
-import datetime
-import dataclasses
-import os
-import concurrent.futures as cf
-
-import sc2ts
-import pandas as pd
-import click
-import tqdm
-
-
-@dataclasses.dataclass
-class MatchWork:
-    ts_path: str
-    ds_path: str
-    sample_id: str
-    num_mismatches: int
-    mismatch_threshold: int
-
-
-def run_match(m):
-    runs = sc2ts.run_hmm(
-        m.ds_path,
-        m.ts_path,
-        strains=[m.sample_id],
-        num_mismatches=m.num_mismatches,
-        mismatch_threshold=m.mismatch_threshold,
-        deletions_as_missing=True,
-        num_threads=0,
-        show_progress=False,
-    )
-    return runs[0]
-
-
-@click.command()
-@click.argument("ts_prefix")
-@click.argument("ds_path", type=click.Path(dir_okay=False, file_okay=True))
-@click.argument("samples_csv_path", type=click.Path(dir_okay=False, file_okay=True))
-@click.argument("output_path")
-@click.option("--num-mismatches", "-k", type=int, multiple=True)
-@click.option("--mismatch-threshold", "-m", type=int, default=100)
-@click.option("--workers", "-w", type=int, default=1)
-def run(
-    ts_prefix,
-    ds_path,
-    samples_csv_path,
-    output_path,
-    num_mismatches,
-    mismatch_threshold,
-    workers,
-):
-
-    ds = sc2ts.Dataset(ds_path, date_field="Date_tree")
-    recomb_df = pd.read_csv(samples_csv_path).set_index("sample_id")
-
-    work = []
-    for s in recomb_df.index:
-        try:
-            date = datetime.datetime.fromisoformat(ds.metadata[s]["Date_tree"])
-        except KeyError:
-            print(f"missing {s}")
-            continue
-        match_date = str(date - datetime.timedelta(days=1)).split()[0]
-        ts_path = f"{ts_prefix}{match_date}.ts"
-        if not os.path.exists(ts_path):
-            raise ValueError(f"Missing path: {ts_path}")
-        for k in num_mismatches:
-            work.append(
-                MatchWork(ts_path, ds.path, s, k, mismatch_threshold=mismatch_threshold)
-            )
-
-    # Clear the file
-    with open(output_path, "w") as f:
-        pass
-
-    with cf.ProcessPoolExecutor(workers) as executor:
-        futures = [executor.submit(run_match, w) for w in work]
[…]

### 2025-08-14 arg_postprocessing/scripts/extract_recombinants.py
WRite the script for rematching recombinants

@@ -1,23 +1,73 @@
+import concurrent.futures as cf
+import pathlib
 import tszip
 import click
 import sc2ts
 import numpy as np
-import pandas as pd
+import dataclasses
+import tqdm
+import subprocess
+import json
+
+
+@dataclasses.dataclass
+class Work:
+    pattern: str
+    node: int
+    date: str
+
+
+def worker(work):
+    cmd = (
+        f"python -m sc2ts rematch-recombinant {work.node} "
+        f"--path-pattern={work.pattern} --date={work.date} -vv"
+    )
+    out = subprocess.check_output(cmd, shell=True)
+    result = json.loads(out.decode())
+    result["recombinant"] = int(work.node)
+    return result
+
+
+def dump(json_data, path):
+    with open(path, "w") as f:
+        json.dump(json_data, f, indent=4)
 
 
 @click.command()
 @click.argument("ts")
+@click.argument("pattern")
 @click.argument("output")
-def run(ts, output):
+@click.option("--cores", type=int)
+def run(ts, pattern, output, cores):
     ts = tszip.load(ts)
 
     recomb_nodes = np.where(ts.nodes_flags & sc2ts.NODE_IS_RECOMBINANT > 0)[0]
     node_mutations = np.bincount(ts.mutations_node)
-    df = pd.DataFrame(
-        {"recombinant": recomb_nodes, "mutations": node_mutations[recomb_nodes]}
-    )
 
-    df.to_csv(output, index=False)
+    easy_work = []
+    hard_work = []
+
+    for u in recomb_nodes:
+        md = ts.node(u).metadata
+        date = md["sc2ts"]["date_added"]
+        muts = node_mutations[u]
+        if muts == 0:
+            easy_work.append(Work(pattern, u, date))
+        else:
+            hard_work.append(Work(pattern, u, date))
+
+    json_data = []
+    with cf.ProcessPoolExecutor(cores) as executor:
+        futures = [executor.submit(worker, work) for work in easy_work]
+        for future in tqdm.tqdm(cf.as_completed(futures), total=len(easy_work)):
+            result = future.result()
+            json_data.append(result)
+            dump(json_data, output)
+
+    for work in tqdm.tqdm(hard_work):
+        result = worker(work)
+        json_data.append(result)
+        dump(json_data, output)
 
 
 if __name__ == "__main__":

### 2025-08-15 latex/main.tex

@@
 \usepackage[right]{lineno} \linenumbers
+
+\graphicspath{ {../figures/} }
 
@@
 
-cell, press, LaTeX, template
+ancestral recombination graphs, recombination, SARS-CoV-2
 
@@
 
-No subheadings, please. Use of abbreviations should be kept to a minimum, and nonstandard abbreviations should be defined when first used in the text.
+%%% No subheadings, please. Use of abbreviations should be kept to a minimum, 
+%%% and nonstandard abbreviations should be defined when first used in the text.
+
+The genomic revolution has brought a paradigm shift in how we study the evolution and spread of pathogens, 
+allowing real-time tracking of pathogens over time and space. 
+During the COVID-19 pandemic, thousands of genome sequences of SARS-CoV-2 samples were generated daily, 
+quickly overwhelming typical phylogenetic approaches. 
+Although real-time phylogenetic methods now exist to cope with this flood of genomic data 
+(e.g., UShER\cite{Turakhia2021,DeMaio2023}), we lack methods to build phylogenies 
+that can integrate recombination in the face of ongoing sampling of infected cases.
+
+UShER is a maximum parsimony method developed during the pandemic to infer phylogenies 
+from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}. 
+UShER has generated phylogenies updated daily \cite{McBroome2021} 
+and proved to be a powerful resource that successfully enabled the world 
+to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}). 
+These phylogenies are used to identify and monitor 
+potentially epidemiologically important “Pango” lineages \cite{Rambaut2020,McBroome2024}, 
+which might have acquired features to become Variants of Concern (VOC), 
+for example, increased transmissibility, disease severity, and immune escape.
+Beyond supporting real-time genomic epidemiology, 
+UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2 
+(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}). 
+UShER uses a tree to represent evolutionary change via mutation, assuming no recombination \cite{Turakhia2021}. 
+Given extensive evidence for recombination in SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2022}, 
+the tree-like representation of UShER is no longer tenable.
+
+Ancestral Recombination Graphs (ARGs) are reticulate genealogies that incorporate both mutation and recombination events \cite{Wong2024}, 
+potentially allowing more comprehensive evolutionary representations of SARS-CoV-2 and other viruses undergoing recombination. 
+Until recently, ARGs lacked software support and sufficiently scalable inference methods to be of practical use. 
+Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023}, 
+with high levels of recombination. 
+These advances now enable evolutionary genealogies to be built quickly and at scale, 
+and there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.
+
+Previous ARG methods are retrospective \cite{Rasmussen2014,Kelleher2019,Speidel2019,Ignatieva2022,Zhang2023}, 
+reconstructing the past from a sample of genome sequences drawn at one moment in time, 
+but scientists are increasingly tracking evolutionary change as it happens, accumulating data forward in time. 
+For example, sequences from daily samples of SARS-CoV-2 were made globally available on a daily basis, 
+leading to over 17 million sequences in the GISAID database \cite{Shu2017}.
+
+Here, we present 
+(1) sc2ts, a new method to infer ARGs in real time by building genealogies forward in time, and 
+(2) an ARG for SARS-CoV-2 with ~2.48 million genome sequences. We used the “Viridian” dataset, 
+which is purged of various systematic errors and artifacts \cite{Hunt2024} 
+and show that this ARG is highly congruent to an UShER phylogeny built from the same dataset. 
+Compared to previously identified recombinants, we show that this ARG contains the vast majority of 
+epidemiologically relevant signals of recombination, while adding significant nuance and texture. 
+This ARG – which captures the major signals of evolution 
+through phylogenetic relationships, point mutations, deletions, and recombination – 
+is a resource of unprecedented detail and precision, 
+and should provide a strong basis for future studies of SARS-CoV-2.
 
@@
 
-The results may be divided with subheadings.
-In our view, good subheadings convey information about the findings, so we encourage you to be specific.
-For example, "Factor X requires factor Y to function in process Z" is better than "Analysis of factors X and Y by approach Q." We recommend that you use similar language in your figure titles for clarity and structural harmony.
-Please do not use numbered subheadings.
-For some formats, ``results and discussion" may be combined under a single heading.
+%%% The results may be divided with subheadings. 
+%%% In our view, good subheadings convey information about the findings, 
+%%% so we encourage you to be specific. For example, "Factor X requires factor Y to function in process Z" 
+%%% is better than "Analysis of factors X and Y by approach Q." 
+%%% We recommend that you use similar language in your figure titles for clarity and structural harmony. 
+%%% Please do not use numbered subheadings. For some formats, ``results and discussion" may be combined under a single heading.
 
-\subsection*{Figures}
-
-For initial submissions, figures and their titles and legends can be run-in with the text.
-For final submissions, the image files must be uploaded as separate files, one complete file per figure.
+%%% For initial submissions, figures and their titles and legends can be run-in with the text.
+%%% For final submissions, the image files must be uploaded as separate files, one complete file per figure.
 The figure titles and legends should be placed toward the end of the main text, after the declaration of interests.
@@
 
-\subsection*{Citations}
+%%% To cite/link to display items (figures and tables) and/or sections of the manuscript, 
+%%% simply write, for example, ``(Figure 1)" or ``see discussion." 
+%%% Our team will do the rest. Please note that we do not number sections or subsections.
 
-To cite/link to display items (figures and tables) and/or sections of the manuscript, simply write, for example, ``(Figure 1)" or ``see discussion." Our team will do the rest.
-Please note that we do not number sections or subsections.
+\subsection*{Brief description of sc2ts}
 
-To cite references, you may use the cite command, e.g., ``Recent articles in \textit{Matter} and \textit{Cell} \cite{wang2020conformational,king2002,cates1984statics}  have shown ..." or  ``Many interesting discoveries have been reported, \cite{radzihovsky1995new,kapoor2022leakage,hantke2018my,alcock2006} which ..."
+Sc2ts reconstructs an ARG by incrementally adding batches of sequences 
+as they become available (Figure 1A; for details, see STAR Methods). 
+First, sc2ts infers likely “copying paths” connecting each sample in the batch to node(s) in the current ARG 
+(allowing for recombination) using the Li and Stephens (LS) model \cite{Li2003}, 
+which is a Hidden Markov Model (HMM) widely used in human genomics 
+to approximate the effects of mutation and recombination on genetic inheritance. 
+As recombination is relatively rare in viruses, 
+we would expect most samples to copy from a single parent in the ARG. 
+Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path 
+between the parents resulting in different genomic segments inherited from the different parents. 
+Next, for clusters within the batch that are inferred to copy from the same ARG node, 
+sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster. 
+Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, 
+and applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.
 
-\subsection*{Equations}
+Sc2ts uses collection dates to build a time-resolved ARG. 
+When examining preliminary ARGs, we found samples that introduced large numbers of mutations (and reversions), 
+likely due to errors in their collection dates 
+(for example, a large batch of samples dated 31 December, 2020 contains samples from Omicron BA.1, 
+which is estimated to have emerged in November, 2021; REF). 
+To preclude adding such “time travelling” samples, we filtered samples by “HMM cost”. 
+We define the “HMM cost” of a sample to be the total penalty cost 
+for switches and mismatches (mutations) in its copying path. 
+We parameterise the LS model such that each mismatch has a penalty cost of one 
+and a switch has a penalty cost of \textit{k} (here, set to 4), 
+which determines how much mutation is favored over recombination when accounting for a sample (Figure 1B). 
+If the HMM cost of a sample exceeds a threshold (here, set to 7), 
+sc2ts does not immediately add the sample to the ARG.
 
-Simple formulae should appear in line with the text whenever possible.
-You can write inline math by enclosing it between \verb|\(| and \verb|\)|, as in this example: \(x^2 + y^2 = z^2\).
-You can also enclose it between dollar signs (\texttt{\$}), as in this example: $E=mc^2$.
+This approach resulted in preliminary ARGs that had noticeably fewer time travellers, 
+but we found that it also excluded the major “saltation” events that characterize many VOCs. 
+These saltation lineages are characterised by a high number of accumulated mutations, 
+and are hypothesised to emerge from chronic infections 
+during which the virus is exposed to individual-specific immune selective pressures \cite{Markov2023}. 
+Though uncommon, such lineages drove large epidemic waves during the pandemic (notably Alpha, Delta, and Omicron). 
+To distinguish time travellers with erroneous dates from true saltation lineages, 
+we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added. 
+If on subsequent days these lineages form a sufficiently large group of samples, 
+sc2ts adds the group back to the ARG. 
+This “retrospective sample grouping” successfully captured the emergence of most saltation lineages (Figure SX). 
+The emergence of the Delta, BA.1 and BA.2 lineages required some manual intervention, 
+in which specific "seed" samples were identified for unconditional inclusion 
+(for details of how seed samples were chosen, see STAR Methods). 
+First arising in geographic regions with low sampling density, 
+the initial origins of these VOC are challenging to infer (REF).
 
-Larger, more complex formulae may appear on a new line, either by enclosing them between \verb|\[| and \verb|\]| or by using the \verb|displaymath| environment:
+\noindent\includegraphics[width=0.85\linewidth]{methodology.png}
 
-\[ x^n + y^n = z^n \]
+\subsection*{Figure 1: Schematic overview of sc2ts.} 
+(A) Procedure to build an ARG by incrementally updating it with samples sorted by collection dates. 
+Sample 0 is used to initialise the ARG as the root. 
+Next, Sample 1 is attached, mapping mutations onto the new branch to explain it. 
+When Sample 4 is detected as a recombinant, 
+a node representing the recombination event is inserted with two edges connecting to its parents, 
+with Sample 4 itself attached to this node as its child. 
+The resulting ARG contains two local trees, one on each side of the inferred breakpoint. 
+(B) Copying paths for Sample 4 inferred under the LS model 
+while allowing recombination (\textit{k} = 4) or disallowing it (\textit{k} = 1000).
 
-\begin{displaymath}
-\sqrt{x^2+1}
-\end{displaymath}
+\subsection*{A comprehensive representation of SARS-CoV-2 evolution}
 
-If any equations or formulae need to be referred to or cited again later in the text, use the \verb|equation| environment to number them. Later, you can cite these as "Equation 1," "Equation 2," etc.
+Using sc2ts, we built an ARG of ~2.48 million-samples spanning the COVID-19 pandemic. 
+Inference took 21 days on a server equipped with dual EPYC (details) 
+and 500 GB of RAM (~5 CPU years; Figure SX). 
+The ARG contains 2,689,054 nodes representing the genealogical relationships 
+among 2,482,157 sampled genomes in the Viridian v04 dataset, 
+which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}. 
+Of the nodes, 929 represent recombination events, 
+of which 386 have a minimum of four nucleotides supporting the non-focal parent on both sides of the breakpoint 
+(we refer to these as “robust” recombination events, Figure S1; see Document S1), 
+reflecting the overwhelming tree-like structure of SARS-CoV-2 evolutionary history.
 
-\begin{equation}
-f(x) = \sum_{i=0}^{n} \frac{a_i}{1+x}
-\end{equation}
+The ARG represents the full evolutionary history of SARS-CoV-2 
+and is in close agreement with the state-of-the-art UShER phylogeny built using the Viridian data \cite{Hunt2024}. 
+First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024} 
+by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods). 
+Tanglegrams reveal that the ARG captures known phylogenetic relationships both within sublineages 
+and at the level of major lineages (including VOCs), 
+with good agreement between the phylogenetic bipartitions in the ARG and the UShER phylogeny (Figure S2). 
+Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian data 
+by pruning both to their intersection of 2,475,418 samples and 27,431 sites. 
+Remarkably, given the differences in inference methods and data processing pipelines, 
+they are almost identically parsimonious, 
+with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts). 
+We also compared the mutational spectra of major VOCs in the ARG 
+against the mutational spectra computed from the UShER phylogeny (STAR Methods), 
+finding a close match to previous reports \cite{Bloom2023} (Figure S3).
+
+Input genome alignments are masked in sc2ts 
+so that deletions and ambiguous bases are treated as missing data by the HMM, 
+and once the samples are incorporated into the ARG, 
+these bases are imputed to carry the values inherited by the node in question. 
+Deletions are an important part of SARS-CoV-2 evolution [cites], however, 
+and to incorporate them into the ARG we identified a set of [XXX] sites most frequently involved in deletions [cite source?] 
+and post-hoc mapped alignment data for these sites to the ARG using parsimony. 
+Although sites are treated independently in this approach, 
+the approximation successfully identified the most frequent deletions, 
+such as the recurrent 21991-21993 (see Document S1 for more details). 
+The node representing the origin of Alpha, with 290,407 descendants, 
+is correctly associated with key lineage-defining deletions: 
+11288-11296 (ORF1A del3675-3677), 
+21765-21770 (Spike del69-70), 
+21991-21993 (Spike del144), and 
+28271 (non-coding). 
+The Spike 69-70 deletion, which causes S-Gene Target Failure (REF), 
+also emerges twice in the origin of Omicron - 
+in the node leading to the BA.1 variant (342,099 descendants), 
+and the node leading to the BA.4 variant (189,746 descendants). 
+To speed up inference and to avoid issues caused by sites enriched for errors [citation?], 
+sc2ts masks out a set of 100 sites identified as the most highly homoplasic 
+by running on a subset of the data (STAR Methods). 
+Of these sites, 9 are in the set of [1XX] “problematic sites” masked by UShER [citation?]. 
+In addition to remapping the data for sites identified as being involved in deletions, 
+we added mutations for XX of these highly recurrent sites 
+that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony. 
+Each node in the ARG was then assigned to one of 2026 Pango lineages using pangolin (STAR Methods). 
+All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment 
+(comparable to the 0.27\% disagreement between pangolin-data versions 1.21 and 1.29 in the source Viridian metadata). 
+A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG, 
+increasing to 1779 (86\%) when we allow for multiple sibling origination nodes (Table SX; Document S1).
+
+\subsection*{Epidemiologically relevant recombinant lineages}
+
+As new data became available during the pandemic, 
+volunteers continually detected recombinants which may develop concerning features, 
+particularly increased transmissibility, disease severity, 
+and ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2022}). 
+The process of gathering evidence for and characterising such recombinants is manual, 
+involving visual inspection of phylogenies for unusual branches 
+caused by forcibly placing recombinant sequences onto them, 
+typically followed by visual analysis of genome sequence alignments 
+to identify parental lineages and breakpoint locations. 
+This process has resulted in the designation of recombinant Pango lineages 
+(whose aliases begin with an “X”), to which we refer as “Pango Xs”. 
+Below, we summarise our analysis of the Pango Xs in the ARG 
+(for extended analyses, see Documents S1 and S2).
+
+We find strong evidence for the recombination events leading to 17 Pango Xs in the ARG (Table 1), 
+which passed the detection-level sc2ts quality control filters (Figure S1). 
+These events are supported by high numbers of averted mutations (Table 1) 
+and high numbers of loci matching their suggested parents (see their copying patterns in Figure SX). 
+Among these Pango Xs are XA, which is the first designated Pango X showing signs of community transmission \cite{Jackson2021}, 
+and XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2022}. 
+The recombinant origin of XA is well explained in the ARG by two parents and one mutation (Figure 2A; Table 1), 
+with sc2ts correctly identifying its parental Pango lineages and breakpoint location.
+
+Furthermore, some of the Pango Xs above originate from shared recombination events in the ARG (Table 1). 
+An example of a set of “nested” closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP, 
+which descend from a single recombination event rather than five independent events in the ARG (Figure 2B). 
+These Pango Xs were designated around the same time. 
+These results highlight that the ARG can help clarify the relationships between recombinants, 
+allowing for more precise interpretation of evidence for recombination.
+
+Sc2ts identified 12 Pango Xs in the ARG as non-recombinants (Table 1), 
+on the basis of the HMM and available data. 
+For all these Pango Xs, the number of recombination-informative sites 
+fell below our threshold of detection (i.e. k = 4 in the HMM). 
+The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG 
+by at most three mutations (Table 1) strongly supports their origins as non-recombinants. 
+For example, the closely related XN and XAU can be explained 
+by two mutations and one mutation in the ARG, respectively (Figure 2C). 
+%%% TODO: Say something about XAJ and XAN.
+However, sc2ts did not detect several Pango Xs (XB, XP, and XAJ) as recombinants 
+because the HMM treated characteristic deletions as missing data.
+
+\noindent\includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}
+
+\subsection*{Figure 2: Illustrative local subgraphs in the ARG.}
+Subgraphs around the XA-associated recombination node (depicted as an encircled R) (A); 
+the recombination node associated with a group of “nested” Pango Xs (B); 
+and the MRCA of the putative non-recombinants XN and XAU (C). 
+For more detailed visualisations of these subgraphs, see Document S2.
+
+\subsection*{Table 1. Recombination events associated with Pango Xs.}
+
+\subsection*{Comparison with other recombination detection methods}
+
+We compared sc2ts against three state-of-the-art (non-phylogenetic) methods 
+that detect inter-lineage recombinants by scanning sequences for mutations associated with different Pango lineages: 
+(1) RecombinHunt \cite{Alfonsi2024}, (2) CovRecomb \cite{Li2024}, and (3) rebar (https://github.com/phac-nml/rebar).
+
+To evaluate the concordance of the methods in identifying the parental Pango lineages 
+and locating the breakpoints of the recombinants, we focus on the Pango Xs, using the data 
+from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison. 
+For RecombinHunt, we took the detection results for the Pango Xs presented in the original study; 
+for CovRecomb, we summarised the results from the supplementary data of the original study (STAR Methods); 
+and for rebar, we obtained results by using it to classify the Pango X-associated recombination events in the ARG (STAR Methods). 
+For each method, we determined whether the inferred parental lineages 
+and breakpoint intervals are concordant with those identified by the community (STAR Methods). 
+We find that the methods performed well overall, 
+with no significant difference in concordance among them (Table 1; p-value = XX, chi-square test), 
+indicating that sc2ts can characterise recombinants as well as the state-of-the-art recombination detection methods.
+
+We also compared the sensitivity of sc2ts and rebar. 
+Of all the “robust” recombination events detected by sc2ts, 
+167 (43\%) were also detected as recombinants by rebar (STAR Methods). 
+We hypothesise that this disagreement arose from differences in sensitivity between these methods. 
+Consistent with this, we find that rebar needs a higher level of divergence 
+between parents to correctly classify a sequence as recombinant, 
+as shown by a shorter time to their MRCA (Figure 3A) 
+and lower distance between their Pango identity (Figure 3B).
+
+\subsection*{Table 2. Concordance among methods in detecting Pango Xs.}
+
+\begin{tabular}{|l | l | l | l|} 
+ \hline
+ \textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bp)} \\ [1ex] 
+ \hline
+ Sc2ts & XX & XX & XX \\ [1ex] 
+ \hline
+ RecombinHunt-GISAID & XX & XX & XX \\ [1ex] 
+ \hline
+ RecombinHunt-Nextstrain & XX & XX & XX \\ [1ex] 
+ \hline
+ CovRecomb & XX & XX & XX \\ [1ex] 
+ \hline
+ Rebar & XX & XX & XX \\ [1ex] 
+ \hline
+\end{tabular}
+
+\subsection*{Quantification of evidence for recombination}
+
+%%% SHZ: We may not need this section at all, but I'm keeping it for now.
+%%% We should probably have a condensed version of the first paragraph in the Pango X section.
+%%% The second and third paragraphs should probably be removed entirely.
+
+Sc2ts provides a method to facilitate the discovery of recombination events, 
+by comparing how well a sample fits the ARG under a model allowing recombination 
+versus a model disallowing recombination, 
+as measured by the number of resulting mutations (Figure 1B). 
+The more mutations are “averted” by allowing recombination, 
+the stronger the evidence for recombination. 
+We calculated the number of averted mutations for all the “robust” recombination events (STAR Methods), 
+which include all the Pango X-associated recombination events (Table 1).
+
+Our analysis of the number of averted mutations reveals a spectrum of plausibility 
+across all the “robust” recombination events (Figure 3C). 
+Examples of highly plausible recombinants are XA and XBB (for an extended analysis, see Document S1). 
+The “copying pattern” of XA visually confirms that a high number of recombination-informative sites 
+support the suggested parents (Figure 3D). 
+XW is an example of a marginally well-supported recombinant. 
+The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent. 
+The copying patterns of all the recombination events are shown in Document S3.
+
+Our results suggest that Delta and Omicron BA.2 might have recombinant origins, 
+but we interpret these results with caution, 
+as these VOCs are undersampled around the time of their emergence (REF). 
+Consistent with this, the presence of mutations and conflicting loci matching non-focal parents 
+observed in the copying patterns of these VOCs (Figure 3D) suggests the existence of 
+many genetically distinct, unsampled ancestors. 
+Future work is needed to test whether these VOCs descended from recombination events, 
+with improved sampling if still possible.
+
+%%% TODO: Old version as placeholder. This figure is not yet finalised.
+\noindent\includegraphics[width=0.85\linewidth]{averted_mutations.png}
+
+\subsection{Figure 3: Evidence for recombination events in the ARG.} 
+The 386 “robust” recombination events detected by sc2ts are colored 
+according to their classification by rebar (recombinant: purple, non-recombinant: gray). 
+(A) Distribution of time to the MRCA of the sc2ts-identified parents. 
+(B) Distribution of the Pango lineage distance between the sc2ts-identified parents. 
+(C) Distribution of the number of mutations averted by recombination (bottom x-axis), 
+defined as the difference in the number of mutations 
+between a copying path allowing recombination (right y-axis) 
+and an alternative copying path disallowing recombination (top x-axis). 
+(D) Copying patterns of five select recombinants.
+
+\subsection*{Recombination breakpoint intervals along the genome}
+
+We explored the spatial distribution of the breakpoint locations of 
+the “robust” recombination events along the genome, 
+and quantified how precisely breakpoints were localized. 
+For each event, we computed an interval estimate of its breakpoint location, 
+using its size as a measure of precision (median, 2,207 bases; STAR Method).
+
+We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome, 
+particularly at the left boundary of the Spike gene (Figure 4A), 
+consistent with the pattern of excess observed in previous studies \cite{Turakhia2022,Li2024}. 
+While this might result from a higher recombination rate in this region, 
+it may also be due to the higher levels of polymorphism in the Spike gene, 
+allowing recombination endpoints to be estimated more precisely. 
+Reassuringly, sites having unusually high mutation counts are absent in this region (Figure 4B), 
+which would confound recombination detection.
+
+Furthermore, we find an intrinsic negative relationship 
+between breakpoint interval size and the amount of divergence between recombinant parents, 
+which is measured by time to their MRCA (Figure 4D). 
+When the two parental lineages have diverged for longer periods of time (the rightmost points), 
+more information becomes available on either side of a breakpoint to pinpoint its location, 
+resulting in shorter intervals. 
+For instance, the breakpoint intervals tend to be the smallest for events 
+involving recombination between the deeply divergent Omicron BA.1 and BA.2 sublineages 
+(the red points in Figure 4D).
+
+%%% TODO: Old version as placeholder. This figure is not yet finalised.
+\noindent\includegraphics[width=0.85\linewidth]{recombination_landscape.png}
+
+\subsection*{Figure 4: Recombination landscape.} 
+(A) Distribution of the breakpoint intervals of the 386 “robust” recombination events along the genome 
+(annotated with gene labels on top). Each line represents an interval colored by its size. 
+(B) Number of overlapping breakpoint intervals (orange; left y-axis) 
+and number of mutations along the genome (blue; right y-axis). 
+The coordinates of the ten sites with the highest mutation counts are labelled. 
+(C) Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations. 
+(D) Relationship between breakpoint interval size and time 
+between the recombination event and the parental MRCA. 
+The black curve shows the theoretical expected relationship between these quantities, 
+estimated by minimizing the sum of squared residuals given a single parameter (the mutation rate) (STAR Methods).
+
+\subsection*{Emergence of recombinants over time}
+
+We examined the frequency of emergence of the recombinants over the time course of the COVID-19 pandemic (Figure 5A). 
+We explored two hypotheses for the timing of recombination. 
+First, we would expect recombination events to occur more often when case counts are high, 
+which increases the chance of co-infection within an individual, allowing recombination. 
+Second, we would expect recombination events to be detected more often when lineage diversity is high, 
+as this increases the chance that two sufficiently distinct viruses co-infected an individual, 
+increasing our power to detect recombination.
+
+We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020} 
+to relate the timing of recombination events with global case numbers. 
+Case numbers (\textit{n}) were binned by week from January, 2020 to March, 2023 
+and sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in Figure 5A). 
+These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset. 
+Alongside the case numbers, the black curve shows the number of recombinants inferred within each week along the ARG, 
+focusing on the 386 “robust” events. 
+Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high. 
+Overall, the number of recombinant events rose with the global number of cases 
+(Figure 5B; linear regression: 1.2 + 2.9 10^{-7} x, p = 3.2  10^{-7}). 
+Recombination events also rose with the diversity of lineages in each week 
+(Figure 5C; linear regression: 1.0 + 5.3 x, p = 1.3 10^{-7}), 
+measured as the chance that two randomly drawn viruses have different Scorpio designations 
+(H, referred to as the “expected heterozygosity”).
+
+We then asked which single measure best explains the number of recombination events inferred each week, 
+considering \textit{n}, \textit{H}, \textit{n H}, \textit{n}^2, \textit{n}^2 \textit{H}. 
+We included the squared number of cases each week thinking that it might better predict co-infections. 
+The best single predictor of recombination events involved the product of cases and expected heterozygosity (\textit{n H}), 
+with an adjusted R^2 of 25.3\%. 
+This measure of case diversity (\textit{n H}) explained 62\% more of the variation in recombination events per week 
+than the next best predictor (\textit{H}, adjusted R^2 of 15.7\%). 
+While consistent with the hypotheses that recombination events occur more often when infection rates are common 
+and are detected more often when viral diversity is high, we note several sources of uncertainty. 
+Global case numbers were underreported, particularly during the Delta and Omicron waves \cite{Wang2025}. 
+Furthermore, there may be substantial uncertainty in the timing of a recombination event, 
+particularly if these occur in chronically infected individuals (as reported by \cite{Burel2022}) 
+with substantial delays before onward cases are detected. 
+Finally, the filters that we applied removed all but the most strongly supported recombination events, 
+although similar conclusions were obtained when all 929 recombinants were considered (STAR Methods).
+
+%%% TODO: Old version as placeholder. This figure is not yet finalised.
+\noindent\includegraphics[width=0.85\linewidth]{recombinants_over_time.png}
+
+\subsection*{Figure 5: Recombination events across the pandemic.} 
+(A) Timing of the recombination events (black curve, right y-axis) 
+relative to the global case numbers (\cite{Dong2020}; left y-axis). 
+Cases are color-coded by their Scorpio labels, 
+with the key showing those groupings with ≥ 1,000 samples in the Viridian dataset. 
+Recombinants more commonly occur (B) in weeks with high case numbers 
+and (C) in weeks with extensive lineage variation, 
+as estimated by high expected heterozygosity (C); 
+solid lines show fits by linear regression, 
+allowing for non-zero intercepts to account for unknown cases 
+and heterozygosity not captured by the Scorpio labels.
+
+\subsection*{Efficient inference and analysis of a pandemic-scale ARG}
+
+The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale, 
+which overwhelmed existing computational approaches, 
+motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}. 
+Although the computational tools used in phylogenetics and population genetics have historically been largely distinct, 
+the functionality required largely overlaps. 
+Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024} 
+and the VCF Zarr data storage format \cite{Czech2025} 
+to describe relationships among virus samples, 
+building an ARG over time as data becomes available. 
+By reusing high-quality and high-performance components of these toolkits, 
+we could develop the sc2ts inference method itself 
+and perform all the pandemic-scale analyses in this article with Python code alone and no compiled languages.
+
+To illustrate the power of these libraries we provide some illustrative examples 
+in which we compare the Usher and sc2ts intersections in tskit format described earlier. 
+First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments. 
+The total aligned dataset for 1,226,587 samples and 27,431 sites shared by the sc2ts and Usher tree sequences 
+represents 31.34 GiB of nucleotide calls. 
+Of the 80,574,518 missing data calls (Ns) in the alignments, 
+sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%). 
+Additionally, 413,303 calls made use of the IUPAC uncertainty codes. 
+Of these sc2ts imputed only 77 (0.02\%) incorrectly (i.e., with a base that is not compatible with the ambiguity code). 
+UShER imputed only 4 calls from this set incorrectly, 
+reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations. 
+This computation, performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees, 
+required 30 lines of Python code and took 2m59s on an intel i7-9700 CPU, with a peak memory usage of about 4.5G 
+(see \cite{Czech2025} for further benchmarks for processing the Viridian data in VCF Zarr format). 
+Performing this analysis using matUtils would require exporting the full dataset to VCF, 
+and converting the VCF to FASTA, and then comparing the alignments one-by-one to the source FASTA alignments. 
+In contrast with matUtils and most phylogenetic software, 
+which provide a command line interface and suite of predefined operations that typically output text files, 
+tskit is a Python API (with C and Rust interfaces also supported) which has many advantages \cite{Baumdicker2022}. 
+As an illustration of the open-ended and efficient nature of the analyses this enables, 
+we computed the number of polytomies (nodes that have > 2 children) 
+and the maximum number of children per node in the Usher and sc2ts outputs. 
+In the Usher tree it took 116ms and 4 lines of code to count polytomies (171,148) 
+and compute the maximum number of children (5918). 
+In the sc2ts ARG, it took 385ms and 6 lines of code to find the maximum number of polytomies (137,001) 
+and maximum number of children (7708) across all 348 local trees.
 
@@
 
-The discussion should explain the significance of the results and place them into a broader context. Subheadings are permitted.
+%%% The discussion should explain the significance of the results 
+%%% and place them into a broader context. Subheadings are permitted.
+
+We present sc2ts, a real-time ARG inference method, 
+and applied it to build a ~2.48 million-sample pandemic-scale ARG of SARS-CoV-2. 
+By identifying specific samples as parents, 
+sc2ts enables more precise interpretation of recombinants than state-of-the-art recombination detection methods, 
+which find recombinants by matching a given sample to pre-defined population-averaged mutational profiles of Pango lineages. 
+Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution, in contrast to RIPPLES \cite{Turakhia2022} 
+which detects recombinants post hoc by searching for unusual branches in a given phylogeny, 
+inevitably resulting in an incomprehensive representation of SARS-CoV-2 evolution. 
+Combined with new methods and tools to evaluate evidence for recombination, 
+sc2ts is a significant technological advance over current methods for the discovery 
+and analysis of recombinants and efficient inference of SARS-CoV-2 genealogy.
+
+The ARG itself is a useful resource to support future studies of SARS-CoV-2. 
+Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution, 
+which are shown in different studies using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}. 
+The ARG displays remarkable phylogenetic congruence with a state-of-the-art phylogeny built using UShER. 
+Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes, 
+allowing for convenient and efficient reuse of this valuable dataset 
+for a variety of evolutionary and genetic analyses beyond those presented here. 
+We hope that the ARG will spur on new research areas, in which recombination is naturally and fully accounted for, 
+and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022} 
+has made on our understanding of human genetic ancestry and demographic past.
+
+We contribute several novel methods and concepts which may find applications beyond sc2ts. 
+First, we devise a simplified LS model, which extends the classic concept of parsimony. 
+In sc2ts, this model allows for accelerated HMM likelihood calculations, 
+enabling reconstruction of the pandemic-scale ARG presented here. 
+Existing HMMs used in statistical genetics may be modified in a similar fashion for significant speedup. 
+Second, we propose the concept of mutations averted by recombination, 
+which leads to an interpretable quantity to weigh 
+the support for the hypothesis that a given sample is a recombinant 
+against the hypothesis that it is a non-recombinant, 
+considering the sampled genetic diversity in an ARG and a specific HMM. 
+Third, we propose the concept of HMM cost to identify poorly explained samples 
+which are either too divergent or have a high number of mutations inconsistent with their collection dates. 
+Here we use HMM cost to reject time travelling samples, 
+which no existing phylogenetic method detects and handles during inference 
+(e.g., in the Viridian UShER phylogeny, parent and child nodes do not necessarily follow time order \cite{Hunt2024}), 
+resulting in a clean dataset freed of samples with erroneous collection dates that may affect downstream analyses. 
+Future work may be estimating dates for the suspected time travelling samples using the ARG provided here 
+and the sc2ts sample matching engine.
+
+Recombination is an important evolutionary force in viruses \cite{Pérez-Losada2015}, 
+but it has been understudied partly owing to a lack of suitable data models, methods, and tools \cite{Neches2020}. 
+While the methods and tools developed here are tailored for SARS-CoV-2, 
+they can be extended and adapted for other viruses.
+A major advantage of the sc2ts methods and tools is 
+that they are built on top of the mature software libraries of tskit and VCF-Zarr,
+which enable pandemic-scale analyses that can be performed on typical laptops. 
+Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem, 
+making them highly adaptable for a broad range of applications. 
+Together with the pandemic-scale ARG, we hope that sc2ts will be highly accessible to many researchers, 
+enabling and accelerating studies of SARS-CoV-2 for years to come.
 
@@
 
-A "limitations" or "limitations of the study" subsection in the discussion is encouraged and may be required for some journals and some article formats.
+%%% A "limitations" or "limitations of the study" subsection in the discussion is encouraged 
+%%% and may be required for some journals and some article formats.
+
+We close by emphasizing several current limitations and potential directions to improve sc2ts. 
+First, more stringent filters to exclude false positive recombinants during inference 
+would obviate post-hoc treatment of these recombinants. 
+Second, indels carry important phylogenetic signals, 
+for example, as lineage-defining mutations of Alpha and Delta (REF). 
+We currently treat multi-base deletions as missing data or multiple independent deletions, 
+but they are in reality single events. 
+More sophisticated treatments of multi-base indels in the HMM could help recover additional signals 
+(and avoid false signals) of recombination. 
+Third, the current HMM assumes a simplistic mutation model, 
+which does not account for documented transition-to-transversion rate bias (REF) 
+and context-dependent variation in mutation rate (REF). 
+A more realistic mutation model may help improve inference of local genetic relationships in the ARG. 
+Fourth, an inherent limitation of detecting recombination from sequence data, 
+which applies to all methods, is that it is impossible to detect events with few (or no) supporting bases, 
+as these cannot be distinguished from mutation events. 
+This limitation makes it particularly challenging to detect events near the termini of linear viral genomes. 
+Fifth, it is challenging to accurately reconstruct global events 
+when there are regions with poor sampling and jurisdictions with long delays 
+before sequences become publicly available (REF). 
+Revisiting the ARG topology periodically 
+(e.g., rewinding the ARG a month and repeating with samples that became available later) 
+and better handling of putative recombinants with suspiciously long branches 
+(e.g., Delta and Omicron BA.2) might improve the accuracy of the genealogy. 
+Finally, future developments may account for geography in the HMM, 
+favouring attachment of samples to geographically close lineages.
 
@@
 
-\noindent{The following Cell Press journals require most research articles to include their methods in a \textbf{METHODS} section, which should appear after the discussion: \textit{Cell Biomaterials}, \textit{Chem}, \textit{Chem Catalysis}, \textit{Cell Reports Physical Science}, \textit{Cell Reports Sustainability}, \textit{Device}, \textit{Joule}, \textit{Matter}, \textit{Newton}, \textit{One Earth}, and \textit{Patterns}.
-For all other Cell Press journals, please \textbf{delete} the contents of this page from the template and refer instead to \textbf{STAR Methods} on page 9.}
-
-\section*{METHODS}
-
-%%%  The methods (also/formerly called "experimental 
-%%%  procedures") should appear 
-%%%  immediately after the discussion. Subheadings 
-%%%  may be customized. Please consult your handling 
-%%%  editor if you have questions about what content 
-%%%  should appear here.
-
-\subsection*{Custom methods subheading 1}
-\subsection*{Custom methods subheading 2}
-\subsection*{Custom methods subheading 3}
-\subsection*{Custom methods subheading 4}
-
-\newpage
-
-%%%  The following components should appear after the 
-%%%  methods. 
+%%%  The following components should appear after the methods. 
 %%%  For journals using STAR Methods, these components 
@@
 
-Requests for further information and resources should be directed to and will be fulfilled by the lead contact, Sally White (s.white@university.edu).
+Requests for further information and resources should be directed to and will be fulfilled by the lead contact, XX (XX@university.edu).
 
@@
 
-Plasmids generated in this study have been deposited to [Addgene, name and catalog number or unique identifier].
+This study did not generate new materials
 
@@
 \begin{itemize}
-    \item Small-molecule crystallography data have been deposited at CCDC under the database identifier ABCDEF and are publicly available as of the date of publication.
-1H NMR spectra have been deposited at Mendeley under the DOI 10.12345/a1b2c3d4e5f6g7.1 and are publicly available as of the date of publication.
-All other data reported in this paper will be shared by the lead contact upon request.
-    \item All original code has been deposited at Zenodo under the DOI 10.56789/a1b2c3d4e5f6g7.1 and is publicly available as of the date of publication.\cite{georgios_rizos_2023_10253149}
-    \item Any additional information required to reanalyze the data reported in this paper is available from the lead contact upon request.    
+    \item The ARG and the sequence alignments have been deposited at Zenodo and are publicly available.
+    \item All original code has been deposited in GitHub and is publicly available.
+    \item Jupyter notebooks to perform the analyses in this paper are available on GitHub.
 \end{itemize}
@@
 
-Conceptualization, S.C.P.
-and S.Y.W.; methodology, A.B., S.C.P., and S.Y.W.; investigation, M.E., A.N.V., N.A.V., S.C.P., and S.Y.W.; writing-–original draft, S.C.P.
-and S.Y.W.; writing-–review \& editing, S.C.P.
-and S.Y.W.; funding acquisition, S.C.P.
-and S.Y.W.; resources, M.E.V and C.K.B.; supervision, A.B., N.L.W., and A.A.D.
+Conceptualization, S.C.P. and S.Y.W.; 
+methodology, A.B., S.C.P., and S.Y.W.; 
+investigation, M.E., A.N.V., N.A.V., S.C.P., and S.Y.W.; 
+writing-–original draft, S.C.P. and S.Y.W.; 
+writing-–review \& editing, S.C.P. and S.Y.W.; 
+funding acquisition, S.C.P. and S.Y.W.; 
+resources, M.E.V and C.K.B.; 
+supervision, A.B., N.L.W., and A.A.D.
 
@@
 
-S.Y.W. is an employee and shareholder of COMPANY. M.E.V. is a founder of COMPANY and a member of its scientific advisory board.
+S.Y.W. is an employee and shareholder of COMPANY.
 
@@
 During the preparation of this work, the author(s) used [NAME OF TOOL OR SERVICE] in order to [REASON]. 
-After using this tool or service, the author(s) reviewed and edited the content as needed and take(s) full responsibility for the content of the publication.
+After using this tool or service, the author(s) reviewed 
+and edited the content as needed and take(s) full responsibility for the content of the publication.
 
@@
 
-%%%  Journals with a general "methods" section: 
-%%%  http://www.cell.com/supplemental-information
-
 %%%  Journals with STAR Methods: 
@@
 \begin{description}
-  \item Figures S1-S5 and their legends in a PDF
+  \item Figures S1-SX and their legends in a PDF
   \item Table S1. A descriptive title for an Excel file that was too large to appear in the PDF
-  \item Table S2. Another descriptive title for a different Excel file
-  \item Data S1. Raw data on x, y, and z
 \end{description}
@@
 
-The table legend (optional) follows the table itself. The legend should be used to provide additional info that relates to the table as a whole.
+The table legend (optional) follows the table itself. 
+The legend should be used to provide additional info that relates to the table as a whole.
 
-\textsuperscript{a}Footnotes can be used to provide additional info on specific content within the table, such as this footnote to the first row (row A). Do not use footnotes in the table title.
+\textsuperscript{a}Footnotes can be used to provide additional info on specific content within the table, 
+such as this footnote to the first row (row A). Do not use footnotes in the table title.
 
@@
 
-All Cell Press \textbf{life and medical science} journals, and the multidisciplinary journal \textbf{iScience}, use the \textbf{\href{https://www.cell.com/star-authors-guide}{STAR Methods}} format for reporting materials and methods.
-Other Cell Press journals use a non-structured \textbf{methods} format; for those journals, please \textbf{delete} the contents of this page from the template and refer instead to the \textbf{methods} template on page 3.
+%%% All Cell Press \textbf{life and medical science} journals, and the multidisciplinary journal \textbf{iScience}, 
+%%% use the \textbf{\href{https://www.cell.com/star-authors-guide}{STAR Methods}} format for reporting materials and methods.
 
-If you are publishing in a STAR Methods journal, please refer to the appropriate guide for authors:
-\begin{itemize}
-\item \textit{Cell} authors should download the guide for \textit{Cell} authors \href{https://www.cell.com/pb-assets/journals/research/cell/methods/Methods_Guide_Cell-1678470557763.pdf}{available here}
-\item All other journal authors should use \href{https://www.cell.com/pb-assets/journals/research/cell/methods/Methods_Guide_general-1678470557763.pdf}{this version of the guide}
-\end{itemize}
+%%% If you are publishing in a STAR Methods journal, please refer to the appropriate guide for authors:
+%%% Cell authors should download the guide https://www.cell.com/pb-assets/journals/research/cell/methods/Methods_Guide_Cell-1678470557763.pdf.
 
@@
 \textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}
-
-\subsection*{Experimental model and study participant details}
-
-%%%  Please list here under separate headings 
-%%%  all the experimental models/study participants 
-%%%  (animals, human participants, plants, microbe 
-%%%  strains, cell lines, primary cell cultures) 
-%%%  used in the study. For each model, provide 
-%%%  information related to their species/strain, 
-%%%  genotype, age/developmental stage, sex (and 
-%%%  gender, ancestry, race, and ethnicity if 
-%%%  reported for human studies), maintenance, 
-%%%  and care, including institutional permission 
-%%%  and oversight information for the studies 
-%%%  the experimental animal/human participant 
-%%%  study. The influence (or association) of sex, 
-%%%  gender, or both on the results of the study 
-%%%  must be reported. In cases where it cannot, 
-%%%  authors should discuss this as a limitation 
-%%%  to their research’s generalizability.
-
-%%%  Please omit this component if your study does 
-%%%  not use experimental models typical in the 
-%%%  life sciences (e.g., if your study is 
-%%%  computational or physical science research).
 
@@
 %%%  subjects).
+
+\subsubsection*{Description of sc2ts}
+Sc2ts (pronounced “scoots”, optionally) 
+is a method for inferring Ancestral Recombination Graphs (ARGs) 
+from densely sampled genome sequence data in real time, 
+in which recombination occurs at a low but significant rate. 
+The basic idea is to incrementally update an ARG regularly 
+with the sequences collected since the previous batch (Figure 1A). 
+We used daily batches for SARS-CoV-2, 
+and use days as the time interval for clarity in what follows. 
+The first step is to find likely “copying paths” under the Li and Stephens (LS) model \cite{Li2003} 
+for each sequence in the daily batch to the current ARG. 
+These copying paths will mostly consist of a new sample sequence copying 
+from a node in the ARG with a small number of mutations, 
+and often many samples from a daily batch will copy from the same ARG node. 
+The second step is then to “resolve” these implied polytomies 
+by using a standard neighbor-joining algorithm \cite{Saitou1987}. 
+This greedy update strategy inevitably leads to unparsimonious topologies, 
+and the third step is to then increase the overall parsimony of the inferred ARG 
+by making some simple topological updates. 
+Recombination is inferred as an integral and ongoing part of this process, 
+requiring only a few additional steps to facilitate later analysis. 
+The result of inference for a given day is then a genealogy recording genetic inheritance 
+as well as mutation and recombination events 
+for all of the samples added to the ARG up to that day.
+
+\subsubsection*{Ancestral Recombination Graphs}
+%%% This paragraph needs to be rewritten.
+Originally introduced (Griffiths, 1991; Griffiths and Marjoram, 1997) 
+to define an alternative formulation of the coalescent with recombination (Hudson, 1983), 
+the term “Ancestral Recombination Graph” (ARG) has come to be used 
+more generally to describe not just realisations of the model, 
+but any recombinant genetic ancestry (Minichiello and Durbin, 2006). 
+While there is some subtlety in the details \cite{Wong2024}, 
+we can think of an ARG as being any graph 
+that encodes the reticulate genetic genealogy of a sample of sequences 
+under the influence of recombination.
+
+The “succinct tree sequence” (STS) is an ARG data structure 
+that is both general (in terms of the types of ancestry that can be described) 
+and computationally efficient \cite{Wong2024}. 
+Initially developed to facilitate large-scale coalescent simulations \cite{Kelleher2016}, 
+STS has been extended and applied to forward-time simulations \cite{Kelleher2018,Haller2019}, 
+calculation of population genetics statistics \cite{Ralph2020} 
+and ARG inference \cite{Kelleher2019,Wohns2022}. 
+The STS is based on a simple tabular representation, 
+which defines a set of nodes, edges, sites and mutations. 
+A node represents a particular genome, which may be an observed sample or an inferred genetic ancestor. 
+The genetic inheritance between a pair of nodes along a segment of genome 
+is defined by the edge (ℓ, r, p, c), 
+which states that child node c inherited its genome from parent node p 
+from left coordinate ℓ to right coordinate r. 
+A site defines a position on the genome and the ancestral state (allele) at that site. 
+A mutation records the site and node IDs where a mutation occurs and the derived state (allele).
+
+Given the node, edge, site and mutation tables, 
+we can efficiently construct the local genealogical trees along the genome (arising from recombination) 
+and perform a range of calculations efficiently 
+by reasoning about the differences between these local trees \cite{Kelleher2016,Ralph2020}. 
+These algorithms have led to performance increases of several orders of magnitude over 
+the state-of-the-art in a range of applications \cite{Kelleher2016,Kelleher2018,Kelleher2019,Ralph2020,Baumdicker2022}. 
+The STS encoding is also very concise, allowing, for example, 
+for millions of complete human genomes to be stored in a few Gigabytes of space \cite{Kelleher2019}, 
+and so it is particularly well suited to representing large SARS-CoV-2 genealogies.
+
+The SARS-CoV-2 ARGs inferred using sc2ts can be conveniently and efficiently analysed 
+using the mature and feature-rich tskit library \cite{Kelleher2018,Ralph2020}. 
+Tskit is widely used in population genetics, and has been incorporated into many downstream applications 
+(e.g., \cite{Haller2019,Speidel2019,TerasakiHart2021,Fan2022,Korfmann2022,Mahmoudi2022,Petr2022,Rasmussen2022,Zhang2023}).
+
+\subsubsection{Li and Stephens haplotype-copying model}
+LS model \cite{Li2003} is an approximation of the coalescent with recombination 
+that captures many of the key features of the joint processes of mutation and recombination. 
+It is widely used in a range of applications in population genetics, including ARG inference \cite{Kelleher2019,Gunnarsson2024}. 
+The LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states), 
+which are probabilistically emitted as an imperfect mosaic of a set of genomes in a reference panel (i.e., the hidden states). 
+The process of switching between the genomes in the reference panel is governed by a transition matrix, 
+and mismatches between a focal genome and a genome of the reference panel (from which it is “copied”) are permitted as emissions.
+
+In the original formulation of the LS model (see Appendix S1 in Li and Stephens, 2003), 
+the probability of switching between the hidden states (i.e., genomes in a reference panel) 
+when going from one site to the next is dependent on 
+(1) the rate of recombination between the sites and 
+(2) the number of genomes in the reference panel (\textit{n}). 
+Sc2ts uses the same formulation except that n is set to 1. 
+This special case of the LS model leads to an intuitive interpretation of the ratio 
+between the probability of switching and the probability of mismatch: 
+When considering only two copying paths, the path having k mismatches 
+and the other path having one switch but no mismatch are equally probable. 
+We call this parameter k the “recombination penalty”. 
+This parameter thus allows us to adjust the relative importance of recombination and mutation 
+when finding the best explanation for a given sample. 
+For simplicity, we assume that any single base mutation causing a mismatch is equally probable, 
+ignoring transition/transversion biases and variation of mutation rate along the genome. 
+As data sampling was dense for SARS-CoV-2, this assumption was considered tolerable.
+
+In our early attempts to build ARGs, we explored using several values of \textit{k}. 
+We ran sc2ts with \textit{k} set to 3, 4, and 5 on the samples collected in 2020 (REF). 
+We would expect that a good value for \textit{k} should lead to parsimonious ARGs 
+that capture plausible recombination events without incurring too many mutations, particularly reversions. 
+A value of \textit{k} too low could result in many artifactual recombination events, 
+whereas a value of \textit{k} too high could miss many genuine recombination events. 
+We found that ARGs built with \textit{k} set to 3 contained many dubious recombination events, 
+but ARGs built with \textit{k} set to 5 failed to include the well supported recombination events documented by Jackson et al. (2022). 
+Hence, we decided to use \textit{k} set to 4 to infer the final ARG described in this study.
+
+A Viterbi algorithm can be used to find the most likely “copying path” of a focal sample 
+through a reference panel of samples under the LS model. 
+Specifically, sc2ts uses a Viterbi algorithm modified to operate on STS, 
+which was originally developed to build genome-wide human genealogies 
+and is capable of scaling to biobank-scale datasets \cite{Kelleher2019}. 
+Using this Viterbi algorithm, sc2ts deterministically infers the copying path of a sample 
+collected on a certain date through the samples collected on earlier dates in a given ARG.
+
+\subsubsection*{Accelerating sample matching}
+The SARS-CoV-2 sequences were determined from samples collected densely in time. 
+SARS-CoV-2 phylogenies have low phylogenetic signals due to high sampling frequency 
+(such that sampling and transmission outpace nucleotide substitution). 
+Therefore, most samples can be explained by a small number of mutations. 
+Leveraging this property of the dataset, we developed a likelihood-thresholding strategy 
+to avoid needing to compute the likelihood of all possible copying paths through an ARG for every single sample.
+%%% TODO: This needs to be explained.
+
+Furthermore, to avoid performing redundant HMM matching on identical samples, new samples 
+which were exactly matched to a (sample or non-sample) node in the ARG were not added immediately. 
+Instead, the copying paths of these samples were recorded, 
+and the samples were eventually added back to the ARG in a post-processing step.
+
+\subsubsection*{Tree inference from daily sample clusters}
+
+With tens of thousands of samples being added to the ARG per day, 
+there are often clusters of hundreds of sequences attaching to the same node 
+(or more generally, the same recombinant path). 
+While some of these samples will require no extra mutations 
+(as they are identical to the attachment node), 
+in general there will be complex patterns of shared mutations among the samples 
+reflecting their evolutionary relationships. 
+A natural way to infer these within-cluster evolutionary relationships 
+is to use standard tree-building algorithms. 
+We can infer a plausible phylogeny relating the samples in a cluster 
+independently of the other samples in a daily batch 
+and then attach the phylogeny (and mutations) to the ARG at the node 
+identified by the LS model.
+
+Sc2ts builds a local phylogeny relating the samples in each cluster as follows. 
+First, it infers a tree using a standard neighbour-joining algorithm \cite{Saitou1987}. 
+Then, mutations are mapped back onto it using the Fitch-Hartigan parsimony algorithm \cite{Fitch1971,Hartigan1973}. 
+Finally, the resulting tree is added to the ARG at an attachment node 
+and the ARG is locally modified using heuristics to improve parsimony (Figure SX). 
+For a recombinant, the attachment node is the inserted (non-sample) recombination node associated with it. 
+For a non-recombinant, it is the common parent in the copying paths of the samples.
+
+\subsubsection*{Handling outlier samples}
+
+Although we expected that most samples can be explained by a small number of mutations in the ARG, 
+we observed two types of outlier samples which had unusually high numbers of mutations: 
+(1) “time travellers” and (2) “saltation” lineages. 
+Time travelling samples are caused by errors in their collection dates. 
+For example, if a sample is actually collected 10 months after its reported collection date, 
+then 20 mutations (instead of, say, 0 or 1) are needed to explain it 
+using the ARG available on its collection date, 
+assuming a substitution rate of 2 events per genome per month. 
+Samples from “saltation” lineages (particularly the major VOCs) are characterised 
+by high numbers of mutations by the time of their detection. 
+Both types of outlier samples are poorly fit to the ARG when they appear on their collection dates. 
+To handle these outlier samples, we introduce three mechanisms: 
+(1) HMM cost-based rejection, (2) retrospective sample grouping, and (3) unconditional seeding.
+
+First, we introduce a rejection mechanism to prevent samples 
+that are generally poorly explained by the ARG 
+using the LS model parameterised by a specific value for the recombination penalty (\textit{k}). 
+For each sample, an HMM cost is calculated based on the number of switches and mismatches 
+in its most likely copying path under the LS model: 
+The sum of the penalty cost for the switches (\textit{k}) 
+and the penalty cost of the mismatches (implicitly, equal to one). 
+To build the ARG described here, we used \textit{k} of 4 and a maximum HMM cost of 7. 
+This combination of parameter values rejected samples which have no switch 
+but 8 mismatches as well as samples which have one switch and 4 mismatches. 
+It is important to note that this does not necessarily mean 
+that samples with more than 8 mismatches 
+or samples with two or more switches cannot be incorporated into the ARG. 
+Such samples can be added back using a separate mechanism (see below). 
+We find that although it effectively removes time travellers with poor sample support (e.g. singletons), 
+the mechanism can filter samples too aggressively. 
+For example, in preliminary ARGs, it led to exclusion of major saltation VOCs (e.g. Alpha).
+
+Second, to prevent exclusion of the important saltation VOCs 
+(and less important but well supported poorly-fit lineages), 
+we introduce “retrospective sample grouping”. 
+The idea is to gather evidence for a lineage within a time window rather than on a single day. 
+After adding a batch of samples, the samples initially rejected on their collection dates 
+due to high HMM costs are reconsidered for attachment to the ARG if 
+(a) they are collected within a retrospective time window (the past 7 days); 
+(b) they have identical copying paths; and 
+(c) they have shared immediate reversions 
+(attaching a new sample to the top or the bottom of a long branch). 
+Sample groups satisfying these criteria are attached to the ARG if 
+(a) they have at least 10 samples; 
+(b) they are collected on at least 3 different dates 
+(to avoid adding batches of time travellers having the same collection date); and 
+(c) they have at least two shared “non-immediate” reversions 
+(so as to prevent unrelated time travellers from being grouped together). 
+Furthermore, phylogenetic tree quality metrics are used to preclude groups of samples 
+that are unlikely to be closely related. 
+Specifically, the tree of a sample group was excluded if 
+(a) it has more than two recurrent mutations or 
+(b) it has more than 5 mutations per sample on average.
+
+Retrospective sample grouping successfully captured most saltation lineages, 
+but integrating Delta and Omicron into the ARG required further manual intervention. 
+To ensure that Delta and Omicron are added, we introduce a third mechanism, “unconditional seeding”. 
+The idea is to choose suitable samples belonging to each VOC, 
+and attach them to the ARG unconditionally (i.e. without filtering using HMM cost). 
+We sought “seed” samples for Delta, Omicron BA.1, and BA.2. 
+For Delta, we used two sets of seeds: 
+(1) 8 samples assigned to B.1.617.2, and 
+(2) 9 additional samples assigned to its sibling lineage B.1.617.1. 
+Specifically, these samples are the earliest samples present in the Viridian v04 dataset 
+that were used for the designation of the VOCs. 
+For Omicron BA.1 and BA.2, we used SRR17041376 and SRR17461792, respectively. 
+We chose SRR17041376 (a sample dated 2021-11-12 from South Africa) 
+as the BA.1 seed (https://github.com/jeromekelleher/sc2ts-paper/issues/264), 
+as it could be well explained as as a non-recombinant given the ARG available on its collection date. 
+For BA.2, we chose SRR17461792 (dated 2021-11-27 from South Africa), 
+which is the earliest BA.2 sample present in the Viridian v04 dataset.
+It is dated later than SRR17041376 (the BA.1 seed), 
+allowing time for the BA.1 lineage to accumulate enough sampled descendants 
+and to get established in the ARG. 
+For the full list of seeds, 
+see https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml.
+
+In practice, it is difficult to find optimal seeds for saltation lineages, 
+which ideally would require no reversion to be added to the ARG 
+and to which sampled descendents would be added with no reversion. 
+Considering poor early sampling and the presence of time travellers for Delta, Omicron BA.1, and BA.2, 
+using suboptimal seeds is likely the best that we can do in practice. 
+The consequence is that there would be some reversions leading to the samples attached to the seed(s). 
+Nevertheless, it is a better alternative to not using seeds, 
+as there would be far more reversions (as seen in preliminary ARGs built without seeding). 
+The caveat of taking this approach is that one must be more careful 
+when analysing the samples around the seed(s) in the ARG.
+
+\subsubsection*{Data pre-processing}
+
+In this study, we used the SARS-CoV-2 sequences reassembled in a recent study \cite{Hunt2024} 
+using an amplicon-aware assembly method (named “Viridian”). 
+The Viridian sequences contain far fewer artifacts compared to their publicly deposited counterparts on GenBank. 
+Consequently, phylogenetic tree inference using the Viridian data led to more parsimonious phylogenies, 
+as seen by a large reduction in the number of likely artifactual reversions \cite{Hunt2024}.
+
+We downloaded 4,484,157 Viridian sequences obtained 
+from samples collected from January, 2020 to March, 2023 (“batch 1”). 
+We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) 
+using MAFFT v7.525 \cite{Katoh2013}, 
+with the option ‘--keeplength’ to retain the reference coordinate system, discarding insertions. 
+We converted the MAFFT alignments and all sample metadata to VCF Zarr format \cite{Czech2025}, 
+which are available for download on Zenodo (URL).
+
+Prior to ARG inference, we pre-processed the remaining sequences as follows. 
+We treated low-quality bases (represented by Ns), gaps and ambiguous characters in the sequences as missing data. 
+When encoded as missing data, the bases do not inform sample matching (see “The Li and Stephens model”).
+
+During inference, we filtered out the samples with low-quality sequences and low-precision sampling dates. 
+We kept the quality check-passing sequences used to infer the Viridian UShER phylogeny \cite{Hunt2024}. 
+We skipped low-quality sequences that had more than 500 missing sites 
+(except when building the “base” ARG, when the maximum number of allowed missing bases was set to 10,000 instead). 
+We also skipped ~40,000 samples which had a collection date of December 31, 2020. 
+The sampling frequency of this date is alarmingly high compared to other dates (at most 15,000 samples per day). 
+We suspect that this batch of samples might be unusually enriched for wrong collection dates, 
+likely due to arbitrary assignment of the originally reported collection date of “2020” to be “31-12-2020”. 
+Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer) 
+confirmed that some of the samples were not collected in 2020.
+
+Sites which are enriched for systematic sequencing errors or frequently experience homoplasy 
+can cause artifactual signals of recombination (REF). 
+The bioinformatics community has identified and curated a set of problematic sites 
+that are recommended to be excluded before phylogenetic inference (REF). 
+Here we assembled a bespoke set of such problematic sites 
+using the information in the Viridian sequences. 
+To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021, 
+without any site masking. 
+Then, we identified the top 100 sites which had the highest mutation counts in the ARG (URL). 
+Only six of these sites overlapped with the “Problematic Sites” 
+downloaded from the UCSC Genome Browser website (URL), 
+which did not have a high number of mutations in the preliminary ARG. 
+To infer the final ARG, we masked the problematic sites 
+in our custom set (https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml).
+
+\subsubsection*{Reconstructing a pandemic-scale ARG}
+%%% TODO: Much of this needs to be rewritten once the post-processing pipeline is finalised.
+
+\subsubsection*{Comparing phylogenetic backbones}
+We used Pango lineage designations to select \textit{representative} samples 
+for the majority of the 2026 Pango lineages in the ARG. 
+We first identify an "origination node" for a focal Pango lineage, 
+defined as the oldest internal node in the ARG 
+whose descendants in any tree include more than 50\% of the Pango samples of the focal type, 
+and where the internal node is itself designated as the focal Pango. 
+The representative sample is then chosen as the oldest sample under the origination node 
+whose own descendants (if any) are entirely of the focal Pango type. 
+We simplify the ARG to include only the representative samples, 
+resulting in an intermediate ARG of 1867 samples, 
+then additionally remove internal samples, 
+known recombinants (i.e. the Pango Xs) and three "singleton recombinant samples" 
+in which a single Pango lineage appears immediately below a recombination node. 
+This results in a simplified ARG of 1667 samples, each with a different Pango designation.
+
+We compare topology in the simplified ARG against the UShER tree, built from the Viridian dataset. 
+To remove time travelling samples from the UShER tree, 
+we reduce it to 2475419 samples shared with the main sc2ts ARG 
+and date the internal nodes by running tsdate, 
+with no rescaling and the root node constrained to the established date of the UShER root (2020-01-16). 
+To compare against the ARG, we simplify this UShER tree down to the 1667 representative ARG samples.
+
+The simplified ARG contains 4 single-breakpoint recombination nodes, 
+corresponding to the origins of Delta, BA.2, BA.5, and a further BA.2 recombinant 
+which we deem artifactual by our standard criteria 
+and treat as non-recombinant for the purposes of plotting 
+(the breakpoint is on the far right at position 27382, 
+and supported by only 4 distinguishing sites / 3 loci). 
+To reduce plot density, we consider only those Pango lineages with fewer than 10 samples, 
+and extract non-recombinant structures from the ARG 
+by creating a "base tree" of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, & BA.5), 
+a Delta subtree of 160 Pangos, 
+a BA.2 subtree of 151 sample (excluding BA.5 descendants) and 
+a BA.5 subtree of 251 samples. 
+Figure SX shows each of these trees compared against the equivalent dated UShER topology, 
+with the left and right parents of Delta and BA.2 highlighted in the base tree. 
+The order of tips in the two trees was chosen to reduce tangling, 
+as calculated using the neighbor-net algorithm from Dendroscope \cite{Scornavacca2011,Huson2012}, 
+with tanglegrams plotted using the SVG outputting facility built into tskit. 
+The entire pipeline is saved as a notebook at URL (?).
+
+The majority of resolvable nodes, including most high-level Scorpio groupings, 
+are shared between the ARG and the UShER tree (magenta nodes). 
+In the few cases where the trees differ, this is often in terms of degree of resolution of polytomies. 
+In these cases, sometimes the sc2ts ARG shows greater resolution, 
+as in the base of the BA.2 subtree, and 
+sometimes the UShER tree shows greater resolution, as in BA.2.9.
+
+\subsubsection*{Calculating Pango lineage distance using pangonet}
+We define a “pseudo-phylogenetic” distance to quantify the divergence between two Pango lineages. 
+The idea is to count the number of edges separating two given Pango lineages in a “pseudo-phylogeny” 
+which relates all the designated Pango lineages. 
+To construct this pseudo-phylogeny, we used pangonet, 
+which mines the information on the Pango lineages in the “alias_key.json” and “lineage_notes.txt” files 
+from the Pango designation issues (https://github.com/cov-lineages/pango-designation). 
+If two Pango lineages are siblings in the pseudo-phylogeny, 
+the distance is defined as the sum of the number of edges from each lineage to their MRCA; 
+if they are non-siblings (i.e. one Pango is a descendant of the other), 
+the distance is defined as the number of edges separating them; 
+if they are identical, the distance is defined as zero.
 

### 2025-08-15 latex/main.tex

@@
 these bases are imputed to carry the values inherited by the node in question. 
-Deletions are an important part of SARS-CoV-2 evolution [cites], however, 
-and to incorporate them into the ARG we identified a set of [XXX] sites most frequently involved in deletions [cite source?] 
+Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however, 
+and to incorporate them into the ARG we identified a set of [XXX] sites 
+most frequently involved in deletions [cite source?] 
 and post-hoc mapped alignment data for these sites to the ARG using parsimony. 
@@
 28271 (non-coding). 
-The Spike 69-70 deletion, which causes S-Gene Target Failure (REF), 
+The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021}, 
 also emerges twice in the origin of Omicron - 

### 2025-08-16 arg_postprocessing/scripts/rematch_recombinants.py
Update for latest sc2ts rematching updates

@@ -41,6 +41,10 @@ def dump(json_data, path):
 def run(ts, pattern, output, cores):
     ts = tszip.load(ts)
 
+    # First, make sure we've pushed up all unary recombinant nodes so we
+    # get accurate information on easy/hard
+    ts = sc2ts.push_up_unary_recombinant_mutations(ts)
+
     recomb_nodes = np.where(ts.nodes_flags & sc2ts.NODE_IS_RECOMBINANT > 0)[0]
     node_mutations = np.bincount(ts.mutations_node)
 
@@ -51,7 +55,7 @@ def run(ts, pattern, output, cores):
         md = ts.node(u).metadata
         date = md["sc2ts"]["date_added"]
         muts = node_mutations[u]
-        if muts == 0:
+        if muts <= 2:
             easy_work.append(Work(pattern, u, date))
         else:
             hard_work.append(Work(pattern, u, date))

### 2025-08-17 arg_postprocessing/scripts/rematch_recombinants.py
Different approach to handling parallelism for rematching

@@ -1,3 +1,5 @@
+import collections
+import random
 import concurrent.futures as cf
 import pathlib
 import tszip
@@ -37,8 +39,7 @@ def dump(json_data, path):
 @click.argument("ts")
 @click.argument("pattern")
 @click.argument("output")
-@click.option("--cores", type=int)
-def run(ts, pattern, output, cores):
+def run(ts, pattern, output):
     ts = tszip.load(ts)
 
     # First, make sure we've pushed up all unary recombinant nodes so we
@@ -48,26 +49,39 @@ def run(ts, pattern, output, cores):
     recomb_nodes = np.where(ts.nodes_flags & sc2ts.NODE_IS_RECOMBINANT > 0)[0]
     node_mutations = np.bincount(ts.mutations_node)
 
-    easy_work = []
-    hard_work = []
+    all_work = collections.defaultdict(list)
 
     for u in recomb_nodes:
         md = ts.node(u).metadata
         date = md["sc2ts"]["date_added"]
         muts = node_mutations[u]
-        if muts <= 2:
-            easy_work.append(Work(pattern, u, date))
-        else:
-            hard_work.append(Work(pattern, u, date))
+        all_work[muts].append(Work(pattern, u, date))
 
+    random.seed(42)
+    for k, v in all_work.items():
+        # Randomise so we're not doing all the later onces (which need more RAM)
+        # at the same time
+        random.shuffle(v)
+        print(f"muts: {k} has {len(v)}")
+
+    cores_for_muts = {0: 20, 1: 10, 2: 5, 3: 2}
     json_data = []
-    with cf.ProcessPoolExecutor(cores) as executor:
-        futures = [executor.submit(worker, work) for work in easy_work]
-        for future in tqdm.tqdm(cf.as_completed(futures), total=len(easy_work)):
-            result = future.result()
-            json_data.append(result)
-            dump(json_data, output)
+    for muts, cores in cores_for_muts.items():
+        easy_work = all_work[muts]
+        print(f"Running {len(easy_work)} for {muts} mutations")
+        with cf.ProcessPoolExecutor(cores) as executor:
+            futures = [executor.submit(worker, work) for work in easy_work]
+            for future in tqdm.tqdm(cf.as_completed(futures), total=len(easy_work)):
+                result = future.result()
+                json_data.append(result)
+                dump(json_data, output)
 
+    hard_work = []
+    for k, v in all_work.items():
+        if k > max(cores_for_muts.keys()):
[…]

### 2025-08-17 arg_postprocessing/scripts/rematch_recombinants.py
Latest attempt at getting to parallelise

@@ -42,50 +42,43 @@ def dump(json_data, path):
 def run(ts, pattern, output):
     ts = tszip.load(ts)
 
-    # First, make sure we've pushed up all unary recombinant nodes so we
-    # get accurate information on easy/hard
-    ts = sc2ts.push_up_unary_recombinant_mutations(ts)
+    recombinants = np.where(ts.nodes_flags & sc2ts.NODE_IS_RECOMBINANT > 0)[0]
 
-    recomb_nodes = np.where(ts.nodes_flags & sc2ts.NODE_IS_RECOMBINANT > 0)[0]
-    node_mutations = np.bincount(ts.mutations_node)
-
-    all_work = collections.defaultdict(list)
-
-    for u in recomb_nodes:
+    all_work = []
+    for u in recombinants:
         md = ts.node(u).metadata
         date = md["sc2ts"]["date_added"]
-        muts = node_mutations[u]
-        all_work[muts].append(Work(pattern, u, date))
+        all_work.append(Work(pattern, u, date))
+
+    # The Delta wave was particularly difficult in terms of memory usage,
+    # so limiting parallelism based on date. Post-Omicron wasn't quite as bad
+    bins = ["2020-01", "2021-09", "2022-01", "2024-01"]
+    cores_per_bin = [20, 4, 8]
+    assert len(cores_per_bin) == len(bins) - 1
 
     random.seed(42)
-    for k, v in all_work.items():
+
+    split_work = {}
+    for j in range(len(bins) - 1):
+        start = bins[j]
+        stop = bins[j + 1]
+        cores = cores_per_bin[j]
+        work = [w for w in all_work if start <= w.date < stop]
         # Randomise so we're not doing all the later onces (which need more RAM)
         # at the same time
-        random.shuffle(v)
-        print(f"muts: {k} has {len(v)}")
+        random.shuffle(work)
+        split_work[cores] = work
 
-    cores_for_muts = {0: 20, 1: 10, 2: 5, 3: 2}
     json_data = []
-    for muts, cores in cores_for_muts.items():
-        easy_work = all_work[muts]
-        print(f"Running {len(easy_work)} for {muts} mutations")
+    for cores in cores_per_bin:
+        work = split_work[cores]
+        print(f"Running {len(work)} on {cores}")
         with cf.ProcessPoolExecutor(cores) as executor:
-            futures = [executor.submit(worker, work) for work in easy_work]
-            for future in tqdm.tqdm(cf.as_completed(futures), total=len(easy_work)):
[…]

### 2025-08-18 latex/docS1.tex
Updates from Shing

\documentclass[12pt,letterpaper]{article}
\usepackage[a4paper, total={7in, 10in}]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{graphicx}
\usepackage{helvet}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{orcidlink} 
\usepackage[super,comma,sort&compress]  
   {natbib}\bibliographystyle{numbered}
\usepackage[right]{lineno} \linenumbers

\graphicspath{ {../figures/} }

\makeatletter
\renewcommand{\maketitle}{\bgroup\setlength{\parindent}{0pt}
\begin{flushleft}
  \textbf{\@title}

  \@author
\end{flushleft}\egroup}
\makeatother

\title{Document S1}
\date{}

\begin{document}

\maketitle

\section*{Supplementary Text}

\subsection*{Identification of robust recombination events}
Before characterizing recombination events in detail, we sought to remove probable artifacts. 
Systematic errors in SARS-CoV-2 sequences associated with sequencing protocols have been documented (Turakhia et al., 2022), 
which can confound recombination detection (De Maio et al., 2024). 
By inspecting copying patterns at sites that differ between recombinant parents, 
we find a convincing signal when treating adjacent and near-adjacent sites as a single locus. 
Artifactual recombination events can be characterized as 
having fewer than four such loci supporting the left or right parent on either side of a putative breakpoint (STAR Methods). 
Of the 929 recombination events in the ARG, this measure identifies 543 as artifactual, 
leaving 386 “robust” events (Figure S1A). 
The copying patterns depicted in Figure S1B illustrate the genomic evidence for 
three recombination events identified as artifactual, compared to a marginally robust one.

Evidence that our simple filter effectively flags artifacts can be seen in the clear association between 
the AmpliSeq V1 kit and recombination events that we identify as artifactual. 
This primer scheme was used to sequence XX of the samples associated with an artifactual recombination event, 
in contrast to 15 of the “robust” events (Figure 2A), strongly implicating it as a common source of problematic data. 
Through further inspection, we observed that many artifactual recombination events 
contain stretches of missing data characteristic of amplicon dropouts (REF).

\subsection*{Analysis of Pango lineage “origination” events}
Notebook

To analyse how well the ARG reflects the phylogenetic structure implicit in the Pango lineage naming system, 
we considered the Pango assignments generated by Pangolin on the alignments for each node in the ARG (STAR Methods). 
If the ARG perfectly reflected the Pango lineage structure, 
each lineage would form a clade descending from a single originating node with that label. 
We identified putative origination nodes among those labeled with a given lineage as the node 
with the maximum number of descendant samples and the earliest time (if there is a tie).

Of the 2058 distinct Pango lineages in the ARG, 
1469 of these (comprising 743922 samples) match perfectly, 
with unique origination events in the ARG 
where all samples assigned a given lineage descend from the first node assigned to that lineage. 
A further 310 lineages (398391 samples) match perfectly 
when we count the descendants of the parent of the first node 
(accounting for polytomies in which multiple originating nodes for a given lineage are siblings). 
We then have 255 lineages (996490 samples) 
where the difference in the number of descendants of the first node's parent is < 100. 
The remaining 25 lineages (344079) are dominated by a few large lineages such as BA.1.1 (155595 samples) and AY.4.2 (54607 samples) 
which have multiple non-sibling origins within the ARG.

\subsection*{Additional analyses of the Pango Xs}
\subsubsection*{Multiple origins}
\subsubsection*{XS}
https://github.com/jeromekelleher/sc2ts-paper/issues/287
In the ARG, the XS samples appear to have originated from two sequential recombination events. 
However, the second event is likely an artifact, 
as the XS sample associated with this event failed the minimum net number of supporting loci. 
In addition, this sample was sequenced using the AmpliSeq V1 kit, 
which was also used to sequence eight other samples descending from the same event. 
We excluded this likely artifactual event from further analysis and 
consider XS to emerge from a single recombinant origin.

\subsubsection*{XM}
https://github.com/jeromekelleher/sc2ts-paper/issues/284
Possibly related to deletions.

\subsubsection*{Discordant parents}
\subsubsection*{XBB}

\subsubsection*{XBF}

\subsubsection*{XBH}

\subsection*{Multiple breakpoints}
Seven of the Pango Xs present in the Viridian v04 dataset have more than one breakpoint, 
with the supporting evidence documented in the associated Pango designation issues. 
Five of these seven Pango Xs were detected as recombinants by sc2ts. 
Three of them (XAY, XBC, XBT) had more than one breakpoint, 
but were not added to the ARG due to high HMM cost and a lack of sample support to be incorporated as retrospective sample groups. 
However, the other two (XAC and XAK) had only one breakpoint.

\subsubsection*{XAY}
https://github.com/jeromekelleher/sc2ts-paper/issues/490

\subsubsection*{XBC}
https://github.com/jeromekelleher/sc2ts-paper/issues/486
Sc2ts detected XBC as a three-breakpoint recombinant, 
correctly identifying the parental Pango lineages. 
The inferred breakpoint intervals mostly overlap with those identified in its designation issue. 
XBC was not added to the ARG due to high HMM costs and inadequate sample support 
(within a seven-day retrospective time window).

\subsubsection*{XBT}
https://github.com/jeromekelleher/sc2ts-paper/issues/487
Sc2ts detected XBT as a two-breakpoint recombinant, 
correctly identifying the parental Pango lineages. 
The inferred breakpoint intervals are identical to those documented in its designation issue. 
XBT was not added to the ARG due to high HMM cost and low sample support 
(only one QC-passing sample in the Viridian v04 dataset).

\subsubsection*{XAC}
https://github.com/jeromekelleher/sc2ts-paper/issues/357
Evidence of the second breakpoint reported in its designation issue relies on 
the presence of a large deletion inherited from BA.2*, which is currently treated as missing data by sc2ts. 
The XAC samples share a recombination node with the samples of four other Pango Xs 
which do not have this deletion (Figure 2C).

\subsubsection*{XAK}
https://github.com/jeromekelleher/sc2ts-paper/issues/484
Our results agree with its designation issue in terms of the parental Pango lineages and breakpoint intervals. 
XAK was not added to the ARG due to HMM costs (9 and 10) and insufficient sample support (only two samples). 
Interestingly, our results suggest that XAK might be a product of two sequential recombination events.

\subsubsection*{Non-recombinants}
\subsubsection*{XN and XAU}
https://github.com/jeromekelleher/sc2ts-paper/issues/358
https://github.com/jeromekelleher/sc2ts-paper/issues/348
Both of these putative recombinants have been defined on the basis of a RHS from BA.2 and a LHS from BA.1. 
In the case of XN, the supposed BA.1 mutations are [C241T, A2832G, C3037T]. 
In the case of XAU, this set also includes C2470T, so [C241T, C2470T, A2832G, C3037T]. 
However, both BA.1 and BA.2 are now defined as also having C241T and C3037T. 
In addition, a very large number of BA.2 samples also have A2832G, and three BA.2-designated samples also have C2470T. 
This means that there is essentially no longer any evidence for a BA.1 parent for XN, and 
only one uncertain mutation that suggests a BA.1 parent for XAU.

\subsubsection*{XAS}
https://github.com/jeromekelleher/sc2ts-paper/issues/340
This was proposed as a recombination between BA.5 (LHS) and BA.2 (RHS). 
However, we find that this is only 2 mutations different from e.g. ERR9929846 which is a BA.4 sample. 
In fact, XAS only differs from the BA.4 root by a single mutation, C27945T, 
which was identified in the original designation issue as an important ORF truncation mutation also present in a few BA.2.65 samples. 
It is possible that C27945T was acquired by recombination, this is clearly below our threshold for considering something to be a recombinant.

\subsubsection*{XB}
https://github.com/jeromekelleher/sc2ts-paper/issues/332
XB was proposed to have originated from recombination between B.1.634 (left parent) and B.1.631 (right parent). 
None of the samples in Viridian v04 belonging to these Pango lineages were integrated into the ARG due to high HMM costs.

\subsubsection*{XP}
https://github.com/jeromekelleher/sc2ts-paper/issues/345

\subsubsection*{XAJ}
https://github.com/jeromekelleher/sc2ts-paper/issues/352

\subsubsection*{XAZ}
https://github.com/jeromekelleher/sc2ts-paper/issues/356
Proposed as a recombinant of BA.2.5 on the left and BA.5 on the right, with 3 supporting sites (C2232T, C3317T, T3358C). 
In sc2ts this is very close to being a recombinant of BA.2.5 (ERR9615610) on the left, 
as not only are the 3 mutations present in this sample, 
but there is another site (C1912T apparently missed in the designation) which supports recombination. 
XAZ has C44T which is lacking in all BA.2.5 samples, 
giving only three net supporting loci for a putative recombinant. 
However, position 44 is often clipped, so this is not strong evidence against a recombinant, and 
this is a very finely balanced case.

\subsubsection*{XAN & XAV}
https://github.com/jeromekelleher/sc2ts-paper/issues/353
Below threshold.

\subsubsection*{XBE}
This was proposed as a recombinant between BA.5,5 and BE.4. 
We find two net supporting loci for a recombination breakpoint to be around 22599.

\subsection*{Analysis of the Jackson et al. (2022) recombinants}
Jackson et al. (2022) conducted detailed analyses on some early recombinants involving B.1.1.7 (Alpha), 
including the first designated Pango X lineage XA. 
The authors first performed a targeted search for candidate recombinants 
by scanning sampled sequences for sequence motifs associated with Alpha, 
followed by an analysis to reduce the list of plausible recombinants displaying signs of onward transmission. 
Here we compared our sc2ts detection results of these recombinants with the detection results reported by Jackson et al. (2022).

Jackson et al. (2022) described four groups of recombinants (named A to D) and four singleton recombinants, 
which have distinct inheritance structures (i.e. distinct parental Pango lineages and/or distinct recombination breakpoint intervals). 
All the four groups and 3 out of the 4 singletons were detected as recombinants in the ARG. 
The other singleton was detected as a recombinant by sc2ts, 
but it was not added to the ARG due to a high HMM cost (two breakpoints with X mutations). 
The sc2ts detection results are overall concordant with the results of Jackson et al. (2022) in terms of 
parental Pango lineages (agreeing to at least three-level resolution) and breakpoint intervals (Table S1).

Table S1: Sc2ts detection results for the recombinants documented by Jackson et al. (2022).
https://github.com/jeromekelleher/sc2ts-paper/issues/293

\section*{Supplementary Figures}

\subsection*{Figure S1: Classification of “robust” and artifactual recombination events.}
(A) Scatterplot of recombination events by the net number of supporting loci for the left parent (x-axis) and for the right parent (y-axis). 
The shaded regions highlight artifactual recombination events, 
which have fewer than four net supporting loci on one side or both sides of the suggested breakpoint. 
Colors indicate the primer scheme used for sequencing. 
The pie charts show breakdowns of the recombination events by primer scheme per quadrant. 
Recombination events associated with the origins of Pango Xs are labeled (classified in Figure 2). 
(B) Copying patterns of an illustrative recombination event picked from each quadrant.

\noindent\includegraphics[width=0.85\linewidth]{FigureS1.png}

\subsection*{Figure S2: Comparison of phylogenetic backbones.}

\noindent\includegraphics[width=0.85\linewidth]{FigureS2.png}

\subsection*{Figure S3: All-site mutational spectra of major VOCs.}
(A) Mutational spectra of the ARG. 
For each major VOC, the relative proportion of each mutation type was calculated from 
the mutations occurring above the nodes with a Scorpio label associated with the VOC 
(e.g. “Alpha-like” encompasses “B.1.1.7-like” and “B.1.1.7-like+E484K”). 
Note that ~98\% of all the mutations in the ARG occur above the nodes 
which have a Scorpio label associated with Alpha, Delta, Omicron BA.1, BA.2, BA.4, or BA.5 
(excluding the nodes with unassigned scorpio labels). 
The Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin. 
(B) Mutational spectra reproduced from the mutation count data analyzed in Bloom et al. (2023). 
The authors extracted mutation counts from a global UShER phylogeny (https://github.com/jbloomlab/SARS2-mut-spectrum/).

\noindent\includegraphics[width=0.85\linewidth]{FigureS3.png}

\end{document}

### 2025-08-18 latex/main.tex
Updates from Shing

@@
 
-The summary (abstract) should consist of a single paragraph of 150 words or fewer.
+%%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.
+%%% This is the SMBE abstract. We can rewrite or work off of it, but it needs to be shorter.
+Widely used phylogenetic methods to study SARS-CoV-2 evolution assume that genetic recombination is negligible and 
+viral relationships can be represented by a single tree. 
+However, important SARS-CoV-2 recombinant lineages have emerged. 
+A more suitable approach is to use ancestral recombination graphs (ARG), 
+which jointly model recombination, mutation, and genetic inheritance in a cohesive framework. 
+Recent computational advances have made ARG inference possible for datasets of millions of genome sequences. 
+Building on these new advances, we have developed sc2ts, a scalable method to simultaneously detect recombinants and 
+reconstruct ARGs from heterochronous SARS-CoV-2 genome sequences. 
+Using sc2ts, we inferred an ARG of 2.48 million genomes sampled from January, 2020 to February, 2023. 
+We find that this pandemic-scale ARG unifies known features of SARS-CoV-2 evolution — 
+the backbone phylogeny, mutational spectra, and major recognized recombinants. 
+This ARG enables us to explore the evolutionary origins of known and novel recombinants. 
+Furthermore, the ARG is compactly stored in the tree sequence format using the tskit library and 
+can be analyzed efficiently and interactively using standard Python tools on a typical laptop. 
+This combination of capabilities makes sc2ts a powerful tool 
+for studying the recombination and mutation processes driving SARS-CoV-2 evolution.
 
@@
 
-ancestral recombination graphs, recombination, SARS-CoV-2
+ancestral recombination graphs, recombination, tree sequence, SARS-CoV-2
 
@@
 
-The genomic revolution has brought a paradigm shift in how we study the evolution and spread of pathogens, 
+The genomic revolution has brought about a paradigm shift 
+in the way we study the evolution and spread of pathogens, 
 allowing real-time tracking of pathogens over time and space. 
-During the COVID-19 pandemic, thousands of genome sequences of SARS-CoV-2 samples were generated daily, 
+During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily, 
 quickly overwhelming typical phylogenetic approaches. 
 Although real-time phylogenetic methods now exist to cope with this flood of genomic data 
-(e.g., UShER\cite{Turakhia2021,DeMaio2023}), we lack methods to build phylogenies 
-that can integrate recombination in the face of ongoing sampling of infected cases.
+(e.g., UShER\cite{Turakhia2021,DeMaio2023}), 
+we lack methods to build phylogenies that can integrate recombination 
+in the face of ongoing sampling of infected cases.
 
@@
 from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}. 
-UShER has generated phylogenies updated daily \cite{McBroome2021} 
-and proved to be a powerful resource that successfully enabled the world 
+UShER has generated phylogenies updated daily \cite{McBroome2021} and 
+has proven to be a powerful resource that successfully enabled the world 
 to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}). 
@@
 with high levels of recombination. 
-These advances now enable evolutionary genealogies to be built quickly and at scale, 
-and there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.
+These advances now enable evolutionary genealogies to be built quickly and at scale, and 
+there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.
 
@@
 (1) sc2ts, a new method to infer ARGs in real time by building genealogies forward in time, and 
-(2) an ARG for SARS-CoV-2 with ~2.48 million genome sequences. We used the “Viridian” dataset, 
-which is purged of various systematic errors and artifacts \cite{Hunt2024} 
-and show that this ARG is highly congruent to an UShER phylogeny built from the same dataset. 
-Compared to previously identified recombinants, we show that this ARG contains the vast majority of 
-epidemiologically relevant signals of recombination, while adding significant nuance and texture. 
+(2) an ARG for SARS-CoV-2 with ~2.48 million genome sequences. 
+We used the “Viridian” dataset, which is purged of various systematic errors and artifacts \cite{Hunt2024} and 
+show that this ARG is highly congruent to an UShER phylogeny built from the same dataset. 
+Compared to previously identified recombinants, 
+we show that this ARG contains the vast majority of epidemiologically relevant signals of recombination, 
+while adding significant nuance and texture. 
 This ARG – which captures the major signals of evolution 
 through phylogenetic relationships, point mutations, deletions, and recombination – 
-is a resource of unprecedented detail and precision, 
-and should provide a strong basis for future studies of SARS-CoV-2.
+is a resource of unprecedented detail and precision, and 
+should provide a strong basis for future studies of SARS-CoV-2.
 
@@
 
-Sc2ts reconstructs an ARG by incrementally adding batches of sequences 
-as they become available (Figure 1A; for details, see STAR Methods). 
+Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available 
+(\ref{fig:methodology}; for details, see STAR Methods). 
 First, sc2ts infers likely “copying paths” connecting each sample in the batch to node(s) in the current ARG 
-(allowing for recombination) using the Li and Stephens (LS) model \cite{Li2003}, 
+(allowing recombination) using the Li and Stephens (LS) model \cite{Li2003}, 
 which is a Hidden Markov Model (HMM) widely used in human genomics 
 to approximate the effects of mutation and recombination on genetic inheritance. 
-As recombination is relatively rare in viruses, 
-we would expect most samples to copy from a single parent in the ARG. 
+As recombination is relatively rare in viruses, we would expect most samples to copy from a single parent in the ARG. 
 Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path 
@@
 sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster. 
-Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, 
-and applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.
+Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and 
+applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.
 
@@
 which is estimated to have emerged in November, 2021; REF). 
-To preclude adding such “time travelling” samples, we filtered samples by “HMM cost”. 
-We define the “HMM cost” of a sample to be the total penalty cost 
-for switches and mismatches (mutations) in its copying path. 
-We parameterise the LS model such that each mismatch has a penalty cost of one 
+To preclude adding such “time traveling” samples, we filtered samples by “HMM cost”. 
+We define the “HMM cost” of a sample to be the total penalty cost for switches and mismatches (mutations) in its copying path. 
+We parameterize the LS model such that each mismatch has a penalty cost of one 
 and a switch has a penalty cost of \textit{k} (here, set to 4), 
-which determines how much mutation is favored over recombination when accounting for a sample (Figure 1B). 
-If the HMM cost of a sample exceeds a threshold (here, set to 7), 
-sc2ts does not immediately add the sample to the ARG.
-
-This approach resulted in preliminary ARGs that had noticeably fewer time travellers, 
+which determines how much mutation is favored over recombination when accounting for a sample (\ref{fig:methodology}). 
+If the HMM cost of a sample exceeds a threshold (here, set to 7), sc2ts does not immediately add the sample to the ARG.
+
+This approach resulted in preliminary ARGs that had noticeably fewer time travelers, 
 but we found that it also excluded the major “saltation” events that characterize many VOCs. 
-These saltation lineages are characterised by a high number of accumulated mutations, 
-and are hypothesised to emerge from chronic infections 
+These saltation lineages are characterized by a high number of accumulated mutations, 
+and are hypothesized to emerge from chronic infections 
 during which the virus is exposed to individual-specific immune selective pressures \cite{Markov2023}. 
 Though uncommon, such lineages drove large epidemic waves during the pandemic (notably Alpha, Delta, and Omicron). 
-To distinguish time travellers with erroneous dates from true saltation lineages, 
+To distinguish time travelers with erroneous dates from true saltation lineages, 
 we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added. 
-If on subsequent days these lineages form a sufficiently large group of samples, 
-sc2ts adds the group back to the ARG. 
+If on subsequent days these lineages form a sufficiently large group of samples, sc2ts adds the group back to the ARG. 
 This “retrospective sample grouping” successfully captured the emergence of most saltation lineages (Figure SX). 
@@
 
-\noindent\includegraphics[width=0.85\linewidth]{methodology.png}
-
-\subsection*{Figure 1: Schematic overview of sc2ts.} 
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.85\linewidth]{methodology.png}
+    \caption{Schematic overview of sc2ts.
 (A) Procedure to build an ARG by incrementally updating it with samples sorted by collection dates.
-Sample 0 is used to initialise the ARG as the root. 
+Sample 0 is used to initialize the ARG as the root.
 Next, Sample 1 is attached, mapping mutations onto the new branch to explain it.
-When Sample 4 is detected as a recombinant, 
-a node representing the recombination event is inserted with two edges connecting to its parents, 
-with Sample 4 itself attached to this node as its child. 
+When Sample 4 is detected as a recombinant, a node representing the recombination event is inserted with two edges connecting to its parents, with Sample 4 itself attached to this node as its child.
 The resulting ARG contains two local trees, one on each side of the inferred breakpoint.
-(B) Copying paths for Sample 4 inferred under the LS model 
-while allowing recombination (\textit{k} = 4) or disallowing it (\textit{k} = 1000).
+(B) Copying paths for Sample 4 inferred under the LS model while allowing recombination (\textit{k} = 4) or disallowing it (\textit{k} = 1000).}
+    \label{fig:methodology}
+\end{figure}
 
@@
 Using sc2ts, we built an ARG of ~2.48 million-samples spanning the COVID-19 pandemic. 
-Inference took 21 days on a server equipped with dual EPYC (details) 
-and 500 GB of RAM (~5 CPU years; Figure SX). 
+Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM (~5 CPU years; Figure SX). 
 The ARG contains 2,689,054 nodes representing the genealogical relationships 
@@
 
-The ARG represents the full evolutionary history of SARS-CoV-2 
-and is in close agreement with the state-of-the-art UShER phylogeny built using the Viridian data \cite{Hunt2024}. 
+The ARG represents the full evolutionary history of SARS-CoV-2 and 
+is in close agreement with the state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}. 
 First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024} 
 by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods). 
-Tanglegrams reveal that the ARG captures known phylogenetic relationships both within sublineages 
-and at the level of major lineages (including VOCs), 
+Tanglegrams reveal that the ARG captures known phylogenetic relationships 
+both within sublineages and at the level of major lineages (including VOCs), 
 with good agreement between the phylogenetic bipartitions in the ARG and the UShER phylogeny (Figure S2). 
-Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian data 
+Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset 
 by pruning both to their intersection of 2,475,418 samples and 27,431 sites. 
@@
 the approximation successfully identified the most frequent deletions, 
-such as the recurrent 21991-21993 (see Document S1 for more details). 
+such as the recurrent 21991-21993 (for more details, see Document S1). 
 The node representing the origin of Alpha, with 290,407 descendants, 
@@
 by running on a subset of the data (STAR Methods). 
-Of these sites, 9 are in the set of [1XX] “problematic sites” masked by UShER [citation?]. 
+Of these sites, 9 are in the set of [1XX] “problematic sites” masked by UShER \cite{Hunt2024}. 
 In addition to remapping the data for sites identified as being involved in deletions, 
@@
 As new data became available during the pandemic, 
-volunteers continually detected recombinants which may develop concerning features, 
-particularly increased transmissibility, disease severity, 
-and ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2022}). 
-The process of gathering evidence for and characterising such recombinants is manual, 
+volunteers continually detected recombinants which might have concerning features, 
+particularly increased transmissibility, disease severity, and 
+ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2022}). 
+The process of gathering evidence for and characterizing such recombinants is manual, 
 involving visual inspection of phylogenies for unusual branches 
@@
 (whose aliases begin with an “X”), to which we refer as “Pango Xs”. 
-Below, we summarise our analysis of the Pango Xs in the ARG 
-(for extended analyses, see Documents S1 and S2).
+Below, we summarize our analysis of the Pango Xs in the ARG 
+(for extended analyses, see Document S1 and Document S2).
 
@@
 which passed the detection-level sc2ts quality control filters (Figure S1). 
-These events are supported by high numbers of averted mutations (Table 1) 
-and high numbers of loci matching their suggested parents (see their copying patterns in Figure SX). 
-Among these Pango Xs are XA, which is the first designated Pango X showing signs of community transmission \cite{Jackson2021}, 
-and XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2022}. 
-The recombinant origin of XA is well explained in the ARG by two parents and one mutation (Figure 2A; Table 1), 
+These events are supported by 
+high numbers of averted mutations (Table 1) and 
+high numbers of loci matching their suggested parents (see their copying patterns in Figure SX). 
+Among these Pango Xs are 
+XA, which is the first designated Pango X that showed signs of community transmission \cite{Jackson2021}, and
+XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2022}. 
+The recombinant origin of XA is well explained in the ARG by two parents and one mutation (\ref{fig:subgraphs}; Table 1), 
 with sc2ts correctly identifying its parental Pango lineages and breakpoint location.
@@
 An example of a set of “nested” closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP, 
-which descend from a single recombination event rather than five independent events in the ARG (Figure 2B). 
+which descend from a single recombination event rather than five independent events in the ARG (\ref{fig:subgraphs}). 
 These Pango Xs were designated around the same time. 
 These results highlight that the ARG can help clarify the relationships between recombinants, 
-allowing for more precise interpretation of evidence for recombination.
+allowing more precise interpretation of evidence for recombination.
 
@@
 For example, the closely related XN and XAU can be explained 
-by two mutations and one mutation in the ARG, respectively (Figure 2C). 
+by two mutations and one mutation in the ARG, respectively (\ref{fig:subgraphs}). 
 %%% TODO: Say something about XAJ and XAN.
@@
 
-\noindent\includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}
-
-\subsection*{Figure 2: Illustrative local subgraphs in the ARG.}
-Subgraphs around the XA-associated recombination node (depicted as an encircled R) (A); 
-the recombination node associated with a group of “nested” Pango Xs (B); 
-and the MRCA of the putative non-recombinants XN and XAU (C). 
-For more detailed visualisations of these subgraphs, see Document S2.
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}
+    \caption{Illustrative subgraphs in the ARG.
+Subgraphs around the XA-associated recombination node (depicted as an encircled R) (A); the recombination node associated with a group of “nested” Pango Xs (B); and the MRCA of the putative non-recombinants XN and XAU (C).
+For more detailed visualizations of these subgraphs, see Document S2.}
+    \label{fig:subgraphs}
+\end{figure}
 
@@
 
-We compared sc2ts against three state-of-the-art (non-phylogenetic) methods 
-that detect inter-lineage recombinants by scanning sequences for mutations associated with different Pango lineages: 
-(1) RecombinHunt \cite{Alfonsi2024}, (2) CovRecomb \cite{Li2024}, and (3) rebar (https://github.com/phac-nml/rebar).
+%%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
+We compared sc2ts against three state-of-the-art (non-phylogenetic) methods that detect inter-lineage recombinants 
+by scanning sequences for mutations associated with different Pango lineages: 
+(1) RecombinHunt \cite{Alfonsi2024}, 
+(2) CovRecomb \cite{Li2024}, and 
+(3) rebar (https://github.com/phac-nml/rebar).
 
@@
 For RecombinHunt, we took the detection results for the Pango Xs presented in the original study; 
-for CovRecomb, we summarised the results from the supplementary data of the original study (STAR Methods); 
+for CovRecomb, we summarized the results from the supplementary data of the original study (STAR Methods); 
 and for rebar, we obtained results by using it to classify the Pango X-associated recombination events in the ARG (STAR Methods). 
-For each method, we determined whether the inferred parental lineages 
-and breakpoint intervals are concordant with those identified by the community (STAR Methods). 
+For each method, we determined whether the inferred parental lineages and breakpoint intervals are 
+concordant with those identified by the community (STAR Methods). 
 We find that the methods performed well overall, 
-with no significant difference in concordance among them (Table 1; p-value = XX, chi-square test), 
-indicating that sc2ts can characterise recombinants as well as the state-of-the-art recombination detection methods.
+with no significant difference in concordance among them (Table 2; p-value = XX, chi-square test), 
+indicating that sc2ts can characterize recombinants as well as the state-of-the-art recombination detection methods.
 
@@
 167 (43\%) were also detected as recombinants by rebar (STAR Methods). 
-We hypothesise that this disagreement arose from differences in sensitivity between these methods. 
-Consistent with this, we find that rebar needs a higher level of divergence 
-between parents to correctly classify a sequence as recombinant, 
-as shown by a shorter time to their MRCA (Figure 3A) 
-and lower distance between their Pango identity (Figure 3B).
+We hypothesize that this disagreement arose from differences in sensitivity between these methods. 
+Consistent with this, we find that rebar needs a higher level of divergence between parents 
+to correctly classify a sequence as recombinant, as shown by 
+a shorter time to their MRCA (\ref{fig:averted_mutations}) and 
+lower distance between their Pango identity (\ref{fig:averted_mutations}).
 
@@
 by comparing how well a sample fits the ARG under a model allowing recombination 
-versus a model disallowing recombination, 
-as measured by the number of resulting mutations (Figure 1B). 
-The more mutations are “averted” by allowing recombination, 
-the stronger the evidence for recombination. 
+versus a model disallowing recombination, as measured by the number of resulting mutations (\ref{fig:methodology}). 
+The more mutations are “averted” by allowing recombination, the stronger the evidence for recombination. 
 We calculated the number of averted mutations for all the “robust” recombination events (STAR Methods), 
@@
 Our analysis of the number of averted mutations reveals a spectrum of plausibility 
-across all the “robust” recombination events (Figure 3C). 
+across all the “robust” recombination events (\ref{fig:averted_mutations}). 
 Examples of highly plausible recombinants are XA and XBB (for an extended analysis, see Document S1). 
-The “copying pattern” of XA visually confirms that a high number of recombination-informative sites 
-support the suggested parents (Figure 3D). 
+The “copying pattern” of XA visually confirms 
+that a high number of recombination-informative sites support the suggested parents (\ref{fig:averted_mutations}). 
 XW is an example of a marginally well-supported recombinant. 
@@
 %%% TODO: Old version as placeholder. This figure is not yet finalised.
-\noindent\includegraphics[width=0.85\linewidth]{averted_mutations.png}
-
-\subsection{Figure 3: Evidence for recombination events in the ARG.} 
-The 386 “robust” recombination events detected by sc2ts are colored 
-according to their classification by rebar (recombinant: purple, non-recombinant: gray). 
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.85\linewidth]{averted_mutations.png}
+    \caption{Evidence for recombination events in the ARG.
+The 386 “robust” recombination events detected by sc2ts are colored according to their classification by rebar (recombinant: purple, non-recombinant: gray).
 (A) Distribution of time to the MRCA of the sc2ts-identified parents.
 (B) Distribution of the Pango lineage distance between the sc2ts-identified parents.
-(C) Distribution of the number of mutations averted by recombination (bottom x-axis), 
-defined as the difference in the number of mutations 
-between a copying path allowing recombination (right y-axis) 
-and an alternative copying path disallowing recombination (top x-axis). 
-(D) Copying patterns of five select recombinants.
+(C) Distribution of the number of mutations averted by recombination (bottom x-axis), defined as the difference in the number of mutations between a copying path allowing recombination (right y-axis) and an alternative copying path disallowing recombination (top x-axis).
+(D) Copying patterns of five select recombinants.}
+    \label{fig:averted_mutations}
+\end{figure}
 
@@
 We explored the spatial distribution of the breakpoint locations of 
-the “robust” recombination events along the genome, 
-and quantified how precisely breakpoints were localized. 
+the “robust” recombination events along the genome, and quantified how precisely breakpoints were localized. 
 For each event, we computed an interval estimate of its breakpoint location, 
@@
 We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome, 
-particularly at the left boundary of the Spike gene (Figure 4A), 
+particularly at the left boundary of the Spike gene (\ref{fig:landscape}), 
 consistent with the pattern of excess observed in previous studies \cite{Turakhia2022,Li2024}. 
@@
 allowing recombination endpoints to be estimated more precisely. 
-Reassuringly, sites having unusually high mutation counts are absent in this region (Figure 4B), 
+Reassuringly, sites having unusually high mutation counts are absent in this region (\ref{fig:landscape}), 
 which would confound recombination detection.
 
-Furthermore, we find an intrinsic negative relationship 
-between breakpoint interval size and the amount of divergence between recombinant parents, 
-which is measured by time to their MRCA (Figure 4D). 
+Furthermore, we find a negative relationship between 
+breakpoint interval size and the amount of divergence between recombinant parents, 
+which is measured by time to their MRCA (\ref{fig:landscape}). 
 When the two parental lineages have diverged for longer periods of time (the rightmost points), 
-more information becomes available on either side of a breakpoint to pinpoint its location, 
-resulting in shorter intervals. 
-For instance, the breakpoint intervals tend to be the smallest for events 
-involving recombination between the deeply divergent Omicron BA.1 and BA.2 sublineages 
-(the red points in Figure 4D).
+more information becomes available on either side of a breakpoint to pinpoint its location, resulting in shorter intervals. 
+For instance, the breakpoint intervals tend to be the smallest for events involving recombination 
+between the deeply divergent Omicron BA.1 and BA.2 sublineages (the red points in \ref{fig:landscape}).
 
 %%% TODO: Old version as placeholder. This figure is not yet finalised.
-\noindent\includegraphics[width=0.85\linewidth]{recombination_landscape.png}
-
-\subsection*{Figure 4: Recombination landscape.} 
-(A) Distribution of the breakpoint intervals of the 386 “robust” recombination events along the genome 
-(annotated with gene labels on top). Each line represents an interval colored by its size. 
-(B) Number of overlapping breakpoint intervals (orange; left y-axis) 
-and number of mutations along the genome (blue; right y-axis). 
-The coordinates of the ten sites with the highest mutation counts are labelled. 
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.85\linewidth]{recombination_landscape.png}
+    \caption{Recombination landscape. (A) Distribution of the breakpoint intervals of the 386 “robust” recombination events along the genome 
+(annotated with gene labels on top).
+Each line represents an interval colored by its size.
+(B) Number of overlapping breakpoint intervals (orange; left y-axis) and number of mutations along the genome (blue; right y-axis).
+The coordinates of the ten sites with the highest mutation counts are labeled.
 (C) Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
-(D) Relationship between breakpoint interval size and time 
-between the recombination event and the parental MRCA. 
-The black curve shows the theoretical expected relationship between these quantities, 
-estimated by minimizing the sum of squared residuals given a single parameter (the mutation rate) (STAR Methods).
+(D) Relationship between breakpoint interval size and time between the recombination event and the parental MRCA.
+The black curve shows the theoretical expected relationship between these quantities, estimated by minimizing the sum of squared residuals given a single parameter (the mutation rate) (STAR Methods).}
+    \label{fig:landscape}
+\end{figure}
 
@@
 
-We examined the frequency of emergence of the recombinants over the time course of the COVID-19 pandemic (Figure 5A). 
+We examined the frequency of emergence of the recombinants over the course of the COVID-19 pandemic (\ref{recombinants_over_time}). 
 We explored two hypotheses for the timing of recombination. 
@@
 to relate the timing of recombination events with global case numbers. 
-Case numbers (\textit{n}) were binned by week from January, 2020 to March, 2023 
-and sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in Figure 5A). 
+Case numbers (\textit{n}) were binned by week from January, 2020 to March, 2023 and 
+sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in \ref{recombinants_over_time}). 
 These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset. 
@@
 %%% TODO: Old version as placeholder. This figure is not yet finalised.
-\noindent\includegraphics[width=0.85\linewidth]{recombinants_over_time.png}
-
-\subsection*{Figure 5: Recombination events across the pandemic.} 
-(A) Timing of the recombination events (black curve, right y-axis) 
-relative to the global case numbers (\cite{Dong2020}; left y-axis). 
-Cases are color-coded by their Scorpio labels, 
-with the key showing those groupings with ≥ 1,000 samples in the Viridian dataset. 
-Recombinants more commonly occur (B) in weeks with high case numbers 
-and (C) in weeks with extensive lineage variation, 
-as estimated by high expected heterozygosity (C); 
-solid lines show fits by linear regression, 
-allowing for non-zero intercepts to account for unknown cases 
-and heterozygosity not captured by the Scorpio labels.
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.85\linewidth]{recombinants_over_time.png}
+    \caption{Recombination events across the pandemic.
+(A) Timing of the recombination events (black curve, right y-axis) relative to the global case numbers (\cite{Dong2020}; left y-axis).
+Cases are color-coded by their Scorpio labels, with the key showing those groupings with \geq 1,000 samples in the Viridian dataset.
+Recombinants more commonly occur (B) in weeks with high case numbers and (C) in weeks with extensive lineage variation, as estimated by high expected heterozygosity (C); solid lines show fits by linear regression, allowing for non-zero intercepts to account for unknown cases and heterozygosity not captured by the Scorpio labels.}
+    \label{fig:recombinants_over_time}
+\end{figure}
 
@@
 and the VCF Zarr data storage format \cite{Czech2025} 
-to describe relationships among virus samples, 
-building an ARG over time as data becomes available. 
+to describe relationships among virus samples, building an ARG over time as data becomes available. 
 By reusing high-quality and high-performance components of these toolkits, 
-we could develop the sc2ts inference method itself 
-and perform all the pandemic-scale analyses in this article with Python code alone and no compiled languages.
+we could develop the sc2ts inference method itself and 
+perform all the pandemic-scale analyses in this article with Python code alone and 
+no compiled languages.
 
@@
 This computation, performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees, 
-required 30 lines of Python code and took 2m59s on an intel i7-9700 CPU, with a peak memory usage of about 4.5G 
+required 30 lines of Python code and took 2m59s on an intel i7-9700 CPU, with a peak memory usage of about 4.5 Gb 
 (see \cite{Czech2025} for further benchmarks for processing the Viridian data in VCF Zarr format). 
@@
 and the maximum number of children per node in the Usher and sc2ts outputs. 
-In the Usher tree it took 116ms and 4 lines of code to count polytomies (171,148) 
+In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148) 
 and compute the maximum number of children (5918). 
-In the sc2ts ARG, it took 385ms and 6 lines of code to find the maximum number of polytomies (137,001) 
-and maximum number of children (7708) across all 348 local trees.
+In the sc2ts ARG, it took 385 ms and 6 lines of code to find the maximum number of polytomies (137,001) and 
+maximum number of children (7708) across all 348 local trees.
 

### 2025-08-18 arg_postprocessing/scripts/rematch_recombinants.py
Update for latest sc2ts version
Do hardest work first

@@ -26,7 +26,6 @@ def worker(work):
     )
     out = subprocess.check_output(cmd, shell=True)
     result = json.loads(out.decode())
-    result["recombinant"] = int(work.node)
     return result
 
 
@@ -52,7 +51,7 @@ def run(ts, pattern, output):
 
     # The Delta wave was particularly difficult in terms of memory usage,
     # so limiting parallelism based on date. Post-Omicron wasn't quite as bad
-    bins = ["2020-01", "2021-09", "2022-01", "2024-01"]
+    bins = ["2020-01", "2021-09", "2022-01", "2030-01"]
     cores_per_bin = [20, 4, 8]
     assert len(cores_per_bin) == len(bins) - 1
 
@@ -70,7 +69,7 @@ def run(ts, pattern, output):
         split_work[cores] = work
 
     json_data = []
-    for cores in cores_per_bin:
+    for cores in reversed(cores_per_bin):
         work = split_work[cores]
         print(f"Running {len(work)} on {cores}")
         with cf.ProcessPoolExecutor(cores) as executor:

### 2025-08-19 arg_postprocessing/scripts/rematch_recombinants.py
Split rematching and update sc2ts

@@ -17,13 +17,20 @@ class Work:
     pattern: str
     node: int
     date: str
+    path: str
 
 
 def worker(work):
-    cmd = (
-        f"python -m sc2ts rematch-recombinant {work.node} "
-        f"--path-pattern={work.pattern} --date={work.date} -vv"
-    )
+    if work.pattern is None:
+        cmd = (
+            f"python -m sc2ts rematch-recombinant-lbs {work.path} "
+            f"{work.node} -vv"
+        )
+    else:
+        cmd = (
+            f"python -m sc2ts rematch-recombinant {work.node} "
+            f"--path-pattern={work.pattern} --date={work.date} -vv"
+        )
     out = subprocess.check_output(cmd, shell=True)
     result = json.loads(out.decode())
     return result
@@ -36,9 +43,10 @@ def dump(json_data, path):
 
 @click.command()
 @click.argument("ts")
-@click.argument("pattern")
 @click.argument("output")
-def run(ts, pattern, output):
+@click.option("--pattern", default=None)
+def run(ts, output, pattern):
+    ts_path = ts
     ts = tszip.load(ts)
 
     recombinants = np.where(ts.nodes_flags & sc2ts.NODE_IS_RECOMBINANT > 0)[0]
@@ -47,12 +55,18 @@ def run(ts, pattern, output):
     for u in recombinants:
         md = ts.node(u).metadata
         date = md["sc2ts"]["date_added"]
-        all_work.append(Work(pattern, u, date))
+        all_work.append(Work(pattern, u, date, ts_path))
+
+    del ts
 
     # The Delta wave was particularly difficult in terms of memory usage,
     # so limiting parallelism based on date. Post-Omicron wasn't quite as bad
     bins = ["2020-01", "2021-09", "2022-01", "2030-01"]
-    cores_per_bin = [20, 4, 8]
+    cores_per_bin = [30, 5, 10]
+    if pattern is not None:
+        # When we're doing the non-recombinant match we have to be
+        # more conservative about memory usage
+        cores_per_bin = [20, 4, 8]
     assert len(cores_per_bin) == len(bins) - 1
 
     random.seed(42)
@@ -71,7 +85,7 @@ def run(ts, pattern, output):
     json_data = []
     for cores in reversed(cores_per_bin):
         work = split_work[cores]
-        print(f"Running {len(work)} on {cores}")
+        print(f"Running {len(work)} on {cores} cores")
[…]

### 2025-08-20 latex/docS1.tex
Updates from Overleaf

@@
 Artifactual recombination events can be characterized as 
-having fewer than four such loci supporting the left or right parent on either side of a putative breakpoint (STAR Methods). 
+having fewer than four such loci supporting the left or right parent on either side of a suggested breakpoint (STAR Methods). 
 Of the 929 recombination events in the ARG, this measure identifies 543 as artifactual, 
@@
 We excluded this likely artifactual event from further analysis and 
-consider XS to emerge from a single recombinant origin.
+consider XS to have emerged from a single recombinant origin.
 
@@
 Three of them (XAY, XBC, XBT) had more than one breakpoint, 
-but were not added to the ARG due to high HMM cost and a lack of sample support to be incorporated as retrospective sample groups. 
+but were not added to the ARG due to high HMM cost and 
+a lack of sample support to be incorporated as retrospective sample groups. 
 However, the other two (XAC and XAK) had only one breakpoint.

### 2025-08-20 latex/main.tex
Updates from Overleaf

@@
 
-ancestral recombination graphs, recombination, tree sequence, SARS-CoV-2
+ancestral recombination graphs, population genetics, recombination, tree sequence, SARS-CoV-2
 
@@
 to approximate the effects of mutation and recombination on genetic inheritance. 
-As recombination is relatively rare in viruses, we would expect most samples to copy from a single parent in the ARG. 
+As recombination is relatively rare in viruses, 
+we would expect most samples to copy from a single parent in the ARG. 
 Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path 
@@
 To preclude adding such “time traveling” samples, we filtered samples by “HMM cost”. 
-We define the “HMM cost” of a sample to be the total penalty cost for switches and mismatches (mutations) in its copying path. 
+We define the “HMM cost” of a sample to be 
+the total penalty cost for switches and mismatches (mutations) in its copying path. 
 We parameterize the LS model such that each mismatch has a penalty cost of one 
-and a switch has a penalty cost of \textit{k} (here, set to 4), 
+and a switch has a penalty cost of $k$ (here, set to 4), 
 which determines how much mutation is favored over recombination when accounting for a sample (\ref{fig:methodology}). 
-If the HMM cost of a sample exceeds a threshold (here, set to 7), sc2ts does not immediately add the sample to the ARG.
+If the HMM cost of a sample exceeds a threshold (here, set to 7), 
+sc2ts does not immediately add the sample to the ARG.
 
@@
 we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added. 
-If on subsequent days these lineages form a sufficiently large group of samples, sc2ts adds the group back to the ARG. 
+If on subsequent days these lineages form a sufficiently large group of samples, 
+sc2ts adds the group back to the ARG. 
 This “retrospective sample grouping” successfully captured the emergence of most saltation lineages (Figure SX). 
@@
 The resulting ARG contains two local trees, one on each side of the inferred breakpoint.
-(B) Copying paths for Sample 4 inferred under the LS model while allowing recombination (\textit{k} = 4) or disallowing it (\textit{k} = 1000).}
+(B) Copying paths for Sample 4 inferred under the LS model while allowing recombination ($k = 4$) or disallowing it ($k = 1000$).}
     \label{fig:methodology}
@@
 For all these Pango Xs, the number of recombination-informative sites 
-fell below our threshold of detection (i.e. k = 4 in the HMM). 
+fell below our threshold of detection (i.e. $k = 4$ in the HMM). 
 The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG 
@@
 to relate the timing of recombination events with global case numbers. 
-Case numbers (\textit{n}) were binned by week from January, 2020 to March, 2023 and 
+Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and 
 sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in \ref{recombinants_over_time}). 
 These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset. 
-Alongside the case numbers, the black curve shows the number of recombinants inferred within each week along the ARG, 
+Alongside the case numbers, 
+the black curve shows the number of recombinants inferred within each week along the ARG, 
 focusing on the 386 “robust” events. 
@@
 Overall, the number of recombinant events rose with the global number of cases 
-(Figure 5B; linear regression: 1.2 + 2.9 10^{-7} x, p = 3.2  10^{-7}). 
+(Figure 5B; linear regression: $1.2 + 2.9 10^{-7} x$, $p = 3.2e^{-7}$). 
 Recombination events also rose with the diversity of lineages in each week 
-(Figure 5C; linear regression: 1.0 + 5.3 x, p = 1.3 10^{-7}), 
+(Figure 5C; linear regression: $1.0 + 5.3 x$, $p = 1.3e^{-7}$), 
 measured as the chance that two randomly drawn viruses have different Scorpio designations 
-(H, referred to as the “expected heterozygosity”).
+($H$, referred to as the “expected heterozygosity”).
 
 We then asked which single measure best explains the number of recombination events inferred each week, 
-considering \textit{n}, \textit{H}, \textit{n H}, \textit{n}^2, \textit{n}^2 \textit{H}. 
+considering $n$, $H$, $n H$, $n^2$, $n^2 H$. 
 We included the squared number of cases each week thinking that it might better predict co-infections. 
-The best single predictor of recombination events involved the product of cases and expected heterozygosity (\textit{n H}), 
-with an adjusted R^2 of 25.3\%. 
-This measure of case diversity (\textit{n H}) explained 62\% more of the variation in recombination events per week 
-than the next best predictor (\textit{H}, adjusted R^2 of 15.7\%). 
+The best single predictor of recombination events involved the product of cases and expected heterozygosity ($n H$), 
+with an adjusted $R^2$ of 25.3\%. 
+This measure of case diversity ($n H$) explained 62\% more of the variation in recombination events per week 
+than the next best predictor ($H$, adjusted $R^2$ of 15.7\%). 
 While consistent with the hypotheses that recombination events occur more often when infection rates are common 
@@
 (A) Timing of the recombination events (black curve, right y-axis) relative to the global case numbers (\cite{Dong2020}; left y-axis).
-Cases are color-coded by their Scorpio labels, with the key showing those groupings with \geq 1,000 samples in the Viridian dataset.
+Cases are color-coded by their Scorpio labels, with the key showing those groupings with $\geq 1,000$ samples in the Viridian dataset.
 Recombinants more commonly occur (B) in weeks with high case numbers and (C) in weeks with extensive lineage variation, as estimated by high expected heterozygosity (C); solid lines show fits by linear regression, allowing for non-zero intercepts to account for unknown cases and heterozygosity not captured by the Scorpio labels.}
@@
 \subsection*{Efficient inference and analysis of a pandemic-scale ARG}
-
 The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale, 
@@
 
-To illustrate the power of these libraries we provide some illustrative examples 
+To illustrate the power of these libraries, we provide some examples 
 in which we compare the Usher and sc2ts intersections in tskit format described earlier. 
@@
 The total aligned dataset for 1,226,587 samples and 27,431 sites shared by the sc2ts and Usher tree sequences 
-represents 31.34 GiB of nucleotide calls. 
+represents 31.34 GB of nucleotide calls. 
 Of the 80,574,518 missing data calls (Ns) in the alignments, 
@@
 This computation, performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees, 
-required 30 lines of Python code and took 2m59s on an intel i7-9700 CPU, with a peak memory usage of about 4.5 Gb 
+required 30 lines of Python code and took 2m59s on an intel i7-9700 CPU, with a peak memory usage of about 4.5 GB 
 (see \cite{Czech2025} for further benchmarks for processing the Viridian data in VCF Zarr format). 
@@
 We present sc2ts, a real-time ARG inference method, 
-and applied it to build a ~2.48 million-sample pandemic-scale ARG of SARS-CoV-2. 
+and applied it to build a \~2.48 million-sample pandemic-scale ARG of SARS-CoV-2. 
 By identifying specific samples as parents, 
@@
 Combined with new methods and tools to evaluate evidence for recombination, 
-sc2ts is a significant technological advance over current methods for the discovery 
-and analysis of recombinants and efficient inference of SARS-CoV-2 genealogy.
+sc2ts is a significant technological advance over current methods for the discovery and 
+analysis of recombinants and efficient inference of SARS-CoV-2 genealogy.
 
@@
 for a variety of evolutionary and genetic analyses beyond those presented here. 
-We hope that the ARG will spur on new research areas, in which recombination is naturally and fully accounted for, 
+We hope that the ARG will spur on new research areas, 
+in which recombination is naturally and fully accounted for, 
 and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022} 
-has made on our understanding of human genetic ancestry and demographic past.
+has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
 
@@
 which are either too divergent or have a high number of mutations inconsistent with their collection dates. 
-Here we use HMM cost to reject time travelling samples, 
+Here we use HMM cost to reject time traveling samples, 
 which no existing phylogenetic method detects and handles during inference 
 (e.g., in the Viridian UShER phylogeny, parent and child nodes do not necessarily follow time order \cite{Hunt2024}), 
-resulting in a clean dataset freed of samples with erroneous collection dates that may affect downstream analyses. 
-Future work may be estimating dates for the suspected time travelling samples using the ARG provided here 
+resulting in a clean dataset rid of samples with erroneous collection dates that may affect downstream analyses. 
+Future work may be estimating dates for the suspected time traveling samples using the ARG provided here 
 and the sc2ts sample matching engine.
@@
 Recombination is an important evolutionary force in viruses \cite{Pérez-Losada2015}, 
-but it has been understudied partly owing to a lack of suitable data models, methods, and tools \cite{Neches2020}. 
+but has been understudied in part due to a lack of suitable data models, methods, and tools \cite{Neches2020}. 
 While the methods and tools developed here are tailored for SARS-CoV-2, 
@@
 \subsubsection*{Description of sc2ts}
+
 Sc2ts (pronounced “scoots”, optionally) 
@@
 The basic idea is to incrementally update an ARG regularly 
-with the sequences collected since the previous batch (Figure 1A). 
+with the sequences collected since the previous batch \ref{fig:methodology}. 
 We used daily batches for SARS-CoV-2, 
@@
 \subsubsection*{Ancestral Recombination Graphs}
+
 %%% This paragraph needs to be rewritten.
@@
 to define an alternative formulation of the coalescent with recombination (Hudson, 1983), 
-the term “Ancestral Recombination Graph” (ARG) has come to be used 
-more generally to describe not just realisations of the model, 
+the term “Ancestral Recombination Graph” has come to be used 
+more generally to describe not just realizations of the model, 
 but any recombinant genetic ancestry (Minichiello and Durbin, 2006). 
@@
 Initially developed to facilitate large-scale coalescent simulations \cite{Kelleher2016}, 
-STS has been extended and applied to forward-time simulations \cite{Kelleher2018,Haller2019}, 
-calculation of population genetics statistics \cite{Ralph2020} 
-and ARG inference \cite{Kelleher2019,Wohns2022}. 
+STS has been extended and applied to 
+forward-time simulations \cite{Kelleher2018,Haller2019}, 
+calculation of population genetics statistics \cite{Ralph2020} and 
+ARG inference \cite{Kelleher2019,Wohns2022}. 
 The STS is based on a simple tabular representation, 
@@
 The genetic inheritance between a pair of nodes along a segment of genome 
-is defined by the edge (ℓ, r, p, c), 
-which states that child node c inherited its genome from parent node p 
-from left coordinate ℓ to right coordinate r. 
+is defined by the edge ($l$, $r$, $p$, $c$), 
+which states that child node $c$ inherited its genome from parent node $p$ 
+from left coordinate $l$ to right coordinate $r$. 
 A site defines a position on the genome and the ancestral state (allele) at that site. 
@@
 The STS encoding is also very concise, allowing, for example, 
-for millions of complete human genomes to be stored in a few Gigabytes of space \cite{Kelleher2019}, 
+for millions of complete human genomes to be stored in a few GB of space \cite{Kelleher2019}, 
 and so it is particularly well suited to representing large SARS-CoV-2 genealogies.
 
-The SARS-CoV-2 ARGs inferred using sc2ts can be conveniently and efficiently analysed 
+The SARS-CoV-2 ARGs inferred using sc2ts can be conveniently and efficiently analyzed 
 using the mature and feature-rich tskit library \cite{Kelleher2018,Ralph2020}. 
@@
 \subsubsection{Li and Stephens haplotype-copying model}
+
 LS model \cite{Li2003} is an approximation of the coalescent with recombination 
@@
 
-In the original formulation of the LS model (see Appendix S1 in Li and Stephens, 2003), 
+In the original formulation of the LS model (see Appendix S1 of \cite{Li2003}), 
 the probability of switching between the hidden states (i.e., genomes in a reference panel) 
@@
 (1) the rate of recombination between the sites and 
-(2) the number of genomes in the reference panel (\textit{n}). 
+(2) the number of genomes in the reference panel ($n$). 
 Sc2ts uses the same formulation except that n is set to 1. 
@@
 For simplicity, we assume that any single base mutation causing a mismatch is equally probable, 
-ignoring transition/transversion biases and variation of mutation rate along the genome. 
+ignoring transition-transversion biases and variation of mutation rate along the genome. 
 As data sampling was dense for SARS-CoV-2, this assumption was considered tolerable.
 
-In our early attempts to build ARGs, we explored using several values of \textit{k}. 
-We ran sc2ts with \textit{k} set to 3, 4, and 5 on the samples collected in 2020 (REF). 
-We would expect that a good value for \textit{k} should lead to parsimonious ARGs 
+In our early attempts to build ARGs, we explored using several values of $k$. 
+We ran sc2ts with $k$ set to 3, 4, and 5 on the samples collected in 2020 \cite{Zhan2023}}(REF). 
+We would expect that a good value for $k$ should lead to parsimonious ARGs 
 that capture plausible recombination events without incurring too many mutations, particularly reversions. 
-A value of \textit{k} too low could result in many artifactual recombination events, 
-whereas a value of \textit{k} too high could miss many genuine recombination events. 
-We found that ARGs built with \textit{k} set to 3 contained many dubious recombination events, 
-but ARGs built with \textit{k} set to 5 failed to include the well supported recombination events documented by Jackson et al. (2022). 
-Hence, we decided to use \textit{k} set to 4 to infer the final ARG described in this study.
+A value of $k$ too low could result in many artifactual recombination events, 
+whereas a value of $k$ too high could miss many genuine recombination events. 
+We found that ARGs built with $k$ set to 3 contained many dubious recombination events, 
+but ARGs built with $k$ set to 5 failed 
+to include the well supported recombination events documented in a previous study \cite{Jackson2021}. 
+Hence, we decided to use $k$ set to 4 to infer the final ARG described in this study.
 
@@
 \subsubsection*{Accelerating sample matching}
+
 The SARS-CoV-2 sequences were determined from samples collected densely in time. 
@@
 Sc2ts builds a local phylogeny relating the samples in each cluster as follows. 
-First, it infers a tree using a standard neighbour-joining algorithm \cite{Saitou1987}. 
+First, it infers a tree using a standard neighbor-joining algorithm \cite{Saitou1987}. 
 Then, mutations are mapped back onto it using the Fitch-Hartigan parsimony algorithm \cite{Fitch1971,Hartigan1973}. 
@@
 we observed two types of outlier samples which had unusually high numbers of mutations: 
-(1) “time travellers” and (2) “saltation” lineages. 
-Time travelling samples are caused by errors in their collection dates. 
+(1) “time travelers” and (2) “saltation” lineages. 
+Time traveling samples are caused by errors in their collection dates. 
 For example, if a sample is actually collected 10 months after its reported collection date, 
@@
 assuming a substitution rate of 2 events per genome per month. 
-Samples from “saltation” lineages (particularly the major VOCs) are characterised 
+Samples from “saltation” lineages (particularly the major VOCs) are characterized 
 by high numbers of mutations by the time of their detection. 
@@
 To handle these outlier samples, we introduce three mechanisms: 
-(1) HMM cost-based rejection, (2) retrospective sample grouping, and (3) unconditional seeding.
+(1) HMM cost-based rejection, 
+(2) retrospective sample grouping, and 
+(3) unconditional seeding.
 
@@
 that are generally poorly explained by the ARG 
-using the LS model parameterised by a specific value for the recombination penalty (\textit{k}). 
+using the LS model parameterised by a specific value for the recombination penalty ($k$). 
 For each sample, an HMM cost is calculated based on the number of switches and mismatches 
 in its most likely copying path under the LS model: 
-The sum of the penalty cost for the switches (\textit{k}) 
+The sum of the penalty cost for the switches ($k$) 
 and the penalty cost of the mismatches (implicitly, equal to one). 
-To build the ARG described here, we used \textit{k} of 4 and a maximum HMM cost of 7. 
+To build the ARG described here, we used $k$ of 4 and a maximum HMM cost of 7. 
 This combination of parameter values rejected samples which have no switch 
@@
 Such samples can be added back using a separate mechanism (see below). 
-We find that although it effectively removes time travellers with poor sample support (e.g. singletons), 
+We find that although it effectively removes time travelers with poor sample support (e.g. singletons), 
 the mechanism can filter samples too aggressively. 
@@
 (b) they are collected on at least 3 different dates 
-(to avoid adding batches of time travellers having the same collection date); and 
+(to avoid adding batches of time travelers having the same collection date); and 
 (c) they have at least two shared “non-immediate” reversions 
-(so as to prevent unrelated time travellers from being grouped together). 
+(so as to prevent unrelated time travelers from being grouped together). 
 Furthermore, phylogenetic tree quality metrics are used to preclude groups of samples 
@@
 
-\subsubsection*{Data pre-processing}
+\subsubsection*{Data preprocessing}
 
@@
 from samples collected from January, 2020 to March, 2023 (“batch 1”). 
-We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) 
-using MAFFT v7.525 \cite{Katoh2013}, 
+We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using MAFFT v7.525 \cite{Katoh2013}, 
 with the option ‘--keeplength’ to retain the reference coordinate system, discarding insertions. 
@@
 
-Prior to ARG inference, we pre-processed the remaining sequences as follows. 
+Prior to ARG inference, we preprocessed the remaining sequences as follows. 
 We treated low-quality bases (represented by Ns), gaps and ambiguous characters in the sequences as missing data. 
@@
 \subsubsection*{Reconstructing a pandemic-scale ARG}
+
 %%% TODO: Much of this needs to be rewritten once the post-processing pipeline is finalised.
+Using sc2ts, we reconstructed the ARG presented here in three stages: 
+(1) initialization, (2) extension, and (3) post-processing.
+
+First, we initialized a base ARG. 
+We added the Wuhan-Hu-1 reference sample (MN908947.3) as the root node, dated December 8, 2019. 
+Then, we ran sc2ts unconditionally (i.e., without excluding samples by HMM cost) up to March 01, 2020 (inclusive). 
+This step is necessary to accumulate sufficient genetic diversity in the ARG 
+for subsequent sample matching and attachment in the next stage.
+
+Second, we extended the base ARG by adding batches of samples collected on and after March 02, 2020. 
+Sample filtering, HMM cost-based rejection, and retrospective sample grouping, as described above, 
+are in full effect during this stage.
+
+Finally, we performed the following post-processing steps 
+to modify certain non-topological properties of the ARG or 
+to enrich the ARG with additional information:
+1. Moving mutations around recombinants? TODO
+2. Adjusting recombination breakpoints. XX.
+3.
+Mapping multi-base deletions.
+Multi-base deletions are challenging to handle, as they violate the Markov property of the HMM.
+In our ARG, insertions are deleted regions that acquire nucleotides (not necessarily the same nucleotides) without altering the genome length, because we discarded length-changing insertions when preprocessing the aligned sequences.
+We treated multi-base indels as multiple separate mutations in the HMM.
+We mapped them back onto the inferred ARG as single events.
+4. Mapping mutations at masked sites. XX.
+5.
+Adding exactly matched samples.
+During sample matching, samples identical to node(s) in a given ARG were not added to the ARG.
+Instead, after inference, we simply added the nodes to the ARG using the saved matching results.
+This allowed for computational savings, as it kept the number of possible parents available for matching to a minimum.
+6.
+Estimating non-sample node ages.
+The ages of the inserted non-sample nodes (the nodes in the local trees and the recombination nodes) were initially assigned to arbitrarily determined values.
+We ran \textit{tsdate} (“variational gamma”, version XX; \cite{Wohns2022} Pope et al., 2025) on the ARG to adjust the ages of these nodes, while fixing the ages of the sample nodes to the reported collection times (defined as fractional years before the collection date of the last sample added to the ARG, which is February 20, 2025 in the ARG described in this study).
+7. Assigning Pango lineages to non-sample nodes.
 
 \subsubsection*{Comparing phylogenetic backbones}
+
 We used Pango lineage designations to select \textit{representative} samples 
@@
 We compare topology in the simplified ARG against the UShER tree, built from the Viridian dataset. 
-To remove time travelling samples from the UShER tree, 
+To remove time traveling samples from the UShER tree, 
 we reduce it to 2475419 samples shared with the main sc2ts ARG 
@@
 
-The simplified ARG contains 4 single-breakpoint recombination nodes, 
+The simplified ARG contains four single-breakpoint recombination nodes, 
 corresponding to the origins of Delta, BA.2, BA.5, and a further BA.2 recombinant 
@@
 and extract non-recombinant structures from the ARG 
-by creating a "base tree" of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, & BA.5), 
+by creating a "base tree" of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, and BA.5), 
 a Delta subtree of 160 Pangos, 
@@
 The order of tips in the two trees was chosen to reduce tangling, 
-as calculated using the neighbor-net algorithm from Dendroscope \cite{Scornavacca2011,Huson2012}, 
+as calculated using the NeighborNet algorithm from Dendroscope \cite{Scornavacca2011,Huson2012}, 
 with tanglegrams plotted using the SVG outputting facility built into tskit. 
@@
 
+\subsubsection{Counting supporting loci for recombinant parents}
+When counting such loci, we considered nearly adjacent sites (separated by at most three bases) as single loci, 
+to avoid over-counting support from correlated loci, which would lead to artificial signals of recombination. 
+Also, when counting supporting loci for a focal parent, 
+we penalized loci where the character(s) in the recombinant matches that in the non-focal parent but not in the focal parent.
+
 \subsubsection*{Calculating Pango lineage distance using pangonet}
@@
 which relates all the designated Pango lineages. 
-To construct this pseudo-phylogeny, we used pangonet, 
+To construct this pseudo-phylogeny, we used \textit{pangonet}, 
 which mines the information on the Pango lineages in the “alias_key.json” and “lineage_notes.txt” files 
@@
 
+\subsubsection{Estimating interval lengths}
+Recombination intervals are imprecise 
+when the recombining parents are identical in sequence at sites spanning the breakpoint. 
+Here we estimate the interval lengths expected around a breakpoint, 
+given a genome-wide mutation rate $\mu$, 
+total time between two samples ($T$), and 
+sufficient support to detect the recombination event. 
+If a breakpoint occurs a fraction $P$ along a genome of length $L$ and 
+$k$ mutations have occurred to the left of the breakpoint, 
+then we expect the nearest mutation to occur $P L / (k + 1)$ bases to the left. 
+If we require $n$ mutations on both sides of the breakpoint, 
+we can calculate the expected value of the interval to the left by summing across a Poisson-distributed number of mutations, 
+conditioned on having at least $n$ mutations:
+
+\[I_{left} = \frac{L}{n T \mu} \frac{\Gamma(1+n) - \Gamma(1+n, P T \mu)}{\Gamma(n) - \Gamma(n, P T \mu)}\]
+
+A similar result holds for the right interval ($I_{right}$), but with $P$ replaced with $1–P$. 
+Summing the two intervals ($I_{left}+I_{right}$) gives the expected interval length. 
+As detectable recombination events are nearer the center of the chromosome, we use $P=½$ in Figure X.
+
+\subsubsection{Analyzing global case counts}
+To explore the timing of emergence of recombinants, 
+case count data were downloaded from 
+the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) 
+at Johns Hopkins University \cite{Dong2020} and binned by week from January, 2020 to March, 2023. 
+The proportion of each lineage $i$ in each week ($p_i$) was then inferred 
+from the Scorpio designation at each node within the ARG for that week. 
+Genetic variation was measured as the “expected heterozygosity” across lineages, 
+which is the chance that two lineages chosen at random 
+from that week had different Scorpio designations ($\sum_{i} \sum_{j} 2 p_i p_j$). 
+The linear regression models reported in \textbf{Figure} \ref{fig:recombinants_over_time} allowed for a non-zero intercept, 
+because there may be unreported cases even without any reported cases. 
+The slopes were highly significant, and 
+the p-values were confirmed to be robust by randomly permuting the predictor variable 10,000 times. 
+ANOVAs were conducted to determine which single predictor ($n$, $H$, $n H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week. 
+These statistical tests and graphics were conducted in Mathematica 14.2 (Wolfram 2024) and available at GITHUB LINK.
+
+In addition to the analysis of the 386 “robust” recombination events with a minimum of four nucleotides supporting each parent, 
+we analyzed the relationship between the full set of 929 recombinants and case numbers. 
+These additional recombinants occurred disproportionately during the Delta and first Omicron wave (July, 2021 to February, 2022). 
+Again, the number of recombinant events rose with the global number of cases 
+(linear regression: $3.4 + 5.8 10-7 x$, $p = 0.0004$) 
+and with the diversity of lineages in each week 
+(linear regression: $0.5 + 19.7 x$, $p = 5.1 10^{-13}$). 
+The single best predictor of recombination events among the predictors tested ($n$, $H$, $n H$, $n^2$, $n^2 H$) 
+was again case diversity ($n H$), 
+with an adjusted $R^2$ of 28.7\%, 
+although this was now followed closely by expected heterozygosity on its own ($H$), 
+with an adjusted $R^2$ of 27.7\%. 
+Thus, similar conclusions are reached with either dataset, 
+although filtering the recombination events to the 386 with better support revealed a stronger relationship with case numbers.
+
 \subsection*{Additional resources}

### 2025-08-21 arg_postprocessing/scripts/add_recombinant_k1000_muts_to_csv.py
Formalise the k1000 mutations addition

@@ -1,0 +1,30 @@
+import click
+import pandas as pd
+import json
+import numpy as np
+
+
+@click.command()
+@click.argument("rematches_json")
+@click.argument("recombinants_csv")
+@click.argument("output")
+def run(rematches_json, recombinants_csv, output):
+    dfr = pd.read_csv(recombinants_csv).set_index("recombinant")
+    no_recomb_match_mutations = {}
+    with open(rematches_json) as f:
+        for r in json.load(f):
+            nrm = r["no_recomb_match"]
+            assert len(nrm["path"]) == 1
+            nrm_muts = len(nrm["mutations"])
+            no_recomb_match_mutations[r["recombinant"]] = nrm_muts
+
+    dfr["k1000_muts"] = no_recomb_match_mutations
+    missing = np.sum(dfr["k1000_muts"].isna())
+    if missing > 0:
+        print(f"WARNING!!! Missing {missing}/{dfr.shape[0]} recombinants from matches")
+
+    dfr.reset_index().to_csv(output, index=False)
+
+
+if __name__ == "__main__":
+    run()

### 2025-08-23 latex/main.tex
Whitespace
Fixup some minor formatting issues

@@
 (\ref{fig:methodology}; for details, see STAR Methods).
-First, sc2ts infers likely “copying paths” connecting each sample in the batch to node(s) in the current ARG 
+First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
 (allowing recombination) using the Li and Stephens (LS) model \cite{Li2003},
@@
 As an illustration of the open-ended and efficient nature of the analyses this enables,
-we computed the number of polytomies (nodes that have > 2 children) 
+we computed the number of polytomies (nodes that have $> 2$ children)
 and the maximum number of children per node in the Usher and sc2ts outputs.
@@
 
-Recombination is an important evolutionary force in viruses \cite{Pérez-Losada2015}, 
+Recombination is an important evolutionary force in viruses \cite{Perez-Losada2015},
 but has been understudied in part due to a lack of suitable data models, methods, and tools \cite{Neches2020}.
@@
 
-\newpage
-
-\section*{MAIN FIGURE TITLES AND LEGENDS}
-
-%%%  At final submission, figure files MUST be 
-%%%  provided separately as high-resolution image 
-%%%  files. All of the panels for a figure should 
-%%%  be in the same file. Figures should have 
-%%%  clear labels/file names (Figure 1, Figure 2, 
-%%%  etc.).
-
-%%%  Figure titles and legends should be placed 
-%%%  at the end of the main text. You do not 
-%%%  need to place the figures, nor their titles 
-%%%  and legends, within the main text. While 
-%%%  typesetting your article, our team will 
-%%%  place each figure in the best location 
-%%%  based on the final layout and on your 
-%%%  figure citations, e.g., (Figures 1A and 1B).
-
-%%%  Please review the figure guidelines before 
-%%%  submitting your final materials: 
-%%%  https://www.cell.com/figureguidelines.
-
-\noindent\includegraphics[width=0.85\linewidth]{Figure1.jpg}
-
-\subsection*{Figure 1. A brief title that describes the entire figure without citing specific panels}
-
-The figure legend can be all one paragraph and describe the images (A), graphs (B), and plots (C), etc., together.
-\newline
-(A) Or each panel or group of panels can be described separately, as shown here and below.
-\newline
-(B) Graph of X, Y, and Z.
-\newline
-(C and D) If panels are grouped like this, please explicitly describe each panel, e.g., “Images showing SEM (C) and TEM (D).”
-\newline
-Please define all scale and error bars, and please review the Cell Press figure guidelines before submission: \href{https://www.cell.com/figureguidelines}{https://www.cell.com/figureguidelines}.
+% \newpage
+
+% \section*{MAIN FIGURE TITLES AND LEGENDS}
+
+% %%%  At final submission, figure files MUST be
+% %%%  provided separately as high-resolution image
+% %%%  files. All of the panels for a figure should
+% %%%  be in the same file. Figures should have
+% %%%  clear labels/file names (Figure 1, Figure 2,
+% %%%  etc.).
+
+% %%%  Figure titles and legends should be placed
+% %%%  at the end of the main text. You do not
+% %%%  need to place the figures, nor their titles
+% %%%  and legends, within the main text. While
+% %%%  typesetting your article, our team will
+% %%%  place each figure in the best location
+% %%%  based on the final layout and on your
+% %%%  figure citations, e.g., (Figures 1A and 1B).
+
+% %%%  Please review the figure guidelines before
+% %%%  submitting your final materials:
+% %%%  https://www.cell.com/figureguidelines.
+
+% \noindent\includegraphics[width=0.85\linewidth]{Figure1.jpg}
+
+% \subsection*{Figure 1. A brief title that describes the entire figure without citing specific panels}
+
+% The figure legend can be all one paragraph and describe the images (A), graphs (B), and plots (C), etc., together.
+% \newline
+% (A) Or each panel or group of panels can be described separately, as shown here and below.
+% \newline
+% (B) Graph of X, Y, and Z.
+% \newline
+% (C and D) If panels are grouped like this, please explicitly describe each panel, e.g., “Images showing SEM (C) and TEM (D).”
+% \newline
+% Please define all scale and error bars, and please review the Cell Press figure guidelines before submission: \href{https://www.cell.com/figureguidelines}{https://www.cell.com/figureguidelines}.
 Example figure created by Cassie Comeau, Cell Press.
 
-\newpage
-
-\section*{MAIN TABLES, INCLUDING TITLES AND LEGENDS}
-
-%%%  Whenever possible, we encourage you to submit 
-%%%  your main-text tables as Microsoft Word documents, 
-%%%  using Word's table function. This will ensure 
-%%%  the best results during conversion. Tables 
-%%%  should be numbered Table 1, Table 2, etc. and 
-%%%  should not include subpanels (do not use Table 1A, 
-%%%  1B, etc.). Give each table a brief descriptive 
-%%%  title. Table legends are optional but encouraged. 
-%%%  Footnotes (superscript lowercase letters) should 
-%%%  be used where necessary to indicate some feature 
-%%%  of the data; please do not use bold, italic, 
-%%%  colored text, or shading for this purpose. Use 
-%%%  separate cells, not line breaks or spaces, for 
-%%%  all discrete data elements. Small embedded 
-%%%  graphics with color are OK.
-
-%%%  Like figures, all tables must be cited within 
-%%%  the main text, and our typesetting team will 
-%%%  place the tables within the typeset paper at 
-%%%  the appropriate locations.
-
-\subsection*{Table 1. A table with clear organization of data}
-
-\begin{tabular}{|l | l | l | l|} 
- \hline
- \textbf{Column 1} & \textbf{Column 2} & \textbf{Column 3} & \textbf{Column 4} \\ [1ex] 
- \hline
- Row A\textsuperscript{a} & 6 & 87,837 & 787 \\ [1ex] 
- \hline
- Row B & 7 & 78 & 5,415 \\ [1ex] 
- \hline
- Row C & 545 & 778 & 7507 \\ [1ex] 
- \hline
- Row D & 545 & 18,744 & 7,560 \\ [1ex] 
- \hline
- Row E & 88 & 788 & 6,344 \\ [1ex] 
- \hline
-\end{tabular}
-
-\bigskip
-
-The table legend (optional) follows the table itself. 
-The legend should be used to provide additional info that relates to the table as a whole.
-
-\textsuperscript{a}Footnotes can be used to provide additional info on specific content within the table, 
-such as this footnote to the first row (row A). Do not use footnotes in the table title.
-
-\textsuperscript{b}More footnotes
+% \newpage
+
+% \section*{MAIN TABLES, INCLUDING TITLES AND LEGENDS}
+
+% %%%  Whenever possible, we encourage you to submit
+% %%%  your main-text tables as Microsoft Word documents,
+% %%%  using Word's table function. This will ensure
+% %%%  the best results during conversion. Tables
+% %%%  should be numbered Table 1, Table 2, etc. and
+% %%%  should not include subpanels (do not use Table 1A,
+% %%%  1B, etc.). Give each table a brief descriptive
+% %%%  title. Table legends are optional but encouraged.
+% %%%  Footnotes (superscript lowercase letters) should
+% %%%  be used where necessary to indicate some feature
+% %%%  of the data; please do not use bold, italic,
+% %%%  colored text, or shading for this purpose. Use
+% %%%  separate cells, not line breaks or spaces, for
+% %%%  all discrete data elements. Small embedded
+% %%%  graphics with color are OK.
+
+% %%%  Like figures, all tables must be cited within
+% %%%  the main text, and our typesetting team will
+% %%%  place the tables within the typeset paper at
+% %%%  the appropriate locations.
+
+% \subsection*{Table 1. A table with clear organization of data}
+
+% \begin{tabular}{|l | l | l | l|}
+%  \hline
+%  \textbf{Column 1} & \textbf{Column 2} & \textbf{Column 3} & \textbf{Column 4} \\ [1ex]
+%  \hline
+%  Row A\textsuperscript{a} & 6 & 87,837 & 787 \\ [1ex]
+%  \hline
+%  Row B & 7 & 78 & 5,415 \\ [1ex]
+%  \hline
+%  Row C & 545 & 778 & 7507 \\ [1ex]
+%  \hline
+%  Row D & 545 & 18,744 & 7,560 \\ [1ex]
+%  \hline
+%  Row E & 88 & 788 & 6,344 \\ [1ex]
+%  \hline
+% \end{tabular}
+
+% \bigskip
+
+% The table legend (optional) follows the table itself.
+% The legend should be used to provide additional info that relates to the table as a whole.
+
+% \textsuperscript{a}Footnotes can be used to provide additional info on specific content within the table,
+% such as this footnote to the first row (row A). Do not use footnotes in the table title.
+
+% \textsuperscript{b}More footnotes
 
@@
 
-\subsubsection{Li and Stephens haplotype-copying model}
+\subsubsection*{Li and Stephens haplotype-copying model}
 
@@
 In our early attempts to build ARGs, we explored using several values of $k$.
-We ran sc2ts with $k$ set to 3, 4, and 5 on the samples collected in 2020 \cite{Zhan2023}}(REF). 
+We ran sc2ts with $k$ set to 3, 4, and 5 on the samples collected in 2020 \cite{Zhan2023}(REF).
 We would expect that a good value for $k$ should lead to parsimonious ARGs
@@
 For the full list of seeds,
-see https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml.
+see
+\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}
 
@@
 To infer the final ARG, we masked the problematic sites
-in our custom set (https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml).
+in our custom set
+\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}.
 
@@
 
-\subsubsection{Counting supporting loci for recombinant parents}
+\subsubsection*{Counting supporting loci for recombinant parents}
 When counting such loci, we considered nearly adjacent sites (separated by at most three bases) as single loci,
@@
 To construct this pseudo-phylogeny, we used \textit{pangonet},
-which mines the information on the Pango lineages in the “alias_key.json” and “lineage_notes.txt” files 
+which mines the information on the Pango lineages in the ``alias\_key.json''
+and ``lineage\_notes.txt'' files
 from the Pango designation issues (https://github.com/cov-lineages/pango-designation).
@@
 
-\subsubsection{Estimating interval lengths}
+\subsubsection*{Estimating interval lengths}
 Recombination intervals are imprecise
@@
 
-\subsubsection{Analyzing global case counts}
+\subsubsection*{Analyzing global case counts}
 To explore the timing of emergence of recombinants,

### 2025-08-23 latex/main.tex
Update recombinant intervals figure

@@
     \centering
-    \includegraphics[width=0.85\linewidth]{recombination_landscape.png}
-    \caption{Recombination landscape. (A) Distribution of the breakpoint intervals of the 386 “robust” recombination events along the genome
-(annotated with gene labels on top).
-Each line represents an interval colored by its size.
-(B) Number of overlapping breakpoint intervals (orange; left y-axis) and number of mutations along the genome (blue; right y-axis).
-The coordinates of the ten sites with the highest mutation counts are labeled.
-(C) Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
-(D) Relationship between breakpoint interval size and time between the recombination event and the parental MRCA.
-The black curve shows the theoretical expected relationship between these quantities, estimated by minimizing the sum of squared residuals given a single parameter (the mutation rate) (STAR Methods).}
+    \includegraphics[width=\linewidth]{recombinant_intervals}
+    \caption{Recombination intervals. (A) Distribution of the breakpoint
+intervals of the 356 ``robust'' recombination events along the genome
+(annotated with gene labels on top). Each line represents an interval colored by its size.
+(B) Number of overlapping breakpoint intervals.
+(C) FIXME Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
+(D) Relationship between breakpoint interval size and time between the recombination event
+and the parental MRCA. The black curve shows the theoretical expected relationship
+between these quantities, estimated by minimizing the sum of squared residuals
+given a single parameter (the mutation rate) (STAR Methods).}
     \label{fig:landscape}

### 2025-08-27 latex/main.tex
Updates from Overleaf

@@
 UShER uses a tree to represent evolutionary change via mutation, assuming no recombination \cite{Turakhia2021}.
-Given extensive evidence for recombination in SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2022},
+However, given extensive evidence for recombination in SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2022},
 the tree-like representation of UShER is no longer tenable.
@@
 This “retrospective sample grouping” successfully captured the emergence of most saltation lineages (Figure SX).
-The emergence of the Delta, BA.1 and BA.2 lineages required some manual intervention,
+The emergence of the Delta, BA.1, and BA.2 lineages required some manual intervention,
 in which specific "seed" samples were identified for unconditional inclusion
@@
 which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
-Of the nodes, 929 represent recombination events,
-of which 386 have a minimum of four nucleotides supporting the non-focal parent on both sides of the breakpoint
-(we refer to these as “robust” recombination events, Figure S1; see Document S1),
+Of the nodes, 855 represent recombination events,
+of which 356 have a minimum of four loci supporting the non-focal parent on both sides of the breakpoint
+(we refer to these as “robust” recombination events, Figure SX; see Document S1),
 reflecting the overwhelming tree-like structure of SARS-CoV-2 evolutionary history.
@@
     \caption{Evidence for recombination events in the ARG.
-The 386 “robust” recombination events detected by sc2ts are colored according to their classification by rebar (recombinant: purple, non-recombinant: gray).
+The 356 “robust” recombination events detected by sc2ts are colored according to their classification by rebar (recombinant: purple, non-recombinant: gray).
 (A) Distribution of time to the MRCA of the sc2ts-identified parents.
@@
 
-We examined the frequency of emergence of the recombinants over the course of the COVID-19 pandemic (\ref{recombinants_over_time}).
+We examined the frequency of emergence of the recombinants over the course of the COVID-19 pandemic (Figure \ref{fig:recombinants_over_time}).
 We explored two hypotheses for the timing of recombination.
@@
 Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
-sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in \ref{recombinants_over_time}).
+sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in \ref{fig:recombinants_over_time}).
 These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset.
@@
 the black curve shows the number of recombinants inferred within each week along the ARG,
-focusing on the 386 “robust” events.
+focusing on the 356 “robust” events.
 Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high.
 Overall, the number of recombinant events rose with the global number of cases
-(Figure 5B; linear regression: $1.2 + 2.9 10^{-7} x$, $p = 3.2e^{-7}$).
+(Figure \ref{fig:recombinants_over_time}B; linear regression: $1.0 + 2.7 \; 10^{-7} x$, $p = 1.1 \; 10^{-7}$).
 Recombination events also rose with the diversity of lineages in each week
-(Figure 5C; linear regression: $1.0 + 5.3 x$, $p = 1.3e^{-7}$),
+(Figure \ref{fig:recombinants_over_time}C; linear regression: $0.9 + 4.8 x$, $p = 3.4\;10^{-7}$),
 measured as the chance that two randomly drawn viruses have different Scorpio designations
@@
 We then asked which single measure best explains the number of recombination events inferred each week,
-considering $n$, $H$, $n H$, $n^2$, $n^2 H$.
+considering $n$, $H$, $n \; H$, $n^2$, $n^2 H$.
 We included the squared number of cases each week thinking that it might better predict co-infections.
-The best single predictor of recombination events involved the product of cases and expected heterozygosity ($n H$),
-with an adjusted $R^2$ of 25.3\%.
-This measure of case diversity ($n H$) explained 62\% more of the variation in recombination events per week
-than the next best predictor ($H$, adjusted $R^2$ of 15.7\%).
+The best single predictor of recombination events involved the product of cases and expected heterozygosity ($n \; H$),
+with an adjusted $R^2$ of 23.7\%.
+This measure of case diversity ($n \; H$) explained 49\% more of the variation in recombination events per week
+than the next best predictor ($H$, adjusted $R^2$ of 15.9\%).
 While consistent with the hypotheses that recombination events occur more often when infection rates are common
@@
 Finally, the filters that we applied removed all but the most strongly supported recombination events,
-although similar conclusions were obtained when all 929 recombinants were considered (STAR Methods).
+although similar conclusions were obtained when all 855 recombinants were considered (STAR Methods).
 
@@
     \centering
-    \includegraphics[width=0.85\linewidth]{recombinants_over_time.png}
+    \includegraphics[width=0.85\linewidth]{recombinants_over_time.pdf}
     \caption{Recombination events across the pandemic.
@@
 Cases are color-coded by their Scorpio labels, with the key showing those groupings with $\geq 1,000$ samples in the Viridian dataset.
-Recombinants more commonly occur (B) in weeks with high case numbers and (C) in weeks with extensive lineage variation, as estimated by high expected heterozygosity (C); solid lines show fits by linear regression, allowing for non-zero intercepts to account for unknown cases and heterozygosity not captured by the Scorpio labels.}
+Recombinants more commonly occur (B) in weeks with high case numbers and (C) in weeks with extensive lineage variation, as estimated by high expected heterozygosity; solid lines show fits by linear regression, allowing for non-zero intercepts to account for unknown cases and heterozygosity not captured by the Scorpio labels.}
     \label{fig:recombinants_over_time}
@@
 which is the chance that two lineages chosen at random
-from that week had different Scorpio designations ($\sum_{i} \sum_{j} 2 p_i p_j$).
+from that week had different Scorpio designations ($\sum_{i} \sum_{j>i} 2 p_i p_j$).
 The linear regression models reported in \textbf{Figure} \ref{fig:recombinants_over_time} allowed for a non-zero intercept,
@@
 The slopes were highly significant, and
-the p-values were confirmed to be robust by randomly permuting the predictor variable 10,000 times.
-ANOVAs were conducted to determine which single predictor ($n$, $H$, $n H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week.
-These statistical tests and graphics were conducted in Mathematica 14.2 (Wolfram 2024) and available at GITHUB LINK.
-
-In addition to the analysis of the 386 “robust” recombination events with a minimum of four nucleotides supporting each parent,
-we analyzed the relationship between the full set of 929 recombinants and case numbers.
+the p-values were confirmed by randomly permuting the predictor variable 10,000 times.
+ANOVAs were conducted to determine which single predictor ($n$, $H$, $n \; H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week.
+These statistical tests and graphics were conducted in \textit{Mathematica} 14.2 (Wolfram 2024) and available at GITHUB LINK.
+
+In addition to the analysis of the 356 “robust” recombination events with a minimum of four nucleotides supporting each parent,
+we analyzed the relationship between the full set of 855 recombinants and case numbers.
 These additional recombinants occurred disproportionately during the Delta and first Omicron wave (July, 2021 to February, 2022).
 Again, the number of recombinant events rose with the global number of cases
-(linear regression: $3.4 + 5.8 10-7 x$, $p = 0.0004$)
+(linear regression: $3.1 + 5.3 \; 10^{-7} x$, $p = 0.0006$)
 and with the diversity of lineages in each week
-(linear regression: $0.5 + 19.7 x$, $p = 5.1 10^{-13}$).
-The single best predictor of recombination events among the predictors tested ($n$, $H$, $n H$, $n^2$, $n^2 H$)
-was again case diversity ($n H$),
-with an adjusted $R^2$ of 28.7\%,
+(linear regression: $0.42 + 18.3 x$, $p = 2 \; 10^{-12}$).
+The single best predictor of recombination events among the predictors tested ($n$, $H$, $n \; H$, $n^2$, $n^2 H$)
+was again the product of cases and diversity ($n \; H$),
+with an adjusted $R^2$ of 26.9\%,
 although this was now followed closely by expected heterozygosity on its own ($H$),
-with an adjusted $R^2$ of 27.7\%.
+with an adjusted $R^2$ of 26.5\%.
 Thus, similar conclusions are reached with either dataset,
-although filtering the recombination events to the 386 with better support revealed a stronger relationship with case numbers.
+although filtering the recombination events to the 356 with better support revealed a stronger relationship with case numbers.
 

### 2025-08-27 notebooks/nb_utils.py
update nb_utils

@@ -1,6 +1,5 @@
 import collections
 from datetime import datetime, timedelta
-import hashlib
 import json
 import os
 import requests
@@ -17,16 +16,7 @@ from matplotlib import pyplot as plt
 from matplotlib.colors import rgb2hex
 import tskit_arg_visualizer as argviz
 
-
-PARENT_COLOURS = [  # Chose to be light enough that black text on top is readable
-    "#8D8",  # First parent: light green
-    "#6AD",  # Second parent: light blue
-    "#B9D",  # Third parent (if any): light purple
-    "#A88",  # Fourth parent (if any): light brown
-]
-
 TSDIR = "../data"
-
 
 def load(filename="sc2ts_viridian_v1.1.trees.tsz"):
     ts = tszip.load(os.path.join(TSDIR, filename))
@@ -339,6 +329,7 @@ class D3ARG_viz:
         )
         lineage_muts = None
         if parent_pangos is not None and self.lineage_consensus_muts is not None:
+            # Could replace with https://github.com/andersen-lab/Freyja/blob/main/freyja/data/lineage_mutations.json
             lm = self.lineage_consensus_muts.get_unique_mutations(p, parent_pangos)
             if lineage_muts is None:
                 lineage_muts = lm
@@ -372,6 +363,7 @@ class D3ARG_viz:
         highlight_mutations=None,
         highlight_nodes=True,  # Can also be a mapping of colour to node IDs
         highlight_colour="plum",
+        positions_file=None,
         oldest_y_label=None,
         node_1_date=datetime(2019, 12, 26),  # date of Wuhan, node #1
         **kwargs,
@@ -384,6 +376,15 @@ class D3ARG_viz:
         or a PosAlt named tuple with positions and derived states (e.g. from
         `MutationContainer.get_mutations(Pango)` )
         """
+        if positions_file is not None:
+            try:
+                self.d3arg.set_node_x_positions(
+                    pos=argviz.extract_x_positions_from_json(json.loads(Path(positions_file).read_text()))
+                )
+            except FileNotFoundError:
+                pass
+                # self.clear_x_01(self.d3arg)
+
         select_nodes = np.isin(self.d3arg.nodes.id, nodes)
         # Find mutations with duplicate position values and the same alleles (could be parallel eg. A->T & A->T, or reversions, e.g. A->T, T->A)
[…]

### 2025-08-27 src/save_pangoX_copying_patterns.py
Create save_pangoX_copying_patterns.py

@@ -1,0 +1,63 @@
+from pathlib import Path
+import numpy as np
+import pandas as pd
+import imgkit  # To convert the HTML table to a PNG. Also needs wkhtmltox to be installed
+import sc2ts
+import tszip
+from PIL import Image
+from io import BytesIO
+
+from tqdm import tqdm
+
+data_dir = Path(__file__).resolve().parent.parent / "data"
+png_dir = Path(__file__).resolve().parent.parent / "figures/static"
+
+
+ts = tszip.load(data_dir / "sc2ts_viridian_v1.1.trees.tsz")
+
+def pangoX_RE_node_labels(exclude_dups=True):
+    """
+    Return a dict mapping RE node IDs to labels for PangoX designations.
+    exclude_dups=False will also return pangoXs that have multiple RE nodes
+    (currently only a few XMs, not including the main XM, which is joint with XAL
+    """
+    # Only label the Pango X origins as listed in Table 1
+    # First find the nodes clearly associated with a RE event 
+    pango_x_events = pd.read_csv(data_dir / "pango_x_events.csv")
+    # Get number of descendants for RE node so we can exclude those with >1e5 descendants, i.e. BA.5
+    recombinants = pd.read_csv(data_dir / "recombinants.csv").set_index('recombinant')
+    closest_recombinant_num_samples = -np.ones(len(pango_x_events), dtype=int)
+    use = (pango_x_events.closest_recombinant >= 0).values
+    closest_recombinant_num_samples[use] = recombinants.loc[
+        pango_x_events.closest_recombinant[use],
+        'num_descendant_samples'
+    ]
+    # Exclude non-recombinants, and ondes where closes RE node is BA.5
+    pango_x = pango_x_events[np.logical_and(
+        closest_recombinant_num_samples > 0,
+        closest_recombinant_num_samples < 1e5
+    )]
+    
+    def makelabel(arr):
+        arr = sorted(arr.values, key=lambda x: (len(x), x))
+        if len(arr) == 1:
+            return arr[0]
+        if len(arr) == 2:
+            return arr[0] + "+" + arr[1]
+        else:
+            return arr[0] + "++"
+            
+    px = pango_x.groupby('closest_recombinant')['root_pango'].apply(makelabel)
+    if exclude_dups:
+        return dict(px.drop_duplicates(keep=False))
+    else:
+        return dict(px)
+
+
[…]

### 2025-08-28 latex/main.tex

@@
 \usepackage{hyperref}
+\usepackage{booktabs}
 \usepackage{amsmath}
@@
 
+% This table is generated by the notebook tab_pango_x_events
+\begin{table}
+\caption{ARG events associated with Pango X lineages.
+Each row corresponds to the oldest node assigned the specified
+Pango lineage and the ancestor of all nodes assigned that label
+(except XBB, XBB.1 and XM; see STAR methods). [MORE DETAILS]
+\label{tab:pango_x_lineages}}
+\begin{tabular}{lrlrrrrr}
+\toprule
+pango & samples & extra & muts & root & recomb & t recomb & p recomb \\
+\midrule
+XA & 39 & \{\} & 1 & 122444 & 122444 & 0 & 0 \\
+XBB & 6452 & \{BA.2:1\} & 14 & 1396207 & 1396207 & 0 & 0 \\
+XBB.1 & 2 & \{\} & 1 & 1429711 & 1429711 & 0 & 0 \\
+XBD & 30 & \{\} & 0 & 1378208 & 1378208 & 0 & 0 \\
+XBF & 185 & \{\} & 3 & 1420385 & 1420385 & 0 & 0 \\
+XBG & 25 & \{\} & 2 & 1291970 & 1291970 & 0 & 0 \\
+XBR & 1 & \{\} & 2 & 1420166 & 1420166 & 0 & 0 \\
+XC & 5 & \{\} & 1 & 414488 & 414488 & 0 & 0 \\
+XF & 16 & \{\} & 1 & 946761 & 946761 & 0 & 0 \\
+XG & 3 & \{\} & 3 & 1083412 & 1083412 & 0 & 0 \\
+XL & 64 & \{\} & 1 & 1034619 & 1034619 & 0 & 0 \\
+XM & 26 & \{BA.2:16, XAL:3\} & 0 & 1003220 & 1003220 & 0 & 0 \\
+XQ & 55 & \{BA.2:37, $\star$\} & 3 & 1058654 & 1058654 & 0 & 0 \\
+XS & 17 & \{\} & 1 & 1000242 & 1000242 & 0 & 0 \\
+XW & 32 & \{\} & 1 & 1159411 & 1159411 & 0 & 0 \\
+XY & 23 & \{\} & 2 & 1187989 & 1187989 & 0 & 0 \\
+\midrule
+XAA & 17 & \{\} & 3 & 1183815 & 1058654 & 33 & 6 \\
+XAE & 9 & \{\} & 7 & 1118099 & 964555 & 47 & 2 \\
+XAF & 1 & \{\} & 8 & 1314603 & 1177107 & 140 & 5 \\
+XAG & 6 & \{\} & 4 & 1248341 & 1058654 & 55 & 6 \\
+XAL & 3 & \{\} & 2 & 1264107 & 1003220 & 66 & 3 \\
+XAM & 21 & \{\} & 3 & 1240312 & 1058654 & 43 & 6 \\
+XAN & 7 & \{\} & 5 & 1301399 & 1189192 & 114 & 4 \\
+XAP & 20 & \{\} & 1 & 1216836 & 964555 & 67 & 4 \\
+XAV & 13 & \{\} & 3 & 1269391 & 1189192 & 106 & 7 \\
+XAZ & 133 & \{\} & 3 & 1285706 & 1189192 & 62 & 2 \\
+XBE & 65 & \{\} & 3 & 1363926 & 1189192 & 136 & 6 \\
+XBH & 2 & \{\} & 2 & 1395926 & 1379419 & 27 & 1 \\
+XBM & 10 & \{\} & 1 & 1358392 & 1348822 & 23 & 1 \\
+XE & 1116 & \{\} & 2 & 965352 & 965353 & 19 & 1 \\
+XH & 2 & \{\} & 7 & 1098084 & 965353 & 55 & 1 \\
+XJ & 68 & \{BA.2:3\} & 2 & 966904 & 966905 & 5 & 1 \\
+XR & 17 & \{BA.2:1\} & 4 & 1148222 & 1058654 & 11 & 3 \\
+XU & 1 & \{\} & 4 & 1243765 & 1058654 & 103 & 6 \\
+XZ & 48 & \{\} & 1 & 1163537 & 964555 & 48 & 4 \\
+\midrule
+XAJ & 18 & \{\} & 11 & 1276376 & -1 & inf & inf \\
+XAS & 77 & \{\} & 1 & 1265115 & -1 & inf & inf \\
+XAU & 8 & \{\} & 1 & 1231548 & -1 & inf & inf \\
+XB & 192 & \{\} & 2 & 223239 & -1 & inf & inf \\
+XBK & 7 & \{\} & 1 & 1425824 & -1 & inf & inf \\
+XBQ & 14 & \{\} & 1 & 1422955 & -1 & inf & inf \\
+XN & 120 & \{BA.2:1\} & 1 & 1061700 & -1 & inf & inf \\
+XP & 45 & \{\} & 1 & 1092789 & -1 & inf & inf \\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \subsection*{Comparison with other recombination detection methods}

### 2025-08-29 src/save_pangoX_copying_patterns.py
Add caption and rescale recombinant-evidence plot

@@ -58,6 +58,10 @@ for node_id, label in tqdm(pangoX_RE_node_labels().items()):
     img_bytes = imgkit.from_string(
     sc2ts.info.CopyingTable(ts, node_id).html(hide_extra_rows=True, hide_labels=True, show_bases=None),
         False,  # return the bytes, rather than saving to file
-        options={"width": 2000, "format": "png", "quiet": ""})
+        options={"width": 2000, "format": "png", "quiet": "", "transparent": "",})
     img = Image.open(BytesIO(img_bytes))
+    img = img.convert('RGBA')
+    # Crop whitespace
+    bbox = img.getbbox()
+    img = img.crop(bbox)
     img.save(png_dir / f"{label}.png", "PNG", optimize=True, compress_level=9)

### 2025-08-29 latex/main.tex

@@
     \centering
-    \includegraphics[width=0.85\linewidth]{averted_mutations.png}
-    \caption{Evidence for recombination events in the ARG.
-The 356 “robust” recombination events detected by sc2ts are colored according to their classification by rebar (recombinant: purple, non-recombinant: gray).
-(A) Distribution of time to the MRCA of the sc2ts-identified parents.
-(B) Distribution of the Pango lineage distance between the sc2ts-identified parents.
-(C) Distribution of the number of mutations averted by recombination (bottom x-axis), defined as the difference in the number of mutations between a copying path allowing recombination (right y-axis) and an alternative copying path disallowing recombination (top x-axis).
-(D) Copying patterns of five select recombinants.}
+    \includegraphics[width=1\linewidth]{fig_recombinant_evidence.pdf}
+    \caption{Evidence for recombination in the ARG. All plots show “robust” sc2ts recombination nodes, colored according
+to their classification by rebar (recombinant: purple, non-recombinant: gray).
+(A) Upper: jittered scatterplot comparing number of mutations when recombination allowed ($y$: right axis)
+versus an alternative copying path disallowing recombination ($x$: top axis);
+recombination nodes in Table \ref{tab:pango_x_lineages} are labelled, point sizes reflect number of descendants;
+Lower: recombination nodes classified by number of mutations averted by allowing recombination (i.e. $x$-$y$).
+(B) Distribution of time to the MRCA of the parents of each robust recombination node.
+(C) Distribution of the Pango lineage distance between parents.
+(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
+each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
     \label{fig:averted_mutations}

### 2025-08-29 notebooks/nb_utils.py
Load in the .vcz file
Even if in .zip format, you load the dataset without the .zip extension

@@ -33,7 +33,7 @@ def load(filename="sc2ts_viridian_v1.1.trees.tsz"):
     return ts
 
 
-def load_dataset(filename="viridian_mafft_2024-10-14_v1.vcz.zip"):
+def load_dataset(filename="viridian_mafft_2024-10-14_v1.vcz"):
     return sc2ts.Dataset(os.path.join(TSDIR, filename), date_field="Date_tree")
 
 

### 2025-08-31 src/makepdf.py
Add a notebook and PDF for the Fig 2 subgraphs

@@ -2,6 +2,7 @@
 import argparse
 import subprocess
 from pathlib import Path
+import os
 from tempfile import NamedTemporaryFile
 
 if __name__ == "__main__":
@@ -12,6 +13,8 @@ if __name__ == "__main__":
         )
     )
     argparser.add_argument("input_ipynb", help="Path to input ts or tsz file", type=str)
+    argparser.add_argument("outdir", nargs="?", default=None, help="Path to output directory", type=str)
+
     args = argparser.parse_args()
     # find absolute path to the input file
     infile = Path(args.input_ipynb).resolve(strict=True)
@@ -41,7 +44,12 @@ if __name__ == "__main__":
             else:
                 output.write(line)
         output.flush()
-        outfile = str(infile.with_suffix(''))
+        if args.outdir is None:
+            args.outdir = infile.parent
+        else:
+            args.outdir = Path(os.getcwd()) / args.outdir
+        outfile = str(infile.with_suffix('').name)
+        print(f"PDF saved to {args.outdir}/{outfile}")
         cmd = [
             "jupyter",
             "nbconvert",
@@ -51,7 +59,7 @@ if __name__ == "__main__":
             "--no-input",
             output.name,
             "--output",
-            outfile,
+            args.outdir / outfile,
         ]
         subprocess.run(cmd, check=True)
 

### 2025-09-02 latex/main.tex
Updates from Overleaf

--- 2025-08-28 latex/main.tex
@@
 particularly increased transmissibility, disease severity, and
-ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2022}).
+ability to escape natural or vaccine-induced immunity (e.g., XBB\cite{Tamura2022}).
 The process of gathering evidence for and characterizing such recombinants is manual,
@@
 This process has resulted in the designation of recombinant Pango lineages
-(whose aliases begin with an “X”), to which we refer as “Pango Xs”.
+(whose aliases begin with an “X”), to which we refer as “Pango X”.
 Below, we summarize our analysis of the Pango Xs in the ARG
@@
 
-Sc2ts identified 12 Pango Xs in the ARG as non-recombinants (Table 1),
-on the basis of the HMM and available data.
-For all these Pango Xs, the number of recombination-informative sites
-fell below our threshold of detection (i.e. $k = 4$ in the HMM).
+Sc2ts identified 12 Pango Xs in the ARG as non-recombinants (Table 1), on the basis of the HMM and available data.
+For all these Pango Xs, the number of recombination-informative sites fell below our threshold of detection (i.e. $k = 4$ in the HMM).
 The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG
 by at most three mutations (Table 1) strongly supports their origins as non-recombinants.
-For example, the closely related XN and XAU can be explained
-by two mutations and one mutation in the ARG, respectively (\ref{fig:subgraphs}).
+For example, the closely related XN and XAU can be explained by two mutations and one mutation in the ARG, respectively (\ref{fig:subgraphs}).
 %%% TODO: Say something about XAJ and XAN.
@@
 \caption{Illustrative subgraphs in the ARG.
-Subgraphs around the XA-associated recombination node (depicted as an encircled R) (A); the recombination node associated with a group of “nested” Pango Xs (B); and the MRCA of the putative non-recombinants XN and XAU (C).
+Subgraphs around the XA-associated recombination node (A); the recombination node associated with a group of “nested” Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C).
 For more detailed visualizations of these subgraphs, see Document S2.}
@@
 \end{figure}
-
-\subsection*{Table 1. Recombination events associated with Pango Xs.}
 
@@
 %%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
-We compared sc2ts against three state-of-the-art (non-phylogenetic) methods that detect inter-lineage recombinants
-by scanning sequences for mutations associated with different Pango lineages:
+We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
 (1) RecombinHunt \cite{Alfonsi2024},
@@
 (3) rebar (https://github.com/phac-nml/rebar).
+These non-phylogenetic methods use the same underlying approach: 
+they scan candidate recombinant sequences for mutations associated with different Pango lineages.
 
@@
 
-\begin{tabular}{|l | l | l | l|}
+\begin{table}
+\begin{tabular}{llll}
  \hline
@@
 \end{tabular}
+\end{table}
 
@@
 We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
-particularly at the left boundary of the Spike gene (\ref{fig:landscape}),
+particularly at the left boundary of the Spike gene (\ref{fig:recombinant_intervals}),
 consistent with the pattern of excess observed in previous studies \cite{Turakhia2022,Li2024}.
@@
 allowing recombination endpoints to be estimated more precisely.
-Reassuringly, sites having unusually high mutation counts are absent in this region (\ref{fig:landscape}),
+Reassuringly, sites having unusually high mutation counts are absent in this region (\ref{fig:recombinant_intervals}),
 which would confound recombination detection.
@@
 breakpoint interval size and the amount of divergence between recombinant parents,
-which is measured by time to their MRCA (\ref{fig:landscape}).
+which is measured by time to their MRCA (\ref{fig:recombinant_intervals}).
 When the two parental lineages have diverged for longer periods of time (the rightmost points),
@@
 For instance, the breakpoint intervals tend to be the smallest for events involving recombination
-between the deeply divergent Omicron BA.1 and BA.2 sublineages (the red points in \ref{fig:landscape}).
+between the deeply divergent Omicron BA.1 and BA.2 sublineages (the red points in \ref{fig:recombinant_intervals}).
 
@@
     \caption{Recombination intervals. (A) Distribution of the breakpoint
-intervals of the 356 ``robust'' recombination events along the genome
+intervals of the 356 robust recombination events along the genome
 (annotated with gene labels on top). Each line represents an interval colored by its size.
@@
 given a single parameter (the mutation rate) (STAR Methods).}
-    \label{fig:landscape}
+    \label{fig:recombinant_intervals}
 \end{figure}
@@
 We contribute several novel methods and concepts which may find applications beyond sc2ts.
-First, we devise a simplified LS model, which extends the classic concept of parsimony.
+First, we devise a simplified LS model that extends the classic concept of parsimony.
 In sc2ts, this model allows for accelerated HMM likelihood calculations,
 enabling reconstruction of the pandemic-scale ARG presented here.
-Existing HMMs used in statistical genetics may be modified in a similar fashion for significant speedup.
+Existing HMMs used in statistical genetics may be modified in a similar way for significant speedup.
 Second, we propose the concept of mutations averted by recombination,
-which leads to an interpretable quantity to weigh
-the support for the hypothesis that a given sample is a recombinant
-against the hypothesis that it is a non-recombinant,
+which leads to an interpretable quantity to weigh the support for 
+the hypothesis that a given sample is a recombinant against the hypothesis that it is a non-recombinant,
 considering the sampled genetic diversity in an ARG and a specific HMM.
@@
 (e.g., in the Viridian UShER phylogeny, parent and child nodes do not necessarily follow time order \cite{Hunt2024}),
-resulting in a clean dataset rid of samples with erroneous collection dates that may affect downstream analyses.
+resulting in a cleaner dataset without samples with wrong collection dates that may affect downstream analyses.
 Future work may be estimating dates for the suspected time traveling samples using the ARG provided here
@@
 Recombination is an important evolutionary force in viruses \cite{Perez-Losada2015},
-but has been understudied in part due to a lack of suitable data models, methods, and tools \cite{Neches2020}.
-While the methods and tools developed here are tailored for SARS-CoV-2,
-they can be extended and adapted for other viruses.
-A major advantage of the sc2ts methods and tools is
-that they are built on top of the mature software libraries of tskit and VCF-Zarr,
+but has been understudied in part due to the lack of suitable data models, methods, and tools \cite{Neches2020}.
+While the methods and tools developed here are tailored for SARS-CoV-2, they can be extended and adapted for other viruses.
+A major advantage of the sc2ts methods and tools is that they are built on top of the mature software libraries of tskit and VCF-Zarr,
 which enable pandemic-scale analyses that can be performed on typical laptops.

### 2025-09-02 latex/docS1.tex
Copy in the old table from the preprint

@@
 \usepackage{orcidlink}
+\usepackage{booktabs}
+\usepackage{tabularx}
+\usepackage{makecell}
 \usepackage[super,comma,sort&compress]
@@
 
-\subsubsection*{XAN & XAV}
+\subsubsection*{XAN and XAV}
 https://github.com/jeromekelleher/sc2ts-paper/issues/353
@@
 
+% FIXME there's too much vspace between the rows here. Probably best to redo
+% the formatting.
+\begin{table} \centering
+\begin{tabular}{l|c|c|c|c|c} \hline
+\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
+\multicolumn{2}{c}{\textbf{sc2ts (Wide ARG)}} \\ \hline
+\textbf{Sample} &
+\textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\
+interval(s)} &
+\textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\
+\hline ALDP-11CF93B & A &
+    \thead{B.1.177 \\ Alpha} & 21,256--21,615 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256--22,228 \\
+ALDP-125C4D7 & A &
+    \thead{B.1.177 \\ Alpha} & 21,256--21,615 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256--22,228 \\
+ALDP-130BB95 & A &
+    \thead{B.1.177 \\ Alpha} & 21,256--21,615 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256--22,228 \\
+LIVE-DFCFFE & A &
+    \thead{B.1.177 \\ Alpha} & 18,999--20,296 &
+    \thead{B.1.177.18 \\ Alpha} & 21,256--22,228 \\
+QEUH-CCCB30 & B &
+    \thead{B.1.36.28 \\ Alpha} & 6,529--6,955 &
+    \thead{B.1.36 \\ Alpha} & 6,529--6,955 \\
+QEUH-CD0F1F & B &
+    \thead{B.1.36.28 \\ Alpha} & 6,529--6,955 &
+    \thead{B.1.36 \\ Alpha} & 6,529--6,955 \\
+MILK-1166F52 & C &
+    \thead{Alpha \\ B.1.221.1} & 25,997--27,443 &
+    \thead{Alpha \\ B.1.221} & 25,997--27,973 \\
+MILK-11C95A6 & C &
+    \thead{Alpha \\ B.1.221.1} & 25,997--27,443 &
+    \thead{Alpha \\ B.1.221} & 25,997--27,973 \\
+QEUH-109B25C & C &
+    \thead{Alpha \\ B.1.221.1} & 25,997--27,443 &
+    \thead{Alpha \\ B.1.221} & 25,997--27,973 \\
+MILK-126FE1F & D &
+    \thead{B.1.36.39 \\ Alpha} & 20,704--23,064 &
+    \thead{B.1.36.39 \\ Alpha} & 22,445--23,064 \\
+RAND-12671E1 & D &
+    \thead{B.1.36.39 \\ Alpha} & 20,704--23,064 &
+    \thead{B.1.36.39 \\ Alpha} & 22,445--23,064 \\
+RAND-128FA33 & D &
+    \thead{B.1.36.39 \\ Alpha} & 20,704--23,064 &
+    \thead{B.1.36.39 \\ Alpha} & 22,445--23,064 \\
+CAMC-CBA018 & n/a &
+    \thead{B.1.177 \\ Alpha} & 20,390--21,256 &
+    \thead{B.1.177 \\ Alpha} & 17,616--21,256 \\
+CAMC-CB7AB3 & n/a &
+    \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,268--4,476 \\ 20,390--21,256} &
+    \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,268--5,389\\17,616--21,256} \\
+MILK-103C712 & n/a &
+    \thead{B.1.177.17 \\ Alpha} & \thead{409--446 \\ 26,802--27,878} &
+    n/a & n/a \\
+QEUH-1067DEF & n/a &
+    \thead{Alpha \\ B.1.177.9} & 10,524--10,871 &
+    \thead{Alpha \\ B.1.177} & 7,729--10,871 \\ \hline
+\end{tabular}
+\caption{\label{tab:jackson_supplement}
+Recombinant sequences involving the
+Alpha (B.1.1.7) variant reported by \cite{Jackson2021-ik} have recombinant
+ancestry in the Wide ARG. The breakpoint intervals and Pango lineage
+assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et
+al., except the Pango lineage assignment of the parents of group B
+recombinants, which were taken from Table 1 (motif-based results).
+3SEQ interval coordinate modifications are described in the caption for
+Table~\ref{tab:jackson}.}
+\end{table}
+
 \section*{Supplementary Figures}
@@
 
-\noindent\includegraphics[width=0.85\linewidth]{FigureS1.png}
+% JK commenting these out as we don't have the figures and want this to compile
+% \noindent\includegraphics[width=0.85\linewidth]{FigureS1.png}
 
@@
 
-\noindent\includegraphics[width=0.85\linewidth]{FigureS2.png}
+% \noindent\includegraphics[width=0.85\linewidth]{FigureS2.png}
 
@@
 
-\noindent\includegraphics[width=0.85\linewidth]{FigureS3.png}
+% \noindent\includegraphics[width=0.85\linewidth]{FigureS3.png}
 

### 2025-09-02 latex/docS1.tex
Update the Jackson et al numbers

--- 2025-08-20 latex/docS1.tex
@@
 
-\subsection*{Analysis of Pango lineage “origination” events}
+\subsection*{Analysis of Pango lineage ``origination'' events}
 Notebook
@@
 
-\subsection*{Analysis of the Jackson et al. (2022) recombinants}
-Jackson et al. (2022) conducted detailed analyses on some early recombinants involving B.1.1.7 (Alpha), 
-including the first designated Pango X lineage XA. 
-The authors first performed a targeted search for candidate recombinants 
-by scanning sampled sequences for sequence motifs associated with Alpha, 
-followed by an analysis to reduce the list of plausible recombinants displaying signs of onward transmission. 
-Here we compared our sc2ts detection results of these recombinants with the detection results reported by Jackson et al. (2022).
+\subsection*{Analysis of the Jackson et al. (2021) recombinants}
+Jackson et al. (2021) conducted detailed analyses on some early recombinants
+involving B.1.1.7 (Alpha), including the first designated Pango X lineage, XA.
+The authors first performed a targeted search for candidate recombinants by
+scanning sampled sequences for sequence motifs associated with Alpha, followed
+by an analysis to reduce the list of plausible recombinants displaying signs of
+onward transmission. Here we compared our sc2ts detection results of these
+recombinants with the detection results reported by Jackson et al.
 
-Jackson et al. (2022) described four groups of recombinants (named A to D) and four singleton recombinants, 
-which have distinct inheritance structures (i.e. distinct parental Pango lineages and/or distinct recombination breakpoint intervals). 
-All the four groups and 3 out of the 4 singletons were detected as recombinants in the ARG. 
-The other singleton was detected as a recombinant by sc2ts, 
-but it was not added to the ARG due to a high HMM cost (two breakpoints with X mutations). 
-The sc2ts detection results are overall concordant with the results of Jackson et al. (2022) in terms of 
-parental Pango lineages (agreeing to at least three-level resolution) and breakpoint intervals (Table S1).
+Jackson et al.\ described four groups of recombinants (named A to D) and
+four singleton recombinants.
+Of the 16 samples analysed by Jackson et al, 12 were included in the ARG;
+the remaining 4 were excluded because the sequences did not pass the 
+QC filtering criteria, or due to a high HMM cost. Groups A-D are summarised 
+in Table~\ref{tab:jackson} and show excellent concordance in terms 
+of the parental Pango lineages and breakpoint intervals. All four groups
+are directly associated with strongly supported recombination events 
+in the ARG in terms of the number of averted mutations 
+(A: 16, B: 6, C: 30, D: 18). 
+The singleton CAMC-CBA018 has parents
+B.1.177+B.1.1.7 and breakpoint interval 16,177--20,133 
+in the ARG 
+and is strongly supported by 14 averted mutations.
+The Jackson et al.\ results are almost identical with the same parent 
+lineages, and breakpoint interval of interval 20,390--21,256.
+The singleton MILK-103C712 
+has parents B.1.177.16+B.1.1.7 and a breakpoint interval of 
+24915--27972 
+in the ARG, again strongly supported by 11 averted mutations.
+The Jackson et al.\ results for this lineage 
+has parents B.1.177.17+B.1.1.7 but two breakpoints
+409--446 and 26,802--27,878.
 
-Table S1: Sc2ts detection results for the recombinants documented by Jackson et al. (2022).
-https://github.com/jeromekelleher/sc2ts-paper/issues/293
+\begin{table} \centering
+\caption{\label{tab:jackson}Comparison of recombination breakpoint
+intervals and parent lineages for Groups A-D
+reported by Jackson et al.\ with the corresponding
+recombination events in the sc2ts ARG.
+The second column gives the number of sequences in the group.
+See the text for details of the groups and the sequences included.
+limited to the samples considered by Jackson et al.
+The 3SEQ coordinates
+reported by Jackson et al.\ have been altered as follows: we
+add one to both left and right coordinates to correspond to the
+\texttt{tskit} definition of inheritance on either side of a breakpoint,
+and add one to the right coordinate to make the intervals
+right-exclusive. }
+\begin{tabular}{lllr@{--}lr@{+}l}
+\toprule
+Group        & Sequences & Method & \multicolumn{2}{c}{Interval}
+    & \multicolumn{2}{r}{Parent lineages} \\
+\midrule
+A (XA)       & 2   & Jackson        &  21,256&21,615 & B.1.177&B.1.1.7 \\
+             &     &\texttt{sc2ts} &  20,411&21,765 & B.1.177.18&B.1.1.7 \\
+\cmidrule{3-7}
+B            & 2   & Jackson        &  6,529&6,955 & B.1.36.28&B.1.1.7  \\
+             &     &\texttt{sc2ts} &   6,529&6,954 & B.1.36.28&B.1.1.7  \\
+\cmidrule{3-7}
+C            & 3   &Jackson         &  25,997&27,443 &  B.1.1.7&B.1.221 \\
+             &     & \texttt{sc2ts} &  25,997&27,972 &  B.1.1.7&B.1.221.1 \\
+\cmidrule{3-7}
+D            & 3   & Jackson        &  21,576&23,064 &  B.1.36.17&B.1.1.7 \\
+             &     & \texttt{sc2ts} &  22,445&23,063 &  B.1.36.39&B.1.1.7 \\
+\bottomrule
+\end{tabular}
+\end{table}
 

### 2025-09-02 latex/main.tex
Updates from Overleaf

@@
 However, important SARS-CoV-2 recombinant lineages have emerged.
-A more suitable approach is to use ancestral recombination graphs (ARG),
+A more suitable approach is to use ancestral recombination graphs (ARGs),
 which jointly model recombination, mutation, and genetic inheritance in a cohesive framework.
@@
 
-Using sc2ts, we built an ARG of ~2.48 million-samples spanning the COVID-19 pandemic.
+Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
 Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM (~5 CPU years; Figure SX).
@@
 (we refer to these as “robust” recombination events, Figure SX; see Document S1),
-reflecting the overwhelming tree-like structure of SARS-CoV-2 evolutionary history.
+reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.
 
@@
 We present sc2ts, a real-time ARG inference method,
-and applied it to build a \~2.48 million-sample pandemic-scale ARG of SARS-CoV-2.
+and applied it to build a $\sim$2.48 million-sample pandemic-scale ARG of SARS-CoV-2.
 By identifying specific samples as parents,
@@
 for a variety of evolutionary and genetic analyses beyond those presented here.
-We hope that the ARG will spur on new research areas,
-in which recombination is naturally and fully accounted for,
-and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
-has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
+% We hope that the ARG will spur on new research areas,
+% in which recombination is naturally and fully accounted for,
+% and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
+% has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
+So far, recent developments in ARG methods have focused almost 
+exclusively on applications to human sequencing data. In that space,
+the availability of ARGs reconstructed from large datasets has already
+had tremendous impact on improving our understanding of human genetic
+ancestry and past demography \cite{Wohns2022}. We expect
+similar leaps to be enabled through tailoring ARG reconstruction methods
+to data from other species and pathogens. Specifically, existing
+inference methods can be adapted to use the sc2ts ARG for the 
+genealogy-based study of selection \cite{hejase2022deep} or the
+geographic spread of the virus through time
+\cite{grundler2025geographic}. This is in addition to simpler
+analyses, such as calculating mutation spectra or querying the number
+or frequency of mutations at a given locus, which the ARG makes 
+significantly more efficient, and ultimately more easily reproducible.
+Since the ARG is a condensed representation of the genetic information
+contained in the sequencing data, it can also be used to understand,
+for instance, how the variability in sequencing efforts across time 
+and location affects our ability to reconstruct the origin of 
+important lineages and detect new recombinants.
+Since reconstructing the ARG is computationally expensive, 
+and checking and improving various aspects of its quality is laborious, 
+we hope that making this resource openly available will save
+significant time and 
+effort for other researchers and enable immediate focus on analysing 
+the information contained within this data. In-depth analysis of the
+ARG is also incredibly useful for identifying important directions for
+further research, both regarding methods (where current models do not
+correctly capture features of the data) and data processing 
+(where gaps in sequencing or data artifacts result in clear
+reconstruction errors).
 

### 2025-09-02 latex/docS1.tex
Rename the jackson recombs

@@
 
+% Data derived from the notebooks/jackson_recombinants.ipynb notebook
 Jackson et al.\ described four groups of recombinants (named A to D) and
@@
 
+\begin{figure} \centering
+\includegraphics[width=0.5\textwidth]{LS_model_schematic}
+\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
+model, in which a focal sequence (bottom) is described as an
+imperfect mosaic of the sequences in a reference panel.
+Black crosses along the focal sequence show sequencing
+errors or mutations.
+In the standard formulation, at site $\ell$, the recombination probability is $r_\ell$,
+the mutation probability is $\mu_\ell$ and $n$
+denotes the size of the reference panel.
+The Viterbi algorithm can be used to find a
+``copying path'' through the reference panel for a given focal sequence that
+maximises the likelihood under these parameters. Unseen states in the reference panel are shown as coloured lines enclosed by
+the grey box. The black arrow describes the true path through the data which leads to the emitted
+focal sequence below. Examples of transition and
+emission probabilities along this trajectory are shown by the red and blue
+arrows, respectively.
+}
+\end{figure}
+
 \end{document}

### 2025-09-03 latex/docS1.tex

@@
 four singleton recombinants.
-Of the 16 samples analysed by Jackson et al, 12 were included in the ARG;
+Of the 16 samples analyzed by Jackson et al, 12 were included in the ARG;
 the remaining 4 were excluded because the sequences did not pass the 
-QC filtering criteria, or due to a high HMM cost. Groups A-D are summarised 
+QC filtering criteria, or due to a high HMM cost. Groups A-D are summarized 
 in Table~\ref{tab:jackson} and show excellent concordance in terms 
@@
 (A: 16, B: 6, C: 30, D: 18).
-The singleton CAMC-CBA018 has parents
-B.1.177+B.1.1.7 and breakpoint interval 16,177--20,133 
+
+The singleton CAMC-CBA018 has 
+the parents B.1.177+B.1.1.7 and the breakpoint interval 16,177--20,133 
 in the ARG 
 and is strongly supported by 14 averted mutations.
-The Jackson et al.\ results are almost identical with the same parent 
-lineages, and breakpoint interval of interval 20,390--21,256.
-The singleton MILK-103C712 
-has parents B.1.177.16+B.1.1.7 and a breakpoint interval of 
-24915--27972 
+The Jackson et al.\ results are almost identical with the same parent lineages, 
+and breakpoint interval of interval 20,388–-21,256.
+
+The singleton MILK-103C712 has 
+the parents B.1.177.16+B.1.1.7 and a breakpoint interval of 24915--27972 
 in the ARG, again strongly supported by 11 averted mutations.
-The Jackson et al.\ results for this lineage 
-has parents B.1.177.17+B.1.1.7 but two breakpoints
-409--446 and 26,802--27,878.
+The Jackson et al.\ results for this lineage has 
+the parents B.1.177.17+B.1.1.7 but two breakpoints 407–-446 and 26,800–-27,878.
 
@@
 limited to the samples considered by Jackson et al.
-The 3SEQ coordinates
-reported by Jackson et al.\ have been altered as follows: we
-add one to both left and right coordinates to correspond to the
-\texttt{tskit} definition of inheritance on either side of a breakpoint,
-and add one to the right coordinate to make the intervals
-right-exclusive. }
+The 3SEQ coordinates reported by Jackson et al.\ have been altered as follows: 
+we subtract one to the left coordinates and add one to the right coordinates 
+to correspond to the \texttt{tskit} definition of inheritance on either side of a breakpoint,
+and add one to the right coordinate to make the intervals right-exclusive. }
 \begin{tabular}{lllr@{--}lr@{+}l}
@@
 \midrule
-A (XA)       & 2   & Jackson        &  21,256&21,615 & B.1.177&B.1.1.7 \\
+A (XA)       & 2   & Jackson        &  21,255&21,613 & B.1.177&B.1.1.7 \\
              &     &\texttt{sc2ts} &  20,411&21,765 & B.1.177.18&B.1.1.7 \\
 \cmidrule{3-7}
-B            & 2   & Jackson        &  6,529&6,955 & B.1.36.28&B.1.1.7  \\
+B            & 2   & Jackson        &  6,528&6,953 & B.1.36.28&B.1.1.7  \\
              &     &\texttt{sc2ts} &   6,529&6,954 & B.1.36.28&B.1.1.7  \\
 \cmidrule{3-7}
-C            & 3   &Jackson         &  25,997&27,443 &  B.1.1.7&B.1.221 \\
+C            & 3   &Jackson         &  25,996&27,441 &  B.1.1.7&B.1.221 \\
              &     & \texttt{sc2ts} &  25,997&27,972 &  B.1.1.7&B.1.221.1 \\
 \cmidrule{3-7}
-D            & 3   & Jackson        &  21,576&23,064 &  B.1.36.17&B.1.1.7 \\
+D            & 3   & Jackson        &  21,575&23,063 &  B.1.36.17&B.1.1.7 \\
              &     & \texttt{sc2ts} &  22,445&23,063 &  B.1.36.39&B.1.1.7 \\

### 2025-09-03 latex/docS1.tex
Updates from Overleaf

@@
 (A: 16, B: 6, C: 30, D: 18).
-
 The singleton CAMC-CBA018 has 
@@
 The Jackson et al.\ results are almost identical with the same parent lineages, 
-and breakpoint interval of interval 20,388–-21,256.
+and breakpoint interval of interval 11,395–21,993.
 
-The singleton MILK-103C712 has 
 the parents B.1.177.16+B.1.1.7 and a breakpoint interval of 24915--27972 
 in the ARG, again strongly supported by 11 averted mutations.
-The Jackson et al.\ results for this lineage has 
-the parents B.1.177.17+B.1.1.7 but two breakpoints 407–-446 and 26,800–-27,878.
+The Jackson et al.\ results for this lineage has the same parents,
+with the breakpoint interval 26,800–-27,974.
+The singleton QEUH-1067DEF has the parents B.1.1.7+B.1.177.9 
+and a breakpoint interval of 7,729--10,870 in the ARG, 
+supported by 11 averted mutations.
+The Jackson et al.\ results for this lineage has the same parents,
+with the breakpoint interval 6,953--10,872.
 
@@
 limited to the samples considered by Jackson et al.
-The 3SEQ coordinates reported by Jackson et al.\ have been altered as follows: 
+The breakpoint coordinates in Table 1 of Jackson et al.\ have been altered as follows: 
 we subtract one to the left coordinates and add one to the right coordinates 
@@
 \midrule
-A (XA)       & 2   & Jackson        &  21,255&21,613 & B.1.177&B.1.1.7 \\
+A (XA)       & 2   & Jackson        &  21,254&21,767 & B.1.177&B.1.1.7 \\
              &     &\texttt{sc2ts} &  20,411&21,765 & B.1.177.18&B.1.1.7 \\
 \cmidrule{3-7}
-B            & 2   & Jackson        &  6,528&6,953 & B.1.36.28&B.1.1.7  \\
+B            & 2   & Jackson        &  6,527&6,956 & B.1.36.28&B.1.1.7  \\
              &     &\texttt{sc2ts} &   6,529&6,954 & B.1.36.28&B.1.1.7  \\
 \cmidrule{3-7}
-C            & 3   &Jackson         &  25,996&27,441 &  B.1.1.7&B.1.221 \\
+C            & 3   &Jackson         &  24,913&28,653 &  B.1.1.7&B.1.221 \\
              &     & \texttt{sc2ts} &  25,997&27,972 &  B.1.1.7&B.1.221.1 \\
 \cmidrule{3-7}
-D            & 3   & Jackson        &  21,575&23,063 &  B.1.36.17&B.1.1.7 \\
+D            & 3   & Jackson        &  21,574&23,065 &  B.1.36.17&B.1.1.7 \\
              &     & \texttt{sc2ts} &  22,445&23,063 &  B.1.36.39&B.1.1.7 \\

### 2025-09-03 notebooks/nb_utils.py
New subgraph layouts, and show the pango for recombinant nodes

@@ -22,8 +22,8 @@ def load(filename="sc2ts_viridian_v1.1.trees.tsz"):
     ts = tszip.load(os.path.join(TSDIR, filename))
     time_zero_date = ts.metadata.get("time_zero_date", "Unknown date")
     print(
-        f"Loaded {ts.nbytes/1e6:0.1f} megabyte SARS-CoV2 ARG to "
-        f"{time_zero_date}: {ts.num_samples} strains"
+        f"Using a {ts.nbytes/1e6:0.1f} megabyte ARG up to "
+        f"{time_zero_date}, with {ts.num_samples} sampled SARS-CoV2 sequences"
     )
     n_RE = np.sum(ts.nodes_flags & sc2ts.NODE_IS_RECOMBINANT != 0)
     print(
@@ -214,18 +214,18 @@ def mutation_p_values(ts, min_time=1, progress=True):
 
 
 class D3ARG_viz:
-    def __init__(self, ts, df, lineage_consensus_muts=None, pangolin_field="pango"):
+    def __init__(self, ts, df, lineage_consensus_muts=None, pangolin_field="pango", progress=True):
         self.ts = ts
         self.df = df
         self.pangolin_field = pangolin_field
         self.lineage_consensus_muts = lineage_consensus_muts
-        self.d3arg = argviz.D3ARG.from_ts(ts, progress=True)
+        self.d3arg = argviz.D3ARG.from_ts(ts, progress=progress)
         self.d3arg.nodes.loc[self.d3arg.nodes.id == 1, "label"] = "Wuhan"
         self.pango_lineage_samples = (
             df[df.is_sample].groupby(pangolin_field)["node_id"].apply(list).to_dict()
         )
 
-    def set_sc2ts_node_labels(self, add_strain_names=True):
+    def set_sc2ts_node_labels(self, add_strain_names=True, progress=True):
         # Set node labels to Pango lineage + strain, if it exists
         def label_lines(row):
             if getattr(row, "sample_id", "") == "Vestigial_ignore":
@@ -235,7 +235,8 @@ class D3ARG_viz:
 
         node_labels = {}
         for row in tqdm(
-            self.df.itertuples(), total=len(self.df), desc="Setting all labels"
+            self.df.itertuples(), total=len(self.df), desc="Setting all labels",
+            disable=not progress,
         ):
             node_labels[row.node_id] = "\n".join(
                 [s for s in label_lines(row) if s not in ("()", "?")]
[…]

### 2025-09-03 src/makepdf.py

@@ -28,14 +28,20 @@ if __name__ == "__main__":
                 start, end = line.rsplit(r'\n', 1)
                 output.write(start + r'.append(\"a\").attr(\"href\", d => \"mut:\" + d.label)\n' + end)
             #  Line labels where the second line starts with (DRR, (ERR, or (SRR
-            #  are turned into a link with the second line as the URL)
-            #  This is EXTEREMLY fragile. 
+            #  are turned into a link with the second line as the URL. Line labels where
+            #  the first line is   
+            #  This is EXTREMELY fragile. 
             elif line.rstrip().endswith(r'const lines = text.split(\"\\n\");\n",'):
                 output.write(r'''       "                let lines = text.split(\"\\n\");\n",
        "                let url = null;\n",
-       "                if ((lines.length == 2) && (lines[1].startsWith('(DRR') || lines[1].startsWith('(ERR') || lines[1].startsWith('(SRR'))) {\n",
-       "                        url = 'sample:' + lines[1].replace(/[()]/g, ''); lines = [lines[0]];\n",
-       "                    }\n",''')
+       "                if (lines.length == 2) {\n",
+       "                  if (lines[1].startsWith('(DRR') || lines[1].startsWith('(ERR') || lines[1].startsWith('(SRR')) {\n",
+       "                    url = 'sample:' + lines[1].replace(/[()]/g, ''); lines = [lines[0]];\n",
+       "                  }\n",
+       "                  else if (lines[1].startsWith('/')) {\n",
+       "                    lines = [lines[1], lines[0]];\n",
+       "                  }\n",
+       "                }\n",''')
             elif line.rstrip().endswith(r'''d3.select(this).selectAll('tspan')\n",'''):
                 output.write(r'''       "                const container = url ? \n",
        "                    d3.select(this).append(\"a\").attr(\"href\", url) :\n",

### 2025-09-03 latex/main.tex
Add brief Jackson explanation to Fig 3 caption

@@
     \centering
-    \includegraphics[width=0.85\linewidth]{methodology.png}
+    \includegraphics[width=0.85\linewidth]{sc2ts_methodology}
     \caption{Schematic overview of sc2ts.
@@
 versus an alternative copying path disallowing recombination ($x$: top axis);
-recombination nodes in Table \ref{tab:pango_x_lineages} are labelled, point sizes reflect number of descendants;
+recombination nodes in Table \ref{tab:pango_x_lineages} are labelled starting with `X' and
+those identified by \cite{Jackson2021} (see Table \ref{tab:jackson}) are prefixed with a `J', e.g.
+Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1). 
+Point sizes reflect number of descendants.
 Lower: recombination nodes classified by number of mutations averted by allowing recombination (i.e. $x$-$y$).

### 2025-09-03 latex/docS1.tex

@@
 
+\subsection{Major deletions in the ARG}
+% Referenced by the main Results.
+Basically, we want to say we did a lot of work to integrate signals of major indels into the ARG,
+making it an even more powerful resource overall.
+Also, indels are encoded by the Mutation-Annotated Tree format used by UShER phylogenies (though PanMAT does this).
+
+% See https://github.com/jeromekelleher/sc2ts-paper/issues/472
+At minimum, we want to show compare the frequencies of the top N indels in the ARG and those reported in Li et al. (2023).
+This can be shown in a table.
+
 \subsection*{Identification of robust recombination events}
@@
 
-To analyse how well the ARG reflects the phylogenetic structure implicit in the Pango lineage naming system,
+To analyze how well the ARG reflects the phylogenetic structure implicit in the Pango lineage naming system,
 we considered the Pango assignments generated by Pangolin on the alignments for each node in the ARG (STAR Methods).
@@
 
-\subsection*{Analysis of the Jackson et al. (2021) recombinants}
-Jackson et al. (2021) conducted detailed analyses on some early recombinants
+\subsection*{Analysis of the Jackson et al. recombinants}
+Jackson et al.\cite{Jackson2021} conducted detailed analyses of some early recombinants
 involving B.1.1.7 (Alpha), including the first designated Pango X lineage, XA.
@@
 % Data derived from the notebooks/jackson_recombinants.ipynb notebook
-Jackson et al.\ described four groups of recombinants (named A to D) and
+% See https://github.com/jeromekelleher/sc2ts-paper/issues/293
+Jackson et al. described four groups of recombinants (named A to D) and
 four singleton recombinants.
@@
 The singleton CAMC-CBA018 has 
-the parents B.1.177+B.1.1.7 and the breakpoint interval 16,177--20,133 
-in the ARG 
+the parents B.1.177+B.1.1.7 and the breakpoint interval 16,177--20,133 in the ARG, 
 and is strongly supported by 14 averted mutations.
@@
 and breakpoint interval of interval 11,395–21,993.
-
+The singleton MILK-103C712 has 
 the parents B.1.177.16+B.1.1.7 and a breakpoint interval of 24915--27972 

### 2025-09-03 latex/main.tex
Some analysis

@@
 \label{tab:pango_x_lineages}}
-\begin{tabular}{lrlrrrrr}
+\begin{tabular}{lrrrrl}
 \toprule
-pango & samples & extra & muts & root & recomb & t recomb & p recomb \\
+\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
+pango & averted & left & right & samples & nested \\
 \midrule
-XA & 39 & \{\} & 1 & 122444 & 122444 & 0 & 0 \\
-XBB & 6452 & \{BA.2:1\} & 14 & 1396207 & 1396207 & 0 & 0 \\
-XBB.1 & 2 & \{\} & 1 & 1429711 & 1429711 & 0 & 0 \\
-XBD & 30 & \{\} & 0 & 1378208 & 1378208 & 0 & 0 \\
-XBF & 185 & \{\} & 3 & 1420385 & 1420385 & 0 & 0 \\
-XBG & 25 & \{\} & 2 & 1291970 & 1291970 & 0 & 0 \\
-XBR & 1 & \{\} & 2 & 1420166 & 1420166 & 0 & 0 \\
-XC & 5 & \{\} & 1 & 414488 & 414488 & 0 & 0 \\
-XF & 16 & \{\} & 1 & 946761 & 946761 & 0 & 0 \\
-XG & 3 & \{\} & 3 & 1083412 & 1083412 & 0 & 0 \\
-XL & 64 & \{\} & 1 & 1034619 & 1034619 & 0 & 0 \\
-XM & 26 & \{BA.2:16, XAL:3\} & 0 & 1003220 & 1003220 & 0 & 0 \\
-XQ & 55 & \{BA.2:37, $\star$\} & 3 & 1058654 & 1058654 & 0 & 0 \\
-XS & 17 & \{\} & 1 & 1000242 & 1000242 & 0 & 0 \\
-XW & 32 & \{\} & 1 & 1159411 & 1159411 & 0 & 0 \\
-XY & 23 & \{\} & 2 & 1187989 & 1187989 & 0 & 0 \\
+XC & 37 & 26768 & 27390 & 5 & \{\} \\
+XBR & 22 & 22034 & 22190 & 1 & \{\} \\
+XA & 16 & 20411 & 21765 & 39 & \{\} \\
+XS & 11 & 9054 & 10449 & 17 & \{\} \\
+XL & 8 & 5925 & 8393 & 64 & \{\} \\
+XBG & 8 & 22305 & 22917 & 25 & \{\} \\
+XBD & 7 & 23020 & 24620 & 30 & \{\} \\
+XQ & 7 & 4322 & 5386 & 55 & \{BA.2:37, $\star$\} \\
+XBB & 6 & 22332 & 22577 & 6452 & \{BA.2:1\} \\
+XM & 6 & 17411 & 21595 & 26 & \{BA.2:16, XAL:3\} \\
+XBF & 5 & 5184 & 9866 & 185 & \{\} \\
+XF & 5 & 5387 & 6402 & 16 & \{\} \\
+XY & 5 & 11538 & 12880 & 23 & \{\} \\
+XG & 4 & 5925 & 6513 & 3 & \{\} \\
+XW & 4 & 2833 & 4321 & 32 & \{\} \\
 \midrule
-XAA & 17 & \{\} & 3 & 1183815 & 1058654 & 33 & 6 \\
-XAE & 9 & \{\} & 7 & 1118099 & 964555 & 47 & 2 \\
-XAF & 1 & \{\} & 8 & 1314603 & 1177107 & 140 & 5 \\
-XAG & 6 & \{\} & 4 & 1248341 & 1058654 & 55 & 6 \\
-XAL & 3 & \{\} & 2 & 1264107 & 1003220 & 66 & 3 \\
-XAM & 21 & \{\} & 3 & 1240312 & 1058654 & 43 & 6 \\
-XAN & 7 & \{\} & 5 & 1301399 & 1189192 & 114 & 4 \\
-XAP & 20 & \{\} & 1 & 1216836 & 964555 & 67 & 4 \\
-XAV & 13 & \{\} & 3 & 1269391 & 1189192 & 106 & 7 \\
-XAZ & 133 & \{\} & 3 & 1285706 & 1189192 & 62 & 2 \\
-XBE & 65 & \{\} & 3 & 1363926 & 1189192 & 136 & 6 \\
-XBH & 2 & \{\} & 2 & 1395926 & 1379419 & 27 & 1 \\
-XBM & 10 & \{\} & 1 & 1358392 & 1348822 & 23 & 1 \\
-XE & 1116 & \{\} & 2 & 965352 & 965353 & 19 & 1 \\
-XH & 2 & \{\} & 7 & 1098084 & 965353 & 55 & 1 \\
-XJ & 68 & \{BA.2:3\} & 2 & 966904 & 966905 & 5 & 1 \\
-XR & 17 & \{BA.2:1\} & 4 & 1148222 & 1058654 & 11 & 3 \\
-XU & 1 & \{\} & 4 & 1243765 & 1058654 & 103 & 6 \\
-XZ & 48 & \{\} & 1 & 1163537 & 964555 & 48 & 4 \\
+\multicolumn{5}{l}{Type II RE event}\\
+alias & averted & left & right & total & by pango \\
 \midrule
-XAJ & 18 & \{\} & 11 & 1276376 & -1 & inf & inf \\
-XAS & 77 & \{\} & 1 & 1265115 & -1 & inf & inf \\
-XAU & 8 & \{\} & 1 & 1231548 & -1 & inf & inf \\
-XB & 192 & \{\} & 2 & 223239 & -1 & inf & inf \\
-XBK & 7 & \{\} & 1 & 1425824 & -1 & inf & inf \\
-XBQ & 14 & \{\} & 1 & 1422955 & -1 & inf & inf \\
-XN & 120 & \{BA.2:1\} & 1 & 1061700 & -1 & inf & inf \\
-XP & 45 & \{\} & 1 & 1092789 & -1 & inf & inf \\
+XAE+ & 6 & 24504 & 26060 & 253 & {BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
+XE/XH & 6 & 10448 & 11283 & 1156 & {XE:1116, BA.2:37, XH:2, XAF:1} \\
+XBH & 6 & 15452 & 22001 & 6 & {BA.2:4, XBH:2} \\
+XBM & 6 & 22305 & 22917 & 12 & {XBM:10, BF.3:2} \\
+XAF & 5 & 10199 & 10447 & 36 & {BA.2:35, XAF:1} \\
+XJ & 5 & 13196 & 17410 & 85 & {XJ:68, BA.2:17} \\
 \bottomrule
+
 \end{tabular}
+
 \end{table}

### 2025-09-03 latex/main.tex
WIP

@@
 \label{tab:pango_x_lineages}}
-\begin{tabular}{lrrrrl}
+\begin{tabular}{lrrrrll}
 \toprule
@@
 XJ & 5 & 13196 & 17410 & 85 & {XJ:68, BA.2:17} \\
+\midrule
+\multicolumn{5}{l}{Nested monophyletic Pango X lineage}\\
+pango & alias & path len & time & mutations & samples & nested \\
+\midrule
+XAP & XAE+ & 4 & 67 & 1 & 20 & \{\} \\
+XZ & XAE+ & 4 & 48 & 1 & 48 & \{\} \\
+XAE & XAE+ & 2 & 47 & 7 & 9 & \{\} \\
+XAF & XAF & 5 & 140 & 8 & 1 & \{\} \\
+XAL & XM & 3 & 66 & 2 & 3 & \{\} \\
+XR & XQ & 3 & 11 & 4 & 17 & \{BA.2:1\} \\
+XU & XQ & 6 & 103 & 4 & 1 & \{\} \\
+XAM & XQ & 6 & 43 & 3 & 21 & \{\} \\
+XAA & XQ & 6 & 33 & 3 & 17 & \{\} \\
+XAG & XQ & 6 & 55 & 4 & 6 & \{\} \\
+
 \bottomrule
-
 \end{tabular}

### 2025-09-04 latex/main.tex

--- 2025-09-03 latex/main.tex
@@
 \label{tab:pango_x_lineages}}
-\begin{tabular}{lrrrrl}
+\begin{footnotesize}
+\begin{tabular}{lrrrllrl}
 \toprule
 \multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
-pango & averted & left & right & samples & nested \\
+pango & avt & left & right & left p & right p & samp & nested \\
 \midrule
-XC & 37 & 26768 & 27390 & 5 & \{\} \\
-XBR & 22 & 22034 & 22190 & 1 & \{\} \\
-XA & 16 & 20411 & 21765 & 39 & \{\} \\
-XS & 11 & 9054 & 10449 & 17 & \{\} \\
-XL & 8 & 5925 & 8393 & 64 & \{\} \\
-XBG & 8 & 22305 & 22917 & 25 & \{\} \\
-XBD & 7 & 23020 & 24620 & 30 & \{\} \\
-XQ & 7 & 4322 & 5386 & 55 & \{BA.2:37, $\star$\} \\
-XBB & 6 & 22332 & 22577 & 6452 & \{BA.2:1\} \\
-XM & 6 & 17411 & 21595 & 26 & \{BA.2:16, XAL:3\} \\
-XBF & 5 & 5184 & 9866 & 185 & \{\} \\
-XF & 5 & 5387 & 6402 & 16 & \{\} \\
-XY & 5 & 11538 & 12880 & 23 & \{\} \\
-XG & 4 & 5925 & 6513 & 3 & \{\} \\
-XW & 4 & 2833 & 4321 & 32 & \{\} \\
+XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
+XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
+XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
+XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
+XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
+XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
+XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
+XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
+XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
+XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
+XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
+XAL:3} \\
+XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
+XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
+XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
+XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
+XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
 \midrule
 \multicolumn{5}{l}{Type II RE event}\\
-alias & averted & left & right & total & by pango \\
+alias & avt & left & right & left p & right p & desc & by pango \\
 \midrule
-XAE+ & 6 & 24504 & 26060 & 253 & {BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
-XE/XH & 6 & 10448 & 11283 & 1156 & {XE:1116, BA.2:37, XH:2, XAF:1} \\
-XBH & 6 & 15452 & 22001 & 6 & {BA.2:4, XBH:2} \\
-XBM & 6 & 22305 & 22917 & 12 & {XBM:10, BF.3:2} \\
-XAF & 5 & 10199 & 10447 & 36 & {BA.2:35, XAF:1} \\
-XJ & 5 & 13196 & 17410 & 85 & {XJ:68, BA.2:17} \\
+XAE+ & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 &
+{BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
+XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
+{XE:1116, BA.2:37, XH:2, XAF:1} \\
+XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {BA.2:4, XBH:2} \\
+XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
+BF.3:2} \\ XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
+XAF:1} \\
+XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
+BA.2:17} \\
 \bottomrule
-
+\multicolumn{5}{l}{Events closely derived from RE}\\
+pango & alias & path len & time & mutations & samples \\
+\midrule
+XR & XQ & 3 & 11 & 4 & 17 \\
+XU & XQ & 6 & 103 & 4 & 1 \\
+XAM & XQ & 6 & 43 & 3 & 21 \\
+XAA & XQ & 6 & 33 & 3 & 17 \\
+XAG & XQ & 6 & 55 & 4 & 6 \\
+XAP & XAE+ & 4 & 67 & 1 & 20 \\
+XZ & XAE+ & 4 & 48 & 1 & 48 \\
+XAE & XAE+ & 2 & 47 & 7 & 9 \\
+XAF & XAF & 5 & 140 & 8 & 1 \\
+XAL & XM & 3 & 66 & 2 & 3 \\
+\midrule
+\multicolumn{5}{l}{Non recombinant}\\
+pango & muts & parent & samples & nested \\
+\midrule
+XP & 1 & \textbf{BA.1.1} & 45 & {} \\
+XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
+XBQ & 1 & CJ.1 & 14 & {} \\
+XBK & 1 & CJ.1 & 7 & {} \\
+XBE & 3 & BA.5.2 & 65 & {} \\
+XB & 2 & B.1 & 192 & {} \\
+XAZ & 3 & \textbf{BA.5} & 133 & {} \\
+XAV & 3 & BA.5.1.24 & 13 & {} \\
+XAU & 1 & BA.2 & 8 & {} \\
+XAS & 1 & BA.4 & 77 & {} \\
+XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
+XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
+\bottomrule
 \end{tabular}
+\end{footnotesize}
+
+% pango & averted & left & right & samples & nested \\
+
+% \midrule
+% XC & 37 & 26768 & 27390 & 5 & \{\} \\
+% XBR & 22 & 22034 & 22190 & 1 & \{\} \\
+% XA & 16 & 20411 & 21765 & 39 & \{\} \\
+% XS & 11 & 9054 & 10449 & 17 & \{\} \\
+% XL & 8 & 5925 & 8393 & 64 & \{\} \\
+% XBG & 8 & 22305 & 22917 & 25 & \{\} \\
+% XBD & 7 & 23020 & 24620 & 30 & \{\} \\
+% XQ & 7 & 4322 & 5386 & 55 & \{BA.2:37, $\star$\} \\
+% XBB & 6 & 22332 & 22577 & 6452 & \{BA.2:1\} \\
+% XM & 6 & 17411 & 21595 & 26 & \{BA.2:16, XAL:3\} \\
+% XBF & 5 & 5184 & 9866 & 185 & \{\} \\
+% XF & 5 & 5387 & 6402 & 16 & \{\} \\
+% XY & 5 & 11538 & 12880 & 23 & \{\} \\
+% XG & 4 & 5925 & 6513 & 3 & \{\} \\
+% XW & 4 & 2833 & 4321 & 32 & \{\} \\
+% \midrule
+% \multicolumn{5}{l}{Type II RE event}\\
+% alias & averted & left & right & total & by pango \\
+% \midrule
+% XAE+ & 6 & 24504 & 26060 & 253 & {BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
+% XE/XH & 6 & 10448 & 11283 & 1156 & {XE:1116, BA.2:37, XH:2, XAF:1} \\
+% XBH & 6 & 15452 & 22001 & 6 & {BA.2:4, XBH:2} \\
+% XBM & 6 & 22305 & 22917 & 12 & {XBM:10, BF.3:2} \\
+% XAF & 5 & 10199 & 10447 & 36 & {BA.2:35, XAF:1} \\
+% XJ & 5 & 13196 & 17410 & 85 & {XJ:68, BA.2:17} \\
+% \midrule
+% \multicolumn{5}{l}{Nested monophyletic Pango X lineage}\\
+% pango & alias & path len & time & mutations & samples & nested \\
+% \midrule
+% XAP & XAE+ & 4 & 67 & 1 & 20 & \{\} \\
+% XZ & XAE+ & 4 & 48 & 1 & 48 & \{\} \\
+% XAE & XAE+ & 2 & 47 & 7 & 9 & \{\} \\
+% XAF & XAF & 5 & 140 & 8 & 1 & \{\} \\
+% XAL & XM & 3 & 66 & 2 & 3 & \{\} \\
+% XR & XQ & 3 & 11 & 4 & 17 & \{BA.2:1\} \\
+% XU & XQ & 6 & 103 & 4 & 1 & \{\} \\
+% XAM & XQ & 6 & 43 & 3 & 21 & \{\} \\
+% XAA & XQ & 6 & 33 & 3 & 17 & \{\} \\
+% XAG & XQ & 6 & 55 & 4 & 6 & \{\} \\
+
+% \bottomrule
+% \end{tabular}
 

### 2025-09-04 notebooks/nb_utils.py
Add cousins of XP
And add the pdf

@@ -214,6 +214,7 @@ def mutation_p_values(ts, min_time=1, progress=True):
 
 
 class D3ARG_viz:
+    colours = ['#332288', '#88CCEE', '#44AA99', '#117733', '#999933', '#DDCC77']  # from https://personal.sron.nl/~pault/
     def __init__(self, ts, df, lineage_consensus_muts=None, pangolin_field="pango", progress=True):
         self.ts = ts
         self.df = df

### 2025-09-05 src/save_copying_patterns.py
Add some extra copying patterns for the supplementary

@@ -53,15 +53,32 @@ def pangoX_RE_node_labels(exclude_dups=True):
     else:
         return dict(px)
 
-
-for node_id, label in tqdm(pangoX_RE_node_labels().items()):
-    img_bytes = imgkit.from_string(
-    sc2ts.info.CopyingTable(ts, node_id).html(hide_extra_rows=True, hide_labels=True, show_bases=None),
-        False,  # return the bytes, rather than saving to file
-        options={"width": 2000, "format": "png", "quiet": "", "transparent": "",})
+def save_copying_pattern_image(node_id, label, ts, save_dir, zoom=None,**kwargs):
+    # NB - do not zoom if no ATCG letters are printed, as boxes will be of varying size
+    html_str = sc2ts.info.CopyingTable(ts, node_id).html(**kwargs)
+    html_str = html_str.replace('transform:', '-webkit-transform:')
+    html_str = html_str.replace('writing-mode:', '-webkit-writing-mode:')
+    options = {"format": "png", "quiet": "", "transparent": ""}  # use transparent so we can crop
+    if zoom is not None:
+        options['zoom'] = zoom 
+    img_bytes = imgkit.from_string(html_str, output_path=False, options=options)
     img = Image.open(BytesIO(img_bytes))
     img = img.convert('RGBA')
     # Crop whitespace
     bbox = img.getbbox()
     img = img.crop(bbox)
-    img.save(png_dir / f"{label}.png", "PNG", optimize=True, compress_level=9)
+    img.save(save_dir / f"{label}.png", "PNG", optimize=True, compress_level=9)
+
+for u, label in tqdm(pangoX_RE_node_labels().items()):
+    save_copying_pattern_image(
+       u, label, ts, png_dir, hide_extra_rows=True, hide_labels=True, show_bases=None, font_family='Verdana')
+
+for u, lab in tqdm([
+    (1372032, "causedby-ERR10306608"),
+    (456023, "causedby-ERR6760245"),
+    (1137450, "causedby-SRR21087725"),
+    (427863, "causedby-SRR15864165"),
+]):
+    save_copying_pattern_image(
+        u, lab, ts, png_dir, hide_extra_rows=False, hide_labels=True, show_bases=True, zoom=4, font_family='Verdana'
+    )

### 2025-09-05 latex/main.tex
Move the supplement into the main document

@@
 
-\subsection*{Additional resources}
+% \subsection*{Additional resources}
 
@@
 
-Cell.com homepage: https://www.cell.com
-\newline Templates for Cell Press authors: https://www.cell.com/article-templates
+% Cell.com homepage: https://www.cell.com
+% \newline Templates for Cell Press authors: https://www.cell.com/article-templates
 
@@
 
+\newpage
+\setcounter{secnumdepth}{2} % Print out appendix section numbers
+
+\section*{Supplementary Material}
+\appendix
+
+\setcounter{table}{0}
+\setcounter{figure}{0}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thesection}{S\arabic{section}}
+
+\section{Major deletions in the ARG}
+% Referenced by the main Results.
+Basically, we want to say we did a lot of work to integrate signals of major
+indels into the ARG, making it an even more powerful resource overall. Also,
+indels are encoded by the Mutation-Annotated Tree format used by UShER
+phylogenies (though PanMAT does this).
+
+% See https://github.com/jeromekelleher/sc2ts-paper/issues/472
+At minimum, we want to show compare the frequencies of the top N indels in the
+ARG and those reported in Li et al. (2023). This can be shown in a table.
+
+\section{Identification of robust recombination events}
+Before characterizing recombination events in detail, we sought to remove
+probable artifacts. Systematic errors in SARS-CoV-2 sequences associated with
+sequencing protocols have been documented (Turakhia et al., 2022), which can
+confound recombination detection (De Maio et al., 2024). By inspecting copying
+patterns at sites that differ between recombinant parents, we find a convincing
+signal when treating adjacent and near-adjacent sites as a single locus.
+Artifactual recombination events can be characterized as having fewer than four
+such loci supporting the left or right parent on either side of a suggested
+breakpoint (STAR Methods). Of the 929 recombination events in the ARG, this
+measure identifies 543 as artifactual, leaving 386 “robust” events (Figure
+S1A). The copying patterns depicted in Figure S1B illustrate the genomic
+evidence for three recombination events identified as artifactual, compared to
+a marginally robust one.
+
+Evidence that our simple filter effectively flags artifacts can be seen in the
+clear association between the AmpliSeq V1 kit and recombination events that we
+identify as artifactual. This primer scheme was used to sequence XX of the
+samples associated with an artifactual recombination event, in contrast to 15
+of the “robust” events (Figure 2A), strongly implicating it as a common source
+of problematic data. Through further inspection, we observed that many
+artifactual recombination events contain stretches of missing data
+characteristic of amplicon dropouts (REF).
+
+\section{Analysis of Pango lineage ``origination'' events}
+
+To analyze how well the ARG reflects the phylogenetic structure implicit in the
+Pango lineage naming system, we considered the Pango assignments generated by
+Pangolin on the alignments for each node in the ARG (STAR Methods). If the ARG
+perfectly reflected the Pango lineage structure, each lineage would form a
+clade descending from a single originating node with that label. We identified
+putative origination nodes among those labeled with a given lineage as the node
+with the maximum number of descendant samples and the earliest time (if there
+is a tie).
+
+Of the 2058 distinct Pango lineages in the ARG, 1469 of these (comprising
+743922 samples) match perfectly, with unique origination events in the ARG
+where all samples assigned a given lineage descend from the first node assigned
+to that lineage. A further 310 lineages (398391 samples) match perfectly when
+we count the descendants of the parent of the first node (accounting for
+polytomies in which multiple originating nodes for a given lineage are
+siblings). We then have 255 lineages (996490 samples) where the difference in
+the number of descendants of the first node's parent is < 100. The remaining 25
+lineages (344079) are dominated by a few large lineages such as BA.1.1 (155595
+samples) and AY.4.2 (54607 samples) which have multiple non-sibling origins
+within the ARG.
+
+\section{Additional analyses of the Pango Xs}
+
+\subsection{XS}
+https://github.com/jeromekelleher/sc2ts-paper/issues/287
+In the ARG, the XS samples appear to have originated from two sequential recombination events.
+However, the second event is likely an artifact,
+as the XS sample associated with this event failed the minimum net number of supporting loci.
+In addition, this sample was sequenced using the AmpliSeq V1 kit,
+which was also used to sequence eight other samples descending from the same event.
+We excluded this likely artifactual event from further analysis and
+consider XS to have emerged from a single recombinant origin.
+
+\subsection{XM}
+https://github.com/jeromekelleher/sc2ts-paper/issues/284
+Possibly related to deletions.
+
+\subsection*{Multiple breakpoints}
+Seven of the Pango Xs present in the Viridian v04 dataset have more than one breakpoint,
+with the supporting evidence documented in the associated Pango designation issues.
+Five of these seven Pango Xs were detected as recombinants by sc2ts.
+Three of them (XAY, XBC, XBT) had more than one breakpoint,
+but were not added to the ARG due to high HMM cost and
+a lack of sample support to be incorporated as retrospective sample groups.
+However, the other two (XAC and XAK) had only one breakpoint.
+
+\subsubsection*{XAY}
+https://github.com/jeromekelleher/sc2ts-paper/issues/490
+
+\subsubsection*{XBC}
+https://github.com/jeromekelleher/sc2ts-paper/issues/486
+Sc2ts detected XBC as a three-breakpoint recombinant,
+correctly identifying the parental Pango lineages.
+The inferred breakpoint intervals mostly overlap with those identified in its designation issue.
+XBC was not added to the ARG due to high HMM costs and inadequate sample support
+(within a seven-day retrospective time window).
+
+\subsubsection*{XBT}
+https://github.com/jeromekelleher/sc2ts-paper/issues/487
+Sc2ts detected XBT as a two-breakpoint recombinant,
+correctly identifying the parental Pango lineages.
+The inferred breakpoint intervals are identical to those documented in its designation issue.
+XBT was not added to the ARG due to high HMM cost and low sample support
+(only one QC-passing sample in the Viridian v04 dataset).
+
+\subsubsection*{XAC}
+https://github.com/jeromekelleher/sc2ts-paper/issues/357
+Evidence of the second breakpoint reported in its designation issue relies on
+the presence of a large deletion inherited from BA.2*, which is currently treated as missing data by sc2ts.
+The XAC samples share a recombination node with the samples of four other Pango Xs
+which do not have this deletion (Figure 2C).
+
+\subsubsection*{XAK}
+https://github.com/jeromekelleher/sc2ts-paper/issues/484
+Our results agree with its designation issue in terms of the parental Pango lineages and breakpoint intervals.
+XAK was not added to the ARG due to HMM costs (9 and 10) and insufficient sample support (only two samples).
+Interestingly, our results suggest that XAK might be a product of two sequential recombination events.
+
+\subsubsection*{Non-recombinants}
+\subsubsection*{XN and XAU}
+https://github.com/jeromekelleher/sc2ts-paper/issues/358
+https://github.com/jeromekelleher/sc2ts-paper/issues/348
+Both of these putative recombinants have been defined on the basis of a RHS from BA.2 and a LHS from BA.1.
+In the case of XN, the supposed BA.1 mutations are [C241T, A2832G, C3037T].
+In the case of XAU, this set also includes C2470T, so [C241T, C2470T, A2832G, C3037T].
+However, both BA.1 and BA.2 are now defined as also having C241T and C3037T.
+In addition, a very large number of BA.2 samples also have A2832G, and three BA.2-designated samples also have C2470T.
+This means that there is essentially no longer any evidence for a BA.1 parent for XN, and
+only one uncertain mutation that suggests a BA.1 parent for XAU.
+
+\subsubsection*{XAS}
+https://github.com/jeromekelleher/sc2ts-paper/issues/340
+This was proposed as a recombination between BA.5 (LHS) and BA.2 (RHS).
+However, we find that this is only 2 mutations different from e.g. ERR9929846 which is a BA.4 sample.
+In fact, XAS only differs from the BA.4 root by a single mutation, C27945T,
+which was identified in the original designation issue as an important ORF truncation mutation also present in a few BA.2.65 samples.
+It is possible that C27945T was acquired by recombination, this is clearly below our threshold for considering something to be a recombinant.
+
+\subsubsection*{XB}
+https://github.com/jeromekelleher/sc2ts-paper/issues/332
+XB was proposed to have originated from recombination between B.1.634 (left parent) and B.1.631 (right parent).
+None of the samples in Viridian v04 belonging to these Pango lineages were integrated into the ARG due to high HMM costs.
+
+\subsubsection*{XP}
+https://github.com/jeromekelleher/sc2ts-paper/issues/345
+
+\subsubsection*{XAJ}
+https://github.com/jeromekelleher/sc2ts-paper/issues/352
+
+\subsubsection*{XAZ}
+https://github.com/jeromekelleher/sc2ts-paper/issues/356
+Proposed as a recombinant of BA.2.5 on the left and BA.5 on the right, with 3 supporting sites (C2232T, C3317T, T3358C).
+In sc2ts this is very close to being a recombinant of BA.2.5 (ERR9615610) on the left,
+as not only are the 3 mutations present in this sample,
+but there is another site (C1912T apparently missed in the designation) which supports recombination.
+XAZ has C44T which is lacking in all BA.2.5 samples,
+giving only three net supporting loci for a putative recombinant.
+However, position 44 is often clipped, so this is not strong evidence against a recombinant, and
+this is a very finely balanced case.
+
+\subsubsection*{XAN and XAV}
+https://github.com/jeromekelleher/sc2ts-paper/issues/353
+Below threshold.
+
+\subsubsection*{XBE}
+This was proposed as a recombinant between BA.5,5 and BE.4.
+We find two net supporting loci for a recombination breakpoint to be around 22599.
+
+\subsection*{Analysis of the Jackson et al. recombinants}
+Jackson et al.\cite{Jackson2021} conducted detailed analyses of some early recombinants
+involving B.1.1.7 (Alpha), including the first designated Pango X lineage, XA.
+The authors first performed a targeted search for candidate recombinants by
+scanning sampled sequences for sequence motifs associated with Alpha, followed
+by an analysis to reduce the list of plausible recombinants displaying signs of
+onward transmission. Here we compared our sc2ts detection results of these
+recombinants with the detection results reported by Jackson et al.
+
+% Data derived from the notebooks/jackson_recombinants.ipynb notebook
+% See https://github.com/jeromekelleher/sc2ts-paper/issues/293
+Jackson et al. described four groups of recombinants (named A to D) and
+four singleton recombinants.
+Of the 16 samples analyzed by Jackson et al, 12 were included in the ARG;
+the remaining 4 were excluded because the sequences did not pass the 
+QC filtering criteria, or due to a high HMM cost. Groups A-D are summarized 
+in Table~\ref{tab:jackson} and show excellent concordance in terms 
+of the parental Pango lineages and breakpoint intervals. All four groups
+are directly associated with strongly supported recombination events 
+in the ARG in terms of the number of averted mutations 
+(A: 16, B: 6, C: 30, D: 18).
+The singleton CAMC-CBA018 has 
+the parents B.1.177+B.1.1.7 and the breakpoint interval 16,177--20,133 in the ARG, 
+and is strongly supported by 14 averted mutations.
+The Jackson et al.\ results are almost identical with the same parent lineages, 
+and breakpoint interval of interval 11,395–21,993.
+The singleton MILK-103C712 has 
+the parents B.1.177.16+B.1.1.7 and a breakpoint interval of 24915--27972 
+in the ARG, again strongly supported by 11 averted mutations.
+The Jackson et al.\ results for this lineage has the same parents,
+with the breakpoint interval 26,800–-27,974.
+The singleton QEUH-1067DEF has the parents B.1.1.7+B.1.177.9 
+and a breakpoint interval of 7,729--10,870 in the ARG, 
+supported by 11 averted mutations.
+The Jackson et al.\ results for this lineage has the same parents,
+with the breakpoint interval 6,953--10,872.
+
+\begin{table} \centering
+\caption{\label{tab:jackson}Comparison of recombination breakpoint
+intervals and parent lineages for Groups A-D
+reported by Jackson et al.\ with the corresponding
+recombination events in the sc2ts ARG.
+The second column gives the number of sequences in the group.
+See the text for details of the groups and the sequences included.
+limited to the samples considered by Jackson et al.
+The breakpoint coordinates in Table 1 of Jackson et al.\ have been altered as follows: 
+we subtract one to the left coordinates and add one to the right coordinates 
+to correspond to the \texttt{tskit} definition of inheritance on either side of a breakpoint,
+and add one to the right coordinate to make the intervals right-exclusive. }
+\begin{tabular}{lllr@{--}lr@{+}l}
+\toprule
+Group        & Sequences & Method & \multicolumn{2}{c}{Interval}
+    & \multicolumn{2}{r}{Parent lineages} \\
+\midrule
+A (XA)       & 2   & Jackson        &  21,254&21,767 & B.1.177&B.1.1.7 \\
+             &     &\texttt{sc2ts} &  20,411&21,765 & B.1.177.18&B.1.1.7 \\
+\cmidrule{3-7}
+B            & 2   & Jackson        &  6,527&6,956 & B.1.36.28&B.1.1.7  \\
+             &     &\texttt{sc2ts} &   6,529&6,954 & B.1.36.28&B.1.1.7  \\
+\cmidrule{3-7}
+C            & 3   &Jackson         &  24,913&28,653 &  B.1.1.7&B.1.221 \\
+             &     & \texttt{sc2ts} &  25,997&27,972 &  B.1.1.7&B.1.221.1 \\
+\cmidrule{3-7}
+D            & 3   & Jackson        &  21,574&23,065 &  B.1.36.17&B.1.1.7 \\
+             &     & \texttt{sc2ts} &  22,445&23,063 &  B.1.36.39&B.1.1.7 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\section*{Supplementary Figures}
+
+\begin{figure}
+    \centering
+    % \includegraphics{sc2ts_methodology}
+    \caption{Classification of ``robust'' and artifactual recombination events.
+(A) Scatterplot of recombination events by the net number of supporting loci for the left parent (x-axis) and for the right parent (y-axis).
+The shaded regions highlight artifactual recombination events,
+which have fewer than four net supporting loci on one side or both sides of the suggested breakpoint.
+Colors indicate the primer scheme used for sequencing.
+The pie charts show breakdowns of the recombination events by primer scheme per quadrant.
+Recombination events associated with the origins of Pango Xs are labeled 
+(classified in Figure 2).
+(B) Copying patterns of an illustrative recombination event picked from each quadrant.
+}\label{fig:recombinant_qc}
+\end{figure}
+
+\begin{figure}
+    \centering
+    % \includegraphics{sc2ts_methodology}
+    \caption{Phylogenetic backbones: TODO}\label{fig:phylo_backbones}
+\end{figure}
+
+\begin{figure}
+    \centering
+    %TODO fix this to make the figures smaller.
+    \includegraphics[width=\linewidth]{mutational_spectra}
+    \caption{All-site mutational spectra of major VOCs.
+(A) Mutational spectra of the ARG.
+For each major VOC, the relative proportion of each mutation type was calculated from
+the mutations occurring above the nodes with a Scorpio label associated with the VOC
+(e.g. “Alpha-like” encompasses “B.1.1.7-like” and “B.1.1.7-like+E484K”).
+Note that $\sim$98\% of all the mutations in the ARG occur above the nodes
+which have a Scorpio label associated with Alpha, Delta, Omicron BA.1, BA.2, BA.4, or BA.5
+(excluding the nodes with unassigned scorpio labels).
+The Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin.
+(B) Mutational spectra reproduced from the mutation count data analyzed in Bloom et al. (2023).
+The authors extracted mutation counts from a global UShER phylogeny
+(https://github.com/jbloomlab/SARS2-mut-spectrum/).
+\label{fig:mutational_spectra}
+}
+\end{figure}
+
+\begin{figure} 
+\centering
+\includegraphics[width=0.5\textwidth]{LS_model_schematic}
+\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
+model, in which a focal sequence (bottom) is described as an
+imperfect mosaic of the sequences in a reference panel.
+Black crosses along the focal sequence show sequencing
+errors or mutations.
+In the standard formulation, at site $\ell$, the recombination probability is $r_\ell$,
+the mutation probability is $\mu_\ell$ and $n$
+denotes the size of the reference panel.
+The Viterbi algorithm can be used to find a
+``copying path'' through the reference panel for a given focal sequence that
+maximises the likelihood under these parameters. Unseen states in the reference panel are shown as coloured lines enclosed by
+the grey box. The black arrow describes the true path through the data which leads to the emitted
+focal sequence below. Examples of transition and
+emission probabilities along this trajectory are shown by the red and blue
+arrows, respectively.
+}
+\end{figure}
+
 \end{document}

### 2025-09-05 latex/main.tex
Add samples per day supplementary figure

@@
 Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available
-(\ref{fig:methodology}; for details, see STAR Methods).
+(Figure~\ref{fig:methodology}; for details, see STAR Methods).
 First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
-(allowing recombination) using the Li and Stephens (LS) model \cite{Li2003},
+(allowing recombination) using the Li and Stephens (LS) model\cite{Li2003},
 which is a Hidden Markov Model (HMM) widely used in human genomics
@@
 and a switch has a penalty cost of $k$ (here, set to 4),
-which determines how much mutation is favored over recombination when accounting for a sample (\ref{fig:methodology}).
+which determines how much mutation is favored over recombination when accounting for a 
+sample (Figure~\ref{fig:methodology}).
 If the HMM cost of a sample exceeds a threshold (here, set to 7),
@@
 This approach resulted in preliminary ARGs that had noticeably fewer time travelers,
-but we found that it also excluded the major “saltation” events that characterize many VOCs.
+but we found that it also excluded the major ``saltation'' events that characterize many VOCs.
 These saltation lineages are characterized by a high number of accumulated mutations,
@@
 sc2ts adds the group back to the ARG.
-This “retrospective sample grouping” successfully captured the emergence of most saltation lineages (Figure SX).
+This ``retrospective sample grouping'' successfully captured the emergence of
+most saltation lineages (Figure~\ref{fig:samples_per_day}).
 The emergence of the Delta, BA.1, and BA.2 lineages required some manual intervention,
@@
 
-\section{Major deletions in the ARG}
+\section{Major deletions in the ARG}\label{sec_supplement_dels}
 % Referenced by the main Results.
@@
 
-\section{Identification of robust recombination events}
+\section{Identification of robust recombination
+events}\label{sec_supplement_robust_recombs}
 Before characterizing recombination events in detail, we sought to remove
@@
 \section*{Supplementary Figures}
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{samples-per-day}
+    \caption{TODO caption}\label{fig:samples_per_day}
+\end{figure}
 

### 2025-09-05 latex/main.tex
Add inference resources figure
Move robust recombs to methods

@@
 Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
-Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM (~5 CPU years; Figure SX).
+Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM 
+(~5 CPU years; Figure~\ref{fig:inference_resources}).
 The ARG contains 2,689,054 nodes representing the genealogical relationships
@@
 of which 356 have a minimum of four loci supporting the non-focal parent on both sides of the breakpoint
-(we refer to these as “robust” recombination events, Figure SX; see Document S1),
+(we refer to these as ``robust'' recombination events,
+Figure~\ref{fig:recombinant_qc}; see STAR Methods)
 reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.
@@
 
+\subsubsection*{Identification of robust recombination events}
+[TODO update refs!]
+
+Before characterizing recombination events in detail, we sought to remove
+probable artifacts. Systematic errors in SARS-CoV-2 sequences associated with
+sequencing protocols have been documented (Turakhia et al., 2022), which can
+confound recombination detection (De Maio et al., 2024). By inspecting copying
+patterns at sites that differ between recombinant parents, we find a convincing
+signal when treating adjacent and near-adjacent sites as a single locus.
+Artifactual recombination events can be characterized as having fewer than four
+such loci supporting the left or right parent on either side of a suggested
+breakpoint (STAR Methods). Of the 929 recombination events in the ARG, this
+measure identifies 543 as artifactual, leaving 386 ``robust'' events (Figure
+S1A). The copying patterns depicted in Figure S1B illustrate the genomic
+evidence for three recombination events identified as artifactual, compared to
+a marginally robust one.
+
+Evidence that our simple filter effectively flags artifacts can be seen in the
+clear association between the AmpliSeq V1 kit and recombination events that we
+identify as artifactual. This primer scheme was used to sequence XX of the
+samples associated with an artifactual recombination event, in contrast to 15
+of the ``robust'' events (Figure 2A), strongly implicating it as a common source
+of problematic data. Through further inspection, we observed that many
+artifactual recombination events contain stretches of missing data
+characteristic of amplicon dropouts (REF).
+
 \subsubsection*{Comparing phylogenetic backbones}
@@
 
-\section{Major deletions in the ARG}\label{sec_supplement_dels}
+\section{Major deletions in the ARG}\label{sec:supplement_dels}
 % Referenced by the main Results.
@@
 ARG and those reported in Li et al. (2023). This can be shown in a table.
-
-\section{Identification of robust recombination
-events}\label{sec_supplement_robust_recombs}
-Before characterizing recombination events in detail, we sought to remove
-probable artifacts. Systematic errors in SARS-CoV-2 sequences associated with
-sequencing protocols have been documented (Turakhia et al., 2022), which can
-confound recombination detection (De Maio et al., 2024). By inspecting copying
-patterns at sites that differ between recombinant parents, we find a convincing
-signal when treating adjacent and near-adjacent sites as a single locus.
-Artifactual recombination events can be characterized as having fewer than four
-such loci supporting the left or right parent on either side of a suggested
-breakpoint (STAR Methods). Of the 929 recombination events in the ARG, this
-measure identifies 543 as artifactual, leaving 386 “robust” events (Figure
-S1A). The copying patterns depicted in Figure S1B illustrate the genomic
-evidence for three recombination events identified as artifactual, compared to
-a marginally robust one.
-
-Evidence that our simple filter effectively flags artifacts can be seen in the
-clear association between the AmpliSeq V1 kit and recombination events that we
-identify as artifactual. This primer scheme was used to sequence XX of the
-samples associated with an artifactual recombination event, in contrast to 15
-of the “robust” events (Figure 2A), strongly implicating it as a common source
-of problematic data. Through further inspection, we observed that many
-artifactual recombination events contain stretches of missing data
-characteristic of amplicon dropouts (REF).
 
@@
 \section*{Supplementary Figures}
+
 \begin{figure}
@@
     \centering
+    \includegraphics[width=\linewidth]{inference-resources}
+    \caption{TODO caption}\label{fig:inference_resources}
+\end{figure}
+
+\begin{figure}
+    \centering
     % \includegraphics{sc2ts_methodology}
     \caption{Classification of ``robust'' and artifactual recombination events.
-(A) Scatterplot of recombination events by the net number of supporting loci for the left parent (x-axis) and for the right parent (y-axis).
-The shaded regions highlight artifactual recombination events,
-which have fewer than four net supporting loci on one side or both sides of the suggested breakpoint.
-Colors indicate the primer scheme used for sequencing.
-The pie charts show breakdowns of the recombination events by primer scheme per quadrant.
+(A) Scatterplot of recombination events by the net number of supporting loci
+for the left parent (x-axis) and for the right parent (y-axis). The shaded
+regions highlight artifactual recombination events, which have fewer than four
+net supporting loci on one side or both sides of the suggested breakpoint.
+Colors indicate the primer scheme used for sequencing. The pie charts show
+breakdowns of the recombination events by primer scheme per quadrant.
 Recombination events associated with the origins of Pango Xs are labeled
-(classified in Figure 2).
-(B) Copying patterns of an illustrative recombination event picked from each quadrant.
+(classified in Figure 2). (B) Copying patterns of an illustrative recombination
+event picked from each quadrant.
 }\label{fig:recombinant_qc}

### 2025-09-05 latex/main.tex
Incorporate recombinant QC fig

@@
     \centering
-    % \includegraphics{sc2ts_methodology}
+    \includegraphics[width=\linewidth]{robust_vs_artificial_classification}
     \caption{Classification of ``robust'' and artifactual recombination events.
-(A) Scatterplot of recombination events by the net number of supporting loci
+Scatterplot of recombination events by the net number of supporting loci
 for the left parent (x-axis) and for the right parent (y-axis). The shaded
@@
 breakdowns of the recombination events by primer scheme per quadrant.
-Recombination events associated with the origins of Pango Xs are labeled
-(classified in Figure 2). (B) Copying patterns of an illustrative recombination
-event picked from each quadrant.
+Recombination events associated with the origins of Pango Xs are labeled.
 }\label{fig:recombinant_qc}

### 2025-09-05 latex/main.tex
Reset the sectioning a bit

@@
 The ARG represents the full evolutionary history of SARS-CoV-2 and
-is in close agreement with the state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}.
-First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
+is in close agreement with the state-of-the-art UShER phylogeny built using the Viridian 
+dataset\cite{Hunt2024}.
+First, we validated the broad structure of the ARG against the UShER phylogeny\cite{Hunt2024}
 by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
@@
 both within sublineages and at the level of major lineages (including VOCs),
-with good agreement between the phylogenetic bipartitions in the ARG and the UShER phylogeny (Figure S2).
+with good agreement between the phylogenetic bipartitions in the ARG and the UShER 
+phylogeny (Figure~\ref{fig:phylo_backbones}).
 Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset
 by pruning both to their intersection of 2,475,418 samples and 27,431 sites.
+% TODO this is out of date and needs to be rewritten along with supplementary 
+% analysis
 Remarkably, given the differences in inference methods and data processing pipelines,
@@
 against the mutational spectra computed from the UShER phylogeny (STAR Methods),
-finding a close match to previous reports \cite{Bloom2023} (Figure S3).
+finding a close match to previous reports\cite{Bloom2023}
+(Figure~\ref{fig:mutational_spectra}).
 
@@
 the approximation successfully identified the most frequent deletions,
-such as the recurrent 21991-21993 (for more details, see Document S1).
+such as the recurrent 21991-21993 (See Document~\ref{sec:supplement_dels}
+for additional analysis).
 The node representing the origin of Alpha, with 290,407 descendants,
@@
 \setcounter{figure}{0}
+\setcounter{section}{1}
 \renewcommand{\thetable}{S\arabic{table}}
@@
 
-\section{Major deletions in the ARG}\label{sec:supplement_dels}
+\subsection{Major deletions in the ARG}\label{sec:supplement_dels}
 % Referenced by the main Results.
@@
 
-\section{Analysis of Pango lineage ``origination'' events}
+\subsection{Analysis of Pango lineage ``origination'' events}
 
@@
 
-\section{Additional analyses of the Pango Xs}
-
-\subsection{XS}
+\subsection{Additional analyses of the Pango Xs}
+
+\subsection*{XS}
 https://github.com/jeromekelleher/sc2ts-paper/issues/287
@@
 
-\subsection{XM}
+\subsection*{XM}
 https://github.com/jeromekelleher/sc2ts-paper/issues/284
@@
 
-\subsection*{Analysis of the Jackson et al. recombinants}
+\subsection{Analysis of the Jackson et al. recombinants}
 Jackson et al.\cite{Jackson2021} conducted detailed analyses of some early recombinants

### 2025-09-05 latex/main.tex
Fix quotes, and XBB reference

@@
 These phylogenies are used to identify and monitor
-potentially epidemiologically important “Pango” lineages \cite{Rambaut2020,McBroome2024},
+potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
 which might have acquired features to become Variants of Concern (VOC),
@@
 UShER uses a tree to represent evolutionary change via mutation, assuming no recombination \cite{Turakhia2021}.
-However, given extensive evidence for recombination in SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2022},
+However, given extensive evidence for recombination in SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
 the tree-like representation of UShER is no longer tenable.
@@
 (2) an ARG for SARS-CoV-2 with ~2.48 million genome sequences.
-We used the “Viridian” dataset, which is purged of various systematic errors and artifacts \cite{Hunt2024} and
+We used the ``Viridian'' dataset, which is purged of various systematic errors and artifacts \cite{Hunt2024} and
 show that this ARG is highly congruent to an UShER phylogeny built from the same dataset.
@@
 which is estimated to have emerged in November, 2021; REF).
-To preclude adding such “time traveling” samples, we filtered samples by “HMM cost”.
-We define the “HMM cost” of a sample to be
+To preclude adding such ``time traveling'' samples, we filtered samples by ``HMM cost''.
+We define the ``HMM cost'' of a sample to be
 the total penalty cost for switches and mismatches (mutations) in its copying path.
@@
 by running on a subset of the data (STAR Methods).
-Of these sites, 9 are in the set of [1XX] “problematic sites” masked by UShER \cite{Hunt2024}.
+Of these sites, 9 are in the set of [1XX] ``problematic sites'' 
+masked by UShER \cite{Hunt2024}.
 In addition to remapping the data for sites identified as being involved in deletions,
@@
 All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment
-(comparable to the 0.27\% disagreement between pangolin-data versions 1.21 and 1.29 in the source Viridian metadata).
+(comparable to the 0.27\% disagreement between pangolin-data 
+versions 1.21 and 1.29 in the source Viridian metadata).
 A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG,
-increasing to 1779 (86\%) when we allow for multiple sibling origination nodes (Table SX; Document S1).
+increasing to 1779 (86\%) when we allow for multiple sibling origination 
+nodes (Document~\ref{sec:supplement_pango_lineages}).
 
@@
 particularly increased transmissibility, disease severity, and
-ability to escape natural or vaccine-induced immunity (e.g., XBB\cite{Tamura2022}).
+ability to escape natural or vaccine-induced immunity (e.g., XBB\cite{Tamura2023}).
 The process of gathering evidence for and characterizing such recombinants is manual,
@@
 This process has resulted in the designation of recombinant Pango lineages
-(whose aliases begin with an “X”), to which we refer as “Pango X”.
+(whose aliases begin with an ``X''), to which we refer as ``Pango X''.
 Below, we summarize our analysis of the Pango Xs in the ARG
@@
 XA, which is the first designated Pango X that showed signs of community transmission \cite{Jackson2021}, and
-XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2022}.
+XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
 The recombinant origin of XA is well explained in the ARG by two parents and one mutation (\ref{fig:subgraphs}; Table 1),
@@
 
-Furthermore, some of the Pango Xs above originate from shared recombination events in the ARG (Table 1).
-An example of a set of “nested” closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
+Furthermore, some of the Pango Xs above originate from shared recombination events in the 
+ARG (Table 1).
+An example of a set of ``nested'' closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
 which descend from a single recombination event rather than five independent events in the ARG (\ref{fig:subgraphs}).
@@
 \caption{Illustrative subgraphs in the ARG.
-Subgraphs around the XA-associated recombination node (A); the recombination node associated with a group of “nested” Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C).
+Subgraphs around the XA-associated recombination node (A); the recombination node associated with a group of ``nested'' Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C).
 For more detailed visualizations of these subgraphs, see Document S2.}
@@
 We also compared the sensitivity of sc2ts and rebar.
-Of all the “robust” recombination events detected by sc2ts,
+Of all the ``robust'' recombination events detected by sc2ts,
 167 (43\%) were also detected as recombinants by rebar (STAR Methods).
@@
 versus a model disallowing recombination, as measured by the number of resulting mutations (\ref{fig:methodology}).
-The more mutations are “averted” by allowing recombination, the stronger the evidence for recombination.
-We calculated the number of averted mutations for all the “robust” recombination events (STAR Methods),
+The more mutations are ``averted'' by allowing recombination, the stronger the evidence for recombination.
+We calculated the number of averted mutations for all the ``robust'' recombination events (STAR Methods),
 which include all the Pango X-associated recombination events (Table 1).
@@
 Our analysis of the number of averted mutations reveals a spectrum of plausibility
-across all the “robust” recombination events (\ref{fig:averted_mutations}).
+across all the ``robust'' recombination events (\ref{fig:averted_mutations}).
 Examples of highly plausible recombinants are XA and XBB (for an extended analysis, see Document S1).
-The “copying pattern” of XA visually confirms
+The ``copying pattern'' of XA visually confirms
 that a high number of recombination-informative sites support the suggested parents (\ref{fig:averted_mutations}).
@@
     \includegraphics[width=1\linewidth]{fig_recombinant_evidence.pdf}
-    \caption{Evidence for recombination in the ARG. All plots show “robust” sc2ts recombination nodes, colored according
+    \caption{Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes, colored according
 to their classification by rebar (recombinant: purple, non-recombinant: gray).
@@
 We explored the spatial distribution of the breakpoint locations of
-the “robust” recombination events along the genome, and quantified how precisely breakpoints were localized.
+the ``robust'' recombination events along the genome, and quantified how precisely breakpoints were localized.
 For each event, we computed an interval estimate of its breakpoint location,
@@
 the black curve shows the number of recombinants inferred within each week along the ARG,
-focusing on the 356 “robust” events.
+focusing on the 356 ``robust'' events.
 Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high.
@@
 measured as the chance that two randomly drawn viruses have different Scorpio designations
-($H$, referred to as the “expected heterozygosity”).
+($H$, referred to as the ``expected heterozygosity'').
 
@@
 %%%  All original research papers must include a
-%%%  comprehensive and accurate “data and code
-%%%  availability” statement within the “resource
-%%%  availability” component of the paper before it
+%%%  comprehensive and accurate ``data and code
+%%%  availability'' statement within the ``resource
+%%%  availability'' component of the paper before it
 %%%  is accepted for publication. These statements
@@
 % \newline
-% (C and D) If panels are grouped like this, please explicitly describe each panel, e.g., “Images showing SEM (C) and TEM (D).”
+% (C and D) If panels are grouped like this, please explicitly describe each panel, e.g., ``Images showing SEM (C) and TEM (D).''
 % \newline
@@
 
-Sc2ts (pronounced “scoots”, optionally)
+Sc2ts (pronounced ``scoots'', optionally)
 is a method for inferring Ancestral Recombination Graphs (ARGs)
@@
 and use days as the time interval for clarity in what follows.
-The first step is to find likely “copying paths” under the Li and Stephens (LS) model \cite{Li2003}
+The first step is to find likely ``copying paths'' under the Li and Stephens (LS) model \cite{Li2003}
 for each sequence in the daily batch to the current ARG.
@@
 and often many samples from a daily batch will copy from the same ARG node.
-The second step is then to “resolve” these implied polytomies
+The second step is then to ``resolve'' these implied polytomies
 by using a standard neighbor-joining algorithm \cite{Saitou1987}.
@@
 to define an alternative formulation of the coalescent with recombination (Hudson, 1983),
-the term “Ancestral Recombination Graph” has come to be used
+the term ``Ancestral Recombination Graph'' has come to be used
 more generally to describe not just realizations of the model,
@@
 
-The “succinct tree sequence” (STS) is an ARG data structure
+The ``succinct tree sequence'' (STS) is an ARG data structure
 that is both general (in terms of the types of ancestry that can be described)
@@
 The process of switching between the genomes in the reference panel is governed by a transition matrix,
-and mismatches between a focal genome and a genome of the reference panel (from which it is “copied”) are permitted as emissions.
+and mismatches between a focal genome and a genome of the reference panel (from which it is ``copied'') are permitted as emissions.
 
@@
 and the other path having one switch but no mismatch are equally probable.
-We call this parameter k the “recombination penalty”.
+We call this parameter k the ``recombination penalty''.
 This parameter thus allows us to adjust the relative importance of recombination and mutation
@@
 
-A Viterbi algorithm can be used to find the most likely “copying path” of a focal sample
+A Viterbi algorithm can be used to find the most likely ``copying path'' of a focal sample
 through a reference panel of samples under the LS model.
@@
 we observed two types of outlier samples which had unusually high numbers of mutations:
-(1) “time travelers” and (2) “saltation” lineages.
+(1) ``time travelers'' and (2) ``saltation'' lineages.
 Time traveling samples are caused by errors in their collection dates.
@@
 assuming a substitution rate of 2 events per genome per month.
-Samples from “saltation” lineages (particularly the major VOCs) are characterized
+Samples from ``saltation'' lineages (particularly the major VOCs) are characterized
 by high numbers of mutations by the time of their detection.
@@
 (and less important but well supported poorly-fit lineages),
-we introduce “retrospective sample grouping”.
+we introduce ``retrospective sample grouping''.
 The idea is to gather evidence for a lineage within a time window rather than on a single day.
@@
 (to avoid adding batches of time travelers having the same collection date); and
-(c) they have at least two shared “non-immediate” reversions
+(c) they have at least two shared ``non-immediate'' reversions
 (so as to prevent unrelated time travelers from being grouped together).
@@
 but integrating Delta and Omicron into the ARG required further manual intervention.
-To ensure that Delta and Omicron are added, we introduce a third mechanism, “unconditional seeding”.
+To ensure that Delta and Omicron are added, we introduce a third mechanism, ``unconditional seeding''.
 The idea is to choose suitable samples belonging to each VOC,
 and attach them to the ARG unconditionally (i.e. without filtering using HMM cost).
-We sought “seed” samples for Delta, Omicron BA.1, and BA.2.
+We sought ``seed'' samples for Delta, Omicron BA.1, and BA.2.
 For Delta, we used two sets of seeds:
@@
 In this study, we used the SARS-CoV-2 sequences reassembled in a recent study \cite{Hunt2024}
-using an amplicon-aware assembly method (named “Viridian”).
+using an amplicon-aware assembly method (named ``Viridian'').
 The Viridian sequences contain far fewer artifacts compared to their publicly deposited counterparts on GenBank.
@@
 We downloaded 4,484,157 Viridian sequences obtained
-from samples collected from January, 2020 to March, 2023 (“batch 1”).
+from samples collected from January, 2020 to March, 2023 (``batch 1'').
 We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using MAFFT v7.525 \cite{Katoh2013},
@@
 We treated low-quality bases (represented by Ns), gaps and ambiguous characters in the sequences as missing data.
-When encoded as missing data, the bases do not inform sample matching (see “The Li and Stephens model”).
+When encoded as missing data, the bases do not inform sample matching (see ``The Li and Stephens model'').
 
@@
 We skipped low-quality sequences that had more than 500 missing sites
-(except when building the “base” ARG, when the maximum number of allowed missing bases was set to 10,000 instead).
+(except when building the ``base'' ARG, when the maximum number of allowed missing bases was set to 10,000 instead).
 We also skipped ~40,000 samples which had a collection date of December 31, 2020.
@@
 We suspect that this batch of samples might be unusually enriched for wrong collection dates,
-likely due to arbitrary assignment of the originally reported collection date of “2020” to be “31-12-2020”.
+likely due to arbitrary assignment of the originally reported collection date of ``2020'' to be ``31-12-2020''.
 Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
@@
 Then, we identified the top 100 sites which had the highest mutation counts in the ARG (URL).
-Only six of these sites overlapped with the “Problematic Sites”
+Only six of these sites overlapped with the ``Problematic Sites''
 downloaded from the UCSC Genome Browser website (URL),
@@
 The ages of the inserted non-sample nodes (the nodes in the local trees and the recombination nodes) were initially assigned to arbitrarily determined values.
-We ran \textit{tsdate} (“variational gamma”, version XX; \cite{Wohns2022} Pope et al., 2025) on the ARG to adjust the ages of these nodes, while fixing the ages of the sample nodes to the reported collection times (defined as fractional years before the collection date of the last sample added to the ARG, which is February 20, 2025 in the ARG described in this study).
+We ran \textit{tsdate} (``variational gamma'', version XX; \cite{Wohns2022} Pope et al., 2025) on the ARG to adjust the ages of these nodes, while fixing the ages of the sample nodes to the reported collection times (defined as fractional years before the collection date of the last sample added to the ARG, which is February 20, 2025 in the ARG described in this study).
 7. Assigning Pango lineages to non-sample nodes.
@@
 \subsubsection*{Calculating Pango lineage distance using pangonet}
-We define a “pseudo-phylogenetic” distance to quantify the divergence between two Pango lineages.
-The idea is to count the number of edges separating two given Pango lineages in a “pseudo-phylogeny”
+We define a ``pseudo-phylogenetic'' distance to quantify the divergence between two Pango lineages.
+The idea is to count the number of edges separating two given Pango lineages in a ``pseudo-phylogeny''
 which relates all the designated Pango lineages.
@@
 from the Scorpio designation at each node within the ARG for that week.
-Genetic variation was measured as the “expected heterozygosity” across lineages,
+Genetic variation was measured as the ``expected heterozygosity'' across lineages,
 which is the chance that two lineages chosen at random
@@
 
-In addition to the analysis of the 356 “robust” recombination events with a minimum of four nucleotides supporting each parent,
+In addition to the analysis of the 356 ``robust'' recombination events with a minimum of four nucleotides supporting each parent,
 we analyzed the relationship between the full set of 855 recombinants and case numbers.
@@
 %%%  relevance for the paper. Please report this
-%%%  information as: “Description: URL.”
+%%%  information as: ``Description: URL.''
 
@@
 \subsection{Analysis of Pango lineage ``origination'' events}
-
+\label{sec:supplement_pango_lineages}
 To analyze how well the ARG reflects the phylogenetic structure implicit in the
@@
 siblings). We then have 255 lineages (996490 samples) where the difference in
-the number of descendants of the first node's parent is < 100. The remaining 25
+the number of descendants of the first node's parent is $<$ 100. The remaining 25
 lineages (344079) are dominated by a few large lineages such as BA.1.1 (155595
@@
 the mutations occurring above the nodes with a Scorpio label associated with the VOC
-(e.g. “Alpha-like” encompasses “B.1.1.7-like” and “B.1.1.7-like+E484K”).
+(e.g. ``Alpha-like'' encompasses ``B.1.1.7-like'' and ``B.1.1.7-like+E484K'').
 Note that $\sim$98\% of all the mutations in the ARG occur above the nodes

### 2025-09-05 latex/main.tex
Batch of updates as far as pango X section

@@
 
-We find strong evidence for the recombination events leading to 17 Pango Xs in the ARG (Table 1),
-which passed the detection-level sc2ts quality control filters (Figure S1).
-These events are supported by
-high numbers of averted mutations (Table 1) and
-high numbers of loci matching their suggested parents (see their copying patterns in Figure SX).
-Among these Pango Xs are
-XA, which is the first designated Pango X that showed signs of community transmission \cite{Jackson2021}, and
-XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
-The recombinant origin of XA is well explained in the ARG by two parents and one mutation (\ref{fig:subgraphs}; Table 1),
-with sc2ts correctly identifying its parental Pango lineages and breakpoint location.
+We find strong evidence for the recombination events leading to 17 Pango Xs in
+the ARG (Table~\ref{tab:pango_x_lineages}), all of which pass the 
+robustness filtering criteria (Figure~\ref{fig:recombinant_qc}). 
+%% JK: Does saying this here help at all? I think we may be better off without 
+% These events are supported by high numbers of averted
+% mutations (Table 1) and high numbers of loci matching their suggested parents
+% (see their copying patterns in Figure SX). 
+Among these Pango Xs are XA, which
+is the first designated Pango X that showed signs of community transmission
+\cite{Jackson2021}, and XBB, which escaped vaccine-induced immunity and gained
+global prevalence \cite{Tamura2023}. The recombinant origin of XA is well
+explained in the ARG by two parents and one mutation (Figure \ref{fig:subgraphs};
+Table~\ref{tab:pango_x_lineages}), 
+with sc2ts correctly identifying its parental Pango lineages and
+breakpoint location.
 
 Furthermore, some of the Pango Xs above originate from shared recombination events in the 
-ARG (Table 1).
+ARG (Table~\ref{tab:pango_x_lineages}).
 An example of a set of ``nested'' closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
@@
 
-Sc2ts identified 12 Pango Xs in the ARG as non-recombinants (Table 1), on the basis of the HMM and available data.
+Sc2ts identified 12 Pango Xs in the ARG as non-recombinants
+(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
 For all these Pango Xs, the number of recombination-informative sites fell below our threshold of detection (i.e. $k = 4$ in the HMM).
 The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG
-by at most three mutations (Table 1) strongly supports their origins as non-recombinants.
+by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
+their origins as non-recombinants.
 For example, the closely related XN and XAU can be explained by two mutations and one mutation in the ARG, respectively (\ref{fig:subgraphs}).
 %%% TODO: Say something about XAJ and XAN.
+%JK: we should weaken this statement a bit, as we're not certain that they 
+% really are recombinants, especially XP.
 However, sc2ts did not detect several Pango Xs (XB, XP, and XAJ) as recombinants
@@
 \includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}
-\caption{Illustrative subgraphs in the ARG.
-Subgraphs around the XA-associated recombination node (A); the recombination node associated with a group of ``nested'' Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C).
+\caption{Illustrative subgraphs in the ARG. Subgraphs around the XA-associated
+recombination node (A); the recombination node associated with a group of
+``nested'' Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C).
 For more detailed visualizations of these subgraphs, see Document S2.}

### 2025-09-05 latex/main.tex
Batch of fixups

@@
 Below, we summarize our analysis of the Pango Xs in the ARG
-(for extended analyses, see Document S1 and Document S2).
+(for extended analyses, see Document~\ref{sec:supplement_pango_lineages} 
+and Document S2).
 
@@
 global prevalence \cite{Tamura2023}. The recombinant origin of XA is well
-explained in the ARG by two parents and one mutation (Figure \ref{fig:subgraphs};
+explained in the ARG by two parents and one mutation
+(Figure~\ref{fig:subgraphs}A;
 Table~\ref{tab:pango_x_lineages}), 
@@
 An example of a set of ``nested'' closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
-which descend from a single recombination event rather than five independent events in the ARG (\ref{fig:subgraphs}).
+which descend from a single recombination event rather than five independent
+events in the ARG (Figure~\ref{fig:subgraphs}B).
 These Pango Xs were designated around the same time.
@@
 their origins as non-recombinants.
-For example, the closely related XN and XAU can be explained by two mutations and one mutation in the ARG, respectively (\ref{fig:subgraphs}).
+For example, the closely related XN and XAU can be explained by two mutations
+and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C).
 %%% TODO: Say something about XAJ and XAN.
@@
 \end{footnotesize}
-
-% pango & averted & left & right & samples & nested \\
-
-% \midrule
-% XC & 37 & 26768 & 27390 & 5 & \{\} \\
-% XBR & 22 & 22034 & 22190 & 1 & \{\} \\
-% XA & 16 & 20411 & 21765 & 39 & \{\} \\
-% XS & 11 & 9054 & 10449 & 17 & \{\} \\
-% XL & 8 & 5925 & 8393 & 64 & \{\} \\
-% XBG & 8 & 22305 & 22917 & 25 & \{\} \\
-% XBD & 7 & 23020 & 24620 & 30 & \{\} \\
-% XQ & 7 & 4322 & 5386 & 55 & \{BA.2:37, $\star$\} \\
-% XBB & 6 & 22332 & 22577 & 6452 & \{BA.2:1\} \\
-% XM & 6 & 17411 & 21595 & 26 & \{BA.2:16, XAL:3\} \\
-% XBF & 5 & 5184 & 9866 & 185 & \{\} \\
-% XF & 5 & 5387 & 6402 & 16 & \{\} \\
-% XY & 5 & 11538 & 12880 & 23 & \{\} \\
-% XG & 4 & 5925 & 6513 & 3 & \{\} \\
-% XW & 4 & 2833 & 4321 & 32 & \{\} \\
-% \midrule
-% \multicolumn{5}{l}{Type II RE event}\\
-% alias & averted & left & right & total & by pango \\
-% \midrule
-% XAE+ & 6 & 24504 & 26060 & 253 & {BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
-% XE/XH & 6 & 10448 & 11283 & 1156 & {XE:1116, BA.2:37, XH:2, XAF:1} \\
-% XBH & 6 & 15452 & 22001 & 6 & {BA.2:4, XBH:2} \\
-% XBM & 6 & 22305 & 22917 & 12 & {XBM:10, BF.3:2} \\
-% XAF & 5 & 10199 & 10447 & 36 & {BA.2:35, XAF:1} \\
-% XJ & 5 & 13196 & 17410 & 85 & {XJ:68, BA.2:17} \\
-% \midrule
-% \multicolumn{5}{l}{Nested monophyletic Pango X lineage}\\
-% pango & alias & path len & time & mutations & samples & nested \\
-% \midrule
-% XAP & XAE+ & 4 & 67 & 1 & 20 & \{\} \\
-% XZ & XAE+ & 4 & 48 & 1 & 48 & \{\} \\
-% XAE & XAE+ & 2 & 47 & 7 & 9 & \{\} \\
-% XAF & XAF & 5 & 140 & 8 & 1 & \{\} \\
-% XAL & XM & 3 & 66 & 2 & 3 & \{\} \\
-% XR & XQ & 3 & 11 & 4 & 17 & \{BA.2:1\} \\
-% XU & XQ & 6 & 103 & 4 & 1 & \{\} \\
-% XAM & XQ & 6 & 43 & 3 & 21 & \{\} \\
-% XAA & XQ & 6 & 33 & 3 & 17 & \{\} \\
-% XAG & XQ & 6 & 55 & 4 & 6 & \{\} \\
-
-% \bottomrule
-% \end{tabular}
-
 \end{table}
@@
 These non-phylogenetic methods use the same underlying approach: 
-they scan candidate recombinant sequences for mutations associated with different Pango lineages.
+they scan candidate recombinant sequences for mutations associated with different 
+Pango lineages.
 
@@
 We find that the methods performed well overall,
-with no significant difference in concordance among them (Table 2; p-value = XX, chi-square test),
+with no significant difference in concordance among them
+(Table~\ref{tab:x_method_concordance}; p-value = XX, chi-square test),
 indicating that sc2ts can characterize recombinants as well as the state-of-the-art recombination detection methods.
@@
 to correctly classify a sequence as recombinant, as shown by
-a shorter time to their MRCA (\ref{fig:averted_mutations}) and
-lower distance between their Pango identity (\ref{fig:averted_mutations}).
-
-\subsection*{Table 2. Concordance among methods in detecting Pango Xs.}
+a shorter time to their MRCA (Figure \ref{fig:averted_mutations}B) and
+lower distance between their Pango identity (Figure
+\ref{fig:averted_mutations}C).
 
 \begin{table}
+\caption{Concordance among methods in detecting Pango Xs.
+\label{tab:x_method_concordance}}
 \begin{tabular}{llll}
@@
 \subsection*{Quantification of evidence for recombination}
-
 %%% SHZ: We may not need this section at all, but I'm keeping it for now.
@@
 %%% The second and third paragraphs should probably be removed entirely.
-
 Sc2ts provides a method to facilitate the discovery of recombination events,
 by comparing how well a sample fits the ARG under a model allowing recombination
-versus a model disallowing recombination, as measured by the number of resulting mutations (\ref{fig:methodology}).
-The more mutations are ``averted'' by allowing recombination, the stronger the evidence for recombination.
-We calculated the number of averted mutations for all the ``robust'' recombination events (STAR Methods),
-which include all the Pango X-associated recombination events (Table 1).
+versus a model disallowing recombination, as measured by the number of resulting 
+mutations (Figure~\ref{fig:methodology}).
+The more mutations are ``averted'' by allowing recombination, the stronger the 
+evidence for recombination.
+We calculated the number of averted mutations for all the ``robust'' recombination 
+events (STAR Methods),
+which include all the Pango X-associated recombination events
+(Table~\ref{tab:pango_x_lineages}) and those studied by Jackson et
+al.\cite{Jackson2021}
 
 Our analysis of the number of averted mutations reveals a spectrum of plausibility
-across all the ``robust'' recombination events (\ref{fig:averted_mutations}).
-Examples of highly plausible recombinants are XA and XBB (for an extended analysis, see Document S1).
+across all the ``robust'' recombination events
+(Figure~\ref{fig:averted_mutations}A).
+Examples of highly plausible recombinants are XA and XBB (for an extended analysis, 
+see Document~\ref{sec:supplement_pango_lineages}).
 The ``copying pattern'' of XA visually confirms
-that a high number of recombination-informative sites support the suggested parents (\ref{fig:averted_mutations}).
+that a high number of recombination-informative sites support the suggested parents 
+(Figure~\ref{fig:averted_mutations}D).
 XW is an example of a marginally well-supported recombinant.
 The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent.
+% FIXME What's document S3?
 The copying patterns of all the recombination events are shown in Document S3.
 
+[FIXME - discuss BA.5 and the BA.2 event instead.]
 Our results suggest that Delta and Omicron BA.2 might have recombinant origins,
@@
 
-%%% TODO: Old version as placeholder. This figure is not yet finalised.
 \begin{figure}[h]

### 2025-09-05 latex/main.tex
Some tex fixups

@@
 (Table~\ref{tab:pango_x_lineages}) and those studied by Jackson et
-al.\cite{Jackson2021}
+al.\cite{Jackson2021} (Table \ref{tab:jackson}).
 
@@
 recombination nodes in Table \ref{tab:pango_x_lineages} are labelled starting with `X' and
-those identified by \cite{Jackson2021} (see Table \ref{tab:jackson}) are prefixed with a `J', e.g.
+those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
 Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1). 
@@
 \subsection*{Recombination breakpoint intervals along the genome}
-
 We explored the spatial distribution of the breakpoint locations of
@@
 For each event, we computed an interval estimate of its breakpoint location,
-using its size as a measure of precision (median, 2,207 bases; STAR Method).
+using its size as a measure of precision (median, 2,207 bases; STAR Methods).
 
 We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
-particularly at the left boundary of the Spike gene (\ref{fig:recombinant_intervals}),
-consistent with the pattern of excess observed in previous studies \cite{Turakhia2022,Li2024}.
+particularly at the left boundary of the Spike gene
+(Figure~\ref{fig:recombinant_intervals}A,B),
+consistent with patterns observed in previous 
+studies\cite{Turakhia2022,Li2024}.
 While this might result from a higher recombination rate in this region,
@@
 allowing recombination endpoints to be estimated more precisely.
-Reassuringly, sites having unusually high mutation counts are absent in this region (\ref{fig:recombinant_intervals}),
-which would confound recombination detection.
 
@@
 breakpoint interval size and the amount of divergence between recombinant parents,
-which is measured by time to their MRCA (\ref{fig:recombinant_intervals}).
+which is measured by time to their MRCA (Figure~\ref{fig:recombinant_intervals}).
 When the two parental lineages have diverged for longer periods of time (the rightmost points),
@@
 For instance, the breakpoint intervals tend to be the smallest for events involving recombination
-between the deeply divergent Omicron BA.1 and BA.2 sublineages (the red points in \ref{fig:recombinant_intervals}).
+between the deeply divergent Omicron BA.1 and BA.2 sublineages 
+(the red points in Figure~\ref{fig:recombinant_intervals}).
 
@@
 Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
-sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in \ref{fig:recombinants_over_time}).
+sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in 
+Figure~\ref{fig:recombinants_over_time}).
 These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset.
@@
 Overall, the number of recombinant events rose with the global number of cases
-(Figure \ref{fig:recombinants_over_time}B; linear regression: $1.0 + 2.7 \; 10^{-7} x$, $p = 1.1 \; 10^{-7}$).
+(Figure~\ref{fig:recombinants_over_time}B; 
+% JK is this correct?? we had 1.0 + 2.7 \; 10^{-7} x$ which I found confusing
+linear regression: $1.0 + 2.7\times 10^{-7} x$, $p = 1.1 \times10^{-7}$).
 Recombination events also rose with the diversity of lineages in each week
-(Figure \ref{fig:recombinants_over_time}C; linear regression: $0.9 + 4.8 x$, $p = 3.4\;10^{-7}$),
+(Figure \ref{fig:recombinants_over_time}C; linear regression: 
+$0.9 + 4.8 x$, $p = 3.4\times10^{-7}$),
 measured as the chance that two randomly drawn viruses have different Scorpio designations
@@
 We then asked which single measure best explains the number of recombination events inferred each week,
-considering $n$, $H$, $n \; H$, $n^2$, $n^2 H$.
+considering $n$, $H$, $n H$, $n^2$, $n^2 H$.
 We included the squared number of cases each week thinking that it might better predict co-infections.
-The best single predictor of recombination events involved the product of cases and expected heterozygosity ($n \; H$),
+The best single predictor of recombination events involved the product of cases and 
+expected heterozygosity ($n H$),
 with an adjusted $R^2$ of 23.7\%.
-This measure of case diversity ($n \; H$) explained 49\% more of the variation in recombination events per week
+This measure of case diversity ($n H$) explained 49\% more of the variation in recombination events per week
 than the next best predictor ($H$, adjusted $R^2$ of 15.9\%).

### 2025-09-05 latex/main.tex

@@
     \includegraphics[width=0.85\linewidth]{recombinants_over_time.pdf}
-    \caption{Recombination events across the pandemic.
-(A) Timing of the recombination events (black curve, right y-axis) relative to the global case numbers (\cite{Dong2020}; left y-axis).
-Cases are color-coded by their Scorpio labels, with the key showing those groupings with $\geq 1,000$ samples in the Viridian dataset.
-Recombinants more commonly occur (B) in weeks with high case numbers and (C) in weeks with extensive lineage variation, as estimated by high expected heterozygosity; solid lines show fits by linear regression, allowing for non-zero intercepts to account for unknown cases and heterozygosity not captured by the Scorpio labels.}
+    \caption{Recombination events across the pandemic. (A) Timing of the
+recombination events (black curve, right y-axis) relative to the global case
+numbers (\cite{Dong2020}; left y-axis). Cases are color-coded by their Scorpio
+labels, with the key showing those groupings with $\geq 1,000$ samples in the
+Viridian dataset. Recombinants more commonly occur (B) in weeks with high case
+numbers and (C) in weeks with extensive lineage variation, as estimated by high
+expected heterozygosity; solid lines show fits by linear regression, allowing
+for non-zero intercepts to account for unknown cases and heterozygosity not
+captured by the Scorpio labels.}
     \label{fig:recombinants_over_time}
@@
 which overwhelmed existing computational approaches,
-motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
+motivating the rapid development of new methods\cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
 Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
@@
 
-\section*{SUPPLEMENTAL INFORMATION INDEX}
-
-%%%  Supplemental information must be uploaded as
-%%%  separate files. In the main text, please list the
-%%%  files to be included in a brief index. For details,
-%%%  please review the supplemental information guidelines:
-
-%%%  Journals with STAR Methods:
-%%%  https://www.cell.com/STAR-supplemental-information
-
-\begin{description}
-  \item Figures S1-SX and their legends in a PDF
-  \item Table S1. A descriptive title for an Excel file that was too large to appear in the PDF
-\end{description}
+% \section*{SUPPLEMENTAL INFORMATION INDEX}
+
+% %%%  Supplemental information must be uploaded as
+% %%%  separate files. In the main text, please list the
+% %%%  files to be included in a brief index. For details,
+% %%%  please review the supplemental information guidelines:
+
+% %%%  Journals with STAR Methods:
+% %%%  https://www.cell.com/STAR-supplemental-information
+
+% \begin{description}
+%   \item Figures S1-SX and their legends in a PDF
+%   \item Table S1. A descriptive title for an Excel file that was too large to appear in the PDF
+% \end{description}
 

### 2025-09-05 latex/main.tex
Indel table

@@
 
+List of indels from Table S1 of Li et al. \cite{Li2023}
+occurring at minimum frequency of 0.01 (out of 9,149,680 GISAID sequences passing filters) and 
+their frequencies in the sc2ts ARG (maximum of max inheritors divided by 2,482,157 sampled sequences).
+
 % See https://github.com/jeromekelleher/sc2ts-paper/issues/472
-At minimum, we want to show compare the frequencies of the top N indels in the
-ARG and those reported in Li et al. (2023). This can be shown in a table.
+\begin{table}
+\caption{Major indels.}
+\label{tab:indels}
+\begin{tabular}{rrcrrrr}
+\toprule & \multicolumn{4}{c}{Li et al.} & \multicolumn{2}{c}{Sc2ts ARG} \\
+\cmidrule(lr){1-5} \cmidrule(lr){6-7}
+Start & Length & Region & Count & Frequency & Count & Frequency \\
+\midrule
+28362 & 9 & N & 3939164 & 0.4305 & 871774 & 0.3512 \\
+28271 & 1 & Intergenic & 3853692 & 0.4212 & 1118363 & 0.4506 \\
+21765 & 6 & S & 3179887 & 0.3475 & 339681 & 0.1368 \\
+11288 & 9 & ORF1ab-nsp6 & 3040576 & 0.3323 & 290408 & 0.1170 \\
+28248 & 6 & ORF8 & 3026186 & 0.3307 & 1114669 & 0.4491 \\
+22029 & 6 & S & 3010665 & 0.3290 & 1112410 & 0.4482 \\
+11283 & 9 & ORF1ab-nsp6 & 2011934 & 0.2199 & 40 & 0.0000 \\
+6513 & 3 & ORF1ab-nsp3 & 1984683 & 0.2169 & 340765 & 0.1373 \\
+21633 & 9 & S & 1941962 & 0.2122 & 532072 & 0.2144 \\
+21987 & 9 & S & 1918094 & 0.2096 & N/A & N/A \\
+22194 & 3 & S & 1833050 & 0.2003 & 312706 & 0.1260 \\
+21991 & 3 & S & 912944 & 0.0998 & 288831 & 0.1164 \\
+686 & 9 & ORF1ab-nsp1 & 115450 & 0.0126 & N/A & N/A \\
+\bottomrule
+\end{tabular}
+\end{table}
 

### 2025-09-05 latex/main.tex
Updates from Shing

@@
 % Referenced by the main Results.
-Basically, we want to say we did a lot of work to integrate signals of major
-indels into the ARG, making it an even more powerful resource overall. Also,
-indels are encoded by the Mutation-Annotated Tree format used by UShER
-phylogenies (though PanMAT does this).
-
-List of indels from Table S1 of Li et al. \cite{Li2023}
-occurring at minimum frequency of 0.01 (out of 9,149,680 GISAID sequences passing filters) and 
-their frequencies in the sc2ts ARG (maximum of max inheritors divided by 2,482,157 sampled sequences).
+Li et al. \cite{Li2023} performed a large-scale analysis of indels 
+using a global GISAID dataset (sampled up to August, 2022). 
+We examined the major indels reported in Table S1 of that study 
+which have a minimum frequency of 0.01 
+(sequence count divided by 9,149,680 filtered sequences).
+Most of these indels (all deletions) are encoded in the ARG (\ref{tab:indels}) 
+and have high frequencies in the ARG
+(sequence count divided by the total number of samples), 
+except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
+During site remapping, the 9-base deletion at 11283 was split up into 
+a 5-base deletion at 11283 (count of 339867) and 
+a 4-base deletion at 11288 (count of 871782).
+The 686 deletion occurs below our threshold for inclusion into the ARG.
 
@@
 \begin{table}
-\caption{Major indels.}
+\caption{Major deletions in Li et al. (2023).}
 \label{tab:indels}

### 2025-09-05 latex/main.tex
Update local tree inference details.

@@
 The basic idea is to incrementally update an ARG regularly
-with the sequences collected since the previous batch \ref{fig:methodology}.
+with the sequences collected since the previous batch (Figure
+\ref{fig:methodology}).
 We used daily batches for SARS-CoV-2,
@@
 that captures many of the key features of the joint processes of mutation and recombination.
-It is widely used in a range of applications in population genetics, including ARG inference \cite{Kelleher2019,Gunnarsson2024}.
-The LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states),
+It is widely used in a range of applications in population genetics, including ARG 
+% Does Threads use LS?
+inference \cite{Kelleher2019,Gunnarsson2024}.
+?he LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states),
 which are probabilistically emitted as an imperfect mosaic of a set of genomes in a reference panel (i.e., the hidden states).
@@
 (2) the number of genomes in the reference panel ($n$).
-Sc2ts uses the same formulation except that n is set to 1.
+Sc2ts uses the same formulation except that $n$ is set to 1.
 This special case of the LS model leads to an intuitive interpretation of the ratio
 between the probability of switching and the probability of mismatch:
-When considering only two copying paths, the path having k mismatches
+When considering only two copying paths, the path having $k$ mismatches
 and the other path having one switch but no mismatch are equally probable.
-We call this parameter k the ``recombination penalty''.
+We call this parameter $k$ the ``recombination penalty''.
+% TODO mention that this can be thought of as extended parsimony.
 This parameter thus allows us to adjust the relative importance of recombination and mutation
@@
 
-In our early attempts to build ARGs, we explored using several values of $k$.
-We ran sc2ts with $k$ set to 3, 4, and 5 on the samples collected in 2020 \cite{Zhan2023}(REF).
+We explored using several values of $k$,
+by running sc2ts with $k$ set to 3, 4, and 5 on the samples collected in 2020.
 We would expect that a good value for $k$ should lead to parsimonious ARGs
@@
 but ARGs built with $k$ set to 5 failed
-to include the well supported recombination events documented in a previous study \cite{Jackson2021}.
+to include the well supported recombination events documented in a previous study\cite{Jackson2021}.
 Hence, we decided to use $k$ set to 4 to infer the final ARG described in this study.
@@
 Sc2ts builds a local phylogeny relating the samples in each cluster as follows.
-First, it infers a tree using a standard neighbor-joining algorithm \cite{Saitou1987}.
-Then, mutations are mapped back onto it using the Fitch-Hartigan parsimony algorithm \cite{Fitch1971,Hartigan1973}.
+First, it infers a tree using a standard neighbor-joining
+algorithm\cite{Saitou1987}. We first compute the Hamming distance between
+all pairs of samples using scipy~\cite{Virtanen2020scipy},
+and then use biotite\cite{Kunzmann2018biotite,Kunzmann2023biotite}
+to compute the neighbor-joining tree.
+Then, mutations are mapped back onto it using the Fitch-Hartigan parsimony 
+algorithm \cite{Fitch1971toward,Hartigan1973minimum} implemented in tskit.
 Finally, the resulting tree is added to the ARG at an attachment node
-and the ARG is locally modified using heuristics to improve parsimony (Figure SX).
-For a recombinant, the attachment node is the inserted (non-sample) recombination node associated with it.
-For a non-recombinant, it is the common parent in the copying paths of the samples.
+and the ARG is locally modified using heuristics to improve parsimony.
+[TODO we need to define these?]
 

### 2025-09-05 latex/main.tex
Reshuffled some text

@@
 STS has been extended and applied to
-forward-time simulations \cite{Kelleher2018,Haller2019},
+forward-time simulations \cite{Kelleher2018efficient,Haller2019},
 calculation of population genetics statistics \cite{Ralph2020} and
@@
 These algorithms have led to performance increases of several orders of magnitude over
-the state-of-the-art in a range of applications \cite{Kelleher2016,Kelleher2018,Kelleher2019,Ralph2020,Baumdicker2022}.
+the state-of-the-art in a range of applications
+\cite{Kelleher2016,Kelleher2018efficient,Kelleher2019,Ralph2020,Baumdicker2022}.
 The STS encoding is also very concise, allowing, for example,
@@
 The SARS-CoV-2 ARGs inferred using sc2ts can be conveniently and efficiently analyzed
-using the mature and feature-rich tskit library \cite{Kelleher2018,Ralph2020}.
+using the mature and feature-rich tskit library \cite{Kelleher2018efficient,Ralph2020}.
 Tskit is widely used in population genetics, and has been incorporated into many downstream applications
@@
 (except when building the ``base'' ARG, when the maximum number of allowed missing bases was set to 10,000 instead).
-We also skipped ~40,000 samples which had a collection date of December 31, 2020.
+We also skipped $\sim$40,000 samples which had a collection date of December 31, 2020.
 The sampling frequency of this date is alarmingly high compared to other dates (at most 15,000 samples per day).
@@
 \subsubsection*{Identification of robust recombination events}
-[TODO update refs!]
+[TODO This section needs a full pass. I've merged in a section that was in the
+supplement with an extremely terse section on counting supporting loci.]
+
+% This para was *{Counting supporting loci for recombinant parents}
+
+When counting such loci, we considered nearly adjacent sites (separated by at
+most three bases) as single loci, to avoid over-counting support from
+correlated loci, which would lead to artificial signals of recombination. Also,
+when counting supporting loci for a focal parent, we penalized loci where the
+character(s) in the recombinant matches that in the non-focal parent but not in
+the focal parent.
 
@@
 
-We used Pango lineage designations to select \textit{representative} samples
+We used Pango lineage designations to select representative samples
 for the majority of the 2026 Pango lineages in the ARG.
@@
 and where the internal node is itself designated as the focal Pango.
-The representative sample is then chosen as the oldest sample under the origination node
+The representative sample is then chosen as the oldest sample descending from the origination node
 whose own descendants (if any) are entirely of the focal Pango type.
-We simplify the ARG to include only the representative samples,
+We ``simplify''\cite{Kelleher2018efficient,Wong2024} 
+the ARG to include only the representative samples,
 resulting in an intermediate ARG of 1867 samples,
 then additionally remove internal samples,
-known recombinants (i.e. the Pango Xs) and three "singleton recombinant samples"
+known recombinants (i.e. the Pango Xs) and three ``singleton recombinant
+samples''
 in which a single Pango lineage appears immediately below a recombination node.
@@
 and extract non-recombinant structures from the ARG
-by creating a "base tree" of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, and BA.5),
+by creating a ``base tree'' of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, and BA.5),
 a Delta subtree of 160 Pangos,
@@
 sometimes the UShER tree shows greater resolution, as in BA.2.9.
-
-\subsubsection*{Counting supporting loci for recombinant parents}
-When counting such loci, we considered nearly adjacent sites (separated by at most three bases) as single loci,
-to avoid over-counting support from correlated loci, which would lead to artificial signals of recombination.
-Also, when counting supporting loci for a focal parent,
-we penalized loci where the character(s) in the recombinant matches that in the non-focal parent but not in the focal parent.
 

### 2025-09-06 latex/main.tex
Fixup

@@
 inference \cite{Kelleher2019,Gunnarsson2024}.
-?he LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states),
+The LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states),
 which are probabilistically emitted as an imperfect mosaic of a set of genomes in a reference panel (i.e., the hidden states).
@@
 (sequence count divided by 9,149,680 filtered sequences).
-Most of these indels (all deletions) are encoded in the ARG (\ref{tab:indels}) 
+Most of these indels (all deletions) are encoded in the ARG (Table~\ref{tab:indels}) 
 and have high frequencies in the ARG

### 2025-09-06 latex/main.tex
Updates from Overleaf
Add citation for pyfaidx
Closes #203

@@
 \begin{table}
-\caption{Concordance among methods in detecting Pango Xs.
-\label{tab:x_method_concordance}}
+\caption{Concordance among methods in detecting Pango X lineages.}
+\label{tab:method_concordance}
 \begin{tabular}{llll}
@@
 \hline
- Sc2ts & XX & XX & XX \\ [1ex]
+Sc2ts & 14 (93.3) & 13 (86.7) & 540.5 (315, 766) \\ [1ex]
 \hline
- RecombinHunt-GISAID & XX & XX & XX \\ [1ex]
+RH-GISAID & 13 (100.0) & 10 (76.9) & 764.0 (4, 1202) \\ [1ex]
 \hline
- RecombinHunt-Nextstrain & XX & XX & XX \\ [1ex]
+RH-Nextstrain & 12 (85.7) & 11 (78.6) & 1160.0 (559, 2785) \\ [1ex]
 \hline
- CovRecomb & XX & XX & XX \\ [1ex]
- \hline
- Rebar & XX & XX & XX \\ [1ex]
+CovRecomb & 7 (87.5) & 4 (50.0) & 1562.5 (217, 13208) \\ [1ex]
 \hline
@@
 
+Sc2ts is implemented in Python building on core PyData infrastructure
+such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
+and Pandas ,
+using pyfaidx\cite{Shirley2015efficient} to access alignment data
+[TODO JK to fill this out]
+
 \subsubsection*{Ancestral Recombination Graphs}
@@
 \label{tab:indels}
+\begin{center}
 \begin{tabular}{rrcrrrr}
@@
 \end{tabular}
+\end{center}
 \end{table}

### 2025-09-06 notebooks/s4.7_merge.R
Remove old preprint notebooks

@@ -1,23 +1,0 @@
-library(data.table)
-
-args = commandArgs(trailingOnly=TRUE)
-
-results_dir <- args[1]
-
-METADATA   <- paste0(results_dir, "/metadata.minimal.tsv")
-NEXTCLADE  <- paste0(results_dir, "/nextclade.minimal.tsv")
-OUTPUT     <- paste0(results_dir, "/minimal.tsv")
-
-print(paste("Loading metadata:", METADATA))
-metadata_df <- data.table::fread(METADATA)
-
-print(paste("Loading nextclade:", NEXTCLADE))
-nextclade_df <- data.table::fread(NEXTCLADE)
-
-print(paste("Merging metadata and nextclade"))
-meta_gisaid <- merge(metadata_df, nextclade_df, by=c("strain"), all.x=TRUE)
-
-print(paste("Saving metadata:", OUTPUT))
-data.table::fwrite(meta_gisaid, file=OUTPUT, sep="\t")
-
-print("Done")

### 2025-09-08 latex/main.tex
Updates from Overleaf

@@
 (3) rebar (https://github.com/phac-nml/rebar).
-These non-phylogenetic methods use the same underlying approach: 
-they scan candidate recombinant sequences for mutations associated with different 
-Pango lineages.
-
-To evaluate the concordance of the methods in identifying the parental Pango lineages
-and locating the breakpoints of the recombinants, we focus on the Pango Xs, using the data
-from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
-For RecombinHunt, we took the detection results for the Pango Xs presented in the original study;
-for CovRecomb, we summarized the results from the supplementary data of the original study (STAR Methods);
-and for rebar, we obtained results by using it to classify the Pango X-associated recombination events in the ARG (STAR Methods).
-For each method, we determined whether the inferred parental lineages and breakpoint intervals are
-concordant with those identified by the community (STAR Methods).
-We find that the methods performed well overall,
-with no significant difference in concordance among them
-(Table~\ref{tab:x_method_concordance}; p-value = XX, chi-square test),
-indicating that sc2ts can characterize recombinants as well as the state-of-the-art recombination detection methods.
-
-We also compared the sensitivity of sc2ts and rebar.
-Of all the ``robust'' recombination events detected by sc2ts,
-167 (43\%) were also detected as recombinants by rebar (STAR Methods).
+These methods use the same underlying approach: 
+they scan candidate recombinant sequences for mutations associated with different Pango lineages 
+without using a phylogeny.
+
+To evaluate the concordance of the methods in 
+identifying the Pango lineages of recombinant parents and locating recombination breakpoints, 
+we focus on the Pango Xs descending from the Type I recombination events (Table \ref{tab:pango_x_lineages}), 
+using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
+For RecombinHunt, we took available detection results for these Pango Xs reported in the original study;
+for CovRecomb, we summarized the results from the original study (STAR Methods);
+For each Pango X, we determined whether the inferred parental lineages and breakpoint intervals 
+are concordant with those suggested by the community (STAR Methods).
+We found that the methods performed well overall, 
+with no evidence for significant difference in the concordance rate among them (Table \ref{tab:x_method_concordance}),
+indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.
+
+Furthermore, we compared the sensitivity of sc2ts and rebar.
+We ran rebar on the ``robust'' recombination events detected by sc2ts (STAR Methods), 
+and found that 167 (43\%) of them were also detected as recombinants by rebar.
 We hypothesize that this disagreement arose from differences in sensitivity between these methods.
-Consistent with this, we find that rebar needs a higher level of divergence between parents
+Consistent with this, we found that rebar needs a higher level of divergence between parents
 to correctly classify a sequence as recombinant, as shown by 
 a shorter time to their MRCA (Figure \ref{fig:averted_mutations}B) and
-lower distance between their Pango identity (Figure
-\ref{fig:averted_mutations}C).
+a lower distance between their Pango identity (Figure \ref{fig:averted_mutations}C).
 
 \begin{table}
-\caption{Concordance among methods in detecting Pango X lineages.}
+\caption{
+Concordance among methods in characterizing Pango X lineages. 
+The Pango X lineages analyzed here are associated with the Type 1 recombination events (related to Table 1).
+For RecombinHunt and CovRecomb, the concordance rates are calculated 
+based on the Pango Xs for which results are available from the original studies.
+Also, for RecombinHunt, there are two sets of detection results, 
+one set obtained using mutational profiles computed from a global GISAID dataset and another from a Nextstrain dataset.
+
+}
 \label{tab:method_concordance}
@@
 \hline
-\textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bp)} \\ [1ex]
+\textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bases)} \\ [1ex]
 \hline

### 2025-09-09 latex/main.tex

@@
 The ARG represents the full evolutionary history of SARS-CoV-2 and 
-is in close agreement with the state-of-the-art UShER phylogeny built using the Viridian 
-dataset\cite{Hunt2024}.
+is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024}.
 First, we validated the broad structure of the ARG against the UShER phylogeny\cite{Hunt2024}
@@
 with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts).
-We also compared the mutational spectra of major VOCs in the ARG
-against the mutational spectra computed from the UShER phylogeny (STAR Methods),
-finding a close match to previous reports\cite{Bloom2023}
-(Figure~\ref{fig:mutational_spectra}).
+We also compared the mutational spectra of major VOCs in the ARG against the mutational spectra 
+from the Viridian UShER phylogeny\cite{Hunt2024} and a previous study\cite{Bloom2023} (STAR Methods),
+finding a close match to them (Figure~\ref{fig:mutational_spectra}).
 
@@
 identifying the Pango lineages of recombinant parents and locating recombination breakpoints, 
-we focus on the Pango Xs descending from the Type I recombination events (Table \ref{tab:pango_x_lineages}), 
+we focus on the Pango Xs associated with the Type I recombination events (Table \ref{tab:pango_x_lineages}), 
 using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
-For RecombinHunt, we took available detection results for these Pango Xs reported in the original study;
-for CovRecomb, we summarized the results from the original study (STAR Methods);
-For each Pango X, we determined whether the inferred parental lineages and breakpoint intervals 
-are concordant with those suggested by the community (STAR Methods).
+For RecombinHunt, we took detection results for these Pango Xs reported in the original study \cite{Alfonsi2024};
+for CovRecomb, we summarized the results from the original study \cite{Li2024} (STAR Methods).
+For each method, we determined the Pango Xs for which the inferred parental lineages and breakpoint intervals 
+are concordant with those identified by the community (STAR Methods).
 We found that the methods performed well overall, 
-with no evidence for significant difference in the concordance rate among them (Table \ref{tab:x_method_concordance}),
+with comparable concordance rates (Table \ref{tab:x_method_concordance}),
 indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.
@@
 based on the Pango Xs for which results are available from the original studies.
-Also, for RecombinHunt, there are two sets of detection results, 
-one set obtained using mutational profiles computed from a global GISAID dataset and another from a Nextstrain dataset.
-
-}
+Also, for RecombinHunt, two sets of detection results are available, 
+one set obtained using mutational profiles computed from a global GISAID dataset and another from a Nextstrain dataset.}
 \label{tab:method_concordance}
@@
 
+\subsubsection{Mutational spectra}
+To validate the mutational information contained in the sc2ts ARG,
+we compared the all-site mutational spectra of major VOCs in the ARG,
+a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024}, and
+a global UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
+Mutational spectra are relative proportions of each type of nucleotide mutation.
+
+First, we calculated the mutational spectra of Alpha, Delta, Omicron BA.1, BA.2, BA.4, and BA.5
+in the sc2ts ARG, considering all the mutations above the sample nodes and internal nodes
+which have Scorpio labels associated with the VOCs (e.g., B.1.1.7-like and B.1.1.7-like+E484K for Alpha).
+These Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin.
+Note that $\sim$98\% of all the mutations in the ARG occur
+above the nodes which have a Scorpio label associated with Alpha, Delta, BA.1, BA.2, BA.4, or BA.5
+(excluding the nodes with unassigned Scorpio labels).
+
+Next, we calculated the mutational spectra of the same VOCs
+using the mutational counts from an UShER phylogeny analyzed in Bloom et al.\cite{Bloom2023},
+which was built using Nextclade and has samples collected globally up to November, 2022.
+We used the following mapping between the WHO labels and the Nextstrain labels:
+Alpha, 20I; Detal, 21I and 21J; BA.1, 21K; BA.2, 21L; BA.4, 22A, and BA.5, 22B.
+The mutation count data are available at https://github.com/jbloomlab/SARS2-mut-spectrum/.
+
+Lastly, we calculated the mutational spectra of the same VOCs
+in the Viridian UShER phylogeny\cite{Hunt2024},
+which was pruned down to the samples present in the sc2ts ARG.
+We used the mutations above the samples based on their Scorpio labels,
+which are available in the Viridian metadata provided by Hunt et al.\cite{Hunt2024}.
+
+\subsubsection{Concordance among recombination detection methods}
+For RecombinHunt, we obtained two sets of detection results. 
+One set is based on mutational profiles calculated from a global GISAID dataset, 
+and the other set from a global Nextstrain dataset. 
+We adjusted the RecombinHunt-identified breakpoint interval 
+so that they are right-exclusive by adding one to the right coordinate.
+
+For CovRecomb, we summarized the detection results for the Pango Xs as follows. 
+First, we obtained the independent recombination events 
+in Table S4 corresponding to a Pango X, if available. 
+When multiple events corresponded to a Pango X, 
+we took the results of the event having the most epidemic recombinants. 
+Then, for other Pango Xs, we collated the data from a file (URL) 
+containing all the detection results produced by the study. 
+For each of these Pango Xs, 
+we took the most commonly observed combination of 
+left parent, right parent, and breakpoint interval.
+We adjusted the CovRecomb-identified breakpoint interval 
+so that they are right-exclusive by adding one to the right coordinate.
+
+We ran rebar on the sampled sequences associated with each robust recombination event. 
+Rebar identifies a sequence as a recombinant if it contains sets of mutations 
+predetermined to be associated with different Pango lineages.
+
+The parents of a Pango X inferred using a method are considered concordant 
+if their Pango lineage assignments are at least as specific as the community-identified parents.
+Also, a breakpoint interval of a Pango X inferred using a method is considered concordant 
+if it overlaps with the community-identified breakpoint interval by at least one base.
+We adjusted the breakpoint intervals from the community 
+by subtract one from the left coordinate and adding two to the right coordinate.
+
 \subsubsection*{Calculating Pango lineage distance using pangonet}
@@
 We examined the major indels reported in Table S1 of that study 
-which have a minimum frequency of 0.01 
+which have a minimum frequency of 1\% 
 (sequence count divided by 9,149,680 filtered sequences).
@@
 \label{tab:indels}
-\begin{center}
 \begin{tabular}{rrcrrrr}
-\toprule & \multicolumn{4}{c}{Li et al.} & \multicolumn{2}{c}{Sc2ts ARG} \\
+\toprule & \multicolumn{4}{c}{Li et al. (2023)} & \multicolumn{2}{c}{Sc2ts ARG} \\
 \cmidrule(lr){1-5} \cmidrule(lr){6-7}
-Start & Length & Region & Count & Frequency & Count & Frequency \\
+Start & Length & Region & Count & Frequency (%) & Count & Frequency (%)\\
 \midrule
-28362 & 9 & N & 3939164 & 0.4305 & 871774 & 0.3512 \\
-28271 & 1 & Intergenic & 3853692 & 0.4212 & 1118363 & 0.4506 \\
-21765 & 6 & S & 3179887 & 0.3475 & 339681 & 0.1368 \\
-11288 & 9 & ORF1ab-nsp6 & 3040576 & 0.3323 & 290408 & 0.1170 \\
-28248 & 6 & ORF8 & 3026186 & 0.3307 & 1114669 & 0.4491 \\
-22029 & 6 & S & 3010665 & 0.3290 & 1112410 & 0.4482 \\
-11283 & 9 & ORF1ab-nsp6 & 2011934 & 0.2199 & 40 & 0.0000 \\
-6513 & 3 & ORF1ab-nsp3 & 1984683 & 0.2169 & 340765 & 0.1373 \\
-21633 & 9 & S & 1941962 & 0.2122 & 532072 & 0.2144 \\
-21987 & 9 & S & 1918094 & 0.2096 & N/A & N/A \\
-22194 & 3 & S & 1833050 & 0.2003 & 312706 & 0.1260 \\
-21991 & 3 & S & 912944 & 0.0998 & 288831 & 0.1164 \\
-686 & 9 & ORF1ab-nsp1 & 115450 & 0.0126 & N/A & N/A \\
+28362 & 9 & N & 3939164 & 43.0525 & 871774 & 35.1216 \\
+28271 & 1 & Intergenic & 3853692 & 42.1183 & 1118363 & 45.0561 \\
+21765 & 6 & S & 3179887 & 34.7541 & 339681 & 13.6849 \\
+11288 & 9 & ORF1ab-nsp6 & 3040576 & 33.2315 & 290408 & 11.6998 \\
+28248 & 6 & ORF8 & 3026186 & 33.0742 & 1114669 & 44.9073 \\
+22029 & 6 & S & 3010665 & 32.9046 & 1112410 & 44.8163 \\
+11283 & 9 & ORF1ab-nsp6 & 2011934 & 21.9891 & 40 & 0.0016 \\
+6513 & 3 & ORF1ab-nsp3 & 1984683 & 21.6913 & 340765 & 13.7286 \\
+21633 & 9 & S & 1941962 & 21.2244 & 532072 & 21.4359 \\
+21987 & 9 & S & 1918094 & 20.9635 & N/A & N/A \\
+22194 & 3 & S & 1833050 & 20.0340 & 312706 & 12.5982 \\
+21991 & 3 & S & 912944 & 9.9779 & 288831 & 11.6363 \\
+686 & 9 & ORF1ab-nsp1 & 115450 & 1.2618 & N/A & N/A \\
 \bottomrule
 \end{tabular}
-\end{center}
 \end{table}
@@
     \includegraphics[width=\linewidth]{mutational_spectra}
-    \caption{All-site mutational spectra of major VOCs.
-(A) Mutational spectra of the ARG.
-For each major VOC, the relative proportion of each mutation type was calculated from
-the mutations occurring above the nodes with a Scorpio label associated with the VOC
-(e.g. ``Alpha-like'' encompasses ``B.1.1.7-like'' and ``B.1.1.7-like+E484K'').
-Note that $\sim$98\% of all the mutations in the ARG occur above the nodes
-which have a Scorpio label associated with Alpha, Delta, Omicron BA.1, BA.2, BA.4, or BA.5
-(excluding the nodes with unassigned scorpio labels).
-The Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin.
-(B) Mutational spectra reproduced from the mutation count data analyzed in Bloom et al. (2023).
-The authors extracted mutation counts from a global UShER phylogeny
-(https://github.com/jbloomlab/SARS2-mut-spectrum/).
+    \caption{All-site mutational spectra of major VOCs calculated from the sc2ts ARG (A),
+    the mutation count data from Bloom et al. (2023) (B), and
+    the Viridian UShER phylogeny from Hunt et al. (2024)\cite{Hunt2024}.
     \label{fig:mutational_spectra}
@@
 ``copying path'' through the reference panel for a given focal sequence that
-maximises the likelihood under these parameters. Unseen states in the reference panel are shown as coloured lines enclosed by
+maximises the likelihood under these parameters. Unseen states in the reference panel are shown as colored lines enclosed by
 the grey box. The black arrow describes the true path through the data which leads to the emitted

### 2025-09-09 latex/main.tex
Fix latex compile

@@
     \centering
-    \includegraphics[width=1\linewidth]{fig_recombinant_evidence.pdf}
+    \includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
     \caption{Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes, colored according
@@
 
-\subsubsection{Mutational spectra}
+\subsubsection*{Mutational spectra}
 To validate the mutational information contained in the sc2ts ARG,
@@
 
-\subsubsection{Concordance among recombination detection methods}
+\subsubsection*{Concordance among recombination detection methods}
 For RecombinHunt, we obtained two sets of detection results. 
@@
 \label{tab:indels}
+\begin{center}
 \begin{tabular}{rrcrrrr}
-\toprule & \multicolumn{4}{c}{Li et al. (2023)} & \multicolumn{2}{c}{Sc2ts ARG} \\
+\toprule 
+& \multicolumn{4}{c}{Li et al. (2023)} & \multicolumn{2}{c}{Sc2ts ARG} \\
 \cmidrule(lr){1-5} \cmidrule(lr){6-7}
-Start & Length & Region & Count & Frequency (%) & Count & Frequency (%)\\
+Start & Length & Region & Count & Frequency (\%) & Count & Frequency (\%)\\
 \midrule

### 2025-09-09 latex/main.tex

--- 2025-09-09 latex/main.tex
@@
 
-\subsection{Major deletions in the ARG}\label{sec:supplement_dels}
+\subsection{Analysis of deletions}\label{sec:supplement_dels}
 % Referenced by the main Results.
-Li et al. \cite{Li2023} performed a large-scale analysis of indels 
+Li et al.\cite{Li2023} performed a large-scale analysis of indels 
 using a global GISAID dataset (sampled up to August, 2022). 
@@
 \begin{table}
-\caption{Major deletions in Li et al. (2023).}
+\caption{Major indels.}
 \label{tab:indels}
@@
 \cmidrule(lr){1-5} \cmidrule(lr){6-7}
-Start & Length & Region & Count & Frequency (%) & Count & Frequency (%)\\
+Start & Length & Region & Count & Frequency (\%) & Count & Frequency (\%)\\
 \midrule

### 2025-09-09 notebooks/nb_utils.py
Move layout files and delete old notebooks

@@ -298,7 +298,7 @@ class D3ARG_viz:
         exclude = set() if exclude is None else set(list(exclude))
         include = set() if include is None else set(list(include))
         if positions_file is None:
-            positions_file = f"subgraph_layouts/{'-'.join(pangos)}.json"
+            positions_file = f"layout_data/{'-'.join(pangos)}.json"
         try:
             self.d3arg.set_node_x_positions(
                 pos=argviz.extract_x_positions_from_json(

### 2025-09-09 latex/main.tex
Updates from Overleaf

@@
 We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
-(1) RecombinHunt \cite{Alfonsi2024},
-(2) CovRecomb \cite{Li2024}, and
-(3) rebar (https://github.com/phac-nml/rebar).
+RecombinHunt \cite{Alfonsi2024},
+CovRecomb \cite{Li2024}, and
+rebar (https://github.com/phac-nml/rebar).
 These methods use the same underlying approach:
@@
 \begin{table}
-\caption{
-Concordance among methods in characterizing Pango X lineages. 
-The Pango X lineages analyzed here are associated with the Type 1 recombination events (related to Table 1).
-For RecombinHunt and CovRecomb, the concordance rates are calculated 
-based on the Pango Xs for which results are available from the original studies.
-Also, for RecombinHunt, two sets of detection results are available, 
-one set obtained using mutational profiles computed from a global GISAID dataset and another from a Nextstrain dataset.}
+\caption{Concordance among methods in characterizing Pango X lineages.}
 \label{tab:method_concordance}
-\begin{tabular}{llll}
-\hline
-\textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bases)} \\ [1ex]
-\hline
-Sc2ts & 14 (93.3) & 13 (86.7) & 540.5 (315, 766) \\ [1ex]
-\hline
-RH-GISAID & 13 (100.0) & 10 (76.9) & 764.0 (4, 1202) \\ [1ex]
-\hline
-RH-Nextstrain & 12 (85.7) & 11 (78.6) & 1160.0 (559, 2785) \\ [1ex]
-\hline
-CovRecomb & 7 (87.5) & 4 (50.0) & 1562.5 (217, 13208) \\ [1ex]
-\hline
+\begin{tabular}{lccc}
+\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
+\cmidrule(lr){2-2} \cmidrule(lr){3-4}
+Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
+\midrule
+Sc2ts & 14 / 15 (93.3) & 13 / 15 (86.7) & 540.5 (315, 766)\\
+RecombinHunt-GISAID & 13 / 13 (100.0) & 10 / 13 (76.9) & 764.0 (4, 1202)\\
+RecombinHunt-Nextstrain & 12 / 14 (85.7) & 11 / 14 (78.6) & 1160.0 (559, 2785)\\
+CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
 \end{tabular}
@@
 For CovRecomb, we summarized the detection results for the Pango Xs as follows. 
-First, we obtained the independent recombination events 
-in Table S4 corresponding to a Pango X, if available. 
+First, we obtained the "independent" recombination events 
+corresponding to the Pango Xs (if available) 
+identified by Li et al.\cite{Li2024} (see Table S4 of the study).
 When multiple events corresponded to a Pango X, 

### 2025-09-09 latex/main.tex
Changes for concordance table

@@
 \centering
-\includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}
+\includegraphics[width=\linewidth]{pango_x_subgraphs}
 \caption{Illustrative subgraphs in the ARG. Subgraphs around the XA-associated
@@
 We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
-RecombinHunt \cite{Alfonsi2024},
-CovRecomb \cite{Li2024}, and
-rebar (https://github.com/phac-nml/rebar).
+RecombinHunt\cite{Alfonsi2024}, CovRecomb\cite{Li2024}, and rebar (https://github.com/phac-nml/rebar).
 These methods use the same underlying approach:
@@
 identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
-we focus on the Pango Xs associated with the Type I recombination events (Table \ref{tab:pango_x_lineages}), 
+we focus on the Pango Xs associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
 using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
-For RecombinHunt, we took detection results for these Pango Xs reported in the original study \cite{Alfonsi2024};
-for CovRecomb, we summarized the results from the original study \cite{Li2024} (STAR Methods).
+For RecombinHunt and CovRecomb, we summarized detection results for these Pango Xs, if available,
+reported in the original studies\cite{Alfonsi2024,Li2024} (STAR Methods).
 For each method, we determined the Pango Xs for which the inferred parental lineages and breakpoint intervals
@@
 We found that the methods performed well overall,
-with comparable concordance rates (Table \ref{tab:x_method_concordance}),
+with comparable concordance rates (Table~\ref{tab:method_concordance}),
 indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.
@@
 \begin{table}
-\caption{Concordance among methods in characterizing Pango X lineages.}
+\caption{
+Concordance among methods in characterizing Pango X lineages.
+Only the Pango X lineages associated with the Type 1 recombination events
+are considered in these comparisons.
+For each method, we calculated the number of Pango X lineages
+which have inferred parent Pango lineages and breakpoint intervals
+concordant with those proposed by the community.
+The denominators indicate the number of Pango X lineages
+for which detection results are available from the original studies.
+For the Pango X lineages with non-overlapping breakpoint intervals,
+the distances between the inferred breakpoint intervals and 
+those proposed by the community are shown (median and range).
+Related to Table \ref{tab:pango_x_lineages}.
+}
 \label{tab:method_concordance}
@@
 CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
+\bottomrule
 \end{tabular}

### 2025-09-09 latex/main.tex

@@
 First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
-(allowing recombination) using the Li and Stephens (LS) model\cite{Li2003},
+(allowing recombination) using the Li and Stephens (LS) model\cite{LiStephens2003},
 which is a Hidden Markov Model (HMM) widely used in human genomics
@@
 We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
-RecombinHunt\cite{Alfonsi2024}, CovRecomb\cite{Li2024}, and rebar (https://github.com/phac-nml/rebar).
+RecombinHunt\cite{Alfonsi2024}, CovRecomb\cite{Li2024CovRecomb}, and rebar (https://github.com/phac-nml/rebar).
 These methods use the same underlying approach:
@@
 For RecombinHunt and CovRecomb, we summarized detection results for these Pango Xs, if available,
-reported in the original studies\cite{Alfonsi2024,Li2024} (STAR Methods).
+reported in the original studies\cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
 For each method, we determined the Pango Xs for which the inferred parental lineages and breakpoint intervals
@@
 Concordance among methods in characterizing Pango X lineages.
-Only the Pango X lineages associated with the Type 1 recombination events
-are considered in these comparisons.
+These comparisons focus on the Pango X lineages associated with
+the Type 1 recombination events (Table \ref{tab:pango_x_lineages})
 For each method, we calculated the number of Pango X lineages
@@
 The denominators indicate the number of Pango X lineages
-for which detection results are available from the original studies.
+for which detection results are available from the original studies\cite{Alfonsi2024,Li2024CovRecomb}.
 For the Pango X lineages with non-overlapping breakpoint intervals,
@@
 those proposed by the community are shown (median and range).
-Related to Table \ref{tab:pango_x_lineages}.
 }
@@
 consistent with patterns observed in previous 
-studies\cite{Turakhia2022,Li2024}.
+studies\cite{Turakhia2022,Li2024CovRecomb}.
 While this might result from a higher recombination rate in this region,
@@
 and use days as the time interval for clarity in what follows.
-The first step is to find likely ``copying paths'' under the Li and Stephens (LS) model \cite{Li2003}
+The first step is to find likely ``copying paths'' under the Li and Stephens (LS) model \cite{LiStephens2003}
 for each sequence in the daily batch to the current ARG.
@@
 
-LS model \cite{Li2003} is an approximation of the coalescent with recombination
+LS model \cite{LiStephens2003} is an approximation of the coalescent with recombination
 that captures many of the key features of the joint processes of mutation and recombination.
@@
 
-In the original formulation of the LS model (see Appendix S1 of \cite{Li2003}),
+In the original formulation of the LS model (see Appendix S1 of \cite{LiStephens2003}),
 the probability of switching between the hidden states (i.e., genomes in a reference panel)
@@
 corresponding to the Pango Xs (if available) 
-identified by Li et al.\cite{Li2024} (see Table S4 of the study).
+identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study).
 When multiple events corresponded to a Pango X, 
@@
 % Referenced by the main Results.
-Li et al.\cite{Li2023} performed a large-scale analysis of indels 
+Li et al.\cite{LiStephens2003} performed a large-scale analysis of indels 
 using a global GISAID dataset (sampled up to August, 2022). 

### 2025-09-09 latex/main.tex
Updates from Overleaf

@@
 Below, we summarize our analysis of the Pango Xs in the ARG
-(for extended analyses, see Document~\ref{sec:supplement_pango_lineages} 
-and Document S2).
+(for extended analyses, see Document~\ref{sec:supplement_pango_lineages} and Document S2).
 
@@
 % (see their copying patterns in Figure SX). 
-Among these Pango Xs are XA, which
-is the first designated Pango X that showed signs of community transmission
-\cite{Jackson2021}, and XBB, which escaped vaccine-induced immunity and gained
-global prevalence \cite{Tamura2023}. The recombinant origin of XA is well
-explained in the ARG by two parents and one mutation
-(Figure~\ref{fig:subgraphs}A;
-Table~\ref{tab:pango_x_lineages}), 
-with sc2ts correctly identifying its parental Pango lineages and
-breakpoint location.
+Among these Pango Xs are XA,
+which is the first designated Pango X that showed signs of onward transmission\cite{Jackson2021}, and
+XBB, which escaped vaccine-induced immunity and gained global prevalence\cite{Tamura2023}.
+The recombinant origin of XA is well explained in the ARG by two parents and one mutation
+(Figure~\ref{fig:subgraphs}A; Table~\ref{tab:pango_x_lineages}), 
+with sc2ts correctly identifying its parental Pango lineages and breakpoint location.
 
@@
 \includegraphics[width=\linewidth]{pango_x_subgraphs}
-\caption{Illustrative subgraphs in the ARG. Subgraphs around the XA-associated
+\caption{
+Illustrative subgraphs in the ARG. Subgraphs around the XA-associated
 recombination node (A); the recombination node associated with a group of
 ``nested'' Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C).
-For more detailed visualizations of these subgraphs, see Document S2.}
+For more detailed visualizations of these subgraphs, see Document S2.
+}
 \label{fig:subgraphs}
@@
 \subsection*{Quantification of evidence for recombination}
+
 %%% SHZ: We may not need this section at all, but I'm keeping it for now.
@@
 \subsection*{Recombination breakpoint intervals along the genome}
+
 We explored the spatial distribution of the breakpoint locations of
@@
 \subsection*{Efficient inference and analysis of a pandemic-scale ARG}
+
 The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
@@
 
-\subsubsection*{Comparing phylogenetic backbones}
+\subsubsection*{Comparison of phylogenetic backbones}
 
@@
 
-\subsubsection*{Mutational spectra}
+\subsubsection*{Comparison of mutational spectra}
 To validate the mutational information contained in the sc2ts ARG,
 we compared the all-site mutational spectra of major VOCs in the ARG,
-a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024}, and
-a global UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
-Mutational spectra are relative proportions of each type of nucleotide mutation.
+a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024},
+and another UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
+Mutational spectra are defined as the relative proportions of each possible type of nucleotide mutation.
 
@@
 in the sc2ts ARG, considering all the mutations above the sample nodes and internal nodes
-which have Scorpio labels associated with the VOCs (e.g., B.1.1.7-like and B.1.1.7-like+E484K for Alpha).
+which have Scorpio labels associated with the VOCs
+(e.g., for Alpha, B.1.1.7-like and B.1.1.7-like+E484K).
 These Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin.
 Note that $\sim$98\% of all the mutations in the ARG occur
-above the nodes which have a Scorpio label associated with Alpha, Delta, BA.1, BA.2, BA.4, or BA.5
+above the nodes that have a Scorpio label associated with these VOCs
 (excluding the nodes with unassigned Scorpio labels).
@@
 which was built using Nextclade and has samples collected globally up to November, 2022.
-We used the following mapping between the WHO labels and the Nextstrain labels:
+We used the following mapping between the WHO labels and the Nextstrain clade labels:
 Alpha, 20I; Detal, 21I and 21J; BA.1, 21K; BA.2, 21L; BA.4, 22A, and BA.5, 22B.
-The mutation count data are available at https://github.com/jbloomlab/SARS2-mut-spectrum/.
+The mutation counts were downloaded from https://github.com/jbloomlab/SARS2-mut-spectrum/.
 
@@
 
-\subsubsection*{Concordance among recombination detection methods}
-For RecombinHunt, we obtained two sets of detection results. 
-One set is based on mutational profiles calculated from a global GISAID dataset, 
-and the other set from a global Nextstrain dataset. 
-We adjusted the RecombinHunt-identified breakpoint interval 
-so that they are right-exclusive by adding one to the right coordinate.
-
-For CovRecomb, we summarized the detection results for the Pango Xs as follows. 
+\subsubsection*{Comparison with recombination detection methods}
+For RecombinHunt, we obtained two sets of detection results of the Pango X lineages
+from the supplementary materials of Alfonso et al.\cite{Alfonsi2024}.
+One set of results is based on mutational profiles calculated from a global GISAID dataset,
+and the other set from a global NextStrain dataset.
+The breakpoint intervals identified using RecombinHunt are defined
+as the left and right recombination-informative sites.
+We adjusted their coordinates by adding one to the right coordinate
+so that the intervals are right-exclusive.
+
+For CovRecomb, we summarized the detection results for the Pango X lineages as follows.
 First, we obtained the "independent" recombination events
-corresponding to the Pango Xs (if available) 
+which correspond to the Pango X lineages (if available)
 identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study).
-When multiple events corresponded to a Pango X, 
-we took the results of the event having the most epidemic recombinants. 
-Then, for other Pango Xs, we collated the data from a file (URL) 
-containing all the detection results produced by the study. 
+When multiple events corresponded to a Pango X lineage,
+we took the results of the event which had the highest number of epidemic recombinants.
+For other Pango X lineages, we downloaded the full list of putative recombinants
+identified in the study
+(https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv)
 For each of these Pango Xs,
 we took the most commonly observed combination of
-left parent, right parent, and breakpoint interval.
-We adjusted the CovRecomb-identified breakpoint interval 
-so that they are right-exclusive by adding one to the right coordinate.
+the left parent, the right parent, and the breakpoint interval.
+The breakpoint intervals are defined as the left and right recombination-informative sites.
+We adjusted the breakpoint intervals by adding one to the right coordinate
+so that they are right-exclusive.
 
@@
 Rebar identifies a sequence as a recombinant if it contains sets of mutations 
-predetermined to be associated with different Pango lineages.
-
-The parents of a Pango X inferred using a method are considered concordant 
-if their Pango lineage assignments are at least as specific as the community-identified parents.
-Also, a breakpoint interval of a Pango X inferred using a method is considered concordant 
-if it overlaps with the community-identified breakpoint interval by at least one base.
-We adjusted the breakpoint intervals from the community 
-by subtract one from the left coordinate and adding two to the right coordinate.
-
-\subsubsection*{Calculating Pango lineage distance using pangonet}
+which are pre-determined to be associated with different Pango lineages.
+
+The parents of a Pango X inferred using a method are deemed concordant
+if they have Pango lineage assignments
+that are at least as specific as those proposed by the community.
+The breakpoint interval of a Pango X inferred using a method is deemed concordant
+if it overlaps with that proposed by the community by at least one base.
+These breakpoint intervals are defined as ranges of possible site positions
+where a breakpoint is inferred to occur.
+To make these breakpoint intervals right-exclusive,
+we adjusted them by subtracting one from the left coordinate and adding two to the right coordinate.
+
+\subsubsection*{Calculation of Pango lineage distance using pangonet}
 We define a ``pseudo-phylogenetic'' distance to quantify the divergence between two Pango lineages.
@@
 \subsection{Analysis of the Jackson et al. recombinants}
-Jackson et al.\cite{Jackson2021} conducted detailed analyses of some early recombinants
+
+Jackson et al.\cite{Jackson2021} conducted detailed analyses of early recombinants
 involving B.1.1.7 (Alpha), including the first designated Pango X lineage, XA.
 The authors first performed a targeted search for candidate recombinants by
-scanning sampled sequences for sequence motifs associated with Alpha, followed
-by an analysis to reduce the list of plausible recombinants displaying signs of
-onward transmission. Here we compared our sc2ts detection results of these
-recombinants with the detection results reported by Jackson et al.
+scanning sampled sequences for sequence motifs associated with Alpha,
+followed by an analysis to reduce the list of plausible recombinants
+displaying signs of onward transmission.
+Here we compared the sc2ts detection results of these recombinants
+with the detection results reported by Jackson et al.
 
@@
 are directly associated with strongly supported recombination events 
-in the ARG in terms of the number of averted mutations 
-(A: 16, B: 6, C: 30, D: 18).
+in the ARG in terms of the number of averted mutations (A: 16, B: 6, C: 30, D: 18).
 The singleton CAMC-CBA018 has 
@@
     \includegraphics[width=\linewidth]{mutational_spectra}
-    \caption{All-site mutational spectra of major VOCs calculated from the sc2ts ARG (A),
-    the mutation count data from Bloom et al. (2023) (B), and
+    \caption{
+    All-site mutational spectra of major VOCs calculated from the sc2ts ARG (A),
+    the mutation count data from Bloom et al. (2023)\cite{Bloom2023} (B), and
     the Viridian UShER phylogeny from Hunt et al. (2024)\cite{Hunt2024}.

### 2025-09-09 latex/main.tex
Rename XAE+ to Xx

@@
 \midrule
-XAE+ & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 &
+Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 &
 {BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
@@
 XAG & XQ & 6 & 55 & 4 & 6 \\
-XAP & XAE+ & 4 & 67 & 1 & 20 \\
-XZ & XAE+ & 4 & 48 & 1 & 48 \\
-XAE & XAE+ & 2 & 47 & 7 & 9 \\
+XAP & Xx & 4 & 67 & 1 & 20 \\
+XZ & Xx & 4 & 48 & 1 & 48 \\
+XAE & Xx & 2 & 47 & 7 & 9 \\
 XAF & XAF & 5 & 140 & 8 & 1 \\

### 2025-09-09 latex/main.tex
Updates from Overleaf

@@
 using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
-For RecombinHunt and CovRecomb, we summarized detection results for these Pango Xs, if available,
+For RecombinHunt and CovRecomb, we summarized the detection results for these Pango X lineages (if available)
 reported in the original studies\cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
-For each method, we determined the Pango Xs for which the inferred parental lineages and breakpoint intervals
+For each method, we determined the Pango X lineages
+for which the inferred parental lineages and breakpoint intervals
 are concordant with those identified by the community (STAR Methods).
@@
 versus an alternative copying path disallowing recombination ($x$: top axis);
-recombination nodes in Table \ref{tab:pango_x_lineages} are labelled starting with `X' and
+recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
 those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
@@
 \subsection{Analysis of deletions}\label{sec:supplement_dels}
+
+% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
 % Referenced by the main Results.
-Li et al.\cite{LiStephens2003} performed a large-scale analysis of indels 
+We found that the ARG captures highly prevalent deletions in SARS-CoV-2 (Table~\ref{tab:major_dels}).
+Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
 using a global GISAID dataset (sampled up to August, 2022).
 We examined the major indels reported in Table S1 of that study
-which have a minimum frequency of 1\% 
-(sequence count divided by 9,149,680 filtered sequences).
+which have a minimum frequency of 1\% (sequence count divided by 9,149,680 filtered sequences).
 Most of these indels (all deletions) are encoded in the ARG (Table~\ref{tab:indels})
-and have high frequencies in the ARG
-(sequence count divided by the total number of samples), 
+and have high frequencies in the ARG (sequence count divided by the total number of samples),
 except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
 During site remapping, the 9-base deletion at 11283 was split up into
-a 5-base deletion at 11283 (count of 339867) and 
-a 4-base deletion at 11288 (count of 871782).
-The 686 deletion occurs below our threshold for inclusion into the ARG.
-
-% See https://github.com/jeromekelleher/sc2ts-paper/issues/472
+a 5-base deletion at 11283 (count of 339867) and a 4-base deletion at 11288 (count of 871782).
+
 \begin{table}
-\caption{Major indels.}
-\label{tab:indels}
-\begin{center}
-\begin{tabular}{rrcrrrr}
-\toprule 
-& \multicolumn{4}{c}{Li et al. (2023)} & \multicolumn{2}{c}{Sc2ts ARG} \\
-\cmidrule(lr){1-5} \cmidrule(lr){6-7}
-Start & Length & Region & Count & Frequency (\%) & Count & Frequency (\%)\\
+\caption{
+Major deletion events in the ARG.
+These events are observed in at least 10,000 samples encoded in the ARG.
+The counts and frequencies of these events from Li et al.\cite{Li2023indels} are also shown.
+}
+\label{tab:major_dels}
+\begin{tabular}{rrcrcrrrr}
+\centering
+\toprule & \multicolumn{6}{c}{Sc2ts ARG} & \multicolumn{2}{c}{Li et al. (2023)} \\
+\cmidrule(lr){1-7} \cmidrule(lr){8-9}
+Start & Length & Region & Node & Pango & Count & Frequency (\%) & Count & Frequency (\%)\\
 \midrule
-28362 & 9 & N & 3939164 & 43.0525 & 871774 & 35.1216 \\
-28271 & 1 & Intergenic & 3853692 & 42.1183 & 1118363 & 45.0561 \\
-21765 & 6 & S & 3179887 & 34.7541 & 339681 & 13.6849 \\
-11288 & 9 & ORF1ab-nsp6 & 3040576 & 33.2315 & 290408 & 11.6998 \\
-28248 & 6 & ORF8 & 3026186 & 33.0742 & 1114669 & 44.9073 \\
-22029 & 6 & S & 3010665 & 32.9046 & 1112410 & 44.8163 \\
-11283 & 9 & ORF1ab-nsp6 & 2011934 & 21.9891 & 40 & 0.0016 \\
-6513 & 3 & ORF1ab-nsp3 & 1984683 & 21.6913 & 340765 & 13.7286 \\
-21633 & 9 & S & 1941962 & 21.2244 & 532072 & 21.4359 \\
-21987 & 9 & S & 1918094 & 20.9635 & N/A & N/A \\
-22194 & 3 & S & 1833050 & 20.0340 & 312706 & 12.5982 \\
-21991 & 3 & S & 912944 & 9.9779 & 288831 & 11.6363 \\
-686 & 9 & ORF1ab-nsp1 & 115450 & 1.2618 & N/A & N/A \\
+28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 & 3853692 & 42.1183 \\
+28248 & 6 & ORF8 & 220186 & B.1.617.2 & 1114669 & 44.9073 & 3026186 & 33.0742 \\
+22029 & 6 & S / S1 - NTD & 200039 & B.1.617.2 & 1112410 & 44.8163 & 3010665 & 32.9046 \\
+11288 & 4 & ORF1AB / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 & N/A & N/A \\
+28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 & 3939164 & 43.0525 \\
+21633 & 9 & S / S1 - NTD & 822854 & BA.2 & 532072 & 21.4359 & 1941962 & 21.2244 \\
+11292 & 5 & ORF1AB / NSP6 & 822854 & BA.2 & 531894 & 21.4287 & N/A & N/A \\
+6513 & 3 & ORF1AB / NSP3 & 851246 & BA.1 & 340765 & 13.7286 & 1984683 & 21.6913 \\
+11283 & 5 & ORF1AB / NSP6 & 851246 & BA.1 & 339867 & 13.6924 & 8360 & 0.0914 \\
+21765 & 6 & S / S1 - NTD & 851246 & BA.1 & 339681 & 13.6849 & 3179887 & 34.7541 \\
+21988 & 8 & S / S1 - NTD & 851246 & BA.1 & 339651 & 13.6837 & 590 & 0.0064 \\
+22194 & 3 & S / S1 - NTD & 851246 & BA.1 & 312706 & 12.5982 & 1833050 & 20.0340 \\
+28271 & 1 & non-coding & 86456 & B.1.1.7 & 290446 & 11.7014 & 3853692 & 42.1183 \\
+11288 & 9 & ORF1AB / NSP6 & 86456 & B.1.1.7 & 290408 & 11.6998 & 3040576 & 33.2315 \\
+21991 & 3 & S / S1 - NTD & 86456 & B.1.1.7 & 288831 & 11.6363 & 912944 & 9.9779 \\
+21765 & 6 & S / S1 - NTD & 86456 & B.1.1.7 & 290438 & 11.7010 & 3179887 & 34.7541 \\
+21765 & 6 & S / S1 - NTD & 1265302 & BA.4 & 190425 & 7.6718 & 3179887 & 34.7541 \\
+28361 & 1 & N & 1246012 & BA.5.2.1 & 20822 & 0.8389 & 287 & 0.0031 \\
 \bottomrule
 \end{tabular}
-\end{center}
+\end{table}
+
+\begin{table}
+\caption{
+Deletion events associated with the Alpha variant in the ARG.
+These events occur above node 86456, which represents the origin of Alpha.
+}
+\label{tab:alpha_dels}
+\centering
+\begin{tabular}{rrrrr}
+\toprule
+Start & Length & Region & Count & Frequency (\%)\\
+\midrule
+28271 & 1 & non-coding & 290446 & 11.7014 \\
+11288 & 9 & ORF1AB / NSP6 & 290408 & 11.6998 \\
+21991 & 3 & S / S1 - NTD & 288831 & 11.6363 \\
+21765 & 6 & S / S1 - NTD & 290438 & 11.7010 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}
+\caption{
+Highly recurrent deletions.
+}
+\label{tab:rec_dels}
+\centering
+\begin{tabular}{rrcrr}
+\toprule
+Start & Length & Region & Occurrences \\
+\midrule
+29781 & 1 & non-coding & 12713 \\
+29762 & 1 & non-coding & 11356 \\
+29779 & 1 & non-coding & 10925 \\
+29769 & 1 & non-coding & 10783 \\
+29774 & 1 & non-coding & 10707 \\
+29700 & 1 & non-coding & 8543 \\
+29614 & 1 & ORF10 & 5041 \\
+29555 & 1 & non-coding & 4816 \\
+29543 & 1 & non-coding & 4806 \\
+29541 & 1 & non-coding & 4804 \\
+203 & 1 & non-coding & 3602 \\
+222 & 1 & non-coding & 3441 \\
+22013 & 6 & S / S1 - NTD & 2111 \\
+21991 & 3 & S / S1 - NTD & 2087 \\
+28254 & 1 & ORF8 & 1967 \\
+27556 & 64 & ORF7a & 1281 \\
+27579 & 3 & ORF7a & 1014 \\
+28093 & 1 & ORF8 & 898 \\
+21765 & 6 & S / S1 - NTD & 478 \\
+27532 & 1 & ORF7a & 403 \\
+\bottomrule
+\end{tabular}
 \end{table}

### 2025-09-09 notebooks/nb_utils.py
Use standard labels in Fig 3

@@ -7,6 +7,7 @@ import fileinput
 from pathlib import Path
 
 import msprime  # only for the NODE_IS_RE_EVENT flag
+import pandas as pd
 import sc2ts
 import tskit
 import tszip
@@ -32,6 +33,77 @@ def load(filename="sc2ts_viridian_v1.1.trees.tsz"):
     )
     return ts
 
+def standard_recombinant_labels(ts, pango_x_events_file):
+    """
+    Return a standard set of labels for "known" recombination nodes
+    (Pango Xs plus hord-coded Jackson recombinants). The
+    pango_x_events_file should be a path to pango_x_events.csv
+    """
+    df = sc2ts.node_data(ts).set_index("sample_id")
+    pango_x_events = pd.read_csv(pango_x_events_file)
+    tree = ts.first()  # Assume the first tree gives  areasonable number of descendant samples for a RE node
+    # First keep all the standard recombinants
+    use = pango_x_events.root_type == "R"
+    # Cases where one PangoX has multiple RE nodes (only XM), use the one with the most descendants
+    for p in np.unique(pango_x_events[use].root_pango):
+        match = pango_x_events.root_pango == p
+        if sum(match) > 1:
+            use[match] = False
+            sorted_indexes = sorted(
+                np.where(match)[0],
+                key=lambda i: tree.num_samples(pango_x_events.closest_recombinant[i])
+            )
+            use[sorted_indexes[-1]] = True
+    
+    labels = {row.closest_recombinant: row.root_pango for row in pango_x_events[use].itertuples()}
+    assert len(labels) == np.sum(use)
+    
+    labels = {row.closest_recombinant: row.root_pango for row in pango_x_events[use].itertuples()}
+    for re_node, rows in pango_x_events[pango_x_events.root_type != "R"].groupby("closest_recombinant"):
+        if re_node >=0:
+            # exclude those with > 100,000 descendants (to exclude BA.5 as a RE node)
+            if re_node not in labels and tree.num_samples(re_node) < 1e5:
+                if len(rows) == 1:
+                    labels[re_node] = rows.iloc[0].root_pango
+                else:
+                    labels[re_node] = "/".join(rows.root_pango.values)
+    
[…]

### 2025-09-09 notebooks/nb_utils.py
Add standard RE labels for robust_vs_artificial_classification.pdf
and skip BQ.1.21

See https://github.com/jeromekelleher/sc2ts-paper/issues/551#issuecomment-3273485995

@@ -77,10 +77,10 @@ def standard_recombinant_labels(ts, pango_x_events_file):
     
     # Add particular pangos that are not X but are recombinants
     
-    for re_pango in ["BQ.1.21"]:
-        potential_re_node = ts.first().mrca(*df.loc[df.pango == re_pango, "node_id"])
-        assert ts.node(potential_re_node).flags & sc2ts.NODE_IS_RECOMBINANT
-        labels[potential_re_node] = re_pango
+    #for re_pango in ["BQ.1.21"]:  # BQ.1.21 is not robust, so skip it
+    #    potential_re_node = ts.first().mrca(*df.loc[df.pango == re_pango, "node_id"])
+    #    assert ts.node(potential_re_node).flags & sc2ts.NODE_IS_RECOMBINANT
+    #    labels[potential_re_node] = re_pango
     
     # Add the Jackson recombinants
     XA = [k for k, v in labels.items() if v == "XA"][0]

### 2025-09-10 latex/main.tex

@@
 \caption{
-Illustrative subgraphs in the ARG. Subgraphs around the XA-associated
-recombination node (A); the recombination node associated with a group of
-``nested'' Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C).
+Illustrative subgraphs in the ARG.
+(A) Subgraph around the recombination event associated with the XA.
+The "Group A" sequences identified by Jackson et al.\cite{Jackson2021}) are attached to this event.
+(B) Subgraph around the recombination event associated with a group of ``nested'' Pango Xs; and
+(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
 For more detailed visualizations of these subgraphs, see Document S2.
@@
 \begin{table}
-\caption{ARG events associated with Pango X lineages.
+\caption{
+ARG events associated with Pango X lineages.
 Each row corresponds to the oldest node assigned the specified
@@
 (except XBB, XBB.1 and XM; see STAR methods). [MORE DETAILS]
-\label{tab:pango_x_lineages}}
+}
+\label{tab:pango_x_lineages}
 \begin{footnotesize}
@@
 identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
-we focus on the Pango Xs associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
+we focus on the Pango X lineages associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
 using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
@@
 and to get established in the ARG.
-For the full list of seeds,
-see
+For the full list of seeds, see
 \url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}
@@
 \subsubsection*{Identification of robust recombination events}
+
 [TODO This section needs a full pass. I've merged in a section that was in the
@@
 \subsubsection*{Comparison of mutational spectra}
+
 To validate the mutational information contained in the sc2ts ARG,
@@
 and another UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
-Mutational spectra are defined as the relative proportions of each possible type of nucleotide mutation.
+Mutational spectra are defined as the relative proportions of
+each possible type of nucleotide change.
 
 First, we calculated the mutational spectra of Alpha, Delta, Omicron BA.1, BA.2, BA.4, and BA.5
-in the sc2ts ARG, considering all the mutations above the sample nodes and internal nodes
+in the sc2ts ARG, considering mutations (ignoring indels) above the sample nodes and internal nodes
 which have Scorpio labels associated with the VOCs
@@
 using the mutational counts from an UShER phylogeny analyzed in Bloom et al.\cite{Bloom2023},
-which was built using Nextclade and has samples collected globally up to November, 2022.
-We used the following mapping between the WHO labels and the Nextstrain clade labels:
-Alpha, 20I; Detal, 21I and 21J; BA.1, 21K; BA.2, 21L; BA.4, 22A, and BA.5, 22B.
+which was built using a Nextclade pipeline and
+has samples collected worldwide up to November, 2022.
+To obtain the WHO label corresponding to a Nextclade-defined clade, we used the following mapping:
+20I, Alpha; 21I and 21J, Delta; 21K, BA.1; 21L, BA.2; 22A, BA.4; and 22B, BA.5.
 The mutation counts were downloaded from https://github.com/jbloomlab/SARS2-mut-spectrum/.
@@
 which was pruned down to the samples present in the sc2ts ARG.
-We used the mutations above the samples based on their Scorpio labels,
-which are available in the Viridian metadata provided by Hunt et al.\cite{Hunt2024}.
+We used mutations above the sample nodes,
+for which Scorpio labels are available
+in the Viridian metadata provided by Hunt et al.\cite{Hunt2024}.
 
 \subsubsection*{Comparison with recombination detection methods}
+
 For RecombinHunt, we obtained two sets of detection results of the Pango X lineages
@@
 First, we obtained the "independent" recombination events
-which correspond to the Pango X lineages (if available)
-identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study).
+identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study)
+which correspond to the Pango X lineages (if available).
 When multiple events corresponded to a Pango X lineage,
@@
 % Referenced by the main Results.
-We found that the ARG captures highly prevalent deletions in SARS-CoV-2 (Table~\ref{tab:major_dels}).
+The ARG captures highly prevalent deletions in SARS-CoV-2 (Table~\ref{tab:major_dels}).
 Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
 using a global GISAID dataset (sampled up to August, 2022).
-We examined the major indels reported in Table S1 of that study
-which have a minimum frequency of 1\% (sequence count divided by 9,149,680 filtered sequences).
-Most of these indels (all deletions) are encoded in the ARG (Table~\ref{tab:indels})
-and have high frequencies in the ARG (sequence count divided by the total number of samples),
+We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
+which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
+All of these indels are deletions and have high frequencies in the ARG
+(sequence count divided by the total number of samples),
 except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
-During site remapping, the 9-base deletion at 11283 was split up into
+During site remapping, the 9-base deletion at 11283 was split into
 a 5-base deletion at 11283 (count of 339867) and a 4-base deletion at 11288 (count of 871782).
 
+<TODO> Highly recurrent deletions (Table~\ref{tab:rec_dels}).
+
 \begin{table}
+\centering
 \caption{
 Major deletion events in the ARG.
-These events are observed in at least 10,000 samples encoded in the ARG.
-The counts and frequencies of these events from Li et al.\cite{Li2023indels} are also shown.
+These events are observed in more than 10,000 samples in the ARG.
 }
 \label{tab:major_dels}
-\begin{tabular}{rrcrcrrrr}
-\centering
-\toprule & \multicolumn{6}{c}{Sc2ts ARG} & \multicolumn{2}{c}{Li et al. (2023)} \\
-\cmidrule(lr){1-7} \cmidrule(lr){8-9}
-Start & Length & Region & Node & Pango & Count & Frequency (\%) & Count & Frequency (\%)\\
+\begin{tabular}{rrcrcrr}
+\toprule
+Start & Length & Region & Node & Pango & Count & Frequency (\%)\\
 \midrule
-28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 & 3853692 & 42.1183 \\
-28248 & 6 & ORF8 & 220186 & B.1.617.2 & 1114669 & 44.9073 & 3026186 & 33.0742 \\
-22029 & 6 & S / S1 - NTD & 200039 & B.1.617.2 & 1112410 & 44.8163 & 3010665 & 32.9046 \\
-11288 & 4 & ORF1AB / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 & N/A & N/A \\
-28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 & 3939164 & 43.0525 \\
-21633 & 9 & S / S1 - NTD & 822854 & BA.2 & 532072 & 21.4359 & 1941962 & 21.2244 \\
-11292 & 5 & ORF1AB / NSP6 & 822854 & BA.2 & 531894 & 21.4287 & N/A & N/A \\
-6513 & 3 & ORF1AB / NSP3 & 851246 & BA.1 & 340765 & 13.7286 & 1984683 & 21.6913 \\
-11283 & 5 & ORF1AB / NSP6 & 851246 & BA.1 & 339867 & 13.6924 & 8360 & 0.0914 \\
-21765 & 6 & S / S1 - NTD & 851246 & BA.1 & 339681 & 13.6849 & 3179887 & 34.7541 \\
-21988 & 8 & S / S1 - NTD & 851246 & BA.1 & 339651 & 13.6837 & 590 & 0.0064 \\
-22194 & 3 & S / S1 - NTD & 851246 & BA.1 & 312706 & 12.5982 & 1833050 & 20.0340 \\
-28271 & 1 & non-coding & 86456 & B.1.1.7 & 290446 & 11.7014 & 3853692 & 42.1183 \\
-11288 & 9 & ORF1AB / NSP6 & 86456 & B.1.1.7 & 290408 & 11.6998 & 3040576 & 33.2315 \\
-21991 & 3 & S / S1 - NTD & 86456 & B.1.1.7 & 288831 & 11.6363 & 912944 & 9.9779 \\
-21765 & 6 & S / S1 - NTD & 86456 & B.1.1.7 & 290438 & 11.7010 & 3179887 & 34.7541 \\
-21765 & 6 & S / S1 - NTD & 1265302 & BA.4 & 190425 & 7.6718 & 3179887 & 34.7541 \\
-28361 & 1 & N & 1246012 & BA.5.2.1 & 20822 & 0.8389 & 287 & 0.0031 \\
+28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
+28248 & 6 & ORF8 & 220186 & B.1.617.2 & 1114669 & 44.9073 \\
+22029 & 6 & S / S1 - NTD & 200039 & B.1.617.2 & 1112410 & 44.8163 \\
+11288 & 4 & ORF1AB / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
+28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 \\
+21633 & 9 & S / S1 - NTD & 822854 & BA.2 & 532072 & 21.4359 \\
+11292 & 5 & ORF1AB / NSP6 & 822854 & BA.2 & 531894 & 21.4287 \\
+6513 & 3 & ORF1AB / NSP3 & 851246 & BA.1 & 340765 & 13.7286 \\
+11283 & 5 & ORF1AB / NSP6 & 851246 & BA.1 & 339867 & 13.6924 \\
+21765 & 6 & S / S1 - NTD & 851246 & BA.1 & 339681 & 13.6849 \\
+21988 & 8 & S / S1 - NTD & 851246 & BA.1 & 339651 & 13.6837 \\
+22194 & 3 & S / S1 - NTD & 851246 & BA.1 & 312706 & 12.5982 \\
+28271 & 1 & non-coding & 86456 & B.1.1.7 & 290446 & 11.7014 \\
+11288 & 9 & ORF1AB / NSP6 & 86456 & B.1.1.7 & 290408 & 11.6998 \\
+21991 & 3 & S / S1 - NTD & 86456 & B.1.1.7 & 288831 & 11.6363 \\
+21765 & 6 & S / S1 - NTD & 86456 & B.1.1.7 & 290438 & 11.7010 \\
+21765 & 6 & S / S1 - NTD & 1265302 & BA.4 & 190425 & 7.6718 \\
+28361 & 1 & N & 1246012 & BA.5.2.1 & 20822 & 0.8389 \\
 \bottomrule
@@
 \begin{table}
+\centering
 \caption{
@@
 \label{tab:alpha_dels}
-\centering
-\begin{tabular}{rrrrr}
+\begin{tabular}{rrcrr}
 \toprule
-Start & Length & Region & Count & Frequency (\%)\\
+Start & Length & Region & Count & Frequency (\%) \\
 \midrule
@@
 \begin{table}
-\caption{
-Highly recurrent deletions.
-}
+\centering
+\caption{Highly recurrent deletions.}
 \label{tab:rec_dels}
-\centering
-\begin{tabular}{rrcrr}
+\begin{tabular}{rrcr}
 \toprule
@@
 \label{sec:supplement_pango_lineages}
+
 To analyze how well the ARG reflects the phylogenetic structure implicit in the
@@
 
-\begin{table} \centering
-\caption{\label{tab:jackson}Comparison of recombination breakpoint
+\begin{table}
+\centering
+\caption{
+Comparison of recombination breakpoint
 intervals and parent lineages for Groups A-D
@@
 to correspond to the \texttt{tskit} definition of inheritance on either side of a breakpoint,
-and add one to the right coordinate to make the intervals right-exclusive. }
+and add one to the right coordinate to make the intervals right-exclusive.
+}
+\label{tab:jackson}
 \begin{tabular}{lllr@{--}lr@{+}l}
 \toprule
-Group        & Sequences & Method & \multicolumn{2}{c}{Interval}
-    & \multicolumn{2}{r}{Parent lineages} \\
+Group & Sequences & Method & \multicolumn{2}{c}{Breakpoint interval} & \multicolumn{2}{r}{Parent lineages} \\
 \midrule
@@
 C            & 3   &Jackson         &  24,913&28,653 &  B.1.1.7&B.1.221 \\
-             &     & \texttt{sc2ts} &  25,997&27,972 &  B.1.1.7&B.1.221.1 \\
+             &     &\texttt{sc2ts} &  25,997&27,972 &  B.1.1.7&B.1.221.1 \\
 \cmidrule{3-7}
 D            & 3   & Jackson        &  21,574&23,065 &  B.1.36.17&B.1.1.7 \\
-             &     & \texttt{sc2ts} &  22,445&23,063 &  B.1.36.39&B.1.1.7 \\
+             &     &\texttt{sc2ts} &  22,445&23,063 &  B.1.36.39&B.1.1.7 \\
 \bottomrule
@@
 \includegraphics[width=\linewidth]{inference-resources}
-    \caption{TODO caption}\label{fig:inference_resources}
+\caption{TODO caption}
+\label{fig:inference_resources}
 \end{figure}
@@
 \includegraphics[width=\linewidth]{robust_vs_artificial_classification}
-    \caption{Classification of ``robust'' and artifactual recombination events.
+\caption{
+Classification of ``robust'' and artifactual recombination events.
 Scatterplot of recombination events by the net number of supporting loci
@@
 Recombination events associated with the origins of Pango Xs are labeled.
-}\label{fig:recombinant_qc}
+}
+\label{fig:recombinant_qc}
 \end{figure}
@@
 % \includegraphics{sc2ts_methodology}
-    \caption{Phylogenetic backbones: TODO}\label{fig:phylo_backbones}
+\caption{
+Phylogenetic backbones: TODO
+}
+\label{fig:phylo_backbones}
 \end{figure}
@@
 the Viridian UShER phylogeny from Hunt et al. (2024)\cite{Hunt2024}.
+}
 \label{fig:mutational_spectra}
-}
 \end{figure}
@@
 \includegraphics[width=0.5\textwidth]{LS_model_schematic}
-\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
+\caption{
+A schematic of the Li and Stephens (LS)
 model, in which a focal sequence (bottom) is described as an
@@
 }
+\label{fig:ls_diagram}
 \end{figure}

### 2025-09-10 latex/main.tex
Tables for deletion analysis

@@
 \centering
-\caption{
-Major deletion events in the ARG.
-These events are observed in more than 10,000 samples in the ARG.
-}
+\caption{Major deletion events in the ARG.}
 \label{tab:major_dels}
@@
 \toprule
-Start & Length & Region & Node & Pango & Count & Frequency (\%)\\
+Start & Length & Region & Node & Pango lineage & Count & Frequency (\%)\\
 \midrule
 28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
-28248 & 6 & ORF8 & 220186 & B.1.617.2 & 1114669 & 44.9073 \\
-22029 & 6 & S / S1 - NTD & 200039 & B.1.617.2 & 1112410 & 44.8163 \\
-11288 & 4 & ORF1AB / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
+28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 \\
+22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112410 & 44.8163 \\
+11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
 28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 \\
-21633 & 9 & S / S1 - NTD & 822854 & BA.2 & 532072 & 21.4359 \\
-11292 & 5 & ORF1AB / NSP6 & 822854 & BA.2 & 531894 & 21.4287 \\
-6513 & 3 & ORF1AB / NSP3 & 851246 & BA.1 & 340765 & 13.7286 \\
-11283 & 5 & ORF1AB / NSP6 & 851246 & BA.1 & 339867 & 13.6924 \\
-21765 & 6 & S / S1 - NTD & 851246 & BA.1 & 339681 & 13.6849 \\
-21988 & 8 & S / S1 - NTD & 851246 & BA.1 & 339651 & 13.6837 \\
-22194 & 3 & S / S1 - NTD & 851246 & BA.1 & 312706 & 12.5982 \\
-28271 & 1 & non-coding & 86456 & B.1.1.7 & 290446 & 11.7014 \\
-11288 & 9 & ORF1AB / NSP6 & 86456 & B.1.1.7 & 290408 & 11.6998 \\
-21991 & 3 & S / S1 - NTD & 86456 & B.1.1.7 & 288831 & 11.6363 \\
-21765 & 6 & S / S1 - NTD & 86456 & B.1.1.7 & 290438 & 11.7010 \\
-21765 & 6 & S / S1 - NTD & 1265302 & BA.4 & 190425 & 7.6718 \\
-28361 & 1 & N & 1246012 & BA.5.2.1 & 20822 & 0.8389 \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}
-\centering
-\caption{
-Deletion events associated with the Alpha variant in the ARG.
-These events occur above node 86456, which represents the origin of Alpha.
-}
-\label{tab:alpha_dels}
-\begin{tabular}{rrcrr}
-\toprule
-Start & Length & Region & Count & Frequency (\%) \\
-\midrule
-28271 & 1 & non-coding & 290446 & 11.7014 \\
-11288 & 9 & ORF1AB / NSP6 & 290408 & 11.6998 \\
-21991 & 3 & S / S1 - NTD & 288831 & 11.6363 \\
-21765 & 6 & S / S1 - NTD & 290438 & 11.7010 \\
+21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 \\
+11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 \\
+6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340765 & 13.7286 \\
+11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 \\
+21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 \\
+21988 & 8 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
+22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 \\
+28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 \\
+11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 \\
+21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 \\
+21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 \\
+21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 \\
+28361 & 1 & N & 1246012 & BA.5.2.1 (Omicron) & 20822 & 0.8389 \\
 \bottomrule
@@
 222 & 1 & non-coding & 3441 \\
-22013 & 6 & S / S1 - NTD & 2111 \\
-21991 & 3 & S / S1 - NTD & 2087 \\
+22013 & 6 & S / S1-NTD & 2111 \\
+21991 & 3 & S / S1-NTD & 2087 \\
 28254 & 1 & ORF8 & 1967 \\
@@
 28093 & 1 & ORF8 & 898 \\
-21765 & 6 & S / S1 - NTD & 478 \\
+21765 & 6 & S / S1-NTD & 478 \\
 27532 & 1 & ORF7a & 403 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}
+\centering
+\caption{Most commonly observed deletions in the ARG or Li et al. (2023).} 
+\label{tab:common_dels}
+\begin{tabular}{rrcrrrr}
+\toprule
+\multicolumn{3}{c}{} & \multicolumn{2}{c}{Sc2ts ARG} & \multicolumn{2}{c}{Li et al. (2023)} \\
+\cmidrule(lr){4-5} \cmidrule(lr){6-7}
+Start & Length & Region & Count & Frequency (\%) & Count & Frequency (\%)\\
+\midrule
+28271 & 1 & non-coding & 1422469 & 57.3078 & 3853692 & 42.1183 \\
+28248 & 6 & ORF8 & 1114712 & 44.9090 & 3026186 & 33.0742 \\
+22029 & 6 & S / S1-NTD & 1114415 & 44.8970 & 3010665 & 32.9046 \\
+11288 & 4 & ORF1ab / NSP6 & 871838 & 35.1242 & N/A & N/A \\
+28362 & 9 & N & 871778 & 35.1218 & 3939164 & 43.0525 \\
+21765 & 6 & S / S1-NTD & 825628 & 33.2625 & 3179887 & 34.7541 \\
+21633 & 9 & S / S1-NTD & 532084 & 21.4364 & 1941962 & 21.2244 \\
+11292 & 5 & ORF1ab / NSP6 & 531943 & 21.4307 & N/A & N/A \\
+6513 & 3 & ORF1ab / NSP3 & 341714 & 13.7668 & 1984683 & 21.6913 \\
+11283 & 5 & ORF1ab / NSP6 & 339935 & 13.6951 & 8360 & 0.0914 \\
+21988 & 8 & S / S1-NTD & 339770 & 13.6885 & 590 & 0.0064 \\
+22194 & 3 & S / S1-NTD & 313878 & 12.6454 & 1833050 & 20.0340 \\
+11288 & 9 & ORF1ab / NSP6 & 311566 & 12.5522 & 3040576 & 33.2315 \\
+21991 & 3 & S / S1-NTD & 308800 & 12.4408 & 912944 & 9.9779 \\
+29762 & 1 & non-coding & 38030 & 1.5321 & N/A & N/A \\
+29781 & 1 & non-coding & 36112 & 1.4549 & N/A & N/A \\
+29769 & 1 & non-coding & 34315 & 1.3825 & N/A & N/A \\
+29774 & 1 & non-coding & 33503 & 1.3498 & N/A & N/A \\
+29779 & 1 & non-coding & 30515 & 1.2294 & N/A & N/A \\
+11283 & 9 & ORF1ab / NSP6 & 60 & 0.0024 & 2011934 & 21.9891 \\
 \bottomrule

### 2025-09-10 latex/main.tex
fix deletions

@@
 Illustrative subgraphs in the ARG.
-(A) Subgraph around the recombination event associated with the XA.
-The "Group A" sequences identified by Jackson et al.\cite{Jackson2021}) are attached to this event.
-(B) Subgraph around the recombination event associated with a group of ``nested'' Pango Xs; and
+(A) Subgraph around the recombination node associated with the XA.
+Two "Group A" recombinant samples reported by Jackson et al.\cite{Jackson2021}) are shown.
+(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages; and
 (C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
+The nodes assigned to a Pango X lineage are colored by that lineage.
+The numbers inside the colored boxes indicate
+the number of samples assigned to the Pango X lineage indicated by the same color.
+The numbers below the vertical dashed lines indicate
+the total number of samples within a collapsed clade.
+For each recombination node, a genome schematic shows
+the location of its inferred breakpoint interval relative to the Spike gene.
+The ancestry above the parents of the recombination nodes or the MRCA is not shown for clarity.
 For more detailed visualizations of these subgraphs, see Document S2.
@@
 \cmidrule(lr){4-5} \cmidrule(lr){6-7}
-Start & Length & Region & Count & Frequency (\%) & Count & Frequency (\%)\\
+Start & Length & Region & Count & Frequency (\%) & Count & Frequency (\%) \\
 \midrule
@@
 11283 & 9 & ORF1ab / NSP6 & 60 & 0.0024 & 2011934 & 21.9891 \\
+686 & 9 & ORF1ab / NSP1 & N/A & N/A & 115450 & 1.2618 \\
+21987 & 9 & S / S1-NTD & N/A & N/A & 1918094 & 20.9635 \\
 \bottomrule

### 2025-09-10 notebooks/nb_utils.py
Sort PangoX labels before joining
See

@@ -66,7 +66,7 @@ def standard_recombinant_labels(ts, pango_x_events_file):
                 if len(rows) == 1:
                     labels[re_node] = rows.iloc[0].root_pango
                 else:
-                    labels[re_node] = "/".join(rows.root_pango.values)
+                    labels[re_node] = "/".join(sorted(rows.root_pango.values))
     
     # Tweak the "XBB.1" RE node label, which should have a bespoke label because it's not
     # reflective of the majority of XBB.1 samples

### 2025-09-10 latex/main.tex
UPdates

@@
 
-The genomic revolution has brought about a paradigm shift
-in the way we study the evolution and spread of pathogens,
+The genomic revolution has transformed how we study the evolution and spread of pathogens,
 allowing real-time tracking of pathogens over time and space.
@@
 Although real-time phylogenetic methods now exist to cope with this flood of genomic data
-(e.g., UShER\cite{Turakhia2021,DeMaio2023}),
+(e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
 we lack methods to build phylogenies that can integrate recombination
@@
 (e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
-UShER uses a tree to represent evolutionary change via mutation, assuming no recombination \cite{Turakhia2021}.
-However, given extensive evidence for recombination in SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
+UShER uses a tree to represent evolutionary change via mutation,
+assuming no recombination \cite{Turakhia2021}.
+However, given extensive evidence for recombination in SARS-CoV-2
+\cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
 the tree-like representation of UShER is no longer tenable.
 
-Ancestral Recombination Graphs (ARGs) are reticulate genealogies that incorporate both mutation and recombination events \cite{Wong2024},
-potentially allowing more comprehensive evolutionary representations of SARS-CoV-2 and other viruses undergoing recombination.
-Until recently, ARGs lacked software support and sufficiently scalable inference methods to be of practical use.
+Ancestral Recombination Graphs (ARGs) are reticulate genealogies
+that incorporate both mutation and recombination events \cite{Wong2024},
+potentially allowing more comprehensive evolutionary representations of SARS-CoV-2
+and other viruses undergoing recombination.
+Until recently, ARGs lacked software support and
+sufficiently scalable inference methods to be of practical use.
 Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023},
@@
 reconstructing the past from a sample of genome sequences drawn at one moment in time,
-but scientists are increasingly tracking evolutionary change as it happens, accumulating data forward in time.
+but scientists are increasingly tracking evolutionary change as it happens,
+accumulating data forward in time.
 For example, sequences from daily samples of SARS-CoV-2 were made globally available on a daily basis,
@@
 (2) an ARG for SARS-CoV-2 with ~2.48 million genome sequences.
-We used the ``Viridian'' dataset, which is purged of various systematic errors and artifacts \cite{Hunt2024} and
+We used the ``Viridian'' dataset,
+which is purged of various systematic errors and artifacts \cite{Hunt2024} and
 show that this ARG is highly congruent to an UShER phylogeny built from the same dataset.
@@
 First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
-(allowing recombination) using the Li and Stephens (LS) model\cite{LiStephens2003},
+(allowing recombination) using the Li and Stephens (LS) model \cite{LiStephens2003},
 which is a Hidden Markov Model (HMM) widely used in human genomics
@@
 Sc2ts uses collection dates to build a time-resolved ARG.
-When examining preliminary ARGs, we found samples that introduced large numbers of mutations (and reversions),
+When examining preliminary ARGs,
+we found samples that introduced large numbers of mutations (and reversions),
 likely due to errors in their collection dates
@@
 and a switch has a penalty cost of $k$ (here, set to 4),
-which determines how much mutation is favored over recombination when accounting for a 
-sample (Figure~\ref{fig:methodology}).
+which determines how much mutation is favored over recombination
+when accounting for a sample (Figure~\ref{fig:methodology}).
 If the HMM cost of a sample exceeds a threshold (here, set to 7),
@@
 \includegraphics[width=0.85\linewidth]{sc2ts_methodology}
-    \caption{Schematic overview of sc2ts.
+\caption{
+Schematic overview of sc2ts.
 (A) Procedure to build an ARG by incrementally updating it with samples sorted by collection dates.
@@
 Next, Sample 1 is attached, mapping mutations onto the new branch to explain it.
-When Sample 4 is detected as a recombinant, a node representing the recombination event is inserted with two edges connecting to its parents, with Sample 4 itself attached to this node as its child.
-The resulting ARG contains two local trees, one on each side of the inferred breakpoint.
-(B) Copying paths for Sample 4 inferred under the LS model while allowing recombination ($k = 4$) or disallowing it ($k = 1000$).}
+When Sample 4 is detected as a recombinant,
+a node representing the recombination event is inserted with two edges connecting to its parents,
+with Sample 4 itself attached to this node as its child.
+The resulting ARG contains two local trees,
+one on each side of the inferred breakpoint.
+(B) Copying paths for Sample 4 inferred under the LS model
+while allowing recombination ($k = 4$) or disallowing it ($k = 1000$).
+}
 \label{fig:methodology}
@@
 The ARG represents the full evolutionary history of SARS-CoV-2 and 
-is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024}.
-First, we validated the broad structure of the ARG against the UShER phylogeny\cite{Hunt2024}
+is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}.
+First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
 by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
@@
 both within sublineages and at the level of major lineages (including VOCs),
-with good agreement between the phylogenetic bipartitions in the ARG and the UShER 
-phylogeny (Figure~\ref{fig:phylo_backbones}).
+with good agreement between the phylogenetic bipartitions
+in the ARG and the UShER phylogeny (Figure~\ref{fig:phylo_backbones}).
 Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset
@@
 We also compared the mutational spectra of major VOCs in the ARG against the mutational spectra 
-from the Viridian UShER phylogeny\cite{Hunt2024} and a previous study\cite{Bloom2023} (STAR Methods),
+from the Viridian UShER phylogeny \cite{Hunt2024} and a previous study \cite{Bloom2023} (STAR Methods),
 finding a close match to them (Figure~\ref{fig:mutational_spectra}).
@@
 particularly increased transmissibility, disease severity, and
-ability to escape natural or vaccine-induced immunity (e.g., XBB\cite{Tamura2023}).
+ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2023}).
 The process of gathering evidence for and characterizing such recombinants is manual,
@@
 This process has resulted in the designation of recombinant Pango lineages
-(whose aliases begin with an ``X''), to which we refer as ``Pango X''.
-Below, we summarize our analysis of the Pango Xs in the ARG
+(whose aliases begin with an ''X'', e.g., XA and XBB).
+Below, we summarize our analysis of the Pango X lineages in the ARG
 (for extended analyses, see Document~\ref{sec:supplement_pango_lineages} and Document S2).
 
-We find strong evidence for the recombination events leading to 17 Pango Xs in
-the ARG (Table~\ref{tab:pango_x_lineages}), all of which pass the 
-robustness filtering criteria (Figure~\ref{fig:recombinant_qc}). 
+We find strong evidence for the recombination events leading to 17 Pango X lineages
+in the ARG (Table~\ref{tab:pango_x_lineages}),
+all of which pass the robustness filtering criteria (Figure~\ref{fig:recombinant_qc}).
 %% JK: Does saying this here help at all? I think we may be better off without 
@@
 % (see their copying patterns in Figure SX). 
-Among these Pango Xs are XA,
-which is the first designated Pango X that showed signs of onward transmission\cite{Jackson2021}, and
-XBB, which escaped vaccine-induced immunity and gained global prevalence\cite{Tamura2023}.
+Among these Pango X lineages are XA,
+which is the first designated Pango X lineage that
+showed signs of onward transmission \cite{Jackson2021}, and
+XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
 The recombinant origin of XA is well explained in the ARG by two parents and one mutation
@@
 
-Furthermore, some of the Pango Xs above originate from shared recombination events in the 
-ARG (Table~\ref{tab:pango_x_lineages}).
+Furthermore, some of the Pango X lineages above originate from
+shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
 An example of a set of ``nested'' closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
@@
 events in the ARG (Figure~\ref{fig:subgraphs}B).
-These Pango Xs were designated around the same time.
+These Pango X lineages were designated around the same time.
 These results highlight that the ARG can help clarify the relationships between recombinants,
@@
 
-Sc2ts identified 12 Pango Xs in the ARG as non-recombinants
+Sc2ts identified 12 Pango X lineages in the ARG as non-recombinants
 (Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
-For all these Pango Xs, the number of recombination-informative sites fell below our threshold of detection (i.e. $k = 4$ in the HMM).
-The fact that most of these Pango Xs (except XAJ and XAN) differ from a single parent node in the ARG
+For all these Pango X lineages, the number of recombination-informative sites
+fell below our threshold of detection (i.e. $k = 4$ in the HMM).
+The fact that most of these Pango X lineages (except XAJ and XAN) differ from a single parent node in the ARG
 by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
@@
 % really are recombinants, especially XP.
-However, sc2ts did not detect several Pango Xs (XB, XP, and XAJ) as recombinants
+However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
 because the HMM treated characteristic deletions as missing data.
@@
 (A) Subgraph around the recombination node associated with the XA.
-Two "Group A" recombinant samples reported by Jackson et al.\cite{Jackson2021}) are shown.
+Two "Group A" recombinant samples reported by Jackson et al. \cite{Jackson2021}) are shown.
 (B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages; and
@@
 We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
-RecombinHunt\cite{Alfonsi2024}, CovRecomb\cite{Li2024CovRecomb}, and rebar (https://github.com/phac-nml/rebar).
+RecombinHunt \cite{Alfonsi2024}, CovRecomb \cite{Li2024CovRecomb}, and rebar (https://github.com/phac-nml/rebar).
 These methods use the same underlying approach:
-they scan candidate recombinant sequences for mutations associated with different Pango lineages
+they scan candidate recombinant sequences for mutations 
+associated with different Pango lineages
 without using a phylogeny.
@@
 using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
-For RecombinHunt and CovRecomb, we summarized the detection results for these Pango X lineages (if available)
-reported in the original studies\cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
+For RecombinHunt and CovRecomb,
+we summarized the detection results for these Pango X lineages (if available)
+reported in the original studies \cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
 For each method, we determined the Pango X lineages
@@
 and found that 167 (43\%) of them were also detected as recombinants by rebar.
-We hypothesize that this disagreement arose from differences in sensitivity between these methods.
+We hypothesize that this disagreement arose from
+differences in sensitivity between these methods.
 Consistent with this, we found that rebar needs a higher level of divergence between parents
@@
 These comparisons focus on the Pango X lineages associated with
-the Type 1 recombination events (Table \ref{tab:pango_x_lineages})
+the Type 1 recombination events (Table~\ref{tab:pango_x_lineages})
 For each method, we calculated the number of Pango X lineages
@@
 The denominators indicate the number of Pango X lineages
-for which detection results are available from the original studies\cite{Alfonsi2024,Li2024CovRecomb}.
+for which detection results are available from the original studies \cite{Alfonsi2024,Li2024CovRecomb}.
 For the Pango X lineages with non-overlapping breakpoint intervals,
@@
 by comparing how well a sample fits the ARG under a model allowing recombination
-versus a model disallowing recombination, as measured by the number of resulting 
-mutations (Figure~\ref{fig:methodology}).
-The more mutations are ``averted'' by allowing recombination, the stronger the 
-evidence for recombination.
-We calculated the number of averted mutations for all the ``robust'' recombination 
-events (STAR Methods),
+versus a model disallowing recombination,
+as measured by the number of resulting mutations (Figure~\ref{fig:methodology}).
+The more mutations are ``averted'' by allowing recombination,
+the stronger the evidence for recombination.
+We calculated the number of averted mutations for
+all the ``robust'' recombination events (STAR Methods),
 which include all the Pango X-associated recombination events
-(Table~\ref{tab:pango_x_lineages}) and those studied by Jackson et
-al.\cite{Jackson2021} (Table \ref{tab:jackson}).
+(Table~\ref{tab:pango_x_lineages})
+and those studied by Jackson et al. \cite{Jackson2021}
+(Table \ref{tab:jackson}).
 
@@
 (Figure~\ref{fig:averted_mutations}A).
-Examples of highly plausible recombinants are XA and XBB (for an extended analysis, 
-see Document~\ref{sec:supplement_pango_lineages}).
+Examples of highly plausible recombinants are XA and XBB
+(for an extended analysis, see Document~\ref{sec:supplement_pango_lineages}).
 The ``copying pattern'' of XA visually confirms
@@
 Future work is needed to test whether these VOCs descended from recombination events,
-with improved sampling if still possible.
+with improved sampling, if possible.
 
@@
 \includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
-\caption{Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes, colored according
-to their classification by rebar (recombinant: purple, non-recombinant: gray).
-(A) Upper: jittered scatterplot comparing number of mutations when recombination allowed ($y$: right axis)
+\caption{
+Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes,
+colored according to their classification by rebar
+(recombinant: purple, non-recombinant: gray).
+(A) Upper: jittered scatterplot comparing number of mutations
+when recombination allowed ($y$: right axis)
 versus an alternative copying path disallowing recombination ($x$: top axis);
 recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
-those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
+those identified by Jackson et al. \cite{Jackson2021} are prefixed with a `J', e.g.
 Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1).
 Point sizes reflect number of descendants.
-Lower: recombination nodes classified by number of mutations averted by allowing recombination (i.e. $x$-$y$).
+Lower: recombination nodes classified by number of mutations averted
+by allowing recombination (i.e. $x$-$y$).
 (B) Distribution of time to the MRCA of the parents of each robust recombination node.
@@
 We explored the spatial distribution of the breakpoint locations of
-the ``robust'' recombination events along the genome, and quantified how precisely breakpoints were localized.
+the ``robust'' recombination events along the genome,
+and quantified how precisely breakpoints were localized.
 For each event, we computed an interval estimate of its breakpoint location,
@@
 We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
-particularly at the left boundary of the Spike gene
-(Figure~\ref{fig:recombinant_intervals}A,B),
-consistent with patterns observed in previous 
-studies\cite{Turakhia2022,Li2024CovRecomb}.
+particularly at the left boundary of the Spike gene (Figure~\ref{fig:recombinant_intervals}A,B),
+consistent with patterns observed in previous studies \cite{Turakhia2022,Li2024CovRecomb}.
 While this might result from a higher recombination rate in this region,
@@
 When the two parental lineages have diverged for longer periods of time (the rightmost points),
-more information becomes available on either side of a breakpoint to pinpoint its location, resulting in shorter intervals.
+more information becomes available on either side of a breakpoint to pinpoint its location,
+resulting in shorter intervals.
 For instance, the breakpoint intervals tend to be the smallest for events involving recombination
@@
 \includegraphics[width=\linewidth]{recombinant_intervals}
-\caption{Recombination intervals. (A) Distribution of the breakpoint
-intervals of the 356 robust recombination events along the genome
-(annotated with gene labels on top). Each line represents an interval colored by its size.
+\caption{
+Recombination intervals.
+(A) Distribution of the breakpoint intervals of the 356 robust recombination events along the genome
+(annotated with gene labels on top).
+Each line represents an interval colored by its size.
 (B) Number of overlapping breakpoint intervals.
@@
 
-We examined the frequency of emergence of the recombinants over the course of the COVID-19 pandemic (Figure \ref{fig:recombinants_over_time}).
+We examined the frequency of emergence of the recombinants
+over the course of the COVID-19 pandemic (Figure \ref{fig:recombinants_over_time}).
 We explored two hypotheses for the timing of recombination.

### 2025-09-11 arg_postprocessing/scripts/generate_sites_for_remapping.py
Update pipeline for Li et al deletions

@@ -1,0 +1,69 @@
+import click
+import pandas as pd
+import numpy as np
+import json
+import tszip
+
+
+@click.command()
+@click.argument("base_ts")
+@click.argument("del_data")
+@click.argument("lineage_data")
+@click.argument("output_csv")
+@click.argument("output_txt")
+def run(base_ts, del_data, lineage_data, output_csv, output_txt):
+
+    base_ts = tszip.load(base_ts)
+    ts_sites = set(base_ts.sites_position.astype(int))
+    print(f"Have {len(ts_sites)} in base ts")
+
+    with open(lineage_data) as f:
+        doc = json.load(f)
+
+    lineage_sites = set()
+    for lineage, mut_list in doc.items():
+        for mut in mut_list:
+            site = int(mut[1:-1])
+            lineage_sites.add(site)
+
+    missing_lineage_sites = lineage_sites - ts_sites
+
+    print(
+        f"Got {len(lineage_sites)} lineage defining sites: "
+        f" {len(missing_lineage_sites)} missing from ts"
+    )
+
+    df = pd.read_excel(del_data)
+    # Denominator here reported in the Text of Li et al.
+    df["frequency"] = df["Count"] / 9_149_680
+    df_1pc = df[df["frequency"] >= 0.01].sort_values("Start")
+    print(df_1pc)
+    del_sites = set()
+    for _, row in df_1pc.iterrows():
+        start = row["Start"]
+        for j in range(start, start + row["Length"]):
+            del_sites.add(j)
+
+    missing_del_sites = del_sites - ts_sites
+    print(
+        f"Got {len(del_sites)} deletion sites"
+        f" {len(missing_del_sites)} missing from ts"
+    )
+
+    # There are two mutually exclusive sets
+    assert len(missing_lineage_sites & del_sites) == 0
+    data = []
+    for pos in missing_lineage_sites:
+        data.append({"position": pos, "reason": "pango"})
+    for pos in del_sites:
+        data.append({"position": pos, "reason": "deletion"})
+    df = pd.DataFrame(data).sort_values("position")
+
+    df.to_csv(output_csv, index=False)
+    # We save to CSV and text here because sc2ts wants a text
+    # file as input and it's not worth changing that.
+    np.savetxt(df.position.values, output_txt)
+
+
+if __name__ == "__main__":
+    run()

### 2025-09-11 arg_postprocessing/scripts/generate_sites_for_remapping.py
Fix some errors

@@ -62,7 +62,7 @@ def run(base_ts, del_data, lineage_data, output_csv, output_txt):
     df.to_csv(output_csv, index=False)
     # We save to CSV and text here because sc2ts wants a text
     # file as input and it's not worth changing that.
-    np.savetxt(df.position.values, output_txt)
+    np.savetxt(output_txt, df.position.values)
 
 
 if __name__ == "__main__":

### 2025-09-11 latex/main.tex
First pass at abstract

@@
 \usepackage{booktabs}
+\usepackage{tabularx}
 \usepackage{amsmath}
@@
 \section*{SUMMARY}
-
-%%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.
-%%% This is the SMBE abstract. We can rewrite or work off of it, but it needs to be shorter.
-Widely used phylogenetic methods to study SARS-CoV-2 evolution assume that genetic recombination is negligible and
-viral relationships can be represented by a single tree.
-However, important SARS-CoV-2 recombinant lineages have emerged.
-A more suitable approach is to use ancestral recombination graphs (ARGs),
-which jointly model recombination, mutation, and genetic inheritance in a cohesive framework.
-Recent computational advances have made ARG inference possible for datasets of millions of genome sequences.
-Building on these new advances, we have developed sc2ts, a scalable method to simultaneously detect recombinants and
-reconstruct ARGs from heterochronous SARS-CoV-2 genome sequences.
-Using sc2ts, we inferred an ARG of 2.48 million genomes sampled from January, 2020 to February, 2023.
-We find that this pandemic-scale ARG unifies known features of SARS-CoV-2 evolution —
-the backbone phylogeny, mutational spectra, and major recognized recombinants.
-This ARG enables us to explore the evolutionary origins of known and novel recombinants.
-Furthermore, the ARG is compactly stored in the tree sequence format using the tskit library and
-can be analyzed efficiently and interactively using standard Python tools on a typical laptop.
-This combination of capabilities makes sc2ts a powerful tool
-for studying the recombination and mutation processes driving SARS-CoV-2 evolution.
+% The dataset is freakin amazing, and will be useful for all sorts of stuff
+% for years to come
+Millions of SARS-CoV-2 whole genome sequences were collected as part of the 
+response to the COVID-19 pandemic, generating a dataset of unprecedented
+depth, richness and potential for informing future studies.
+% Trees are a useful abstraction of the data
+Estimated genealogies, describing the likely inheritance structure among
+the sampled sequences and their putative ancestors, are fundamental to
+% JK: Not happy with this specific phrase, but it's good to emphasise the point
+% that data is not insight, and the sheer volume of data here requires
+% special effort.
+gleaning information from this ocean of data, 
+and form the primary input to many downstream analyses.
+A basic assumption of methods to infer such genealogies from viral 
+genetic sequence data, however, is that recombination is negligible and 
+the genealogy can be described as a tree.
+% Recombination *is* important.
+Recombinant lineages have risen to global prevalance in SARS-CoV-2,
+and therefore such single-tree summaries of the data are necessarily
+incomplete and potentially misleading.
+% What do we do?
+We present sc2ts, a method to infer reticulate genealogies for SARS-CoV-2
+that automaticaly detects and fully incorporates recombination, 
+and apply it to 2.48 million genomes.
+We show that this structure captures a broad range of evolutionary
+features documented in the literature and provides novel 
+insights into the relationships between previously identified
+recombinant lineages.
+% We need to emphasise the *resource*
+Together, the inference method, inferred genealogy and 
+a suite of efficient and flexible open-source analysis tools, 
+represent a powerful resource upon which future research 
+can build.
 
@@
 This ``retrospective sample grouping'' successfully captured the emergence of
-most saltation lineages (Figure~\ref{fig:samples_per_day}).
+many saltation lineages such as Alpha (Figure~\ref{fig:subgraph_alpha}).
 The emergence of the Delta, BA.1, and BA.2 lineages required some manual intervention,
-in which specific "seed" samples were identified for unconditional inclusion
+in which specific ``seed'' samples were identified for unconditional inclusion
 (for details of how seed samples were chosen, see STAR Methods).
@@
 the initial origins of these VOC are challenging to infer (REF).
+With the addition of these manual seeds, we successfully incorporated 
+these challenging events (Figures~\ref{fig:subgraph_delta},
+\ref{fig:subgraph_ba1_ba2}) and 
+all other major lineages (Figure~\ref{fig:samples_per_day}) 
+into the ARG.
 
@@
 First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
-by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
-Tanglegrams reveal that the ARG captures known phylogenetic relationships
-both within sublineages and at the level of major lineages (including VOCs),
-with good agreement between the phylogenetic bipartitions
-in the ARG and the UShER phylogeny (Figure~\ref{fig:phylo_backbones}).
-Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset
-by pruning both to their intersection of 2,475,418 samples and 27,431 sites.
-% TODO this is out of date and needs to be rewritten along with supplementary 
-% analysis
-Remarkably, given the differences in inference methods and data processing pipelines,
-they are almost identically parsimonious,
-with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts).
-We also compared the mutational spectra of major VOCs in the ARG against the mutational spectra 
+by pruning both down to representative samples for 1097 deeply-sampled 
+lineages (STAR Methods).
+Tanglegrams show that the ARG closely matches the phylogenetic
+structure of the UShER tree (Figures~\ref{fig:tanglegram_base},
+\ref{fig:tanglegram_delta},
+\ref{fig:tanglegram_ba2},
+\ref{fig:tanglegram_ba5}).
+Next, we compared the sc2ts and UShER inferences 
+from the Viridian dataset
+by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
+% JK: this feels a bit wishy washy, but there really is no simple answer to
+% this and I don't want to claim we're more parsimonious.
+and found they were very similar in terms of parsimony
+(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
+and imputation of missing and ambiguous bases
+(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
+We also compared the mutational spectra of major VOCs in the ARG against 
+the mutational spectra 
 from the Viridian UShER phylogeny \cite{Hunt2024} and a previous study \cite{Bloom2023} (STAR Methods),
@@
 (whose aliases begin with an ''X'', e.g., XA and XBB).
-Below, we summarize our analysis of the Pango X lineages in the ARG
-(for extended analyses, see Document~\ref{sec:supplement_pango_lineages} and Document S2).
 
@@
 because the HMM treated characteristic deletions as missing data.
+
+[NOTE: putting this here for now, until we find the right place to slot it
+in next draft.]
+See Document~\ref{sec:supplement_pango_lineages} for extended analyses,
+Figure~\ref{fig:pango_x_copying_patterns} for copying patterns,
+and Document S2 for ARG visualations of all Pango X lineages.
 
@@
 [TODO This section needs a full pass. I've merged in a section that was in the
-supplement with an extremely terse section on counting supporting loci.]
+supplement with an extremely terse section on counting supporting loci.
+Reference Figure~\ref{fig:recombinant_qc} for the scatter plot and 
+labelled quadrants, and Figure~\ref{fig:copying_pattern_examples}
+for examples of the corresponding copying patterns. Also reference
+any additional examples that we include beyond A-D.]
 
@@
 
+\subsubsection*{Comparison with UShER}
+To facilitate comparison with the sc2ts ARG, we first converted the Viridian 
+UShER tree from MAT~\cite{McBroome2021} to tskit format, which has similar
+properties but is more general (see the section ``Ancestral Recombination
+Graphs``).
+We first downloaded the tree in JSON format (tree.all\_viridian.202409.jsonl)
+from Figshare
+and converted the ARG topology. We then downloaded the protobuf formatted
+tree (tree.all\_viridian.202409.pb.gz), and exported the mutations to 
+nucleotide format using \texttt{matUtils summary --translate}
+(usher version 0.6.3), and translated
+these mutations into tskit format. The resulting tszip file 
+(usher\_viridian\_v1.0.trees.tsz) contains 5,345,019 nodes, 3,364,841
+mutations and requires 29MB of storage space.
+We validated that the data was correctly encoded by comparing the 
+variant data for the UShER tskit tree sequence against our alignments.
+There was an exact match between the non-ambiguous nucleotide calls 
+at 27469 sites. At the remaining 39 sites there were a significant 
+number of gap characters (deletion calls), suggesting that the mismatches
+could be due to differences in the alignments used as input to UShER.
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/validate\_usher\_tree.ipynb}{validate\_usher\_tree.ipynb}
+for details.
+
+We then computed intersection ARGs for sc2ts and UShER by 
+``simplifying''\cite{Kelleher2018efficient,Wong2024} 
+both to the 2,475,418 samples and 27,507 sites that are shared.
+
+All conversions were performed as part of the reproducible 
+postprocessing pipeline described in [section?]. The full UShER
+tree and the intersection files are available for download on Zenodo
+in tszip format.
+
 \subsubsection*{Comparison of phylogenetic backbones}
@@
 for the majority of the 2026 Pango lineages in the ARG.
-We first identify an "origination node" for a focal Pango lineage,
+We first identify an ``origination node" for a focal Pango lineage,
 defined as the oldest internal node in the ARG
@@
 % See https://github.com/jeromekelleher/sc2ts-paper/issues/628
-% Referenced by the main Results.
-The ARG captures highly prevalent deletions in SARS-CoV-2 (Table~\ref{tab:major_dels}).
+By mapping deletions to the ARG, we identified
+important deletion events (Table~\ref{tab:major_dels}),
+many of which are associated with the origins of major VOCs,
+for example, the deletions in ORF1ab and
+S1-NTD (the N-terminal domain of Spike)
+associated with the Alpha lineage.
+We also identified highly recurrent deletions (Table~\ref{tab:rec_dels}),
+for example, two known recurrent deletions in S1-NTD (REF).
+
 Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
@@
 except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
-During site remapping, the 9-base deletion at 11283 was split into
-a 5-base deletion at 11283 (count of 339867) and a 4-base deletion at 11288 (count of 871782).
-
-<TODO> Highly recurrent deletions (Table~\ref{tab:rec_dels}).
 
@@
 \centering
-\caption{Major deletion events in the ARG.}
+\caption{Major deletion events in the ARG.
+These events are inherited by at least 10,000 samples in the ARG.
+Count is the number of samples inheriting a deletion.
+The Pango lineage assigned to the node above which an event occurs is shown.
+Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in S1 subunit.}
 \label{tab:major_dels}
@@
 28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 \\
-22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112410 & 44.8163 \\
+22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 \\
 11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
@@
 11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 \\
-6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340765 & 13.7286 \\
+6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 \\
 11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 \\
 21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 \\
-21988 & 8 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
+21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
 22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 \\
@@
 21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 \\
-28361 & 1 & N & 1246012 & BA.5.2.1 (Omicron) & 20822 & 0.8389 \\
+686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 \\
 \bottomrule
@@
 \centering
-\caption{Highly recurrent deletions.}
+\caption{Highly recurrent deletions.
+These are the top 20 deletions in terms of the number of occurrences in the ARG.
+Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in S1 subunit.}
 \label{tab:rec_dels}
@@
 \midrule
-29781 & 1 & non-coding & 12713 \\
-29762 & 1 & non-coding & 11356 \\
-29779 & 1 & non-coding & 10925 \\
-29769 & 1 & non-coding & 10783 \\
-29774 & 1 & non-coding & 10707 \\
-29700 & 1 & non-coding & 8543 \\
-29614 & 1 & ORF10 & 5041 \\
-29555 & 1 & non-coding & 4816 \\
-29543 & 1 & non-coding & 4806 \\
-29541 & 1 & non-coding & 4804 \\
-203 & 1 & non-coding & 3602 \\
-222 & 1 & non-coding & 3441 \\
-22013 & 6 & S / S1-NTD & 2111 \\
-21991 & 3 & S / S1-NTD & 2087 \\
-28254 & 1 & ORF8 & 1967 \\
-27556 & 64 & ORF7a & 1281 \\
-27579 & 3 & ORF7a & 1014 \\
-28093 & 1 & ORF8 & 898 \\
-21765 & 6 & S / S1-NTD & 478 \\
-27532 & 1 & ORF7a & 403 \\
+29781 & 1 & non-coding & 12754 \\
+29762 & 1 & non-coding & 11383 \\
+29779 & 1 & non-coding & 10951 \\
+29769 & 1 & non-coding & 10809 \\
+29774 & 1 & non-coding & 10735 \\
+29700 & 1 & non-coding & 8562 \\
+686 & 9 & ORF1ab / NSP1 & 6436 \\
+29614 & 1 & ORF10 & 5039 \\
+29555 & 1 & non-coding & 4814 \\
+29543 & 1 & non-coding & 4804 \\
+29541 & 1 & non-coding & 4802 \\
+203 & 1 & non-coding & 3604 \\
+222 & 1 & non-coding & 3442 \\
+21991 & 3 & S / S1-NTD & 2088 \\
+28254 & 1 & ORF8 & 1969 \\
+28093 & 1 & ORF8 & 900 \\
+21765 & 6 & S / S1-NTD & 482 \\
+27532 & 1 & ORF7a & 404 \\
+335 & 1 & ORF1ab / NSP1 & 370 \\
+337 & 1 & ORF1ab / NSP1 & 346 \\
 \bottomrule
@@
 
-\begin{table}
-\centering
-\caption{Most commonly observed deletions in the ARG or Li et al. (2023).} 
-\label{tab:common_dels}
-\begin{tabular}{rrcrrrr}
-\toprule
-\multicolumn{3}{c}{} & \multicolumn{2}{c}{Sc2ts ARG} & \multicolumn{2}{c}{Li et al. (2023)} \\
-\cmidrule(lr){4-5} \cmidrule(lr){6-7}
-Start & Length & Region & Count & Frequency (\%) & Count & Frequency (\%) \\
-\midrule
-28271 & 1 & non-coding & 1422469 & 57.3078 & 3853692 & 42.1183 \\
-28248 & 6 & ORF8 & 1114712 & 44.9090 & 3026186 & 33.0742 \\
-22029 & 6 & S / S1-NTD & 1114415 & 44.8970 & 3010665 & 32.9046 \\
-11288 & 4 & ORF1ab / NSP6 & 871838 & 35.1242 & N/A & N/A \\
-28362 & 9 & N & 871778 & 35.1218 & 3939164 & 43.0525 \\
-21765 & 6 & S / S1-NTD & 825628 & 33.2625 & 3179887 & 34.7541 \\
-21633 & 9 & S / S1-NTD & 532084 & 21.4364 & 1941962 & 21.2244 \\
-11292 & 5 & ORF1ab / NSP6 & 531943 & 21.4307 & N/A & N/A \\
-6513 & 3 & ORF1ab / NSP3 & 341714 & 13.7668 & 1984683 & 21.6913 \\
-11283 & 5 & ORF1ab / NSP6 & 339935 & 13.6951 & 8360 & 0.0914 \\
-21988 & 8 & S / S1-NTD & 339770 & 13.6885 & 590 & 0.0064 \\
-22194 & 3 & S / S1-NTD & 313878 & 12.6454 & 1833050 & 20.0340 \\
-11288 & 9 & ORF1ab / NSP6 & 311566 & 12.5522 & 3040576 & 33.2315 \\
-21991 & 3 & S / S1-NTD & 308800 & 12.4408 & 912944 & 9.9779 \\
-29762 & 1 & non-coding & 38030 & 1.5321 & N/A & N/A \\
-29781 & 1 & non-coding & 36112 & 1.4549 & N/A & N/A \\
-29769 & 1 & non-coding & 34315 & 1.3825 & N/A & N/A \\
-29774 & 1 & non-coding & 33503 & 1.3498 & N/A & N/A \\
-29779 & 1 & non-coding & 30515 & 1.2294 & N/A & N/A \\
-11283 & 9 & ORF1ab / NSP6 & 60 & 0.0024 & 2011934 & 21.9891 \\
-686 & 9 & ORF1ab / NSP1 & N/A & N/A & 115450 & 1.2618 \\
-21987 & 9 & S / S1-NTD & N/A & N/A & 1918094 & 20.9635 \\
-\bottomrule
-\end{tabular}
-\end{table}
+\subsection{Parsimony comparison of sc2ts and UShER}
+\label{sec:supplement_sc2ts_usher_parsimony}
+
+We performed an indepth analysis of the sc2ts and UShER ARGs 
+(we refer to the UShER tree as an ARG for simplicity here as a single 
+tree is also an ARG) in terms of parsimony.
+A direct comparison is difficult because of the different aligments 
+used as input, the different sets of sites used for inference and 
+the fact that we are including deletions for a subset of sites in
+the sc2ts ARG.
+
+When we compare the intersection of the base sc2ts ARG (pre 
+site-remapping), sc2ts is very slightly more parsimonious 
+% TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/684
+with XX mutations over the YYY sites shared between them.
+After site remapping, and including all mutations involving 
+deletion characters, sc2ts has 143,690 mutations more 
+(2,078,370 vs 1,934,680). If we exclude mutations involving 
+deletions in sc2ts, this is reduced to 46,371 more (2.3\%)
+than UShER.
+If we restrict to the 26,192 sites that were not remapped, 
+sc2ts has 4,785 (0.26\%) fewer mutations than UShER,
+and has around 2.5 times more mutations than UShER on the 
+163 sites that were post-hoc remapped.
+
+To provide a more objective measure of the overall parsimony 
+of the inferred ARGs with respect to the Viridian alignments,
+we remapped all 29903 sites in both ARGs 
+using Fitch-Hartigan parsimony. Here the nucleotides
+ACGT and ``-'' were treated as distinct characters, with all 
+other values regarded as missing data. In this case, the sc2ts 
+ARG has 629,864 (5.5\%) fewer mutations (10,840,687 
+vs 11,470,551). Some of this difference may be due to the 
+post-hoc incorporation of phylogenetic signal from deletions 
+by sc2ts, but more work is required to fully understand 
+the differences between the two approaches.
+
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_usher\_sc2ts.ipynb}{analysis\_usher\_sc2ts.ipynb}
+for details.
+
+\subsection{Imputation comparison of sc2ts and UShER}
+\label{sec:supplement_sc2ts_usher_imputation}
+Both sc2ts and UShER support missing data in input alignments, and effectively
+``hard call'' these values when the samples are integrated into the tree/ARG. 
+In doing so, we impute the missing and ambiguous data for these samples.
+We evaluate the imputation performance of sc2ts and UShER by comparing their
+outputs on the Viridian alignments.
+
+The total aligned dataset for 2,475,418 samples and 27,507 sites 
+in the sc2ts and UShER intersection ARGs represents 63.41GiB of nucleotide calls. 
+Of the 162,838,461 missing data calls (Ns) in the alignments, sc2ts and UShER
+disagreed in their imputed values for only 203,579 (0.125\%). 
+Additionally, 956,859 calls made use of the IUPAC uncertainty codes. 
+Of these sc2ts imputed 435 (0.05\%) incorrectly (i.e., 
+with a base that is not compatible with the
+ambiguity code). 
+UShER imputed only 7 calls (0.007\%) from this set incorrectly, likely 
+due to special handling of ambiguous bases in the parsimony calculations.
+
+This computation is a good example of type of integrated large-scale data analysis 
+that can be easily performed using the combination of tskit and VCF Zarr.
+The entire analysis is performed in a short Jupyter notebook, with the 
+core calculation requiring about 30 lines of Python code. Retrieving 
+data from the base alignments and ARGs for 2.48 million samples 
+at the same time over all 27,507 sites took only 6 minutes, with a peak memory
+usage of about 5GiB of RAM on a standard desktop computer.
+
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/example\_data\_processing.ipynb}{example\_data\_processing.ipynb}
+for details.
 
@@
 
+\newcommand{\cprheight}{4.0em}
+
 \begin{figure}
 \centering
-% \includegraphics{sc2ts_methodology}
-\caption{
-Phylogenetic backbones: TODO
-}
-\label{fig:phylo_backbones}
+% NOTE the size of the second "m" column doesn't matter, but we have to fix the
+% size in order for vertical centering to work :sigh:
+% The layout here isn't ideal, but it's incredibly fiddly to get center
+% alignment veritically with cells to work, so this is good enough.
+\begin{tabularx}{0.85\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
+% FIXME: these samples are chosen randomly and so almost certainly do not
+% belong to the correct quadrants.
+A & \includegraphics[height=\cprheight]{static/causedby-ERR10306608.png}\\
+B & \includegraphics[height=\cprheight]{static/causedby-ERR6760245.png}\\
+C & \includegraphics[height=\cprheight]{static/causedby-SRR15864165.png}\\
+D & \includegraphics[height=\cprheight]{static/causedby-SRR21087725.png}\\
+\end{tabularx} 
+\caption{Examples of problematic copying patterns. A-D are examples drawn from
+the labelled quadrants in Figure~\ref{fig:recombinant_qc}.}
+\label{fig:copying_pattern_examples}
 \end{figure}
+
+\renewcommand{\cprheight}{1.5em}
+% Add some more space between rows. Note this it would be sligntly nicer if the 
+% label was aligned vertically in the center of the row, but this is good
+% enough for now.
+\renewcommand{\arraystretch}{2}
 
@@
 \centering
-%TODO fix this to make the figures smaller.
+\begin{tabularx}{\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
+XC & \includegraphics[height=\cprheight]{static/XC.png}\\
+XBR & \includegraphics[height=\cprheight]{static/XBR.png}\\
+XA & \includegraphics[height=\cprheight]{static/XA.png}\\
+XS & \includegraphics[height=\cprheight]{static/XS.png}\\
+XL & \includegraphics[height=\cprheight]{static/XL.png}\\
+XBG &\includegraphics[height=\cprheight]{static/XBG.png}\\
+XBD & \includegraphics[height=\cprheight]{static/XBD.png}\\
+XQ & \includegraphics[height=\cprheight]{static/XQ++.png}\\
+XBB & \includegraphics[height=\cprheight]{static/XBB.png}\\
+XM & \includegraphics[height=\cprheight]{static/XM+XAL.png}\\
+XBF & \includegraphics[height=\cprheight]{static/XBF.png}\\
+XF & \includegraphics[height=\cprheight]{static/XF.png}\\
+XY & \includegraphics[height=\cprheight]{static/XY.png}\\
+XG & \includegraphics[height=\cprheight]{static/XG.png}\\
+XW & \includegraphics[height=\cprheight]{static/XW.png}\\
+Xx & \includegraphics[height=\cprheight]{static/XZ++.png}\\
+XE/XH  & \includegraphics[height=\cprheight]{static/XE+XH.png}\\
+XBH & \includegraphics[height=\cprheight]{static/XBH.png}\\
+XBM & \includegraphics[height=\cprheight]{static/XBM.png}\\
+XJ & \includegraphics[height=\cprheight]{static/XJ.png}\\
+\end{tabularx}
+\caption{Copying patterns for the 20 recombination events associated with Pango 
+X lineages listed in Table~\ref{tab:pango_x_lineages} (in the same order)}
+\label{fig:pango_x_copying_patterns}
+\end{figure}
+
+% Reset this in case we make any more tables beyond this
+\renewcommand{\arraystretch}{1}
+
+\begin{figure}
+\centering
+\centering
+\includegraphics[width=\linewidth]{tanglegram_base_tree.pdf}
+\caption{
+Tanglegram comparing phylogenetic backbones of sc2ts and UShER.
+}
+\label{fig:tanglegram_base}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{tanglegram_Delta_subtree.pdf}
+\caption{
+Tanglegram comparing sc2ts and UShER on the Delta subtree.
+}
+\label{fig:tanglegram_delta}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{tanglegram_BA.2_subtree.pdf}
+\caption{
+Tanglegram comparing sc2ts and UShER on the BA.2 subtree.
+}
+\label{fig:tanglegram_ba2}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{tanglegram_BA.5_subtree.pdf}
+\caption{
+Tanglegram comparing sc2ts and UShER on the BA.5 subtree.
+}
+\label{fig:tanglegram_ba5}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.85\linewidth]{subgraph-Alpha.pdf}
+\caption{
+Subgraph of Alpha. [Some notes to help the reader know this 
+reflects known events]
+}
+\label{fig:subgraph_alpha}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.85\linewidth]{subgraph-Delta.pdf}
+\caption{
+Subgraph of Delta. [Some notes to help the reader know this 
+reflects known events]
+}
+\label{fig:subgraph_delta}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.85\linewidth]{subgraph-BA.1+BA.2.pdf}
+\caption{
+Subgraph of BA.1 and BA.2.
+[Some notes to help the reader know this 
+reflects known events]
+}
+\label{fig:subgraph_ba1_ba2}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics[width=\linewidth]{mutational_spectra}

### 2025-09-11 latex/main.tex
Trim down a bit

@@
 
+%%% The summary (abstract) should consist of a single paragraph of 150 words or
+%%% fewer.
+
 \section*{SUMMARY}
-% The dataset is freakin amazing, and will be useful for all sorts of stuff
-% for years to come
-Millions of SARS-CoV-2 whole genome sequences were collected as part of the 
-response to the COVID-19 pandemic, generating a dataset of unprecedented
-depth, richness and potential for informing future studies.
-% Trees are a useful abstraction of the data
-Estimated genealogies, describing the likely inheritance structure among
-the sampled sequences and their putative ancestors, are fundamental to
-% JK: Not happy with this specific phrase, but it's good to emphasise the point
-% that data is not insight, and the sheer volume of data here requires
-% special effort.
+
+Millions of SARS-CoV-2 genome sequences were collected in
+response to the COVID-19 pandemic, forming a dataset of unprecedented
+depth and richness.
+Estimated genealogies are fundamental to
 gleaning information from this ocean of data, 
 and form the primary input to many downstream analyses.
-A basic assumption of methods to infer such genealogies from viral 
-genetic sequence data, however, is that recombination is negligible and 
-the genealogy can be described as a tree.
-% Recombination *is* important.
+A basic assumption of methods to infer genealogies from viral 
+genetic sequence data is that recombination is negligible and 
+the genealogy is a tree.
 Recombinant lineages have risen to global prevalance in SARS-CoV-2,
-and therefore such single-tree summaries of the data are necessarily
+and therefore such single-tree representations are necessarily
 incomplete and potentially misleading.
-% What do we do?
 We present sc2ts, a method to infer reticulate genealogies for SARS-CoV-2
-that automaticaly detects and fully incorporates recombination, 
+that automaticaly detects and incorporates recombination, 
 and apply it to 2.48 million genomes.
-We show that this structure captures a broad range of evolutionary
-features documented in the literature and provides novel 
+We show that the inferred genealogy captures a broad range of 
+known evolutionary features and provides novel 
 insights into the relationships between previously identified
 recombinant lineages.
-% We need to emphasise the *resource*
 Together, the inference method, inferred genealogy and 
-a suite of efficient and flexible open-source analysis tools, 
-represent a powerful resource upon which future research 
-can build.
+suite of open-source analysis tools, 
+represent a powerful resource for future research.
 
@@
 
-ancestral recombination graphs, population genetics, recombination, tree sequence, SARS-CoV-2
+Ancestral recombination graphs, population genetics, recombination, tree sequence, SARS-CoV-2
 

### 2025-09-11 src/makepdf.py
Allow makepdf to convert SVGs as well as ipynb

@@ -5,68 +5,93 @@ from pathlib import Path
 import os
 from tempfile import NamedTemporaryFile
 
+from playwright.sync_api import sync_playwright
+
+
+def convert_file(infile, outdir):
+    infile = Path(infile).resolve(strict=True)
+    if outdir is None:
+        outdir = infile.parent
+    else:
+        outdir = Path(os.getcwd()) / outdir
+    outfile = outdir / str(infile.with_suffix('').name)
+
+    if infile.suffix == ".svg":
+        print(f"Converting {infile} to {outfile}")
+        with sync_playwright() as playwright:
+            browser = playwright.chromium.launch(headless=True)
+            context = browser.new_context()
+            page = context.new_page()
+            page.goto("file://" + str(infile))
+            page.emulate_media(media="print")
+            page.pdf(path=outfile.with_suffix('.pdf'))
+            browser.close()
+
+    elif infile.suffix == ".ipynb":
+        with open(infile, "r") as input, NamedTemporaryFile(mode="w") as output:
+            # Replace the javascript in the notebook code
+            # to change mutation symbols and node labels to contain URLs.
+            # This is a complete hack that fakes tooltips in a pdf, as URLS appear on hover.
+            for line in input.readlines():
+                ## Mutation symbols have a hover that shows the position and state change
+                if line.rstrip().endswith(r'var mut_symbol_rect = mut_symbol\n",'):
+                    start, end = line.rsplit(r'\n', 1)
+                    output.write(start + r'.append(\"a\").attr(\"href\", d => \"mut:\" + d.label)\n' + end)
+                #  Line labels where the second line starts with (DRR, (ERR, or (SRR
+                #  are turned into a link with the second line as the URL. Line labels where
+                #  the first line is   
+                #  This is EXTREMELY fragile. 
+                elif line.rstrip().endswith(r'const lines = text.split(\"\\n\");\n",'):
+                    output.write(r'''       "                let lines = text.split(\"\\n\");\n",
+        "                let url = null;\n",
+        "                if (lines.length == 2) {\n",
+        "                  if (lines[1].startsWith('(DRR') || lines[1].startsWith('(ERR') || lines[1].startsWith('(SRR')) {\n",
+        "                    url = 'sample:' + lines[1].replace(/[()]/g, ''); lines = [lines[0]];\n",
+        "                  }\n",
+        "                  else if (lines[1].startsWith('/')) {\n",
+        "                    lines = [lines[1], lines[0]];\n",
+        "                  }\n",
[…]

### 2025-09-12 latex/main.tex
Add some methods for UShER conversion

--- 2025-09-10 latex/main.tex
@@
 The ARG captures highly prevalent deletions in SARS-CoV-2 (Table~\ref{tab:major_dels}).
+
+<TODO> Highly recurrent deletions (Table~\ref{tab:rec_dels}).
+
 Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
@@
 \midrule
-29781 & 1 & non-coding & 12713 \\
-29762 & 1 & non-coding & 11356 \\
-29779 & 1 & non-coding & 10925 \\
-29769 & 1 & non-coding & 10783 \\
-29774 & 1 & non-coding & 10707 \\
-29700 & 1 & non-coding & 8543 \\
-29614 & 1 & ORF10 & 5041 \\
-29555 & 1 & non-coding & 4816 \\
-29543 & 1 & non-coding & 4806 \\
-29541 & 1 & non-coding & 4804 \\
-203 & 1 & non-coding & 3602 \\
-222 & 1 & non-coding & 3441 \\
-22013 & 6 & S / S1-NTD & 2111 \\
-21991 & 3 & S / S1-NTD & 2087 \\
-28254 & 1 & ORF8 & 1967 \\
-27556 & 64 & ORF7a & 1281 \\
-27579 & 3 & ORF7a & 1014 \\
-28093 & 1 & ORF8 & 898 \\
-21765 & 6 & S / S1-NTD & 478 \\
-27532 & 1 & ORF7a & 403 \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}
-\centering
-\caption{Most commonly observed deletions in the ARG or Li et al. (2023).} 
-\label{tab:common_dels}
-\begin{tabular}{rrcrrrr}
-\toprule
-\multicolumn{3}{c}{} & \multicolumn{2}{c}{Sc2ts ARG} & \multicolumn{2}{c}{Li et al. (2023)} \\
-\cmidrule(lr){4-5} \cmidrule(lr){6-7}
-Start & Length & Region & Count & Frequency (\%) & Count & Frequency (\%) \\
-\midrule
-28271 & 1 & non-coding & 1422469 & 57.3078 & 3853692 & 42.1183 \\
-28248 & 6 & ORF8 & 1114712 & 44.9090 & 3026186 & 33.0742 \\
-22029 & 6 & S / S1-NTD & 1114415 & 44.8970 & 3010665 & 32.9046 \\
-11288 & 4 & ORF1ab / NSP6 & 871838 & 35.1242 & N/A & N/A \\
-28362 & 9 & N & 871778 & 35.1218 & 3939164 & 43.0525 \\
-21765 & 6 & S / S1-NTD & 825628 & 33.2625 & 3179887 & 34.7541 \\
-21633 & 9 & S / S1-NTD & 532084 & 21.4364 & 1941962 & 21.2244 \\
-11292 & 5 & ORF1ab / NSP6 & 531943 & 21.4307 & N/A & N/A \\
-6513 & 3 & ORF1ab / NSP3 & 341714 & 13.7668 & 1984683 & 21.6913 \\
-11283 & 5 & ORF1ab / NSP6 & 339935 & 13.6951 & 8360 & 0.0914 \\
-21988 & 8 & S / S1-NTD & 339770 & 13.6885 & 590 & 0.0064 \\
-22194 & 3 & S / S1-NTD & 313878 & 12.6454 & 1833050 & 20.0340 \\
-11288 & 9 & ORF1ab / NSP6 & 311566 & 12.5522 & 3040576 & 33.2315 \\
-21991 & 3 & S / S1-NTD & 308800 & 12.4408 & 912944 & 9.9779 \\
-29762 & 1 & non-coding & 38030 & 1.5321 & N/A & N/A \\
-29781 & 1 & non-coding & 36112 & 1.4549 & N/A & N/A \\
-29769 & 1 & non-coding & 34315 & 1.3825 & N/A & N/A \\
-29774 & 1 & non-coding & 33503 & 1.3498 & N/A & N/A \\
-29779 & 1 & non-coding & 30515 & 1.2294 & N/A & N/A \\
-11283 & 9 & ORF1ab / NSP6 & 60 & 0.0024 & 2011934 & 21.9891 \\
-686 & 9 & ORF1ab / NSP1 & N/A & N/A & 115450 & 1.2618 \\
-21987 & 9 & S / S1-NTD & N/A & N/A & 1918094 & 20.9635 \\
+29781 & 1 & non-coding & 12754 \\
+29762 & 1 & non-coding & 11383 \\
+29779 & 1 & non-coding & 10951 \\
+29769 & 1 & non-coding & 10809 \\
+29774 & 1 & non-coding & 10735 \\
+29700 & 1 & non-coding & 8562 \\
+686 & 9 & ORF1ab / NSP1 & 6436 \\
+29614 & 1 & ORF10 & 5039 \\
+29555 & 1 & non-coding & 4814 \\
+29543 & 1 & non-coding & 4804 \\
+29541 & 1 & non-coding & 4802 \\
+203 & 1 & non-coding & 3604 \\
+222 & 1 & non-coding & 3442 \\
+21991 & 3 & S / S1-NTD & 2088 \\
+28254 & 1 & ORF8 & 1969 \\
+28093 & 1 & ORF8 & 900 \\
+21765 & 6 & S / S1-NTD & 482 \\
+27532 & 1 & ORF7a & 404 \\
+335 & 1 & ORF1ab / NSP1 & 370 \\
+337 & 1 & ORF1ab / NSP1 & 346 \\
 \bottomrule

### 2025-09-12 notebooks/nb_utils.py
Add notbook and plots for non-pangoX subgraphs
Alpha, Delta, etc.

@@ -17,10 +17,9 @@ from matplotlib import pyplot as plt
 from matplotlib.colors import rgb2hex
 import tskit_arg_visualizer as argviz
 
-TSDIR = "../data"
+DATA_DIR = Path("../data")
 
-def load(filename="sc2ts_viridian_v1.1.trees.tsz"):
-    ts = tszip.load(os.path.join(TSDIR, filename))
+def print_info(ts):
     time_zero_date = ts.metadata.get("time_zero_date", "Unknown date")
     print(
         f"Using a {ts.nbytes/1e6:0.1f} megabyte ARG up to "
@@ -32,6 +31,11 @@ def load(filename="sc2ts_viridian_v1.1.trees.tsz"):
         f"over {ts.sequence_length}bp with {n_RE} recomb. events)"
     )
     return ts
+
+def load_dataset(filename="viridian_mafft_2024-10-14_v1.vcz.zip"):
+    return sc2ts.Dataset(os.path.join(DATA_DIR, filename), date_field="Date_tree")
+
+
 
 def standard_recombinant_labels(ts, pango_x_events_file):
     """
@@ -104,9 +108,6 @@ def standard_recombinant_labels(ts, pango_x_events_file):
         labels[int(u)] = label
     
     return labels
-
-def load_dataset(filename="viridian_mafft_2024-10-14_v1.vcz.zip"):
-    return sc2ts.Dataset(os.path.join(TSDIR, filename), date_field="Date_tree")
 
 
 def date(ts, node_id):

### 2025-09-12 src/makepdf.py

@@ -24,7 +24,7 @@ def convert_file(infile, outdir):
             page = context.new_page()
             page.goto("file://" + str(infile))
             page.emulate_media(media="print")
-            page.pdf(path=outfile.with_suffix('.pdf'))
+            page.pdf(path=str(outfile) + '.pdf')
             browser.close()
 
     elif infile.suffix == ".ipynb":

### 2025-09-12 latex/main.tex
Write up the parsimony analysis

@@
 
+\subsection{Parsimony comparison of sc2ts and UShER}
+\label{sec:supplement_sc2ts_usher_parsimony}
+We performed an indepth analysis of the sc2ts and UShER ARGs 
+(we refer to the UShER tree as an ARG for simplicity here as a single 
+tree is also an ARG) in terms of parsimony.
+A direct comparison is difficult because of the different aligments 
+used as input, the different sets of sites used for inference and 
+the fact that we are including deletions for a subset of sites in
+the sc2ts ARG.
+
+When we compare the intersection of the base sc2ts ARG (pre 
+site-remapping), sc2ts is very slightly more parsimonious 
+% TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/684
+with XX mutations over the YYY sites shared between them.
+After site remapping, and including all mutations involving 
+deletion characters, sc2ts has 143,690 mutations more 
+(2,078,370 vs 1,934,680). If we exclude mutations involving 
+deletions in sc2ts, this is reduced to 46,371 more (2.3\%)
+than UShER.
+If we restrict to the 26,192 sites that were not remapped, 
+sc2ts has 4,785 (0.26\%) fewer mutations than UShER,
+and has around 2.5 times more mutations than UShER on the 
+163 sites that were post-hoc remapped.
+
+To provide a more objective measure of the overall parsimony 
+of the inferred ARGs with respect to the Viridian alignments,
+we remapped all 29903 sites in both ARGs 
+using Fitch-Hartigan parsimony. Here the nucleotides
+ACGT and ``-'' were treated as distinct characters, with all 
+other values regarded as missing data. In this case, the sc2ts 
+ARG has 629,864 (5.5\%) fewer mutations (10,840,687 
+vs 11,470,551). Some of this difference may be due to the 
+post-hoc incorporation of phylogenetic signal from deletions 
+by sc2ts, but more work is required to fully understand 
+the differences between the two approaches.
+
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_usher\_sc2ts.ipynb}{analysis\_usher\_sc2ts.ipynb}
+for details.
+
 \subsection{Analysis of Pango lineage ``origination'' events}

### 2025-09-12 latex/main.tex
Update main text

@@
 in the ARG and the UShER phylogeny (Figure~\ref{fig:phylo_backbones}).
-Next, we compared the high-level properties of the sc2ts and UShER inferences from the Viridian dataset
-by pruning both to their intersection of 2,475,418 samples and 27,431 sites.
-% TODO this is out of date and needs to be rewritten along with supplementary 
-% analysis
-Remarkably, given the differences in inference methods and data processing pipelines,
-they are almost identically parsimonious,
-with sc2ts requiring 4,208 (0.23\%) fewer mutations than UShER (1,843,132 for UShER vs 1,838,924 for sc2ts).
-We also compared the mutational spectra of major VOCs in the ARG against the mutational spectra 
+Next, we compared the sc2ts and UShER inferences 
+from the Viridian dataset
+by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
+% JK: this feels a bit wishy washy, but there really is no simple answer to
+% this and I don't want to claim we're more parsimonious.
+and found they were very similar in terms of parsimony
+when accounting for various differences in the data represented
+(Document~\ref{sec:supplement_sc2ts_usher_parsimony}).
+We also compared the mutational spectra of major VOCs in the ARG against 
+the mutational spectra 
 from the Viridian UShER phylogeny \cite{Hunt2024} and a previous study \cite{Bloom2023} (STAR Methods),

### 2025-09-12 latex/main.tex
Updates from Overleaf

--- 2025-09-12 latex/main.tex
@@
 
-\subsubsection*{Comparison with UShER}
-To facilitate comparison with the sc2ts ARG, we first converted the Viridian 
-UShER tree from MAT~\cite{McBroome2021} to tskit format, which has similar
-properties but is more general (see the section ``Ancestral Recombination
-Graphs``).
-We first downloaded the tree in JSON format (tree.all\_viridian.202409.jsonl)
-from Figshare
-and converted the ARG topology. We then downloaded the protobuf formatted
-tree (tree.all\_viridian.202409.pb.gz), and exported the mutations to 
-nucleotide format using \texttt{matUtils summary --translate}
-(usher version 0.6.3), and translated
-these mutations into tskit format. The resulting tszip file 
-(usher\_viridian\_v1.0.trees.tsz) contains 5,345,019 nodes, 3,364,841
-mutations and requires 29MB of storage space.
-We validated that the data was correctly encoded by comparing the 
-variant data for the UShER tskit tree sequence against our alignments.
-There was an exact match between the non-ambiguous nucleotide calls 
-at 27469 sites. At the remaining 39 sites there were a significant 
-number of gap characters (deletion calls), suggesting that the mismatches
-could be due to differences in the alignments used as input to UShER.
-See notebook
-\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/validate\_usher\_tree.ipynb}{validate\_usher\_tree.ipynb}
-for details.
-
-We then computed intersection ARGs for sc2ts and UShER by 
-``simplifying''\cite{Kelleher2018efficient,Wong2024} 
-both to the 2,475,418 samples and 27,507 sites that are shared.
-
-All conversions were performed as part of the reproducible 
-postprocessing pipeline described in [section?]. The full UShER
-tree and the intersection files are available for download on Zenodo
-in tszip format.
-
 \subsubsection*{Comparison of phylogenetic backbones}
@@
 The ARG captures highly prevalent deletions in SARS-CoV-2 (Table~\ref{tab:major_dels}).
-
-<TODO> Highly recurrent deletions (Table~\ref{tab:rec_dels}).
-
 Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
@@
 except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
+During site remapping, the 9-base deletion at 11283 was split into
+a 5-base deletion at 11283 (count of 339867) and a 4-base deletion at 11288 (count of 871782).
+
+<TODO> Highly recurrent deletions (Table~\ref{tab:rec_dels}).
 

### 2025-09-12 latex/main.tex
Updates from Overleaf
Add imputation analysis

--- 2025-09-12 latex/main.tex
@@
 and found they were very similar in terms of parsimony
-when accounting for various differences in the data represented
-(Document~\ref{sec:supplement_sc2ts_usher_parsimony}).
+(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
+and imputation of missing and ambiguous bases
+(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
 We also compared the mutational spectra of major VOCs in the ARG against 
@@
 
+\subsection{Imputation comparison of sc2ts and UShER}
+\label{sec:supplement_sc2ts_usher_imputation}
+Both sc2ts and UShER support missing data in input alignments, and effectively
+``hard call'' these values when the samples are integrated into the tree/ARG. 
+In doing so, we impute the missing and ambiguous data for these samples.
+We evaluate the imputation performance of sc2ts and UShER by comparing their
+outputs on the Viridian alignments.
+
+The total aligned dataset for 2,475,418 samples and 27,507 sites 
+in the sc2ts and UShER intersection ARGs represents 63.41GiB of nucleotide calls. 
+Of the 162,838,461 missing data calls (Ns) in the alignments, sc2ts and UShER
+disagreed in their imputed values for only 203,579 (0.125\%). 
+Additionally, 956,859 calls made use of the IUPAC uncertainty codes. 
+Of these sc2ts imputed 435 (0.05\%) incorrectly (i.e., 
+with a base that is not compatible with the
+ambiguity code). 
+UShER imputed only 7 calls (0.007\%) from this set incorrectly, likely 
+due to special handling of ambiguous bases in the parsimony calculations.
+
+This computation is a good example of type of integrated large-scale data analysis 
+that can be easily performed using the combination of tskit and VCF Zarr.
+The entire analysis is performed in a short Jupyter notebook, with the 
+core calculation requiring about 30 lines of Python code. Retrieving 
+data from the base alignments and ARGs for 2.48 million samples 
+at the same time over all 27,507 sites took only 6 minutes, with a peak memory
+usage of about 5GiB of RAM on a standard desktop computer.
+
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/example\_data\_processing.ipynb}{example\_data\_processing.ipynb}
+for details.
+
 \subsection{Analysis of Pango lineage ``origination'' events}

### 2025-09-12 latex/main.tex
Incoroporate the tanglegrams

@@
 First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
-by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
-Tanglegrams reveal that the ARG captures known phylogenetic relationships
-both within sublineages and at the level of major lineages (including VOCs),
-with good agreement between the phylogenetic bipartitions
-in the ARG and the UShER phylogeny (Figure~\ref{fig:phylo_backbones}).
+by pruning both down to representative samples for 1097 deeply-sampled 
+lineages (STAR Methods).
+Tanglegrams show that the ARG closely matches the phylogenetic
+structure of the UShER tree (Figures~\ref{fig:tanglegram_base},
+\ref{fig:tanglegram_delta},
+\ref{fig:tanglegram_ba2},
+\ref{fig:tanglegram_ba5}).
 Next, we compared the sc2ts and UShER inferences 
@@
 % Referenced by the main Results.
-The ARG captures highly prevalent deletions in SARS-CoV-2 (Table~\ref{tab:major_dels}).
-
-<TODO> Highly recurrent deletions (Table~\ref{tab:rec_dels}).
+The ARG captures important deletion events (Table~\ref{tab:major_dels}) and highly recurrent deletion events (Table~\ref{tab:rec_dels}).
 
@@
 \centering
-\caption{Major deletion events in the ARG.}
+\caption{Major deletion events in the ARG.
+These events are inherited by at least 10,000 samples in the ARG.
+Count is the number of samples inheriting a deletion.
+The Pango lineage assigned to the node above which an event occurs is shown.
+Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in Subunit 1.}
 \label{tab:major_dels}
@@
 \centering
-\caption{Highly recurrent deletions.}
+\caption{Highly recurrent deletions.
+These are the top 20 deletions in terms of the number of occurrences in the ARG.
+Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in Subunit 1.}
 \label{tab:rec_dels}
@@
 \centering
-% \includegraphics{sc2ts_methodology}
+\centering
+\includegraphics[width=\linewidth]{tanglegram_base_tree.pdf}
 \caption{
-Phylogenetic backbones: TODO
+Tanglegram comparing phylogenetic backbones of sc2ts and UShER.
 }
-\label{fig:phylo_backbones}
+\label{fig:tanglegram_base}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{tanglegram_Delta_subtree.pdf}
+\caption{
+Tanglegram comparing sc2ts and UShER on the Delta subtree.
+}
+\label{fig:tanglegram_delta}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{tanglegram_BA.2_subtree.pdf}
+\caption{
+Tanglegram comparing sc2ts and UShER on the BA.2 subtree.
+}
+\label{fig:tanglegram_ba2}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{tanglegram_BA.5_subtree.pdf}
+\caption{
+Tanglegram comparing sc2ts and UShER on the BA.5 subtree.
+}
+\label{fig:tanglegram_ba5}
+\end{figure}
+
+\begin{figure}
+\centering
+\centering
+\includegraphics[width=\linewidth]{subgraph-Alpha.pdf}
+\caption{
+Subgraph of Alpha
+}
+\label{fig:subgraph_alpha}
+\end{figure}
+
+\begin{figure}
+\centering
+\centering
+\includegraphics[width=\linewidth]{subgraph-Delta.pdf}
+\caption{
+Subgraph of Delta
+}
+\label{fig:subgraph_delta}
+\end{figure}
+
+\begin{figure}
+\centering
+\centering
+\includegraphics[width=\linewidth]{subgraph-BA.1+BA.2.pdf}
+\caption{
+Subgraph of BA.1 and BA.2
+}
+\label{fig:subgraph_ba1_ba2}
 \end{figure}

### 2025-09-12 latex/main.tex
Update the text to incorporate subgraphs.

@@
 \centering
-\centering
-\includegraphics[width=\linewidth]{subgraph-Alpha.pdf}
+\includegraphics[width=0.85\linewidth]{subgraph-Alpha.pdf}
 \caption{
-Subgraph of Alpha
+Subgraph of Alpha. [Some notes to help the reader know this 
+reflects known events]
 }
@@
 \centering
-\centering
-\includegraphics[width=\linewidth]{subgraph-Delta.pdf}
+\includegraphics[width=0.85\linewidth]{subgraph-Delta.pdf}
 \caption{
-Subgraph of Delta
+Subgraph of Delta. [Some notes to help the reader know this 
+reflects known events]
 }
@@
 \centering
-\centering
-\includegraphics[width=\linewidth]{subgraph-BA.1+BA.2.pdf}
+\includegraphics[width=0.85\linewidth]{subgraph-BA.1+BA.2.pdf}
 \caption{
-Subgraph of BA.1 and BA.2
+Subgraph of BA.1 and BA.2.
+[Some notes to help the reader know this 
+reflects known events]
 }
@@
 \centering
-%TODO fix this to make the figures smaller.
 \includegraphics[width=\linewidth]{mutational_spectra}

### 2025-09-12 latex/main.tex
Initial pass on tables for copying patterns

--- 2025-09-12 latex/main.tex
@@
 \usepackage{booktabs}
+\usepackage{tabularray}
 \usepackage{amsmath}
@@
 % See https://github.com/jeromekelleher/sc2ts-paper/issues/628
-% Referenced by the main Results.
-The ARG captures important deletion events (Table~\ref{tab:major_dels}) and highly recurrent deletion events (Table~\ref{tab:rec_dels}).
+By mapping deletions to the ARG, we identified
+important deletion events (Table~\ref{tab:major_dels}),
+many of which are associated with the origins of major VOCs,
+for example, the deletions in ORF1ab and
+S1-NTD (the N-terminal domain of Spike)
+associated with the Alpha lineage.
+We also identified highly recurrent deletions (Table~\ref{tab:rec_dels}),
+for example, two known recurrent deletions in S1-NTD (REF).
 
@@
 The Pango lineage assigned to the node above which an event occurs is shown.
-Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in Subunit 1.}
+Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in S1 subunit.}
 \label{tab:major_dels}
@@
 These are the top 20 deletions in terms of the number of occurrences in the ARG.
-Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in Subunit 1.}
+Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in S1 subunit.}
 \label{tab:rec_dels}
@@
 \label{sec:supplement_sc2ts_usher_parsimony}
+
 We performed an indepth analysis of the sc2ts and UShER ARGs 
@@
 
+\newcommand{\cprheight}{5.0em}
+
+\begin{figure}
+\centering
+\begin{tblr}{Q[l,t]Q[m]}
+  % cells={valign=m,halign=c},
+  % rowsep=2pt,
+  % colspec={Q[r,b]Q[m]},   
+% }
+% \begin{tabular}{ll}
+A & \includegraphics[height=\cprheight]{static/causedby-ERR10306608.png}\\
+B & \includegraphics[height=\cprheight]{static/causedby-ERR6760245.png}\\
+C & \includegraphics[height=\cprheight]{static/causedby-SRR15864165.png}\\
+D & \includegraphics[height=\cprheight]{static/causedby-SRR21087725.png}\\
+\end{tblr}
+% \end{tabular}
+\caption{}
+\label{fig:copying_pattern_examples}
+\end{figure}
+
+\renewcommand{\cprheight}{1.5em}
+% Add some more space between rows. Note this it would be sligntly nicer if the 
+% label was aligned vertically in the center of the row, but this is good
+% enough for now.
+\renewcommand{\arraystretch}{2}
+
+\begin{figure}
+\centering
+\begin{tabular}{ll}
+XC & \includegraphics[height=\cprheight]{static/XC.png}\\
+XBR & \includegraphics[height=\cprheight]{static/XBR.png}\\
+XA & \includegraphics[height=\cprheight]{static/XA.png}\\
+XS & \includegraphics[height=\cprheight]{static/XS.png}\\
+XL & \includegraphics[height=\cprheight]{static/XL.png}\\
+XBG &\includegraphics[height=\cprheight]{static/XBG.png}\\
+XBD & \includegraphics[height=\cprheight]{static/XBD.png}\\
+XQ & \includegraphics[height=\cprheight]{static/XQ++.png}\\
+XBB & \includegraphics[height=\cprheight]{static/XBB.png}\\
+XM & \includegraphics[height=\cprheight]{static/XM+XAL.png}\\
+XBF & \includegraphics[height=\cprheight]{static/XBF.png}\\
+XF & \includegraphics[height=\cprheight]{static/XF.png}\\
+XY & \includegraphics[height=\cprheight]{static/XY.png}\\
+XG & \includegraphics[height=\cprheight]{static/XG.png}\\
+XW & \includegraphics[height=\cprheight]{static/XW.png}\\
+Xx & \includegraphics[height=\cprheight]{static/XZ++.png}\\
+XE/XH  & \includegraphics[height=\cprheight]{static/XE+XH.png}\\
+XBH & \includegraphics[height=\cprheight]{static/XBH.png}\\
+XBM & \includegraphics[height=\cprheight]{static/XBM.png}\\
+XJ & \includegraphics[height=\cprheight]{static/XJ.png}\\
+\end{tabular}
+\caption{Copying patterns for the 20 recombination events associated with Pango 
+X lineages listed in Table~\ref{tab:pango_x_lineages} (in the same order)}
+\label{fig:pango_x_copying_patterns}
+\end{figure}
+
+% Reset this in case we make any more tables beyond this
+\renewcommand{\arraystretch}{1}
+
 \begin{figure}

### 2025-09-12 latex/main.tex
Updates from Overleaf

--- 2025-09-12 latex/main.tex
@@
 % Referenced by the main Results.
-The ARG captures important deletion events (Table~\ref{tab:major_dels}) and highly recurrent deletion events (Table~\ref{tab:rec_dels}).
-
-Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
-using a global GISAID dataset (sampled up to August, 2022).
-We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
-which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
-All of these indels are deletions and have high frequencies in the ARG
-(sequence count divided by the total number of samples),
-except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
+The ARG captures important deletion events (Table~\ref{tab:major_dels}), many of which are associated with the origins of major VOCs, for example, the deletions in ORF1ab and the N terminal domain of the S1 subunit associated with the Alpha lineage.
+and highly recurrent deletion events (Table~\ref{tab:rec_dels}).
 
@@
 The Pango lineage assigned to the node above which an event occurs is shown.
-Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in Subunit 1.}
+Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N terminal domain in S1 subunit.}
 \label{tab:major_dels}
@@
 These are the top 20 deletions in terms of the number of occurrences in the ARG.
-Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in Subunit 1.}
+Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N terminal domain in S1 subunit.}
 \label{tab:rec_dels}

### 2025-09-13 latex/main.tex
Deletion updates from Shing

@@
 % See https://github.com/jeromekelleher/sc2ts-paper/issues/628
-% Referenced by the main Results.
-The ARG captures important deletion events (Table~\ref{tab:major_dels}), many of which are associated with the origins of major VOCs, for example, the deletions in ORF1ab and the N terminal domain of the S1 subunit associated with the Alpha lineage.
-and highly recurrent deletion events (Table~\ref{tab:rec_dels}).
+By mapping deletions to the ARG, we identified
+important deletion events (Table~\ref{tab:major_dels}),
+many of which are associated with the origins of major VOCs,
+for example, the deletions in ORF1ab and
+S1-NTD (the N-terminal domain of Spike)
+associated with the Alpha lineage.
+We also identified highly recurrent deletions (Table~\ref{tab:rec_dels}),
+for example, two known recurrent deletions in S1-NTD (REF).
+
+Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
+using a global GISAID dataset (sampled up to August, 2022).
+We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
+which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
+All of these indels are deletions and have high frequencies in the ARG
+(sequence count divided by the total number of samples),
+except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
 
@@
 The Pango lineage assigned to the node above which an event occurs is shown.
-Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N terminal domain in S1 subunit.}
+Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in S1 subunit.}
 \label{tab:major_dels}
@@
 These are the top 20 deletions in terms of the number of occurrences in the ARG.
-Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N terminal domain in S1 subunit.}
+Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in S1 subunit.}
 \label{tab:rec_dels}

### 2025-09-13 latex/main.tex
Fixup table alignment

@@
 
+\newcommand{\cprheight}{4.0em}
+
+\begin{figure}
+\centering
+% NOTE the size of the second "m" column doesn't matter, but we have to fix the
+% size in order for vertical centering to work :sigh:
+% The layout here isn't ideal, but it's incredibly fiddly to get center
+% alignment veritically with cells to work, so this is good enough.
+\begin{tabularx}{0.85\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
+% FIXME: these samples are chosen randomly and so almost certainly do not
+% belong to the correct quadrants.
+A & \includegraphics[height=\cprheight]{static/causedby-ERR10306608.png}\\
+B & \includegraphics[height=\cprheight]{static/causedby-ERR6760245.png}\\
+C & \includegraphics[height=\cprheight]{static/causedby-SRR15864165.png}\\
+D & \includegraphics[height=\cprheight]{static/causedby-SRR21087725.png}\\
+\end{tabularx} 
+\caption{Examples of problematic copying patterns. A-D are examples drawn from
+the labelled quadrants in Figure~\ref{fig:recombinant_qc}.}
+\label{fig:copying_pattern_examples}
+\end{figure}
+
+\renewcommand{\cprheight}{1.5em}
+% Add some more space between rows. Note this it would be sligntly nicer if the 
+% label was aligned vertically in the center of the row, but this is good
+% enough for now.
+\renewcommand{\arraystretch}{2}
+
+\begin{figure}
+\centering
+\begin{tabularx}{\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
+XC & \includegraphics[height=\cprheight]{static/XC.png}\\
+XBR & \includegraphics[height=\cprheight]{static/XBR.png}\\
+XA & \includegraphics[height=\cprheight]{static/XA.png}\\
+XS & \includegraphics[height=\cprheight]{static/XS.png}\\
+XL & \includegraphics[height=\cprheight]{static/XL.png}\\
+XBG &\includegraphics[height=\cprheight]{static/XBG.png}\\
+XBD & \includegraphics[height=\cprheight]{static/XBD.png}\\
+XQ & \includegraphics[height=\cprheight]{static/XQ++.png}\\
+XBB & \includegraphics[height=\cprheight]{static/XBB.png}\\
+XM & \includegraphics[height=\cprheight]{static/XM+XAL.png}\\
+XBF & \includegraphics[height=\cprheight]{static/XBF.png}\\
+XF & \includegraphics[height=\cprheight]{static/XF.png}\\
+XY & \includegraphics[height=\cprheight]{static/XY.png}\\
+XG & \includegraphics[height=\cprheight]{static/XG.png}\\
+XW & \includegraphics[height=\cprheight]{static/XW.png}\\
+Xx & \includegraphics[height=\cprheight]{static/XZ++.png}\\
+XE/XH  & \includegraphics[height=\cprheight]{static/XE+XH.png}\\
+XBH & \includegraphics[height=\cprheight]{static/XBH.png}\\
+XBM & \includegraphics[height=\cprheight]{static/XBM.png}\\
+XJ & \includegraphics[height=\cprheight]{static/XJ.png}\\
+\end{tabularx}
+\caption{Copying patterns for the 20 recombination events associated with Pango 
+X lineages listed in Table~\ref{tab:pango_x_lineages} (in the same order)}
+\label{fig:pango_x_copying_patterns}
+\end{figure}
+
+% Reset this in case we make any more tables beyond this
+\renewcommand{\arraystretch}{1}
+
 \begin{figure}

### 2025-09-13 src/plot.py
Remove main_arg function from plotting

@@ -10,12 +10,6 @@ import tszip
 import sc2ts
 
 
-def main_arg():
-    return tszip.load(
-        "data/sc2ts_v1_2023-02-21_pp_dated_remapped_bps_pango_mmps.trees.tsz"
-    )
-
-
 def samples_csv():
     return pd.read_csv("arg_postprocessing/sc2ts_v1_2023-02-21_samples.csv")
 

### 2025-09-15 latex/main.tex

@@
 
+%%% The summary (abstract) should consist of a single paragraph of 150 words or
+%%% fewer.
+
 \section*{SUMMARY}
 
-%%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.
-%%% This is the SMBE abstract. We can rewrite or work off of it, but it needs to be shorter.
-Widely used phylogenetic methods to study SARS-CoV-2 evolution assume that genetic recombination is negligible and
-viral relationships can be represented by a single tree.
-However, important SARS-CoV-2 recombinant lineages have emerged.
-A more suitable approach is to use ancestral recombination graphs (ARGs),
-which jointly model recombination, mutation, and genetic inheritance in a cohesive framework.
-Recent computational advances have made ARG inference possible for datasets of millions of genome sequences.
-Building on these new advances, we have developed sc2ts, a scalable method to simultaneously detect recombinants and
-reconstruct ARGs from heterochronous SARS-CoV-2 genome sequences.
-Using sc2ts, we inferred an ARG of 2.48 million genomes sampled from January, 2020 to February, 2023.
-We find that this pandemic-scale ARG unifies known features of SARS-CoV-2 evolution —
-the backbone phylogeny, mutational spectra, and major recognized recombinants.
-This ARG enables us to explore the evolutionary origins of known and novel recombinants.
-Furthermore, the ARG is compactly stored in the tree sequence format using the tskit library and
-can be analyzed efficiently and interactively using standard Python tools on a typical laptop.
-This combination of capabilities makes sc2ts a powerful tool
-for studying the recombination and mutation processes driving SARS-CoV-2 evolution.
+% NOTE: currently 155 words
+
+Millions of SARS-CoV-2 genome sequences were collected in
+response to the COVID-19 pandemic, forming a dataset of unprecedented
+depth and richness.
+Estimated genealogies are fundamental to
+gleaning information from this ocean of data, 
+and form the primary input to many downstream analyses.
+A basic assumption of methods to infer genealogies from viral 
+genetic sequence data is that recombination is negligible and 
+the genealogy is a tree.
+Recombinant lineages have risen to global prevalance in SARS-CoV-2,
+and therefore such single-tree representations are necessarily
+incomplete and potentially misleading.
+We present sc2ts, a method to infer reticulate genealogies for SARS-CoV-2
+that automaticaly detects and incorporates recombination, 
+and apply it to 2.48 million genomes.
+We show that the inferred genealogy captures a broad range of 
+known evolutionary features and provides novel 
+insights into the relationships between previously identified
+recombinant lineages.
+Together, the inference method, inferred genealogy and 
+suite of open-source analysis tools, 
+represent a powerful resource for future research.
 

### 2025-09-15 latex/main.tex
Tweak abstract

@@
 
-% NOTE: currently 155 words
+% NOTE: currently 160 words
 
@@
 the genealogy is a tree.
-Recombinant lineages have risen to global prevalance in SARS-CoV-2,
+Recombinant lineages have risen to global prevalence in SARS-CoV-2,
 and therefore such single-tree representations are necessarily
@@
 We present sc2ts, a method to infer reticulate genealogies for SARS-CoV-2
-that automaticaly detects and incorporates recombination, 
-and apply it to 2.48 million genomes.
-We show that the inferred genealogy captures a broad range of 
+that automatically detects and incorporates recombination 
+in real time.  
+We use sc2ts to infer a genealogy of 2.48 million genomes,
+and show that it captures a broad range of 
 known evolutionary features and provides novel 

### 2025-09-15 latex/main.tex
Update fig 1 caption
Closes #666

@@
 Schematic overview of sc2ts.
-(A) Procedure to build an ARG by incrementally updating it with samples sorted by collection dates.
+(A) Process to build an ARG by incrementally updating it with samples sorted by collection dates.
+(B) The ARG resulting from the inference process.
 Sample 0 is used to initialize the ARG as the root.
@@
 with Sample 4 itself attached to this node as its child.
-The resulting ARG contains two local trees,
-one on each side of the inferred breakpoint.
-(B) Copying paths for Sample 4 inferred under the LS model
+(C) Two local trees contained in the resulting ARG,
+one on each side of the inferred breakpoint (A), which occurs at position 50.
+The genomic coordinates of the segments inherited from different recombinant parents
+are denoted by two half-open intervals (B).
+(D) Copying paths for Sample 4 inferred under the LS model
 while allowing recombination ($k = 4$) or disallowing it ($k = 1000$).
@@
 % belong to the correct quadrants.
-A & \includegraphics[height=\cprheight]{static/causedby-ERR10306608.png}\\
-B & \includegraphics[height=\cprheight]{static/causedby-ERR6760245.png}\\
-C & \includegraphics[height=\cprheight]{static/causedby-SRR15864165.png}\\
-D & \includegraphics[height=\cprheight]{static/causedby-SRR21087725.png}\\
+Q1 & \includegraphics[height=\cprheight]{static/causedby-ERR10306608.png}\\
+Q2 & \includegraphics[height=\cprheight]{static/causedby-SRR15864165.png}\\
+Q3 & \includegraphics[height=\cprheight]{static/causedby-ERR6760245.png}\\
+Q4 & \includegraphics[height=\cprheight]{static/causedby-SRR21087725.png}\\
 \end{tabularx}
 \caption{Examples of problematic copying patterns. A-D are examples drawn from
-the labelled quadrants in Figure~\ref{fig:recombinant_qc}.}
+the labeled quadrants in Figure~\ref{fig:recombinant_qc}.}
 \label{fig:copying_pattern_examples}

### 2025-09-15 latex/main.tex
Updates from Overleaf
Fix fig reference

@@
 \centering
-\includegraphics[width=\linewidth]{robust_vs_artificial_classification}
+\includegraphics[width=\linewidth]{figures/recombinant_qc}
 \caption{
@@
 \begin{tabularx}{0.85\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
-% FIXME: these samples are chosen randomly and so almost certainly do not
-% belong to the correct quadrants.
 Q1 & \includegraphics[height=\cprheight]{static/causedby-ERR10306608.png}\\
@@
 \end{tabularx}
-\caption{Examples of problematic copying patterns. A-D are examples drawn from
-the labeled quadrants in Figure~\ref{fig:recombinant_qc}.}
+\caption{
+Examples of copying patterns.
+Q1 to Q4 are examples drawn from
+the labeled quadrants in Figure~\ref{fig:recombinant_qc}.
+Q2 illustrates a marginally good copying pattern.
+Q1, Q3, and Q4 illustrate copying patterns that fail the QC thresholds.
+}
 \label{fig:copying_pattern_examples}

### 2025-09-15 latex/main.tex
Start on finalising mapping details

@@
 that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
-Each node in the ARG was then assigned to one of 2026 Pango lineages using pangolin (STAR Methods).
-All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment
+Each node in the ARG was then assigned to one of 2058 Pango lineages using pangolin (STAR Methods).
+All but 21,680 sampled sequences (0.87\%) agreed with the source sequences in terms of Pango lineage assignment
 (comparable to the 0.27\% disagreement between pangolin-data 
 versions 1.21 and 1.29 in the source Viridian metadata).
-A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG,
+A majority of these lineages (1473 of 2058) have monophyletic origins within the ARG,
 increasing to 1779 (86\%) when we allow for multiple sibling origination 
@@
 shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
-An example of a set of ``nested'' closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
+An example of a set of ``nested'' closely related lineages consists of XZ, XAC, XAD, XAE, and XAP,
 which descend from a single recombination event rather than five independent
@@
 in next draft.]
-See Document~\ref{sec:supplement_pango_lineages} for extended analyses,
+See Document~\ref{sec:supplement_indepth_pango_x_lineages} for extended analyses,
 Figure~\ref{fig:pango_x_copying_patterns} for copying patterns,
@@
 XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
-XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
 XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
 XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
-XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
+XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
 XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
 XAL:3} \\
+XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {} \\
+XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
+XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
+XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
 XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
-XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
-XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
+XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
 XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
-XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
 \midrule
@@
 \midrule
-Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 &
-{BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
+Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 & {BA.2:156,
+XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
 XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
 {XE:1116, BA.2:37, XH:2, XAF:1} \\
-XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {BA.2:4, XBH:2} \\
 XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
-BF.3:2} \\ XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
+BF.3:2} \\
+XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
 XAF:1} \\
@@
 XR & XQ & 3 & 11 & 4 & 17 \\
-XU & XQ & 6 & 103 & 4 & 1 \\
-XAM & XQ & 6 & 43 & 3 & 21 \\
+XU & XQ & 6 & 103 & 13 & 1 \\
+XAM & XQ & 6 & 40 & 3 & 21 \\
 XAA & XQ & 6 & 33 & 3 & 17 \\
-XAG & XQ & 6 & 55 & 4 & 6 \\
-XAP & Xx & 4 & 67 & 1 & 20 \\
+XAG & XQ & 6 & 53 & 4 & 6 \\
+XAP & Xx & 4 & 66 & 1 & 20 \\
 XZ & Xx & 4 & 48 & 1 & 48 \\
@@
 XAF & XAF & 5 & 140 & 8 & 1 \\
-XAL & XM & 3 & 66 & 2 & 3 \\
+XAL & XM & 3 & 65 & 2 & 3 \\
 \midrule
@@
 \midrule
-Sc2ts & 14 / 15 (93.3) & 13 / 15 (86.7) & 540.5 (315, 766)\\
-RecombinHunt-GISAID & 13 / 13 (100.0) & 10 / 13 (76.9) & 764.0 (4, 1202)\\
-RecombinHunt-Nextstrain & 12 / 14 (85.7) & 11 / 14 (78.6) & 1160.0 (559, 2785)\\
+Sc2ts & 14 / 16 (87.5) & 14 / 16 (87.5) & 540.5 (315, 766)\\
+RecombinHunt-GISAID & 13 / 15 (86.7) & 10 / 15 (66.7) & 764.0 (4, 1202)\\
+RecombinHunt-Nextstrain & 12 / 15 (80.0) & 11 / 15 (73.3) & 1160.0 (559, 2785)\\
 CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
@@
 Examples of highly plausible recombinants are XA and XBB
-(for an extended analysis, see Document~\ref{sec:supplement_pango_lineages}).
+(for an extended analysis, see Document~\ref{sec:supplement_indepth_pango_x_lineages}).
 The ``copying pattern'' of XA visually confirms
@@
 \begin{itemize}
+\item Snakemake workflow for aligning and converting the Viridian dataset to
+VCF Zarr:
+\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/viridian\_dataset/}
+\end{itemize}
+
+TODO add these to the list above with actual URLs
+\begin{itemize}
     \item The ARG and the sequence alignments have been deposited at Zenodo and are publicly available.
@@
 
-\subsection*{Key resources table}
-
-\textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}
+% \subsection*{Key resources table}
+
+% \textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}
 
@@
 such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
-and Pandas ,
+and Pandas~\cite{Mckinney2010data} ,
 using pyfaidx\cite{Shirley2015efficient} to access alignment data
+and Numba~\cite{Lam2015numba} to accelerate certain alignment processing
+tasks.
 [TODO JK to fill this out]
@@
 Tskit is widely used in population genetics, and has been incorporated into many downstream applications
-(e.g., \cite{Haller2019,Speidel2019,TerasakiHart2021,Fan2022,Korfmann2022,Mahmoudi2022,Petr2022,Rasmussen2022,Zhang2023}).
+(e.g., \cite{Haller2019,Speidel2019,TerasakiHart2021,Fan2022,Korfmann2023,Mahmoudi2022,Petr2023,Rasmussen2023,Zhang2023}).
 
@@
 
-\subsubsection*{Data preprocessing}
-
-In this study, we used the SARS-CoV-2 sequences reassembled in a recent study \cite{Hunt2024}
-using an amplicon-aware assembly method (named ``Viridian'').
-The Viridian sequences contain far fewer artifacts compared to their publicly deposited counterparts on GenBank.
-Consequently, phylogenetic tree inference using the Viridian data led to more parsimonious phylogenies,
-as seen by a large reduction in the number of likely artifactual reversions \cite{Hunt2024}.
-
-We downloaded 4,484,157 Viridian sequences obtained
-from samples collected from January, 2020 to March, 2023 (``batch 1'').
-We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using MAFFT v7.525 \cite{Katoh2013},
-with the option ‘--keeplength’ to retain the reference coordinate system, discarding insertions.
-We converted the MAFFT alignments and all sample metadata to VCF Zarr format \cite{Czech2025},
-which are available for download on Zenodo (URL).
-
-Prior to ARG inference, we preprocessed the remaining sequences as follows.
-We treated low-quality bases (represented by Ns), gaps and ambiguous characters in the sequences as missing data.
-When encoded as missing data, the bases do not inform sample matching (see ``The Li and Stephens model'').
-
-During inference, we filtered out the samples with low-quality sequences and low-precision sampling dates.
-We kept the quality check-passing sequences used to infer the Viridian UShER phylogeny \cite{Hunt2024}.
-We skipped low-quality sequences that had more than 500 missing sites
-(except when building the ``base'' ARG, when the maximum number of allowed missing bases was set to 10,000 instead).
-We also skipped $\sim$40,000 samples which had a collection date of December 31, 2020.
-The sampling frequency of this date is alarmingly high compared to other dates (at most 15,000 samples per day).
-We suspect that this batch of samples might be unusually enriched for wrong collection dates,
-likely due to arbitrary assignment of the originally reported collection date of ``2020'' to be ``31-12-2020''.
-Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
+\subsubsection*{Pandemic scale alignment storage}
+To provide the efficient access to alignment data required for inference with sc2ts, we
+developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
+Briefly, Zarr is a storage format for scientific data,
+originally developed for the
+\textit{Anopheles gambiae}
+1000 Genomes Project~\cite{Anopheles2017genetic}, and
+is now seeing widespread adoption across the 
+sciences~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
+Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
+Zarr is essentially a simple mechanism for storing n-dimensional array
+data as a regular grid of compressed chunks, which provides 
+high levels of data compression, efficient access across multiple
+dimensions, and is ideally suited to modern cloud-based
+deployments~\cite{Czech2025}.
+
+We converted the Viridian v05 dataset\cite{Hunt2024}, consisting of 4,484,157 SARS-CoV-2
+consensus sequences (125GiB over 48 FASTA files) and 30 metadata fields (1.4
+GiB TSV file) to VCF Zarr using a 
+reproducible Snakemake\cite{Koster2012snakemake} pipeline.
+We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using 
+MAFFT v7.475 \cite{Katoh2013},
+with the option \texttt{--keeplength} to retain the reference coordinate system.
+We then used \texttt{sc2ts import-alignments} and 
+\texttt{sc2ts import-metadata} to convert all alignments and metadata to VCF
+Zarr format, resulting in a single 401 MiB file 
+(around 77X smaller than gzip compressed FASTA\cite{Czech2025}).
+
+A classical tradeoff in bioinformatics is whether to store data 
+in sample-major (FASTA) or variant-major (VCF) format, determining
+whether the data can be accessed efficiently either by row or column
+of the variant matrix. Because VCF Zarr is chunked in two dimensions,
+we can now access either sample alignments or the variation
+data for a given site across all samples efficiently.
+While the VCF Zarr encoding of the MAFFT-aligned Viridian sequences
+and metadata can be accessed via native Zarr libraries in several
+popular programming languages (see Czech et al.\cite{Czech2025} 
+for an overview), the sc2ts Python library provides a convenient interface. 
+The \texttt{sc2ts.Dataset}
+class provides efficient access to alignments for samples and 
+variant data for sites along the genome, as well as methods to export
+subsets of the data to FASTA if required.
+Document~\ref{sec:supplement_sc2ts_usher_imputation} gives an example
+of using this interface, where we retrieve alignment data for millions 
+of samples in a few minutes on a standard desktop computer with minimal
+memory overhead. See Czech et al.\cite{Czech2025} for more details 
+and benchmarks.
+
+Both the pipeline and converted Viridian dataset are available for download;
+see Data and code availability for details.
+
+\subsubsection*{Sample filtering and preprocessing}
+The collection date for a sequence is an important piece of information
+for sc2ts. We used the Viridian
+consensus date (``Date\_tree``), based on information from COG-UK, GISAID, and
+ENA/SRA as our source. Samples that have missing or incomplete dates (e.g., 2020-01)
+were omitted, as were
+the $\sim$40,000 samples which had a collection date of December 31, 2020.
+As the sample count for this date is a major outlier for this period of
+the pandemic, it seems likely that this 
+batch is highly enriched for incorrect collection dates.
+Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
 confirmed that some of the samples were not collected in 2020.
 
-Sites which are enriched for systematic sequencing errors or frequently experience homoplasy
-can cause artifactual signals of recombination (REF).
-The bioinformatics community has identified and curated a set of problematic sites
-that are recommended to be excluded before phylogenetic inference (REF).
-Here we assembled a bespoke set of such problematic sites
-using the information in the Viridian sequences.
+The bioinformatics community has curated a set of problematic sites
+for SARS-CoV-2 that are enriched for systematic sequencing errors or 
+high levels of homoplasy~\cite{DeMaio2020issues}. 
+Although these sites are recommended to be masked for UShER
+inference~\cite{Turakhia2020},
+the list has not been updated to take into account the systematic improvements 
+made by the Viridian dataset, and we therefore assembled a custom set of sites to mask. 
 To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
 without any site masking.
-Then, we identified the top 100 sites which had the highest mutation counts in the ARG (URL).
-Only six of these sites overlapped with the ``Problematic Sites''
-downloaded from the UCSC Genome Browser website (URL),
-which did not have a high number of mutations in the preliminary ARG.
-To infer the final ARG, we masked the problematic sites
-in our custom set
-\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}.
+Then, we identified the top 100 sites in terms of mutation counts in the ARG.
+Only 11 of these sites 
+(635, 8835, 11074, 11083, 15521, 16887, 21304, 21305, 21575, 21987 and 28253)
+were present in the problematic sites file
+(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}),
+and most sites in this list were typical in terms of the numbers of mutations.
+We therefore masked our 100 most-mutagenic sites to remove sequencing errors and improve
+matching performance in the LS HMM.
+
+Alignments read from the VCZ Zarr file were first preprocessed to mask any 
+non nucleotide characters (A,C,G or T) as missing data. 
+Then, we filtered 
+any samples that had missing data at more than 500 sites, excluding
+the 100 sites identified as problematic (this was 
+set to 10,000 sites for the initial period up to 2020-03-01 where sampling
+density is low).
 
@@
 
-\subsubsection*{Identification of robust recombination events}
+\subsubsection*{Post hoc parsimony mapping}
+After the main inference step we mapped the variation data for 163 sites 
+in the alignments to the ARG using the Fitch-Hartigan 
+algorithm\cite{Fitch1971toward,Hartigan1973minimum} 
+implemented in tskit. In contrast to the main inference process where 
+deletion characters are masked as missing data, we allow for 
+5 different states in these parsimony calculations: A,C,G,T and -.
+By doing this, we can approximate the mutation events causing
+deletions, but with the clear caveat that each site is treated 
+independently. 
+Using data provided by Li et al.\cite{Li2023indels} (Table S1) 
+we chose 75 sites to remap
+
+performed a large-scale analysis of indels
+using a global GISAID dataset (sampled up to August, 2022).
+We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
+which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
+All of these indels are deletions and have high frequencies in the ARG
+(sequence count divided by the total number of samples),
+except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
+
+Based on the data of Table S4 of Li2023indels
+
+\subsubsection*{Quality control of recombinants}
 
@@
 Reference Figure~\ref{fig:recombinant_qc} for the scatter plot and 
-labelled quadrants, and Figure~\ref{fig:copying_pattern_examples}
+labeled quadrants, and Figure~\ref{fig:copying_pattern_examples}
 for examples of the corresponding copying patterns. Also reference
@@
 
-% This para was *{Counting supporting loci for recombinant parents}
-
-When counting such loci, we considered nearly adjacent sites (separated by at
-most three bases) as single loci, to avoid over-counting support from
-correlated loci, which would lead to artificial signals of recombination. Also,
-when counting supporting loci for a focal parent, we penalized loci where the
-character(s) in the recombinant matches that in the non-focal parent but not in
-the focal parent.
-
-Before characterizing recombination events in detail, we sought to remove
-probable artifacts. Systematic errors in SARS-CoV-2 sequences associated with
-sequencing protocols have been documented (Turakhia et al., 2022), which can
-confound recombination detection (De Maio et al., 2024). By inspecting copying
-patterns at sites that differ between recombinant parents, we find a convincing
-signal when treating adjacent and near-adjacent sites as a single locus.
-Artifactual recombination events can be characterized as having fewer than four
-such loci supporting the left or right parent on either side of a suggested
-breakpoint (STAR Methods). Of the 929 recombination events in the ARG, this
-measure identifies 543 as artifactual, leaving 386 ``robust'' events (Figure
-S1A). The copying patterns depicted in Figure S1B illustrate the genomic
-evidence for three recombination events identified as artifactual, compared to
-a marginally robust one.
-
-Evidence that our simple filter effectively flags artifacts can be seen in the
-clear association between the AmpliSeq V1 kit and recombination events that we
-identify as artifactual. This primer scheme was used to sequence XX of the
-samples associated with an artifactual recombination event, in contrast to 15
-of the ``robust'' events (Figure 2A), strongly implicating it as a common source
-of problematic data. Through further inspection, we observed that many
-artifactual recombination events contain stretches of missing data
-characteristic of amplicon dropouts (REF).
+Before characterizing recombination events in detail,
+we sought to remove probable artifactual recombinants
+due to correlated bases
+being treated as individual sites by the HMM.
+Correlated bases can be caused by
+systematic errors in genome sequences
+associated with sequencing protocols \cite{Turakhia2020},
+multi-nucleotide substitutions \cite{DeMaio2024},
+genuine deletions, and
+alignment errors (e.g., deletions due to MAFFT misalignment \cite{Hunt2024}).
+
+We characterized artifactual recombination events
+as having fewer than four such loci supporting the left or right parent
+on either side of a suggested breakpoint.
+We treated adjacent and near-adjacent sites
+(separated by at most three bases) as single loci.
+Also, when counting loci supporting a focal parent,
+we penalized loci where the character(s)
+in the recombinant matches that in the non-focal parent
+but not in the focal parent.
+
+Of the 929 recombination events in the ARG,
+this measure identifies 543 as artifactual,
+leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
+Evidence that this filter effectively flagged artifacts can be seen in the
+clear association between the AmpliSeq V1 kit and recombination events
+that we identify as artifactual.
+This primer scheme was used to sequence XX of the samples
+associated with an artifactual recombination event,
+in contrast to 15 of the ``robust'' events (Figure~\ref{fig:recombinant_qc}),
+strongly implicating it as a common source of problematic data.
+
+Three copying patterns in Figure~\ref{fig:copying_pattern_examples}
+Q1, 686
+Q3, 22029, Delta lineage-defining deletion (REF)
+There are others associated with a known amplicon dropout caused by this particular deletion.
+Q4, 22194
+True MNS
+Alignment error
 
@@
 then additionally remove internal samples,
-known recombinants (i.e. the Pango Xs) and three ``singleton recombinant
+known recombinants (i.e.\ the Pango X lineages) and three ``singleton recombinant
 samples''
@@
 (https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv)
-For each of these Pango Xs,
+For each of these Pango X lineages,
 we took the most commonly observed combination of
@@
 that can be easily performed using the combination of tskit and VCF Zarr.
-The entire analysis is performed in a short Jupyter notebook, with the 
+The entire analysis is performed in a short Jupyter
+notebook~\cite{kluyver2016jupyter}, with the 
 core calculation requiring about 30 lines of Python code. Retrieving 
@@
 \label{sec:supplement_pango_lineages}
-
 To analyze how well the ARG reflects the phylogenetic structure implicit in the
@@
 
-Of the 2058 distinct Pango lineages in the ARG, 1469 of these (comprising
-743922 samples) match perfectly, with unique origination events in the ARG
+Of the 2058 distinct pango lineages in the ARG, 1473 of these (comprising
+717798 samples) match perfectly, with unique origination events in the ARG
 where all samples assigned a given lineage descend from the first node assigned
-to that lineage. A further 310 lineages (398391 samples) match perfectly when
-we count the descendants of the parent of the first node (accounting for
+that lineage. A further 245 lineages (589197 samples) match perfectly when we
+count the descendants of the parent of the first node (accounting for
 polytomies in which multiple originating nodes for a given lineage are
-siblings). We then have 255 lineages (996490 samples) where the difference in
-the number of descendants of the first node's parent is $<$ 100. The remaining 25
-lineages (344079) are dominated by a few large lineages such as BA.1.1 (155595
-samples) and AY.4.2 (54607 samples) which have multiple non-sibling origins
-within the ARG.
-
-\subsection{Additional analyses of the Pango Xs}
+siblings). We then have 306 lineages (755840 samples) where the difference in
+the number descendants of the first node's parent is $<$ 100. The remaining 35
+lineages (420047 samples) are dominated by a few large lineages such as BA.1.1
+(147271 samples) and AY.4.2 (54607 samples) which have multiple non-sibling
+origins within the ARG.
+
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_pango\_events\_in\_arg.ipynb}
+{analysis\_pango\_events\_in\_arg.ipynb}
+for details.
+
+\subsection{Additional analyses of the Pango X lineages}
+\label{sec:supplement_indepth_pango_x_lineages}
 
@@
 \subsection*{Multiple breakpoints}
-Seven of the Pango Xs present in the Viridian v04 dataset have more than one breakpoint,
+Seven of the Pango X lineages present in the Viridian v04 dataset have more than one breakpoint,
 with the supporting evidence documented in the associated Pango designation issues.
-Five of these seven Pango Xs were detected as recombinants by sc2ts.
+Five of these seven X lineages were detected as recombinants by sc2ts.
 Three of them (XAY, XBC, XBT) had more than one breakpoint,
@@
 the presence of a large deletion inherited from BA.2*, which is currently treated as missing data by sc2ts.
-The XAC samples share a recombination node with the samples of four other Pango Xs
+The XAC samples share a recombination node with the samples of four other 
+X lineages
 which do not have this deletion (Figure 2C).
@@
 \centering
-\includegraphics[width=\linewidth]{figures/recombinant_qc}
+\includegraphics[width=\linewidth]{recombinant_qc}
 \caption{
@@
 breakdowns of the recombination events by primer scheme per quadrant.
-Recombination events associated with the origins of Pango Xs are labeled.
+Recombination events associated with the origins of Pango X lineages are labeled.
 }

### 2025-09-15 latex/main.tex
Updates from Overleaf

--- 2025-09-15 latex/main.tex
@@
 
-\subsubsection*{Identification of robust recombination events}
+\subsubsection*{Quality control of recombinants}
 
@@
 
-% This para was *{Counting supporting loci for recombinant parents}
-
-When counting such loci, we considered nearly adjacent sites (separated by at
-most three bases) as single loci, to avoid over-counting support from
-correlated loci, which would lead to artificial signals of recombination. Also,
-when counting supporting loci for a focal parent, we penalized loci where the
-character(s) in the recombinant matches that in the non-focal parent but not in
-the focal parent.
-
-Before characterizing recombination events in detail, we sought to remove
-probable artifacts. Systematic errors in SARS-CoV-2 sequences associated with
-sequencing protocols have been documented (Turakhia et al., 2022), which can
-confound recombination detection (De Maio et al., 2024). By inspecting copying
-patterns at sites that differ between recombinant parents, we find a convincing
-signal when treating adjacent and near-adjacent sites as a single locus.
-Artifactual recombination events can be characterized as having fewer than four
-such loci supporting the left or right parent on either side of a suggested
-breakpoint (STAR Methods). Of the 929 recombination events in the ARG, this
-measure identifies 543 as artifactual, leaving 386 ``robust'' events (Figure
-S1A). The copying patterns depicted in Figure S1B illustrate the genomic
-evidence for three recombination events identified as artifactual, compared to
-a marginally robust one.
+Before characterizing recombination events in detail,
+we sought to remove probable artifacts.
+Systematic errors in SARS-CoV-2 sequences
+associated with sequencing protocols \cite{Turakhia2020}
+can complicate recombination detection \cite{DeMaio2024}.
+% Explain to readers what convincing genomic evidence looks like.
+By inspecting copying patterns at sites that differ between recombinant parents,
+we find a convincing signal when treating adjacent and near-adjacent sites as a single locus.
+Artifactual recombination events can be characterized
+as having fewer than four such loci supporting the left or right parent
+on either side of a suggested breakpoint.
+When counting such loci, we considered nearly adjacent sites
+(separated by at most three bases) as single loci,
+to avoid over-counting support from correlated loci,
+which could lead to artificial signals of recombination.
+Also, when counting supporting loci for a focal parent,
+we penalized loci where the character(s)
+in the recombinant matches that in the non-focal parent
+but not in the focal parent.
+
+Of the 929 recombination events in the ARG,
+this measure identifies 543 as artifactual,
+leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
+The copying patterns depicted Figure~\ref{fig:copying_pattern_examples}
+illustrate the genomic evidence for three recombination events identified as artifactual,
+compared to a marginally robust one.
 
@@
 of the ``robust'' events (Figure 2A), strongly implicating it as a common source
-of problematic data. Through further inspection, we observed that many
-artifactual recombination events contain stretches of missing data
-characteristic of amplicon dropouts (REF).
+of problematic data.
 

### 2025-09-15 latex/main.tex

@@
 
-Of the 2058 distinct Pango lineages in the ARG, 1469 of these (comprising
-743922 samples) match perfectly, with unique origination events in the ARG
+Of the 2058 distinct pango lineages in the ARG, 1473 of these (comprising
+717798 samples) match perfectly, with unique origination events in the ARG
 where all samples assigned a given lineage descend from the first node assigned
-to that lineage. A further 310 lineages (398391 samples) match perfectly when
-we count the descendants of the parent of the first node (accounting for
+that lineage. A further 245 lineages (589197 samples) match perfectly when we
+count the descendants of the parent of the first node (accounting for
 polytomies in which multiple originating nodes for a given lineage are
-siblings). We then have 255 lineages (996490 samples) where the difference in
-the number of descendants of the first node's parent is $<$ 100. The remaining 25
-lineages (344079) are dominated by a few large lineages such as BA.1.1 (155595
-samples) and AY.4.2 (54607 samples) which have multiple non-sibling origins
-within the ARG.
+siblings). We then have 306 lineages (755840 samples) where the difference in
+the number descendants of the first node's parent is $<$ 100. The remaining 35
+lineages (420047 samples) are dominated by a few large lineages such as BA.1.1
+(147271 samples) and AY.4.2 (54607 samples) which have multiple non-sibling
+origins within the ARG.
+
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_pango\_events\_in\_arg.ipynb}
+{analysis\_pango\_events\_in\_arg.ipynb}
+for details.
 
 \subsection{Additional analyses of the Pango Xs}
+\label{sec:supplement_indepth_pango_x_lineages}
 

### 2025-09-15 latex/main.tex
Update the pango x events notebook and table

@@
 XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
-XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
 XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
 XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
-XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
+XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
+XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {} \\
 XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
 XAL:3} \\
+XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
+XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
+XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
 XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
-XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
-XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
+XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
 XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
-XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
 \midrule
@@
 \midrule
-Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 &
-{BA.2:156, XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
+Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 & {BA.2:156,
+XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
 XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
 {XE:1116, BA.2:37, XH:2, XAF:1} \\
-XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {BA.2:4, XBH:2} \\
+XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {XBH:6} \\
 XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
-BF.3:2} \\ XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
+BF.3:2} \\
+XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
 XAF:1} \\

### 2025-09-15 latex/main.tex
Fixup XBH

@@
 XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
-XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {} \\
 XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
 XAL:3} \\
+XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {} \\
 XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
@@
 {XE:1116, BA.2:37, XH:2, XAF:1} \\
-XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {XBH:6} \\
 XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,

### 2025-09-16 latex/main.tex
edits from Shing

@@
 \midrule
-Sc2ts & 14 / 15 (93.3) & 13 / 15 (86.7) & 540.5 (315, 766)\\
+Sc2ts & 14 / 16 (87.5) & 14 / 16 (87.5) & 540.5 (315, 766)\\
 RecombinHunt-GISAID & 13 / 13 (100.0) & 10 / 13 (76.9) & 764.0 (4, 1202)\\
@@
 we find a convincing signal when treating adjacent and near-adjacent sites as a single locus.
+
+Three copying patterns in Figure~\ref{fig:copying_pattern_examples}
+illustrate the evidence for three recombination events identified as artifactual.
+Explain the three bad patterns:
+
+Compared to a marginally robust one.
+
 Artifactual recombination events can be characterized
@@
 on either side of a suggested breakpoint.
+
 When counting such loci, we considered nearly adjacent sites
@@
 leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
-The copying patterns depicted Figure~\ref{fig:copying_pattern_examples}
-illustrate the genomic evidence for three recombination events identified as artifactual,
-compared to a marginally robust one.
-
-Evidence that our simple filter effectively flags artifacts can be seen in the
-clear association between the AmpliSeq V1 kit and recombination events that we
-identify as artifactual. This primer scheme was used to sequence XX of the
-samples associated with an artifactual recombination event, in contrast to 15
-of the ``robust'' events (Figure 2A), strongly implicating it as a common source
-of problematic data.
+Evidence that our simple filter effectively flagged artifacts can be seen in the
+clear association between the AmpliSeq V1 kit and recombination events
+that we identify as artifactual.
+This primer scheme was used to sequence XX of the samples
+associated with an artifactual recombination event,
+in contrast to 15 of the ``robust'' events (Figure~\ref{fig:recombinant_qc}),
+strongly implicating it as a common source of problematic data.
 

### 2025-09-16 latex/main.tex
Pango Xs to Pango X lineages
or just X lineages

Closes  #652

@@
 
-\subsection{Additional analyses of the Pango Xs}
+\subsection{Additional analyses of the Pango X lineages}
 \label{sec:supplement_indepth_pango_x_lineages}

### 2025-09-16 latex/main.tex

@@
 Sc2ts & 14 / 16 (87.5) & 14 / 16 (87.5) & 540.5 (315, 766)\\
-RecombinHunt-GISAID & 13 / 13 (100.0) & 10 / 13 (76.9) & 764.0 (4, 1202)\\
-RecombinHunt-Nextstrain & 12 / 14 (85.7) & 11 / 14 (78.6) & 1160.0 (559, 2785)\\
+RecombinHunt-GISAID & 13 / 15 (86.7) & 10 / 15 (66.7) & 764.0 (4, 1202)\\
+RecombinHunt-Nextstrain & 12 / 15 (80.0) & 11 / 15 (73.3) & 1160.0 (559, 2785)\\
 CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\

### 2025-09-16 latex/main.tex
Firt draft of Zarr section

@@
 
+\subsubsection*{Pandemic scale alignment storage}
+To provide the efficient access to alignment data required for inference with sc2ts, we
+developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
+Briefly, Zarr is a data storage format designed for scientific data applications
+originally developed to store genetic variation data for the
+\textit{Anopheles gambiae}
+1000 Genomes Project~\cite{Anopheles2017genetic}, and
+is now seeing adoption across a wide range of 
+applications~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
+Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
+Zarr is essentially a simple mechanism for storing n-dimensional array
+data as a regular grid of compressed chunks, which provides 
+high levels of data compression, efficient access across multiple
+dimensions, and is ideally suited to modern cloud-based
+deployments~\cite{Czech2025}.
+
+We converted the Viridian v05 dataset\cite{Hunt2024}, consisting of 4,484,157 SARS-CoV-2
+consensus sequences (125GiB over 48 FASTA files) and 30 metadata fields (1.4
+GiB TSV file) to VCF Zarr using a 
+reproducible Snakemake\cite{Koster2012snakemake} pipeline.
+We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using 
+MAFFT v7.475 \cite{Katoh2013},
+with the option \texttt{--keeplength} to retain the reference coordinate system.
+We then used \texttt{sc2ts import-alignments} and 
+\texttt{sc2ts import-metadata} to convert all alignments and metadata to VCF
+Zarr format, resulting in a single 401 MiB file 
+(around 77X smaller than gzip compressed FASTA\cite{Czech2025}).
+
+A classical tradeoff in bioinformatics is whether to store data 
+in sample-major (FASTA) or variant-major (VCF) format, determining
+whether the data can be accessed efficiently either by row or column
+of the variant matrix. Because VCF Zarr is chunked in two dimensions,
+we can now access either sample aligments or the variation
+data for a given site across all samples efficiently.
+See Czech et al.\cite{Czech2025} for more details and benchmarks on 
+this dataset. 
+Another advantage of Zarr over formats such as VCF is that is
+is inherently ``appendable'', and large datasets can be updated
+in real time
+
+Both the pipeline and converted Viridian dataset are available for download;
+see Data and code availability for details.
+
 \subsubsection*{Data preprocessing}
@@
 as seen by a large reduction in the number of likely artifactual reversions \cite{Hunt2024}.
-
-We downloaded 4,484,157 Viridian sequences obtained
-from samples collected from January, 2020 to March, 2023 (``batch 1'').
-We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using MAFFT v7.525 \cite{Katoh2013},
-with the option ‘--keeplength’ to retain the reference coordinate system, discarding insertions.
-We converted the MAFFT alignments and all sample metadata to VCF Zarr format \cite{Czech2025},
-which are available for download on Zenodo (URL).
 
@@
 Before characterizing recombination events in detail,
-we sought to remove probable artifacts.
-Systematic errors in SARS-CoV-2 sequences
-associated with sequencing protocols \cite{Turakhia2020}
-can complicate recombination detection \cite{DeMaio2024}.
-% Explain to readers what convincing genomic evidence looks like.
-By inspecting copying patterns at sites that differ between recombinant parents,
-we find a convincing signal when treating adjacent and near-adjacent sites as a single locus.
-
-Three copying patterns in Figure~\ref{fig:copying_pattern_examples}
-illustrate the evidence for three recombination events identified as artifactual.
-Explain the three bad patterns:
-
-Compared to a marginally robust one.
-
-Artifactual recombination events can be characterized
+we sought to remove probable artifactual recombinants,
+which can be caused by correlated bases
+that are treated independently by the HMM.
+Correlated changes can be caused by
+systematic errors in genome sequences
+associated with sequencing protocols \cite{Turakhia2020},
+multi-nucleotide substitutions \cite{DeMaio2024}, and
+alignment errors (e.g., deletions due to MAFFT misalignment \cite{Hunt2024}).
+
+We characterized artifactual recombination events
 as having fewer than four such loci supporting the left or right parent
 on either side of a suggested breakpoint.
-
-When counting such loci, we considered nearly adjacent sites
-(separated by at most three bases) as single loci,
-to avoid over-counting support from correlated loci,
-which could lead to artificial signals of recombination.
-Also, when counting supporting loci for a focal parent,
+We treated adjacent and near-adjacent sites
+(separated by at most three bases) as single loci.
+Also, when counting loci supporting a focal parent,
 we penalized loci where the character(s)
@@
 leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
-Evidence that our simple filter effectively flagged artifacts can be seen in the
+Evidence that this filter effectively flagged artifacts can be seen in the
 clear association between the AmpliSeq V1 kit and recombination events
@@
 strongly implicating it as a common source of problematic data.
+
+Three copying patterns in Figure~\ref{fig:copying_pattern_examples}
+AmpliSeq
+True MNS
+Alignment error
 

### 2025-09-16 latex/main.tex
Rework the data preprocessing Methods
Closes #688

@@
 is inherently ``appendable'', and large datasets can be updated
-in real time
+in real time.
 
@@
 
-\subsubsection*{Data preprocessing}
-
-In this study, we used the SARS-CoV-2 sequences reassembled in a recent study \cite{Hunt2024}
-using an amplicon-aware assembly method (named ``Viridian'').
-The Viridian sequences contain far fewer artifacts compared to their publicly deposited counterparts on GenBank.
-Consequently, phylogenetic tree inference using the Viridian data led to more parsimonious phylogenies,
-as seen by a large reduction in the number of likely artifactual reversions \cite{Hunt2024}.
-
-Prior to ARG inference, we preprocessed the remaining sequences as follows.
-We treated low-quality bases (represented by Ns), gaps and ambiguous characters in the sequences as missing data.
-When encoded as missing data, the bases do not inform sample matching (see ``The Li and Stephens model'').
-
-During inference, we filtered out the samples with low-quality sequences and low-precision sampling dates.
-We kept the quality check-passing sequences used to infer the Viridian UShER phylogeny \cite{Hunt2024}.
-We skipped low-quality sequences that had more than 500 missing sites
-(except when building the ``base'' ARG, when the maximum number of allowed missing bases was set to 10,000 instead).
-We also skipped $\sim$40,000 samples which had a collection date of December 31, 2020.
-The sampling frequency of this date is alarmingly high compared to other dates (at most 15,000 samples per day).
-We suspect that this batch of samples might be unusually enriched for wrong collection dates,
-likely due to arbitrary assignment of the originally reported collection date of ``2020'' to be ``31-12-2020''.
+\subsubsection*{Sample preprocessing}
+The collection date for a sequence is an important piece of information
+for sc2ts. We used the Viridian
+consensus date (``Date\_tree``), based on information from COG-UK, GISAID, and
+ENA/SRA as our source. Samples that have missing or incomplete dates (e.g., 2020-01)
+were omitted, as were
+the $\sim$40,000 samples which had a collection date of December 31, 2020.
+As the  sampling frequency of this date is much higher than other
+(at most 15,000 samples per day), it seems likely that this 
+batch is highly enriched for wrong collection dates.
 Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
@@
 
-Sites which are enriched for systematic sequencing errors or frequently experience homoplasy
-can cause artifactual signals of recombination (REF).
 The bioinformatics community has identified and curated a set of problematic sites
-that are recommended to be excluded before phylogenetic inference (REF).
-Here we assembled a bespoke set of such problematic sites
-using the information in the Viridian sequences.
+for SARS-CoV-2 that are enriched for systematic sequencing errors or 
+high levels of homoplasy~\cite{DeMaio2020issues}. 
+As these sites had not been updated to
+take into account the systematic improvements made by the Viridian dataset,
+we assembled a custom set of sites to mask. 
 To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
 without any site masking.
-Then, we identified the top 100 sites which had the highest mutation counts in the ARG (URL).
-Only six of these sites overlapped with the ``Problematic Sites''
-downloaded from the UCSC Genome Browser website (URL),
-which did not have a high number of mutations in the preliminary ARG.
-To infer the final ARG, we masked the problematic sites
-in our custom set
-\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}.
+Then, we identified the top 100 sites which had the highest mutation counts in
+the ARG.
+Only 11 of these sites 
+(635, 8835, 11074, 11083, 15521, 16887, 21304, 21305, 21575, 21987 and 28253)
+were present in the problematic sites file
+(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}).
+
+Alignments read from the VCZ Zarr file were first preprocessed to mask any 
+non nucleotide characters (A,C,G or T) as missing data. 
+Then, we filtered 
+any samples that had missing data at more than 500 sites, excluding
+the 100 sites identified as problematic (this was 
+set to 10,000 sites for the period up to 2020-03-01).
 

### 2025-09-17 latex/main.tex
Better version of Zarr section

@@
 developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
-Briefly, Zarr is a data storage format designed for scientific data applications
-originally developed to store genetic variation data for the
+Briefly, Zarr is a storage format for scientific data,
+originally developed for the
 \textit{Anopheles gambiae}
 1000 Genomes Project~\cite{Anopheles2017genetic}, and
-is now seeing adoption across a wide range of 
-applications~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
+is now seeing widespread adoption across the 
+sciences~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
 Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
@@
 of the variant matrix. Because VCF Zarr is chunked in two dimensions,
-we can now access either sample aligments or the variation
+we can now access either sample alignments or the variation
 data for a given site across all samples efficiently.
-See Czech et al.\cite{Czech2025} for more details and benchmarks on 
-this dataset.  
-Another advantage of Zarr over formats such as VCF is that is
-is inherently ``appendable'', and large datasets can be updated
-in real time.
+While the VCF Zarr encoding of the MAFFT-aligned Viridian sequences
+and metadata can be accessed via native Zarr libraries in several
+popular programming languages (see Czech et al.\cite{Czech2025} 
+for an overview), the sc2ts Python library provides a convenient interface. 
+The \texttt{sc2ts.Dataset}
+class provides efficient access to alignments for samples and 
+variant data for sites along the genome, as well as methods to export
+subsets of the data to FASTA if required.
+Document~\ref{sec:supplement_sc2ts_usher_imputation} gives an example
+of using this interface, where we retrieve alignment data for millions 
+of samples in a few minutes on a standard desktop computer with minimal
+memory overhead. See Czech et al.\cite{Czech2025} for more details 
+and benchmarks.
 

### 2025-09-17 latex/main.tex
Decent version of first part of methods

@@
 
-\subsubsection*{Sample preprocessing}
+\subsubsection*{Sample filtering and preprocessing}
 The collection date for a sequence is an important piece of information
@@
 the $\sim$40,000 samples which had a collection date of December 31, 2020.
-As the  sampling frequency of this date is much higher than other
-(at most 15,000 samples per day), it seems likely that this 
-batch is highly enriched for wrong collection dates.
-Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
+As the sample count for this date is a major outlier for this period of
+the pandemic, it seems likely that this 
+batch is highly enriched for incorrect collection dates.
+Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
 confirmed that some of the samples were not collected in 2020.
 
-The bioinformatics community has identified and curated a set of problematic sites
+The bioinformatics community has curated a set of problematic sites
 for SARS-CoV-2 that are enriched for systematic sequencing errors or 
 high levels of homoplasy~\cite{DeMaio2020issues}. 
-As these sites had not been updated to
-take into account the systematic improvements made by the Viridian dataset,
-we assembled a custom set of sites to mask. 
+Although these sites are recommended to be masked for UShER
+inference~\cite{Turakhia2020},
+the list has not been updated to take into account the systematic improvements 
+made by the Viridian dataset, and we therefore assembled a custom set of sites to mask. 
 To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
 without any site masking.
-Then, we identified the top 100 sites which had the highest mutation counts in
-the ARG.
+Then, we identified the top 100 sites in terms of mutation counts in the ARG.
 Only 11 of these sites 
@@
 were present in the problematic sites file
-(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}).
+(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}),
+and most sites in this list were typical in terms of the numbers of mutations.
+We therefore masked our 100 most-mutagenic sites to remove sequencing errors and improve
+matching performance in the LS HMM.
 
@@
 the 100 sites identified as problematic (this was 
-set to 10,000 sites for the period up to 2020-03-01).
+set to 10,000 sites for the initial period up to 2020-03-01 where sampling
+density is low).
 

### 2025-09-17 latex/main.tex
Updates from Overleaf

--- 2025-09-16 latex/main.tex
@@
 \begin{itemize}
-\item Snakemake workflow for aligning and converting the Viridian dataset to
-VCF Zarr:
-\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/viridian\_dataset/}
-\end{itemize}
-
-TODO add these to the list above with actual URLs
-\begin{itemize}
     \item The ARG and the sequence alignments have been deposited at Zenodo and are publicly available.
@@
 
-% \subsection*{Key resources table}
-
-% \textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}
+\subsection*{Key resources table}
+
+\textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}
 
@@
 such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
-and Pandas~\cite{Mckinney2010data} ,
+and Pandas ,
 using pyfaidx\cite{Shirley2015efficient} to access alignment data
-and Numba~\cite{Lam2015numba} to accelerate certain alignment processing
-tasks.
 [TODO JK to fill this out]
@@
 
-\subsubsection*{Pandemic scale alignment storage}
-To provide the efficient access to alignment data required for inference with sc2ts, we
-developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
-Briefly, Zarr is a data storage format designed for scientific data applications
-originally developed to store genetic variation data for the
-\textit{Anopheles gambiae}
-1000 Genomes Project~\cite{Anopheles2017genetic}, and
-is now seeing adoption across a wide range of 
-applications~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
-Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
-Zarr is essentially a simple mechanism for storing n-dimensional array
-data as a regular grid of compressed chunks, which provides 
-high levels of data compression, efficient access across multiple
-dimensions, and is ideally suited to modern cloud-based
-deployments~\cite{Czech2025}.
-
-We converted the Viridian v05 dataset\cite{Hunt2024}, consisting of 4,484,157 SARS-CoV-2
-consensus sequences (125GiB over 48 FASTA files) and 30 metadata fields (1.4
-GiB TSV file) to VCF Zarr using a 
-reproducible Snakemake\cite{Koster2012snakemake} pipeline.
-We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using 
-MAFFT v7.475 \cite{Katoh2013},
-with the option \texttt{--keeplength} to retain the reference coordinate system.
-We then used \texttt{sc2ts import-alignments} and 
-\texttt{sc2ts import-metadata} to convert all alignments and metadata to VCF
-Zarr format, resulting in a single 401 MiB file 
-(around 77X smaller than gzip compressed FASTA\cite{Czech2025}).
-
-A classical tradeoff in bioinformatics is whether to store data 
-in sample-major (FASTA) or variant-major (VCF) format, determining
-whether the data can be accessed efficiently either by row or column
-of the variant matrix. Because VCF Zarr is chunked in two dimensions,
-we can now access either sample aligments or the variation
-data for a given site across all samples efficiently.
-See Czech et al.\cite{Czech2025} for more details and benchmarks on 
-this dataset.  
-Another advantage of Zarr over formats such as VCF is that is
-is inherently ``appendable'', and large datasets can be updated
-in real time.
-
-Both the pipeline and converted Viridian dataset are available for download;
-see Data and code availability for details.
-
-\subsubsection*{Sample preprocessing}
-The collection date for a sequence is an important piece of information
-for sc2ts. We used the Viridian
-consensus date (``Date\_tree``), based on information from COG-UK, GISAID, and
-ENA/SRA as our source. Samples that have missing or incomplete dates (e.g., 2020-01)
-were omitted, as were
-the $\sim$40,000 samples which had a collection date of December 31, 2020.
-As the  sampling frequency of this date is much higher than other
-(at most 15,000 samples per day), it seems likely that this 
-batch is highly enriched for wrong collection dates.
+\subsubsection*{Data preprocessing}
+
+In this study, we used the SARS-CoV-2 sequences reassembled in a recent study \cite{Hunt2024}
+using an amplicon-aware assembly method (named ``Viridian'').
+The Viridian sequences contain far fewer artifacts compared to their publicly deposited counterparts on GenBank.
+Consequently, phylogenetic tree inference using the Viridian data led to more parsimonious phylogenies,
+as seen by a large reduction in the number of likely artifactual reversions \cite{Hunt2024}.
+
+We downloaded 4,484,157 Viridian sequences obtained
+from samples collected from January, 2020 to March, 2023 (``batch 1'').
+We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using MAFFT v7.525 \cite{Katoh2013},
+with the option ‘--keeplength’ to retain the reference coordinate system, discarding insertions.
+We converted the MAFFT alignments and all sample metadata to VCF Zarr format \cite{Czech2025},
+which are available for download on Zenodo (URL).
+
+Prior to ARG inference, we preprocessed the remaining sequences as follows.
+We treated low-quality bases (represented by Ns), gaps and ambiguous characters in the sequences as missing data.
+When encoded as missing data, the bases do not inform sample matching (see ``The Li and Stephens model'').
+
+During inference, we filtered out the samples with low-quality sequences and low-precision sampling dates.
+We kept the quality check-passing sequences used to infer the Viridian UShER phylogeny \cite{Hunt2024}.
+We skipped low-quality sequences that had more than 500 missing sites
+(except when building the ``base'' ARG, when the maximum number of allowed missing bases was set to 10,000 instead).
+We also skipped $\sim$40,000 samples which had a collection date of December 31, 2020.
+The sampling frequency of this date is alarmingly high compared to other dates (at most 15,000 samples per day).
+We suspect that this batch of samples might be unusually enriched for wrong collection dates,
+likely due to arbitrary assignment of the originally reported collection date of ``2020'' to be ``31-12-2020''.
 Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
@@
 
+Sites which are enriched for systematic sequencing errors or frequently experience homoplasy
+can cause artifactual signals of recombination (REF).
 The bioinformatics community has identified and curated a set of problematic sites
-for SARS-CoV-2 that are enriched for systematic sequencing errors or 
-high levels of homoplasy~\cite{DeMaio2020issues}. 
-As these sites had not been updated to
-take into account the systematic improvements made by the Viridian dataset,
-we assembled a custom set of sites to mask. 
+that are recommended to be excluded before phylogenetic inference (REF).
+Here we assembled a bespoke set of such problematic sites
+using the information in the Viridian sequences.
 To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
 without any site masking.
-Then, we identified the top 100 sites which had the highest mutation counts in
-the ARG.
-Only 11 of these sites 
-(635, 8835, 11074, 11083, 15521, 16887, 21304, 21305, 21575, 21987 and 28253)
-were present in the problematic sites file
-(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}).
-
-Alignments read from the VCZ Zarr file were first preprocessed to mask any 
-non nucleotide characters (A,C,G or T) as missing data. 
-Then, we filtered 
-any samples that had missing data at more than 500 sites, excluding
-the 100 sites identified as problematic (this was 
-set to 10,000 sites for the period up to 2020-03-01).
+Then, we identified the top 100 sites which had the highest mutation counts in the ARG (URL).
+Only six of these sites overlapped with the ``Problematic Sites''
+downloaded from the UCSC Genome Browser website (URL),
+which did not have a high number of mutations in the preliminary ARG.
+To infer the final ARG, we masked the problematic sites
+in our custom set
+\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}.
 
@@
 that can be easily performed using the combination of tskit and VCF Zarr.
-The entire analysis is performed in a short Jupyter
-notebook~\cite{kluyver2016jupyter}, with the 
+The entire analysis is performed in a short Jupyter notebook, with the 
 core calculation requiring about 30 lines of Python code. Retrieving 

### 2025-09-17 latex/main.tex
Move redundant Methods sc2ts overview text

--- 2025-09-17 latex/main.tex
@@
 \end{figure}
+
+%% JK: This is copy of old Methods section "overview of sc2ts" which is basically
+%% the same as this section. TODO Merge the useful text in here.
+
+% Sc2ts (pronounced ``scoots'', optionally)
+% is a method for inferring Ancestral Recombination Graphs (ARGs)
+% from densely sampled genome sequence data in real time,
+% in which recombination occurs at a low but significant rate.
+% The basic idea is to incrementally update an ARG regularly
+% with the sequences collected since the previous batch (Figure
+% \ref{fig:methodology}).
+% We used daily batches for SARS-CoV-2,
+% and use days as the time interval for clarity in what follows.
+% The first step is to find likely ``copying paths'' under the Li and Stephens (LS) model \cite{LiStephens2003}
+% for each sequence in the daily batch to the current ARG.
+% These copying paths will mostly consist of a new sample sequence copying
+% from a node in the ARG with a small number of mutations,
+% and often many samples from a daily batch will copy from the same ARG node.
+% The second step is then to ``resolve'' these implied polytomies
+% by using a standard neighbor-joining algorithm \cite{Saitou1987}.
+% This greedy update strategy inevitably leads to unparsimonious topologies,
+% and the third step is to then increase the overall parsimony of the inferred ARG
+% by making some simple topological updates.
+% Recombination is inferred as an integral and ongoing part of this process,
+% requiring only a few additional steps to facilitate later analysis.
+% The result of inference for a given day is then a genealogy recording genetic inheritance
+% as well as mutation and recombination events
+% for all of the samples added to the ARG up to that day.
+
+% Sc2ts is implemented in Python building on core PyData infrastructure
+% such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
+% and Pandas~\cite{Mckinney2010data} ,
+% using pyfaidx\cite{Shirley2015efficient} to access alignment data
+% and Numba~\cite{Lam2015numba} to accelerate certain alignment processing
+% tasks.
+% [TODO JK to fill this out]
 
@@
 
-\subsubsection*{Description of sc2ts}
-
-Sc2ts (pronounced ``scoots'', optionally)
-is a method for inferring Ancestral Recombination Graphs (ARGs)
-from densely sampled genome sequence data in real time,
-in which recombination occurs at a low but significant rate.
-The basic idea is to incrementally update an ARG regularly
-with the sequences collected since the previous batch (Figure
-\ref{fig:methodology}).
-We used daily batches for SARS-CoV-2,
-and use days as the time interval for clarity in what follows.
-The first step is to find likely ``copying paths'' under the Li and Stephens (LS) model \cite{LiStephens2003}
-for each sequence in the daily batch to the current ARG.
-These copying paths will mostly consist of a new sample sequence copying
-from a node in the ARG with a small number of mutations,
-and often many samples from a daily batch will copy from the same ARG node.
-The second step is then to ``resolve'' these implied polytomies
-by using a standard neighbor-joining algorithm \cite{Saitou1987}.
-This greedy update strategy inevitably leads to unparsimonious topologies,
-and the third step is to then increase the overall parsimony of the inferred ARG
-by making some simple topological updates.
-Recombination is inferred as an integral and ongoing part of this process,
-requiring only a few additional steps to facilitate later analysis.
-The result of inference for a given day is then a genealogy recording genetic inheritance
-as well as mutation and recombination events
-for all of the samples added to the ARG up to that day.
-
-Sc2ts is implemented in Python building on core PyData infrastructure
-such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
-and Pandas~\cite{Mckinney2010data} ,
-using pyfaidx\cite{Shirley2015efficient} to access alignment data
-and Numba~\cite{Lam2015numba} to accelerate certain alignment processing
-tasks.
-[TODO JK to fill this out]
-
 \subsubsection*{Ancestral Recombination Graphs}
@@
 
+\subsubsection*{Post hoc parsimony mapping}
+After the main inference step we mapped the variation data for 163 sites 
+in the alignments to the ARG using the Fitch-Hartigan 
+algorithm\cite{Fitch1971toward,Hartigan1973minimum} 
+implemented in tskit. In contrast to the main inference process where 
+deletion characters are masked as missing data, we allow for 
+5 different states in these parsimony calculations: A,C,G,T and -.
+By doing this, we can approximate the mutation events causing
+deletions, but with the clear caveat that each site is treated 
+independently. 
+Using data provided by Li et al.\cite{Li2023indels} (Table S1) 
+we chose 75 sites to remap
+
+performed a large-scale analysis of indels
+using a global GISAID dataset (sampled up to August, 2022).
+We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
+which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
+All of these indels are deletions and have high frequencies in the ARG
+(sequence count divided by the total number of samples),
+except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
+
+Based on the data of Table S4 of Li2023indels
+
 \subsubsection*{Quality control of recombinants}
@@
 Before characterizing recombination events in detail,
-we sought to remove probable artifactual recombinants,
-which can be caused by correlated bases
-that are treated independently by the HMM.
-Correlated changes can be caused by
+we sought to remove probable artifactual recombinants
+due to correlated bases
+being treated as individual sites by the HMM.
+Correlated bases can be caused by
 systematic errors in genome sequences
 associated with sequencing protocols \cite{Turakhia2020},
-multi-nucleotide substitutions \cite{DeMaio2024}, and
+multi-nucleotide substitutions \cite{DeMaio2024},
+genuine deletions, and
 alignment errors (e.g., deletions due to MAFFT misalignment \cite{Hunt2024}).
@@
 Three copying patterns in Figure~\ref{fig:copying_pattern_examples}
-AmpliSeq
+Q1, 686
+Q3, 22029, Delta lineage-defining deletion (REF)
+There are others associated with a known amplicon dropout caused by this particular deletion.
+Q4, 22194
 True MNS

### 2025-09-17 latex/main.tex
Move up the alignment storage and preprocessing bits
remove "accelerated sample matcing" ssection

@@
 
+\subsubsection*{Pandemic scale alignment storage}
+To provide the efficient access to alignment data required for inference with sc2ts, we
+developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
+Briefly, Zarr is a storage format for scientific data,
+originally developed for the
+\textit{Anopheles gambiae}
+1000 Genomes Project~\cite{Anopheles2017genetic}, and
+is now seeing widespread adoption across the 
+sciences~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
+Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
+Zarr is essentially a simple mechanism for storing n-dimensional array
+data as a regular grid of compressed chunks, which provides 
+high levels of data compression, efficient access across multiple
+dimensions, and is ideally suited to modern cloud-based
+deployments~\cite{Czech2025}.
+
+We converted the Viridian v05 dataset\cite{Hunt2024}, consisting of 4,484,157 SARS-CoV-2
+consensus sequences (125GiB over 48 FASTA files) and 30 metadata fields (1.4
+GiB TSV file) to VCF Zarr using a 
+reproducible Snakemake\cite{Koster2012snakemake} pipeline.
+We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using 
+MAFFT v7.475 \cite{Katoh2013},
+with the option \texttt{--keeplength} to retain the reference coordinate system.
+We then used \texttt{sc2ts import-alignments} and 
+\texttt{sc2ts import-metadata} to convert all alignments and metadata to VCF
+Zarr format, resulting in a single 401 MiB file 
+(around 77X smaller than gzip compressed FASTA\cite{Czech2025}).
+
+A classical tradeoff in bioinformatics is whether to store data 
+in sample-major (FASTA) or variant-major (VCF) format, determining
+whether the data can be accessed efficiently either by row or column
+of the variant matrix. Because VCF Zarr is chunked in two dimensions,
+we can now access either sample alignments or the variation
+data for a given site across all samples efficiently.
+While the VCF Zarr encoding of the MAFFT-aligned Viridian sequences
+and metadata can be accessed via native Zarr libraries in several
+popular programming languages (see Czech et al.\cite{Czech2025} 
+for an overview), the sc2ts Python library provides a convenient interface. 
+The \texttt{sc2ts.Dataset}
+class provides efficient access to alignments for samples and 
+variant data for sites along the genome, as well as methods to export
+subsets of the data to FASTA if required.
+Document~\ref{sec:supplement_sc2ts_usher_imputation} gives an example
+of using this interface, where we retrieve alignment data for millions 
+of samples in a few minutes on a standard desktop computer with minimal
+memory overhead. See Czech et al.\cite{Czech2025} for more details 
+and benchmarks.
+
+Both the pipeline and converted Viridian dataset are available for download;
+see Data and code availability for details.
+
+\subsubsection*{Sample filtering and preprocessing}
+The collection date for a sequence is an important piece of information
+for sc2ts. We used the Viridian
+consensus date (``Date\_tree``), based on information from COG-UK, GISAID, and
+ENA/SRA as our source. Samples that have missing or incomplete dates (e.g., 2020-01)
+were omitted, as were
+the $\sim$40,000 samples which had a collection date of December 31, 2020.
+As the sample count for this date is a major outlier for this period of
+the pandemic, it seems likely that this 
+batch is highly enriched for incorrect collection dates.
+Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
+confirmed that some of the samples were not collected in 2020.
+
+The bioinformatics community has curated a set of problematic sites
+for SARS-CoV-2 that are enriched for systematic sequencing errors or 
+high levels of homoplasy~\cite{DeMaio2020issues}. 
+Although these sites are recommended to be masked for UShER
+inference~\cite{Turakhia2020},
+the list has not been updated to take into account the systematic improvements 
+made by the Viridian dataset, and we therefore assembled a custom set of sites to mask. 
+To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
+without any site masking.
+Then, we identified the top 100 sites in terms of mutation counts in the ARG.
+Only 11 of these sites 
+(635, 8835, 11074, 11083, 15521, 16887, 21304, 21305, 21575, 21987 and 28253)
+were present in the problematic sites file
+(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}),
+and most sites in this list were typical in terms of the numbers of mutations.
+We therefore masked our 100 most-mutagenic sites to remove sequencing errors and improve
+matching performance in the LS HMM.
+
+Alignments read from the VCZ Zarr file were first preprocessed to mask any 
+non nucleotide characters (A,C,G or T) as missing data. 
+Then, we filtered 
+any samples that had missing data at more than 500 sites, excluding
+the 100 sites identified as problematic (this was 
+set to 10,000 sites for the initial period up to 2020-03-01 where sampling
+density is low).
+
 \subsubsection*{Li and Stephens haplotype-copying model}
 
-LS model \cite{LiStephens2003} is an approximation of the coalescent with recombination
+The LS model \cite{LiStephens2003} is an approximation of the coalescent with recombination
 that captures many of the key features of the joint processes of mutation and recombination.
@@
 collected on a certain date through the samples collected on earlier dates in a given ARG.
-
-\subsubsection*{Accelerating sample matching}
 
@@
 when analysing the samples around the seed(s) in the ARG.
-
-\subsubsection*{Pandemic scale alignment storage}
-To provide the efficient access to alignment data required for inference with sc2ts, we
-developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
-Briefly, Zarr is a storage format for scientific data,
-originally developed for the
-\textit{Anopheles gambiae}
-1000 Genomes Project~\cite{Anopheles2017genetic}, and
-is now seeing widespread adoption across the 
-sciences~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
-Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
-Zarr is essentially a simple mechanism for storing n-dimensional array
-data as a regular grid of compressed chunks, which provides 
-high levels of data compression, efficient access across multiple
-dimensions, and is ideally suited to modern cloud-based
-deployments~\cite{Czech2025}.
-
-We converted the Viridian v05 dataset\cite{Hunt2024}, consisting of 4,484,157 SARS-CoV-2
-consensus sequences (125GiB over 48 FASTA files) and 30 metadata fields (1.4
-GiB TSV file) to VCF Zarr using a 
-reproducible Snakemake\cite{Koster2012snakemake} pipeline.
-We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using 
-MAFFT v7.475 \cite{Katoh2013},
-with the option \texttt{--keeplength} to retain the reference coordinate system.
-We then used \texttt{sc2ts import-alignments} and 
-\texttt{sc2ts import-metadata} to convert all alignments and metadata to VCF
-Zarr format, resulting in a single 401 MiB file 
-(around 77X smaller than gzip compressed FASTA\cite{Czech2025}).
-
-A classical tradeoff in bioinformatics is whether to store data 
-in sample-major (FASTA) or variant-major (VCF) format, determining
-whether the data can be accessed efficiently either by row or column
-of the variant matrix. Because VCF Zarr is chunked in two dimensions,
-we can now access either sample alignments or the variation
-data for a given site across all samples efficiently.
-While the VCF Zarr encoding of the MAFFT-aligned Viridian sequences
-and metadata can be accessed via native Zarr libraries in several
-popular programming languages (see Czech et al.\cite{Czech2025} 
-for an overview), the sc2ts Python library provides a convenient interface. 
-The \texttt{sc2ts.Dataset}
-class provides efficient access to alignments for samples and 
-variant data for sites along the genome, as well as methods to export
-subsets of the data to FASTA if required.
-Document~\ref{sec:supplement_sc2ts_usher_imputation} gives an example
-of using this interface, where we retrieve alignment data for millions 
-of samples in a few minutes on a standard desktop computer with minimal
-memory overhead. See Czech et al.\cite{Czech2025} for more details 
-and benchmarks.
-
-Both the pipeline and converted Viridian dataset are available for download;
-see Data and code availability for details.
-
-\subsubsection*{Sample filtering and preprocessing}
-The collection date for a sequence is an important piece of information
-for sc2ts. We used the Viridian
-consensus date (``Date\_tree``), based on information from COG-UK, GISAID, and
-ENA/SRA as our source. Samples that have missing or incomplete dates (e.g., 2020-01)
-were omitted, as were
-the $\sim$40,000 samples which had a collection date of December 31, 2020.
-As the sample count for this date is a major outlier for this period of
-the pandemic, it seems likely that this 
-batch is highly enriched for incorrect collection dates.
-Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
-confirmed that some of the samples were not collected in 2020.
-
-The bioinformatics community has curated a set of problematic sites
-for SARS-CoV-2 that are enriched for systematic sequencing errors or 
-high levels of homoplasy~\cite{DeMaio2020issues}. 
-Although these sites are recommended to be masked for UShER
-inference~\cite{Turakhia2020},
-the list has not been updated to take into account the systematic improvements 
-made by the Viridian dataset, and we therefore assembled a custom set of sites to mask. 
-To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
-without any site masking.
-Then, we identified the top 100 sites in terms of mutation counts in the ARG.
-Only 11 of these sites 
-(635, 8835, 11074, 11083, 15521, 16887, 21304, 21305, 21575, 21987 and 28253)
-were present in the problematic sites file
-(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}),
-and most sites in this list were typical in terms of the numbers of mutations.
-We therefore masked our 100 most-mutagenic sites to remove sequencing errors and improve
-matching performance in the LS HMM.
-
-Alignments read from the VCZ Zarr file were first preprocessed to mask any 
-non nucleotide characters (A,C,G or T) as missing data. 
-Then, we filtered 
-any samples that had missing data at more than 500 sites, excluding
-the 100 sites identified as problematic (this was 
-set to 10,000 sites for the initial period up to 2020-03-01 where sampling
-density is low).
 

### 2025-09-17 latex/main.tex
Minor tweaking on the Handling outlier samples section

@@
 
-Although we expected that most samples can be explained by a small number of mutations in the ARG,
-we observed two types of outlier samples which had unusually high numbers of mutations:
-(1) ``time travelers'' and (2) ``saltation'' lineages.
+Although most samples can be explained by a small number of mutations in the ARG,
+there are two types of outlier samples which have unusually high numbers of mutations:
+``time travelers'' and ``saltation'' lineages.
 Time traveling samples are caused by errors in their collection dates.

### 2025-09-17 latex/main.tex
Some details on the postprocessing

@@
 \subsection*{Data and code availability}
+\label{sec:data_and_code_availability}
 
@@
 
+[TODO fix the typesetting once all resources are compiled]
+
 \begin{itemize}
@@
 \url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/viridian\_dataset/}
+\item Complete configuration for running sc2ts base ARG inference:
+\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}
+\item Snakemake workflow for ARG post-processing
+\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/Snakefile}
 \end{itemize}
@@
 %%% TODO: This needs to be explained.
+
+[TODO explain Fig S1 and S2]
 
@@
 
-\subsubsection*{Reconstructing a pandemic-scale ARG}
-
-%%% TODO: Much of this needs to be rewritten once the post-processing pipeline is finalised.
-Using sc2ts, we reconstructed the ARG presented here in three stages:
-(1) initialization, (2) extension, and (3) post-processing.
-
-First, we initialized a base ARG.
-We added the Wuhan-Hu-1 reference sample (MN908947.3) as the root node, dated December 8, 2019.
-Then, we ran sc2ts unconditionally (i.e., without excluding samples by HMM cost) up to March 01, 2020 (inclusive).
-This step is necessary to accumulate sufficient genetic diversity in the ARG
-for subsequent sample matching and attachment in the next stage.
-
-Second, we extended the base ARG by adding batches of samples collected on and after March 02, 2020.
-Sample filtering, HMM cost-based rejection, and retrospective sample grouping, as described above,
-are in full effect during this stage.
-
-Finally, we performed the following post-processing steps
-to modify certain non-topological properties of the ARG or
-to enrich the ARG with additional information:
-1. Moving mutations around recombinants? TODO
-2. Adjusting recombination breakpoints. XX.
-3.
-Mapping multi-base deletions.
-Multi-base deletions are challenging to handle, as they violate the Markov property of the HMM.
-In our ARG, insertions are deleted regions that acquire nucleotides (not necessarily the same nucleotides) without altering the genome length, because we discarded length-changing insertions when preprocessing the aligned sequences.
-We treated multi-base indels as multiple separate mutations in the HMM.
-We mapped them back onto the inferred ARG as single events.
-4. Mapping mutations at masked sites. XX.
-5.
-Adding exactly matched samples.
-During sample matching, samples identical to node(s) in a given ARG were not added to the ARG.
-Instead, after inference, we simply added the nodes to the ARG using the saved matching results.
-This allowed for computational savings, as it kept the number of possible parents available for matching to a minimum.
-6.
-Estimating non-sample node ages.
-The ages of the inserted non-sample nodes (the nodes in the local trees and the recombination nodes) were initially assigned to arbitrarily determined values.
-We ran \textit{tsdate} (``variational gamma'', version XX; \cite{Wohns2022} Pope et al., 2025) on the ARG to adjust the ages of these nodes, while fixing the ages of the sample nodes to the reported collection times (defined as fractional years before the collection date of the last sample added to the ARG, which is February 20, 2025 in the ARG described in this study).
-7. Assigning Pango lineages to non-sample nodes.
-
-\subsubsection*{Post hoc parsimony mapping}
-After the main inference step we mapped the variation data for 163 sites 
-in the alignments to the ARG using the Fitch-Hartigan 
+\subsubsection*{Primary ARG inference}
+The ``base ARG'' we infer with sc2ts is the starting point for later analyses, 
+and this step dominates overall processing time. To summarise, we first 
+initialise the ARG using the Wuhan-Hu-1 reference sample (MN908947.3) as the
+root node, and then ran the algorithm day-by-day using the parameters 
+described in the preceding sections and documented in the configuration file 
+(see \nameref{sec:data_and_code_availability}).
+As discussed above,
+computational resources are dominated by the LS model
+(Figures~\ref{fig:samples_per_day},\ref{fig:inference_resources}) with all 
+other aspects contributing negligibly.
+
+\subsubsection*{ARG postprocessing}
+The first step in ARG postprocessing is to examine putative recombinants to
+determine whether a more parsimonious explanation is possible. This is done 
+by rematching the inferred recombinant sequence against the ARG pruned back
+to include only samples before the date of the recombinant. Additionally,
+ each recombination node is examined to determine if any mutations 
+are shared [TODO Yan write this up].
+We refer to this process as ``long branch splitting''. 
+% Data from
+% https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/logs/sc2ts_v1_2023-02-21_rewire-recombinants-report.ipynb
+From this, 58 recombinants were identified as being more parsimoniously 
+explained by matching to existing nodes in the ARG (inserted by 
+subsequent parsimony improving heuristics) and 16 by long branch 
+splitting. Of these 16 recombinants, two are noteworthy as 
+being the origins of the Delta and BA.2 lineages in the ARG.
+Both result in substantially fewer mutations (15 fewer for 
+Delta, 6 for BA.2) than the previous recombinant HMM solution.
+These major saltational lineages highlight the challenges 
+of using the LS HMM with highly diverged samples.
+We then updated the ARG to remove these 74 artefactual 
+recombinants by ``rewiring'' the recombination node to the 
+more parsimonious non-recombinant solution. This 
+resulted in 16 additional nodes, 182 fewer mutations and reduced
+the number of trees along the genome from 348 to 317.
+
+% and requires a lot of off-topic explanation.
+% It is also noteworthy that the origin of the BA.5
+% lineage is identified as non-recombinant when long-branch
+% splitting is applied to the daily ARG immediately after 
+% its inclusion---for this reason, we have identif
+
+The second postprocessing step performed is to add 1,252,208 samples
+detected as exact matches during primary inference back
+into the ARG. These are not included in the ARG during 
+primary inference to reduce memory usage and improve 
+LS HMM performance.
+
+During primary inference, gap characters (``-'') are masked 
+out as missing data, but this excludes both important information
+about deletions from the ARG as well as potentially informative 
+phylogenetic signal. 
+We therefore performed post-hoc parsinony mapping at 163 sites to include 
+the effects of deletions and to ensure that all Pango-informative
+sites are included in the ARG.
+Using deletions with a 
+a minimum frequency of 1\% 
+(count of sequences divided by 9,149,680 sequences analyzed)
+from Table S1 of Li et al.\cite{Li2023indels},
+we chose 75 sites to remap.
+In addition, we remapped data at 88 sites that were excluded
+from primary inference and are defined as Pango lineage 
+defining sites by
+ Freyja\cite{Karthikeyan2022}
+(\url{https://raw.githubusercontent.com/andersen-lab/Freyja/refs/heads/main/freyja/data/lineage\_mutations.json}).
+Alignment data for these 163 sites was then mapped to the 
+ARG using the Fitch-Hartigan 
 algorithm\cite{Fitch1971toward,Hartigan1973minimum} 
-implemented in tskit. In contrast to the main inference process where 
-deletion characters are masked as missing data, we allow for 
-5 different states in these parsimony calculations: A,C,G,T and -.
-By doing this, we can approximate the mutation events causing
-deletions, but with the clear caveat that each site is treated 
-independently. 
-Using data provided by Li et al.\cite{Li2023indels} (Table S1) 
-we chose 75 sites to remap
-
-performed a large-scale analysis of indels
-using a global GISAID dataset (sampled up to August, 2022).
-We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
-which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
-All of these indels are deletions and have high frequencies in the ARG
-(sequence count divided by the total number of samples),
-except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
-
-Based on the data of Table S4 of Li2023indels
+implemented in tskit, including the gap character as a state 
+along with the 4 nucleotides.
+Following this, we iteratively applied the node parsimony 
+heuristics in sc2ts until there were no further reductions in 
+the numbers of mutations.
+
+% Shifting breakpoints is a minor point, not worth mentioning?
+We computed Pango assignments for all 2,747,985 nodes in the ARG by first 
+exporting their aligned sequences to a 77GiB FASTA file and 
+then running pangolin (version 4.3.1; pangolin-data version 1.29).
+The resulting Pango and Scorpio designations were then stored as 
+metadata associated with each node in the ARG.
+
+When then estimated non-sample node dates using 
+the ``variational-gamma'' algorithm in tsdate\cite{Wohns2022}
+version 0.2.3. [TODO Say something above the node ages?]
+
+Finally, we reduced the metadata stored in the ARG to a minimal
+set (node sample IDs, along with computed Pango and Scorpio 
+designations), encoded to enable efficient columnar access,
+and compressed the file using tszip (version 0.2.5). The final 
+file requires 32MiB of storage.
+
+This full post-processing pipeline is encoded as a reproducible Snakemake
+workflow (see \nameref{sec:data_and_code_availability}).
 

### 2025-09-17 latex/main.tex
Updates from Overleaf

--- 2025-09-17 latex/main.tex
@@
 \end{figure}
-
-%% JK: This is copy of old Methods section "overview of sc2ts" which is basically
-%% the same as this section. TODO Merge the useful text in here.
-
-% Sc2ts (pronounced ``scoots'', optionally)
-% is a method for inferring Ancestral Recombination Graphs (ARGs)
-% from densely sampled genome sequence data in real time,
-% in which recombination occurs at a low but significant rate.
-% The basic idea is to incrementally update an ARG regularly
-% with the sequences collected since the previous batch (Figure
-% \ref{fig:methodology}).
-% We used daily batches for SARS-CoV-2,
-% and use days as the time interval for clarity in what follows.
-% The first step is to find likely ``copying paths'' under the Li and Stephens (LS) model \cite{LiStephens2003}
-% for each sequence in the daily batch to the current ARG.
-% These copying paths will mostly consist of a new sample sequence copying
-% from a node in the ARG with a small number of mutations,
-% and often many samples from a daily batch will copy from the same ARG node.
-% The second step is then to ``resolve'' these implied polytomies
-% by using a standard neighbor-joining algorithm \cite{Saitou1987}.
-% This greedy update strategy inevitably leads to unparsimonious topologies,
-% and the third step is to then increase the overall parsimony of the inferred ARG
-% by making some simple topological updates.
-% Recombination is inferred as an integral and ongoing part of this process,
-% requiring only a few additional steps to facilitate later analysis.
-% The result of inference for a given day is then a genealogy recording genetic inheritance
-% as well as mutation and recombination events
-% for all of the samples added to the ARG up to that day.
-
-% Sc2ts is implemented in Python building on core PyData infrastructure
-% such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
-% and Pandas~\cite{Mckinney2010data} ,
-% using pyfaidx\cite{Shirley2015efficient} to access alignment data
-% and Numba~\cite{Lam2015numba} to accelerate certain alignment processing
-% tasks.
-% [TODO JK to fill this out]
 
@@
 
+\subsubsection*{Description of sc2ts}
+
+Sc2ts (pronounced ``scoots'', optionally)
+is a method for inferring Ancestral Recombination Graphs (ARGs)
+from densely sampled genome sequence data in real time,
+in which recombination occurs at a low but significant rate.
+The basic idea is to incrementally update an ARG regularly
+with the sequences collected since the previous batch (Figure
+\ref{fig:methodology}).
+We used daily batches for SARS-CoV-2,
+and use days as the time interval for clarity in what follows.
+The first step is to find likely ``copying paths'' under the Li and Stephens (LS) model \cite{LiStephens2003}
+for each sequence in the daily batch to the current ARG.
+These copying paths will mostly consist of a new sample sequence copying
+from a node in the ARG with a small number of mutations,
+and often many samples from a daily batch will copy from the same ARG node.
+The second step is then to ``resolve'' these implied polytomies
+by using a standard neighbor-joining algorithm \cite{Saitou1987}.
+This greedy update strategy inevitably leads to unparsimonious topologies,
+and the third step is to then increase the overall parsimony of the inferred ARG
+by making some simple topological updates.
+Recombination is inferred as an integral and ongoing part of this process,
+requiring only a few additional steps to facilitate later analysis.
+The result of inference for a given day is then a genealogy recording genetic inheritance
+as well as mutation and recombination events
+for all of the samples added to the ARG up to that day.
+
+Sc2ts is implemented in Python building on core PyData infrastructure
+such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
+and Pandas~\cite{Mckinney2010data} ,
+using pyfaidx\cite{Shirley2015efficient} to access alignment data
+and Numba~\cite{Lam2015numba} to accelerate certain alignment processing
+tasks.
+[TODO JK to fill this out]
+
 \subsubsection*{Ancestral Recombination Graphs}
@@
 
-\subsubsection*{Pandemic scale alignment storage}
-To provide the efficient access to alignment data required for inference with sc2ts, we
-developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
-Briefly, Zarr is a storage format for scientific data,
-originally developed for the
-\textit{Anopheles gambiae}
-1000 Genomes Project~\cite{Anopheles2017genetic}, and
-is now seeing widespread adoption across the 
-sciences~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
-Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
-Zarr is essentially a simple mechanism for storing n-dimensional array
-data as a regular grid of compressed chunks, which provides 
-high levels of data compression, efficient access across multiple
-dimensions, and is ideally suited to modern cloud-based
-deployments~\cite{Czech2025}.
-
-We converted the Viridian v05 dataset\cite{Hunt2024}, consisting of 4,484,157 SARS-CoV-2
-consensus sequences (125GiB over 48 FASTA files) and 30 metadata fields (1.4
-GiB TSV file) to VCF Zarr using a 
-reproducible Snakemake\cite{Koster2012snakemake} pipeline.
-We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using 
-MAFFT v7.475 \cite{Katoh2013},
-with the option \texttt{--keeplength} to retain the reference coordinate system.
-We then used \texttt{sc2ts import-alignments} and 
-\texttt{sc2ts import-metadata} to convert all alignments and metadata to VCF
-Zarr format, resulting in a single 401 MiB file 
-(around 77X smaller than gzip compressed FASTA\cite{Czech2025}).
-
-A classical tradeoff in bioinformatics is whether to store data 
-in sample-major (FASTA) or variant-major (VCF) format, determining
-whether the data can be accessed efficiently either by row or column
-of the variant matrix. Because VCF Zarr is chunked in two dimensions,
-we can now access either sample alignments or the variation
-data for a given site across all samples efficiently.
-While the VCF Zarr encoding of the MAFFT-aligned Viridian sequences
-and metadata can be accessed via native Zarr libraries in several
-popular programming languages (see Czech et al.\cite{Czech2025} 
-for an overview), the sc2ts Python library provides a convenient interface. 
-The \texttt{sc2ts.Dataset}
-class provides efficient access to alignments for samples and 
-variant data for sites along the genome, as well as methods to export
-subsets of the data to FASTA if required.
-Document~\ref{sec:supplement_sc2ts_usher_imputation} gives an example
-of using this interface, where we retrieve alignment data for millions 
-of samples in a few minutes on a standard desktop computer with minimal
-memory overhead. See Czech et al.\cite{Czech2025} for more details 
-and benchmarks.
-
-Both the pipeline and converted Viridian dataset are available for download;
-see Data and code availability for details.
-
-\subsubsection*{Sample filtering and preprocessing}
-The collection date for a sequence is an important piece of information
-for sc2ts. We used the Viridian
-consensus date (``Date\_tree``), based on information from COG-UK, GISAID, and
-ENA/SRA as our source. Samples that have missing or incomplete dates (e.g., 2020-01)
-were omitted, as were
-the $\sim$40,000 samples which had a collection date of December 31, 2020.
-As the sample count for this date is a major outlier for this period of
-the pandemic, it seems likely that this 
-batch is highly enriched for incorrect collection dates.
-Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
-confirmed that some of the samples were not collected in 2020.
-
-The bioinformatics community has curated a set of problematic sites
-for SARS-CoV-2 that are enriched for systematic sequencing errors or 
-high levels of homoplasy~\cite{DeMaio2020issues}. 
-Although these sites are recommended to be masked for UShER
-inference~\cite{Turakhia2020},
-the list has not been updated to take into account the systematic improvements 
-made by the Viridian dataset, and we therefore assembled a custom set of sites to mask. 
-To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
-without any site masking.
-Then, we identified the top 100 sites in terms of mutation counts in the ARG.
-Only 11 of these sites 
-(635, 8835, 11074, 11083, 15521, 16887, 21304, 21305, 21575, 21987 and 28253)
-were present in the problematic sites file
-(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}),
-and most sites in this list were typical in terms of the numbers of mutations.
-We therefore masked our 100 most-mutagenic sites to remove sequencing errors and improve
-matching performance in the LS HMM.
-
-Alignments read from the VCZ Zarr file were first preprocessed to mask any 
-non nucleotide characters (A,C,G or T) as missing data. 
-Then, we filtered 
-any samples that had missing data at more than 500 sites, excluding
-the 100 sites identified as problematic (this was 
-set to 10,000 sites for the initial period up to 2020-03-01 where sampling
-density is low).
-
 \subsubsection*{Li and Stephens haplotype-copying model}
 
-The LS model \cite{LiStephens2003} is an approximation of the coalescent with recombination
+LS model \cite{LiStephens2003} is an approximation of the coalescent with recombination
 that captures many of the key features of the joint processes of mutation and recombination.
@@
 collected on a certain date through the samples collected on earlier dates in a given ARG.
+
+\subsubsection*{Accelerating sample matching}
 
@@
 
+\subsubsection*{Pandemic scale alignment storage}
+To provide the efficient access to alignment data required for inference with sc2ts, we
+developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
+Briefly, Zarr is a storage format for scientific data,
+originally developed for the
+\textit{Anopheles gambiae}
+1000 Genomes Project~\cite{Anopheles2017genetic}, and
+is now seeing widespread adoption across the 
+sciences~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
+Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
+Zarr is essentially a simple mechanism for storing n-dimensional array
+data as a regular grid of compressed chunks, which provides 
+high levels of data compression, efficient access across multiple
+dimensions, and is ideally suited to modern cloud-based
+deployments~\cite{Czech2025}.
+
+We converted the Viridian v05 dataset\cite{Hunt2024}, consisting of 4,484,157 SARS-CoV-2
+consensus sequences (125GiB over 48 FASTA files) and 30 metadata fields (1.4
+GiB TSV file) to VCF Zarr using a 
+reproducible Snakemake\cite{Koster2012snakemake} pipeline.
+We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using 
+MAFFT v7.475 \cite{Katoh2013},
+with the option \texttt{--keeplength} to retain the reference coordinate system.
+We then used \texttt{sc2ts import-alignments} and 
+\texttt{sc2ts import-metadata} to convert all alignments and metadata to VCF
+Zarr format, resulting in a single 401 MiB file 
+(around 77X smaller than gzip compressed FASTA\cite{Czech2025}).
+
+A classical tradeoff in bioinformatics is whether to store data 
+in sample-major (FASTA) or variant-major (VCF) format, determining
+whether the data can be accessed efficiently either by row or column
+of the variant matrix. Because VCF Zarr is chunked in two dimensions,
+we can now access either sample alignments or the variation
+data for a given site across all samples efficiently.
+While the VCF Zarr encoding of the MAFFT-aligned Viridian sequences
+and metadata can be accessed via native Zarr libraries in several
+popular programming languages (see Czech et al.\cite{Czech2025} 
+for an overview), the sc2ts Python library provides a convenient interface. 
+The \texttt{sc2ts.Dataset}
+class provides efficient access to alignments for samples and 
+variant data for sites along the genome, as well as methods to export
+subsets of the data to FASTA if required.
+Document~\ref{sec:supplement_sc2ts_usher_imputation} gives an example
+of using this interface, where we retrieve alignment data for millions 
+of samples in a few minutes on a standard desktop computer with minimal
+memory overhead. See Czech et al.\cite{Czech2025} for more details 
+and benchmarks.
+
+Both the pipeline and converted Viridian dataset are available for download;
+see Data and code availability for details.
+
+\subsubsection*{Sample filtering and preprocessing}
+The collection date for a sequence is an important piece of information
+for sc2ts. We used the Viridian
+consensus date (``Date\_tree``), based on information from COG-UK, GISAID, and
+ENA/SRA as our source. Samples that have missing or incomplete dates (e.g., 2020-01)
+were omitted, as were
+the $\sim$40,000 samples which had a collection date of December 31, 2020.
+As the sample count for this date is a major outlier for this period of
+the pandemic, it seems likely that this 
+batch is highly enriched for incorrect collection dates.
+Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
+confirmed that some of the samples were not collected in 2020.
+
+The bioinformatics community has curated a set of problematic sites
+for SARS-CoV-2 that are enriched for systematic sequencing errors or 
+high levels of homoplasy~\cite{DeMaio2020issues}. 
+Although these sites are recommended to be masked for UShER
+inference~\cite{Turakhia2020},
+the list has not been updated to take into account the systematic improvements 
+made by the Viridian dataset, and we therefore assembled a custom set of sites to mask. 
+To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
+without any site masking.
+Then, we identified the top 100 sites in terms of mutation counts in the ARG.
+Only 11 of these sites 
+(635, 8835, 11074, 11083, 15521, 16887, 21304, 21305, 21575, 21987 and 28253)
+were present in the problematic sites file
+(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}),
+and most sites in this list were typical in terms of the numbers of mutations.
+We therefore masked our 100 most-mutagenic sites to remove sequencing errors and improve
+matching performance in the LS HMM.
+
+Alignments read from the VCZ Zarr file were first preprocessed to mask any 
+non nucleotide characters (A,C,G or T) as missing data. 
+Then, we filtered 
+any samples that had missing data at more than 500 sites, excluding
+the 100 sites identified as problematic (this was 
+set to 10,000 sites for the initial period up to 2020-03-01 where sampling
+density is low).
+
 \subsubsection*{Reconstructing a pandemic-scale ARG}
@@
 7. Assigning Pango lineages to non-sample nodes.
-
-\subsubsection*{Post hoc parsimony mapping}
-After the main inference step we mapped the variation data for 163 sites 
-in the alignments to the ARG using the Fitch-Hartigan 
-algorithm\cite{Fitch1971toward,Hartigan1973minimum} 
-implemented in tskit. In contrast to the main inference process where 
-deletion characters are masked as missing data, we allow for 
-5 different states in these parsimony calculations: A,C,G,T and -.
-By doing this, we can approximate the mutation events causing
-deletions, but with the clear caveat that each site is treated 
-independently. 
-Using data provided by Li et al.\cite{Li2023indels} (Table S1) 
-we chose 75 sites to remap
-
-performed a large-scale analysis of indels
-using a global GISAID dataset (sampled up to August, 2022).
-We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
-which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
-All of these indels are deletions and have high frequencies in the ARG
-(sequence count divided by the total number of samples),
-except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
-
-Based on the data of Table S4 of Li2023indels
 

### 2025-09-17 latex/main.tex
Updates from Overleaf

--- 2025-09-17 latex/main.tex
@@
 we sought to remove probable artifactual recombinants
-due to correlated bases
-being treated as individual sites by the HMM.
-Correlated bases can be caused by
+due to correlated bases, which can be caused by
 systematic errors in genome sequences
 associated with sequencing protocols \cite{Turakhia2020},
-multi-nucleotide substitutions \cite{DeMaio2024},
-genuine deletions, and
+multi-nucleotide substitutions (MNS) \cite{DeMaio2024}, and
 alignment errors (e.g., deletions due to MAFFT misalignment \cite{Hunt2024}).
@@
 but not in the focal parent.
+Illustrative copying patterns in Figure~\ref{fig:copying_pattern_examples}
+show the evidence for some QC-failing recombinants:
+(1) Q1, 686;
+(2) Q3, MNS at 13423-13424 \cite{DeMaio2024};
+(3) Q4, 22194 (possibly MAFFT alignment error \cite{Hunt2024}).
 
@@
 leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
-Evidence that this filter effectively flagged artifacts can be seen in the
-clear association between the AmpliSeq V1 kit and recombination events
-that we identify as artifactual.
-This primer scheme was used to sequence XX of the samples
-associated with an artifactual recombination event,
-in contrast to 15 of the ``robust'' events (Figure~\ref{fig:recombinant_qc}),
-strongly implicating it as a common source of problematic data.
-
-Three copying patterns in Figure~\ref{fig:copying_pattern_examples}
-Q1, 686
-Q3, 22029, Delta lineage-defining deletion (REF)
-There are others associated with a known amplicon dropout caused by this particular deletion.
-Q4, 22194
-True MNS
-Alignment error
+Evidence that this filter effectively flagged artifacts can be seen
+in the clear association between the AmpliSeq V1 kit and
+recombination events that we identified as artifactual.
 
@@
 \caption{
-Classification of ``robust'' and artifactual recombination events.
+Quality control of recombination events.
 Scatterplot of recombination events by the net number of supporting loci

### 2025-09-18 latex/main.tex
Updates to copying patterns and QC

@@
 multi-nucleotide substitutions (MNS) \cite{DeMaio2024}, and
-alignment errors (e.g., deletions due to MAFFT misalignment \cite{Hunt2024}).
+alignment errors \cite{Hunt2024}).
 
@@
 but not in the focal parent.
-Illustrative copying patterns in Figure~\ref{fig:copying_pattern_examples}
-show the evidence for some QC-failing recombinants:
+
+Of the 929 recombination events in the ARG,
+this measure identifies 543 as artifactual,
+leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
+Evidence that this filter effectively removed artifacts can be seen
+in the clear association between the AmpliSeq V1 kit and
+recombination events that we identified as artifactual.
+
+Different types of recombinants successfully removed are depicted
+by the illustrative copying patterns in Figure~\ref{fig:copying_pattern_examples}:
 (1) Q1, 686;
@@
 (3) Q4, 22194 (possibly MAFFT alignment error \cite{Hunt2024}).
-
-Of the 929 recombination events in the ARG,
-this measure identifies 543 as artifactual,
-leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
-Evidence that this filter effectively flagged artifacts can be seen
-in the clear association between the AmpliSeq V1 kit and
-recombination events that we identified as artifactual.
 

### 2025-09-18 latex/main.tex
Shorten section on daily sample cluster inference.

@@
 \subsubsection*{Tree inference from daily sample clusters}
-
 With tens of thousands of samples being added to the ARG per day,
-there are often clusters of hundreds of sequences attaching to the same node
-(or more generally, the same recombinant path).
-While some of these samples will require no extra mutations
+there are often clusters of hundreds of sequences with the same
+LS Viterbi solution. That is, large clusters of samples attach
+to the same node LS copying (or more generally, the same recombinant path)
+in the ARG. While some of these samples will require no extra mutations
 (as they are identical to the attachment node),
@@
 reflecting their evolutionary relationships.
-A natural way to infer these within-cluster evolutionary relationships
-is to use standard tree-building algorithms.
-We can infer a plausible phylogeny relating the samples in a cluster
-independently of the other samples in a daily batch
-and then attach the phylogeny (and mutations) to the ARG at the node
-identified by the LS model.
-
-Sc2ts builds a local phylogeny relating the samples in each cluster as follows.
-First, it infers a tree using a standard neighbor-joining
-algorithm\cite{Saitou1987}. We first compute the Hamming distance between
+We use a standard neighbor-joining algorithm\cite{Saitou1987}
+to infer these within-cluster evolutionary relationships.
+We first compute the Hamming distance between
 all pairs of samples using scipy~\cite{Virtanen2020scipy},
@@
 to compute the neighbor-joining tree.
-Then, mutations are mapped back onto it using the Fitch-Hartigan parsimony 
-algorithm \cite{Fitch1971toward,Hartigan1973minimum} implemented in tskit.
-Finally, the resulting tree is added to the ARG at an attachment node
-and the ARG is locally modified using heuristics to improve parsimony.
-[TODO we need to define these?]
+Then, mutations are mapped back onto this daily sample cluster tree using 
+the Fitch-Hartigan parsimony 
+algorithm\cite{Fitch1971toward,Hartigan1973minimum} implemented in tskit.
+Finally, the resulting tree and mutations are added to the ARG at the attachment node(s)
+specified by the shared LS copying path.
+
+\subsubsection*{Parsimony improving heuristics}
 

### 2025-09-18 latex/main.tex
Add parsimony hueristics section from preep.

@@
 \subsubsection*{Parsimony improving heuristics}
+Attaching trees built from the clusters of samples that copy from
+a particular node is an inherently greedy strategy
+and can produce inferences that are clearly unparsimonious.
+The final step in adding a daily batch of samples to the ARG
+is therefore to perform some local updates that target specific
+types of parsimony violations in the just-updated regions of the
+ARG. There are currently two parsimony-increasing operations
+applied, which we refer to as ``mutation collapsing'' and ``reversion
+pushing''.
+
+Given a newly attached node, mutation collapsing inspects its siblings from
+previous sample days to check if any of them share (a subset of) the mutations
+that
+it carries. If so, we increase the overall parsimony of the inference
+by creating a new node representing the ancestor that carried
+those shared mutations and make that new node the parent of the
+siblings carrying those shared mutations.
+The patterns of
+shared mutations between siblings can be complex, and the current
+implementation uses a simple greedy strategy for choosing
+the particular mutations to collapse.
+
+The reversion push operation inspects a newly added node to see
+if any of its mutations are ``immediate reversions''; that is,
+are reversions of a mutation that occurred on the new node's
+immediate parent. We increase the overall parsimony of the
+inference by ``pushing in'' a new node which descends from the
+original parent, and carries all its mutations except those
+causing the reversions on the newly added node.
 

### 2025-09-18 latex/main.tex
Updates from Overleaf

--- 2025-09-18 latex/main.tex
@@
 due to correlated bases, which can be caused by
-systematic errors in genome sequences
-associated with sequencing protocols \cite{Turakhia2020},
+bioinformatics errors in genome assembly \cite{Turakhia2020},
 multi-nucleotide substitutions (MNS) \cite{DeMaio2024}, and
@@
 this measure identifies 543 as artifactual,
-leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
+leaving 354 QC-passing events (Figure~\ref{fig:recombinant_qc}).
 Evidence that this filter effectively removed artifacts can be seen
@@
 
-Different types of recombinants successfully removed are depicted
-by the illustrative copying patterns in Figure~\ref{fig:copying_pattern_examples}:
-(1) Q1, 686;
-(2) Q3, MNS at 13423-13424 \cite{DeMaio2024};
-(3) Q4, 22194 (possibly MAFFT alignment error \cite{Hunt2024}).
+Furthermore,
+the illustrative copying patterns in Figure~\ref{fig:copying_pattern_examples}
+depict different types of artefacts that were successfully removed.
+Q1 shows an example of a recombinant caused by an MNS (A507T, T508C, G509A),
+which is a documented sequencing error due to incomplete read trimming \cite{DeMaio2024}.
+Q4 shows an example of a recombinant caused by a Delta lineage-defining deletion.
+The vast majority of the QC-failing recombinants has this deletion, and
+many cases are affected by a known amplicon dropout caused by this particular deletion.
+Q3 shows an example artefact caused by a combination of an MNS and a deletion.
 
@@
 Scatterplot of recombination events by the net number of supporting loci
-for the left parent (x-axis) and for the right parent (y-axis). The shaded
-regions highlight artifactual recombination events, which have fewer than four
-net supporting loci on one side or both sides of the suggested breakpoint.
-Colors indicate the primer scheme used for sequencing. The pie charts show
-breakdowns of the recombination events by primer scheme per quadrant.
-Recombination events associated with the origins of Pango X lineages are labeled.
+for the left parent (x-axis) and for the right parent (y-axis).
+The shaded regions highlight artifactual recombination events,
+which have fewer than four net supporting loci
+on one side or both sides of the suggested breakpoint.
+Colors indicate the primer scheme used for sequencing.
+The pie charts show breakdowns of the recombination events
+by primer scheme per quadrant.
+Recombination events associated with
+the origins of Pango X lineages are labeled.
 }
@@
 \begin{tabularx}{0.85\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
-Q1 & \includegraphics[height=\cprheight]{static/causedby-ERR10306608.png}\\
-Q2 & \includegraphics[height=\cprheight]{static/causedby-SRR15864165.png}\\
-Q3 & \includegraphics[height=\cprheight]{static/causedby-ERR6760245.png}\\
-Q4 & \includegraphics[height=\cprheight]{static/causedby-SRR21087725.png}\\
+Q1 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-663484.png}\\
+Q2 & \includegraphics[height=\cprheight]{static/RE_node-QCpass-427863.png}\\
+Q3 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-411345.png}\\
+Q4 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-748991.png}\\
 \end{tabularx}
@@
 Q2 illustrates a marginally good copying pattern.
-Q1, Q3, and Q4 illustrate copying patterns that fail the QC thresholds.
+Q1, Q3, and Q4 illustrate copying patterns that fail QC.
 }

### 2025-09-18 latex/main.tex
Fix typo

--- 2025-09-18 latex/main.tex
@@
 LS Viterbi solution. That is, large clusters of samples attach
-to the same node LS copying (or more generally, the same recombinant path)
+to the same node (or more generally, the same recombinant path)
 in the ARG. While some of these samples will require no extra mutations

### 2025-09-18 latex/main.tex
Add section on time travellers

@@
 
+\subsubsection*{Filtering time travellers}
+Time-travelling samples---those in which the recorded collection date differs
+substantially from their true collection date---pose significant problems
+for sc2ts, and are present at a significant frequency. For example, of the 
+43 Scorpio designations present in the Viridian dataset studied, 32 have 
+samples present with sampling dates in 2020, including lineages that do not
+arise until late 2022. Inserting these time-travellers into the ARG 
+at the claimed sampling date causes major issues, and they must therefore be 
+filtered out. We use the simple approach of filtering out samples that 
+exceed a given ``HMM cost'', i.e., requiring too many mutations or
+recombination breakpoints to plausibly represent the steady accumulation
+of diversity expected with dense sampling. 
+The HMM cost for the Viterbi solution for a sample under the LS model
+is $k$ times the number of recombination events plus the number of 
+mismatches, where $k$ is the recombination penalty (here, $k=4$).
+After some experimentation, we set the recombination penalty to 7, 
+disallowing samples that require 8 or more mutations (or, e.g., 1 recombination
+and 4 mutations, or 2 recombinations) from immediate inclusion (see next
+section) in the ARG. This approach has the benefit of filtering
+samples enriched for errors as well as time-travellers.
+
 \subsubsection*{Handling outlier samples}
-
 Although most samples can be explained by a small number of mutations in the ARG,
-there are two types of outlier samples which have unusually high numbers of mutations:
+there are two types of outlier samples which result in unusually high numbers of mutations:
 ``time travelers'' and ``saltation'' lineages.
@@
 leaving 354 QC-passing events (Figure~\ref{fig:recombinant_qc}).
+
+Furthermore,
+the example copying patterns in Figure~\ref{fig:copying_pattern_examples}
+illustrate different types of artefacts removed using this filter.
+Q1 shows an example of a recombinant caused by an MNS (A507T, T508C, G509A),
+which is a documented sequencing error due to incomplete read trimming \cite{DeMaio2024}.
+Q4 shows an example of a recombinant caused by a Delta lineage-defining deletion.
+The vast majority (XX, or XX\%) of the QC-failing recombinants has this deletion, and
+many cases are affected by a known amplicon dropout caused by this particular deletion.
+Q3 shows an example artefact caused by a combination of an MNS and a deletion.
+
 Evidence that this filter effectively removed artifacts can be seen
@@
 recombination events that we identified as artifactual.
-
-Furthermore,
-the illustrative copying patterns in Figure~\ref{fig:copying_pattern_examples}
-depict different types of artefacts that were successfully removed.
-Q1 shows an example of a recombinant caused by an MNS (A507T, T508C, G509A),
-which is a documented sequencing error due to incomplete read trimming \cite{DeMaio2024}.
-Q4 shows an example of a recombinant caused by a Delta lineage-defining deletion.
-The vast majority of the QC-failing recombinants has this deletion, and
-many cases are affected by a known amplicon dropout caused by this particular deletion.
-Q3 shows an example artefact caused by a combination of an MNS and a deletion.
 

### 2025-09-18 latex/main.tex

--- 2025-09-18 latex/main.tex
@@
 applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.
-
-Sc2ts uses collection dates to build a time-resolved ARG.
-When examining preliminary ARGs,
-we found samples that introduced large numbers of mutations (and reversions),
-likely due to errors in their collection dates
-(for example, a large batch of samples dated 31 December, 2020 contains samples from Omicron BA.1,
-which is estimated to have emerged in November, 2021; REF).
-To preclude adding such ``time traveling'' samples, we filtered samples by ``HMM cost''.
-We define the ``HMM cost'' of a sample to be
-the total penalty cost for switches and mismatches (mutations) in its copying path.
-We parameterize the LS model such that each mismatch has a penalty cost of one
-and a switch has a penalty cost of $k$ (here, set to 4),
-which determines how much mutation is favored over recombination
-when accounting for a sample (Figure~\ref{fig:methodology}).
-If the HMM cost of a sample exceeds a threshold (here, set to 7),
-sc2ts does not immediately add the sample to the ARG.
-
-This approach resulted in preliminary ARGs that had noticeably fewer time travelers,
-but we found that it also excluded the major ``saltation'' events that characterize many VOCs.
-These saltation lineages are characterized by a high number of accumulated mutations,
-and are hypothesized to emerge from chronic infections
-during which the virus is exposed to individual-specific immune selective pressures \cite{Markov2023}.
-Though uncommon, such lineages drove large epidemic waves during the pandemic (notably Alpha, Delta, and Omicron).
-To distinguish time travelers with erroneous dates from true saltation lineages,
-we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added.
-If on subsequent days these lineages form a sufficiently large group of samples,
-sc2ts adds the group back to the ARG.
-This ``retrospective sample grouping'' successfully captured the emergence of
-many saltation lineages such as Alpha (Figure~\ref{fig:subgraph_alpha}).
-The emergence of the Delta, BA.1, and BA.2 lineages required some manual intervention,
-in which specific ``seed'' samples were identified for unconditional inclusion
-(for details of how seed samples were chosen, see STAR Methods).
-First arising in geographic regions with low sampling density,
-the initial origins of these VOC are challenging to infer (REF).
-With the addition of these manual seeds, we successfully incorporated 
-these challenging events (Figures~\ref{fig:subgraph_delta},
-\ref{fig:subgraph_ba1_ba2}) and 
-all other major lineages (Figure~\ref{fig:samples_per_day}) 
-into the ARG.
+[TODO add brief summary of sample filtering and how we handle VOCs, 
+refing Methods for details]
 
@@
 
-\subsubsection*{Handling outlier samples}
-Although most samples can be explained by a small number of mutations in the ARG,
-there are two types of outlier samples which result in unusually high numbers of mutations:
-``time travelers'' and ``saltation'' lineages.
-Time traveling samples are caused by errors in their collection dates.
-For example, if a sample is actually collected 10 months after its reported collection date,
-then 20 mutations (instead of, say, 0 or 1) are needed to explain it
-using the ARG available on its collection date,
-assuming a substitution rate of 2 events per genome per month.
-Samples from ``saltation'' lineages (particularly the major VOCs) are characterized
-by high numbers of mutations by the time of their detection.
-Both types of outlier samples are poorly fit to the ARG when they appear on their collection dates.
-To handle these outlier samples, we introduce three mechanisms:
-(1) HMM cost-based rejection,
-(2) retrospective sample grouping, and
-(3) unconditional seeding.
-
-First, we introduce a rejection mechanism to prevent samples
-that are generally poorly explained by the ARG
-using the LS model parameterised by a specific value for the recombination penalty ($k$).
-For each sample, an HMM cost is calculated based on the number of switches and mismatches
-in its most likely copying path under the LS model:
-The sum of the penalty cost for the switches ($k$)
-and the penalty cost of the mismatches (implicitly, equal to one).
-To build the ARG described here, we used $k$ of 4 and a maximum HMM cost of 7.
-This combination of parameter values rejected samples which have no switch
-but 8 mismatches as well as samples which have one switch and 4 mismatches.
-It is important to note that this does not necessarily mean
-that samples with more than 8 mismatches
-or samples with two or more switches cannot be incorporated into the ARG.
-Such samples can be added back using a separate mechanism (see below).
-We find that although it effectively removes time travelers with poor sample support (e.g. singletons),
-the mechanism can filter samples too aggressively.
-For example, in preliminary ARGs, it led to exclusion of major saltation VOCs (e.g. Alpha).
-
-Second, to prevent exclusion of the important saltation VOCs
-(and less important but well supported poorly-fit lineages),
-we introduce ``retrospective sample grouping''.
-The idea is to gather evidence for a lineage within a time window rather than on a single day.
-After adding a batch of samples, the samples initially rejected on their collection dates
-due to high HMM costs are reconsidered for attachment to the ARG if
-(a) they are collected within a retrospective time window (the past 7 days);
-(b) they have identical copying paths; and
-(c) they have shared immediate reversions
-(attaching a new sample to the top or the bottom of a long branch).
-Sample groups satisfying these criteria are attached to the ARG if
-(a) they have at least 10 samples;
-(b) they are collected on at least 3 different dates
-(to avoid adding batches of time travelers having the same collection date); and
-(c) they have at least two shared ``non-immediate'' reversions
-(so as to prevent unrelated time travelers from being grouped together).
-Furthermore, phylogenetic tree quality metrics are used to preclude groups of samples
-that are unlikely to be closely related.
-Specifically, the tree of a sample group was excluded if
-(a) it has more than two recurrent mutations or
-(b) it has more than 5 mutations per sample on average.
-
-Retrospective sample grouping successfully captured most saltation lineages,
-but integrating Delta and Omicron into the ARG required further manual intervention.
-To ensure that Delta and Omicron are added, we introduce a third mechanism, ``unconditional seeding''.
-The idea is to choose suitable samples belonging to each VOC,
-and attach them to the ARG unconditionally (i.e. without filtering using HMM cost).
-We sought ``seed'' samples for Delta, Omicron BA.1, and BA.2.
-For Delta, we used two sets of seeds:
-(1) 8 samples assigned to B.1.617.2, and
-(2) 9 additional samples assigned to its sibling lineage B.1.617.1.
-Specifically, these samples are the earliest samples present in the Viridian v04 dataset
-that were used for the designation of the VOCs.
+\subsubsection*{Inserting saltation lineages}
+The HMM cost threshold approach outlined in the previous section is 
+effective at removing time-travelling samples, but it also filters out 
+important evolutionary events such as 
+the major ``saltations'' that characterize many VOCs.
+These lineages are characterized by a high number of accumulated mutations,
+and are hypothesized to emerge from chronic infections
+during which the virus is exposed to individual-specific immune selective
+% Other refs here for this Sally?
+pressures\cite{Otto2021origins,Markov2023}.
+To distinguish time travelers from true saltation lineages,
+we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added.
+If these samples form a sufficiently plausible group over successive days (under criteria
+described in the next paragraph) sc2ts adds the group to the ARG as a 
+``retrospective sample group'' using same mechanisms as standard
+daily sample clusters.
+
+Distinguishing a truly emerging outbreak from different forms of
+correlated error is non-trivial, and sc2ts has 5 parameters to fine
+tune the behaviour. Firstly, we require at least 10 samples in the
+group to ensure there is sufficient support over a time window of 
+7 days. Then, we build a local tree of the samples and consider
+some tree metrics to evaluate whether the samples within the 
+group are closely related. If we have at least 2 mutations 
+shared by all samples, there's at most 2 recurrent mutations
+within the group and at most 5 mutations unique to each sample, 
+the retrospective group is accepted for inclusion.
+Although this process is highly hueristic, it successfully 
+captured the majority of saltational lineages
+such as Alpha (Figure~\ref{fig:subgraph_alpha}), and the 
+vast majority of samples were added following the standard
+daily sample grouping mechanisms. 
+% https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/logs/sc2ts_v1_2023-02-21_pr_pp_mp_aph_bps_pango_dated_recombinants_report.ipynb
+In total, 882,398 daily 
+sample groups were added to the ARG, while only 100 were
+added via this retrospective grouping mechanism.
+
+The Delta, BA.1, and BA.2 lineages were 
+particularly challenging, however, as the initial emergences of 
+the saltation lineages were not densely sampled. In this case,
+there is little information for the retrospective sampling grouping mechanism
+to work with and it is very difficult to distinguish short-term time-travellers
+from true outbreak samples. For these cases, we chose specific
+high-quality samples with confident dates as ``seeds`` which are 
+unconditionally included in the ARG on the specified date.
+
+% We used more seeds, but in practise it doesn't matter and it's just
+% complicating the story.
+For Delta, we used two seeds: one for B.1.617.1 and 1 samples assigned to
+B.1.617.2, [with more details and some references to back up our assertions.
+There's some specific info in the Viridian paper on this, I'm sure. 
+We should mention long branch splitting here too.]
+The resulting subgraph can be see in Figure~\ref{fig:subgraph_delta}.
+
+[TODO revise this paragraph to give some background info, and reference
+the Viridian work]
 For Omicron BA.1 and BA.2, we used SRR17041376 and SRR17461792, respectively.
@@
 and to get established in the ARG.
-For the full list of seeds, see
-\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}
-
-In practice, it is difficult to find optimal seeds for saltation lineages,
-which ideally would require no reversion to be added to the ARG
-and to which sampled descendents would be added with no reversion.
-Considering poor early sampling and the presence of time travellers for Delta, Omicron BA.1, and BA.2,
-using suboptimal seeds is likely the best that we can do in practice.
-The consequence is that there would be some reversions leading to the samples attached to the seed(s).
-Nevertheless, it is a better alternative to not using seeds,
-as there would be far more reversions (as seen in preliminary ARGs built without seeding).
-The caveat of taking this approach is that one must be more careful
-when analysing the samples around the seed(s) in the ARG.
+See Figure~\ref{fig:subgraph_ba1_ba2} for an overview of the resulting 
+subgraph.
+
+With the combination of these mechanisms---gradual accumulation 
+of diversity through daily sample groups, retrospective group
+inclusion and manual seeding---we successfully incorporated all major
+lineages in the dataset into the ARG (Figure~\ref{fig:samples_per_day}).
 

### 2025-09-18 latex/main.tex

@@
 applied, which we refer to as ``mutation collapsing'' and ``reversion
-pushing''.
+pushing'' (Figure~\ref{fig:parsimony_ops}).
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=0.85\linewidth]{tree_ops}
+\caption{
+Parsimony improving heuristics. (A) Mutation collapsing. Mutation A100G is
+shared by two siblings, and we create a new node to represent the ancestor
+on which this mutation occured. (B) Reversion pushing. Mutation A128C is
+immediately reverted by C128A, and we create a new node to represent the 
+ancestor that did not carry A128C.
+}
+\label{fig:parsimony_ops}
+\end{figure}
+
 \end{document}

### 2025-09-19 latex/main.tex
Add counts for parsimony
QC section pass

@@
 the particular mutations to collapse. 
+119,844 nodes were 
+added to the ARG during primary inference by mutation collapsing.
 
@@
 causing the reversions on the newly added node.
+52,336 nodes were added to the ARG during primary inference by
+reversion pushing.
 
@@
 by rematching the inferred recombinant sequence against the ARG pruned back
-to include only samples before the date of the recombinant. Additionally,
- each recombination node is examined to determine if any mutations 
-are shared [TODO Yan write this up].
-We refer to this process as ``long branch splitting''. 
+to include only samples before the date of the recombinant.
+
+Where the recent ancestry of a recombinant includes a branch containing
+large numbers of mutations, it is possible
+that a non-recombinant solution could be found by postulating an intermediate
+node along that branch, that contains only a subset of those mutations.
+In particular, for a given mutation-rich branch, artefactual recombinants may
+be found in which one recombinant region matches to
+a node like the parent whereas another region matches to a node like the child
+of that branch. Such recombinants can be identified by the
+presence of recurrent or reversion mutations in the left or right parent branches.
+To help remove these recombination events from the ARG, we note the genomic positions
+of mutations associated with the recombinant, and identify branches
+in their ancestry which have mutations at the same positions. Of these branches,
+we chose the one that contains the largest total number of mutations as
+a candidate for intermediate node creation. This is done by sorting the
+mutations on the branch such that the ones that match the allelic state in the
+recombinant are placed above the intermediate node, and the others are left
+below the node. We refer to this process as ``long branch splitting'', and it
+creates additional nodes which are available for recombinant rematching.
+
 % Data from
 % https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/logs/sc2ts_v1_2023-02-21_rewire-recombinants-report.ipynb
-From this, 58 recombinants were identified as being more parsimoniously 
+Using recombinant-rematching, 58 recombinants were identified as being more parsimoniously 
 explained by matching to existing nodes in the ARG (inserted by 
@@
 resulted in 16 additional nodes, 182 fewer mutations and reduced
-the number of trees along the genome from 348 to 317.
+the number of trees along the genome from 348 to 317. The long branches
+involved in the origins of Delta and BA.2 originally joined
+B.1 to B.1.167 and B.1.1 to BA.1 respectively; they can be seen
+in Figure \ref{fig:subgraph_delta} and \ref{fig:subgraph_ba1_ba2}
+with inserted intermediate nodes leading to the Delta and BA.2
+origination nodes.
 
@@
 
-\subsubsection*{Quality control of recombinants}
+\subsubsection*{Quality control of recombination events}
 
@@
 
-Before characterizing recombination events in detail,
-we sought to remove probable artifactual recombinants
-due to correlated bases, which can be caused by
-bioinformatics errors in genome assembly \cite{Turakhia2020},
-multi-nucleotide substitutions (MNS) \cite{DeMaio2024}, and
-alignment errors \cite{Hunt2024}).
-
-We characterized artifactual recombination events
-as having fewer than four such loci supporting the left or right parent
-on either side of a suggested breakpoint.
-We treated adjacent and near-adjacent sites
-(separated by at most three bases) as single loci.
-Also, when counting loci supporting a focal parent,
-we penalized loci where the character(s)
-in the recombinant matches that in the non-focal parent
-but not in the focal parent.
-
-Of the 929 recombination events in the ARG,
-this measure identifies 543 as artifactual,
-leaving 354 QC-passing events (Figure~\ref{fig:recombinant_qc}).
-
-Furthermore,
-the example copying patterns in Figure~\ref{fig:copying_pattern_examples}
-illustrate different types of artefacts removed using this filter.
-Q1 shows an example of a recombinant caused by an MNS (A507T, T508C, G509A),
-which is a documented sequencing error due to incomplete read trimming \cite{DeMaio2024}.
-Q4 shows an example of a recombinant caused by a Delta lineage-defining deletion.
-The vast majority (XX, or XX\%) of the QC-failing recombinants has this deletion, and
+Sequencing errors in genome assemblies \cite{Turakhia2020},
+read misalignments \cite{Hunt2024}, and
+multi-nucleotide substitutions (MNS \cite{DeMaio2020issues})
+can cause short regions of a sample genome to match more closely
+to an alternative genome rather than the true parent.
+Such genomic regions can be mistaken for evidence of recombination,
+because the LS HMM counts each base separately.
+Fortunately, artefactual recombinants caused by this effect can be identified
+because they affect multiple adjacent or near-adjacent bases simultaneously.
+
+We devised a quality control (QC) measure based on this observation, and
+applied it to each putative recombination event in the ARG.
+We listed the genomic positions at which the suggested parents differed in allelic state.
+Where such sites clustered closely (3 or fewer bases apart),
+we treated the entire cluster of sites as a single \emph{locus}.
+For each genomic region identified by the HMM as having a single parent,
+we calculated the "net number of supporting loci" by counting the number of loci in the region
+that supported the suggested parent minus the number that did not support it.
+Finally, with all the recombination events in the ARG having a single breakpoint,
+we calculated the net number of supporting loci on the left and right sides of each breakpoint, and
+used this to plot the 855 recombination events in Figure~\ref{fig:recombinant_qc}.
+
+This simple QC measure effectively differentiates
+artefactual recombination events from those that are well supported.
+All the sc2ts-detected recombination events associated with the Pango X lineages (section XX)
+occur in the upper right of Figure~\ref{fig:recombinant_qc},
+with high numbers of net supporting loci on both the right and left parents.
+In contrast, the other regions of the plot, where one or both parents are less well supported,
+are dominated by recombination events associated with samples sequenced using a specific protocol (AmpliSeq).
+Coupled with closer inspection of copying patterns (below),
+we chose a QC cutoff of four or more net supporting loci on both the left and right side of the breakpoint.
+This leads to 354 QC-passing recombination events (Figure~\ref{fig:recombinant_qc}, upper right quadrant),
+compared to 501 recombination events for which there is less support and which may be artefactual (shaded quadrants).
+
+Inspection of the copying patterns for all the recombination events
+clearly identified associations between recombination events
+that failed this QC check and known problematic loci in the SARS-CoV-2 genome.
+Figure~\ref{fig:copying_pattern_examples} illustrates examples of these copying patterns (Q1 to Q4),
+in which adjacent sites are underlined in red, and near-adjacent sites (within 3 bases) in orange.
+A continuous red+orange underline thus indicates a region treated as a single locus.
+Q1 shows a well supported recombination event that passed QC (albeit marginally),
+having a net number of supporting loci of 4 on both sides of the breakpoint.
+Q2 shows an artefactual recombination event caused by a 9-base Delta lineage-defining deletion.
+There are XX (XX\%) of the recombination events involving this deletion that failed QC, and
 many cases are affected by a known amplicon dropout caused by this particular deletion.
-Q3 shows an example artefact caused by a combination of an MNS and a deletion.
-
-Evidence that this filter effectively removed artifacts can be seen
-in the clear association between the AmpliSeq V1 kit and
-recombination events that we identified as artifactual.
+Q3 shows an artefactual recombination event caused by a combination of
+this Delta lineage-defining deletion and a known MNS (see below).
+Finally, Q4 shows an artifactual recombination event associated with the A507T-T508C-G509A MNS,
+which has been identified as a sequencing error due to incomplete read trimming \cite{DeMaio2024}.
+There are XX recombination events involving this particular MNS, of which XX\% failed QC.
 

### 2025-09-19 latex/main.tex

@@
 applied it to each putative recombination event in the ARG.
-We listed the genomic positions at which the suggested parents differed in allelic state.
+We examined the genomic positions at which the suggested parents differed in allelic state.
 Where such sites clustered closely (3 or fewer bases apart),
@@
 artefactual recombination events from those that are well supported.
-All the sc2ts-detected recombination events associated with the Pango X lineages (section XX)
+All the sc2ts-detected recombination events associated with the Pango X lineages (Table~\ref{tab:pango_x_lineages})
 occur in the upper right of Figure~\ref{fig:recombinant_qc},
@@
 are dominated by recombination events associated with samples sequenced using a specific protocol (AmpliSeq).
-Coupled with closer inspection of copying patterns (below),
+Coupled with closer inspection of copying patterns (see below),
 we chose a QC cutoff of four or more net supporting loci on both the left and right side of the breakpoint.
 This leads to 354 QC-passing recombination events (Figure~\ref{fig:recombinant_qc}, upper right quadrant),
-compared to 501 recombination events for which there is less support and which may be artefactual (shaded quadrants).
+compared to 501 recombination events
+for which there is less support and which may be artefactual (shaded quadrants).
 
 Inspection of the copying patterns for all the recombination events
-clearly identified associations between recombination events
-that failed this QC check and known problematic loci in the SARS-CoV-2 genome.
-Figure~\ref{fig:copying_pattern_examples} illustrates examples of these copying patterns (Q1 to Q4),
+clearly identified associations between recombination events that failed QC and
+known problematic loci in the SARS-CoV-2 genome.
+Figure~\ref{fig:copying_pattern_examples} shows illustrative copying patterns
+(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}),
 in which adjacent sites are underlined in red, and near-adjacent sites (within 3 bases) in orange.
@@
 Q2 shows an artefactual recombination event caused by a 9-base Delta lineage-defining deletion.
-There are XX (XX\%) of the recombination events involving this deletion that failed QC, and
-many cases are affected by a known amplicon dropout caused by this particular deletion.
+There are XX (XX\%) of the QC-failing recombination events involving this deletion, and
+many cases are affected by a known amplicon dropout caused by this particular deletion \cite{Sanderson2021}
 Q3 shows an artefactual recombination event caused by a combination of
@@
 Scatterplot of recombination events by the net number of supporting loci
-for the left parent (x-axis) and for the right parent (y-axis).
+on the left parent (x-axis) and the right parent (y-axis).
 The shaded regions highlight artifactual recombination events,
-which have fewer than four net supporting loci
+which have fewer than 4 net supporting loci
 on one side or both sides of the suggested breakpoint.
@@
 The pie charts show breakdowns of the recombination events
-by primer scheme per quadrant.
+by primer scheme per quadrant (labeled Q1 to Q4).
 Recombination events associated with
@@
 \begin{tabularx}{0.85\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
-Q1 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-663484.png}\\
-Q2 & \includegraphics[height=\cprheight]{static/RE_node-QCpass-427863.png}\\
+Q1 & \includegraphics[height=\cprheight]{static/RE_node-QCpass-427863.png}\\
+Q2 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-748991.png}\\
 Q3 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-411345.png}\\
-Q4 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-748991.png}\\
+Q4 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-663484.png}\\
 \end{tabularx}
 \caption{
-Examples of copying patterns.
-Q1 to Q4 are examples drawn from
-the labeled quadrants in Figure~\ref{fig:recombinant_qc}.
-Q2 illustrates a marginally good copying pattern.
-Q1, Q3, and Q4 illustrate copying patterns that fail QC.
+Illustrative copying patterns.
+Q1 to Q4 are examples drawn from the labeled quadrants in Figure~\ref{fig:recombinant_qc}.
+Q1 depicts the copying pattern of a recombination event that marginally passed QC,
+whereas the others depict the copying patterns of recombination events that failed QC.
 }

### 2025-09-19 latex/main.tex
Update methods/comparison with UShER

--- 2025-09-19 latex/main.tex
@@
 \subsubsection*{Ancestral Recombination Graphs}
-
+\label{sec:args}
 %%% This paragraph needs to be rewritten.
@@
 \subsubsection*{ARG postprocessing}
+\label{sec:arg_postprocessing}
 The first step in ARG postprocessing is to examine putative recombinants to
@@
 UShER tree from MAT~\cite{McBroome2021} to tskit format, which has similar
-properties but is more general (see the section ``Ancestral Recombination
-Graphs``).
+properties but is more general 
+(see the \nameref{sec:args} section).
 We first downloaded the tree in JSON format (tree.all\_viridian.202409.jsonl)
-from Figshare
-and converted the ARG topology. We then downloaded the protobuf formatted
-tree (tree.all\_viridian.202409.pb.gz), and exported the mutations to 
+and converted the tree topology to tskit node and edge table 
+descriptions. We then downloaded the protobuf formatted
+tree (tree.all\_viridian.202409.pb.gz), exported the mutations to 
 nucleotide format using \texttt{matUtils summary --translate}
 (usher version 0.6.3), and translated
-these mutations into tskit format. The resulting tszip file 
+these mutations into tskit site and mutation tables. 
+The resulting tszip file 
 (usher\_viridian\_v1.0.trees.tsz) contains 5,345,019 nodes, 3,364,841
@@
 There was an exact match between the non-ambiguous nucleotide calls 
-at 27469 sites. At the remaining 39 sites there were a significant 
+at 27,469 sites. 
+% Usher tree has 27,508 vs sc2ts 29,893, or 2385 fewer sites 
+At the remaining 39 sites for which the UShER tree 
+had mutations there were a significant 
 number of gap characters (deletion calls), suggesting that the mismatches
@@
 
-We then computed intersection ARGs for sc2ts and UShER by 
-``simplifying''\cite{Kelleher2018efficient,Wong2024} 
-both to the 2,475,418 samples and 27,507 sites that are shared.
+We then computed intersection ARGs for sc2ts
+(sc2ts\_viridian\_inter\_v1.2.trees.tsz) and UShER 
+(usher\_viridian\_inter\_v1.2.trees.tsz) 
+by 
+simplifying\cite{Kelleher2018efficient,Wong2024} 
+both to the 2,475,418 samples and 27,507 sites that are shared
+(the UShER tree has mutations at 27,508 sites and sc2ts ARG has 29,893).
+As there was no documented means of deriving node dates from 
+the UShER tree, and we wish to visually compare 
+the phylogenetic backbones, we
+also dated the intersection tree by fixing
+the sample dates using the ``Date\_tree'' metadata field, 
+and using tsdate (version 0.2.3) to date internal nodes.
+We did not date the main UShER tree in this way because
+of the time-travelling samples present.
 
 All conversions were performed as part of the reproducible 
-postprocessing pipeline described in [section?]. The full UShER
-tree and the intersection files are available for download on Zenodo
-in tszip format.
+pipeline described in the \nameref{sec:arg_postprocessing} section. 
+The full UShER
+tree and the intersection files are available for download 
+in tszip format (see \nameref{sec:data_and_code_availability}).
 
@@
 for the majority of the 2026 Pango lineages in the ARG.
-We first identify an ``origination node" for a focal Pango lineage,
-defined as the oldest internal node in the ARG
-whose descendants in any tree include more than 50\% of the Pango samples of the focal type,
-and where the internal node is itself designated as the focal Pango.
+We first identify an ``origination node'' for each Pango lineage,
+defined as the oldest node in the ARG with that Pango designation
+whose descendants in any tree include more than 50\% of the Pango samples of
+the focal type.
 The representative sample is then chosen as the oldest sample descending from the origination node
@@
 
-We compare topology in the simplified ARG against the UShER tree, built from the Viridian dataset.
-To remove time traveling samples from the UShER tree,
-we reduce it to 2475419 samples shared with the main sc2ts ARG
-and date the internal nodes by running tsdate,
-with no rescaling and the root node constrained to the established date of the UShER root (2020-01-16).
-To compare against the ARG, we simplify this UShER tree down to the 1667 representative ARG samples.
+We compare topology of the simplified ARG against the intersection UShER tree
+(which has been dated) simplified down to the same 1667 samples.
 

### 2025-09-19 latex/main.tex
Updates from Overleaf

--- 2025-09-19 latex/main.tex
@@
 many cases are affected by a known amplicon dropout caused by this particular deletion \cite{Sanderson2021}
-Q3 shows an artefactual recombination event caused by a combination of
-this Delta lineage-defining deletion and a known MNS (see below).
+Q3 shows an artefactual recombination event caused by
+a combination of this Delta deletion and a known MNS (see below).
 Finally, Q4 shows an artifactual recombination event associated with the A507T-T508C-G509A MNS,

### 2025-09-19 latex/main.tex

--- 2025-09-19 latex/main.tex
@@
 Q2 shows an artefactual recombination event caused by a 9-base Delta lineage-defining deletion.
-There are XX (XX\%) of the recombination events involving this deletion that failed QC, and
-many cases are affected by a known amplicon dropout caused by this particular deletion.
-Q3 shows an artefactual recombination event caused by a combination of
-this Delta lineage-defining deletion and a known MNS (see below).
+There are XX (XX\%) of the QC-failing recombination events involving this deletion, and
+many cases are affected by a known amplicon dropout caused by this particular deletion \cite{Sanderson2021}
+Q3 shows an artefactual recombination event caused by
+a combination of this Delta deletion and a known MNS (see below).
 Finally, Q4 shows an artifactual recombination event associated with the A507T-T508C-G509A MNS,

### 2025-09-19 latex/main.tex
Add snappy summary of ARGs to intro

@@
 
+Ancestral recombination graphs (ARGs) describe the reticulate 
+genealogies of sampled sequences in a recombining species\cite{Wong2024}.
+While ARGs have been of theoretical interest in population genetics 
+for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
+recent breakthroughs in simulation\cite{Kelleher2016,Kelleher2018efficient}
+and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
+Wohns2022,Zhang2023,Gunnarsson2024,Deng2024robust} have led to a surge 
+of interest in their application\cite{Brandt2024,Lewanski2024,Nielsen2025}.
+
+% TODO merge this in
 Ancestral Recombination Graphs (ARGs) are reticulate genealogies
@@
 \label{sec:args}
+
+describe how a set of sample sequences
+relate to each other at each position along the genome in a recombining
+species 
+lots of interest
+
+, and are currently the subject of intense research interest 
+(Brandt et al. 2024, Lewanski et al. 2024, Wong et al. 2024, Nielsen et al. 2025).
+
 %%% This paragraph needs to be rewritten.
@@
 under the influence of recombination.
+
+Until recently, ARGs lacked software support and
+sufficiently scalable inference methods to be of practical use.
+Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023},
+with high levels of recombination.
+These advances now enable evolutionary genealogies to be built quickly and at scale, and
+there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.
 

### 2025-09-19 latex/main.tex
Pass through ARG section

--- 2025-09-19 latex/main.tex
@@
 
+Ancestral recombination graphs (ARGs) describe the reticulate 
+genealogies of sampled sequences in a recombining species\cite{Wong2024}.
+While ARGs have been of theoretical interest in population genetics 
+for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
+recent breakthroughs in
+simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019}
+and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
+Wohns2022,Zhang2023,Gunnarsson2024,Deng2024robust} have led to a surge 
+of interest in their application\cite{Brandt2024,Lewanski2024,Nielsen2025}.
+
+% TODO merge this in
 Ancestral Recombination Graphs (ARGs) are reticulate genealogies
@@
 geographic spread of the virus through time
-\cite{grundler2025geographic}. This is in addition to simpler
+\cite{Grundler2025geographic}. This is in addition to simpler
 analyses, such as calculating mutation spectra or querying the number
@@
 \label{sec:args}
-%%% This paragraph needs to be rewritten.
-Originally introduced (Griffiths, 1991; Griffiths and Marjoram, 1997)
-to define an alternative formulation of the coalescent with recombination (Hudson, 1983),
-the term ``Ancestral Recombination Graph'' has come to be used
-more generally to describe not just realizations of the model,
-but any recombinant genetic ancestry (Minichiello and Durbin, 2006).
-While there is some subtlety in the details \cite{Wong2024},
-we can think of an ARG as being any graph
-that encodes the reticulate genetic genealogy of a sample of sequences
-under the influence of recombination.
-
-The ``succinct tree sequence'' (STS) is an ARG data structure
-that is both general (in terms of the types of ancestry that can be described)
-and computationally efficient \cite{Wong2024}.
-Initially developed to facilitate large-scale coalescent simulations \cite{Kelleher2016},
-STS has been extended and applied to
-forward-time simulations \cite{Kelleher2018efficient,Haller2019},
-calculation of population genetics statistics \cite{Ralph2020} and
-ARG inference \cite{Kelleher2019,Wohns2022}.
-The STS is based on a simple tabular representation,
-which defines a set of nodes, edges, sites and mutations.
-A node represents a particular genome, which may be an observed sample or an inferred genetic ancestor.
-The genetic inheritance between a pair of nodes along a segment of genome
-is defined by the edge ($l$, $r$, $p$, $c$),
-which states that child node $c$ inherited its genome from parent node $p$
-from left coordinate $l$ to right coordinate $r$.
-A site defines a position on the genome and the ancestral state (allele) at that site.
-A mutation records the site and node IDs where a mutation occurs and the derived state (allele).
-
-Given the node, edge, site and mutation tables,
-we can efficiently construct the local genealogical trees along the genome (arising from recombination)
-and perform a range of calculations efficiently
-by reasoning about the differences between these local trees \cite{Kelleher2016,Ralph2020}.
-These algorithms have led to performance increases of several orders of magnitude over
-the state-of-the-art in a range of applications
-\cite{Kelleher2016,Kelleher2018efficient,Kelleher2019,Ralph2020,Baumdicker2022}.
-The STS encoding is also very concise, allowing, for example,
-for millions of complete human genomes to be stored in a few GB of space \cite{Kelleher2019},
-and so it is particularly well suited to representing large SARS-CoV-2 genealogies.
-
-The SARS-CoV-2 ARGs inferred using sc2ts can be conveniently and efficiently analyzed
-using the mature and feature-rich tskit library \cite{Kelleher2018efficient,Ralph2020}.
-Tskit is widely used in population genetics, and has been incorporated into many downstream applications
-(e.g., \cite{Haller2019,Speidel2019,TerasakiHart2021,Fan2022,Korfmann2023,Mahmoudi2022,Petr2023,Rasmussen2023,Zhang2023}).
+The topology of an Ancestral Recombination Graph (ARG) can be defined 
+% genome seems better than "haplotype" here, and will be clear what it means
+% in the viral context?
+as a set of \emph{nodes}, representing a genome that existed at some specific time,
+and \emph{edges} that describe the genetic inheritance relationships between those 
+genomes~\cite{Wong2024}. 
+Each edge defines the parent and child nodes and a genomic interval 
+over which inheritance occurs. The ``succinct tree sequence'' ARG encoding defines 
+these relationships in a simple and concise tabular format, 
+sufficient to represent arbitrarily complex patterns of
+recombination~\cite{Wong2024} and providing the basis for a range of highly
+efficient data processing
+algorithms~\cite{Kelleher2016,Ralph2020,Kelleher2020coalescent,Grundler2025geographic,Lehmann2025on}.
+The mature and feature-rich tskit library has interfaces in Python, C and Rust,
+extensive documentation,
+and is the basis of a growing software ecosystem of 
+simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
+visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
+inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022}
+% More here? I'm sure there's others
+and data analysis~\cite{Fan2022,Nowbandegani2023extremely}
+tools.
+
+Similarly to the Mutation Annotated Tree (MAT)~\cite{McBroome2021} format used 
+by UShER, tskit incorporates mutational information directly in the data 
+model. This is done using the site (defining the genome position and 
+ancestral state) and mutation (defining the site, node, and derived state)
+tables. In addition, tskit provides a flexible system for attaching metadata
+to all aspects of the data model, which can either be in JSON format 
+for flexibility (used heavily by sc2ts for debugging information) 
+or in a vectorisable packed binary format for efficiency (used to attach
+sample identifiers and pango assignments in the final ARGs). 
+Tskit also provides a ``provenance'' table, which provides a mechanism
+for recording software version and parameter information through each
+step of complex processing pipelines.
 

### 2025-09-19 src/save_copying_patterns.py
Label copying patterns with node IDs

@@ -71,7 +71,15 @@ def save_copying_pattern_image(node_id, label, ts, save_dir, zoom=None,**kwargs)
 
 for u, label in tqdm(pangoX_RE_node_labels().items()):
     save_copying_pattern_image(
-       u, label, ts, png_dir, hide_extra_rows=True, hide_labels=True, show_bases=None, font_family='Verdana')
+       u,
+       label,
+       ts,
+       png_dir,
+        hide_extra_rows=True,
+        hide_labels=True,
+        show_bases=None,
+       font_family='Verdana',
+    )
 
 for u, lab in tqdm([
     (427863, "RE_node-QCpass-427863"),  # Q1
@@ -80,5 +88,14 @@ for u, lab in tqdm([
     (663484, "RE_node-QCfail-663484"),  # Q4
 ]):
     save_copying_pattern_image(
-        u, lab, ts, png_dir, hide_extra_rows=False, hide_labels=True, show_bases=True, zoom=4, font_family='Verdana'
+        u,
+        lab,
+        ts,
+        png_dir,
+        hide_extra_rows=False,
+        hide_labels=False,
+        show_bases=True,
+        child_label=f"{u}",
+        zoom=4,
+        font_family='Verdana',
     )

### 2025-09-19 latex/main.tex
Update to clarify who uses MAT

@@
 Similarly to the Mutation Annotated Tree (MAT)~\cite{McBroome2021} format used 
-by UShER, tskit incorporates mutational information directly in the data 
+by UShER (RIPPLES~\cite{Turakhia2022} and matOptimize~\cite{DeMaio2023} are 
+included in the UShER software distribution), 
+tskit incorporates mutational information directly in the data 
 model. This is done using the site (defining the genome position and 

### 2025-09-19 latex/main.tex
Updates from Overleaf

--- 2025-09-19 latex/main.tex
@@
 
-Ancestral recombination graphs (ARGs) describe the reticulate 
-genealogies of sampled sequences in a recombining species\cite{Wong2024}.
-While ARGs have been of theoretical interest in population genetics 
-for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
-recent breakthroughs in simulation\cite{Kelleher2016,Kelleher2018efficient}
-and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
-Wohns2022,Zhang2023,Gunnarsson2024,Deng2024robust} have led to a surge 
-of interest in their application\cite{Brandt2024,Lewanski2024,Nielsen2025}.
-
-% TODO merge this in
 Ancestral Recombination Graphs (ARGs) are reticulate genealogies
@@
 \label{sec:args}
-
-describe how a set of sample sequences
-relate to each other at each position along the genome in a recombining
-species 
-lots of interest
-
-, and are currently the subject of intense research interest 
-(Brandt et al. 2024, Lewanski et al. 2024, Wong et al. 2024, Nielsen et al. 2025).
-
 %%% This paragraph needs to be rewritten.
@@
 under the influence of recombination.
-
-Until recently, ARGs lacked software support and
-sufficiently scalable inference methods to be of practical use.
-Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023},
-with high levels of recombination.
-These advances now enable evolutionary genealogies to be built quickly and at scale, and
-there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.
 
@@
 applied it to each putative recombination event in the ARG.
-We examined the genomic positions at which the suggested parents differed in allelic state.
+We examined the genomic positions
+at which the suggested parents differed in allelic state.
 Where such sites clustered closely (3 or fewer bases apart),
@@
 For each genomic region identified by the HMM as having a single parent,
-we calculated the "net number of supporting loci" by counting the number of loci in the region
+we calculated the "net number of supporting loci"
+by counting the number of loci in the region
 that supported the suggested parent minus the number that did not support it.
 Finally, with all the recombination events in the ARG having a single breakpoint,
-we calculated the net number of supporting loci on the left and right sides of each breakpoint, and
-used this to plot the 855 recombination events in Figure~\ref{fig:recombinant_qc}.
+we calculated the net number of supporting loci
+on the left and right sides of each breakpoint, and
+used this to plot all the 855 recombination events in Figure~\ref{fig:recombinant_qc}.
 
@@
 artefactual recombination events from those that are well supported.
-All the sc2ts-detected recombination events associated with the Pango X lineages (Table~\ref{tab:pango_x_lineages})
+All the sc2ts-detected recombination events
+associated with the Pango X lineages (Table~\ref{tab:pango_x_lineages})
 occur in the upper right of Figure~\ref{fig:recombinant_qc},
 with high numbers of net supporting loci on both the right and left parents.
-In contrast, the other regions of the plot, where one or both parents are less well supported,
-are dominated by recombination events associated with samples sequenced using a specific protocol (AmpliSeq).
+In contrast, the other regions of the plot,
+where one or both parents are less well supported,
+are dominated by recombination events associated with
+samples sequenced using the Ion AmpliSeq protocol.
 Coupled with closer inspection of copying patterns (see below),
-we chose a QC cutoff of four or more net supporting loci on both the left and right side of the breakpoint.
-This leads to 354 QC-passing recombination events (Figure~\ref{fig:recombinant_qc}, upper right quadrant),
+we chose a QC cutoff of 4 or more net supporting loci
+on both the left and right side of the breakpoint.
+This leads to 354 QC-passing recombination events
+(Figure~\ref{fig:recombinant_qc}, upper right quadrant),
 compared to 501 recombination events
@@
 (Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}),
-in which adjacent sites are underlined in red, and near-adjacent sites (within 3 bases) in orange.
+in which adjacent sites are underlined in red, and
+near-adjacent sites (within 3 bases) in orange.
 A continuous red+orange underline thus indicates a region treated as a single locus.
@@
 having a net number of supporting loci of 4 on both sides of the breakpoint.
-Q2 shows an artefactual recombination event caused by a 9-base Delta lineage-defining deletion.
-There are XX (XX\%) of the QC-failing recombination events involving this deletion, and
-many cases are affected by a known amplicon dropout caused by this particular deletion \cite{Sanderson2021}
+Q2 shows an artefactual recombination event
+caused by a 6-base Delta lineage-defining deletion.
+There are 71 (14\%) QC-failing recombination events involving this deletion, and
+many cases are affected by a known amplicon dropout
+caused by this particular deletion \cite{Sanderson2021}
 Q3 shows an artefactual recombination event caused by
 a combination of this Delta deletion and a known MNS (see below).
-Finally, Q4 shows an artifactual recombination event associated with the A507T-T508C-G509A MNS,
-which has been identified as a sequencing error due to incomplete read trimming \cite{DeMaio2024}.
-There are XX recombination events involving this particular MNS, of which XX\% failed QC.
+Finally, Q4 shows an artifactual recombination event
+associated with the A507T-T508C-G509A MNS,
+which has been identified as a sequencing error
+due to incomplete read trimming \cite{DeMaio2024}.
+There are 9 recombination events involving this particular MNS, of which 6 failed QC.
 

### 2025-09-19 latex/main.tex
Pass at trying to jazz up intro

@@
 \section*{INTRODUCTION}
-
-%%% No subheadings, please. Use of abbreviations should be kept to a minimum,
-%%% and nonstandard abbreviations should be defined when first used in the text.
-
-The genomic revolution has transformed how we study the evolution and spread of pathogens,
-allowing real-time tracking of pathogens over time and space.
-During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
-quickly overwhelming typical phylogenetic approaches.
-Although real-time phylogenetic methods now exist to cope with this flood of genomic data
-(e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
-we lack methods to build phylogenies that can integrate recombination
-in the face of ongoing sampling of infected cases.
-
-UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
-from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
-UShER has generated phylogenies updated daily \cite{McBroome2021} and
-has proven to be a powerful resource that successfully enabled the world
-to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
-These phylogenies are used to identify and monitor
-potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
-which might have acquired features to become Variants of Concern (VOC),
-for example, increased transmissibility, disease severity, and immune escape.
-Beyond supporting real-time genomic epidemiology,
-UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
-(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
-UShER uses a tree to represent evolutionary change via mutation,
-assuming no recombination \cite{Turakhia2021}.
-However, given extensive evidence for recombination in SARS-CoV-2
+% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
+% in terms of the basic goal, assumptions about samples and how recombination
+% can be treated.
+Classical phylogenetics is largely concerned with inferring the evolutionary history 
+of species~\cite{Felsenstein2004}, with 
+% Not sure about this assertion! Yan?
+samples separated by millions of years of evolution.
+[This is rough outline text]
+Complex models are required to fill in the gaps in our knowledge,
+incorporating prior information about sequence evoltion etc etc.
+In this context, recombination can be treated as a nuisance factor 
+because we cannot hope to observe it over these timescales [or something
+similar].
+Phylogenetics has continued along this vein, 
+although with various challenges on how to accomodate 
+more closely related samples and to handle potential recombination met.
+
+The data collected during the COVID-19 pandemic as part of global
+surveillance fundamentally changed these assumptions. 
+At the peak in [month] 202X, there were over X thousand samples 
+per day deposited in GISAID\cite{Shu2017}, with a total of 
+over 17 million as of Sep 2025. This superabundance of data 
+overwhelmed classical phylogenetic methods, both in terms 
+of the number of sequences that could be processed, but also
+more fundamentally in terms of the underlying assumption of 
+retrospective data sparsity. [TODO rephrase]
+The UShER method \cite{Turakhia2021} conclusively demonstrated 
+the value of simpler models that build a tree iteratively
+in real-time, quickly becoming an indispensible element 
+of the pandemic response\cite{Hinrichs2024}.
+
+Recombination continues to be treated as a nuisance factor, however~\cite{Neches2020}. 
+Overwhelming evidence demonstrates the existance of recombination in SARS-CoV-2
 \cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
-the tree-like representation of UShER is no longer tenable.
+with multiple deep recombination events now known to be present in 
+genealogy of the global population [cites??].
+Despite this, phylogenies used as the basis for innumerable studies 
+continue to depict the genealogy of SARS-CoV-2 as a single
+tree~\cite{McBroome2021,Hadfield2018}.
+
+% The genomic revolution has transformed how we study the evolution and spread of pathogens,
+% allowing real-time tracking of pathogens over time and space.
+% During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
+% quickly overwhelming typical phylogenetic approaches.
+% Although real-time phylogenetic methods now exist to cope with this flood of genomic data
+
+% (e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
+% we lack methods to build phylogenies that can integrate recombination
+% in the face of ongoing sampling of infected cases.
+
+% UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
+% from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
+% UShER has generated phylogenies updated daily \cite{McBroome2021} and
+% has proven to be a powerful resource that successfully enabled the world
+% to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
+% These phylogenies are used to identify and monitor
+% potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
+% which might have acquired features to become Variants of Concern (VOC),
+% for example, increased transmissibility, disease severity, and immune escape.
+% Beyond supporting real-time genomic epidemiology,
+% UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
+% (e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
+% UShER uses a tree to represent evolutionary change via mutation,
+% assuming no recombination \cite{Turakhia2021}.
 
@@
 of interest in their application\cite{Brandt2024,Lewanski2024,Nielsen2025}.
-
-% TODO merge this in
-Ancestral Recombination Graphs (ARGs) are reticulate genealogies
-that incorporate both mutation and recombination events \cite{Wong2024},
-potentially allowing more comprehensive evolutionary representations of SARS-CoV-2
-and other viruses undergoing recombination.
-Until recently, ARGs lacked software support and
-sufficiently scalable inference methods to be of practical use.
-Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023},
-with high levels of recombination.
-These advances now enable evolutionary genealogies to be built quickly and at scale, and
-there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.
 
@@
 Of the nodes, 855 represent recombination events,
-of which 356 have a minimum of four loci supporting the non-focal parent on both sides of the breakpoint
-(we refer to these as ``robust'' recombination events,
-Figure~\ref{fig:recombinant_qc}; see STAR Methods)
+of which 354 pass quality control (QC) filters
+(see STAR Methods, Figure~\ref{fig:recombinant_qc}),
 reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.
@@
 in the ARG (Table~\ref{tab:pango_x_lineages}),
-all of which pass the robustness filtering criteria (Figure~\ref{fig:recombinant_qc}).
+all of which pass QC (STAR Methods).
 %% JK: Does saying this here help at all? I think we may be better off without 
@@
 which descend from a single recombination event rather than five independent
-events in the ARG (Figure~\ref{fig:subgraphs}B).
+events in the ARG (Figure~\ref{fig:subgraphs}B), and which we tentatively group using the unused label Xx.
 These Pango X lineages were designated around the same time.
@@
 For example, the closely related XN and XAU can be explained by two mutations
-and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C).
+and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C), and we
+believe the original recombinant designation to be based on erroneous data
+(Supplementary document S1.XXX).
 %%% TODO: Say something about XAJ and XAN.
@@
 Illustrative subgraphs in the ARG.
-(A) Subgraph around the recombination node associated with the XA.
+Boxes identify specific samples, colored by their Pango X classification;
+smaller squares list extra samples descending from a node, hidden for brevity,
+with all hidden samples taking the same Pango designation as their parent.
+For the two recombination nodes, a genome bar above shows
+the location of the inferred breakpoint interval relative to the Spike gene.
+For simplicity, ancestral lineages above the recombination parents or the MRCA are omitted.
+(A) Subgraph around the recombination node associated with XA.
 Two "Group A" recombinant samples reported by Jackson et al. \cite{Jackson2021}) are shown.
-(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages; and
+(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages: we use the label Xx as a shorthand for this related group of recombinants; and
 (C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
-The nodes assigned to a Pango X lineage are colored by that lineage.
-The numbers inside the colored boxes indicate
-the number of samples assigned to the Pango X lineage indicated by the same color.
-The numbers below the vertical dashed lines indicate
-the total number of samples within a collapsed clade.
-For each recombination node, a genome schematic shows
-the location of its inferred breakpoint interval relative to the Spike gene.
-The ancestry above the parents of the recombination nodes or the MRCA is not shown for clarity.
 For more detailed visualizations of these subgraphs, see Document S2.
@@
 Pango lineage and the ancestor of all nodes assigned that label
-(except XBB, XBB.1 and XM; see STAR methods). [MORE DETAILS]
+(except XBB, XBB.x and XM; see STAR methods). [MORE DETAILS]
 }
@@
 
-Furthermore, we compared the sensitivity of sc2ts and rebar.
-We ran rebar on the ``robust'' recombination events detected by sc2ts (STAR Methods),
-and found that 167 (43\%) of them were also detected as recombinants by rebar.
+% https://github.com/jeromekelleher/sc2ts-paper/issues/717
+167 (43\%) of QC-passing recombination events were also detected as recombinants by rebar.
 We hypothesize that this disagreement arose from
@@
 and those studied by Jackson et al. \cite{Jackson2021}
-(Table \ref{tab:jackson}).
+(Table \ref{tab:jackson}, Document~\ref{sec:jackson_recombs}).
 
@@
 (C) Distribution of the Pango lineage distance between parents.
-(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
-each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
+(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
+leading to XA, Xx 
+(the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an
+undesignated recombination event under XBB, which we here refer to as  XBB.x.}
 \label{fig:averted_mutations}
@@
 particularly at the left boundary of the Spike gene (Figure~\ref{fig:recombinant_intervals}A,B),
-consistent with patterns observed in previous studies \cite{Turakhia2022,Li2024CovRecomb}.
+consistent with patterns observed in previous studies
+\cite{Bobay2020recombination,Turakhia2022,Li2024CovRecomb}.
 While this might result from a higher recombination rate in this region,
@@
 Recombination intervals.
-(A) Distribution of the breakpoint intervals of the 356 robust recombination events along the genome
-(annotated with gene labels on top).
+(A) Distribution of the breakpoint intervals of 354 recombination events 
+along the genome (annotated with gene labels on top).
 Each line represents an interval colored by its size.
@@
 the black curve shows the number of recombinants inferred within each week along the ARG,
-focusing on the 356 ``robust'' events.
+focusing on the 354 events passing QC filters.
 Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high.
@@
 analyses, such as calculating mutation spectra or querying the number
-or frequency of mutations at a given locus, which the ARG makes 
+or frequency of mutations at a given position, which the ARG makes 
 significantly more efficient, and ultimately more easily reproducible.
@@
 recombination~\cite{Wong2024} and providing the basis for a range of highly
-efficient data processing
+efficient 
 algorithms~\cite{Kelleher2016,Ralph2020,Kelleher2020coalescent,Grundler2025geographic,Lehmann2025on}.
@@
 visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
-inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022}
+inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
 % More here? I'm sure there's others
-and data analysis~\cite{Fan2022,Nowbandegani2023extremely}
+and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
 tools.
@@
 the retrospective group is accepted for inclusion.
-Although this process is highly hueristic, it successfully 
+Although this process is highly heuristic, it successfully 
 captured the majority of saltational lineages
@@
 \subsubsection*{Quality control of recombination events}
-
-[TODO This section needs a full pass. I've merged in a section that was in the
-supplement with an extremely terse section on counting supporting loci.
-Reference Figure~\ref{fig:recombinant_qc} for the scatter plot and 
-labeled quadrants, and Figure~\ref{fig:copying_pattern_examples}
-for examples of the corresponding copying patterns. Also reference
-any additional examples that we include beyond A-D.]
-
 Sequencing errors in genome assemblies \cite{Turakhia2020},
@@
 
-Inspection of the copying patterns for all the recombination events
-clearly identified associations between recombination events that failed QC and
+We developed a SnpIt-like~\cite{OToole2024} visualisation 
+that shows the informative genomic positions between 
+a recombinant and its parent sequences, and highlights 
+potentially problematic loci
+(See Fig~\ref{fig:copying_pattern_examples} caption for details).
+Inspection of these copying patterns for all the recombination events
+identified associations between recombination events that failed QC and
 known problematic loci in the SARS-CoV-2 genome.
 Figure~\ref{fig:copying_pattern_examples} shows illustrative copying patterns
-(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}),
-in which adjacent sites are underlined in red, and
-near-adjacent sites (within 3 bases) in orange.
-A continuous red+orange underline thus indicates a region treated as a single locus.
+(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}).
 Q1 shows a well supported recombination event that passed QC (albeit marginally),
@@
 many cases are affected by a known amplicon dropout
-caused by this particular deletion \cite{Sanderson2021}
+caused by this particular deletion\cite{Sanderson2021}.
 Q3 shows an artefactual recombination event caused by
@@
 
-In addition to the analysis of the 356 ``robust'' recombination events with a minimum of four nucleotides supporting each parent,
+In addition to the analysis of the 354 recombination events 
+that pass QC,
 we analyzed the relationship between the full set of 855 recombinants and case numbers.
@@
 Thus, similar conclusions are reached with either dataset,
-although filtering the recombination events to the 356 with better support revealed a stronger relationship with case numbers.
+although filtering the recombination events to the 354 with better support revealed a stronger relationship with case numbers.
 
@@
 
-\subsection*{XS}
+\subsubsection*{XS}
 https://github.com/jeromekelleher/sc2ts-paper/issues/287
-In the ARG, the XS samples appear to have originated from two sequential recombination events.
+In the ARG, the XS samples appear to have originated from
+two sequential recombination events.
 However, the second event is likely an artifact,
-as the XS sample associated with this event failed the minimum net number of supporting loci.
+as the XS sample associated with this event failed
+the minimum net number of supporting loci.
 In addition, this sample was sequenced using the AmpliSeq V1 kit,
@@
 
-\subsection*{XM}
+\subsubsection*{XM}
 https://github.com/jeromekelleher/sc2ts-paper/issues/284
@@
 correctly identifying the parental Pango lineages.
-The inferred breakpoint intervals mostly overlap with those identified in its designation issue.
-XBC was not added to the ARG due to high HMM costs and inadequate sample support
+The inferred breakpoint intervals mostly overlap with
+those identified in its designation issue.
+XBC was not added to the ARG due to high HMM costs and
+inadequate sample support
 (within a seven-day retrospective time window).
@@
 https://github.com/jeromekelleher/sc2ts-paper/issues/357
+
 Evidence of the second breakpoint reported in its designation issue relies on
-the presence of a large deletion inherited from BA.2*, which is currently treated as missing data by sc2ts.
-The XAC samples share a recombination node with the samples of four other 
-X lineages
+the presence of a large deletion inherited from BA.2*,
+which is currently treated as missing data by sc2ts.
+The XAC samples share a recombination node with
+the samples of four other Pango X lineages,
 which do not have this deletion (Figure 2C).
@@
 \subsubsection*{XN and XAU}
+
+https://github.com/jeromekelleher/sc2ts-paper/issues/348
 https://github.com/jeromekelleher/sc2ts-paper/issues/358
-https://github.com/jeromekelleher/sc2ts-paper/issues/348
-Both of these putative recombinants have been defined on the basis of a RHS from BA.2 and a LHS from BA.1.
+
+Both of these putative recombinants have been defined
+on the basis of an RHS from BA.2 and a LHS from BA.1.
 In the case of XN, the supposed BA.1 mutations are [C241T, A2832G, C3037T].
@@
 However, both BA.1 and BA.2 are now defined as also having C241T and C3037T.
-In addition, a very large number of BA.2 samples also have A2832G, and three BA.2-designated samples also have C2470T.
-This means that there is essentially no longer any evidence for a BA.1 parent for XN, and
+In addition, a very large number of BA.2 samples also have A2832G, and
+three BA.2-designated samples also have C2470T.
+This means that there is no longer any evidence for a BA.1 parent for XN, and
 only one uncertain mutation that suggests a BA.1 parent for XAU.
@@
 This was proposed as a recombination between BA.5 (LHS) and BA.2 (RHS).
-However, we find that this is only 2 mutations different from e.g. ERR9929846 which is a BA.4 sample.
+However, we find that this is only 2 mutations different from
+e.g. ERR9929846 which is a BA.4 sample.
 In fact, XAS only differs from the BA.4 root by a single mutation, C27945T,
-which was identified in the original designation issue as an important ORF truncation mutation also present in a few BA.2.65 samples.
-It is possible that C27945T was acquired by recombination, this is clearly below our threshold for considering something to be a recombinant.
+which was identified in the original designation issue
+as an important ORF truncation mutation also present in a few BA.2.65 samples.
+It is possible that C27945T was acquired by recombination,
+this is clearly below our threshold for considering something to be a recombinant.
 
@@
 
+\subsection{Sc2ts recombination detection results for additional Pango X lineages}
+
+\begin{table}
+\centering
+\caption{
+Sc2ts recombination detection results for Pango X lineages not added to the ARG.
+}
+\label{tab:absent_pango_x}
+\begin{tabular}{crrr}
+\toprule
+Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
+\midrule
+XBJ & 2 & 1 & 1\\
+XBP & 8 & 2 & 1\\
+XBS & 19 & 15 & 1\\
+XBW & 1 & 1 & 1\\
+XCA & 11 & 4 & 1\\
+XAK & 2 & 2 & 1\\
+XAY & 15 & 11 & 6\\
+XBC & 36 & 21 & 3\\
+XBL & 2 & 2 & 0\\
+XBT & 1 & 1 & 2\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \subsection{Analysis of the Jackson et al. recombinants}
+\label{sec:jackson_recombs}
 
@@
 
-% Data derived from the notebooks/jackson_recombinants.ipynb notebook
-% See https://github.com/jeromekelleher/sc2ts-paper/issues/293
-Jackson et al. described four groups of recombinants (named A to D) and
-four singleton recombinants.
-Of the 16 samples analyzed by Jackson et al, 12 were included in the ARG;
-the remaining 4 were excluded because the sequences did not pass the 
-QC filtering criteria, or due to a high HMM cost. Groups A-D are summarized 
-in Table~\ref{tab:jackson} and show excellent concordance in terms 
-of the parental Pango lineages and breakpoint intervals. All four groups
-are directly associated with strongly supported recombination events 
+Jackson et al. described
+four groups of recombinants (named A to D) and four singleton recombinants.
+Of the 16 samples analyzed by Jackson et al.\
+12 were included in the ARG;
+the remaining 4 were excluded
+because the sequences did not pass the QC filtering criteria or
+because of a high HMM cost.
+Groups A-D are summarized in Table~\ref{tab:jackson} and show excellent concordance
+in terms of the parental Pango lineages and breakpoint intervals.
+All four groups are directly associated with strongly supported recombination events
 in the ARG in terms of the number of averted mutations (A: 16, B: 6, C: 30, D: 18).
-The singleton CAMC-CBA018 has 
-the parents B.1.177+B.1.1.7 and the breakpoint interval 16,177--20,133 in the ARG, 
+In the ARG, the singleton CAMC-CBA018 has
+the parents B.1.177+B.1.1.7 and
+the breakpoint interval 16,177--20,133,
 and is strongly supported by 14 averted mutations.
-The Jackson et al.\ results are almost identical with the same parent lineages, 
-and breakpoint interval of interval 11,395–21,993.
-The singleton MILK-103C712 has 
-the parents B.1.177.16+B.1.1.7 and a breakpoint interval of 24915--27972 
-in the ARG, again strongly supported by 11 averted mutations.
-The Jackson et al.\ results for this lineage has the same parents,
+In the Jackson et al.\ results, this singleton is
+almost identical with the same parent lineages and
+a breakpoint interval of 11,395–-21,993.
+In the ARG, the singleton MILK-103C712 has
+the parents B.1.177.16+B.1.1.7 and
+a breakpoint interval of 24915--27972,
+again strongly supported by 11 averted mutations.
+In the Jackson et al.\ results,
+this lineage has the same parents,
 with the breakpoint interval 26,800–-27,974.
-The singleton QEUH-1067DEF has the parents B.1.1.7+B.1.177.9 
-and a breakpoint interval of 7,729--10,870 in the ARG, 
+In the ARG, the singleton QEUH-1067DEF has
+the parents B.1.1.7+B.1.177.9 and
+a breakpoint interval of 7,729--10,870,
 supported by 11 averted mutations.
-The Jackson et al.\ results for this lineage has the same parents,
+In the Jackson et al.\ results,
+this lineage has the same parents,
 with the breakpoint interval 6,953--10,872.
+
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_jackson\_recombinants.ipynb}
+{analysis\_jackson\_recombinants.ipynb}
+for details.
 
@@
 \caption{
-Illustrative copying patterns.
-Q1 to Q4 are examples drawn from the labeled quadrants in Figure~\ref{fig:recombinant_qc}.
-Q1 depicts the copying pattern of a recombination event that marginally passed QC,
-whereas the others depict the copying patterns of recombination events that failed QC.
+Illustrative copying patterns
+drawn from the quadrants labeled in Figure~\ref{fig:recombinant_qc}. 
+Each copying pattern shows the 
+the positions
+where the allelic state in a recombinant
+(middle row, labelled by node ID)
+matches that of the left (P0: upper row, coloured green) 
+or right parent (P1: lower row, coloured blue),
+or where the recombination event requires a de-novo mutation
+(gold, with mutational change below).
+Parental states that correspond to the reference but are not 
+inherited by the recombinant are shown with a gray background.
+Genome position (``pos'') and reference allele (``ref'') are
+shown for each column.
+Underneath the copying pattern, adjacent genomic positions are 
+underlined in red, and
+near-adjacent sites (within 3 bases) in orange.
 }

### 2025-09-19 latex/main.tex
Updates from Overleaf

--- 2025-09-19 latex/main.tex
@@
 Of the nodes, 855 represent recombination events,
-of which 356 have a minimum of four loci supporting the non-focal parent on both sides of the breakpoint
+of which 356 have at least four loci supporting the non-focal parent on both sides of the breakpoint
 (we refer to these as ``robust'' recombination events,
@@
 Illustrative copying patterns.
-Q1 to Q4 are examples drawn from the labeled quadrants in Figure~\ref{fig:recombinant_qc}.
+Q1 to Q4 are examples drawn from the labeled quadrants in
+Figure~\ref{fig:recombinant_qc}.
 Q1 depicts the copying pattern of a recombination event that marginally passed QC,
 whereas the others depict the copying patterns of recombination events that failed QC.
+The node ID of a recombination event is shown in the middle row.
 }

### 2025-09-20 latex/main.tex
Updates from Overleaf

@@
 Figure~\ref{fig:copying_pattern_examples} shows illustrative copying patterns
-(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}),
-in which adjacent sites are underlined in red, and
-near-adjacent sites (within 3 bases) in orange.
-A continuous red+orange underline thus indicates a region treated as a single locus.
+(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}).
 Q1 shows a well supported recombination event that passed QC (albeit marginally),
@@
 Illustrative copying patterns.
-Q1 to Q4 are examples drawn from the labeled quadrants in
-Figure~\ref{fig:recombinant_qc}.
-Q1 depicts the copying pattern of a recombination event that marginally passed QC,
-whereas the others depict the copying patterns of recombination events that failed QC.
-The node ID of a recombination event is shown in the middle row.
+These copying patterns are examples
+drawn from the quadrants labeled in Figure~\ref{fig:recombinant_qc}.
+Each pattern indicates the genomic positions
+where the allelic state in a recombinant
+(middle row, with the label showing the node ID of the recombination event)
+matches that of the suggested left or right parent (P0 and P1, respectively)
+or where the allelic state differs from that of both the parents.
+For reference, the allelic states of the Wuhan-Hu-1 genome
+(which is the assumed root of the ARG)
+are shown on top of the left parent.
+The allelic states are colored as follows:
+green for those matching the left parent;
+blue for those matching the right parent;
+yellow for those differing from both the parents; and
+grey for those matching the reference.
+Adjacent sites are underlined in red, and
+near-adjacent sites (within 3 bases) in orange.
+A continuous red+orange underline thus indicates
+a region treated as a single locus.
+For further context, mutations above a recombination event
+are shown on the bottom.
 }

### 2025-09-20 latex/main.tex

@@
 https://github.com/jeromekelleher/sc2ts-paper/issues/287
-In the ARG, the XS samples appear to have originated from two sequential recombination events.
+In the ARG, the XS samples appear to have originated from
+two sequential recombination events.
 However, the second event is likely an artifact,
-as the XS sample associated with this event failed the minimum net number of supporting loci.
+as the XS sample associated with this event failed
+the minimum net number of supporting loci.
 In addition, this sample was sequenced using the AmpliSeq V1 kit,
@@
 \subsubsection*{XN and XAU}
+
 https://github.com/jeromekelleher/sc2ts-paper/issues/358
 https://github.com/jeromekelleher/sc2ts-paper/issues/348
-Both of these putative recombinants have been defined on the basis of a RHS from BA.2 and a LHS from BA.1.
+
+Both of these putative recombinants have been defined
+on the basis of an RHS from BA.2 and a LHS from BA.1.
 In the case of XN, the supposed BA.1 mutations are [C241T, A2832G, C3037T].
@@
 \subsection{Analysis of the Jackson et al. recombinants}
+\label{sec:jackson_recombs}
 

### 2025-09-22 latex/main.tex
Add some tskit refs
Closes #761

@@
 Illustrative subgraphs in the ARG.
-(A) Subgraph around the recombination node associated with the XA.
+(A) Subgraph around the recombination node associated with XA.
 Two "Group A" recombinant samples reported by Jackson et al. \cite{Jackson2021}) are shown.
-(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages; and
+(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages: we use the label Xx as a shorthand for this related group of recombinants; and
 (C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
@@
 (C) Distribution of the Pango lineage distance between parents.
-(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
+(D) Copying patterns leading to XA, Xx (the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an undesignated recombination event under XBB, which we here refer to as  XBB.x; See Document S1.XXX for a description of these SnpIt-like\cite{xxx} plots, in which
 each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
@@
 
-\subsection*{XS}
+\subsubsection*{XS}
 https://github.com/jeromekelleher/sc2ts-paper/issues/287

### 2025-09-22 latex/main.tex
Updates from Overleaf

--- 2025-09-22 latex/main.tex
@@
 recombination~\cite{Wong2024} and providing the basis for a range of highly
-efficient 
+efficient data processing
 algorithms~\cite{Kelleher2016,Ralph2020,Kelleher2020coalescent,Grundler2025geographic,Lehmann2025on}.
@@
 visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
-inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
+inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022}
 % More here? I'm sure there's others
-and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
+and data analysis~\cite{Fan2022,Nowbandegani2023extremely}
 tools.

### 2025-09-22 latex/main.tex
Updates from yan and shing

--- 2025-09-22 latex/main.tex
@@
 (C) Distribution of the Pango lineage distance between parents.
-(D) Copying patterns leading to XA, Xx (the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an undesignated recombination event under XBB, which we here refer to as  XBB.x; See Document S1.XXX for a description of these SnpIt-like\cite{xxx} plots, in which
-each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
+(D) Copying patterns leading to XA, Xx (the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an undesignated recombination event under XBB, which we here refer to as  XBB.x; See Fig~\ref{fig:copying_pattern_examples} for a description of these SnpIt-like~\cite{OToole2024} plots, in which
+recombinants are shown  as a mosaic of variants inherited from the left vs right parents (green vs blue, de-novo mutations in gold).}
 \label{fig:averted_mutations}
@@
 
+https://github.com/jeromekelleher/sc2ts-paper/issues/348
 https://github.com/jeromekelleher/sc2ts-paper/issues/358
-https://github.com/jeromekelleher/sc2ts-paper/issues/348
 
@@
 \caption{
-Illustrative copying patterns.
-These copying patterns are examples
+Illustrative copying patterns
 drawn from the quadrants labeled in Figure~\ref{fig:recombinant_qc}.
-Each pattern indicates the genomic positions
+Q1-Q4 correspond to copying patterns for recombination nodes 427863, 728991, 411345, and 663484 (none of which are associated with designated Pango recombinant lineages).
+The SnpIt-like~\cite{OToole2024} schematic highlights the set of genomic positions
 where the allelic state in a recombinant
-(middle row, with the label showing the node ID of the recombination event)
-matches that of the suggested left or right parent (P0 and P1, respectively)
-or where the allelic state differs from that of both the parents.
-For reference, the allelic states of the Wuhan-Hu-1 genome
-(which is the assumed root of the ARG)
-are shown on top of the left parent.
-The allelic states are colored as follows:
-green for those matching the left parent;
-blue for those matching the right parent;
-yellow for those differing from both the parents; and
-grey for those matching the reference.
-Adjacent sites are underlined in red, and
+(middle row, labelled by node ID)
+matches that of the suggested left (P0, above, coloured green) or right parent (P1, below, coloured blue),
+or where the recombination event requires a de-novo mutation (gold).
+Genomic positions are shown above each column, with the allelic state of the Wuhan-Hu-1 reference genome (``ref'') immediately below;  parental states that are not inherited by the recombinant and correspond to the reference are shown with a grey background.
+At the Adjacent genomic positions are underlined in red, and
 near-adjacent sites (within 3 bases) in orange.

### 2025-09-22 latex/main.tex
354 vs 356
Closes #700

@@
 Of the nodes, 855 represent recombination events,
-of which 356 have at least four loci supporting the non-focal parent on both sides of the breakpoint
-(we refer to these as ``robust'' recombination events,
-Figure~\ref{fig:recombinant_qc}; see STAR Methods)
+of which 354 pass quality control (QC) filters
+(see STAR Methods, Figure~\ref{fig:recombinant_qc}),
 reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.

### 2025-09-22 latex/main.tex
Updates from Overleaf

--- 2025-09-22 latex/main.tex
@@
 (C) Distribution of the Pango lineage distance between parents.
-(D) Copying patterns leading to XA, Xx (the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an undesignated recombination event under XBB, which we here refer to as  XBB.x; See Fig~\ref{fig:copying_pattern_examples} for a description of these SnpIt-like~\cite{OToole2024} plots, in which
-recombinants are shown  as a mosaic of variants inherited from the left vs right parents (green vs blue, de-novo mutations in gold).}
+(D) Copying patterns leading to XA, Xx (the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an
+undesignated recombination event under XBB, which we here refer to as  XBB.x;
+See Fig~\ref{fig:copying_pattern_examples} for a description of these
+SnpIt-like~\cite{OToole2024} plots, in which recombinants are shown
+as a mosaic of variants inherited from the left vs right parents
+(green vs blue, de-novo mutations in gold).}
 \label{fig:averted_mutations}
@@
 drawn from the quadrants labeled in Figure~\ref{fig:recombinant_qc}.
-Q1-Q4 correspond to copying patterns for recombination nodes 427863, 728991, 411345, and 663484 (none of which are associated with designated Pango recombinant lineages).
-The SnpIt-like~\cite{OToole2024} schematic highlights the set of genomic positions
+Q1-Q4 correspond to copying patterns for recombination nodes 427863, 728991, 411345, and 663484, none of which are associated with designated Pango recombinant lineages.
+The SnpIt-like~\cite{OToole2024} schematic highlights the specific genomic positions
 where the allelic state in a recombinant
 (middle row, labelled by node ID)
-matches that of the suggested left (P0, above, coloured green) or right parent (P1, below, coloured blue),
-or where the recombination event requires a de-novo mutation (gold).
-Genomic positions are shown above each column, with the allelic state of the Wuhan-Hu-1 reference genome (``ref'') immediately below;  parental states that are not inherited by the recombinant and correspond to the reference are shown with a grey background.
-At the Adjacent genomic positions are underlined in red, and
+matches that of the suggested left (P0: upper row, coloured green) or right parent (P1: lower row, coloured blue),
+or where the recombination event requires a de-novo mutation (gold, with mutational change at bottom).
+Genome position (``pos'') is shown above each column, with the allelic state of the Wuhan-Hu-1 reference genome (``ref'') immediately below;  parental states that correspond to the reference but are not inherited by the recombinant are shown with a gray background.
+Underneath each copying pattern, adjacent genomic positions are underlined in red, and
 near-adjacent sites (within 3 bases) in orange.
-A continuous red+orange underline thus indicates
-a region treated as a single locus.
-For further context, mutations above a recombination event
-are shown on the bottom.
+A continuous red-and-orange underline thus indicates
+a region treated as a single ``locus'' for the purposes of quality control (see Supplementary S1.XXX).
 }

### 2025-09-22 latex/main.tex
Tidy up copying pattern description
Closes #698

@@
 (C) Distribution of the Pango lineage distance between parents.
-(D) Copying patterns leading to XA, Xx (the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an
-undesignated recombination event under XBB, which we here refer to as  XBB.x;
-See Fig~\ref{fig:copying_pattern_examples} for a description of these
-SnpIt-like~\cite{OToole2024} plots, in which recombinants are shown
-as a mosaic of variants inherited from the left vs right parents
-(green vs blue, de-novo mutations in gold).}
+(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
+leading to XA, Xx 
+(the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an
+undesignated recombination event under XBB, which we here refer to as  XBB.x.}
 \label{fig:averted_mutations}
@@
 
-Inspection of the copying patterns for all the recombination events
-clearly identified associations between recombination events that failed QC and
+We developed a SnpIt-like~\cite{OToole2024} visualisation 
+that shows the informative genomic positions between 
+a recombinant and its parent sequences, and highlights 
+potentially problematic loci
+(See Fig~\ref{fig:copying_pattern_examples} caption for details).
+Inspection of these copying patterns for all the recombination events
+identified associations between recombination events that failed QC and
 known problematic loci in the SARS-CoV-2 genome.
@@
 drawn from the quadrants labeled in Figure~\ref{fig:recombinant_qc}. 
-Q1-Q4 correspond to copying patterns for recombination nodes 427863, 728991, 411345, and 663484, none of which are associated with designated Pango recombinant lineages.
-The SnpIt-like~\cite{OToole2024} schematic highlights the specific genomic positions
+Each copying pattern shows the 
+the positions
 where the allelic state in a recombinant
 (middle row, labelled by node ID)
-matches that of the suggested left (P0: upper row, coloured green) or right parent (P1: lower row, coloured blue),
-or where the recombination event requires a de-novo mutation (gold, with mutational change at bottom).
-Genome position (``pos'') is shown above each column, with the allelic state of the Wuhan-Hu-1 reference genome (``ref'') immediately below;  parental states that correspond to the reference but are not inherited by the recombinant are shown with a gray background.
-Underneath each copying pattern, adjacent genomic positions are underlined in red, and
+matches that of the left (P0: upper row, coloured green) 
+or right parent (P1: lower row, coloured blue),
+or where the recombination event requires a de-novo mutation
+(gold, with mutational change below).
+Parental states that correspond to the reference but are not 
+inherited by the recombinant are shown with a gray background.
+Genome position (``pos'') and reference allele (``ref'') are
+shown for each column.
+Underneath the copying pattern, adjacent genomic positions are 
+underlined in red, and
 near-adjacent sites (within 3 bases) in orange.
-A continuous red-and-orange underline thus indicates
-a region treated as a single ``locus'' for the purposes of quality control (see Supplementary S1.XXX).
 }

### 2025-09-22 latex/main.tex
Add Muller cite
Closes #759

@@
 \section*{INTRODUCTION}
-
-%%% No subheadings, please. Use of abbreviations should be kept to a minimum,
-%%% and nonstandard abbreviations should be defined when first used in the text.
-
-The genomic revolution has transformed how we study the evolution and spread of pathogens,
-allowing real-time tracking of pathogens over time and space.
-During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
-quickly overwhelming typical phylogenetic approaches.
-Although real-time phylogenetic methods now exist to cope with this flood of genomic data
-(e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
-we lack methods to build phylogenies that can integrate recombination
-in the face of ongoing sampling of infected cases.
-
-UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
-from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
-UShER has generated phylogenies updated daily \cite{McBroome2021} and
-has proven to be a powerful resource that successfully enabled the world
-to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
-These phylogenies are used to identify and monitor
-potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
-which might have acquired features to become Variants of Concern (VOC),
-for example, increased transmissibility, disease severity, and immune escape.
-Beyond supporting real-time genomic epidemiology,
-UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
-(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
-UShER uses a tree to represent evolutionary change via mutation,
-assuming no recombination \cite{Turakhia2021}.
-However, given extensive evidence for recombination in SARS-CoV-2
-\cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
-the tree-like representation of UShER is no longer tenable.
+% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
+% in terms of the basic goal, assumptions about samples and how recombination
+% can be treated.
+Classical phylogenetics is largely concerned with inferring the evolutionary history 
+of species~\cite{Felsenstein2004}, with 
+% Not sure about this assertion! Yan?
+samples separated by millions of years of evolution.
+[This is rough outline text]
+Complex models are required to fill in the gaps in our knowledge,
+incorporating prior information about sequence evoltion etc etc.
+In this context, recombination can be treated as a nuisance factor 
+because we cannot hope to observe it over these timescales [or something
+similar].
+Phylogenetics has continued along this vein, 
+although with various challenges on how to accomodate 
+more closely related samples and to handle potential recombination met.
+
+The data collected during the COVID-19 pandemic as part of global
+surveillance fundamentally changed these assumptions. 
+At the peak in [month] 202X, there were over X thousand samples 
+per day deposited in GISAID\cite{Shu2017}, with a total of 
+over 17 million as of Sep 2025. This superabundance of data 
+overwhelmed classical phylogenetic methods, both in terms 
+of the number of sequences that could be processed, but also
+more fundamentally in terms of the underlying assumption of 
+retrospective data sparsity. [TODO rephrase]
+The UShER method \cite{Turakhia2021} conclusively demonstrated 
+the value of simpler models that build a tree iteratively
+in real-time, quickly becoming an indispensible element 
+of the pandemic response\cite{Hinrichs2024}.
+
+Recombination continues to be treated as a nuisance factor, however~\cite{Neches2020}. 
+Overwhelming evidence demonstrates the existance of recombination in SARS-CoV-2
+\cite{Jackson2021,VanInsberghe2021,Muller2022bayesian,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
+with multiple deep recombination events now known to be present in 
+genealogy of the global population [cites??].
+Despite this, phylogenies used as the basis for innumerable studies 
+continue to depict the genealogy of SARS-CoV-2 as a single
+tree~\cite{McBroome2021,Hadfield2018}.
+
+% The genomic revolution has transformed how we study the evolution and spread of pathogens,
+% allowing real-time tracking of pathogens over time and space.
+% During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
+% quickly overwhelming typical phylogenetic approaches.
+% Although real-time phylogenetic methods now exist to cope with this flood of genomic data
+
+% (e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
+% we lack methods to build phylogenies that can integrate recombination
+% in the face of ongoing sampling of infected cases.
+
+% UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
+% from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
+% UShER has generated phylogenies updated daily \cite{McBroome2021} and
+% has proven to be a powerful resource that successfully enabled the world
+% to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
+% These phylogenies are used to identify and monitor
+% potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
+% which might have acquired features to become Variants of Concern (VOC),
+% for example, increased transmissibility, disease severity, and immune escape.
+% Beyond supporting real-time genomic epidemiology,
+% UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
+% (e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
+% UShER uses a tree to represent evolutionary change via mutation,
+% assuming no recombination \cite{Turakhia2021}.
 
@@
 Illustrative subgraphs in the ARG.
+Boxes identify specific samples, colored by their Pango X classification;
+smaller squares list extra samples descending from a node, hidden for brevity,
+with all hidden samples taking the same Pango designation as their parent.
+For the two recombination nodes, a genome bar above shows
+the location of the inferred breakpoint interval relative to the Spike gene.
+For simplicity, ancestral lineages above the recombination parents or the MRCA are omitted.
 (A) Subgraph around the recombination node associated with XA.
@@
 (C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
-The nodes assigned to a Pango X lineage are colored by that lineage.
-The numbers inside the colored boxes indicate
-the number of samples assigned to the Pango X lineage indicated by the same color.
-The numbers below the vertical dashed lines indicate
-the total number of samples within a collapsed clade.
-For each recombination node, a genome schematic shows
-the location of its inferred breakpoint interval relative to the Spike gene.
-The ancestry above the parents of the recombination nodes or the MRCA is not shown for clarity.
 For more detailed visualizations of these subgraphs, see Document S2.
@@
 
+\subsection{Sc2ts recombination detection results for additional Pango X lineages}
+
+\begin{table}
+\centering
+\caption{
+Sc2ts recombination detection results for Pango X lineages not added to the ARG.
+}
+\label{tab:absent_pango_x}
+\begin{tabular}{crrr}
+\toprule
+Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
+\midrule
+XBJ & 2 & 1 & 1\\
+XBP & 8 & 2 & 1\\
+XBS & 19 & 15 & 1\\
+XBW & 1 & 1 & 1\\
+XCA & 11 & 4 & 1\\
+XAK & 2 & 2 & 1\\
+XAY & 15 & 11 & 6\\
+XBC & 36 & 21 & 3\\
+XBL & 2 & 2 & 0\\
+XBT & 1 & 1 & 2\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \subsection{Analysis of the Jackson et al. recombinants}

### 2025-09-22 latex/main.tex
Add some cites for ARG-like phylo methods

@@
 
+Recombination breaks things\cite{Posada2002effect}
+
+ARG/phylonetwork -like
+methods~\cite{Didelot2010inference,Vaughan2017inferring,
+Muller2020bayesian,Muller2022bayesian}
+
+Distinction between recombination detection --- identifying that a 
+given sequences is a likely recombinant --- and incorporation
+of recombination.
+
+Recombination has been incorporated in a number of models
+
 The data collected during the COVID-19 pandemic as part of global
@@
 Overwhelming evidence demonstrates the existance of recombination in SARS-CoV-2
-\cite{Jackson2021,VanInsberghe2021,Muller2022bayesian,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
+\cite{Jackson2021,VanInsberghe2021,Muller2022bayesian,Focosi2022,
+Gutierrez2022,Ignatieva2022,Tamura2023},
 with multiple deep recombination events now known to be present in 

### 2025-09-22 latex/main.tex
Updates

@@
 Classical phylogenetics is largely concerned with inferring the evolutionary history 
-of species~\cite{Felsenstein2004}, with 
-% Not sure about this assertion! Yan?
-samples separated by millions of years of evolution.
+of species~\cite{Felsenstein2004}, 
+where recombination
+is a problem. 
+In this context, recombination is something to be minimised,
+with species trees inferred over 
+non-recombining regions~\cite{Rannala2025recombination}.
+
+Modern pathogen datasets are not sparsely sampled from different species,
+but many isolates from the same species~\cite{Cryptic2022data}
+% Cryptic used neighbor-joining tree
+% A pairwise genetic distance matrix was constructed for 15,211 isolates by
+% comparing pairs of regenotyped VCF files (see “Genomic data processing and
+% variant calling” section above for more details). A neighbourhood-joining tree
+% was constructed from the distance matrix using quicktree [53].
+
 [This is rough outline text]

### 2025-09-23 latex/main.tex
Updates from Overleaf

--- 2025-09-22 latex/main.tex
@@
 \section*{INTRODUCTION}
-% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
-% in terms of the basic goal, assumptions about samples and how recombination
-% can be treated.
-Classical phylogenetics is largely concerned with inferring the evolutionary history 
-of species~\cite{Felsenstein2004}, with 
-% Not sure about this assertion! Yan?
-samples separated by millions of years of evolution.
-[This is rough outline text]
-Complex models are required to fill in the gaps in our knowledge,
-incorporating prior information about sequence evoltion etc etc.
-In this context, recombination can be treated as a nuisance factor 
-because we cannot hope to observe it over these timescales [or something
-similar].
-Phylogenetics has continued along this vein, 
-although with various challenges on how to accomodate 
-more closely related samples and to handle potential recombination met.
-
-The data collected during the COVID-19 pandemic as part of global
-surveillance fundamentally changed these assumptions. 
-At the peak in [month] 202X, there were over X thousand samples 
-per day deposited in GISAID\cite{Shu2017}, with a total of 
-over 17 million as of Sep 2025. This superabundance of data 
-overwhelmed classical phylogenetic methods, both in terms 
-of the number of sequences that could be processed, but also
-more fundamentally in terms of the underlying assumption of 
-retrospective data sparsity. [TODO rephrase]
-The UShER method \cite{Turakhia2021} conclusively demonstrated 
-the value of simpler models that build a tree iteratively
-in real-time, quickly becoming an indispensible element 
-of the pandemic response\cite{Hinrichs2024}.
-
-Recombination continues to be treated as a nuisance factor, however~\cite{Neches2020}. 
-Overwhelming evidence demonstrates the existance of recombination in SARS-CoV-2
-\cite{Jackson2021,VanInsberghe2021,Muller2022bayesian,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
-with multiple deep recombination events now known to be present in 
-genealogy of the global population [cites??].
-Despite this, phylogenies used as the basis for innumerable studies 
-continue to depict the genealogy of SARS-CoV-2 as a single
-tree~\cite{McBroome2021,Hadfield2018}.
-
-% The genomic revolution has transformed how we study the evolution and spread of pathogens,
-% allowing real-time tracking of pathogens over time and space.
-% During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
-% quickly overwhelming typical phylogenetic approaches.
-% Although real-time phylogenetic methods now exist to cope with this flood of genomic data
-
-% (e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
-% we lack methods to build phylogenies that can integrate recombination
-% in the face of ongoing sampling of infected cases.
-
-% UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
-% from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
-% UShER has generated phylogenies updated daily \cite{McBroome2021} and
-% has proven to be a powerful resource that successfully enabled the world
-% to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
-% These phylogenies are used to identify and monitor
-% potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
-% which might have acquired features to become Variants of Concern (VOC),
-% for example, increased transmissibility, disease severity, and immune escape.
-% Beyond supporting real-time genomic epidemiology,
-% UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
-% (e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
-% UShER uses a tree to represent evolutionary change via mutation,
-% assuming no recombination \cite{Turakhia2021}.
+
+%%% No subheadings, please. Use of abbreviations should be kept to a minimum,
+%%% and nonstandard abbreviations should be defined when first used in the text.
+
+The genomic revolution has transformed how we study the evolution and spread of pathogens,
+allowing real-time tracking of pathogens over time and space.
+During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
+quickly overwhelming typical phylogenetic approaches.
+Although real-time phylogenetic methods now exist to cope with this flood of genomic data
+(e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
+we lack methods to build phylogenies that can integrate recombination
+in the face of ongoing sampling of infected cases.
+
+UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
+from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
+UShER has generated phylogenies updated daily \cite{McBroome2021} and
+has proven to be a powerful resource that successfully enabled the world
+to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
+These phylogenies are used to identify and monitor
+potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
+which might have acquired features to become Variants of Concern (VOC),
+for example, increased transmissibility, disease severity, and immune escape.
+Beyond supporting real-time genomic epidemiology,
+UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
+(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
+UShER uses a tree to represent evolutionary change via mutation,
+assuming no recombination \cite{Turakhia2021}.
+However, given extensive evidence for recombination in SARS-CoV-2
+\cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
+the tree-like representation of UShER is no longer tenable.
 
@@
 of interest in their application\cite{Brandt2024,Lewanski2024,Nielsen2025}.
+
+% TODO merge this in
+Ancestral Recombination Graphs (ARGs) are reticulate genealogies
+that incorporate both mutation and recombination events \cite{Wong2024},
+potentially allowing more comprehensive evolutionary representations of SARS-CoV-2
+and other viruses undergoing recombination.
+Until recently, ARGs lacked software support and
+sufficiently scalable inference methods to be of practical use.
+Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023},
+with high levels of recombination.
+These advances now enable evolutionary genealogies to be built quickly and at scale, and
+there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.
 
@@
 
-\subsection{Sc2ts recombination detection results for additional Pango X lineages}
-
-\begin{table}
-\centering
-\caption{
-Sc2ts recombination detection results for Pango X lineages not added to the ARG.
-}
-\label{tab:absent_pango_x}
-\begin{tabular}{crrr}
-\toprule
-Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
-\midrule
-XBJ & 2 & 1 & 1\\
-XBP & 8 & 2 & 1\\
-XBS & 19 & 15 & 1\\
-XBW & 1 & 1 & 1\\
-XCA & 11 & 4 & 1\\
-XAK & 2 & 2 & 1\\
-XAY & 15 & 11 & 6\\
-XBC & 36 & 21 & 3\\
-XBL & 2 & 2 & 0\\
-XBT & 1 & 1 & 2\\
-\bottomrule
-\end{tabular}
-\end{table}
-
 \subsection{Analysis of the Jackson et al. recombinants}

### 2025-09-23 latex/main.tex
Decent pass on first para

--- 2025-09-22 latex/main.tex
@@
 Classical phylogenetics is largely concerned with inferring the evolutionary history 
-of species~\cite{Felsenstein2004}, with 
-% Not sure about this assertion! Yan?
-samples separated by millions of years of evolution.
-[This is rough outline text]
-Complex models are required to fill in the gaps in our knowledge,
-incorporating prior information about sequence evoltion etc etc.
-In this context, recombination can be treated as a nuisance factor 
-because we cannot hope to observe it over these timescales [or something
-similar].
-Phylogenetics has continued along this vein, 
-although with various challenges on how to accomodate 
-more closely related samples and to handle potential recombination met.
-
-Recombination breaks things\cite{Posada2002effect}
-
-ARG/phylonetwork -like
-methods~\cite{Didelot2010inference,Vaughan2017inferring,
-Muller2020bayesian,Muller2022bayesian}
-
-Distinction between recombination detection --- identifying that a 
-given sequences is a likely recombinant --- and incorporation
-of recombination.
-
-Recombination has been incorporated in a number of models
+of species~\cite{Felsenstein2004}, with small numbers of widely diverged
+samples. In this context, recombination is a nuisance which must be 
+carefully accounted for~\cite{Rannala2025recombination},
+affecting the inference of species trees~\cite{Posada2002effect}
+and biasing downstream
+analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
+Modern pathogen datasets are not sparsely sampled from different species,
+but consist of tens to hundreds of thousands of samples from 
+a single species~\cite{
+Shu2017,% GISAID ~270k FLU + others
+Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
+Apetrei2021hiv,% HIV 1,026,217 sequences
+Cryptic2022data,% TB 12k
+Dyer2025enterobase% 1 million, various things like E-coli
+}.
+In contrast to the sophisticated model-based approaches 
+used to perform species-level tree 
+inference~\cite{Suchard2018bayesian,Bouckaert2019beast},
+comprehensive phylogenies of these datasets are typically
+built with simpler distance-matrix methods [citation?]
+or approximations~\cite{Price2009fasttree}, primarily
+due to computational scaling limitations, but also because of
+complexity of modelling such datasets.
+While several methods have been proposed to incorporate
+recombination explicitly into model-based phylogenetic 
+inference~\cite{Didelot2010inference,%
+Vaughan2017inferring,%
+Muller2020bayesian,%
+Muller2022bayesian}, 
+scalability 
+is a major issue and practical application has been limited.
 

### 2025-09-23 latex/main.tex
Add VirusRecomb cite
Closes #311

@@
 used to perform species-level tree 
-inference~\cite{Suchard2018bayesian,Bouckaert2019beast},
+inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
 comprehensive phylogenies of these datasets are typically
@@
 due to computational scaling limitations, but also because of
-complexity of modelling such datasets.
+the complexity of modelling such datasets.
 While several methods have been proposed to incorporate
@@
 is a major issue and practical application has been limited.
+Software support in explicitly \emph{using} recombination-aware
+inference is also extremely limited [TODO add some concrete 
+details].
 
 The data collected during the COVID-19 pandemic as part of global
-surveillance fundamentally changed these assumptions. 
+surveillance presents fundamental challenges existing appraoches
+in two ways. Firstly, the sheer scale of data collected 
+overwhelmed even approximate distance-matrix based methods.
 At the peak in [month] 202X, there were over X thousand samples 
 per day deposited in GISAID\cite{Shu2017}, with a total of 
-over 17 million as of Sep 2025. This superabundance of data 
-overwhelmed classical phylogenetic methods, both in terms 
-of the number of sequences that could be processed, but also
-more fundamentally in terms of the underlying assumption of 
-retrospective data sparsity. [TODO rephrase]
+over 17 million as of Sep 2025. % TODO check numbers
+Regularly reinferring trees de-novo from rapidly 
+growing datasets is untenable.
 The UShER method \cite{Turakhia2021} conclusively demonstrated 
@@
 of the pandemic response\cite{Hinrichs2024}.
+The second challenge posed is that recombination is 
+an essential element of SARS-CoV-2 
+evolution\cite{Jackson2021,VanInsberghe2021,Muller2022bayesian,Focosi2022,
+Gutierrez2022,Ignatieva2022,Tamura2023},
+with multiple deep recombination events now known to be present in 
+genealogy of the global population [cites??].
+Although several methods have been proposed to \emph{detect}
+recombinant samples~\cite{Zhou2023virusrecom}
 

### 2025-09-23 latex/main.tex
Overview of SC2 inference + recomb
Closes #626

@@
 Although several methods have been proposed to \emph{detect}
-recombinant samples~\cite{Zhou2023virusrecom}
-
-Recombination continues to be treated as a nuisance factor, however~\cite{Neches2020}. 
-Overwhelming evidence demonstrates the existance of recombination in SARS-CoV-2
-\cite{Jackson2021,VanInsberghe2021,Muller2022bayesian,Focosi2022,
-Gutierrez2022,Ignatieva2022,Tamura2023},
-with multiple deep recombination events now known to be present in 
-genealogy of the global population [cites??].
-Despite this, phylogenies used as the basis for innumerable studies 
-continue to depict the genealogy of SARS-CoV-2 as a single
-tree~\cite{McBroome2021,Hadfield2018}.
+recombinant samples~\cite{
+Turakhia2022,%RIPPLES
+Zhou2023virusrecom,% CovRecomb
+Alfonsi2024,%RecombinHunt 
+Li2024CovRecomb%CovRecomb 
+},
+this information is not incorporated into the phylogenies used 
+in innumerable studies~\cite{McBroome2021,Hadfield2018},
+and recombination continues to be treated as a 
+nuisance factor~\cite{Neches2020}.
 

### 2025-09-23 latex/main.tex
Improve refs in second para

@@
 growing datasets is untenable.
-The UShER method \cite{Turakhia2021} conclusively demonstrated 
-the value of simpler models that build a tree iteratively
-in real-time, quickly becoming an indispensible element 
+The UShER\cite{Turakhia2021} method is based on adding 
+samples to an existing tree using maximum parsimony, 
+with periodic rearrangements to counter the inherent 
+greediness of the approach\cite{DeMaio2023}.
+The benefits of this simple model and the value of 
+a comprehensive genealogy updated in real-time~\cite{McBroome2021}
+were conclusively demonstrated, as the resource
+quickly becoming an indispensible element 
 of the pandemic response\cite{Hinrichs2024}.
@@
 an essential element of SARS-CoV-2 
-evolution\cite{Jackson2021,VanInsberghe2021,Muller2022bayesian,Focosi2022,
+evolution\cite{Jackson2021,VanInsberghe2021,Focosi2022,
 Gutierrez2022,Ignatieva2022,Tamura2023},
@@
 nuisance factor~\cite{Neches2020}.
-
-% The genomic revolution has transformed how we study the evolution and spread of pathogens,
-% allowing real-time tracking of pathogens over time and space.
-% During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
-% quickly overwhelming typical phylogenetic approaches.
-% Although real-time phylogenetic methods now exist to cope with this flood of genomic data
 

### 2025-09-23 latex/main.tex
Decent pass at para 3

@@
 
-% (e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
-% we lack methods to build phylogenies that can integrate recombination
-% in the face of ongoing sampling of infected cases.
-
-% UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
-% from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
-% UShER has generated phylogenies updated daily \cite{McBroome2021} and
-% has proven to be a powerful resource that successfully enabled the world
-% to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
-% These phylogenies are used to identify and monitor
-% potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
-% which might have acquired features to become Variants of Concern (VOC),
-% for example, increased transmissibility, disease severity, and immune escape.
-% Beyond supporting real-time genomic epidemiology,
-% UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
-% (e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
-% UShER uses a tree to represent evolutionary change via mutation,
-% assuming no recombination \cite{Turakhia2021}.
-
 Ancestral recombination graphs (ARGs) describe the reticulate 
-genealogies of sampled sequences in a recombining species\cite{Wong2024}.
+genealogies of sampled sequences in a recombining species\cite{Wong2024},
+and provide a natural and efficient means of incorporating recombination 
+into the study of SARS-CoV-2.
 While ARGs have been of theoretical interest in population genetics 
@@
 Wohns2022,Zhang2023,Gunnarsson2024,Deng2024robust} have led to a surge 
-of interest in their application\cite{Brandt2024,Lewanski2024,Nielsen2025}.
-
-Previous ARG methods are retrospective \cite{Rasmussen2014,Kelleher2019,Speidel2019,Ignatieva2022,Zhang2023},
-reconstructing the past from a sample of genome sequences drawn at one moment in time,
-but scientists are increasingly tracking evolutionary change as it happens,
-accumulating data forward in time.
-For example, sequences from daily samples of SARS-CoV-2 were made globally available on a daily basis,
-leading to over 17 million sequences in the GISAID database \cite{Shu2017}.
-
+of interest in their application in population and statical 
+genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
+% This is necessary because a bunch of the phylo literature refers
+% to ARGs, but it's all in the "coalescent with recombination" sense
+% Rannala2025 dismisses ARG-based methods at a stroke because they're
+% all based on the assumption of a single panmictic population 
+A subtle but important point distinguishing recent work on ARGs
+from the classical literature is that the ARG 
+``data structure''\cite{Minichiello2006mapping,Wong2024} is no longer
+tightly coupled to the generative
+process\cite{Hudson1983,Griffiths1997ancestral}, substantially 
+broadening the range of potential applications~\cite{Bisschop2025likelihoods}.
+Another important point is that strong software support for ARGs now
+exists, and the tskit library~\cite{Ralph2020} is the basis of a 
+rapidly growing 
+software ecosystem of 
+simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
+visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
+inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
+% More here? I'm sure there's others
+and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
+tools.
+
+[TODO rewrite this paragraph]
 Here, we present

### 2025-09-23 latex/main.tex

--- 2025-09-23 latex/main.tex
@@
 \section*{INTRODUCTION}
-% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
-% in terms of the basic goal, assumptions about samples and how recombination
-% can be treated.
-Classical phylogenetics is largely concerned with inferring the evolutionary history 
-of species~\cite{Felsenstein2004}, with small numbers of widely diverged
-samples. In this context, recombination is a nuisance which must be 
-carefully accounted for~\cite{Rannala2025recombination},
-affecting the inference of species trees~\cite{Posada2002effect}
-and biasing downstream
-analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
-Modern pathogen datasets are not sparsely sampled from different species,
-but consist of tens to hundreds of thousands of samples from 
-a single species~\cite{
-Shu2017,% GISAID ~270k FLU + others
-Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
-Apetrei2021hiv,% HIV 1,026,217 sequences
-Cryptic2022data,% TB 12k
-Dyer2025enterobase% 1 million, various things like E-coli
-}.
-In contrast to the sophisticated model-based approaches 
-used to perform species-level tree 
-inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
-comprehensive phylogenies of these datasets are typically
-built with simpler distance-matrix methods [citation?]
-or approximations~\cite{Price2009fasttree}, primarily
-due to computational scaling limitations, but also because of
-the complexity of modelling such datasets.
-While several methods have been proposed to incorporate
-recombination explicitly into model-based phylogenetic 
-inference~\cite{Didelot2010inference,%
-Vaughan2017inferring,%
-Muller2020bayesian,%
-Muller2022bayesian}, 
-scalability 
-is a major issue and practical application has been limited.
-Software support in explicitly \emph{using} recombination-aware
-inference is also extremely limited [TODO add some concrete 
-details].
-
-The data collected during the COVID-19 pandemic as part of global
-surveillance presents fundamental challenges existing appraoches
-in two ways. Firstly, the sheer scale of data collected 
-overwhelmed even approximate distance-matrix based methods.
-At the peak in [month] 202X, there were over X thousand samples 
-per day deposited in GISAID\cite{Shu2017}, with a total of 
-over 17 million as of Sep 2025. % TODO check numbers
-Regularly reinferring trees de-novo from rapidly 
-growing datasets is untenable.
-The UShER\cite{Turakhia2021} method is based on adding 
-samples to an existing tree using maximum parsimony, 
-with periodic rearrangements to counter the inherent 
-greediness of the approach\cite{DeMaio2023}.
-The benefits of this simple model and the value of 
-a comprehensive genealogy updated in real-time~\cite{McBroome2021}
-were conclusively demonstrated, as the resource
-quickly becoming an indispensible element 
-of the pandemic response\cite{Hinrichs2024}.
-The second challenge posed is that recombination is 
-an essential element of SARS-CoV-2 
-evolution\cite{Jackson2021,VanInsberghe2021,Focosi2022,
-Gutierrez2022,Ignatieva2022,Tamura2023},
-with multiple deep recombination events now known to be present in 
-genealogy of the global population [cites??].
-Although several methods have been proposed to \emph{detect}
-recombinant samples~\cite{
-Turakhia2022,%RIPPLES
-Zhou2023virusrecom,% CovRecomb
-Alfonsi2024,%RecombinHunt 
-Li2024CovRecomb%CovRecomb 
-},
-this information is not incorporated into the phylogenies used 
-in innumerable studies~\cite{McBroome2021,Hadfield2018},
-and recombination continues to be treated as a 
-nuisance factor~\cite{Neches2020}.
-
-% (e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
-% we lack methods to build phylogenies that can integrate recombination
-% in the face of ongoing sampling of infected cases.
-
-% UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
-% from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
-% UShER has generated phylogenies updated daily \cite{McBroome2021} and
-% has proven to be a powerful resource that successfully enabled the world
-% to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
-% These phylogenies are used to identify and monitor
-% potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
-% which might have acquired features to become Variants of Concern (VOC),
-% for example, increased transmissibility, disease severity, and immune escape.
-% Beyond supporting real-time genomic epidemiology,
-% UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
-% (e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
-% UShER uses a tree to represent evolutionary change via mutation,
-% assuming no recombination \cite{Turakhia2021}.
+
+%%% No subheadings, please. Use of abbreviations should be kept to a minimum,
+%%% and nonstandard abbreviations should be defined when first used in the text.
+
+The genomic revolution has transformed how we study the evolution and spread of pathogens,
+allowing real-time tracking of pathogens over time and space.
+During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
+quickly overwhelming typical phylogenetic approaches.
+Although real-time phylogenetic methods now exist to cope with this flood of genomic data
+(e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
+we lack methods to build phylogenies that can integrate recombination
+in the face of ongoing sampling of infected cases.
+
+UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
+from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
+UShER has generated phylogenies updated daily \cite{McBroome2021} and
+has proven to be a powerful resource that successfully enabled the world
+to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
+These phylogenies are used to identify and monitor
+potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
+which might have acquired features to become Variants of Concern (VOC),
+for example, increased transmissibility, disease severity, and immune escape.
+Beyond supporting real-time genomic epidemiology,
+UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
+(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
+UShER uses a tree to represent evolutionary change via mutation,
+assuming no recombination \cite{Turakhia2021}.
+However, given extensive evidence for recombination in SARS-CoV-2
+\cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
+the tree-like representation of UShER is no longer tenable.
 

### 2025-09-23 latex/main.tex
Fix Singer citation

--- 2025-09-23 latex/main.tex
@@
 and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
-Wohns2022,Zhang2023,Gunnarsson2024,Deng2024robust} have led to a surge 
+Wohns2022,Zhang2023,Gunnarsson2024,Deng2025robust} have led to a surge 
 of interest in their application in population and statical 
@@
 shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
-An example of a set of ``nested'' closely related lineages consists of XZ, XAC, XAD, XAE, and XAP,
+An example of a set of ``nested'' closely related lineages
+consists of XZ, XAC, XAD, XAE, and XAP,
 which descend from a single recombination event rather than five independent
-events in the ARG (Figure~\ref{fig:subgraphs}B), and which we tentatively group using the unused label Xx.
+events in the ARG (Figure~\ref{fig:subgraphs}B), and
+which we tentatively group using the unused label Xx.
 These Pango X lineages were designated around the same time.
-These results highlight that the ARG can help clarify the relationships between recombinants,
+These results highlight that the ARG can help
+clarify the relationships between recombinants,
 allowing more precise interpretation of evidence for recombination.
@@
 fell below our threshold of detection (i.e. $k = 4$ in the HMM).
-The fact that most of these Pango X lineages (except XAJ and XAN) differ from a single parent node in the ARG
+The fact that most of these Pango X lineages (except XAJ and XAN)
+differ from a single parent node in the ARG
 by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
@@
 which increases the chance of co-infection within an individual, allowing recombination.
-Second, we would expect recombination events to be detected more often when lineage diversity is high,
-as this increases the chance that two sufficiently distinct viruses co-infected an individual,
+Second, we would expect recombination events to be detected
+more often when lineage diversity is high,
+as this increases the chance that
+two sufficiently distinct viruses co-infected an individual,
 increasing our power to detect recombination.
@@
 Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
-sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in 
-Figure~\ref{fig:recombinants_over_time}).
-These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset.
+sub-divided by the proportion in each Scorpio label inferred within the ARG
+(shading in Figure~\ref{fig:recombinants_over_time}).
+These lineages correspond loosely to a group
+of related lineages within a major VOC in the Viridian dataset.
 Alongside the case numbers,
-the black curve shows the number of recombinants inferred within each week along the ARG,
+the black curve shows the number of recombinants
+inferred within each week along the ARG,
 focusing on the 354 events passing QC filters.
-Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high.
+Recombination events were inferred more often
+during the Delta and Omicron waves, when case counts were high.
 Overall, the number of recombinant events rose with the global number of cases
@@
 Recombination events also rose with the diversity of lineages in each week
-(Figure \ref{fig:recombinants_over_time}C; linear regression: 
+(Figure~\ref{fig:recombinants_over_time}C; linear regression: 
 $0.9 + 4.8 x$, $p = 3.4\times10^{-7}$),
@@
 
-We then asked which single measure best explains the number of recombination events inferred each week,
+We then asked which single measure best explains
+the number of recombination events inferred each week,
 considering $n$, $H$, $n H$, $n^2$, $n^2 H$.
-We included the squared number of cases each week thinking that it might better predict co-infections.
-The best single predictor of recombination events involved the product of cases and 
-expected heterozygosity ($n H$),
+We included the squared number of cases each week
+thinking that it might better predict co-infections.
+The best single predictor of recombination events involved
+the product of cases and expected heterozygosity ($n H$),
 with an adjusted $R^2$ of 23.7\%.
-This measure of case diversity ($n H$) explained 49\% more of the variation in recombination events per week
+This measure of case diversity ($n H$)
+explained 49\% more of the variation in recombination events per week
 than the next best predictor ($H$, adjusted $R^2$ of 15.9\%).
-While consistent with the hypotheses that recombination events occur more often when infection rates are common
-and are detected more often when viral diversity is high, we note several sources of uncertainty.
-Global case numbers were underreported, particularly during the Delta and Omicron waves \cite{Wang2025}.
+While consistent with the hypotheses
+that recombination events occur more often when infection rates are common
+and are detected more often when viral diversity is high,
+we note several sources of uncertainty.
+Global case numbers were underreported,
+particularly during the Delta and Omicron waves \cite{Wang2025}.
 Furthermore, there may be substantial uncertainty in the timing of a recombination event,
-particularly if these occur in chronically infected individuals (as reported by \cite{Burel2022})
+particularly if these occur in chronically infected individuals
+(as reported by \cite{Burel2022})
 with substantial delays before onward cases are detected.
-Finally, the filters that we applied removed all but the most strongly supported recombination events,
-although similar conclusions were obtained when all 855 recombinants were considered (STAR Methods).
+Finally, the filters that we applied removed all
+but the most strongly supported recombination events,
+although similar conclusions were obtained
+when all 855 recombinants were considered (STAR Methods).
 
@@
 \centering
-\includegraphics[width=0.85\linewidth]{recombinants_over_time.pdf}
-\caption{Recombination events across the pandemic. (A) Timing of the
-recombination events (black curve, right y-axis) relative to the global case
-numbers (\cite{Dong2020}; left y-axis). Cases are color-coded by their Scorpio
-labels, with the key showing those groupings with $\geq 1,000$ samples in the
-Viridian dataset. Recombinants more commonly occur (B) in weeks with high case
-numbers and (C) in weeks with extensive lineage variation, as estimated by high
-expected heterozygosity; solid lines show fits by linear regression, allowing
-for non-zero intercepts to account for unknown cases and heterozygosity not
-captured by the Scorpio labels.}
+\includegraphics[width=0.85\linewidth]{recombinants_over_time}
+\caption{
+Recombination events across the pandemic.
+(A) Timing of the recombination events (black curve, right y-axis)
+relative to the global case numbers (\cite{Dong2020}; left y-axis).
+Cases are color-coded by their Scorpio labels,
+with the key showing those groupings
+with $\geq 1,000$ samples in the Viridian dataset.
+Recombinants more commonly occur
+(B) in weeks with high case numbers and
+(C) in weeks with extensive lineage variation,
+as estimated by high expected heterozygosity;
+solid lines show fits by linear regression,
+allowing for non-zero intercepts to account for
+unknown cases and heterozygosity not captured by the Scorpio labels.
+}
 \label{fig:recombinants_over_time}
@@
 which overwhelmed existing computational approaches,
-motivating the rapid development of new methods\cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
+motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
 Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
@@
 and the VCF Zarr data storage format \cite{Czech2025}
-to describe relationships among virus samples, building an ARG over time as data becomes available.
+to describe relationships among virus samples,
+building an ARG over time as data becomes available.
 By reusing high-quality and high-performance components of these toolkits,
 we could develop the sc2ts inference method itself and
-perform all the pandemic-scale analyses in this article with Python code alone and
-no compiled languages.
+perform all the pandemic-scale analyses in this article
+with Python code alone and no compiled languages.
 
@@
 First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments.
-The total aligned dataset for 1,226,587 samples and 27,431 sites shared by the sc2ts and Usher tree sequences
-represents 31.34 GB of nucleotide calls.
+The total aligned dataset for 1,226,587 samples and 27,431 sites
+shared by the sc2ts and Usher tree sequences represents 31.34 GB of nucleotide calls.
 Of the 80,574,518 missing data calls (Ns) in the alignments,
@@
 Additionally, 413,303 calls made use of the IUPAC uncertainty codes.
-Of these sc2ts imputed only 77 (0.02\%) incorrectly (i.e., with a base that is not compatible with the ambiguity code).
+Of these sc2ts imputed only 77 (0.02\%) incorrectly
+(i.e., with a base that is not compatible with the ambiguity code).
 UShER imputed only 4 calls from this set incorrectly,
 reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations.
-This computation, performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
-required 30 lines of Python code and took 2m59s on an intel i7-9700 CPU, with a peak memory usage of about 4.5 GB
-(see \cite{Czech2025} for further benchmarks for processing the Viridian data in VCF Zarr format).
-Performing this analysis using matUtils would require exporting the full dataset to VCF,
-and converting the VCF to FASTA, and then comparing the alignments one-by-one to the source FASTA alignments.
+This computation,
+performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
+required 30 lines of Python code and took 2m 59s on an intel i7-9700 CPU,
+with a peak memory usage of about 4.5 GB
+(see \cite{Czech2025} for further benchmarks for
+processing the Viridian data in VCF Zarr format).
+Performing this analysis using matUtils would require
+exporting the full dataset to VCF, and
+converting the VCF to FASTA, and
+then comparing the alignments one-by-one to the source FASTA alignments.
 In contrast with matUtils and most phylogenetic software,
 which provide a command line interface and suite of predefined operations that typically output text files,
-tskit is a Python API (with C and Rust interfaces also supported) which has many advantages \cite{Baumdicker2022}.
+tskit is a Python API (with C and Rust interfaces also supported)
+which has many advantages \cite{Baumdicker2022}.
 As an illustration of the open-ended and efficient nature of the analyses this enables,
@@
 and compute the maximum number of children (5918).
-In the sc2ts ARG, it took 385 ms and 6 lines of code to find the maximum number of polytomies (137,001) and
+In the sc2ts ARG, it took 385 ms and 6 lines of code
+to find the maximum number of polytomies (137,001) and
 maximum number of children (7708) across all 348 local trees.
@@
 which no existing phylogenetic method detects and handles during inference
-(e.g., in the Viridian UShER phylogeny, parent and child nodes do not necessarily follow time order \cite{Hunt2024}),
-resulting in a cleaner dataset without samples with wrong collection dates that may affect downstream analyses.
-Future work may be estimating dates for the suspected time traveling samples using the ARG provided here
-and the sc2ts sample matching engine.
+(e.g., in the Viridian UShER phylogeny, parent and child nodes
+do not necessarily follow time order \cite{Hunt2024}),
+resulting in a cleaner dataset without samples
+with wrong collection dates that may affect downstream analyses.
+Future work may be estimating dates
+for the suspected time traveling samples using the ARG provided here and
+the sc2ts sample matching engine.
 
 Recombination is an important evolutionary force in viruses \cite{Perez-Losada2015},
-but has been understudied in part due to the lack of suitable data models, methods, and tools \cite{Neches2020}.
-While the methods and tools developed here are tailored for SARS-CoV-2, they can be extended and adapted for other viruses.
-A major advantage of the sc2ts methods and tools is that they are built on top of the mature software libraries of tskit and VCF-Zarr,
+but has been understudied in part
+due to the lack of suitable data models, methods, and tools \cite{Neches2020}.
+While the methods and tools developed here are tailored for SARS-CoV-2,
+they can be extended and adapted for other viruses.
+A major advantage of the sc2ts methods and tools is that
+they are built on top of the mature software libraries of tskit and VCF-Zarr,
 which enable pandemic-scale analyses that can be performed on typical laptops.
@@
 making them highly adaptable for a broad range of applications.
-Together with the pandemic-scale ARG, we hope that sc2ts will be highly accessible to many researchers,
+Together with the pandemic-scale ARG,
+we hope that sc2ts will be highly accessible to many researchers,
 enabling and accelerating studies of SARS-CoV-2 for years to come.
@@
 
-This work was funded by [FUNDER] via grant [GRANT NO.]. The authors thank all members of the lab for their support.
+This work was funded by [FUNDER] via grant [GRANT NO.].
+The authors thank all members of the lab for their support.
 
@@
 % https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/logs/sc2ts_v1_2023-02-21_rewire-recombinants-report.ipynb
-Using recombinant-rematching, 58 recombinants were identified as being more parsimoniously 
+Using recombinant-rematching,
+58 recombinants were identified as being more parsimoniously 
 explained by matching to existing nodes in the ARG (inserted by 
@@
 sites are included in the ARG.
-Using deletions with a 
-a minimum frequency of 1\% 
+Using deletions with a minimum frequency of 1\%
 (count of sequences divided by 9,149,680 sequences analyzed)
@@
 from primary inference and are defined as Pango lineage 
-defining sites by
- Freyja\cite{Karthikeyan2022}
+defining sites by Freyja\cite{Karthikeyan2022}
 (\url{https://raw.githubusercontent.com/andersen-lab/Freyja/refs/heads/main/freyja/data/lineage\_mutations.json}).
 Alignment data for these 163 sites was then mapped to the 
-ARG using the Fitch-Hartigan 
-algorithm\cite{Fitch1971toward,Hartigan1973minimum} 
+ARG using the Fitch-Hartigan algorithm\cite{Fitch1971toward,Hartigan1973minimum}
 implemented in tskit, including the gap character as a state 
-along with the 4 nucleotides.
+along with the four nucleotides.
 Following this, we iteratively applied the node parsimony 
@@
 potentially problematic loci
-(See Fig~\ref{fig:copying_pattern_examples} caption for details).
+(See Figure~\ref{fig:copying_pattern_examples} caption for details).
 Inspection of these copying patterns for all the recombination events
@@
 many cases are affected by a known amplicon dropout
-caused by this particular deletion\cite{Sanderson2021}.
+caused by this particular deletion \cite{Sanderson2021}.
 Q3 shows an artefactual recombination event caused by
@@
 due to incomplete read trimming \cite{DeMaio2024}.
-There are 9 recombination events involving this particular MNS, of which 6 failed QC.
+There are 9 recombination events involving this particular MNS,
+of which 6 failed QC.
 
@@
 
-We then computed intersection ARGs for sc2ts
-(sc2ts\_viridian\_inter\_v1.2.trees.tsz) and UShER 
-(usher\_viridian\_inter\_v1.2.trees.tsz) 
-by 
-simplifying\cite{Kelleher2018efficient,Wong2024} 
+We then computed intersection ARGs for
+sc2ts (sc2ts\_viridian\_inter\_v1.2.trees.tsz) and
+UShER (usher\_viridian\_inter\_v1.2.trees.tsz)
+by simplifying \cite{Kelleher2018efficient,Wong2024}
 both to the 2,475,418 samples and 27,507 sites that are shared
@@
 the UShER tree, and we wish to visually compare
-the phylogenetic backbones, we
-also dated the intersection tree by fixing
+the phylogenetic backbones,
+we also dated the intersection tree by fixing
 the sample dates using the ``Date\_tree'' metadata field,
@@
 pipeline described in the \nameref{sec:arg_postprocessing} section.
-The full UShER
-tree and the intersection files are available for download 
+The full UShER tree and the intersection files are available for download
 in tszip format (see \nameref{sec:data_and_code_availability}).
@@
 ANOVAs were conducted to determine which single predictor ($n$, $H$, $n \; H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week.
-These statistical tests and graphics were conducted in \textit{Mathematica} 14.2 (Wolfram 2024) and available at GITHUB LINK.
+These statistical tests and graphics were conducted
+in \textit{Mathematica} 14.2 (Wolfram 2024) and available at GITHUB LINK.
 
@@
 We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
-which have a minimum frequency of 1\% (count of sequences divided by 9,149,680 sequences analyzed).
+which have a minimum frequency of 1\%
+(count of sequences divided by 9,149,680 sequences analyzed).
 All of these indels are deletions and have high frequencies in the ARG
 (sequence count divided by the total number of samples),
-except a 9-base deletion at 11283, a 9-based deletion at 21987, and a 9-base deletion at 686.
+except a 9-base deletion at 11283, a 9-based deletion at 21987, and
+a 9-base deletion at 686.
 
@@
 
-\subsubsection*{XS}
+\subsubsection{XA}
+(also illustrated in Figure 2a / 3d)
+XA traces to a well supported recombination node, with no reversions.
+Note that the “causal” sample ERR5308556
+(which triggers the initial recombination event)
+is identical to the recombination node, and
+therefore lacks C8090T,
+which is one of the XA consensus mutations
+(identified as those shared by over 90\% of XA samples).
+As mentioned elsewhere,
+the sc2ts detection results in terms of
+the Pango lineages of the parents and the breakpoint intervals
+agree with the results reported by Jackson et al. (\ref{tab:jackson}).
+Two of the Jackson et al. XA samples
+are present in the ARG (ERR5308556 and ERR5414941);
+these samples are specifically marked in Figure~\ref{fig:subgraphs},
+which is a summary of this subgraph.
+
+\subsubsection{XB}
+https://github.com/jeromekelleher/sc2ts-paper/issues/332
+
+The displayed subgraph shows a small selection of the 192 XB samples.
+XB was originally defined as a recombinant between
+the closely related lineages B.1.634 (left parent) and B.1.631 (right parent)
+from North and Central America \cite{Gutierrez2022}.
+However, these parent lineages were not added to the ARG due to a high HMM cost.
+Together with other lineages such as B.1.627
+(sister to B.1.631 \cite{Gutierrez2022}),
+they are likely to attach somewhere along the long branch
+in the subgraph visualization (with >30 mutations),
+which also contains the key "R2" deletion
+associated with XB / B.1.634 \cite{Gutierrez2022}.
+
+We hypothesise that there are a set of samples
+from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
+which, should they be added to the ARG, would break up the long branch,
+with XB being a recombination between two of them.
+In the absence of these, sc2ts cannot detect XB as a recombinant.
+Also, note that the XB consensus mutations all occur on or above the long branch.
+We suspect that the child node of the long branch
+represents the (potentially recombinant) origin of XB, and
+the B.1 samples at the bottom left of the subgraph
+may be misclassified by pangolin.
+
+Further work is required to determine
+why these additional samples were not gradually added
+to the ARG by the daily sample matching process or
+retrospective sample grouping.
+
+\subsubsection{XC}
+The five XC samples descend from a well supported recombination node.
+The two parents that are both identical to actual samples
+(ERR4908740 on the left and DRR321274 on the right).
+There is one reversion (G21987A) on the left side of the breakpoint,
+but the site 21987 was masked during ARG inference
+because it was identified to be experiencing
+frequent mutation in a preliminary ARG.
+
+\subsubsection{XE/XH}
+https://github.com/jeromekelleher/sc2ts-paper/issues/337
+
+The XE samples descend from a well supported recombination node,
+with one mutation and one reversion (T22792C).
+This reversion was mapped onto the ARG as a part of postprocessing, and
+could be placed more parsimoniously by a single C22792T that groups together
+XH and the lineage leading to the further BA.2 recombination nodes
+displayed on the left.
+Nevertheless, the exact arrangement of the mutations at position 22792
+should not affect the proposed breakpoint,
+which is very distant, at position 11283.
+
+*** Should we describe the sequential recombination nodes to the left,
+here: note that the lower one is the RE node for XAF. ***
+
+Note that there might be some misalignment immediately
+to the right side of the breakpoint
+due to a repeated homopolymeric sequence involving 5-base deletions.
+Although this problem might obscure the breakpoint position,
+there nevertheless is ample evidence for the right parent.
+Furthermore, the only two XH samples in the ARG descend
+from the same recombination node,
+albeit with a number of mutations.
+
+\subsubsection{XF}
+XF is a well-supported recombinant between Delta and Omicron BA.1,
+with the breakpoint on the far left of the genome.
+All XF samples are the unique descendants of
+a well supported recombination node,
+with both parent nodes (ERR6193241: AY.4 and ERR8042040: BA.1) and
+the single child node (ERR8076129) being samples.
+
+\subsubsection{XG}
+The three XG samples descend from a reconstructed internal node,
+which in turn descends from a well supported recombination node.
+The left parent is a sample (SRR20179863), and
+the right parent is a reconstructed internal node
+assigned to Omicron BA.2,
+which is supported by a sample (ERR9124067) that is identical to it.
+
+\subsubsection{XJ}
+All the XJ samples descend from a well supported recombination node.
+There are 14 samples (including e.g. ERR8683190)
+which sc2ts identified as descendants of the same recombination node
+but are labeled BA.2 by pangolin (displayed on the left), and
+which do not possess the XJ consensus mutations C22792T and C27945T.
+These BA.2 samples could be worthy of further investigation.
+
+\subsubsection{XL}
+All the XL samples descend from a well supported recombination node,
+with three child samples which are identical to it.
+The left parent is a sample (ERR7989424), and
+the right parent is a reconstructed internal node
+assigned to BA.2.5 with a large number of descendant samples.
+
+\subsubsection{XM/XAL}
+https://github.com/jeromekelleher/sc2ts-paper/issues/284
+
+XM is a recombinant of BA.1 (left parent) and BA.2 (right parent).
+Sc2ts identifies multiple recombinant origins of XM.
+One of these origins leads to a dominant clade
+(node 1003220 at position 21,595),
+which also contains the only three XAL samples.
+
+Independent evidence for multiple XM origins
+comes from consideration of a 9-base deletion
+in the Spike gene at position 21633,
+which is characteristic of BA.2 but not BA.1
+(https://github.com/cov-lineages/pango-designation/issues/361).
+Most of the XM-labeled samples in ARG have this deletion,
+but three (SRR19024311, ERR9447529, and SRR18775874) do not,
+suggesting that XM samples trace to at least two recombination events,
+one to the right of the deletion
+(which leads to the majority of the XM samples), and one to the right.
+In the subgraph, it can be seen that one of the samples, SRR18775874,
+has been (presumably wrongly) placed
+within the main XM deletion-containing clade.
+As sc2ts does not use deletions for inference,
+deletion remapping (section XXXX) has reverted the deletion in this case,
+creating many insertions on the branch above the sample (magenta).
+A better use of deletions in sc2ts would likely result
+in placing this sample within one of the two XM clades
+which do not have the deletion.
+
+\subsubsection{XN/XAU}
+(also illustrated in Figure 2c)
+https://github.com/jeromekelleher/sc2ts-paper/issues/348
+https://github.com/jeromekelleher/sc2ts-paper/issues/358
+
+In the sc2ts ARG, the initial XN sample (ERR8986821)
+is well explained as a non-recombinant,
+copying from an inferred internal node
+assigned to BA.2 with only one mutation.
+This internal node is directly supported by the sample ERR9502469.
+All the other XN samples descend from ERR8986821.
+A small clade containing XAU is the sister to XN, and
+there are a series of samples (ERR9453882, ERR9502469, and ERR9823912)
+which support the gradual emergence of both groups
+from a BA.2 ancestral lineage via accumulation of single mutations.
+
+We believe that the explanation of XN and XAU
+in the sc2ts ARG is correct, and
+the evidence supporting the original designation of
+XN and XAU as recombinants
+comes from inaccurate or problematic data sources.
+In the case of XN, support for a recombinant BA.1+BA.2 origin
+is from three mutations that are claimed to be BA-1-specific:
+C241T, A2832G, and C3037T
+(https://github.com/cov-lineages/pango-designation/issues/480).
+However, the BA.1/BA.2 split
+(https://github.com/cov-lineages/pango-designation/issues/361)
+does not list C241T and C3037T as BA.1-specific,
+meaning that these two mutations are not supportive of recombination.
+Moreover, although the remaining mutation
+claimed to support a recombinant origin, A2832G,
+is typical of BA.1 and not BA.2,
+it is seen in several BA.2 samples (e.g. ERR9453882),
+supporting the idea that it has independently arisen within BA.2.
+Hence, none of the sites used to designate XN as a recombinant reliably do so.
+
+In the case of XAU, a recombinant origin was proposed
+on the basis of the same three sites, plus C2470T
+(https://github.com/cov-lineages/pango-designation/issues/894).
+However, C2470T is not BA.1-specific
+(https://github.com/cov-lineages/pango-designation/issues/361).
+Moreover, some BA.2-designated samples such as ERR9823912 also have C2470T,
+so we do not believe this to be a recombinant-supporting mutation either.
+
+This case illustrates the value of automated ARG reconstruction
+for correcting the inevitable errors that arise
+when designating recombinants by hand and
+using different (and inconsistent) data sources of
+Pango lineage-associated mutational constellations.
+
+\subsubsection{XP}
+https://github.com/jeromekelleher/sc2ts-paper/issues/345
+
+This is a case where the Pango X lineage is designated
+on the basis of two loci on the far right of the genome:
+a single nucleotide mutation (A29510C) and
+a rare, large deletion at position 29734 in the S2M step loop
+(https://github.com/cov-lineages/pango-designation/issues/481),
+which is not common enough to map in sc2ts during postprocessing
+and hence not shown here.
+Sc2ts requires at least four loci on either side
+to infer a QC-passing recombinant (section XXXX).
+Therefore, unless this deletion were given more weight than a single locus,
+sc2ts would not identify XP as a recombinant,
+even if deletions were used during ARG inference.
+If this deletion were recurrent, the evidence from sc2ts suggests
+that XP might not be a recombinant,
+because it is only one mutation different from the sample ERR8628084.
+
+\subsubsection{XQ/XR/XU/XAA/XAG/XAM}
+https://github.com/jeromekelleher/sc2ts-paper/issues/338
+
+In the ARG, XR, XU, XAA, XAG, and XAM
+are nested under the XQ recombinant lineage.
+However, upon closer inspection,
+the right hand branch above the recombination node appears problematic,
+as it displays three reversion mutations,
+which revert mutations at positions 22792, 24023, and 25624
+in the ancestral lineage just above.
+These sites were excluded during ARG inference, and
+remapped onto the ARG by parsimony during ARG postprocessing.
+We posit that if these three sites were used during ARG inference,
+the right parent of the recombination node
+would be chosen to be the ancestral BA.2 sample
+on the right-hand path with 18 child nodes.
+A more parsimonious reconstruction
+would then have the XQ clade as a sibling to
+a group of (monophyletic) XAA, XAM, XAG, and XR lineages.
+This would remove between 4 and 5 reversion mutations.
+
+\subsubsection{XS}
 https://github.com/jeromekelleher/sc2ts-paper/issues/287
+
 In the ARG, the XS samples appear to have originated from
-two sequential recombination events.
-However, the second event is likely an artifact,
-as the XS sample associated with this event failed
-the minimum net number of supporting loci.
-In addition, this sample was sequenced using the AmpliSeq V1 kit,
-which was also used to sequence eight other samples descending from the same event.
+two sequential recombination events (separated by a single branch).
+However, the later event is likely an artifact,
+because the XS sample (SRR18343220) associated with
+this event failed detection-level QC
+(i.e., the minimum net number of supporting loci;
+see the lower copying pattern).
+In addition, this sample was sequenced using the Ion AmpliSeq V1 kit,
+which was also used to sequence eight other samples
+descending from the same event
+(clustered due to a shared sequencing error due to AmpliSeq).
+These eight samples would likely attach to the nodes
+descending from the first recombination event.
 We excluded this likely artifactual event from further analysis and
@@
 
-\subsubsection*{XM}
-https://github.com/jeromekelleher/sc2ts-paper/issues/284
-Possibly related to deletions.
-
-\subsection*{Multiple breakpoints}
-Seven of the Pango X lineages present in the Viridian v04 dataset have more than one breakpoint,
-with the supporting evidence documented in the associated Pango designation issues.
-Five of these seven X lineages were detected as recombinants by sc2ts.
-Three of them (XAY, XBC, XBT) had more than one breakpoint,
-but were not added to the ARG due to high HMM cost and
-a lack of sample support to be incorporated as retrospective sample groups.
-However, the other two (XAC and XAK) had only one breakpoint.
-
-\subsubsection*{XAY}
-https://github.com/jeromekelleher/sc2ts-paper/issues/490
-
-\subsubsection*{XBC}
-https://github.com/jeromekelleher/sc2ts-paper/issues/486
-Sc2ts detected XBC as a three-breakpoint recombinant,
-correctly identifying the parental Pango lineages.
-The inferred breakpoint intervals mostly overlap with
-those identified in its designation issue.
-XBC was not added to the ARG due to high HMM costs and
-inadequate sample support
-(within a seven-day retrospective time window).
-
-\subsubsection*{XBT}
-https://github.com/jeromekelleher/sc2ts-paper/issues/487
-Sc2ts detected XBT as a two-breakpoint recombinant,
-correctly identifying the parental Pango lineages.
-The inferred breakpoint intervals are identical to those documented in its designation issue.
-XBT was not added to the ARG due to high HMM cost and low sample support
-(only one QC-passing sample in the Viridian v04 dataset).
-
-\subsubsection*{XAC}
-https://github.com/jeromekelleher/sc2ts-paper/issues/357
-
-Evidence of the second breakpoint reported in its designation issue relies on
-the presence of a large deletion inherited from BA.2*,
-which is currently treated as missing data by sc2ts.
-The XAC samples share a recombination node with
-the samples of four other Pango X lineages,
-which do not have this deletion (Figure 2C).
-
-\subsubsection*{XAK}
-https://github.com/jeromekelleher/sc2ts-paper/issues/484
-Our results agree with its designation issue in terms of the parental Pango lineages and breakpoint intervals.
-XAK was not added to the ARG due to HMM costs (9 and 10) and insufficient sample support (only two samples).
-Interestingly, our results suggest that XAK might be a product of two sequential recombination events.
-
-\subsubsection*{Non-recombinants}
-\subsubsection*{XN and XAU}
-
-https://github.com/jeromekelleher/sc2ts-paper/issues/348
-https://github.com/jeromekelleher/sc2ts-paper/issues/358
-
-Both of these putative recombinants have been defined
-on the basis of an RHS from BA.2 and a LHS from BA.1.
-In the case of XN, the supposed BA.1 mutations are [C241T, A2832G, C3037T].
-In the case of XAU, this set also includes C2470T, so [C241T, C2470T, A2832G, C3037T].
-However, both BA.1 and BA.2 are now defined as also having C241T and C3037T.
-In addition, a very large number of BA.2 samples also have A2832G, and
-three BA.2-designated samples also have C2470T.
-This means that there is no longer any evidence for a BA.1 parent for XN, and
-only one uncertain mutation that suggests a BA.1 parent for XAU.
-
-\subsubsection*{XAS}
-https://github.com/jeromekelleher/sc2ts-paper/issues/340
-This was proposed as a recombination between BA.5 (LHS) and BA.2 (RHS).
-However, we find that this is only 2 mutations different from
-e.g. ERR9929846 which is a BA.4 sample.
-In fact, XAS only differs from the BA.4 root by a single mutation, C27945T,
-which was identified in the original designation issue
-as an important ORF truncation mutation also present in a few BA.2.65 samples.
-It is possible that C27945T was acquired by recombination,
-this is clearly below our threshold for considering something to be a recombinant.
-
-\subsubsection*{XB}
-https://github.com/jeromekelleher/sc2ts-paper/issues/332
-XB was proposed to have originated from recombination between B.1.634 (left parent) and B.1.631 (right parent).
-None of the samples in Viridian v04 belonging to these Pango lineages were integrated into the ARG due to high HMM costs.
-
-\subsubsection*{XP}
-https://github.com/jeromekelleher/sc2ts-paper/issues/345
-
-\subsubsection*{XAJ}
-https://github.com/jeromekelleher/sc2ts-paper/issues/352
-
-\subsubsection*{XAZ}
-https://github.com/jeromekelleher/sc2ts-paper/issues/356
-Proposed as a recombinant of BA.2.5 on the left and BA.5 on the right, with 3 supporting sites (C2232T, C3317T, T3358C).
-In sc2ts this is very close to being a recombinant of BA.2.5 (ERR9615610) on the left,
-as not only are the 3 mutations present in this sample,
-but there is another site (C1912T apparently missed in the designation) which supports recombination.
-XAZ has C44T which is lacking in all BA.2.5 samples,
-giving only three net supporting loci for a putative recombinant.
-However, position 44 is often clipped, so this is not strong evidence against a recombinant, and
-this is a very finely balanced case.
-
-\subsubsection*{XAN and XAV}
-https://github.com/jeromekelleher/sc2ts-paper/issues/353
-Below threshold.
-
-\subsubsection*{XBE}
-This was proposed as a recombinant between BA.5,5 and BE.4.
-We find two net supporting loci for a recombination breakpoint to be around 22599.
+\subsubsection{XW}
+All the XW samples descend from a well supported recombination node.
+Its left parent is a BA.1.1.15-labelled sample (ERR8069658), and
+its right parent is a BA.2-labelled sample (ERR9086669).
+
+\subsubsection{XY}
+All the XY samples descend from a well supported recombination node,
+through a shared reconstructed internal node.
+Its left parent is a reconstructed internal node assigned to BA.1.1,
+which is supported by a sample (ERR8627048) identical to it.
+Its right parent is a BA.2-labeled sample (ERR9399983).
+There are a pair of (orange) reversion mutations
+within the XY clade at position 23948:
+these were not used for inference
+but remapped using parsimony during postprocessing, and
+indicate that the topology within the XY clade
+could potentially be improved
+by alternative phylogenetic reconstruction methods.
 

### 2025-09-23 latex/main.tex
Add first pass at authorship

@@
 
-\author[1,\orcidlink{0000-0000-0000-0000}]{First Last}
-\author[1,2]{First Middle Last}
-\author[2,3]{First M.M. Last}
-\author[3,*]{First Last, Jr.}
-\author[3,4,5,**]{F. Middle Last}
+% \author[1,\orcidlink{0000-0000-0000-0000}]{First Last}
+% \author[1,2]{First Middle Last}
+% \author[2,3]{First M.M. Last}
+% \author[3,*]{First Last, Jr.}
+% \author[3,4,5,**]{F. Middle Last}
+
+\author[1]{Shing~H.~Zhan}
+% TODO add joint second here?
+\author[2,3]{Anastasia~Ignatieva}
+\author[1]{Yan~Wong}
+\author[4]{Katherine~Eaton}
+\author[1]{Isobel~Guthrie}
+\author[1]{Benjamin~Jeffery}
+\author[1]{Duncan~S.~Palmer}
+\author[4]{Carmen~Lia~Murall}
+\author[5]{Sarah~P.~Otto}
+\author[1\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}
 
@@
 
-\affil[1]{University A, London SW7 2AZ, UK}
-\affil[2]{Department B, University B, Toronto, ON M5S 3H6, Canada}
-\affil[3]{Division C, Department C, University C, Cambridge, MA, USA}
-
-\affil[4]{Senior author}
-\affil[5]{Lead contact}
+\affil[1]{Big Data Institute, Li Ka Shing Centre for Health
+Information and Discovery, University of Oxford, United Kingdom}
+\affil[2]{Department of Statistics, University of Oxford, United Kingdom}
+\affil[3]{School of Mathematics and Statistics, University of Glasgow, United Kingdom}
+\affil[4]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
+\affil[5]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Canada}
+
+% \affil[1]{University A, London SW7 2AZ, UK}
+% \affil[2]{Department B, University B, Toronto, ON M5S 3H6, Canada}
+% \affil[3]{Division C, Department C, University C, Cambridge, MA, USA}
+
+% \affil[4]{Senior author}
+% \affil[5]{Lead contact}
 
@@
 
+%% TODO add correspodning - Shing and Jerome?
 \affil[*]{Correspondence: first.last.jr@university.edu}

### 2025-09-24 latex/main.tex
Updates on pango lineages from yan and shing

--- 2025-09-23 latex/main.tex
@@
 and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
-Wohns2022,Zhang2023,Gunnarsson2024,Deng2025robust} have led to a surge 
+Wohns2022,Zhang2023,Gunnarsson2024,Deng2024robust} have led to a surge 
 of interest in their application in population and statical 

### 2025-09-24 latex/main.tex
Update title and fill out remaining metadata

--- 2025-09-23 latex/main.tex
@@
 
-\title{Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time}
+\title{A comprehensive Ancestral Recombination Graph for SARS-CoV-2 inferred in real time}
 \date{}
@@
 
-\author[1]{Shing~H.~Zhan}
+\author[1,6]{Shing~H.~Zhan}
 % TODO add joint second here?
-\author[2,3]{Anastasia~Ignatieva}
-\author[1]{Yan~Wong}
-\author[4]{Katherine~Eaton}
+\author[1,5]{Yan~Wong}
+\author[2,5]{Anastasia~Ignatieva}
+\author[3]{Katherine~Eaton}
 \author[1]{Isobel~Guthrie}
@@
 \author[1]{Duncan~S.~Palmer}
-\author[4]{Carmen~Lia~Murall}
-\author[5]{Sarah~P.~Otto}
+\author[3]{Carmen~Lia~Murall}
+\author[4]{Sarah~P.~Otto}
 \author[1\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}
@@
 \affil[2]{Department of Statistics, University of Oxford, United Kingdom}
-\affil[3]{School of Mathematics and Statistics, University of Glasgow, United Kingdom}
-\affil[4]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
-\affil[5]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Canada}
+\affil[3]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
+\affil[4]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Canada}
 
@@
 
-% \affil[4]{Senior author}
-% \affil[5]{Lead contact}
+\affil[5]{Joint second author}
+\affil[6]{Lead contact}
 
@@
 
-%% TODO add correspodning - Shing and Jerome?
-\affil[*]{Correspondence: first.last.jr@university.edu}
-\affil[**]{Correspondence: f.middle.last@university.edu}
+\affil[*]{Correspondence: shing.zhan@bdi.ox.ac.uk}
+\affil[**]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
 

### 2025-09-24 notebooks/nb_utils.py
Update subgraphs

@@ -75,7 +75,7 @@ def standard_recombinant_labels(ts, pango_x_events_file):
     # Tweak the "XBB.1" RE node label, which should have a bespoke label because it's not
     # reflective of the majority of XBB.1 samples
     XBB_1 = [k for k, v in labels.items() if v == "XBB.1"][0]
-    labels[XBB_1] = "XBB.x1"
+    labels[XBB_1] = "XBB.x"
     Xx = [k for k, v in labels.items() if "XZ" in v][0]
     labels[Xx] = "Xx"
     
@@ -287,6 +287,7 @@ def mutation_p_values(ts, min_time=1, progress=True):
 
 
 class D3ARG_viz:
+    highlight_colour = "plum"
     colours = ['#332288', '#88CCEE', '#44AA99', '#117733', '#999933', '#DDCC77']  # from https://personal.sron.nl/~pault/
     def __init__(self, ts, df, lineage_consensus_muts=None, pangolin_field="pango", progress=True):
         self.ts = ts
@@ -421,6 +422,7 @@ class D3ARG_viz:
             highlight_nodes=highlight_nodes,
             parent_levels=parent_levels,
             child_levels=child_levels,
+            save_filename="-".join(pangos),
             **kwargs,
         )
 
@@ -437,10 +439,11 @@ class D3ARG_viz:
         label_mutations=False,
         highlight_mutations=None,
         highlight_nodes=True,  # Can also be a mapping of colour to node IDs
-        highlight_colour="plum",
+        highlight_colour=None,
         positions_file=None,
         oldest_y_label=None,
         node_1_date=datetime(2019, 12, 26),  # date of Wuhan, node #1
+        save_filename=None,
         **kwargs,
     ):
         """
@@ -451,6 +454,8 @@ class D3ARG_viz:
         or a PosAlt named tuple with positions and derived states (e.g. from
         `MutationContainer.get_mutations(Pango)` )
         """
+        if highlight_colour is None:
+            highlight_colour = self.highlight_colour
         if positions_file is not None:
             try:
                 self.d3arg.set_node_x_positions(
@@ -573,16 +578,27 @@ class D3ARG_viz:
                     oldest_y_label, zero_date - timedelta(days=times.min())
                 )
             }
-
-        return self.d3arg.draw_nodes(
-            nodes,
-            depth=(parent_levels, child_levels),
-            show_mutations=True,
-            y_axis_scale=y_axis_scale,
[…]

### 2025-09-24 src/makepdf.py

@@ -67,6 +67,8 @@ def convert_file(infile, outdir):
                 "webpdf",
                 "--no-prompt",
                 "--TagRemovePreprocessor.remove_cell_tags='{\"remove_cell\"}'",
+                "--PDFExporter.margin_left=0.2cm",
+                "--PDFExporter.margin_right=0.2cm",
                 "--no-input",
                 output.name,
                 "--output",

### 2025-09-25 latex/main.tex
Updates from Overleaf

@@
 
-[FIXME - discuss BA.5 and the BA.2 event instead.]
-Our results suggest that Delta and Omicron BA.2 might have recombinant origins,
+[FIXME - discuss BA.5 (kept as a recombination event but only because we can't 
+rewire it post-hoc, owning to subsequent tree building) and the BA.2 event (doesn't
+pass QC.]
+[Our results suggest that Delta and Omicron BA.2 might have recombinant origins [no longer true],
 but we interpret these results with caution,
-as these VOCs are undersampled around the time of their emergence (REF).
-Consistent with this, the presence of mutations and conflicting loci matching non-focal parents
-observed in the copying patterns of these VOCs (Figure 3D) suggests the existence of
-many genetically distinct, unsampled ancestors.
+as these VOCs are undersampled around the time of their emergence (REF).]
+Note that the presence of many de-novo mutations
+observed in the copying patterns of XBB (Figure 3D) suggests the existence of
+many genetically distinct, unsampled ancestors [FIXME: in this case, probably BJ.1, see
+https://github.com/jeromekelleher/sc2ts-paper/issues/336].
 Future work is needed to test whether these VOCs descended from recombination events,
@@
 
-We hypothesise that there are a set of samples
+We hypothesize that there are a set of samples
 from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
@@
 All the XW samples descend from a well supported recombination node.
-Its left parent is a BA.1.1.15-labelled sample (ERR8069658), and
-its right parent is a BA.2-labelled sample (ERR9086669).
+Its left parent is a BA.1.1.15-labeled sample (ERR8069658), and
+its right parent is a BA.2-labeled sample (ERR9086669).
 
@@
 by alternative phylogenetic reconstruction methods.
+
+\subsubsection{Xx: XZ/XAC/XAD/XAE/XAP}
+WIP
+
+\subsubsection{XAF}
+https://github.com/jeromekelleher/sc2ts-paper/discussions/638
+
+Of the 21 XAF samples in the Viridian v04 dataset,
+only XAF sample passed our QC filters and was added to the ARG:
+many have missing Ns in their consensus sequences (> 500) or
+require a large number of mutations to incorporate into the ARG
+(i.e. have a high HMM cost).
+However, we find a small recombinant BA1.1/BA.2 clade of 37 samples,
+associated with the BA.2-labeled sample ERR9755541,
+within which the single XAF sample is nested.
+The inferred parents and breakpoint at 10,447
+aligns with that proposed for XAF
+(https://github.com/cov-lineages/pango-designation/issues/676).
+Inspection of the omitted XAF samples with high HMM cost shows that
+they would attach as close sisters to the included XAF sample,
+albeit on branches with many new mutations.
+Hence we believe the current XAF topology in the ARG is reasonable,
+although somewhat tentative, given the number of included samples.
+
+The mutation-rich branch in the subgraph with
+a 5-base deletion (11283-11287) and insertion (11292-11296)
+is likely to be a simple deletion alignment error.
+
+\subsubsection{XAJ}
+https://github.com/jeromekelleher/sc2ts-paper/discussions/634
+
+In the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/826),
+XAJ is suggested to be a recombinant between
+BA.2.12.1 (left parent) and BA.4 (right parent).
+XAJ was detected as a non-recombinant by sc2ts.
+There are only three loci supporting
+the right BA.4 parent in the ARG (T22917G, T23018G, and G23040A),
+which means that XAJ fell below our threshold of detection.
+There is also a deletion (H69 in Spike) supporting
+the BA.4 parent (see the Pango designation issue),
+but it was treated as missing data during ARG inference.
+If this deletion were accounted for,
+XAJ should have been detected as a recombinant.
+
+\subsubsection{XAN/XAV}
+https://github.com/jeromekelleher/sc2ts-paper/discussions/783
+https://github.com/jeromekelleher/sc2ts-paper/discussions/785
+
+XAN and XAV are shown in this subgraph, as they share some mutations.
+In the Pango designation issues
+(https://github.com/cov-lineages/pango-designation/issues/771;
+https://github.com/cov-lineages/pango-designation/issues/911),
+XAN and XAV are suggested to be a recombinant between
+BA.2 (left parent) and BA.5 (right parent).
+Both these Pango lineages were detected as non-recombinants by sc2ts,
+because the number of supporting loci fell below our threshold of detection.
+In the ARG, we found only one mutation (C9866T)
+supporting the suggested BA.2 parent for both XAN and XAV.
+
+\subsubsection{XAS}
+https://github.com/jeromekelleher/sc2ts-paper/discussions/784
+
+XAS was proposed as a recombination between
+BA.5 (left parent) and BA.2 (right parent).
+However, we find that this is only two mutations
+different from e.g. ERR9929846, which is a BA.4 sample.
+In fact, XAS only differs from the BA.4 root by a single mutation,
+C27945T, which was identified in the original Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/882)
+as an important ORF truncation mutation
+which is also present in a few BA.2.65 samples.
+It is possible that C27945T was acquired by recombination,
+but this is clearly below our threshold of detection.
+
+\subsubsection{XAZ}
+https://github.com/jeromekelleher/sc2ts-paper/discussions/786
+
+In the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/797),
+XAZ is suggested as a two-breakpoint recombinant between
+BA.2.5 and BA.5 (the middle parent),
+with three sites supporting the BA.2.5 parent
+on the left end (C2232T, C3317T, and T3358C).
+In the ARG, this is very close to being a recombinant of
+BA.2.5 (ERR9615610) on the left side of the breakpoint,
+as not only are the three mutations present in this sample,
+but there is another site which supports recombination 
+C1912T, which is not reported in the Pango designation issue).
+XAZ has C44T which is lacking in all BA.2.5 samples,
+leaving only three net supporting loci for a putative recombinant.
+However, position 44 is often clipped,
+so this is not strong evidence against a recombinant, and
+therefore XAZ is a marginal case of recombination.
+We found no evidence for the third segment
+inherited from a BA2.5 parent on the right end.
+In the ARG, all the XAZ samples form a clade.
+Here we display only the first 20 XAZ samples.
+
+\subsubsection{XBB}
+WIP
+
+\subsubsection{XBB.1}
+WIP
+
+\subsubsection{XBD}
+All the XBD samples descend from a well supported recombination node.
+Its left and right parents are reconstructed internal nodes,
+both of which are directly supported by samples
+(SRR21382561 and ERR9825609, respectively)
+that are identical to them.
+
+\subsubsection{XBE}
+WIP
+
+\subsubsection{XBF}
+All the XBF samples descend from a well supported recombination node,
+with three mutations on the left of the breakpoint.
+Its left parent is a sample assigned to BA.5.2.1 (SRR20305553).
+Its right parent is a reconstructed internal node
+that is supported by a CJ.1-labeled sample (SRR21948068),
+which differs by three mutations.
+
+\subsubsection{XBG}
+All the XBG samples descend from a well supported recombination node,
+with two mutations, one on each side of the breakpoint.
+Both its left and right parents are samples
+(SRR19920084 and ERR9852351, respectively).
+
+\subsubsection{XBH}
+All the XBH samples descend from a well supported recombination node,
+despite two reversions (T222C and T12789C, respectively),
+which occur at excluded sites and
+were remapped by parsimony during postprocessing.
+If these two sites had been used during ARG inference,
+the branch attached to BA.2.1 would have been attached to
+the BA.2.5 shown on the top left instead,
+leading to a more parsimonious local topology
+with two mutations on the left branch.
+
+\subsubsection{XBQ/XBK/XBK.1/CJ.1.3}
+WIP
+
+\subsubsection{XBM}
+All the XBM samples descend from a well supported recombination node.
+Its left parent is a reconstructed internal node
+that is directly supported by a BA.2.76-labeled sample identical to it.
+Its right parent is a BF.3-labeled sample.
+
+\subsubsection{XBR}
+This one XBR sample descends from a well supported recombination node.
+Its left parent is a BN.3.1-labeled sample, and
+its right parent is a reconstructed internal node
+that is directly supported by a BQ.1.25-labeled sample identical to it.
 

### 2025-09-25 latex/main.tex
Update title and abstract accordingly

@@
 
-\title{A comprehensive Ancestral Recombination Graph for SARS-CoV-2 inferred in real time}
+\title{A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2}
 \date{}
@@
 
-% NOTE: currently 160 words
+% NOTE: currently 163 words
 
@@
 incomplete and potentially misleading.
-We present sc2ts, a method to infer reticulate genealogies for SARS-CoV-2
-that automatically detects and incorporates recombination 
-in real time.  
-We use sc2ts to infer a genealogy of 2.48 million genomes,
+We present sc2ts, a method to infer reticulate genealogies 
+as an Ancestral Recombination Graph (ARG) for SARS-CoV-2
+that automatically detects and incorporates recombination.  
+We use sc2ts to infer an ARG of 2.48 million genomes,
 and show that it captures a broad range of 
@@
 recombinant lineages.
-Together, the inference method, inferred genealogy and 
+Together, the inference method, infererred ARG and 
 suite of open-source analysis tools, 

### 2025-09-25 latex/main.tex
Good draft

@@
 The data collected during the COVID-19 pandemic as part of global
-surveillance presents fundamental challenges existing appraoches
+surveillance presents fundamental challenges to existing approaches
 in two ways. Firstly, the sheer scale of data collected 
@@
 over 17 million as of Sep 2025. % TODO check numbers
-Regularly reinferring trees de-novo from rapidly 
+Regularly reinferring trees de-novo from such rapidly 
 growing datasets is untenable.
@@
 and recombination continues to be treated as a 
-nuisance factor~\cite{Neches2020}.
+nuisance factor.
 
@@
 genealogies of sampled sequences in a recombining species\cite{Wong2024},
-and provide a natural and efficient means of incorporating recombination 
+and provide a natural and efficient means of 
+systematically incorporating recombination 
 into the study of SARS-CoV-2.
@@
 Wohns2022,Zhang2023,Gunnarsson2024,Deng2025robust} have led to a surge 
-of interest in their application in population and statical 
+of interest in their application in population and statistical 
 genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
@@
 
-[TODO rewrite this paragraph]
-Here, we present
-(1) sc2ts, a new method to infer ARGs in real time by building genealogies forward in time, and
-(2) an ARG for SARS-CoV-2 with ~2.48 million genome sequences.
-We used the ``Viridian'' dataset,
-which is purged of various systematic errors and artifacts \cite{Hunt2024} and
-show that this ARG is highly congruent to an UShER phylogeny built from the same dataset.
-Compared to previously identified recombinants,
-we show that this ARG contains the vast majority of epidemiologically relevant signals of recombination,
-while adding significant nuance and texture.
-This ARG – which captures the major signals of evolution
-through phylogenetic relationships, point mutations, deletions, and recombination –
-is a resource of unprecedented detail and precision, and
-should provide a strong basis for future studies of SARS-CoV-2.
+Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
+that incrementally adds samples over time in a similar manner 
+to UShER, but also automatically detects and incorporates recombination
+in real-time. Based on the Viridian dataset, a freely available  
+resource purged of various systematic errors and artifacts \cite{Hunt2024},
+we infer an ARG with 2.48 million whole genome sequences.
+We show that this ARG is highly congruent with the UShER phylogeny built 
+from the same dataset as well as the Pango lineage naming system\cite{Rambaut2020},
+and also accurately captures evolutionary features such as mutational 
+spectra\cite{Bloom2023} and deletions\cite{Li2023indels}.
+In addition, we show that sc2ts automatically and accurately detects details about
+known recombinant lineages that were painstakingly collated in a community 
+process, and provides novel insights into the relationships between these
+recombinant lineages. By fully incorporating these and other novel
+recombination events, the resulting data structure---representing
+phylogenetic relationships, point mutations, deletions, and 
+recombination---provides a basis for the study of SARS-CoV-2
+that systematically accounts for all of these evolutionary forces.
+This ARG, coupled with the rich suite of supporting software
+in the tskit\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
+enables pandemic-scale analysis on a laptop. 
+% Dunno about this! Feels important to finish on a wider note, helping
+% the reader understand the significance of all of this?
+These resources
+will help us to learn from the vast troves of data collected during 
+pandemic, and to prepare for the next.
 

### 2025-09-25 latex/main.tex
Updates from Overleaf
Pass on "Overview of sc2ts"

@@
 recombinant lineages.
-Together, the inference method, infererred ARG and 
+Together, the inference method, inferred ARG and 
 suite of open-source analysis tools, 
@@
 These resources
-will help us to learn from the vast troves of data collected during 
+will help us to learn from the vast troves of data collected during the
 pandemic, and to prepare for the next.
@@
 
-\subsection*{Brief description of sc2ts}
-
-Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available
-(Figure~\ref{fig:methodology}; for details, see STAR Methods).
-First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
-(allowing recombination) using the Li and Stephens (LS) model \cite{LiStephens2003},
-which is a Hidden Markov Model (HMM) widely used in human genomics
-to approximate the effects of mutation and recombination on genetic inheritance.
-As recombination is relatively rare in viruses,
-we would expect most samples to copy from a single parent in the ARG.
-Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path
-between the parents resulting in different genomic segments inherited from the different parents.
-Next, for clusters within the batch that are inferred to copy from the same ARG node,
-sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster.
-Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
-applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.
-[TODO add brief summary of sample filtering and how we handle VOCs, 
-refing Methods for details]
+\subsection*{Overview of sc2ts}
 
@@
 
-%% JK: This is copy of old Methods section "overview of sc2ts" which is basically
-%% the same as this section. TODO Merge the useful text in here.
-
-% Sc2ts (pronounced ``scoots'', optionally)
-% is a method for inferring Ancestral Recombination Graphs (ARGs)
-% from densely sampled genome sequence data in real time,
-% in which recombination occurs at a low but significant rate.
-% The basic idea is to incrementally update an ARG regularly
-% with the sequences collected since the previous batch (Figure
-% \ref{fig:methodology}).
-% We used daily batches for SARS-CoV-2,
-% and use days as the time interval for clarity in what follows.
-% The first step is to find likely ``copying paths'' under the Li and Stephens (LS) model \cite{LiStephens2003}
-% for each sequence in the daily batch to the current ARG.
-% These copying paths will mostly consist of a new sample sequence copying
-% from a node in the ARG with a small number of mutations,
-% and often many samples from a daily batch will copy from the same ARG node.
-% The second step is then to ``resolve'' these implied polytomies
-% by using a standard neighbor-joining algorithm \cite{Saitou1987}.
-% This greedy update strategy inevitably leads to unparsimonious topologies,
-% and the third step is to then increase the overall parsimony of the inferred ARG
-% by making some simple topological updates.
-% Recombination is inferred as an integral and ongoing part of this process,
-% requiring only a few additional steps to facilitate later analysis.
-% The result of inference for a given day is then a genealogy recording genetic inheritance
-% as well as mutation and recombination events
-% for all of the samples added to the ARG up to that day.
-
-% Sc2ts is implemented in Python building on core PyData infrastructure
-% such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
-% and Pandas~\cite{Mckinney2010data} ,
-% using pyfaidx\cite{Shirley2015efficient} to access alignment data
-% and Numba~\cite{Lam2015numba} to accelerate certain alignment processing
-% tasks.
-% [TODO JK to fill this out]
+Sc2ts reconstructs an ARG by incrementally adding daily batches of 
+sequences (Figure~\ref{fig:methodology}).
+First, sc2ts infers likely ``copying paths'' connecting each sample in the 
+batch to node(s) in the current ARG
+(allowing recombination) using a simplified version of the Li and Stephens (LS) model
+\cite{LiStephens2003}. 
+% Important to go into a bit of detail about LS here as it's so fundamental
+% and novel in the area.
+The LS model is a Hidden Markov Model (HMM) widely used in human genomics
+to approximate the effects of mutation and recombination on genetic
+inheritance\citep{Delaneau2019-wl,Browning2018-nk},
+typically parameterised by population-scaled mutation and
+recombination rates varying along the genome~\cite{Donnelly2010-coalescent}.
+In the interest of interpretability, we simplified this model to 
+use a single parameter $k$ which essentially 
+controls the number of recurrent mutations that is preferred to 
+switching to a different parent via recombination. We used the 
+highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
+with some additional optimisations enabled by our simplified model,
+to find exact Viterbi solutions of these copying paths for each sample
+given the current ARG.
+Note that this is a form of parsimony, with the recombination
+penalty $k$ determining the relative parsimoniousness of recombinant
+vs non-recombinant solutions to the HMM.
+The per-sample copying paths produced by the HMM (most of which do not 
+involve recombination) form clusters of samples within a batch, which 
+are then resolved using standard phylogenetic techniques.
+Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
+applies parsimony-improving heuristics to address issues introduced 
+by the inherent greediness of this strategy.
+Sc2ts is implemented in Python,
+building on core PyData infrastructure~\cite{Harris2020array,Virtanen2020scipy,
+Mckinney2010data,Lam2015numba} and bioinformatics 
+packages\cite{Shirley2015efficient,Kunzmann2018biotite,Kunzmann2023biotite}.
+See STAR Methods for details on all aspects of the method.
 
@@
 consists of XZ, XAC, XAD, XAE, and XAP,
-which descend from a single recombination event rather than five independent
-events in the ARG (Figure~\ref{fig:subgraphs}B), and
+which descend from a single recombination event in the ARG, rather than five independent
+ones (Figure~\ref{fig:subgraphs}B), and
 which we tentatively group using the unused label Xx.

### 2025-09-25 latex/main.tex
Update to overview

@@
 Schematic overview of sc2ts.
-(A) Process to build an ARG by incrementally updating it with samples sorted by collection dates.
+(A) Process to build an ARG by incrementally updating it with samples 
+sorted by collection dates.
 (B) The ARG resulting from the inference process.
@@
 When Sample 4 is detected as a recombinant,
-a node representing the recombination event is inserted with two edges connecting to its parents,
+a node representing the recombination event is inserted with two edges 
+connecting to its parents,
 with Sample 4 itself attached to this node as its child.
 (C) Two local trees contained in the resulting ARG,
-one on each side of the inferred breakpoint (A), which occurs at position 50.
+one on each side of the inferred breakpoint which occurs at position 50.
 The genomic coordinates of the segments inherited from different recombinant parents
-are denoted by two half-open intervals (B).
+are denoted by two half-open intervals.
 (D) Copying paths for Sample 4 inferred under the LS model
@@
 controls the number of recurrent mutations that is preferred to 
-switching to a different parent via recombination. We used the 
-highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
+switching to a different parent via recombination (we use $k=4$ here).
+We used the highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
 with some additional optimisations enabled by our simplified model,

### 2025-09-25 latex/main.tex

--- 2025-09-25 latex/main.tex
@@
 
-Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
-Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM 
-(~5 CPU years; Figure~\ref{fig:inference_resources}).
-The ARG contains 2,689,054 nodes representing the genealogical relationships
-among 2,482,157 sampled genomes in the Viridian v04 dataset,
-which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
+Using sc2ts, we built an ARG from 2,482,157 samples from the Viridian v04
+dataset\cite{Hunt2024}, collected between 2020-01-01 and 2023-02-20.
+Inference took 21 days on a XX core server with 500 GB of RAM 
+($\sim 5$ CPU years; Figure~\ref{fig:inference_resources}).
+After postprocessing, the ARG 
+contains 2,689,054 nodes with estimated dates, 
+2,748,838 edges, and 2,285,344 mutations at 29,893 sites. 
+Including sample IDs and the Pango and Scorpio designations computed by
+Pangolin~\cite{OToole2021assignment} as metadata for each node,
+the ARG is stored in a 32MiB tszip format file.
 Of the nodes, 855 represent recombination events, 
-of which 354 pass quality control (QC) filters
-(see STAR Methods, Figure~\ref{fig:recombinant_qc}),
-reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.
-
-The ARG represents the full evolutionary history of SARS-CoV-2 and 
-is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}.
-First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
+resulting in 316 distinct trees along the genome (some
+events occured at the same breakpoints).
+% It's confusing to talk about recombinant QC here as we're talking 
+% about the full ARG for the rest of the section.
+
+The ARG is a comprehesive record of the evolutionary history of SARS-CoV-2 and 
+is in close agreement with a state-of-the-art UShER phylogeny also built 
+from the Viridian dataset \cite{Hunt2024}.
+First, we validated the broad structure of the ARG against 
+the UShER phylogeny \cite{Hunt2024}
 by pruning both down to representative samples for 1097 deeply-sampled 
@@
 the mutational spectra 
-from the Viridian UShER phylogeny \cite{Hunt2024} and a previous study \cite{Bloom2023} (STAR Methods),
-finding a close match to them (Figure~\ref{fig:mutational_spectra}).
-
+from the Viridian UShER phylogeny and a previous study \cite{Bloom2023} (STAR Methods),
+finding close agreement (Figure~\ref{fig:mutational_spectra}).
+
+% FIXME this is too detailed, and too methodsy. Abstract out and summarise
+% results
 Input genome alignments are masked in sc2ts

### 2025-09-25 latex/main.tex
Decent first paragraph of pango X section
Updata

Update

@@
 
-The ARG is a comprehesive record of the evolutionary history of SARS-CoV-2 and 
+The ARG is a comprehensive record of the evolutionary history of SARS-CoV-2 and 
 is in close agreement with a state-of-the-art UShER phylogeny also built 
@@
 the UShER phylogeny \cite{Hunt2024}
-by pruning both down to representative samples for 1097 deeply-sampled 
-lineages (STAR Methods).
+by pruning both down to representative samples
+for 1097 deeply-sampled lineages (STAR Methods).
 Tanglegrams show that the ARG closely matches the phylogenetic
@@
 (Document~\ref{sec:supplement_sc2ts_usher_imputation}).
-We also compared the mutational spectra of major VOCs in the ARG against 
-the mutational spectra 
-from the Viridian UShER phylogeny and a previous study \cite{Bloom2023} (STAR Methods),
+We also compared the mutational spectra of major VOCs in the ARG
+against the mutational spectra from
+the Viridian UShER phylogeny and
+a previous study \cite{Bloom2023} (STAR Methods),
 finding close agreement (Figure~\ref{fig:mutational_spectra}).
@@
 28271 (non-coding).
-The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021},
+The Spike 69-70 deletion,
+which causes S-Gene Target Failure \cite{Walker2021},
 also emerges twice in the origin of Omicron -
@@
 and the node leading to the BA.4 variant (189,746 descendants).
-To speed up inference and to avoid issues caused by sites enriched for errors [citation?],
-sc2ts masks out a set of 100 sites identified as the most highly homoplasic
+To speed up inference and to avoid issues
+caused by sites enriched for errors [citation?],
+sc2ts masks out a set of 100 sites identified
+as the most highly homoplasic
 by running on a subset of the data (STAR Methods).
@@
 masked by UShER \cite{Hunt2024}.
-In addition to remapping the data for sites identified as being involved in deletions,
+In addition to remapping the data for sites identified
+as being involved in deletions,
 we added mutations for XX of these highly recurrent sites
 that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
-Each node in the ARG was then assigned to one of 2058 Pango lineages using pangolin (STAR Methods).
-All but 21,680 sampled sequences (0.87\%) agreed with the source sequences in terms of Pango lineage assignment
+Each node in the ARG was then assigned
+to one of 2058 Pango lineages using pangolin (STAR Methods).
+All but 21,680 sampled sequences (0.87\%) agreed
+with the source sequences in terms of Pango lineage assignment
 (comparable to the 0.27\% disagreement between pangolin-data
 versions 1.21 and 1.29 in the source Viridian metadata).
-A majority of these lineages (1473 of 2058) have monophyletic origins within the ARG,
-increasing to 1779 (86\%) when we allow for multiple sibling origination 
-nodes (Document~\ref{sec:supplement_pango_lineages}).
+A majority of these lineages (1473 of 2058)
+have monophyletic origins within the ARG,
+increasing to 1779 (86\%) when we allow
+for multiple sibling origination nodes
+(Document~\ref{sec:supplement_pango_lineages}).
 
@@
 
-As new data became available during the pandemic,
-volunteers continually detected recombinants which might have concerning features,
-particularly increased transmissibility, disease severity, and
-ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2023}).
-The process of gathering evidence for and characterizing such recombinants is manual,
-involving visual inspection of phylogenies for unusual branches
-caused by forcibly placing recombinant sequences onto them,
-typically followed by visual analysis of genome sequence alignments
-to identify parental lineages and breakpoint locations.
-This process has resulted in the designation of recombinant Pango lineages
-(whose aliases begin with an ''X'', e.g., XA and XBB).
-
-We find strong evidence for the recombination events leading to 17 Pango X lineages
-in the ARG (Table~\ref{tab:pango_x_lineages}),
-all of which pass QC (STAR Methods).
-%% JK: Does saying this here help at all? I think we may be better off without 
-% These events are supported by high numbers of averted
-% mutations (Table 1) and high numbers of loci matching their suggested parents
-% (see their copying patterns in Figure SX). 
-Among these Pango X lineages are XA,
-which is the first designated Pango X lineage that
-showed signs of onward transmission \cite{Jackson2021}, and
-XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
-The recombinant origin of XA is well explained in the ARG by two parents and one mutation
-(Figure~\ref{fig:subgraphs}A; Table~\ref{tab:pango_x_lineages}), 
-with sc2ts correctly identifying its parental Pango lineages and breakpoint location.
+[TODO narrative is muddled here: the point of the para is to say that 
+pango X lineage designation is manual and subjective.]
+% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
+% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
+% ‘Pangolin’ nomenclature system. "
+Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
+can be attributed to the clarity of the rules under which new lineages 
+are designated and their explicitly phylogenetic basis. 
+The first recombinant lineage XA was designated in 2021 based on the 
+analysis of Jackson et al\.~\cite{Jackson2021}, which established 
+that recombination with onward transmission had occured [weak FIXME].
+As of pango-designation version v1.33.1 (2025-03-31), a total of 146 distinct 
+X lineages have been designated.
+%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
+% 146
+Although the process of incorporating recombinant lineages in the Pango 
+nomenclature system is well defined,
+the identification of such events is much more subjective, requiring
+the synthesis of multiple lines of evidence and often the 
+manual inspection of phylogenies and sequence alignments
+by volunteers.
+
+By automatically detecting and incorporating recombination events
+into the ARG, sc2ts allows us to model the origins of these 
+X lineages as nodes in the graph. 
+Table~\ref{tab:pango_x_lineages} summarises
+these results for the XX Pango X lineages present in
+the ARG. 
+For 16 Pango X lineages the oldest node that has been assigned
+the focal Pango lineage is a recombination node, and that node 
+is the ancestor of all samples assigned that (or a descendant) Pango 
+lineage. These are the Type I events, in which the ARG reflects the
+expected phylogenetic structure of Pango designation.
+Among these events, concordance with the community-consensus 
+parent lineages and breakpoint intervals is excellent (see next 
+section). 13 of the Type I events are monophyletic for the focal
+pango; Figure~\ref{fig:subgraphs}A shows XA as an example.
+The remaining 3 Type I events (XBB, XBQ and XM) have descendants not assigned to the 
+focal pango, many of which are the generic BA.2 label
+[FIXME presumably due to uncertainty? Look up the results CSV? Be precise].
+
+The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
+are recombination events closely associated with Pango X lineages, 
+but that do not meet the strict origination criteria for Type I.
+Type III events are then nodes in the ARG associated with the 
+orin
+
+Of these, XAF and XJ are relatively straightforwardo
 
@@
 % really are recombinants, especially XP.
+% YW: we now have a better in-depth breakdown of these recombinants
+% to be pasted into doc S1 - e.g. XB is missing the parent lineages, XP is defined on the basis
+% of 2 loci, and XAJ on the basis of 3.
 However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
@@
 
-\subsection{Analysis of deletions}\label{sec:supplement_dels}
+\subsection{Analysis of deletions}
+\label{sec:supplement_dels}
 
 % See https://github.com/jeromekelleher/sc2ts-paper/issues/628
-By mapping deletions to the ARG, we identified
-important deletion events (Table~\ref{tab:major_dels}),
-many of which are associated with the origins of major VOCs,
-for example, the deletions in ORF1ab and
-S1-NTD (the N-terminal domain of Spike)
-associated with the Alpha lineage.
-We also identified highly recurrent deletions (Table~\ref{tab:rec_dels}),
-for example, two known recurrent deletions in S1-NTD (REF).
-
-Li et al.\cite{Li2023indels} performed a large-scale analysis of indels
-using a global GISAID dataset (sampled up to August, 2022).
-We examined the major indels reported in Table S1 of Li et al.\cite{Li2023indels}
-which have a minimum frequency of 1\%
-(count of sequences divided by 9,149,680 sequences analyzed).
-All of these indels are deletions and have high frequencies in the ARG
-(sequence count divided by the total number of samples),
-except a 9-base deletion at 11283, a 9-based deletion at 21987, and
-a 9-base deletion at 686.
+We reconstructed a more comprehensive mutational landscape
+of SARS-CoV-2 by mapping deletions onto the ARG
+using post-hoc parsimony.
+We identified important deletion events
+(which are inherited by at least 10,000 samples;
+Table~\ref{tab:major_dels}).
+All these deletions are known to occur major VOCs
+(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
+\cite{Akaishi2023,Li2023indels}).
+Eight of these deletions occur in S1-NTD,
+a known hotspot for recurrent deletions \cite{McCarthy2021}.
+There are three contiguous deletions in NSP6
+(4-base at 11288, 5-base at 11283, and 5-base at 11292),
+which should have been mapped as two 9-base deletions,
+one above the BA.1 origin node and
+the other above the BA.2 origin node in the ARG
+(Figure~\ref{fig:subgraph_ba1_ba2}),
+but they were broken up
+during deletion remapping because they had
+overlapping four sites at 11288-11292
+(which was mapped as the 4-base deletion event).
+Furthermore, the placement of these deletion events in the ARG
+reflects that of the VOC-associated deletion events
+inferred in a previous study \cite{Akaishi2023}.
+
+We also found highly recurrent deletions using the ARG.
+The top 20 deletions in the number of occurrences in the ARG
+are shown in Table~\ref{tab:rec_dels}.
+About half of these deletions are one-base,
+occur in non-coding regions, and
+are concentrated in the 3' end of the genome.
+Four one-base deletions occur
+in accessory genes (ORF7a, ORF8, and ORF10).
+The two deletions in S1-NTD are known recurrent deletions
+associated with antibody escape \cite{McCarthy2021}.
 
@@
 \centering
-\caption{Major deletion events in the ARG.
+\caption{
+Major deletion events in the ARG.
 These events are inherited by at least 10,000 samples in the ARG.
 Count is the number of samples inheriting a deletion.
-The Pango lineage assigned to the node above which an event occurs is shown.
-Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in S1 subunit.}
+The Pango lineage assigned to the node
+above which an event occurs is shown.
+Abbreviations: NSP, non-structural protein;
+ORF, open reading frame;
+S, Spike;
+S1-NTD, N-terminal domain in S1 subunit of Spike.
+}
 \label{tab:major_dels}
@@
 \centering
-\caption{Highly recurrent deletions.
-These are the top 20 deletions in terms of the number of occurrences in the ARG.
-Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in S1 subunit.}
+\caption{
+Highly recurrent deletions.
+These are the top 20 deletions
+in terms of the number of occurrences in the ARG.
+Abbreviations:
+NSP, non-structural protein;
+ORF, open reading frame;
+S, Spike;
+S1-NTD, N-terminal domain in S1 subunit in Spike.
+}
 \label{tab:rec_dels}
@@
 \subsubsection{XE/XH}
-https://github.com/jeromekelleher/sc2ts-paper/issues/337
+https://github.com/jeromekelleher/sc2ts-paper/discussions/633
 
@@
 with one mutation and one reversion (T22792C).
-This reversion was mapped onto the ARG as a part of postprocessing, and
-could be placed more parsimoniously by a single C22792T that groups together
-XH and the lineage leading to the further BA.2 recombination nodes
-displayed on the left.
+This reversion was mapped onto the ARG
+as a part of postprocessing,
+and could be placed more parsimoniously by a single C22792T
+that groups together XH
+and the lineage leading to the BA.2 recombination nodes
+displayed on the lower left.
 Nevertheless, the exact arrangement of the mutations at position 22792
@@
 which is very distant, at position 11283.
-
-*** Should we describe the sequential recombination nodes to the left,
-here: note that the lower one is the RE node for XAF. ***
-
-Note that there might be some misalignment immediately
-to the right side of the breakpoint
-due to a repeated homopolymeric sequence involving 5-base deletions.
-Although this problem might obscure the breakpoint position,
-there nevertheless is ample evidence for the right parent.
-Furthermore, the only two XH samples in the ARG descend
-from the same recombination node,
-albeit with a number of mutations.
+This breakpoint coincides with a repeated homopolymeric sequence
+involving 5-base deletions,
+so that alignment errors could have a small effect
+on the breakpoint position.
+Note that the two lower recombination nodes are both well supported,
+the youngest being postulated as the origin of XAF (see below);
+we leave investigation of this suggested series of
+sequential recombination events to future work.
 
@@
 \subsubsection{Xx: XZ/XAC/XAD/XAE/XAP}
-WIP
+https://github.com/jeromekelleher/sc2ts-paper/discussions/631
+
+This nested set of Pango X lineages descends
+from a single recombination node in the ARG, and
+we suggest that these Pango X lineages
+may be aliases of a single recombinant
+which we tentatively refer to as "Xx".
+Note that all were designated around the same time,
+with their samples collected
+roughly within a three-month period (Mar., 2022 to Jul., 2022).
+Several samples labeled as BA.2 by pangolin also lie
+within this recombinant clade,
+including two that are identical
+to the proposed originating recombinant (ERR8146303 and ERR8163061):
+we suggest that researchers examining the emergence of
+these Pango X lineages could direct efforts
+into examination of these and closely related samples.
+The detection of Xx illustrates
+the ability of sc2ts to clarify relationships between
+samples during the course of a pandemic,
+informing designation of emerging recombinants.
+
+Exact relationships within the Xx clade are somewhat tentative.
+For example, there are three highly mutable sites (22792, 28877, and 28878)
+which were not used during inference,
+but were remapped onto the ARG during postprocessing.
+The inferred topology places these sites
+as recurrent (colored orange, grey, and faded yellow),
+but a more parsimonious arrangement would,
+for example, group the two XAD samples as siblings
+(removing the need to postulate mutations
+A1250G, A28877T, and G28878C to be recurrent).
+It is possible that a more parsimonious arrangement would also
+reduce the recurrent mutations and reversions seen at
+sites 22,792 (four orange mutations) and
+26,060 (two parallel brown mutations),
+but we do not show such a reconstruction,
+which is complicated by the presence of
+additional (non-displayed) samples descending from
+internal nodes in the plotted subgraph.
+
+For XAC, in particular, the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/590) suggests
+it to be a recombinant with two breakpoints:
+BA.2 on the left, BA.1 in the middle, and
+a switch back to a very small region of BA.2 on the right.
+This contrasts with the one-breakpoint BA.2+BA.1 origin inferred in the ARG.
+However, only two loci support the right hand BA.2 assignment:
+A29510C and a large deletion
+(deletions are treated as missing data during ARG inference).
+Sc2ts does not count this as enough evidence
+to insert an additional breakpoint,
+which would require at least four supporting variants.
 
@@
 BA.2.12.1 (left parent) and BA.4 (right parent).
-XAJ was detected as a non-recombinant by sc2ts.
-There are only three loci supporting
-the right BA.4 parent in the ARG (T22917G, T23018G, and G23040A),
-which means that XAJ fell below our threshold of detection.
-There is also a deletion (H69 in Spike) supporting
-the BA.4 parent (see the Pango designation issue),
-but it was treated as missing data during ARG inference.
-If this deletion were accounted for,
-XAJ should have been detected as a recombinant.
+However, XAJ is not inferred to be a recombinant lineage in the ARG.
+From the subgraph, which shows the 18 XAJ samples
+plus a small number of relevant BA.4 samples,
+it can be seen that 9 sites leading to XAJ indeed recur
+in the lineage to the closest BA.4 sample (SRR18848050),
+which strongly indicates that this is a recombinant
+that is missed by sc2ts.
+The reason that is is not a recombinant in the ARG is
+that 6 sites represent a 6bp deletion and so are not used for inference.
+The remaining three sites (T22917G, T23018G, and G23040A)
+are not enough to exceed our k=4 threshold  for recombinant designation.
+Including deletions in inference would certainly
+lead to XAJ being detected as a recombinant,
+and with a lower threshold of k=3
+the distribution of mutations on the ARG topology suggests
+that it is is even possible that XAJ is a 2-breakpoint recombinant
+with the region from 21765-23535
+coming from the BA.4 labeled sample SRR18848050.
 
@@
 \subsubsection{XBB}
-WIP
+https://github.com/jeromekelleher/sc2ts-paper/issues/336
+
+Most of the XBB samples descend from a well supported recombination node.
+In the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/1058),
+the Pango lineages of the suggested left and right parents are
+BJ.1 and BM.1.1.1, respectively.
+In the ARG, the Pango lineage of the right parent is identical,
+but that of the left parent is a related ancestral Pango lineage
+(BA.2.10 instead of BA.2.10.1.1 = BJ.1).
+This is likely due to the absence of BJ.1 in the ARG:
+only four BJ.1 samples exist in the Viridian v04 dataset,
+a single one passed sequence-level QC, and
+that was then rejected due to high HMM cost.
+Note that BJ.1 differs from BA.2.10 by over 20 mutations
+(https://github.com/cov-lineages/pango-designation/issues/915),
+explaining the large number of (gold) \emph{de novo} mutations
+in the XBB copying pattern.
+Had BJ.1 samples been present in the ARG,
+we strongly suspect that the left branch of
+the XBB-associated recombination node
+would have been attached to a BJ.1-labeled node, and
+the origin of XBB would not require postulating as many new mutations.
+The fact that sc2ts has deduced
+the XBB left parent to a close relative of BJ.1,
+even though BJ.1 is missing from the ARG,
+shows our approach to be relatively robust to missing un-sampled lineages:
+an important consideration given global variation in pathogen surveillance.
 
 \subsubsection{XBB.1}
-WIP
+XBB.1 is characterized by an additional mutation in Spike, G22317T (G252V) and
+is not designated as a recombinant of XBB in the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/1088).
+The ARG agrees with this and places the characteristic mutation
+above the root of a single non-recombinant XBB.1 clade.
+We include this non-recombinant lineage
+in order to highlight some well supported subsequent recombination nodes
+within XBB.1, such as the (pink) one leading to two XBB.1-labeled samples
+at the bottom left of the subgraph which we refer to as XBB.x.
+XBB.x is well supported: its left parent is a reconstructed internal node,
+which is directly supported by a BQ.1-labeled sample, and
+its right parent is a BB.1.5.18-labeled sample (ERR10792363).
+This highlights the automatic ability of sc2ts
+to infer sequential recombination events
+(i.e. recombinants-of-recombinants),
+even those involving few samples.
+A few other recombination events occur within XBB.1,
+but they either fail QC checks, or
+(as the other one shown here, to the bottom right)
+lead to a single sample so may be
+e.g. caused by accidental mixing of samples during sequencing.
 
@@
 \subsubsection{XBE}
-WIP
+https://github.com/jeromekelleher/sc2ts-paper/issues/351
+
+In the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/1246),
+XBE was suggested as a recombinant between
+BA.5.2 (left parent; most of the genome) and
+BE.4 (right parent; somewhere from 23,609 onwards). 
+owever, sc2ts treats the 65 XBE samples
+as non-recombinant descendants of BA.5.2
+because the number of supporting sites
+fell below our threshold of detection.
+We have included the BE.4 lineage on the subgraph,
+revealing two mutations (C27800T, and G28681T)
+which occur in parallel above the root of the XBE clade
+and along the path ascending from the displayed BE.4 sample,
+suggesting that if we used k=2
+as the recombinant threshold,
+sc2ts would indeed detect XBE as a recombinant.
 
@@
 \subsubsection{XBQ/XBK/XBK.1/CJ.1.3}
-WIP
+https://github.com/jeromekelleher/sc2ts-paper/issues/349
+https://github.com/jeromekelleher/sc2ts-paper/issues/350
+
+XBK and XBQ are siblings, and neither appear to be recombinants in the ARG,
+as they are closely related to the non-recombinant node that originates CJ.1.3. 
+At the time of designating XBK
+a hypothesis was made that a missing common ancestor could exist 
+(https://github.com/cov-lineages/pango-designation/issues/1440\#issuecomment-1355264973),
+making XBK non-recombinant.
+Exactly such a missing ancestor
+is found as the upper internal node labeled CJ.1 in the ARG,
+constructed by sc2ts as part of its tree building steps.
+We suggest that sc2ts can therefore play an epidemiological role
+both in generating hypotheses involving recombination, and
+quantifying mutational evidence to support them.
 

### 2025-09-25 latex/main.tex
Updates from Overleaf

--- 2025-09-25 latex/main.tex
@@
 \subsubsection{XBE}
-WIP
+https://github.com/jeromekelleher/sc2ts-paper/issues/351
+
+In the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/1246),
+XBE was suggested as a recombinant between
+BA.5.2 (left parent; most of the genome) and
+BE.4 (right parent; somewhere from 23609 onwards).
+However, sc2ts treats the 65 XBE samples
+as non-recombinant descendants of BA.5.2
+because the number of supporting sites fell
+below our threshold of detection.
+We have included the BE.4 lineage on the subgraph,
+revealing three mutations (G22599C, C27800T, and G28681T)
+which occur in parallel above the root of the XBE clade and
+along the path ascending from the BE.4 sample,
+suggesting that if we used k=3 as the recombinant threshold,
+sc2ts would indeed detect XBE as a recombinant.
 

### 2025-09-26 latex/main.tex
Updates from Overleaf

--- 2025-09-25 latex/main.tex
@@
 
-[TODO narrative is muddled here: the point of the para is to say that 
-pango X lineage designation is manual and subjective.]
-% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
-% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
-% ‘Pangolin’ nomenclature system. "
-Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
-can be attributed to the clarity of the rules under which new lineages 
-are designated and their explicitly phylogenetic basis. 
-The first recombinant lineage XA was designated in 2021 based on the 
-analysis of Jackson et al\.~\cite{Jackson2021}, which established 
-that recombination with onward transmission had occured [weak FIXME].
-As of pango-designation version v1.33.1 (2025-03-31), a total of 146 distinct 
-X lineages have been designated.
-%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
-% 146
-Although the process of incorporating recombinant lineages in the Pango 
-nomenclature system is well defined,
-the identification of such events is much more subjective, requiring
-the synthesis of multiple lines of evidence and often the 
-manual inspection of phylogenies and sequence alignments
-by volunteers.
-
-By automatically detecting and incorporating recombination events
-into the ARG, sc2ts allows us to model the origins of these 
-X lineages as nodes in the graph. 
-Table~\ref{tab:pango_x_lineages} summarises
-these results for the XX Pango X lineages present in
-the ARG. 
-For 16 Pango X lineages the oldest node that has been assigned
-the focal Pango lineage is a recombination node, and that node 
-is the ancestor of all samples assigned that (or a descendant) Pango 
-lineage. These are the Type I events, in which the ARG reflects the
-expected phylogenetic structure of Pango designation.
-Among these events, concordance with the community-consensus 
-parent lineages and breakpoint intervals is excellent (see next 
-section). 13 of the Type I events are monophyletic for the focal
-pango; Figure~\ref{fig:subgraphs}A shows XA as an example.
-The remaining 3 Type I events (XBB, XBQ and XM) have descendants not assigned to the 
-focal pango, many of which are the generic BA.2 label
-[FIXME presumably due to uncertainty? Look up the results CSV? Be precise].
-
-The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
-are recombination events closely associated with Pango X lineages, 
-but that do not meet the strict origination criteria for Type I.
-Type III events are then nodes in the ARG associated with the 
-orin
-
-Of these, XAF and XJ are relatively straightforwardo
+As new data became available during the pandemic,
+volunteers continually detected recombinants which might have concerning features,
+particularly increased transmissibility, disease severity, and
+ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2023}).
+The process of gathering evidence for and characterizing such recombinants is manual,
+involving visual inspection of phylogenies for unusual branches
+caused by forcibly placing recombinant sequences onto them,
+typically followed by visual analysis of genome sequence alignments
+to identify parental lineages and breakpoint locations.
+This process has resulted in the designation of recombinant Pango lineages
+(whose aliases begin with an ''X'', e.g., XA and XBB).
+
+We find strong evidence for the recombination events leading to 17 Pango X lineages
+in the ARG (Table~\ref{tab:pango_x_lineages}),
+all of which pass QC (STAR Methods).
+%% JK: Does saying this here help at all? I think we may be better off without 
+% These events are supported by high numbers of averted
+% mutations (Table 1) and high numbers of loci matching their suggested parents
+% (see their copying patterns in Figure SX). 
+Among these Pango X lineages are XA,
+which is the first designated Pango X lineage that
+showed signs of onward transmission \cite{Jackson2021}, and
+XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
+The recombinant origin of XA is well explained in the ARG by two parents and one mutation
+(Figure~\ref{fig:subgraphs}A; Table~\ref{tab:pango_x_lineages}), 
+with sc2ts correctly identifying its parental Pango lineages and breakpoint location.
 
@@
 Table~\ref{tab:major_dels}).
-All these deletions are known to occur major VOCs
-(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
-\cite{Akaishi2023,Li2023indels}).
-Eight of these deletions occur in S1-NTD,
-a known hotspot for recurrent deletions \cite{McCarthy2021}.
-There are three contiguous deletions in NSP6
-(4-base at 11288, 5-base at 11283, and 5-base at 11292),
-which should have been mapped as two 9-base deletions,
-one above the BA.1 origin node and
-the other above the BA.2 origin node in the ARG
-(Figure~\ref{fig:subgraph_ba1_ba2}),
-but they were broken up
-during deletion remapping because they had
-overlapping four sites at 11288-11292
-(which was mapped as the 4-base deletion event).
-Furthermore, the placement of these deletion events in the ARG
-reflects that of the VOC-associated deletion events
-inferred in a previous study \cite{Akaishi2023}.
+All of these deletions
+(except a one-base deletion at 28,271
+above a B.1-labeled node)
+are lineage-defining mutations of major VOCs
+(Alpha, Delta, Omicron BA.1, BA.2, and BA.4).
+Many of these deletions occur in S1-NTD,
+a known hotspot of in-frame deletions (REF).
+There is a recurrent 1-base deletion at 28,271,
+which is in a non-coding region.
+<TODO: The 5-base dels and 4-base del are split up.
+It should be a 9-base del (11,288–11,296).
+I don't see excluded sites among these deleted sites.>
 
 We also found highly recurrent deletions using the ARG.
-The top 20 deletions in the number of occurrences in the ARG
+The top 20 deletions in the number of occurrences
 are shown in Table~\ref{tab:rec_dels}.
-About half of these deletions are one-base,
-occur in non-coding regions, and
-are concentrated in the 3' end of the genome.
-Four one-base deletions occur
-in accessory genes (ORF7a, ORF8, and ORF10).
-The two deletions in S1-NTD are known recurrent deletions
-associated with antibody escape \cite{McCarthy2021}.
+About half of these deletions affect a single base
+and occur in non-coding regions
+(concentrated in the 3' end of the genome);
+some occur in accessory genes ORF7a and ORF8.
+<TODO: Are any of these assembly errors or misalignments?>.
+Two of these are known recurrent deletions in S1-NTD,
+which are associated with the origins of major VOCs (REF).
 

### 2025-09-26 latex/main.tex
Updates from Overleaf

@@
 Table~\ref{tab:major_dels}).
-All of these deletions
-(except a one-base deletion at 28,271
-above a B.1-labeled node)
-are lineage-defining mutations of major VOCs
-(Alpha, Delta, Omicron BA.1, BA.2, and BA.4).
-Many of these deletions occur in S1-NTD,
-a known hotspot of in-frame deletions (REF).
-There is a recurrent 1-base deletion at 28,271,
-which is in a non-coding region.
-<TODO: The 5-base dels and 4-base del are split up.
-It should be a 9-base del (11,288–11,296).
-I don't see excluded sites among these deleted sites.>
+All these deletions are known to occur major VOCs
+(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
+\cite{Akaishi2023,Li2023indels}).
+Eight of these deletions occur in S1-NTD,
+a known hotspot for recurrent deletions \cite{McCarthy2021}.
+""" The 5-base dels and 4-base del are split up as separate events.
+It should be a 9-base del (11,288–11,296)."""
+Furthermore, these deletion events reflect
+the VOC-associated deletion events
+inferred in a previous study \cite{Akaishi2023}.
 
 We also found highly recurrent deletions using the ARG.
-The top 20 deletions in the number of occurrences
+The top 20 deletions in the number of occurrences in the ARG
 are shown in Table~\ref{tab:rec_dels}.
-About half of these deletions affect a single base
-and occur in non-coding regions
-(concentrated in the 3' end of the genome);
-some occur in accessory genes ORF7a and ORF8.
-<TODO: Are any of these assembly errors or misalignments?>.
-Two of these are known recurrent deletions in S1-NTD,
-which are associated with the origins of major VOCs (REF).
+About half of these deletions are one-base,
+occur in non-coding regions, and
+are concentrated in the 3' end of the genome.
+Four one-base deletions occur
+in accessory genes (ORF7a, ORF8, and ORF10).
+The two deletions in S1-NTD are known recurrent deletions
+associated with antibody escape \cite{McCarthy2021}.
 

### 2025-09-26 latex/main.tex
Update hte pango X events table

@@
 
-% This table is generated by the notebook tab_pango_x_events
 \begin{table}
@@
 ARG events associated with Pango X lineages.
-Each row corresponds to the oldest node assigned the specified
-Pango lineage and the ancestor of all nodes assigned that label
-(except XBB, XBB.x and XM; see STAR methods). [MORE DETAILS]
+Each row for Type I, III and IV events corresponds to the oldest node 
+assigned the specified Pango lineage and is the ancestor of all 
+nodes assigned that label. Type II events are recombination events
+associated with Pango X lineages not meeting Type I criteria.
+For Type I and II recombination 
+events, the averted shows the number of additional mutations required
+under a recombination-free HMM solution. Parent lineages shown in boldface
+are nodes exactly matched by sampled sequences. The descendants column
+shows the descendant samples broken down by Pango lineage (grouped 
+by root lineage, where appropriate).
 }
 \label{tab:pango_x_lineages}
-\begin{footnotesize}
-\begin{tabular}{lrrrllrl}
+% scriptsize is quite small here, and we could probably cram it all 
+% in with footnotesize if we worked a bit harder on the intercol
+% spacing
+\begin{scriptsize}
+\begin{tabular}{lrrlrll}
 \toprule
-\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
-pango & avt & left & right & left p & right p & samp & nested \\
+\multicolumn{7}{l}{Type I event: Recombination coinciding with Pango origin node}\\
+&&\multicolumn{2}{c}{Parent lineage} & \multicolumn{2}{c}{Interval}\\
+pango & averted & left & right & left & right & descendants \\
 \midrule
-XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
-XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
-XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
-XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
-XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
-XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
-XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
+
+XC & 37 & \textbf{AY.29} & \textbf{B.1.1.7} & 26768 & 27390 & {XC:5} \\
+XBR & 22 & \textbf{BN.3.1} & BQ.1.25 & 22034 & 22190 & {XBR:1} \\
+XA & 16 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 20411 & 21765 & {XA:39} \\
+XS & 11 & AY.103 & \textbf{BA.1.1} & 9054 & 10449 & {XS:17} \\
+XL & 8 & \textbf{BA.1.17.2} & \textbf{BA.2.5} & 5925 & 8393 & {XL:64} \\
+XBG & 8 & \textbf{BA.2.76} & \textbf{BA.5.2} & 22305 & 22917 & {XBG:25} \\
+XQ & 7 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 4322 & 5386 & {XQ:55, BA.2:37,
 XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
-XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
-XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
+XBD & 7 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 23020 & 24620 & {XBD:30} \\
+XM & 6 & \textbf{BA.1.1} & \textbf{BA.2} & 17411 & 21595 & {XM:26, BA.2:16,
 XAL:3} \\
-XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {} \\
-XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
-XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
-XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
-XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
-XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
-XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
+XBH & 6 & \textbf{BA.2.1} & BA.2.75.2 & 15452 & 22001 & {XBH:6} \\
+XBB & 6 & \textbf{BA.2.10} & BM.1.1.1 & 22332 & 22577 & {XBB:6452, BA.2:1} \\
+XY & 5 & \textbf{BA.1.1} & \textbf{BA.2} & 11538 & 12880 & {XY:23} \\
+XF & 5 & \textbf{AY.4} & \textbf{BA.1} & 5387 & 6402 & {XF:16} \\
+XBF & 5 & \textbf{BA.5.2.1} & CJ.1 & 5184 & 9866 & {XBF:185} \\
+XW & 4 & \textbf{BA.1.1.15} & \textbf{BA.2} & 2833 & 4321 & {XW:32} \\
+XG & 4 & \textbf{BA.1.17} & \textbf{BA.2} & 5925 & 6513 & {XG:3} \\
+
 \midrule
-\multicolumn{5}{l}{Type II RE event}\\
-alias & avt & left & right & left p & right p & desc & by pango \\
+\multicolumn{7}{l}{Type II event: Recombination closely associated with pango lineage(s)}\\
+&&\multicolumn{2}{c}{Parent lineage} & \multicolumn{2}{c}{Interval}\\
+name & averted & left & right & left & right & descendants \\
 \midrule
-Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 & {BA.2:156,
-XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
-XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
-{XE:1116, BA.2:37, XH:2, XAF:1} \\
-XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
-BF.3:2} \\
-XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
-XAF:1} \\
-XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
-BA.2:17} \\
-\bottomrule
-\multicolumn{5}{l}{Events closely derived from RE}\\
-pango & alias & path len & time & mutations & samples \\
+
+Xx & 6 & \textbf{BA.2} & \textbf{BA.1.17.2} & 24504 & 26060 & {BA.2:156, XZ:48,
+XAP:20, XAC:18, XAE:9, XAD:2} \\
+XE/XH & 6 & \textbf{BA.1.17.2} & \textbf{BA.2} & 10448 & 11283 & {XE:1116,
+BA.2:37, XH:2, XAF:1} \\
+XBM & 6 & \textbf{BA.2.76} & \textbf{BF.3} & 22305 & 22917 & {XBM:10, BF.3:2}
+\\
+XJ & 5 & \textbf{BA.1.17.2} & \textbf{BA.2} & 13196 & 17410 & {XJ:68, BA.2:17}
+\\
+XAF & 5 & \textbf{BA.1.1} & \textbf{BA.2} & 10199 & 10447 & {BA.2:35, XAF:1} \\
+
 \midrule
-XR & XQ & 3 & 11 & 4 & 17 \\
-XU & XQ & 6 & 103 & 13 & 1 \\
-XAM & XQ & 6 & 40 & 3 & 21 \\
-XAA & XQ & 6 & 33 & 3 & 17 \\
-XAG & XQ & 6 & 53 & 4 & 6 \\
-XAP & Xx & 4 & 66 & 1 & 20 \\
-XZ & Xx & 4 & 48 & 1 & 48 \\
-XAE & Xx & 2 & 47 & 7 & 9 \\
-XAF & XAF & 5 & 140 & 8 & 1 \\
-XAL & XM & 3 & 65 & 2 & 3 \\
+\multicolumn{7}{l}{Type III events: Pango origin nodes derived from 
+Type I and II events}\\
+pango & name & path len & time & mutations & samples \\
 \midrule
-\multicolumn{5}{l}{Non recombinant}\\
+
+XAF & XAF & 5 & 140 & 8   && {XAF:1} \\
+XBM & XBM & 1 & 23 & 1    && {XBM:10} \\
+XE  & XE/XH & 1 & 19 & 2  && {XE:1116} \\
+XH  & XE/XH & 1 & 55 & 7  && {XH:2} \\
+XJ  & XJ & 1 & 5 & 2      && {XJ:68, BA.2:3} \\
+XAL & XM & 3 & 65 & 2     && {XAL:3} \\
+XAA & XQ & 6 & 33 & 3     && {XAA:17} \\
+XAG & XQ & 6 & 53 & 4     && {XAG:6} \\
+XAM & XQ & 6 & 40 & 3     && {XAM:21} \\
+XR  & XQ & 3 & 11 & 4     && {XR:17, BA.2:1} \\
+XU  & XQ & 6 & 103 & 13   && {XU:1} \\
+XAE & Xx & 2 & 47 & 7     && {XAE:9} \\
+XAP & Xx & 4 & 66 & 1     && {XAP:20} \\
+XZ  & Xx & 4 & 48 & 1     && {XZ:48} \\
+
+\midrule
+\multicolumn{7}{l}{Type IV events: non recombinants}\\
 pango & muts & parent & samples & nested \\
 \midrule
-XP & 1 & \textbf{BA.1.1} & 45 & {} \\
-XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
-XBQ & 1 & CJ.1 & 14 & {} \\
-XBK & 1 & CJ.1 & 7 & {} \\
-XBE & 3 & BA.5.2 & 65 & {} \\
-XB & 2 & B.1 & 192 & {} \\
-XAZ & 3 & \textbf{BA.5} & 133 & {} \\
-XAV & 3 & BA.5.1.24 & 13 & {} \\
-XAU & 1 & BA.2 & 8 & {} \\
-XAS & 1 & BA.4 & 77 & {} \\
-XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
-XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
+
+XP  & 1  &&&& \textbf{BA.1.1} & {XP:45} \\
+XN  & 1  &&&& \textbf{BA.2} & {XN:120, BA.2:1} \\
+XBQ & 1  &&&& CJ.1 & {XBQ:14} \\
+XBK & 1  &&&& CJ.1 & {XBK:7} \\
+XBE & 3  &&&& BA.5.2 & {XBE:65} \\
+XB  & 2  &&&& B.1 & {XB:192} \\
+XAZ & 3  &&&& \textbf{BA.5} & {XAZ:133} \\
+XAV & 3  &&&& BA.5.1.24 & {XAV:13} \\
+XAU & 1  &&&& BA.2 & {XAU:8} \\
+XAS & 1  &&&& BA.4 & {XAS:77} \\
+XAN & 5  &&&& \textbf{BA.5.1} & {XAN:7} \\
+XAJ & 11 &&&& \textbf{BA.2.12} & {XAJ:18} \\
+
 \bottomrule
 \end{tabular}
-\end{footnotesize}
+\end{scriptsize}
 \end{table}
@@
 a known hotspot for recurrent deletions \cite{McCarthy2021}.
-""" The 5-base dels and 4-base del are split up as separate events.
-It should be a 9-base del (11,288–11,296)."""
-Furthermore, these deletion events reflect
-the VOC-associated deletion events
+There are three contiguous deletions in NSP6
+(4-base at 11288, 5-base at 11283, and 5-base at 11292),
+which should have been mapped as two 9-base deletions,
+one above the BA.1 origin node and
+the other above the BA.2 origin node in the ARG
+(Figure~\ref{fig:subgraph_ba1_ba2}),
+but they were broken up
+during deletion remapping because they had
+overlapping four sites at 11288-11292
+(which was mapped as the 4-base deletion event).
+Furthermore, the placement of these deletion events in the ARG
+reflects that of the VOC-associated deletion events
 inferred in a previous study \cite{Akaishi2023}.

### 2025-09-26 latex/main.tex
Updates from Overleaf

--- 2025-09-26 latex/main.tex
@@
 
-[TODO narrative is muddled here: the point of the para is to say that 
-pango X lineage designation is manual and subjective.]
-% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
-% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
-% ‘Pangolin’ nomenclature system. "
-Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
-can be attributed to the clarity of the rules under which new lineages 
-are designated and their explicitly phylogenetic basis. 
-The first recombinant lineage XA was designated in 2021 based on the 
-analysis of Jackson et al\.~\cite{Jackson2021}, which established 
-that recombination with onward transmission had occured [weak FIXME].
-As of pango-designation version v1.33.1 (2025-03-31), a total of 146 distinct 
-X lineages have been designated.
-%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
-% 146
-Although the process of incorporating recombinant lineages in the Pango 
-nomenclature system is well defined,
-the identification of such events is much more subjective, requiring
-the synthesis of multiple lines of evidence and often the 
-manual inspection of phylogenies and sequence alignments
-by volunteers.
-
-By automatically detecting and incorporating recombination events
-into the ARG, sc2ts allows us to model the origins of these 
-X lineages as nodes in the graph. 
-Table~\ref{tab:pango_x_lineages} summarises
-these results for the XX Pango X lineages present in
-the ARG. 
-For 16 Pango X lineages the oldest node that has been assigned
-the focal Pango lineage is a recombination node, and that node 
-is the ancestor of all samples assigned that (or a descendant) Pango 
-lineage. These are the Type I events, in which the ARG reflects the
-expected phylogenetic structure of Pango designation.
-Among these events, concordance with the community-consensus 
-parent lineages and breakpoint intervals is excellent (see next 
-section). 13 of the Type I events are monophyletic for the focal
-pango; Figure~\ref{fig:subgraphs}A shows XA as an example.
-The remaining 3 Type I events (XBB, XBQ and XM) have descendants not assigned to the 
-focal pango, many of which are the generic BA.2 label
-[FIXME presumably due to uncertainty? Look up the results CSV? Be precise].
-
-The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
-are recombination events closely associated with Pango X lineages, 
-but that do not meet the strict origination criteria for Type I.
-Type III events are then nodes in the ARG associated with the 
-orin
-
-Of these, XAF and XJ are relatively straightforwardo
-
-Furthermore, some of the Pango X lineages above originate from
-shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
-An example of a set of ``nested'' closely related lineages
-consists of XZ, XAC, XAD, XAE, and XAP,
-which descend from a single recombination event in the ARG, rather than five independent
-ones (Figure~\ref{fig:subgraphs}B), and
-which we tentatively group using the unused label Xx.
-These Pango X lineages were designated around the same time.
-These results highlight that the ARG can help
-clarify the relationships between recombinants,
-allowing more precise interpretation of evidence for recombination.
-
-Sc2ts identified 12 Pango X lineages in the ARG as non-recombinants
-(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
-For all these Pango X lineages, the number of recombination-informative sites
-fell below our threshold of detection (i.e. $k = 4$ in the HMM).
-The fact that most of these Pango X lineages (except XAJ and XAN)
-differ from a single parent node in the ARG
-by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
-their origins as non-recombinants.
-For example, the closely related XN and XAU can be explained by two mutations
-and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C), and we
-believe the original recombinant designation to be based on erroneous data
-(Supplementary document S1.XXX).
-%%% TODO: Say something about XAJ and XAN.
-%JK: we should weaken this statement a bit, as we're not certain that they 
-% really are recombinants, especially XP.
-% YW: we now have a better in-depth breakdown of these recombinants
-% to be pasted into doc S1 - e.g. XB is missing the parent lineages, XP is defined on the basis
-% of 2 loci, and XAJ on the basis of 3.
-However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
-because the HMM treated characteristic deletions as missing data.
-
-[NOTE: putting this here for now, until we find the right place to slot it
-in next draft.]
-See Document~\ref{sec:supplement_indepth_pango_x_lineages} for extended analyses,
-Figure~\ref{fig:pango_x_copying_patterns} for copying patterns,
-and Document S2 for ARG visualations of all Pango X lineages.
-
 \begin{figure}[h]
@@
 \end{table}
+
+[TODO narrative is muddled here: the point of the para is to say that 
+pango X lineage designation is manual and subjective.]
+% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
+% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
+% ‘Pangolin’ nomenclature system. "
+Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
+can be attributed to the clarity of the rules under which new lineages 
+are designated and their explicitly phylogenetic basis. 
+The first recombinant lineage XA was designated in 2021 based on the 
+analysis of Jackson et al\.~\cite{Jackson2021}, which established 
+that recombination with onward transmission had occured [weak FIXME].
+As of pango-designation version v1.33.1 (2025-03-31), a total of 146 distinct 
+X lineages have been designated.
+%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
+% 146
+Although the process of incorporating recombinant lineages in the Pango 
+nomenclature system is well defined,
+the identification of such events is much more subjective, requiring
+the synthesis of multiple lines of evidence and often the 
+manual inspection of phylogenies and sequence alignments
+by volunteers.
+
+By automatically detecting and incorporating recombination events
+into the ARG, sc2ts allows us to model the origins of these 
+X lineages as nodes in the graph. 
+Table~\ref{tab:pango_x_lineages} summarises
+these results for the XX Pango X lineages present in
+the ARG. 
+For 16 Pango X lineages the oldest node that has been assigned
+the focal Pango lineage is a recombination node, and that node 
+is the ancestor of all samples assigned that (or a descendant) Pango 
+lineage. These are the Type I events, in which the ARG reflects the
+expected phylogenetic structure of Pango designation.
+Among these events, concordance with the community-consensus 
+parent lineages and breakpoint intervals is excellent (see next 
+section). 13 of the Type I events are monophyletic for the focal
+pango; Figure~\ref{fig:subgraphs}A shows XA as an example.
+The remaining 3 Type I events (XBB, XBQ and XM) have descendants not assigned to the 
+focal pango, many of which are the generic BA.2 label
+[FIXME presumably due to uncertainty? Look up the results CSV? Be precise].
+
+The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
+are recombination events closely associated with Pango X lineages, 
+but that do not meet the strict origination criteria for Type I.
+Type III events are then nodes in the ARG associated with the 
+orin
+
+Of these, XAF and XJ are relatively straightforwardo
+
+Furthermore, some of the Pango X lineages above originate from
+shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
+An example of a set of ``nested'' closely related lineages
+consists of XZ, XAC, XAD, XAE, and XAP,
+which descend from a single recombination event in the ARG, rather than five independent
+ones (Figure~\ref{fig:subgraphs}B), and
+which we tentatively group using the unused label Xx.
+These Pango X lineages were designated around the same time.
+These results highlight that the ARG can help
+clarify the relationships between recombinants,
+allowing more precise interpretation of evidence for recombination.
+
+Sc2ts identified 12 Pango X lineages in the ARG as non-recombinants
+(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
+For all these Pango X lineages, the number of recombination-informative sites
+fell below our threshold of detection (i.e. $k = 4$ in the HMM).
+The fact that most of these Pango X lineages (except XAJ and XAN)
+differ from a single parent node in the ARG
+by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
+their origins as non-recombinants.
+For example, the closely related XN and XAU can be explained by two mutations
+and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C), and we
+believe the original recombinant designation to be based on erroneous data
+(Supplementary document S1.XXX).
+%%% TODO: Say something about XAJ and XAN.
+%JK: we should weaken this statement a bit, as we're not certain that they 
+% really are recombinants, especially XP.
+% YW: we now have a better in-depth breakdown of these recombinants
+% to be pasted into doc S1 - e.g. XB is missing the parent lineages, XP is defined on the basis
+% of 2 loci, and XAJ on the basis of 3.
+However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
+because the HMM treated characteristic deletions as missing data.
+
+[NOTE: putting this here for now, until we find the right place to slot it
+in next draft.]
+See Document~\ref{sec:supplement_indepth_pango_x_lineages} for extended analyses,
+Figure~\ref{fig:pango_x_copying_patterns} for copying patterns,
+and Document S2 for ARG visualations of all Pango X lineages.
 
@@
 Table~\ref{tab:major_dels}).
-All these deletions are known to occur major VOCs
+All these deletions are known to occur in major VOCs
 (https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;

### 2025-09-26 latex/main.tex
Pass at the X events section

@@
 shows the descendant samples broken down by Pango lineage (grouped 
-by root lineage, where appropriate).
+by root lineage, where appropriate).[FIXME Typesetting to be finished,
+describe cols for Type III and IV]
 }
@@
 Type I and II events}\\
-pango & name & path len & time & mutations & samples \\
+pango & name & path len & time & mutations & & descendants\\
 \midrule
@@
 \multicolumn{7}{l}{Type IV events: non recombinants}\\
-pango & muts & parent & samples & nested \\
+pango & muts &&&& parent & descendants \\
 \midrule
 
+XAS & 1  &&&& BA.4 & {XAS:77} \\
+XAU & 1  &&&& BA.2 & {XAU:8} \\
+XBK & 1  &&&& CJ.1 & {XBK:7} \\
+XBQ & 1  &&&& CJ.1 & {XBQ:14} \\
+XN  & 1  &&&& \textbf{BA.2} & {XN:120, BA.2:1} \\
 XP  & 1  &&&& \textbf{BA.1.1} & {XP:45} \\
-XN  & 1  &&&& \textbf{BA.2} & {XN:120, BA.2:1} \\
-XBQ & 1  &&&& CJ.1 & {XBQ:14} \\
-XBK & 1  &&&& CJ.1 & {XBK:7} \\
+XB  & 2  &&&& B.1 & {XB:192} \\
+XAV & 3  &&&& BA.5.1.24 & {XAV:13} \\
+XAZ & 3  &&&& \textbf{BA.5} & {XAZ:133} \\
 XBE & 3  &&&& BA.5.2 & {XBE:65} \\
-XB  & 2  &&&& B.1 & {XB:192} \\
-XAZ & 3  &&&& \textbf{BA.5} & {XAZ:133} \\
-XAV & 3  &&&& BA.5.1.24 & {XAV:13} \\
-XAU & 1  &&&& BA.2 & {XAU:8} \\
-XAS & 1  &&&& BA.4 & {XAS:77} \\
 XAN & 5  &&&& \textbf{BA.5.1} & {XAN:7} \\
@@
 By automatically detecting and incorporating recombination events
-into the ARG, sc2ts allows us to model the origins of these 
-X lineages as nodes in the graph. 
+into the ARG, sc2ts provides us with a systematic means of 
+discovering and assessing evidence around these events.
 Table~\ref{tab:pango_x_lineages} summarises
-these results for the XX Pango X lineages present in
-the ARG. 
+these results for the 44 Pango X lineages present in
+the ARG (XAC and XAD have been omitted, and XM 
+partially represented, corresponding to 23 samples in total; see STAR Methods).
 For 16 Pango X lineages the oldest node that has been assigned
@@
 section). 13 of the Type I events are monophyletic for the focal
-pango; Figure~\ref{fig:subgraphs}A shows XA as an example.
-The remaining 3 Type I events (XBB, XBQ and XM) have descendants not assigned to the 
-focal pango, many of which are the generic BA.2 label
-[FIXME presumably due to uncertainty? Look up the results CSV? Be precise].
+pango (i.e., the descendants are entirely from that lineage); 
+Figure~\ref{fig:subgraphs}A shows XA as an example (see also 
+Document~\ref{sec:jackson_recombs} for further analysis 
+of the Jackson et al.\ recombinants).
 
@@
 but that do not meet the strict origination criteria for Type I.
-Type III events are then nodes in the ARG associated with the 
-orin
-
-Of these, XAF and XJ are relatively straightforwardo
-
-Furthermore, some of the Pango X lineages above originate from
-shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
-An example of a set of ``nested'' closely related lineages
-consists of XZ, XAC, XAD, XAE, and XAP,
-which descend from a single recombination event in the ARG, rather than five independent
-ones (Figure~\ref{fig:subgraphs}B), and
-which we tentatively group using the unused label Xx.
-These Pango X lineages were designated around the same time.
-These results highlight that the ARG can help
-clarify the relationships between recombinants,
-allowing more precise interpretation of evidence for recombination.
-
-Sc2ts identified 12 Pango X lineages in the ARG as non-recombinants
-(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
-For all these Pango X lineages, the number of recombination-informative sites
-fell below our threshold of detection (i.e. $k = 4$ in the HMM).
-The fact that most of these Pango X lineages (except XAJ and XAN)
-differ from a single parent node in the ARG
-by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
-their origins as non-recombinants.
-For example, the closely related XN and XAU can be explained by two mutations
-and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C), and we
-believe the original recombinant designation to be based on erroneous data
-(Supplementary document S1.XXX).
-%%% TODO: Say something about XAJ and XAN.
-%JK: we should weaken this statement a bit, as we're not certain that they 
-% really are recombinants, especially XP.
-% YW: we now have a better in-depth breakdown of these recombinants
-% to be pasted into doc S1 - e.g. XB is missing the parent lineages, XP is defined on the basis
-% of 2 loci, and XAJ on the basis of 3.
-However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
-because the HMM treated characteristic deletions as missing data.
-
-[NOTE: putting this here for now, until we find the right place to slot it
-in next draft.]
-See Document~\ref{sec:supplement_indepth_pango_x_lineages} for extended analyses,
-Figure~\ref{fig:pango_x_copying_patterns} for copying patterns,
-and Document S2 for ARG visualations of all Pango X lineages.
+Type III events are Pango origin nodes in the ARG that are 
+derived from Type I or Type II recombination events. There is significant
+nesting among the Pango X lineage origin nodes within the ARG,
+with two major clusters of particular note. Although XQ is 
+directly associated with a recombination event in the ARG, 
+XAA, XAG, XAM, XR and XU are all closely derived, suggesting 
+that they are more parsimoniously explained as sublineages of XQ
+than de-novo recombinant lineages.
+The second cluster is based around a putative undocumented
+recombinant that is the ancestor of lineages 
+XZ, XAC, XAD, XAE, and XAP, which we refer to as ``Xx''
+(Figure~\ref{fig:subgraphs}B).
+
+The final class of event (Type IV) in Table~\ref{tab:pango_x_lineages} 
+correspond to nodes that are not associated with a recombination
+event. These vary in their plausibility of being non-recombinant.
+For example, XN (Figure~\ref{fig:subgraphs}C) is only one mutation
+distant from a BA.2 sample and has little additional evidence to 
+suggest recombination [ref sup section]. XB, on the other hand,
+has been thoroughly characterised\cite{Gutierrez2022} and 
+may have misclassified by sc2ts due to limitations in handling
+deletions or matching against long branches (see Discussion).
+See Document~\ref{sec:supplement_indepth_pango_x_lineages} for extended 
+analyses on all of the Pango X lineages above,
+Figure~\ref{fig:pango_x_copying_patterns} for copying patterns 
+associated with Table~\ref{tab:pango_x_lineages},
+and Document S2 for ARG visualisations.
+
+The nested samples not assigned to the focal pango lineage are
+either from different pango X lineages (see next paragraph), 
+or given a generic BA.2 assignment. Of these 59 generic assignents
+6 had conflicts in their UShER placements and could have been
+assigned an X lineage (XJ: 3, XU: 2, XN: 1), and 
+25 were cases in which Scorpio~\cite{Colquhoun2023scorpio}
+failed to find a specific lineage (accounting for 7\%
+of the 357 ``Omicron (Unassigned)`` Scorpio calls across
+2,747,985 alignments).
 
@@
 
-\subsection{Analysis of deletions}
+\subsection{Analysis of deletion events}
 \label{sec:supplement_dels}
@@
 are shown in Table~\ref{tab:rec_dels}.
-About half of these deletions are one-base,
-occur in non-coding regions, and
-are concentrated in the 3' end of the genome.
+About half of these deletions are one-base and
+occur in the non-coding regions in the 3' end of the genome
+between the accessory genes.
 Four one-base deletions occur
-in accessory genes (ORF7a, ORF8, and ORF10).
+in the accessory genes ORF7a, ORF8, and ORF10.
 The two deletions in S1-NTD are known recurrent deletions
@@
 Major deletion events in the ARG.
-These events are inherited by at least 10,000 samples in the ARG.
+These events are inherited by
+at least 10,000 samples in the ARG.
 Count is the number of samples inheriting a deletion.
@@
 S, Spike;
-S1-NTD, N-terminal domain in S1 subunit of Spike.
+S1-NTD, N-terminal domain in Spike S1 subunit.
 }
@@
 22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 \\
-11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
-28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 \\
+11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 (Omicron) & 871782 & 35.1220 \\
+28362 & 9 & N & 1436802 & B.1.1.529 (Omicron) & 871774 & 35.1216 \\
 21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 \\
@@
 S, Spike;
-S1-NTD, N-terminal domain in S1 subunit in Spike.
+S1-NTD, N-terminal domain in Spike S1 subunit.
 }
@@
 
+\subsection{Major saltation lineages}
+
+% https://github.com/jeromekelleher/sc2ts-paper/issues/772
+
 \subsection{Additional analyses of the Pango X lineages}
@@
 \subsubsection{XQ/XR/XU/XAA/XAG/XAM}
+\label{sec:supplement_indepth_pango_x_lineages_xq}
 https://github.com/jeromekelleher/sc2ts-paper/issues/338

### 2025-09-26 latex/main.tex
Update

@@
 (A) Subgraph around the recombination node associated with XA.
-Two "Group A" recombinant samples reported by Jackson et al. \cite{Jackson2021}) are shown.
-(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages: we use the label Xx as a shorthand for this related group of recombinants; and
+Two ``Group A'' recombinant samples reported by Jackson 
+et al.\cite{Jackson2021} are shown.
+(B) Subgraph around the recombination node associated with a 
+group of ``nested'' Pango X lineages: we use the 
+label Xx as a shorthand for this related group of recombinants; and
 (C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
@@
 The first recombinant lineage XA was designated in 2021 based on the 
-analysis of Jackson et al\.~\cite{Jackson2021}, which established 
+analysis of Jackson et al.\cite{Jackson2021}, which established 
 that recombination with onward transmission had occured [weak FIXME].
@@
 the ARG (XAC and XAD have been omitted, and XM 
-partially represented, corresponding to 23 samples in total; see STAR Methods).
+partially represented, corresponding to 23 samples in total; 
+see Document~\ref{sec:supplement_x_pango_lineages}).
 For 16 Pango X lineages the oldest node that has been assigned
@@
 and Document S2 for ARG visualisations.
-
-The nested samples not assigned to the focal pango lineage are
-either from different pango X lineages (see next paragraph), 
-or given a generic BA.2 assignment. Of these 59 generic assignents
-6 had conflicts in their UShER placements and could have been
-assigned an X lineage (XJ: 3, XU: 2, XN: 1), and 
-25 were cases in which Scorpio~\cite{Colquhoun2023scorpio}
-failed to find a specific lineage (accounting for 7\%
-of the 357 ``Omicron (Unassigned)`` Scorpio calls across
-2,747,985 alignments).
 
@@
 for sc2ts, and are present at a significant frequency. For example, of the 
-43 Scorpio designations present in the Viridian dataset studied, 32 have 
+43 Scorpio~\cite{Colquhoun2023scorpio}
+designations present in the Viridian dataset studied, 32 have 
 samples present with sampling dates in 2020, including lineages that do not
@@
 
-\subsection{Analysis of Pango lineage ``origination'' events}
+\subsection{Pango lineage ``origination'' events}
 \label{sec:supplement_pango_lineages}
@@
 for details.
+
+\subsection{Pango X lineage events}
+\label{sec:supplement_x_pango_lineages}
+[TODO DOCUMENT DETAILS of deriving events]
+
+The nested samples not assigned to the focal pango lineage are
+either from different pango X lineages (see next paragraph), 
+or given a generic BA.2 assignment. Of these 59 generic assignents
+6 had conflicts in their UShER placements and could have been
+assigned an X lineage (XJ: 3, XU: 2, XN: 1), and 
+25 were cases in which Scorpio~\cite{Colquhoun2023scorpio}
+failed to find a specific lineage (accounting for 7\%
+of the 357 ``Omicron (Unassigned)`` Scorpio calls across
+2,747,985 alignments).
 

### 2025-09-26 latex/main.tex
Updates from Overleaf

--- 2025-09-26 latex/main.tex
@@
 for details.
+
+\subsection{Major saltation lineages}
+
+% https://github.com/jeromekelleher/sc2ts-paper/issues/772
 

### 2025-09-28 latex/main.tex

--- 2025-09-26 latex/main.tex
@@
 
-We also found highly recurrent deletions using the ARG.
-The top 20 deletions in the number of occurrences in the ARG
-are shown in Table~\ref{tab:rec_dels}.
+We found highly recurrent deletions using the ARG,
+for example, the top 20 deletions in terms of
+the number of occurrences in the ARG
+shown in Table~\ref{tab:rec_dels}.
 About half of these deletions are one-base and
-occur in the non-coding regions in the 3' end of the genome
-between the accessory genes.
-Four one-base deletions occur
-in the accessory genes ORF7a, ORF8, and ORF10.
-The two deletions in S1-NTD are known recurrent deletions
+occur in the 5' or 3' UTR of the SARS-CoV-2 genome,
+which are enriched for poor-quality sequencing data.
+Genuine recurrent deletions are also captured in the ARG,
+for example, two known recurrent deletions in S1-NTD
 associated with antibody escape \cite{McCarthy2021}.
+In-depth analysis of the deletions in the ARG
+is beyond the scope of this study,
+but we believe that the ARG will greatly
+facilitate future evolutionary studies of
+deletions in SARS-CoV-2.
 

### 2025-09-28 latex/main.tex
Rejig intro section on the ARG

@@
 recombinant lineages.
-Together, the inference method, inferred ARG and 
+Together, the ARG, inference method and 
 suite of open-source analysis tools, 
@@
 were conclusively demonstrated, as the resource
-quickly becoming an indispensible element 
+quickly became an indispensible element 
 of the pandemic response\cite{Hinrichs2024}.
 The second challenge posed is that recombination is 
-an essential element of SARS-CoV-2 
+an important force in SARS-CoV-2 
 evolution\cite{Jackson2021,VanInsberghe2021,Focosi2022,
@@
 Note that this is a form of parsimony, with the recombination
-penalty $k$ determining the relative parsimoniousness of recombinant
+penalty $k$ defining the relative parsimoniousness of recombinant
 vs non-recombinant solutions to the HMM.
@@
 After postprocessing, the ARG 
-contains 2,689,054 nodes with estimated dates, 
+contains 2,689,054 nodes,  % Note the estimated dates 
 2,748,838 edges, and 2,285,344 mutations at 29,893 sites. 
@@
 the ARG is stored in a 32MiB tszip format file.
+% It's confusing to talk about recombinant QC here as we're talking 
+% about the full ARG for the rest of the section.
+[Do we need to talk about recombination?]
 Of the nodes, 855 represent recombination events, 
 resulting in 316 distinct trees along the genome (some
-events occured at the same breakpoints).
-% It's confusing to talk about recombinant QC here as we're talking 
-% about the full ARG for the rest of the section.
-
-The ARG is a comprehensive record of the evolutionary history of SARS-CoV-2 and 
-is in close agreement with a state-of-the-art UShER phylogeny also built 
-from the Viridian dataset \cite{Hunt2024}.
-First, we validated the broad structure of the ARG against 
-the UShER phylogeny \cite{Hunt2024}
-by pruning both down to representative samples
-for 1097 deeply-sampled lineages (STAR Methods).
-Tanglegrams show that the ARG closely matches the phylogenetic
-structure of the UShER tree (Figures~\ref{fig:tanglegram_base},
-\ref{fig:tanglegram_delta},
-\ref{fig:tanglegram_ba2},
-\ref{fig:tanglegram_ba5}).
-Next, we compared the sc2ts and UShER inferences 
-from the Viridian dataset
-by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
+events occured at the same breakpoints [vague, rephrase]).
+[TODO note about loading up the time it takes to get the node 
+data dataframes in a notebook?]
+
+% PARA: we compared against the UShER tree in a bunch of ways
+% NOTE: Trying to clarify that this of the pandemic period and we're 
+% not trying to cover *all* history.
+The ARG is a comprehensive record of the pandemic phase 
+of SARS-CoV-2 evolutionary history, and accurately reflects
+a range of results from the literature.
+First, we compared the sc2ts ARG and with the state-of-the-art UShER 
+tree inferred from the same dataset
+by pruning both to their intersection of 2,475,418 samples and 27,507 sites
+(STAR Methods)
 % JK: this feels a bit wishy washy, but there really is no simple answer to
@@
 We also compared the mutational spectra of major VOCs in the ARG
-against the mutational spectra from
-the Viridian UShER phylogeny and
+against the UShER tree and 
 a previous study \cite{Bloom2023} (STAR Methods),
 finding close agreement (Figure~\ref{fig:mutational_spectra}).
+We validated that the broad phylogenetic structure of the ARG and 
+UShER tree closely match by generating tanglegrams for a set of representative
+samples from 1097 deeply-sampled lineages (STAR Methods; 
+Figures~\ref{fig:tanglegram_base},
+\ref{fig:tanglegram_delta},
+\ref{fig:tanglegram_ba2},
+\ref{fig:tanglegram_ba5}).
 
@@
 these bases are imputed to carry the values inherited by the node in question.
+
 Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
@@
 Each node in the ARG was then assigned
+
 to one of 2058 Pango lineages using pangolin (STAR Methods).

### 2025-09-28 latex/main.tex
Pass through dels section

@@
 
-% FIXME this is too detailed, and too methodsy. Abstract out and summarise
-% results
-Input genome alignments are masked in sc2ts
-so that deletions and ambiguous bases are treated as missing data by the HMM,
-and once the samples are incorporated into the ARG,
-these bases are imputed to carry the values inherited by the node in question.
-
-Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
-and to incorporate them into the ARG we identified a set of [XXX] sites
-most frequently involved in deletions [cite source?]
-and post-hoc mapped alignment data for these sites to the ARG using parsimony.
+Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
+but because they are difficult to disintinguish from bioinformatics errors 
+and their multi-base nature presents challenges to the Markovian structure of 
+the LS HMM, we masked out all gap characters as missing data during primary
+inference. To incorporate the most important deletions into the ARG, we 
+remapped data using posthoc parsimony at 75 sites involved in deletions 
+with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
+during postprocessing (STAR Methods).
 Although sites are treated independently in this approach,
-the approximation successfully identified the most frequent deletions,
-such as the recurrent 21991-21993 (See Document~\ref{sec:supplement_dels}
-for additional analysis).
-The node representing the origin of Alpha, with 290,407 descendants,
-is correctly associated with key lineage-defining deletions:
-11288-11296 (ORF1A del3675-3677),
-21765-21770 (Spike del69-70),
-21991-21993 (Spike del144), and
-28271 (non-coding).
-The Spike 69-70 deletion,
+runs of mutations to the gap character at adjacent sites
+correctly identified many known 
+high frequency (Table~\ref{tab:major_dels}) and 
+recurrent (Table~\ref{tab:rec_dels}) deletions.
+For example, the 4 characteristic deletions associated with 
+% FIXME: this is just my notation - is this something recognised more 
+% widely? If not, fix to something that is?
+Alpha (11288+9, 21765+6, 21991+3, 28271+1) are correctly
+identified as having a single origin [CITATION?].
+The 21765+6 deletion (Spike 69-70)
 which causes S-Gene Target Failure \cite{Walker2021},
-also emerges twice in the origin of Omicron -
-in the node leading to the BA.1 variant (342,099 descendants),
-and the node leading to the BA.4 variant (189,746 descendants).
-To speed up inference and to avoid issues
+emerges two more times at the origins of the 
+BA.1 and BA.4 variants.
+Although most highly recurrent deletions are likely errors,
+some known truly recurrent deletions are present,
+for example, two deletions in S1-NTD [WHICH TWO???]
+associated with antibody escape \cite{McCarthy2021}
+See Document\ref{sec:supplement_dels} for further analysis.
+
 caused by sites enriched for errors [citation?],
@@
 \label{sec:supplement_dels}
+[TODO This is too terse and too vague. Be concrete and precise. WHICH 
+deletions are we talking about?]
 
@@
 a known hotspot for recurrent deletions \cite{McCarthy2021}.
+
+[This is a separate paramgraph discussing where the model has 
+broken down a bit]
 There are three contiguous deletions in NSP6
@@
 (which was mapped as the 4-base deletion event).
-Furthermore, the placement of these deletion events in the ARG
+
+Furthermore, the placement of these [Which events? All of them? just 
+the ones we didn't properly place?] deletion events in the ARG
 reflects that of the VOC-associated deletion events

### 2025-09-28 latex/main.tex
Pass through the section on high-level results and dels

@@
 After postprocessing, the ARG 
-contains 2,689,054 nodes,  % Note the estimated dates 
+contains 2,689,054 nodes,  % Note the estimated dates ?
 2,748,838 edges, and 2,285,344 mutations at 29,893 sites. 
@@
 % about the full ARG for the rest of the section.
-[Do we need to talk about recombination?]
 Of the nodes, 855 represent recombination events, 
@@
 \ref{fig:tanglegram_ba5}).
-
+We also evaluated how well the ARG reflects the structure 
+of the Pango nomenclature system~\cite{Rambaut2020} by computing 
+Pango assignments for all 2.7 million nodes in the ARG using 
+Pangolin~\cite{OToole2021assignment} (STAR Methods).
+All but 21,680 of the exported sample alignments (0.87\%) agreed
+with the source consensus sequences in terms of Pango lineage assignment
+(comparable to the 0.27\% disagreement between pangolin-data
+versions 1.21 and 1.29 in the source Viridian metadata).
+A majority of these lineages (1473 of 2058)
+have monophyletic origins within the ARG,
+increasing to 1779 (86\%) when we allow
+for multiple sibling origination nodes
+(Document~\ref{sec:supplement_pango_lineages}).
+
+[TODO we should mention UShER here - what do they do with dels?]
 Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
@@
 the LS HMM, we masked out all gap characters as missing data during primary
-inference. To incorporate the most important deletions into the ARG, we 
+inference. 
+To incorporate the most important deletions into the ARG, we 
 remapped data using posthoc parsimony at 75 sites involved in deletions 
@@
 Alpha (11288+9, 21765+6, 21991+3, 28271+1) are correctly
-identified as having a single origin [CITATION?].
+identified as having a single origin [CITATION?]
+(Figure~\ref{fig:subgraph_alpha}).
 The 21765+6 deletion (Spike 69-70)
@@
 for example, two deletions in S1-NTD [WHICH TWO???]
-associated with antibody escape \cite{McCarthy2021}
+associated with antibody escape\cite{McCarthy2021}.
 See Document\ref{sec:supplement_dels} for further analysis.
-
-caused by sites enriched for errors [citation?],
-sc2ts masks out a set of 100 sites identified
-as the most highly homoplasic
-by running on a subset of the data (STAR Methods).
-Of these sites, 9 are in the set of [1XX] ``problematic sites''
-masked by UShER \cite{Hunt2024}.
-In addition to remapping the data for sites identified
-as being involved in deletions,
-we added mutations for XX of these highly recurrent sites
-that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
-Each node in the ARG was then assigned
-
-to one of 2058 Pango lineages using pangolin (STAR Methods).
-All but 21,680 sampled sequences (0.87\%) agreed
-with the source sequences in terms of Pango lineage assignment
-(comparable to the 0.27\% disagreement between pangolin-data
-versions 1.21 and 1.29 in the source Viridian metadata).
-A majority of these lineages (1473 of 2058)
-have monophyletic origins within the ARG,
-increasing to 1779 (86\%) when we allow
-for multiple sibling origination nodes
-(Document~\ref{sec:supplement_pango_lineages}).
 

### 2025-09-29 latex/main.tex

@@
 using post-hoc parsimony.
-We identified important deletion events
+We identified major deletion events
 (which are inherited by at least 10,000 samples;
@@
 
-[This is a separate paramgraph discussing where the model has 
-broken down a bit]
-There are three contiguous deletions in NSP6
-(4-base at 11288, 5-base at 11283, and 5-base at 11292),
+The placement of the following major deletion events
+in the ARG matches the phylogenetic placement of
+VOC-associated deletion events
+suggested in a previous study \cite{Akaishi2023}:
+a 3-base deletion 21,991 above the Alpha origin node;
+two 6-base deletions at 22,029 and 28,248 above the Delta origin node;
+a 9-base deletion at 28,362 above Omicron (B.1.1.529) origin node;
+a 3-base deletion at 6,513,
+a 9-base deletion at 21,987, and
+a 3-base deletion at 22,194 above the BA.1 origin node;
+a 9-base deletion at 21,633 above the BA.2 origin node.
+
+However, the placement of several major deletion events
+in the ARG differs from the phylogenetic placement of
+the corresponding deletion events suggested
+in Akaishi and Fujiwara \cite{Akaishi2023}.
+That study \cite{Akaishi2023} suggests that
+a 6-base deletion at 21,765 occurred
+before the divergence of Alpha and Omicron,
+but in the ARG this deletion appears to be recurrent,
+occurring above the origin nodes of Alpha, BA.1, and BA.4.
+Similarly, that study \cite{Akaishi2023} suggests that
+a 1-base deletion at 28,271 occurred
+before the divergence of Alpha, Delta, and Omicron,
+but it is mapped as a recurrent deletion in the ARG
+(at the Alpha origin node and above node 1436808).
+These differences suggest that
+some key deletions carry important phylogenetic signals
+ignored by the HMM used by sc2ts.
+
+% This is a separate paragraph discussing where the model has 
+% broken down a bit.
+Furthermore, there are three contiguous deletions in NSP6
+(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292),
 which should have been mapped as two 9-base deletions,
-one above the BA.1 origin node and
-the other above the BA.2 origin node in the ARG
-(Figure~\ref{fig:subgraph_ba1_ba2}),
-but they were broken up
-during deletion remapping because they had
-overlapping four sites at 11288-11292
+one above the BA.1 origin node (starting at 11,283) and
+the other above the BA.2 origin node (starting at 11,288)
+in the ARG (Figure~\ref{fig:subgraph_ba1_ba2}),
+but they were broken up during deletion remapping
+because they had four overlapping sites at 11,288-11,292
 (which was mapped as the 4-base deletion event).
-
-Furthermore, the placement of these [Which events? All of them? just 
-the ones we didn't properly place?] deletion events in the ARG
-reflects that of the VOC-associated deletion events
-inferred in a previous study \cite{Akaishi2023}.
-
-We found highly recurrent deletions using the ARG,
+Furthermore, the analysis of
+Akaishi and Fujiwara \cite{Akaishi2023} suggests that
+a 9-base deletion at 11,288
+occurred before the divergence of Alpha and Omicron.
+In the ARG, however, this deletion is mapped
+above the Alpha origin node.
+
+By mapping deletion events on the ARG,
+we also identified highly recurrent deletions,
 for example, the top 20 deletions in terms of
@@
 shown in Table~\ref{tab:rec_dels}.
-About half of these deletions are one-base and
-occur in the 5' or 3' UTR of the SARS-CoV-2 genome,
+About half of these deletions are 1-base and
+occur in the 5' or 3' UTR of the SARS-CoV-2 genome
+(from the start to 250 and from 29,700 to the end),
 which are enriched for poor-quality sequencing data.
-Genuine recurrent deletions are also captured in the ARG,
+Genuine recurrent deletions are captured in the ARG,
 for example, two known recurrent deletions in S1-NTD
 associated with antibody escape \cite{McCarthy2021}.
-In-depth analysis of the deletions in the ARG
+There are also 1-base recurrent deletions
+at 29,541, 29,543, and 29,555
+in a non-coding region (between N and ORF10)
+which are plausibly homoplasic as they might arise due to
+polymerase skipping during genome replication and
+this regions is likely free from strong selective pressure.
+An in-depth analysis of all the recurrent deletions in the ARG
 is beyond the scope of this study,
-but we believe that the ARG will greatly
-facilitate future evolutionary studies of
+but we believe that the ARG will serve as a rich resource
+to facilitate future evolutionary studies of
 deletions in SARS-CoV-2.
@@
 leading to a more parsimonious local topology
-with two mutations on the left branch.
-
-\subsubsection{XBQ/XBK/XBK.1/CJ.1.3}
+with two mutations on the left branch. This would also remove the two
+blue lines on the left and side of the copying pattern.
+
+\subsubsection{XBQ/XBK/XBK.1}
 https://github.com/jeromekelleher/sc2ts-paper/issues/349

### 2025-09-29 latex/main.tex
Updates from Overleaf

@@
 only XAF sample passed our QC filters and was added to the ARG:
-many have missing Ns in their consensus sequences (> 500) or
+many have greater than 500 missing Ns in their consensus sequences or
 require a large number of mutations to incorporate into the ARG

### 2025-09-29 latex/main.tex
Updates from Overleaf

@@
 
-[TODO we should mention UShER here - what do they do with dels?]
+[TODO we should mention UShER here: they state multiple times in the GitHub issues 
+that they don't handle them at all, so I think we can just say this in some nice
+way.]
 Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},

### 2025-09-29 latex/main.tex
Updates from Overleaf

--- 2025-09-29 latex/main.tex
@@
 
-[TODO we should mention UShER here - what do they do with dels?]
+[TODO we should mention UShER here: they state multiple times in the GitHub issues 
+that they don't handle them at all, so I think we can just say this in some nice
+way.]
 Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
-but because they are difficult to disintinguish from bioinformatics errors 
+but because they are difficult to distinguish from bioinformatics errors 
 and their multi-base nature presents challenges to the Markovian structure of 

### 2025-09-29 latex/main.tex
Move comparison with other methods to supplement
Closes #654

@@
 Among these events, concordance with the community-consensus 
-parent lineages and breakpoint intervals is excellent (see next 
-section). 13 of the Type I events are monophyletic for the focal
+parent lineages and breakpoint intervals is excellent,
+with similar performance to
+the RecombinHunt \cite{Alfonsi2024}
+and CovRecomb \cite{Li2024CovRecomb} detection methods
+(Table~\ref{tab:method_concordance}).
+Of the Type I events, 13 are monophyletic for the focal
 pango (i.e., the descendants are entirely from that lineage); 
@@
 and Document S2 for ARG visualisations.
-
-\subsection*{Comparison with other recombination detection methods}
-
-%%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
-We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
-RecombinHunt \cite{Alfonsi2024}, CovRecomb \cite{Li2024CovRecomb}, and rebar (https://github.com/phac-nml/rebar).
-These methods use the same underlying approach:
-they scan candidate recombinant sequences for mutations 
-associated with different Pango lineages
-without using a phylogeny.
-
-To evaluate the concordance of the methods in
-identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
-we focus on the Pango X lineages associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
-using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
-For RecombinHunt and CovRecomb,
-we summarized the detection results for these Pango X lineages (if available)
-reported in the original studies \cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
-For each method, we determined the Pango X lineages
-for which the inferred parental lineages and breakpoint intervals
-are concordant with those identified by the community (STAR Methods).
-We found that the methods performed well overall,
-with comparable concordance rates (Table~\ref{tab:method_concordance}),
-indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.
-
-% https://github.com/jeromekelleher/sc2ts-paper/issues/717
-167 (43\%) of QC-passing recombination events were also detected as recombinants by rebar.
-We hypothesize that this disagreement arose from
-differences in sensitivity between these methods.
-Consistent with this, we found that rebar needs a higher level of divergence between parents
-to correctly classify a sequence as recombinant, as shown by
-a shorter time to their MRCA (Figure \ref{fig:averted_mutations}B) and
-a lower distance between their Pango identity (Figure \ref{fig:averted_mutations}C).
-
-\begin{table}
-\caption{
-Concordance among methods in characterizing Pango X lineages.
-These comparisons focus on the Pango X lineages associated with
-the Type 1 recombination events (Table~\ref{tab:pango_x_lineages})
-For each method, we calculated the number of Pango X lineages
-which have inferred parent Pango lineages and breakpoint intervals
-concordant with those proposed by the community.
-The denominators indicate the number of Pango X lineages
-for which detection results are available from the original studies \cite{Alfonsi2024,Li2024CovRecomb}.
-For the Pango X lineages with non-overlapping breakpoint intervals,
-the distances between the inferred breakpoint intervals and 
-those proposed by the community are shown (median and range).
-}
-\label{tab:method_concordance}
-\begin{tabular}{lccc}
-\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
-\cmidrule(lr){2-2} \cmidrule(lr){3-4}
-Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
-\midrule
-Sc2ts & 14 / 16 (87.5) & 14 / 16 (87.5) & 540.5 (315, 766)\\
-RecombinHunt-GISAID & 13 / 15 (86.7) & 10 / 15 (66.7) & 764.0 (4, 1202)\\
-RecombinHunt-Nextstrain & 12 / 15 (80.0) & 11 / 15 (73.3) & 1160.0 (559, 2785)\\
-CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
-\bottomrule
-\end{tabular}
-\end{table}
 
@@
 \end{figure}
+
+% https://github.com/jeromekelleher/sc2ts-paper/issues/717
+167 (43\%) of QC-passing recombination events were also detected as recombinants by rebar.
+We hypothesize that this disagreement arose from
+differences in sensitivity between these methods.
+Consistent with this, we found that rebar needs a higher level of divergence between parents
+to correctly classify a sequence as recombinant, as shown by
+a shorter time to their MRCA (Figure \ref{fig:averted_mutations}B) and
+a lower distance between their Pango identity (Figure \ref{fig:averted_mutations}C).
 
@@
 \subsubsection*{Comparison with recombination detection methods}
+[TODO Merge in this text that was in the Results]
+
+To evaluate the concordance of the methods in
+identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
+we focus on the Pango X lineages associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
+using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
+For RecombinHunt and CovRecomb,
+we summarized the detection results for these Pango X lineages (if available)
+reported in the original studies \cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
+For each method, we determined the Pango X lineages
+for which the inferred parental lineages and breakpoint intervals
+are concordant with those identified by the community (STAR Methods).
+We found that the methods performed well overall,
+with comparable concordance rates (Table~\ref{tab:method_concordance}),
+indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.
 
@@
 
+\begin{table}
+\caption{
+Concordance among methods in characterizing Pango X lineages.
+These comparisons focus on the Pango X lineages associated with
+the Type 1 recombination events (Table~\ref{tab:pango_x_lineages})
+For each method, we calculated the number of Pango X lineages
+which have inferred parent Pango lineages and breakpoint intervals
+concordant with those proposed by the community.
+The denominators indicate the number of Pango X lineages
+for which detection results are available from 
+the original studies \cite{Alfonsi2024,Li2024CovRecomb}.
+For the Pango X lineages with non-overlapping breakpoint intervals,
+the distances between the inferred breakpoint intervals and 
+those proposed by the community are shown (median and range).
+}
+\label{tab:method_concordance}
+\centering
+\begin{tabular}{lccl}
+\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
+\cmidrule(lr){2-2} \cmidrule(lr){3-4}
+Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
+\midrule
+Sc2ts & 14 / 16 (87.5) & 14 / 16 (87.5) & 540.5 (315, 766)\\
+RecombinHunt-GISAID & 13 / 15 (86.7) & 10 / 15 (66.7) & 764.0 (4, 1202)\\
+RecombinHunt-Nextstrain & 12 / 15 (80.0) & 11 / 15 (73.3) & 1160.0 (559, 2785)\\
+CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \subsection{Parsimony comparison of sc2ts and UShER}

### 2025-09-29 latex/main.tex

@@
 and CovRecomb \cite{Li2024CovRecomb} detection methods
-(Table~\ref{tab:method_concordance}).
+(STAR Methods; Table~\ref{tab:method_concordance}).
 Of the Type I events, 13 are monophyletic for the focal
@@
 
+\subsection*{Quality control of recombination events}
+Sequencing errors in genome assemblies \cite{Turakhia2020},
+read misalignments \cite{Hunt2024}, and
+multi-nucleotide substitutions \cite{DeMaio2020issues} (MNSs)
+can cause short regions of a sample genome to match more closely
+to an alternative genome rather than the true parent.
+Such genomic regions can be mistaken for evidence of recombination,
+because the LS HMM counts each base separately.
+We developed a quality control (QC) measure 
+which counts the number of ``loci'' (clusters of closely spaced sites)
+supporting the left and right parents of a recombinant (STAR Methods),
+and effectively identifies a range of issues. 
+Figure~\ref{fig:recombinant_qc} plots the net number of supporting
+loci for the left and right parents for all 855 recombination
+events. The 501 recombinants with fewer than 4 net loci supporting 
+each parent are highly enriched for the AmpliSeq V1 sequencing
+protocol [TODO concrete numbers?], 
+and 71 (14\%) are associated with 6-base deletion in the 
+Delta lineage, known to be affected 
+by amplicon dropout\cite{Sanderson2021}.
+The 354 remaining recombination events that pass this QC filter
+contain all of the recombinants associated with Pango X lineages
+(Table~\ref{tab:pango_x_lineages}) and 
+the Jackson et al recombinants (Document\ref{sec:jackson_recombs}).
+
+To assist in the identification of such issues
+we developed a SnpIt-like~\cite{OToole2024} visualisation 
+that shows the informative genomic positions between 
+a recombinant and its parent sequences, and highlights 
+potentially problematic loci.
+Each copying pattern shows the positions
+where the allelic state in a recombinant
+(middle row) matches that of the left (upper row, coloured green) 
+or right parent (lower row, coloured blue),
+or where the recombination event requires a de-novo mutation
+(gold, with mutational change below).
+See Figures~\ref{fig:averted_mutations}D and
+\ref{fig:pango_x_copying_patterns} for copying patterns of the 
+Pango X lineages, and 
+Figure~\ref{fig:copying_pattern_examples} for copying patterns
+highlighting specific types of QC problem (and more details on the 
+visualisation).
+
 \subsection*{Quantification of evidence for recombination}
 
-%%% SHZ: We may not need this section at all, but I'm keeping it for now.
-%%% We should probably have a condensed version of the first paragraph in the Pango X section.
-%%% The second and third paragraphs should probably be removed entirely.
+The LS HMM at the heart of sc2ts provides a simple and powerful means 
+of quantifying the plausibility of a given recombination event.
+To do this, we force the HMM to find a Viterbi solution for the 
+putative recombinant sequence with no recombination by specifying
+a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
+The difference between the numbers of mutations between 
+the two solutions---the number of 
+mutations ``averted'' by recombination---allows us to quantify
+the plausibility.
+
 Sc2ts provides a method to facilitate the discovery of recombination events,
@@
 versus a model disallowing recombination,
-as measured by the number of resulting mutations (Figure~\ref{fig:methodology}).
+as measured by the number of resulting mutations 
 The more mutations are ``averted'' by allowing recombination,
@@
 \subsubsection*{Quality control of recombination events}
+[TODO rewrite this to account for new Results section]
+
 Sequencing errors in genome assemblies \cite{Turakhia2020},
@@
 indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.
+
+[TODO explain why we restrict to the Type I events]
 

### 2025-09-29 src/save_copying_patterns.py
New copying patterns
And copying_patterns.html file

@@ -32,7 +32,7 @@ def pangoX_RE_node_labels(exclude_dups=True):
         pango_x_events.closest_recombinant[use],
         'num_descendant_samples'
     ]
-    # Exclude non-recombinants, and ondes where closes RE node is BA.5
+    # Exclude non-recombinants, and nodes where closes RE node is BA.5
     pango_x = pango_x_events[np.logical_and(
         closest_recombinant_num_samples > 0,
         closest_recombinant_num_samples < 1e5
@@ -53,11 +53,7 @@ def pangoX_RE_node_labels(exclude_dups=True):
     else:
         return dict(px)
 
-def save_copying_pattern_image(node_id, label, ts, save_dir, zoom=None,**kwargs):
-    # NB - do not zoom if no ATCG letters are printed, as boxes will be of varying size
-    html_str = sc2ts.info.CopyingTable(ts, node_id).html(**kwargs)
-    html_str = html_str.replace('transform:', '-webkit-transform:')
-    html_str = html_str.replace('writing-mode:', '-webkit-writing-mode:')
+def save_copying_pattern_image(html_str, label, save_dir, zoom=None):
     options = {"format": "png", "quiet": "", "transparent": ""}  # use transparent so we can crop
     if zoom is not None:
         options['zoom'] = zoom 
@@ -69,33 +65,104 @@ def save_copying_pattern_image(node_id, label, ts, save_dir, zoom=None,**kwargs)
     img = img.crop(bbox)
     img.save(save_dir / f"{label}.png", "PNG", optimize=True, compress_level=9)
 
+def get_copying_table(ts, node_id, **kwargs):
+    html_str = sc2ts.info.CopyingTable(ts, node_id).html(**kwargs)
+    html_str = html_str.replace('transform:', '-webkit-transform:')
+    return html_str.replace('writing-mode:', '-webkit-writing-mode:')
+
+def save_copying_table_image(ts, node_id, label, save_dir, zoom=None,**kwargs):
+    # NB - do not zoom if no ATCG letters are printed, as boxes will be of varying size
+    html_str = get_copying_table(ts, node_id, **kwargs)
+    html_str = html_str.replace('<style>', '<style>.copying-table th {text-align: right; font-weight: normal;}')
+    save_copying_pattern_image(html_str, label, save_dir, zoom=zoom)
+
+rec = pd.read_csv(data_dir / "recombinants.csv").set_index('recombinant')
+
+####
+# The pango X versions
+####
[…]

### 2025-09-30 latex/main.tex
Updates from Overleaf

@@
 \label{sec:supplement_dels}
-[TODO This is too terse and too vague. Be concrete and precise. WHICH 
-deletions are we talking about?]
+% [TODO This is too terse and too vague.
+% Be concrete and precise.
+%WHICH deletions are we talking about?]
 
@@
 (at the Alpha origin node and above node 1436808).
-These differences suggest that
-some key deletions carry important phylogenetic signals
-ignored by the HMM used by sc2ts.
 
@@
 \caption{
-Tanglegram comparing phylogenetic backbones of sc2ts and UShER.
+Tanglegram comparing the basic phylogenetic backbones of sc2ts and UShER.
 }

### 2025-09-30 latex/main.tex

@@
 UShER tree closely match by generating tanglegrams for a set of representative
-samples from 1097 deeply-sampled lineages (STAR Methods; 
+samples from 814 deeply-sampled lineages (STAR Methods; 
 Figures~\ref{fig:tanglegram_base},
@@
 By identifying specific samples as parents,
-sc2ts enables more precise interpretation of recombinants than state-of-the-art recombination detection methods,
-which find recombinants by matching a given sample to pre-defined population-averaged mutational profiles of Pango lineages.
-Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution, in contrast to RIPPLES \cite{Turakhia2022}
+sc2ts enables more precise interpretation of recombinants
+than state-of-the-art recombination detection methods,
+which find recombinants by matching a given sample
+to pre-defined population-averaged mutational profiles of Pango lineages.
+Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution,
+in contrast to RIPPLES \cite{Turakhia2022}
 which detects recombinants post hoc by searching for unusual branches in a given phylogeny,
@@
 Combined with new methods and tools to evaluate evidence for recombination,
-sc2ts is a significant technological advance over current methods for the discovery and
-analysis of recombinants and efficient inference of SARS-CoV-2 genealogy.
+sc2ts is a significant technological advance over current methods
+for the discovery and analysis of recombinants and
+efficient inference of SARS-CoV-2 genealogy.
 
@@
 Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
-which are shown in different studies using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
-The ARG displays remarkable phylogenetic congruence with a state-of-the-art phylogeny built using UShER.
+which are shown in different studies
+using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
+The ARG displays remarkable phylogenetic congruence
+with a state-of-the-art phylogeny built using UShER.
 Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
@@
 So far, recent developments in ARG methods have focused almost 
-exclusively on applications to human sequencing data. In that space,
-the availability of ARGs reconstructed from large datasets has already
-had tremendous impact on improving our understanding of human genetic
-ancestry and past demography \cite{Wohns2022}. We expect
-similar leaps to be enabled through tailoring ARG reconstruction methods
-to data from other species and pathogens. Specifically, existing
-inference methods can be adapted to use the sc2ts ARG for the 
-genealogy-based study of selection \cite{hejase2022deep} or the
-geographic spread of the virus through time
-\cite{Grundler2025geographic}. This is in addition to simpler
-analyses, such as calculating mutation spectra or querying the number
-or frequency of mutations at a given position, which the ARG makes 
-significantly more efficient, and ultimately more easily reproducible.
+exclusively on applications to human sequencing data.
+In that space, the availability of ARGs reconstructed from large datasets
+has already had tremendous impact on improving our understanding of
+human genetic ancestry and past demography \cite{Wohns2022}.
+We expect similar leaps to be enabled
+through tailoring ARG reconstruction methods
+to data from other species and pathogens.
+Specifically, existing inference methods can be adapted
+to use the sc2ts ARG for the genealogy-based study of selection \cite{hejase2022deep}
+or the geographic spread of the virus through time \cite{Grundler2025geographic}.
+This is in addition to simpler analyses,
+such as calculating mutation spectra
+or querying the number or frequency of mutations at a given position,
+which the ARG makes significantly more efficient, and
+ultimately more easily reproducible.
 Since the ARG is a condensed representation of the genetic information
@@
 and checking and improving various aspects of its quality is laborious, 
-we hope that making this resource openly available will save
-significant time and 
-effort for other researchers and enable immediate focus on analysing 
+we hope that making this resource openly available will save significant time
+and effort for other researchers and enable immediate focus on analyzing
 the information contained within this data. In-depth analysis of the
@@
 
-We contribute several novel methods and concepts which may find applications beyond sc2ts.
+We contribute several novel methods and concepts
+which may find applications beyond sc2ts.
 First, we devise a simplified LS model that extends the classic concept of parsimony.
@@
 enabling reconstruction of the pandemic-scale ARG presented here.
-Existing HMMs used in statistical genetics may be modified in a similar way for significant speedup.
+Existing HMMs used in statistical genetics may be modified
+in a similar way for significant speedup.
 Second, we propose the concept of mutations averted by recombination,
 which leads to an interpretable quantity to weigh the support for 
-the hypothesis that a given sample is a recombinant against the hypothesis that it is a non-recombinant,
+the hypothesis that a given sample is a recombinant against
+the hypothesis that it is a non-recombinant,
 considering the sampled genetic diversity in an ARG and a specific HMM.
-Third, we propose the concept of HMM cost to identify poorly explained samples
-which are either too divergent or have a high number of mutations inconsistent with their collection dates.
-Here we use HMM cost to reject time traveling samples,
+Third, we propose the concept of HMM cost
+to identify poorly explained samples
+which are either too divergent or
+have a high number of mutations inconsistent with their collection dates.
+Here we use HMM cost to reject time-traveling samples,
 which no existing phylogenetic method detects and handles during inference
@@
 with wrong collection dates that may affect downstream analyses.
-Future work may be estimating dates
+Future work may be to estimate dates
 for the suspected time traveling samples using the ARG provided here and
@@
 
-We used Pango lineage designations to select representative samples
-for the majority of the 2026 Pango lineages in the ARG.
-We first identify an ``origination node'' for each Pango lineage,
-defined as the oldest node in the ARG with that Pango designation
-whose descendants in any tree include more than 50\% of the Pango samples of
-the focal type.
-The representative sample is then chosen as the oldest sample descending from the origination node
+We compare the non-recombinant parts of the ARG with the UShER topology
+by simplifying the phylogenies to a smaller number of
+Pango representative samples.
+We focus on the 1473 Pango lineages that have perfect ``origination events'' 
+as described in \ref{sec:supplement_pango_lineages}, and
+omit Pango designations starting with "X",
+to avoid recombinants, leaving 1329 Pango lineages.
+To ensure that each sample is a tip (rather than an internal node)
+in the simplified topologies,
+a representative sample is then chosen as the oldest sample
+descending from the origination node
 whose own descendants (if any) are entirely of the focal Pango type.
-We ``simplify''\cite{Kelleher2018efficient,Wong2024} 
-the ARG to include only the representative samples,
-resulting in an intermediate ARG of 1867 samples,
-then additionally remove internal samples,
-known recombinants (i.e.\ the Pango X lineages) and three ``singleton recombinant
-samples''
-in which a single Pango lineage appears immediately below a recombination node.
-This results in a simplified ARG of 1667 samples, each with a different Pango designation.
-
-We compare topology of the simplified ARG against the intersection UShER tree
-(which has been dated) simplified down to the same 1667 samples.
-
-The simplified ARG contains four single-breakpoint recombination nodes,
-corresponding to the origins of Delta, BA.2, BA.5, and a further BA.2 recombinant
-which we deem artifactual by our standard criteria
-and treat as non-recombinant for the purposes of plotting
-(the breakpoint is on the far right at position 27382,
-and supported by only 4 distinguishing sites / 3 loci).
+We then further remove 4 samples
+which are singleton descendants of a recombination event in the ARG
+(representing potentially missed Pango recombinant lineages),
+as well as 43 samples that are descendants of the recombinant XBB
+but which do not start with an "X".
+This results in a simplified ARG of 1282 samples,
+each sample having a different Pango designation.
+
+The simplified ARG contains two single-breakpoint recombination nodes. The first
+corresponds to a BA.2 recombinant whose breakpoint is on the far right (position 27382) supported by only 4 sites / 3 loci, anch which therefore does not pass our recombinant quality control measure.
+The second is the origin of BA.5, which (the ).
 To reduce plot density, we consider only those Pango lineages with fewer than 10 samples,
@@
 where a breakpoint is inferred to occur.
-To make these breakpoint intervals right-exclusive,
-we adjusted them by subtracting one from the left coordinate and adding two to the right coordinate.
+To make these breakpoint intervals right-exclusive, we adjusted them
+by subtracting one from the left coordinate and adding two to the right coordinate.
 
 \subsubsection*{Calculation of Pango lineage distance using pangonet}
-We define a ``pseudo-phylogenetic'' distance to quantify the divergence between two Pango lineages.
-The idea is to count the number of edges separating two given Pango lineages in a ``pseudo-phylogeny''
+We define a ``pseudo-phylogenetic'' distance
+to quantify the divergence between two Pango lineages.
+The idea is to count the number of edges
+separating two given Pango lineages in a pseudo-phylogeny
 which relates all the designated Pango lineages.
 To construct this pseudo-phylogeny, we used \textit{pangonet},
-which mines the information on the Pango lineages in the ``alias\_key.json''
-and ``lineage\_notes.txt'' files
-from the Pango designation issues (https://github.com/cov-lineages/pango-designation).
+which mines the information on the Pango lineages
+in the ``alias\_key.json'' and ``lineage\_notes.txt'' files
+from the Pango designation issues
+(\url{https://github.com/cov-lineages/pango-designation}).
 If two Pango lineages are siblings in the pseudo-phylogeny,
@@
 If we require $n$ mutations on both sides of the breakpoint,
-we can calculate the expected value of the interval to the left by summing across a Poisson-distributed number of mutations,
+we can calculate the expected value of the interval to the left
+by summing across a Poisson-distributed number of mutations,
 conditioned on having at least $n$ mutations:
@@
 Summing the two intervals ($I_{left}+I_{right}$) gives the expected interval length.
-As detectable recombination events are nearer the center of the chromosome, we use $P=½$ in Figure X.
+As detectable recombination events are nearer the center of the chromosome,
+we use $P=½$ in Figure X.
 
@@
 % See https://github.com/jeromekelleher/sc2ts-paper/issues/628
-We reconstructed a more comprehensive mutational landscape
-of SARS-CoV-2 by mapping deletions onto the ARG
-using post-hoc parsimony.
+We reconstructed
+a more comprehensive mutational landscape of SARS-CoV-2
+by mapping deletions onto the ARG using post-hoc parsimony.
 We identified major deletion events
@@
 All these deletions are known to occur in major VOCs 
-(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
-\cite{Akaishi2023,Li2023indels}).
-Eight of these deletions occur in S1-NTD,
-a known hotspot for recurrent deletions \cite{McCarthy2021}.
+(\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
+\url{https://github.com/cov-lineages/pango-designation/issues/361};
+\cite{Stern2021,Akaishi2023,Li2023indels},
+except the 4-base deletion and two 5-base deletions in ORF1ab,
+which are likely artefacts of deletion remapping (see below).
+Notably, eight of these deletions occur
+in the N-terminal domain of the S1 unit in Spike,
+a known hotspot for recurrent deletions \cite{McCarthy2021}
+(also see Table~\ref{tab:rec_dels}).
 
 The placement of the following major deletion events
-in the ARG matches the phylogenetic placement of
+in the ARG aligns well with the phylogenetic placement of
 VOC-associated deletion events
@@
 two 6-base deletions at 22,029 and 28,248 above the Delta origin node;
-a 9-base deletion at 28,362 above Omicron (B.1.1.529) origin node;
+a 9-base deletion at 28,362 above the B.1.1.529 origin node;
 a 3-base deletion at 6,513,
@@
 
-However, the placement of several major deletion events
-in the ARG differs from the phylogenetic placement of
-the corresponding deletion events suggested
-in Akaishi and Fujiwara \cite{Akaishi2023}.
-That study \cite{Akaishi2023} suggests that
-a 6-base deletion at 21,765 occurred
-before the divergence of Alpha and Omicron,
-but in the ARG this deletion appears to be recurrent,
-occurring above the origin nodes of Alpha, BA.1, and BA.4.
-Similarly, that study \cite{Akaishi2023} suggests that
-a 1-base deletion at 28,271 occurred
-before the divergence of Alpha, Delta, and Omicron,
-but it is mapped as a recurrent deletion in the ARG
-(at the Alpha origin node and above node 1436808).
-
-% This is a separate paragraph discussing where the model has 
-% broken down a bit.
-Furthermore, there are three contiguous deletions in NSP6
+However, the ARG suggests that two major deletions
+(a 6-base deletion at 21,765 and a 1-base deletion at 28,271;
+Table~\ref{tab:major_dels}) are recurrent
+(i.e., acquired multiple times),
+but Akaishi and Fujiwara \cite{Akaishi2023} suggest that
+these deletions might have been acquired
+before the divergence of certain VOCs,
+based on an analysis of a small set of 20 samples.
+That study suggests that the 6-base deletion
+occurred before the divergence of Alpha and Omicron,
+but in the ARG this deletion occurs three times,
+above the origin nodes of Alpha, BA.1, and BA.4.
+Similarly, that study suggests that the 1-base deletion
+occurred before the divergence of Alpha, Delta, and Omicron,
+but in the ARG this deletion occurs
+above the Alpha origin node and above node 1436808.
+
+There are three contiguous deletions in NSP6
 (5-base at 11,283, 4-base at 11,288, and 5-base at 11,292),
@@
 but they were broken up during deletion remapping
-because they had four overlapping sites at 11,288-11,292
+as they had four overlapping sites at 11,288-11,292
 (which was mapped as the 4-base deletion event).
-Furthermore, the analysis of
-Akaishi and Fujiwara \cite{Akaishi2023} suggests that
-a 9-base deletion at 11,288
-occurred before the divergence of Alpha and Omicron.
-In the ARG, however, this deletion is mapped
+Furthermore, Akaishi and Fujiwara \cite{Akaishi2023}
+suggest that a 9-base deletion at 11,288 occurred
+before the divergence of Alpha and Omicron.
+In the ARG, however, this deletion occurs
 above the Alpha origin node.
 
-By mapping deletion events on the ARG,
-we also identified highly recurrent deletions,
+We identified highly recurrent deletions,
 for example, the top 20 deletions in terms of
-the number of occurrences in the ARG
-shown in Table~\ref{tab:rec_dels}.
-About half of these deletions are 1-base and
-occur in the 5' or 3' UTR of the SARS-CoV-2 genome
-(from the start to 250 and from 29,700 to the end),
-which are enriched for poor-quality sequencing data.
-Genuine recurrent deletions are captured in the ARG,
-for example, two known recurrent deletions in S1-NTD
-associated with antibody escape \cite{McCarthy2021}.
-There are also 1-base recurrent deletions
-at 29,541, 29,543, and 29,555
-in a non-coding region (between N and ORF10)
-which are plausibly homoplasic as they might arise due to
-polymerase skipping during genome replication and
-this regions is likely free from strong selective pressure.
-An in-depth analysis of all the recurrent deletions in the ARG
-is beyond the scope of this study,
-but we believe that the ARG will serve as a rich resource
-to facilitate future evolutionary studies of
-deletions in SARS-CoV-2.
+the number of occurrences in the ARG shown in Table~\ref{tab:rec_dels}.
+These include known recurrent deletions, for example,
+6-base deletion at 21,765,
+3-base deletion at 21,991, and
+3-base deletion 22,194 in S1-NTD
+associated with antibody escape \cite{McCarthy2021,Meng2021}.
+The out-of-frame deletions in S1-NTD
+(e.g., the 2-base deletion at 21,765),
+N, ORF1ab, and ORF8
+are implausible and are likely artefacts caused by
+sequencing errors or sequence misalignments.
+Future improvements in
+data pre-processing and deletion remapping
+should remove such deletions.
 
@@
 \midrule
-29781 & 1 & non-coding & 12754 \\
-29762 & 1 & non-coding & 11383 \\
-29779 & 1 & non-coding & 10951 \\
-29769 & 1 & non-coding & 10809 \\
-29774 & 1 & non-coding & 10735 \\
-29700 & 1 & non-coding & 8562 \\
 686 & 9 & ORF1ab / NSP1 & 6436 \\
-29614 & 1 & ORF10 & 5039 \\
-29555 & 1 & non-coding & 4814 \\
-29543 & 1 & non-coding & 4804 \\
-29541 & 1 & non-coding & 4802 \\
-203 & 1 & non-coding & 3604 \\
-222 & 1 & non-coding & 3442 \\
 21991 & 3 & S / S1-NTD & 2088 \\
-28254 & 1 & ORF8 & 1969 \\
-28093 & 1 & ORF8 & 900 \\
 21765 & 6 & S / S1-NTD & 482 \\
-27532 & 1 & ORF7a & 404 \\
-335 & 1 & ORF1ab / NSP1 & 370 \\
-337 & 1 & ORF1ab / NSP1 & 346 \\
+22029 & 6 & S / S1-NTD & 275 \\
+22194 & 3 & S / S1-NTD & 200 \\
+6513 & 3 & ORF1ab / NSP3 & 181 \\
+21765 & 2 & S / S1-NTD & 140 \\
+21995 & 1 & S / S1-NTD & 90 \\
+21988 & 3 & S / S1-NTD & 78 \\
+21987 & 7 & S / S1-NTD & 78 \\
+21993 & 3 & S / S1-NTD & 77 \\
+21987 & 9 & S / S1-NTD & 70 \\
+21994 & 2 & S / S1-NTD & 65 \\
+28370 & 1 & N & 61 \\
+11292 & 3 & ORF1ab / NSP6 & 48 \\
+21990 & 1 & S / S1-NTD & 46 \\
+21987 & 4 & S / S1-NTD & 46 \\
+21990 & 6 & S / S1-NTD & 45 \\
+686 & 7 & ORF1ab / NSP1 & 43 \\
+28252 & 2 & ORF8 & 41 \\
 \bottomrule
@@
 
-We performed an indepth analysis of the sc2ts and UShER ARGs 
-(we refer to the UShER tree as an ARG for simplicity here as a single 
-tree is also an ARG) in terms of parsimony.
-A direct comparison is difficult because of the different aligments 
-used as input, the different sets of sites used for inference and 
-the fact that we are including deletions for a subset of sites in
-the sc2ts ARG.
-
-When we compare the intersection of the base sc2ts ARG (pre 
-site-remapping), sc2ts is very slightly more parsimonious 
+We performed an in-depth analysis of the sc2ts and UShER ARGs
+(we refer to the UShER tree as an ARG for simplicity here,
+as a single tree is also an ARG) in terms of parsimony.
+A direct comparison is difficult because of
+the different alignments used as input,
+the different sets of sites used for inference and
+the fact that we are including deletions
+for a subset of sites in the sc2ts ARG.
+
+When we compare the intersection of the base sc2ts ARG
+(pre site-remapping), sc2ts is very slightly more parsimonious
 % TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/684
@@
 (2,078,370 vs 1,934,680). If we exclude mutations involving
-deletions in sc2ts, this is reduced to 46,371 more (2.3\%)
-than UShER.
+deletions in sc2ts, this is reduced to 46,371 more (2.3\%) than UShER.
 If we restrict to the 26,192 sites that were not remapped,
 sc2ts has 4,785 (0.26\%) fewer mutations than UShER,
-and has around 2.5 times more mutations than UShER on the 
-163 sites that were post-hoc remapped.
+and has around 2.5 times more mutations than UShER
+on the 163 sites that were post-hoc remapped.
 
@@
 in the sc2ts and UShER intersection ARGs represents 63.41GiB of nucleotide calls. 
-Of the 162,838,461 missing data calls (Ns) in the alignments, sc2ts and UShER
-disagreed in their imputed values for only 203,579 (0.125\%). 
+Of the 162,838,461 missing data calls (Ns) in the alignments,
+sc2ts and UShER disagreed in their imputed values for only 203,579 (0.125\%).
 Additionally, 956,859 calls made use of the IUPAC uncertainty codes. 
-Of these sc2ts imputed 435 (0.05\%) incorrectly (i.e., 
-with a base that is not compatible with the
-ambiguity code). 
-UShER imputed only 7 calls (0.007\%) from this set incorrectly, likely 
-due to special handling of ambiguous bases in the parsimony calculations.
-
-This computation is a good example of type of integrated large-scale data analysis 
+Of these sc2ts imputed 435 (0.05\%) incorrectly
+(i.e., with a base that is not compatible with the ambiguity code).
+UShER imputed only 7 calls (0.007\%) from this set incorrectly,
+likely due to special handling of
+ambiguous bases in the parsimony calculations.
+
+This computation is a good example of
+type of integrated large-scale data analysis
 that can be easily performed using the combination of tskit and VCF Zarr.
 The entire analysis is performed in a short Jupyter
-notebook~\cite{kluyver2016jupyter}, with the 
-core calculation requiring about 30 lines of Python code. Retrieving 
-data from the base alignments and ARGs for 2.48 million samples 
-at the same time over all 27,507 sites took only 6 minutes, with a peak memory
-usage of about 5GiB of RAM on a standard desktop computer.
+notebook~\cite{kluyver2016jupyter},
+with the core calculation requiring about 30 lines of Python code.
+Retrieving data from the base alignments and ARGs for 2.48 million samples 
+at the same time over all 27,507 sites took only 6 minutes,
+with a peak memory usage of about 5GiB of RAM
+on a standard desktop computer.
 

### 2025-10-01 latex/main.tex
Updates from Overleaf

@@
 
+Akaishi and Fujiwara \cite{Akaishi2023} have hypothesized that XX.
+We agree on XX, but not others.
+Could be limitation of the fact that they didn't analyse many samples.
+
 The placement of the following major deletion events
@@
 \caption{Copying patterns for the 20 recombination events associated with Pango 
-X lineages listed in Table~\ref{tab:pango_x_lineages} (in the same order)}
+X lineages listed in Table~\ref{tab:pango_x_lineages} (in the same order). See also Document S3 for exact positions and nucleotide bases}
 \label{fig:pango_x_copying_patterns}

### 2025-10-01 latex/main.tex
Updates from Overleaf

@@
 
-Akaishi and Fujiwara \cite{Akaishi2023} have hypothesized that XX.
-We agree on XX, but not others.
-Could be limitation of the fact that they didn't analyse many samples.
-
-The placement of the following major deletion events
-in the ARG aligns well with the phylogenetic placement of
-VOC-associated deletion events
-suggested in a previous study \cite{Akaishi2023}:
+Based on a phylogenetic analysis of 25 randomly chosen samples,
+Akaishi and Fujiwara \cite{Akaishi2023} hypothesized
+the emergence of commonly observed indels in SARS-CoV-2.
+We found support in the ARG for the the phylogenetic placement of
+these VOC-associated deletion events suggested in that study:
 a 3-base deletion 21,991 above the Alpha origin node;
@@
 a 9-base deletion at 21,633 above the BA.2 origin node.
-
 However, the ARG suggests that two major deletions
@@
 
+We identified highly recurrent deletions,
+for example, the top 20 deletions in terms of
+the number of occurrences in the ARG shown in Table~\ref{tab:rec_dels}.
+These include known recurrent deletions, for example,
+6-base deletion at 21,765,
+3-base deletion at 21,991, and
+3-base deletion 22,194 in S1-NTD
+associated with antibody escape \cite{McCarthy2021,Meng2021}.
+
+% SHZ: Talk about the agreement before getting into improvements.
+
 There are three contiguous deletions in NSP6
@@
 above the Alpha origin node.
-
-We identified highly recurrent deletions,
-for example, the top 20 deletions in terms of
-the number of occurrences in the ARG shown in Table~\ref{tab:rec_dels}.
-These include known recurrent deletions, for example,
-6-base deletion at 21,765,
-3-base deletion at 21,991, and
-3-base deletion 22,194 in S1-NTD
-associated with antibody escape \cite{McCarthy2021,Meng2021}.
 The out-of-frame deletions in S1-NTD
@@
 they are likely to attach somewhere along the long branch
-in the subgraph visualization (with >30 mutations),
+in the subgraph visualization (with > 30 mutations),
 which also contains the key "R2" deletion

### 2025-10-01 latex/main.tex

@@
 manual inspection of phylogenies and sequence alignments
-by volunteers.
+by volunteers~\cite{Smith2023tracking}.
 
@@
 read misalignments \cite{Hunt2024}, and
-multi-nucleotide substitutions \cite{DeMaio2020issues} (MNSs)
+multi-nucleotide substitutions \cite{DeMaio2020issues} 
 can cause short regions of a sample genome to match more closely
@@
 
-To assist in the identification of such issues
-we developed a SnpIt-like~\cite{OToole2024} visualisation 
-that shows the informative genomic positions between 
-a recombinant and its parent sequences, and highlights 
-potentially problematic loci.
-Each copying pattern shows the positions
-where the allelic state in a recombinant
-(middle row) matches that of the left (upper row, coloured green) 
-or right parent (lower row, coloured blue),
-or where the recombination event requires a de-novo mutation
-(gold, with mutational change below).
-See Figures~\ref{fig:averted_mutations}D and
-\ref{fig:pango_x_copying_patterns} for copying patterns of the 
-Pango X lineages, and 
-Figure~\ref{fig:copying_pattern_examples} for copying patterns
-highlighting specific types of QC problem (and more details on the 
-visualisation).
-
 \subsection*{Quantification of evidence for recombination}
@@
 the two solutions---the number of 
-mutations ``averted'' by recombination---allows us to quantify
-the plausibility.
-
-Sc2ts provides a method to facilitate the discovery of recombination events,
-by comparing how well a sample fits the ARG under a model allowing recombination
-versus a model disallowing recombination,
-as measured by the number of resulting mutations 
-The more mutations are ``averted'' by allowing recombination,
-the stronger the evidence for recombination.
-We calculated the number of averted mutations for
-all the ``robust'' recombination events (STAR Methods),
-which include all the Pango X-associated recombination events
-(Table~\ref{tab:pango_x_lineages})
-and those studied by Jackson et al. \cite{Jackson2021}
-(Table \ref{tab:jackson}, Document~\ref{sec:jackson_recombs}).
-
-Our analysis of the number of averted mutations reveals a spectrum of plausibility
-across all the ``robust'' recombination events
-(Figure~\ref{fig:averted_mutations}A).
-Examples of highly plausible recombinants are XA and XBB
-(for an extended analysis, see Document~\ref{sec:supplement_indepth_pango_x_lineages}).
-The ``copying pattern'' of XA visually confirms
-that a high number of recombination-informative sites support the suggested parents 
-(Figure~\ref{fig:averted_mutations}D).
-XW is an example of a marginally well-supported recombinant.
-The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent.
-% FIXME What's document S3?
-The copying patterns of all the recombination events are shown in Document S3.
-
-[FIXME - discuss BA.5 (kept as a recombination event but only because we can't 
-rewire it post-hoc, owning to subsequent tree building) and the BA.2 event (doesn't
-pass QC.]
-[Our results suggest that Delta and Omicron BA.2 might have recombinant origins [no longer true],
-but we interpret these results with caution,
-as these VOCs are undersampled around the time of their emergence (REF).]
-Note that the presence of many de-novo mutations
-observed in the copying patterns of XBB (Figure 3D) suggests the existence of
-many genetically distinct, unsampled ancestors [FIXME: in this case, probably BJ.1, see
-https://github.com/jeromekelleher/sc2ts-paper/issues/336].
-Future work is needed to test whether these VOCs descended from recombination events,
-with improved sampling, if possible.
+mutations ``averted'' by recombination---is a measure of the evidence
+in favour of recombination. 
+Table~\ref{tab:pango_x_lineages} shows
+the number of averted mutations for each of the 21 recombination events
+associated with Pango X lineages, with support ranging from very
+strong in the case of XC (the most parsimonious non-recombinant 
+placement would require 37 additional mutations) to marginal 
+for XG and XW, which have the minimal possible number of averted 
+mutations.
 
@@
 
-% https://github.com/jeromekelleher/sc2ts-paper/issues/717
-167 (43\%) of QC-passing recombination events were also detected as recombinants by rebar.
-We hypothesize that this disagreement arose from
-differences in sensitivity between these methods.
-Consistent with this, we found that rebar needs a higher level of divergence between parents
-to correctly classify a sequence as recombinant, as shown by
-a shorter time to their MRCA (Figure \ref{fig:averted_mutations}B) and
-a lower distance between their Pango identity (Figure \ref{fig:averted_mutations}C).
+Figure~\ref{fig:averted_mutations}A plots the number of mutations required
+in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
+and the resulting distribution of the number of 
+averted mutations (lower histogram)
+over all 354 recombination events that pass QC. 
+Of these, only 24 
+(21 associated with Pango X lineages in Table~\ref{tab:pango_x_lineages},
+and 3 additional from Jackon et al in Table~\ref{tab:jackson}) have 
+% Slightly risky, this, as we could get asked to incorporate 
+% novel recombs found by RIPPLEs? This could be done if we could get 
+% Viridian sample IDs for their list.
+been thoroughly characterised. 
+The figure also shows the classification status for each recombinant
+according to rebar, a method that classifies sequences
+as recombinant or not based on groups of mutations associated with
+Pango lineages (STAR Methods). 
+Of the 38 recombinants that are strongly supported (at least 10 averted
+mutations), 29 are novel. The known recombinants are XA, XC, XS, XBR, 
+and 5 from Jackson et al. All but 2 of the 38 are classified as
+recombinant by rebar. The majority (20) of these novel strongly supported
+recombinants are singletons, with 9 having further descendants in the ARG.
+Of the remaining 316 recombinants with less than 10 averted mutations,
+only 151 are classified as recombinant by rebar. 
+While some are likely to 
+be false positives, it is also interesting to note that rebar validation
+rate is much lower for recombinants with closely related 
+parents (Figure~\ref{fig:averted_mutations}B,C) suggesting that 
+sc2ts may have a higher sensitivity.
+
+Recombination is rare in SARS-CoV-2, and ultimately judgements about
+plausibility require expert analysis. 
+To aid such interpretation and help identify quality control
+issues we developed a visualisation 
+based on SnpIt~\cite{OToole2024} and 
+the recombinant view from RIVET~\cite{Smith2023tracking},
+showing the informative genomic positions between 
+a recombinant and its parent sequences, and highlights 
+potentially problematic loci.
+Each copying pattern shows the positions
+where the allelic state in a recombinant
+(middle row) matches that of the left (upper row, coloured green) 
+or right parent (lower row, coloured blue),
+or where the recombination event requires a de-novo mutation
+(gold, with mutational change below).
+Figure~\ref{fig:averted_mutations}D shows some example copying patterns,
+illustrating this spectrum of plausibility.
+XA is an example of a highly plausible recombinant, shown by the 
+large number of supporting loci on the left and right. As we have 
+only one de-novo mutation, we can be sure that the parents are closely
+observed [Some details about XA here, how well the parents were sampled etc].
+XBB is also strongly supported by many loci on the left and right, 
+but the presence of many de-novo mutations
+observed in the copying patterns suggests the existence of
+many genetically distinct, unsampled ancestors 
+[FIXME: in this case, probably BJ.1, see issue 336],
+and some caution should be exercised in interpreting them. 
+See Figures~\ref{fig:averted_mutations}D and
+\ref{fig:pango_x_copying_patterns} for copying patterns of the 
+Pango X lineages, and 
+Figure~\ref{fig:copying_pattern_examples} for copying patterns
+highlighting specific types of QC problem (and more details on the 
+visualisation).
+
+Two further recombinants in the sc2ts ARG that are likely to be 
+spurious are important to discuss.
+[FIXME - discuss BA.5 (kept as a recombination event but only because we can't 
+rewire it post-hoc, owning to subsequent tree building) and the BA.2 event (doesn't
+pass QC.]
 
@@
 consistent with patterns observed in previous studies
-\cite{Bobay2020recombination,Turakhia2022,Li2024CovRecomb}.
+\cite{Bobay2020recombination,Turakhia2022,Smith2023tracking,Li2024CovRecomb}.
 While this might result from a higher recombination rate in this region,

### 2025-10-02 latex/main.tex
Updates from Overleaf

@@
 
-\title{A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2}
+\title{A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2}
 \date{}
@@
 
-\affil[*]{Correspondence: shing.zhan@bdi.ox.ac.uk}
+\affil[*]{Correspondence: shing.zhan@ndph.ox.ac.uk}
 \affil[**]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
@@
 some known truly recurrent deletions are present,
-for example, two deletions in S1-NTD [WHICH TWO???]
+for example, three deletions in S1-NTD
+(21,765+6, 21,991+3, and 22,194+3)
 associated with antibody escape\cite{McCarthy2021}.
@@
 
-This work was funded by [FUNDER] via grant [GRANT NO.].
-The authors thank all members of the lab for their support.
+SHZ is supported by an Intermediate Research Fellowship
+from the Nuffield Department of Population Health,
+University of Oxford.
 
@@
 splitting. Of these 16 recombinants, two are noteworthy as 
-being the origins of the Delta and BA.2 lineages in the ARG.
+being the origins of the Delta (B.1.617.2) and Omicron-BA.2 lineages in the ARG.
 Both result in substantially fewer mutations (15 fewer for 
@@
 B.1 to B.1.167 and B.1.1 to BA.1 respectively; they can be seen
-in Figure \ref{fig:subgraph_delta} and \ref{fig:subgraph_ba1_ba2}
+in Figure \ref{fig:subgraph_delta} and \ref{fig:subgraph_omicron}
 with inserted intermediate nodes leading to the Delta and BA.2
@@
 because the LS HMM counts each base separately.
-Fortunately, artefactual recombinants caused by this effect can be identified
-because they affect multiple adjacent or near-adjacent bases simultaneously.
+Fortunately, artefactual recombinants caused by this effect are usually
+identifiable because they affect multiple adjacent or near-adjacent
+bases simultaneously.
 
@@
 of which 6 failed QC.
+
+\subsubsection*{Likely spurious influential recombinants}
+In the sc2ts ARG, there are two recombination events which have a large number of
+descendants that are not classified as recombinants by Pango. The first of these
+corresponds to recombination event within BA.2 whose breakpoint is on the far right
+(position 27382) supported by only 4 sites / 3 loci, and which therefore does not
+pass our recombinant quality control measure. The second of these involves a single
+breakpoint at corresponds to
+the origin of the BA.5 lineage
 
@@
 whose own descendants (if any) are entirely of the focal Pango type.
-We then further remove 4 samples
+We then further remove 4 lineages
 which are singleton descendants of a recombination event in the ARG
@@
 The simplified ARG contains two single-breakpoint recombination nodes. The first
-corresponds to a BA.2 recombinant whose breakpoint is on the far right (position 27382) supported by only 4 sites / 3 loci, anch which therefore does not pass our recombinant quality control measure.
-The second is the origin of BA.5, which (the ).
+ The second is the origin of BA.5 which we also suggest may be artefactual (see Supp FIXME).
 To reduce plot density, we consider only those Pango lineages with fewer than 10 samples,
@@
 by mapping deletions onto the ARG using post-hoc parsimony.
-We identified major deletion events
-(which are inherited by at least 10,000 samples;
-Table~\ref{tab:major_dels}).
+We identified major deletion events, for example,
+the deletions inherited by at least 10,000 samples
+shown in Table~\ref{tab:major_dels}.
+% This just says that the deletions are found in sequences of VOCs.
 All these deletions are known to occur in major VOCs 
@@
 \url{https://github.com/cov-lineages/pango-designation/issues/361};
-\cite{Stern2021,Akaishi2023,Li2023indels},
+\cite{Stern2021,Tegally2022,Li2023indels}),
 except the 4-base deletion and two 5-base deletions in ORF1ab,
-which are likely artefacts of deletion remapping (see below).
+which are likely artefacts of deletion remapping or
+sequence misalignment (see below).
 Notably, eight of these deletions occur
@@
 
-Based on a phylogenetic analysis of 25 randomly chosen samples,
-Akaishi and Fujiwara \cite{Akaishi2023} hypothesized
-the emergence of commonly observed indels in SARS-CoV-2.
-We found support in the ARG for the the phylogenetic placement of
-these VOC-associated deletion events suggested in that study:
-a 3-base deletion 21,991 above the Alpha origin node;
-two 6-base deletions at 22,029 and 28,248 above the Delta origin node;
+% Placement of deletion events linked to VOCs
+In the ARG, we found that the following deletion events
+are associated with the origins of major VOCs
+(Table~\ref{tab:major_dels}):
+a 9-base deletion at 11288,
+a 6-base deletion at 21,765, and
+a 3-base deletion at 21,991 above the Alpha origin node;
+a 6-base deletion at 22,029 above the Delta origin node;
+a 6-base deletion at 28,248 above a child of the Delta origin node;
 a 9-base deletion at 28,362 above the B.1.1.529 origin node;
 a 3-base deletion at 6,513,
-a 9-base deletion at 21,987, and
-a 3-base deletion at 22,194 above the BA.1 origin node;
-a 9-base deletion at 21,633 above the BA.2 origin node.
-However, the ARG suggests that two major deletions
-(a 6-base deletion at 21,765 and a 1-base deletion at 28,271;
-Table~\ref{tab:major_dels}) are recurrent
-(i.e., acquired multiple times),
-but Akaishi and Fujiwara \cite{Akaishi2023} suggest that
-these deletions might have been acquired
-before the divergence of certain VOCs,
-based on an analysis of a small set of 20 samples.
-That study suggests that the 6-base deletion
-occurred before the divergence of Alpha and Omicron,
-but in the ARG this deletion occurs three times,
-above the origin nodes of Alpha, BA.1, and BA.4.
-Similarly, that study suggests that the 1-base deletion
-occurred before the divergence of Alpha, Delta, and Omicron,
-but in the ARG this deletion occurs
-above the Alpha origin node and above node 1436808.
-
-We identified highly recurrent deletions,
+a 3-base deletion at 22,194,
+a 6-base deletion at 21,765, and
+a 9-base deletion at 21,987 above the BA.1 origin node;
+a 9-base deletion at 21,633 above the BA.2 origin node; and
+a 6-base deletion at 21,633 above the BA.4 origin node.
+All these deletion events have been proposed previously
+to be linked with the emergence of these VOCs
+(\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
+\url{https://github.com/cov-lineages/pango-designation/issues/361};
+\cite{Stern2021,Tegally2022}).
+
+% Known recurrent deletions
+Using the ARG, we identified highly recurrent deletions,
 for example, the top 20 deletions in terms of
-the number of occurrences in the ARG shown in Table~\ref{tab:rec_dels}.
+the number of occurrences in the ARG
+shown in Table~\ref{tab:rec_dels}.
 These include known recurrent deletions, for example,
-6-base deletion at 21,765,
-3-base deletion at 21,991, and
-3-base deletion 22,194 in S1-NTD
-associated with antibody escape \cite{McCarthy2021,Meng2021}.
-
-% SHZ: Talk about the agreement before getting into improvements.
-
-There are three contiguous deletions in NSP6
-(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292),
-which should have been mapped as two 9-base deletions,
-one above the BA.1 origin node (starting at 11,283) and
-the other above the BA.2 origin node (starting at 11,288)
-in the ARG (Figure~\ref{fig:subgraph_ba1_ba2}),
-but they were broken up during deletion remapping
-as they had four overlapping sites at 11,288-11,292
-(which was mapped as the 4-base deletion event).
-Furthermore, Akaishi and Fujiwara \cite{Akaishi2023}
-suggest that a 9-base deletion at 11,288 occurred
-before the divergence of Alpha and Omicron.
-In the ARG, however, this deletion occurs
-above the Alpha origin node.
-The out-of-frame deletions in S1-NTD
-(e.g., the 2-base deletion at 21,765),
-N, ORF1ab, and ORF8
-are implausible and are likely artefacts caused by
-sequencing errors or sequence misalignments.
-Future improvements in
-data pre-processing and deletion remapping
-should remove such deletions.
+a 6-base deletion at 21,765,
+a 3-base deletion at 21,991, and
+a 3-base deletion 22,194 in S1-NTD
+associated with antibody escape \cite{McCarthy2021,Meng2021},
+which are also caused by major deletion events in the ARG
+(Table~\ref{tab:major_dels}).
+
+% Problems and future improvement
+Detection of deletions and inference of deletion events
+are difficult, especially when the dataset is enormous.
+Here we used a naive procedure to map deletions onto the ARG,
+treating each site involved in a deletion
+as independent of the other sites.
+Consequently, some deletion events might be split up and
+appear as multiple smaller deletion events,
+which are likely artefactual.
+The three contiguous out-of-frame deletions in NSP6
+(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292)
+are examples of such artefacts (Table~\ref{tab:major_dels}).
+In the sequence alignments, there are two 9-base deletions,
+one at 11,283 which commonly occur in BA.1 samples and
+the other at 11,288 which commonly occur in BA.2 samples.
+Assuming that the alignments are correct,
+there should be two deletion events,
+but they were mapped as three deletion events
+because they had four overlapping sites from 11,288 to 11,292,
+which were inferred by parsimony
+as the 4-base deletion event at 11,288.
+Furthermore, sequencing errors or sequence misalignments
+can also cause artefactual deletions,
+for example, the out-of-frame recurrent deletions
+in N, ORF1ab, ORF8, and particularly S1-NTD
+(e.g., the 2-base deletion at 21,765)
+(Table~\ref{tab:rec_dels}).
+Future improvements in deletion remapping and data pre-processing
+should remove such artefactual deletions.
 
@@
 \centering
-\includegraphics[width=0.85\linewidth]{subgraph-BA.1+BA.2.pdf}
+\includegraphics[width=0.85\linewidth]{subgraph-Omicron.pdf}
 \caption{
-Subgraph of BA.1 and BA.2.
+Subgraph of the BA.1 and BA.2 Omicron lineages. BA.1 samples are shown 
 [Some notes to help the reader know this 
@@
 }
-\label{fig:subgraph_ba1_ba2}
+\label{fig:subgraph_omicron}
 \end{figure}

### 2025-10-02 latex/main.tex
Fixup references
Updates from Overleaf

@@
 and to get established in the ARG.
-See Figure~\ref{fig:subgraph_ba1_ba2} for an overview of the resulting 
+See Figure~\ref{fig:subgraph_omicron} for an overview of the resulting 
 subgraph.
@@
 except the 4-base deletion and two 5-base deletions in ORF1ab,
-which are likely artefacts of deletion remapping or
-sequence misalignment (see below).
+which are likely artefacts of
+deletion remapping or sequence misalignment (see below).
 Notably, eight of these deletions occur
@@
 (also see Table~\ref{tab:rec_dels}).
-
-% Placement of deletion events linked to VOCs
-In the ARG, we found that the following deletion events
+In the ARG, we found that most of the major deletion events
 are associated with the origins of major VOCs
-(Table~\ref{tab:major_dels}):
-a 9-base deletion at 11288,
-a 6-base deletion at 21,765, and
-a 3-base deletion at 21,991 above the Alpha origin node;
-a 6-base deletion at 22,029 above the Delta origin node;
-a 6-base deletion at 28,248 above a child of the Delta origin node;
-a 9-base deletion at 28,362 above the B.1.1.529 origin node;
-a 3-base deletion at 6,513,
-a 3-base deletion at 22,194,
-a 6-base deletion at 21,765, and
-a 9-base deletion at 21,987 above the BA.1 origin node;
-a 9-base deletion at 21,633 above the BA.2 origin node; and
-a 6-base deletion at 21,633 above the BA.4 origin node.
-All these deletion events have been proposed previously
-to be linked with the emergence of these VOCs
+(Table~\ref{tab:major_dels}).
+They have been linked previously with the emergence of these VOCs
 (\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
@@
 
-% Known recurrent deletions
 Using the ARG, we identified highly recurrent deletions,
@@
 
-% Problems and future improvement
 Detection of deletions and inference of deletion events
@@
 }
-\label{tab:major_dels}
-\begin{tabular}{rrcrcrr}
+\begin{tabular}{rrcrcrrc}
 \toprule
-Start & Length & Region & Node & Pango lineage & Count & Frequency (\%)\\
+Start & Length & Region & Node & Pango lineage & Count & Frequency (\%) & Ref\\
 \midrule
-28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
-28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 \\
-22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 \\
-11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 (Omicron) & 871782 & 35.1220 \\
-28362 & 9 & N & 1436802 & B.1.1.529 (Omicron) & 871774 & 35.1216 \\
-21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 \\
-11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 \\
-6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 \\
-11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 \\
-21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 \\
-21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
-22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 \\
-28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 \\
-11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 \\
-21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 \\
-21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 \\
-21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 \\
-686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 \\
+28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 &  \\
+28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 & B \\
+22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 & B \\
+11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 (Omicron) & 871782 & 35.1220 &  \\
+28362 & 9 & N & 1436802 & B.1.1.529 (Omicron) & 871774 & 35.1216 & C \\
+21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 & D \\
+11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 &  \\
+6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 & D \\
+11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 &  \\
+21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 & D \\
+21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 & D \\
+22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 & D \\
+28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 &  \\
+11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 & A \\
+21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 & A \\
+21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 & A \\
+21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 & E \\
+686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 &  \\
 \bottomrule

### 2025-10-02 latex/main.tex
Updates from Overleaf

@@
 All these deletions are known to occur in major VOCs
-(\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
-\url{https://github.com/cov-lineages/pango-designation/issues/361};
-\cite{Stern2021,Tegally2022,Li2023indels}),
+\cite{RambautAlpha2020,Stern2021,RambautOmicron2021,
+Tegally2022,Li2023indels},
 except the 4-base deletion and two 5-base deletions in ORF1ab,
@@
 are associated with the origins of major VOCs
-(Table~\ref{tab:major_dels}).
-They have been linked previously with the emergence of these VOCs
-(\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
-\url{https://github.com/cov-lineages/pango-designation/issues/361};
-\cite{Stern2021,Tegally2022}).
+(Table~\ref{tab:major_dels}), and
+they have been linked previously with the emergence of these VOCs
+\cite{RambautAlpha2020,Stern2021,RambautOmicron2021,Tegally2022}.
 
@@
 }
+\label{tab:major_dels}
 \begin{tabular}{rrcrcrrc}
@@
 28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 &  \\
-28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 & B \\
-22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 & B \\
+28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 & \cite{Stern2021} \\
+22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 & \cite{Stern2021} \\
 11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 (Omicron) & 871782 & 35.1220 &  \\

### 2025-10-02 latex/main.tex
Updates from Overleaf

@@
 \label{tab:major_dels}
-\begin{tabular}{rrcrcrrc}
+\begin{tabular}{rccrcrrc}
 \toprule
-Start & Length & Region & Node & Pango lineage & Count & Frequency (\%) & Ref\\
+Start & Length & Region & Node & Pango lineage & Frequency (\%) & Ref\\
 \midrule
-28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 &  \\
-28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 & \cite{Stern2021} \\
-22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 & \cite{Stern2021} \\
-11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 (Omicron) & 871782 & 35.1220 &  \\
-28362 & 9 & N & 1436802 & B.1.1.529 (Omicron) & 871774 & 35.1216 & C \\
-21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 & D \\
-11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 &  \\
-6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 & D \\
-11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 &  \\
-21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 & D \\
-21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 & D \\
-22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 & D \\
-28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 &  \\
-11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 & A \\
-21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 & A \\
-21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 & A \\
-21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 & E \\
-686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 &  \\
+28271 & 1 & non-coding & 1436808 & B.1 & 45.06 &  \\
+28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 44.91 & \cite{Stern2021} \\
+22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 44.82 & \cite{Stern2021} \\
+11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 (Omicron) & 35.12 &  \\
+28362 & 9 & N & 1436802 & B.1.1.529 (Omicron) & 35.12 & \cite{RambautOmicron2021} \\
+21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 21.44 & \cite{RambautOmicron2021} \\
+11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 21.43 &  \\
+6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 13.73 & \cite{RambautOmicron2021} \\
+11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 13.69 &  \\
+21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 13.68 & \cite{RambautOmicron2021} \\
+21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 13.68 & \cite{RambautOmicron2021} \\
+22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 12.60 & \cite{RambautOmicron2021} \\
+28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 11.70 &  \\
+11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 11.70 & \cite{RambautAlpha2020} \\
+21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 11.64 & \cite{RambautAlpha2020} \\
+21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 11.70 & \cite{RambautAlpha2020} \\
+21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 7.67 & \cite{Tegally2022} \\
+686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 0.94 &  \\
 \bottomrule
@@
 % https://github.com/jeromekelleher/sc2ts-paper/issues/772
+\subsubsection{Alpha}
+XX
+
+\subsubsection{Delta}
+XX
+
+\subsubsection{Omicron BA.1 and BA.2}
+XX
 
@@
 \caption{
-Sc2ts recombination detection results for Pango X lineages not added to the ARG.
+Sc2ts recombination detection results for Pango X lineages which were not added to the ARG.
 }

### 2025-10-03 notebooks/nb_utils.py
Update subgraphs to remove titles
Also add a text description to each

@@ -350,6 +350,7 @@ class D3ARG_viz:
         pangos,
         extra_html=None,
         *,
+        show_title=True,
         width=750,
         height=950,
         parent_levels=20,
@@ -400,10 +401,10 @@ class D3ARG_viz:
         nodes = list((used_pango_samples | include) - exclude)
         used = self.d3arg.subset_graph(nodes, depth=(parent_levels, child_levels))
         title = (
-            f'Subgraph of {self.pangolin_field} {"/".join(pangos)}: '
+            f'Subgraph of {self.pangolin_field.capitalize()} {"/".join(pangos)}: '
             f'({len(pango_samples)} sample{"" if len(pango_samples) == 1 else "s"},'
             f" {len(used_pango_samples)} shown)"
-        )
+        ) if show_title else None
         if highlight_mutations is None:
             if parent_pangos is not None and self.lineage_consensus_muts is not None:
                 # Could replace with https://github.com/andersen-lab/Freyja/blob/main/freyja/data/lineage_mutations.json

### 2025-10-03 latex/main.tex

@@
 \subsubsection{Alpha}
-XX
+The B.1.1.7 (Alpha) variant is the first major VOC, and
+it is a saltational lineage
+characterized by a long branch preceding its origin.
+The origin of Alpha is well reconstructed in the ARG.
+Most of the characteristic Alpha mutations (\cite{RambautAlpha2020})
+(20 out of 23) in the ARG
+occur immediately above the Alpha root.
+However, a characteristic three-base correlated mutation
+is instead associated with the parent node (labeled B.1.1) of the Alpha root,
+suggesting that this mutation was acquired prior to the emergence of Alpha.
+This parent node has an additional 5 children
+(leading to 12 samples),
+of which we display one (ERR4413600)
+with no associated mutations, meaning that it has an identical sequence to the inferred ancestor of Alpha.
+Below the root of Alpha, a single sample (ERR5178290)
+appears without the characteristic mutation C3267T
+(or any additional mutations).
+However, this is only a single sample and
+given that it also appears two months after the first Alpha samples
+we treat its placement with caution.
+Note that the list of mutations in \cite{RambautAlpha2020}
+includes a T26801C in the M gene
+which should no longer be considered characteristic of Alpha,
+as the Wuhan-Hu-1 reference at 26,801 has since been updated to C.
+In the ARG, a one-base non-coding deletion (position 28,271)
+is also associated with the Alpha root;
+although this is not in the original list of
+characteristic Alpha mutations,
+it has been suggested elsewhere as an Alpha-defining mutation 
+(\cite{Li2023indels}).
 
@@
 \caption{
-Subgraph of Alpha. [Some notes to help the reader know this 
-reflects known events]
+Subgraph illustrating the saltational origin of B.1.1.7 (Alpha).
+Squares represent sample nodes,
+circles represent reconstructed internal nodes;
+nodes classified as B.1.1.7 by Pangolin v4.3.1 are in pink.
+Mutations are shown as rectangles along branches,
+ordered by position; deletions are black.
+Characteristic mutations
+associated with the emergence of Alpha
+are highlighted in pink borders.
 }
@@
 \caption{
-Subgraph of Delta. [Some notes to help the reader know this 
-reflects known events]
+Subgraph illustrating the saltational origin of B.1.617.2 (Delta).
+Symbols as in \ref{fig:subgraph_alpha}, with characteristic Delta mutations as pink boxes.
+Sites associated with multiple mutations in the subgraph have their mutations assigned a unique fill colour (orange, blue, green, red, etc); reversions are further highlighted with a black outline.
+Sc2ts seed samples (see STAR methods) for B.1.617, B.1.617.1, and B.1.617.2 (Delta) are plotted as red squares.
 }

### 2025-10-03 arg_postprocessing/scripts/process_ripples.py
Add script to process ripples data

@@ -1,0 +1,92 @@
+import click
+import pandas as pd
+import numpy as np
+import sc2ts
+import tszip
+import dataclasses
+import collections
+import tqdm
+
+
+@dataclasses.dataclass
+class UsherEvent:
+    usher_node: str
+    descendants: list[str]
+    pango_counts: dict
+    sc2ts_mrca: int = -1
+    sc2ts_num_descendants: int = 0
+    sc2ts_closest_recombinant: int = -1
+    sc2ts_closest_recombinant_path_len: int = 2**31
+
+    def asdict(self):
+        d = dataclasses.asdict(self)
+        del d["descendants"]
+        d["usher_num_descendants"] = len(self.descendants)
+        return d
+
+
+@click.command()
+@click.argument("ripples_dir")
+@click.argument("sc2ts_ts")
+@click.argument("output")
+def run(ripples_dir, sc2ts_ts, output):
+
+    descendants = {}
+    with open(f"{ripples_dir}/descendants.tsv") as f:
+        for line in f:
+            if not line.startswith("#"):
+                split = line.strip().split("\t")
+                descs = split[1].split(",")
+                if descs[-1] == "":
+                    descs = descs[:-1]
+                descendants[split[0]] = descs
+    len("Read {len(descendants)} events")
+
+    ts = tszip.load(sc2ts_ts)
+    dfn = sc2ts.node_data(ts, inheritance_stats=False)
+    df_sample = dfn[dfn.is_sample].set_index("sample_id")
+
+    print("Computing mappings")
+    sample_id_to_node = df_sample["node_id"].to_dict()
+    sample_id_to_pango = df_sample["pango"].to_dict()
+    tree = ts.first()
+
+    seen_descendants = set()
+    events = []
+    for usher_node, desc in tqdm.tqdm(descendants.items()):
+        desc = frozenset(desc)
+        if desc in seen_descendants:
+            # We just pick the first one if we see repeated descendant sets
+            print(f"Skipping {len(desc)} descendant set already seen")
+            continue
+        seen_descendants.add(desc)
+        event = UsherEvent(usher_node, desc, {"unclassified": -1})
+        if len(desc) < 10_000:
+            nodes = [sample_id_to_node[sid] for sid in desc]
+            if len(nodes) == 1:
+                mrca = int(nodes[0])
+            else:
+                mrca = tree.mrca(*nodes)
+            event.sc2ts_mrca = mrca
[…]

### 2025-10-03 arg_postprocessing/scripts/process_ripples.py
ADd the ripples p4 data

@@ -48,6 +48,9 @@ def run(ripples_dir, sc2ts_ts, output):
 
     print("Computing mappings")
     sample_id_to_node = df_sample["node_id"].to_dict()
+    # Note: possibly we should use the original source metadata
+    # for sample Pango mappings here, but it really shouldn't
+    # matter.
     sample_id_to_pango = df_sample["pango"].to_dict()
     tree = ts.first()
 
@@ -85,7 +88,7 @@ def run(ripples_dir, sc2ts_ts, output):
     for e in events:
         data.append(e.asdict())
     df = pd.DataFrame(data)
-    df.reset_index().to_csv(output, index=False)
+    df.to_csv(output, index=False)
 
 
 if __name__ == "__main__":

### 2025-10-03 latex/main.tex
Updates from Overleaf

@@
 % See https://github.com/jeromekelleher/sc2ts-paper/issues/628
-We reconstructed
-a more comprehensive mutational landscape of SARS-CoV-2
-by mapping deletions onto the ARG using post-hoc parsimony.
-We identified major deletion events, for example,
-the deletions inherited by at least 10,000 samples
-shown in Table~\ref{tab:major_dels}.
+To reconstruct
+a more comprehensive mutational landscape of SARS-CoV-2,
+we mapped deletions onto the ARG using post-hoc parsimony.
+We identified major deletion events in the ARG,
+which are inherited by at least 10,000 samples
+(Table~\ref{tab:major_dels}).
 % This just says that the deletions are found in sequences of VOCs.
-All these deletions are known to occur in major VOCs
-\cite{RambautAlpha2020,Stern2021,RambautOmicron2021,
-Tegally2022,Li2023indels},
+These deletions are known to occur in major VOCs
+\cite{RambautAlpha2020,
+Stern2021,
+RambautOmicron2021,
+Tegally2022,
+Li2023indels},
 except the 4-base deletion and two 5-base deletions in ORF1ab,
-which are likely artefacts of
-deletion remapping or sequence misalignment (see below).
+which are likely artefacts of deletion remapping (see below).
 Notably, eight of these deletions occur
-in the N-terminal domain of the S1 unit in Spike,
+in the N-terminal domain of the S1 unit of Spike (S1-NTD),
 a known hotspot for recurrent deletions \cite{McCarthy2021}
 (also see Table~\ref{tab:rec_dels}).
-In the ARG, we found that most of the major deletion events
+In the ARG, we found that most of these major deletion events
 are associated with the origins of major VOCs
 (Table~\ref{tab:major_dels}), and
-they have been linked previously with the emergence of these VOCs
+they have been linked previously to the emergence of these VOCs
 \cite{RambautAlpha2020,Stern2021,RambautOmicron2021,Tegally2022}.
 
+The ARG enales large-scale detection of recurrent deletions.
 Using the ARG, we identified highly recurrent deletions,
-for example, the top 20 deletions in terms of
-the number of occurrences in the ARG
-shown in Table~\ref{tab:rec_dels}.
+for example, the top five deletions
+in the number of occurrences in the ARG
+(Table~\ref{tab:rec_dels}).
 These include known recurrent deletions, for example,
-a 6-base deletion at 21,765,
-a 3-base deletion at 21,991, and
-a 3-base deletion 22,194 in S1-NTD
-associated with antibody escape \cite{McCarthy2021,Meng2021},
+a 6-base deletion at 21,765 \cite{Meng2021} and
+a 3-base deletion at 21,991 \cite{McCarthy2021} in S1-NTD,
 which are also caused by major deletion events in the ARG
@@
 Detection of deletions and inference of deletion events
-are difficult, especially when the dataset is enormous.
-Here we used a naive procedure to map deletions onto the ARG,
+remain challenging, especially when the dataset is enormous.
+Here we took a naive approach to map deletions onto the ARG,
 treating each site involved in a deletion
@@
 In the sequence alignments, there are two 9-base deletions,
-one at 11,283 which commonly occur in BA.1 samples and
-the other at 11,288 which commonly occur in BA.2 samples.
+one at 11,283 and the other at 11,288,
+which are commonly found in BA.1 and BA.2 samples, respectively.
 Assuming that the alignments are correct,
@@
 but they were mapped as three deletion events
-because they had four overlapping sites from 11,288 to 11,292,
+as they had four overlapping sites from 11,288 to 11,292,
 which were inferred by parsimony
@@
 Furthermore, sequencing errors or sequence misalignments
-can also cause artefactual deletions,
-for example, the out-of-frame recurrent deletions
-in N, ORF1ab, ORF8, and particularly S1-NTD
-(e.g., the 2-base deletion at 21,765)
+can also cause artefactual deletions, for example,
+the out-of-frame 4-base recurrent deletion at 21,767 in S1-NTD
 (Table~\ref{tab:rec_dels}).
@@
 \caption{
-Highly recurrent deletions.
-These are the top 20 deletions
-in terms of the number of occurrences in the ARG.
+Recurrent deletions.
+These deletions occur at least 5 times in the ARG and
+are found in at least 1,000 samples.
 Abbreviations:
@@
 \label{tab:rec_dels}
-\begin{tabular}{rrcr}
+\begin{tabular}{rrcrr}
 \toprule
-Start & Length & Region & Occurrences \\
+Start & Length & Region & Occurrences & Samples\\
 \midrule
-686 & 9 & ORF1ab / NSP1 & 6436 \\
-21991 & 3 & S / S1-NTD & 2088 \\
-21765 & 6 & S / S1-NTD & 482 \\
-22029 & 6 & S / S1-NTD & 275 \\
-22194 & 3 & S / S1-NTD & 200 \\
-6513 & 3 & ORF1ab / NSP3 & 181 \\
-21765 & 2 & S / S1-NTD & 140 \\
-21995 & 1 & S / S1-NTD & 90 \\
-21988 & 3 & S / S1-NTD & 78 \\
-21987 & 7 & S / S1-NTD & 78 \\
-21993 & 3 & S / S1-NTD & 77 \\
-21987 & 9 & S / S1-NTD & 70 \\
-21994 & 2 & S / S1-NTD & 65 \\
-28370 & 1 & N & 61 \\
-11292 & 3 & ORF1ab / NSP6 & 48 \\
-21990 & 1 & S / S1-NTD & 46 \\
-21987 & 4 & S / S1-NTD & 46 \\
-21990 & 6 & S / S1-NTD & 45 \\
-686 & 7 & ORF1ab / NSP1 & 43 \\
-28252 & 2 & ORF8 & 41 \\
+686 & 9 & ORF1ab / NSP1 & 6436 & 43354\\
+21991 & 3 & S / S1-NTD & 2088 & 309629\\
+21765 & 6 & S / S1-NTD & 482 & 820194\\
+22029 & 6 & S / S1-NTD & 275 & 1107941\\
+22194 & 3 & S / S1-NTD & 200 & 310349\\
+6513 & 3 & ORF1ab / NSP3 & 181 & 341380\\
+21987 & 9 & S / S1-NTD & 70 & 338633\\
+28248 & 6 & ORF8 & 28 & 1113392\\
+28271 & 1 & non-coding & 24 & 1331409\\
+11288 & 9 & ORF1ab / NSP6 & 19 & 843291\\
+11283 & 9 & ORF1ab / NSP6 & 13 & 339810\\
+21767 & 4 & S / S1-NTD & 10 & 2144\\
+28362 & 9 & N & 7 & 850483\\
+21633 & 9 & S / S1-NTD & 7 & 531973\\
 \bottomrule

### 2025-10-03 latex/main.tex
Fix up x lineage copying table
Closes #798

@@
 XBG &\includegraphics[height=\cprheight]{static/XBG.png}\\
+XQ & \includegraphics[height=\cprheight]{static/XQ++.png}\\
 XBD & \includegraphics[height=\cprheight]{static/XBD.png}\\
-XQ & \includegraphics[height=\cprheight]{static/XQ++.png}\\
+XM & \includegraphics[height=\cprheight]{static/XM+XAL.png}\\
+XBH & \includegraphics[height=\cprheight]{static/XBH.png}\\
 XBB & \includegraphics[height=\cprheight]{static/XBB.png}\\
-XM & \includegraphics[height=\cprheight]{static/XM+XAL.png}\\
+XY & \includegraphics[height=\cprheight]{static/XY.png}\\
+XF & \includegraphics[height=\cprheight]{static/XF.png}\\
 XBF & \includegraphics[height=\cprheight]{static/XBF.png}\\
-XF & \includegraphics[height=\cprheight]{static/XF.png}\\
-XY & \includegraphics[height=\cprheight]{static/XY.png}\\
+XW & \includegraphics[height=\cprheight]{static/XW.png}\\
 XG & \includegraphics[height=\cprheight]{static/XG.png}\\
-XW & \includegraphics[height=\cprheight]{static/XW.png}\\
+\hline
 Xx & \includegraphics[height=\cprheight]{static/XZ++.png}\\
 XE/XH  & \includegraphics[height=\cprheight]{static/XE+XH.png}\\
-XBH & \includegraphics[height=\cprheight]{static/XBH.png}\\
 XBM & \includegraphics[height=\cprheight]{static/XBM.png}\\
 XJ & \includegraphics[height=\cprheight]{static/XJ.png}\\
+XAF & \includegraphics[height=\cprheight]{static/XAF.png}\\
 \end{tabularx}
-\caption{Copying patterns for the 20 recombination events associated with Pango 
-X lineages listed in Table~\ref{tab:pango_x_lineages} (in the same order). See also Document S3 for exact positions and nucleotide bases}
+\caption{Copying patterns for the 21 recombination events associated with Pango 
+X lineages listed in Table~\ref{tab:pango_x_lineages} (in the same order). 
+The horizontal line separates Type I and Type II events.
+See also Document S3 for exact positions and nucleotide bases.}
 \label{fig:pango_x_copying_patterns}

### 2025-10-03 latex/main.tex
Cite bolotie
Closes #826

@@
 recombinant samples~\cite{
+Varabyou2021rapid,% Bolotie
 Turakhia2022,%RIPPLES

### 2025-10-03 latex/main.tex

@@
 recent breakthroughs in
-simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019}
+simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019,Gopalan2025bridging}
 and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,

### 2025-10-03 latex/main.tex
Updates

--- 2025-10-03 latex/main.tex
@@
 % See https://github.com/jeromekelleher/sc2ts-paper/issues/628
-To reconstruct
-a more comprehensive mutational landscape of SARS-CoV-2,
-we mapped deletions onto the ARG using post-hoc parsimony.
-We identified major deletion events in the ARG,
-which are inherited by at least 10,000 samples
-(Table~\ref{tab:major_dels}).
-% This just says that the deletions are found in sequences of VOCs.
-These deletions are known to occur in major VOCs
-\cite{RambautAlpha2020,
-Stern2021,
-RambautOmicron2021,
-Tegally2022,
-Li2023indels},
-except the 4-base deletion and two 5-base deletions in ORF1ab,
-which are likely artefacts of deletion remapping (see below).
-Notably, eight of these deletions occur
+Mapping deletions onto the ARG using post-hoc parsimony
+allowed us to identify important deletions and
+to clarify their nature in SARS-CoV-2 evolution.
+In Table~\ref{tab:major_dels},
+we list a subset of deletion events in the ARG
+which are inherited by at least 10,000 samples.
+We consider these deletions to be major events
+in the evolutionary history of SARS-CoV-2.
+Most of these deletion events
+(besides the 1-base deletion at 28,271
+inherited by 45\% of the samples)
+are associated with the origin nodes of major VOCs in the ARG
+(indicated by the Pango lineages assigned to them,
+e.g., node 86456 is the Alpha origin node;
+also see Figure~\ref{fig:subgraph_alpha}).
+Most of them are characteristic mutations
+that are linked to the emergence of these VOCs
+(see the references in Table~\ref{tab:major_dels}
+pointing to the analyses
+which originally identified them as characteristic mutations).
+They are also highly prevalent
+(all except two events in this table
+are inherited by at least 11\% of the samples).
+Notably, eight of these deletion events occur
 in the N-terminal domain of the S1 unit of Spike (S1-NTD),
-a known hotspot for recurrent deletions \cite{McCarthy2021}
-(also see Table~\ref{tab:rec_dels}).
-In the ARG, we found that most of these major deletion events
-are associated with the origins of major VOCs
-(Table~\ref{tab:major_dels}), and
-they have been linked previously to the emergence of these VOCs
-\cite{RambautAlpha2020,Stern2021,RambautOmicron2021,Tegally2022}.
-
-The ARG enales large-scale detection of recurrent deletions.
-Using the ARG, we identified highly recurrent deletions,
-for example, the top five deletions
-in the number of occurrences in the ARG
-(Table~\ref{tab:rec_dels}).
-These include known recurrent deletions, for example,
+a known hotspot of recurrent deletions \cite{McCarthy2021};
+These are all highly recurrent (Table~\ref{tab:rec_dels}), and
+include known recurrent deletions, for example,
 a 6-base deletion at 21,765 \cite{Meng2021} and
-a 3-base deletion at 21,991 \cite{McCarthy2021} in S1-NTD,
-which are also caused by major deletion events in the ARG
-(Table~\ref{tab:major_dels}).
+a 3-base deletion at 21,991 \cite{McCarthy2021}.
 
@@
 Major deletion events in the ARG.
-These events are inherited by
-at least 10,000 samples in the ARG.
-Count is the number of samples inheriting a deletion.
+The deletions caused by these events
+are inherited by at least 10,000 samples.
+The frequencies of these deletions in the ARG are shown.
 The Pango lineage assigned to the node
-above which an event occurs is shown.
+where an event occurs is shown.
+The references point to analyses
+which identified a deletion
+to be associated with a major VOC assigned to its node.
 Abbreviations: NSP, non-structural protein;
@@
 characterized by a long branch preceding its origin.
-The origin of Alpha is well reconstructed in the ARG.
+Figure \ref{fig:subgraph_alpha} shows that the origin of Alpha is well reconstructed in the ARG.
 Most of the characteristic Alpha mutations (\cite{RambautAlpha2020})
@@
 \subsubsection{Delta}
-XX
+ \ref{fig:subgraph_delta}
 
@@
 XBG &\includegraphics[height=\cprheight]{static/XBG.png}\\
+XBD & \includegraphics[height=\cprheight]{static/XBD.png}\\
 XQ & \includegraphics[height=\cprheight]{static/XQ++.png}\\
-XBD & \includegraphics[height=\cprheight]{static/XBD.png}\\
+XBB & \includegraphics[height=\cprheight]{static/XBB.png}\\
 XM & \includegraphics[height=\cprheight]{static/XM+XAL.png}\\
-XBH & \includegraphics[height=\cprheight]{static/XBH.png}\\
-XBB & \includegraphics[height=\cprheight]{static/XBB.png}\\
+XBF & \includegraphics[height=\cprheight]{static/XBF.png}\\
+XF & \includegraphics[height=\cprheight]{static/XF.png}\\
 XY & \includegraphics[height=\cprheight]{static/XY.png}\\
-XF & \includegraphics[height=\cprheight]{static/XF.png}\\
-XBF & \includegraphics[height=\cprheight]{static/XBF.png}\\
+XG & \includegraphics[height=\cprheight]{static/XG.png}\\
 XW & \includegraphics[height=\cprheight]{static/XW.png}\\
-XG & \includegraphics[height=\cprheight]{static/XG.png}\\
-\hline
 Xx & \includegraphics[height=\cprheight]{static/XZ++.png}\\
 XE/XH  & \includegraphics[height=\cprheight]{static/XE+XH.png}\\
+XBH & \includegraphics[height=\cprheight]{static/XBH.png}\\
 XBM & \includegraphics[height=\cprheight]{static/XBM.png}\\
 XJ & \includegraphics[height=\cprheight]{static/XJ.png}\\
-XAF & \includegraphics[height=\cprheight]{static/XAF.png}\\
 \end{tabularx}
-\caption{Copying patterns for the 21 recombination events associated with Pango 
-X lineages listed in Table~\ref{tab:pango_x_lineages} (in the same order). 
-The horizontal line separates Type I and Type II events.
-See also Document S3 for exact positions and nucleotide bases.}
+\caption{Copying patterns for the 20 recombination events associated with Pango 
+X lineages listed in Table~\ref{tab:pango_x_lineages} (in the same order). See also Document S3 for exact positions and nucleotide bases}
 \label{fig:pango_x_copying_patterns}
@@
 associated with the emergence of Alpha
-are highlighted in pink borders.
+are highlighted with a pink outline.
 }
@@
 Subgraph illustrating the saltational origin of B.1.617.2 (Delta).
-Symbols as in \ref{fig:subgraph_alpha}, with characteristic Delta mutations as pink boxes.
+Symbols as in \ref{fig:subgraph_alpha}, with characteristic Delta mutations outlined in pink.
 Sites associated with multiple mutations in the subgraph have their mutations assigned a unique fill colour (orange, blue, green, red, etc); reversions are further highlighted with a black outline.

### 2025-10-03 latex/main.tex
Updates from Overleaf

@@
 those identified by Jackson et al. \cite{Jackson2021} are prefixed with a `J', e.g.
-Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1).
+Jackson group B (JB) and Jackson singleton recombinant 1 (J1).
 Point sizes reflect number of descendants.
@@
 (D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
-leading to XA, Xx 
-(the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an
+leading to XA, XW, XBB, and an
 undesignated recombination event under XBB, which we here refer to as  XBB.x.}
@@
 
-\subsubsection*{Likely spurious influential recombinants}
+\subsubsection*{Likely spurious ancestrally-important recombinants}
+\label{sec:spurious-ancestrally-important-recombinants}
 In the sc2ts ARG, there are two recombination events which have a large number of
-descendants that are not classified as recombinants by Pango. The first of these
-corresponds to recombination event within BA.2 whose breakpoint is on the far right
-(position 27382) supported by only 4 sites / 3 loci, and which therefore does not
-pass our recombinant quality control measure. The second of these involves a single
-breakpoint at corresponds to
-the origin of the BA.5 lineage
+descendants but which are not classified as recombinants by Pango. The first of these
+corresponds to recombination event within BA.2 whose breakpoint is at position 27382
+on the far right of the genome (see supplementary document S3, node 1030562). Although
+supported by 5 sites, three of them are adjacent so that they only represent 3 loci,
+meaning that this event does not pass our recombinant quality control measure.
+
+The second of these involves a single breakpoint at position 26858, and is the origin
+of the BA.5 lineage in the ARG (see supplementary document S3, node 1189192). Although
+supported by 5 loci on the right, this recombinant also requires 3 de-novo mutations,
+some of which are adjacent to the supporting loci. Furthermore, when the first BA.5
+sample is added to the ARG, it is possible to construct a more parsimonious
+non-recombinant origin by long branch splitting (see the \nameref{sec:arg_postprocessing}
+section), although subsequent tree building does not make this a parsimonious
+way to rewire the final ARG. We therefore suggest that this too could be an artifactual
+recombination event.
 
@@
 as described in \ref{sec:supplement_pango_lineages}, and
-omit Pango designations starting with "X",
+omit Pango designations starting with "X"
 to avoid recombinants, leaving 1329 Pango lineages.
@@
 
-The simplified ARG contains two single-breakpoint recombination nodes. The first
- The second is the origin of BA.5 which we also suggest may be artefactual (see Supp FIXME).
-To reduce plot density, we consider only those Pango lineages with fewer than 10 samples,
-and extract non-recombinant structures from the ARG
-by creating a ``base tree'' of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, and BA.5),
-a Delta subtree of 160 Pangos,
-a BA.2 subtree of 151 sample (excluding BA.5 descendants) and
-a BA.5 subtree of 251 samples.
-Figure SX shows each of these trees compared against the equivalent dated UShER topology,
-with the left and right parents of Delta and BA.2 highlighted in the base tree.
-The order of tips in the two trees was chosen to reduce tangling,
+The simplified ARG contains only two recombination nodes, which
+correspond to those on the far right of the genome identified as
+artefactual in the \ref{sec:spurious-ancestrally-important-recombinants} section.
+We therefore compare the Usher topology against the first tree in the ARG (covering
+positions 1-26858; the majority of the genome). To reduce plot density, we consider
+only those Pango lineages with greater than 10 samples, and to aid visualization,
+we divide resulting tree into 4 parts: a BA.5 subtree of 187 Pangos,
+a Delta subtree of 104 Pangos, a BA.2 subtree (excluding BA.5 descendants) of 118 Pangos,
+and a base tree (i.e. excluding all descendants of Delta and BA.2) of 405 Pangos,
+Figures \ref{fig:tanglegram_base}-\ref{fig:tanglegram_ba5} show these trees compared
+against the equivalent dated UShER topologies.
+The order of tips in each plot was chosen to reduce tangling,
 as calculated using the NeighborNet algorithm from Dendroscope \cite{Scornavacca2011,Huson2012},
 with tanglegrams plotted using the SVG outputting facility built into tskit.
-The entire pipeline is saved as a notebook at URL (?).
-
-The majority of resolvable nodes, including most high-level Scorpio groupings,
-are shared between the ARG and the UShER tree (magenta nodes).
+See \href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/supp\_pdf-cophylogeny-tanglegrams.ipynb}{supp\_pdf-cophylogeny-tanglegrams.ipynb}
+for details and the pipeline used.
+
+Inspection of Figures~\ref{fig:tanglegram_base}-\ref{fig:tanglegram_ba5} reveals that the majority of resolvable nodes,
+including most high-level Scorpio groupings,
+are shared between the ARG and the UShER tree (magenta nodes), and there is extremely close (usually perfect) agreement on topological relations between the Pango lineages.
 In the few cases where the trees differ, this is often in terms of degree of resolution of polytomies.
@@
 as in the base of the BA.2 subtree, and
-sometimes the UShER tree shows greater resolution, as in BA.2.9.
+sometimes the UShER tree shows greater resolution, as in the close relatives of Beta (B.1.351).
 
@@
 % See https://github.com/jeromekelleher/sc2ts-paper/issues/628
-Mapping deletions onto the ARG using post-hoc parsimony
-allowed us to identify important deletions and
-to clarify their nature in SARS-CoV-2 evolution.
+Although uncommon relative to single nucleotide mutations,
+deletions are an important component of SARS-CoV-2 evolution,
+contributing to the emergence of VOCs by enhancing,
+e.g., efficiency for cell entry \cite{Cantoni2022} and
+potential for antibody escape \cite{McCarthy2021}.
+Using the ARG, we identified major deletion events
+(which are propagated to many descendants)
+as well as recurrent deletions
+(which occur on at least two branches) in the ARG.
 In Table~\ref{tab:major_dels},
-we list a subset of deletion events in the ARG
-which are inherited by at least 10,000 samples.
-We consider these deletions to be major events
-in the evolutionary history of SARS-CoV-2.
-Most of these deletion events
-(besides the 1-base deletion at 28,271
-inherited by 45\% of the samples)
-are associated with the origin nodes of major VOCs in the ARG
-(indicated by the Pango lineages assigned to them,
-e.g., node 86456 is the Alpha origin node;
-also see Figure~\ref{fig:subgraph_alpha}).
-Most of them are characteristic mutations
-that are linked to the emergence of these VOCs
-(see the references in Table~\ref{tab:major_dels}
-pointing to the analyses
-which originally identified them as characteristic mutations).
-They are also highly prevalent
-(all except two events in this table
-are inherited by at least 11\% of the samples).
-Notably, eight of these deletion events occur
-in the N-terminal domain of the S1 unit of Spike (S1-NTD),
-a known hotspot of recurrent deletions \cite{McCarthy2021};
-These are all highly recurrent (Table~\ref{tab:rec_dels}), and
-include known recurrent deletions, for example,
-a 6-base deletion at 21,765 \cite{Meng2021} and
-a 3-base deletion at 21,991 \cite{McCarthy2021}.
+we list the major deletion events in the ARG
+which are each inherited by at least 10,000 descendant samples; and
+in Table~\ref{tab:rec_dels},
+we list the recurrent deletions in the ARG
+which are each observed by at least 10,000 (0.4\%) samples.
+Most of the deletions in these tables
+occur in coding regions
+(except the 1-base deletions at 28,271) and
+do not cause frame shifts, involving 3, 6, or 9 sites
+(except the 4-base deletion and the two 5-base deletions,
+which are likely artefacts; see below).
+All the deletions in these tables
+(besides the artefactual cases)
+have been linked to the emergence of major VOCs
+(Alpha \cite{RambautAlpha2020},
+Delta \cite{Stern2021},
+Omicron B.1.1.529 \cite{RambautOmicron2021},
+BA.1 \cite{RambautOmicron2021},
+BA.2 \cite{RambautOmicron2021}, and
+BA.4 \cite{Tegally2022})
+and/or reported to be prevalent
+in previous studies of deletions \cite{Li2023indels}.
+All these major deletion events
+(again, besides the artefactual cases)
+are associated with the origins of major VOCs in the ARG
+(see the Alpha subgraph in Figure~\ref{fig:subgraph_alpha};
+the Delta subgraph in Figure~\ref{fig:subgraph_delta}; and
+the Omicron subgraph in Figure~\ref{fig:subgraph_omicron}),
+except the 1-base deletion at 28,271 on node 1436808
+and the 9-base deletion at 686 on the BA.4-labeled node 2698748
+(Table~\ref{tab:major_dels}).
+Most of these deletions have been identified
+as characteristic mutations of the major VOCs
+in the studies or analysis cited in Table~\ref{tab:major_dels}.
+Notably, eight of the major deletion events occur
+in the N-terminal domain of the S1 unit of Spike
+(a known hotspot of recurrent deletions \cite{McCarthy2021}
+that plays a substantial role in antigenicity \cite{Harvey2021}),
+which are all highly recurrent in the ARG
+(Table~\ref{tab:rec_dels}).
+These include known recurrent deletions:
+the 6-base deletion at 21,765 and
+the 3-base deletion at 21,991,
+which are associated with
+increased efficiency in cell entry \cite{Meng2021} and
+antibody escape \cite{McCarthy2021}, respectively.
+In addition, we found recurrent deletions in ORF8 and NSP3,
+which have been suggested to be indel hotspots
+\cite{Akaishi2022}.
+Together, these findings demonstrate that
+the ARG captures genuine deletion events
+in SARS-CoV-2 evolution.
+A comprehensive evolutionary analysis of
+the deletion events in the ARG
+is beyond the scope of the current study;
+however, we believe that the ARG, as a resource,
+will enable and facilitate future studies
+to better understand the role of deletions
+in SARS-CoV-2 evolution
+(alongside recombination and single nucleotide mutations),
+which to-date has been understudied
+in great part due to
+a lack of suitable data models, methods, and tools.
 
@@
 one at 11,283 and the other at 11,288,
-which are commonly found in BA.1 and BA.2 samples, respectively.
+which are found in
+339,810 BA.1 samples and 843,291 BA.2 samples
+in the Viridian v04 dataset, respectively.
 Assuming that the alignments are correct,
@@
 as the 4-base deletion event at 11,288.
-Furthermore, sequencing errors or sequence misalignments
-can also cause artefactual deletions, for example,
-the out-of-frame 4-base recurrent deletion at 21,767 in S1-NTD
-(Table~\ref{tab:rec_dels}).
-Future improvements in deletion remapping and data pre-processing
+Future improvements in deletion remapping
 should remove such artefactual deletions.
@@
 are inherited by at least 10,000 samples.
-The frequencies of these deletions in the ARG are shown.
-The Pango lineage assigned to the node
-where an event occurs is shown.
-The references point to analyses
-which identified a deletion
-to be associated with a major VOC assigned to its node.
+These deletions are characterized by
+a site position in the genome (start) and
+the number of subsequence sites involved (length), and
+they are annotated by the gene in which they occur
+(non-coding otherwise).
+The IDs of the nodes in the ARG
+above which the deletions occur and
+the Pango lineages assigned to these nodes are indicated.
+The percentage of samples inheriting each deletion
+is shown as frequencies.
+If a deletion has been identified
+as a characteristic mutation of a VOC by a study (or an analysis),
+a reference to the study (or the analysis) is provided.
 Abbreviations: NSP, non-structural protein;
@@
 \caption{
-Recurrent deletions.
-These deletions occur at least 5 times in the ARG and
-are found in at least 1,000 samples.
+Recurrent deletions in the ARG.
+These deletions occur on at least two branches in the ARG and
+are found in the alignments of at least 10,000 (0.4\%) samples
+in the Viridian v04 dataset.
+For each deletion, the number of occurrences in the ARG and
+the number of samples where it is observed are shown.
+For further explanation,
+see the caption of Table~\ref{tab:major_dels}.
 Abbreviations:
@@
 11283 & 9 & ORF1ab / NSP6 & 13 & 339810\\
-21767 & 4 & S / S1-NTD & 10 & 2144\\
 28362 & 9 & N & 7 & 850483\\
@@
 \subsubsection{Alpha}
-The B.1.1.7 (Alpha) variant is the first major VOC, and
-it is a saltational lineage
-characterized by a long branch preceding its origin.
-Figure \ref{fig:subgraph_alpha} shows that the origin of Alpha is well reconstructed in the ARG.
+\label{supp:alpha}
+The B.1.1.7 (Alpha) variant emerged in the UK as the first major VOC.
+Is well reconstructed in the ARG with Figure \ref{fig:subgraph_alpha} showing it to be a saltational lineage characterised by a long branch preceding its origin.
 Most of the characteristic Alpha mutations (\cite{RambautAlpha2020})
@@
 \subsubsection{Delta}
+\label{supp:delta}
+
  \ref{fig:subgraph_delta}
@@
 \subsubsection{Omicron BA.1 and BA.2}
-XX
+\label{supp:omicron}
+
+\ref{fig:subgraph_omicron}
 
@@
 on the left parent (x-axis) and the right parent (y-axis).
-The shaded regions highlight artifactual recombination events,
+The shaded regions highlight potentially artifactual recombination events,
 which have fewer than 4 net supporting loci
@@
 Recombination events associated with
-the origins of Pango X lineages are labeled.
+the origins of Pango X and Jackson\cite{Jackson2021} lineages are labeled.
 }
@@
 Sites associated with multiple mutations in the subgraph have their mutations assigned a unique fill colour (orange, blue, green, red, etc); reversions are further highlighted with a black outline.
-Sc2ts seed samples (see STAR methods) for B.1.617, B.1.617.1, and B.1.617.2 (Delta) are plotted as red squares.
+Sc2ts seed samples (see STAR methods) for B.1.617, B.1.617.1, and B.1.617.2 are plotted as red squares.
 }
@@
 \caption{
-Subgraph of the BA.1 and BA.2 Omicron lineages. BA.1 samples are shown 
-[Some notes to help the reader know this 
-reflects known events]
+Subgraph illustrating the saltational origin of the major Omicron lineages BA.1 and BA.2.
+Symbol and multiple-colouring schemes as in \ref{fig:subgraph_delta}, except BA.1 samples are filled in light blue (also used to outline BA.1 specific mutations), BA.2 samples in light yellow (also used to outline BA.2 specific mutations), and B.1.1.529 mutations not seen in either BA.1 or BA.1 are outlined in green.
+These colours correspond to those used in the Pango designation issue\cite{RambautOmicron2021}.
 }

### 2025-10-03 latex/main.tex
Updates from Overleaf

@@
 as described in \ref{sec:supplement_pango_lineages}, and
-omit Pango designations starting with "X"
+omit Pango designations starting with ``X''
 to avoid recombinants, leaving 1329 Pango lineages.
@@
 as well as 43 samples that are descendants of the recombinant XBB
-but which do not start with an "X".
+but which do not start with an ``X''.
 This results in a simplified ARG of 1282 samples,
@@
 correspond to those on the far right of the genome identified as
-artefactual in the \ref{sec:spurious-ancestrally-important-recombinants} section.
+artefactual in the FIXME Xref \ref{sec:spurious-ancestrally-important-recombinants} section.
 We therefore compare the Usher topology against the first tree in the ARG (covering
@@
 which are each observed by at least 10,000 (0.4\%) samples.
+[FIXME define how we measured this].
+
 Most of the deletions in these tables
@@
 which have been suggested to be indel hotspots
+[FIXME broken reference]
 \cite{Akaishi2022}.
@@
 to better understand the role of deletions
-in SARS-CoV-2 evolution
-(alongside recombination and single nucleotide mutations),
+(alongside recombination and single nucleotide mutations)
+in SARS-CoV-2 evolution,
 which to-date has been understudied

### 2025-10-04 arg_postprocessing/scripts/process_ripples.py
Loosen restriction on what nodes we consider

@@ -64,7 +64,7 @@ def run(ripples_dir, sc2ts_ts, output):
             continue
         seen_descendants.add(desc)
         event = UsherEvent(usher_node, desc, {"unclassified": -1})
-        if len(desc) < 10_000:
+        if len(desc) < 100_000:
             nodes = [sample_id_to_node[sid] for sid in desc]
             if len(nodes) == 1:
                 mrca = int(nodes[0])
@@ -81,6 +81,8 @@ def run(ripples_dir, sc2ts_ts, output):
             event.sc2ts_closest_recombinant = u
             if u != -1:
                 event.sc2ts_closest_recombinant_path_len = tree.path_length(mrca, u)
+        else:
+            print(f"Skipping classification for node with {len(desc)} descendants")
 
         events.append(event)
 

### 2025-10-06 latex/main.tex
Updates from Overleaf

@@
 we list the recurrent deletions in the ARG
-which are each observed by at least 10,000 (0.4\%) samples.
-[FIXME define how we measured this].
+which are each observed in at least 10,000 (0.4\%) samples
+% [FIXME define how we measured this].
+(i.e., for a deletion caused by multiple events in the ARG,
+the sum of all the samples inheriting the deletion
+across all the events).
 
@@
 are inherited by at least 10,000 samples.
-These deletions are characterized by
+The deletions are characterized by
 a site position in the genome (start) and
-the number of subsequence sites involved (length), and
+the number of sites involved (length);
 they are annotated by the gene in which they occur
@@
 above which the deletions occur and
-the Pango lineages assigned to these nodes are indicated.
-The percentage of samples inheriting each deletion
-is shown as frequencies.
+the Pango lineages assigned to these nodes are shown.
+Frequency is the percentage of samples inheriting a deletion.
 If a deletion has been identified
-as a characteristic mutation of a VOC by a study (or an analysis),
+as a characteristic mutation of a VOC
+by a study (or an analysis),
 a reference to the study (or the analysis) is provided.

### 2025-10-06 src/oldest_node.py
Node time check notebook

@@ -1,0 +1,102 @@
+# -*- coding: utf-8 -*-
+"""
+This script parses a Nextstrain NEXUS tree file to find the oldest
+node for each defined clade and extracts its metadata.
+"""
+
+import csv
+import re
+import io
+from Bio import Phylo
+
+def parse_nexus_comments(comment_str):
+    if not comment_str:
+        return {}
+    pattern = re.compile(r'([^,=&]+)=({[^}]+}|[^,]+)')
+    attributes = dict(pattern.findall(comment_str))
+    return attributes
+
+def find_oldest_clades(tree_file):
+    try:
+        with open(tree_file, 'r', encoding='utf-8') as f:
+            nexus_content = f.read()
+
+        # To handle "(", ")" inside metadata
+        def sanitize_comments(match):
+            return match.group(0).replace('(', '_').replace(')', '_')
+
+        sanitized_content = re.sub(r'\[&[^\]]*\]', sanitize_comments, nexus_content)
+        handle = io.StringIO(sanitized_content)
+        tree = Phylo.read(handle, 'nexus')
+        print("Successfully parsed the tree file.")
+
+    except FileNotFoundError:
+        print(f"Error: File '{tree_file}' not found.")
+        return None
+    except Exception as e:
+        print(f"Error reading or parsing the tree file: {e}")
+        return None
+
+    oldest_clades = {}
+    nodes_processed = 0
+
+    for node in tree.find_clades():
+        nodes_processed += 1
+        if node.comment:
+            attributes = parse_nexus_comments(node.comment)
+
+            clade_name = attributes.get('clade_membership')
+            date_str = attributes.get('num_date')
+            node_id = node.name if node.name else 'NA'
+
+            if clade_name and date_str:
+                try:
+                    date = float(date_str)
+                    
+                    if clade_name not in oldest_clades or date < oldest_clades[clade_name]['date']:
+                        oldest_clades[clade_name] = {
+                            'node_id': node_id,
+                            'date': date,
+                            'date_CI': attributes.get('num_date_CI', 'NA')
+                        }
+                except (ValueError, TypeError):
+                    continue
+    
+    return oldest_clades
+
+def write_output_tsv(data, output_file):
+    if not data:
+        print("No clade data found.")
+        return
+        
+    sorted_clade_names = sorted(data.keys())
+
+    try:
[…]

### 2025-10-06 latex/main.tex
Work in progress

@@
 
-Using sc2ts, we built an ARG from 2,482,157 samples from the Viridian v04
-dataset\cite{Hunt2024}, collected between 2020-01-01 and 2023-02-20.
+Using sc2ts, we built an ARG from 2,482,157 samples
+from the Viridian v04 dataset \cite{Hunt2024},
+collected between 2020-01-01 and 2023-02-20.
 Inference took 21 days on a XX core server with 500 GB of RAM
@@
 Of the nodes, 855 represent recombination events,
-resulting in 316 distinct trees along the genome (some
-events occured at the same breakpoints [vague, rephrase]).
+resulting in 316 distinct trees along the genome
+(some events occurred at the same breakpoints [vague, rephrase]).
 [TODO note about loading up the time it takes to get the node 
@@
 the location of the inferred breakpoint interval relative to the Spike gene.
-For simplicity, ancestral lineages above the recombination parents or the MRCA are omitted.
+For simplicity, ancestral lineages above the recombinant parents or the MRCA are omitted.
 (A) Subgraph around the recombination node associated with XA.
-Two ``Group A'' recombinant samples reported by Jackson 
-et al.\cite{Jackson2021} are shown.
+Two ``Group A'' recombinant samples reported by Jackson et al.\cite{Jackson2021} are shown.
 (B) Subgraph around the recombination node associated with a 
@@
 are nodes exactly matched by sampled sequences. The descendants column
-shows the descendant samples broken down by Pango lineage (grouped 
-by root lineage, where appropriate).[FIXME Typesetting to be finished,
-describe cols for Type III and IV]
+shows the descendant samples broken down by Pango lineage
+(grouped by root lineage, where appropriate).
+[FIXME Typesetting to be finished, describe cols for Type III and IV]
 }
@@
 % 146
-Although the process of incorporating recombinant lineages in the Pango 
-nomenclature system is well defined,
-the identification of such events is much more subjective, requiring
-the synthesis of multiple lines of evidence and often the 
-manual inspection of phylogenies and sequence alignments
+Although the process of incorporating recombinant lineages
+in the Pango nomenclature system is well defined,
+the identification of such events is much more subjective,
+requiring the synthesis of multiple lines of evidence and
+often the manual inspection of phylogenies and sequence alignments
 by volunteers~\cite{Smith2023tracking}.
 
-By automatically detecting and incorporating recombination events
-into the ARG, sc2ts provides us with a systematic means of 
+By automatically detecting and incorporating recombination events into the ARG,
+sc2ts provides us with a systematic means of
 discovering and assessing evidence around these events.
-Table~\ref{tab:pango_x_lineages} summarises
-these results for the 44 Pango X lineages present in
-the ARG (XAC and XAD have been omitted, and XM 
-partially represented, corresponding to 23 samples in total; 
+Table~\ref{tab:pango_x_lineages} summarizes
+these results for the 44 Pango X lineages present in the ARG
+(XAC and XAD have been omitted, and XM partially represented,
+corresponding to 23 samples in total;
 see Document~\ref{sec:supplement_x_pango_lineages}).
-For 16 Pango X lineages the oldest node that has been assigned
-the focal Pango lineage is a recombination node, and that node 
-is the ancestor of all samples assigned that (or a descendant) Pango 
-lineage. These are the Type I events, in which the ARG reflects the
+For 16 Pango X lineages, the oldest node
+that has been assigned to the focal Pango lineage is a recombination node, and
+that node is the ancestor of all samples assigned
+to that (or a descendant) Pango lineage.
+These are the Type I events, in which the ARG reflects the
 expected phylogenetic structure of Pango designation.
@@
 (STAR Methods; Table~\ref{tab:method_concordance}).
-Of the Type I events, 13 are monophyletic for the focal
-pango (i.e., the descendants are entirely from that lineage); 
-Figure~\ref{fig:subgraphs}A shows XA as an example (see also 
-Document~\ref{sec:jackson_recombs} for further analysis 
-of the Jackson et al.\ recombinants).
+Of the Type I events, 13 are monophyletic for the focal Pango lineage
+(i.e., the descendants are entirely from that lineage);
+Figure~\ref{fig:subgraphs}A shows XA as an example
+(see also Document~\ref{sec:jackson_recombs}
+for further analysis of the Jackson et al.\ recombinants).
 
@@
 and Document S2 for ARG visualisations.
+
+\subsection*{Comparison with RIPPLES}
+
+\begin{table}
+\caption{RIPPLES pango X lineage events.} 
+\label{tab:ripples_x_lineages}
+\centering
+% \begin{scriptsize}
+\begin{tabular}{llrr}
+\toprule
+\multicolumn{4}{c}{Detected at p3 and p4}\\
+ripples & sc2ts & usher descs& scts descs \\
+\midrule
+XBR & XBR & 1 & 1 \\
+XAC & novel & 18 & 18 \\
+XC & XC & 5 & 5 \\
+XA & XA & 39 & 39 \\
+XS & XS & 17 & 17 \\
+XM & XM3 & 1 & 1 \\
+\midrule
+\multicolumn{4}{c}{Detected at p3}\\
+\midrule
+XY & XY & 23 & 23 \\
+XE & novel & 1113 & 1116 \\
+XL & XL & 64 & 64 \\
+XAM & XQ & 1 & 1 \\
+XW & XW & 32 & 32 \\
+XBB & XBB & 6452 & 6453 \\
+XAM & XQ & 21 & 21 \\
+XAE & novel & 9 & 9 \\
+\end{tabular}
+% \end{scriptsize}
+\end{table}
 
@@
 mutations ``averted'' by recombination---is a measure of the evidence
-in favour of recombination. 
+in favor of recombination. 
 Table~\ref{tab:pango_x_lineages} shows
@@
 
+\subsubsection*{Checking inferred node dates}
+
+To validate the accuracy of times for non-sample internal nodes inserted 
+into the ARG
+(inferred using tsdate), we compared the dates of first emergence of
+Nextstrain clades against the Nextstrain SARS-CoV-2 tree (downloaded on 
+2023-01-21). Figure \ref{fig:pango_vs_nextstrain_node_times} shows relatively
+close agreement for most clades, with only three 
+disagreeing by more than 28 days: 20C (an early lineage from 2020), 20E (Pango B.1.177 also known as EU1, emerging mid-2020), and 21B (Pango B.1.617.1 also known as Kappa, emerging in late 2020 to early 2021).
+
+Then, to gauge internal consistency of the time estimation procedure, we
+selected a subset of sample nodes, matching the distribution of node times 
+and average number of children of the non-sample internal nodes. These
+selected sample nodes were then re-dated, and the inferred dates compared 
+against their recorded collection dates. The results in Figure 
+\ref{fig:node_time_testing} demonstrate that although the inferred and true
+times are generally very close, there is a clear bias with re-estimated times
+being consistently older (by an average of 37 days). This is in line with 
+the findings of other recent
+work \cite{sanderson2021chronumental}, and is likely to be a consequence
+of edges carrying more mutations than would be expected under a neutral
+molecular clock prior (either due to errors in tree-building or genuinely varying mutation rates), as well as inaccuracies in the recording of 
+sample dates.
+
 \subsubsection*{Comparison of mutational spectra}
@@
 (i.e., for a deletion caused by multiple events in the ARG,
-the sum of all the samples inheriting the deletion
-across all the events).
+the sum of all the samples inheriting the deletion across all the events).
 
@@
 Together, these findings demonstrate that
-the ARG captures genuine deletion events
-in SARS-CoV-2 evolution.
-A comprehensive evolutionary analysis of
-the deletion events in the ARG
+the ARG captures genuine deletion events in SARS-CoV-2 evolution.
+A comprehensive evolutionary analysis of the deletion events in the ARG
 is beyond the scope of the current study;
@@
 will enable and facilitate future studies
-to better understand the role of deletions
-(alongside recombination and single nucleotide mutations)
-in SARS-CoV-2 evolution,
-which to-date has been understudied
-in great part due to
+to better understand the role of deletions in SARS-CoV-2 evolution
+(alongside recombination and single nucleotide mutations),
+which to-date has been understudied in great part due to
 a lack of suitable data models, methods, and tools.
@@
 
-\subsection{Major saltation lineages}
+\subsection{Origins of major VOCs}
 
@@
 \subsubsection{Alpha}
-\label{supp:alpha}
+\label{sec:alpha}
+
 The B.1.1.7 (Alpha) variant emerged in the UK as the first major VOC.
-Is well reconstructed in the ARG with Figure \ref{fig:subgraph_alpha} showing it to be a saltational lineage characterised by a long branch preceding its origin.
+Its origin is well reconstructed in the ARG,
+with Figure \ref{fig:subgraph_alpha}
+showing it to be a saltational lineage
+characterized by a long branch preceding its origin.
 Most of the characteristic Alpha mutations (\cite{RambautAlpha2020})
@@
 of which we display one (ERR4413600)
-with no associated mutations, meaning that it has an identical sequence to the inferred ancestor of Alpha.
+with no associated mutations,
+meaning that it has an identical sequence
+to the inferred ancestor of Alpha.
 Below the root of Alpha, a single sample (ERR5178290)
@@
 \subsubsection{Delta}
-\label{supp:delta}
-
- \ref{fig:subgraph_delta}
+\label{sec:delta}
+
+The B.1.617.2 lineage (Delta) was first detected in India \cite{ECDCreport}.
+The subgraph in Figure~\ref{fig:subgraph_delta} shows
+a mutation-rich branch leading to the Delta root,
+which contains 11 single nucleotide mutations and a 6-base deletion.
+The placement of this branch in the ARG
+is a result of long branch splitting of the lineage above B.1.617.
+
+Delta belongs to the “B.1.617” clade,
+which also includes Kappa (B.1.617.1) and B.1.617.3 \cite{ECDCreport}.
+Previous phylogenetic analyses suggest that
+Delta and B.1.617.3 are more closely related to each other
+than either is to Kappa (\cite{Cherian2021,Farinholt2021}).
+In the ARG, however, Delta and Kappa appear as sibling lineages
+(see the subgraph in Figure~\ref{fig:subgraph_delta}).
+This reflects the closest approximation to the relationships
+suggested in the previous studies that could be inferred,
+given that no B.1.617.3 samples passed our QC filters and
+no seed B.1.617.3 samples were specified for ARG inference.
+
+In the ARG, all six of the shared Delta-Kappa mutations (red)
+occur above the B.1-labeled common ancestor
+of Kappa (left branch) and Delta (right branch).
+All nine of the Kappa-specific mutations (blue)
+occur above the B.1.617-labeled node leading to Kappa samples.
+Seven out of nine Delta-specific mutations (purple)
+occur above the Delta root
+(the other two are above the MRCA of Kappa and Delta).
+We further examined the mutations associated with two major branches of Delta.
+Using a phylogenetic analysis, Stern et al. \cite{Stern2021} suggest that
+the Delta sublineages can be classified into distinct clades and
+identified the characteristic mutations of these clades.
+In the Nextstrain phylogeny shown in Stern et al. \cite{Stern2021},
+the first major split separates clades A to D from clade E;
+this topology is reflected in the ARG.
+All three of the clade A-D mutations
+(G15451A, C16466T, and a 6-base deletion at 28,248)
+occur on the uppermost branch of Delta (which corresponds to clades A to D).
+There are ten clade E mutations:
+five occur on the right branch of Delta (which corresponds to clade E) and
+three (C1191T, C5184T, and C28253T) occur above the Delta root.
+The placement of these mutations in the ARG generally reflects
+what is proposed in Stern et al. \cite{Stern2021}.
+
+There are four positions with multiple mutations
+(orange at position 1267, blue at 210, green at 20396, and red at 25276).
+Of these, the only clear topological change
+that could make the graph more parsimonious would be
+to attach the node above SRR12316669 one position higher up the tree,
+which would allow the placement of two recurrent red mutations
+(a C25276T above SRR12316669 and another above ERR4561562 / B.1.617)
+rather than one C25276T mutation followed by two reversions.
+The other recurrent mutations can only be made more parsimonious
+by topological rearrangements that would cause recurrence at other sites.
+We therefore believe that the inferred Delta origins
+in the sc2ts ARG are largely correct.
 
 \subsubsection{Omicron BA.1 and BA.2}
-\label{supp:omicron}
-
-\ref{fig:subgraph_omicron}
+\label{sec:omicron}
+
+The B.1.1.529 lineage (Omicron)
+is believed to have emerged in southern Africa, and
+it is a saltational lineage with two highly successful sublineages,
+BA.1 and BA.2 \cite{Dor2025}.
+BA.1 and BA.2 are genetically distinct from each other and
+from their parental B.1.1.529 lineage \cite{RambautOmicron2021}.
+In the subgraph shown in Figure~\ref{fig:subgraph_omicron},
+BA.1 and BA.2 differ from the non-Omicron B.1.1 ancestor
+(above the B.1.1.529-labeled MRCA of BA.1 and BA.2)
+by 27 mutations (including two deletions),
+reflecting its saltational origin.
+The ARG captures the split between BA.1 and BA.2
+which has been shown in previous phylogenetic analysis \cite{Dor2025}.
+The 5-base deletion highlighted in green above the BA.2 root
+is an artefact of our naive deletion remapping procedure,
+which splits two 9-base deletions in BA.1 and BA.2 into three deletions
+(one in green at 11288, one in blue at 11283, and one in orange 11292;
+see Document~\ref{sec:supplement_dels}).
+Also, there is uncertainty in the sequence alignments
+leading to ambiguity in the start position of these deletions
+(if these deletions occur at 11283,
+there would be a single deletion event before the BA.1/BA.2 split),
+as noted in the Pango designation issue \cite{RambautOmicron2021}.
+
+We selected a BA.1 seed sample that led to the fewest mutations.
+This strategy resulted in five mutations
+highlighted in blue/orange/green/red/purple
+which are absent in the chosen
+but present in its inferred descendants.
+In retrospect, SRR17051902 should have been a better choice for a seed,
+because it would have resulted in a topology
+where those five mutations would be placed
+above the split between BA.1 and BA.2.
+
+\subsection{Sc2ts detection results of rejected Pango X lineages}
+\label{sec:supplement_rejected_pango_x_lineages}
+
+Ten Pango X lineages which are present in the Viridian v04 dataset
+were not integrated into the ARG
+due to high HMM costs and/or insufficient sample support,
+but their samples were matched against the ARG during inference.
+% https://github.com/jeromekelleher/sc2ts-paper/issues/366
+Table~\ref{tab:extra_hmm_results} summarizes
+the HMM matching results of these Pango X lineages.
+All were identified as recombinants by sc2ts, except XBL.
+(For XBL, the evidence for recombination
+proposed in the Pango designation issue \cite{PangoIssueXBL}
+involves only three informative sites
+supporting the suggested rightmost parent, and
+therefore this likely fell below our threshold of detection.)
+% https://github.com/jeromekelleher/sc2ts-paper/issues/379
+Importantly, these results demonstrate that sc2ts can detect
+recombinants with multiple breakpoints (XAY, XBC, and XBT).
+% https://github.com/jeromekelleher/sc2ts-paper/discussions/776
+% https://github.com/jeromekelleher/sc2ts-paper/discussions/788
+XBC and XBT are relatively clean examples of multiple-breakpoint recombinants,
+with the Pango lineages of the parents and the breakpoint intervals
+concordant with those suggested in the Pango designation issues,
+but XAY is more complex, possibly involving up to six breakpoints as per sc2ts.
+Exploring how to integrate these and other recombinants
+with complex recombination patterns \cite{Roemer2023}
+should be an important direction for future work.
+
+\begin{table}
+\centering
+\caption{
+Sc2ts detection results for Pango X lineages not added to the ARG.
+The number of samples in Viridian v04 dataset
+assigned to these Pango X lineages,
+the number of samples which passed sequence-level quality control, and
+the number of recombination breakpoints
+in the Viterbi solutions of the samples
+are shown.
+}
+\label{tab:extra_hmm_results}
+\begin{tabular}{crrr}
+\toprule
+Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
+\midrule
+XBJ & 2 & 1 & 1\\
+XBP & 8 & 2 & 1\\
+XBS & 19 & 15 & 1\\
+XBW & 1 & 1 & 1\\
+XCA & 11 & 4 & 1\\
+XAK & 2 & 2 & 1\\   % Multiple breakpoint Pango X
+XAY & 15 & 11 & 6\\ % Multiple breakpoint Pango X
+XBC & 36 & 21 & 3\\ % Multiple breakpoint Pango X
+XBL & 2 & 2 & 0\\   % Multiple breakpoint Pango X
+XBT & 1 & 1 & 2\\   % Multiple breakpoint Pango X
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 
-\subsection{Sc2ts recombination detection results for additional Pango X lineages}
-
-\begin{table}
-\centering
-\caption{
-Sc2ts recombination detection results for Pango X lineages which were not added to the ARG.
-}
-\label{tab:absent_pango_x}
-\begin{tabular}{crrr}
-\toprule
-Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
-\midrule
-XBJ & 2 & 1 & 1\\
-XBP & 8 & 2 & 1\\
-XBS & 19 & 15 & 1\\
-XBW & 1 & 1 & 1\\
-XCA & 11 & 4 & 1\\
-XAK & 2 & 2 & 1\\
-XAY & 15 & 11 & 6\\
-XBC & 36 & 21 & 3\\
-XBL & 2 & 2 & 0\\
-XBT & 1 & 1 & 2\\
-\bottomrule
-\end{tabular}
-\end{table}
-
 \subsection{Analysis of the Jackson et al. recombinants}
@@
 
-Jackson et al.\cite{Jackson2021} conducted detailed analyses of early recombinants
-involving B.1.1.7 (Alpha), including the first designated Pango X lineage, XA.
+Jackson et al.\cite{Jackson2021} conducted a detailed analysis of
+early recombinants involving B.1.1.7 (Alpha),
+including the first designated Pango X lineage, XA.
 The authors first performed a targeted search for candidate recombinants by
@@
 four groups of recombinants (named A to D) and four singleton recombinants.
-Of the 16 samples analyzed by Jackson et al.\
-12 were included in the ARG;
+Of the 16 samples analyzed by Jackson et al., 12 were included in the ARG;
 the remaining 4 were excluded
@@
 because of a high HMM cost.
-Groups A-D are summarized in Table~\ref{tab:jackson} and show excellent concordance
+Groups A-D are summarized in Table~\ref{tab:jackson} and
+show excellent concordance
 in terms of the parental Pango lineages and breakpoint intervals.
-All four groups are directly associated with strongly supported recombination events
-in the ARG in terms of the number of averted mutations (A: 16, B: 6, C: 30, D: 18).
+All four groups are directly associated with
+strongly supported recombination events in the ARG
+in terms of the number of averted mutations (A: 16, B: 6, C: 30, D: 18).
 In the ARG, the singleton CAMC-CBA018 has
 the parents B.1.177+B.1.1.7 and
-the breakpoint interval 16,177--20,133,
-and is strongly supported by 14 averted mutations.
-In the Jackson et al.\ results, this singleton is
+the breakpoint interval 16,177--20,133, and
+is strongly supported by 14 averted mutations.
+In the Jackson et al. results, this singleton is
 almost identical with the same parent lineages and
@@
 again strongly supported by 11 averted mutations.
-In the Jackson et al.\ results,
+In the Jackson et al. results,
 this lineage has the same parents,
@@
 \centering
+\includegraphics[width=0.85\linewidth]{pango_vs_nextstrain_node_times.pdf}
+\caption{
+Inferred times of internal nodes corresponding to origins of Nextclade clades ($y$-axis) against those estimated in the Nextclade tree ($x$-axis).
+Confidence intervals shown as horizontal bars.
+Clades are labelled where dates differ by more than 28 days.
+}
+\label{fig:pango_vs_nextstrain_node_times}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.85\linewidth]{node_time_testing.pdf}
+\caption{
+Heat map showing differences (in days) between actual and re-estimated sample dates for a subset of internal sample nodes. 
+}
+\label{fig:node_time_testing}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics[width=0.85\linewidth]{subgraph-Alpha.pdf}
@@
 \centering
-\includegraphics[width=0.85\linewidth]{subgraph-Delta.pdf}
+\includegraphics[width=0.85\linewidth]{subgraph-Delta_recolored.pdf}
 \caption{
-Subgraph illustrating the saltational origin of B.1.617.2 (Delta).
-Symbols as in \ref{fig:subgraph_alpha}, with characteristic Delta mutations outlined in pink.
-Sites associated with multiple mutations in the subgraph have their mutations assigned a unique fill colour (orange, blue, green, red, etc); reversions are further highlighted with a black outline.
-Sc2ts seed samples (see STAR methods) for B.1.617, B.1.617.1, and B.1.617.2 are plotted as red squares.
+Subgraph illustrating the origin of B.1.617.2 (Delta).
+Symbols are described in \ref{fig:subgraph_alpha}, except that 
+we highlight the characteristic mutations of Kappa and Delta:
+those in blue which are listed
+in the Pango designation issue \cite{PangoIssueKappa}
+or the Kappa constellation \cite{ScorpioKappa},
+but absent in the Delta constellation;
+those in purple which are listed
+in the Delta constellation \cite{ScorpioDelta}
+or are Delta-specific deletions reported in Stern et al. \cite{Stern2021},
+but absent in the Kappa constellation; and
+those in red which are listed
+in both the Kappa and Delta constellations.
+We also highlight
+in orange the mutations associated with the lineage containing clades A to D and
+in green those associated with lineage containing clade E.
+Sites associated with multiple mutations in the subgraph
+have their mutations assigned a unique fill color
+(orange, blue, green, red, etc);
+reversions are further highlighted with a black outline.
+Sc2ts seed samples (see STAR methods) for
+B.1.617, B.1.617.1, and B.1.617.2 are plotted as red squares.
 }
@@
 \centering
-\includegraphics[width=0.85\linewidth]{subgraph-Omicron.pdf}
+\includegraphics[width=0.85\linewidth]{subgraph-Omicron_recolored.pdf}
 \caption{
-Subgraph illustrating the saltational origin of the major Omicron lineages BA.1 and BA.2.
-Symbol and multiple-colouring schemes as in \ref{fig:subgraph_delta}, except BA.1 samples are filled in light blue (also used to outline BA.1 specific mutations), BA.2 samples in light yellow (also used to outline BA.2 specific mutations), and B.1.1.529 mutations not seen in either BA.1 or BA.1 are outlined in green.
-These colours correspond to those used in the Pango designation issue\cite{RambautOmicron2021}.
+Subgraph illustrating the saltational origin of
+the major Omicron lineages BA.1 and BA.2.
+Symbols are described in Figure~\ref{fig:subgraph_delta},
+except that BA.1 samples are filled in light blue
+(also used to outline BA.1 characteristic mutations),
+BA.2 samples in light yellow
+(also used to outline BA.2 characteristic mutations), and
+B.1.1.529 mutations
+which are neither BA.1 or BA.2 characteristic mutations are
+outlined in green.
+These colors correspond to those used
+in the Pango designation issue \cite{RambautOmicron2021}.
 }

### 2025-10-06 latex/main.tex
Note date

--- 2025-10-06 latex/main.tex
@@
 sometimes the UShER tree shows greater resolution, as in the close relatives of Beta (B.1.351).
+
+\subsubsection*{Checking inferred node dates}
+
+To validate the accuracy of times for non-sample internal nodes inserted 
+into the ARG
+(inferred using tsdate), we compared the dates of first emergence of
+Nextstrain clades against the Nextstrain SARS-CoV-2 tree (downloaded on 
+2023-01-21). Figure \ref{fig:pango_vs_nextstrain_node_times} shows relatively
+close agreement for most clades, with only three (20C, 20E and 21B)
+disagreeing by more than 28 days.
+
+Then, to gauge internal consistency of the time estimation procedure, we
+selected a subset of sample nodes, matching the distribution of node times 
+and average number of children of the non-sample internal nodes. These
+selected sample nodes were then re-dated, and the inferred dates compared 
+against their known, true sampling dates. The results in Figure 
+\ref{fig:node_time_testing} demonstrate that although the inferred and true
+times are generally very close, there is a clear bias with re-estimated times
+being consistently older (by an average of 37 days). This is in line with 
+the findings of other recent
+work \cite{sanderson2021chronumental}, and is likely to be a consequence
+of edges carrying more mutations than would be expected under a neutral
+molecular clock prior, as well as inaccuracies in the recording of 
+sample dates.
 
@@
 \centering
+\includegraphics[width=0.85\linewidth]{node_time_testing.pdf}
+\caption{
+Heat map showing differences (in days) between actual and re-estimated sample dates for a subset of internal sample nodes. 
+}
+\label{fig:node_time_testing}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.85\linewidth]{pango_vs_nextstrain_node_times.pdf}
+\caption{
+Inferred times of internal nodes corresponding to origins of Nextclade clades ($y$-axis) against those estimated in the Nextclade tree ($x$-axis).
+Confidence intervals shown as horizontal bars.
+Clades are labelled where dates differ by more than 28 days.
+}
+\label{fig:pango_vs_nextstrain_node_times}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics[width=0.85\linewidth]{subgraph-Alpha.pdf}

### 2025-10-07 latex/main.tex

@@
 \subsubsection{Delta}
-\label{supp:delta}
-
- \ref{fig:subgraph_delta}
+\label{sec:delta}
+
+The B.1.617.2 lineage (Delta) was first detected in India \cite{ECDCreport}.
+The subgraph in Figure~\ref{fig:subgraph_delta} shows
+a mutation-rich branch leading to the Delta root,
+which contains 11 single nucleotide mutations and a 6-base deletion.
+The placement of this branch in the ARG
+is a result of long branch splitting of the lineage above B.1.617.
+
+Delta belongs to the “B.1.617” clade,
+which also includes Kappa (B.1.617.1) and B.1.617.3 \cite{ECDCreport}.
+Previous phylogenetic analyses suggest that
+Delta and B.1.617.3 are more closely related to each other
+than either is to Kappa (\cite{Cherian2021,Farinholt2021}).
+In the ARG, however, Delta and Kappa appear as sibling lineages
+(see the subgraph in Figure~\ref{fig:subgraph_delta}).
+This reflects the closest approximation to the relationships
+suggested in the previous studies that could be inferred,
+given that no B.1.617.3 samples passed our QC filters and
+no seed B.1.617.3 samples were specified for ARG inference.
+
+In the ARG, all six of the shared Delta-Kappa mutations (red)
+occur above the B.1-labeled common ancestor
+of Kappa (left branch) and Delta (right branch).
+All nine of the Kappa-specific mutations (blue)
+occur above the B.1.617-labeled node leading to Kappa samples.
+Seven out of nine Delta-specific mutations (purple)
+occur above the Delta root
+(the other two are above the MRCA of Kappa and Delta).
+We further examined the mutations associated with two major branches of Delta.
+Using a phylogenetic analysis, Stern et al. \cite{Stern2021} suggest that
+the Delta sublineages can be classified into distinct clades and
+identified the characteristic mutations of these clades.
+In the Nextstrain phylogeny shown in Stern et al. \cite{Stern2021},
+the first major split separates clades A to D from clade E;
+this topology is reflected in the ARG.
+All three of the clade A-D mutations
+(G15451A, C16466T, and a 6-base deletion at 28,248)
+occur on the uppermost branch of Delta (which corresponds to clades A to D).
+There are ten clade E mutations:
+five occur on the right branch of Delta (which corresponds to clade E) and
+three (C1191T, C5184T, and C28253T) occur above the Delta root.
+The placement of these mutations in the ARG generally reflects
+what is proposed in Stern et al. \cite{Stern2021}.
+
+There are four positions with multiple mutations
+(orange at position 1267, blue at 210, green at 20396, and red at 25276).
+Of these, the only clear topological change
+that could make the graph more parsimonious would be
+to attach the node above SRR12316669 one position higher up the tree,
+which would allow the placement of two recurrent red mutations
+(a C25276T above SRR12316669 and another above ERR4561562 / B.1.617)
+rather than one C25276T mutation followed by two reversions.
+The other recurrent mutations can only be made more parsimonious
+by topological rearrangements that would cause recurrence at other sites.
+We therefore believe that the inferred Delta origins
+in the sc2ts ARG are largely correct.
 
 \subsubsection{Omicron BA.1 and BA.2}
-\label{supp:omicron}
-
-\ref{fig:subgraph_omicron}
+\label{sec:omicron}
+
+The B.1.1.529 lineage (Omicron) is believed to have emerged in southern Africa, and
+it is a saltational lineage with two highly successful sublineages, BA.1 and BA.2 \cite{Dor2025}.
+BA.1 and BA.2 are genetically distinct from each other and
+from their parental B.1.1.529 lineage \cite{RambautOmicron2021}.
+In the subgraph shown in Figure~\ref{fig:subgraph_omicron},
+BA.1 and BA.2 differ from the non-Omicron B.1.1 ancestor
+(above the B.1.1.529-labeled MRCA of BA.1 and BA.2)
+by more than ten mutations, including six deleted regions.
+The ARG captures the split between BA.1 and BA.2,
+which has been shown in previous phylogenetic analysis \cite{Dor2025}.
+The 5-base deletion highlighted in green above the BA.2 root
+is an artefact of our naive deletion remapping procedure,
+which splits two 9-base deletions in BA.1 and BA.2 into three deletions
+(one in green at 11288, one in blue at 11283, and one in orange 11292;
+see Document~\ref{sec:supplement_dels}).
+Also, there is uncertainty in the sequence alignments
+leading to ambiguity in the start position of these deletions,
+as noted in the Pango designation issue \cite{RambautOmicron2021}.
+
+We selected a BA.1 seed sample that led to the fewest mutations.
+This strategy resulted in five mutations highlighted
+in blue/orange/green/red/purple
+which the chosen seed does not have
+but are present in its inferred descendants.
+In retrospect, SRR17051902 should have been a better choice for a seed,
+because it would have resulted in a topology
+where those five mutations would be placed by parsimony
+above the split between BA.1 and BA.2.
 
@@
 \caption{
-Subgraph illustrating the saltational origin of the major Omicron lineages BA.1 and BA.2.
-Symbol and multiple-colouring schemes as in \ref{fig:subgraph_delta}, except BA.1 samples are filled in light blue (also used to outline BA.1 specific mutations), BA.2 samples in light yellow (also used to outline BA.2 specific mutations), and B.1.1.529 mutations not seen in either BA.1 or BA.1 are outlined in green.
-These colours correspond to those used in the Pango designation issue\cite{RambautOmicron2021}.
+Subgraph illustrating the saltational origin of
+the major Omicron lineages BA.1 and BA.2.
+Symbols are described in Figure~\ref{fig:subgraph_delta},
+except that BA.1 samples are filled in light blue
+(also used to outline BA.1 characteristic mutations),
+BA.2 samples in light yellow
+(also used to outline BA.2 characteristic mutations), and
+B.1.1.529 mutations
+which are neither BA.1 or BA.2 characteristic mutations are outlined in green.
+These colours correspond to those used
+in the Pango designation issue \cite{RambautOmicron2021}.
 }

### 2025-10-07 latex/main.tex

@@
 2023-01-21). Figure \ref{fig:pango_vs_nextstrain_node_times} shows relatively
-close agreement for most clades, with only three (20C, 20E and 21B)
-disagreeing by more than 28 days.
+close agreement for most clades, with only three 
+disagreeing by more than 28 days: 20C (an early lineage from 2020), 20E (Pango B.1.177 also known as EU1, emerging mid-2020), and 21B (Pango B.1.617.1 also known as Kappa, emerging in late 2020 to early 2021).
 
@@
 selected sample nodes were then re-dated, and the inferred dates compared 
-against their known, true sampling dates. The results in Figure 
+against their recorded collection dates. The results in Figure 
 \ref{fig:node_time_testing} demonstrate that although the inferred and true
@@
 of edges carrying more mutations than would be expected under a neutral
-molecular clock prior, as well as inaccuracies in the recording of 
+molecular clock prior (either due to errors in tree-building or genuinely varying mutation rates), as well as inaccuracies in the recording of 
 sample dates.
@@
 
-The B.1.1.529 lineage (Omicron) is believed to have emerged in southern Africa, and
-it is a saltational lineage with two highly successful sublineages, BA.1 and BA.2 \cite{Dor2025}.
+The B.1.1.529 lineage (Omicron)
+is believed to have emerged in southern Africa, and
+it is a saltational lineage with two highly successful sublineages,
+BA.1 and BA.2 \cite{Dor2025}.
 BA.1 and BA.2 are genetically distinct from each other and
@@
 (above the B.1.1.529-labeled MRCA of BA.1 and BA.2)
-by more than ten mutations, including six deleted regions.
-The ARG captures the split between BA.1 and BA.2,
+by 27 mutations (including two deletions),
+reflecting its saltational origin.
+The ARG captures the split between BA.1 and BA.2
 which has been shown in previous phylogenetic analysis \cite{Dor2025}.
@@
 Also, there is uncertainty in the sequence alignments
-leading to ambiguity in the start position of these deletions,
+leading to ambiguity in the start position of these deletions
+(if these deletions occur at 11283,
+there would be a single deletion event before the BA.1/BA.2 split),
 as noted in the Pango designation issue \cite{RambautOmicron2021}.
@@
 We selected a BA.1 seed sample that led to the fewest mutations.
-This strategy resulted in five mutations highlighted
-in blue/orange/green/red/purple
-which the chosen seed does not have
-but are present in its inferred descendants.
+This strategy resulted in five mutations
+highlighted in blue/orange/green/red/purple
+which are absent in the chosen
+but present in its inferred descendants.
 In retrospect, SRR17051902 should have been a better choice for a seed,
 because it would have resulted in a topology
-where those five mutations would be placed by parsimony
+where those five mutations would be placed
 above the split between BA.1 and BA.2.
@@
 \label{sec:supplement_indepth_pango_x_lineages}
+
+% Sc2ts detection results in matches
+% https://github.com/jeromekelleher/sc2ts-paper/issues/366
+
+% Multiple-breakpoint recombinants
+% https://github.com/jeromekelleher/sc2ts-paper/issues/379
+
+\begin{table}
+\centering
+\caption{
+Sc2ts detection results for Pango X lineages not added to the ARG.
+Some Pango X lineages present in the Viridian v04 dataset
+were not integrated into the ARG absent in the ARG.
+The number of samples in Viridian v04 dataset
+assigned to a Pango X lineage and
+the number of samples which passed sequence-level quality control
+are shown here.
+}
+\label{tab:absent_pango_x}
+\begin{tabular}{crrr}
+\toprule
+Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
+\midrule
+XBJ & 2 & 1 & 1\\
+XBP & 8 & 2 & 1\\
+XBS & 19 & 15 & 1\\
+XBW & 1 & 1 & 1\\
+XCA & 11 & 4 & 1\\
+XAK & 2 & 2 & 1\\   % Multiple breakpoint Pango X
+XAY & 15 & 11 & 6\\ % Multiple breakpoint Pango X
+XBC & 36 & 21 & 3\\ % Multiple breakpoint Pango X
+XBL & 2 & 2 & 0\\   % Multiple breakpoint Pango X
+XBT & 1 & 1 & 2\\   % Multiple breakpoint Pango X
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 that is directly supported by a BQ.1.25-labeled sample identical to it.
-
-\subsection{Sc2ts recombination detection results for additional Pango X lineages}
-
-\begin{table}
-\centering
-\caption{
-Sc2ts recombination detection results for Pango X lineages which were not added to the ARG.
-}
-\label{tab:absent_pango_x}
-\begin{tabular}{crrr}
-\toprule
-Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
-\midrule
-XBJ & 2 & 1 & 1\\
-XBP & 8 & 2 & 1\\
-XBS & 19 & 15 & 1\\
-XBW & 1 & 1 & 1\\
-XCA & 11 & 4 & 1\\
-XAK & 2 & 2 & 1\\
-XAY & 15 & 11 & 6\\
-XBC & 36 & 21 & 3\\
-XBL & 2 & 2 & 0\\
-XBT & 1 & 1 & 2\\
-\bottomrule
-\end{tabular}
-\end{table}
 
@@
 \centering
-\includegraphics[width=0.85\linewidth]{node_time_testing.pdf}
-\caption{
-Heat map showing differences (in days) between actual and re-estimated sample dates for a subset of internal sample nodes. 
-}
-\label{fig:node_time_testing}
-\end{figure}
-
-\begin{figure}
-\centering
 \includegraphics[width=0.85\linewidth]{pango_vs_nextstrain_node_times.pdf}
@@
 \label{fig:pango_vs_nextstrain_node_times}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.85\linewidth]{node_time_testing.pdf}
+\caption{
+Heat map showing differences (in days) between actual and re-estimated sample dates for a subset of internal sample nodes. 
+}
+\label{fig:node_time_testing}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.85\linewidth]{subgraph-Omicron.pdf}
+\includegraphics[width=0.85\linewidth]{subgraph-Omicron_recolored.pdf}
 \caption{
@@
 B.1.1.529 mutations
-which are neither BA.1 or BA.2 characteristic mutations are outlined in green.
-These colours correspond to those used
+which are neither BA.1 or BA.2 characteristic mutations are
+outlined in green.
+These colors correspond to those used
 in the Pango designation issue \cite{RambautOmicron2021}.

### 2025-10-07 latex/main.tex
Updates from Overleaf

@@
 Some Pango X lineages present in the Viridian v04 dataset
-were not integrated into the ARG absent in the ARG.
+were not integrated into the ARG due to high HMM costs
 The number of samples in Viridian v04 dataset

### 2025-10-07 latex/main.tex
Updates from Overleaf

--- 2025-10-07 latex/main.tex
@@
 % https://github.com/jeromekelleher/sc2ts-paper/issues/379
+Ten Pango X lineages present in the Viridian v04 dataset
+were not integrated into the ARG
+due to high HMM costs and/or insufficient sample support,
+but their samples were matched against the ARG during inference.
+Table~\ref{tab:absent_pango_x} summarizes
+the HMM matching results of these Pango X lineages.
+All were identified as recombinants by sc2ts, except XBL
+for which the evidence for recombination
+supporting the suggested rightmost parent
+involves only three informative sites,
+which fall below our threshold of detection.
+These results demonstrate that sc2ts can detect
+recombinants with multiple breakpoints (XAY, XBC, and XBT).
+The recombination patterns of these Pango X lineages
+can be complex; exploring how to integrate such recombinants
+should an important direction for future work.
 
@@
 Sc2ts detection results for Pango X lineages not added to the ARG.
-Some Pango X lineages present in the Viridian v04 dataset
-were not integrated into the ARG absent in the ARG.
-The number of samples in Viridian v04 dataset
-assigned to a Pango X lineage and
-the number of samples which passed sequence-level quality control
+The number of samples assigned to these Pango X lineages
+in Viridian v04 dataset,
+the number of samples which passed sequence-level quality control, and
+the number of recombination breakpoints in the Viterbi solutions of the samples
 are shown here.

### 2025-10-07 latex/main.tex
Updates from Overleaf

@@
 the HMM matching results of these Pango X lineages.
-All were identified as recombinants by sc2ts, except XBL
-for which the evidence for recombination
+All were identified as recombinants by sc2ts, except XBL.
+The evidence for recombination
 supporting the suggested rightmost parent

### 2025-10-07 latex/main.tex
Updates from Overleaf

--- 2025-10-07 latex/main.tex
@@
 Using sc2ts, we built an ARG from 2,482,157 samples from the Viridian v04
-dataset\cite{Hunt2024}, collected between 2020-01-01 and 2023-02-20.
+dataset \cite{Hunt2024}, collected between 2020-01-01 and 2023-02-20.
 Inference took 21 days on a XX core server with 500 GB of RAM 
@@
 resulting in 316 distinct trees along the genome (some
-events occured at the same breakpoints [vague, rephrase]).
+events occurred at the same breakpoints [vague, rephrase]).
 [TODO note about loading up the time it takes to get the node 
@@
 the location of the inferred breakpoint interval relative to the Spike gene.
-For simplicity, ancestral lineages above the recombination parents or the MRCA are omitted.
+For simplicity, ancestral lineages above the recombinant parents or the MRCA are omitted.
 (A) Subgraph around the recombination node associated with XA.
@@
 shows the descendant samples broken down by Pango lineage (grouped 
-by root lineage, where appropriate).[FIXME Typesetting to be finished,
-describe cols for Type III and IV]
+by root lineage, where appropriate).
+[FIXME Typesetting to be finished, describe cols for Type III and IV]
 }
@@
 % 146
-Although the process of incorporating recombinant lineages in the Pango 
-nomenclature system is well defined,
-the identification of such events is much more subjective, requiring
-the synthesis of multiple lines of evidence and often the 
-manual inspection of phylogenies and sequence alignments
+Although the process of incorporating recombinant lineages
+in the Pango nomenclature system is well defined,
+the identification of such events is much more subjective,
+requiring the synthesis of multiple lines of evidence and
+often the manual inspection of phylogenies and sequence alignments
 by volunteers~\cite{Smith2023tracking}.
@@
 discovering and assessing evidence around these events.
-Table~\ref{tab:pango_x_lineages} summarises
-these results for the 44 Pango X lineages present in
-the ARG (XAC and XAD have been omitted, and XM 
-partially represented, corresponding to 23 samples in total; 
+Table~\ref{tab:pango_x_lineages} summarizes
+these results for the 44 Pango X lineages present in the ARG
+(XAC and XAD have been omitted, and
+XM partially represented, corresponding to 23 samples in total; 
 see Document~\ref{sec:supplement_x_pango_lineages}).
@@
 
-\subsection{Additional analyses of the Pango X lineages}
-\label{sec:supplement_indepth_pango_x_lineages}
-
-% Sc2ts detection results in matches
+\subsection{Sc2ts detection results of rejected Pango X lineages}
+\label{sec:supplement_rejected_pango_x_lineages}
+
+Ten Pango X lineages which are present in the Viridian v04 dataset
+were not integrated into the ARG
+due to high HMM costs and/or insufficient sample support,
+but their samples were matched against the ARG during inference.
 % https://github.com/jeromekelleher/sc2ts-paper/issues/366
-
-% Multiple-breakpoint recombinants
+Table~\ref{tab:extra_hmm_results} summarizes
+the HMM matching results of these Pango X lineages.
+All were identified as recombinants by sc2ts, except XBL.
+(For XBL, the evidence for recombination
+proposed in the Pango designation issue \cite{PangoIssueXBL}
+involves only three informative sites
+supporting the suggested rightmost parent, and
+therefore this likely fell below our threshold of detection.)
 % https://github.com/jeromekelleher/sc2ts-paper/issues/379
+Importantly, these results demonstrate that sc2ts can detect
+recombinants with multiple breakpoints (XAY, XBC, and XBT).
+% https://github.com/jeromekelleher/sc2ts-paper/discussions/776
+% https://github.com/jeromekelleher/sc2ts-paper/discussions/788
+XBC and XBT are clean examples of two-breakpoint recombinants,
+with the Pango lineages of the parents and the breakpoint intervals
+concordant with those suggested in the Pango designation issues.
+The recombination patterns of XAY, however, is complex.
+Exploring how to integrate these recombinants and
+other multiple-breakpoint recombinants
+should be an important direction for future work.
 
@@
 Sc2ts detection results for Pango X lineages not added to the ARG.
-Some Pango X lineages present in the Viridian v04 dataset
-were not integrated into the ARG due to high HMM costs
 The number of samples in Viridian v04 dataset
-assigned to a Pango X lineage and
-the number of samples which passed sequence-level quality control
-are shown here.
+assigned to these Pango X lineages,
+the number of samples which passed sequence-level quality control, and
+the number of recombination breakpoints
+in the Viterbi solutions of the samples
+are shown.
 }
-\label{tab:absent_pango_x}
+\label{tab:extra_hmm_results}
 \begin{tabular}{crrr}
@@
 \end{table}
+
+\subsection{Additional analyses of the Pango X lineages}
+\label{sec:supplement_indepth_pango_x_lineages}
 

### 2025-10-07 latex/main.tex
Updates from Overleaf

@@
 
-Using sc2ts, we built an ARG from 2,482,157 samples from the Viridian v04
-dataset \cite{Hunt2024}, collected between 2020-01-01 and 2023-02-20.
+Using sc2ts, we built an ARG from 2,482,157 samples
+from the Viridian v04 dataset \cite{Hunt2024},
+collected between 2020-01-01 and 2023-02-20.
 Inference took 21 days on a XX core server with 500 GB of RAM
@@
 Of the nodes, 855 represent recombination events,
-resulting in 316 distinct trees along the genome (some
-events occurred at the same breakpoints [vague, rephrase]).
+resulting in 316 distinct trees along the genome
+(some events occurred at the same breakpoints [vague, rephrase]).
 [TODO note about loading up the time it takes to get the node 
@@
 (A) Subgraph around the recombination node associated with XA.
-Two ``Group A'' recombinant samples reported by Jackson 
-et al.\cite{Jackson2021} are shown.
+Two ``Group A'' recombinant samples reported by Jackson et al.\cite{Jackson2021} are shown.
 (B) Subgraph around the recombination node associated with a 
@@
 are nodes exactly matched by sampled sequences. The descendants column
-shows the descendant samples broken down by Pango lineage (grouped 
-by root lineage, where appropriate).
+shows the descendant samples broken down by Pango lineage
+(grouped by root lineage, where appropriate).
 [FIXME Typesetting to be finished, describe cols for Type III and IV]
@@
 
-By automatically detecting and incorporating recombination events
-into the ARG, sc2ts provides us with a systematic means of 
+By automatically detecting and incorporating recombination events into the ARG,
+sc2ts provides us with a systematic means of
 discovering and assessing evidence around these events.
@@
 these results for the 44 Pango X lineages present in the ARG
-(XAC and XAD have been omitted, and
-XM partially represented, corresponding to 23 samples in total; 
+(XAC and XAD have been omitted, and XM partially represented,
+corresponding to 23 samples in total;
 see Document~\ref{sec:supplement_x_pango_lineages}).
-For 16 Pango X lineages the oldest node that has been assigned
-the focal Pango lineage is a recombination node, and that node 
-is the ancestor of all samples assigned that (or a descendant) Pango 
-lineage. These are the Type I events, in which the ARG reflects the
+For 16 Pango X lineages, the oldest node
+that has been assigned to the focal Pango lineage is a recombination node, and
+that node is the ancestor of all samples assigned
+to that (or a descendant) Pango lineage.
+These are the Type I events, in which the ARG reflects the
 expected phylogenetic structure of Pango designation.
@@
 % https://github.com/jeromekelleher/sc2ts-paper/discussions/788
-XBC and XBT are clean examples of two-breakpoint recombinants,
+XBC and XBT are relatively clean examples of multiple-breakpoint recombinants,
 with the Pango lineages of the parents and the breakpoint intervals
-concordant with those suggested in the Pango designation issues.
-The recombination patterns of XAY, however, is complex.
-Exploring how to integrate these recombinants and
-other multiple-breakpoint recombinants
+concordant with those suggested in the Pango designation issues,
+but XAY is more complex, possibly involving up to six breakpoints as per sc2ts.
+Exploring how to integrate these and other recombinants
+with complex recombination patterns \cite{Roemer2023}
 should be an important direction for future work.

### 2025-10-07 latex/main.tex
Good enough ripples pango X table

@@
 and Document S2 for ARG visualisations.
+
+\subsection*{Comparison with RIPPLES}
+
+\begin{table}
+\caption{RIPPLES pango X lineage events.} 
+\label{tab:ripples_x_lineages}
+\centering
+\begin{scriptsize}
+
+\begin{tabular}{lrrlrl}
+\toprule
+pango & in p4 & UShER desc & sc2ts class & clade diff & descendants \\
+\midrule
+XBR & $\checkmark$ & 1 & I & 0 & {XBR:1} \\
+XM3 & $\checkmark$ & 1 & I & 0 & {XM:1} \\
+XC & $\checkmark$ & 5 & I & 0 & {XC:5} \\
+XBM & $\checkmark$ & 12 & I & 0 & {XBM:10, BF.3:2} \\
+XS & $\checkmark$ & 17 & I & 0 & {XS:17} \\
+XY &  & 23 & I & 0 & {XY:23} \\
+XW &  & 32 & I & 0 & {XW:32} \\
+XA & $\checkmark$ & 39 & I & 0 & {XA:39} \\
+XL &  & 64 & I & 0 & {XL:64} \\
+XBB &  & 6452 & I & 1 & {XBB:2361} \\
+\midrule
+XQ &  & 1 & II & 0 & {XAM:1} \\
+XM & $\checkmark$ & 8 & II & 0 & {BA.2:4, XM:4} \\
+Xx &  & 9 & II & 0 & {XAE:9} \\
+Xx & $\checkmark$ & 18 & II & 0 & {XAC:18} \\
+XQ &  & 21 & II & 0 & {XAM:21} \\
+XE/XH &  & 1113 & II & 3 & {XE:1113} \\
+\midrule
+XB & $\checkmark$ & 411 & IV & 0 & {B.1:219, XB:192} \\
+\midrule
+NA & $\checkmark$ & 19 & NA & 45190 & {XAJ:18, BA.4:1} \\
+NA &  & 20 & NA & 1061775 & {BA.1.15:1, XS:17, BA.1:1, BA.1.1:1} \\
+NA & $\checkmark$ & 35 & NA & 7147 & {XBD:30, BA.2.75:5} \\
+NA & $\checkmark$ & 36 & NA & 7146 & {XBD:30, XBR:1, BA.2.75:5} \\
+NA &  & 1321 & NA & 340501 & {XE:1113, XH:2, XJ:68, XY:23, BA.2:82, XM:29,
+XAL:3, XAF:1} \\
+\bottomrule
+\end{tabular}
+
+\end{scriptsize}
+\end{table}
 

### 2025-10-07 latex/main.tex
Add draft figures and document table

@@
 
+Table~\ref{tab:ripples_x_lineages}
+Figure \ref{fig:ripples_sc2ts_events}
+
 \begin{table}
-\caption{RIPPLES pango X lineage events.} 
+\caption{RIPPLES Pango X lineage events. Each row is a recombination event
+identified by RIPPLEs (at p=3) in which X lineages appear, along with the 
+corresponding sc2ts event from Table~\ref{tab:pango_x_lineages}, where
+applicable, grouped by class. 
+Events also present at p=4 are indicated. Clades are matched 
+by finding the most recent common ancestor
+of the samples associated with each UShER/RIPPLES event in 
+the sc2ts ARG. The clade differences indicates how close this match
+is. The final group of rows shows events that cannot be reasonably
+mapped between UShER and sc2ts, where placement of one or more samples
+differed significantly.}
 \label{tab:ripples_x_lineages}
@@
 XQ &  & 1 & II & 0 & {XAM:1} \\
+XQ &  & 21 & II & 0 & {XAM:21} \\
 XM & $\checkmark$ & 8 & II & 0 & {BA.2:4, XM:4} \\
@@
 Xx & $\checkmark$ & 18 & II & 0 & {XAC:18} \\
-XQ &  & 21 & II & 0 & {XAM:21} \\
 XE/XH &  & 1113 & II & 3 & {XE:1113} \\
@@
 \centering
+\begin{tabular}{cc}
+\includegraphics[]{ripples_p4_sc2ts_events.pdf}&
+\includegraphics[]{ripples_p3_sc2ts_events.pdf}\\
+\end{tabular}
+\caption{TODO caption}
+\label{fig:ripples_sc2ts_events}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics[width=\linewidth]{recombinant_qc}

### 2025-10-07 latex/main.tex
Updates from Overleaf

@@
 [TODO we should mention UShER here: they state multiple times in the GitHub issues 
-that they don't handle them at all, so I think we can just say this in some nice
-way.]
+that they don't handle them at all, so I think we can just say this in some nice way.]
 Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
 but because they are difficult to distinguish from bioinformatics errors 
-and their multi-base nature presents challenges to the Markovian structure of 
-the LS HMM, we masked out all gap characters as missing data during primary
-inference. 
-To incorporate the most important deletions into the ARG, we 
-remapped data using posthoc parsimony at 75 sites involved in deletions 
+and their multi-base nature presents challenges
+to the Markovian structure of the LS HMM,
+we masked out all gap characters as missing data during primary inference. 
+To incorporate the most important deletions into the ARG,
+we remapped data using post hoc parsimony at 75 sites involved in deletions
 with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
-during postprocessing (STAR Methods).
+during post-processing (STAR Methods).
 Although sites are treated independently in this approach,
@@
 % widely? If not, fix to something that is?
+% According to Isobel, there is no standard notation.
 Alpha (11288+9, 21765+6, 21991+3, 28271+1) are correctly
@@
 which causes S-Gene Target Failure \cite{Walker2021},
-emerges two more times at the origins of the 
-BA.1 and BA.4 variants.
+emerges two more times at the origins of the BA.1 and BA.4 variants.
 Although most highly recurrent deletions are likely errors,
 some known truly recurrent deletions are present,
-for example, three deletions in S1-NTD
-(21,765+6, 21,991+3, and 22,194+3)
+for example, three deletions in S1-NTD (21765+6, 21991+3, and 22194+3)
 associated with antibody escape\cite{McCarthy2021}.
@@
 \caption{
-ARG events associated with Pango X lineages.
-Each row for Type I, III and IV events corresponds to the oldest node
-assigned the specified Pango lineage and is the ancestor of all
-nodes assigned that label. Type II events are recombination events
+Events in the ARG associated with Pango X lineages.
+Each row for Type I, III and IV events corresponds to
+the oldest node assigned to the specified Pango lineage and
+is the ancestor of all the nodes assigned to that label.
+Type II events are recombination events
 associated with Pango X lineages not meeting Type I criteria.
-For Type I and II recombination
-events, the averted shows the number of additional mutations required
-under a recombination-free HMM solution. Parent lineages shown in boldface
-are nodes exactly matched by sampled sequences. The descendants column
-shows the descendant samples broken down by Pango lineage
+For Type I and II recombination events,
+the averted shows the number of additional mutations required
+under a recombination-free Viterbi solution.
+Parent lineages shown in boldface are nodes exactly matched by sampled sequences.
+The descendants column shows the descendant samples broken down by Pango lineage
 (grouped by root lineage, where appropriate).
@@
 are designated and their explicitly phylogenetic basis.
-The first recombinant lineage XA was designated in 2021 based on the 
-analysis of Jackson et al.\cite{Jackson2021}, which established 
-that recombination with onward transmission had occured [weak FIXME].
-As of pango-designation version v1.33.1 (2025-03-31), a total of 146 distinct 
-X lineages have been designated.
+The first recombinant lineage XA was designated in 2021
+based on the analysis of Jackson et al.\cite{Jackson2021},
+which established that recombination with onward transmission had occurred [weak FIXME].
+As of pango-designation version v1.33.1 (2025-03-31),
+a total of 146 distinct Pango X lineages have been designated.
 %$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
@@
 that has been assigned to the focal Pango lineage is a recombination node, and
-that node is the ancestor of all samples assigned
-to that (or a descendant) Pango lineage.
+that node is the ancestor of all samples
+assigned to that (or a descendant) Pango lineage.
 These are the Type I events, in which the ARG reflects the
@@
 (see also Document~\ref{sec:jackson_recombs}
-for further analysis of the Jackson et al.\ recombinants).
+for further analysis of the recombinants
+identified by Jackson et al.\cite{Jackson2021}).
 
@@
 
-The final class of event (Type IV) in Table~\ref{tab:pango_x_lineages} 
-correspond to nodes that are not associated with a recombination
-event. These vary in their plausibility of being non-recombinant.
+The final class of events (Type IV) in Table~\ref{tab:pango_x_lineages}
+corresponds to nodes that are not associated with a recombination event.
+These vary in their plausibility of being non-recombinant.
 For example, XN (Figure~\ref{fig:subgraphs}C) is only one mutation
-distant from a BA.2 sample and has little additional evidence to 
-suggest recombination [ref sup section]. XB, on the other hand,
-has been thoroughly characterised\cite{Gutierrez2022} and 
-may have misclassified by sc2ts due to limitations in handling
-deletions or matching against long branches (see Discussion).
-See Document~\ref{sec:supplement_indepth_pango_x_lineages} for extended 
-analyses on all of the Pango X lineages above,
+distant from a BA.2 sample and has little additional evidence
+to suggest recombination [ref sup section].
+XB, on the other hand,
+has been thoroughly characterized \cite{Gutierrez2022} and
+may have misclassified by sc2ts due to limitations in
+handling deletions or matching against long branches (see Discussion).
+See Document~\ref{sec:supplement_indepth_pango_x_lineages}
+for extended analyses on all of the Pango X lineages above,
 Figure~\ref{fig:pango_x_copying_patterns} for copying patterns 
-associated with Table~\ref{tab:pango_x_lineages},
-and Document S2 for ARG visualisations.
+associated with Table~\ref{tab:pango_x_lineages}, and
+Document S2 for ARG visualizations.
 
@@
 \begin{table}
-\caption{RIPPLES Pango X lineage events. Each row is a recombination event
-identified by RIPPLEs (at p=3) in which X lineages appear, along with the 
-corresponding sc2ts event from Table~\ref{tab:pango_x_lineages}, where
-applicable, grouped by class. 
-Events also present at p=4 are indicated. Clades are matched 
-by finding the most recent common ancestor
-of the samples associated with each UShER/RIPPLES event in 
-the sc2ts ARG. The clade differences indicates how close this match
-is. The final group of rows shows events that cannot be reasonably
-mapped between UShER and sc2ts, where placement of one or more samples
-differed significantly.}
+\caption{
+RIPPLES events associated with Pango X lineages.
+Each row is a recombination event identified by RIPPLEs (at p = 3)
+in which Pango X lineages appear,
+along with the corresponding sc2ts event from Table~\ref{tab:pango_x_lineages},
+where applicable, grouped by class.
+Events also present at p = 4 are indicated.
+Clades are matched by finding the most recent common ancestor of
+the samples associated with each UShER/RIPPLES event in the sc2ts ARG.
+The clade differences indicates how close this match is.
+The final group of rows shows events
+that cannot be reasonably mapped between UShER and sc2ts,
+where placement of one or more samples differed significantly.
+}
 \label{tab:ripples_x_lineages}
@@
 \subsection*{Quality control of recombination events}
+
 Sequencing errors in genome assemblies \cite{Turakhia2020},
@@
 and effectively identifies a range of issues.
-Figure~\ref{fig:recombinant_qc} plots the net number of supporting
-loci for the left and right parents for all 855 recombination
-events. The 501 recombinants with fewer than 4 net loci supporting 
-each parent are highly enriched for the AmpliSeq V1 sequencing
-protocol [TODO concrete numbers?], 
-and 71 (14\%) are associated with 6-base deletion in the 
-Delta lineage, known to be affected 
-by amplicon dropout\cite{Sanderson2021}.
+Figure~\ref{fig:recombinant_qc} plots
+the net number of supporting loci for the left and right parents
+for all 855 recombination events.
+The 501 recombinants having fewer than 4 net loci supporting each parent
+are highly enriched for samples sequenced using the AmpliSeq V1 protocol
+% https://github.com/jeromekelleher/sc2ts-paper/issues/808
+(70\% of the recombinants despite making up 2\% of the Viridian dataset), and
+71 (14\%) are associated with 6-base deletion in the Delta lineage,
+known to be affected by amplicon dropout\cite{Sanderson2021}.
 The 354 remaining recombination events that pass this QC filter
@@
 (Table~\ref{tab:pango_x_lineages}) and
-the Jackson et al recombinants (Document\ref{sec:jackson_recombs}).
+the Jackson et al. recombinants (Document\ref{sec:jackson_recombs}).
 
@@
 The difference between the numbers of mutations between
-the two solutions---the number of 
-mutations ``averted'' by recombination---is a measure of the evidence
-in favor of recombination. 
-Table~\ref{tab:pango_x_lineages} shows
-the number of averted mutations for each of the 21 recombination events
-associated with Pango X lineages, with support ranging from very
-strong in the case of XC (the most parsimonious non-recombinant 
-placement would require 37 additional mutations) to marginal 
-for XG and XW, which have the minimal possible number of averted 
-mutations.
+the two solutions---the number of mutations
+``averted'' by recombination---is a measure of
+the evidence in favor of recombination.
+Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
+for each of the 21 recombination events associated with Pango X lineages,
+with support ranging from very strong in the case of XC
+(the most parsimonious non-recombinant placement would
+require 37 additional mutations) to marginal for XG and XW,
+which have the minimal possible number of averted mutations.
 
@@
 \caption{
-Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes,
+Evidence for recombination in the ARG.
+All plots show ``robust'' sc2ts recombination nodes,
 colored according to their classification by rebar
@@
 versus an alternative copying path disallowing recombination ($x$: top axis);
-recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
-those identified by Jackson et al. \cite{Jackson2021} are prefixed with a `J', e.g.
+recombination nodes in Table~\ref{tab:pango_x_lineages} are labeled starting with `X' and
+those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
 Jackson group B (JB) and Jackson singleton recombinant 1 (J1).
@@
 characterized by a long branch preceding its origin.
-Most of the characteristic Alpha mutations (\cite{RambautAlpha2020})
+Most of the characteristic Alpha mutations \cite{RambautAlpha2020}
 (20 out of 23) in the ARG
@@
 \begin{tabular}{cc}
-\includegraphics[]{ripples_p4_sc2ts_events.pdf}&
-\includegraphics[]{ripples_p3_sc2ts_events.pdf}\\
+\includegraphics[width=0.5\linewidth]{ripples_p4_sc2ts_events.pdf}&
+\includegraphics[width=0.5\linewidth]{ripples_p3_sc2ts_events.pdf}\\
 \end{tabular}
-\caption{TODO caption}
+\caption{Intersection of the RIPPLES and sc2ts recombination events at
+different values of the RIPPLES parsimony parameter, p.}
 \label{fig:ripples_sc2ts_events}

### 2025-10-07 latex/main.tex
Updates
Update recombinants.csv with RIPPLEs

Fix on figs

Update the caption

Updates

@@
 overwhelmed even approximate distance-matrix based methods.
-At the peak in [month] 202X, there were over X thousand samples 
-per day deposited in GISAID\cite{Shu2017}, with a total of 
-over 17 million as of Sep 2025. % TODO check numbers
+% Accessed GISAID/EpiCov on 08 Oct., 2025
+At the peak in February, 2022, about 80 thousand samples
+were submitted on average per day to GISAID\cite{Shu2017},
+accumulating to a total of over 17.5 million samples
+as of October, 2025.
 Regularly reinferring trees de-novo from such rapidly 
@@
 were conclusively demonstrated, as the resource
-quickly became an indispensible element 
+quickly became an indispensable element 
 of the pandemic response\cite{Hinrichs2024}.
@@
 exists, and the tskit library~\cite{Ralph2020} is the basis of a 
-rapidly growing 
-software ecosystem of 
+rapidly growing software ecosystem of
 simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
-visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
+visualization~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
 inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
@@
 recombination events, the resulting data structure---representing
-phylogenetic relationships, point mutations, deletions, and 
+genealogical relationships, point mutations, deletions, and 
 recombination---provides a basis for the study of SARS-CoV-2
@@
 Figure~\ref{fig:subgraphs}A shows XA as an example
-(see also Document~\ref{sec:jackson_recombs}
-for further analysis of the recombinants
-identified by Jackson et al.\cite{Jackson2021}).
+(for further analysis of the other recombinants
+found by Jackson et al.\cite{Jackson2021},
+see Table~\ref{tab:jackson} in Document~\ref{sec:jackson_recombs},
+which shows that sc2ts identified the parents and breakpoints
+concordant with those reported by Jackson et al.\cite{Jackson2021}).
 
@@
 \subsection*{Quantification of evidence for recombination}
-
 The LS HMM at the heart of sc2ts provides a simple and powerful means 
@@
 \caption{
-Evidence for recombination in the ARG.
-All plots show ``robust'' sc2ts recombination nodes,
-colored according to their classification by rebar
-(recombinant: purple, non-recombinant: gray).
+Properties of recombination events. A,B,C 
+show summaries of the 354 sc2ts recombination nodes
+passing quality control filters
+colored by CovRecomb classification (recombinant: purple, non-recombinant: gray).
 (A) Upper: jittered scatterplot comparing number of mutations
@@
 by allowing recombination (i.e. $x$-$y$).
-(B) Distribution of time to the MRCA of the parents of each robust recombination node.
+(B) Distribution of time to the MRCA of the parents of each QC-passing recombination node.
 (C) Distribution of the Pango lineage distance between parents.
 (D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
-leading to XA, XW, XBB, and an
-undesignated recombination event under XBB, which we here refer to as  XBB.x.}
-\label{fig:averted_mutations}
+for XA, XW, XBB, and an
+undesignated recombination event under XBB, which we refer to as XBB.x.}
+\label{fig:recombinant_evidence}
 \end{figure}
 
-Figure~\ref{fig:averted_mutations}A plots the number of mutations required
+Figure~\ref{fig:recombinant_evidence}A plots the number of mutations required
 in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
@@
 rate is much lower for recombinants with closely related 
-parents (Figure~\ref{fig:averted_mutations}B,C) suggesting that 
+parents (Figure~\ref{fig:recombinant_evidence}B,C) suggesting that 
 sc2ts may have a higher sensitivity.
+
+See Figure~\ref{fig:recombinant_evidence_rebar} for rebar,
+Figure~\ref{fig:recombinant_evidence_ripples_p4} for RIPPLES p4
+and Figure~\ref{fig:recombinant_evidence_ripples_p4} for RIPPLES p3.
 
@@
 (gold, with mutational change below).
-Figure~\ref{fig:averted_mutations}D shows some example copying patterns,
+Figure~\ref{fig:recombinant_evidence}D shows some example copying patterns,
 illustrating this spectrum of plausibility.
@@
 and some caution should be exercised in interpreting them. 
-See Figures~\ref{fig:averted_mutations}D and
+See Figures~\ref{fig:recombinant_evidence}D and
 \ref{fig:pango_x_copying_patterns} for copying patterns of the 
@@
 
-\subsection*{Recombination breakpoint intervals along the genome}
+\subsection*{Recombination intervals}
 
 We explored the spatial distribution of the breakpoint locations of
-the ``robust'' recombination events along the genome,
+the QC-passing recombination events along the genome,
 and quantified how precisely breakpoints were localized.
@@
 
-We ran rebar on the sampled sequences associated with each robust recombination event. 
+We ran rebar on the sampled sequences associated with each QC-passing recombination event.
 Rebar identifies a sequence as a recombinant if it contains sets of mutations 
@@
 \includegraphics[width=\linewidth]{samples-per-day}
-\caption{TODO caption}\label{fig:samples_per_day}
+\caption{
+TODO caption
+}\label{fig:samples_per_day}
 \end{figure}
@@
 \includegraphics[width=\linewidth]{inference-resources}
-\caption{TODO caption}
+\caption{
+TODO caption
+}
 \label{fig:inference_resources}
@@
 \label{fig:ripples_sc2ts_events}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{recombinant_evidence_rebar}
+\caption{Things}
+\label{fig:recombinant_evidence_rebar}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{recombinant_evidence_ripples_p4}
+\caption{Things}
+\label{fig:recombinant_evidence_ripples_p4}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{recombinant_evidence_ripples_p3}
+\caption{Things}
+\label{fig:recombinant_evidence_ripples_p3}
 \end{figure}

### 2025-10-07 latex/main.tex
Updates from Overleaf

--- 2025-10-07 latex/main.tex
@@
 Evidence for recombination in the ARG.
-All plots show ``robust'' sc2ts recombination nodes,
+All plots show QC-passing sc2ts recombination nodes,
 colored according to their classification by rebar
@@
 by allowing recombination (i.e. $x$-$y$).
-(B) Distribution of time to the MRCA of the parents of each robust recombination node.
+(B) Distribution of time to the MRCA of the parents of each QC-passing recombination node.
 (C) Distribution of the Pango lineage distance between parents.
@@
 We explored the spatial distribution of the breakpoint locations of
-the ``robust'' recombination events along the genome,
+the QC-passing recombination events along the genome,
 and quantified how precisely breakpoints were localized.

### 2025-10-08 latex/main.tex

@@
 
-\subsection*{Recombination breakpoint intervals along the genome}
+\subsection*{Recombination intervals}
 

### 2025-10-08 latex/main.tex
Updates from Overleaf

@@
 recurrent (Table~\ref{tab:rec_dels}) deletions.
-For example, the 4 characteristic deletions associated with 
+For example, the 4 characteristic deletions
+associated with Alpha \cite{RambautAlpha2020}
 % FIXME: this is just my notation - is this something recognised more 
 % widely? If not, fix to something that is?
-% According to Isobel, there is no standard notation.
-Alpha (11288+9, 21765+6, 21991+3, 28271+1) are correctly
-identified as having a single origin [CITATION?]
-(Figure~\ref{fig:subgraph_alpha}).
-The 21765+6 deletion (Spike 69-70)
+% SHZ: According to Isobel, there is no standard notation.
+% So, I'm just following what Rambaut did in this post.
+(11288-11296, 21765-21770, 21991-21993)
+occur at the origin of Alpha in the ARG (Figure~\ref{fig:subgraph_alpha}).
+The 6-base 21765-21770 deletion (Spike $\Delta$H69/V70),
 which causes S-Gene Target Failure \cite{Walker2021},
-emerges two more times at the origins of the BA.1 and BA.4 variants.
-Although most highly recurrent deletions are likely errors,
-some known truly recurrent deletions are present,
+emerges two more times at the origins of BA.1 and BA.4 (Table~\ref{tab:major_dels}).
+The ARG also capture genuine recurrent deletions,
 for example, three deletions in S1-NTD (21765+6, 21991+3, and 22194+3)
@@
 Although uncommon relative to single nucleotide mutations,
-deletions are an important component of SARS-CoV-2 evolution,
+deletions are an important component of
+SARS-CoV-2 evolution \cite{Jeronimo2023},
 contributing to the emergence of VOCs by enhancing,
@@
 (i.e., for a deletion caused by multiple events in the ARG,
-the sum of all the samples inheriting the deletion across all the events).
+the total of all the samples inheriting the deletion across those events).
 

### 2025-10-08 latex/main.tex

@@
 \caption{
-Evidence for recombination in the ARG.
-All plots show QC-passing sc2ts recombination nodes,
-colored according to their classification by rebar
-(recombinant: purple, non-recombinant: gray).
+Properties of recombination events. A,B,C 
+show summaries of the 354 sc2ts recombination nodes
+passing quality control filters
+colored by CovRecomb classification (recombinant: purple, non-recombinant: gray).
 (A) Upper: jittered scatterplot comparing number of mutations
@@
 (D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
-leading to XA, XW, XBB, and an
-undesignated recombination event under XBB, which we here refer to as  XBB.x.}
-\label{fig:averted_mutations}
+for XA, XW, XBB, and an
+undesignated recombination event under XBB, which we refer to as XBB.x.}
+\label{fig:recombinant_evidence}
 \end{figure}
 
-Figure~\ref{fig:averted_mutations}A plots the number of mutations required
+Figure~\ref{fig:recombinant_evidence}A plots the number of mutations required
 in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
@@
 \centering
+\includegraphics[width=\linewidth]{recombinant_evidence_rebar}
+\caption{Things}
+\label{fig:recombinant_evidence_rebar}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{recombinant_evidence_ripples_p4}
+\caption{Things}
+\label{fig:recombinant_evidence_ripples_p4}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{recombinant_evidence_ripples_p3}
+\caption{Things}
+\label{fig:recombinant_evidence_ripples_p3}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics[width=\linewidth]{recombinant_qc}

### 2025-10-08 latex/main.tex
Updates from Overleaf

--- 2025-10-08 latex/main.tex
@@
 (11288-11296, 21765-21770, 21991-21993)
-occur at the origin of Alpha in the ARG (Figure~\ref{fig:subgraph_alpha}).
-The 6-base 21765-21770 deletion (Spike $\Delta$H69/V70),
-which causes S-Gene Target Failure \cite{Walker2021},
-emerges two more times at the origins of BA.1 and BA.4 (Table~\ref{tab:major_dels}).
-The ARG also capture genuine recurrent deletions,
-for example, three deletions in S1-NTD (21765+6, 21991+3, and 22194+3)
-associated with antibody escape\cite{McCarthy2021}.
+occur at the Alpha origin in the ARG (Figure~\ref{fig:subgraph_alpha}).
+The ARG also captures genuine recurrent deletions.
+For example, the known recurrent deletions in S1-NTD:
+the 6-base 21765-21770 deletion (Spike $\Delta$H69/V70),
+which causes S-Gene Target Failure \cite{Walker2021};
+(21765+6, 21991+3, and 22194+3)
+associated with antibody escape \cite{McCarthy2021}.
+(Table~\ref{tab:major_dels}).
 See Document\ref{sec:supplement_dels} for further analysis.

### 2025-10-08 latex/main.tex

@@
 of the pandemic response\cite{Hinrichs2024}.
-The second challenge posed is that recombination is 
-an important force in SARS-CoV-2 
-evolution\cite{Jackson2021,VanInsberghe2021,Focosi2022,
+The second challenge posed is that recombination is an important force
+in SARS-CoV-2 evolution \cite{Jackson2021,VanInsberghe2021,Focosi2022,
 Gutierrez2022,Ignatieva2022,Tamura2023},
-with multiple deep recombination events now known to be present in 
-genealogy of the global population [cites??].
+with multiple deep recombination events now known to be present
+in the genealogy of the global population \cite{Roemer2023}.
 Although several methods have been proposed to \emph{detect}
@@
 Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
-but because they are difficult to distinguish from bioinformatics errors 
+but because they can be difficult to distinguish from bioinformatics errors
 and their multi-base nature presents challenges
@@
 correctly identified many known
-high frequency (Table~\ref{tab:major_dels}) and 
-recurrent (Table~\ref{tab:rec_dels}) deletions.
-For example, the 4 characteristic deletions
-associated with Alpha \cite{RambautAlpha2020}
-% FIXME: this is just my notation - is this something recognised more 
-% widely? If not, fix to something that is?
-% SHZ: According to Isobel, there is no standard notation.
-% So, I'm just following what Rambaut did in this post.
-(11288-11296, 21765-21770, 21991-21993)
-occur at the Alpha origin in the ARG (Figure~\ref{fig:subgraph_alpha}).
-The ARG also captures genuine recurrent deletions.
-For example, the known recurrent deletions in S1-NTD:
-the 6-base 21765-21770 deletion (Spike $\Delta$H69/V70),
-which causes S-Gene Target Failure \cite{Walker2021};
-(21765+6, 21991+3, and 22194+3)
-associated with antibody escape \cite{McCarthy2021}.
-(Table~\ref{tab:major_dels}).
+high-frequency deletions (Table~\ref{tab:major_dels}) and
+recurrent deletions (Table~\ref{tab:rec_dels}).
+For example, the three characteristic deletions of Alpha \cite{RambautAlpha2020}
+(11288+9, 21765+6, 21991+3,
+where the first and second numbers are
+the start position and length, respectively)
+are correctly mapped above the Alpha origin in the ARG
+(Figure~\ref{fig:subgraph_alpha}).
+The ARG captures known recurrent deletions
+in the N-terminal domain of the S1 unit of Spike
+(which is a deletion hotspot \cite{McCarthy2021})
+for example,
+the 21765+6 deletion ($\Delta$H69/V70),
+which causes S-Gene Target Failure \cite{Walker2021} and
+is associated with increased efficiency in cell entry \cite{Meng2021}, and
+the 21991+3 deletion
+which is associated with antibody escape \cite{McCarthy2021}.
 See Document\ref{sec:supplement_dels} for further analysis.

### 2025-10-08 latex/main.tex
Updates from Overleaf

@@
 inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
-comprehensive phylogenies of these datasets are typically
-built with simpler distance-matrix methods [citation?]
-or approximations~\cite{Price2009fasttree}, primarily
-due to computational scaling limitations, but also because of
-the complexity of modelling such datasets.
+comprehensive phylogenies of these datasets are typically built with
+simpler distance methods [citation?]
+or approximations~\cite{Price2009fasttree},
+primarily due to computational scaling limitations,
+but also because of the complexity of modeling such datasets.
 While several methods have been proposed to incorporate
@@
 Software support in explicitly \emph{using} recombination-aware
-inference is also extremely limited [TODO add some concrete 
-details].
+inference is also extremely limited
+[TODO add some concrete details].
 
@@
 The ARG captures the split between BA.1 and BA.2
-which has been shown in previous phylogenetic analysis \cite{Dor2025}.
+shown in previous phylogenetic analysis \cite{Dor2025}.
 The 5-base deletion highlighted in green above the BA.2 root
@@
 We selected a BA.1 seed sample that led to the fewest mutations.
-This strategy resulted in five mutations
+This strategy resulted in five recurrent mutations
 highlighted in blue/orange/green/red/purple
-which are absent in the chosen
-but present in its inferred descendants.
-In retrospect, SRR17051902 should have been a better choice for a seed,
-because it would have resulted in a topology
-where those five mutations would be placed
-above the split between BA.1 and BA.2.
+which are absent in the chosen seed
+but present in many of its inferred descendants.
+Inspection of the subgraph reveals that an alternative choice for a seed would be the later sample SRR17051902. Seeding with this sample would be likely to generate a topology
+in which those five recurrent mutations are shared between BA.1 and BA.2 and subsequently reverted in a number of the deep branching BA.1 lineages such as those to ERR7671333 and ERR7858953.
 
@@
 we highlight the characteristic mutations of Kappa and Delta:
-those in blue which are listed
+those with a blue outline are listed
 in the Pango designation issue \cite{PangoIssueKappa}
@@
 but absent in the Delta constellation;
-those in purple which are listed
+those with a purple outline are listed
 in the Delta constellation \cite{ScorpioDelta}
@@
 but absent in the Kappa constellation; and
-those in red which are listed
+those with a red outline are listed
 in both the Kappa and Delta constellations.
-We also highlight
+We also outline
 in orange the mutations associated with the lineage containing clades A to D and
@@
 These colors correspond to those used
-in the Pango designation issue \cite{RambautOmicron2021}.
+in the BA.1/BA.2 Pango designation issue \cite{RambautOmicron2021}.
 }

### 2025-10-08 latex/main.tex

@@
 are recombination events closely associated with Pango X lineages, 
-but that do not meet the strict origination criteria for Type I.
+but do not meet the strict origination criteria for Type I.
 Type III events are Pango origin nodes in the ARG that are 
@@
 \subsection*{Comparison with RIPPLES}
-
-Table~\ref{tab:ripples_x_lineages}
-Figure \ref{fig:ripples_sc2ts_events}
+RIPPLES~\cite{Turakhia2022} is a method for detecting 
+recombination events in a tree inferred by UShER. Recombination is 
+detected by first identifying plausible branches carrying multiple
+mutations (at least 3, by default). For each candidate branch,
+RIPPLES then systematically splits the sequence into 2 or 3
+segments, and finds parsimonious placements for each segment independently. 
+A recombination event at that node is then flagged if the overall parsimony of these multiple 
+placements sufficiently improves over the original single-parent
+origin in the tree (the ``parsimony improvement'' parameter, $p=3$, by default).
+RIVET~\cite{Smith2023tracking} is a tool for quality control
+and visualisation of these recombination events.
 
@@
 The clade differences indicates how close this match is.
-The final group of rows shows events
-that cannot be reasonably mapped between UShER and sc2ts,
-where placement of one or more samples differed significantly.
 }
@@
 \toprule
-pango & in p4 & UShER desc & sc2ts class & clade diff & descendants \\
+sc2ts event & in p4 & UShER desc & sc2ts class & clade diff & descendants \\
 \midrule
+\multicolumn{6}{l}{Equivalent Sc2ts and RIPPLEs recombination events}\\
 XBR & $\checkmark$ & 1 & I & 0 & {XBR:1} \\
-XM3 & $\checkmark$ & 1 & I & 0 & {XM:1} \\
+% XM3 & $\checkmark$ & 1 & I & 0 & {XM:1} \\
 XC & $\checkmark$ & 5 & I & 0 & {XC:5} \\
-XBM & $\checkmark$ & 12 & I & 0 & {XBM:10, BF.3:2} \\
 XS & $\checkmark$ & 17 & I & 0 & {XS:17} \\
@@
 XBB &  & 6452 & I & 1 & {XBB:2361} \\
+XBM & $\checkmark$ & 12 & II & 0 & {XBM:10, BF.3:2} \\
 \midrule
-XQ &  & 1 & II & 0 & {XAM:1} \\
-XQ &  & 21 & II & 0 & {XAM:21} \\
-XM & $\checkmark$ & 8 & II & 0 & {BA.2:4, XM:4} \\
-Xx &  & 9 & II & 0 & {XAE:9} \\
-Xx & $\checkmark$ & 18 & II & 0 & {XAC:18} \\
+\multicolumn{6}{l}{RIPPLEs recombination event close to sc2ts recombination event}\\
 XE/XH &  & 1113 & II & 3 & {XE:1113} \\
+XQ &  & 1 & III & 0 & {XAM:1} \\
+XQ &  & 21 & III & 0 & {XAM:21} \\
+Xx &  & 9 & III & 0 & {XAE:9} \\
+Xx & $\checkmark$ & 18 & III & 0 & {XAC:18} \\
+XM & $\checkmark$ & 8 & III & 0 & {BA.2:4, XM:4} \\
 \midrule
+\multicolumn{6}{l}{RIPPLEs recombination event equivalant to sc2ts non-recombination event}\\
 XB & $\checkmark$ & 411 & IV & 0 & {B.1:219, XB:192} \\
 \midrule
+\multicolumn{6}{l}{RIPPLEs recombination event not comparable to sc2ts}\\
 NA & $\checkmark$ & 19 & NA & 45190 & {XAJ:18, BA.4:1} \\
@@
 \end{table}
+
+We ran fast-RIPPLES\cite{Smith2023tracking} on the Viridian UShER 
+tree subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
+We first ran RIPPLES with parameter $p=4$, as being roughly comparable to 
+the recombination penalty used in sc2ts ($k=4$; STAR Methods).
+This resulted in 1,174 recombination events reported, of which 460
+could be mapped to one of the 855 recombination events in the sc2ts ARG (STAR Methods;
+Figure \ref{fig:ripples_sc2ts_events}A). Analysis of the descendant 
+samples showed that many Pango X lineages were not associated with
+recombination events (including XBB), and we therefore ran
+RIPPLES with the more lenient value of $p=3$ (the default).
+This resulted in a total of 4111 recombination events
+of which 595 have corresponding recombination events in the 
+sc2ts ARG (Figure \ref{fig:ripples_sc2ts_events}B).
+
+Table~\ref{tab:ripples_x_lineages} shows a summary of all 
+recombination events associated with Pango X lineages detected by RIPPLES,
+and their relationship with the sc2ts ARG. 
+Of the 16 Class I events 
+(Pango X lineages unequivocally associated with a recombination node in
+the sc2ts ARG) 
+in Table~\ref{tab:pango_x_lineages}, 8
+have corresponding recombination events (4, at $p=4$). Of the 5 Type II
+events (recombination nodes with more complex assocation with X lineages),
+two have close matches. XBM matches exactly (including 2 descending BF.3 samples),
+and XE/XH is similar (but without including the 2 XH sample in the clade).
+UShER and sc2ts agree on creating two separate XAM clades, but while sc2ts
+makes these both descendants of the Xx event, RIPPLEs creates two independent
+recombinations. Similarly, XAE and have exactly matching clades, but RIPPLEs
+infers to independent recombination events while sc2ts groups them both
+under the Xx event. XB is notable as a likely sc2ts false negative that 
+identified by RIPPLES at both $p=3$ and $p=4$. The remaining events 
+associated with X lineages output by RIPPLES cannot be easily mapped 
+to the sc2ts ARG as one or more descendants is distantly placed.
+XAJ is an outlier for sc2ts, being the Pango X lineage identified as a 
+non-recombinant with the largest number of mutations 
+(Table~\ref{tab:pango_x_lineages}), and a likely false negative.
+The remaining 4 events are mixtures of multiple lineages that are 
+difficult to reconcile with either the other events in the table 
+or the sc2ts ARG.
 
@@
 mismatches, where $k$ is the recombination penalty (here, $k=4$).
-After some experimentation, we set the recombination penalty to 7, 
+After some experimentation, we set the HMM cost threshold to 7, 
 disallowing samples that require 8 or more mutations (or, e.g., 1 recombination
@@
 \includegraphics[width=0.5\linewidth]{ripples_p3_sc2ts_events.pdf}\\
+(A) & (B) \\
 \end{tabular}

### 2025-10-08 latex/main.tex
Averted muts similar to ripples parsimony improvement
Closes #827

@@
 ``averted'' by recombination---is a measure of
-the evidence in favor of recombination.
+the evidence in favor of recombination  (similar to 
+idea of ``parsimony improvement'' in RIPPLES.)
 Table~\ref{tab:pango_x_lineages} shows the number of averted mutations

### 2025-10-08 latex/main.tex
First draft of RIPPLES numbers summary

@@
 
+[Rough first draft]
+This estimate for effective recombination in SARS-CoV-2 is much lower 
+than previous studies have suggested.
+% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
+% recombinant origins out of the first 87,695 genomes to be released
+In an early analysis, 225 potentially recombinant sequences out 
+of 87,695 samples [WHAT DATE?]\cite{Varabyou2021rapid}.
+Over the same time period, XX events are present in the sc2ts ARG,
+representing a total of XX samples.
+In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
+589 unique recombination events in a tree of 1.6 million samples
+up to May 2021 were identified after extensive quality control.
+These nodes had a combined total of 43,104 descendants, suggesting
+that approximately 2.7\% of genomes were of recombinant origin.
+Over the same time period, the sc2ts ARG contains 14 recombinants
+passing QC with a total of 172 descendants. The ARG contains a total
+501,235 samples over this time period, giving an overall total fraction of
+0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
+events and 79-fold descendant samples over the same time period.
+The numbers are not directly comparable as the published RIPPLES 
+estimate is based on GISAID data which is enriched for
+artefacts~\cite{Hunt2024}, but given the observations in 
+the previous section on running RIPPLES on the Viridian data
+it seems likely that RIPPLES is overestimating the prevalence
+of recombination in SARS-CoV-2.
+
+% CovRecomb:  135,567 putative recombinants across the entirety of 14.5 million
+% accessed SARS-CoV-2 genomes
+
+% Although multiple studies have 
+% published lists of putatitive novel recombinant 
+% sequences~\cite{,Turakhia2022,Li2024CovRecomb} 
+% these are for GISAID data and not trival to reconcile with 
+% the Viridian data or the specific recombination events in the ARG.
+
+% RIPPLES: In a 1.6 million sample tree from May 2021, we identify 589
+% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
+% recombinant origins out of the first 87,695 genomes to be released
+% recombination events, 
+% RecombinHunt just looks at X lineages (I think?)
+
+To assess the evidence for these putative novel recombinants
+
 \subsection*{Quantification of evidence for recombination}
@@
 and 3 additional from Jackon et al in Table~\ref{tab:jackson}) have 
-% Slightly risky, this, as we could get asked to incorporate 
-% novel recombs found by RIPPLEs? This could be done if we could get 
-% Viridian sample IDs for their list.
 been thoroughly characterised.
+
+Although multiple studies have 
+published lists of putatitive novel recombinant 
+sequences~\cite{Varabyou2021rapid,Turakhia2022,Li2024CovRecomb} 
+these are for GISAID data and not trival to reconcile with 
+the Viridian data or the specific recombination events in the ARG.
+
+% CovRecomb:  135,567 putative recombinants across the entirety of 14.5 million
+% accessed SARS-CoV-2 genomes
+% RIPPLES: In a 1.6 million sample tree from May 2021, we identify 589
+% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
+% recombinant origins out of the first 87,695 genomes to be released
+% recombination events, 
+% RecombinHunt just looks at X lineages (I think?)
+
+To assess the evidence for these putative novel recombinants
+
 The figure also shows the classification status for each recombinant

### 2025-10-08 latex/main.tex

@@
 
-To assess the evidence for these putative novel recombinants
-
 \subsection*{Quantification of evidence for recombination}
@@
 been thoroughly characterised.  
-
-Although multiple studies have 
-published lists of putatitive novel recombinant 
-sequences~\cite{Varabyou2021rapid,Turakhia2022,Li2024CovRecomb} 
-these are for GISAID data and not trival to reconcile with 
-the Viridian data or the specific recombination events in the ARG.
-
-% CovRecomb:  135,567 putative recombinants across the entirety of 14.5 million
-% accessed SARS-CoV-2 genomes
-% RIPPLES: In a 1.6 million sample tree from May 2021, we identify 589
-% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
-% recombinant origins out of the first 87,695 genomes to be released
-% recombination events, 
-% RecombinHunt just looks at X lineages (I think?)
-
-To assess the evidence for these putative novel recombinants
-
-The figure also shows the classification status for each recombinant
-according to rebar, a method that classifies sequences
-as recombinant or not based on groups of mutations associated with
-Pango lineages (STAR Methods). 
+To assess the evidence for these putative novel recombination events we 
+ran CovRecomb~\cite{Li2024CovRecomb} and rebar on the corresponding sequences 
+(STAR Methods). Of the 354 QC-passing sc2ts recombination events, 
+90 are classified as 
+recombinant by CovRecomb (all but 2 of the 510 QC-failing recombinants 
+were classified as non-recombinant).
+Rebar classified 203 of the QC-passing sc2ts recombination events, 
+as recombinant, with only 10 of those failing QC being classified 
+as recombinant.
+
+[WIP]
 Of the 38 recombinants that are strongly supported (at least 10 averted

### 2025-10-08 latex/main.tex

@@
 with multiple deep recombination events now known to be present
-in the genealogy of the global population \cite{Roemer2023}.
+in the genealogy of the global population \cite{Roemer2023,Shiraz2023enhanced}.
 Although several methods have been proposed to \emph{detect}

### 2025-10-08 latex/main.tex
Updates from Overleaf

@@
 is a major issue and practical application has been limited.
-Software support in explicitly \emph{using} recombination-aware
-inference is also extremely limited
-[TODO add some concrete details].
+Software support in explicitly \emph{using}
+recombination-aware inference is limited to
+visualization of small-scale phylogenetic
+networks~\cite{Huson2012,Vaughan2017,Solís-Lemus2017} and
+using tanglegrams to visualize tree
+changes (e.g.~\cite{Rasmussen2023}),
+typically using the Extended Newick format for
+interchange~\cite{Cardona2008}.
 
@@
 which is associated with antibody escape \cite{McCarthy2021}.
-See Document\ref{sec:supplement_dels} for further analysis.
+See Document~\ref{sec:supplement_dels} for further analysis.
 
@@
 RIPPLES~\cite{Turakhia2022} is a method for detecting 
-recombination events in a tree inferred by UShER. Recombination is 
+recombination events using a tree inferred by UShER. Recombination is 
 detected by first identifying plausible branches carrying multiple
@@
 two have close matches. XBM matches exactly (including 2 descending BF.3 samples),
-and XE/XH is similar (but without including the 2 XH sample in the clade).
+and XE/XH is similar (but without including the 2 XH samples in the clade).
 UShER and sc2ts agree on creating two separate XAM clades, but while sc2ts
-makes these both descendants of the Xx event, RIPPLEs creates two independent
-recombinations. Similarly, XAE and have exactly matching clades, but RIPPLEs
-infers to independent recombination events while sc2ts groups them both
-under the Xx event. XB is notable as a likely sc2ts false negative that 
+makes these both descendants of the Xx event, RIPPLES creates two independent
+recombinations. Similarly, XAE and ??? have exactly matching clades, but RIPPLES
+infers two independent recombination events while sc2ts groups them both
+under the Xx event. XB is notable as a likely sc2ts false negative that is 
 identified by RIPPLES at both $p=3$ and $p=4$. The remaining events 
@@
 and effectively identifies a range of issues.
-Figure~\ref{fig:recombinant_qc} plots
+Figure~\ref{fig:recombinant_qc} shows
 the net number of supporting loci for the left and right parents
@@
 
-\subsubsection*{Sample filtering and preprocessing}
-The collection date for a sequence is an important piece of information
-for sc2ts. We used the Viridian
-consensus date (``Date\_tree``), based on information from COG-UK, GISAID, and
-ENA/SRA as our source. Samples that have missing or incomplete dates (e.g., 2020-01)
-were omitted, as were
-the $\sim$40,000 samples which had a collection date of December 31, 2020.
-As the sample count for this date is a major outlier for this period of
-the pandemic, it seems likely that this 
-batch is highly enriched for incorrect collection dates.
+\subsubsection*{Sample filtering and alignment preprocessing}
+
+The collection date of a sample
+is an important piece of information for sc2ts.
+We used the Viridian consensus date (``Date\_tree``),
+based on information from COG-UK, GISAID, and ENA/SRA as our source.
+Samples which have missing or incomplete dates (e.g., 2020-01) were omitted,
+as were the $\sim$40,000 samples
+which had a collection date of December 31, 2020.
+As the sample count for this date is a major outlier
+for this period of the pandemic,
+it seems likely that this batch is highly enriched
+for incorrect collection dates.
 Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
@@
 inference~\cite{Turakhia2020},
-the list has not been updated to take into account the systematic improvements 
-made by the Viridian dataset, and we therefore assembled a custom set of sites to mask. 
-To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
+the list has not been updated to take into account
+the systematic improvements made by the Viridian dataset, and
+we therefore assembled a custom set of sites to mask. 
+To do so, we first built a preliminary ARG
+from the samples collected up to June 30, 2021,
 without any site masking.
-Then, we identified the top 100 sites in terms of mutation counts in the ARG.
+Then, we identified the top 100 sites
+in terms of mutation counts in the ARG.
 Only 11 of these sites
-(635, 8835, 11074, 11083, 15521, 16887, 21304, 21305, 21575, 21987 and 28253)
+(635, 8835, 11074, 11083, 15521, 16887,
+21304, 21305, 21575, 21987 and 28253)
 were present in the problematic sites file
-(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}),
-and most sites in this list were typical in terms of the numbers of mutations.
-We therefore masked our 100 most-mutagenic sites to remove sequencing errors and improve
-matching performance in the LS HMM.
-
-Alignments read from the VCZ Zarr file were first preprocessed to mask any 
-non nucleotide characters (A,C,G or T) as missing data. 
-Then, we filtered 
-any samples that had missing data at more than 500 sites, excluding
-the 100 sites identified as problematic (this was 
-set to 10,000 sites for the initial period up to 2020-03-01 where sampling
-density is low).
+(\url{https://github.com/W-L/ProblematicSites_SARS-CoV2}),
+and most sites in this list were typical
+in terms of the numbers of mutations.
+We therefore masked our 100 mutation-rich sites
+to remove sequencing errors and
+improve matching performance in the LS HMM.
+
+Alignments read from the VCZ Zarr file were first preprocessed
+to mask any non nucleotide characters (A,C,G or T) as missing data.
+% https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml
+Then, we filtered any samples that had missing data at more than 500 sites,
+excluding the 100 sites identified as problematic
+(this was set to 10,000 sites for the initial period up to 2020-03-01
+where sampling density is low).
 
@@
 
-The LS model \cite{LiStephens2003} is an approximation of the coalescent with recombination
-that captures many of the key features of the joint processes of mutation and recombination.
-It is widely used in a range of applications in population genetics, including ARG 
+The LS model \cite{LiStephens2003} is an approximation of
+the coalescent with recombination
+that captures many of the key features of
+the joint processes of mutation and recombination.
+It is widely used in a range of applications in population genetics,
 % Does Threads use LS?
-inference \cite{Kelleher2019,Gunnarsson2024}.
-The LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states),
-which are probabilistically emitted as an imperfect mosaic of a set of genomes in a reference panel (i.e., the hidden states).
-The process of switching between the genomes in the reference panel is governed by a transition matrix,
-and mismatches between a focal genome and a genome of the reference panel (from which it is ``copied'') are permitted as emissions.
-
-In the original formulation of the LS model (see Appendix S1 of \cite{LiStephens2003}),
-the probability of switching between the hidden states (i.e., genomes in a reference panel)
+including ARG inference \cite{Kelleher2019,Gunnarsson2024}.
+The LS model is an HMM that models a focal genome
+as a sequence of nucleotides (i.e. the observed states),
+which are probabilistically emitted
+as an imperfect mosaic of a set of genomes
+in a reference panel (i.e., the hidden states).
+The process of switching between the genomes in the reference panel
+is governed by a transition matrix,
+and mismatches between a focal genome and a genome of the reference panel
+(from which it is ``copied'') are permitted as emissions.
+
+In the original formulation of the LS model
+(see Appendix S1 of \cite{LiStephens2003}),
+the probability of switching
+between the hidden states (i.e., genomes in a reference panel)
 when going from one site to the next is dependent on
@@
 Sc2ts uses the same formulation except that $n$ is set to 1.
-This special case of the LS model leads to an intuitive interpretation of the ratio
-between the probability of switching and the probability of mismatch:
+This special case of the LS model leads to an intuitive interpretation of
+the ratio between the probability of switching and the probability of mismatch:
 When considering only two copying paths, the path having $k$ mismatches
@@
 % TODO mention that this can be thought of as extended parsimony.
-This parameter thus allows us to adjust the relative importance of recombination and mutation
+This parameter thus allows us to adjust
+the relative importance of recombination and mutation
 when finding the best explanation for a given sample.
-For simplicity, we assume that any single base mutation causing a mismatch is equally probable,
-ignoring transition-transversion biases and variation of mutation rate along the genome.
-As data sampling was dense for SARS-CoV-2, this assumption was considered tolerable.
+For simplicity, we assume that any single base mutation
+causing a mismatch is equally probable,
+ignoring transition-transversion biases and
+variation of mutation rate along the genome.
+As data sampling was dense for SARS-CoV-2,
+this assumption was considered tolerable.
 
 We explored using several values of $k$,
-by running sc2ts with $k$ set to 3, 4, and 5 on the samples collected in 2020.
-We would expect that a good value for $k$ should lead to parsimonious ARGs
-that capture plausible recombination events without incurring too many mutations, particularly reversions.
+by running sc2ts with $k$ set to 3, 4, and 5
+on the samples collected in 2020.
+We would expect that a good value for $k$
+should lead to parsimonious ARGs
+that capture plausible recombination events
+without incurring too many mutations, particularly reversions.
 A value of $k$ too low could result in many artifactual recombination events,
 whereas a value of $k$ too high could miss many genuine recombination events.
-We found that ARGs built with $k$ set to 3 contained many dubious recombination events,
+We found that ARGs built with $k$ set to 3
+contained many dubious recombination events,
 but ARGs built with $k$ set to 5 failed
-to include the well supported recombination events documented in a previous study\cite{Jackson2021}.
-Hence, we decided to use $k$ set to 4 to infer the final ARG described in this study.
-
-A Viterbi algorithm can be used to find the most likely ``copying path'' of a focal sample
+to include the well supported recombination events documented
+in a previous study\cite{Jackson2021}.
+Hence, we decided to use $k$ set to 4
+to infer the final ARG described in this study.
+
+A Viterbi algorithm can be used
+to find the most likely ``copying path'' of a focal sample
 through a reference panel of samples under the LS model.
 Specifically, sc2ts uses a Viterbi algorithm modified to operate on STS,
-which was originally developed to build genome-wide human genealogies
-and is capable of scaling to biobank-scale datasets \cite{Kelleher2019}.
-Using this Viterbi algorithm, sc2ts deterministically infers the copying path of a sample
-collected on a certain date through the samples collected on earlier dates in a given ARG.
-
-The SARS-CoV-2 sequences were determined from samples collected densely in time.
-SARS-CoV-2 phylogenies have low phylogenetic signals due to high sampling frequency
+which was originally developed to build genome-wide human genealogies and
+is capable of scaling to biobank-scale datasets \cite{Kelleher2019}.
+Using this Viterbi algorithm,
+sc2ts deterministically infers the copying path of a sample
+collected on a certain date through the samples
+collected on earlier dates in a given ARG.
+
+The SARS-CoV-2 sequences were determined
+from samples collected densely in time.
+SARS-CoV-2 phylogenies have low phylogenetic signals
+due to high sampling frequency
 (such that sampling and transmission outpace nucleotide substitution).
 Therefore, most samples can be explained by a small number of mutations.
-Leveraging this property of the dataset, we developed a likelihood-thresholding strategy
-to avoid needing to compute the likelihood of all possible copying paths through an ARG for every single sample.
+Leveraging this property of the dataset,
+we developed a likelihood-thresholding strategy
+to avoid needing to compute the likelihood of
+all possible copying paths through an ARG for every single sample.
 %%% TODO: This needs to be explained.

### 2025-10-08 latex/main.tex

@@
 visualization of small-scale phylogenetic
-networks~\cite{Huson2012,Vaughan2017,Solís-Lemus2017} and
+networks~\cite{Huson2012,Vaughan2017,Solis-Lemus2017} and
 using tanglegrams to visualize tree
@@
 
-The bioinformatics community has curated a set of problematic sites
+A set of problematic sites
 for SARS-CoV-2 that are enriched for systematic sequencing errors or 
-high levels of homoplasy~\cite{DeMaio2020issues}. 
+high levels of homoplasy was compiled and updated during 
+the pandemic~\cite{DeMaio2020issues,Turakhia2020stability}. 
 Although these sites are recommended to be masked for UShER

### 2025-10-08 latex/main.tex

@@
 XZ, XAC, XAD, XAE, and XAP, which we refer to as ``Xx''
-(Figure~\ref{fig:subgraphs}B).
+(Figure~\ref{fig:subgraphs}B, Document \ref{sec:dd_xx}).
 
@@
 distant from a BA.2 sample and has little additional evidence
-to suggest recombination [ref sup section].
+to suggest recombination (Document \ref{sec:dd_xn}).
 XB, on the other hand,
@@
 \newpage
-\setcounter{secnumdepth}{2} % Print out appendix section numbers
+\setcounter{secnumdepth}{3} % Print out appendix section numbers
 
@@
 \subsubsection{XA}
+% mneumonic: deep dive XA
+\label{sec:dd_xa}
 (also illustrated in Figure 2a / 3d)
@@
 \subsubsection{XB}
+\label{sec:dd_xb}
 https://github.com/jeromekelleher/sc2ts-paper/issues/332
@@
 \subsubsection{XC}
+\label{sec:dd_xc}
 The five XC samples descend from a well supported recombination node.
@@
 \subsubsection{XE/XH}
+\label{sec:dd_xexh}
 https://github.com/jeromekelleher/sc2ts-paper/discussions/633
@@
 \subsubsection{XF}
+\label{sec:dd_xf}
 XF is a well-supported recombinant between Delta and Omicron BA.1,
@@
 \subsubsection{XG}
+\label{sec:dd_xg}
 The three XG samples descend from a reconstructed internal node,
@@
 \subsubsection{XJ}
+\label{sec:dd_xj}
 All the XJ samples descend from a well supported recombination node.
@@
 \subsubsection{XL}
+\label{sec:dd_xl}
 All the XL samples descend from a well supported recombination node,
@@
 \subsubsection{XM/XAL}
+\label{sec:dd_xm}
 https://github.com/jeromekelleher/sc2ts-paper/issues/284
@@
 \subsubsection{XN/XAU}
+\label{sec:dd_xn}
+
 (also illustrated in Figure 2c)
@@
 \subsubsection{XP}
+\label{sec:dd_xp}
+
 https://github.com/jeromekelleher/sc2ts-paper/issues/345
@@
 \subsubsection{XQ/XR/XU/XAA/XAG/XAM}
-\label{sec:supplement_indepth_pango_x_lineages_xq}
+\label{sec:dd_xq}
+
 https://github.com/jeromekelleher/sc2ts-paper/issues/338
@@
 \subsubsection{XS}
+\label{sec:dd_xs}
+
 https://github.com/jeromekelleher/sc2ts-paper/issues/287
@@
 \subsubsection{XW}
+\label{sec:dd_xw}
+
 All the XW samples descend from a well supported recombination node.
@@
 \subsubsection{XY}
+\label{sec:dd_xy}
+
 All the XY samples descend from a well supported recombination node,
@@
 \subsubsection{Xx: XZ/XAC/XAD/XAE/XAP}
+\label{sec:dd_xx}
+
 https://github.com/jeromekelleher/sc2ts-paper/discussions/631
@@
 from a single recombination node in the ARG, and
-we suggest that these Pango X lineages
+These Pango X lineages
 may be aliases of a single recombinant
-which we tentatively refer to as "Xx".
+which we tentatively refer to as ``Xx''.
 Note that all were designated around the same time,
@@
 including two that are identical
-to the proposed originating recombinant (ERR8146303 and ERR8163061):
-we suggest that researchers examining the emergence of
-these Pango X lineages could direct efforts
-into examination of these and closely related samples.
-The detection of Xx illustrates
-the ability of sc2ts to clarify relationships between
-samples during the course of a pandemic,
-informing designation of emerging recombinants.
+to the proposed originating recombinant (ERR8146303 and ERR8163061).
 
@@
 \subsubsection{XAF}
+\label{sec:dd_xaf}
+
 https://github.com/jeromekelleher/sc2ts-paper/discussions/638
@@
 \subsubsection{XAJ}
+\label{sec:dd_xaj}
+
 https://github.com/jeromekelleher/sc2ts-paper/discussions/634
@@
 \subsubsection{XAN/XAV}
+\label{sec:dd_xan}
+
 https://github.com/jeromekelleher/sc2ts-paper/discussions/783
@@
 \subsubsection{XAS}
+\label{sec:dd_xas}
+
 https://github.com/jeromekelleher/sc2ts-paper/discussions/784
@@
 \subsubsection{XAZ}
+\label{sec:dd_xaz}
+
 https://github.com/jeromekelleher/sc2ts-paper/discussions/786
@@
 \subsubsection{XBB}
+\label{sec:dd_xbb}
+
 https://github.com/jeromekelleher/sc2ts-paper/issues/336
@@
 \subsubsection{XBB.1}
+\label{sec:dd_xbb.1}
+
 XBB.1 is characterized by an additional mutation in Spike, G22317T (G252V) and
@@
 \subsubsection{XBD}
+\label{sec:dd_xbd}
+
 All the XBD samples descend from a well supported recombination node.
@@
 \subsubsection{XBE}
+\label{sec:dd_xbe}
+
 https://github.com/jeromekelleher/sc2ts-paper/issues/351
@@
 \subsubsection{XBF}
+\label{sec:dd_xbf}
+
 All the XBF samples descend from a well supported recombination node,
@@
 \subsubsection{XBG}
+\label{sec:dd_xbg}
+
 All the XBG samples descend from a well supported recombination node,
@@
 \subsubsection{XBH}
+\label{sec:dd_xbh}
 All the XBH samples descend from a well supported recombination node,
@@
 \subsubsection{XBQ/XBK/XBK.1}
+\label{sec:dd_xbq}
+
 https://github.com/jeromekelleher/sc2ts-paper/issues/349
@@
 constructed by sc2ts as part of its tree building steps.
-We suggest that sc2ts can therefore play an epidemiological role
-both in generating hypotheses involving recombination, and
-quantifying mutational evidence to support them.
 
 \subsubsection{XBM}
+\label{sec:dd_xbm}
+
 All the XBM samples descend from a well supported recombination node.
@@
 \subsubsection{XBR}
+\label{sec:dd_xbr}
 This one XBR sample descends from a well supported recombination node.

### 2025-10-09 latex/main.tex
Updates from Overleaf

@@
 \subsection*{Comparison with RIPPLES}
+
 RIPPLES~\cite{Turakhia2022} is a method for detecting 
-recombination events using a tree inferred by UShER. Recombination is 
-detected by first identifying plausible branches carrying multiple
-mutations (at least 3, by default). For each candidate branch,
-RIPPLES then systematically splits the sequence into 2 or 3
-segments, and finds parsimonious placements for each segment independently. 
-A recombination event at that node is then flagged if the overall parsimony of these multiple 
-placements sufficiently improves over the original single-parent
-origin in the tree (the ``parsimony improvement'' parameter, $p=3$, by default).
+recombination events using a tree inferred by UShER.
+Recombination is detected by first identifying plausible branches
+carrying multiple mutations (at least 3, by default).
+For each candidate branch,
+RIPPLES then systematically splits the sequence into 2 or 3 segments, and
+finds parsimonious placements for each segment independently.
+A recombination event at that node is then flagged if
+the overall parsimony of these multiple placements
+sufficiently improves over the original single-parent origin in the tree
+(the ``parsimony improvement'' parameter, $p=3$, by default).
 RIVET~\cite{Smith2023tracking} is a tool for quality control
-and visualisation of these recombination events.
+and visualization of these recombination events.
 
@@
 
-We ran fast-RIPPLES\cite{Smith2023tracking} on the Viridian UShER 
-tree subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
-We first ran RIPPLES with parameter $p=4$, as being roughly comparable to 
-the recombination penalty used in sc2ts ($k=4$; STAR Methods).
-This resulted in 1,174 recombination events reported, of which 460
-could be mapped to one of the 855 recombination events in the sc2ts ARG (STAR Methods;
+We ran fast-RIPPLES\cite{Smith2023tracking} on the Viridian UShER tree
+subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
+We first ran RIPPLES with parameter $p=4$, as being roughly comparable
+to the recombination penalty used in sc2ts ($k=4$; STAR Methods).
+This resulted in 1,174 recombination events reported, of which 460 could be mapped
+to one of the 855 recombination events in the sc2ts ARG (STAR Methods;
 Figure \ref{fig:ripples_sc2ts_events}A). Analysis of the descendant 
@@
 and their relationship with the sc2ts ARG.
-Of the 16 Class I events 
-(Pango X lineages unequivocally associated with a recombination node in
-the sc2ts ARG) 
-in Table~\ref{tab:pango_x_lineages}, 8
-have corresponding recombination events (4, at $p=4$). Of the 5 Type II
-events (recombination nodes with more complex assocation with X lineages),
-two have close matches. XBM matches exactly (including 2 descending BF.3 samples),
+Of the 16 Class I events (Pango X lineages unequivocally associated with
+a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
+8 have corresponding recombination events (4, at $p=4$).
+Of the 5 Type II events
+(recombination nodes with more complex assocation with X lineages),
+two have close matches.
+XBM matches exactly (including 2 descending BF.3 samples),
 and XE/XH is similar (but without including the 2 XH samples in the clade).
 UShER and sc2ts agree on creating two separate XAM clades, but while sc2ts
-makes these both descendants of the Xx event, RIPPLES creates two independent
-recombinations. Similarly, XAE and ??? have exactly matching clades, but RIPPLES
+makes these both descendants of the Xx event,
+RIPPLES creates two independent recombinations.
+Similarly, XAE and ??? have exactly matching clades, but RIPPLES
 infers two independent recombination events while sc2ts groups them both
@@
 
-\subsection*{Recombination intervals}
+\subsection*{Recombination breakpoint intervals}
 
@@
 
-We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
-particularly at the left boundary of the Spike gene (Figure~\ref{fig:recombinant_intervals}A,B),
-consistent with patterns observed in previous studies
-\cite{Bobay2020recombination,Turakhia2022,Smith2023tracking,Li2024CovRecomb}.
+We find an overall preponderance of breakpoint intervals
+towards the 3’ end of the genome,
+particularly at the left boundary of the Spike gene
+(Figure~\ref{fig:recombinant_intervals}A,B),
+consistent with patterns observed in previous
+studies\cite{Bobay2020recombination,Turakhia2022,
+Smith2023tracking,Li2024CovRecomb}.
 While this might result from a higher recombination rate in this region,

### 2025-10-09 latex/main.tex
Tweak recomb section of intro

@@
 Muller2020bayesian,%
+Guo2022recombination,
 Muller2022bayesian},
@@
 Software support in explicitly \emph{using}
-recombination-aware inference is limited to
-visualization of small-scale phylogenetic
-networks~\cite{Huson2012,Vaughan2017,Solis-Lemus2017} and
-using tanglegrams to visualize tree
-changes (e.g.~\cite{Rasmussen2023}),
-typically using the Extended Newick format for
-interchange~\cite{Cardona2008}.
+recombination-aware inference is mostly limited to
+visualization, for example of small-scale phylogenetic
+networks~\cite{Huson2012,Vaughan2017,Solis-Lemus2017}
+or tree changes along the genome~\cite{Rasmussen2023}.
 

### 2025-10-09 latex/main.tex
Pass through intro incorporating RIPPLES

@@
 spectra\cite{Bloom2023} and deletions\cite{Li2023indels}.
-In addition, we show that sc2ts automatically and accurately detects details about
+We show that sc2ts automatically and accurately detects details about
 known recombinant lineages that were painstakingly collated in a community 
 process, and provides novel insights into the relationships between these
-recombinant lineages. By fully incorporating these and other novel
-recombination events, the resulting data structure---representing
+recombinant lineages. We compare the recombination events detected by 
+sc2ts with the state-of-the-art RIPPLES method~\cite{Turakhia2022}, showing
+that sc2ts is more sensitive and has a substantially lower false positive rate.
+The output of sc2ts---incorporating
 genealogical relationships, point mutations, deletions, and 

### 2025-10-09 latex/main.tex
Hardware
Closes #853

@@
 enables pandemic-scale analysis on a laptop. 
-% Dunno about this! Feels important to finish on a wider note, helping
-% the reader understand the significance of all of this?
 These resources
@@
 batch to node(s) in the current ARG
-(allowing recombination) using a simplified version of the Li and Stephens (LS) model
-\cite{LiStephens2003}. 
+using a simplified version of the Li and Stephens (LS) model\cite{LiStephens2003}. 
 % Important to go into a bit of detail about LS here as it's so fundamental
@@
 collected between 2020-01-01 and 2023-02-20.
-Inference took 21 days on a XX core server with 500 GB of RAM
-($\sim 5$ CPU years; Figure~\ref{fig:inference_resources}).
+Inference took 21 days on a server with 64 cores (2X AMD EPYC 7502)
+and 512 GB of RAM ($\sim 5$ CPU years; Figure~\ref{fig:inference_resources}).
 After postprocessing, the ARG

### 2025-10-09 latex/main.tex
Incorporate new validation work into Results

@@
 
-% PARA: we compared against the UShER tree in a bunch of ways
+% PARA: we did a BUNCH of validation work. There's a lot in here, but the
+% point is to summarise a mountain of validation and provide links to the 
+% details.
+
 % NOTE: Trying to clarify that this of the pandemic period and we're 
@@
 (Document~\ref{sec:supplement_pango_lineages}).
+The ARG is time-resolved, with internal node dates estimated by
+a non-parametric method~\cite{Pope2025} (STAR Methods). 
+We show that the dates agree reasonably with NextStrain 
+estimates (Figure \ref{fig:pango_vs_nextstrain_node_times})
+and overall accuracy to another large-scale
+method~\cite{sanderson2021chronumental} (Document TODO).
+Finally, we verified
+that the origins of the 
+Alpha (Figure~\ref{fig:subgraph_alpha}, Document~\ref{sec:dd_alpha}),
+Delta (Figure~\ref{fig:subgraph_delta}, Document~\ref{sec:dd_delta})
+and
+Omicron (Figure~\ref{fig:subgraph_omicron}, Document~\ref{sec:dd_omicron})
+VOCs reflect expectations from the literature\cite{RambautAlpha2020,ECDCreport,
+Cherian2021,Farinholt2021,Stern2021,Dor2025,RambautOmicron2021}.
 
@@
 \subsubsection{Alpha}
-\label{sec:alpha}
+\label{sec:dd_alpha}
 
@@
 \subsubsection{Delta}
-\label{sec:delta}
+\label{sec:dd_delta}
 
@@
 
-Delta belongs to the “B.1.617” clade,
+Delta belongs to the ``B.1.617'' clade,
 which also includes Kappa (B.1.617.1) and B.1.617.3 \cite{ECDCreport}.
@@
 \subsubsection{Omicron BA.1 and BA.2}
-\label{sec:omicron}
+\label{sec:dd_omicron}
 

### 2025-10-09 latex/main.tex
Pass through first para of Results

@@
 \author[1,6]{Shing~H.~Zhan}
-% TODO add joint second here?
 \author[1,5]{Yan~Wong}
@@
 contains 2,689,054 nodes,  % Note the estimated dates ?
-2,748,838 edges, and 2,285,344 mutations at 29,893 sites.
-Including sample IDs and the Pango and Scorpio designations computed by
-Pangolin~\cite{OToole2021assignment} as metadata for each node,
-the ARG is stored in a 32MiB tszip format file.
-% It's confusing to talk about recombinant QC here as we're talking 
-% about the full ARG for the rest of the section.
+2,748,838 edges, and 2,285,344 mutations at 29,893 sites,
+and is stored in a 32MiB file.
 Of the nodes, 855 represent recombination events,
@@
 (some events occurred at the same breakpoints [vague, rephrase]).
-[TODO note about loading up the time it takes to get the node 
-data dataframes in a notebook?]
+% NOTE: if we keep this we should redo on Yan's laptop and quote those numbers
+Using the tskit Python API, the ARG can be loaded into 
+memory in less than a second [Concrete numbers?], and the 
+vectorised representation integrates tightly with 
+powerful data-science tools such as NumPy~\cite{Harris2020array}.
+For example, a tabular summary for the 2.7m nodes, including
+sample IDs and Pango designations (stored as metadata in the ARG),
+can be derived as a Pandas dataframe~\cite{Mckinney2010data} in 0.6s.
+The ARG can therefore be easily and efficiently analysed using familiar tools
+with minimal hardware requirements.
 
@@
 \caption{
-Recombination intervals.
-(A) Distribution of the breakpoint intervals of 354 recombination events 
-along the genome (annotated with gene labels on top).
+Recombination breakpoint intervals.
+(A) Distribution of the breakpoint intervals of
+354 recombination events along the genome
+(annotated with gene labels on top).
 Each line represents an interval colored by its size.
 (B) Number of overlapping breakpoint intervals.
-(C) FIXME Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
-(D) Relationship between breakpoint interval size and time between the recombination event
-and the parental MRCA. The black curve shows the theoretical expected relationship
-between these quantities, estimated by minimizing the sum of squared residuals
+(C) Distribution of breakpoint interval size (bases).
+(D) Relationship between breakpoint interval size and
+time between the recombination event and the parental MRCA.
+The black curve shows the theoretical expected relationship
+between these quantities,
+estimated by minimizing the sum of squared residuals
 given a single parameter (the mutation rate) (STAR Methods).}
+(E) Distribution of time interval size (months).
 \label{fig:recombinant_intervals}

### 2025-10-09 latex/main.tex
Update

@@
 We show that this ARG is highly congruent with the UShER phylogeny built 
-from the same dataset as well as the Pango lineage naming system\cite{Rambaut2020},
+from the same dataset as well as the Pango nomenclature system\cite{Rambaut2020},
 and also accurately captures evolutionary features such as mutational 
@@
 
+% The ARG is big, took a decent chunk of time to infer (but not crazy)
+% and is very easy to analyse.
 Using sc2ts, we built an ARG from 2,482,157 samples
@@
 powerful data-science tools such as NumPy~\cite{Harris2020array}.
-For example, a tabular summary for the 2.7m nodes, including
-sample IDs and Pango designations (stored as metadata in the ARG),
+For example, a tabular summary for the 2.75m nodes, including
+sample IDs and Pango assignments (stored as metadata in the ARG),
 can be derived as a Pandas dataframe~\cite{Mckinney2010data} in 0.6s.
-The ARG can therefore be easily and efficiently analysed using familiar tools
-with minimal hardware requirements.
+The ARG can therefore be easily accessed using familiar tools
+with minimal hardware requirements, enabling diverse bespoke analyses.
 
@@
 We also evaluated how well the ARG reflects the structure
-of the Pango nomenclature system~\cite{Rambaut2020} by computing
-Pango assignments for all 2.7 million nodes in the ARG using
+of the Pango nomenclature system by computing
+Pango assignments for all 2.75 million nodes in the ARG using
 Pangolin~\cite{OToole2021assignment} (STAR Methods).
@@
 and overall accuracy to another large-scale
-method~\cite{sanderson2021chronumental} (Document TODO).
+method~\cite{sanderson2021chronumental} (Document \ref{sec:evaluating_node_dates}).
 Finally, we verified
@@
 
-[TODO we should mention UShER here: they state multiple times in the GitHub issues 
-that they don't handle them at all, so I think we can just say this in some nice way.]
 Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
-but because they can be difficult to distinguish from bioinformatics errors
+but are not incorporated into UShER trees and can be difficult
+to account for. Because deletions are difficult to distinguish 
+from bioinformatics errors
 and their multi-base nature presents challenges
@@
 we masked out all gap characters as missing data during primary inference.
-To incorporate the most important deletions into the ARG,
+To approximate the effects of the most important deletions into the ARG,
 we remapped data using post hoc parsimony at 75 sites involved in deletions
@@
 sometimes the UShER tree shows greater resolution, as in the close relatives of Beta (B.1.351).
-
-\subsubsection*{Checking inferred node dates}
-
-To validate the accuracy of times for non-sample internal nodes inserted 
-into the ARG
-(inferred using tsdate), we compared the dates of first emergence of
-Nextstrain clades against the Nextstrain SARS-CoV-2 tree (downloaded on 
-2023-01-21). Figure \ref{fig:pango_vs_nextstrain_node_times} shows relatively
-close agreement for most clades, with only three 
-disagreeing by more than 28 days: 20C (an early lineage from 2020), 20E (Pango B.1.177 also known as EU1, emerging mid-2020), and 21B (Pango B.1.617.1 also known as Kappa, emerging in late 2020 to early 2021).
-
-Then, to gauge internal consistency of the time estimation procedure, we
-selected a subset of sample nodes, matching the distribution of node times 
-and average number of children of the non-sample internal nodes. These
-selected sample nodes were then re-dated, and the inferred dates compared 
-against their recorded collection dates. The results in Figure 
-\ref{fig:node_time_testing} demonstrate that although the inferred and true
-times are generally very close, there is a clear bias with re-estimated times
-being consistently older (by an average of 37 days). This is in line with 
-the findings of other recent
-work \cite{sanderson2021chronumental}, and is likely to be a consequence
-of edges carrying more mutations than would be expected under a neutral
-molecular clock prior (either due to errors in tree-building or genuinely varying mutation rates), as well as inaccuracies in the recording of 
-sample dates.
 
@@
 
+\subsection*{Evaluating inferred node dates}
+\label{sec:evaluating_node_dates}
+
+To validate the accuracy of times for non-sample internal nodes inserted 
+into the ARG
+(inferred using tsdate), we compared the dates of first emergence of
+Nextstrain clades against the Nextstrain SARS-CoV-2 tree (downloaded on 
+2023-01-21). Figure \ref{fig:pango_vs_nextstrain_node_times} shows relatively
+close agreement for most clades, with only three 
+disagreeing by more than 28 days: 20C (an early lineage from 2020), 20E (Pango B.1.177 also known as EU1, emerging mid-2020), and 21B (Pango B.1.617.1 also known as Kappa, emerging in late 2020 to early 2021).
+
+Then, to gauge internal consistency of the time estimation procedure, we
+selected a subset of sample nodes, matching the distribution of node times 
+and average number of children of the non-sample internal nodes. These
+selected sample nodes were then re-dated, and the inferred dates compared 
+against their recorded collection dates. The results in Figure 
+\ref{fig:node_time_testing} demonstrate that although the inferred and true
+times are generally very close, there is a clear bias with re-estimated times
+being consistently older (by an average of 37 days). This is in line with 
+the findings of other recent
+work \cite{sanderson2021chronumental}, and is likely to be a consequence
+of edges carrying more mutations than would be expected under a neutral
+molecular clock prior (either due to errors in tree-building or genuinely varying mutation rates), as well as inaccuracies in the recording of 
+sample dates.
+
 \subsection{Pango X lineage events}

### 2025-10-09 latex/main.tex
Tweaks

@@
 % not trying to cover *all* history.
-The ARG is a comprehensive record of the pandemic phase 
-of SARS-CoV-2 evolutionary history, and accurately reflects
+The ARG is a comprehensive and deeply validated record of the pandemic phase 
+of SARS-CoV-2 evolutionary history, accurately reflecting
 a range of results from the literature.
@@
 
-[TODO narrative is muddled here: the point of the para is to say that 
-pango X lineage designation is manual and subjective.]
 % " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
@@
 which established that recombination with onward transmission had occurred [weak FIXME].
-As of pango-designation version v1.33.1 (2025-03-31),
-a total of 146 distinct Pango X lineages have been designated.
-%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
-% 146
 Although the process of incorporating recombinant lineages
@@
 by volunteers~\cite{Smith2023tracking}.
+% Debatable whether this adds or substracts from narrative
+As of pango-designation version v1.33.1 (2025-03-31),
+a total of 146 distinct Pango X lineages have been designated.
+%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
+% 146
 
@@
 (XAC and XAD have been omitted, and XM partially represented,
-corresponding to 23 samples in total;
+corresponding to 23 omitted samples in total;
 see Document~\ref{sec:supplement_x_pango_lineages}).
@@
 and CovRecomb \cite{Li2024CovRecomb} detection methods
-(STAR Methods; Table~\ref{tab:method_concordance}).
+(STAR Methods; Table~\ref{tab:method_concordance}; see also
+Document~\ref{sec:jackson_recombs} for analysis of the 
+Jackson et al.\ recombinants).
 Of the Type I events, 13 are monophyletic for the focal Pango lineage
 (i.e., the descendants are entirely from that lineage);
-Figure~\ref{fig:subgraphs}A shows XA as an example
-(for further analysis of the other recombinants
-found by Jackson et al.\cite{Jackson2021},
-see Table~\ref{tab:jackson} in Document~\ref{sec:jackson_recombs},
-which shows that sc2ts identified the parents and breakpoints
-concordant with those reported by Jackson et al.\cite{Jackson2021}).
+Figure~\ref{fig:subgraphs}A shows XA as an example.
 
@@
 that they are more parsimoniously explained as sublineages of XQ
-than de-novo recombinant lineages.
+than de-novo recombinant lineages (Document \ref{sec:dd_xq}). 
 The second cluster is based around a putative undocumented
@@
 may have misclassified by sc2ts due to limitations in
-handling deletions or matching against long branches (see Discussion).
+handling deletions or matching against long branches (Document \ref{sec:dd_xb}). 
 See Document~\ref{sec:supplement_indepth_pango_x_lineages}
-for extended analyses on all of the Pango X lineages above,
+for extended analyses on all of the Pango X lineages,
 Figure~\ref{fig:pango_x_copying_patterns} for copying patterns
 associated with Table~\ref{tab:pango_x_lineages}, and
-Document S2 for ARG visualizations.
+Document S2 for ARG visualizations of the Pango X lineages.
 

### 2025-10-09 latex/main.tex
WIP on RIPPLES

@@
 \subsection*{Comparison with RIPPLES}
-
 RIPPLES~\cite{Turakhia2022} is a method for detecting 
-recombination events using a tree inferred by UShER.
+recombination events using a tree inferred by UShER,
+and is the only method directly comparable with sc2ts.
 Recombination is detected by first identifying plausible branches
-carrying multiple mutations (at least 3, by default).
+carrying multiple mutations .
 For each candidate branch,
@@
 
-\begin{table}
-\caption{
-RIPPLES events associated with Pango X lineages.
-Each row is a recombination event identified by RIPPLEs (at p = 3)
-in which Pango X lineages appear,
-along with the corresponding sc2ts event from Table~\ref{tab:pango_x_lineages},
-where applicable, grouped by class.
-Events also present at p = 4 are indicated.
-Clades are matched by finding the most recent common ancestor of
-the samples associated with each UShER/RIPPLES event in the sc2ts ARG.
-The clade differences indicates how close this match is.
-}
-\label{tab:ripples_x_lineages}
-\centering
-\begin{scriptsize}
-
-\begin{tabular}{lrrlrl}
-\toprule
-sc2ts event & in p4 & UShER desc & sc2ts class & clade diff & descendants \\
-\midrule
-\multicolumn{6}{l}{Equivalent Sc2ts and RIPPLEs recombination events}\\
-XBR & $\checkmark$ & 1 & I & 0 & {XBR:1} \\
-% XM3 & $\checkmark$ & 1 & I & 0 & {XM:1} \\
-XC & $\checkmark$ & 5 & I & 0 & {XC:5} \\
-XS & $\checkmark$ & 17 & I & 0 & {XS:17} \\
-XY &  & 23 & I & 0 & {XY:23} \\
-XW &  & 32 & I & 0 & {XW:32} \\
-XA & $\checkmark$ & 39 & I & 0 & {XA:39} \\
-XL &  & 64 & I & 0 & {XL:64} \\
-XBB &  & 6452 & I & 1 & {XBB:2361} \\
-XBM & $\checkmark$ & 12 & II & 0 & {XBM:10, BF.3:2} \\
-\midrule
-\multicolumn{6}{l}{RIPPLEs recombination event close to sc2ts recombination event}\\
-XE/XH &  & 1113 & II & 3 & {XE:1113} \\
-XQ &  & 1 & III & 0 & {XAM:1} \\
-XQ &  & 21 & III & 0 & {XAM:21} \\
-Xx &  & 9 & III & 0 & {XAE:9} \\
-Xx & $\checkmark$ & 18 & III & 0 & {XAC:18} \\
-XM & $\checkmark$ & 8 & III & 0 & {BA.2:4, XM:4} \\
-\midrule
-\multicolumn{6}{l}{RIPPLEs recombination event equivalant to sc2ts non-recombination event}\\
-XB & $\checkmark$ & 411 & IV & 0 & {B.1:219, XB:192} \\
-\midrule
-\multicolumn{6}{l}{RIPPLEs recombination event not comparable to sc2ts}\\
-NA & $\checkmark$ & 19 & NA & 45190 & {XAJ:18, BA.4:1} \\
-NA &  & 20 & NA & 1061775 & {BA.1.15:1, XS:17, BA.1:1, BA.1.1:1} \\
-NA & $\checkmark$ & 35 & NA & 7147 & {XBD:30, BA.2.75:5} \\
-NA & $\checkmark$ & 36 & NA & 7146 & {XBD:30, XBR:1, BA.2.75:5} \\
-NA &  & 1321 & NA & 340501 & {XE:1113, XH:2, XJ:68, XY:23, BA.2:82, XM:29,
-XAL:3, XAF:1} \\
-\bottomrule
-\end{tabular}
-
-\end{scriptsize}
-\end{table}
-
 We ran fast-RIPPLES\cite{Smith2023tracking} on the Viridian UShER tree
 subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
-We first ran RIPPLES with parameter $p=4$, as being roughly comparable
-to the recombination penalty used in sc2ts ($k=4$; STAR Methods).
-This resulted in 1,174 recombination events reported, of which 460 could be mapped
-to one of the 855 recombination events in the sc2ts ARG (STAR Methods;
+At $p=3$, this resulted in a total of 4111 recombination events
+of which 595 could be mapped to one of the 855 recombination events in the 
+sc2ts ARG.
+
 Figure \ref{fig:ripples_sc2ts_events}A). Analysis of the descendant 
@@
 sc2ts ARG (Figure \ref{fig:ripples_sc2ts_events}B).
+Analysis of the descending samples for each event shows that 
+10 Pango of the 44 X lineages present in the dataset
+% NOTE: listing them like this for sorting and manipulation purposes
+(%
+XA, 
+XAC, 
+XAE, 
+XAJ,
+XB, 
+XBB, 
+XBM, 
+XC, 
+XL, 
+XW,%
+)
+are associated unequivocally with a single recombination event.
+
+XBR, XBD 
+XS appears twice
+
+Another event suggests that 
+XAF 
+XAL, 
+XE, 
+XH, 
+XJ, 
+XM, 
+XY, 
+and 
+82 BA.2 samples share a recombination event [With what breakpoint and 
+parents?].
 
@@
 or the sc2ts ARG.
+
+Figure \ref{fig:ripples_sc2ts_events}A). 
+See Document\ref{sec:detailed_ripples} and Table~\ref{tab:ripples_x_lineages}
+for detailed analysis.
 
@@
 
+\begin{table}
+\caption{
+RIPPLES events associated with Pango X lineages.
+Each row is a recombination event identified by RIPPLEs (at p = 3)
+in which Pango X lineages appear,
+along with the corresponding sc2ts event from Table~\ref{tab:pango_x_lineages},
+where applicable, grouped by class.
+Events also present at p = 4 are indicated.
+Clades are matched by finding the most recent common ancestor of
+the samples associated with each UShER/RIPPLES event in the sc2ts ARG.
+The clade differences indicates how close this match is.
+}
+\label{tab:ripples_x_lineages}
+\centering
+\begin{scriptsize}
+
+\begin{tabular}{lrrlrl}
+\toprule
+sc2ts event & in p4 & UShER desc & sc2ts class & clade diff & descendants \\
+\midrule
+\multicolumn{6}{l}{Equivalent Sc2ts and RIPPLEs recombination events}\\
+XBR & $\checkmark$ & 1 & I & 0 & {XBR:1} \\
+% XM3 & $\checkmark$ & 1 & I & 0 & {XM:1} \\
+XC & $\checkmark$ & 5 & I & 0 & {XC:5} \\
+XS & $\checkmark$ & 17 & I & 0 & {XS:17} \\
+XY &  & 23 & I & 0 & {XY:23} \\
+XW &  & 32 & I & 0 & {XW:32} \\
+XA & $\checkmark$ & 39 & I & 0 & {XA:39} \\
+XL &  & 64 & I & 0 & {XL:64} \\
+XBB &  & 6452 & I & 1 & {XBB:2361} \\
+XBM & $\checkmark$ & 12 & II & 0 & {XBM:10, BF.3:2} \\
+\midrule
+\multicolumn{6}{l}{RIPPLEs recombination event close to sc2ts recombination event}\\
+XE/XH &  & 1113 & II & 3 & {XE:1113} \\
+XQ &  & 1 & III & 0 & {XAM:1} \\
+XQ &  & 21 & III & 0 & {XAM:21} \\
+Xx &  & 9 & III & 0 & {XAE:9} \\
+Xx & $\checkmark$ & 18 & III & 0 & {XAC:18} \\
+XM & $\checkmark$ & 8 & III & 0 & {BA.2:4, XM:4} \\
+\midrule
+\multicolumn{6}{l}{RIPPLEs recombination event equivalant to sc2ts non-recombination event}\\
+XB & $\checkmark$ & 411 & IV & 0 & {B.1:219, XB:192} \\
+\midrule
+\multicolumn{6}{l}{RIPPLEs recombination event not comparable to sc2ts}\\
+NA & $\checkmark$ & 19 & NA & 45190 & {XAJ:18, BA.4:1} \\
+NA &  & 20 & NA & 1061775 & {BA.1.15:1, XS:17, BA.1:1, BA.1.1:1} \\
+NA & $\checkmark$ & 35 & NA & 7147 & {XBD:30, BA.2.75:5} \\
+NA & $\checkmark$ & 36 & NA & 7146 & {XBD:30, XBR:1, BA.2.75:5} \\
+NA &  & 1321 & NA & 340501 & {XE:1113, XH:2, XJ:68, XY:23, BA.2:82, XM:29,
+XAL:3, XAF:1} \\
+\bottomrule
+\end{tabular}
+
+\end{scriptsize}
+\end{table}
+
+\subsection{Detailed RIPPLES analysis}
+\label{sec:detailed_ripples}
+
+We ran fast-RIPPLES\cite{Smith2023tracking} on the Viridian UShER tree
+subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
+We first ran RIPPLES with parameter $p=4$, as being roughly comparable
+to the recombination penalty used in sc2ts ($k=4$; STAR Methods).
+This resulted in 1,174 recombination events reported, of which 460 could be mapped
+to one of the 855 recombination events in the sc2ts ARG (STAR Methods;
+Figure \ref{fig:ripples_sc2ts_events}A). Analysis of the descendant 
+samples showed that many Pango X lineages were not associated with
+recombination events (including XBB), and we therefore ran
+RIPPLES with the more lenient value of $p=3$ (the default).
+This resulted in a total of 4111 recombination events
+of which 595 have corresponding recombination events in the 
+sc2ts ARG (Figure \ref{fig:ripples_sc2ts_events}B).
+
+Table~\ref{tab:ripples_x_lineages} shows a summary of all
+recombination events associated with Pango X lineages detected by RIPPLES,
+and their relationship with the sc2ts ARG.
+Of the 16 Class I events (Pango X lineages unequivocally associated with
+a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
+8 have corresponding recombination events (4, at $p=4$).
+Of the 5 Type II events
+(recombination nodes with more complex assocation with X lineages),
+two have close matches.
+XBM matches exactly (including 2 descending BF.3 samples),
+and XE/XH is similar (but without including the 2 XH samples in the clade).
+UShER and sc2ts agree on creating two separate XAM clades, but while sc2ts
+makes these both descendants of the Xx event,
+RIPPLES creates two independent recombinations.
+Similarly, XAE and ??? have exactly matching clades, but RIPPLES
+infers two independent recombination events while sc2ts groups them both
+under the Xx event. XB is notable as a likely sc2ts false negative that is 
+identified by RIPPLES at both $p=3$ and $p=4$. The remaining events 
+associated with X lineages output by RIPPLES cannot be easily mapped 
+to the sc2ts ARG as one or more descendants is distantly placed.
+XAJ is an outlier for sc2ts, being the Pango X lineage identified as a 
+non-recombinant with the largest number of mutations 
+(Table~\ref{tab:pango_x_lineages}), and a likely false negative.
+The remaining 4 events are mixtures of multiple lineages that are 
+difficult to reconcile with either the other events in the table 
+or the sc2ts ARG.
+
 \subsection{Origins of major VOCs}

### 2025-10-09 latex/main.tex
Updates from Overleaf

--- 2025-10-09 latex/main.tex
@@
 \author[1,6]{Shing~H.~Zhan}
+% TODO add joint second here?
 \author[1,5]{Yan~Wong}
@@
 We show that this ARG is highly congruent with the UShER phylogeny built 
-from the same dataset as well as the Pango nomenclature system\cite{Rambaut2020},
+from the same dataset as well as the Pango lineage naming system\cite{Rambaut2020},
 and also accurately captures evolutionary features such as mutational 
@@
 
-% The ARG is big, took a decent chunk of time to infer (but not crazy)
-% and is very easy to analyse.
 Using sc2ts, we built an ARG from 2,482,157 samples
@@
 contains 2,689,054 nodes,  % Note the estimated dates ?
-2,748,838 edges, and 2,285,344 mutations at 29,893 sites,
-and is stored in a 32MiB file.
+2,748,838 edges, and 2,285,344 mutations at 29,893 sites.
+Including sample IDs and the Pango and Scorpio designations computed by
+Pangolin~\cite{OToole2021assignment} as metadata for each node,
+the ARG is stored in a 32MiB tszip format file.
+% It's confusing to talk about recombinant QC here as we're talking 
+% about the full ARG for the rest of the section.
 Of the nodes, 855 represent recombination events,
@@
 (some events occurred at the same breakpoints [vague, rephrase]).
-% NOTE: if we keep this we should redo on Yan's laptop and quote those numbers
-Using the tskit Python API, the ARG can be loaded into 
-memory in less than a second [Concrete numbers?], and the 
-vectorised representation integrates tightly with 
-powerful data-science tools such as NumPy~\cite{Harris2020array}.
-For example, a tabular summary for the 2.75m nodes, including
-sample IDs and Pango assignments (stored as metadata in the ARG),
-can be derived as a Pandas dataframe~\cite{Mckinney2010data} in 0.6s.
-The ARG can therefore be easily accessed using familiar tools
-with minimal hardware requirements, enabling diverse bespoke analyses.
+[TODO note about loading up the time it takes to get the node 
+data dataframes in a notebook?]
 
@@
 We also evaluated how well the ARG reflects the structure
-of the Pango nomenclature system by computing
-Pango assignments for all 2.75 million nodes in the ARG using
+of the Pango nomenclature system~\cite{Rambaut2020} by computing
+Pango assignments for all 2.7 million nodes in the ARG using
 Pangolin~\cite{OToole2021assignment} (STAR Methods).
@@
 and overall accuracy to another large-scale
-method~\cite{sanderson2021chronumental} (Document \ref{sec:evaluating_node_dates}).
+method~\cite{sanderson2021chronumental} (Document TODO).
 Finally, we verified
@@
 
+[TODO we should mention UShER here: they state multiple times in the GitHub issues 
+that they don't handle them at all, so I think we can just say this in some nice way.]
 Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
-but are not incorporated into UShER trees and can be difficult
-to account for. Because deletions are difficult to distinguish 
-from bioinformatics errors
+but because they can be difficult to distinguish from bioinformatics errors
 and their multi-base nature presents challenges
@@
 we masked out all gap characters as missing data during primary inference.
-To approximate the effects of the most important deletions into the ARG,
+To incorporate the most important deletions into the ARG,
 we remapped data using post hoc parsimony at 75 sites involved in deletions
@@
 sometimes the UShER tree shows greater resolution, as in the close relatives of Beta (B.1.351).
+
+\subsubsection*{Checking inferred node dates}
+
+To validate the accuracy of times for non-sample internal nodes inserted 
+into the ARG
+(inferred using tsdate), we compared the dates of first emergence of
+Nextstrain clades against the Nextstrain SARS-CoV-2 tree (downloaded on 
+2023-01-21). Figure \ref{fig:pango_vs_nextstrain_node_times} shows relatively
+close agreement for most clades, with only three 
+disagreeing by more than 28 days: 20C (an early lineage from 2020), 20E (Pango B.1.177 also known as EU1, emerging mid-2020), and 21B (Pango B.1.617.1 also known as Kappa, emerging in late 2020 to early 2021).
+
+Then, to gauge internal consistency of the time estimation procedure, we
+selected a subset of sample nodes, matching the distribution of node times 
+and average number of children of the non-sample internal nodes. These
+selected sample nodes were then re-dated, and the inferred dates compared 
+against their recorded collection dates. The results in Figure 
+\ref{fig:node_time_testing} demonstrate that although the inferred and true
+times are generally very close, there is a clear bias with re-estimated times
+being consistently older (by an average of 37 days). This is in line with 
+the findings of other recent
+work \cite{sanderson2021chronumental}, and is likely to be a consequence
+of edges carrying more mutations than would be expected under a neutral
+molecular clock prior (either due to errors in tree-building or genuinely varying mutation rates), as well as inaccuracies in the recording of 
+sample dates.
 
@@
 
-\subsection*{Evaluating inferred node dates}
-\label{sec:evaluating_node_dates}
-
-To validate the accuracy of times for non-sample internal nodes inserted 
-into the ARG
-(inferred using tsdate), we compared the dates of first emergence of
-Nextstrain clades against the Nextstrain SARS-CoV-2 tree (downloaded on 
-2023-01-21). Figure \ref{fig:pango_vs_nextstrain_node_times} shows relatively
-close agreement for most clades, with only three 
-disagreeing by more than 28 days: 20C (an early lineage from 2020), 20E (Pango B.1.177 also known as EU1, emerging mid-2020), and 21B (Pango B.1.617.1 also known as Kappa, emerging in late 2020 to early 2021).
-
-Then, to gauge internal consistency of the time estimation procedure, we
-selected a subset of sample nodes, matching the distribution of node times 
-and average number of children of the non-sample internal nodes. These
-selected sample nodes were then re-dated, and the inferred dates compared 
-against their recorded collection dates. The results in Figure 
-\ref{fig:node_time_testing} demonstrate that although the inferred and true
-times are generally very close, there is a clear bias with re-estimated times
-being consistently older (by an average of 37 days). This is in line with 
-the findings of other recent
-work \cite{sanderson2021chronumental}, and is likely to be a consequence
-of edges carrying more mutations than would be expected under a neutral
-molecular clock prior (either due to errors in tree-building or genuinely varying mutation rates), as well as inaccuracies in the recording of 
-sample dates.
-
 \subsection{Pango X lineage events}

### 2025-10-09 latex/main.tex

--- 2025-10-09 latex/main.tex
@@
 subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
-At $p=3$, this resulted in a total of 4111 recombination events
-of which 595 could be mapped to one of the 855 recombination events in the 
-sc2ts ARG.
-
-Figure \ref{fig:ripples_sc2ts_events}A). Analysis of the descendant 
-samples showed that many Pango X lineages were not associated with
-recombination events (including XBB), and we therefore ran
-RIPPLES with the more lenient value of $p=3$ (the default).
-This resulted in a total of 4111 recombination events
-of which 595 have corresponding recombination events in the 
-sc2ts ARG (Figure \ref{fig:ripples_sc2ts_events}B).
+At $p=3$, this resulted in a total of 4111 recombination events.
+Most of these subtend $<10$ samples (2616 singletons), with 4 events subtending 
+more than 100,000 samples.
 Analysis of the descending samples for each event shows that 
-10 Pango of the 44 X lineages present in the dataset
+10 of the 44 Pango X lineages present in the dataset
 % NOTE: listing them like this for sorting and manipulation purposes
@@
 XL, 
-XW,%
+XW%
 )
 are associated unequivocally with a single recombination event.
-
-XBR, XBD 
-XS appears twice
-
-Another event suggests that 
-XAF 
+7 lineages 
+(%
+XAL,
+XAM,
+XBD, 
+XBR, 
+XE,
+XS,
+XY%
+)
+are associated with two events. Finally, 
+one event subtending 1321 samples suggests that 
+XAF,
 XAL, 
@@
 parents?].
-
+We also ran RIPPLES with $p=4$, which reduced the overall number of 
+events to 1174 (roughly 4-fold) but also approximately halved the number of 
+events involving X lineages, including well-characterised events 
+such as XBB (Table~\ref{tab:ripples_x_lineages}).
+
+To compare these results with the sc2ts ARG, we first matched clades
+by finding the most recent common ancestor (MRCA) of the samples
+descending from each RIPPLES event. Overall, clade agreement was 
+high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG where 
+descendants match exactly. Of these nodes, 68\% had no recombinant
+ancestry. Of the remainder, 595 could be matched with one of the 
+855 recombination events in the sc2ts ARG
+(Figure \ref{fig:ripples_sc2ts_events}).
 Table~\ref{tab:ripples_x_lineages} shows a summary of all
@@
 (recombination nodes with more complex assocation with X lineages),
-two have close matches.
-XBM matches exactly (including 2 descending BF.3 samples),
+two have close matches, with XBM matching exactly (including 2 descending BF.3 samples),
 and XE/XH is similar (but without including the 2 XH samples in the clade).
-UShER and sc2ts agree on creating two separate XAM clades, but while sc2ts
-makes these both descendants of the Xx event,
-RIPPLES creates two independent recombinations.
-Similarly, XAE and ??? have exactly matching clades, but RIPPLES
-infers two independent recombination events while sc2ts groups them both
-under the Xx event. XB is notable as a likely sc2ts false negative that is 
-identified by RIPPLES at both $p=3$ and $p=4$. The remaining events 
-associated with X lineages output by RIPPLES cannot be easily mapped 
-to the sc2ts ARG as one or more descendants is distantly placed.
-XAJ is an outlier for sc2ts, being the Pango X lineage identified as a 
-non-recombinant with the largest number of mutations 
-(Table~\ref{tab:pango_x_lineages}), and a likely false negative.
-The remaining 4 events are mixtures of multiple lineages that are 
-difficult to reconcile with either the other events in the table 
-or the sc2ts ARG.
-
-Figure \ref{fig:ripples_sc2ts_events}A). 
-See Document\ref{sec:detailed_ripples} and Table~\ref{tab:ripples_x_lineages}
-for detailed analysis.
+Two events in the RIPPLES set correspond to likely false negatives 
+in sc2ts: XB (Document~\ref{sec:dd_xb}) 
+and XAJ (Document~\ref{sec:dd_xaj}).
+See Document~\ref{sec:detailed_ripples} for more detailed analysis.
 
@@
 the Jackson et al. recombinants (Document\ref{sec:jackson_recombs}).
-
-[Rough first draft]
-This estimate for effective recombination in SARS-CoV-2 is much lower 
-than previous studies have suggested.
-% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
-% recombinant origins out of the first 87,695 genomes to be released
-In an early analysis, 225 potentially recombinant sequences out 
-of 87,695 samples [WHAT DATE?]\cite{Varabyou2021rapid}.
-Over the same time period, XX events are present in the sc2ts ARG,
-representing a total of XX samples.
-In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
-589 unique recombination events in a tree of 1.6 million samples
-up to May 2021 were identified after extensive quality control.
-These nodes had a combined total of 43,104 descendants, suggesting
-that approximately 2.7\% of genomes were of recombinant origin.
-Over the same time period, the sc2ts ARG contains 14 recombinants
-passing QC with a total of 172 descendants. The ARG contains a total
-501,235 samples over this time period, giving an overall total fraction of
-0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
-events and 79-fold descendant samples over the same time period.
-The numbers are not directly comparable as the published RIPPLES 
-estimate is based on GISAID data which is enriched for
-artefacts~\cite{Hunt2024}, but given the observations in 
-the previous section on running RIPPLES on the Viridian data
-it seems likely that RIPPLES is overestimating the prevalence
-of recombination in SARS-CoV-2.
-
-% CovRecomb:  135,567 putative recombinants across the entirety of 14.5 million
-% accessed SARS-CoV-2 genomes
-
-% Although multiple studies have 
-% published lists of putatitive novel recombinant 
-% sequences~\cite{,Turakhia2022,Li2024CovRecomb} 
-% these are for GISAID data and not trival to reconcile with 
-% the Viridian data or the specific recombination events in the ARG.
-
-% RIPPLES: In a 1.6 million sample tree from May 2021, we identify 589
-% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
-% recombinant origins out of the first 87,695 genomes to be released
-% recombination events, 
-% RecombinHunt just looks at X lineages (I think?)
 
@@
 enabling and accelerating studies of SARS-CoV-2 for years to come.
+
+\subsection*{Prevalance of recombination in SARS-CoV-2}
+[Rough first draft taken from earlier location. Sticking in a heading to
+make it clear it's distinct from other paragraphs.]
+
+This estimate for effective recombination in SARS-CoV-2 is much lower 
+than previous studies have suggested.
+% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
+% recombinant origins out of the first 87,695 genomes to be released
+In an early analysis, 225 potentially recombinant sequences out 
+of 87,695 samples [WHAT DATE?]\cite{Varabyou2021rapid}.
+Over the same time period, XX events are present in the sc2ts ARG,
+representing a total of XX samples.
+In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
+589 unique recombination events in a tree of 1.6 million samples
+up to May 2021 were identified after extensive quality control.
+These nodes had a combined total of 43,104 descendants, suggesting
+that approximately 2.7\% of genomes were of recombinant origin.
+Over the same time period, the sc2ts ARG contains 14 recombinants
+passing QC with a total of 172 descendants. The ARG contains a total
+501,235 samples over this time period, giving an overall total fraction of
+0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
+events and 79-fold descendant samples over the same time period.
+The numbers are not directly comparable as the published RIPPLES 
+estimate is based on GISAID data which is enriched for
+artefacts~\cite{Hunt2024}, but given the observations in 
+the previous section on running RIPPLES on the Viridian data
+it seems likely that RIPPLES is overestimating the prevalence
+of recombination in SARS-CoV-2.
+
+% CovRecomb:  135,567 putative recombinants across the entirety of 14.5 million
+% accessed SARS-CoV-2 genomes
+
+% Although multiple studies have 
+% published lists of putatitive novel recombinant 
+% sequences~\cite{,Turakhia2022,Li2024CovRecomb} 
+% these are for GISAID data and not trival to reconcile with 
+% the Viridian data or the specific recombination events in the ARG.
+
+% RIPPLES: In a 1.6 million sample tree from May 2021, we identify 589
+% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
+% recombinant origins out of the first 87,695 genomes to be released
+% recombination events, 
+% RecombinHunt just looks at X lineages (I think?)
 
@@
 
+\subsection{Detailed RIPPLES analysis}
+\label{sec:detailed_ripples}
+
 \begin{table}
@@
 
-\subsection{Detailed RIPPLES analysis}
-\label{sec:detailed_ripples}
+[NOTE: this is an earlier draft of main text which explains the table in
+detail]
 

### 2025-10-09 latex/main.tex
Updates from Overleaf

@@
 Classical phylogenetics is largely concerned with inferring the evolutionary history 
-of species~\cite{Felsenstein2004}, with small numbers of widely diverged
-samples. In this context, recombination is a nuisance which must be 
+of species~\cite{Felsenstein2004}, using small numbers of widely diverged
+genomes. In this context, recombination is a nuisance which must be 
 carefully accounted for~\cite{Rannala2025recombination},
@@
 Modern pathogen datasets are not sparsely sampled from different species,
-but consist of tens to hundreds of thousands of samples from 
+but consist of tens to hundreds of thousands of genomes sampled from 
 a single species~\cite{
@@
 is a major issue and practical application has been limited.
-Software support in explicitly \emph{using}
+Software support that explicitly \emph{uses}
 recombination-aware inference is mostly limited to
@@
 
-The data collected during the COVID-19 pandemic as part of global
-surveillance presents fundamental challenges to existing approaches
-in two ways. Firstly, the sheer scale of data collected 
+The global
+surveillance data collected during the COVID-19 pandemic presents fundamental challenges to existing approaches
+in two ways. Firstly, the sheer amount of data 
 overwhelmed even approximate distance-matrix based methods.
@@
 Inference took 21 days on a server with 64 cores (2X AMD EPYC 7502)
-and 512 GB of RAM ($\sim 5$ CPU years; Figure~\ref{fig:inference_resources}).
+and 512 GB of RAM ($\sim$5 CPU years; Figure~\ref{fig:inference_resources}).
 After postprocessing, the ARG
@@
 XAA, XAG, XAM, XR and XU are all closely derived, suggesting 
-that they are more parsimoniously explained as sublineages of XQ
+that they are more parsimoniously explained as sublineages or sisters of XQ
 than de-novo recombinant lineages (Document \ref{sec:dd_xq}). 
@@
 events to 1174 (roughly 4-fold) but also approximately halved the number of 
-events involving X lineages, including well-characterised events 
+events involving X lineages, missing well-characterised events 
 such as XBB (Table~\ref{tab:ripples_x_lineages}).
@@
 (Table~\ref{tab:pango_x_lineages}) and
-the Jackson et al. recombinants (Document\ref{sec:jackson_recombs}).
+the Jackson et al. recombinants (Document~\ref{sec:jackson_recombs}).
 
@@
 showing the informative genomic positions between 
-a recombinant and its parent sequences, and highlights 
+a recombinant and its parent sequences, and highlighting 
 potentially problematic loci.
@@
 rewire it post-hoc, owning to subsequent tree building) and the BA.2 event (doesn't
-pass QC.]
+pass QC.) - note this is discussed in more detail in document \ref{sec:spurious-ancestrally-important-recombinants}]
 
@@
 
-We explored the spatial distribution of the breakpoint locations of
-the QC-passing recombination events along the genome,
-and quantified how precisely breakpoints were localized.
-For each event, we computed an interval estimate of its breakpoint location,
-using its size as a measure of precision (median, 2,207 bases; STAR Methods).
-
-We find an overall preponderance of breakpoint intervals
+Although the exact location of a recombination breakpoint
+is impossible to pinpoint in practice,
+bounds can be placed by examining the parental and recombinant sequences.
+% Cite VirusRecomb
+In methods \cite{Alfonsi2024,Li2024CovRecomb}, and
+approaches (e.g., Pango designation)
+based on lineage-associated mutations,
+the breakpoints are limited in precision to these mutations.
+Similarly, methods based on probabilistic representations of
+% Cite Bolotie
+mutational profiles (e.g., Bolotie) suffer from this limitation.
+% Does RIPPLES qualify as a genealogical method?
+Genealogical methods, such as sc2ts and RIPPLES,
+based on pinpointing exact parental \emph{sequences},
+are in principle maximally precise,
+assuming that the parental sequences
+are well represented in the ARG.
+
+We explored the genomic distribution of the breakpoint intervals of
+the QC-passing recombination events in the ARG.
+We found an overall preponderance of breakpoint intervals
 towards the 3’ end of the genome,
@@
 (Figure~\ref{fig:recombinant_intervals}A,B),
-consistent with patterns observed in previous
-studies\cite{Bobay2020recombination,Turakhia2022,
+consistent with the patterns seen in previous
+studies \cite{Bobay2020recombination,Turakhia2022,
 Smith2023tracking,Li2024CovRecomb}.
@@
 
-Furthermore, we find a negative relationship between
-breakpoint interval size and the amount of divergence between recombinant parents,
-which is measured by time to their MRCA (Figure~\ref{fig:recombinant_intervals}).
-When the two parental lineages have diverged for longer periods of time (the rightmost points),
-more information becomes available on either side of a breakpoint to pinpoint its location,
+Using the size of a breakpoint interval as a proxy,
+we quantified how precisely breakpoints could be
+localized along the genome.
+Across the QC-passing recombination events,
+we found the median interval size to be 2,207 bases
+(Figure~\ref{fig:recombinant_intervals}C),
+indicating that the breakpoints of half of the events
+could be narrowed to a small region (less than ~7\% of the genome),
+while the uncertainty around the breakpoints of the other events
+is considerably greater.
+We further found a negative relationship between
+breakpoint interval size and
+the amount of divergence between recombinant parents,
+which is measured by time to their MRCA
+(Figure~\ref{fig:recombinant_intervals}D).
+When the two parental lineages have diverged
+for longer periods of time (the rightmost points),
+more information becomes available
+on either side of a breakpoint to pinpoint its location,
 resulting in shorter intervals.
-For instance, the breakpoint intervals tend to be the smallest for events involving recombination
-between the deeply divergent Omicron BA.1 and BA.2 sublineages
-(the red points in Figure~\ref{fig:recombinant_intervals}).
+For instance, the breakpoint intervals tend to be the smallest
+for events involving recombination between
+the deeply divergent Omicron BA.1 and BA.2 sublineages
+(the red points in Figure~\ref{fig:recombinant_intervals}D).
+
+% Not interesting enough
+Another key feature of genealogical methods
+is that they can also provide \emph{time} intervals
+reflecting uncertainty in the timing of recombination events.
+We calculated the time interval of a recombination node
+as the difference between the age of the youngest parent and
+the age of the oldest child of the recombination node in the ARG
+(either given as collection dates or estimated using tsdate).
+We found that for three quarters of the QC-passing events,
+the time interval is less than ~2 months
+(Figure~\ref{fig:recombinant_intervals}D),
+% Do we want to refer to the CIs around the node dates?
+% Maybe we could use this result to lead to the next section?
+% Connect to epidemiology a bit?
+indicating that the ARG can precisely estimate
+the emergence of recombination events.
 

### 2025-10-09 latex/main.tex
Updates from Overleaf

@@
 response to the COVID-19 pandemic, forming a dataset of unprecedented
-depth and richness.
+richness.
 Estimated genealogies are fundamental to
@@
 Recombinant lineages have risen to global prevalence in SARS-CoV-2,
-and therefore such single-tree representations are necessarily
+and therefore simple tree representations are necessarily
 incomplete and potentially misleading.
@@
 These are the Type I events, in which the ARG reflects the
-expected phylogenetic structure of Pango designation.
+hierarchical structure of Pango designation.
 Among these events, concordance with the community-consensus 

### 2025-10-09 latex/main.tex

@@
 
+\newcommand{\scghdiscussion}[1]{%
+See \href{https://github.com/jeromekelleher/sc2ts-paper/discussions/#1}%
+{github:sc2ts-paper/discussions/#1} for further details.}
+
 \subsubsection{XA}
@@
 \label{sec:dd_xa}
-(also illustrated in Figure 2a / 3d)
 XA traces to a well supported recombination node, with no reversions.
-Note that the “causal” sample ERR5308556
+Note that the ``causal'' sample ERR5308556
 (which triggers the initial recombination event)
@@
 \label{sec:dd_xb}
-https://github.com/jeromekelleher/sc2ts-paper/issues/332
 
@@
 they are likely to attach somewhere along the long branch
-in the subgraph visualization (with > 30 mutations),
-which also contains the key "R2" deletion
+in the subgraph visualization (with $> 30$ mutations),
+which also contains the key ``R2'' deletion
 associated with XB / B.1.634 \cite{Gutierrez2022}.
@@
 
+\scghdiscussion{780}
+
 \subsubsection{XC}
@@
 \label{sec:dd_xexh}
-https://github.com/jeromekelleher/sc2ts-paper/discussions/633
 
@@
 
+\scghdiscussion{633}
+
 \subsubsection{XF}
@@
 \label{sec:dd_xm}
-https://github.com/jeromekelleher/sc2ts-paper/issues/284
 
@@
 which is characteristic of BA.2 but not BA.1
-(https://github.com/cov-lineages/pango-designation/issues/361).
+\url{https://github.com/cov-lineages/pango-designation/issues/361}.
 Most of the XM-labeled samples in ARG have this deletion,
@@
 
+\scghdiscussion{656}
+
 \subsubsection{XN/XAU}
 \label{sec:dd_xn}
-
-(also illustrated in Figure 2c)
-https://github.com/jeromekelleher/sc2ts-paper/issues/348
-https://github.com/jeromekelleher/sc2ts-paper/issues/358
 
@@
 Pango lineage-associated mutational constellations.
+
+https://github.com/jeromekelleher/sc2ts-paper/issues/348
+https://github.com/jeromekelleher/sc2ts-paper/issues/358
 

### 2025-10-10 latex/main.tex
Updates from Overleaf

@@
 
-Although the exact location of a recombination breakpoint
-is impossible to pinpoint in practice,
-bounds can be placed by examining the parental and recombinant sequences.
-% Cite VirusRecomb
-In methods \cite{Alfonsi2024,Li2024CovRecomb}, and
-approaches (e.g., Pango designation)
+Although in practice
+the exact location of a recombination breakpoint
+is impossible to pinpoint,
+bounds can be placed by
+examining the parental and recombinant sequences.
+In methods \cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb},
+and approaches (e.g., Pango designation),
 based on lineage-associated mutations,
@@
 Similarly, methods based on probabilistic representations of
-% Cite Bolotie
-mutational profiles (e.g., Bolotie) suffer from this limitation.
-% Does RIPPLES qualify as a genealogical method?
-Genealogical methods, such as sc2ts and RIPPLES,
-based on pinpointing exact parental \emph{sequences},
+mutational profiles (e.g.\cite{Varabyou2021rapid})
+suffer from this limitation.
+However, genealogical methods, such as sc2ts and RIPPLES,
+based on identifying exact parental \emph{sequences},
 are in principle maximally precise,
-assuming that the parental sequences
-are well represented in the ARG.
+assuming that the parental sequences are well represented in the ARG.
 
@@
 
-Using the size of a breakpoint interval as a proxy,
-we quantified how precisely breakpoints could be
-localized along the genome.
+Using the size of a breakpoint interval,
+we measured how precisely breakpoints could
+be localized along the genome.
 Across the QC-passing recombination events,
-we found the median interval size to be 2,207 bases
+we found the median breakpoint interval size to be 2,207 bases
 (Figure~\ref{fig:recombinant_intervals}C),
 indicating that the breakpoints of half of the events
-could be narrowed to a small region (less than ~7\% of the genome),
+could be narrowed to a small region (less than $\sim$7\% of the genome),
 while the uncertainty around the breakpoints of the other events
-is considerably greater.
-We further found a negative relationship between
-breakpoint interval size and
+is considerably large.
+Further, we found a negative relationship between
+the size of a breakpoint interval and
 the amount of divergence between recombinant parents,
@@
 
-% Not interesting enough
 Another key feature of genealogical methods
 is that they can also provide \emph{time} intervals
-reflecting uncertainty in the timing of recombination events.
-We calculated the time interval of a recombination node
-as the difference between the age of the youngest parent and
-the age of the oldest child of the recombination node in the ARG
+which reflect uncertainty
+in the estimated timing of recombination events.
+The time interval of a recombination node in the ARG is bounded by
+the age of the youngest parent and
+the age of the oldest child of the recombination node
 (either given as collection dates or estimated using tsdate).
 We found that for three quarters of the QC-passing events,
-the time interval is less than ~2 months
+the time interval spans less than $\sim$2 months
 (Figure~\ref{fig:recombinant_intervals}D),
+with long time intervals suggesting that
+a parental lineage or the recombinant lineage
+might be unsampled for an extended period of time.
 % Do we want to refer to the CIs around the node dates?
-% Maybe we could use this result to lead to the next section?
-% Connect to epidemiology a bit?
-indicating that the ARG can precisely estimate
-the emergence of recombination events.
+This result indicates that the ARG can pinpoint
+when new recombinant lineages emerge with fine resolution.
 
@@
 Each line represents an interval colored by its size.
-(B) Number of overlapping breakpoint intervals.
+(B) Number of overlapping breakpoint intervals per site.
 (C) Distribution of breakpoint interval size (bases).
@@
 estimated by minimizing the sum of squared residuals
-given a single parameter (the mutation rate) (STAR Methods).}
+given a single parameter (the mutation rate) (STAR Methods).
 (E) Distribution of time interval size (months).
+}
 \label{fig:recombinant_intervals}

### 2025-10-10 latex/main.tex
Basic usage of 3SEQ data
Update

Update figures for new 3SEQ data

Update WIP

@@
 the Jackson et al. recombinants (Document~\ref{sec:jackson_recombs}).
-
-\subsection*{Quantification of evidence for recombination}
+This QC filter also correlates strongly with results of other 
+recombination detection methods when run on the putative 
+recombinant sequences (see next section).
+
+\subsection*{Validating and evaluating plausibility of recombination events}
+[THIS IS VERY MUCH WIP]
+
+We validated the recombination events in the sc2ts ARG by running 
+several recombination detection programs on the putative recombinant
+sequences. [TODO generate supplemental table summarising results of 
+valiation.]
+
+We first ran 3SEQ~\cite{Lam2018improved} [etc etc].
+
+To assess the evidence for these putative novel recombination events we 
+ran CovRecomb~\cite{Li2024CovRecomb} and rebar on the corresponding sequences 
+(STAR Methods). Of the 354 QC-passing sc2ts recombination events, 
+90 are classified as 
+recombinant by CovRecomb (all but 2 of the 510 QC-failing recombinants 
+were classified as non-recombinant).
+Rebar classified 203 of the QC-passing sc2ts recombination events, 
+as recombinant, with only 10 of those failing QC being classified 
+as recombinant.
+
 The LS HMM at the heart of sc2ts provides a simple and powerful means 
@@
 
+See Figure~\ref{fig:recombinant_evidence_cov_recomb} CovRecomb
 See Figure~\ref{fig:recombinant_evidence_rebar} for rebar,
@@
 
-\subsection*{Recombination breakpoint intervals}
-
-Although in practice
-the exact location of a recombination breakpoint
-is impossible to pinpoint,
-bounds can be placed by
+\subsection*{Precision of recombination event times and positions}
+
+The location of a recombination breakpoint is generally
+impossible to pinpoint exactly, and instead an interval
+containing the breakpoint can be inferred by
 examining the parental and recombinant sequences.
-In methods \cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb},
-and approaches (e.g., Pango designation),
-based on lineage-associated mutations,
-the breakpoints are limited in precision to these mutations.
+Some methods \cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}
+and approaches (e.g., Pango designation) infer this
+as the genomic distance between lineage-associated
+mutations that differentiate the two parental \emph{lineages}.
 Similarly, methods based on probabilistic representations of
-mutational profiles (e.g.\cite{Varabyou2021rapid})
-suffer from this limitation.
-However, genealogical methods, such as sc2ts and RIPPLES,
+mutational profiles \cite{Varabyou2021rapid}
+can only be as precise as examination of the two
+parent lineages allows.
+However, genealogical methods (such as sc2ts and RIPPLES) are
 based on identifying exact parental \emph{sequences},
-are in principle maximally precise,
-assuming that the parental sequences are well represented in the ARG.
-
-We explored the genomic distribution of the breakpoint intervals of
-the QC-passing recombination events in the ARG.
-We found an overall preponderance of breakpoint intervals
+so are in principle maximally precise.
+The breakpoint interval of a recombinant is bounded by
+two positions: a left position where the recombinant
+matches the left parent sequence but not the right, and
+a right position where the recombinant matches the right
+parent sequence but not the left. 
+How well the time and position of a recombination event
+can be inferred thus depends on the divergence between the
+parent sequences (measured as time to their MRCA).
+
+The distribution of breakpoint intervals of
+the QC-passing recombination events
+shows an overall enrichment of breakpoints
 towards the 3’ end of the genome,
@@
 it may also be due to the higher levels of polymorphism in the Spike gene,
-allowing recombination endpoints to be estimated more precisely.
-
-Using the size of a breakpoint interval,
-we measured how precisely breakpoints could
-be localized along the genome.
-Across the QC-passing recombination events,
-we found the median breakpoint interval size to be 2,207 bases
+making recombination events more likely to be detectable, and 
+allowing breakpoint intervals to be estimated more precisely.
+We found the median breakpoint interval size to be 2,207 bases
 (Figure~\ref{fig:recombinant_intervals}C),
@@
 while the uncertainty around the breakpoints of the other events
-is considerably large.
-Further, we found a negative relationship between
+is considerably larger (up to $\sim$54\% of the genome).
+As expected, we found a negative relationship between
 the size of a breakpoint interval and
-the amount of divergence between recombinant parents,
-which is measured by time to their MRCA
-(Figure~\ref{fig:recombinant_intervals}D).
-When the two parental lineages have diverged
-for longer periods of time (the rightmost points),
-more information becomes available
-on either side of a breakpoint to pinpoint its location,
-resulting in shorter intervals.
-For instance, the breakpoint intervals tend to be the smallest
+divergence between recombinant parents
+(Figure~\ref{fig:recombinant_intervals}D),
+which follows theoretical expectation (STAR Methods).
+Breakpoint intervals tend to be the smallest
 for events involving recombination between
@@
 
-Another key feature of genealogical methods
-is that they can also provide \emph{time} intervals
-which reflect uncertainty
-in the estimated timing of recombination events.
 The time interval of a recombination node in the ARG is bounded by
@@
 (either given as collection dates or estimated using tsdate).
-We found that for three quarters of the QC-passing events,
-the time interval spans less than $\sim$2 months
-(Figure~\ref{fig:recombinant_intervals}D),
-with long time intervals suggesting that
-a parental lineage or the recombinant lineage
-might be unsampled for an extended period of time.
-% Do we want to refer to the CIs around the node dates?
-This result indicates that the ARG can pinpoint
-when new recombinant lineages emerge with fine resolution.
+We found that for $\sim$75\% of the QC-passing recombination events,
+the time interval size is less than $\sim$2 months
+(Figure~\ref{fig:recombinant_intervals}D);
+% Jackson samples in <= 7 day range:
+% Group B ERR5058070 (5 day), singleton ERR5054123 (7 days)
+for example, the time intervals of
+the events associated with the Jackson et al.\ samples
+span 5 to 48 days.
+% Q3: 60 days, max: 259 days
+Longer time intervals imply that a parental or a recombinant lineage
+might have been unsampled for an extended period of time.
 
@@
 \caption{
-Recombination breakpoint intervals.
+Recombination intervals.
 (A) Distribution of the breakpoint intervals of
@@
 (annotated with gene labels on top).
-Each line represents an interval colored by its size.
+Each line represents a breakpoint interval colored by its size.
 (B) Number of overlapping breakpoint intervals per site.
-(C) Distribution of breakpoint interval size (bases).
+(C) Length distribution of breakpoint intervals.
 (D) Relationship between breakpoint interval size and
 time between the recombination event and the parental MRCA.
-The black curve shows the theoretical expected relationship
-between these quantities,
-estimated by minimizing the sum of squared residuals
-given a single parameter (the mutation rate) (STAR Methods).
-(E) Distribution of time interval size (months).
+The black curve shows
+the theoretical expected relationship between these quantities,
+which is calculated assuming a nucleotide substitution rate of
+% N.B. This is a pre-VOC estimate.
+% This study (https://pmc.ncbi.nlm.nih.gov/articles/PMC8807201/#sup1)
+% estimated background rate including VOCs.
+% 0.00065 * 30000 / 12 = 1.625 per month
+$\sim2$ x $10^{-6}$ per site per day
+\cite{Duchene2020,Ghafari2022,Markov2023},
+or $\sim$0.06 per day (STAR Methods).
+(E) Length distribution of time intervals around recombination events.
 }
@@
 We examined the frequency of emergence of the recombinants
-over the course of the COVID-19 pandemic (Figure \ref{fig:recombinants_over_time}).
+over the course of the COVID-19 pandemic
+(Figure \ref{fig:recombinants_over_time}).
 We explored two hypotheses for the timing of recombination.
@@
 
-\subsubsection*{Pandemic scale alignment storage}
-To provide the efficient access to alignment data required for inference with sc2ts, we
-developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
+\subsubsection*{Pandemic-scale alignment storage}
+To provide the efficient access to alignment data required for inference with sc2ts,
+we developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
 Briefly, Zarr is a storage format for scientific data,
@@
 sciences~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
-Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
+Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,
+Marconato2024spatialdata,Ruan2024image}.
 Zarr is essentially a simple mechanism for storing n-dimensional array
@@
 
-\subsubsection*{Sample filtering and alignment preprocessing}
+\subsubsection*{Sample filtering and alignment pre-processing}
 
@@
 \subsubsection*{Calculation of Pango lineage distance using pangonet}
+
 We define a ``pseudo-phylogenetic'' distance
@@
 If two Pango lineages are siblings in the pseudo-phylogeny,
-the distance is defined as the sum of the number of edges from each lineage to their MRCA;
-if they are non-siblings (i.e. one Pango is a descendant of the other),
-the distance is defined as the number of edges separating them;
+the distance is defined as
+the sum of the number of edges from each lineage to their MRCA;
+if they are non-siblings (i.e., one Pango is a descendant of the other),
+the distance is defined as
+the number of edges separating them;
 if they are identical, the distance is defined as zero.
+
+\subsubsection*{Expected lengths of recombination breakpoint intervals}
+
+Recombination breakpoint intervals are imprecise
+when the recombining parents are identical in sequence
+at sites spanning the breakpoint.
+Here, we estimate the interval lengths expected around a breakpoint,
+given a genome-wide mutation rate $\mu$,
+total time between two samples ($T$), and
+sufficient support to detect the recombination event.
+If a breakpoint occurs a fraction $P$ along a genome of length $L$ and
+$k$ mutations have occurred to the left of the breakpoint,
+then we expect the nearest mutation to occur $P L / (k + 1)$ bases to the left.
+If we require $n$ mutations on both sides of the breakpoint,
+we can calculate the expected value of the interval to the left
+by summing across a Poisson-distributed number of mutations,
+conditioned on having at least $n$ mutations:
+
+\[I_{left} = \frac{L}{n T \mu} \frac{\Gamma(1+n) - \Gamma(1+n, P T \mu)}{\Gamma(n) - \Gamma(n, P T \mu)}\]
+
+A similar result holds for the right interval ($I_{right}$),
+but with $P$ replaced with $1–P$.
+Summing the two intervals ($I_{left}+I_{right}$)
+gives the expected interval length.
+As detectable recombination events
+tend to be closer to the center of the genome,
+we use $P=0.5$ in Figure~\ref{fig:recombinant_intervals}D.
 
@@
 
-\subsubsection*{Estimating interval lengths}
-Recombination intervals are imprecise
-when the recombining parents are identical in sequence at sites spanning the breakpoint.
-Here we estimate the interval lengths expected around a breakpoint,
-given a genome-wide mutation rate $\mu$,
-total time between two samples ($T$), and
-sufficient support to detect the recombination event.
-If a breakpoint occurs a fraction $P$ along a genome of length $L$ and
-$k$ mutations have occurred to the left of the breakpoint,
-then we expect the nearest mutation to occur $P L / (k + 1)$ bases to the left.
-If we require $n$ mutations on both sides of the breakpoint,
-we can calculate the expected value of the interval to the left
-by summing across a Poisson-distributed number of mutations,
-conditioned on having at least $n$ mutations:
-
-\[I_{left} = \frac{L}{n T \mu} \frac{\Gamma(1+n) - \Gamma(1+n, P T \mu)}{\Gamma(n) - \Gamma(n, P T \mu)}\]
-
-A similar result holds for the right interval ($I_{right}$), but with $P$ replaced with $1–P$.
-Summing the two intervals ($I_{left}+I_{right}$) gives the expected interval length.
-As detectable recombination events are nearer the center of the chromosome,
-we use $P=½$ in Figure X.
-
 \subsubsection*{Analyzing global case counts}
+
 To explore the timing of emergence of recombinants,
@@
 \label{sec:supplement_dels}
-% [TODO This is too terse and too vague.
-% Be concrete and precise.
-%WHICH deletions are we talking about?]
 
@@
 % [FIXME define how we measured this].
-(i.e., for a deletion caused by multiple events in the ARG,
-the total of all the samples inheriting the deletion across those events).
+(i.e., for a deletion resulting from multiple events in the ARG,
+this is the total samples inheriting the deletion).
 
@@
 \label{sec:supplement_sc2ts_usher_imputation}
+
 Both sc2ts and UShER support missing data in input alignments, and effectively
@@
 The total aligned dataset for 2,475,418 samples and 27,507 sites 
-in the sc2ts and UShER intersection ARGs represents 63.41GiB of nucleotide calls. 
+in the sc2ts and UShER intersection ARGs represents
+63.41 GiB of nucleotide calls.
 Of the 162,838,461 missing data calls (Ns) in the alignments,
@@
 \centering
+\includegraphics[width=\linewidth]{recombinant_evidence_cov_recomb}
+\caption{Things}
+\label{fig:recombinant_evidence_cov_recomb}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics[width=\linewidth]{recombinant_evidence_rebar}

### 2025-10-10 latex/main.tex
Updates from Overleaf

@@
 the Jackson et al. recombinants (Document~\ref{sec:jackson_recombs}).
-This QC filter also correlates strongly with results of other 
-recombination detection methods when run on the putative 
-recombinant sequences (see next section).
-
-\subsection*{Validating and evaluating plausibility of recombination events}
-[THIS IS VERY MUCH WIP]
-
-We validated the recombination events in the sc2ts ARG by running 
-several recombination detection programs on the putative recombinant
-sequences. [TODO generate supplemental table summarising results of 
-valiation.]
-
-We first ran 3SEQ~\cite{Lam2018improved} [etc etc].
-
-To assess the evidence for these putative novel recombination events we 
-ran CovRecomb~\cite{Li2024CovRecomb} and rebar on the corresponding sequences 
-(STAR Methods). Of the 354 QC-passing sc2ts recombination events, 
-90 are classified as 
-recombinant by CovRecomb (all but 2 of the 510 QC-failing recombinants 
-were classified as non-recombinant).
-Rebar classified 203 of the QC-passing sc2ts recombination events, 
-as recombinant, with only 10 of those failing QC being classified 
-as recombinant.
-
+
+\subsection*{Quantification of evidence for recombination}
 The LS HMM at the heart of sc2ts provides a simple and powerful means 
@@
 
-See Figure~\ref{fig:recombinant_evidence_cov_recomb} CovRecomb
 See Figure~\ref{fig:recombinant_evidence_rebar} for rebar,
@@
 
-\subsection*{Precision of recombination event times and positions}
+\subsection*{Recombination breakpoint intervals}
 
@@
 impossible to pinpoint exactly, and instead an interval
-containing the breakpoint can be inferred by
+containing the breakpoint can be inferred through
 examining the parental and recombinant sequences. 
@@
 as the genomic distance between lineage-associated
-mutations that differentiate the two parental \emph{lineages}.
+mutations that differentiate the two parental strains.
 Similarly, methods based on probabilistic representations of
 mutational profiles \cite{Varabyou2021rapid}
-can only be as precise as examination of the two
+can only be as accurate as examination of the two
 parent lineages allows.
@@
 so are in principle maximally precise.
+
 The breakpoint interval of a recombinant is bounded by
@@
 
-The distribution of breakpoint intervals of
-the QC-passing recombination events
-shows an overall enrichment of breakpoints
+First, we explored the genomic distribution of the breakpoint intervals of
+the QC-passing recombination events in the ARG.
+
+Between these two positions, there is no genetic variation
+to distinguish the parental source of inheritance.
+We found an overall preponderance of breakpoint intervals
 towards the 3’ end of the genome,
@@
 it may also be due to the higher levels of polymorphism in the Spike gene,
-making recombination events more likely to be detectable, and 
-allowing breakpoint intervals to be estimated more precisely.
-We found the median breakpoint interval size to be 2,207 bases
+allowing recombination endpoints to be estimated more precisely.
+
+Next, using the size of a breakpoint interval,
+we measured how precisely breakpoints could
+be localized along the genome.
+Across the QC-passing recombination events,
+we found the median breakpoint interval size to be 2,207 bases
 (Figure~\ref{fig:recombinant_intervals}C),
@@
 is considerably larger (up to $\sim$54\% of the genome).
-As expected, we found a negative relationship between
+Further, we found a negative relationship between
 the size of a breakpoint interval and
-divergence between recombinant parents
-(Figure~\ref{fig:recombinant_intervals}D),
+the amount of divergence between recombinant parents
+(measured by time to their MRCA),
+(Figure~\ref{fig:recombinant_intervals}D).
 which follows theoretical expectation (STAR Methods).
-Breakpoint intervals tend to be the smallest
+When the two parental lineages have diverged
+for longer periods of time (the rightmost points),
+more information becomes available
+on either side of a breakpoint to pinpoint its location,
+resulting in shorter intervals.
+For instance, the breakpoint intervals tend to be the smallest
 for events involving recombination between
@@
 
+Another key feature of genealogical methods
+is that they can also provide \emph{time} intervals
+which reflect uncertainty in the timing of recombination events.
 The time interval of a recombination node in the ARG is bounded by
@@
 might have been unsampled for an extended period of time.
+% Do we want to refer to the CIs around the node dates?
+These results show that the ARG can finely pinpoint
+the emergence of new recombinant lineages in time.
 
@@
 which is calculated assuming a nucleotide substitution rate of
-% N.B. This is a pre-VOC estimate.
-% This study (https://pmc.ncbi.nlm.nih.gov/articles/PMC8807201/#sup1)
-% estimated background rate including VOCs.
-% 0.00065 * 30000 / 12 = 1.625 per month
-$\sim2$ x $10^{-6}$ per site per day
-\cite{Duchene2020,Ghafari2022,Markov2023},
+$\sim$2 per site per day \cite{Duchene2020,Ghafari2022,Markov2023},
 or $\sim$0.06 per day (STAR Methods).
@@
 We examined the frequency of emergence of the recombinants
-over the course of the COVID-19 pandemic
-(Figure \ref{fig:recombinants_over_time}).
+over the course of the COVID-19 pandemic (Figure \ref{fig:recombinants_over_time}).
 We explored two hypotheses for the timing of recombination.
@@
 
-\subsubsection*{Pandemic-scale alignment storage}
-To provide the efficient access to alignment data required for inference with sc2ts,
-we developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
+\subsubsection*{Pandemic scale alignment storage}
+To provide the efficient access to alignment data required for inference with sc2ts, we
+developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
 Briefly, Zarr is a storage format for scientific data,
@@
 sciences~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
-Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,
-Marconato2024spatialdata,Ruan2024image}.
+Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
 Zarr is essentially a simple mechanism for storing n-dimensional array
@@
 
-\subsubsection*{Sample filtering and alignment pre-processing}
+\subsubsection*{Sample filtering and alignment preprocessing}
 
@@
 As detectable recombination events
-tend to be closer to the center of the genome,
+tend to be closer to the middle of the genome,
 we use $P=0.5$ in Figure~\ref{fig:recombinant_intervals}D.
@@
 \centering
-\includegraphics[width=\linewidth]{recombinant_evidence_cov_recomb}
-\caption{Things}
-\label{fig:recombinant_evidence_cov_recomb}
-\end{figure}
-
-\begin{figure}
-\centering
 \includegraphics[width=\linewidth]{recombinant_evidence_rebar}

### 2025-10-10 latex/main.tex

--- 2025-10-10 latex/main.tex
@@
 the Jackson et al. recombinants (Document~\ref{sec:jackson_recombs}).
-This QC filter also correlates strongly with results of other 
-recombination detection methods when run on the putative 
-recombinant sequences (see next section).
-
-\subsection*{Validating and evaluating plausibility of recombination events}
-[THIS IS VERY MUCH WIP]
-
-We validated the recombination events in the sc2ts ARG by running 
-several recombination detection programs on the putative recombinant
-sequences. [TODO generate supplemental table summarising results of 
-valiation.]
-
-We first ran 3SEQ~\cite{Lam2018improved} [etc etc].
-
-To assess the evidence for these putative novel recombination events we 
-ran CovRecomb~\cite{Li2024CovRecomb} and rebar on the corresponding sequences 
-(STAR Methods). Of the 354 QC-passing sc2ts recombination events, 
-90 are classified as 
-recombinant by CovRecomb (all but 2 of the 510 QC-failing recombinants 
-were classified as non-recombinant).
-Rebar classified 203 of the QC-passing sc2ts recombination events, 
-as recombinant, with only 10 of those failing QC being classified 
-as recombinant.
+
+\subsection*{Quantification of evidence for recombination}
 

### 2025-10-11 arg_postprocessing/scripts/run_3seq.py
Implement some basic conversion stuff for RIPPLES/Usher

@@ -8,9 +8,11 @@ import pathlib
 import subprocess
 import tempfile
 import os.path
+import dataclasses
 
 
-def run_single_3seq(fasta_file):
+def _old_run_single_3seq(fasta_file):
+
     exe = os.path.abspath("./tmp/3seq/3seq")
     fasta_file = os.path.abspath(fasta_file)
     with tempfile.TemporaryDirectory() as tempdir:
@@ -20,6 +22,31 @@ def run_single_3seq(fasta_file):
             f"yes | {exe} -full {fasta_file}", shell=True, cwd=tempdir
         )
         output = pd.read_csv(pathlib.Path(tempdir) / "3s.rec.csv")
+    return output
+
+def run_single_3seq(child_fasta, parent_fastas):
+
+    exe = os.path.abspath("./tmp/3seq/3seq")
+    # with tempfile.TemporaryDirectory() as tempdir:
+        # tempdir = pathlib.Path(tempdir)
+    if True:
+        tempdir = pathlib.Path("tmp/")
+        parents_file = (tempdir / "parents.fasta").absolute()
+        with open(parents_file, "w") as fw:
+            for j, parent_file in enumerate(parent_fastas):
+                with open(parent_file) as fr:
+                    for line in fr.read():
+                        if line.startswith(">"):
+                            print(f">parent_{j}", file=fw)
+                        else:
+                            print(line, file=fw)
+
+        # NOTE: I can't get 3seq to run in -triplet mode whatever I do,
+        # but full seems to work. Same thing, ultimately?
+        subprocess.check_output(
+            f"yes | {exe} -full {parents_file} {child_fasta}", shell=True, cwd=tempdir
+        )
+        output = pd.read_csv(tempdir / "3s.rec.csv")
     return output
 
 
@@ -82,11 +109,65 @@ def generate_fasta(ts, recombinants_csv, output_dir):
                 f.write(v + "\n")
 
 
+@click.command()
+@click.argument("ripples_file")
+@click.argument("output")
+def generate_ripples_sample_list(ripples_file, output):
+    df = pd.read_csv(ripples_file, sep="\t")
+    samples = set(df["#recomb_node_id"]) | set(df["donor_node_id"]) | set(df["acceptor_node_id"])
+    samples = np.array(list(samples))
+    # We to do this messing around to chunk the VCF to FASTA conversion up
[…]

### 2025-10-11 arg_postprocessing/scripts/run_3seq.py
Run 3seq for RIPPLES

@@ -24,32 +24,37 @@ def _old_run_single_3seq(fasta_file):
         output = pd.read_csv(pathlib.Path(tempdir) / "3s.rec.csv")
     return output
 
+
 def run_single_3seq(child_fasta, parent_fastas):
 
     exe = os.path.abspath("./tmp/3seq/3seq")
-    # with tempfile.TemporaryDirectory() as tempdir:
-        # tempdir = pathlib.Path(tempdir)
-    if True:
-        tempdir = pathlib.Path("tmp/")
+    # if True:
+    #     tempdir = pathlib.Path("tmp/")
+    with tempfile.TemporaryDirectory() as tempdir:
+        tempdir = pathlib.Path(tempdir)
         parents_file = (tempdir / "parents.fasta").absolute()
         with open(parents_file, "w") as fw:
             for j, parent_file in enumerate(parent_fastas):
                 with open(parent_file) as fr:
-                    for line in fr.read():
+                    for line in fr.readlines():
                         if line.startswith(">"):
                             print(f">parent_{j}", file=fw)
                         else:
                             print(line, file=fw)
-
         # NOTE: I can't get 3seq to run in -triplet mode whatever I do,
         # but full seems to work. Same thing, ultimately?
-        subprocess.check_output(
-            f"yes | {exe} -full {parents_file} {child_fasta}", shell=True, cwd=tempdir
-        )
-        output = pd.read_csv(tempdir / "3s.rec.csv")
+        cmd = f"{exe} -full {parents_file} {child_fasta}"
+        subprocess.check_output(f"yes | {cmd}", shell=True, cwd=tempdir)
+        # Note: this emits a parser warning when 3SEQ suggests multiple
+        # breakpoints, as it uses commas to separate them (sigh). So, we
+        # thrown away that information here. This is better than the
+        # alternative, which gets column alignment completely wrong.
+        output = pd.read_csv(tempdir / "3s.rec.csv", index_col=False)
     return output
 
 
+# TODO rejigger this to use the new run_single_3seq function above and
+# use the different approach to putting fastas together.
 @click.command()
 @click.argument("recombinants_csv")
 @click.argument("fasta_dir")
@@ -114,21 +119,18 @@ def generate_fasta(ts, recombinants_csv, output_dir):
 @click.argument("output")
[…]

### 2025-10-13 latex/main.tex

--- 2025-10-10 latex/main.tex
@@
 \subsection*{Comparison with RIPPLES}
+
 RIPPLES~\cite{Turakhia2022} is a method for detecting 
@@
 Recombination is detected by first identifying plausible branches
-carrying multiple mutations .
+carrying multiple mutations.
 For each candidate branch,
@@
 allowing breakpoint intervals to be estimated more precisely.
-We found the median breakpoint interval size to be 2,207 bases
+We found the median breakpoint interval length to be 2,207 bases
 (Figure~\ref{fig:recombinant_intervals}C),
@@
 As expected, we found a negative relationship between
-the size of a breakpoint interval and
+the length of a breakpoint interval and
 divergence between recombinant parents
@@
 \caption{
-Recombination intervals.
+Recombination breakpoint and time intervals.
 (A) Distribution of the breakpoint intervals of
@@
 (annotated with gene labels on top).
-Each line represents a breakpoint interval colored by its size.
+Each line represents a breakpoint interval colored by its length.
 (B) Number of overlapping breakpoint intervals per site.
 (C) Length distribution of breakpoint intervals.
-(D) Relationship between breakpoint interval size and
+(D) Relationship between breakpoint interval length and
 time between the recombination event and the parental MRCA.
@@
 the theoretical expected relationship between these quantities,
-which is calculated assuming a nucleotide substitution rate of
-% N.B. This is a pre-VOC estimate.
-% This study (https://pmc.ncbi.nlm.nih.gov/articles/PMC8807201/#sup1)
-% estimated background rate including VOCs.
-% 0.00065 * 30000 / 12 = 1.625 per month
-$\sim2$ x $10^{-6}$ per site per day
-\cite{Duchene2020,Ghafari2022,Markov2023},
-or $\sim$0.06 per day (STAR Methods).
+which is calculated assuming
+a genome-wide nucleotide substitution rate of
+$\sim$0.06 per day (STAR Methods).
 (E) Length distribution of time intervals around recombination events.
@@
 Here, we estimate the interval lengths expected around a breakpoint,
-given a genome-wide mutation rate $\mu$,
+given a genome-wide nucleotide substitution rate $\mu$,
 total time between two samples ($T$), and
@@
 we use $P=0.5$ in Figure~\ref{fig:recombinant_intervals}D.
+
+In Figure~\ref{fig:recombinant_intervals}D,
+we calculated the expected lengths of breakpoint intervals
+by setting $\mu$ to $\sim$0.0598 per day.
+This is based on published estimates of
+the per-site nucleotide substitution rate of
+% N.B. This is a pre-VOC estimate.
+% This study 
+% (https://pmc.ncbi.nlm.nih.gov/articles/PMC8807201/#sup1)
+% estimated background rate including VOCs.
+% 0.00065 * 30000 / 12 = 1.625 per month
+$\sim2$ x $10^{-6}$ per site per day
+\cite{Duchene2020,Ghafari2022,Markov2023}.
 
@@
 
-\subsection*{Evaluating inferred node dates}
+\subsection{Evaluating inferred node dates}
 \label{sec:evaluating_node_dates}

### 2025-10-13 arg_postprocessing/scripts/run_3seq.py
Stash some partial snakemake updates in case we need to revert

@@ -28,21 +28,15 @@ def _old_run_single_3seq(fasta_file):
 def run_single_3seq(child_fasta, parent_fastas):
 
     exe = os.path.abspath("./tmp/3seq/3seq")
-    # if True:
-    #     tempdir = pathlib.Path("tmp/")
+    if True:
+        tempdir = pathlib.Path("tmp/")
     with tempfile.TemporaryDirectory() as tempdir:
         tempdir = pathlib.Path(tempdir)
         parents_file = (tempdir / "parents.fasta").absolute()
         with open(parents_file, "w") as fw:
             for j, parent_file in enumerate(parent_fastas):
                 with open(parent_file) as fr:
-                    for line in fr.readlines():
-                        if line.startswith(">"):
-                            print(f">parent_{j}", file=fw)
-                        else:
-                            print(line, file=fw)
-        # NOTE: I can't get 3seq to run in -triplet mode whatever I do,
-        # but full seems to work. Same thing, ultimately?
+                    fw.write(fr.read())
         cmd = f"{exe} -full {parents_file} {child_fasta}"
         subprocess.check_output(f"yes | {cmd}", shell=True, cwd=tempdir)
         # Note: this emits a parser warning when 3SEQ suggests multiple

### 2025-10-13 arg_postprocessing/scripts/run_3seq.py
Add 3SEQ results for p=4 and p=3

@@ -28,8 +28,8 @@ def _old_run_single_3seq(fasta_file):
 def run_single_3seq(child_fasta, parent_fastas):
 
     exe = os.path.abspath("./tmp/3seq/3seq")
-    if True:
-        tempdir = pathlib.Path("tmp/")
+    # if True:
+    #     tempdir = pathlib.Path("tmp/")
     with tempfile.TemporaryDirectory() as tempdir:
         tempdir = pathlib.Path(tempdir)
         parents_file = (tempdir / "parents.fasta").absolute()
@@ -118,13 +118,8 @@ def generate_ripples_sample_list(ripples_file, output):
         | set(df["donor_node_id"])
         | set(df["acceptor_node_id"])
     )
-    samples = np.array(list(samples))
-    # We to do this messing around to chunk the VCF to FASTA conversion up
-    n = len(samples) // 900  # Ensure we have no more than 1000
-    splits = np.array_split(samples, n)
-    for j, a in enumerate(splits):
-        np.savetxt(f"{output}_{j}.txt", a, fmt="%s")
-    print(len(splits))
+    samples = np.array(sorted(list(samples)))
+    np.savetxt(f"{output}", samples, fmt="%s")
 
 
 @click.command()
@@ -140,9 +135,6 @@ def run_ripples_3seq(ripples_file, fasta_dir, output):
     # parsimony score). If there's several, we pick one arbitrarily.
     df = df.loc[df.groupby(["recomb_node_id"])["recomb_parsimony"].idxmin()]
     print(f"Have {df.shape[0]} unique recombination events")
-    # samples = set(df["#recomb_node_id"]) | set(df["donor_node_id"]) | set(df["acceptor_node_id"])
-    # samples = np.array(list(samples))
-    # np.savetxt(output, samples, fmt="%s")
 
     def fasta_file(name):
         return os.path.abspath(pathlib.Path(fasta_dir) / f"{name}_NC_045512:0.fa")

### 2025-10-14 latex/main.tex

@@
 
-RIPPLES~\cite{Turakhia2022} is a method for detecting 
-recombination events using a tree inferred by UShER,
-and is the only method directly comparable with sc2ts.
-Recombination is detected by first identifying plausible branches
-carrying multiple mutations.
+RIPPLES~\cite{Turakhia2022} is a method for identifying
+recombination \emph{events} using a phylogeny inferred by UShER.
+It is the only \emph{genealogical} method built for this purpose and
+therefore the only method directly comparable to sc2ts.
+A recombination event is detected by first
+identifying plausible branches carrying multiple mutations.
 For each candidate branch,
-RIPPLES then systematically splits the sequence into 2 or 3 segments, and
+RIPPLES systematically splits the sequence into 2 or 3 segments, and
 finds parsimonious placements for each segment independently.
@@
 the overall parsimony of these multiple placements
-sufficiently improves over the original single-parent origin in the tree
+sufficiently improves over
+the original single-parent origin in the phylogeny
 (the ``parsimony improvement'' parameter, $p=3$, by default).
-RIVET~\cite{Smith2023tracking} is a tool for quality control
-and visualization of these recombination events.
-
-We ran fast-RIPPLES\cite{Smith2023tracking} on the Viridian UShER tree
+RIVET~\cite{Smith2023tracking} is a tool for quality control and
+visualization of these recombination events.
+
+We ran fast-RIPPLES\cite{Smith2023tracking}
+on the Viridian UShER phylogeny
 subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
 At $p=3$, this resulted in a total of 4111 recombination events.
-Most of these subtend $<10$ samples (2616 singletons), with 4 events subtending 
-more than 100,000 samples.
+Most of these subtend $<10$ samples (2616 singletons),
+with 4 events subtending more than 100,000 samples.
 Analysis of the descending samples for each event shows that
@@
 and 
-82 BA.2 samples share a recombination event [With what breakpoint and 
-parents?].
+82 BA.2 samples share a recombination event [With what breakpoint and parents?].
 We also ran RIPPLES with $p=4$, which reduced the overall number of
 events to 1174 (roughly 4-fold) but also approximately halved the number of
-events involving X lineages, missing well-characterised events 
+events involving X lineages, missing well-characterized events,
 such as XBB (Table~\ref{tab:ripples_x_lineages}).
@@
 descending from each RIPPLES event. Overall, clade agreement was
-high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG where 
-descendants match exactly. Of these nodes, 68\% had no recombinant
-ancestry. Of the remainder, 595 could be matched with one of the 
-855 recombination events in the sc2ts ARG
-(Figure \ref{fig:ripples_sc2ts_events}).
-Table~\ref{tab:ripples_x_lineages} shows a summary of all
-recombination events associated with Pango X lineages detected by RIPPLES,
+high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG
+where descendants match exactly. Of these nodes, 68\% had no recombinant
+ancestry. Of the remainder, 595 could be matched with
+one of the 855 recombination events in the sc2ts ARG
+(Figure~\ref{fig:ripples_sc2ts_events}).
+Table~\ref{tab:ripples_x_lineages} shows a summary of
+all recombination events associated with Pango X lineages detected by RIPPLES,
 and their relationship with the sc2ts ARG.
@@
 (recombination nodes with more complex assocation with X lineages),
-two have close matches, with XBM matching exactly (including 2 descending BF.3 samples),
+two have close matches, with XBM matching exactly
+(including 2 descending BF.3 samples),
 and XE/XH is similar (but without including the 2 XH samples in the clade).
@@
 The denominators indicate the number of Pango X lineages
-for which detection results are available from 
-the original studies \cite{Alfonsi2024,Li2024CovRecomb}.
+for which detection results are available from the original
+studies\cite{Alfonsi2024,Li2024CovRecomb}.
 For the Pango X lineages with non-overlapping breakpoint intervals,

### 2025-10-14 latex/main.tex
Updates from Overleaf

@@
 \caption{
-TODO caption
+Daily statistics of samples analyzed
+during the ARG inference process.
+(A) The number of samples
+processed from the input sequence alignments,
+exactly matched to a node in the current ARG, and
+added to the ARG
+all peak in December, 2022
+and subsequently around February, 2022.
+(B) The fraction of exactly matched samples and
+the fraction of samples added to the ARG
+mirror each other.
+Around the peak of sample volume,
+the fraction of exactly matched samples
+is relatively high (above 0.50).
+As expected, this fraction later decreases
+when sampling effort decreases.
+(C) The mean HMM cost per sample is relatively high
+at the beginning as the ARG had not accumulated
+enough genetic diversity yet.
+(D) Proportion of samples classified
+by Scorpio designation.
+These proportions reflect waves of infections
+driven by the emergence of Alpha, Delta, and Omicron.
 }\label{fig:samples_per_day}
@@
 \caption{
-TODO caption
+Usage of computational resources
+during the ARG inference process.
+(A) The amount of time to process a daily batch of samples
+peaks when the sample volume is the highest.
+(B) Mean time to process a sample remained low.
+(C) Max RAM usage also peaks around the time
+that sampling effort is the highest.
 }

### 2025-10-14 latex/main.tex

@@
 
-\subsection*{Comparison with RIPPLES}
-
-RIPPLES~\cite{Turakhia2022} is a method for identifying
-recombination \emph{events} using a phylogeny inferred by UShER.
-It is the only \emph{genealogical} method built for this purpose and
-therefore the only method directly comparable to sc2ts.
-A recombination event is detected by first
-identifying plausible branches carrying multiple mutations.
-For each candidate branch,
-RIPPLES systematically splits the sequence into 2 or 3 segments, and
-finds parsimonious placements for each segment independently.
-A recombination event at that node is then flagged if
-the overall parsimony of these multiple placements
-sufficiently improves over
-the original single-parent origin in the phylogeny
-(the ``parsimony improvement'' parameter, $p=3$, by default).
-RIVET~\cite{Smith2023tracking} is a tool for quality control and
-visualization of these recombination events.
-
-We ran fast-RIPPLES\cite{Smith2023tracking}
-on the Viridian UShER phylogeny
-subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
-At $p=3$, this resulted in a total of 4111 recombination events.
-Most of these subtend $<10$ samples (2616 singletons),
-with 4 events subtending more than 100,000 samples.
-Analysis of the descending samples for each event shows that
-10 of the 44 Pango X lineages present in the dataset
-% NOTE: listing them like this for sorting and manipulation purposes
-(%
-XA, 
-XAC, 
-XAE, 
-XAJ,
-XB, 
-XBB, 
-XBM, 
-XC, 
-XL, 
-XW%
-)
-are associated unequivocally with a single recombination event.
-7 lineages 
-(%
-XAL,
-XAM,
-XBD, 
-XBR, 
-XE,
-XS,
-XY%
-)
-are associated with two events. Finally, 
-one event subtending 1321 samples suggests that 
-XAF,
-XAL, 
-XE, 
-XH, 
-XJ, 
-XM, 
-XY, 
-and 
-82 BA.2 samples share a recombination event [With what breakpoint and parents?].
-We also ran RIPPLES with $p=4$, which reduced the overall number of
-events to 1174 (roughly 4-fold) but also approximately halved the number of
-events involving X lineages, missing well-characterized events,
-such as XBB (Table~\ref{tab:ripples_x_lineages}).
-
-To compare these results with the sc2ts ARG, we first matched clades
-by finding the most recent common ancestor (MRCA) of the samples
-descending from each RIPPLES event. Overall, clade agreement was
-high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG
-where descendants match exactly. Of these nodes, 68\% had no recombinant
-ancestry. Of the remainder, 595 could be matched with
-one of the 855 recombination events in the sc2ts ARG
-(Figure~\ref{fig:ripples_sc2ts_events}).
-Table~\ref{tab:ripples_x_lineages} shows a summary of
-all recombination events associated with Pango X lineages detected by RIPPLES,
-and their relationship with the sc2ts ARG.
-Of the 16 Class I events (Pango X lineages unequivocally associated with
-a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
-8 have corresponding recombination events (4, at $p=4$).
-Of the 5 Type II events
-(recombination nodes with more complex assocation with X lineages),
-two have close matches, with XBM matching exactly
-(including 2 descending BF.3 samples),
-and XE/XH is similar (but without including the 2 XH samples in the clade).
-Two events in the RIPPLES set correspond to likely false negatives
-in sc2ts: XB (Document~\ref{sec:dd_xb})
-and XAJ (Document~\ref{sec:dd_xaj}).
-See Document~\ref{sec:detailed_ripples} for more detailed analysis.
-
-\subsection*{Quality control of recombination events}
+\subsection*{Quality control and validation of recombination events}
 
@@
 because the LS HMM counts each base separately.
-We developed a quality control (QC) measure
-which counts the number of ``loci'' (clusters of closely spaced sites)
+We use a simple quality control (QC) measure
+that counts the number of ``loci'' (clusters of closely spaced sites)
 supporting the left and right parents of a recombinant (STAR Methods),
@@
 (Table~\ref{tab:pango_x_lineages}) and
-the Jackson et al. recombinants (Document~\ref{sec:jackson_recombs}).
-This QC filter also correlates strongly with results of other 
-recombination detection methods when run on the putative 
-recombinant sequences (see next section).
-
-\subsection*{Validating and evaluating plausibility of recombination events}
-[THIS IS VERY MUCH WIP]
+the Jackson et al.\ recombinants (Document~\ref{sec:jackson_recombs}).
 
 We validated the recombination events in the sc2ts ARG by running 
-several recombination detection programs on the putative recombinant
-sequences. [TODO generate supplemental table summarising results of 
-valiation.]
-
-We first ran 3SEQ~\cite{Lam2018improved} [etc etc].
-
-To assess the evidence for these putative novel recombination events we 
-ran CovRecomb~\cite{Li2024CovRecomb} and rebar on the corresponding sequences 
-(STAR Methods). Of the 354 QC-passing sc2ts recombination events, 
-90 are classified as 
-recombinant by CovRecomb (all but 2 of the 510 QC-failing recombinants 
-were classified as non-recombinant).
-Rebar classified 203 of the QC-passing sc2ts recombination events, 
-as recombinant, with only 10 of those failing QC being classified 
-as recombinant.
-
+the 3SEQ~\cite{Lam2018improved}, CovRecomb~\cite{Li2024CovRecomb} and rebar 
+recombination detection methods on the putative recombinant sequences
+(STAR Methods) summarised in Table~\ref{tab:recombinant_validation}.
+Overall, validation among the QC-passing recombinants is highest
+at with 3SEQ at 88\%, and is lowest for CovRecomb at xx\%. 
+CovRecomb has been previously shown to be 
+less sensitive than other methods~\cite{Li2024CovRecomb}, 
+and is specifically intended 
+to detect interlineage recombinants. The 12\% of recombinants 
+not passing 3SEQ validation are also the most marginally supported
+(see next section; Figure~\ref{fig:recombinant_evidence}).
+All of the validation metrics also correlate strongly with the 
+QC filter, with very few of the QC-failing recombinants
+passing external validation (Table~\ref{tab:recombinant_validation}).
+
+\subsection*{Assessing evidence for recombination events}
 The LS HMM at the heart of sc2ts provides a simple and powerful means 
@@
 over all 354 recombination events that pass QC. 
-Of these, only 24 
-(21 associated with Pango X lineages in Table~\ref{tab:pango_x_lineages},
-and 3 additional from Jackon et al in Table~\ref{tab:jackson}) have 
-been thoroughly characterised.  
-To assess the evidence for these putative novel recombination events we 
-ran CovRecomb~\cite{Li2024CovRecomb} and rebar on the corresponding sequences 
-(STAR Methods). Of the 354 QC-passing sc2ts recombination events, 
-90 are classified as 
-recombinant by CovRecomb (all but 2 of the 510 QC-failing recombinants 
-were classified as non-recombinant).
-Rebar classified 203 of the QC-passing sc2ts recombination events, 
-as recombinant, with only 10 of those failing QC being classified 
-as recombinant.
-
-[WIP]
-Of the 38 recombinants that are strongly supported (at least 10 averted
-mutations), 29 are novel. The known recombinants are XA, XC, XS, XBR, 
-and 5 from Jackson et al. All but 2 of the 38 are classified as
-recombinant by rebar. The majority (20) of these novel strongly supported
-recombinants are singletons, with 9 having further descendants in the ARG.
-Of the remaining 316 recombinants with less than 10 averted mutations,
-only 151 are classified as recombinant by rebar. 
-While some are likely to 
-be false positives, it is also interesting to note that rebar validation
-rate is much lower for recombinants with closely related 
-parents (Figure~\ref{fig:recombinant_evidence}B,C) suggesting that 
-sc2ts may have a higher sensitivity.
-
-See Figure~\ref{fig:recombinant_evidence_cov_recomb} CovRecomb
-See Figure~\ref{fig:recombinant_evidence_rebar} for rebar,
-Figure~\ref{fig:recombinant_evidence_ripples_p4} for RIPPLES p4
-and Figure~\ref{fig:recombinant_evidence_ripples_p4} for RIPPLES p3.
+Also shown is the 3SEQ classification status of these recombinants.
+All of the 12\% of recombinants that do not pass 3SEQ validation are 
+from the most weakly supported, with 4 or 5 averted mutations.
+Figures~\ref{fig:recombinant_evidence_cov_recomb} and
+~\ref{fig:recombinant_evidence_rebar} 
+same the same information for CovRecomb and rebar validation,
+showing that their validation status tends to be function 
+of the divergence between the recombinant parents.
 
@@
 
-Two further recombinants in the sc2ts ARG that are likely to be 
-spurious are important to discuss.
-[FIXME - discuss BA.5 (kept as a recombination event but only because we can't 
-rewire it post-hoc, owning to subsequent tree building) and the BA.2 event (doesn't
-pass QC.) - note this is discussed in more detail in document \ref{sec:spurious-ancestrally-important-recombinants}]
+\subsection*{Comparison with UShER+RIPPLES}
+
+RIPPLES~\cite{Turakhia2022} is a method for identifying
+recombination events using a phylogeny inferred by UShER.
+This distinguishes RIPPLES from other recombination detection
+methods such as 3SEQ, which classify particular \emph{sequences}
+as recombinant or not and can only identify potential onward 
+transmission post-hoc. The combination of UShER and RIPPLES
+is therefore the only method that is directly comparable 
+to sc2ts. It works by first identifying plausible long 
+branches in the input tree.
+For each candidate branch,
+RIPPLES systematically splits the sequence into 2 or 3 segments, and
+finds parsimonious placements for each segment independently.
+A recombination event at that node is then flagged if
+the overall parsimony of these multiple placements
+sufficiently improves over
+the original single-parent origin in the phylogeny
+(the ``parsimony improvement'' parameter, $p=3$, by default).
+RIVET~\cite{Smith2023tracking} is a tool for quality control and
+visualization of these recombination events.
+
+To compare the accuracy of sc2ts with UShER+RIPPLES we ran
+RIPPLES the Viridian UShER phylogeny
+subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
+We used the RIPPLES parameter $p=4$ as is equivalent to the 
+sc2ts $k=4$ parameter used to build the ARG, which resulted 
+in 11[XX] distinct recombination events. We then ran 3SEQ on
+each of the candidate trios, giving an overall validation
+rate of XX\% (STAR Methods). Comparing events identified 
+in the UShER tree and sc2ts ARG is not straightforward in 
+general, and so we focus comparison on the singleton recombinants
+returned by both. Although these are likely to be enriched for 
+artefacts, both methods face the same challenges and the 
+events can be compared like-for like. Of the 11XX recombination
+events returned by RIPPLES, 757 are singletons (resp 525/855 for 
+sc2ts). Of the RIPPLES singletons, 705 are also singletons 
+in the sc2ts ARG. [There's a complicated story to tell here 
+about extra singletons in the RIPPLES set which pass
+3SEQ validation, but have parents from the future.]
+Thus, 216 of these recombinants  pass 3SEQ validation, have 
+at least one parent that is on average 63 days \emph{younger}
+than the sequence itself. These parents were not available
+to sc2ts as potential matches, as it enforces strict 
+time ordering based on sample metadata.
+Further study is required to determine
+if the metadata is incorrect in these cases, or if these 
+should be flagged as false positives. However, the 
+key point is that RIPPLES is dependent on signals
+embedded in the UShER tree, and therefore we cannot 
+decouple the finer details of post-hoc recombinant detection
+from the tree building algorithm.
+
+We also compared the recombination events output by RIPPLES with
+the Pango X lineages. At p=4, we found that only XX of the 
+X lineages were present and many well-characterized events 
+such as XBB were not present. We therefore reran at p=3,
+resulting in XXXX events overall. At p=4, events involving
+XX of the 44 Pango X lineages present in the dataset,
+with unequivocal origins for 10 lineages, and 
+7 lineages other lineages associated with two competing events. 
+[More summary]
+
+[Paragraph summing things up]
+See Document~\ref{sec:detailed_ripples} for more detailed analysis 
+of the UShER+RIPPLES results.
 
@@
 
+\begin{table}
+\caption{Validation of sc2ts recombinants by 3SEQ, CovRecomb and rebar over 
+all 855 recombinants (total) and XXX passing and XXX failing QC.
+TODO FILL OUT DETAILS.
+}
+\label{tab:recombinant_validation}
+\centering
+\begin{tabular}{llll}
+\toprule
+Method & Overall & QC Pass & QC Fail \\
+\midrule
+3SEQ & xx & 88\% \\
+CovRecomb & \\
+rebar & \\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \subsection{Parsimony comparison of sc2ts and UShER}
@@
 \label{sec:detailed_ripples}
+
+At $p=3$, this resulted in a total of 4111 recombination events.
+Most of these subtend $<10$ samples (2616 singletons),
+with 4 events subtending more than 100,000 samples.
+Analysis of the descending samples for each event shows that
+10 of the 44 Pango X lineages present in the dataset
+% NOTE: listing them like this for sorting and manipulation purposes
+(%
+XA, 
+XAC, 
+XAE, 
+XAJ,
+XB, 
+XBB, 
+XBM, 
+XC, 
+XL, 
+XW%
+)
+are associated unequivocally with a single recombination event.
+7 lineages 
+(%
+XAL,
+XAM,
+XBD, 
+XBR, 
+XE,
+XS,
+XY%
+)
+are associated with two events. Finally, 
+one event subtending 1321 samples suggests that 
+XAF,
+XAL, 
+XE, 
+XH, 
+XJ, 
+XM, 
+XY, 
+and 
+82 BA.2 samples share a recombination event [With what breakpoint and parents?].
+We also ran RIPPLES with $p=4$, which reduced the overall number of
+events to 1174 (roughly 4-fold) but also approximately halved the number of
+events involving X lineages, missing well-characterized events,
+such as XBB (Table~\ref{tab:ripples_x_lineages}).
+
+To compare these results with the sc2ts ARG, we first matched clades
+by finding the most recent common ancestor (MRCA) of the samples
+descending from each RIPPLES event. Overall, clade agreement was
+high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG
+where descendants match exactly. Of these nodes, 68\% had no recombinant
+ancestry. Of the remainder, 595 could be matched with
+one of the 855 recombination events in the sc2ts ARG
+(Figure~\ref{fig:ripples_sc2ts_events}).
+Table~\ref{tab:ripples_x_lineages} shows a summary of
+all recombination events associated with Pango X lineages detected by RIPPLES,
+and their relationship with the sc2ts ARG.
+Of the 16 Class I events (Pango X lineages unequivocally associated with
+a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
+8 have corresponding recombination events (4, at $p=4$).
+Of the 5 Type II events
+(recombination nodes with more complex assocation with X lineages),
+two have close matches, with XBM matching exactly
+(including 2 descending BF.3 samples),
+and XE/XH is similar (but without including the 2 XH samples in the clade).
+Two events in the RIPPLES set correspond to likely false negatives
+in sc2ts: XB (Document~\ref{sec:dd_xb})
+and XAJ (Document~\ref{sec:dd_xaj}).
+See Document~\ref{sec:detailed_ripples} for more detailed analysis.
+
+[TODO Discuss  the large effect recombinants inferred by both]
+
+Two further recombinants in the sc2ts ARG that are likely to be 
+spurious are important to discuss.
+[FIXME - discuss BA.5 (kept as a recombination event but only because we can't 
+rewire it post-hoc, owning to subsequent tree building) and the BA.2 event (doesn't
+pass QC.) - note this is discussed in more detail in document \ref{sec:spurious-ancestrally-important-recombinants}]
 

### 2025-10-15 latex/main.tex
Updates from Overleaf

@@
 recombination detection methods on the putative recombinant sequences
-(STAR Methods) summarised in Table~\ref{tab:recombinant_validation}.
-Overall, validation among the QC-passing recombinants is highest
-at with 3SEQ at 88\%, and is lowest for CovRecomb at xx\%. 
+(STAR Methods) summarized in Table~\ref{tab:recombinant_validation}.
+Overall, validation among the QC-passing recombinants is
+the highest for 3SEQ at 88\% and
+the lowest for CovRecomb at 25\%.
 CovRecomb has been previously shown to be
 less sensitive than other methods~\cite{Li2024CovRecomb}, 
-and is specifically intended 
-to detect interlineage recombinants. The 12\% of recombinants 
+and is specifically intended to detect interlineage recombinants.
+As expected, CovRecomb missed most of the recombinants that
+involve closely related parental Pango lineages
+(Figure~\ref{fig:recombinant_evidence_cov_recomb}C),
+in contrast to RIPPLES
+which can detect intralineage recombinants
+(see Figure~\ref{fig:recombinant_evidence_ripples_p4} and
+Figure~\ref{fig:recombinant_evidence_ripples_p3}).
+The 12\% of recombinants
 not passing 3SEQ validation are also the most marginally supported
@@
 \subsection*{Assessing evidence for recombination events}
+
 The LS HMM at the heart of sc2ts provides a simple and powerful means 
@@
 
-Recombination is rare in SARS-CoV-2, and ultimately judgements about
-plausibility require expert analysis. 
-To aid such interpretation and help identify quality control
-issues we developed a visualisation 
-based on SnpIt~\cite{OToole2024} and 
-the recombinant view from RIVET~\cite{Smith2023tracking},
-showing the informative genomic positions between 
-a recombinant and its parent sequences, and highlighting 
-potentially problematic loci.
+Recombination is rare in SARS-CoV-2, and
+ultimately judgments about plausibility require expert analysis.
+To aid such interpretation and help identify quality control issues
+we developed a visualization tool
+inspired by SnpIt~\cite{OToole2024} and 
+the recombinant view of RIVET~\cite{Smith2023tracking},
+showing the recombination-informative sites
+between a recombinant and its parent sequences, and
+highlighting potentially problematic loci.
 Each copying pattern shows the positions
 where the allelic state in a recombinant
-(middle row) matches that of the left (upper row, coloured green) 
-or right parent (lower row, coloured blue),
-or where the recombination event requires a de-novo mutation
+(middle row) matches that of the left (upper row, colored green) 
+or right parent (lower row, colored blue),
+or where the recombination event requires a \emph{de novo} mutation
 (gold, with mutational change below).
@@
 large number of supporting loci on the left and right. As we have 
-only one de-novo mutation, we can be sure that the parents are closely
+only one \emph{de novo} mutation, we can be sure that the parents are closely
 observed [Some details about XA here, how well the parents were sampled etc].
 XBB is also strongly supported by many loci on the left and right, 
-but the presence of many de-novo mutations
+but the presence of many \emph{de novo} mutations
 observed in the copying patterns suggests the existence of
@@
 See Figures~\ref{fig:recombinant_evidence}D and
-\ref{fig:pango_x_copying_patterns} for copying patterns of the 
-Pango X lineages, and 
+\ref{fig:pango_x_copying_patterns}
+for copying patterns of the Pango X lineages, and
 Figure~\ref{fig:copying_pattern_examples} for copying patterns
-highlighting specific types of QC problem (and more details on the 
-visualisation).
+highlighting specific types of QC problems
+(and more details on the visualizations).
 
@@
 
-To compare the accuracy of sc2ts with UShER+RIPPLES we ran
-RIPPLES the Viridian UShER phylogeny
+To compare the accuracy of sc2ts with UShER+RIPPLES
+we ran RIPPLES the Viridian UShER phylogeny
 subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
 We used the RIPPLES parameter $p=4$ as is equivalent to the
-sc2ts $k=4$ parameter used to build the ARG, which resulted 
-in 11[XX] distinct recombination events. We then ran 3SEQ on
-each of the candidate trios, giving an overall validation
-rate of XX\% (STAR Methods). Comparing events identified 
-in the UShER tree and sc2ts ARG is not straightforward in 
-general, and so we focus comparison on the singleton recombinants
-returned by both. Although these are likely to be enriched for 
-artefacts, both methods face the same challenges and the 
+sc2ts $k=4$ parameter used to build the ARG,
+which resulted in 11[XX] distinct recombination events.
+We then ran 3SEQ on each of the candidate trios,
+giving an overall validation rate of XX\% (STAR Methods).
+Comparing events identified in the UShER tree and sc2ts ARG
+is not straightforward in general, and
+so we focus comparison on the singleton recombinants
+returned by both.
+Although these are likely to be enriched for artefacts,
+both methods face the same challenges and the 
 events can be compared like-for like. Of the 11XX recombination
@@
 (A) The amount of time to process a daily batch of samples
-peaks when the sample volume is the highest.
-(B) Mean time to process a sample remained low.
+peaks when the sampling effort is the highest.
+(B) Mean time to process a sample remained low throughout,
+except towards the very end when there were few exact matches.
+The HMM cost was moderate initially,
+then noticeably decreased when sampling effort is the highest
+before increasing again towards the end.
 (C) Max RAM usage also peaks around the time
-that sampling effort is the highest.
+when sampling effort is the highest.
 }

### 2025-10-15 latex/main.tex
Tweaks

@@
 Recombinant lineages have risen to global prevalence in SARS-CoV-2,
-and therefore simple tree representations are necessarily
+and simple tree representations are therefore necessarily
 incomplete and potentially misleading.
@@
 insights into the relationships between previously identified
-recombinant lineages.
+recombinant lineages. [NOTE: mention overall estimate
+of recombination compared with UShER+RIPPLES?]
 Together, the ARG, inference method and 
@@
 % Accessed GISAID/EpiCov on 08 Oct., 2025
-At the peak in February, 2022, about 80 thousand samples
-were submitted on average per day to GISAID\cite{Shu2017},
+At the peak in February 2022 about 80 thousand samples
+per day  were submitted to GISAID\cite{Shu2017},
 accumulating to a total of over 17.5 million samples
-as of October, 2025.
+as of October 2025.
 Regularly reinferring trees de-novo from such rapidly
@@
 recombinant lineages. We compare the recombination events detected by 
-sc2ts with the state-of-the-art RIPPLES method~\cite{Turakhia2022}, showing
-that sc2ts is more sensitive and has a substantially lower false positive rate.
+sc2ts with the state-of-the-art combination of UShER and 
+RIPPLES ~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
+detecting more of the well-characterized recombination events 
+curated by the Pango lineage designation community.
 The output of sc2ts---incorporating
@@
 controls the number of recurrent mutations that is preferred to 
-switching to a different parent via recombination (we use $k=4$ here).
+switching to a different parent via recombination (we use $k=4$ here
+[TODO cite support for this, like CovRecomb etc]).
 We used the highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
@@
 UShER tree closely match by generating tanglegrams for a set of representative
-samples from 814 deeply-sampled lineages (STAR Methods; 
+samples from 814 well-sampled lineages (STAR Methods; 
 Figures~\ref{fig:tanglegram_base},
@@
 estimates (Figure \ref{fig:pango_vs_nextstrain_node_times})
-and overall accuracy to another large-scale
-method~\cite{sanderson2021chronumental} (Document \ref{sec:evaluating_node_dates}).
+and overall accuracy comparable to 
+Chronumental~\cite{sanderson2021chronumental} applied to 
+an UShER tree (Document \ref{sec:evaluating_node_dates}).
 Finally, we verified
@@
 Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
-but are not incorporated into UShER trees and can be difficult
-to account for. Because deletions are difficult to distinguish 
+but are not incorporated into UShER trees and must therefore
+be accounted for separately.
+Because deletions are difficult to distinguish 
 from bioinformatics errors

### 2025-10-15 arg_postprocessing/scripts/process_ripples.py
Update ripples data to include dates, parents, etc

@@ -6,17 +6,25 @@ import tszip
 import dataclasses
 import collections
 import tqdm
+import bte
 
 
 @dataclasses.dataclass
 class UsherEvent:
-    usher_node: str
+    recombinant: str
+    parent1: str
+    parent2: str
+    recombinant_date: str
+    parent1_date: str
+    parent2_date: str
     descendants: list[str]
     pango_counts: dict
     sc2ts_mrca: int = -1
     sc2ts_num_descendants: int = 0
     sc2ts_closest_recombinant: int = -1
     sc2ts_closest_recombinant_path_len: int = 2**31
+    sc2ts_num_mutations: int = -1
+    x_lineage_involved: bool = False
 
     def asdict(self):
         d = dataclasses.asdict(self)
@@ -27,20 +35,37 @@ class UsherEvent:
 
 @click.command()
 @click.argument("ripples_dir")
+@click.argument("usher_tree")
 @click.argument("sc2ts_ts")
 @click.argument("output")
-def run(ripples_dir, sc2ts_ts, output):
+def run(ripples_dir, usher_tree, sc2ts_ts, output):
+
+    mt = bte.MATree(usher_tree)
 
     descendants = {}
     with open(f"{ripples_dir}/descendants.tsv") as f:
         for line in f:
             if not line.startswith("#"):
                 split = line.strip().split("\t")
+                node_id = split[0]
                 descs = split[1].split(",")
                 if descs[-1] == "":
                     descs = descs[:-1]
-                descendants[split[0]] = descs
+                descendants[node_id] = descs
+                # Check that we're looking at the same things here
+                assert mt.count_leaves(node_id) == len(descs)
+
     len("Read {len(descendants)} events")
+
+    df = pd.read_csv(f"{ripples_dir}/recombination.tsv", sep="\t")
+    df["recomb_node_id"] = df["#recomb_node_id"]
+    print(f"Staring with {df.shape[0]} records")
+    # The dataframe can contain multiple events for each recombination node. We
+    # pick the one with the "maximum" parsimony (i.e., the one with the lowest
+    # parsimony score). If there's several, we pick one arbitrarily.
+    df_events = df.loc[df.groupby(["recomb_node_id"])["recomb_parsimony"].idxmin()]
+    df_events = df_events.set_index("recomb_node_id")
+    print(f"Have {df_events.shape[0]} unique recombination events")
 
     ts = tszip.load(sc2ts_ts)
     dfn = sc2ts.node_data(ts, inheritance_stats=False)
[…]

### 2025-10-15 latex/main.tex
Updates from Overleaf

@@
 of quantifying the plausibility of a given recombination event.
-To do this, we force the HMM to find a Viterbi solution for the 
-putative recombinant sequence with no recombination by specifying
+To do this, we force the HMM to find a Viterbi solution
+for the putative recombinant sequence
+with no recombination by specifying
 a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
@@
 ``averted'' by recombination---is a measure of
-the evidence in favor of recombination  (similar to 
-idea of ``parsimony improvement'' in RIPPLES.)
+the evidence in favor of recombination
+(similar to the idea of ``parsimony improvement'' in RIPPLES.)
 Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
@@
 
-Figure~\ref{fig:recombinant_evidence}A plots the number of mutations required
-in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
+Figure~\ref{fig:recombinant_evidence}A plots
+the number of mutations required
+in the recombinant versus non-recombinant
+Viterbi solutions (upper scatter plot)
 and the resulting distribution of the number of
@@
 ~\ref{fig:recombinant_evidence_rebar} 
-same the same information for CovRecomb and rebar validation,
+plot the same information for CovRecomb and rebar validation,
 showing that their validation status tends to be function
@@
 Although these are likely to be enriched for artefacts,
-both methods face the same challenges and the 
-events can be compared like-for like. Of the 11XX recombination
-events returned by RIPPLES, 757 are singletons (resp 525/855 for 
-sc2ts). Of the RIPPLES singletons, 705 are also singletons 
-in the sc2ts ARG. [There's a complicated story to tell here 
+both methods face the same challenges and
+the events can be compared like-for like.
+Of the 11XX recombination events returned by RIPPLES,
+757 are singletons (respectively, 525/855 for sc2ts).
+Of the RIPPLES singletons, 705 are also singletons in the sc2ts ARG.
+[There's a complicated story to tell here
 about extra singletons in the RIPPLES set which pass
 3SEQ validation, but have parents from the future.]
-Thus, 216 of these recombinants pass 3SEQ validation, have 
-at least one parent that is on average 63 days \emph{younger}
-than the sequence itself. These parents were not available
-to sc2ts as potential matches, as it enforces strict 
-time ordering based on sample metadata.
+Thus, 216 of these recombinants pass 3SEQ validation,
+have at least one parent that is on average
+63 days \emph{younger} than the sequence itself.
+These parents were not available to sc2ts as potential matches,
+as it enforces strict time ordering based on sample metadata.
 Further study is required to determine
-if the metadata is incorrect in these cases, or if these 
-should be flagged as false positives. However, the 
-key point is that RIPPLES is dependent on signals
+if the metadata is incorrect in these cases, or
+if these should be flagged as false positives.
+However, the key point is that RIPPLES is dependent on signals
 embedded in the UShER tree, and therefore we cannot
@@
 
-We also compared the recombination events output by RIPPLES with
-the Pango X lineages. At p=4, we found that only XX of the 
-X lineages were present and many well-characterized events 
-such as XBB were not present. We therefore reran at p=3,
-resulting in XXXX events overall. At p=4, events involving
-XX of the 44 Pango X lineages present in the dataset,
-with unequivocal origins for 10 lineages, and 
+We also compared the recombination events
+output by RIPPLES with the Pango X lineages.
+At p=4, we found that only XX of the Pango X lineages were present and
+many well-characterized events such as XBB were not present.
+We therefore reran at p=3, resulting in XXXX events overall.
+At p=4, events involving XX of the 44 Pango X lineages
+present in the dataset, with unequivocal origins for 10 lineages, and
 7 lineages other lineages associated with two competing events.

### 2025-10-15 latex/main.tex
Move the USHER+RIPPLES section to the tend
Also comment out section on compute

Update

@@
 
-\subsection*{Comparison with UShER+RIPPLES}
-
-RIPPLES~\cite{Turakhia2022} is a method for identifying
-recombination events using a phylogeny inferred by UShER.
-This distinguishes RIPPLES from other recombination detection
-methods such as 3SEQ, which classify particular \emph{sequences}
-as recombinant or not and can only identify potential onward 
-transmission post-hoc. The combination of UShER and RIPPLES
-is therefore the only method that is directly comparable 
-to sc2ts. It works by first identifying plausible long 
-branches in the input tree.
-For each candidate branch,
-RIPPLES systematically splits the sequence into 2 or 3 segments, and
-finds parsimonious placements for each segment independently.
-A recombination event at that node is then flagged if
-the overall parsimony of these multiple placements
-sufficiently improves over
-the original single-parent origin in the phylogeny
-(the ``parsimony improvement'' parameter, $p=3$, by default).
-RIVET~\cite{Smith2023tracking} is a tool for quality control and
-visualization of these recombination events.
-
-To compare the accuracy of sc2ts with UShER+RIPPLES
-we ran RIPPLES the Viridian UShER phylogeny
-subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
-We used the RIPPLES parameter $p=4$ as is equivalent to the
-sc2ts $k=4$ parameter used to build the ARG,
-which resulted in 11[XX] distinct recombination events.
-We then ran 3SEQ on each of the candidate trios,
-giving an overall validation rate of XX\% (STAR Methods).
-Comparing events identified in the UShER tree and sc2ts ARG
-is not straightforward in general, and
-so we focus comparison on the singleton recombinants
-returned by both.
-Although these are likely to be enriched for artefacts,
-both methods face the same challenges and
-the events can be compared like-for like.
-Of the 11XX recombination events returned by RIPPLES,
-757 are singletons (respectively, 525/855 for sc2ts).
-Of the RIPPLES singletons, 705 are also singletons in the sc2ts ARG.
-[There's a complicated story to tell here
-about extra singletons in the RIPPLES set which pass
-3SEQ validation, but have parents from the future.]
-Thus, 216 of these recombinants pass 3SEQ validation,
-have at least one parent that is on average
-63 days \emph{younger} than the sequence itself.
-These parents were not available to sc2ts as potential matches,
-as it enforces strict time ordering based on sample metadata.
-Further study is required to determine
-if the metadata is incorrect in these cases, or
-if these should be flagged as false positives.
-However, the key point is that RIPPLES is dependent on signals
-embedded in the UShER tree, and therefore we cannot
-decouple the finer details of post-hoc recombinant detection
-from the tree building algorithm.
-
-We also compared the recombination events
-output by RIPPLES with the Pango X lineages.
-At p=4, we found that only XX of the Pango X lineages were present and
-many well-characterized events such as XBB were not present.
-We therefore reran at p=3, resulting in XXXX events overall.
-At p=4, events involving XX of the 44 Pango X lineages
-present in the dataset, with unequivocal origins for 10 lineages, and
-7 lineages other lineages associated with two competing events.
-[More summary]
-
-[Paragraph summing things up]
-See Document~\ref{sec:detailed_ripples} for more detailed analysis 
-of the UShER+RIPPLES results.
-
 \subsection*{Precision of recombination event times and positions}
@@
 
-\subsection*{Efficient inference and analysis of a pandemic-scale ARG}
-
-The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
-which overwhelmed existing computational approaches,
-motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
-Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
-the functionality required largely overlaps.
-Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
-and the VCF Zarr data storage format \cite{Czech2025}
-to describe relationships among virus samples,
-building an ARG over time as data becomes available.
-By reusing high-quality and high-performance components of these toolkits,
-we could develop the sc2ts inference method itself and
-perform all the pandemic-scale analyses in this article
-with Python code alone and no compiled languages.
-
-To illustrate the power of these libraries, we provide some examples
-in which we compare the Usher and sc2ts intersections in tskit format described earlier.
-First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments.
-The total aligned dataset for 1,226,587 samples and 27,431 sites
-shared by the sc2ts and Usher tree sequences represents 31.34 GB of nucleotide calls.
-Of the 80,574,518 missing data calls (Ns) in the alignments,
-sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%).
-Additionally, 413,303 calls made use of the IUPAC uncertainty codes.
-Of these sc2ts imputed only 77 (0.02\%) incorrectly
-(i.e., with a base that is not compatible with the ambiguity code).
-UShER imputed only 4 calls from this set incorrectly,
-reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations.
-This computation,
-performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
-required 30 lines of Python code and took 2m 59s on an intel i7-9700 CPU,
-with a peak memory usage of about 4.5 GB
-(see \cite{Czech2025} for further benchmarks for
-processing the Viridian data in VCF Zarr format).
-Performing this analysis using matUtils would require
-exporting the full dataset to VCF, and
-converting the VCF to FASTA, and
-then comparing the alignments one-by-one to the source FASTA alignments.
-In contrast with matUtils and most phylogenetic software,
-which provide a command line interface and suite of predefined operations that typically output text files,
-tskit is a Python API (with C and Rust interfaces also supported)
-which has many advantages \cite{Baumdicker2022}.
-As an illustration of the open-ended and efficient nature of the analyses this enables,
-we computed the number of polytomies (nodes that have $> 2$ children)
-and the maximum number of children per node in the Usher and sc2ts outputs.
-In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148)
-and compute the maximum number of children (5918).
-In the sc2ts ARG, it took 385 ms and 6 lines of code
-to find the maximum number of polytomies (137,001) and
-maximum number of children (7708) across all 348 local trees.
+\subsection*{Comparison with UShER+RIPPLES}
+RIPPLES~\cite{Turakhia2022} is a method for identifying
+recombination events using a phylogeny inferred by UShER.
+This distinguishes RIPPLES from other recombination detection
+methods such as 3SEQ, which classify particular \emph{sequences}
+as recombinant or not and can only identify potential onward transmission post-hoc. 
+The combination of UShER and RIPPLES
+is therefore the only method that is directly comparable to sc2ts. 
+RIPPLES works by first identifying long 
+branches in the input tree, by default those carrying 3 or more 
+mutations. For each candidate branch,
+RIPPLES systematically splits the sequence into 2 or 3 segments, and
+finds parsimonious placements for each segment independently.
+A recombination event at that node is then flagged if
+the overall parsimony of these multiple placements
+sufficiently improves over
+the original single-parent origin in the phylogeny.
+This ``parsimony improvement'' parameter ($p$)
+is equivalent to the sc2ts recombination penalty ($k$).
+
+To compare the performance of sc2ts and UShER+RIPPLES 
+in terms of recombinant detection, we ran RIPPLES the Viridian UShER phylogeny
+subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
+Using the RIPPLES parameter $p=4$ (equivalent to $k=4$ used 
+to build the sc2ts ARG) resulted in 1174 reported recombination events.
+We ran 3SEQ on each of the candidate trios (STAR Methods) and 601 
+were classified as recombinant, giving an overall validation rate of 51\%.
+Aligning the events reported by both methods using the sets of descendant
+samples (STAR Methods), 468 of the events are shared by both, 
+with the intersection comprising 40\% of the RIPPLES events
+and 55\% of the sc2ts events (Figure~\ref{fig:ripples_sc2ts_events}). 
+The 706 recombination events output by RIPPLES that do not 
+correspond to a recombination in the sc2ts ARG can largely
+be explained by a combination of two factors.
+Firstly, 408 of these RIPPLES identified recombination events
+have one or more parents that are \emph{younger} than the
+recombinant sequence based on metadata. Thus, the parental
+sequences were simply not available to match against for 
+sc2ts (which enforces strict time ordering) when the 
+recombinant sequence was inserted. Secondly, of the 179
+remaining events with matching clades and in which the 
+sc2ts node has more than 4 mutations, of the total
+1,167 mutations, 452 of these occured at sites that were 
+remapped during post-hoc parsimony. The remaining ``missing''
+events are likely attributable to the finer details of the 
+differences in tree building approaches between UShER
+and sc2ts.
+
+The 387 events detected by sc2ts and not by UShER+RIPPLES
+would require a detailed dissection of the UShER tree
+to explain, which is beyond the scope of this work.
+However, the set does contain some well characterised events such as XBB.
+We also compared the recombination events
+output by RIPPLES with the Pango X lineages.
+At p=4, we found that only XX of the Pango X lineages were present and
+many well-characterized events such as XBB were not present.
+We therefore reran at p=3, resulting in 4111 events overall,
+and increasing the intersection of events with sc2ts to 595 (70\%).
+At p=3, events involving XX of the 44 Pango X lineages
+present in the dataset, with unequivocal origins for 10 lineages, and
+7 lineages other lineages associated with two competing events.
+[More summary]
+
+[Paragraph summing things up. It seems like RIPPLES is doing its
+job, and the question is really whether UShER is giving 
+the correct signals for it to detect. Sc2ts detects more of the 
+well-known events. Needs careful explanation.].
+See Document~\ref{sec:detailed_ripples} for more detailed analysis 
+of the UShER+RIPPLES results.
+
+% TODO convert this into a discussion pargraph..
+% \subsection*{Efficient inference and analysis of a pandemic-scale ARG}
+
+% The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
+% which overwhelmed existing computational approaches,
+% motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
+% Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
+% the functionality required largely overlaps.
+% Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
+% and the VCF Zarr data storage format \cite{Czech2025}
+% to describe relationships among virus samples,
+% building an ARG over time as data becomes available.
+% By reusing high-quality and high-performance components of these toolkits,
+% we could develop the sc2ts inference method itself and
+% perform all the pandemic-scale analyses in this article
+% with Python code alone and no compiled languages.
+
+% To illustrate the power of these libraries, we provide some examples
+% in which we compare the Usher and sc2ts intersections in tskit format described earlier.
+% First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments.
+% The total aligned dataset for 1,226,587 samples and 27,431 sites
+% shared by the sc2ts and Usher tree sequences represents 31.34 GB of nucleotide calls.
+% Of the 80,574,518 missing data calls (Ns) in the alignments,
+% sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%).
+% Additionally, 413,303 calls made use of the IUPAC uncertainty codes.
+% Of these sc2ts imputed only 77 (0.02\%) incorrectly
+% (i.e., with a base that is not compatible with the ambiguity code).
+% UShER imputed only 4 calls from this set incorrectly,
+% reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations.
+% This computation,
+% performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
+% required 30 lines of Python code and took 2m 59s on an intel i7-9700 CPU,
+% with a peak memory usage of about 4.5 GB
+% (see \cite{Czech2025} for further benchmarks for
+% processing the Viridian data in VCF Zarr format).
+% Performing this analysis using matUtils would require
+% exporting the full dataset to VCF, and
+% converting the VCF to FASTA, and
+% then comparing the alignments one-by-one to the source FASTA alignments.
+% In contrast with matUtils and most phylogenetic software,
+% which provide a command line interface and suite of predefined operations that typically output text files,
+% tskit is a Python API (with C and Rust interfaces also supported)
+% which has many advantages \cite{Baumdicker2022}.
+% As an illustration of the open-ended and efficient nature of the analyses this enables,
+% we computed the number of polytomies (nodes that have $> 2$ children)
+% and the maximum number of children per node in the Usher and sc2ts outputs.
+% In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148)
+% and compute the maximum number of children (5918).
+% In the sc2ts ARG, it took 385 ms and 6 lines of code
+% to find the maximum number of polytomies (137,001) and
+% maximum number of children (7708) across all 348 local trees.
 

### 2025-10-16 latex/main.tex
Updates from Overleaf

@@
 To compare the performance of sc2ts and UShER+RIPPLES 
-in terms of recombinant detection, we ran RIPPLES the Viridian UShER phylogeny
+in terms of recombinant detection, we ran RIPPLES on the Viridian UShER phylogeny
 subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
@@
 sc2ts node has more than 4 mutations, of the total
-1,167 mutations, 452 of these occured at sites that were 
+1,167 mutations, 452 of these occurred at sites that were 
 remapped during post-hoc parsimony. The remaining ``missing''
@@
 to explain, which is beyond the scope of this work.
-However, the set does contain some well characterised events such as XBB.
+However, the set does contain some well characterized events such as XBB.
 We also compared the recombination events

### 2025-10-16 latex/main.tex
Add ARGscape ref

@@
 simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
-visualization~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
+visualization~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit,Talbot2025argscape},
 inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}

### 2025-10-16 latex/main.tex
Updates from Overleaf

--- 2025-10-16 latex/main.tex
@@
 The 706 recombination events output by RIPPLES that do not 
-correspond to a recombination in the sc2ts ARG can largely
+correspond to a recombination event in the sc2ts ARG can largely
 be explained by a combination of two factors.

### 2025-10-16 latex/main.tex
Comment in the discussion

@@
 controls the number of recurrent mutations that is preferred to 
-switching to a different parent via recombination (we use $k=4$ here
-[TODO cite support for this, like CovRecomb etc]).
+switching to a different parent via recombination 
+% This is quite a big thing really, and ultimately what determines
+% the number of recombinants detected.
+(we use $k=4$ here; see STAR Methods and Discussion).
 We used the highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
@@
 methods such as 3SEQ, which classify particular \emph{sequences}
-as recombinant or not and can only identify potential onward transmission post-hoc. 
+as recombinant or not and can only identify 
+potential onward transmission post-hoc. 
 The combination of UShER and RIPPLES
@@
 
-This estimate for effective recombination in SARS-CoV-2 is much lower 
-than previous studies have suggested.
+The number of recombination events that we detect is largely 
+driven by the choice of recombination penalty, $k$ (or $p$
+for RIPPLES). 
+At $k=4$ we detect fewer recombinants overall than the combination of 
+UShER and RIPPLES at the same threshold. 
+However, most of the extra recombinants detected by UShER+RIPPLES
+involve violations of time ordering; whether this is due to incorrect
+metadata or these are false positives would require further study.
+However, at $p=4$, UShER+RIPPLES detects very few of the well-characterised
+Pango X-lineages which are detected by sc2ts. At the more lenient
+threshold of $p=3$, more of these events are detected (but still fewer
+than sc2ts) but at the cost of nearly 4 times more recombination
+events overall. Our detailed analysis shows that RIPPLES has a well
+defined task, which it performs well, finding the potential events
+in an exhaustive search. However, it can only operate within the
+input that it is provided, and it is the UShER tree provided with
+the Viridian data that does not have clear signals for recombination
+of X lineages such as XBB at the $p=4$ threshold. 
+The inference of recombination is therefore intrinsically linked
+with the underlying genealogy inference method.
+
+[Whether k=3 or k=4 is ``correct'' is a matter for debate, and feeds
+into the next point.]
+
+The estimate for effective recombination in SARS-CoV-2 that we obtain
+at k=4 is much lower 
+than previous studies have suggbalanced
 % Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
@@
 In an early analysis, 225 potentially recombinant sequences out 
-of 87,695 samples [WHAT DATE?]\cite{Varabyou2021rapid}.
+of 87,695 samples collected before October 2020\cite{Varabyou2021rapid}.
 Over the same time period, XX events are present in the sc2ts ARG,
@@
 
-We developed a SnpIt-like~\cite{OToole2024} visualisation 
+We developed a SnipIt-like~\cite{OToole2024} visualisation 
 that shows the informative genomic positions between 

### 2025-10-16 latex/main.tex
Add recombinant evidence notebook
Closes #898

@@
 For example, the three characteristic deletions of Alpha \cite{RambautAlpha2020}
-(11288+9, 21765+6, 21991+3,
-where the first and second numbers are
-the start position and length, respectively)
+(11288:9, 21765:6, 21991:3)
 are correctly mapped above the Alpha origin in the ARG
@@
 for example,
-the 21765+6 deletion ($\Delta$H69/V70),
+the 21765:6 deletion ($\Delta$H69/V70),
 which causes S-Gene Target Failure \cite{Walker2021} and
 is associated with increased efficiency in cell entry \cite{Meng2021}, and
-the 21991+3 deletion
+the 21991:3 deletion
 which is associated with antibody escape \cite{McCarthy2021}.
@@
 based on the analysis of Jackson et al.\cite{Jackson2021},
-which established that recombination with onward transmission had occurred [weak FIXME].
+which established that recombination with onward transmission had occurred.
 Although the process of incorporating recombinant lineages
@@
 XAA, XAG, XAM, XR and XU are all closely derived, suggesting 
-that they are more parsimoniously explained as sublineages or sisters of XQ
+% NOTE: deleted the "sister" bit because it's confusing when
+% reading the document at this level. Keep the detailed hyptothese
+% to the deep dives.
+that they are more parsimoniously explained as sublineages of XQ
 than de-novo recombinant lineages (Document \ref{sec:dd_xq}). 
@@
 We validated the recombination events in the sc2ts ARG by running 
-the 3SEQ~\cite{Lam2018improved}, CovRecomb~\cite{Li2024CovRecomb} and rebar 
-recombination detection methods on the putative recombinant sequences
-(STAR Methods) summarized in Table~\ref{tab:recombinant_validation}.
-Overall, validation among the QC-passing recombinants is
-the highest for 3SEQ at 88\% and
-the lowest for CovRecomb at 25\%.
-CovRecomb has been previously shown to be
-less sensitive than other methods~\cite{Li2024CovRecomb}, 
-and is specifically intended to detect interlineage recombinants.
-As expected, CovRecomb missed most of the recombinants that
-involve closely related parental Pango lineages
-(Figure~\ref{fig:recombinant_evidence_cov_recomb}C),
-in contrast to RIPPLES
-which can detect intralineage recombinants
-(see Figure~\ref{fig:recombinant_evidence_ripples_p4} and
-Figure~\ref{fig:recombinant_evidence_ripples_p3}).
-The 12\% of recombinants
-not passing 3SEQ validation are also the most marginally supported
+3SEQ~\cite{Lam2018improved} on the putative recombinant 
+sequence trios (STAR Methods), 
+and 383 of the 855 (45\%) were classified as recombinant.
+Validation status correlates strongly with our QC filter, with 
+88\% of the 354 QC-passing recombinants 
+and 14\% of the QC-failing recombinants
+being classified as recombinant.
+The 12\% of QC-passing recombinants
+not validated by 3SEQ are also among the most marginally supported
 (see next section; Figure~\ref{fig:recombinant_evidence}).
-All of the validation metrics also correlate strongly with the 
-QC filter, with very few of the QC-failing recombinants
-passing external validation (Table~\ref{tab:recombinant_validation}).
+In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
+and rebar recombination detection methods on the putative recombinant 
+sequences, summarized in Table~\ref{tab:recombinant_validation}.
+CovRecomb has a substantially lower validation rate at 25\% 
+among the QC-passing recombinants. This is expected, as 
+CovRecomb and is specifically intended to detect interlineage recombinants
+and been previously shown to be
+less sensitive than other methods~\cite{Li2024CovRecomb}.
+Rebar validation is intermediate, at 57\% among QC-passing recombinants.
+(See also Figures~\ref{fig:recombinant_evidence_cov_recomb},
+and \ref{fig:recombinant_evidence_rebar}.)
+% We don't talk about RIPPLES results yet as we haven't discussed it, a
+% and that would be confusing.
 
@@
 \caption{Validation of sc2ts recombinants by 3SEQ, CovRecomb and rebar over 
-all 855 recombinants (total) and XXX passing and XXX failing QC.
-TODO FILL OUT DETAILS.
+all 855 recombinants and the 354 passing and 501 failing QC.
 }
@@
 \toprule
-Method & Overall & QC Pass & QC Fail \\
+Method & Total & QC Pass & QC Fail \\
 \midrule
-3SEQ & xx & 88\% \\
-CovRecomb & \\
-rebar & \\
+3SEQ & 383 & 71 & 312 \\
+CovRecomb & 92& 90 & 2\\
+rebar & 213& 203 & 10 \\
 \bottomrule

### 2025-10-16 arg_postprocessing/scripts/run_3seq.py
Finalise sc2ts 3SEQ results

@@ -47,22 +47,24 @@ def run_single_3seq(child_fasta, parent_fastas):
     return output
 
 
-# TODO rejigger this to use the new run_single_3seq function above and
-# use the different approach to putting fastas together.
 @click.command()
 @click.argument("recombinants_csv")
 @click.argument("fasta_dir")
 @click.argument("output")
-def run_3seq(recombinants_csv, fasta_dir, output):
+def run_sc2ts_3seq(recombinants_csv, fasta_dir, output):
     fasta_dir = pathlib.Path(fasta_dir)
     dfr = pd.read_csv(recombinants_csv)
 
+    def fasta_file(node_id):
+        return (fasta_dir / f"{node_id}.fasta").absolute()
+
     data = []
-    for recombinant in tqdm.tqdm(dfr["recombinant"].values):
-        fasta = fasta_dir / f"{recombinant}.fasta"
-        out = run_single_3seq(fasta)
+    for _, row in tqdm.tqdm(list(dfr.iterrows())):
+        child_fasta = fasta_file(row["recombinant"])
+        parent_fastas = [fasta_file(row[p]) for p in ["parent_left", "parent_right"]]
+        out = run_single_3seq(child_fasta.absolute(), parent_fastas)
         if len(out) > 0:
-            out["recombinant"] = recombinant
+            out["recombinant"] = row["recombinant"]
             data.append(out)
     # print(data)
     df = pd.concat(data)
@@ -73,7 +75,7 @@ def run_3seq(recombinants_csv, fasta_dir, output):
 @click.argument("ts")
 @click.argument("recombinants_csv")
 @click.argument("output_dir")
-def generate_fasta(ts, recombinants_csv, output_dir):
+def generate_sc2ts_fasta(ts, recombinants_csv, output_dir):
     output_dir = pathlib.Path(output_dir)
     ts = tszip.load(ts)
     dfr = pd.read_csv(recombinants_csv)
@@ -87,7 +89,7 @@ def generate_fasta(ts, recombinants_csv, output_dir):
     if not np.all(ts.nodes_flags == tskit.NODE_IS_SAMPLE):
         raise ValueError("Not all the nodes are samples.")
 
-    labels = ["sample", "parent_left", "parent_right"]
+    labels = ["recombinant", "parent_left", "parent_right"]
     samples = []
     for _, row in dfr.iterrows():
         samples.extend([row[label] for label in labels])
[…]

### 2025-10-16 arg_postprocessing/scripts/run_3seq.py
Finalise code for running 3SEQ in the pipeline

@@ -11,23 +11,9 @@ import os.path
 import dataclasses
 
 
-def _old_run_single_3seq(fasta_file):
-
-    exe = os.path.abspath("./tmp/3seq/3seq")
-    fasta_file = os.path.abspath(fasta_file)
-    with tempfile.TemporaryDirectory() as tempdir:
-        # NOTE: I can't get 3seq to run in -triplet mode whatever I do,
-        # but full seems to work. Same thing, ultimately?
-        subprocess.check_output(
-            f"yes | {exe} -full {fasta_file}", shell=True, cwd=tempdir
-        )
-        output = pd.read_csv(pathlib.Path(tempdir) / "3s.rec.csv")
-    return output
-
-
 def run_single_3seq(child_fasta, parent_fastas):
 
-    exe = os.path.abspath("./tmp/3seq/3seq")
+    exe = os.path.abspath("./3seq/3seq")
     # if True:
     #     tempdir = pathlib.Path("tmp/")
     with tempfile.TemporaryDirectory() as tempdir:
@@ -112,12 +98,8 @@ def generate_sc2ts_fasta(ts, recombinants_csv, output_dir):
 @click.argument("ripples_file")
 @click.argument("output")
 def generate_ripples_sample_list(ripples_file, output):
-    df = pd.read_csv(ripples_file, sep="\t")
-    samples = (
-        set(df["#recomb_node_id"])
-        | set(df["donor_node_id"])
-        | set(df["acceptor_node_id"])
-    )
+    df = pd.read_csv(ripples_file)
+    samples = set(df["recombinant"]) | set(df["parent1"]) | set(df["parent2"])
     samples = np.array(sorted(list(samples)))
     np.savetxt(f"{output}", samples, fmt="%s")
 
@@ -127,14 +109,8 @@ def generate_ripples_sample_list(ripples_file, output):
 @click.argument("fasta_dir")
 @click.argument("output")
 def run_ripples_3seq(ripples_file, fasta_dir, output):
-    df = pd.read_csv(ripples_file, sep="\t")
-    df["recomb_node_id"] = df["#recomb_node_id"]
-    print(f"Staring with {df.shape[0]} records")
-    # The dataframe can contain multiple events for each recombination node. We
-    # pick the one with the "maximum" parsimony (i.e., the one with the lowest
-    # parsimony score). If there's several, we pick one arbitrarily.
-    df = df.loc[df.groupby(["recomb_node_id"])["recomb_parsimony"].idxmin()]
[…]

### 2025-10-16 latex/main.tex
Minor  edits

--- 2025-10-16 latex/main.tex
@@
 ``averted'' by recombination---is a measure of
-the evidence in favor of recombination
-(similar to the idea of ``parsimony improvement'' in RIPPLES.)
+the evidence in favor of recombination.
+[FIXME need to insert a sentence saying this is similar to 
+metrics used by RIPPLES and other methods, contrasting
+recombinant vs non-recombinant explanations.]
 Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
@@
 passing quality control filters
-colored by CovRecomb classification (recombinant: purple, non-recombinant: gray).
+colored by 3SEQ classification (recombinant: purple, non-recombinant: gray).
 (A) Upper: jittered scatterplot comparing number of mutations
@@
 Lower: recombination nodes classified by number of mutations averted
-by allowing recombination (i.e. $x$-$y$).
+by allowing recombination (i.e. $x-y$).
 (B) Distribution of time to the MRCA of the parents of each QC-passing recombination node.
@@
 At p=3, events involving XX of the 44 Pango X lineages
-present in the dataset, with unequivocal origins for 10 lineages, and
+are present in the dataset, with unequivocal origins for 10 lineages, and
 7 lineages other lineages associated with two competing events.

### 2025-10-16 latex/main.tex
First details on 3SEQ methods

@@
 \subsubsection*{Quality control of recombination events}
-[TODO rewrite this to account for new Results section]
-
-Sequencing errors in genome assemblies \cite{Turakhia2020},
-read misalignments \cite{Hunt2024}, and
-multi-nucleotide substitutions (MNS \cite{DeMaio2020issues})
-can cause short regions of a sample genome to match more closely
-to an alternative genome rather than the true parent.
-Such genomic regions can be mistaken for evidence of recombination,
-because the LS HMM counts each base separately.
-Fortunately, artefactual recombinants caused by this effect are usually
-identifiable because they affect multiple adjacent or near-adjacent
-bases simultaneously.
-
-We devised a quality control (QC) measure based on this observation, and
-applied it to each putative recombination event in the ARG.
+
 We examined the genomic positions
@@
 For each genomic region identified by the HMM as having a single parent,
-we calculated the "net number of supporting loci"
+we calculated the ``net number of supporting loci''
 by counting the number of loci in the region
@@
 of which 6 failed QC.
+
+\subsubsection*{Validation of recombination events}
+
+We validated recombination events in the sc2ts ARG 
+using 3SEQ~\cite{Lam2018improved} as follows.
+We first exported the alignments for the recombination node and both 
+parents to FASTA format, and then rand 3SEQ in ``full mode''
+independently for each recombinant. Specifically, we 
+ran \texttt{3seq -full  parents.fa recombinant.fa} using default 
+thresholds.
+
+We used 3SEQ source code downloaded from \url{https://gitlab.com/lamhm/3seq}
+at git hash e45ee3b4 for all analyses.
 

### 2025-10-16 latex/main.tex
Updates from Overleaf

--- 2025-10-16 latex/main.tex
@@
 ``averted'' by recombination---is a measure of
-the evidence in favor of recombination.
-[FIXME need to insert a sentence saying this is similar to 
-metrics used by RIPPLES and other methods, contrasting
-recombinant vs non-recombinant explanations.]
+the evidence in favor of recombination
+(similar to the idea of ``parsimony improvement'' in RIPPLES.)
 Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
@@
 passing quality control filters
-colored by 3SEQ classification (recombinant: purple, non-recombinant: gray).
+colored by CovRecomb classification (recombinant: purple, non-recombinant: gray).
 (A) Upper: jittered scatterplot comparing number of mutations
@@
 Lower: recombination nodes classified by number of mutations averted
-by allowing recombination (i.e. $x-y$).
+by allowing recombination (i.e. $x$-$y$).
 (B) Distribution of time to the MRCA of the parents of each QC-passing recombination node.

### 2025-10-16 latex/main.tex
Update for new 3SEQ sc2ts results

--- 2025-10-16 latex/main.tex
@@
 sequence trios (STAR Methods), 
-and 383 of the 855 (45\%) were classified as recombinant.
+and 484 of the 855 (57\%) were classified as recombinant.
 Validation status correlates strongly with our QC filter, with 
-88\% of the 354 QC-passing recombinants 
-and 14\% of the QC-failing recombinants
+95\% of the QC-passing recombinants 
+and 29\% of the QC-failing recombinants
 being classified as recombinant.
-The 12\% of QC-passing recombinants
-not validated by 3SEQ are also among the most marginally supported
-(see next section; Figure~\ref{fig:recombinant_evidence}).
 In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
@@
 over all 354 recombination events that pass QC.
+% The 5\% of QC-passing recombinants
+% not validated by 3SEQ are also among the most marginally supported
+% (see next section; Figure~\ref{fig:recombinant_evidence}).
 Also shown is the 3SEQ classification status of these recombinants.
@@
 \midrule
-3SEQ & 383 & 71 & 312 \\
-CovRecomb & 92& 90 & 2\\
-rebar & 213& 203 & 10 \\
+3SEQ        & 484 & 338     & 146 \\
+CovRecomb   & 92  & 90      & 2\\
+rebar       & 213 & 203     & 10 \\
 \bottomrule

### 2025-10-16 notebooks/nb_utils.py
Rerun for new 3SEQ results and remove XBB.x
Closes #905

@@ -58,10 +58,10 @@ def standard_recombinant_labels(ts, pango_x_events_file):
                 key=lambda i: tree.num_samples(pango_x_events.closest_recombinant[i])
             )
             use[sorted_indexes[-1]] = True
-    
+
     labels = {row.closest_recombinant: row.root_pango for row in pango_x_events[use].itertuples()}
     assert len(labels) == np.sum(use)
-    
+
     labels = {row.closest_recombinant: row.root_pango for row in pango_x_events[use].itertuples()}
     for re_node, rows in pango_x_events[pango_x_events.root_type != "R"].groupby("closest_recombinant"):
         if re_node >=0:
@@ -71,25 +71,27 @@ def standard_recombinant_labels(ts, pango_x_events_file):
                     labels[re_node] = rows.iloc[0].root_pango
                 else:
                     labels[re_node] = "/".join(sorted(rows.root_pango.values))
-    
+
     # Tweak the "XBB.1" RE node label, which should have a bespoke label because it's not
     # reflective of the majority of XBB.1 samples
     XBB_1 = [k for k, v in labels.items() if v == "XBB.1"][0]
     labels[XBB_1] = "XBB.x"
+    # Remove this from the list, as it's not interesting enough to point out.
+    del labels[XBB_1]
     Xx = [k for k, v in labels.items() if "XZ" in v][0]
     labels[Xx] = "Xx"
-    
+
     # Add particular pangos that are not X but are recombinants
-    
+
     #for re_pango in ["BQ.1.21"]:  # BQ.1.21 is not robust, so skip it
     #    potential_re_node = ts.first().mrca(*df.loc[df.pango == re_pango, "node_id"])
     #    assert ts.node(potential_re_node).flags & sc2ts.NODE_IS_RECOMBINANT
     #    labels[potential_re_node] = re_pango
-    
+
     # Add the Jackson recombinants
     XA = [k for k, v in labels.items() if v == "XA"][0]
     labels[XA] = "XA(JA)"
-    
+
     jackson_recombs = {
         "JB": "ERR5058070",
         "JC": "ERR5232711",
@@ -98,7 +100,7 @@ def standard_recombinant_labels(ts, pango_x_events_file):
         "J2": "ERR5304348", # MILK-103C712
         "J3": "ERR5238288", # QEUH-1067DEF
     }
-    
+
     for label, sample_id in jackson_recombs.items():
         u = df.loc[sample_id, "node_id"]
[…]

### 2025-10-16 latex/main.tex
Updates from Overleaf

@@
 (D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
-for XA, XW, XBB, and an
-undesignated recombination event under XBB, which we refer to as XBB.x.}
+for XA, XW and XBB.}
 \label{fig:recombinant_evidence}
@@
 over all 354 recombination events that pass QC.
-% The 5\% of QC-passing recombinants
-% not validated by 3SEQ are also among the most marginally supported
-% (see next section; Figure~\ref{fig:recombinant_evidence}).
 Also shown is the 3SEQ classification status of these recombinants.
-All of the 12\% of recombinants that do not pass 3SEQ validation are
-from the most weakly supported, with 4 or 5 averted mutations.
+The 5\% of recombinants that do not pass 3SEQ validation are
+among the most weakly supported, with 4 or 5 averted mutations.
+[TODO discuss B and C]
 Figures~\ref{fig:recombinant_evidence_cov_recomb} and

### 2025-10-16 latex/main.tex

@@
 subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
+Note that we are interested in the direct output of sc2ts and UShER+RIPPLES
+here, and do not perform QC filtering of either.
 Using the RIPPLES parameter $p=4$ (equivalent to $k=4$ used 
@@
 events are likely attributable to the finer details of the 
-differences in tree building approaches between UShER
-and sc2ts.
+differences in tree building approaches between UShER and sc2ts.
 
@@
 We also compared the recombination events
-output by RIPPLES with the Pango X lineages.
-At p=4, we found that only XX of the Pango X lineages were present and
-many well-characterized events such as XBB were not present.
+output by RIPPLES with the Pango X lineages
+(Table~\ref{tab:ripples_x_lineages}).
+At p=4, we found that only 9 of the Pango X lineages were present and
+with unequivocal recombinant origin for 6 (contrast with 
+16 Type I events in Table~\ref{tab:pango_x_lineages}).
 We therefore reran at p=3, resulting in 4111 events overall,
@@
 [More summary]
-
-[Paragraph summing things up. It seems like RIPPLES is doing its
-job, and the question is really whether UShER is giving 
-the correct signals for it to detect. Sc2ts detects more of the 
-well-known events. Needs careful explanation.].
+[Note somewhere here that XB and XAJ are found to be recombinants 
+at p=4, showing that UShER is doing better here for some reason]
+
+In terms of \emph{detecting} recombination events, 
+sc2ts and the combination of UShER and RIPPLES have much in common, 
+and it is likely that very similar results could be obtained if run
+in the same way. That is, if we constructed an UShER tree by 
+incrementally adding the same daily batches as used for sc2ts
+and examined any newly-added long branches each day
+using RIPPLES, we would likely get very similar results to sc2ts
+in terms recombinants detected.
+For a given matching operation, RIPPLES and the LS HMM 
+(under its current parameterisation) 
+are deterministically solving the same well-defined problem
+(albeit with radically different implementations) and they 
+must therefore produce similar answers. 
+Where sc2ts and UShER+RIPPLES diverge, however, is in terms 
+of how well such signals of recombination can be preserved
+over longer timescales. Continuing our thought experiment
+in which we run UShER and RIPPLES in daily batches, suppose
+that on a given day a branch with 4 mutations is identified 
+as being a recombinant. Unless this branch is specifically 
+``protected'' as representing a recombinant it may 
+be broken up by subsequent parsimony operations, thus 
+erasing the signal of recombination. This would appear
+to be the mostly likely explanation for the ``missing'' 
+X-lineages like XBB (detected by sc2ts at $k=4$ but 
+not by UShER+RIPPLES at $p=4$).
+
 See Document~\ref{sec:detailed_ripples} for more detailed analysis 
 of the UShER+RIPPLES results.
+[
 
@@
 \section*{DISCUSSION}
+
+[NOTE: the discussion needs to be reworked, and hasn't been updated
+to account for changes. See the new draft subsection on prevalence of recombination,
+though]
 
@@
 with the underlying genealogy inference method.
+
+[Same point for us with XB. We don't pick up XB because of a long 
+branch and we don't have the right parents present.]
 

### 2025-10-17 latex/main.tex

@@
 The mature and feature-rich tskit library has interfaces in Python, C and Rust,
-extensive documentation,
-and is the basis of a growing software ecosystem of 
-simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
-visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
-inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
-% More here? I'm sure there's others
-and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
-tools.
-
-Similarly to the Mutation Annotated Tree (MAT)~\cite{McBroome2021} format used 
-by UShER (RIPPLES~\cite{Turakhia2022} and matOptimize~\cite{DeMaio2023} are 
-included in the UShER software distribution), 
+extensive documentation, and is used in numerous downstream 
+applications~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,
+Tagami2024tstrait,Karthikeyan2025tsbrowse,Kitchens2025tskit,Kelleher2019,
+Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting,
+Fan2022,Nowbandegani2023extremely,Fritze2024forest}.
+
+Similarly to the Mutation Annotated Tree (MAT)~\cite{McBroome2021} format 
+(used 
+by UShER\cite{Turakhia2021}, matOptimize~\cite{DeMaio2023}, 
+RIPPLES~\cite{Turakhia2022} and 
+RIVET~\cite{Smith2023tracking}),
 tskit incorporates mutational information directly in the data 
-model. This is done using the site (defining the genome position and 
+model. 
+This is done using the site (defining the genome position and 
 ancestral state) and mutation (defining the site, node, and derived state)
@@
 the joint processes of mutation and recombination.
-It is widely used in a range of applications in population genetics,
-% Does Threads use LS?
-including ARG inference \cite{Kelleher2019,Gunnarsson2024}.
+It is widely used in a range of applications in population 
+genetics~\cite{mcvean2019linkage}.
 The LS model is an HMM that models a focal genome
-as a sequence of nucleotides (i.e. the observed states),
+as a sequence of nucleotides (the observed states),
 which are probabilistically emitted
 as an imperfect mosaic of a set of genomes
-in a reference panel (i.e., the hidden states).
+in a reference panel (the hidden states).
 The process of switching between the genomes in the reference panel
@@
 
-In the original formulation of the LS model
-(see Appendix S1 of \cite{LiStephens2003}),
+In the standard formulations of the LS model
+\cite{LiStephens2003,Donnelly2010-coalescent}
 the probability of switching
@@
 We call this parameter $k$ the ``recombination penalty''.
-% TODO mention that this can be thought of as extended parsimony.
 This parameter thus allows us to adjust
@@
 variation of mutation rate along the genome.
-As data sampling was dense for SARS-CoV-2,
-this assumption was considered tolerable.
+
+A Viterbi algorithm can be used
+to find the most likely ``copying path'' of a focal sample
+through a reference panel of samples under the LS model.
+Specifically, sc2ts uses a Viterbi algorithm modified to operate on STS,
+which was originally developed to build genome-wide human genealogies and
+is capable of scaling to biobank-scale datasets \cite{Kelleher2019}.
+Using this Viterbi algorithm,
+sc2ts deterministically infers the copying path of a sample
+collected on a certain date through the samples
+collected on earlier dates in a given ARG.
+
+The SARS-CoV-2 sequences were determined
+from samples collected densely in time.
+SARS-CoV-2 phylogenies have low phylogenetic signals
+due to high sampling frequency
+(such that sampling and transmission outpace nucleotide substitution).
+Therefore, most samples can be explained by a small number of mutations.
+Leveraging this property of the dataset,
+we developed a likelihood-thresholding strategy
+to avoid needing to compute the likelihood of
+all possible copying paths through an ARG for every single sample.
+%%% TODO: This needs to be explained.
+
+[TODO explain Fig S1 and S2]
+
+Furthermore, to avoid performing redundant HMM matching on identical samples, new samples
+which were exactly matched to a (sample or non-sample) node in the ARG were not added immediately.
+Instead, the copying paths of these samples were recorded,
+and the samples were eventually added back to the ARG in a post-processing step.
+
+\subsubsection*{Choice of k paramter}
+% This is a key decision that we will be referencing the discussion, so 
+% deserves its own section
 
@@
 to infer the final ARG described in this study.
-
-A Viterbi algorithm can be used
-to find the most likely ``copying path'' of a focal sample
-through a reference panel of samples under the LS model.
-Specifically, sc2ts uses a Viterbi algorithm modified to operate on STS,
-which was originally developed to build genome-wide human genealogies and
-is capable of scaling to biobank-scale datasets \cite{Kelleher2019}.
-Using this Viterbi algorithm,
-sc2ts deterministically infers the copying path of a sample
-collected on a certain date through the samples
-collected on earlier dates in a given ARG.
-
-The SARS-CoV-2 sequences were determined
-from samples collected densely in time.
-SARS-CoV-2 phylogenies have low phylogenetic signals
-due to high sampling frequency
-(such that sampling and transmission outpace nucleotide substitution).
-Therefore, most samples can be explained by a small number of mutations.
-Leveraging this property of the dataset,
-we developed a likelihood-thresholding strategy
-to avoid needing to compute the likelihood of
-all possible copying paths through an ARG for every single sample.
-%%% TODO: This needs to be explained.
-
-[TODO explain Fig S1 and S2]
-
-Furthermore, to avoid performing redundant HMM matching on identical samples, new samples
-which were exactly matched to a (sample or non-sample) node in the ARG were not added immediately.
-Instead, the copying paths of these samples were recorded,
-and the samples were eventually added back to the ARG in a post-processing step.
 

### 2025-10-20 latex/main.tex
Various updates suggested through ChatGPT discussions

@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%  Welcome to the Cell Press LaTeX template,
-%%%  version 1.10. This is a minimalist template
-%%%  to help you organize your article for
-%%%  publication at Cell Press. PLEASE NOTE:
-
-%%%  (1) If you submit your final manuscript materials
-%%%  in LaTeX format, our typesetters will prepare
-%%%  a Word file for use in production. This conversion
-%%%  process allows us to copyedit your paper, fix
-%%%  any typos, and add formatting and tagging. The
-%%%  conversion process will add approximately 3
-%%%  business days to the production timeline.
-%%%  Authors using LaTeX should keep this in mind
-%%%  when considering deadlines.
-
-%%%  (2) Keep your LaTeX files as simple as possible.
-%%%  Avoid the use of elaborate local macros and/or
-%%%  customized external style files. If you need
-%%%  additional macros, please keep them simple and
-%%%  include them in the actual .tex document preamble.
-%%%  Source code should be set up so that all .sty
-%%%  and .bst files called by the main .tex file are
-%%%  in the same directory as the main .tex file.
-
-%%%  (3) Cell Press publishes more than 40 journals,
-%%%  some of which may have different or additional
-%%%  formatting requirements not specified in this
-%%%  template. When revising your paper before
-%%%  acceptance, please review the formatting
-%%%  guidelines, including the Final Files Requirements,
-%%%  for the journal you are publishing with.
-
-%%%  Please send feedback on this template to lshipp@cell.com.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Notes on format:
+% The total word count of an article must be under 7,000 (including main figure
+% legends but excluding STAR Methods text, supplemental item legends, and
+% references section), and there should be no more than seven figures and/or
+% tables.
 
@@
 
+% \section*{DISCUSSION}
+
+% NOTES: paragraph structure. Lead with the biological insights first 
+% to prevent the ``this is a methods paper'' criticism
+
+% \begin{itemize}
+% \item We present a landmark ARG, which has a comprehensive representation
+% of evolutionary history.
+% \item This is a conservative lower-bound on the occurance of recombination.
+% Very high validation rate post-QC.
+% Substantially lower than previous estimate. 
+% The first order term of this 
+% is the choice of $p$ parameter. With p=4 we show that the ARG contains
+% more of the previously identified recombinants.
+% \item Truth data for recombination is difficult to find. The closest we have is the 
+% pango X lineages. No methods agrees with all of the designations being 
+% true recombinants. Our analysis suggests that there is significant over-naming
+% with the same underlying recombination event getting associated with 
+% multiple X lienages at least twice. Also many lineages are not suggested 
+% as being recombinant. This resolves long-standing complexity in 
+% the Omicron evolution and simplifies our understanding.
+% \item Based on this cleaned up and conservative estimate of recombination
+% prevalance. Supports the idea that  recombination frequency in SARS-CoV-2 
+% is driven jointly by viral
+% load in the population and lineage diversity, implying predictable
+% recombination ``hot periods'' during pandemics
+% \item Also found evidence that breakpoints are enriched near spike, 
+% supporting findings of earlier studies.
+% Suggests functional constraints or template-switching
+% biases that may shape SARS-CoV-2 evolution.
+% Spike gene junctions are recurrent recombination zones, likely under both
+% mechanistic and selective pressures.
+% \item This ARG is base on a new method, sc2ts. Major methodological 
+% breakthrough in terms of handling recombination. We validate the method
+% by using results from the literature. We don't use simulations because 
+% there's no real point in this case. We're just focusing on SARS-CoV-2.
+% We could create some simulations and generate some metrics about how
+% well sc2ts picks up simulated recombination events on clean data, 
+% but what would that tell us ultimately? (Similarly for tree reconstruction.)
+% Certainly standardised simulations would help to refine methods, but
+% we have moved the bar forwards massively here and made a great starting
+% point for subsequent method development.
+% \item There is great potential for using this ARG in future studies,
+% and using sc2ts as a baseline to judge future methods.
+% \item The tskit and VCF Zarr toolchains enable this research.
+% \item We have shown that recombination can be handled, provided
+% feasible methods to infer it, and shown that a preexisting tool
+% chain can provide quick insignts. The future is bright.
+
+% \end{itemize}
+
 \section*{DISCUSSION}
 
-[NOTE: the discussion needs to be reworked, and hasn't been updated
-to account for changes. See the new draft subsection on prevalence of recombination,
-though]
+We present here the first pandemic-scale Ancestral Recombination Graph (ARG)
+for SARS-CoV-2, reconstructed from 2.48 million genomes. This resource provides
+a comprehensive, recombination-aware view of the evolutionary history of the
+virus, capturing not only point mutations but also deletions and recombination
+events in a single coherent framework. By representing SARS-CoV-2 evolution as
+a graph rather than a tree, the ARG unifies the relationships among lineages
+that could not previously be reconciled by tree-based phylogenies. The
+resulting structure serves as a long-term reference for studies of viral
+evolution and as a foundation for future recombination-aware surveillance.
+
+Our inference of recombination represents a conservative lower bound on the
+true frequency of recombination in SARS-CoV-2. This conservatism arises from
+stringent quality control criteria, including the requirement for multiple
+supporting loci for each parental assignment. Even under these strict
+thresholds, the ARG contains hundreds of high-confidence recombination events
+with a validation rate exceeding 90\% among events passing QC filters. The
+overall prevalence of recombination inferred here is substantially lower than
+early estimates based on smaller or less curated datasets, but this is
+consistent with our aim to prioritise specificity over sensitivity.
+Importantly, at the default parsimony penalty ($p = 4$), the ARG recovers the
+majority of previously recognised recombinant lineages, including those most
+relevant to global transmission dynamics.
+
+Comparisons with the Pango X designations reveal that several nominally
+distinct recombinant lineages share common ancestral recombination events. This
+over-splitting has led to an inflated perception of recombinant diversity.
+Within Omicron, for example, multiple X lineages are better explained as
+descendants of a single ancestral recombinant rather than as independent
+crossover events. Resolving these redundancies clarifies long-standing
+complexity in the Omicron genealogy and simplifies the interpretation of
+SARS-CoV-2’s reticulate evolution. Conversely, several designated X lineages
+show little evidence of recombination in the ARG, emphasising that not all
+phylogenetically unusual lineages are true recombinants. Together, these
+results provide a more parsimonious and internally consistent model of
+recombinant evolution.
+
+Using the dated ARG, we examined the temporal and epidemiological correlates of
+recombination. The frequency of inferred recombination events correlates
+strongly with both global case numbers and the diversity of circulating
+lineages. This pattern supports the hypothesis that recombination is driven
+jointly by viral load in the population, which increases opportunities for
+co-infection, and by lineage diversity, which provides distinct parental
+genomes. These findings imply predictable ``hot periods'' for recombinant
+emergence during pandemics, particularly when multiple antigenically divergent
+variants co-circulate at high prevalence. Understanding these dynamics may aid
+future outbreak forecasting and genomic surveillance strategies.
+
+Analysis of genomic breakpoint locations revealed enrichment of recombination
+near the boundaries of the Spike gene, consistent with previous observations
+from smaller datasets. The repeated occurrence of breakpoints in these regions
+suggests that structural or functional constraints may influence where viable
+recombinants can arise. Mechanistic biases in template switching, combined with
+selective pressures on Spike for receptor binding and immune escape, may
+together shape the landscape of recombination. The ARG thus not only captures
+recombination events but also points to underlying biological mechanisms worthy
+of experimental follow-up.
+
+Beyond its biological insights, this study represents a methodological advance
+in real-time, recombination-aware phylogenetic inference. The \texttt{sc2ts}
+method integrates an efficient, simplified Li--Stephens model with incremental
+sample addition, allowing scalable ARG construction from millions of genomes.
+We validated the method against empirical data and literature-curated
+recombinant lineages rather than simulated datasets, because the global
+SARS-CoV-2 collection offers an unparalleled real-world benchmark.
+Nevertheless, standardised simulations will remain valuable for refining models
+and comparing future implementations. Our results demonstrate that the
+\texttt{sc2ts} framework captures the essential genealogical structure of
+SARS-CoV-2 while accommodating recombination in a computationally tractable
+way.
+
+The ARG produced here is embedded within the \texttt{tskit} and
+\texttt{VCF-Zarr} ecosystems, enabling compact storage, rapid analysis, and
+open sharing. This integration ensures that the resource is immediately usable
+by the broader community for tasks such as mapping selective sweeps, tracing
+recombinant origins, and evaluating lineage dynamics. The dataset and
+accompanying workflows provide a transparent foundation for benchmarking future
+methods and for extending ARG-based analyses to other pathogens.
+
+In summary, this work demonstrates that recombination in SARS-CoV-2 can be
+robustly inferred and represented at scale, transforming how we study the
+evolution of rapidly diversifying viruses. The ARG provides a unified,
+validated view of SARS-CoV-2 evolution that resolves the relationships among
+recombinant lineages, links recombination frequency to epidemiological
+dynamics, and highlights mechanistic constraints on recombination. More
+broadly, it establishes that scalable ARG inference is now a practical and
+powerful tool for pandemic genomics.
+
+\subsection*{OUTLOOK}
+
+The pandemic-scale ARG presented here establishes a new foundation for studying
+viral evolution in real time. Beyond SARS-CoV-2, the \texttt{sc2ts} framework
+can be directly adapted to other recombining pathogens such as influenza, HIV,
+and rotaviruses, where complex genealogies have long limited evolutionary
+analysis. Integrating the ARG with epidemiological and immunological data will
+enable quantitative studies of how recombination interacts with selection,
+immune escape, and transmission dynamics. As sequencing becomes increasingly
+routine in global surveillance, efficient ARG inference will allow real-time
+tracking of recombinant emergence and fitness, supporting faster risk
+assessment and lineage designation. The combination of scalable inference, open
+data standards, and community tools transforms recombination from a nuisance
+factor into an analyzable dimension of viral evolution. We anticipate that this
+resource will not only clarify the history of SARS-CoV-2 but also guide the
+design of next-generation computational and experimental approaches to
+understanding recombination-driven adaptation in RNA viruses.
+
+% [NOTE: the discussion needs to be reworked, and hasn't been updated
+% to account for changes. See the new draft subsection on prevalence of recombination,
+% though]
 
@@
 
-We present sc2ts, a real-time ARG inference method,
-and applied it to build a $\sim$2.48 million-sample pandemic-scale ARG of SARS-CoV-2.
-By identifying specific samples as parents,
-sc2ts enables more precise interpretation of recombinants
-than state-of-the-art recombination detection methods,
-which find recombinants by matching a given sample
-to pre-defined population-averaged mutational profiles of Pango lineages.
-Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution,
-in contrast to RIPPLES \cite{Turakhia2022}
-which detects recombinants post hoc by searching for unusual branches in a given phylogeny,
-inevitably resulting in an incomprehensive representation of SARS-CoV-2 evolution.
-Combined with new methods and tools to evaluate evidence for recombination,
-sc2ts is a significant technological advance over current methods
-for the discovery and analysis of recombinants and
-efficient inference of SARS-CoV-2 genealogy.
-
-The ARG itself is a useful resource to support future studies of SARS-CoV-2.
-Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
-which are shown in different studies
-using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
-The ARG displays remarkable phylogenetic congruence
-with a state-of-the-art phylogeny built using UShER.
-Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
-allowing for convenient and efficient reuse of this valuable dataset
-for a variety of evolutionary and genetic analyses beyond those presented here.
-% We hope that the ARG will spur on new research areas,
-% in which recombination is naturally and fully accounted for,
-% and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
-% has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
-So far, recent developments in ARG methods have focused almost 
-exclusively on applications to human sequencing data.
-In that space, the availability of ARGs reconstructed from large datasets
-has already had tremendous impact on improving our understanding of
-human genetic ancestry and past demography \cite{Wohns2022}.
-We expect similar leaps to be enabled
-through tailoring ARG reconstruction methods
-to data from other species and pathogens.
-Specifically, existing inference methods can be adapted
-to use the sc2ts ARG for the genealogy-based study of selection \cite{hejase2022deep}
-or the geographic spread of the virus through time \cite{Grundler2025geographic}.
-This is in addition to simpler analyses,
-such as calculating mutation spectra
-or querying the number or frequency of mutations at a given position,
-which the ARG makes significantly more efficient, and
-ultimately more easily reproducible.
-Since the ARG is a condensed representation of the genetic information
-contained in the sequencing data, it can also be used to understand,
-for instance, how the variability in sequencing efforts across time 
-and location affects our ability to reconstruct the origin of 
-important lineages and detect new recombinants.
-Since reconstructing the ARG is computationally expensive, 
-and checking and improving various aspects of its quality is laborious, 
-we hope that making this resource openly available will save significant time
-and effort for other researchers and enable immediate focus on analyzing
-the information contained within this data. In-depth analysis of the
-ARG is also incredibly useful for identifying important directions for
-further research, both regarding methods (where current models do not
-correctly capture features of the data) and data processing 
-(where gaps in sequencing or data artifacts result in clear
-reconstruction errors).
-
-We contribute several novel methods and concepts
-which may find applications beyond sc2ts.
-First, we devise a simplified LS model that extends the classic concept of parsimony.
-In sc2ts, this model allows for accelerated HMM likelihood calculations,
-enabling reconstruction of the pandemic-scale ARG presented here.
-Existing HMMs used in statistical genetics may be modified
-in a similar way for significant speedup.
-Second, we propose the concept of mutations averted by recombination,
-which leads to an interpretable quantity to weigh the support for 
-the hypothesis that a given sample is a recombinant against
-the hypothesis that it is a non-recombinant,
-considering the sampled genetic diversity in an ARG and a specific HMM.
-Third, we propose the concept of HMM cost
-to identify poorly explained samples
-which are either too divergent or
-have a high number of mutations inconsistent with their collection dates.
-Here we use HMM cost to reject time-traveling samples,
-which no existing phylogenetic method detects and handles during inference
-(e.g., in the Viridian UShER phylogeny, parent and child nodes
-do not necessarily follow time order \cite{Hunt2024}),
-resulting in a cleaner dataset without samples
-with wrong collection dates that may affect downstream analyses.
-Future work may be to estimate dates
-for the suspected time traveling samples using the ARG provided here and
-the sc2ts sample matching engine.
-
-Recombination is an important evolutionary force in viruses \cite{Perez-Losada2015},
-but has been understudied in part
-due to the lack of suitable data models, methods, and tools \cite{Neches2020}.
-While the methods and tools developed here are tailored for SARS-CoV-2,
-they can be extended and adapted for other viruses.
-A major advantage of the sc2ts methods and tools is that
-they are built on top of the mature software libraries of tskit and VCF-Zarr,
-which enable pandemic-scale analyses that can be performed on typical laptops.
-Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem,
-making them highly adaptable for a broad range of applications.
-Together with the pandemic-scale ARG,
-we hope that sc2ts will be highly accessible to many researchers,
-enabling and accelerating studies of SARS-CoV-2 for years to come.
-
-\subsection*{Prevalance of recombination in SARS-CoV-2}
-[Rough first draft taken from earlier location. Sticking in a heading to
-make it clear it's distinct from other paragraphs.]
-
-The number of recombination events that we detect is largely 
-driven by the choice of recombination penalty, $k$ (or $p$
-for RIPPLES). 
-At $k=4$ we detect fewer recombinants overall than the combination of 
-UShER and RIPPLES at the same threshold. 
-However, most of the extra recombinants detected by UShER+RIPPLES
-involve violations of time ordering; whether this is due to incorrect
-metadata or these are false positives would require further study.
-However, at $p=4$, UShER+RIPPLES detects very few of the well-characterised
-Pango X-lineages which are detected by sc2ts. At the more lenient
-threshold of $p=3$, more of these events are detected (but still fewer
-than sc2ts) but at the cost of nearly 4 times more recombination
-events overall. Our detailed analysis shows that RIPPLES has a well
-defined task, which it performs well, finding the potential events
-in an exhaustive search. However, it can only operate within the
-input that it is provided, and it is the UShER tree provided with
-the Viridian data that does not have clear signals for recombination
-of X lineages such as XBB at the $p=4$ threshold. 
-The inference of recombination is therefore intrinsically linked
-with the underlying genealogy inference method.
-
-[Same point for us with XB. We don't pick up XB because of a long 
-branch and we don't have the right parents present.]
-
-[Whether k=3 or k=4 is ``correct'' is a matter for debate, and feeds
-into the next point.]
-
-The estimate for effective recombination in SARS-CoV-2 that we obtain
-at k=4 is much lower 
-than previous studies have suggbalanced
-% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
-% recombinant origins out of the first 87,695 genomes to be released
-In an early analysis, 225 potentially recombinant sequences out 
-of 87,695 samples collected before October 2020\cite{Varabyou2021rapid}.
-Over the same time period, XX events are present in the sc2ts ARG,
-representing a total of XX samples.
-In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
-589 unique recombination events in a tree of 1.6 million samples
-up to May 2021 were identified after extensive quality control.
-These nodes had a combined total of 43,104 descendants, suggesting
-that approximately 2.7\% of genomes were of recombinant origin.
-Over the same time period, the sc2ts ARG contains 14 recombinants
-passing QC with a total of 172 descendants. The ARG contains a total
-501,235 samples over this time period, giving an overall total fraction of
-0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
-events and 79-fold descendant samples over the same time period.
-The numbers are not directly comparable as the published RIPPLES 
-estimate is based on GISAID data which is enriched for
-artefacts~\cite{Hunt2024}, but given the observations in 
-the previous section on running RIPPLES on the Viridian data
-it seems likely that RIPPLES is overestimating the prevalence
-of recombination in SARS-CoV-2.
+% We present sc2ts, a real-time ARG inference method,
+% and applied it to build a $\sim$2.48 million-sample pandemic-scale ARG of SARS-CoV-2.
+% By identifying specific samples as parents,
+% sc2ts enables more precise interpretation of recombinants
+% than state-of-the-art recombination detection methods,
+% which find recombinants by matching a given sample
+% to pre-defined population-averaged mutational profiles of Pango lineages.
+% Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution,
+% in contrast to RIPPLES \cite{Turakhia2022}
+% which detects recombinants post hoc by searching for unusual branches in a given phylogeny,
+% inevitably resulting in an incomprehensive representation of SARS-CoV-2 evolution.
+% Combined with new methods and tools to evaluate evidence for recombination,
+% sc2ts is a significant technological advance over current methods
+% for the discovery and analysis of recombinants and
+% efficient inference of SARS-CoV-2 genealogy.
+
+% The ARG itself is a useful resource to support future studies of SARS-CoV-2.
+% Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
+% which are shown in different studies
+% using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
+% The ARG displays remarkable phylogenetic congruence
+% with a state-of-the-art phylogeny built using UShER.
+% Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
+% allowing for convenient and efficient reuse of this valuable dataset
+% for a variety of evolutionary and genetic analyses beyond those presented here.
+% % We hope that the ARG will spur on new research areas,
+% % in which recombination is naturally and fully accounted for,
+% % and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
+% % has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
+% So far, recent developments in ARG methods have focused almost 
+% exclusively on applications to human sequencing data.
+% In that space, the availability of ARGs reconstructed from large datasets
+% has already had tremendous impact on improving our understanding of
+% human genetic ancestry and past demography \cite{Wohns2022}.
+% We expect similar leaps to be enabled
+% through tailoring ARG reconstruction methods
+% to data from other species and pathogens.
+% Specifically, existing inference methods can be adapted
+% to use the sc2ts ARG for the genealogy-based study of selection \cite{hejase2022deep}
+% or the geographic spread of the virus through time \cite{Grundler2025geographic}.
+% This is in addition to simpler analyses,
+% such as calculating mutation spectra
+% or querying the number or frequency of mutations at a given position,
+% which the ARG makes significantly more efficient, and
+% ultimately more easily reproducible.
+% Since the ARG is a condensed representation of the genetic information
+% contained in the sequencing data, it can also be used to understand,
+% for instance, how the variability in sequencing efforts across time 
+% and location affects our ability to reconstruct the origin of 
+% important lineages and detect new recombinants.
+% Since reconstructing the ARG is computationally expensive, 
+% and checking and improving various aspects of its quality is laborious, 
+% we hope that making this resource openly available will save significant time
+% and effort for other researchers and enable immediate focus on analyzing
+% the information contained within this data. In-depth analysis of the
+% ARG is also incredibly useful for identifying important directions for
+% further research, both regarding methods (where current models do not
+% correctly capture features of the data) and data processing 
+% (where gaps in sequencing or data artifacts result in clear
+% reconstruction errors).
+
+% We contribute several novel methods and concepts
+% which may find applications beyond sc2ts.
+% First, we devise a simplified LS model that extends the classic concept of parsimony.
+% In sc2ts, this model allows for accelerated HMM likelihood calculations,
+% enabling reconstruction of the pandemic-scale ARG presented here.
+% Existing HMMs used in statistical genetics may be modified
+% in a similar way for significant speedup.
+% Second, we propose the concept of mutations averted by recombination,
+% which leads to an interpretable quantity to weigh the support for 
+% the hypothesis that a given sample is a recombinant against
+% the hypothesis that it is a non-recombinant,
+% considering the sampled genetic diversity in an ARG and a specific HMM.
+% Third, we propose the concept of HMM cost
+% to identify poorly explained samples
+% which are either too divergent or
+% have a high number of mutations inconsistent with their collection dates.
+% Here we use HMM cost to reject time-traveling samples,
+% which no existing phylogenetic method detects and handles during inference
+% (e.g., in the Viridian UShER phylogeny, parent and child nodes
+% do not necessarily follow time order \cite{Hunt2024}),
+% resulting in a cleaner dataset without samples
+% with wrong collection dates that may affect downstream analyses.
+% Future work may be to estimate dates
+% for the suspected time traveling samples using the ARG provided here and
+% the sc2ts sample matching engine.
+
+% Recombination is an important evolutionary force in viruses \cite{Perez-Losada2015},
+% but has been understudied in part
+% due to the lack of suitable data models, methods, and tools \cite{Neches2020}.
+% While the methods and tools developed here are tailored for SARS-CoV-2,
+% they can be extended and adapted for other viruses.
+% A major advantage of the sc2ts methods and tools is that
+% they are built on top of the mature software libraries of tskit and VCF-Zarr,
+% which enable pandemic-scale analyses that can be performed on typical laptops.
+% Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem,
+% making them highly adaptable for a broad range of applications.
+% Together with the pandemic-scale ARG,
+% we hope that sc2ts will be highly accessible to many researchers,
+% enabling and accelerating studies of SARS-CoV-2 for years to come.
+
+% \subsection*{Prevalance of recombination in SARS-CoV-2}
+% [Rough first draft taken from earlier location. Sticking in a heading to
+% make it clear it's distinct from other paragraphs.]
+
+% The number of recombination events that we detect is largely 
+% driven by the choice of recombination penalty, $k$ (or $p$
+% for RIPPLES). 
+% At $k=4$ we detect fewer recombinants overall than the combination of 
+% UShER and RIPPLES at the same threshold. 
+% However, most of the extra recombinants detected by UShER+RIPPLES
+% involve violations of time ordering; whether this is due to incorrect
+% metadata or these are false positives would require further study.
+% However, at $p=4$, UShER+RIPPLES detects very few of the well-characterised
+% Pango X-lineages which are detected by sc2ts. At the more lenient
+% threshold of $p=3$, more of these events are detected (but still fewer
+% than sc2ts) but at the cost of nearly 4 times more recombination
+% events overall. Our detailed analysis shows that RIPPLES has a well
+% defined task, which it performs well, finding the potential events
+% in an exhaustive search. However, it can only operate within the
+% input that it is provided, and it is the UShER tree provided with
+% the Viridian data that does not have clear signals for recombination
+% of X lineages such as XBB at the $p=4$ threshold. 
+% The inference of recombination is therefore intrinsically linked
+% with the underlying genealogy inference method.
+
+% [Same point for us with XB. We don't pick up XB because of a long 
+% branch and we don't have the right parents present.]
+
+% [Whether k=3 or k=4 is ``correct'' is a matter for debate, and feeds
+% into the next point.]
+
+% The estimate for effective recombination in SARS-CoV-2 that we obtain
+% at k=4 is much lower 
+% than previous studies have suggbalanced
+% % Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
+% % recombinant origins out of the first 87,695 genomes to be released
+% In an early analysis, 225 potentially recombinant sequences out 
+% of 87,695 samples collected before October 2020\cite{Varabyou2021rapid}.
+% Over the same time period, XX events are present in the sc2ts ARG,
+% representing a total of XX samples.
+% In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
+% 589 unique recombination events in a tree of 1.6 million samples
+% up to May 2021 were identified after extensive quality control.
+% These nodes had a combined total of 43,104 descendants, suggesting
+% that approximately 2.7\% of genomes were of recombinant origin.
+% Over the same time period, the sc2ts ARG contains 14 recombinants
+% passing QC with a total of 172 descendants. The ARG contains a total
+% 501,235 samples over this time period, giving an overall total fraction of
+% 0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
+% events and 79-fold descendant samples over the same time period.
+% The numbers are not directly comparable as the published RIPPLES 
+% estimate is based on GISAID data which is enriched for
+% artefacts~\cite{Hunt2024}, but given the observations in 
+% the previous section on running RIPPLES on the Viridian data
+% it seems likely that RIPPLES is overestimating the prevalence
+% of recombination in SARS-CoV-2.
 

### 2025-10-20 latex/main.tex
Draft of second paragraph

@@
 of the UShER+RIPPLES results.
-[
-
-% TODO convert this into a discussion pargraph..
-% \subsection*{Efficient inference and analysis of a pandemic-scale ARG}
-
-% The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
-% which overwhelmed existing computational approaches,
-% motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
-% Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
-% the functionality required largely overlaps.
-% Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
-% and the VCF Zarr data storage format \cite{Czech2025}
-% to describe relationships among virus samples,
-% building an ARG over time as data becomes available.
-% By reusing high-quality and high-performance components of these toolkits,
-% we could develop the sc2ts inference method itself and
-% perform all the pandemic-scale analyses in this article
-% with Python code alone and no compiled languages.
-
-% To illustrate the power of these libraries, we provide some examples
-% in which we compare the Usher and sc2ts intersections in tskit format described earlier.
-% First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments.
-% The total aligned dataset for 1,226,587 samples and 27,431 sites
-% shared by the sc2ts and Usher tree sequences represents 31.34 GB of nucleotide calls.
-% Of the 80,574,518 missing data calls (Ns) in the alignments,
-% sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%).
-% Additionally, 413,303 calls made use of the IUPAC uncertainty codes.
-% Of these sc2ts imputed only 77 (0.02\%) incorrectly
-% (i.e., with a base that is not compatible with the ambiguity code).
-% UShER imputed only 4 calls from this set incorrectly,
-% reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations.
-% This computation,
-% performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
-% required 30 lines of Python code and took 2m 59s on an intel i7-9700 CPU,
-% with a peak memory usage of about 4.5 GB
-% (see \cite{Czech2025} for further benchmarks for
-% processing the Viridian data in VCF Zarr format).
-% Performing this analysis using matUtils would require
-% exporting the full dataset to VCF, and
-% converting the VCF to FASTA, and
-% then comparing the alignments one-by-one to the source FASTA alignments.
-% In contrast with matUtils and most phylogenetic software,
-% which provide a command line interface and suite of predefined operations that typically output text files,
-% tskit is a Python API (with C and Rust interfaces also supported)
-% which has many advantages \cite{Baumdicker2022}.
-% As an illustration of the open-ended and efficient nature of the analyses this enables,
-% we computed the number of polytomies (nodes that have $> 2$ children)
-% and the maximum number of children per node in the Usher and sc2ts outputs.
-% In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148)
-% and compute the maximum number of children (5918).
-% In the sc2ts ARG, it took 385 ms and 6 lines of code
-% to find the maximum number of polytomies (137,001) and
-% maximum number of children (7708) across all 348 local trees.
 
@@
 
-We present here the first pandemic-scale Ancestral Recombination Graph (ARG)
-for SARS-CoV-2, reconstructed from 2.48 million genomes. This resource provides
-a comprehensive, recombination-aware view of the evolutionary history of the
-virus, capturing not only point mutations but also deletions and recombination
-events in a single coherent framework. By representing SARS-CoV-2 evolution as
-a graph rather than a tree, the ARG unifies the relationships among lineages
-that could not previously be reconciled by tree-based phylogenies. The
-resulting structure serves as a long-term reference for studies of viral
-evolution and as a foundation for future recombination-aware surveillance.
-
-Our inference of recombination represents a conservative lower bound on the
-true frequency of recombination in SARS-CoV-2. This conservatism arises from
-stringent quality control criteria, including the requirement for multiple
-supporting loci for each parental assignment. Even under these strict
-thresholds, the ARG contains hundreds of high-confidence recombination events
-with a validation rate exceeding 90\% among events passing QC filters. The
-overall prevalence of recombination inferred here is substantially lower than
-early estimates based on smaller or less curated datasets, but this is
-consistent with our aim to prioritise specificity over sensitivity.
-Importantly, at the default parsimony penalty ($p = 4$), the ARG recovers the
-majority of previously recognised recombinant lineages, including those most
-relevant to global transmission dynamics.
-
-Comparisons with the Pango X designations reveal that several nominally
-distinct recombinant lineages share common ancestral recombination events. This
-over-splitting has led to an inflated perception of recombinant diversity.
-Within Omicron, for example, multiple X lineages are better explained as
-descendants of a single ancestral recombinant rather than as independent
-crossover events. Resolving these redundancies clarifies long-standing
-complexity in the Omicron genealogy and simplifies the interpretation of
-SARS-CoV-2’s reticulate evolution. Conversely, several designated X lineages
-show little evidence of recombination in the ARG, emphasising that not all
-phylogenetically unusual lineages are true recombinants. Together, these
-results provide a more parsimonious and internally consistent model of
-recombinant evolution.
-
-Using the dated ARG, we examined the temporal and epidemiological correlates of
-recombination. The frequency of inferred recombination events correlates
-strongly with both global case numbers and the diversity of circulating
-lineages. This pattern supports the hypothesis that recombination is driven
-jointly by viral load in the population, which increases opportunities for
-co-infection, and by lineage diversity, which provides distinct parental
-genomes. These findings imply predictable ``hot periods'' for recombinant
-emergence during pandemics, particularly when multiple antigenically divergent
-variants co-circulate at high prevalence. Understanding these dynamics may aid
-future outbreak forecasting and genomic surveillance strategies.
-
-Analysis of genomic breakpoint locations revealed enrichment of recombination
-near the boundaries of the Spike gene, consistent with previous observations
-from smaller datasets. The repeated occurrence of breakpoints in these regions
-suggests that structural or functional constraints may influence where viable
-recombinants can arise. Mechanistic biases in template switching, combined with
-selective pressures on Spike for receptor binding and immune escape, may
-together shape the landscape of recombination. The ARG thus not only captures
-recombination events but also points to underlying biological mechanisms worthy
-of experimental follow-up.
-
-Beyond its biological insights, this study represents a methodological advance
-in real-time, recombination-aware phylogenetic inference. The \texttt{sc2ts}
-method integrates an efficient, simplified Li--Stephens model with incremental
-sample addition, allowing scalable ARG construction from millions of genomes.
-We validated the method against empirical data and literature-curated
-recombinant lineages rather than simulated datasets, because the global
-SARS-CoV-2 collection offers an unparalleled real-world benchmark.
-Nevertheless, standardised simulations will remain valuable for refining models
-and comparing future implementations. Our results demonstrate that the
-\texttt{sc2ts} framework captures the essential genealogical structure of
-SARS-CoV-2 while accommodating recombination in a computationally tractable
-way.
-
-The ARG produced here is embedded within the \texttt{tskit} and
-\texttt{VCF-Zarr} ecosystems, enabling compact storage, rapid analysis, and
-open sharing. This integration ensures that the resource is immediately usable
-by the broader community for tasks such as mapping selective sweeps, tracing
-recombinant origins, and evaluating lineage dynamics. The dataset and
-accompanying workflows provide a transparent foundation for benchmarking future
-methods and for extending ARG-based analyses to other pathogens.
-
-In summary, this work demonstrates that recombination in SARS-CoV-2 can be
-robustly inferred and represented at scale, transforming how we study the
-evolution of rapidly diversifying viruses. The ARG provides a unified,
-validated view of SARS-CoV-2 evolution that resolves the relationships among
-recombinant lineages, links recombination frequency to epidemiological
-dynamics, and highlights mechanistic constraints on recombination. More
-broadly, it establishes that scalable ARG inference is now a practical and
-powerful tool for pandemic genomics.
-
-\subsection*{OUTLOOK}
-
-The pandemic-scale ARG presented here establishes a new foundation for studying
-viral evolution in real time. Beyond SARS-CoV-2, the \texttt{sc2ts} framework
-can be directly adapted to other recombining pathogens such as influenza, HIV,
-and rotaviruses, where complex genealogies have long limited evolutionary
-analysis. Integrating the ARG with epidemiological and immunological data will
-enable quantitative studies of how recombination interacts with selection,
-immune escape, and transmission dynamics. As sequencing becomes increasingly
-routine in global surveillance, efficient ARG inference will allow real-time
-tracking of recombinant emergence and fitness, supporting faster risk
-assessment and lineage designation. The combination of scalable inference, open
-data standards, and community tools transforms recombination from a nuisance
-factor into an analyzable dimension of viral evolution. We anticipate that this
-resource will not only clarify the history of SARS-CoV-2 but also guide the
-design of next-generation computational and experimental approaches to
-understanding recombination-driven adaptation in RNA viruses.
+% P1. The ARG. Unique resource. Emphasise utility of UShER and how 
+% this has more.
+
+We present the first pandemic-scale ancestral recombination graph (ARG) for
+SARS-CoV-2, reconstructed from 2.48 million genomes using \texttt{sc2ts}. This
+ARG represents a step change in how the evolutionary history of this virus can
+be represented and analysed. Like the UShER phylogeny, it provides a
+comprehensive and continuously updatable record of global genomic diversity,
+but it extends this paradigm by embedding recombination directly into the
+genealogy. The ARG unifies point mutations, deletions, and recombination events
+within a single, compact data structure, enabling analyses that
+were previously fragmented across multiple methods and data models. By encoding
+reticulate ancestry explicitly rather than assuming a single tree, the ARG
+provides a complete and internally consistent representation of SARS-CoV-2
+evolution. In this way, the
+ARG offers the same accessibility and scalability that made the UShER tree
+indispensable during the pandemic, while expanding its scope to encompass the
+full spectrum of evolutionary processes shaping SARS-CoV-2.
+
+% P2. This is a conservative lower-bound on recombination. 
+% Discuss other estimates.
+
+The recombination frequency estimated from the \texttt{sc2ts} ARG is
+substantially lower than that implied by the combination of UShER+RIPPLES,
+largely reflecting differences in model formulation and parameter choice.
+RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
+reported 589 unique recombination events with 43,104 descendant
+samples---corresponding to roughly 2.7\% of all genomes being of recombinant
+origin.  Using the same time interval, our inference at \(k=4\) identifies only
+14 high-confidence recombination events with 172 descendants among 501,235
+genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
+explained by the stricter parsimony penalty adopted here: increasing the
+penalty from \(p=3\) (as used in the RIPPLES analysis) 
+to \(k=4\) greatly reduces false
+positives while retaining nearly all verified recombinant lineages, 
+including those missed by RIPPLES at equivalent thresholds.  
+We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
+of recombination as
+tree rearrangements proceed, thereby requiring the more lenient 
+threshold in order to find known recombinats.
+These comparisons suggest that our estimate represents a conservative 
+lower bound. Further work will be required to determine
+the true frequency of recombination in SARS-CoV-2 evolution.
+
+% P3. Clarifies relationships between pango X lineages, simplifying
+% the Omicron sublineage.
+
+% P4. Results on global case numbers and frequency of recomb.
+
+% P5. Confirm breakpoints around Spike. suggests further study.
+
+% P6. Sc2ts method is a major advance. We validate with real data
+% rather than simulations. Forward looking statement about simulations.
+
+% P7. The software ecosystem enables diverse besplike analysis.
+
+% P8. Outlook
+
+\begin{itemize}
+\item We present a pandemic-scale ancestral recombination graph (ARG) that
+provides the first comprehensive, recombination-aware reconstruction of
+SARS-CoV-2 evolutionary history. This ARG unifies mutational, deletional,
+and recombinational processes within a single coherent framework and
+serves as a long-term reference for viral evolution.
+
+\item Our estimate of recombination frequency represents a conservative
+lower bound, reflecting stringent quality control and a high validation
+rate of detected events. Even at this conservative threshold, sc2ts
+recovers the majority of previously recognized recombinant lineages,
+indicating that recombination has been a recurrent yet tractable feature
+of SARS-CoV-2 evolution.
+
+\item By comparing to the Pango X designations, we show that several
+recombinant lineages share common ancestral recombination events,
+revealing systematic over-splitting of designations and clarifying the
+relationships among Omicron sublineages. This synthesis simplifies the
+previously complex picture of Omicron evolution and supports a model of
+hierarchical recombination.
+
+\item Using this cleaned and validated ARG, we show that the frequency of
+recombination events is correlated with both global case numbers and viral
+lineage diversity. These results suggest that recombination frequency is
+driven jointly by viral load and co-circulating diversity, leading to
+predictable ``hot periods'' during large outbreaks.
+
+\item Breakpoints are enriched near the boundaries of the Spike gene,
+consistent with previous reports and suggesting that structural and
+functional constraints shape where viable recombinants arise. These
+recurrent junctions point to mechanistic or selective pressures on Spike
+that warrant further experimental study.
+
+\item The sc2ts method represents a methodological advance that enables
+real-time, recombination-aware genealogy inference at pandemic scale. We
+validated its performance against empirical data and literature-curated
+events rather than simulations, because the global SARS-CoV-2 dataset
+provides a uniquely comprehensive real-world benchmark. Nonetheless,
+standardized simulations will be valuable for future methodological
+refinement.
+
+\item The resulting ARG, built within the tskit and VCF-Zarr ecosystems,
+is an open, efficient, and extensible resource that can underpin studies
+of selection, recombination mechanisms, and global viral spread.
+It provides a foundation for benchmarking future inference methods and
+for integrative analyses of viral evolution.
+
+\item Together, these results demonstrate that recombination in
+SARS-CoV-2 can be robustly inferred and represented at scale, and that
+such representations are critical for understanding the emergence and
+spread of major variants. Scalable ARG inference has now become a
+practical tool for pandemic genomics.
+\end{itemize}
+
+Our catalogue of recombination events is intentionally conservative. Stringent
+quality-control filters (requiring multiple supporting loci per parent) produce
+a high validation rate among QC-passing events (>\;90\%), and the set of
+recombinants we report should be interpreted as a lower bound on true
+recombination. At the default parsimony penalty (\(p=4\)) \texttt{sc2ts}
+recovers the majority of well-characterised recombinant lineages, while
+avoiding many false positives that arise in less-curated datasets. Differences
+between our prevalence estimates and earlier reports largely reflect these
+choices of data curation and thresholds, and underscore the importance of
+rigorous QC when estimating recombination at pandemic scale.
+
+Comparisons to Pango X designations clarify previously confusing features of
+SARS-CoV-2 recombinant diversity. Several nominally distinct X lineages are
+parsimoniously explained as descendants of the same ancestral recombination
+event, indicating systematic over-splitting in some parts of the nomenclature.
+Conversely, a number of designated X lineages show little evidence of
+recombination in the ARG. Taken together, these observations simplify the
+genealogy of Omicron and related groups: rather than many independent
+crossovers, a smaller number of ancestral recombination events gave rise to
+nested clusters of descendant lineages.
+
+The dated ARG enables an explicit connection between epidemiology and
+recombination. The weekly count of inferred recombination events tracks both
+global case numbers and lineage diversity; the product of cases and expected
+heterozygosity (nH) is the strongest single predictor of event counts. This
+joint dependence is consistent with the mechanistic expectation that
+recombination requires both frequent co-infections (high prevalence) and
+genetically distinct co-infecting viruses (high diversity), and implies
+predictable ‘‘hot periods’’ for recombinant emergence during major waves.
+
+Genomically, breakpoints are enriched near the Spike boundaries, a pattern
+observed previously at smaller scales. The repeated occurrence of junctions
+around Spike suggests that mechanistic biases in template switching and
+selective constraints on Spike jointly shape where viable recombinants arise.
+These hotspot locations therefore provide concrete hypotheses for experimental
+work aimed at understanding how recombination and selection interact to produce
+successful hybrid genomes.
+
+Methodologically, \texttt{sc2ts} demonstrates that scalable, real-time ARG
+inference is tractable for pandemic-scale datasets. By combining a simplified
+Li and Stephens HMM with incremental sample addition 
+we detect and incorporate recombination in a timely manner. The alternative
+of using tree-only updates with post-hoc detection of recombination leads
+to both inaccurate trees and the loss of recombination signal over time.
+We validated \texttt{sc2ts} against empirical benchmarks
+and curated recombinant lineages; while standardised simulations remain useful
+for method comparisons and stress-testing, the global SARS-CoV-2 collection
+provides an unparalleled real-world testbed and motivated our emphasis on
+empirical validation and stringent QC.
+
+The ARG and the \texttt{sc2ts} workflows are made available within the
+\texttt{tskit} and VCF-Zarr ecosystems, enabling efficient storage, rapid
+querying, and wide reuse. This integration makes the resource immediately
+actionable for tasks such as tracing recombinant origins, mapping selective
+sweeps, and benchmarking future inference methods. As pathogen genomic
+surveillance grows, ARG-based approaches can be incorporated into routine
+pipelines to provide real-time, recombination-aware situational awareness.
+
+More broadly, the present work shows that recombination — long treated as a
+nuisance in large-scale pathogen genomics — can be robustly inferred and
+represented at scale. The ARG turns recombination into an analyzable dimension
+of viral evolution and lays a practical foundation for future studies that
+integrate genomic, epidemiological and immunological data to quantify how
+recombination interacts with selection, immunity and transmission. We
+anticipate that \texttt{sc2ts} and the ARG will catalyse both methodological
+development and biological discovery for SARS-CoV-2 and other recombining
+pathogens.
+
+% \section*{DISCUSSION}
+
+% We present here the first pandemic-scale Ancestral Recombination Graph (ARG)
+% for SARS-CoV-2, reconstructed from 2.48 million genomes. This resource provides
+% a comprehensive, recombination-aware view of the evolutionary history of the
+% virus, capturing not only point mutations but also deletions and recombination
+% events in a single coherent framework. By representing SARS-CoV-2 evolution as
+% a graph rather than a tree, the ARG unifies the relationships among lineages
+% that could not previously be reconciled by tree-based phylogenies. The
+% resulting structure serves as a long-term reference for studies of viral
+% evolution and as a foundation for future recombination-aware surveillance.
+
+% Our inference of recombination represents a conservative lower bound on the
+% true frequency of recombination in SARS-CoV-2. This conservatism arises from
+% stringent quality control criteria, including the requirement for multiple
+% supporting loci for each parental assignment. Even under these strict
+% thresholds, the ARG contains hundreds of high-confidence recombination events
+% with a validation rate exceeding 90\% among events passing QC filters. The
+% overall prevalence of recombination inferred here is substantially lower than
+% early estimates based on smaller or less curated datasets, but this is
+% consistent with our aim to prioritise specificity over sensitivity.
+% Importantly, at the default parsimony penalty ($p = 4$), the ARG recovers the
+% majority of previously recognised recombinant lineages, including those most
+% relevant to global transmission dynamics.
+
+% Comparisons with the Pango X designations reveal that several nominally
+% distinct recombinant lineages share common ancestral recombination events. This
+% over-splitting has led to an inflated perception of recombinant diversity.
+% Within Omicron, for example, multiple X lineages are better explained as
+% descendants of a single ancestral recombinant rather than as independent
+% crossover events. Resolving these redundancies clarifies long-standing
+% complexity in the Omicron genealogy and simplifies the interpretation of
+% SARS-CoV-2’s reticulate evolution. Conversely, several designated X lineages
+% show little evidence of recombination in the ARG, emphasising that not all
+% phylogenetically unusual lineages are true recombinants. Together, these
+% results provide a more parsimonious and internally consistent model of
+% recombinant evolution.
+
+% Using the dated ARG, we examined the temporal and epidemiological correlates of
+% recombination. The frequency of inferred recombination events correlates
+% strongly with both global case numbers and the diversity of circulating
+% lineages. This pattern supports the hypothesis that recombination is driven
+% jointly by viral load in the population, which increases opportunities for
+% co-infection, and by lineage diversity, which provides distinct parental
+% genomes. These findings imply predictable ``hot periods'' for recombinant
+% emergence during pandemics, particularly when multiple antigenically divergent
+% variants co-circulate at high prevalence. Understanding these dynamics may aid
+% future outbreak forecasting and genomic surveillance strategies.
+
+% Analysis of genomic breakpoint locations revealed enrichment of recombination
+% near the boundaries of the Spike gene, consistent with previous observations
+% from smaller datasets. The repeated occurrence of breakpoints in these regions
+% suggests that structural or functional constraints may influence where viable
+% recombinants can arise. Mechanistic biases in template switching, combined with
+% selective pressures on Spike for receptor binding and immune escape, may
+% together shape the landscape of recombination. The ARG thus not only captures
+% recombination events but also points to underlying biological mechanisms worthy
+% of experimental follow-up.
+
+% Beyond its biological insights, this study represents a methodological advance
+% in real-time, recombination-aware phylogenetic inference. The \texttt{sc2ts}
+% method integrates an efficient, simplified Li--Stephens model with incremental
+% sample addition, allowing scalable ARG construction from millions of genomes.
+% We validated the method against empirical data and literature-curated
+% recombinant lineages rather than simulated datasets, because the global
+% SARS-CoV-2 collection offers an unparalleled real-world benchmark.
+% Nevertheless, standardised simulations will remain valuable for refining models
+% and comparing future implementations. Our results demonstrate that the
+% \texttt{sc2ts} framework captures the essential genealogical structure of
+% SARS-CoV-2 while accommodating recombination in a computationally tractable
+% way.
+
+% The ARG produced here is embedded within the \texttt{tskit} and
+% \texttt{VCF-Zarr} ecosystems, enabling compact storage, rapid analysis, and
+% open sharing. This integration ensures that the resource is immediately usable
+% by the broader community for tasks such as mapping selective sweeps, tracing
+% recombinant origins, and evaluating lineage dynamics. The dataset and
+% accompanying workflows provide a transparent foundation for benchmarking future
+% methods and for extending ARG-based analyses to other pathogens.
+
+% In summary, this work demonstrates that recombination in SARS-CoV-2 can be
+% robustly inferred and represented at scale, transforming how we study the
+% evolution of rapidly diversifying viruses. The ARG provides a unified,
+% validated view of SARS-CoV-2 evolution that resolves the relationships among
+% recombinant lineages, links recombination frequency to epidemiological
+% dynamics, and highlights mechanistic constraints on recombination. More
+% broadly, it establishes that scalable ARG inference is now a practical and
+% powerful tool for pandemic genomics.
+
+% \subsection*{OUTLOOK}
+
+% The pandemic-scale ARG presented here establishes a new foundation for studying
+% viral evolution in real time. Beyond SARS-CoV-2, the \texttt{sc2ts} framework
+% can be directly adapted to other recombining pathogens such as influenza, HIV,
+% and rotaviruses, where complex genealogies have long limited evolutionary
+% analysis. Integrating the ARG with epidemiological and immunological data will
+% enable quantitative studies of how recombination interacts with selection,
+% immune escape, and transmission dynamics. As sequencing becomes increasingly
+% routine in global surveillance, efficient ARG inference will allow real-time
+% tracking of recombinant emergence and fitness, supporting faster risk
+% assessment and lineage designation. The combination of scalable inference, open
+% data standards, and community tools transforms recombination from a nuisance
+% factor into an analyzable dimension of viral evolution. We anticipate that this
+% resource will not only clarify the history of SARS-CoV-2 but also guide the
+% design of next-generation computational and experimental approaches to
+% understanding recombination-driven adaptation in RNA viruses.
 

### 2025-10-20 latex/main.tex
Update discussion narrative

--- 2025-10-17 latex/main.tex
@@
 of the UShER+RIPPLES results.
-[
-
-% TODO convert this into a discussion pargraph..
-% \subsection*{Efficient inference and analysis of a pandemic-scale ARG}
-
-% The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
-% which overwhelmed existing computational approaches,
-% motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
-% Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
-% the functionality required largely overlaps.
-% Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
-% and the VCF Zarr data storage format \cite{Czech2025}
-% to describe relationships among virus samples,
-% building an ARG over time as data becomes available.
-% By reusing high-quality and high-performance components of these toolkits,
-% we could develop the sc2ts inference method itself and
-% perform all the pandemic-scale analyses in this article
-% with Python code alone and no compiled languages.
-
-% To illustrate the power of these libraries, we provide some examples
-% in which we compare the Usher and sc2ts intersections in tskit format described earlier.
-% First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments.
-% The total aligned dataset for 1,226,587 samples and 27,431 sites
-% shared by the sc2ts and Usher tree sequences represents 31.34 GB of nucleotide calls.
-% Of the 80,574,518 missing data calls (Ns) in the alignments,
-% sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%).
-% Additionally, 413,303 calls made use of the IUPAC uncertainty codes.
-% Of these sc2ts imputed only 77 (0.02\%) incorrectly
-% (i.e., with a base that is not compatible with the ambiguity code).
-% UShER imputed only 4 calls from this set incorrectly,
-% reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations.
-% This computation,
-% performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
-% required 30 lines of Python code and took 2m 59s on an intel i7-9700 CPU,
-% with a peak memory usage of about 4.5 GB
-% (see \cite{Czech2025} for further benchmarks for
-% processing the Viridian data in VCF Zarr format).
-% Performing this analysis using matUtils would require
-% exporting the full dataset to VCF, and
-% converting the VCF to FASTA, and
-% then comparing the alignments one-by-one to the source FASTA alignments.
-% In contrast with matUtils and most phylogenetic software,
-% which provide a command line interface and suite of predefined operations that typically output text files,
-% tskit is a Python API (with C and Rust interfaces also supported)
-% which has many advantages \cite{Baumdicker2022}.
-% As an illustration of the open-ended and efficient nature of the analyses this enables,
-% we computed the number of polytomies (nodes that have $> 2$ children)
-% and the maximum number of children per node in the Usher and sc2ts outputs.
-% In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148)
-% and compute the maximum number of children (5918).
-% In the sc2ts ARG, it took 385 ms and 6 lines of code
-% to find the maximum number of polytomies (137,001) and
-% maximum number of children (7708) across all 348 local trees.
+
+% \section*{DISCUSSION}
+
+% NOTES: paragraph structure. Lead with the biological insights first 
+% to prevent the ``this is a methods paper'' criticism
+
+% \begin{itemize}
+% \item We present a landmark ARG, which has a comprehensive representation
+% of evolutionary history.
+% \item This is a conservative lower-bound on the occurance of recombination.
+% Very high validation rate post-QC.
+% Substantially lower than previous estimate. 
+% The first order term of this 
+% is the choice of $p$ parameter. With p=4 we show that the ARG contains
+% more of the previously identified recombinants.
+% \item Truth data for recombination is difficult to find. The closest we have is the 
+% pango X lineages. No methods agrees with all of the designations being 
+% true recombinants. Our analysis suggests that there is significant over-naming
+% with the same underlying recombination event getting associated with 
+% multiple X lienages at least twice. Also many lineages are not suggested 
+% as being recombinant. This resolves long-standing complexity in 
+% the Omicron evolution and simplifies our understanding.
+% \item Based on this cleaned up and conservative estimate of recombination
+% prevalance. Supports the idea that  recombination frequency in SARS-CoV-2 
+% is driven jointly by viral
+% load in the population and lineage diversity, implying predictable
+% recombination ``hot periods'' during pandemics
+% \item Also found evidence that breakpoints are enriched near spike, 
+% supporting findings of earlier studies.
+% Suggests functional constraints or template-switching
+% biases that may shape SARS-CoV-2 evolution.
+% Spike gene junctions are recurrent recombination zones, likely under both
+% mechanistic and selective pressures.
+% \item This ARG is base on a new method, sc2ts. Major methodological 
+% breakthrough in terms of handling recombination. We validate the method
+% by using results from the literature. We don't use simulations because 
+% there's no real point in this case. We're just focusing on SARS-CoV-2.
+% We could create some simulations and generate some metrics about how
+% well sc2ts picks up simulated recombination events on clean data, 
+% but what would that tell us ultimately? (Similarly for tree reconstruction.)
+% Certainly standardised simulations would help to refine methods, but
+% we have moved the bar forwards massively here and made a great starting
+% point for subsequent method development.
+% \item There is great potential for using this ARG in future studies,
+% and using sc2ts as a baseline to judge future methods.
+% \item The tskit and VCF Zarr toolchains enable this research.
+% \item We have shown that recombination can be handled, provided
+% feasible methods to infer it, and shown that a preexisting tool
+% chain can provide quick insignts. The future is bright.
+
+% \end{itemize}
 
@@
 
-[NOTE: the discussion needs to be reworked, and hasn't been updated
-to account for changes. See the new draft subsection on prevalence of recombination,
-though]
-
-%%% The discussion should explain the significance of the results
-%%% and place them into a broader context. Subheadings are permitted.
-
-We present sc2ts, a real-time ARG inference method,
-and applied it to build a $\sim$2.48 million-sample pandemic-scale ARG of SARS-CoV-2.
-By identifying specific samples as parents,
-sc2ts enables more precise interpretation of recombinants
-than state-of-the-art recombination detection methods,
-which find recombinants by matching a given sample
-to pre-defined population-averaged mutational profiles of Pango lineages.
-Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution,
-in contrast to RIPPLES \cite{Turakhia2022}
-which detects recombinants post hoc by searching for unusual branches in a given phylogeny,
-inevitably resulting in an incomprehensive representation of SARS-CoV-2 evolution.
-Combined with new methods and tools to evaluate evidence for recombination,
-sc2ts is a significant technological advance over current methods
-for the discovery and analysis of recombinants and
-efficient inference of SARS-CoV-2 genealogy.
-
-The ARG itself is a useful resource to support future studies of SARS-CoV-2.
-Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
-which are shown in different studies
-using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
-The ARG displays remarkable phylogenetic congruence
-with a state-of-the-art phylogeny built using UShER.
-Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
-allowing for convenient and efficient reuse of this valuable dataset
-for a variety of evolutionary and genetic analyses beyond those presented here.
-% We hope that the ARG will spur on new research areas,
-% in which recombination is naturally and fully accounted for,
-% and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
-% has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
-So far, recent developments in ARG methods have focused almost 
-exclusively on applications to human sequencing data.
-In that space, the availability of ARGs reconstructed from large datasets
-has already had tremendous impact on improving our understanding of
-human genetic ancestry and past demography \cite{Wohns2022}.
-We expect similar leaps to be enabled
-through tailoring ARG reconstruction methods
-to data from other species and pathogens.
-Specifically, existing inference methods can be adapted
-to use the sc2ts ARG for the genealogy-based study of selection \cite{hejase2022deep}
-or the geographic spread of the virus through time \cite{Grundler2025geographic}.
-This is in addition to simpler analyses,
-such as calculating mutation spectra
-or querying the number or frequency of mutations at a given position,
-which the ARG makes significantly more efficient, and
-ultimately more easily reproducible.
-Since the ARG is a condensed representation of the genetic information
-contained in the sequencing data, it can also be used to understand,
-for instance, how the variability in sequencing efforts across time 
-and location affects our ability to reconstruct the origin of 
-important lineages and detect new recombinants.
-Since reconstructing the ARG is computationally expensive, 
-and checking and improving various aspects of its quality is laborious, 
-we hope that making this resource openly available will save significant time
-and effort for other researchers and enable immediate focus on analyzing
-the information contained within this data. In-depth analysis of the
-ARG is also incredibly useful for identifying important directions for
-further research, both regarding methods (where current models do not
-correctly capture features of the data) and data processing 
-(where gaps in sequencing or data artifacts result in clear
-reconstruction errors).
-
-We contribute several novel methods and concepts
-which may find applications beyond sc2ts.
-First, we devise a simplified LS model that extends the classic concept of parsimony.
-In sc2ts, this model allows for accelerated HMM likelihood calculations,
-enabling reconstruction of the pandemic-scale ARG presented here.
-Existing HMMs used in statistical genetics may be modified
-in a similar way for significant speedup.
-Second, we propose the concept of mutations averted by recombination,
-which leads to an interpretable quantity to weigh the support for 
-the hypothesis that a given sample is a recombinant against
-the hypothesis that it is a non-recombinant,
-considering the sampled genetic diversity in an ARG and a specific HMM.
-Third, we propose the concept of HMM cost
-to identify poorly explained samples
-which are either too divergent or
-have a high number of mutations inconsistent with their collection dates.
-Here we use HMM cost to reject time-traveling samples,
-which no existing phylogenetic method detects and handles during inference
-(e.g., in the Viridian UShER phylogeny, parent and child nodes
-do not necessarily follow time order \cite{Hunt2024}),
-resulting in a cleaner dataset without samples
-with wrong collection dates that may affect downstream analyses.
-Future work may be to estimate dates
-for the suspected time traveling samples using the ARG provided here and
-the sc2ts sample matching engine.
-
-Recombination is an important evolutionary force in viruses \cite{Perez-Losada2015},
-but has been understudied in part
-due to the lack of suitable data models, methods, and tools \cite{Neches2020}.
-While the methods and tools developed here are tailored for SARS-CoV-2,
-they can be extended and adapted for other viruses.
-A major advantage of the sc2ts methods and tools is that
-they are built on top of the mature software libraries of tskit and VCF-Zarr,
-which enable pandemic-scale analyses that can be performed on typical laptops.
-Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem,
-making them highly adaptable for a broad range of applications.
-Together with the pandemic-scale ARG,
-we hope that sc2ts will be highly accessible to many researchers,
-enabling and accelerating studies of SARS-CoV-2 for years to come.
-
-\subsection*{Prevalance of recombination in SARS-CoV-2}
-[Rough first draft taken from earlier location. Sticking in a heading to
-make it clear it's distinct from other paragraphs.]
-
-The number of recombination events that we detect is largely 
-driven by the choice of recombination penalty, $k$ (or $p$
-for RIPPLES). 
-At $k=4$ we detect fewer recombinants overall than the combination of 
-UShER and RIPPLES at the same threshold. 
-However, most of the extra recombinants detected by UShER+RIPPLES
-involve violations of time ordering; whether this is due to incorrect
-metadata or these are false positives would require further study.
-However, at $p=4$, UShER+RIPPLES detects very few of the well-characterised
-Pango X-lineages which are detected by sc2ts. At the more lenient
-threshold of $p=3$, more of these events are detected (but still fewer
-than sc2ts) but at the cost of nearly 4 times more recombination
-events overall. Our detailed analysis shows that RIPPLES has a well
-defined task, which it performs well, finding the potential events
-in an exhaustive search. However, it can only operate within the
-input that it is provided, and it is the UShER tree provided with
-the Viridian data that does not have clear signals for recombination
-of X lineages such as XBB at the $p=4$ threshold. 
-The inference of recombination is therefore intrinsically linked
-with the underlying genealogy inference method.
-
-[Same point for us with XB. We don't pick up XB because of a long 
-branch and we don't have the right parents present.]
-
-[Whether k=3 or k=4 is ``correct'' is a matter for debate, and feeds
-into the next point.]
-
-The estimate for effective recombination in SARS-CoV-2 that we obtain
-at k=4 is much lower 
-than previous studies have suggbalanced
-% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
-% recombinant origins out of the first 87,695 genomes to be released
-In an early analysis, 225 potentially recombinant sequences out 
-of 87,695 samples collected before October 2020\cite{Varabyou2021rapid}.
-Over the same time period, XX events are present in the sc2ts ARG,
-representing a total of XX samples.
-In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
-589 unique recombination events in a tree of 1.6 million samples
-up to May 2021 were identified after extensive quality control.
-These nodes had a combined total of 43,104 descendants, suggesting
-that approximately 2.7\% of genomes were of recombinant origin.
-Over the same time period, the sc2ts ARG contains 14 recombinants
-passing QC with a total of 172 descendants. The ARG contains a total
-501,235 samples over this time period, giving an overall total fraction of
-0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
-events and 79-fold descendant samples over the same time period.
-The numbers are not directly comparable as the published RIPPLES 
-estimate is based on GISAID data which is enriched for
-artefacts~\cite{Hunt2024}, but given the observations in 
-the previous section on running RIPPLES on the Viridian data
-it seems likely that RIPPLES is overestimating the prevalence
-of recombination in SARS-CoV-2.
-
-% CovRecomb:  135,567 putative recombinants across the entirety of 14.5 million
-% accessed SARS-CoV-2 genomes
-
-% Although multiple studies have 
-% published lists of putatitive novel recombinant 
-% sequences~\cite{,Turakhia2022,Li2024CovRecomb} 
-% these are for GISAID data and not trival to reconcile with 
-% the Viridian data or the specific recombination events in the ARG.
-
-% RIPPLES: In a 1.6 million sample tree from May 2021, we identify 589
-% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
-% recombinant origins out of the first 87,695 genomes to be released
-% recombination events, 
-% RecombinHunt just looks at X lineages (I think?)
+% P1. The ARG. Unique resource. Emphasise utility of UShER and how 
+% this has more.
+
+We present the first pandemic-scale ancestral recombination graph (ARG) for
+SARS-CoV-2, reconstructed from 2.48 million genomes using \texttt{sc2ts}. This
+ARG represents a step change in how the evolutionary history of this virus can
+be represented and analysed. Like the UShER phylogeny, it provides a
+comprehensive and continuously updatable record of global genomic diversity,
+but it extends this paradigm by embedding recombination directly into the
+genealogy. The ARG unifies point mutations, deletions, and recombination events
+within a single, compact data structure, enabling analyses that
+were previously fragmented across multiple methods and data models. By encoding
+reticulate ancestry explicitly rather than assuming a single tree, the ARG
+provides a complete and internally consistent representation of SARS-CoV-2
+evolution. In this way, the
+ARG offers the same accessibility and scalability that made the UShER tree
+indispensable during the pandemic, while expanding its scope to encompass the
+full spectrum of evolutionary processes shaping SARS-CoV-2.
+
+% P2. This is a conservative lower-bound on recombination. 
+% Discuss other estimates.
+
+The recombination frequency estimated from the \texttt{sc2ts} ARG is
+substantially lower than that implied by the combination of UShER+RIPPLES,
+largely reflecting differences in model formulation and parameter choice.
+RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
+reported 589 unique recombination events with 43,104 descendant
+samples---corresponding to roughly 2.7\% of all genomes being of recombinant
+origin.  Using the same time interval, our inference at \(k=4\) identifies only
+14 high-confidence recombination events with 172 descendants among 501,235
+genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
+explained by the stricter parsimony penalty adopted here: increasing the
+penalty from \(p=3\) (as used in the RIPPLES analysis) 
+to \(k=4\) greatly reduces false
+positives while retaining nearly all verified recombinant lineages, 
+including those missed by RIPPLES at equivalent thresholds.  
+We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
+of recombination as
+tree rearrangements proceed, thereby requiring the more lenient 
+threshold in order to find known recombinats.
+These comparisons suggest that our estimate represents a conservative 
+lower bound. Further work will be required to determine
+the true frequency of recombination in SARS-CoV-2 evolution.
+
+% P3. Clarifies relationships between pango X lineages, simplifying
+% the Omicron sublineage.
+
+The Pango X lineage designations represent the community’s best attempt to
+catalogue recombinant SARS-CoV-2 lineages, but they arise from a largely
+manual, volunteer-driven process that synthesizes multiple lines of evidence
+and expert judgment rather than a single automated criterion.  As a result, the
+system provides the closest available approximation to “truth data,” but
+inevitably contains inconsistencies and over-splitting.  Within the
+\texttt{sc2ts} ARG, 44 Pango X lineages are represented, of which 16 correspond
+directly to recombination nodes (Type I events).  Examination of their
+genealogical relationships shows that many of these 44 lineages share common
+ancestral recombination events rather than arising independently.  In
+particular, five lineages—XAA, XAG, XAM, XR, and XU—descend from the same
+recombinant ancestor associated with XQ, while another five—XZ, XAC, XAD, XAE,
+and XAP—share a second common ancestor that we refer to as “Xx.”  These
+clusters account for roughly half of all X designations in the dataset,
+indicating that the community process has at times produced multiple labels for
+the descendants of a single recombinant lineage.  Conversely, several [TODO 
+be more concrete]
+designated X lineages show little or no evidence of recombination in the ARG,
+suggesting that some may reflect convergent mutation or limited sampling rather
+than hybrid origin.  Together, these results provide a data-driven synthesis in
+which the 44 currently recognized recombinant lineages collapse into a smaller
+number of well-supported ancestral events, simplifying the complex picture of
+Omicron-era evolution and bringing the community nomenclature into closer
+alignment with the underlying genealogy.
+
+% P4. Results on global case numbers and frequency of recomb.
+
+[FIXME: This is direct output from ChatGPT.  Review!]
+The dated ARG allows us to place recombination within its epidemiological
+context, revealing clear temporal structure in when recombinant lineages arise.
+The frequency of inferred recombination events follows global case numbers
+through the pandemic, with peaks during the Delta and Omicron waves when both
+infection prevalence and genetic diversity were high.  Quantitatively, the best
+single predictor of weekly recombinant counts is the product of global cases
+and lineage heterozygosity (\(nH\)), which explains nearly a quarter of the
+variation in event frequency (\(R^{2}_{\mathrm{adj}} = 0.24\)).  This
+relationship is consistent with the mechanistic expectation that recombination
+requires both frequent co-infections (driven by high case numbers) and
+genetically distinct co-infecting viruses (driven by lineage diversity).  In
+other words, recombination becomes most detectable when epidemiological and
+evolutionary opportunity coincide.  These results imply that the emergence of
+novel recombinants is not uniform through time but concentrated into
+predictable ``hot periods'' during large, antigenically diverse outbreaks, when
+the ecological conditions for viral recombination are most favourable.
+
+% P5. Confirm breakpoints around Spike.
+
+[FIXME: This is directish output from ChatGPT. Review!]
+The genomic distribution of breakpoints in the ARG is highly uneven, with a
+pronounced enrichment near the $5'$ boundary of the Spike gene and across the
+S1–S2 junction.  This pattern closely mirrors that observed in the
+UShER+RIPPLES dataset \citep{Turakhia2022} and in earlier 
+analyses \citep{Jackson2021} [more cites], but here it is recovered
+independently over a much longer sampling period and under a more conservative
+inference regime.  Under our more conservative estimate of the overall
+prevalence of recombination, we find that this enrichment persists despite a substantial
+reduction in the total number of inferred recombination events, indicating that
+it reflects a genuine and stable feature of SARS-CoV-2 evolution rather than a
+by-product of detection sensitivity.  Across all high-confidence events, the
+median breakpoint interval length is $\sim 2.2$kb and declines with parental
+divergence, consistent with expectations from sequence-based localization
+accuracy.  These results show that even with a conservative lower bound on
+recombination frequency, the concentration of breakpoints around Spike remains
+a robust signature of coronavirus genome biology—likely reflecting a
+combination of mechanistic biases in template switching and selective
+constraints on Spike function.
+
+% P6. Sc2ts method is a major advance. We validate with real data
+% rather than simulations. Forward looking statement about simulations.
+
+In the broader context of pathogen genomics, \texttt{sc2ts} demonstrates that
+it is now possible to reconstruct recombination-aware genealogies at the scale
+of entire pandemics.  Although recombination is recognised as a major
+evolutionary force in many pathogens—including RNA viruses such as influenza
+and HIV, and numerous bacterial species—genealogical analyses have typically
+relied on tree-based approximations that ignore or collapse recombination.
+This simplification has limited the accuracy of downstream evolutionary and
+epidemiological inference, even in well-studied systems.  \texttt{Sc2ts}
+overcomes this barrier by directly inferring an ancestral recombination graph
+from millions of viral genomes, producing the first coherent reticulate
+genealogy for any pathogen dataset of comparable size.  The ability to
+integrate recombination into genome-wide genealogies at this scale represents a
+major conceptual step forward: it transforms recombination from a confounding
+complication into an analyzable component of pathogen evolution.
+
+% P7. The software ecosystem enables diverse besplike analysis.
+
+[TODO: This is close, needs revision]
+The practical value of the ARG is amplified by its integration into the broader
+software ecosystems built around \texttt{tskit} and VCF~Zarr.  These mature,
+open-source libraries provide highly efficient storage and analysis of tree
+sequences and variant data, allowing millions of genomes to be explored
+interactively within familiar environments such as Python and Jupyter
+notebooks.  Analyses that previously required bespoke phylogenetic
+pipelines—such as extracting local genealogies, summarising ancestral
+relationships, mapping mutations to branches, or tracking recombinant ancestry
+through time—can now be performed with concise, reproducible code on standard
+hardware.  Because \texttt{tskit} and VCF~Zarr share a common data model, the
+ARG integrates seamlessly with widely used tools in the PyData ecosystem,
+enabling users to combine genealogical information with epidemiological,
+geographic, or phenotypic metadata in the same workflow.  This interoperability
+transforms what can be done with large viral genome collections, lowering the
+barrier to sophisticated evolutionary analysis and making recombination-aware
+genealogy inference accessible to a wide community of researchers.
+
+% P8. Outlook
+
+[TODO a bit muddled, refine]
+Together, these results demonstrate that recombination in SARS-CoV-2 can be
+systematically incorporated into genealogical inference at pandemic scale. The
+\texttt{sc2ts} ARG captures the complexity of viral evolution in a form that is
+both biologically realistic and computationally tractable, showing that
+recombination can be modelled rather than ignored. Embedded within an
+interoperable software ecosystem, it provides a comprehensive record of
+SARS-CoV-2 ancestry and a practical framework for real-time,
+recombination-aware genomic surveillance—turning large sequence archives into
+dynamic evolutionary models that can be queried with concise, reproducible code. 
+More broadly, this work marks a transition from
+tree-based pandemic phylogenetics to recombination-aware genealogy inference.
+While our focus is SARS-CoV-2, the approach offers a blueprint and a starting
+point for other recombining pathogens: the core ideas are portable,
+even though effective deployment will require pathogen-specific adjustments. 
+We anticipate that this resource will clarify the recent history of
+SARS-CoV-2 and catalyse methodological refinement and targeted biological
+studies of recombination across systems where such extensions prove feasible.
 
@@
 %%% and may be required for some journals and some article formats.
+
+[TODO]
 
@@
 
-During the preparation of this work, the author(s) used [NAME OF TOOL OR SERVICE] in order to [REASON].
+During the preparation of this work, the author(s) used ChatGPT  
+in order to obtain editorial input on the text.
 After using this tool or service, the author(s) reviewed
@@
 workflow (see \nameref{sec:data_and_code_availability}).
+
+\subsubsection*{Evaluating the accuracy of sc2ts}
+
+[Section laying out why we don't use simulation.]
+Benchmarking
+against extensive empirical results for SARS-CoV-2 rather than simulations
+leverages a uniquely rich literature and dataset, ensuring that the resulting
+ARG reflects both biological realism and computational reproducibility.
 

### 2025-10-22 latex/main.tex
Updates from Overleaf

@@
 assigned to that (or a descendant) Pango lineage.
-These are the Type I events, in which the ARG reflects the
+We designate these as Type I events, where the ARG reflects the
 hierarchical structure of Pango designation.
@@
 CovRecomb and is specifically intended to detect interlineage recombinants
-and been previously shown to be
+and has previously been shown to be
 less sensitive than other methods~\cite{Li2024CovRecomb}.
@@
 or where the recombination event requires a \emph{de novo} mutation
-(gold, with mutational change below).
+(gold, with mutational change below). [should we also mention the red/orange lines under the patterns that identify adjacent sites?]
 Figure~\ref{fig:recombinant_evidence}D shows some example copying patterns,
@@
 using RIPPLES, we would likely get very similar results to sc2ts
-in terms recombinants detected.
+in terms of recombinants detected.
 For a given matching operation, RIPPLES and the LS HMM 
@@
 % Discuss other estimates.
-
+% 
+% SHZ: Why should we care about getting this number/frequency right?
+% Sure, popgen folks love estimating this, but to other people,
+% existing estimates already tell us that
+% recombination in SARS-CoV-2 is (qualitatively) rare,
+% even without the advances of sc2ts.
+% Also, we are talking about the frequency of detectable recombination
+% in the last sentence, right?
+% The true frequencey of recombination is unknowable from this type of data.
+% So, the best we are ever going to get is a better lower bound.
 The recombination frequency estimated from the \texttt{sc2ts} ARG is
@@
 % the Omicron sublineage.
+% 
+% SHZ: I don't think knowing the relationships between Pango Xs is what's really important.
+% Sure, phylogeneticists would get a kick out of it,
+% but I don't think the majority of Cell readership would.
+% I think what is important is knowing
+% how many emerging recombinants carrying new combinations of mutations of concern
+% we are actually dealing with in real time.
+% Clarifying the relationships between Pango Xs make this possible.
 
@@
 % P4. Results on global case numbers and frequency of recomb.
-
+% SHZ: Should we mention how the reliability of global case number data
+% and variation in sampling effort over time and across countries
+% might affect our interpretation of these findings?
 [FIXME: This is direct output from ChatGPT.  Review!]
@@
 % P5. Confirm breakpoints around Spike. 
-
+% SHZ: Maybe we can say it's a known (instead of robust)
+% pattern of detectable recombination in CoVs in general?
+% There are some citations for template switching hotspots in SC2.
+% Also, it's not only about selection favoring new combinations of mutations in Spike.
+% It's about combinations of mutations in, say, ORF1ab, getting selected against.
+% Maybe mention epistasis?
 [FIXME: This is directish output from ChatGPT. Review!]
@@
 % rather than simulations. Forward looking statement about simulations.
-
+% 
+% SHZ: Could at least some of this go into the Introduction?
 In the broader context of pathogen genomics, \texttt{sc2ts} demonstrates that
@@
 Second, indels carry important phylogenetic signals,
-for example, as lineage-defining mutations of Alpha and Delta (REF).
+for example, as defining mutations of Alpha and Delta (REF).
 We currently treat multi-base deletions as missing data or multiple independent deletions,
@@
 Finally, future developments may account for geography in the HMM,
-favouring attachment of samples to geographically close lineages.
+favoring attachment of samples to geographically close lineages.
 

### 2025-10-22 latex/main.tex
Minor tweaks

@@
 % how many emerging recombinants carrying new combinations of mutations of concern
-% we are actually dealing with in real time.
+% we are actually dealing with.
 % Clarifying the relationships between Pango Xs make this possible.
@@
 
-Alignments read from the VCZ Zarr file were first preprocessed
+Alignments read from the VCF Zarr file were first preprocessed
 to mask any non nucleotide characters (A,C,G or T) as missing data.
@@
 the joint processes of mutation and recombination.
-It is widely used in a range of applications in population 
-genetics~\cite{mcvean2019linkage}.
 The LS model is an HMM that models a focal genome
@@
 
-Furthermore, to avoid performing redundant HMM matching on identical samples, new samples
-which were exactly matched to a (sample or non-sample) node in the ARG were not added immediately.
+Furthermore, to avoid performing redundant HMM matching on identical samples,
+new samples which were exactly matched to a (sample or non-sample) node in the ARG
+were not added immediately.
 Instead, the copying paths of these samples were recorded,
@@
 
-\subsubsection*{Choice of k paramter}
+\subsubsection*{Choice of k parameter}
 % This is a key decision that we will be referencing the discussion, so 

### 2025-10-22 latex/main.tex
Move recombinant rematching to own section

@@
 
-The LS model \cite{LiStephens2003} is an approximation of
+The Li and Stephens (LS) model \cite{LiStephens2003} is an approximation of
 the coalescent with recombination
@@
 the joint processes of mutation and recombination.
+It is a Hidden Markov Model (HMM) in which a focal genome
+is modelled as a
+sequence of nucleotides that are probabilistically emitted as
+an imperfect mosaic of a set of reference genomes
+(Figure~\ref{fig:ls_diagram}).
 The LS model is an HMM that models a focal genome
@@
 
-\subsubsection*{ARG postprocessing}
-\label{sec:arg_postprocessing}
+\subsubsection*{Recombinant rematching}
 The first step in ARG postprocessing is to examine putative recombinants to
@@
 to include only samples before the date of the recombinant. 
-
 Where the recent ancestry of a recombinant includes a branch containing
@@
 Delta, 6 for BA.2) than the previous recombinant HMM solution.
-These major saltational lineages highlight the challenges 
-of using the LS HMM with highly diverged samples.
 We then updated the ARG to remove these 74 artefactual 
@@
 
-The second postprocessing step performed is to add 1,252,208 samples
-detected as exact matches during primary inference back
-into the ARG. These are not included in the ARG during 
-primary inference to reduce memory usage and improve 
-LS HMM performance.
+\subsubsection*{Postprocessing pipeline}
+\label{sec:arg_postprocessing}
+This section described the various postprocessing steps that 
+we performed after primary inference, 
+which is encoded as a reproducible Snakemake
+workflow (see \nameref{sec:data_and_code_availability}).
+
+The first step is to perform recombinant rematching as described in the previous 
+section, to find any recombinants that can be more parsimoniously explained
+either by matching to nodes subsequently added or by targeted 
+splitting of long branches.
+
+For performance reasons, samples that match exactly to an existing
+node in the ARG are not added during primary inference, but are 
+stored in a secondary ``match database''.
+During postprocessing, we added 1,252,208 of these exact match samples 
+into the ARG using the stored information.
 
@@
 When then estimated non-sample node dates using 
-the ``variational-gamma'' algorithm in tsdate\cite{Wohns2022}
-version 0.2.3. [TODO Say something above the node ages?]
+the ``variational-gamma'' algorithm in tsdate\cite{Wohns2022,Pope2025}
+version 0.2.3. See Document \ref{sec:evaluating_node_dates} for 
+more details and analysis of these dates.
 
 Finally, we reduced the metadata stored in the ARG to a minimal
-set (node sample IDs, along with computed Pango and Scorpio 
+set (node sample IDs, and the computed Pango and Scorpio 
 designations), encoded to enable efficient columnar access,
@@
 file requires 32MiB of storage.
-
-This full post-processing pipeline is encoded as a reproducible Snakemake
-workflow (see \nameref{sec:data_and_code_availability}).
 

### 2025-10-22 latex/main.tex
Add section on evaluation via sims

@@
 
-[Section laying out why we don't use simulation.]
-Benchmarking
-against extensive empirical results for SARS-CoV-2 rather than simulations
-leverages a uniquely rich literature and dataset, ensuring that the resulting
-ARG reflects both biological realism and computational reproducibility.
+% NOTES: This section is a bit weird as it's like a discussion point, but is 
+% too technical and negative to go in there. Nonetheless, we do need to address
+% this point somewhere, as it'll be something that would surely come up.
+% We might get asked to do some simulations, but let's try and steer the 
+% reviewer into asking us to do something simple like crudely recombining
+% different lineages and seeing what we get.
+
+[NOTE: first draft]
+Sc2ts is a new method to infer ARGs for SARS-CoV-2, and the standard practise 
+is to evaluate new inference methods via a simulation analysis.  
+We have, instead, focused our validation efforts on comparing the output 
+produced by sc2ts to known results.
+Benchmarking against these extensive empirical results rather 
+than simulations leverages the uniquely rich literature, 
+ensuring that the resulting ARG (the primary contribution of this work)
+reflects biological reality.
+While specialised simulation packages for SARS-CoV-2 data
+% JK: did a search for VGSim, no recombination API included.
+exist~\cite{Shchur2022vgsim,DeMaio2022phastsim} there is 
+no support for recombination.
+While bespoke simulation models can be created by manually recombining 
+sequences in order to find false-positive and false-negative
+rates, it is unclear how well this models the complexities 
+of real data. Further work is needed to clarify the processes 
+and frequency of recombination in SARS-CoV-2 in order to 
+create simulations models that would inform this.
 

### 2025-10-22 latex/main.tex
Updates from Overleaf

--- 2025-10-22 latex/main.tex
@@
 
-Alignments read from the VCF Zarr file were first preprocessed
+Alignments read from the VCZ Zarr file were first preprocessed
 to mask any non nucleotide characters (A,C,G or T) as missing data.
@@
 the joint processes of mutation and recombination.
+It is widely used in a range of applications in population 
+genetics~\cite{mcvean2019linkage}.
 The LS model is an HMM that models a focal genome

### 2025-10-23 latex/main.tex
Removed section on sims

--- 2025-10-22 latex/main.tex
@@
 
-\subsubsection*{Evaluating the accuracy of sc2ts}
-
-[Section laying out why we don't use simulation.]
-Benchmarking
-against extensive empirical results for SARS-CoV-2 rather than simulations
-leverages a uniquely rich literature and dataset, ensuring that the resulting
-ARG reflects both biological realism and computational reproducibility.
+%%% 
+% This is an early draft which was trying to tackle the problem of why
+% we don't compare against sims. Consensus was that it's better to just
+% ignore entirely and see what reviewers say.
+%%%
+% \subsubsection*{Evaluating the accuracy of sc2ts}
+
+% [NOTE: first draft]
+% Sc2ts is a new method to infer ARGs for SARS-CoV-2, and the standard practise 
+% is to evaluate new inference methods via a simulation analysis.  
+% We have, instead, focused our validation efforts on comparing the output 
+% produced by sc2ts to known results.
+% Benchmarking against these extensive empirical results rather 
+% than simulations leverages the uniquely rich literature, 
+% ensuring that the resulting ARG (the primary contribution of this work)
+% reflects biological reality.
+% While specialised simulation packages for SARS-CoV-2 data
+% % JK: did a search for VGSim, no recombination API included.
+% exist~\cite{Shchur2022vgsim,DeMaio2022phastsim} there is 
+% no support for recombination.
+% While bespoke simulation models can be created by manually recombining 
+% sequences in order to find false-positive and false-negative
+% rates, it is unclear how well this models the complexities 
+% of real data. Further work is needed to clarify the processes 
+% and frequency of recombination in SARS-CoV-2 in order to 
+% create simulations models that would inform this.
 
@@
 
-We developed a SnipIt-like~\cite{OToole2024} visualisation 
-that shows the informative genomic positions between 
-a recombinant and its parent sequences, and highlights 
-potentially problematic loci
-(See Figure~\ref{fig:copying_pattern_examples} caption for details).
-Inspection of these copying patterns for all the recombination events
+Inspection of the copying patterns for all the recombination events
 identified associations between recombination events that failed QC and

### 2025-10-23 latex/main.tex
Update validation section, move spurious recombs to supp

@@
 We first exported the alignments for the recombination node and both 
-parents to FASTA format, and then rand 3SEQ in ``full mode''
+parents to FASTA format, and then ran 3SEQ in ``full mode''
 independently for each recombinant. Specifically, we 
@@
 thresholds.
-
 We used 3SEQ source code downloaded from \url{https://gitlab.com/lamhm/3seq}
-at git hash e45ee3b4 for all analyses.
-
-\subsubsection*{Likely spurious ancestrally-important recombinants}
-\label{sec:spurious-ancestrally-important-recombinants}
-In the sc2ts ARG, there are two recombination events which have a large number of
-descendants but which are not classified as recombinants by Pango. The first of these
-corresponds to recombination event within BA.2 whose breakpoint is at position 27382
-on the far right of the genome (see supplementary document S3, node 1030562). Although
-supported by 5 sites, three of them are adjacent so that they only represent 3 loci,
-meaning that this event does not pass our recombinant quality control measure.
-
-The second of these involves a single breakpoint at position 26858, and is the origin
-of the BA.5 lineage in the ARG (see supplementary document S3, node 1189192). Although
-supported by 5 loci on the right, this recombinant also requires 3 de-novo mutations,
-some of which are adjacent to the supporting loci. Furthermore, when the first BA.5
-sample is added to the ARG, it is possible to construct a more parsimonious
-non-recombinant origin by long branch splitting (see the \nameref{sec:arg_postprocessing}
-section), although subsequent tree building does not make this a parsimonious
-way to rewire the final ARG. We therefore suggest that this too could be an artifactual
-recombination event.
+at git hash e45ee3b4.
+
+For CovRecomb\cite{Zhou2023virusrecom}, we obtained the original alignments 
+for the ``causal'' samples for each recombination node, and ran
+CovRecomb-Local-Version with default settings. We used source code 
+downloaded from \url{https://github.com/wuaipinglab/CovRecomb} at 
+git hash 1fbf6b2.
+
+We ran rebar in a similar manner, exporting the original alignments for the 
+causal samples to FASTA and using default parameter values. We used
+rebar version 0.2.1 from bioconda~\cite{Gruning2018bioconda}, and 
+rebar dataset version ``2025-01-28''.
+
+These steps are included in the Snakemake postprocessing
+workflow (see \nameref{sec:data_and_code_availability}).
 
@@
 so that they are right-exclusive.
-
-We ran rebar on the sampled sequences associated with each QC-passing recombination event.
-Rebar identifies a sequence as a recombinant if it contains sets of mutations 
-which are pre-determined to be associated with different Pango lineages.
 
@@
 
+\subsection*{Likely spurious recombinants with many descendants}
+\label{sec:spurious-ancestrally-important-recombinants}
+In the sc2ts ARG, there are two recombination events which have a large number of
+descendants but which are not classified as recombinants by Pango. The first of these
+corresponds to recombination event within BA.2 whose breakpoint is at position 27382
+on the far right of the genome (see supplementary document S3, node 1030562). Although
+supported by 5 sites, three of them are adjacent so that they only represent 3 loci,
+meaning that this event does not pass our recombinant quality control measure.
+
+The second of these involves a single breakpoint at position 26858, and is the origin
+of the BA.5 lineage in the ARG (see supplementary document S3, node 1189192). Although
+supported by 5 loci on the right, this recombinant also requires 3 de-novo mutations,
+some of which are adjacent to the supporting loci. Furthermore, when the first BA.5
+sample is added to the ARG, it is possible to construct a more parsimonious
+non-recombinant origin by long branch splitting, 
+although subsequent tree building does not make this a parsimonious
+way to rewire the final ARG.
+
 \subsection{Origins of major VOCs}

### 2025-10-23 latex/main.tex
Update phylo backbones section

@@
 
-\subsubsection*{Comparison with UShER}
+\subsubsection*{Conversion of UShER tree to tskit}
 To facilitate comparison with the sc2ts ARG, we first converted the Viridian 
-UShER tree from MAT~\cite{McBroome2021} to tskit format, which has similar
-properties but is more general 
-(see the \nameref{sec:args} section).
+UShER tree from MAT~\cite{McBroome2021} to tskit format.
 We first downloaded the tree in JSON format (tree.all\_viridian.202409.jsonl)
@@
 nucleotide format using \texttt{matUtils summary --translate}
-(usher version 0.6.3), and translated
+(usher version 0.6.3 from bioconda), and translated
 these mutations into tskit site and mutation tables. 
-The resulting tszip file 
-(usher\_viridian\_v1.0.trees.tsz) contains 5,345,019 nodes, 3,364,841
-mutations and requires 29MB of storage space.
-We validated that the data was correctly encoded by comparing the 
-variant data for the UShER tskit tree sequence against our alignments.
+The resulting file 
+ contains 5,345,019 nodes, 3,364,841
+mutations and requires 29MB of storage space in tszip format (usher\_viridian\_v1.0.trees.tsz).
+We validated that the data encoding by comparing the 
+variant data for the UShER tskit tree sequence against our MAFFT alignments.
 There was an exact match between the non-ambiguous nucleotide calls 
@@
 and using tsdate (version 0.2.3) to date internal nodes.
-We did not date the main UShER tree in this way because
-of the time-travelling samples present.
 
@@
 \subsubsection*{Comparison of phylogenetic backbones}
-
 We compare the non-recombinant parts of the ARG with the UShER topology
@@
 We focus on the 1473 Pango lineages that have perfect ``origination events'' 
-as described in \ref{sec:supplement_pango_lineages}, and
-omit Pango designations starting with ``X''
-to avoid recombinants, leaving 1329 Pango lineages.
-To ensure that each sample is a tip (rather than an internal node)
-in the simplified topologies,
+(Document \ref{sec:supplement_pango_lineages}). We then omit
+omit X lineages, leaving 1329 lineages.
+To ensure that each sample is a tip in the simplified topologies,
 a representative sample is then chosen as the oldest sample
@@
 We then further remove 4 lineages
-which are singleton descendants of a recombination event in the ARG
-(representing potentially missed Pango recombinant lineages),
+which are singleton descendants of a recombination event in the ARG,
+% NOTE: would have been simpler to say we just got rid of XBB and 
+% descendants.
 as well as 43 samples that are descendants of the recombinant XBB
@@
 This results in a simplified ARG of 1282 samples,
-each sample having a different Pango designation.
+each sample from a different Pango lineage.
 
 The simplified ARG contains only two recombination nodes, which
-correspond to those on the far right of the genome identified as
-artefactual in the FIXME Xref \ref{sec:spurious-ancestrally-important-recombinants} section.
+are the two likely artefactual events discussed in 
+Document \ref{sec:spurious-ancestrally-important-recombinants}, 
+both of which have breakpoints near the 5' end of genome.
 We therefore compare the Usher topology against the first tree in the ARG (covering
 positions 1-26858; the majority of the genome). To reduce plot density, we consider
-only those Pango lineages with greater than 10 samples, and to aid visualization,
-we divide resulting tree into 4 parts: a BA.5 subtree of 187 Pangos,
-a Delta subtree of 104 Pangos, a BA.2 subtree (excluding BA.5 descendants) of 118 Pangos,
-and a base tree (i.e. excluding all descendants of Delta and BA.2) of 405 Pangos,
-Figures \ref{fig:tanglegram_base}-\ref{fig:tanglegram_ba5} show these trees compared
-against the equivalent dated UShER topologies.
+only those Pango lineages with greater than 10 samples.
+Figure \ref{fig:tanglegram_base} shows the resulting tree compared with 
+the equivalent dated UShER tree as a tanglegram.
+We also show subtrees for Delta (Figure~\ref{fig:tanglegram_delta}),
+BA.2 (Figure~\ref{fig:tanglegram_ba2}),
+and BA.5 (Figure~\ref{fig:tanglegram_ba5}).
+
 The order of tips in each plot was chosen to reduce tangling,
-as calculated using the NeighborNet algorithm from Dendroscope \cite{Scornavacca2011,Huson2012},
+as calculated using the NeighborNet algorithm from 
+Dendroscope \cite{Scornavacca2011,Huson2012},
 with tanglegrams plotted using the SVG outputting facility built into tskit.
 See \href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/supp\_pdf-cophylogeny-tanglegrams.ipynb}{supp\_pdf-cophylogeny-tanglegrams.ipynb}
-for details and the pipeline used.
-
-Inspection of Figures~\ref{fig:tanglegram_base}-\ref{fig:tanglegram_ba5} reveals that the majority of resolvable nodes,
-including most high-level Scorpio groupings,
-are shared between the ARG and the UShER tree (magenta nodes), and there is extremely close (usually perfect) agreement on topological relations between the Pango lineages.
-In the few cases where the trees differ, this is often in terms of degree of resolution of polytomies.
-In these cases, sometimes the sc2ts ARG shows greater resolution,
-as in the base of the BA.2 subtree, and
-sometimes the UShER tree shows greater resolution, as in the close relatives of Beta (B.1.351).
+for details.
 

### 2025-10-23 latex/main.tex
Mutation spectra

@@
 \subsubsection*{Comparison of mutational spectra}
-
-To validate the mutational information contained in the sc2ts ARG,
-we compared the all-site mutational spectra of major VOCs in the ARG,
-a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024},
-and another UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
-Mutational spectra are defined as the relative proportions of
-each possible type of nucleotide change.
-
-First, we calculated the mutational spectra of Alpha, Delta, Omicron BA.1, BA.2, BA.4, and BA.5
-in the sc2ts ARG, considering mutations (ignoring indels) above the sample nodes and internal nodes
+% To validate the mutational information contained in the sc2ts ARG,
+% we compared the all-site mutational spectra of major VOCs in the ARG,
+% a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024},
+% and another UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
+% Mutational spectra are defined as the relative proportions of
+% each possible type of nucleotide change.
+
+We calculated the mutational spectra of Alpha, Delta, Omicron BA.1, BA.2, BA.4, and BA.5
+in the sc2ts ARG, considering mutations (ignoring indels) above the sample 
+nodes and internal nodes
 which have Scorpio labels associated with the VOCs
 (e.g., for Alpha, B.1.1.7-like and B.1.1.7-like+E484K).
-These Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin.
 Note that $\sim$98\% of all the mutations in the ARG occur
@@
 
+We calculated the mutational spectra of the same VOCs
+in the intersection UShER tree in tskit format using a similar approach.
+Here, we used the Scorpio sample assignments provided in the Viridian metadata,
+and therefore restricted to mutations above sample nodes only.
+
 Next, we calculated the mutational spectra of the same VOCs
-using the mutational counts from an UShER phylogeny analyzed in Bloom et al.\cite{Bloom2023},
+using the mutational counts from an UShER phylogeny 
+analyzed by Bloom et al.\cite{Bloom2023},
 which was built using a Nextclade pipeline and
@@
 20I, Alpha; 21I and 21J, Delta; 21K, BA.1; 21L, BA.2; 22A, BA.4; and 22B, BA.5.
-The mutation counts were downloaded from https://github.com/jbloomlab/SARS2-mut-spectrum/.
-
-Lastly, we calculated the mutational spectra of the same VOCs
-in the Viridian UShER phylogeny\cite{Hunt2024},
-which was pruned down to the samples present in the sc2ts ARG.
-We used mutations above the sample nodes,
-for which Scorpio labels are available
-in the Viridian metadata provided by Hunt et al.\cite{Hunt2024}.
+The mutation counts were downloaded from
+\url{https://github.com/jbloomlab/SARS2-mut-spectrum/}.
+
+See
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/fig\_mutational\_spectra.ipynb}{fig\_mutational\_spectra.ipynb}
+notebook for details.
 

### 2025-10-23 latex/main.tex
Update concordance methods section

@@
 
-See
+See t he
 \href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/fig\_mutational\_spectra.ipynb}{fig\_mutational\_spectra.ipynb}
@@
 
-\subsubsection*{Comparison with recombination detection methods}
-[TODO Merge in this text that was in the Results]
-
-To evaluate the concordance of the methods in
-identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
-we focus on the Pango X lineages associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
-using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
-For RecombinHunt and CovRecomb,
-we summarized the detection results for these Pango X lineages (if available)
-reported in the original studies \cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
-For each method, we determined the Pango X lineages
-for which the inferred parental lineages and breakpoint intervals
-are concordant with those identified by the community (STAR Methods).
-We found that the methods performed well overall,
-with comparable concordance rates (Table~\ref{tab:method_concordance}),
-indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.
-
-[TODO explain why we restrict to the Type I events]
+\subsubsection*{Concordance of recombination events}
+We compare concordance rates of sc2ts in identifying the 
+Pango lineages of recombinant parents and the location
+of breakpoint intervals with RecombinHunt\cite{Alfonsi2024} and 
+CovRecomb\cite{Zhou2023virusrecom}.
+We use these two methods because detection
+results on the Pango X lineages were reported and made available 
+in the original publications. We do not compare here with 
+UShER+RIPPLES+RIVET because curated data on the X lineages is 
+not available for this pipeline (see Document \ref{sec:detailed_ripples}
+and Table~\ref{tab:ripples_x_lineages}, however).
+We use the data from the Pango designation community 
+(\url{https://github.com/cov-lineages/pango-designation/})
+as the ``truth'' when comparing performance in identifying
+parent Pango lineages and breakpoints between the methods.
+We focus on the 16 Type I recombination events 
+(Table~\ref{tab:pango_x_lineages}) identified by sc2ts for simplicity.
+While this is certainly biased in favour of sc2ts (i.e., by 
+ignoring likely sc2ts false negatives such as XB), it helps to 
+understand the probable accuracy of sc2ts on the true positives.
+It is important to emphasise that the ``true'' truth set among
+the Pango X lineages is currently unknown, and there is significant 
+disagreement among methods beyond a core set of lineages 
+(see Discussion).
 
@@
 as the left and right recombination-informative sites.
-We adjusted their coordinates by adding one to the right coordinate
-so that the intervals are right-exclusive.
 
 For CovRecomb, we summarized the detection results for the Pango X lineages as follows.
-First, we obtained the "independent" recombination events
+First, we obtained the ``independent'' recombination events
 identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study)
@@
 identified in the study
-(https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv)
+(\url{https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv}).
 For each of these Pango X lineages,
@@
 The breakpoint intervals are defined as the left and right recombination-informative sites.
-We adjusted the breakpoint intervals by adding one to the right coordinate
-so that they are right-exclusive.
 
@@
 if they have Pango lineage assignments
-that are at least as specific as those proposed by the community.
+that are at least as specific as those proposed by the Pango designation community.
 The breakpoint interval of a Pango X inferred using a method is deemed concordant
 if it overlaps with that proposed by the community by at least one base.
-These breakpoint intervals are defined as ranges of possible site positions
-where a breakpoint is inferred to occur.
-To make these breakpoint intervals right-exclusive, we adjusted them
-by subtracting one from the left coordinate and adding two to the right coordinate.
+
+The results are reported in Table~\ref{tab:method_concordance}.
+See the
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/tab\_methods\_concordance.ipynb}{tab\_methods\_concordance.ipynb}
+notebook for details.
 

### 2025-10-23 latex/main.tex
Tidy up pangonet

@@
 which relates all the designated Pango lineages.
-To construct this pseudo-phylogeny, we used \textit{pangonet},
+To construct this pseudo-phylogeny, we used 
+pangonet (\url{https://github.com/phac-nml/pangonet})
 which mines the information on the Pango lineages
 in the ``alias\_key.json'' and ``lineage\_notes.txt'' files
-from the Pango designation issues
+defining the designations
 (\url{https://github.com/cov-lineages/pango-designation}).
-If two Pango lineages are siblings in the pseudo-phylogeny,
-the distance is defined as
-the sum of the number of edges from each lineage to their MRCA;
-if they are non-siblings (i.e., one Pango is a descendant of the other),
-the distance is defined as
-the number of edges separating them;
-if they are identical, the distance is defined as zero.
+The distance between two Pango lineages is then the number of edges
+separating the corresponding nodes in this graph.
 

### 2025-10-23 latex/main.tex
Start on RIPPLES methods section

@@
 \cite{Duchene2020,Ghafari2022,Markov2023}.
+
+\subsubsection*{Comparison with UShER+RIPPLES}
+We began with the Viridian UShER tree downloaded from FigShare.
+To simplify the comparison with sc2ts, we subset this 
+file down to the 2,475,418 samples shared with the sc2ts ARG 
+by running \texttt{matUtils extract}.
+We then ran \texttt{ripples-fast}~\cite{Smith2023tracking}
+on this file, using \texttt{-n1} so that all detected 
+recombinants were reported (equivalent to sc2ts). We ran
+ripples with the parsimony improvement parameter $p$
+equal to 3 (the default) and 4.
+
+The output of RIPPLES is two text files,
+
+To compare these results with the sc2ts ARG, we first matched clades
+by finding the most recent common ancestor (MRCA) of the samples
+descending from each RIPPLES event. Overall, clade agreement was
+high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG
+where descendants match exactly. Of these nodes, 68\% had no recombinant
+ancestry. Of the remainder, 595 could be matched with
+one of the 855 recombination events in the sc2ts ARG
 

### 2025-10-23 latex/main.tex
More work on ripples

@@
 
-See t he
+See the
 \href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/fig\_mutational\_spectra.ipynb}{fig\_mutational\_spectra.ipynb}
@@
 recombinants were reported (equivalent to sc2ts). We ran
-ripples with the parsimony improvement parameter $p$
+RIPPLES with the parsimony improvement parameter $p$
 equal to 3 (the default) and 4.
 
-The output of RIPPLES is two text files,
+The output of RIPPLES is two text files, ``recombination.tsv``
+which lists details of putative recombination events and ``descendants.tsv``
+which lists the descendant samples of all recombination events.
 
 To compare these results with the sc2ts ARG, we first matched clades
-by finding the most recent common ancestor (MRCA) of the samples
+bynding the most recent common ancestor (MRCA) of the samples
 descending from each RIPPLES event. Overall, clade agreement was

### 2025-10-23 latex/main.tex
Comparison with RIPPLES methods

--- 2025-10-23 latex/main.tex
@@
 recombinants were reported (equivalent to sc2ts). We ran
-ripples with the parsimony improvement parameter $p$
+RIPPLES with the parsimony improvement parameter $p$
 equal to 3 (the default) and 4.
 
-The output of RIPPLES is two text files,
-
-To compare these results with the sc2ts ARG, we first matched clades
-by finding the most recent common ancestor (MRCA) of the samples
-descending from each RIPPLES event. Overall, clade agreement was
-high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG
-where descendants match exactly. Of these nodes, 68\% had no recombinant
-ancestry. Of the remainder, 595 could be matched with
-one of the 855 recombination events in the sc2ts ARG
+The output of RIPPLES is two text files, ``recombination.tsv''
+and ``descendants.tsv``. We use the ``descendants.tsv'' file,
+which lists the descendant samples of all recombination events
+keyed by the recombinant node ID, to match events with sc2ts.
+Specifically, for each RIPPLES event identified a list of 
+descendant samples, we find the MRCA of 
+that set of samples in the first tree of the sc2ts ARG. 
+The clades are then said to agree if the number of additional 
+samples descending from the sc2ts node is $<2$
+(Figure~\ref{fig:ripples_sc2ts_events}).
+
+We ran 3SEQ on the RIPPLES recombinants by extracting the 
+alignments for the child and parent sequences for each 
+recombination event. We first obtained the parent
+node IDs from the ``recombination.tsv'' file by choosing 
+the most parsimonious event associated with each recombinant
+node,  arbitrarily choosing among the possibilities in 
+the case of ties. We then used this list of child and 
+parent node IDs as 
+input to \texttt{matUtils extract -v} to extract their
+sequences in VCF format. Following this, 
+we used vcflib~\cite{Garrison2022spectrum} to convert
+to alignments in FASTA format, as recommended by
+the UShER documentation (\texttt{vcf2fasta -f NC\_045512v2.fa}).
+We then ran 3SEQ in ``full mode''
+independently for each trio, in the same manner as described
+when validating the sc2ts recombinants.
+
+The UShER tree is not dated, and so to obtain dates for 
+parent nodes we used sample metadata 
+(Date\_Tree from the Viridian metadata). For the minority 
+of parent sequences that are not samples, we used the oldest
+sample descending from a given internal node as a proxy for 
+its date. We used the BTE Python library 
+(\url{https://jmcbroome.github.io/BTE/build/html/index.html})
+to perform the required traversals on the UShER tree.
+
+The full workflow is encoded in the Snakemake postprocessing 
+pipeline.
+See Document~\ref{sec:detailed_ripples} for detailed analysis
+of the RIPPLES events, 
+and the
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_ripples.ipynb}{analysis\_ripples.ipynb}
+notebook for full methodological details of the analysis.
 

### 2025-10-23 latex/main.tex
Define deletion presense
Closes #886

@@
 we list the recurrent deletions in the ARG
-which are each observed in at least 10,000 (0.4\%) samples
-% [FIXME define how we measured this].
-(i.e., for a deletion resulting from multiple events in the ARG,
-this is the total samples inheriting the deletion).
+which are each observed in at least 10,000 (0.4\%) samples.
+A deletion is said to be observed in a sample here 
+if the gap character is present at the specified positions,
+and the immediately adjacent positions are not the gap
+character.
 

### 2025-10-23 latex/main.tex
Added a note on node dates to limitations bit
Move some stuff around

@@
 Document S2 for ARG visualizations of the Pango X lineages.
+[TODO ref Document \ref{sec:supplement_rejected_pango_x_lineages}
+to mention multi breakpoints and close issue 379]
 
@@
 favoring attachment of samples to geographically close lineages.
+
+TODO: add a short discussion of the node dates and some references
+to the literature, closing issue 725
 
@@
 
-\subsection*{Likely spurious recombinants with many descendants}
+\subsection{Likely spurious recombinants with many descendants}
 \label{sec:spurious-ancestrally-important-recombinants}
@@
 
-\subsection{Sc2ts detection results of rejected Pango X lineages}
+\subsection{Additional analyses of the Pango X lineages}
+\label{sec:supplement_indepth_pango_x_lineages}
+
+\newcommand{\scghdiscussion}[1]{%
+See \href{https://github.com/jeromekelleher/sc2ts-paper/discussions/#1}%
+{github:sc2ts-paper/discussions/#1} for further details.}
+
+\subsubsection{XA}
+% mneumonic: deep dive XA
+\label{sec:dd_xa}
+XA traces to a well supported recombination node, with no reversions.
+Note that the ``causal'' sample ERR5308556
+(which triggers the initial recombination event)
+is identical to the recombination node, and
+therefore lacks C8090T,
+which is one of the XA consensus mutations
+(identified as those shared by over 90\% of XA samples).
+As mentioned elsewhere,
+the sc2ts detection results in terms of
+the Pango lineages of the parents and the breakpoint intervals
+agree with the results reported by Jackson et al. (\ref{tab:jackson}).
+Two of the Jackson et al. XA samples
+are present in the ARG (ERR5308556 and ERR5414941);
+these samples are specifically marked in Figure~\ref{fig:subgraphs},
+which is a summary of this subgraph.
+
+\subsubsection{XB}
+\label{sec:dd_xb}
+
+The displayed subgraph shows a small selection of the 192 XB samples.
+XB was originally defined as a recombinant between
+the closely related lineages B.1.634 (left parent) and B.1.631 (right parent)
+from North and Central America \cite{Gutierrez2022}.
+However, these parent lineages were not added to the ARG due to a high HMM cost.
+Together with other lineages such as B.1.627
+(sister to B.1.631 \cite{Gutierrez2022}),
+they are likely to attach somewhere along the long branch
+in the subgraph visualization (with $> 30$ mutations),
+which also contains the key ``R2'' deletion
+associated with XB / B.1.634 \cite{Gutierrez2022}.
+
+We hypothesize that there are a set of samples
+from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
+which, should they be added to the ARG, would break up the long branch,
+with XB being a recombination between two of them.
+In the absence of these, sc2ts cannot detect XB as a recombinant.
+Also, note that the XB consensus mutations all occur on or above the long branch.
+We suspect that the child node of the long branch
+represents the (potentially recombinant) origin of XB, and
+the B.1 samples at the bottom left of the subgraph
+may be misclassified by pangolin.
+
+Further work is required to determine
+why these additional samples were not gradually added
+to the ARG by the daily sample matching process or
+retrospective sample grouping.
+
+\scghdiscussion{780}
+
+\subsubsection{XC}
+\label{sec:dd_xc}
+The five XC samples descend from a well supported recombination node.
+The two parents that are both identical to actual samples
+(ERR4908740 on the left and DRR321274 on the right).
+There is one reversion (G21987A) on the left side of the breakpoint,
+but the site 21987 was masked during ARG inference
+because it was identified to be experiencing
+frequent mutation in a preliminary ARG.
+
+\subsubsection{XE/XH}
+\label{sec:dd_xexh}
+
+The XE samples descend from a well supported recombination node,
+with one mutation and one reversion (T22792C).
+This reversion was mapped onto the ARG
+as a part of postprocessing,
+and could be placed more parsimoniously by a single C22792T
+that groups together XH
+and the lineage leading to the BA.2 recombination nodes
+displayed on the lower left.
+Nevertheless, the exact arrangement of the mutations at position 22792
+should not affect the proposed breakpoint,
+which is very distant, at position 11283.
+This breakpoint coincides with a repeated homopolymeric sequence
+involving 5-base deletions,
+so that alignment errors could have a small effect
+on the breakpoint position.
+Note that the two lower recombination nodes are both well supported,
+the youngest being postulated as the origin of XAF (see below);
+we leave investigation of this suggested series of
+sequential recombination events to future work.
+
+\scghdiscussion{633}
+
+\subsubsection{XF}
+\label{sec:dd_xf}
+XF is a well-supported recombinant between Delta and Omicron BA.1,
+with the breakpoint on the far left of the genome.
+All XF samples are the unique descendants of
+a well supported recombination node,
+with both parent nodes (ERR6193241: AY.4 and ERR8042040: BA.1) and
+the single child node (ERR8076129) being samples.
+
+\subsubsection{XG}
+\label{sec:dd_xg}
+The three XG samples descend from a reconstructed internal node,
+which in turn descends from a well supported recombination node.
+The left parent is a sample (SRR20179863), and
+the right parent is a reconstructed internal node
+assigned to Omicron BA.2,
+which is supported by a sample (ERR9124067) that is identical to it.
+
+\subsubsection{XJ}
+\label{sec:dd_xj}
+All the XJ samples descend from a well supported recombination node.
+There are 14 samples (including e.g. ERR8683190)
+which sc2ts identified as descendants of the same recombination node
+but are labeled BA.2 by pangolin (displayed on the left), and
+which do not possess the XJ consensus mutations C22792T and C27945T.
+These BA.2 samples could be worthy of further investigation.
+
+\subsubsection{XL}
+\label{sec:dd_xl}
+All the XL samples descend from a well supported recombination node,
+with three child samples which are identical to it.
+The left parent is a sample (ERR7989424), and
+the right parent is a reconstructed internal node
+assigned to BA.2.5 with a large number of descendant samples.
+
+\subsubsection{XM/XAL}
+\label{sec:dd_xm}
+
+XM is a recombinant of BA.1 (left parent) and BA.2 (right parent).
+Sc2ts identifies multiple recombinant origins of XM.
+One of these origins leads to a dominant clade
+(node 1003220 at position 21,595),
+which also contains the only three XAL samples.
+
+Independent evidence for multiple XM origins
+comes from consideration of a 9-base deletion
+in the Spike gene at position 21633,
+which is characteristic of BA.2 but not BA.1
+\url{https://github.com/cov-lineages/pango-designation/issues/361}.
+Most of the XM-labeled samples in ARG have this deletion,
+but three (SRR19024311, ERR9447529, and SRR18775874) do not,
+suggesting that XM samples trace to at least two recombination events,
+one to the right of the deletion
+(which leads to the majority of the XM samples), and one to the right.
+In the subgraph, it can be seen that one of the samples, SRR18775874,
+has been (presumably wrongly) placed
+within the main XM deletion-containing clade.
+As sc2ts does not use deletions for inference,
+deletion remapping (section XXXX) has reverted the deletion in this case,
+creating many insertions on the branch above the sample (magenta).
+A better use of deletions in sc2ts would likely result
+in placing this sample within one of the two XM clades
+which do not have the deletion.
+
+\scghdiscussion{656}
+
+\subsubsection{XN/XAU}
+\label{sec:dd_xn}
+
+In the sc2ts ARG, the initial XN sample (ERR8986821)
+is well explained as a non-recombinant,
+copying from an inferred internal node
+assigned to BA.2 with only one mutation.
+This internal node is directly supported by the sample ERR9502469.
+All the other XN samples descend from ERR8986821.
+A small clade containing XAU is the sister to XN, and
+there are a series of samples (ERR9453882, ERR9502469, and ERR9823912)
+which support the gradual emergence of both groups
+from a BA.2 ancestral lineage via accumulation of single mutations.
+
+We believe that the explanation of XN and XAU
+in the sc2ts ARG is correct, and
+the evidence supporting the original designation of
+XN and XAU as recombinants
+comes from inaccurate or problematic data sources.
+In the case of XN, support for a recombinant BA.1+BA.2 origin
+is from three mutations that are claimed to be BA-1-specific:
+C241T, A2832G, and C3037T
+(https://github.com/cov-lineages/pango-designation/issues/480).
+However, the BA.1/BA.2 split
+(https://github.com/cov-lineages/pango-designation/issues/361)
+does not list C241T and C3037T as BA.1-specific,
+meaning that these two mutations are not supportive of recombination.
+Moreover, although the remaining mutation
+claimed to support a recombinant origin, A2832G,
+is typical of BA.1 and not BA.2,
+it is seen in several BA.2 samples (e.g. ERR9453882),
+supporting the idea that it has independently arisen within BA.2.
+Hence, none of the sites used to designate XN as a recombinant reliably do so.
+
+In the case of XAU, a recombinant origin was proposed
+on the basis of the same three sites, plus C2470T
+(https://github.com/cov-lineages/pango-designation/issues/894).
+However, C2470T is not BA.1-specific
+(https://github.com/cov-lineages/pango-designation/issues/361).
+Moreover, some BA.2-designated samples such as ERR9823912 also have C2470T,
+so we do not believe this to be a recombinant-supporting mutation either.
+
+This case illustrates the value of automated ARG reconstruction
+for correcting the inevitable errors that arise
+when designating recombinants by hand and
+using different (and inconsistent) data sources of
+Pango lineage-associated mutational constellations.
+
+https://github.com/jeromekelleher/sc2ts-paper/issues/348
+https://github.com/jeromekelleher/sc2ts-paper/issues/358
+
+\subsubsection{XP}
+\label{sec:dd_xp}
+
+https://github.com/jeromekelleher/sc2ts-paper/issues/345
+
+This is a case where the Pango X lineage is designated
+on the basis of two loci on the far right of the genome:
+a single nucleotide mutation (A29510C) and
+a rare, large deletion at position 29734 in the S2M step loop
+(https://github.com/cov-lineages/pango-designation/issues/481),
+which is not common enough to map in sc2ts during postprocessing
+and hence not shown here.
+Sc2ts requires at least four loci on either side
+to infer a QC-passing recombinant (section XXXX).
+Therefore, unless this deletion were given more weight than a single locus,
+sc2ts would not identify XP as a recombinant,
+even if deletions were used during ARG inference.
+If this deletion were recurrent, the evidence from sc2ts suggests
+that XP might not be a recombinant,
+because it is only one mutation different from the sample ERR8628084.
+
+\subsubsection{XQ/XR/XU/XAA/XAG/XAM}
+\label{sec:dd_xq}
+
+https://github.com/jeromekelleher/sc2ts-paper/issues/338
+
+In the ARG, XR, XU, XAA, XAG, and XAM
+are nested under the XQ recombinant lineage.
+However, upon closer inspection,
+the right hand branch above the recombination node appears problematic,
+as it displays three reversion mutations,
+which revert mutations at positions 22792, 24023, and 25624
+in the ancestral lineage just above.
+These sites were excluded during ARG inference, and
+remapped onto the ARG by parsimony during ARG postprocessing.
+We posit that if these three sites were used during ARG inference,
+the right parent of the recombination node
+would be chosen to be the ancestral BA.2 sample
+on the right-hand path with 18 child nodes.
+A more parsimonious reconstruction
+would then have the XQ clade as a sibling to
+a group of (monophyletic) XAA, XAM, XAG, and XR lineages.
+This would remove between 4 and 5 reversion mutations.
+
+\subsubsection{XS}
+\label{sec:dd_xs}
+
+https://github.com/jeromekelleher/sc2ts-paper/issues/287
+
+In the ARG, the XS samples appear to have originated from
+two sequential recombination events (separated by a single branch).
+However, the later event is likely an artifact,
+because the XS sample (SRR18343220) associated with
+this event failed detection-level QC
+(i.e., the minimum net number of supporting loci;
+see the lower copying pattern).
+In addition, this sample was sequenced using the Ion AmpliSeq V1 kit,
+which was also used to sequence eight other samples
+descending from the same event
+(clustered due to a shared sequencing error due to AmpliSeq).
+These eight samples would likely attach to the nodes
+descending from the first recombination event.
+We excluded this likely artifactual event from further analysis and
+consider XS to have emerged from a single recombinant origin.
+
+\subsubsection{XW}
+\label{sec:dd_xw}
+
+All the XW samples descend from a well supported recombination node.
+Its left parent is a BA.1.1.15-labeled sample (ERR8069658), and
+its right parent is a BA.2-labeled sample (ERR9086669).
+
+\subsubsection{XY}
+\label{sec:dd_xy}
+
+All the XY samples descend from a well supported recombination node,
+through a shared reconstructed internal node.
+Its left parent is a reconstructed internal node assigned to BA.1.1,
+which is supported by a sample (ERR8627048) identical to it.
+Its right parent is a BA.2-labeled sample (ERR9399983).
+There are a pair of (orange) reversion mutations
+within the XY clade at position 23948:
+these were not used for inference
+but remapped using parsimony during postprocessing, and
+indicate that the topology within the XY clade
+could potentially be improved
+by alternative phylogenetic reconstruction methods.
+
+\subsubsection{Xx: XZ/XAC/XAD/XAE/XAP}
+\label{sec:dd_xx}
+
+https://github.com/jeromekelleher/sc2ts-paper/discussions/631
+
+This nested set of Pango X lineages descends
+from a single recombination node in the ARG, and
+These Pango X lineages
+may be aliases of a single recombinant
+which we tentatively refer to as ``Xx''.
+Note that all were designated around the same time,
+with their samples collected
+roughly within a three-month period (Mar., 2022 to Jul., 2022).
+Several samples labeled as BA.2 by pangolin also lie
+within this recombinant clade,
+including two that are identical
+to the proposed originating recombinant (ERR8146303 and ERR8163061).
+
+Exact relationships within the Xx clade are somewhat tentative.
+For example, there are three highly mutable sites (22792, 28877, and 28878)
+which were not used during inference,
+but were remapped onto the ARG during postprocessing.
+The inferred topology places these sites
+as recurrent (colored orange, grey, and faded yellow),
+but a more parsimonious arrangement would,
+for example, group the two XAD samples as siblings
+(removing the need to postulate mutations
+A1250G, A28877T, and G28878C to be recurrent).
+It is possible that a more parsimonious arrangement would also
+reduce the recurrent mutations and reversions seen at
+sites 22,792 (four orange mutations) and
+26,060 (two parallel brown mutations),
+but we do not show such a reconstruction,
+which is complicated by the presence of
+additional (non-displayed) samples descending from
+internal nodes in the plotted subgraph.
+
+For XAC, in particular, the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/590) suggests
+it to be a recombinant with two breakpoints:
+BA.2 on the left, BA.1 in the middle, and
+a switch back to a very small region of BA.2 on the right.
+This contrasts with the one-breakpoint BA.2+BA.1 origin inferred in the ARG.
+However, only two loci support the right hand BA.2 assignment:
+A29510C and a large deletion
+(deletions are treated as missing data during ARG inference).
+Sc2ts does not count this as enough evidence
+to insert an additional breakpoint,
+which would require at least four supporting variants.
+
+\subsubsection{XAF}
+\label{sec:dd_xaf}
+
+https://github.com/jeromekelleher/sc2ts-paper/discussions/638
+
+Of the 21 XAF samples in the Viridian v04 dataset,
+only XAF sample passed our QC filters and was added to the ARG:
+many have greater than 500 missing Ns in their consensus sequences or
+require a large number of mutations to incorporate into the ARG
+(i.e. have a high HMM cost).
+However, we find a small recombinant BA1.1/BA.2 clade of 37 samples,
+associated with the BA.2-labeled sample ERR9755541,
+within which the single XAF sample is nested.
+The inferred parents and breakpoint at 10,447
+aligns with that proposed for XAF
+(https://github.com/cov-lineages/pango-designation/issues/676).
+Inspection of the omitted XAF samples with high HMM cost shows that
+they would attach as close sisters to the included XAF sample,
+albeit on branches with many new mutations.
+Hence we believe the current XAF topology in the ARG is reasonable,
+although somewhat tentative, given the number of included samples.
+
+The mutation-rich branch in the subgraph with
+a 5-base deletion (11283-11287) and insertion (11292-11296)
+is likely to be a simple deletion alignment error.
+
+\subsubsection{XAJ}
+\label{sec:dd_xaj}
+
+https://github.com/jeromekelleher/sc2ts-paper/discussions/634
+
+In the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/826),
+XAJ is suggested to be a recombinant between
+BA.2.12.1 (left parent) and BA.4 (right parent).
+However, XAJ is not inferred to be a recombinant lineage in the ARG.
+From the subgraph, which shows the 18 XAJ samples
+plus a small number of relevant BA.4 samples,
+it can be seen that 9 sites leading to XAJ indeed recur
+in the lineage to the closest BA.4 sample (SRR18848050),
+which strongly indicates that this is a recombinant
+that is missed by sc2ts.
+The reason that is is not a recombinant in the ARG is
+that 6 sites represent a 6bp deletion and so are not used for inference.
+The remaining three sites (T22917G, T23018G, and G23040A)
+are not enough to exceed our k=4 threshold  for recombinant designation.
+Including deletions in inference would certainly
+lead to XAJ being detected as a recombinant,
+and with a lower threshold of k=3
+the distribution of mutations on the ARG topology suggests
+that it is is even possible that XAJ is a 2-breakpoint recombinant
+with the region from 21765-23535
+coming from the BA.4 labeled sample SRR18848050.
+
+\subsubsection{XAN/XAV}
+\label{sec:dd_xan}
+
+https://github.com/jeromekelleher/sc2ts-paper/discussions/783
+https://github.com/jeromekelleher/sc2ts-paper/discussions/785
+
+XAN and XAV are shown in this subgraph, as they share some mutations.
+In the Pango designation issues
+(https://github.com/cov-lineages/pango-designation/issues/771;
+https://github.com/cov-lineages/pango-designation/issues/911),
+XAN and XAV are suggested to be a recombinant between
+BA.2 (left parent) and BA.5 (right parent).
+Both these Pango lineages were detected as non-recombinants by sc2ts,
+because the number of supporting loci fell below our threshold of detection.
+In the ARG, we found only one mutation (C9866T)
+supporting the suggested BA.2 parent for both XAN and XAV.
+
+\subsubsection{XAS}
+\label{sec:dd_xas}
+
+https://github.com/jeromekelleher/sc2ts-paper/discussions/784
+
+XAS was proposed as a recombination between
+BA.5 (left parent) and BA.2 (right parent).
+However, we find that this is only two mutations
+different from e.g. ERR9929846, which is a BA.4 sample.
+In fact, XAS only differs from the BA.4 root by a single mutation,
+C27945T, which was identified in the original Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/882)
+as an important ORF truncation mutation
+which is also present in a few BA.2.65 samples.
+It is possible that C27945T was acquired by recombination,
+but this is clearly below our threshold of detection.
+
+\subsubsection{XAZ}
+\label{sec:dd_xaz}
+
+https://github.com/jeromekelleher/sc2ts-paper/discussions/786
+
+In the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/797),
+XAZ is suggested as a two-breakpoint recombinant between
+BA.2.5 and BA.5 (the middle parent),
+with three sites supporting the BA.2.5 parent
+on the left end (C2232T, C3317T, and T3358C).
+In the ARG, this is very close to being a recombinant of
+BA.2.5 (ERR9615610) on the left side of the breakpoint,
+as not only are the three mutations present in this sample,
+but there is another site which supports recombination 
+C1912T, which is not reported in the Pango designation issue).
+XAZ has C44T which is lacking in all BA.2.5 samples,
+leaving only three net supporting loci for a putative recombinant.
+However, position 44 is often clipped,
+so this is not strong evidence against a recombinant, and
+therefore XAZ is a marginal case of recombination.
+We found no evidence for the third segment
+inherited from a BA2.5 parent on the right end.
+In the ARG, all the XAZ samples form a clade.
+Here we display only the first 20 XAZ samples.
+
+\subsubsection{XBB}
+\label{sec:dd_xbb}
+
+https://github.com/jeromekelleher/sc2ts-paper/issues/336
+
+Most of the XBB samples descend from a well supported recombination node.
+In the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/1058),
+the Pango lineages of the suggested left and right parents are
+BJ.1 and BM.1.1.1, respectively.
+In the ARG, the Pango lineage of the right parent is identical,
+but that of the left parent is a related ancestral Pango lineage
+(BA.2.10 instead of BA.2.10.1.1 = BJ.1).
+This is likely due to the absence of BJ.1 in the ARG:
+only four BJ.1 samples exist in the Viridian v04 dataset,
+a single one passed sequence-level QC, and
+that was then rejected due to high HMM cost.
+Note that BJ.1 differs from BA.2.10 by over 20 mutations
+(https://github.com/cov-lineages/pango-designation/issues/915),
+explaining the large number of (gold) \emph{de novo} mutations
+in the XBB copying pattern.
+Had BJ.1 samples been present in the ARG,
+we strongly suspect that the left branch of
+the XBB-associated recombination node
+would have been attached to a BJ.1-labeled node, and
+the origin of XBB would not require postulating as many new mutations.
+The fact that sc2ts has deduced
+the XBB left parent to a close relative of BJ.1,
+even though BJ.1 is missing from the ARG,
+shows our approach to be relatively robust to missing un-sampled lineages:
+an important consideration given global variation in pathogen surveillance.
+
+\subsubsection{XBB.1}
+\label{sec:dd_xbb.1}
+
+XBB.1 is characterized by an additional mutation in Spike, G22317T (G252V) and
+is not designated as a recombinant of XBB in the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/1088).
+The ARG agrees with this and places the characteristic mutation
+above the root of a single non-recombinant XBB.1 clade.
+We include this non-recombinant lineage
+in order to highlight some well supported subsequent recombination nodes
+within XBB.1, such as the (pink) one leading to two XBB.1-labeled samples
+at the bottom left of the subgraph which we refer to as XBB.x.
+XBB.x is well supported: its left parent is a reconstructed internal node,
+which is directly supported by a BQ.1-labeled sample, and
+its right parent is a BB.1.5.18-labeled sample (ERR10792363).
+This highlights the automatic ability of sc2ts
+to infer sequential recombination events
+(i.e. recombinants-of-recombinants),
+even those involving few samples.
+A few other recombination events occur within XBB.1,
+but they either fail QC checks, or
+(as the other one shown here, to the bottom right)
+lead to a single sample so may be
+e.g. caused by accidental mixing of samples during sequencing.
+
+\subsubsection{XBD}
+\label{sec:dd_xbd}
+
+All the XBD samples descend from a well supported recombination node.
+Its left and right parents are reconstructed internal nodes,
+both of which are directly supported by samples
+(SRR21382561 and ERR9825609, respectively)
+that are identical to them.
+
+\subsubsection{XBE}
+\label{sec:dd_xbe}
+
+https://github.com/jeromekelleher/sc2ts-paper/issues/351
+
+In the Pango designation issue
+(https://github.com/cov-lineages/pango-designation/issues/1246),
+XBE was suggested as a recombinant between
+BA.5.2 (left parent; most of the genome) and
+BE.4 (right parent; somewhere from 23,609 onwards). 
+owever, sc2ts treats the 65 XBE samples
+as non-recombinant descendants of BA.5.2
+because the number of supporting sites
+fell below our threshold of detection.
+We have included the BE.4 lineage on the subgraph,
+revealing two mutations (C27800T, and G28681T)
+which occur in parallel above the root of the XBE clade
+and along the path ascending from the displayed BE.4 sample,
+suggesting that if we used k=2
+as the recombinant threshold,
+sc2ts would indeed detect XBE as a recombinant.
+
+\subsubsection{XBF}
+\label{sec:dd_xbf}
+
+All the XBF samples descend from a well supported recombination node,
+with three mutations on the left of the breakpoint.
+Its left parent is a sample assigned to BA.5.2.1 (SRR20305553).
+Its right parent is a reconstructed internal node
+that is supported by a CJ.1-labeled sample (SRR21948068),
+which differs by three mutations.
+
+\subsubsection{XBG}
+\label{sec:dd_xbg}
+
+All the XBG samples descend from a well supported recombination node,
+with two mutations, one on each side of the breakpoint.
+Both its left and right parents are samples
+(SRR19920084 and ERR9852351, respectively).
+
+\subsubsection{XBH}
+\label{sec:dd_xbh}
+All the XBH samples descend from a well supported recombination node,
+despite two reversions (T222C and T12789C, respectively),
+which occur at excluded sites and
+were remapped by parsimony during postprocessing.
+If these two sites had been used during ARG inference,
+the branch attached to BA.2.1 would have been attached to
+the BA.2.5 shown on the top left instead,
+leading to a more parsimonious local topology
+with two mutations on the left branch. This would also remove the two
+blue lines on the left and side of the copying pattern.
+
+\subsubsection{XBQ/XBK/XBK.1}
+\label{sec:dd_xbq}
+
+https://github.com/jeromekelleher/sc2ts-paper/issues/349
+https://github.com/jeromekelleher/sc2ts-paper/issues/350
+
+XBK and XBQ are siblings, and neither appear to be recombinants in the ARG,
+as they are closely related to the non-recombinant node that originates CJ.1.3. 
+At the time of designating XBK
+a hypothesis was made that a missing common ancestor could exist 
+(https://github.com/cov-lineages/pango-designation/issues/1440\#issuecomment-1355264973),
+making XBK non-recombinant.
+Exactly such a missing ancestor
+is found as the upper internal node labeled CJ.1 in the ARG,
+constructed by sc2ts as part of its tree building steps.
+
+\subsubsection{XBM}
+\label{sec:dd_xbm}
+
+All the XBM samples descend from a well supported recombination node.
+Its left parent is a reconstructed internal node
+that is directly supported by a BA.2.76-labeled sample identical to it.
+Its right parent is a BF.3-labeled sample.
+
+\subsubsection{XBR}
+\label{sec:dd_xbr}
+This one XBR sample descends from a well supported recombination node.
+Its left parent is a BN.3.1-labeled sample, and
+its right parent is a reconstructed internal node
+that is directly supported by a BQ.1.25-labeled sample identical to it.
+
+\subsection{Pango X lineages not present in the ARG}
 \label{sec:supplement_rejected_pango_x_lineages}
@@
 but XAY is more complex, possibly involving up to six breakpoints as per sc2ts.
-Exploring how to integrate these and other recombinants
-with complex recombination patterns \cite{Roemer2023}
-should be an important direction for future work.
 
@@
 \end{table}
-
-\subsection{Additional analyses of the Pango X lineages}
-\label{sec:supplement_indepth_pango_x_lineages}
-
-\newcommand{\scghdiscussion}[1]{%
-See \href{https://github.com/jeromekelleher/sc2ts-paper/discussions/#1}%
-{github:sc2ts-paper/discussions/#1} for further details.}
-
-\subsubsection{XA}
-% mneumonic: deep dive XA
-\label{sec:dd_xa}
-XA traces to a well supported recombination node, with no reversions.
-Note that the ``causal'' sample ERR5308556
-(which triggers the initial recombination event)
-is identical to the recombination node, and
-therefore lacks C8090T,
-which is one of the XA consensus mutations
-(identified as those shared by over 90\% of XA samples).
-As mentioned elsewhere,
-the sc2ts detection results in terms of
-the Pango lineages of the parents and the breakpoint intervals
-agree with the results reported by Jackson et al. (\ref{tab:jackson}).
-Two of the Jackson et al. XA samples
-are present in the ARG (ERR5308556 and ERR5414941);
-these samples are specifically marked in Figure~\ref{fig:subgraphs},
-which is a summary of this subgraph.
-
-\subsubsection{XB}
-\label{sec:dd_xb}
-
-The displayed subgraph shows a small selection of the 192 XB samples.
-XB was originally defined as a recombinant between
-the closely related lineages B.1.634 (left parent) and B.1.631 (right parent)
-from North and Central America \cite{Gutierrez2022}.
-However, these parent lineages were not added to the ARG due to a high HMM cost.
-Together with other lineages such as B.1.627
-(sister to B.1.631 \cite{Gutierrez2022}),
-they are likely to attach somewhere along the long branch
-in the subgraph visualization (with $> 30$ mutations),
-which also contains the key ``R2'' deletion
-associated with XB / B.1.634 \cite{Gutierrez2022}.
-
-We hypothesize that there are a set of samples
-from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
-which, should they be added to the ARG, would break up the long branch,
-with XB being a recombination between two of them.
-In the absence of these, sc2ts cannot detect XB as a recombinant.
-Also, note that the XB consensus mutations all occur on or above the long branch.
-We suspect that the child node of the long branch
-represents the (potentially recombinant) origin of XB, and
-the B.1 samples at the bottom left of the subgraph
-may be misclassified by pangolin.
-
-Further work is required to determine
-why these additional samples were not gradually added
-to the ARG by the daily sample matching process or
-retrospective sample grouping.
-
-\scghdiscussion{780}
-
-\subsubsection{XC}
-\label{sec:dd_xc}
-The five XC samples descend from a well supported recombination node.
-The two parents that are both identical to actual samples
-(ERR4908740 on the left and DRR321274 on the right).
-There is one reversion (G21987A) on the left side of the breakpoint,
-but the site 21987 was masked during ARG inference
-because it was identified to be experiencing
-frequent mutation in a preliminary ARG.
-
-\subsubsection{XE/XH}
-\label{sec:dd_xexh}
-
-The XE samples descend from a well supported recombination node,
-with one mutation and one reversion (T22792C).
-This reversion was mapped onto the ARG
-as a part of postprocessing,
-and could be placed more parsimoniously by a single C22792T
-that groups together XH
-and the lineage leading to the BA.2 recombination nodes
-displayed on the lower left.
-Nevertheless, the exact arrangement of the mutations at position 22792
-should not affect the proposed breakpoint,
-which is very distant, at position 11283.
-This breakpoint coincides with a repeated homopolymeric sequence
-involving 5-base deletions,
-so that alignment errors could have a small effect
-on the breakpoint position.
-Note that the two lower recombination nodes are both well supported,
-the youngest being postulated as the origin of XAF (see below);
-we leave investigation of this suggested series of
-sequential recombination events to future work.
-
-\scghdiscussion{633}
-
-\subsubsection{XF}
-\label{sec:dd_xf}
-XF is a well-supported recombinant between Delta and Omicron BA.1,
-with the breakpoint on the far left of the genome.
-All XF samples are the unique descendants of
-a well supported recombination node,
-with both parent nodes (ERR6193241: AY.4 and ERR8042040: BA.1) and
-the single child node (ERR8076129) being samples.
-
-\subsubsection{XG}
-\label{sec:dd_xg}
-The three XG samples descend from a reconstructed internal node,
-which in turn descends from a well supported recombination node.
-The left parent is a sample (SRR20179863), and
-the right parent is a reconstructed internal node
-assigned to Omicron BA.2,
-which is supported by a sample (ERR9124067) that is identical to it.
-
-\subsubsection{XJ}
-\label{sec:dd_xj}
-All the XJ samples descend from a well supported recombination node.
-There are 14 samples (including e.g. ERR8683190)
-which sc2ts identified as descendants of the same recombination node
-but are labeled BA.2 by pangolin (displayed on the left), and
-which do not possess the XJ consensus mutations C22792T and C27945T.
-These BA.2 samples could be worthy of further investigation.
-
-\subsubsection{XL}
-\label{sec:dd_xl}
-All the XL samples descend from a well supported recombination node,
-with three child samples which are identical to it.
-The left parent is a sample (ERR7989424), and
-the right parent is a reconstructed internal node
-assigned to BA.2.5 with a large number of descendant samples.
-
-\subsubsection{XM/XAL}
-\label{sec:dd_xm}
-
-XM is a recombinant of BA.1 (left parent) and BA.2 (right parent).
-Sc2ts identifies multiple recombinant origins of XM.
-One of these origins leads to a dominant clade
-(node 1003220 at position 21,595),
-which also contains the only three XAL samples.
-
-Independent evidence for multiple XM origins
-comes from consideration of a 9-base deletion
-in the Spike gene at position 21633,
-which is characteristic of BA.2 but not BA.1
-\url{https://github.com/cov-lineages/pango-designation/issues/361}.
-Most of the XM-labeled samples in ARG have this deletion,
-but three (SRR19024311, ERR9447529, and SRR18775874) do not,
-suggesting that XM samples trace to at least two recombination events,
-one to the right of the deletion
-(which leads to the majority of the XM samples), and one to the right.
-In the subgraph, it can be seen that one of the samples, SRR18775874,
-has been (presumably wrongly) placed
-within the main XM deletion-containing clade.
-As sc2ts does not use deletions for inference,
-deletion remapping (section XXXX) has reverted the deletion in this case,
-creating many insertions on the branch above the sample (magenta).
-A better use of deletions in sc2ts would likely result
-in placing this sample within one of the two XM clades
-which do not have the deletion.
-
-\scghdiscussion{656}
-
-\subsubsection{XN/XAU}
-\label{sec:dd_xn}
-
-In the sc2ts ARG, the initial XN sample (ERR8986821)
-is well explained as a non-recombinant,
-copying from an inferred internal node
-assigned to BA.2 with only one mutation.
-This internal node is directly supported by the sample ERR9502469.
-All the other XN samples descend from ERR8986821.
-A small clade containing XAU is the sister to XN, and
-there are a series of samples (ERR9453882, ERR9502469, and ERR9823912)
-which support the gradual emergence of both groups
-from a BA.2 ancestral lineage via accumulation of single mutations.
-
-We believe that the explanation of XN and XAU
-in the sc2ts ARG is correct, and
-the evidence supporting the original designation of
-XN and XAU as recombinants
-comes from inaccurate or problematic data sources.
-In the case of XN, support for a recombinant BA.1+BA.2 origin
-is from three mutations that are claimed to be BA-1-specific:
-C241T, A2832G, and C3037T
-(https://github.com/cov-lineages/pango-designation/issues/480).
-However, the BA.1/BA.2 split
-(https://github.com/cov-lineages/pango-designation/issues/361)
-does not list C241T and C3037T as BA.1-specific,
-meaning that these two mutations are not supportive of recombination.
-Moreover, although the remaining mutation
-claimed to support a recombinant origin, A2832G,
-is typical of BA.1 and not BA.2,
-it is seen in several BA.2 samples (e.g. ERR9453882),
-supporting the idea that it has independently arisen within BA.2.
-Hence, none of the sites used to designate XN as a recombinant reliably do so.
-
-In the case of XAU, a recombinant origin was proposed
-on the basis of the same three sites, plus C2470T
-(https://github.com/cov-lineages/pango-designation/issues/894).
-However, C2470T is not BA.1-specific
-(https://github.com/cov-lineages/pango-designation/issues/361).
-Moreover, some BA.2-designated samples such as ERR9823912 also have C2470T,
-so we do not believe this to be a recombinant-supporting mutation either.
-
-This case illustrates the value of automated ARG reconstruction
-for correcting the inevitable errors that arise
-when designating recombinants by hand and
-using different (and inconsistent) data sources of
-Pango lineage-associated mutational constellations.
-
-https://github.com/jeromekelleher/sc2ts-paper/issues/348
-https://github.com/jeromekelleher/sc2ts-paper/issues/358
-
-\subsubsection{XP}
-\label{sec:dd_xp}
-
-https://github.com/jeromekelleher/sc2ts-paper/issues/345
-
-This is a case where the Pango X lineage is designated
-on the basis of two loci on the far right of the genome:
-a single nucleotide mutation (A29510C) and
-a rare, large deletion at position 29734 in the S2M step loop
-(https://github.com/cov-lineages/pango-designation/issues/481),
-which is not common enough to map in sc2ts during postprocessing
-and hence not shown here.
-Sc2ts requires at least four loci on either side
-to infer a QC-passing recombinant (section XXXX).
-Therefore, unless this deletion were given more weight than a single locus,
-sc2ts would not identify XP as a recombinant,
-even if deletions were used during ARG inference.
-If this deletion were recurrent, the evidence from sc2ts suggests
-that XP might not be a recombinant,
-because it is only one mutation different from the sample ERR8628084.
-
-\subsubsection{XQ/XR/XU/XAA/XAG/XAM}
-\label{sec:dd_xq}
-
-https://github.com/jeromekelleher/sc2ts-paper/issues/338
-
-In the ARG, XR, XU, XAA, XAG, and XAM
-are nested under the XQ recombinant lineage.
-However, upon closer inspection,
-the right hand branch above the recombination node appears problematic,
-as it displays three reversion mutations,
-which revert mutations at positions 22792, 24023, and 25624
-in the ancestral lineage just above.
-These sites were excluded during ARG inference, and
-remapped onto the ARG by parsimony during ARG postprocessing.
-We posit that if these three sites were used during ARG inference,
-the right parent of the recombination node
-would be chosen to be the ancestral BA.2 sample
-on the right-hand path with 18 child nodes.
-A more parsimonious reconstruction
-would then have the XQ clade as a sibling to
-a group of (monophyletic) XAA, XAM, XAG, and XR lineages.
-This would remove between 4 and 5 reversion mutations.
-
-\subsubsection{XS}
-\label{sec:dd_xs}
-
-https://github.com/jeromekelleher/sc2ts-paper/issues/287
-
-In the ARG, the XS samples appear to have originated from
-two sequential recombination events (separated by a single branch).
-However, the later event is likely an artifact,
-because the XS sample (SRR18343220) associated with
-this event failed detection-level QC
-(i.e., the minimum net number of supporting loci;
-see the lower copying pattern).
-In addition, this sample was sequenced using the Ion AmpliSeq V1 kit,
-which was also used to sequence eight other samples
-descending from the same event
-(clustered due to a shared sequencing error due to AmpliSeq).
-These eight samples would likely attach to the nodes
-descending from the first recombination event.
-We excluded this likely artifactual event from further analysis and
-consider XS to have emerged from a single recombinant origin.
-
-\subsubsection{XW}
-\label{sec:dd_xw}
-
-All the XW samples descend from a well supported recombination node.
-Its left parent is a BA.1.1.15-labeled sample (ERR8069658), and
-its right parent is a BA.2-labeled sample (ERR9086669).
-
-\subsubsection{XY}
-\label{sec:dd_xy}
-
-All the XY samples descend from a well supported recombination node,
-through a shared reconstructed internal node.
-Its left parent is a reconstructed internal node assigned to BA.1.1,
-which is supported by a sample (ERR8627048) identical to it.
-Its right parent is a BA.2-labeled sample (ERR9399983).
-There are a pair of (orange) reversion mutations
-within the XY clade at position 23948:
-these were not used for inference
-but remapped using parsimony during postprocessing, and
-indicate that the topology within the XY clade
-could potentially be improved
-by alternative phylogenetic reconstruction methods.
-
-\subsubsection{Xx: XZ/XAC/XAD/XAE/XAP}
-\label{sec:dd_xx}
-
-https://github.com/jeromekelleher/sc2ts-paper/discussions/631
-
-This nested set of Pango X lineages descends
-from a single recombination node in the ARG, and
-These Pango X lineages
-may be aliases of a single recombinant
-which we tentatively refer to as ``Xx''.
-Note that all were designated around the same time,
-with their samples collected
-roughly within a three-month period (Mar., 2022 to Jul., 2022).
-Several samples labeled as BA.2 by pangolin also lie
-within this recombinant clade,
-including two that are identical
-to the proposed originating recombinant (ERR8146303 and ERR8163061).
-
-Exact relationships within the Xx clade are somewhat tentative.
-For example, there are three highly mutable sites (22792, 28877, and 28878)
-which were not used during inference,
-but were remapped onto the ARG during postprocessing.
-The inferred topology places these sites
-as recurrent (colored orange, grey, and faded yellow),
-but a more parsimonious arrangement would,
-for example, group the two XAD samples as siblings
-(removing the need to postulate mutations
-A1250G, A28877T, and G28878C to be recurrent).
-It is possible that a more parsimonious arrangement would also
-reduce the recurrent mutations and reversions seen at
-sites 22,792 (four orange mutations) and
-26,060 (two parallel brown mutations),
-but we do not show such a reconstruction,
-which is complicated by the presence of
-additional (non-displayed) samples descending from
-internal nodes in the plotted subgraph.
-
-For XAC, in particular, the Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/590) suggests
-it to be a recombinant with two breakpoints:
-BA.2 on the left, BA.1 in the middle, and
-a switch back to a very small region of BA.2 on the right.
-This contrasts with the one-breakpoint BA.2+BA.1 origin inferred in the ARG.
-However, only two loci support the right hand BA.2 assignment:
-A29510C and a large deletion
-(deletions are treated as missing data during ARG inference).
-Sc2ts does not count this as enough evidence
-to insert an additional breakpoint,
-which would require at least four supporting variants.
-
-\subsubsection{XAF}
-\label{sec:dd_xaf}
-
-https://github.com/jeromekelleher/sc2ts-paper/discussions/638
-
-Of the 21 XAF samples in the Viridian v04 dataset,
-only XAF sample passed our QC filters and was added to the ARG:
-many have greater than 500 missing Ns in their consensus sequences or
-require a large number of mutations to incorporate into the ARG
-(i.e. have a high HMM cost).
-However, we find a small recombinant BA1.1/BA.2 clade of 37 samples,
-associated with the BA.2-labeled sample ERR9755541,
-within which the single XAF sample is nested.
-The inferred parents and breakpoint at 10,447
-aligns with that proposed for XAF
-(https://github.com/cov-lineages/pango-designation/issues/676).
-Inspection of the omitted XAF samples with high HMM cost shows that
-they would attach as close sisters to the included XAF sample,
-albeit on branches with many new mutations.
-Hence we believe the current XAF topology in the ARG is reasonable,
-although somewhat tentative, given the number of included samples.
-
-The mutation-rich branch in the subgraph with
-a 5-base deletion (11283-11287) and insertion (11292-11296)
-is likely to be a simple deletion alignment error.
-
-\subsubsection{XAJ}
-\label{sec:dd_xaj}
-
-https://github.com/jeromekelleher/sc2ts-paper/discussions/634
-
-In the Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/826),
-XAJ is suggested to be a recombinant between
-BA.2.12.1 (left parent) and BA.4 (right parent).
-However, XAJ is not inferred to be a recombinant lineage in the ARG.
-From the subgraph, which shows the 18 XAJ samples
-plus a small number of relevant BA.4 samples,
-it can be seen that 9 sites leading to XAJ indeed recur
-in the lineage to the closest BA.4 sample (SRR18848050),
-which strongly indicates that this is a recombinant
-that is missed by sc2ts.
-The reason that is is not a recombinant in the ARG is
-that 6 sites represent a 6bp deletion and so are not used for inference.
-The remaining three sites (T22917G, T23018G, and G23040A)
-are not enough to exceed our k=4 threshold  for recombinant designation.
-Including deletions in inference would certainly
-lead to XAJ being detected as a recombinant,
-and with a lower threshold of k=3
-the distribution of mutations on the ARG topology suggests
-that it is is even possible that XAJ is a 2-breakpoint recombinant
-with the region from 21765-23535
-coming from the BA.4 labeled sample SRR18848050.
-
-\subsubsection{XAN/XAV}
-\label{sec:dd_xan}
-
-https://github.com/jeromekelleher/sc2ts-paper/discussions/783
-https://github.com/jeromekelleher/sc2ts-paper/discussions/785
-
-XAN and XAV are shown in this subgraph, as they share some mutations.
-In the Pango designation issues
-(https://github.com/cov-lineages/pango-designation/issues/771;
-https://github.com/cov-lineages/pango-designation/issues/911),
-XAN and XAV are suggested to be a recombinant between
-BA.2 (left parent) and BA.5 (right parent).
-Both these Pango lineages were detected as non-recombinants by sc2ts,
-because the number of supporting loci fell below our threshold of detection.
-In the ARG, we found only one mutation (C9866T)
-supporting the suggested BA.2 parent for both XAN and XAV.
-
-\subsubsection{XAS}
-\label{sec:dd_xas}
-
-https://github.com/jeromekelleher/sc2ts-paper/discussions/784
-
-XAS was proposed as a recombination between
-BA.5 (left parent) and BA.2 (right parent).
-However, we find that this is only two mutations
-different from e.g. ERR9929846, which is a BA.4 sample.
-In fact, XAS only differs from the BA.4 root by a single mutation,
-C27945T, which was identified in the original Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/882)
-as an important ORF truncation mutation
-which is also present in a few BA.2.65 samples.
-It is possible that C27945T was acquired by recombination,
-but this is clearly below our threshold of detection.
-
-\subsubsection{XAZ}
-\label{sec:dd_xaz}
-
-https://github.com/jeromekelleher/sc2ts-paper/discussions/786
-
-In the Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/797),
-XAZ is suggested as a two-breakpoint recombinant between
-BA.2.5 and BA.5 (the middle parent),
-with three sites supporting the BA.2.5 parent
-on the left end (C2232T, C3317T, and T3358C).
-In the ARG, this is very close to being a recombinant of
-BA.2.5 (ERR9615610) on the left side of the breakpoint,
-as not only are the three mutations present in this sample,
-but there is another site which supports recombination 
-C1912T, which is not reported in the Pango designation issue).
-XAZ has C44T which is lacking in all BA.2.5 samples,
-leaving only three net supporting loci for a putative recombinant.
-However, position 44 is often clipped,
-so this is not strong evidence against a recombinant, and
-therefore XAZ is a marginal case of recombination.
-We found no evidence for the third segment
-inherited from a BA2.5 parent on the right end.
-In the ARG, all the XAZ samples form a clade.
-Here we display only the first 20 XAZ samples.
-
-\subsubsection{XBB}
-\label{sec:dd_xbb}
-
-https://github.com/jeromekelleher/sc2ts-paper/issues/336
-
-Most of the XBB samples descend from a well supported recombination node.
-In the Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/1058),
-the Pango lineages of the suggested left and right parents are
-BJ.1 and BM.1.1.1, respectively.
-In the ARG, the Pango lineage of the right parent is identical,
-but that of the left parent is a related ancestral Pango lineage
-(BA.2.10 instead of BA.2.10.1.1 = BJ.1).
-This is likely due to the absence of BJ.1 in the ARG:
-only four BJ.1 samples exist in the Viridian v04 dataset,
-a single one passed sequence-level QC, and
-that was then rejected due to high HMM cost.
-Note that BJ.1 differs from BA.2.10 by over 20 mutations
-(https://github.com/cov-lineages/pango-designation/issues/915),
-explaining the large number of (gold) \emph{de novo} mutations
-in the XBB copying pattern.
-Had BJ.1 samples been present in the ARG,
-we strongly suspect that the left branch of
-the XBB-associated recombination node
-would have been attached to a BJ.1-labeled node, and
-the origin of XBB would not require postulating as many new mutations.
-The fact that sc2ts has deduced
-the XBB left parent to a close relative of BJ.1,
-even though BJ.1 is missing from the ARG,
-shows our approach to be relatively robust to missing un-sampled lineages:
-an important consideration given global variation in pathogen surveillance.
-
-\subsubsection{XBB.1}
-\label{sec:dd_xbb.1}
-
-XBB.1 is characterized by an additional mutation in Spike, G22317T (G252V) and
-is not designated as a recombinant of XBB in the Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/1088).
-The ARG agrees with this and places the characteristic mutation
-above the root of a single non-recombinant XBB.1 clade.
-We include this non-recombinant lineage
-in order to highlight some well supported subsequent recombination nodes
-within XBB.1, such as the (pink) one leading to two XBB.1-labeled samples
-at the bottom left of the subgraph which we refer to as XBB.x.
-XBB.x is well supported: its left parent is a reconstructed internal node,
-which is directly supported by a BQ.1-labeled sample, and
-its right parent is a BB.1.5.18-labeled sample (ERR10792363).
-This highlights the automatic ability of sc2ts
-to infer sequential recombination events
-(i.e. recombinants-of-recombinants),
-even those involving few samples.
-A few other recombination events occur within XBB.1,
-but they either fail QC checks, or
-(as the other one shown here, to the bottom right)
-lead to a single sample so may be
-e.g. caused by accidental mixing of samples during sequencing.
-
-\subsubsection{XBD}
-\label{sec:dd_xbd}
-
-All the XBD samples descend from a well supported recombination node.
-Its left and right parents are reconstructed internal nodes,
-both of which are directly supported by samples
-(SRR21382561 and ERR9825609, respectively)
-that are identical to them.
-
-\subsubsection{XBE}
-\label{sec:dd_xbe}
-
-https://github.com/jeromekelleher/sc2ts-paper/issues/351
-
-In the Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/1246),
-XBE was suggested as a recombinant between
-BA.5.2 (left parent; most of the genome) and
-BE.4 (right parent; somewhere from 23,609 onwards). 
-owever, sc2ts treats the 65 XBE samples
-as non-recombinant descendants of BA.5.2
-because the number of supporting sites
-fell below our threshold of detection.
-We have included the BE.4 lineage on the subgraph,
-revealing two mutations (C27800T, and G28681T)
-which occur in parallel above the root of the XBE clade
-and along the path ascending from the displayed BE.4 sample,
-suggesting that if we used k=2
-as the recombinant threshold,
-sc2ts would indeed detect XBE as a recombinant.
-
-\subsubsection{XBF}
-\label{sec:dd_xbf}
-
-All the XBF samples descend from a well supported recombination node,
-with three mutations on the left of the breakpoint.
-Its left parent is a sample assigned to BA.5.2.1 (SRR20305553).
-Its right parent is a reconstructed internal node
-that is supported by a CJ.1-labeled sample (SRR21948068),
-which differs by three mutations.
-
-\subsubsection{XBG}
-\label{sec:dd_xbg}
-
-All the XBG samples descend from a well supported recombination node,
-with two mutations, one on each side of the breakpoint.
-Both its left and right parents are samples
-(SRR19920084 and ERR9852351, respectively).
-
-\subsubsection{XBH}
-\label{sec:dd_xbh}
-All the XBH samples descend from a well supported recombination node,
-despite two reversions (T222C and T12789C, respectively),
-which occur at excluded sites and
-were remapped by parsimony during postprocessing.
-If these two sites had been used during ARG inference,
-the branch attached to BA.2.1 would have been attached to
-the BA.2.5 shown on the top left instead,
-leading to a more parsimonious local topology
-with two mutations on the left branch. This would also remove the two
-blue lines on the left and side of the copying pattern.
-
-\subsubsection{XBQ/XBK/XBK.1}
-\label{sec:dd_xbq}
-
-https://github.com/jeromekelleher/sc2ts-paper/issues/349
-https://github.com/jeromekelleher/sc2ts-paper/issues/350
-
-XBK and XBQ are siblings, and neither appear to be recombinants in the ARG,
-as they are closely related to the non-recombinant node that originates CJ.1.3. 
-At the time of designating XBK
-a hypothesis was made that a missing common ancestor could exist 
-(https://github.com/cov-lineages/pango-designation/issues/1440\#issuecomment-1355264973),
-making XBK non-recombinant.
-Exactly such a missing ancestor
-is found as the upper internal node labeled CJ.1 in the ARG,
-constructed by sc2ts as part of its tree building steps.
-
-\subsubsection{XBM}
-\label{sec:dd_xbm}
-
-All the XBM samples descend from a well supported recombination node.
-Its left parent is a reconstructed internal node
-that is directly supported by a BA.2.76-labeled sample identical to it.
-Its right parent is a BF.3-labeled sample.
-
-\subsubsection{XBR}
-\label{sec:dd_xbr}
-This one XBR sample descends from a well supported recombination node.
-Its left parent is a BN.3.1-labeled sample, and
-its right parent is a reconstructed internal node
-that is directly supported by a BQ.1.25-labeled sample identical to it.
 

### 2025-10-30 src/makepdf.py
Update to newest visualizer

@@ -32,9 +32,13 @@ def convert_file(infile, outdir):
             # Replace the javascript in the notebook code
             # to change mutation symbols and node labels to contain URLs.
             # This is a complete hack that fakes tooltips in a pdf, as URLS appear on hover.
+            add_mut_links = False
             for line in input.readlines():
                 ## Mutation symbols have a hover that shows the position and state change
-                if line.rstrip().endswith(r'var mut_symbol_rect = mut_symbol\n",'):
+                if "var mut_symbol = svg" in line:
+                    add_mut_links = True
+                if add_mut_links and ".enter()" in line:
+                    add_mut_links = False
                     start, end = line.rsplit(r'\n', 1)
                     output.write(start + r'.append(\"a\").attr(\"href\", d => \"mut:\" + d.label)\n' + end)
                 #  Line labels where the second line starts with (DRR, (ERR, or (SRR

### 2025-10-31 latex/main.tex
Updates from Overleaf

@@
 \section*{INTRODUCTION}
+% SHZ: The first paragraph should explain
+% why recombination "can" be important in pathogens.
+% Then, we explain why we need recombination-aware methods for pathogens.
+% This should set the stage for why this work is really important,
+% not just for SC2 but more broadly for the field of pathogen genomics.
+
 % Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
@@
 See Document~\ref{sec:supplement_indepth_pango_x_lineages}
-for extended analyses on all of the Pango X lineages,
+for extended analyses on all the Pango X lineages;
 Figure~\ref{fig:pango_x_copying_patterns} for copying patterns
-associated with Table~\ref{tab:pango_x_lineages}, and
-Document S2 for ARG visualizations of the Pango X lineages.
-[TODO ref Document \ref{sec:supplement_rejected_pango_x_lineages}
-to mention multi breakpoints and close issue 379]
+associated with Table~\ref{tab:pango_x_lineages};
+Document S2 for ARG visualizations of the Pango X lineages; and
+% TODO: ref Document \ref{sec:supplement_rejected_pango_x_lineages}
+% to mention multi breakpoints and close issue 379
+Document \ref{sec:supplement_rejected_pango_x_lineages}
+for an analysis of the Pango X lineages
+which were detected as multiple-breakpoint recombinants by sc2ts
+but were not added to the ARG
+due to high HMM costs and/or poor sample support.
 
@@
 We explored two hypotheses for the timing of recombination.
-First, we would expect recombination events to occur more often when case counts are high,
-which increases the chance of co-infection within an individual, allowing recombination.
+First, we would expect recombination events
+to occur more often when case counts are high,
+which increases the chance of co-infection within an individual,
+allowing recombination.
 Second, we would expect recombination events to be detected
@@
 during the Delta and Omicron waves, when case counts were high.
-Overall, the number of recombinant events rose with the global number of cases
+Overall, the number of recombination events rose with the global case count
 (Figure~\ref{fig:recombinants_over_time}B; 
@@
 % this has more.
-
+% SHZ: This makes it sound like we are just making analyses
+% (which people can do already using existing methods and tools)
+% easier to do
+% rather than making a leap forward from anything else out there.
 We present the first pandemic-scale ancestral recombination graph (ARG) for
@@
 % existing estimates already tell us that
-% recombination in SARS-CoV-2 is (qualitatively) rare,
+% detectable recombination in SARS-CoV-2 is (qualitatively) rare,
 % even without the advances of sc2ts.
@@
 % we are actually dealing with.
-% Clarifying the relationships between Pango Xs make this possible.
+% Clarifying the relationships between Pango Xs helps with this task.
 
@@
 % Also, it's not only about selection favoring new combinations of mutations in Spike.
-% It's about combinations of mutations in, say, ORF1ab, getting selected against.
+% It's about combinations of mutations in, say, ORF1ab, getting selected against
+% (or just low mutation rates).
 % Maybe mention epistasis?
@@
 
+%%% WIP: SHZ
+We present the first pandemic-scale
+ancestral recombination graph (ARG) for SARS-CoV-2,
+reconstructed from $\sim$2.48 million genomes using \textit{sc2ts}.
+We demonstrate how this ARG, for the first time,
+enables systematic identification and characterization of
+recombination events in fine-grained
+detail--in terms of their parents, breakpoints, and timing of
+emergence--providing richer and more nuanced
+genealogical, genomic, and epidemiological context
+than existing phylogenetic approaches.
+Together, sc2ts and the ARG represent a new paradigm in
+how we study the evolution of pathogens in the face of ongoing sampling.
+
+Accounting for recombination is a key challenge in pathogen genomics, and
+recombination has been largely ignored in analysis of viral genomes.
+Existing phylogenetic approaches for pathogens detect recombination
+\textit{post hoc} (e.g., UShER+RIPPLES),
+treat it as a nuance (e.g., XX), or
+discard its signals from downstream analyses (e.g., XX).
+Sc2ts addresses this challenge
+by using the ARG as the underlying data model,
+explicitly representing recombination events as individual nodes and
+relating them to specific parents in the ARG.
+The ARG provides a coherent framework to determine and represent
+the relationships between recombinants and their specific parents
+(which are not possible using a tree-based data model,
+e.g., the Mutation Annotated Tree format of UShER).
+The ARG reveals and clarifies the relationships between
+individual samples of the same recombinant origin,
+avoiding issues such as over-naming recombinants
+(e.g., the samples assigned to
+five differently labeled recombinant Pango lineages
+descending from the “Xx” recombination event in the ARG).
+Our results show how an ARG can considerably simplify
+efforts to identify and monitor emerging recombinants,
+facilitating communication of genomic surveillance results.
+
+% SHZ: A bit vague, but is this worth discussing?
+% What else to say about the timing of these events?
+% TODO: Example with say XBB?
+% TODO: Say we don’t explore this fully, but it could be a follow-up study?
+The timing of recombination events is of key interest in
+pathogen genomic epidemiology,
+because this information may yield insights into
+% Probably A bad idea to say this.
+how the emergence of a new recombinant might be
+linked with changes in public health policies.
+Using the ARG, we reconstructed a historical timeline of
+recombination events during the COVID-19 pandemic,
+showing that it is possible to pinpoint the occurrence of
+recombination events precisely in time,
+within windows of X to X months
+(given the intensity of sampling in the COVID-19 dataset).
+
+The overall frequency of detectable recombination events
+estimated from the sc2ts ARG is substantially lower than
+that from UShER+RIPPLES,
+largely reflecting differences in model formulation and parameter choice.
+When applied \textit{post hoc} to an UShER phylogeny of
+$\sim$1.6 million genomes sampled up to May, 2021,
+RIPPLES reported 589 unique recombination events with 43,104 descendant
+samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
+Using the same time interval, sc2ts \(k=4\) identifies
+14 high-confidence recombination events with
+172 descendant samples among 501,235 samples \(0.034\%\),
+a 40- to 80-fold reduction.
+Much of this difference is explained by
+the stricter parsimony penalty used here:
+increasing the penalty from p=3 (as used in the RIPPLES analysis) to
+k=4 greatly reduces false positives
+while retaining nearly all verified recombinant lineages,
+including those missed by RIPPLES at equivalent thresholds.
+These indirect comparisons suggest that
+our estimate represents a conservative lower bound.
+In future studies, dissecting the methodological differences
+between sc2ts and UShER+RIPPLES may
+inspire new ideas to develop or improve ARG building methods
+for SARS-CoV-2 (and potentially other pathogens).
+
+The ARG itself is a powerful resource
+to support future studies of SARS-CoV-2.
+Our results show that the ARG captures and unifies
+important features of SARS-CoV-2 evolution,
+which are shown in different studies
+using various methods, tools, and approaches.
+The ARG also compactly encodes genetic variation in
+millions of SARS-CoV-2 genomes,
+allowing for convenient and efficient reuse of
+this valuable dataset for a range of
+genetic and evolutionary analyses beyond those presented here,
+for instance, to understand
+how the variability in sequencing efforts
+across time and location affects our ability
+to reconstruct the origin of important recombinant lineages and
+to estimate when they emerge.
+Because building the ARG is computationally expensive, and
+checking and improving various aspects of its quality is laborious,
+we hope that making this resource openly available will
+save significant time and effort for other researchers,
+enabling immediate focus on analyzing the information
+contained within the COVID-19 dataset.
+% TODO: Briefly discussion software tools here or in a subsequent paragraph.
+
+Recent developments in ARG methods have focused almost exclusively on
+applications to human sequencing data.
+The availability of human ARGs reconstructed from large datasets
+has already had a tremendous impact on improving
+our understanding of human genetic ancestry and past demography.
+We anticipate similar advances to be made
+by adapting existing inference methods for human ARGs
+to use the sc2ts ARG for the genealogy-based study of
+natural selection or the geographic spread of SARS-CoV-2 through time.
+
+The unprecedented throughput and pace of genome sequencing are
+creating ample opportunity to understand the impacts of recombination
+on pathogen evolution and epidemiology.
+We hope that this study will spur
+development of ARG methods for pathogens,
+ultimately making recombination-aware genome analysis
+routine for all sorts of pathogens beyond SARS-CoV-2.
+
 \subsection*{Limitations of the study}
@@
 
-[TODO]
-
-We close by emphasizing several current limitations and potential directions to improve sc2ts.
-First, more stringent filters to exclude false positive recombinants during inference
-would obviate post-hoc treatment of these recombinants.
-Second, indels carry important phylogenetic signals,
-for example, as defining mutations of Alpha and Delta (REF).
-We currently treat multi-base deletions as missing data or multiple independent deletions,
-but they are in reality single events.
-More sophisticated treatments of multi-base indels in the HMM could help recover additional signals
+We close by emphasizing several current limitations
+as well as potential directions to improve sc2ts,
+which we hope will also inspire future research
+to develop ARG methods for other pathogens.
+
+Deletions which are reverted can be strong signals of recombination,
+because bases which are lost cannot be easily regained,
+except through recombination involving
+a lineage that has bases at the deleted sites.
+We currently treat multi-base deletions as missing data.
+More sophisticated treatments of such indels in the HMM
+may help recover additional signals
 (and avoid false signals) of recombination.
-Third, the current HMM assumes a simplistic mutation model,
-which does not account for documented transition-to-transversion rate bias (REF)
-and context-dependent variation in mutation rate (REF).
-A more realistic mutation model may help improve inference of local genetic relationships in the ARG.
-Fourth, an inherent limitation of detecting recombination from sequence data,
-which applies to all methods, is that it is impossible to detect events with few (or no) supporting bases,
+
+The current HMM assumes a simplistic mutation model
+(where mutation to any other allele is equally probable) and
+implicitly assumes a uniform mutation rate along the SARS-CoV-2 genome.
+However, genome analyses have revealed
+transition-to-transversion rate bias in SARS-CoV-2 \cite{Bloom2023} and
+sequence context-dependent variation
+in the mutation rate along the SARS-CoV-2 genome (REF).
+A more realistic mutation model may help improve
+inference of local genetic relationships in the ARG, and
+accounting for mutation rate heterogeneity
+during sample matching or in the HMM cost
+(e.g., down-weighing sites in highly mutable regions)
+may help reduce false positive recombinants.
+
+We developed and used several methods
+to exclude poor-quality recombination events post hoc, including
+filtering by supporting loci per parent and
+long-branch splitting.
+These methods could instead be run immediately
+after sample matching during ARG inference,
+obviating the need for post-processing of the ARG.
+
+An inherent limitation of detecting recombination from sequence data,
+which applies to all methods,
+is that it is impossible to detect events with few (or no) supporting bases,
 as these cannot be distinguished from mutation events.
-This limitation makes it particularly challenging to detect events near the termini of linear viral genomes.
-Fifth, it is challenging to accurately reconstruct global events
-when there are regions with poor sampling and jurisdictions with long delays
-before sequences become publicly available (REF).
+This limitation makes it particularly challenging
+to detect events near the termini of linear viral genomes.
+
+It is challenging to accurately reconstruct global events
+when there are regions with poor sampling and jurisdictions
+with long delays before sequences become publicly available (REF).
 Revisiting the ARG topology periodically
-(e.g., rewinding the ARG a month and repeating with samples that became available later)
-and better handling of putative recombinants with suspiciously long branches
-(e.g., Delta and Omicron BA.2) might improve the accuracy of the genealogy.
+(e.g., rewinding the ARG a month and repeating
+with samples that became available later) and
+better handling of putative recombinants
+with suspiciously long branches (e.g., Delta and Omicron BA.2)
+might improve the accuracy of the genealogy.
+
+% TODO: Add a short discussion of the node dates and
+% some references to the literature, closing issue 725.
+
 Finally, future developments may account for geography in the HMM,
 favoring attachment of samples to geographically close lineages.
-
-TODO: add a short discussion of the node dates and some references
-to the literature, closing issue 725
 
@@
 independently for each recombinant. Specifically, we 
-ran \texttt{3seq -full  parents.fa recombinant.fa} using default 
-thresholds.
+ran \texttt{3seq -full parents.fa recombinant.fa}
+% TODO: Should we mention the default p-value threshold is 0.05?
+using default thresholds.
 We used 3SEQ source code downloaded from \url{https://gitlab.com/lamhm/3seq}
@@
 Here, we estimate the interval lengths expected around a breakpoint,
-given a genome-wide nucleotide substitution rate $\mu$,
+given a genome-wide nucleotide substitution rate ($\mu$),
 total time between two samples ($T$), and
 sufficient support to detect the recombination event.
-If a breakpoint occurs a fraction $P$ along a genome of length $L$ and
+If a breakpoint occurs a fraction $P$
+along a genome of length $L$ and
 $k$ mutations have occurred to the left of the breakpoint,
-then we expect the nearest mutation to occur $P L / (k + 1)$ bases to the left.
+then we expect the nearest mutation
+to occur $P L / (k + 1)$ bases to the left.
 If we require $n$ mutations on both sides of the breakpoint,
@@
 by summing across a Poisson-distributed number of mutations,
-conditioned on having at least $n$ mutations:
+conditioned on having at least $n$ mutations,
+using the following equation
+(where $\Gamma(x)$ is the gamma function, and
+$\Gamma(x, y)$ is the upper incomplete gamma function):
 
@@
 
-To explore the timing of emergence of recombinants,
+To explore the timing of emergence of recombination events,
 case count data were downloaded from
 the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE)
-at Johns Hopkins University \cite{Dong2020} and binned by week from January, 2020 to March, 2023.
+at Johns Hopkins University \cite{Dong2020} and
+binned by week from January, 2020 to March, 2023.
 The proportion of each lineage $i$ in each week ($p_i$) was then inferred
@@
 from that week had different Scorpio designations ($\sum_{i} \sum_{j>i} 2 p_i p_j$).
-The linear regression models reported in \textbf{Figure} \ref{fig:recombinants_over_time} allowed for a non-zero intercept,
+The linear regression models reported in \textbf{Figure} \ref{fig:recombinants_over_time}
+allowed for a non-zero intercept,
 because there may be unreported cases even without any reported cases.
@@
 the p-values were confirmed by randomly permuting the predictor variable 10,000 times.
-ANOVAs were conducted to determine which single predictor ($n$, $H$, $n \; H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week.
+ANOVAs were conducted to determine which single predictor
+($n$, $H$, $n \; H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week.
 These statistical tests and graphics were conducted
+% TODO: The link!
 in \textit{Mathematica} 14.2 (Wolfram 2024) and available at GITHUB LINK.
 
-In addition to the analysis of the 354 recombination events 
-that pass QC,
+In addition to the analysis of the 354 recombination events that pass QC,
 we analyzed the relationship between the full set of 855 recombinants and case numbers.
-These additional recombinants occurred disproportionately during the Delta and first Omicron wave (July, 2021 to February, 2022).
-Again, the number of recombinant events rose with the global number of cases
+These additional recombinants occurred disproportionately
+during the Delta and first Omicron wave (July, 2021 to February, 2022).
+Again, the number of recombinant events rose with the global case count
 (linear regression: $3.1 + 5.3 \; 10^{-7} x$, $p = 0.0006$)
@@
 (linear regression: $0.42 + 18.3 x$, $p = 2 \; 10^{-12}$).
-The single best predictor of recombination events among the predictors tested ($n$, $H$, $n \; H$, $n^2$, $n^2 H$)
+The single best predictor of recombination events
+among the predictors tested ($n$, $H$, $n \; H$, $n^2$, $n^2 H$)
 was again the product of cases and diversity ($n \; H$),
@@
 Thus, similar conclusions are reached with either dataset,
-although filtering the recombination events to the 354 with better support revealed a stronger relationship with case numbers.
+although filtering the recombination events to the 354
+with better support revealed a stronger relationship with case numbers.
 
@@
 
-\subsection{Pango X lineages not present in the ARG}
+\subsection{Pango X lineages not added to the ARG}
 \label{sec:supplement_rejected_pango_x_lineages}

### 2025-10-31 latex/main.tex
Updates on complex recombinants
Closes #366
Closes #379

@@
 handling deletions or matching against long branches (Document \ref{sec:dd_xb}).
+
+The remaining 10 X lineages (97 samples) present in the dataset are not 
+included in the ARG because the HMM cost exceeds the inclusion threshold
+and there was insufficient support for retrospective inclusion (STAR Methods).
+These include ``complex recombinants'' such as XBC and XAY~\cite{Roemer2023}
+which are confirmed by sc2ts to have multiple breakpoints; see 
+Document \ref{sec:supplement_rejected_pango_x_lineages} and Table 
+\ref{tab:extra_hmm_results} for details.
 See Document~\ref{sec:supplement_indepth_pango_x_lineages}
-for extended analyses on all the Pango X lineages;
+for extended analyses on all the Pango X lineages in the ARG;
 Figure~\ref{fig:pango_x_copying_patterns} for copying patterns
 associated with Table~\ref{tab:pango_x_lineages};
-Document S2 for ARG visualizations of the Pango X lineages; and
-% TODO: ref Document \ref{sec:supplement_rejected_pango_x_lineages}
-% to mention multi breakpoints and close issue 379
-Document \ref{sec:supplement_rejected_pango_x_lineages}
-for an analysis of the Pango X lineages
-which were detected as multiple-breakpoint recombinants by sc2ts
-but were not added to the ARG
-due to high HMM costs and/or poor sample support.
+and Document S2 for ARG visualizations of the Pango X lineages.
 
@@
 
+% 2 + 8 + 19+ 1 + 11+ 2 + 15+ 36+ 2 + 1
+% 97
+
 \subsection{Analysis of the Jackson et al. recombinants}

### 2025-10-31 latex/main.tex
Updates from Overleaf

--- 2025-10-31 latex/main.tex
@@
 However, genome analyses have revealed
-transition-to-transversion rate bias in SARS-CoV-2 \cite{Bloom2023} and
+transition-to-transversion rate bias \cite{Bloom2023,Simmonds2024} and
 sequence context-dependent variation
-in the mutation rate along the SARS-CoV-2 genome (REF).
+in the mutation rate in the SARS-CoV-2 genome \cite{Haddox2025}.
 A more realistic mutation model may help improve
@@
 
-We developed and used several methods
-to exclude poor-quality recombination events post hoc, including
-filtering by supporting loci per parent and
-long-branch splitting.
+Finally, future developments may account for geography in the HMM,
+favoring attachment of samples to geographically close lineages.
+
+% TODO: Add a short discussion of the node dates and
+% some references to the literature, closing issue 725.
+
+It is challenging to accurately reconstruct global events
+when there are regions with poor sampling and
+jurisdictions with long delays
+before sequences become publicly available (REF).
+For example, revisiting the ARG topology periodically
+(e.g., rewinding the ARG a month and repeating
+with samples that became available later)
+might improve the accuracy of the genealogy.
+
+Several methods are developed and used
+to exclude probable false positive recombination events \textit{post hoc},
+including filtering by supporting loci per parent and
+splitting of suspiciously long branches.
 These methods could instead be run immediately
@@
 
-An inherent limitation of detecting recombination from sequence data,
+Finally, an inherent limitation of detecting recombination from sequence data,
 which applies to all methods,
@@
 to detect events near the termini of linear viral genomes.
-
-It is challenging to accurately reconstruct global events
-when there are regions with poor sampling and jurisdictions
-with long delays before sequences become publicly available (REF).
-Revisiting the ARG topology periodically
-(e.g., rewinding the ARG a month and repeating
-with samples that became available later) and
-better handling of putative recombinants
-with suspiciously long branches (e.g., Delta and Omicron BA.2)
-might improve the accuracy of the genealogy.
-
-% TODO: Add a short discussion of the node dates and
-% some references to the literature, closing issue 725.
-
-Finally, future developments may account for geography in the HMM,
-favoring attachment of samples to geographically close lineages.
 
@@
 
-SHZ is supported by an Intermediate Research Fellowship
-from the Nuffield Department of Population Health,
-University of Oxford.
+SHZ is supported by an NDPH Intermediate Research Fellowship
+from Oxford Population Health, University of Oxford.
 
@@
 
-S.Y.W. is an employee and shareholder of COMPANY.
+% TODO: Close https://github.com/jeromekelleher/sc2ts-paper/issues/922
+The authors declare no competing interests.
 

### 2025-10-31 latex/main.tex
Updates from Overleaf

--- 2025-10-31 latex/main.tex
@@
 
-%%% WIP: SHZ
+%%% SHZ: WIP alternative.
 We present the first pandemic-scale
@@
 the relationships between recombinants and their specific parents
-(which are not possible using a tree-based data model,
-e.g., the Mutation Annotated Tree format of UShER).
+(which are not possible using a tree-based data model, e.g., the MAT format of UShER).
 The ARG reveals and clarifies the relationships between
@@
 
-% SHZ: A bit vague, but is this worth discussing?
-% What else to say about the timing of these events?
-% TODO: Example with say XBB?
-% TODO: Say we don’t explore this fully, but it could be a follow-up study?
+% SHZ: Very WIP
+% Discuss the new preprint from Turakhia.
 The timing of recombination events is of key interest in
-pathogen genomic epidemiology,
+pathogen genomic epidemiology.
 because this information may yield insights into
-% Probably A bad idea to say this.
 how the emergence of a new recombinant might be
 linked with changes in public health policies.
-Using the ARG, we reconstructed a historical timeline of
-recombination events during the COVID-19 pandemic,
-showing that it is possible to pinpoint the occurrence of
-recombination events precisely in time,
-within windows of X to X months
-(given the intensity of sampling in the COVID-19 dataset).
 
@@
 contained within the COVID-19 dataset.
+
 % TODO: Briefly discussion software tools here or in a subsequent paragraph.

### 2025-10-31 latex/main.tex
Updates from Overleaf

@@
 which we hope will also inspire future research
-to develop ARG methods for other pathogens.
-
-Deletions which are reverted can be strong signals of recombination,
-because bases which are lost cannot be easily regained,
-except through recombination involving
-a lineage that has bases at the deleted sites.
-We currently treat multi-base deletions as missing data.
-More sophisticated treatments of such indels in the HMM
-may help recover additional signals
-(and avoid false signals) of recombination.
+to develop ARG methods for SARS-CoV-2 and other pathogens.
+
+Reversions of deletions can indicate recombination,
+because bases that are lost are unlikely to be regained,
+except via recombination with a lineage that has those bases.
+We presently treat multi-base deletions as missing data.
+Explicitly modeling deletions in the HMM
+could help recover additional signals of recombination
+and reduce false positives.
 
@@
 implicitly assumes a uniform mutation rate along the SARS-CoV-2 genome.
-However, genome analyses have revealed
+However, genomic analyses have revealed
 transition-to-transversion rate bias \cite{Bloom2023,Simmonds2024} and
-sequence context-dependent variation
-in the mutation rate in the SARS-CoV-2 genome \cite{Haddox2025}.
-A more realistic mutation model may help improve
-inference of local genetic relationships in the ARG, and
+sequence context-dependent mutation rate variation
+across the SARS-CoV-2 genome \cite{Haddox2025}.
+Using a more realistic mutation model and
 accounting for mutation rate heterogeneity
-during sample matching or in the HMM cost
-(e.g., down-weighing sites in highly mutable regions)
-may help reduce false positive recombinants.
-
-Finally, future developments may account for geography in the HMM,
+could improve
+inference of local genetic relationships in the ARG and
+reduce spurious recombinants.
+
+Future developments may account for geography in the HMM,
 favoring attachment of samples to geographically close lineages.
@@
 % some references to the literature, closing issue 725.
+
+We have developed and used filters
+to detect time traveling samples in the Viridian dataset and
+to exclude them from the ARG,
+yielding a set of samples with reliable collection dates.
+Future studies could leverage the ARG
+to flag wrongly dated samples or
+to infer the dates of samples
+with missing or incomplete collection dates in the metadata.
 
@@
 jurisdictions with long delays
-before sequences become publicly available (REF).
+before sequences become publicly
+available \cite{Lancet2021,Kalia2021,Chen2022}.
 For example, revisiting the ARG topology periodically
@@
 
-Several methods are developed and used
-to exclude probable false positive recombination events \textit{post hoc},
+We have developed and used several methods
+to exclude spurious recombination events \textit{post hoc},
 including filtering by supporting loci per parent and
-splitting of suspiciously long branches.
+splitting of suspiciously long branches (e.g., Delta and Omicron BA.2).
 These methods could instead be run immediately
 after sample matching during ARG inference,
-obviating the need for post-processing of the ARG.
+reducing the number of ARG post-processing steps.
 

### 2025-11-03 latex/main.tex
Notes on the discussion

@@
 Schematic overview of sc2ts.
-(A) Process to build an ARG by incrementally updating it with samples 
-sorted by collection dates.
+(A) Process to build an ARG by incrementally updating it
+with samples sorted by collection dates.
 (B) The ARG resulting from the inference process.
 Sample 0 is used to initialize the ARG as the root.
-Next, Sample 1 is attached, mapping mutations onto the new branch to explain it.
+Next, Sample 1 is attached,
+mapping mutations onto the new branch to explain it.
 When Sample 4 is detected as a recombinant,
-a node representing the recombination event is inserted with two edges 
-connecting to its parents,
+a node representing the recombination event is inserted with
+two edges connecting to its parents,
 with Sample 4 itself attached to this node as its child.
@@
 one on each side of the inferred breakpoint which occurs at position 50.
-The genomic coordinates of the segments inherited from different recombinant parents
+The genomic coordinates of the segments
+inherited from different recombinant parents
 are denoted by two half-open intervals.
@@
 
-Sc2ts reconstructs an ARG by incrementally adding daily batches of 
-sequences (Figure~\ref{fig:methodology}).
-First, sc2ts infers likely ``copying paths'' connecting each sample in the 
-batch to node(s) in the current ARG
-using a simplified version of the Li and Stephens (LS) model\cite{LiStephens2003}. 
+Sc2ts reconstructs an ARG by incrementally adding
+daily batches of sequences (Figure~\ref{fig:methodology}).
+First, sc2ts infers likely ``copying paths'' connecting each sample
+in the batch to node(s) in the current ARG
+using a simplified version of the Li and Stephens (LS) model~\cite{LiStephens2003}.
 % Important to go into a bit of detail about LS here as it's so fundamental
@@
 The LS model is a Hidden Markov Model (HMM) widely used in human genomics
-to approximate the effects of mutation and recombination on genetic
-inheritance\citep{Delaneau2019-wl,Browning2018-nk},
-typically parameterised by population-scaled mutation and
+to approximate the effects of mutation and recombination
+on genetic inheritance~\cite{Delaneau2019-wl,Browning2018-nk},
+typically parameterized by population-scaled mutation and
 recombination rates varying along the genome~\cite{Donnelly2010-coalescent}.
-In the interest of interpretability, we simplified this model to 
-use a single parameter $k$ which essentially 
-controls the number of recurrent mutations that is preferred to 
-switching to a different parent via recombination 
+In the interest of interpretability, we simplified this model
+to use a single parameter $k$
+which controls the number of recurrent mutations
+that is preferred to switching to a different parent via recombination
 % This is quite a big thing really, and ultimately what determines
@@
 (we use $k=4$ here; see STAR Methods and Discussion).
-We used the highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
-with some additional optimisations enabled by our simplified model,
+We used the highly efficient LS HMM implementation in tsinfer~\cite{Kelleher2019},
+with some additional optimizations enabled by our simplified model,
 to find exact Viterbi solutions of these copying paths for each sample
 given the current ARG.
-Note that this is a form of parsimony, with the recombination
-penalty $k$ defining the relative parsimoniousness of recombinant
-vs non-recombinant solutions to the HMM.
-The per-sample copying paths produced by the HMM (most of which do not 
-involve recombination) form clusters of samples within a batch, which 
-are then resolved using standard phylogenetic techniques.
+Note that this is a form of parsimony,
+with the recombination penalty $k$
+defining the relative parsimoniousness of
+recombinant versus non-recombinant solutions to the HMM.
+The per-sample copying paths produced by the HMM
+(most of which do not involve recombination)
+form clusters of samples within a batch,
+which are then resolved using standard phylogenetic techniques.
 Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
@@
 building on core PyData infrastructure~\cite{Harris2020array,Virtanen2020scipy,
-Mckinney2010data,Lam2015numba} and bioinformatics 
-packages\cite{Shirley2015efficient,Kunzmann2018biotite,Kunzmann2023biotite}.
+Mckinney2010data,Lam2015numba} and
+bioinformatics packages~\cite{Shirley2015efficient,Kunzmann2018biotite,Kunzmann2023biotite}.
 See STAR Methods for details on all aspects of the method.
@@
 Using sc2ts, we built an ARG from 2,482,157 samples
-from the Viridian v04 dataset \cite{Hunt2024},
+from the Viridian v04 dataset~\cite{Hunt2024},
 collected between 2020-01-01 and 2023-02-20.
-Inference took 21 days on a server with 64 cores (2X AMD EPYC 7502)
-and 512 GB of RAM ($\sim$5 CPU years; Figure~\ref{fig:inference_resources}).
-After postprocessing, the ARG
-contains 2,689,054 nodes,  % Note the estimated dates ?
-2,748,838 edges, and 2,285,344 mutations at 29,893 sites,
-and is stored in a 32MiB file.
+Inference took 21 days on a server with
+64 cores (2X AMD EPYC 7502) and 512 GB of RAM
+($\sim$5 CPU years; Figure~\ref{fig:inference_resources}).
+After post-processing, the ARG contains
+2,689,054 nodes,  % Note the estimated dates ?
+2,748,838 edges, and
+2,285,344 mutations
+at 29,893 sites,
+and is stored in a 32 MB file.
 Of the nodes, 855 represent recombination events,
@@
 % NOTE: if we keep this we should redo on Yan's laptop and quote those numbers
-Using the tskit Python API, the ARG can be loaded into 
-memory in less than a second [Concrete numbers?], and the 
-vectorised representation integrates tightly with 
-powerful data-science tools such as NumPy~\cite{Harris2020array}.
-For example, a tabular summary for the 2.75m nodes, including
-sample IDs and Pango assignments (stored as metadata in the ARG),
-can be derived as a Pandas dataframe~\cite{Mckinney2010data} in 0.6s.
+Using the tskit Python API,
+% TODO: Concrete numbers?
+the ARG can be loaded into memory in less than a second,
+and the vectorized representation integrates tightly
+with powerful data science tools, such as NumPy~\cite{Harris2020array}.
+For example, a tabular summary for the 2.75 million nodes,
+including sample IDs and Pango assignments (stored as metadata in the ARG),
+can be derived as a Pandas data frame~\cite{Mckinney2010data} in 0.6 second.
 The ARG can therefore be easily accessed using familiar tools
@@
 
-% NOTE: Trying to clarify that this of the pandemic period and we're 
-% not trying to cover *all* history.
-The ARG is a comprehensive and deeply validated record of the pandemic phase 
-of SARS-CoV-2 evolutionary history, accurately reflecting
-a range of results from the literature.
-First, we compared the sc2ts ARG and with the state-of-the-art UShER 
-tree inferred from the same dataset
-by pruning both to their intersection of 2,475,418 samples and 27,507 sites
-(STAR Methods)
+% NOTE: Trying to clarify that this of the pandemic period and
+% we're not trying to cover *all* history.
+The ARG is a comprehensive and deeply validated record of
+the pandemic phase of SARS-CoV-2 evolutionary history,
+accurately reflecting a range of results from the literature.
+First, we compared the sc2ts ARG
+with the state-of-the-art UShER tree inferred from the same dataset
+by pruning both to their intersection of
+2,475,418 samples and 27,507 sites (STAR Methods).
 % JK: this feels a bit wishy washy, but there really is no simple answer to
 % this and I don't want to claim we're more parsimonious.
-and found they were very similar in terms of parsimony
+We found that they were very similar in terms of parsimony
 (Document~\ref{sec:supplement_sc2ts_usher_parsimony})
@@
 We also compared the mutational spectra of major VOCs in the ARG
-against the UShER tree and 
-a previous study \cite{Bloom2023} (STAR Methods),
+against the UShER tree and a previous study \cite{Bloom2023} (STAR Methods),
 finding close agreement (Figure~\ref{fig:mutational_spectra}).
-We validated that the broad phylogenetic structure of the ARG and 
-UShER tree closely match by generating tanglegrams for a set of representative
-samples from 814 well-sampled lineages (STAR Methods; 
-Figures~\ref{fig:tanglegram_base},
+We validated that the broad phylogenetic structure of
+the ARG and UShER tree closely match by generating tanglegrams
+for a set of representative samples from 814 well-sampled lineages
+(STAR Methods; Figures~\ref{fig:tanglegram_base},
 \ref{fig:tanglegram_delta},
@@
 \ref{fig:tanglegram_ba5}).
-We also evaluated how well the ARG reflects the structure
-of the Pango nomenclature system by computing
-Pango assignments for all 2.75 million nodes in the ARG using
-Pangolin~\cite{OToole2021assignment} (STAR Methods).
+We also evaluated how well the ARG reflects the structure of
+the Pango nomenclature system
+by computing Pango assignments for all 2.75 million nodes in the ARG
+using Pangolin~\cite{OToole2021assignment} (STAR Methods).
 All but 21,680 of the exported sample alignments (0.87\%) agreed
@@
 A majority of these lineages (1473 of 2058)
-have monophyletic origins within the ARG,
+have monophyletic origins in the ARG,
 increasing to 1779 (86\%) when we allow
@@
 (Document~\ref{sec:supplement_pango_lineages}).
-The ARG is time-resolved, with internal node dates estimated by
-a non-parametric method~\cite{Pope2025} (STAR Methods). 
-We show that the dates agree reasonably with NextStrain 
-estimates (Figure \ref{fig:pango_vs_nextstrain_node_times})
+The ARG is time-resolved, with internal node dates estimated
+by a non-parametric method~\cite{Pope2025} (STAR Methods).
+We show that the dates agree reasonably well with
+NextStrain estimates (Figure~\ref{fig:pango_vs_nextstrain_node_times})
 and overall accuracy comparable to 
@@
 Omicron (Figure~\ref{fig:subgraph_omicron}, Document~\ref{sec:dd_omicron})
-VOCs reflect expectations from the literature\cite{RambautAlpha2020,ECDCreport,
+VOCs reflect expectations from the literature~\cite{RambautAlpha2020,ECDCreport,
 Cherian2021,Farinholt2021,Stern2021,Dor2025,RambautOmicron2021}.
 
-Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
-but are not incorporated into UShER trees and must therefore
-be accounted for separately.
+Deletions are an important part of SARS-CoV-2 evolution~\cite{Jeronimo2023},
+but are not incorporated into UShER trees and
+therefore must be accounted for separately.
 Because deletions are difficult to distinguish
+% SHZ: Sequencing errors or misleading alignment.
 from bioinformatics errors
@@
 To approximate the effects of the most important deletions into the ARG,
-we remapped data using post hoc parsimony at 75 sites involved in deletions
-with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
+we remapped data using \textit{post hoc} parsimony at 75 sites
+involved in deletions with a frequency of
+$\geq 1\%$ reported by Li et al.\cite{Li2023indels}
 during post-processing (STAR Methods).
@@
 recurrent deletions (Table~\ref{tab:rec_dels}).
-For example, the three characteristic deletions of Alpha \cite{RambautAlpha2020}
+For example,
+the three defining deletions of Alpha \cite{RambautAlpha2020}
 (11288:9, 21765:6, 21991:3)
@@
 in the N-terminal domain of the S1 unit of Spike
-(which is a deletion hotspot \cite{McCarthy2021})
+(which is a deletion hot spot \cite{McCarthy2021}),
 for example,
@@
 
-Sequencing errors in genome assemblies \cite{Turakhia2020},
-read misalignments \cite{Hunt2024}, and
-multi-nucleotide substitutions \cite{DeMaio2020issues}
+Sequencing errors in genome assemblies~\cite{Turakhia2020},
+incorrect read alignments~\cite{Hunt2024}, and
+multi-nucleotide substitutions~\cite{DeMaio2020issues}
 can cause short regions of a sample genome to match more closely
@@
 71 (14\%) are associated with 6-base deletion in the Delta lineage,
-known to be affected by amplicon dropout\cite{Sanderson2021}.
+known to be affected by amplicon dropout~\cite{Sanderson2021}.
 The 354 remaining recombination events that pass this QC filter
@@
 
-We validated the recombination events in the sc2ts ARG by running 
-3SEQ~\cite{Lam2018improved} on the putative recombinant 
-sequence trios (STAR Methods), 
+We validated the recombination events in the sc2ts ARG
+by running 3SEQ~\cite{Lam2018improved}
+on the putative recombinant sequence trios (STAR Methods),
 and 484 of the 855 (57\%) were classified as recombinant.
-Validation status correlates strongly with our QC filter, with 
-95\% of the QC-passing recombinants 
+Validation status correlates strongly with our QC filter,
+with 95\% of the QC-passing recombinants
 and 29\% of the QC-failing recombinants
@@
 In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
-and rebar recombination detection methods on the putative recombinant 
-sequences, summarized in Table~\ref{tab:recombinant_validation}.
-CovRecomb has a substantially lower validation rate at 25\% 
-among the QC-passing recombinants. This is expected, as 
-CovRecomb and is specifically intended to detect interlineage recombinants
+and rebar recombination detection methods
+on the putative recombinant sequences,
+summarized in Table~\ref{tab:recombinant_validation}.
+CovRecomb has a substantially lower validation rate
+at 25\% among the QC-passing recombinants.
+This is expected, as CovRecomb is specifically intended
+to detect interlineage recombinants
 and has previously been shown to be
@@
 the evidence in favor of recombination.
+% TODO
 [FIXME need to insert a sentence saying this is similar to
@@
 \caption{
-Properties of recombination events. A,B,C 
-show summaries of the 354 sc2ts recombination nodes
-passing quality control filters
+Properties of recombination events.
+A,B,C show summaries of the 354 sc2ts recombination nodes passing QC filters
 colored by 3SEQ classification (recombinant: purple, non-recombinant: gray).
@@
 recombination nodes in Table~\ref{tab:pango_x_lineages} are labeled starting with `X' and
-those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
-Jackson group B (JB) and Jackson singleton recombinant 1 (J1).
+those identified by Jackson et al.~\cite{Jackson2021} are prefixed with a `J',
+e.g., Jackson group B (JB) and Jackson singleton recombinant 1 (J1).
 Point sizes reflect number of descendants.
@@
 (C) Distribution of the Pango lineage distance between parents.
-(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
-for XA, XW and XBB.}
+(D) Copying patterns
+(STAR Methods, Figure~\ref{fig:copying_pattern_examples}) for XA, XW and XBB.}
 \label{fig:recombinant_evidence}
@@
 
+% SHZ: We should be more careful with the wording.
+% Recombination is likely frequent,
+% but detectable recombination
+% leading to lineages that are competitive enough to be sampled
+% compared to their (VOC) parents
+% is rare.
 Recombination is rare in SARS-CoV-2, and
 ultimately judgments about plausibility require expert analysis.
-To aid such interpretation and help identify quality control issues
+To aid such interpretation and help identify QC issues,
 we developed a visualization tool
@@
 Each copying pattern shows the positions
-where the allelic state in a recombinant
-(middle row) matches that of the left (upper row, colored green) 
+where the allelic state
+in a recombinant (middle row) matches that of
+the left (upper row, colored green) 
 or right parent (lower row, colored blue),
-or where the recombination event requires a \emph{de novo} mutation
-(gold, with mutational change below). [should we also mention the red/orange lines under the patterns that identify adjacent sites?]
+or where the recombination event requires
+a \textit{de novo} mutation (gold, with mutational change below).
+[FIXME: Should we also mention the red/orange lines
+under the patterns that identify adjacent sites?]
 Figure~\ref{fig:recombinant_evidence}D shows some example copying patterns,
-illustrating this spectrum of plausibility.
-XA is an example of a highly plausible recombinant, shown by the 
-large number of supporting loci on the left and right. As we have 
-only one \emph{de novo} mutation, we can be sure that the parents are closely
-observed [Some details about XA here, how well the parents were sampled etc].
+illustrating a spectrum of plausibility.
+XA is an example of a highly plausible recombinant,
+shown by the large number of supporting loci on the left and right.
+As we have only one \textit{de novo} mutation,
+we can be sure that the parents are closely observed
+[FIXME: Some details about XA here, how well the parents were sampled, etc.].
 XBB is also strongly supported by many loci on the left and right, 
-but the presence of many \emph{de novo} mutations
+but the presence of many \textit{de novo} mutations
 observed in the copying patterns suggests the existence of
-many genetically distinct, unsampled ancestors 
+many genetically distinct, un-sampled ancestors
 [FIXME: in this case, probably BJ.1, see issue 336],
-and some caution should be exercised in interpreting them. 
+and some caution should be exercised when interpreting them.
 See Figures~\ref{fig:recombinant_evidence}D and
@@
 \section*{DISCUSSION}
+[NOTE: this is two different WIP versions of the Discussion. 
+The ChatGPT based stuff was only ever intended as a starting point,
+and not to be taken literally.]
 
@@
 
+[NOTE: This is Shing's WIP alternative]
+
 %%% SHZ: WIP alternative.
@@
 
+% SHZ: Should we do point form?
 We close by emphasizing several current limitations
@@
 
-Reversions of deletions can indicate recombination,
+Reversion of deletions can indicate recombination,
 because bases that are lost are unlikely to be regained,
@@
 We presently treat multi-base deletions as missing data.
-Explicitly modeling deletions in the HMM
+Modeling deletions (and more generally insertions) in the HMM
 could help recover additional signals of recombination
@@
 accounting for mutation rate heterogeneity
-could improve
-inference of local genetic relationships in the ARG and
+could improve inference of local genetic relationships in the ARG and
 reduce spurious recombinants.
@@
 
-Requests for further information and resources should be directed to and will be fulfilled by the lead contact, XX (XX@university.edu).
+Requests for further information and resources
+should be directed to and will be fulfilled
+by the lead contact, XX (XX@university.edu).
 
@@
 
-This study did not generate new materials
+This study did not generate new materials.
 
@@
 SHZ is supported by an NDPH Intermediate Research Fellowship
-from Oxford Population Health, University of Oxford.
+(Oxford Population Health, University of Oxford).
 
@@
 Similarly to the Mutation Annotated Tree (MAT)~\cite{McBroome2021} format 
-(used 
-by UShER\cite{Turakhia2021}, matOptimize~\cite{DeMaio2023}, 
-RIPPLES~\cite{Turakhia2022} and 
+(used by
+UShER\cite{Turakhia2021},
+matOptimize~\cite{DeMaio2023},
+RIPPLES~\cite{Turakhia2022}, and
 RIVET~\cite{Smith2023tracking}),
-tskit incorporates mutational information directly in the data 
-model. 
-This is done using the site (defining the genome position and 
-ancestral state) and mutation (defining the site, node, and derived state)
-tables. In addition, tskit provides a flexible system for attaching metadata
+tskit incorporates mutational information directly in the data model.
+This is done using
+the site (defining the genome position and ancestral state) and
+mutation (defining the site, node, and derived state)
+tables.
+In addition, tskit provides a flexible system for attaching metadata
 to all aspects of the data model, which can either be in JSON format
 for flexibility (used heavily by sc2ts for debugging information)
-or in a vectorisable packed binary format for efficiency (used to attach
-sample identifiers and pango assignments in the final ARGs). 
-Tskit also provides a ``provenance'' table, which provides a mechanism
-for recording software version and parameter information through each
-step of complex processing pipelines.
+or in a vectorisable packed binary format for efficiency
+(used to attach sample identifiers and pango assignments in the final ARGs).
+Tskit also provides a ``provenance'' table,
+which provides a mechanism for recording software version and
+parameter information through each step of complex processing pipelines.
 
@@
 To explore the timing of emergence of recombination events,
-case count data were downloaded from
-the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE)
+global case count data were downloaded from
+the COVID-19 Data Repository by the Center
+for Systems Science and Engineering (CSSE)
 at Johns Hopkins University \cite{Dong2020} and
 binned by week from January, 2020 to March, 2023.
-The proportion of each lineage $i$ in each week ($p_i$) was then inferred
-from the Scorpio designation at each node within the ARG for that week.
-Genetic variation was measured as the ``expected heterozygosity'' across lineages,
+The proportion of each lineage $i$ in each week ($p_i$)
+was then inferred from the Scorpio designation
+at each node within the ARG for that week.
+Genetic variation was measured
+as the ``expected heterozygosity'' across lineages,
 which is the chance that two lineages chosen at random
-from that week had different Scorpio designations ($\sum_{i} \sum_{j>i} 2 p_i p_j$).
-The linear regression models reported in \textbf{Figure} \ref{fig:recombinants_over_time}
+from that week had different Scorpio designations
+($\sum_{i} \sum_{j>i} 2 p_i p_j$).
+The linear regression models reported in
+Figure~\ref{fig:recombinants_over_time}
 allowed for a non-zero intercept,
@@
 The slopes were highly significant, and
-the p-values were confirmed by randomly permuting the predictor variable 10,000 times.
+the p-values were confirmed
+by randomly permuting the predictor variable 10,000 times.
 ANOVAs were conducted to determine which single predictor
-($n$, $H$, $n \; H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week.
+($n$, $H$, $n \; H$, $n^2$, $n^2 H$)
+explained the most variation in recombination events by week.
 These statistical tests and graphics were conducted
 % TODO: The link!
-in \textit{Mathematica} 14.2 (Wolfram 2024) and available at GITHUB LINK.
+in \textit{Mathematica} 14.2 (Wolfram 2024) and
+available at GITHUB LINK.
 

### 2025-11-03 latex/main.tex
Tidy up supp typesetting

@@
 comprehensive phylogenies of these datasets are typically built with
-simpler distance methods [citation?]
+simpler distance methods % citation for this? 
 or approximations~\cite{Price2009fasttree},
@@
 The benefits of this simple model and the value of
-a comprehensive genealogy updated in real-time~\cite{McBroome2021}
+a comprehensive genealogy updated in real-time~\cite{McBroome2021,Kramer2023}
 were conclusively demonstrated, as the resource
@@
 in real-time. Based on the Viridian dataset, a freely available  
-resource purged of various systematic errors and artifacts \cite{Hunt2024},
+resource purged of various systematic errors and artifacts~\cite{Hunt2024},
 we infer an ARG with 2.48 million whole genome sequences.
 We show that this ARG is highly congruent with the UShER phylogeny built 
-from the same dataset as well as the Pango nomenclature system\cite{Rambaut2020},
+from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020},
 and also accurately captures evolutionary features such as mutational 
-spectra\cite{Bloom2023} and deletions\cite{Li2023indels}.
+spectra~\cite{Bloom2023} and deletions\cite{Li2023indels}.
 We show that sc2ts automatically and accurately detects details about
@@
 sc2ts with the state-of-the-art combination of UShER and 
-RIPPLES ~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
+RIPPLES~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
 detecting more of the well-characterized recombination events 
@@
 This ARG, coupled with the rich suite of supporting software
-in the tskit\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
+in the tskit~\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
 enables pandemic-scale analysis on a laptop. 
@@
 We also compared the mutational spectra of major VOCs in the ARG
-against the UShER tree and a previous study \cite{Bloom2023} (STAR Methods),
+against the UShER tree and a previous study~\cite{Bloom2023} (STAR Methods),
 finding close agreement (Figure~\ref{fig:mutational_spectra}).
@@
 Chronumental~\cite{sanderson2021chronumental} applied to 
-an UShER tree (Document \ref{sec:evaluating_node_dates}).
+an UShER tree (Document~\ref{sec:evaluating_node_dates}).
 Finally, we verified
@@
 involved in deletions with a frequency of
-$\geq 1\%$ reported by Li et al.\cite{Li2023indels}
+$\geq 1\%$ reported by Li et al.~\cite{Li2023indels}
 during post-processing (STAR Methods).
@@
 For example,
-the three defining deletions of Alpha \cite{RambautAlpha2020}
+the three defining deletions of Alpha~\cite{RambautAlpha2020}
 (11288:9, 21765:6, 21991:3)
@@
 in the N-terminal domain of the S1 unit of Spike
-(which is a deletion hot spot \cite{McCarthy2021}),
+(which is a deletion hot spot~\cite{McCarthy2021}),
 for example,
@@
 which causes S-Gene Target Failure \cite{Walker2021} and
-is associated with increased efficiency in cell entry \cite{Meng2021}, and
+is associated with increased efficiency in cell entry~\cite{Meng2021}, and
 the 21991:3 deletion
-which is associated with antibody escape \cite{McCarthy2021}.
+which is associated with antibody escape~\cite{McCarthy2021}.
 See Document~\ref{sec:supplement_dels} for further analysis.
@@
 (A) Subgraph around the recombination node associated with XA.
-Two ``Group A'' recombinant samples reported by Jackson et al.\cite{Jackson2021} are shown.
+Two ``Group A'' recombinant samples
+reported by Jackson et al.~\cite{Jackson2021} are shown.
 (B) Subgraph around the recombination node associated with a 
@@
 The first recombinant lineage XA was designated in 2021
-based on the analysis of Jackson et al.\cite{Jackson2021},
+based on the analysis of Jackson et al.~\cite{Jackson2021},
 which established that recombination with onward transmission had occurred.
@@
 with similar performance to
-the RecombinHunt \cite{Alfonsi2024}
-and CovRecomb \cite{Li2024CovRecomb} detection methods
+the RecombinHunt~\cite{Alfonsi2024}
+and CovRecomb~\cite{Li2024CovRecomb} detection methods
 (STAR Methods; Table~\ref{tab:method_concordance}; see also
@@
 that they are more parsimoniously explained as sublineages of XQ
-than de-novo recombinant lineages (Document \ref{sec:dd_xq}). 
+than de-novo recombinant lineages (Document~\ref{sec:dd_xq}). 
 The second cluster is based around a putative undocumented
@@
 XZ, XAC, XAD, XAE, and XAP, which we refer to as ``Xx''
-(Figure~\ref{fig:subgraphs}B, Document \ref{sec:dd_xx}).
+(Figure~\ref{fig:subgraphs}B, Document~\ref{sec:dd_xx}).
 
@@
 distant from a BA.2 sample and has little additional evidence
-to suggest recombination (Document \ref{sec:dd_xn}).
+to suggest recombination (Document~\ref{sec:dd_xn}).
 XB, on the other hand,
-has been thoroughly characterized \cite{Gutierrez2022} and
+has been thoroughly characterized~\cite{Gutierrez2022} and
 may have misclassified by sc2ts due to limitations in
@@
 examining the parental and recombinant sequences.
-Some methods \cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}
+Some methods~\cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}
 and approaches (e.g., Pango designation) infer this
@@
 consistent with the patterns seen in previous
-studies \cite{Bobay2020recombination,Turakhia2022,
+studies~\cite{Bobay2020recombination,Turakhia2022,
 Smith2023tracking,Li2024CovRecomb}.
@@
 Global case numbers were underreported,
-particularly during the Delta and Omicron waves \cite{Wang2025}.
+particularly during the Delta and Omicron waves~\cite{Wang2025}.
 Furthermore, there may be substantial uncertainty in the timing of a recombination event,
@@
 {github:sc2ts-paper/discussions/#1} for further details.}
-
+\newcommand{\scghdiscussiontwo}[2]{%
+See \href{https://github.com/jeromekelleher/sc2ts-paper/discussions/#1}%
+{github:sc2ts-paper/discussions/#1} for further details on #2.}
 \subsubsection{XA}
@@
 
-We believe that the explanation of XN and XAU
-in the sc2ts ARG is correct, and
-the evidence supporting the original designation of
-XN and XAU as recombinants
-comes from inaccurate or problematic data sources.
+The evidence supporting the original designation of
+XN and XAU as recombinants may come
+from inaccurate or problematic data sources.
 In the case of XN, support for a recombinant BA.1+BA.2 origin
@@
 C241T, A2832G, and C3037T
-(https://github.com/cov-lineages/pango-designation/issues/480).
+(\url{https://github.com/cov-lineages/pango-designation/issues/480}).
 However, the BA.1/BA.2 split
-(https://github.com/cov-lineages/pango-designation/issues/361)
+(\url{https://github.com/cov-lineages/pango-designation/issues/361})
 does not list C241T and C3037T as BA.1-specific,
@@
 on the basis of the same three sites, plus C2470T
-(https://github.com/cov-lineages/pango-designation/issues/894).
+(\url{https://github.com/cov-lineages/pango-designation/issues/894}).
 However, C2470T is not BA.1-specific
-(https://github.com/cov-lineages/pango-designation/issues/361).
+(\url{https://github.com/cov-lineages/pango-designation/issues/361}).
 Moreover, some BA.2-designated samples such as ERR9823912 also have C2470T,
@@
 
-This case illustrates the value of automated ARG reconstruction
-for correcting the inevitable errors that arise
-when designating recombinants by hand and
-using different (and inconsistent) data sources of
-Pango lineage-associated mutational constellations.
-
-https://github.com/jeromekelleher/sc2ts-paper/issues/348
-https://github.com/jeromekelleher/sc2ts-paper/issues/358
+\scghdiscussiontwo{781}{XN}
+
+\scghdiscussiontwo{928}{XAU}
 
@@
 \label{sec:dd_xp}
-
-https://github.com/jeromekelleher/sc2ts-paper/issues/345
 
@@
 a rare, large deletion at position 29734 in the S2M step loop
-(https://github.com/cov-lineages/pango-designation/issues/481),
+(\url{https://github.com/cov-lineages/pango-designation/issues/481}),
 which is not common enough to map in sc2ts during postprocessing
 and hence not shown here.
-Sc2ts requires at least four loci on either side
-to infer a QC-passing recombinant (section XXXX).
 Therefore, unless this deletion were given more weight than a single locus,
@@
 
+\scghdiscussion{775}
+
 \subsubsection{XQ/XR/XU/XAA/XAG/XAM}
 \label{sec:dd_xq}
-
-https://github.com/jeromekelleher/sc2ts-paper/issues/338
 
@@
 remapped onto the ARG by parsimony during ARG postprocessing.
-We posit that if these three sites were used during ARG inference,
+If these three sites were used during primary ARG inference,
 the right parent of the recombination node
-would be chosen to be the ancestral BA.2 sample
+may have been chosen to be the ancestral BA.2 sample
 on the right-hand path with 18 child nodes.
@@
 
+\scghdiscussion{655}
+
 \subsubsection{XS}
 \label{sec:dd_xs}
-
-https://github.com/jeromekelleher/sc2ts-paper/issues/287
 
@@
 
+\scghdiscussion{782}
+
 \subsubsection{XW}
@@
 
-https://github.com/jeromekelleher/sc2ts-paper/discussions/631
-
 This nested set of Pango X lineages descends
-from a single recombination node in the ARG, and
-These Pango X lineages
-may be aliases of a single recombinant
+from a single recombination node in the ARG,
 which we tentatively refer to as ``Xx''.
@@
 roughly within a three-month period (Mar., 2022 to Jul., 2022).
-Several samples labeled as BA.2 by pangolin also lie
+Several samples labeled as BA.2 by Pangolin also lie
 within this recombinant clade,
@@
 For XAC, in particular, the Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/590) suggests
+(\url{https://github.com/cov-lineages/pango-designation/issues/590}) suggests
 it to be a recombinant with two breakpoints:
@@
 A29510C and a large deletion
-(deletions are treated as missing data during ARG inference).
+(deletions are treated as missing data during primary ARG inference).
 Sc2ts does not count this as enough evidence
-to insert an additional breakpoint,
-which would require at least four supporting variants.
+to insert an additional breakpoint.
+
+\scghdiscussion{631}
 
@@
 
-https://github.com/jeromekelleher/sc2ts-paper/discussions/638
-
 Of the 21 XAF samples in the Viridian v04 dataset,
-only XAF sample passed our QC filters and was added to the ARG:
-many have greater than 500 missing Ns in their consensus sequences or
-require a large number of mutations to incorporate into the ARG
-(i.e. have a high HMM cost).
+only one XAF sample passed our QC filters and was added to the ARG.
+% many have greater than 500 missing Ns in their consensus sequences or
+% require a large number of mutations to incorporate into the ARG
+% (i.e. have a high HMM cost).
 However, we find a small recombinant BA1.1/BA.2 clade of 37 samples,
@@
 aligns with that proposed for XAF
-(https://github.com/cov-lineages/pango-designation/issues/676).
+(\url{https://github.com/cov-lineages/pango-designation/issues/676}).
 Inspection of the omitted XAF samples with high HMM cost shows that
@@
 albeit on branches with many new mutations.
-Hence we believe the current XAF topology in the ARG is reasonable,
-although somewhat tentative, given the number of included samples.
+% Hence we believe the current XAF topology in the ARG is reasonable,
+% although somewhat tentative, given the number of included samples.
 
@@
 a 5-base deletion (11283-11287) and insertion (11292-11296)
-is likely to be a simple deletion alignment error.
+may be due to a simple deletion alignment error.
+
+\scghdiscussion{638}
 
@@
 
-https://github.com/jeromekelleher/sc2ts-paper/discussions/634
-
 In the Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/826),
+(\url{https://github.com/cov-lineages/pango-designation/issues/826}),
 XAJ is suggested to be a recombinant between
@@
 
+\scghdiscussion{634}
+
 \subsubsection{XAN/XAV}
@@
 
-https://github.com/jeromekelleher/sc2ts-paper/discussions/783
-https://github.com/jeromekelleher/sc2ts-paper/discussions/785
-
 XAN and XAV are shown in this subgraph, as they share some mutations.
 In the Pango designation issues
-(https://github.com/cov-lineages/pango-designation/issues/771;
-https://github.com/cov-lineages/pango-designation/issues/911),
+(\url{https://github.com/cov-lineages/pango-designation/issues/771};
+\url{https://github.com/cov-lineages/pango-designation/issues/911}),
 XAN and XAV are suggested to be a recombinant between
@@
 
+\scghdiscussiontwo{783}{XAN}
+
+\scghdiscussiontwo{785}{XAV}
+
 \subsubsection{XAS}
 \label{sec:dd_xas}
-
-https://github.com/jeromekelleher/sc2ts-paper/discussions/784
 
@@
 C27945T, which was identified in the original Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/882)
+(\url{https://github.com/cov-lineages/pango-designation/issues/882})
 as an important ORF truncation mutation
@@
 
+\scghdiscussion{784}
+
 \subsubsection{XAZ}
@@
 
-https://github.com/jeromekelleher/sc2ts-paper/discussions/786
-
 In the Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/797),
+(\url{https://github.com/cov-lineages/pango-designation/issues/797}),
 XAZ is suggested as a two-breakpoint recombinant between
@@
 
+\scghdiscussion{786}
+
 \subsubsection{XBB}
@@
 
-https://github.com/jeromekelleher/sc2ts-paper/issues/336
-
 Most of the XBB samples descend from a well supported recombination node.
 In the Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/1058),
+(\url{https://github.com/cov-lineages/pango-designation/issues/1058}),
 the Pango lineages of the suggested left and right parents are
@@
 Note that BJ.1 differs from BA.2.10 by over 20 mutations
-(https://github.com/cov-lineages/pango-designation/issues/915),
-explaining the large number of (gold) \emph{de novo} mutations
+(\url{https://github.com/cov-lineages/pango-designation/issues/915}),
+explaining the large number of (gold) de novo mutations
 in the XBB copying pattern.
 Had BJ.1 samples been present in the ARG,
-we strongly suspect that the left branch of
+the left branch of
 the XBB-associated recombination node
-would have been attached to a BJ.1-labeled node, and
-the origin of XBB would not require postulating as many new mutations.
-The fact that sc2ts has deduced
-the XBB left parent to a close relative of BJ.1,
-even though BJ.1 is missing from the ARG,
-shows our approach to be relatively robust to missing un-sampled lineages:
-an important consideration given global variation in pathogen surveillance.
-
-\subsubsection{XBB.1}
-\label{sec:dd_xbb.1}
-
-XBB.1 is characterized by an additional mutation in Spike, G22317T (G252V) and
-is not designated as a recombinant of XBB in the Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/1088).
-The ARG agrees with this and places the characteristic mutation
-above the root of a single non-recombinant XBB.1 clade.
-We include this non-recombinant lineage
-in order to highlight some well supported subsequent recombination nodes
-within XBB.1, such as the (pink) one leading to two XBB.1-labeled samples
-at the bottom left of the subgraph which we refer to as XBB.x.
-XBB.x is well supported: its left parent is a reconstructed internal node,
-which is directly supported by a BQ.1-labeled sample, and
-its right parent is a BB.1.5.18-labeled sample (ERR10792363).
-This highlights the automatic ability of sc2ts
-to infer sequential recombination events
-(i.e. recombinants-of-recombinants),
-even those involving few samples.
-A few other recombination events occur within XBB.1,
-but they either fail QC checks, or
-(as the other one shown here, to the bottom right)
-lead to a single sample so may be
-e.g. caused by accidental mixing of samples during sequencing.
+may have attached to a BJ.1-labeled node, and
+the origin of XBB would required fewer mutations.
+
+\scghdiscussion{929}
+
+%% JK: commented this out as it's confusing the relationship between XBB.1 and
+%% XBB.x. Note that most methods pick out XBB.x as recombinant, so it's not that 
+%% remarkable.
+
+% \subsubsection{XBB.1}
+% \label{sec:dd_xbb.1}
+
+% XBB.1 is characterized by an additional mutation in Spike, G22317T (G252V) and
+% is not designated as a recombinant of XBB in the Pango designation issue
+% (https://github.com/cov-lineages/pango-designation/issues/1088).
+% The ARG agrees with this and places the characteristic mutation
+% above the root of a single non-recombinant XBB.1 clade.
+% We include this non-recombinant lineage
+% in order to highlight some well supported subsequent recombination nodes
+% within XBB.1, such as the (pink) one leading to two XBB.1-labeled samples
+% at the bottom left of the subgraph which we refer to as XBB.x.
+% XBB.x is well supported: its left parent is a reconstructed internal node,
+% which is directly supported by a BQ.1-labeled sample, and
+% its right parent is a BB.1.5.18-labeled sample (ERR10792363).
+% This highlights the automatic ability of sc2ts
+% to infer sequential recombination events
+% (i.e. recombinants-of-recombinants),
+% even those involving few samples.
+% A few other recombination events occur within XBB.1,
+% but they either fail QC checks, or
+% (as the other one shown here, to the bottom right)
+% lead to a single sample so may be
+% e.g. caused by accidental mixing of samples during sequencing.
 
@@
 
-https://github.com/jeromekelleher/sc2ts-paper/issues/351
-
 In the Pango designation issue
-(https://github.com/cov-lineages/pango-designation/issues/1246),
+(\url{https://github.com/cov-lineages/pango-designation/issues/1246}),
 XBE was suggested as a recombinant between
@@
 BE.4 (right parent; somewhere from 23,609 onwards). 
-owever, sc2ts treats the 65 XBE samples
+However, sc2ts treats the 65 XBE samples
 as non-recombinant descendants of BA.5.2
@@
 sc2ts would indeed detect XBE as a recombinant.
+
+\scghdiscussion{930}
 
@@
 
-https://github.com/jeromekelleher/sc2ts-paper/issues/349
-https://github.com/jeromekelleher/sc2ts-paper/issues/350
-
 XBK and XBQ are siblings, and neither appear to be recombinants in the ARG,
@@
 a hypothesis was made that a missing common ancestor could exist 
-(https://github.com/cov-lineages/pango-designation/issues/1440\#issuecomment-1355264973),
+(\url{https://github.com/cov-lineages/pango-designation/issues/1440\#issuecomment-1355264973}),
 making XBK non-recombinant.
@@
 constructed by sc2ts as part of its tree building steps.
+
+\scghdiscussiontwo{931}{XBQ}
+
+\scghdiscussiontwo{932}{XBK}
 

### 2025-11-03 latex/main.tex
Updates from Overleaf

--- 2025-11-03 latex/main.tex
@@
 
-To validate the accuracy of times for non-sample internal nodes inserted 
-into the ARG
-(inferred using tsdate), we compared the dates of first emergence of
-Nextstrain clades against the Nextstrain SARS-CoV-2 tree (downloaded on 
-2023-01-21). Figure \ref{fig:pango_vs_nextstrain_node_times} shows relatively
-close agreement for most clades, with only three 
-disagreeing by more than 28 days: 20C (an early lineage from 2020), 20E (Pango B.1.177 also known as EU1, emerging mid-2020), and 21B (Pango B.1.617.1 also known as Kappa, emerging in late 2020 to early 2021).
-
-Then, to gauge internal consistency of the time estimation procedure, we
-selected a subset of sample nodes, matching the distribution of node times 
-and average number of children of the non-sample internal nodes. These
-selected sample nodes were then re-dated, and the inferred dates compared 
-against their recorded collection dates. The results in Figure 
-\ref{fig:node_time_testing} demonstrate that although the inferred and true
-times are generally very close, there is a clear bias with re-estimated times
-being consistently older (by an average of 37 days). This is in line with 
-the findings of other recent
-work \cite{sanderson2021chronumental}, and is likely to be a consequence
-of edges carrying more mutations than would be expected under a neutral
-molecular clock prior (either due to errors in tree-building or genuinely varying mutation rates), as well as inaccuracies in the recording of 
-sample dates.
+To validate the accuracy of times for non-sample internal nodes
+inserted into the ARG (inferred using tsdate),
+we compared the dates of first emergence of
+Nextstrain clades against the Nextstrain SARS-CoV-2 tree
+(downloaded on 2023-01-21).
+Figure~\ref{fig:pango_vs_nextstrain_node_times} shows
+relatively close agreement for most clades,
+with only three disagreeing by more than 28 days:
+20C (an early lineage from 2020),
+20E (Pango B.1.177 also known as EU1, emerging mid-2020), and
+21B (Pango B.1.617.1 also known as Kappa, emerging in late 2020 to early 2021).
+
+Then, to gauge internal consistency of the time estimation procedure,
+we selected a subset of sample nodes,
+matching the distribution of node times  and
+average number of children of the non-sample internal nodes.
+These selected sample nodes were then re-dated, and
+the inferred dates compared against their recorded collection dates.
+The results in Figure~\ref{fig:node_time_testing} demonstrate that
+although the inferred and true times are generally very close,
+there is a clear bias with re-estimated times
+being consistently older (by an average of 37 days).
+This is in line with the findings of
+other recent work \cite{sanderson2021chronumental}, and
+is likely to be a consequence of edges
+carrying more mutations than would be expected
+under a neutral molecular clock prior
+(either due to errors in tree-building or genuinely varying mutation rates),
+as well as inaccuracies in the recording of sample dates.
 

### 2025-11-03 latex/main.tex
Fixup captions for resources
Closes #767
Closes #768

@@
 exactly matched to a node in the current ARG, and
-added to the ARG
-all peak in December, 2022
-and subsequently around February, 2022.
+added to the ARG per day.
 (B) The fraction of exactly matched samples and
@@
 mirror each other.
-Around the peak of sample volume,
-the fraction of exactly matched samples
-is relatively high (above 0.50).
-As expected, this fraction later decreases
-when sampling effort decreases.
-(C) The mean HMM cost per sample is relatively high
-at the beginning as the ARG had not accumulated
-enough genetic diversity yet.
-(D) Proportion of samples classified
-by Scorpio designation.
-These proportions reflect waves of infections
-driven by the emergence of Alpha, Delta, and Omicron.
+The fraction of exact matches is a function of 
+both sampling depth and composition of the global population.
+(C) The mean HMM cost per sample is a useful metric.
+(D) Proportion of samples classified by (simplified) Scorpio designation.
 }\label{fig:samples_per_day}
@@
 \caption{
-Usage of computational resources
-during the ARG inference process.
+Computational resources used during the ARG inference process.
 (A) The amount of time to process a daily batch of samples
-peaks when the sampling effort is the highest.
-(B) Mean time to process a sample remained low throughout,
-except towards the very end when there were few exact matches.
-The HMM cost was moderate initially,
-then noticeably decreased when sampling effort is the highest
-before increasing again towards the end.
-(C) Max RAM usage also peaks around the time
-when sampling effort is the highest.
+and the number of samples processed per day.
+(B) Mean time to process a sample is less than 5s except
+when the number of samples is small. 
+(C) RAM usage is the most difficult resource to manage, and depends
+on the sample composition. The sharp drop in 2022-01 corresponds to the 
+sweep in which Omicron BA.1 replaced Delta and there was very little sample
+diversity. The drop in RAM usage around 2022-02 corresponds to a change
+in configuration where we reduced the number of HMM threads from 128 to 80
+to better fit into available memory. 
 }

### 2025-11-03 src/plot.py

@@ -76,6 +76,10 @@ def inference_resources():
     ax[2].plot(x, df.max_memory / 1024**3)
     ax[2].set_ylabel("Max memory (GiB)")
 
+    ax[0].text(x=-0.02, y=1.0, s="A", size=20, transform=ax[0].transAxes)
+    ax[1].text(x=-0.02, y=1.0, s="B", size=20, transform=ax[1].transAxes)
+    ax[2].text(x=-0.02, y=1.0, s="C", size=20, transform=ax[2].transAxes)
+
     for a in ax:
         a.grid()
     savefig("inference-resources")
@@ -151,6 +155,11 @@ def samples_per_day():
         labels=[" ".join(s.split("_")) for s in df_scorpio],
     )
     ax4.legend(loc="upper left", ncol=2)
+
+    ax1.text(x=-0.02, y=1.0, s="A", size=20, transform=ax1.transAxes)
+    ax2.text(x=-0.02, y=1.0, s="B", size=20, transform=ax2.transAxes)
+    ax3.text(x=-0.02, y=1.0, s="C", size=20, transform=ax3.transAxes)
+    ax4.text(x=-0.02, y=1.0, s="D", size=20, transform=ax4.transAxes)
 
     savefig("samples-per-day")
 

### 2025-11-03 latex/main.tex

@@
 
+[FIXME Going in too fast here]
 Sc2ts reconstructs an ARG by incrementally adding
@@
 to use a single parameter $k$
-which controls the number of recurrent mutations
-that is preferred to switching to a different parent via recombination
+that controls the number of recurrent mutations
+that is allowed before switching to a different parent via recombination
 % This is quite a big thing really, and ultimately what determines
@@
 given the current ARG.
+% TODO note the parsimony is an OK model here, as Sally suggests?
+% Could cite paper showing parsimony works well too.
 Note that this is a form of parsimony, 
@@
 resulting in 316 distinct trees along the genome
-(some events occurred at the same breakpoints [vague, rephrase]).
+(some events occurred at the same breakpoints).
 % NOTE: if we keep this we should redo on Yan's laptop and quote those numbers
@@
 NextStrain estimates (Figure~\ref{fig:pango_vs_nextstrain_node_times})
-and overall accuracy comparable to 
+and overall accuracy is comparable to 
 Chronumental~\cite{sanderson2021chronumental} applied to 
@@
 
-Deletions are an important part of SARS-CoV-2 evolution~\cite{Jeronimo2023},
+Deletions are an important part of SARS-CoV-2 evolution~\cite{Jeronimo2023}
 but are not incorporated into UShER trees and
@@
 % SHZ: Sequencing errors or misleading alignment.
-from bioinformatics errors
+from bioinformatic errors
 and their multi-base nature presents challenges
@@
 \midrule
-\multicolumn{7}{l}{Type IV events: non recombinants}\\
+\multicolumn{7}{l}{Type IV events: Non recombinants in the ARG}\\
 pango & muts &&&& parent & descendants \\
@@
 these results for the 44 Pango X lineages present in the ARG
-(XAC and XAD have been omitted, and XM partially represented,
-corresponding to 23 omitted samples in total;
-see Document~\ref{sec:supplement_x_pango_lineages}).
+% (XAC and XAD have been omitted, and XM partially represented,
+% corresponding to 23 omitted samples in total;
+(see Document~\ref{sec:supplement_x_pango_lineages}).
 For 16 Pango X lineages, the oldest node
-that has been assigned to the focal Pango lineage is a recombination node, and
+that has been assigned to the focal Pango lineage is a recombination node and
 that node is the ancestor of all samples
 assigned to that (or a descendant) Pango lineage.
-We designate these as Type I events, where the ARG reflects the
+We consider these to be fully concordant (Type I) events,
+where the ARG reflects the
 hierarchical structure of Pango designation.
@@
 \includegraphics[width=\linewidth]{recombinant_evidence_cov_recomb}
-\caption{Things}
-\label{fig:recombinant_evidence_cov_recomb}
+\caption{Properties of sc2ts recombination events, colored by CovRecomb 
+classification. All other details as per Figure~\ref{fig:recombinant_evidence}.
+\label{fig:recombinant_evidence_cov_recomb}}
 \end{figure}
@@
 \includegraphics[width=\linewidth]{recombinant_evidence_rebar}
-\caption{Things}
-\label{fig:recombinant_evidence_rebar}
+\caption{Properties of sc2ts recombination events, colored by rebar
+classification. All other details as per Figure~\ref{fig:recombinant_evidence}.
+\label{fig:recombinant_evidence_rebar}}
 \end{figure}
 
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{recombinant_evidence_ripples_p4}
-\caption{Things}
-\label{fig:recombinant_evidence_ripples_p4}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{recombinant_evidence_ripples_p3}
-\caption{Things}
-\label{fig:recombinant_evidence_ripples_p3}
-\end{figure}
+% \begin{figure}
+% \centering
+% \includegraphics[width=\linewidth]{recombinant_evidence_ripples_p4}
+% \caption{Things}
+% \label{fig:recombinant_evidence_ripples_p4}
+% \end{figure}
+
+% \begin{figure}
+% \centering
+% \includegraphics[width=\linewidth]{recombinant_evidence_ripples_p3}
+% \caption{Things}
+% \label{fig:recombinant_evidence_ripples_p3}
+% \end{figure}
 

### 2025-11-04 latex/main.tex
Various minor updates

@@
 \usepackage[right]{lineno} \linenumbers
+\usepackage[normalem]{ulem}
 
@@
 % this and I don't want to claim we're more parsimonious.
-We found that they were very similar in terms of parsimony
+Eliminating recombination nodes and their descendants (which we expected to differ), we found very similar relationships among  lineages in terms of parsimony
 (Document~\ref{sec:supplement_sc2ts_usher_parsimony})
@@
 by running 3SEQ~\cite{Lam2018improved}
-on the putative recombinant sequence trios (STAR Methods),
+on the putative recombinant sequence trios (STAR Methods)\sout{,
 and 484 of the 855 (57\%) were classified as recombinant.
@@
 and 29\% of the QC-failing recombinants
-being classified as recombinant.
+being classified as recombinant.}
+.
+Of the 855 recombination events identified by sc2ts, 3SEQ verified 57\% of them, rising to 95\% verification among the 354 recombinants that passed our QC filter. 
 In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
@@
 In addition, we found recurrent deletions in ORF8 and NSP3,
-which have been suggested to be indel hotspots
-[FIXME broken reference]
-\cite{Akaishi2022}.
+which have been suggested to be indel hotspots\cite{Akaishi2022}.
 Together, these findings demonstrate that
@@
 we treat its placement with caution.
-Note that the list of mutations in \cite{RambautAlpha2020}
+Note that the list of mutations in ref\cite{RambautAlpha2020}
 includes a T26801C in the M gene
@@
 characteristic Alpha mutations,
-it has been suggested elsewhere as an Alpha-defining mutation 
-(\cite{Li2023indels}).
+it has been suggested elsewhere as an Alpha-defining 
+mutation\cite{Li2023indels}.
 
@@
 Delta and B.1.617.3 are more closely related to each other
-than either is to Kappa (\cite{Cherian2021,Farinholt2021}).
+than either is to Kappa\cite{Cherian2021,Farinholt2021}.
 In the ARG, however, Delta and Kappa appear as sibling lineages
@@
 \label{sec:supplement_indepth_pango_x_lineages}
+In this section we provide some additional analysis on Pango X lineages
+in the ARG. These analyses are intended to be read in conjunction with
+the subgraph illustrations in Document S2.
 
@@
 the Pango lineages of the parents and the breakpoint intervals
-agree with the results reported by Jackson et al. (\ref{tab:jackson}).
+agree with the results reported by Jackson et al. (Table \ref{tab:jackson}).
 Two of the Jackson et al. XA samples
@@
 from North and Central America \cite{Gutierrez2022}.
-However, these parent lineages were not added to the ARG due to a high HMM cost.
+However, these parent lineages were not added to the ARG due to high HMM costs.
 Together with other lineages such as B.1.627
@@
 they are likely to attach somewhere along the long branch
-in the subgraph visualization (with $> 30$ mutations),
-which also contains the key ``R2'' deletion
-associated with XB / B.1.634 \cite{Gutierrez2022}.
-
-We hypothesize that there are a set of samples
-from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
-which, should they be added to the ARG, would break up the long branch,
-with XB being a recombination between two of them.
-In the absence of these, sc2ts cannot detect XB as a recombinant.
-Also, note that the XB consensus mutations all occur on or above the long branch.
-We suspect that the child node of the long branch
-represents the (potentially recombinant) origin of XB, and
-the B.1 samples at the bottom left of the subgraph
-may be misclassified by pangolin.
-
-Further work is required to determine
-why these additional samples were not gradually added
-to the ARG by the daily sample matching process or
-retrospective sample grouping.
+in the subgraph visualization (with $> 30$ mutations).
+If samples from these parental lineages were available for attachment,
+then XB would be detected as a recombinant by sc2ts.
+% which also contains the key ``R2'' deletion
+% associated with XB / B.1.634 \cite{Gutierrez2022}.
+
+% We hypothesize that there are a set of samples
+% from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
+% which, should they be added to the ARG, would break up the long branch,
+% with XB being a recombination between two of them.
+% In the absence of these, sc2ts cannot detect XB as a recombinant.
+% Also, note that the XB consensus mutations all occur on or above the long branch.
+% We suspect that the child node of the long branch
+% represents the (potentially recombinant) origin of XB, and
+% the B.1 samples at the bottom left of the subgraph
+% may be misclassified by pangolin.
+
+% Further work is required to determine
+% why these additional samples were not gradually added
+% to the ARG by the daily sample matching process or
+% retrospective sample grouping.
 

### 2025-11-04 latex/main.tex
WIP for discussion

@@
 
-% \section*{DISCUSSION}
-
-% NOTES: paragraph structure. Lead with the biological insights first 
-% to prevent the ``this is a methods paper'' criticism
-
-% \begin{itemize}
-% \item We present a landmark ARG, which has a comprehensive representation
-% of evolutionary history.
-% \item This is a conservative lower-bound on the occurance of recombination.
-% Very high validation rate post-QC.
-% Substantially lower than previous estimate. 
-% The first order term of this 
-% is the choice of $p$ parameter. With p=4 we show that the ARG contains
-% more of the previously identified recombinants.
-% \item Truth data for recombination is difficult to find. The closest we have is the 
-% pango X lineages. No methods agrees with all of the designations being 
-% true recombinants. Our analysis suggests that there is significant over-naming
-% with the same underlying recombination event getting associated with 
-% multiple X lienages at least twice. Also many lineages are not suggested 
-% as being recombinant. This resolves long-standing complexity in 
-% the Omicron evolution and simplifies our understanding.
-% \item Based on this cleaned up and conservative estimate of recombination
-% prevalance. Supports the idea that  recombination frequency in SARS-CoV-2 
-% is driven jointly by viral
-% load in the population and lineage diversity, implying predictable
-% recombination ``hot periods'' during pandemics
-% \item Also found evidence that breakpoints are enriched near spike, 
-% supporting findings of earlier studies.
-% Suggests functional constraints or template-switching
-% biases that may shape SARS-CoV-2 evolution.
-% Spike gene junctions are recurrent recombination zones, likely under both
-% mechanistic and selective pressures.
-% \item This ARG is base on a new method, sc2ts. Major methodological 
-% breakthrough in terms of handling recombination. We validate the method
-% by using results from the literature. We don't use simulations because 
-% there's no real point in this case. We're just focusing on SARS-CoV-2.
-% We could create some simulations and generate some metrics about how
-% well sc2ts picks up simulated recombination events on clean data, 
-% but what would that tell us ultimately? (Similarly for tree reconstruction.)
-% Certainly standardised simulations would help to refine methods, but
-% we have moved the bar forwards massively here and made a great starting
-% point for subsequent method development.
-% \item There is great potential for using this ARG in future studies,
-% and using sc2ts as a baseline to judge future methods.
-% \item The tskit and VCF Zarr toolchains enable this research.
-% \item We have shown that recombination can be handled, provided
-% feasible methods to infer it, and shown that a preexisting tool
-% chain can provide quick insignts. The future is bright.
-
-% \end{itemize}
-
 \section*{DISCUSSION}
-[NOTE: this is two different WIP versions of the Discussion. 
-The ChatGPT based stuff was only ever intended as a starting point,
-and not to be taken literally.]
-
-% P1. The ARG. Unique resource. Emphasise utility of UShER and how 
-% this has more.
-% SHZ: This makes it sound like we are just making analyses
-% (which people can do already using existing methods and tools)
-% easier to do
-% rather than making a leap forward from anything else out there.
-We present the first pandemic-scale ancestral recombination graph (ARG) for
-SARS-CoV-2, reconstructed from 2.48 million genomes using \texttt{sc2ts}. This
-ARG represents a step change in how the evolutionary history of this virus can
-be represented and analysed. Like the UShER phylogeny, it provides a
-comprehensive and continuously updatable record of global genomic diversity,
-but it extends this paradigm by embedding recombination directly into the
-genealogy. The ARG unifies point mutations, deletions, and recombination events
-within a single, compact data structure, enabling analyses that
-were previously fragmented across multiple methods and data models. By encoding
-reticulate ancestry explicitly rather than assuming a single tree, the ARG
-provides a complete and internally consistent representation of SARS-CoV-2
-evolution. In this way, the
-ARG offers the same accessibility and scalability that made the UShER tree
-indispensable during the pandemic, while expanding its scope to encompass the
-full spectrum of evolutionary processes shaping SARS-CoV-2.
-
-% P2. This is a conservative lower-bound on recombination. 
-% Discuss other estimates.
-% 
-% SHZ: Why should we care about getting this number/frequency right?
-% Sure, popgen folks love estimating this, but to other people,
-% existing estimates already tell us that
-% detectable recombination in SARS-CoV-2 is (qualitatively) rare,
-% even without the advances of sc2ts.
-% Also, we are talking about the frequency of detectable recombination
-% in the last sentence, right?
-% The true frequencey of recombination is unknowable from this type of data.
-% So, the best we are ever going to get is a better lower bound.
-The recombination frequency estimated from the \texttt{sc2ts} ARG is
-substantially lower than that implied by the combination of UShER+RIPPLES,
-largely reflecting differences in model formulation and parameter choice.
-RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
-reported 589 unique recombination events with 43,104 descendant
-samples---corresponding to roughly 2.7\% of all genomes being of recombinant
-origin.  Using the same time interval, our inference at \(k=4\) identifies only
-14 high-confidence recombination events with 172 descendants among 501,235
-genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
-explained by the stricter parsimony penalty adopted here: increasing the
-penalty from \(p=3\) (as used in the RIPPLES analysis) 
-to \(k=4\) greatly reduces false
-positives while retaining nearly all verified recombinant lineages, 
-including those missed by RIPPLES at equivalent thresholds.  
-We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
-of recombination as
-tree rearrangements proceed, thereby requiring the more lenient 
-threshold in order to find known recombinats.
-These comparisons suggest that our estimate represents a conservative 
-lower bound. Further work will be required to determine
-the true frequency of recombination in SARS-CoV-2 evolution.
-
-% P3. Clarifies relationships between pango X lineages, simplifying
-% the Omicron sublineage.
-% 
-% SHZ: I don't think knowing the relationships between Pango Xs is what's really important.
-% Sure, phylogeneticists would get a kick out of it,
-% but I don't think the majority of Cell readership would.
-% I think what is important is knowing
-% how many emerging recombinants carrying new combinations of mutations of concern
-% we are actually dealing with.
-% Clarifying the relationships between Pango Xs helps with this task.
-
-The Pango X lineage designations represent the community’s best attempt to
-catalogue recombinant SARS-CoV-2 lineages, but they arise from a largely
-manual, volunteer-driven process that synthesizes multiple lines of evidence
-and expert judgment rather than a single automated criterion.  As a result, the
-system provides the closest available approximation to “truth data,” but
-inevitably contains inconsistencies and over-splitting.  Within the
-\texttt{sc2ts} ARG, 44 Pango X lineages are represented, of which 16 correspond
-directly to recombination nodes (Type I events).  Examination of their
-genealogical relationships shows that many of these 44 lineages share common
-ancestral recombination events rather than arising independently.  In
-particular, five lineages—XAA, XAG, XAM, XR, and XU—descend from the same
-recombinant ancestor associated with XQ, while another five—XZ, XAC, XAD, XAE,
-and XAP—share a second common ancestor that we refer to as “Xx.”  These
-clusters account for roughly half of all X designations in the dataset,
-indicating that the community process has at times produced multiple labels for
-the descendants of a single recombinant lineage.  Conversely, several [TODO 
-be more concrete]
-designated X lineages show little or no evidence of recombination in the ARG,
-suggesting that some may reflect convergent mutation or limited sampling rather
-than hybrid origin.  Together, these results provide a data-driven synthesis in
-which the 44 currently recognized recombinant lineages collapse into a smaller
-number of well-supported ancestral events, simplifying the complex picture of
-Omicron-era evolution and bringing the community nomenclature into closer
-alignment with the underlying genealogy.
-
-% P4. Results on global case numbers and frequency of recomb.
-% SHZ: Should we mention how the reliability of global case number data
-% and variation in sampling effort over time and across countries
-% might affect our interpretation of these findings?
-[FIXME: This is direct output from ChatGPT.  Review!]
-The dated ARG allows us to place recombination within its epidemiological
-context, revealing clear temporal structure in when recombinant lineages arise.
-The frequency of inferred recombination events follows global case numbers
-through the pandemic, with peaks during the Delta and Omicron waves when both
-infection prevalence and genetic diversity were high.  Quantitatively, the best
-single predictor of weekly recombinant counts is the product of global cases
-and lineage heterozygosity (\(nH\)), which explains nearly a quarter of the
-variation in event frequency (\(R^{2}_{\mathrm{adj}} = 0.24\)).  This
-relationship is consistent with the mechanistic expectation that recombination
-requires both frequent co-infections (driven by high case numbers) and
-genetically distinct co-infecting viruses (driven by lineage diversity).  In
-other words, recombination becomes most detectable when epidemiological and
-evolutionary opportunity coincide.  These results imply that the emergence of
-novel recombinants is not uniform through time but concentrated into
-predictable ``hot periods'' during large, antigenically diverse outbreaks, when
-the ecological conditions for viral recombination are most favourable.
-
-% P5. Confirm breakpoints around Spike. 
-% SHZ: Maybe we can say it's a known (instead of robust)
-% pattern of detectable recombination in CoVs in general?
-% There are some citations for template switching hotspots in SC2.
-% Also, it's not only about selection favoring new combinations of mutations in Spike.
-% It's about combinations of mutations in, say, ORF1ab, getting selected against
-% (or just low mutation rates).
-% Maybe mention epistasis?
-[FIXME: This is directish output from ChatGPT. Review!]
-The genomic distribution of breakpoints in the ARG is highly uneven, with a
-pronounced enrichment near the $5'$ boundary of the Spike gene and across the
-S1–S2 junction.  This pattern closely mirrors that observed in the
-UShER+RIPPLES dataset \citep{Turakhia2022} and in earlier 
-analyses \citep{Jackson2021} [more cites], but here it is recovered
-independently over a much longer sampling period and under a more conservative
-inference regime.  Under our more conservative estimate of the overall
-prevalence of recombination, we find that this enrichment persists despite a substantial
-reduction in the total number of inferred recombination events, indicating that
-it reflects a genuine and stable feature of SARS-CoV-2 evolution rather than a
-by-product of detection sensitivity.  Across all high-confidence events, the
-median breakpoint interval length is $\sim 2.2$kb and declines with parental
-divergence, consistent with expectations from sequence-based localization
-accuracy.  These results show that even with a conservative lower bound on
-recombination frequency, the concentration of breakpoints around Spike remains
-a robust signature of coronavirus genome biology—likely reflecting a
-combination of mechanistic biases in template switching and selective
-constraints on Spike function.
-
-% P6. Sc2ts method is a major advance. We validate with real data
-% rather than simulations. Forward looking statement about simulations.
-% 
-% SHZ: Could at least some of this go into the Introduction?
-In the broader context of pathogen genomics, \texttt{sc2ts} demonstrates that
-it is now possible to reconstruct recombination-aware genealogies at the scale
-of entire pandemics.  Although recombination is recognised as a major
-evolutionary force in many pathogens—including RNA viruses such as influenza
-and HIV, and numerous bacterial species—genealogical analyses have typically
-relied on tree-based approximations that ignore or collapse recombination.
-This simplification has limited the accuracy of downstream evolutionary and
-epidemiological inference, even in well-studied systems.  \texttt{Sc2ts}
-overcomes this barrier by directly inferring an ancestral recombination graph
-from millions of viral genomes, producing the first coherent reticulate
-genealogy for any pathogen dataset of comparable size.  The ability to
-integrate recombination into genome-wide genealogies at this scale represents a
-major conceptual step forward: it transforms recombination from a confounding
-complication into an analyzable component of pathogen evolution.
-
-% P7. The software ecosystem enables diverse besplike analysis.
-
-[TODO: This is close, needs revision]
-The practical value of the ARG is amplified by its integration into the broader
-software ecosystems built around \texttt{tskit} and VCF~Zarr.  These mature,
-open-source libraries provide highly efficient storage and analysis of tree
-sequences and variant data, allowing millions of genomes to be explored
-interactively within familiar environments such as Python and Jupyter
-notebooks.  Analyses that previously required bespoke phylogenetic
-pipelines—such as extracting local genealogies, summarising ancestral
-relationships, mapping mutations to branches, or tracking recombinant ancestry
-through time—can now be performed with concise, reproducible code on standard
-hardware.  Because \texttt{tskit} and VCF~Zarr share a common data model, the
-ARG integrates seamlessly with widely used tools in the PyData ecosystem,
-enabling users to combine genealogical information with epidemiological,
-geographic, or phenotypic metadata in the same workflow.  This interoperability
-transforms what can be done with large viral genome collections, lowering the
-barrier to sophisticated evolutionary analysis and making recombination-aware
-genealogy inference accessible to a wide community of researchers.
-
-% P8. Outlook
-
-[TODO a bit muddled, refine]
-Together, these results demonstrate that recombination in SARS-CoV-2 can be
-systematically incorporated into genealogical inference at pandemic scale. The
-\texttt{sc2ts} ARG captures the complexity of viral evolution in a form that is
-both biologically realistic and computationally tractable, showing that
-recombination can be modelled rather than ignored. Embedded within an
-interoperable software ecosystem, it provides a comprehensive record of
-SARS-CoV-2 ancestry and a practical framework for real-time,
-recombination-aware genomic surveillance—turning large sequence archives into
-dynamic evolutionary models that can be queried with concise, reproducible code. 
-More broadly, this work marks a transition from
-tree-based pandemic phylogenetics to recombination-aware genealogy inference.
-While our focus is SARS-CoV-2, the approach offers a blueprint and a starting
-point for other recombining pathogens: the core ideas are portable,
-even though effective deployment will require pathogen-specific adjustments. 
-We anticipate that this resource will clarify the recent history of
-SARS-CoV-2 and catalyse methodological refinement and targeted biological
-studies of recombination across systems where such extensions prove feasible.
-
-[NOTE: This is Shing's WIP alternative]
-
-%%% SHZ: WIP alternative.
-We present the first pandemic-scale
-ancestral recombination graph (ARG) for SARS-CoV-2,
-reconstructed from $\sim$2.48 million genomes using \textit{sc2ts}.
-We demonstrate how this ARG, for the first time,
-enables systematic identification and characterization of
-recombination events in fine-grained
-detail--in terms of their parents, breakpoints, and timing of
-emergence--providing richer and more nuanced
-genealogical, genomic, and epidemiological context
-than existing phylogenetic approaches.
-Together, sc2ts and the ARG represent a new paradigm in
-how we study the evolution of pathogens in the face of ongoing sampling.
-
-Accounting for recombination is a key challenge in pathogen genomics, and
-recombination has been largely ignored in analysis of viral genomes.
-Existing phylogenetic approaches for pathogens detect recombination
-\textit{post hoc} (e.g., UShER+RIPPLES),
-treat it as a nuance (e.g., XX), or
-discard its signals from downstream analyses (e.g., XX).
-Sc2ts addresses this challenge
-by using the ARG as the underlying data model,
-explicitly representing recombination events as individual nodes and
-relating them to specific parents in the ARG.
-The ARG provides a coherent framework to determine and represent
-the relationships between recombinants and their specific parents
-(which are not possible using a tree-based data model, e.g., the MAT format of UShER).
-The ARG reveals and clarifies the relationships between
-individual samples of the same recombinant origin,
-avoiding issues such as over-naming recombinants
-(e.g., the samples assigned to
-five differently labeled recombinant Pango lineages
-descending from the “Xx” recombination event in the ARG).
-Our results show how an ARG can considerably simplify
-efforts to identify and monitor emerging recombinants,
-facilitating communication of genomic surveillance results.
-
-% SHZ: Very WIP
-% Discuss the new preprint from Turakhia.
-The timing of recombination events is of key interest in
-pathogen genomic epidemiology.
-because this information may yield insights into
-how the emergence of a new recombinant might be
-linked with changes in public health policies.
-
-The overall frequency of detectable recombination events
-estimated from the sc2ts ARG is substantially lower than
-that from UShER+RIPPLES,
-largely reflecting differences in model formulation and parameter choice.
-When applied \textit{post hoc} to an UShER phylogeny of
-$\sim$1.6 million genomes sampled up to May, 2021,
-RIPPLES reported 589 unique recombination events with 43,104 descendant
-samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
-Using the same time interval, sc2ts \(k=4\) identifies
-14 high-confidence recombination events with
-172 descendant samples among 501,235 samples \(0.034\%\),
-a 40- to 80-fold reduction.
-Much of this difference is explained by
-the stricter parsimony penalty used here:
-increasing the penalty from p=3 (as used in the RIPPLES analysis) to
-k=4 greatly reduces false positives
-while retaining nearly all verified recombinant lineages,
-including those missed by RIPPLES at equivalent thresholds.
-These indirect comparisons suggest that
-our estimate represents a conservative lower bound.
-In future studies, dissecting the methodological differences
-between sc2ts and UShER+RIPPLES may
-inspire new ideas to develop or improve ARG building methods
-for SARS-CoV-2 (and potentially other pathogens).
-
-The ARG itself is a powerful resource
-to support future studies of SARS-CoV-2.
-Our results show that the ARG captures and unifies
-important features of SARS-CoV-2 evolution,
-which are shown in different studies
-using various methods, tools, and approaches.
-The ARG also compactly encodes genetic variation in
-millions of SARS-CoV-2 genomes,
-allowing for convenient and efficient reuse of
-this valuable dataset for a range of
-genetic and evolutionary analyses beyond those presented here,
-for instance, to understand
-how the variability in sequencing efforts
-across time and location affects our ability
-to reconstruct the origin of important recombinant lineages and
-to estimate when they emerge.
-Because building the ARG is computationally expensive, and
-checking and improving various aspects of its quality is laborious,
-we hope that making this resource openly available will
-save significant time and effort for other researchers,
-enabling immediate focus on analyzing the information
-contained within the COVID-19 dataset.
-
-% TODO: Briefly discussion software tools here or in a subsequent paragraph.
-
-Recent developments in ARG methods have focused almost exclusively on
-applications to human sequencing data.
-The availability of human ARGs reconstructed from large datasets
-has already had a tremendous impact on improving
-our understanding of human genetic ancestry and past demography.
-We anticipate similar advances to be made
-by adapting existing inference methods for human ARGs
-to use the sc2ts ARG for the genealogy-based study of
-natural selection or the geographic spread of SARS-CoV-2 through time.
-
-The unprecedented throughput and pace of genome sequencing are
-creating ample opportunity to understand the impacts of recombination
-on pathogen evolution and epidemiology.
-We hope that this study will spur
-development of ARG methods for pathogens,
-ultimately making recombination-aware genome analysis
-routine for all sorts of pathogens beyond SARS-CoV-2.
+There are two primary contributions of this work. The first is a
+Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
+Through multiple layers of validation against results from the 
+literature, we have demonstrated that this ARG successfully synthesises
+the evolutionary features of genetic inheritance, point mutations,
+deletions and recombination in a single cohesive structure.
+The ARG is freely available and easily accessible using the 
+tskit software ecosystem, a rich and efficient platform for 
+recombination-aware analysis widely used in population genomics.
+This combination of a deeply validated ARG with a pre-existing
+and highly capable software platform forms a unique resource.
+Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
+have helped spur the development of numerous ARG-aware 
+methods in human 
+genomics~\cite{Tagami2024tstrait,Mejia2025using} [NOTE add more refs 
+if we like this].
+We hope that the ARG provided here will similarly
+enable a range recombination-aware analyses of SARS-CoV-2, and 
+catalyse a wave of methodological improvements fully integrating
+recombination into pathogen genomics.
+
+The second primary contribution of this work is the method used to 
+infer the ARG, sc2ts. [WIP]
+
+% Old discussions
+
+% % P1. The ARG. Unique resource. Emphasise utility of UShER and how 
+% % this has more.
+% % SHZ: This makes it sound like we are just making analyses
+% % (which people can do already using existing methods and tools)
+% % easier to do
+% % rather than making a leap forward from anything else out there.
+% We present the first pandemic-scale ancestral recombination graph (ARG) for
+% SARS-CoV-2, reconstructed from 2.48 million genomes using \texttt{sc2ts}. This
+% ARG represents a step change in how the evolutionary history of this virus can
+% be represented and analysed. Like the UShER phylogeny, it provides a
+% comprehensive and continuously updatable record of global genomic diversity,
+% but it extends this paradigm by embedding recombination directly into the
+% genealogy. The ARG unifies point mutations, deletions, and recombination events
+% within a single, compact data structure, enabling analyses that
+% were previously fragmented across multiple methods and data models. By encoding
+% reticulate ancestry explicitly rather than assuming a single tree, the ARG
+% provides a complete and internally consistent representation of SARS-CoV-2
+% evolution. In this way, the
+% ARG offers the same accessibility and scalability that made the UShER tree
+% indispensable during the pandemic, while expanding its scope to encompass the
+% full spectrum of evolutionary processes shaping SARS-CoV-2.
+
+% % P2. This is a conservative lower-bound on recombination. 
+% % Discuss other estimates.
+% % 
+% % SHZ: Why should we care about getting this number/frequency right?
+% % Sure, popgen folks love estimating this, but to other people,
+% % existing estimates already tell us that
+% % detectable recombination in SARS-CoV-2 is (qualitatively) rare,
+% % even without the advances of sc2ts.
+% % Also, we are talking about the frequency of detectable recombination
+% % in the last sentence, right?
+% % The true frequencey of recombination is unknowable from this type of data.
+% % So, the best we are ever going to get is a better lower bound.
+% The recombination frequency estimated from the \texttt{sc2ts} ARG is
+% substantially lower than that implied by the combination of UShER+RIPPLES,
+% largely reflecting differences in model formulation and parameter choice.
+% RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
+% reported 589 unique recombination events with 43,104 descendant
+% samples---corresponding to roughly 2.7\% of all genomes being of recombinant
+% origin.  Using the same time interval, our inference at \(k=4\) identifies only
+% 14 high-confidence recombination events with 172 descendants among 501,235
+% genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
+% explained by the stricter parsimony penalty adopted here: increasing the
+% penalty from \(p=3\) (as used in the RIPPLES analysis) 
+% to \(k=4\) greatly reduces false
+% positives while retaining nearly all verified recombinant lineages, 
+% including those missed by RIPPLES at equivalent thresholds.  
+% We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
+% of recombination as
+% tree rearrangements proceed, thereby requiring the more lenient 
+% threshold in order to find known recombinats.
+% These comparisons suggest that our estimate represents a conservative 
+% lower bound. Further work will be required to determine
+% the true frequency of recombination in SARS-CoV-2 evolution.
+
+% % P3. Clarifies relationships between pango X lineages, simplifying
+% % the Omicron sublineage.
+% % 
+% % SHZ: I don't think knowing the relationships between Pango Xs is what's really important.
+% % Sure, phylogeneticists would get a kick out of it,
+% % but I don't think the majority of Cell readership would.
+% % I think what is important is knowing
+% % how many emerging recombinants carrying new combinations of mutations of concern
+% % we are actually dealing with.
+% % Clarifying the relationships between Pango Xs helps with this task.
+
+% The Pango X lineage designations represent the community’s best attempt to
+% catalogue recombinant SARS-CoV-2 lineages, but they arise from a largely
+% manual, volunteer-driven process that synthesizes multiple lines of evidence
+% and expert judgment rather than a single automated criterion.  As a result, the
+% system provides the closest available approximation to “truth data,” but
+% inevitably contains inconsistencies and over-splitting.  Within the
+% \texttt{sc2ts} ARG, 44 Pango X lineages are represented, of which 16 correspond
+% directly to recombination nodes (Type I events).  Examination of their
+% genealogical relationships shows that many of these 44 lineages share common
+% ancestral recombination events rather than arising independently.  In
+% particular, five lineages—XAA, XAG, XAM, XR, and XU—descend from the same
+% recombinant ancestor associated with XQ, while another five—XZ, XAC, XAD, XAE,
+% and XAP—share a second common ancestor that we refer to as “Xx.”  These
+% clusters account for roughly half of all X designations in the dataset,
+% indicating that the community process has at times produced multiple labels for
+% the descendants of a single recombinant lineage.  Conversely, several [TODO 
+% be more concrete]
+% designated X lineages show little or no evidence of recombination in the ARG,
+% suggesting that some may reflect convergent mutation or limited sampling rather
+% than hybrid origin.  Together, these results provide a data-driven synthesis in
+% which the 44 currently recognized recombinant lineages collapse into a smaller
+% number of well-supported ancestral events, simplifying the complex picture of
+% Omicron-era evolution and bringing the community nomenclature into closer
+% alignment with the underlying genealogy.
+
+% % P4. Results on global case numbers and frequency of recomb.
+% % SHZ: Should we mention how the reliability of global case number data
+% % and variation in sampling effort over time and across countries
+% % might affect our interpretation of these findings?
+% [FIXME: This is direct output from ChatGPT.  Review!]
+% The dated ARG allows us to place recombination within its epidemiological
+% context, revealing clear temporal structure in when recombinant lineages arise.
+% The frequency of inferred recombination events follows global case numbers
+% through the pandemic, with peaks during the Delta and Omicron waves when both
+% infection prevalence and genetic diversity were high.  Quantitatively, the best
+% single predictor of weekly recombinant counts is the product of global cases
+% and lineage heterozygosity (\(nH\)), which explains nearly a quarter of the
+% variation in event frequency (\(R^{2}_{\mathrm{adj}} = 0.24\)).  This
+% relationship is consistent with the mechanistic expectation that recombination
+% requires both frequent co-infections (driven by high case numbers) and
+% genetically distinct co-infecting viruses (driven by lineage diversity).  In
+% other words, recombination becomes most detectable when epidemiological and
+% evolutionary opportunity coincide.  These results imply that the emergence of
+% novel recombinants is not uniform through time but concentrated into
+% predictable ``hot periods'' during large, antigenically diverse outbreaks, when
+% the ecological conditions for viral recombination are most favourable.
+
+% % P5. Confirm breakpoints around Spike. 
+% % SHZ: Maybe we can say it's a known (instead of robust)
+% % pattern of detectable recombination in CoVs in general?
+% % There are some citations for template switching hotspots in SC2.
+% % Also, it's not only about selection favoring new combinations of mutations in Spike.
+% % It's about combinations of mutations in, say, ORF1ab, getting selected against
+% % (or just low mutation rates).
+% % Maybe mention epistasis?
+% [FIXME: This is directish output from ChatGPT. Review!]
+% The genomic distribution of breakpoints in the ARG is highly uneven, with a
+% pronounced enrichment near the $5'$ boundary of the Spike gene and across the
+% S1–S2 junction.  This pattern closely mirrors that observed in the
+% UShER+RIPPLES dataset \citep{Turakhia2022} and in earlier 
+% analyses \citep{Jackson2021} [more cites], but here it is recovered
+% independently over a much longer sampling period and under a more conservative
+% inference regime.  Under our more conservative estimate of the overall
+% prevalence of recombination, we find that this enrichment persists despite a substantial
+% reduction in the total number of inferred recombination events, indicating that
+% it reflects a genuine and stable feature of SARS-CoV-2 evolution rather than a
+% by-product of detection sensitivity.  Across all high-confidence events, the
+% median breakpoint interval length is $\sim 2.2$kb and declines with parental
+% divergence, consistent with expectations from sequence-based localization
+% accuracy.  These results show that even with a conservative lower bound on
+% recombination frequency, the concentration of breakpoints around Spike remains
+% a robust signature of coronavirus genome biology—likely reflecting a
+% combination of mechanistic biases in template switching and selective
+% constraints on Spike function.
+
+% % P6. Sc2ts method is a major advance. We validate with real data
+% % rather than simulations. Forward looking statement about simulations.
+% % 
+% % SHZ: Could at least some of this go into the Introduction?
+% In the broader context of pathogen genomics, \texttt{sc2ts} demonstrates that
+% it is now possible to reconstruct recombination-aware genealogies at the scale
+% of entire pandemics.  Although recombination is recognised as a major
+% evolutionary force in many pathogens—including RNA viruses such as influenza
+% and HIV, and numerous bacterial species—genealogical analyses have typically
+% relied on tree-based approximations that ignore or collapse recombination.
+% This simplification has limited the accuracy of downstream evolutionary and
+% epidemiological inference, even in well-studied systems.  \texttt{Sc2ts}
+% overcomes this barrier by directly inferring an ancestral recombination graph
+% from millions of viral genomes, producing the first coherent reticulate
+% genealogy for any pathogen dataset of comparable size.  The ability to
+% integrate recombination into genome-wide genealogies at this scale represents a
+% major conceptual step forward: it transforms recombination from a confounding
+% complication into an analyzable component of pathogen evolution.
+
+% % P7. The software ecosystem enables diverse besplike analysis.
+
+% [TODO: This is close, needs revision]
+% The practical value of the ARG is amplified by its integration into the broader
+% software ecosystems built around \texttt{tskit} and VCF~Zarr.  These mature,
+% open-source libraries provide highly efficient storage and analysis of tree
+% sequences and variant data, allowing millions of genomes to be explored
+% interactively within familiar environments such as Python and Jupyter
+% notebooks.  Analyses that previously required bespoke phylogenetic
+% pipelines—such as extracting local genealogies, summarising ancestral
+% relationships, mapping mutations to branches, or tracking recombinant ancestry
+% through time—can now be performed with concise, reproducible code on standard
+% hardware.  Because \texttt{tskit} and VCF~Zarr share a common data model, the
+% ARG integrates seamlessly with widely used tools in the PyData ecosystem,
+% enabling users to combine genealogical information with epidemiological,
+% geographic, or phenotypic metadata in the same workflow.  This interoperability
+% transforms what can be done with large viral genome collections, lowering the
+% barrier to sophisticated evolutionary analysis and making recombination-aware
+% genealogy inference accessible to a wide community of researchers.
+
+% % P8. Outlook
+
+% [TODO a bit muddled, refine]
+% Together, these results demonstrate that recombination in SARS-CoV-2 can be
+% systematically incorporated into genealogical inference at pandemic scale. The
+% \texttt{sc2ts} ARG captures the complexity of viral evolution in a form that is
+% both biologically realistic and computationally tractable, showing that
+% recombination can be modelled rather than ignored. Embedded within an
+% interoperable software ecosystem, it provides a comprehensive record of
+% SARS-CoV-2 ancestry and a practical framework for real-time,
+% recombination-aware genomic surveillance—turning large sequence archives into
+% dynamic evolutionary models that can be queried with concise, reproducible code. 
+% More broadly, this work marks a transition from
+% tree-based pandemic phylogenetics to recombination-aware genealogy inference.
+% While our focus is SARS-CoV-2, the approach offers a blueprint and a starting
+% point for other recombining pathogens: the core ideas are portable,
+% even though effective deployment will require pathogen-specific adjustments. 
+% We anticipate that this resource will clarify the recent history of
+% SARS-CoV-2 and catalyse methodological refinement and targeted biological
+% studies of recombination across systems where such extensions prove feasible.
+
+% [NOTE: This is Shing's WIP alternative]
+
+% %%% SHZ: WIP alternative.
+% We present the first pandemic-scale
+% ancestral recombination graph (ARG) for SARS-CoV-2,
+% reconstructed from $\sim$2.48 million genomes using \textit{sc2ts}.
+% We demonstrate how this ARG, for the first time,
+% enables systematic identification and characterization of
+% recombination events in fine-grained
+% detail--in terms of their parents, breakpoints, and timing of
+% emergence--providing richer and more nuanced
+% genealogical, genomic, and epidemiological context
+% than existing phylogenetic approaches.
+% Together, sc2ts and the ARG represent a new paradigm in
+% how we study the evolution of pathogens in the face of ongoing sampling.
+
+% Accounting for recombination is a key challenge in pathogen genomics, and
+% recombination has been largely ignored in analysis of viral genomes.
+% Existing phylogenetic approaches for pathogens detect recombination
+% \textit{post hoc} (e.g., UShER+RIPPLES),
+% treat it as a nuance (e.g., XX), or
+% discard its signals from downstream analyses (e.g., XX).
+% Sc2ts addresses this challenge
+% by using the ARG as the underlying data model,
+% explicitly representing recombination events as individual nodes and
+% relating them to specific parents in the ARG.
+% The ARG provides a coherent framework to determine and represent
+% the relationships between recombinants and their specific parents
+% (which are not possible using a tree-based data model, e.g., the MAT format of UShER).
+% The ARG reveals and clarifies the relationships between
+% individual samples of the same recombinant origin,
+% avoiding issues such as over-naming recombinants
+% (e.g., the samples assigned to
+% five differently labeled recombinant Pango lineages
+% descending from the “Xx” recombination event in the ARG).
+% Our results show how an ARG can considerably simplify
+% efforts to identify and monitor emerging recombinants,
+% facilitating communication of genomic surveillance results.
+
+% % SHZ: Very WIP
+% % Discuss the new preprint from Turakhia.
+% The timing of recombination events is of key interest in
+% pathogen genomic epidemiology.
+% because this information may yield insights into
+% how the emergence of a new recombinant might be
+% linked with changes in public health policies.
+
+% The overall frequency of detectable recombination events
+% estimated from the sc2ts ARG is substantially lower than
+% that from UShER+RIPPLES,
+% largely reflecting differences in model formulation and parameter choice.
+% When applied \textit{post hoc} to an UShER phylogeny of
+% $\sim$1.6 million genomes sampled up to May, 2021,
+% RIPPLES reported 589 unique recombination events with 43,104 descendant
+% samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
+% Using the same time interval, sc2ts \(k=4\) identifies
+% 14 high-confidence recombination events with
+% 172 descendant samples among 501,235 samples \(0.034\%\),
+% a 40- to 80-fold reduction.
+% Much of this difference is explained by
+% the stricter parsimony penalty used here:
+% increasing the penalty from p=3 (as used in the RIPPLES analysis) to
+% k=4 greatly reduces false positives
+% while retaining nearly all verified recombinant lineages,
+% including those missed by RIPPLES at equivalent thresholds.
+% These indirect comparisons suggest that
+% our estimate represents a conservative lower bound.
+% In future studies, dissecting the methodological differences
+% between sc2ts and UShER+RIPPLES may
+% inspire new ideas to develop or improve ARG building methods
+% for SARS-CoV-2 (and potentially other pathogens).
+
+% The ARG itself is a powerful resource
+% to support future studies of SARS-CoV-2.
+% Our results show that the ARG captures and unifies
+% important features of SARS-CoV-2 evolution,
+% which are shown in different studies
+% using various methods, tools, and approaches.
+% The ARG also compactly encodes genetic variation in
+% millions of SARS-CoV-2 genomes,
+% allowing for convenient and efficient reuse of
+% this valuable dataset for a range of
+% genetic and evolutionary analyses beyond those presented here,
+% for instance, to understand
+% how the variability in sequencing efforts
+% across time and location affects our ability
+% to reconstruct the origin of important recombinant lineages and
+% to estimate when they emerge.
+% Because building the ARG is computationally expensive, and
+% checking and improving various aspects of its quality is laborious,
+% we hope that making this resource openly available will
+% save significant time and effort for other researchers,
+% enabling immediate focus on analyzing the information
+% contained within the COVID-19 dataset.
+
+% % TODO: Briefly discussion software tools here or in a subsequent paragraph.
+
+% Recent developments in ARG methods have focused almost exclusively on
+% applications to human sequencing data.
+% The availability of human ARGs reconstructed from large datasets
+% has already had a tremendous impact on improving
+% our understanding of human genetic ancestry and past demography.
+% We anticipate similar advances to be made
+% by adapting existing inference methods for human ARGs
+% to use the sc2ts ARG for the genealogy-based study of
+% natural selection or the geographic spread of SARS-CoV-2 through time.
+
+% The unprecedented throughput and pace of genome sequencing are
+% creating ample opportunity to understand the impacts of recombination
+% on pathogen evolution and epidemiology.
+% We hope that this study will spur
+% development of ARG methods for pathogens,
+% ultimately making recombination-aware genome analysis
+% routine for all sorts of pathogens beyond SARS-CoV-2.
 
@@
 
-% SHZ: Should we do point form?
+% SHZ: WIP
 We close by emphasizing several current limitations
-as well as potential directions to improve sc2ts,
+and potential directions to improve sc2ts,
 which we hope will also inspire future research
-to develop ARG methods for SARS-CoV-2 and other pathogens.
-
-Reversion of deletions can indicate recombination,
+to develop ARG inference methods for pathogens beyond SARS-CoV-2.
+
+We suggest several ways to extend the HMM.
+First, reversion of deletions can indicate recombination,
 because bases that are lost are unlikely to be regained,
 except via recombination with a lineage that has those bases.
-We presently treat multi-base deletions as missing data.
-Modeling deletions (and more generally insertions) in the HMM
-could help recover additional signals of recombination
-and reduce false positives.
-
-The current HMM assumes a simplistic mutation model
+We presently treated multi-base deletions as missing data, and
+added deletions to the ARG \textit{post hoc}.
+Instead, by explicitly modeling deletions (and more generally insertions) in the HMM,
+we should be able to recover additional signals of recombination
+and reduce spurious recombinants.
+Second, the current HMM assumes a simplistic mutation model
 (where mutation to any other allele is equally probable) and
-implicitly assumes a uniform mutation rate along the SARS-CoV-2 genome.
+a uniform mutation rate along the SARS-CoV-2 genome.
 However, genomic analyses have revealed
-transition-to-transversion rate bias \cite{Bloom2023,Simmonds2024} and
+transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024} and
 sequence context-dependent mutation rate variation
-across the SARS-CoV-2 genome \cite{Haddox2025}.
+across the SARS-CoV-2 genome~\cite{Haddox2025}.
 Using a more realistic mutation model and
 accounting for mutation rate heterogeneity
-could improve inference of local genetic relationships in the ARG and
-reduce spurious recombinants.
-
-Future developments may account for geography in the HMM,
+could improve inference of local genetic relationships in the ARG.
+Third, geography can be accounted for in the HMM,
 favoring attachment of samples to geographically close lineages.
+This should also improve precise identification of recombinant parents,
+because parents sampled in the same region or geographically close regions
+are expected to be more plausible
+than parents sampled in geographically distant regions.
 
@@
 before sequences become publicly
-available \cite{Lancet2021,Kalia2021,Chen2022}.
+available~\cite{Lancet2021,Kalia2021,Chen2022}.
 For example, revisiting the ARG topology periodically

### 2025-11-04 latex/main.tex
Update

--- 2025-11-04 latex/main.tex
@@
 The second primary contribution of this work is the method used to 
-infer the ARG, sc2ts. [WIP]
+infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
+genomes is a step change in scale over existing ARG 
+inference methods for pathogen data\cite{Didelot2010inference,%
+Vaughan2017inferring,%
+Muller2020bayesian,%
+Ignatieva2022,%
+Muller2022bayesian}.
+Similarly, detecting and fully incorporating recombination 
+in real-time at pandemic scale extends the 
+demonstrated utility of a comprehensive
+genealogy~\cite{Hinrichs2024,McBroome2024}, clarifying the 
+role of recombination and enabling timely analysis.
+Given the novelty of sc2ts and the inferred ARG, we have focused 
+significant attention here on methodological issues. 
+Through these detailed analyses we validate the ARG itself,
+ensuring that it is a trustworthy basis for future research,
+and deepen our understanding of the inference method, and 
+how it may be adapted and improved (see Limitations of Study).
+
+The secondary contributions of this work are the initial 
+insights into the evolution of SARS-CoV-2 obtained by studying
+the sc2ts ARG. Firstly, we have clarified the relationships among
+the recombinant Pango X-lineages, suggesting significant 
+clustering of named lineages under shared recombination events.
+Secondly, we have used the ARG to study breakpoint intervals 
+for recombination events, adding a novel temporal dimension
+and demonstrating its theoretical relationship with the 
+size of the interval in base pairs. Thirdly,
+
+These underscore the utility of the ARG itself, and suggest
+several important directions for future research.
 

### 2025-11-04 latex/main.tex
Disco update

@@
 \section*{DISCUSSION}
-There are two primary contributions of this work. The first is a
+There are two primary contributions of this work. The first is an
 Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
@@
 and deepen our understanding of the inference method, and 
-how it may be adapted and improved (see Limitations of Study).
+how it may be adapted and improved (see Limitations of the Study).
 
@@
 for recombination events, adding a novel temporal dimension
-and demonstrating its theoretical relationship with the 
-size of the interval in base pairs. Thirdly,
-
-These underscore the utility of the ARG itself, and suggest
+and demonstrating a simple relationship with the 
+size of the interval in base pairs derived from theory.
+Thirdly, we demonstrate a clear relationship between
+genetic diversity and the rate of detectable recombination,
+with the best predictor of the number of recombination
+events over time being the product of global case numbers
+and lineage heterozygosity.
+Finally, and perhaps most fundamentally, through our 
+conservative parameter choices and real-time handling of 
+recombination events, we have provided a stringent lower
+bound on the amount of recombination that occured during
+the pandemic phase of SARS-CoV-2.
+Together, these  underscore the utility of the ARG itself, and suggest
 several important directions for future research.
 
-% Old discussions
+% Old discussions(s)
 

### 2025-11-04 latex/main.tex
Fixup pango X lineage table
Closes #580

@@
 Events in the ARG associated with Pango X lineages.
-Each row for Type I, III and IV events corresponds to
-the oldest node assigned to the specified Pango lineage and
-is the ancestor of all the nodes assigned to that label.
-Type II events are recombination events
-associated with Pango X lineages not meeting Type I criteria.
+Each row for Type I, III and IV events (see main text for definitions) 
+corresponds to
+the origin node for a given Pango lineage.
 For Type I and II recombination events,
-the averted shows the number of additional mutations required
-under a recombination-free Viterbi solution.
+the ``averted'' column shows the number of additional mutations required
+without recombination (see main text).
 Parent lineages shown in boldface are nodes exactly matched by sampled sequences.
-The descendants column shows the descendant samples broken down by Pango lineage
-(grouped by root lineage, where appropriate).
-[FIXME Typesetting to be finished, describe cols for Type III and IV]
+The descendants column shows the descendant samples broken down by Pango lineage.
+For Type III events, the focal pango, event (Type I or II) it is derived 
+from and the distance (path length and time in days) from that event are shown,
+as well as the number of mutations directly ancestral to the pango origin
+node. For Type IV events, the pango designation of the parent node and number
+of mutations are shown.
 }
@@
 \toprule
-\multicolumn{7}{l}{Type I event: Recombination coinciding with Pango origin node}\\
+\multicolumn{7}{l}{\textbf{Type I event: Recombination coinciding with Pango
+origin node}}\\
 &&\multicolumn{2}{c}{Parent lineage} & \multicolumn{2}{c}{Interval}\\
@@
 \midrule
-\multicolumn{7}{l}{Type II event: Recombination closely associated with pango lineage(s)}\\
+\multicolumn{7}{l}{\textbf{Type II event: Recombination closely associated with
+pango lineage(s)}}\\
 &&\multicolumn{2}{c}{Parent lineage} & \multicolumn{2}{c}{Interval}\\
@@
 \midrule
-\multicolumn{7}{l}{Type III events: Pango origin nodes derived from 
-Type I and II events}\\
-pango & name & path len & time & mutations & & descendants\\
+\multicolumn{7}{l}{\textbf{Type III events: Pango origin nodes derived from 
+Type I and II events}}\\
+pango & event & path len & time & mutations & & descendants\\
 \midrule
@@
 \midrule
-\multicolumn{7}{l}{Type IV events: Non recombinants in the ARG}\\
-pango & muts &&&& parent & descendants \\
+\multicolumn{7}{l}{\textbf{Type IV events: Non recombinants in the ARG}}\\
+pango & parent &&& mutations && descendants \\
 \midrule
 
-XAS & 1  &&&& BA.4 & {XAS:77} \\
-XAU & 1  &&&& BA.2 & {XAU:8} \\
-XBK & 1  &&&& CJ.1 & {XBK:7} \\
-XBQ & 1  &&&& CJ.1 & {XBQ:14} \\
-XN  & 1  &&&& \textbf{BA.2} & {XN:120, BA.2:1} \\
-XP  & 1  &&&& \textbf{BA.1.1} & {XP:45} \\
-XB  & 2  &&&& B.1 & {XB:192} \\
-XAV & 3  &&&& BA.5.1.24 & {XAV:13} \\
-XAZ & 3  &&&& \textbf{BA.5} & {XAZ:133} \\
-XBE & 3  &&&& BA.5.2 & {XBE:65} \\
-XAN & 5  &&&& \textbf{BA.5.1} & {XAN:7} \\
-XAJ & 11 &&&& \textbf{BA.2.12} & {XAJ:18} \\
-
+XAS &  BA.4            &&& 1 && {XAS:77} \\
+XAU &  BA.2            &&& 1 && {XAU:8} \\
+XBK &  CJ.1            &&& 1 && {XBK:7} \\
+XBQ &  CJ.1            &&& 1 && {XBQ:14} \\
+XN  &  \textbf{BA.2}   &&& 1 && {XN:120, BA.2:1} \\
+XP  &  \textbf{BA.1.1} &&& 1 && {XP:45} \\
+XB  &  B.1             &&& 2 && {XB:192} \\
+XAV &  BA.5.1.24       &&& 3 && {XAV:13} \\
+XAZ &  \textbf{BA.5}   &&& 3 && {XAZ:133} \\
+XBE &  BA.5.2          &&& 3 && {XBE:65} \\
+XAN &  \textbf{BA.5.1} &&& 5 && {XAN:7} \\
+XAJ &  \textbf{BA.2.12}&&& 11&& {XAJ:18} \\
 \bottomrule
@@
 The first recombinant lineage XA was designated in 2021
-based on the analysis of Jackson et al.~\cite{Jackson2021},
-which established that recombination with onward transmission had occurred.
+based on the analysis of Jackson et al.~\cite{Jackson2021}.
 Although the process of incorporating recombinant lineages
@@
 
-The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
-are recombination events closely associated with Pango X lineages, 
-but do not meet the strict origination criteria for Type I.
+There are 5 partially concordant events (Type II) 
+in Table~\ref{tab:pango_x_lineages}, which 
+are recombination events closely associated with Pango X lineages
+that do not meet the strict origination criteria for Type I.
 Type III events are Pango origin nodes in the ARG that are 
@@
 XAA, XAG, XAM, XR and XU are all closely derived, suggesting 
-% NOTE: deleted the "sister" bit because it's confusing when
-% reading the document at this level. Keep the detailed hyptothese
-% to the deep dives.
-that they are more parsimoniously explained as sublineages of XQ
+that they may be more parsimoniously explained as sublineages of XQ
 than de-novo recombinant lineages (Document~\ref{sec:dd_xq}). 
@@
 The final class of events (Type IV) in Table~\ref{tab:pango_x_lineages}
-corresponds to nodes that are not associated with a recombination event.
+corresponds to nodes that are not associated with a recombination event
+in the ARG.
 These vary in their plausibility of being non-recombinant.
@@
 We further examined the mutations associated with two major branches of Delta.
-Using a phylogenetic analysis, Stern et al. \cite{Stern2021} suggest that
+Using a phylogenetic analysis, Stern et al.\cite{Stern2021} suggest that
 the Delta sublineages can be classified into distinct clades and
@@
 If samples from these parental lineages were available for attachment,
-then XB would be detected as a recombinant by sc2ts.
+then XB would likely be detected as a recombinant by sc2ts.
 % which also contains the key ``R2'' deletion
@@
 Note that the two lower recombination nodes are both well supported,
-the youngest being postulated as the origin of XAF (see below);
-we leave investigation of this suggested series of
-sequential recombination events to future work.
+the youngest being postulated as the origin of XAF (see below).
+% we leave investigation of this suggested series of
+% sequential recombination events to future work.
 

### 2025-11-04 latex/main.tex
Clear up details on Pango X lineage events

@@
 \label{sec:supplement_x_pango_lineages}
-[TODO DOCUMENT DETAILS of deriving events]
-
-The nested samples not assigned to the focal pango lineage are
-either from different pango X lineages (see next paragraph), 
+Most of the Pango X lineages had straightforward origination events 
+as defined in Section~\ref{sec:supplement_pango_lineages} 
+and are represented in Table~\ref{tab:pango_x_lineages}.
+Three lineages were treated specially, 
+corresponding to 23 omitted samples in total.
+
+XM has 4 separate origins, 3 singletons and one dominant node
+with 26 descendants. We include the dominant node in 
+Table~\ref{tab:pango_x_lineages} and omit the singletons. See
+Document~\ref{sec:dd_xm} for further details.
+XAC has 4 separate origins, and has been omitted from 
+Table~\ref{tab:pango_x_lineages}. However, as Figure~\ref{fig:subgraphs}B
+shows, these are all descending from a single BA.2 classified node 
+descending from the putative Xx event. 
+Similarly, XAD has two independent origins descending from the Xx
+event, and we omit it for simplicity.
+See Document~\ref{sec:dd_xx} for more details on the Xx event.
+
+In cases where nested samples not assigned to the focal pango lineage are
+either from different pango X lineages 
 or given a generic BA.2 assignment. Of these 59 generic assignents
@@
 2,747,985 alignments).
+
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/tab\_pango\_x\_events.ipynb}
+{tab\_pango\_x\_events.ipynb} for details.
 

### 2025-11-04 latex/main.tex

@@
 In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
-and rebar recombination detection methods
+and rebar (\url{https://github.com/phac-nml/rebar}) recombination detection methods
 on the putative recombinant sequences,

### 2025-11-04 latex/main.tex
Updates from Overleaf

@@
 are nested under the XQ recombinant lineage.
-However, upon closer inspection,
-the right hand branch above the recombination node appears problematic,
+These lineages were designated around the same time,
+roughly within a five-month period (Mar. to Jul., 2022).
+Also, their breakpoint intervals of these lineages,
+as suggested in the Pango designation issues,
+all occur at the 5' end of the genome
+within a $\sim$5,000 base segment ($\sim$4,300 to $\sim$9,400).
+
+Upon closer inspection,
+the right-hand branch above the recombination node appears problematic,
 as it displays three reversion mutations,

### 2025-11-04 latex/main.tex
Updates from Overleaf

@@
 \usepackage{amssymb}
+\usepackage{placeins}
 \usepackage{orcidlink}
@@
 ``averted'' by recombination---is a measure of
-the evidence in favor of recombination.
-% TODO
-[FIXME need to insert a sentence saying this is similar to
-metrics used by RIPPLES and other methods, contrasting
-recombinant vs non-recombinant explanations.]
+the evidence in favor of recombination (similar to the  
+``parsimony improvement'' score in RIPPLES~\cite{Turakhia2022}).
 Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
@@
 among the most weakly supported, with 4 or 5 averted mutations.
-[TODO discuss B and C]
+Figures~\ref{fig:recombinant_evidence}B and C shows the 
+distribution of recombination events as a function the 
+divergence between the parents.
 Figures~\ref{fig:recombinant_evidence_cov_recomb} and
 ~\ref{fig:recombinant_evidence_rebar} 
-plot the same information for CovRecomb and rebar validation,
-showing that their validation status tends to be function
-of the divergence between the recombinant parents.
-
-% SHZ: We should be more careful with the wording.
-% Recombination is likely frequent,
-% but detectable recombination
-% leading to lineages that are competitive enough to be sampled
-% compared to their (VOC) parents
-% is rare.
-Recombination is rare in SARS-CoV-2, and
+plot the same information as Figure~\ref{fig:recombinant_evidence} 
+broken down by CovRecomb and rebar validation status.
+
+Detectable recombination is rare in SARS-CoV-2, and
 ultimately judgments about plausibility require expert analysis.
@@
 or where the recombination event requires
-a \textit{de novo} mutation (gold, with mutational change below).
-[FIXME: Should we also mention the red/orange lines
-under the patterns that identify adjacent sites?]
+a de novo mutation (gold, with mutational change below).
 Figure~\ref{fig:recombinant_evidence}D shows some example copying patterns,
@@
 shown by the large number of supporting loci on the left and right.
-As we have only one \textit{de novo} mutation,
-we can be sure that the parents are closely observed
-[FIXME: Some details about XA here, how well the parents were sampled, etc.].
 XBB is also strongly supported by many loci on the left and right, 
-but the presence of many \textit{de novo} mutations
+but the presence of many de novo mutations
 observed in the copying patterns suggests the existence of
 many genetically distinct, un-sampled ancestors
-[FIXME: in this case, probably BJ.1, see issue 336],
-and some caution should be exercised when interpreting them.
+and some caution should be exercised when interpreting them
+(see Document~\ref{sec:dd_xbb} for more details).
 See Figures~\ref{fig:recombinant_evidence}D and
@@
 here, and do not perform QC filtering of either.
-Using the RIPPLES parameter $p=4$ (equivalent to $k=4$ used 
-to build the sc2ts ARG) resulted in 1174 reported recombination events.
+Using the RIPPLES parameter $p=4$ (matching $k=4$ in the sc2ts ARG) resulted in 1174 reported recombination events.
 We ran 3SEQ on each of the candidate trios (STAR Methods) and 601 
@@
 
+\FloatBarrier
+
 \section*{Supplementary Figures}

### 2025-11-04 latex/main.tex
Updates from Overleaf

--- 2025-11-04 latex/main.tex
@@
 It is important to emphasise that the ``true'' truth set among
-the Pango X lineages is currently unknown, and there is significant 
-disagreement among methods beyond a core set of lineages 
-(see Discussion).
+the Pango X lineages is currently unknown.
 

### 2025-11-05 latex/main.tex
Updates from Overleaf

@@
 These lineages were designated around the same time,
-roughly within a five-month period (Mar. to Jul., 2022).
-Also, their breakpoint intervals of these lineages,
+approximately within a five-month period (Mar. to Jul., 2022).
+The breakpoint intervals of these lineages,
 as suggested in the Pango designation issues,
-all occur at the 5' end of the genome
-within a $\sim$5,000 base segment ($\sim$4,300 to $\sim$9,400).
+all occur at the 5' end of the genome,
+within a $\sim$5 kb segment ($\sim$4,300 to $\sim$9,400):
+% TODO: Add URLs to Pango issues?
+XQ (4321-5387), % https://github.com/cov-lineages/pango-designation/issues/468
+XAG (6515-8394), % https://github.com/cov-lineages/pango-designation/issues/709
+XU (6517-9345), % https://github.com/cov-lineages/pango-designation/issues/522
+XAM (8087-9193), % https://github.com/cov-lineages/pango-designation/issues/759
+and XAA (8392-9345). % https://github.com/cov-lineages/pango-designation/issues/664
 

### 2025-11-05 latex/main.tex
Major rejigging for first results section

@@
 64 cores (2X AMD EPYC 7502) and 512 GB of RAM
-($\sim$5 CPU years; Figure~\ref{fig:inference_resources}).
+($\sim$5 CPU years;
+Figures~\ref{fig:inference_resources}~\ref{fig:samples_per_day}).
 After post-processing, the ARG contains
@@
 2,475,418 samples and 27,507 sites (STAR Methods).
-% JK: this feels a bit wishy washy, but there really is no simple answer to
-% this and I don't want to claim we're more parsimonious.
-Eliminating recombination nodes and their descendants (which we expected to differ), we found very similar relationships among  lineages in terms of parsimony
-(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
-and imputation of missing and ambiguous bases
-(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
-We also compared the mutational spectra of major VOCs in the ARG
-against the UShER tree and a previous study~\cite{Bloom2023} (STAR Methods),
-finding close agreement (Figure~\ref{fig:mutational_spectra}).
-We validated that the broad phylogenetic structure of
-the ARG and UShER tree closely match by generating tanglegrams
+Eliminating recombination nodes and their descendants (which we expected to differ), 
+we found very similar relationships among lineages in
+the ARG and UShER tree based on tanglegrams
 for a set of representative samples from 814 well-sampled lineages
@@
 \ref{fig:tanglegram_ba5}).
+% JK: this feels a bit wishy washy, but there really is no simple answer to
+% this and I don't want to claim we're more parsimonious.
+Overall, we also found close similarity between the ARG
+and UShER tree in terms of parsimoniousness 
+(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
+and imputation of missing and ambiguous bases
+(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
+We compared the mutational spectra of major VOCs in the ARG
+against the UShER tree and a previous study~\cite{Bloom2023} (STAR Methods),
+finding close agreement (Figure~\ref{fig:mutational_spectra}).
 We also evaluated how well the ARG reflects the structure of
@@
 Chronumental~\cite{sanderson2021chronumental} applied to 
-an UShER tree (Document~\ref{sec:evaluating_node_dates}).
+an UShER tree (Figure~\ref{fig:node_time_testing},
+Document~\ref{sec:evaluating_node_dates}).
 Finally, we verified
@@
 \renewcommand{\thesection}{S\arabic{section}}
+
+\subsection{Parsimony comparison of sc2ts and UShER}
+\label{sec:supplement_sc2ts_usher_parsimony}
+
+We performed an in-depth analysis of the sc2ts and UShER ARGs
+(we refer to the UShER tree as an ARG for simplicity here,
+as a single tree is also an ARG) in terms of parsimony.
+A direct comparison is difficult because of
+the different alignments used as input,
+the different sets of sites used for inference and
+the fact that we are including deletions
+for a subset of sites in the sc2ts ARG.
+
+When we compare the intersection of the base sc2ts ARG
+(pre site-remapping), sc2ts is very slightly more parsimonious
+% TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/684
+with XX mutations over the YYY sites shared between them.
+After site remapping, and including all mutations involving
+deletion characters, sc2ts has 143,690 mutations more
+(2,078,370 vs 1,934,680). If we exclude mutations involving
+deletions in sc2ts, this is reduced to 46,371 more (2.3\%) than UShER.
+If we restrict to the 26,192 sites that were not remapped,
+sc2ts has 4,785 (0.26\%) fewer mutations than UShER,
+and has around 2.5 times more mutations than UShER
+on the 163 sites that were post-hoc remapped.
+
+To provide a more objective measure of the overall parsimony 
+of the inferred ARGs with respect to the Viridian alignments,
+we remapped all 29903 sites in both ARGs 
+using Fitch-Hartigan parsimony. Here the nucleotides
+ACGT and ``-'' were treated as distinct characters, with all 
+other values regarded as missing data. In this case, the sc2ts 
+ARG has 629,864 (5.5\%) fewer mutations (10,840,687 
+vs 11,470,551). Some of this difference may be due to the 
+post-hoc incorporation of phylogenetic signal from deletions 
+by sc2ts, but more work is required to fully understand 
+the differences between the two approaches.
+
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_usher\_sc2ts.ipynb}{analysis\_usher\_sc2ts.ipynb}
+for details.
+
+\subsection{Imputation comparison of sc2ts and UShER}
+\label{sec:supplement_sc2ts_usher_imputation}
+
+Both sc2ts and UShER support missing data in input alignments, and effectively
+``hard call'' these values when the samples are integrated into the tree/ARG. 
+In doing so, we impute the missing and ambiguous data for these samples.
+We evaluate the imputation performance of sc2ts and UShER by comparing their
+outputs on the Viridian alignments.
+
+The total aligned dataset for 2,475,418 samples and 27,507 sites 
+in the sc2ts and UShER intersection ARGs represents
+63.41 GiB of nucleotide calls.
+Of the 162,838,461 missing data calls (Ns) in the alignments,
+sc2ts and UShER disagreed in their imputed values for only 203,579 (0.125\%).
+Additionally, 956,859 calls made use of the IUPAC uncertainty codes. 
+Of these sc2ts imputed 435 (0.05\%) incorrectly
+(i.e., with a base that is not compatible with the ambiguity code).
+UShER imputed only 7 calls (0.007\%) from this set incorrectly,
+likely due to special handling of
+ambiguous bases in the parsimony calculations.
+
+This computation is a good example of
+type of integrated large-scale data analysis
+that can be easily performed using the combination of tskit and VCF Zarr.
+The entire analysis is performed in a short Jupyter
+notebook~\cite{kluyver2016jupyter},
+with the core calculation requiring about 30 lines of Python code.
+Retrieving data from the base alignments and ARGs for 2.48 million samples 
+at the same time over all 27,507 sites took only 6 minutes,
+with a peak memory usage of about 5GiB of RAM
+on a standard desktop computer.
+
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/example\_data\_processing.ipynb}{example\_data\_processing.ipynb}
+for details.
+
+\subsection{Pango lineage ``origination'' events}
+\label{sec:supplement_pango_lineages}
+
+To analyze how well the ARG reflects the phylogenetic structure implicit in the
+Pango lineage naming system, we considered the Pango assignments generated by
+Pangolin on the alignments for each node in the ARG (STAR Methods). If the ARG
+perfectly reflected the Pango lineage structure, each lineage would form a
+clade descending from a single originating node with that label. We identified
+putative origination nodes among those labeled with a given lineage as the node
+with the maximum number of descendant samples and the earliest time (if there
+is a tie).
+
+Of the 2058 distinct pango lineages in the ARG, 1473 of these (comprising
+717798 samples) match perfectly, with unique origination events in the ARG
+where all samples assigned a given lineage descend from the first node assigned
+that lineage. A further 245 lineages (589197 samples) match perfectly when we
+count the descendants of the parent of the first node (accounting for
+polytomies in which multiple originating nodes for a given lineage are
+siblings). We then have 306 lineages (755840 samples) where the difference in
+the number descendants of the first node's parent is $<$ 100. The remaining 35
+lineages (420047 samples) are dominated by a few large lineages such as BA.1.1
+(147271 samples) and AY.4.2 (54607 samples) which have multiple non-sibling
+origins within the ARG.
+
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_pango\_events\_in\_arg.ipynb}
+{analysis\_pango\_events\_in\_arg.ipynb}
+for details.
+
+\subsection{Evaluating inferred node dates}
+\label{sec:evaluating_node_dates}
+
+To validate the accuracy of times for non-sample internal nodes
+inserted into the ARG (inferred using tsdate),
+we compared the dates of first emergence of
+Nextstrain clades against the Nextstrain SARS-CoV-2 tree
+(downloaded on 2023-01-21).
+Figure~\ref{fig:pango_vs_nextstrain_node_times} shows
+relatively close agreement for most clades,
+with only three disagreeing by more than 28 days:
+20C (an early lineage from 2020),
+20E (Pango B.1.177 also known as EU1, emerging mid-2020), and
+21B (Pango B.1.617.1 also known as Kappa, emerging in late 2020 to early 2021).
+
+Then, to gauge internal consistency of the time estimation procedure,
+we selected a subset of sample nodes,
+matching the distribution of node times  and
+average number of children of the non-sample internal nodes.
+These selected sample nodes were then re-dated, and
+the inferred dates compared against their recorded collection dates.
+The results in Figure~\ref{fig:node_time_testing} demonstrate that
+although the inferred and true times are generally very close,
+there is a clear bias with re-estimated times
+being consistently older (by an average of 37 days).
+This is in line with the findings of
+other recent work \cite{sanderson2021chronumental}, and
+is likely to be a consequence of edges
+carrying more mutations than would be expected
+under a neutral molecular clock prior
+(either due to errors in tree-building or genuinely varying mutation rates),
+as well as inaccuracies in the recording of sample dates.
+
+\subsection{Origins of major VOCs}
+
+% https://github.com/jeromekelleher/sc2ts-paper/issues/772
+\subsubsection{Alpha}
+\label{sec:dd_alpha}
+
+The B.1.1.7 (Alpha) variant emerged in the UK as the first major VOC.
+Its origin is well reconstructed in the ARG,
+with Figure \ref{fig:subgraph_alpha}
+showing it to be a saltational lineage
+characterized by a long branch preceding its origin.
+Most of the characteristic Alpha mutations \cite{RambautAlpha2020}
+(20 out of 23) in the ARG
+occur immediately above the Alpha root.
+However, a characteristic three-base correlated mutation
+is instead associated with the parent node (labeled B.1.1) of the Alpha root,
+suggesting that this mutation was acquired prior to the emergence of Alpha.
+This parent node has an additional 5 children
+(leading to 12 samples),
+of which we display one (ERR4413600)
+with no associated mutations,
+meaning that it has an identical sequence
+to the inferred ancestor of Alpha.
+Below the root of Alpha, a single sample (ERR5178290)
+appears without the characteristic mutation C3267T
+(or any additional mutations).
+However, this is only a single sample and
+given that it also appears two months after the first Alpha samples
+we treat its placement with caution.
+Note that the list of mutations in ref\cite{RambautAlpha2020}
+includes a T26801C in the M gene
+which should no longer be considered characteristic of Alpha,
+as the Wuhan-Hu-1 reference at 26,801 has since been updated to C.
+In the ARG, a one-base non-coding deletion (position 28,271)
+is also associated with the Alpha root;
+although this is not in the original list of
+characteristic Alpha mutations,
+it has been suggested elsewhere as an Alpha-defining 
+mutation\cite{Li2023indels}.
+
+\subsubsection{Delta}
+\label{sec:dd_delta}
+
+The B.1.617.2 lineage (Delta) was first detected in India \cite{ECDCreport}.
+The subgraph in Figure~\ref{fig:subgraph_delta} shows
+a mutation-rich branch leading to the Delta root,
+which contains 11 single nucleotide mutations and a 6-base deletion.
+The placement of this branch in the ARG
+is a result of long branch splitting of the lineage above B.1.617.
+
+Delta belongs to the ``B.1.617'' clade,
+which also includes Kappa (B.1.617.1) and B.1.617.3 \cite{ECDCreport}.
+Previous phylogenetic analyses suggest that
+Delta and B.1.617.3 are more closely related to each other
+than either is to Kappa\cite{Cherian2021,Farinholt2021}.
+In the ARG, however, Delta and Kappa appear as sibling lineages
+(see the subgraph in Figure~\ref{fig:subgraph_delta}).
+This reflects the closest approximation to the relationships
+suggested in the previous studies that could be inferred,
+given that no B.1.617.3 samples passed our QC filters and
+no seed B.1.617.3 samples were specified for ARG inference.
+
+In the ARG, all six of the shared Delta-Kappa mutations (red)
+occur above the B.1-labeled common ancestor
+of Kappa (left branch) and Delta (right branch).
+All nine of the Kappa-specific mutations (blue)
+occur above the B.1.617-labeled node leading to Kappa samples.
+Seven out of nine Delta-specific mutations (purple)
+occur above the Delta root
+(the other two are above the MRCA of Kappa and Delta).
+We further examined the mutations associated with two major branches of Delta.
+Using a phylogenetic analysis, Stern et al.\cite{Stern2021} suggest that
+the Delta sublineages can be classified into distinct clades and
+identified the characteristic mutations of these clades.
+In the Nextstrain phylogeny shown in Stern et al. \cite{Stern2021},
+the first major split separates clades A to D from clade E;
+this topology is reflected in the ARG.
+All three of the clade A-D mutations
+(G15451A, C16466T, and a 6-base deletion at 28,248)
+occur on the uppermost branch of Delta (which corresponds to clades A to D).
+There are ten clade E mutations:
+five occur on the right branch of Delta (which corresponds to clade E) and
+three (C1191T, C5184T, and C28253T) occur above the Delta root.
+The placement of these mutations in the ARG generally reflects
+what is proposed in Stern et al. \cite{Stern2021}.
+
+There are four positions with multiple mutations
+(orange at position 1267, blue at 210, green at 20396, and red at 25276).
+Of these, the only clear topological change
+that could make the graph more parsimonious would be
+to attach the node above SRR12316669 one position higher up the tree,
+which would allow the placement of two recurrent red mutations
+(a C25276T above SRR12316669 and another above ERR4561562 / B.1.617)
+rather than one C25276T mutation followed by two reversions.
+The other recurrent mutations can only be made more parsimonious
+by topological rearrangements that would cause recurrence at other sites.
+We therefore believe that the inferred Delta origins
+in the sc2ts ARG are largely correct.
+
+\subsubsection{Omicron BA.1 and BA.2}
+\label{sec:dd_omicron}
+
+The B.1.1.529 lineage (Omicron)
+is believed to have emerged in southern Africa, and
+it is a saltational lineage with two highly successful sublineages,
+BA.1 and BA.2 \cite{Dor2025}.
+BA.1 and BA.2 are genetically distinct from each other and
+from their parental B.1.1.529 lineage \cite{RambautOmicron2021}.
+In the subgraph shown in Figure~\ref{fig:subgraph_omicron},
+BA.1 and BA.2 differ from the non-Omicron B.1.1 ancestor
+(above the B.1.1.529-labeled MRCA of BA.1 and BA.2)
+by 27 mutations (including two deletions),
+reflecting its saltational origin.
+The ARG captures the split between BA.1 and BA.2
+shown in previous phylogenetic analysis \cite{Dor2025}.
+The 5-base deletion highlighted in green above the BA.2 root
+is an artefact of our naive deletion remapping procedure,
+which splits two 9-base deletions in BA.1 and BA.2 into three deletions
+(one in green at 11288, one in blue at 11283, and one in orange 11292;
+see Document~\ref{sec:supplement_dels}).
+Also, there is uncertainty in the sequence alignments
+leading to ambiguity in the start position of these deletions
+(if these deletions occur at 11283,
+there would be a single deletion event before the BA.1/BA.2 split),
+as noted in the Pango designation issue \cite{RambautOmicron2021}.
+
+We selected a BA.1 seed sample that led to the fewest mutations.
+This strategy resulted in five recurrent mutations
+highlighted in blue/orange/green/red/purple
+which are absent in the chosen seed
+but present in many of its inferred descendants.
+Inspection of the subgraph reveals that an alternative choice for a seed would be the later sample SRR17051902. Seeding with this sample would be likely to generate a topology
+in which those five recurrent mutations are shared between BA.1 and BA.2 and subsequently reverted in a number of the deep branching BA.1 lineages such as those to ERR7671333 and ERR7858953.
 
@@
 
-\subsection{Parsimony comparison of sc2ts and UShER}
-\label{sec:supplement_sc2ts_usher_parsimony}
-
-We performed an in-depth analysis of the sc2ts and UShER ARGs
-(we refer to the UShER tree as an ARG for simplicity here,
-as a single tree is also an ARG) in terms of parsimony.
-A direct comparison is difficult because of
-the different alignments used as input,
-the different sets of sites used for inference and
-the fact that we are including deletions
-for a subset of sites in the sc2ts ARG.
-
-When we compare the intersection of the base sc2ts ARG
-(pre site-remapping), sc2ts is very slightly more parsimonious
-% TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/684
-with XX mutations over the YYY sites shared between them.
-After site remapping, and including all mutations involving
-deletion characters, sc2ts has 143,690 mutations more
-(2,078,370 vs 1,934,680). If we exclude mutations involving
-deletions in sc2ts, this is reduced to 46,371 more (2.3\%) than UShER.
-If we restrict to the 26,192 sites that were not remapped,
-sc2ts has 4,785 (0.26\%) fewer mutations than UShER,
-and has around 2.5 times more mutations than UShER
-on the 163 sites that were post-hoc remapped.
-
-To provide a more objective measure of the overall parsimony 
-of the inferred ARGs with respect to the Viridian alignments,
-we remapped all 29903 sites in both ARGs 
-using Fitch-Hartigan parsimony. Here the nucleotides
-ACGT and ``-'' were treated as distinct characters, with all 
-other values regarded as missing data. In this case, the sc2ts 
-ARG has 629,864 (5.5\%) fewer mutations (10,840,687 
-vs 11,470,551). Some of this difference may be due to the 
-post-hoc incorporation of phylogenetic signal from deletions 
-by sc2ts, but more work is required to fully understand 
-the differences between the two approaches.
-
-See notebook
-\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_usher\_sc2ts.ipynb}{analysis\_usher\_sc2ts.ipynb}
-for details.
-
-\subsection{Imputation comparison of sc2ts and UShER}
-\label{sec:supplement_sc2ts_usher_imputation}
-
-Both sc2ts and UShER support missing data in input alignments, and effectively
-``hard call'' these values when the samples are integrated into the tree/ARG. 
-In doing so, we impute the missing and ambiguous data for these samples.
-We evaluate the imputation performance of sc2ts and UShER by comparing their
-outputs on the Viridian alignments.
-
-The total aligned dataset for 2,475,418 samples and 27,507 sites 
-in the sc2ts and UShER intersection ARGs represents
-63.41 GiB of nucleotide calls.
-Of the 162,838,461 missing data calls (Ns) in the alignments,
-sc2ts and UShER disagreed in their imputed values for only 203,579 (0.125\%).
-Additionally, 956,859 calls made use of the IUPAC uncertainty codes. 
-Of these sc2ts imputed 435 (0.05\%) incorrectly
-(i.e., with a base that is not compatible with the ambiguity code).
-UShER imputed only 7 calls (0.007\%) from this set incorrectly,
-likely due to special handling of
-ambiguous bases in the parsimony calculations.
-
-This computation is a good example of
-type of integrated large-scale data analysis
-that can be easily performed using the combination of tskit and VCF Zarr.
-The entire analysis is performed in a short Jupyter
-notebook~\cite{kluyver2016jupyter},
-with the core calculation requiring about 30 lines of Python code.
-Retrieving data from the base alignments and ARGs for 2.48 million samples 
-at the same time over all 27,507 sites took only 6 minutes,
-with a peak memory usage of about 5GiB of RAM
-on a standard desktop computer.
-
-See notebook
-\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/example\_data\_processing.ipynb}{example\_data\_processing.ipynb}
-for details.
-
-\subsection{Pango lineage ``origination'' events}
-\label{sec:supplement_pango_lineages}
-
-To analyze how well the ARG reflects the phylogenetic structure implicit in the
-Pango lineage naming system, we considered the Pango assignments generated by
-Pangolin on the alignments for each node in the ARG (STAR Methods). If the ARG
-perfectly reflected the Pango lineage structure, each lineage would form a
-clade descending from a single originating node with that label. We identified
-putative origination nodes among those labeled with a given lineage as the node
-with the maximum number of descendant samples and the earliest time (if there
-is a tie).
-
-Of the 2058 distinct pango lineages in the ARG, 1473 of these (comprising
-717798 samples) match perfectly, with unique origination events in the ARG
-where all samples assigned a given lineage descend from the first node assigned
-that lineage. A further 245 lineages (589197 samples) match perfectly when we
-count the descendants of the parent of the first node (accounting for
-polytomies in which multiple originating nodes for a given lineage are
-siblings). We then have 306 lineages (755840 samples) where the difference in
-the number descendants of the first node's parent is $<$ 100. The remaining 35
-lineages (420047 samples) are dominated by a few large lineages such as BA.1.1
-(147271 samples) and AY.4.2 (54607 samples) which have multiple non-sibling
-origins within the ARG.
-
-See notebook
-\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_pango\_events\_in\_arg.ipynb}
-{analysis\_pango\_events\_in\_arg.ipynb}
-for details.
-
-\subsection{Evaluating inferred node dates}
-\label{sec:evaluating_node_dates}
-
-To validate the accuracy of times for non-sample internal nodes
-inserted into the ARG (inferred using tsdate),
-we compared the dates of first emergence of
-Nextstrain clades against the Nextstrain SARS-CoV-2 tree
-(downloaded on 2023-01-21).
-Figure~\ref{fig:pango_vs_nextstrain_node_times} shows
-relatively close agreement for most clades,
-with only three disagreeing by more than 28 days:
-20C (an early lineage from 2020),
-20E (Pango B.1.177 also known as EU1, emerging mid-2020), and
-21B (Pango B.1.617.1 also known as Kappa, emerging in late 2020 to early 2021).
-
-Then, to gauge internal consistency of the time estimation procedure,
-we selected a subset of sample nodes,
-matching the distribution of node times  and
-average number of children of the non-sample internal nodes.
-These selected sample nodes were then re-dated, and
-the inferred dates compared against their recorded collection dates.
-The results in Figure~\ref{fig:node_time_testing} demonstrate that
-although the inferred and true times are generally very close,
-there is a clear bias with re-estimated times
-being consistently older (by an average of 37 days).
-This is in line with the findings of
-other recent work \cite{sanderson2021chronumental}, and
-is likely to be a consequence of edges
-carrying more mutations than would be expected
-under a neutral molecular clock prior
-(either due to errors in tree-building or genuinely varying mutation rates),
-as well as inaccuracies in the recording of sample dates.
-
 \subsection{Pango X lineage events}
@@
 
-\subsection{Origins of major VOCs}
-
-% https://github.com/jeromekelleher/sc2ts-paper/issues/772
-\subsubsection{Alpha}
-\label{sec:dd_alpha}
-
-The B.1.1.7 (Alpha) variant emerged in the UK as the first major VOC.
-Its origin is well reconstructed in the ARG,
-with Figure \ref{fig:subgraph_alpha}
-showing it to be a saltational lineage
-characterized by a long branch preceding its origin.
-Most of the characteristic Alpha mutations \cite{RambautAlpha2020}
-(20 out of 23) in the ARG
-occur immediately above the Alpha root.
-However, a characteristic three-base correlated mutation
-is instead associated with the parent node (labeled B.1.1) of the Alpha root,
-suggesting that this mutation was acquired prior to the emergence of Alpha.
-This parent node has an additional 5 children
-(leading to 12 samples),
-of which we display one (ERR4413600)
-with no associated mutations,
-meaning that it has an identical sequence
-to the inferred ancestor of Alpha.
-Below the root of Alpha, a single sample (ERR5178290)
-appears without the characteristic mutation C3267T
-(or any additional mutations).
-However, this is only a single sample and
-given that it also appears two months after the first Alpha samples
-we treat its placement with caution.
-Note that the list of mutations in ref\cite{RambautAlpha2020}
-includes a T26801C in the M gene
-which should no longer be considered characteristic of Alpha,
-as the Wuhan-Hu-1 reference at 26,801 has since been updated to C.
-In the ARG, a one-base non-coding deletion (position 28,271)
-is also associated with the Alpha root;
-although this is not in the original list of
-characteristic Alpha mutations,
-it has been suggested elsewhere as an Alpha-defining 
-mutation\cite{Li2023indels}.
-
-\subsubsection{Delta}
-\label{sec:dd_delta}
-
-The B.1.617.2 lineage (Delta) was first detected in India \cite{ECDCreport}.
-The subgraph in Figure~\ref{fig:subgraph_delta} shows
-a mutation-rich branch leading to the Delta root,
-which contains 11 single nucleotide mutations and a 6-base deletion.
-The placement of this branch in the ARG
-is a result of long branch splitting of the lineage above B.1.617.
-
-Delta belongs to the ``B.1.617'' clade,
-which also includes Kappa (B.1.617.1) and B.1.617.3 \cite{ECDCreport}.
-Previous phylogenetic analyses suggest that
-Delta and B.1.617.3 are more closely related to each other
-than either is to Kappa\cite{Cherian2021,Farinholt2021}.
-In the ARG, however, Delta and Kappa appear as sibling lineages
-(see the subgraph in Figure~\ref{fig:subgraph_delta}).
-This reflects the closest approximation to the relationships
-suggested in the previous studies that could be inferred,
-given that no B.1.617.3 samples passed our QC filters and
-no seed B.1.617.3 samples were specified for ARG inference.
-
-In the ARG, all six of the shared Delta-Kappa mutations (red)
-occur above the B.1-labeled common ancestor
-of Kappa (left branch) and Delta (right branch).
-All nine of the Kappa-specific mutations (blue)
-occur above the B.1.617-labeled node leading to Kappa samples.
-Seven out of nine Delta-specific mutations (purple)
-occur above the Delta root
-(the other two are above the MRCA of Kappa and Delta).
-We further examined the mutations associated with two major branches of Delta.
-Using a phylogenetic analysis, Stern et al.\cite{Stern2021} suggest that
-the Delta sublineages can be classified into distinct clades and
-identified the characteristic mutations of these clades.
-In the Nextstrain phylogeny shown in Stern et al. \cite{Stern2021},
-the first major split separates clades A to D from clade E;
-this topology is reflected in the ARG.
-All three of the clade A-D mutations
-(G15451A, C16466T, and a 6-base deletion at 28,248)
-occur on the uppermost branch of Delta (which corresponds to clades A to D).
-There are ten clade E mutations:
-five occur on the right branch of Delta (which corresponds to clade E) and
-three (C1191T, C5184T, and C28253T) occur above the Delta root.
-The placement of these mutations in the ARG generally reflects
-what is proposed in Stern et al. \cite{Stern2021}.
-
-There are four positions with multiple mutations
-(orange at position 1267, blue at 210, green at 20396, and red at 25276).
-Of these, the only clear topological change
-that could make the graph more parsimonious would be
-to attach the node above SRR12316669 one position higher up the tree,
-which would allow the placement of two recurrent red mutations
-(a C25276T above SRR12316669 and another above ERR4561562 / B.1.617)
-rather than one C25276T mutation followed by two reversions.
-The other recurrent mutations can only be made more parsimonious
-by topological rearrangements that would cause recurrence at other sites.
-We therefore believe that the inferred Delta origins
-in the sc2ts ARG are largely correct.
-
-\subsubsection{Omicron BA.1 and BA.2}
-\label{sec:dd_omicron}
-
-The B.1.1.529 lineage (Omicron)
-is believed to have emerged in southern Africa, and
-it is a saltational lineage with two highly successful sublineages,
-BA.1 and BA.2 \cite{Dor2025}.
-BA.1 and BA.2 are genetically distinct from each other and
-from their parental B.1.1.529 lineage \cite{RambautOmicron2021}.
-In the subgraph shown in Figure~\ref{fig:subgraph_omicron},
-BA.1 and BA.2 differ from the non-Omicron B.1.1 ancestor
-(above the B.1.1.529-labeled MRCA of BA.1 and BA.2)
-by 27 mutations (including two deletions),
-reflecting its saltational origin.
-The ARG captures the split between BA.1 and BA.2
-shown in previous phylogenetic analysis \cite{Dor2025}.
-The 5-base deletion highlighted in green above the BA.2 root
-is an artefact of our naive deletion remapping procedure,
-which splits two 9-base deletions in BA.1 and BA.2 into three deletions
-(one in green at 11288, one in blue at 11283, and one in orange 11292;
-see Document~\ref{sec:supplement_dels}).
-Also, there is uncertainty in the sequence alignments
-leading to ambiguity in the start position of these deletions
-(if these deletions occur at 11283,
-there would be a single deletion event before the BA.1/BA.2 split),
-as noted in the Pango designation issue \cite{RambautOmicron2021}.
-
-We selected a BA.1 seed sample that led to the fewest mutations.
-This strategy resulted in five recurrent mutations
-highlighted in blue/orange/green/red/purple
-which are absent in the chosen seed
-but present in many of its inferred descendants.
-Inspection of the subgraph reveals that an alternative choice for a seed would be the later sample SRR17051902. Seeding with this sample would be likely to generate a topology
-in which those five recurrent mutations are shared between BA.1 and BA.2 and subsequently reverted in a number of the deep branching BA.1 lineages such as those to ERR7671333 and ERR7858953.
-
 \subsection{Additional analyses of the Pango X lineages}
@@
 \centering
+\includegraphics[width=\linewidth]{inference-resources}
+\caption{
+Computational resources used during the ARG inference process.
+(A) The amount of time to process a daily batch of samples
+and the number of samples processed per day.
+(B) Mean time to process a sample is less than 5s except
+when the number of samples is small. 
+(C) RAM usage is the most difficult resource to manage, and depends
+on the sample composition. The sharp drop in 2022-01 corresponds to the 
+sweep in which Omicron BA.1 replaced Delta and there was very little sample
+diversity. The drop in RAM usage around 2022-02 corresponds to a change
+in configuration where we reduced the number of HMM threads from 128 to 80
+to better fit into available memory. 
+}
+\label{fig:inference_resources}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics[width=\linewidth]{samples-per-day}
@@
 \centering
-\includegraphics[width=\linewidth]{inference-resources}
+\centering
+\includegraphics[width=\linewidth]{tanglegram_base_tree.pdf}
 \caption{
-Computational resources used during the ARG inference process.
-(A) The amount of time to process a daily batch of samples
-and the number of samples processed per day.
-(B) Mean time to process a sample is less than 5s except
-when the number of samples is small. 
-(C) RAM usage is the most difficult resource to manage, and depends
-on the sample composition. The sharp drop in 2022-01 corresponds to the 
-sweep in which Omicron BA.1 replaced Delta and there was very little sample
-diversity. The drop in RAM usage around 2022-02 corresponds to a change
-in configuration where we reduced the number of HMM threads from 128 to 80
-to better fit into available memory. 
+Tanglegram comparing the basic phylogenetic backbones of sc2ts and UShER.
 }
-\label{fig:inference_resources}
+\label{fig:tanglegram_base}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{tanglegram_Delta_subtree.pdf}
+\caption{
+Tanglegram comparing sc2ts and UShER on the Delta subtree.
+}
+\label{fig:tanglegram_delta}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{tanglegram_BA.2_subtree.pdf}
+\caption{
+Tanglegram comparing sc2ts and UShER on the BA.2 subtree.
+}
+\label{fig:tanglegram_ba2}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{tanglegram_BA.5_subtree.pdf}
+\caption{
+Tanglegram comparing sc2ts and UShER on the BA.5 subtree.
+}
+\label{fig:tanglegram_ba5}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{mutational_spectra}
+\caption{
+All-site mutational spectra of major VOCs calculated from the sc2ts ARG (A),
+the mutation count data from Bloom et al. (2023)\cite{Bloom2023} (B), and
+the Viridian UShER phylogeny from Hunt et al. (2024)\cite{Hunt2024}.
+}
+\label{fig:mutational_spectra}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.85\linewidth]{pango_vs_nextstrain_node_times.pdf}
+\caption{
+Inferred times of internal nodes corresponding to origins of Nextclade clades ($y$-axis) against those estimated in the Nextclade tree ($x$-axis).
+Confidence intervals shown as horizontal bars.
+Clades are labelled where dates differ by more than 28 days.
+}
+\label{fig:pango_vs_nextstrain_node_times}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.85\linewidth]{node_time_testing.pdf}
+\caption{
+Heat map showing differences (in days) between actual and re-estimated sample dates for a subset of internal sample nodes. 
+}
+\label{fig:node_time_testing}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.85\linewidth]{subgraph-Alpha.pdf}
+\caption{
+Subgraph illustrating the saltational origin of B.1.1.7 (Alpha).
+Squares represent sample nodes,
+circles represent reconstructed internal nodes;
+nodes classified as B.1.1.7 by Pangolin v4.3.1 are in pink.
+Mutations are shown as rectangles along branches,
+ordered by position; deletions are black.
+Characteristic mutations
+associated with the emergence of Alpha
+are highlighted with a pink outline.
+}
+\label{fig:subgraph_alpha}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.85\linewidth]{subgraph-Delta_recolored.pdf}
+\caption{
+Subgraph illustrating the origin of B.1.617.2 (Delta).
+Symbols are described in \ref{fig:subgraph_alpha}, except that 
+we highlight the characteristic mutations of Kappa and Delta:
+those with a blue outline are listed
+in the Pango designation issue \cite{PangoIssueKappa}
+or the Kappa constellation \cite{ScorpioKappa},
+but absent in the Delta constellation;
+those with a purple outline are listed
+in the Delta constellation \cite{ScorpioDelta}
+or are Delta-specific deletions reported in Stern et al. \cite{Stern2021},
+but absent in the Kappa constellation; and
+those with a red outline are listed
+in both the Kappa and Delta constellations.
+We also outline
+in orange the mutations associated with the lineage containing clades A to D and
+in green those associated with lineage containing clade E.
+Sites associated with multiple mutations in the subgraph
+have their mutations assigned a unique fill color
+(orange, blue, green, red, etc);
+reversions are further highlighted with a black outline.
+Sc2ts seed samples (see STAR methods) for
+B.1.617, B.1.617.1, and B.1.617.2 are plotted as red squares.
+}
+\label{fig:subgraph_delta}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.85\linewidth]{subgraph-Omicron_recolored.pdf}
+\caption{
+Subgraph illustrating the saltational origin of
+the major Omicron lineages BA.1 and BA.2.
+Symbols are described in Figure~\ref{fig:subgraph_delta},
+except that BA.1 samples are filled in light blue
+(also used to outline BA.1 characteristic mutations),
+BA.2 samples in light yellow
+(also used to outline BA.2 characteristic mutations), and
+B.1.1.529 mutations
+which are neither BA.1 or BA.2 characteristic mutations are
+outlined in green.
+These colors correspond to those used
+in the BA.1/BA.2 Pango designation issue \cite{RambautOmicron2021}.
+}
+\label{fig:subgraph_omicron}
 \end{figure}
@@
 \centering
-\centering
-\includegraphics[width=\linewidth]{tanglegram_base_tree.pdf}
-\caption{
-Tanglegram comparing the basic phylogenetic backbones of sc2ts and UShER.
-}
-\label{fig:tanglegram_base}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{tanglegram_Delta_subtree.pdf}
-\caption{
-Tanglegram comparing sc2ts and UShER on the Delta subtree.
-}
-\label{fig:tanglegram_delta}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{tanglegram_BA.2_subtree.pdf}
-\caption{
-Tanglegram comparing sc2ts and UShER on the BA.2 subtree.
-}
-\label{fig:tanglegram_ba2}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{tanglegram_BA.5_subtree.pdf}
-\caption{
-Tanglegram comparing sc2ts and UShER on the BA.5 subtree.
-}
-\label{fig:tanglegram_ba5}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.85\linewidth]{pango_vs_nextstrain_node_times.pdf}
-\caption{
-Inferred times of internal nodes corresponding to origins of Nextclade clades ($y$-axis) against those estimated in the Nextclade tree ($x$-axis).
-Confidence intervals shown as horizontal bars.
-Clades are labelled where dates differ by more than 28 days.
-}
-\label{fig:pango_vs_nextstrain_node_times}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.85\linewidth]{node_time_testing.pdf}
-\caption{
-Heat map showing differences (in days) between actual and re-estimated sample dates for a subset of internal sample nodes. 
-}
-\label{fig:node_time_testing}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.85\linewidth]{subgraph-Alpha.pdf}
-\caption{
-Subgraph illustrating the saltational origin of B.1.1.7 (Alpha).
-Squares represent sample nodes,
-circles represent reconstructed internal nodes;
-nodes classified as B.1.1.7 by Pangolin v4.3.1 are in pink.
-Mutations are shown as rectangles along branches,
-ordered by position; deletions are black.
-Characteristic mutations
-associated with the emergence of Alpha
-are highlighted with a pink outline.
-}
-\label{fig:subgraph_alpha}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.85\linewidth]{subgraph-Delta_recolored.pdf}
-\caption{
-Subgraph illustrating the origin of B.1.617.2 (Delta).
-Symbols are described in \ref{fig:subgraph_alpha}, except that 
-we highlight the characteristic mutations of Kappa and Delta:
-those with a blue outline are listed
-in the Pango designation issue \cite{PangoIssueKappa}
-or the Kappa constellation \cite{ScorpioKappa},
-but absent in the Delta constellation;
-those with a purple outline are listed
-in the Delta constellation \cite{ScorpioDelta}
-or are Delta-specific deletions reported in Stern et al. \cite{Stern2021},
-but absent in the Kappa constellation; and
-those with a red outline are listed
-in both the Kappa and Delta constellations.
-We also outline
-in orange the mutations associated with the lineage containing clades A to D and
-in green those associated with lineage containing clade E.
-Sites associated with multiple mutations in the subgraph
-have their mutations assigned a unique fill color
-(orange, blue, green, red, etc);
-reversions are further highlighted with a black outline.
-Sc2ts seed samples (see STAR methods) for
-B.1.617, B.1.617.1, and B.1.617.2 are plotted as red squares.
-}
-\label{fig:subgraph_delta}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.85\linewidth]{subgraph-Omicron_recolored.pdf}
-\caption{
-Subgraph illustrating the saltational origin of
-the major Omicron lineages BA.1 and BA.2.
-Symbols are described in Figure~\ref{fig:subgraph_delta},
-except that BA.1 samples are filled in light blue
-(also used to outline BA.1 characteristic mutations),
-BA.2 samples in light yellow
-(also used to outline BA.2 characteristic mutations), and
-B.1.1.529 mutations
-which are neither BA.1 or BA.2 characteristic mutations are
-outlined in green.
-These colors correspond to those used
-in the BA.1/BA.2 Pango designation issue \cite{RambautOmicron2021}.
-}
-\label{fig:subgraph_omicron}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{mutational_spectra}
-\caption{
-All-site mutational spectra of major VOCs calculated from the sc2ts ARG (A),
-the mutation count data from Bloom et al. (2023)\cite{Bloom2023} (B), and
-the Viridian UShER phylogeny from Hunt et al. (2024)\cite{Hunt2024}.
-}
-\label{fig:mutational_spectra}
-\end{figure}
-
-\begin{figure} 
-\centering
 \includegraphics[width=0.5\textwidth]{LS_model_schematic}

### 2025-11-05 latex/main.tex
More
Updates from Overleaf

@@
 \section*{DISCUSSION}
-There are two primary contributions of this work. The first is an
+There are two major contributions of this work. The first is an
 Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
@@
 
-The second primary contribution of this work is the method used to 
+The second major contribution is the method used to 
 infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
@@
 Muller2022bayesian}.
-Similarly, detecting and fully incorporating recombination 
-in real-time at pandemic scale extends the 
-demonstrated utility of a comprehensive
-genealogy~\cite{Hinrichs2024,McBroome2024}, clarifying the 
-role of recombination and enabling timely analysis.
-Given the novelty of sc2ts and the inferred ARG, we have focused 
-significant attention here on methodological issues. 
+By capturing recombination in real time at a global scale, our approach extends the established utility of comprehensive genealogies~\cite{Hinrichs2024,McBroome2024}, offers clear insight into recombination dynamics, and supports timely investigation of .
+The novelty of sc2ts and the inferred ARG motivates our  focus on methodological details. 
 Through these detailed analyses we validate the ARG itself,
@@
 
-The secondary contributions of this work are the initial 
+Study of the sc2ts ARG provides insights in the  contributions of this work are provided by our 
 insights into the evolution of SARS-CoV-2 obtained by studying
@@
 
+\subsection{Analysis of the Jackson et al. recombinants}
+\label{sec:jackson_recombs}
+
+Jackson et al.\cite{Jackson2021} conducted a detailed analysis of
+early recombinants involving B.1.1.7 (Alpha),
+including the first designated Pango X lineage, XA.
+The authors first performed a targeted search for candidate recombinants by
+scanning sampled sequences for sequence motifs associated with Alpha,
+followed by an analysis to reduce the list of plausible recombinants
+displaying signs of onward transmission.
+Here we compared the sc2ts detection results of these recombinants
+with the detection results reported by Jackson et al.
+
+Jackson et al. described
+four groups of recombinants (named A to D) and four singleton recombinants.
+Of the 16 samples analyzed by Jackson et al., 12 were included in the ARG;
+the remaining 4 were excluded
+because the sequences did not pass the QC filtering criteria or
+because of a high HMM cost.
+Groups A-D are summarized in Table~\ref{tab:jackson} and
+show excellent concordance
+in terms of the parental Pango lineages and breakpoint intervals.
+All four groups are directly associated with
+strongly supported recombination events in the ARG
+in terms of the number of averted mutations (A: 16, B: 6, C: 30, D: 18).
+In the ARG, the singleton CAMC-CBA018 has
+the parents B.1.177+B.1.1.7 and
+the breakpoint interval 16,177--20,133, and
+is strongly supported by 14 averted mutations.
+In the Jackson et al. results, this singleton is
+almost identical with the same parent lineages and
+a breakpoint interval of 11,395–-21,993.
+In the ARG, the singleton MILK-103C712 has
+the parents B.1.177.16+B.1.1.7 and
+a breakpoint interval of 24915--27972,
+again strongly supported by 11 averted mutations.
+In the Jackson et al. results,
+this lineage has the same parents,
+with the breakpoint interval 26,800--27,974.
+In the ARG, the singleton QEUH-1067DEF has
+the parents B.1.1.7+B.1.177.9 and
+a breakpoint interval of 7,729--10,870,
+supported by 11 averted mutations.
+In the Jackson et al.\ results,
+this lineage has the same parents,
+with the breakpoint interval 6,953--10,872.
+
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_jackson\_recombinants.ipynb}
+{analysis\_jackson\_recombinants.ipynb}
+for details.
+
+\begin{table}
+\centering
+\caption{
+Comparison of recombination breakpoint
+intervals and parent lineages for Groups A-D
+reported by Jackson et al.\ with the corresponding
+recombination events in the sc2ts ARG.
+The second column gives the number of sequences in the group.
+See the text for details of the groups and the sequences included.
+limited to the samples considered by Jackson et al.
+The breakpoint coordinates in Table 1 of Jackson et al.\ have been altered as follows: 
+we subtract one to the left coordinates and add one to the right coordinates 
+to correspond to the \texttt{tskit} definition of inheritance on either side of a breakpoint,
+and add one to the right coordinate to make the intervals right-exclusive.
+}
+\label{tab:jackson}
+\begin{tabular}{lllr@{--}lr@{+}l}
+\toprule
+Group & Sequences & Method & \multicolumn{2}{c}{Breakpoint interval} & \multicolumn{2}{r}{Parent lineages} \\
+\midrule
+A (XA)       & 2   & Jackson        &  21,254&21,767 & B.1.177&B.1.1.7 \\
+             &     &\texttt{sc2ts} &  20,411&21,765 & B.1.177.18&B.1.1.7 \\
+\cmidrule{3-7}
+B            & 2   & Jackson        &  6,527&6,956 & B.1.36.28&B.1.1.7  \\
+             &     &\texttt{sc2ts} &   6,529&6,954 & B.1.36.28&B.1.1.7  \\
+\cmidrule{3-7}
+C            & 3   &Jackson         &  24,913&28,653 &  B.1.1.7&B.1.221 \\
+             &     &\texttt{sc2ts} &  25,997&27,972 &  B.1.1.7&B.1.221.1 \\
+\cmidrule{3-7}
+D            & 3   & Jackson        &  21,574&23,065 &  B.1.36.17&B.1.1.7 \\
+             &     &\texttt{sc2ts} &  22,445&23,063 &  B.1.36.39&B.1.1.7 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\subsection{Additional analyses of the Pango X lineages}
+\label{sec:supplement_indepth_pango_x_lineages}
+In this section we provide some additional analysis on Pango X lineages
+in the ARG. These analyses are intended to be read in conjunction with
+the subgraph illustrations in Document S2.
+
+\newcommand{\scghdiscussion}[1]{%
+See \href{https://github.com/jeromekelleher/sc2ts-paper/discussions/#1}%
+{github:sc2ts-paper/discussions/#1} for further details.}
+\newcommand{\scghdiscussiontwo}[2]{%
+See \href{https://github.com/jeromekelleher/sc2ts-paper/discussions/#1}%
+{github:sc2ts-paper/discussions/#1} for further details on #2.}
+\subsubsection{XA}
+% mneumonic: deep dive XA
+\label{sec:dd_xa}
+XA traces to a well supported recombination node, with no reversions.
+Note that the ``causal'' sample ERR5308556
+(which triggers the initial recombination event)
+is identical to the recombination node, and
+therefore lacks C8090T,
+which is one of the XA consensus mutations
+(identified as those shared by over 90\% of XA samples).
+As mentioned elsewhere,
+the sc2ts detection results in terms of
+the Pango lineages of the parents and the breakpoint intervals
+agree with the results reported by Jackson et al. (Table \ref{tab:jackson}).
+Two of the Jackson et al. XA samples
+are present in the ARG (ERR5308556 and ERR5414941);
+these samples are specifically marked in Figure~\ref{fig:subgraphs},
+which is a summary of this subgraph.
+
+\subsubsection{XB}
+\label{sec:dd_xb}
+
+The displayed subgraph shows a small selection of the 192 XB samples.
+XB was originally defined as a recombinant between
+the closely related lineages B.1.634 (left parent) and B.1.631 (right parent)
+from North and Central America \cite{Gutierrez2022}.
+However, these parent lineages were not added to the ARG due to high HMM costs.
+Together with other lineages such as B.1.627
+(sister to B.1.631 \cite{Gutierrez2022}),
+they are likely to attach somewhere along the long branch
+in the subgraph visualization (with $> 30$ mutations).
+If samples from these parental lineages were available for attachment,
+then XB would likely be detected as a recombinant by sc2ts.
+% which also contains the key ``R2'' deletion
+% associated with XB / B.1.634 \cite{Gutierrez2022}.
+
+% We hypothesize that there are a set of samples
+% from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
+% which, should they be added to the ARG, would break up the long branch,
+% with XB being a recombination between two of them.
+% In the absence of these, sc2ts cannot detect XB as a recombinant.
+% Also, note that the XB consensus mutations all occur on or above the long branch.
+% We suspect that the child node of the long branch
+% represents the (potentially recombinant) origin of XB, and
+% the B.1 samples at the bottom left of the subgraph
+% may be misclassified by pangolin.
+
+% Further work is required to determine
+% why these additional samples were not gradually added
+% to the ARG by the daily sample matching process or
+% retrospective sample grouping.
+
+\scghdiscussion{780}
+
+\subsubsection{XC}
+\label{sec:dd_xc}
+The five XC samples descend from a well supported recombination node.
+The two parents that are both identical to actual samples
+(ERR4908740 on the left and DRR321274 on the right).
+There is one reversion (G21987A) on the left side of the breakpoint,
+but the site 21987 was masked during ARG inference
+because it was identified to be experiencing
+frequent mutation in a preliminary ARG.
+
+\subsubsection{XE/XH}
+\label{sec:dd_xexh}
+
+The XE samples descend from a well supported recombination node,
+with one mutation and one reversion (T22792C).
+This reversion was mapped onto the ARG
+as a part of postprocessing,
+and could be placed more parsimoniously by a single C22792T
+that groups together XH
+and the lineage leading to the BA.2 recombination nodes
+displayed on the lower left.
+Nevertheless, the exact arrangement of the mutations at position 22792
+should not affect the proposed breakpoint,
+which is very distant, at position 11283.
+This breakpoint coincides with a repeated homopolymeric sequence
+involving 5-base deletions,
+so that alignment errors could have a small effect
+on the breakpoint position.
+Note that the two lower recombination nodes are both well supported,
+the youngest being postulated as the origin of XAF (see below).
+% we leave investigation of this suggested series of
+% sequential recombination events to future work.
+
+\scghdiscussion{633}
+
+\subsubsection{XF}
+\label{sec:dd_xf}
+XF is a well-supported recombinant between Delta and Omicron BA.1,
+with the breakpoint on the far left of the genome.
+All XF samples are the unique descendants of
+a well supported recombination node,
+with both parent nodes (ERR6193241: AY.4 and ERR8042040: BA.1) and
+the single child node (ERR8076129) being samples.
+
+\subsubsection{XG}
+\label{sec:dd_xg}
+The three XG samples descend from a reconstructed internal node,
+which in turn descends from a well supported recombination node.
+The left parent is a sample (SRR20179863), and
+the right parent is a reconstructed internal node
+assigned to Omicron BA.2,
+which is supported by a sample (ERR9124067) that is identical to it.
+
+\subsubsection{XJ}
+\label{sec:dd_xj}
+All the XJ samples descend from a well supported recombination node.
+There are 14 samples (including e.g. ERR8683190)
+which sc2ts identified as descendants of the same recombination node
+but are labeled BA.2 by pangolin (displayed on the left), and
+which do not possess the XJ consensus mutations C22792T and C27945T.
+These BA.2 samples could be worthy of further investigation.
+
+\subsubsection{XL}
+\label{sec:dd_xl}
+All the XL samples descend from a well supported recombination node,
+with three child samples which are identical to it.
+The left parent is a sample (ERR7989424), and
+the right parent is a reconstructed internal node
+assigned to BA.2.5 with a large number of descendant samples.
+
+\subsubsection{XM/XAL}
+\label{sec:dd_xm}
+
+XM is a recombinant of BA.1 (left parent) and BA.2 (right parent).
+Sc2ts identifies multiple recombinant origins of XM.
+One of these origins leads to a dominant clade
+(node 1003220 at position 21,595),
+which also contains the only three XAL samples.
+
+Independent evidence for multiple XM origins
+comes from consideration of a 9-base deletion
+in the Spike gene at position 21633,
+which is characteristic of BA.2 but not BA.1
+\url{https://github.com/cov-lineages/pango-designation/issues/361}.
+Most of the XM-labeled samples in ARG have this deletion,
+but three (SRR19024311, ERR9447529, and SRR18775874) do not,
+suggesting that XM samples trace to at least two recombination events,
+one to the right of the deletion
+(which leads to the majority of the XM samples), and one to the right.
+In the subgraph, it can be seen that one of the samples, SRR18775874,
+has been (presumably wrongly) placed
+within the main XM deletion-containing clade.
+As sc2ts does not use deletions for inference,
+deletion remapping (section XXXX) has reverted the deletion in this case,
+creating many insertions on the branch above the sample (magenta).
+A better use of deletions in sc2ts would likely result
+in placing this sample within one of the two XM clades
+which do not have the deletion.
+
+\scghdiscussion{656}
+
+\subsubsection{XN/XAU}
+\label{sec:dd_xn}
+
+In the sc2ts ARG, the initial XN sample (ERR8986821)
+is well explained as a non-recombinant,
+copying from an inferred internal node
+assigned to BA.2 with only one mutation.
+This internal node is directly supported by the sample ERR9502469.
+All the other XN samples descend from ERR8986821.
+A small clade containing XAU is the sister to XN, and
+there are a series of samples (ERR9453882, ERR9502469, and ERR9823912)
+which support the gradual emergence of both groups
+from a BA.2 ancestral lineage via accumulation of single mutations.
+
+The evidence supporting the original designation of
+XN and XAU as recombinants may come
+from inaccurate or problematic data sources.
+In the case of XN, support for a recombinant BA.1+BA.2 origin
+is from three mutations that are claimed to be BA-1-specific:
+C241T, A2832G, and C3037T
+(\url{https://github.com/cov-lineages/pango-designation/issues/480}).
+However, the BA.1/BA.2 split
+(\url{https://github.com/cov-lineages/pango-designation/issues/361})
+does not list C241T and C3037T as BA.1-specific,
+meaning that these two mutations are not supportive of recombination.
+Moreover, although the remaining mutation
+claimed to support a recombinant origin, A2832G,
+is typical of BA.1 and not BA.2,
+it is seen in several BA.2 samples (e.g. ERR9453882),
+supporting the idea that it has independently arisen within BA.2.
+Hence, none of the sites used to designate XN as a recombinant reliably do so.
+
+In the case of XAU, a recombinant origin was proposed
+on the basis of the same three sites, plus C2470T
+(\url{https://github.com/cov-lineages/pango-designation/issues/894}).
+However, C2470T is not BA.1-specific
+(\url{https://github.com/cov-lineages/pango-designation/issues/361}).
+Moreover, some BA.2-designated samples such as ERR9823912 also have C2470T,
+so we do not believe this to be a recombinant-supporting mutation either.
+
+\scghdiscussiontwo{781}{XN}
+
+\scghdiscussiontwo{928}{XAU}
+
+\subsubsection{XP}
+\label{sec:dd_xp}
+
+This is a case where the Pango X lineage is designated
+on the basis of two loci on the far right of the genome:
+a single nucleotide mutation (A29510C) and
+a rare, large deletion at position 29734 in the S2M step loop
+(\url{https://github.com/cov-lineages/pango-designation/issues/481}),
+which is not common enough to map in sc2ts during postprocessing
+and hence not shown here.
+Therefore, unless this deletion were given more weight than a single locus,
+sc2ts would not identify XP as a recombinant,
+even if deletions were used during ARG inference.
+If this deletion were recurrent, the evidence from sc2ts suggests
+that XP might not be a recombinant,
+because it is only one mutation different from the sample ERR8628084.
+
+\scghdiscussion{775}
+
+\subsubsection{XQ/XR/XU/XAA/XAG/XAM}
+\label{sec:dd_xq}
+
+In the ARG, XR, XU, XAA, XAG, and XAM
+are nested under the XQ recombinant lineage.
+These lineages were designated around the same time,
+approximately within a five-month period (Mar. to Jul., 2022).
+The breakpoint intervals of these lineages,
+as suggested in the Pango designation issues,
+all occur at the 5' end of the genome,
+within a $\sim$5 kb segment ($\sim$4,300 to $\sim$9,400):
+% TODO: Add URLs to Pango issues?
+XQ (4321-5387), % https://github.com/cov-lineages/pango-designation/issues/468
+XAG (6515-8394), % https://github.com/cov-lineages/pango-designation/issues/709
+XU (6517-9345), % https://github.com/cov-lineages/pango-designation/issues/522
+XAM (8087-9193), % https://github.com/cov-lineages/pango-designation/issues/759
+and XAA (8392-9345). % https://github.com/cov-lineages/pango-designation/issues/664
+
+Upon closer inspection,
+the right-hand branch above the recombination node appears problematic,
+as it displays three reversion mutations,
+which revert mutations at positions 22792, 24023, and 25624
+in the ancestral lineage just above.
+These sites were excluded during ARG inference, and
+remapped onto the ARG by parsimony during ARG postprocessing.
+If these three sites were used during primary ARG inference,
+the right parent of the recombination node
+may have been chosen to be the ancestral BA.2 sample
+on the right-hand path with 18 child nodes.
+A more parsimonious reconstruction
+would then have the XQ clade as a sibling to
+a group of (monophyletic) XAA, XAM, XAG, and XR lineages.
+This would remove between 4 and 5 reversion mutations.
+
+\scghdiscussion{655}
+
+\subsubsection{XS}
+\label{sec:dd_xs}
+
+In the ARG, the XS samples appear to have originated from
+two sequential recombination events (separated by a single branch).
+However, the later event is likely an artifact,
+because the XS sample (SRR18343220) associated with
+this event failed detection-level QC
+(i.e., the minimum net number of supporting loci;
+see the lower copying pattern).
+In addition, this sample was sequenced using the Ion AmpliSeq V1 kit,
+which was also used to sequence eight other samples
+descending from the same event
+(clustered due to a shared sequencing error due to AmpliSeq).
+These eight samples would likely attach to the nodes
+descending from the first recombination event.
+We excluded this likely artifactual event from further analysis and
+consider XS to have emerged from a single recombinant origin.
+
+\scghdiscussion{782}
+
+\subsubsection{XW}
+\label{sec:dd_xw}
+
+All the XW samples descend from a well supported recombination node.
+Its left parent is a BA.1.1.15-labeled sample (ERR8069658), and
+its right parent is a BA.2-labeled sample (ERR9086669).
+
+\subsubsection{XY}
+\label{sec:dd_xy}
+
+All the XY samples descend from a well supported recombination node,
+through a shared reconstructed internal node.
+Its left parent is a reconstructed internal node assigned to BA.1.1,
+which is supported by a sample (ERR8627048) identical to it.
+Its right parent is a BA.2-labeled sample (ERR9399983).
+There are a pair of (orange) reversion mutations
+within the XY clade at position 23948:
+these were not used for inference
+but remapped using parsimony during postprocessing, and
+indicate that the topology within the XY clade
+could potentially be improved
+by alternative phylogenetic reconstruction methods.
+
+\subsubsection{Xx: XZ/XAC/XAD/XAE/XAP}
+\label{sec:dd_xx}
+
+This nested set of Pango X lineages descends
+from a single recombination node in the ARG,
+which we tentatively refer to as ``Xx''.
+Note that all were designated around the same time,
+with their samples collected
+roughly within a three-month period (Mar., 2022 to Jul., 2022).
+Several samples labeled as BA.2 by Pangolin also lie
+within this recombinant clade,
+including two that are identical
+to the proposed originating recombinant (ERR8146303 and ERR8163061).
+
+Exact relationships within the Xx clade are somewhat tentative.
+For example, there are three highly mutable sites (22792, 28877, and 28878)
+which were not used during inference,
+but were remapped onto the ARG during postprocessing.
+The inferred topology places these sites
+as recurrent (colored orange, grey, and faded yellow),
+but a more parsimonious arrangement would,
+for example, group the two XAD samples as siblings
+(removing the need to postulate mutations
+A1250G, A28877T, and G28878C to be recurrent).
+It is possible that a more parsimonious arrangement would also
+reduce the recurrent mutations and reversions seen at
+sites 22,792 (four orange mutations) and
+26,060 (two parallel brown mutations),
+but we do not show such a reconstruction,
+which is complicated by the presence of
+additional (non-displayed) samples descending from
+internal nodes in the plotted subgraph.
+
+For XAC, in particular, the Pango designation issue
+(\url{https://github.com/cov-lineages/pango-designation/issues/590}) suggests
+it to be a recombinant with two breakpoints:
+BA.2 on the left, BA.1 in the middle, and
+a switch back to a very small region of BA.2 on the right.
+This contrasts with the one-breakpoint BA.2+BA.1 origin inferred in the ARG.
+However, only two loci support the right hand BA.2 assignment:
+A29510C and a large deletion
+(deletions are treated as missing data during primary ARG inference).
+Sc2ts does not count this as enough evidence
+to insert an additional breakpoint.
+
+\scghdiscussion{631}
+
+\subsubsection{XAF}
+\label{sec:dd_xaf}
+
+Of the 21 XAF samples in the Viridian v04 dataset,
+only one XAF sample passed our QC filters and was added to the ARG.
+% many have greater than 500 missing Ns in their consensus sequences or
+% require a large number of mutations to incorporate into the ARG
+% (i.e. have a high HMM cost).
+However, we find a small recombinant BA1.1/BA.2 clade of 37 samples,
+associated with the BA.2-labeled sample ERR9755541,
+within which the single XAF sample is nested.
+The inferred parents and breakpoint at 10,447
+aligns with that proposed for XAF
+(\url{https://github.com/cov-lineages/pango-designation/issues/676}).
+Inspection of the omitted XAF samples with high HMM cost shows that
+they would attach as close sisters to the included XAF sample,
+albeit on branches with many new mutations.
+% Hence we believe the current XAF topology in the ARG is reasonable,
+% although somewhat tentative, given the number of included samples.
+
+The mutation-rich branch in the subgraph with
+a 5-base deletion (11283-11287) and insertion (11292-11296)
+may be due to a simple deletion alignment error.
+
+\scghdiscussion{638}
+
+\subsubsection{XAJ}
+\label{sec:dd_xaj}
+
+In the Pango designation issue
+(\url{https://github.com/cov-lineages/pango-designation/issues/826}),
+XAJ is suggested to be a recombinant between
+BA.2.12.1 (left parent) and BA.4 (right parent).
+However, XAJ is not inferred to be a recombinant lineage in the ARG.
+From the subgraph, which shows the 18 XAJ samples
+plus a small number of relevant BA.4 samples,
+it can be seen that 9 sites leading to XAJ indeed recur
+in the lineage to the closest BA.4 sample (SRR18848050),
+which strongly indicates that this is a recombinant
+that is missed by sc2ts.
+The reason that is is not a recombinant in the ARG is
+that 6 sites represent a 6bp deletion and so are not used for inference.
+The remaining three sites (T22917G, T23018G, and G23040A)
+are not enough to exceed our k=4 threshold  for recombinant designation.
+Including deletions in inference would certainly
+lead to XAJ being detected as a recombinant,
+and with a lower threshold of k=3
+the distribution of mutations on the ARG topology suggests
+that it is is even possible that XAJ is a 2-breakpoint recombinant
+with the region from 21765-23535
+coming from the BA.4 labeled sample SRR18848050.
+
+\scghdiscussion{634}
+
+\subsubsection{XAN/XAV}
+\label{sec:dd_xan}
+
+XAN and XAV are shown in this subgraph, as they share some mutations.
+In the Pango designation issues
+(\url{https://github.com/cov-lineages/pango-designation/issues/771};
+\url{https://github.com/cov-lineages/pango-designation/issues/911}),
+XAN and XAV are suggested to be a recombinant between
+BA.2 (left parent) and BA.5 (right parent).
+Both these Pango lineages were detected as non-recombinants by sc2ts,
+because the number of supporting loci fell below our threshold of detection.
+In the ARG, we found only one mutation (C9866T)
+supporting the suggested BA.2 parent for both XAN and XAV.
+
+\scghdiscussiontwo{783}{XAN}
+
+\scghdiscussiontwo{785}{XAV}
+
+\subsubsection{XAS}
+\label{sec:dd_xas}
+
+XAS was proposed as a recombination between
+BA.5 (left parent) and BA.2 (right parent).
+However, we find that this is only two mutations
+different from e.g. ERR9929846, which is a BA.4 sample.
+In fact, XAS only differs from the BA.4 root by a single mutation,
+C27945T, which was identified in the original Pango designation issue
+(\url{https://github.com/cov-lineages/pango-designation/issues/882})
+as an important ORF truncation mutation
+which is also present in a few BA.2.65 samples.
+It is possible that C27945T was acquired by recombination,
+but this is clearly below our threshold of detection.
+
+\scghdiscussion{784}
+
+\subsubsection{XAZ}
+\label{sec:dd_xaz}
+
+In the Pango designation issue
+(\url{https://github.com/cov-lineages/pango-designation/issues/797}),
+XAZ is suggested as a two-breakpoint recombinant between
+BA.2.5 and BA.5 (the middle parent),
+with three sites supporting the BA.2.5 parent
+on the left end (C2232T, C3317T, and T3358C).
+In the ARG, this is very close to being a recombinant of
+BA.2.5 (ERR9615610) on the left side of the breakpoint,
+as not only are the three mutations present in this sample,
+but there is another site which supports recombination 
+C1912T, which is not reported in the Pango designation issue).
+XAZ has C44T which is lacking in all BA.2.5 samples,
+leaving only three net supporting loci for a putative recombinant.
+However, position 44 is often clipped,
+so this is not strong evidence against a recombinant, and
+therefore XAZ is a marginal case of recombination.
+We found no evidence for the third segment
+inherited from a BA2.5 parent on the right end.
+In the ARG, all the XAZ samples form a clade.
+Here we display only the first 20 XAZ samples.
+
+\scghdiscussion{786}
+
+\subsubsection{XBB}
+\label{sec:dd_xbb}
+
+Most of the XBB samples descend from a well supported recombination node.
+In the Pango designation issue
+(\url{https://github.com/cov-lineages/pango-designation/issues/1058}),
+the Pango lineages of the suggested left and right parents are
+BJ.1 and BM.1.1.1, respectively.
+In the ARG, the Pango lineage of the right parent is identical,
+but that of the left parent is a related ancestral Pango lineage
+(BA.2.10 instead of BA.2.10.1.1 = BJ.1).
+This is likely due to the absence of BJ.1 in the ARG:
+only four BJ.1 samples exist in the Viridian v04 dataset,
+a single one passed sequence-level QC, and
+that was then rejected due to high HMM cost.
+Note that BJ.1 differs from BA.2.10 by over 20 mutations
+(\url{https://github.com/cov-lineages/pango-designation/issues/915}),
+explaining the large number of (gold) de novo mutations
+in the XBB copying pattern.
+Had BJ.1 samples been present in the ARG,
+the left branch of
+the XBB-associated recombination node
+may have attached to a BJ.1-labeled node, and
+the origin of XBB would required fewer mutations.
+
+\scghdiscussion{929}
+
+%% JK: commented this out as it's confusing the relationship between XBB.1 and
+%% XBB.x. Note that most methods pick out XBB.x as recombinant, so it's not that 
+%% remarkable.
+
+% \subsubsection{XBB.1}
+% \label{sec:dd_xbb.1}
+
+% XBB.1 is characterized by an additional mutation in Spike, G22317T (G252V) and
+% is not designated as a recombinant of XBB in the Pango designation issue
+% (https://github.com/cov-lineages/pango-designation/issues/1088).
+% The ARG agrees with this and places the characteristic mutation
+% above the root of a single non-recombinant XBB.1 clade.
+% We include this non-recombinant lineage
+% in order to highlight some well supported subsequent recombination nodes
+% within XBB.1, such as the (pink) one leading to two XBB.1-labeled samples
+% at the bottom left of the subgraph which we refer to as XBB.x.
+% XBB.x is well supported: its left parent is a reconstructed internal node,
+% which is directly supported by a BQ.1-labeled sample, and
+% its right parent is a BB.1.5.18-labeled sample (ERR10792363).
+% This highlights the automatic ability of sc2ts
+% to infer sequential recombination events
+% (i.e. recombinants-of-recombinants),
+% even those involving few samples.
+% A few other recombination events occur within XBB.1,
+% but they either fail QC checks, or
+% (as the other one shown here, to the bottom right)
+% lead to a single sample so may be
+% e.g. caused by accidental mixing of samples during sequencing.
+
+\subsubsection{XBD}
+\label{sec:dd_xbd}
+
+All the XBD samples descend from a well supported recombination node.
+Its left and right parents are reconstructed internal nodes,
+both of which are directly supported by samples
+(SRR21382561 and ERR9825609, respectively)
+that are identical to them.
+
+\subsubsection{XBE}
+\label{sec:dd_xbe}
+
+In the Pango designation issue
+(\url{https://github.com/cov-lineages/pango-designation/issues/1246}),
+XBE was suggested as a recombinant between
+BA.5.2 (left parent; most of the genome) and
+BE.4 (right parent; somewhere from 23,609 onwards). 
+However, sc2ts treats the 65 XBE samples
+as non-recombinant descendants of BA.5.2
+because the number of supporting sites
+fell below our threshold of detection.
+We have included the BE.4 lineage on the subgraph,
+revealing two mutations (C27800T, and G28681T)
+which occur in parallel above the root of the XBE clade
+and along the path ascending from the displayed BE.4 sample,
+suggesting that if we used k=2
+as the recombinant threshold,
+sc2ts would indeed detect XBE as a recombinant.
+
+\scghdiscussion{930}
+
+\subsubsection{XBF}
+\label{sec:dd_xbf}
+
+All the XBF samples descend from a well supported recombination node,
+with three mutations on the left of the breakpoint.
+Its left parent is a sample assigned to BA.5.2.1 (SRR20305553).
+Its right parent is a reconstructed internal node
+that is supported by a CJ.1-labeled sample (SRR21948068),
+which differs by three mutations.
+
+\subsubsection{XBG}
+\label{sec:dd_xbg}
+
+All the XBG samples descend from a well supported recombination node,
+with two mutations, one on each side of the breakpoint.
+Both its left and right parents are samples
+(SRR19920084 and ERR9852351, respectively).
+
+\subsubsection{XBH}
+\label{sec:dd_xbh}
+All the XBH samples descend from a well supported recombination node,
+despite two reversions (T222C and T12789C, respectively),
+which occur at excluded sites and
+were remapped by parsimony during postprocessing.
+If these two sites had been used during ARG inference,
+the branch attached to BA.2.1 would have been attached to
+the BA.2.5 shown on the top left instead,
+leading to a more parsimonious local topology
+with two mutations on the left branch. This would also remove the two
+blue lines on the left and side of the copying pattern.
+
+\subsubsection{XBQ/XBK/XBK.1}
+\label{sec:dd_xbq}
+
+XBK and XBQ are siblings, and neither appear to be recombinants in the ARG,
+as they are closely related to the non-recombinant node that originates CJ.1.3. 
+At the time of designating XBK
+a hypothesis was made that a missing common ancestor could exist 
+(\url{https://github.com/cov-lineages/pango-designation/issues/1440\#issuecomment-1355264973}),
+making XBK non-recombinant.
+Exactly such a missing ancestor
+is found as the upper internal node labeled CJ.1 in the ARG,
+constructed by sc2ts as part of its tree building steps.
+
+\scghdiscussiontwo{931}{XBQ}
+
+\scghdiscussiontwo{932}{XBK}
+
+\subsubsection{XBM}
+\label{sec:dd_xbm}
+
+All the XBM samples descend from a well supported recombination node.
+Its left parent is a reconstructed internal node
+that is directly supported by a BA.2.76-labeled sample identical to it.
+Its right parent is a BF.3-labeled sample.
+
+\subsubsection{XBR}
+\label{sec:dd_xbr}
+This one XBR sample descends from a well supported recombination node.
+Its left parent is a BN.3.1-labeled sample, and
+its right parent is a reconstructed internal node
+that is directly supported by a BQ.1.25-labeled sample identical to it.
+
+\subsection{Pango X lineages not added to the ARG}
+\label{sec:supplement_rejected_pango_x_lineages}
+
+Ten Pango X lineages which are present in the Viridian v04 dataset
+were not integrated into the ARG
+due to high HMM costs and/or insufficient sample support,
+but their samples were matched against the ARG during inference.
+% https://github.com/jeromekelleher/sc2ts-paper/issues/366
+Table~\ref{tab:extra_hmm_results} summarizes
+the HMM matching results of these Pango X lineages.
+All were identified as recombinants by sc2ts, except XBL.
+(For XBL, the evidence for recombination
+proposed in the Pango designation issue \cite{PangoIssueXBL}
+involves only three informative sites
+supporting the suggested rightmost parent, and
+therefore this likely fell below our threshold of detection.)
+% https://github.com/jeromekelleher/sc2ts-paper/issues/379
+Importantly, these results demonstrate that sc2ts can detect
+recombinants with multiple breakpoints (XAY, XBC, and XBT).
+% https://github.com/jeromekelleher/sc2ts-paper/discussions/776
+% https://github.com/jeromekelleher/sc2ts-paper/discussions/788
+XBC and XBT are relatively clean examples of multiple-breakpoint recombinants,
+with the Pango lineages of the parents and the breakpoint intervals
+concordant with those suggested in the Pango designation issues,
+but XAY is more complex, possibly involving up to six breakpoints as per sc2ts.
+
+\begin{table}
+\centering
+\caption{
+Sc2ts detection results for Pango X lineages not added to the ARG.
+The number of samples in Viridian v04 dataset
+assigned to these Pango X lineages,
+the number of samples which passed sequence-level quality control, and
+the number of recombination breakpoints
+in the Viterbi solutions of the samples
+are shown.
+}
+\label{tab:extra_hmm_results}
+\begin{tabular}{crrr}
+\toprule
+Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
+\midrule
+XBJ & 2 & 1 & 1\\
+XBP & 8 & 2 & 1\\
+XBS & 19 & 15 & 1\\
+XBW & 1 & 1 & 1\\
+XCA & 11 & 4 & 1\\
+XAK & 2 & 2 & 1\\   % Multiple breakpoint Pango X
+XAY & 15 & 11 & 6\\ % Multiple breakpoint Pango X
+XBC & 36 & 21 & 3\\ % Multiple breakpoint Pango X
+XBL & 2 & 2 & 0\\   % Multiple breakpoint Pango X
+XBT & 1 & 1 & 2\\   % Multiple breakpoint Pango X
+\bottomrule
+\end{tabular}
+\end{table}
+
+% 2 + 8 + 19+ 1 + 11+ 2 + 15+ 36+ 2 + 1
+% 97
+
 \subsection{Detailed RIPPLES analysis}
@@
 
-\subsection{Additional analyses of the Pango X lineages}
-\label{sec:supplement_indepth_pango_x_lineages}
-In this section we provide some additional analysis on Pango X lineages
-in the ARG. These analyses are intended to be read in conjunction with
-the subgraph illustrations in Document S2.
-
-\newcommand{\scghdiscussion}[1]{%
-See \href{https://github.com/jeromekelleher/sc2ts-paper/discussions/#1}%
-{github:sc2ts-paper/discussions/#1} for further details.}
-\newcommand{\scghdiscussiontwo}[2]{%
-See \href{https://github.com/jeromekelleher/sc2ts-paper/discussions/#1}%
-{github:sc2ts-paper/discussions/#1} for further details on #2.}
-\subsubsection{XA}
-% mneumonic: deep dive XA
-\label{sec:dd_xa}
-XA traces to a well supported recombination node, with no reversions.
-Note that the ``causal'' sample ERR5308556
-(which triggers the initial recombination event)
-is identical to the recombination node, and
-therefore lacks C8090T,
-which is one of the XA consensus mutations
-(identified as those shared by over 90\% of XA samples).
-As mentioned elsewhere,
-the sc2ts detection results in terms of
-the Pango lineages of the parents and the breakpoint intervals
-agree with the results reported by Jackson et al. (Table \ref{tab:jackson}).
-Two of the Jackson et al. XA samples
-are present in the ARG (ERR5308556 and ERR5414941);
-these samples are specifically marked in Figure~\ref{fig:subgraphs},
-which is a summary of this subgraph.
-
-\subsubsection{XB}
-\label{sec:dd_xb}
-
-The displayed subgraph shows a small selection of the 192 XB samples.
-XB was originally defined as a recombinant between
-the closely related lineages B.1.634 (left parent) and B.1.631 (right parent)
-from North and Central America \cite{Gutierrez2022}.
-However, these parent lineages were not added to the ARG due to high HMM costs.
-Together with other lineages such as B.1.627
-(sister to B.1.631 \cite{Gutierrez2022}),
-they are likely to attach somewhere along the long branch
-in the subgraph visualization (with $> 30$ mutations).
-If samples from these parental lineages were available for attachment,
-then XB would likely be detected as a recombinant by sc2ts.
-% which also contains the key ``R2'' deletion
-% associated with XB / B.1.634 \cite{Gutierrez2022}.
-
-% We hypothesize that there are a set of samples
-% from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
-% which, should they be added to the ARG, would break up the long branch,
-% with XB being a recombination between two of them.
-% In the absence of these, sc2ts cannot detect XB as a recombinant.
-% Also, note that the XB consensus mutations all occur on or above the long branch.
-% We suspect that the child node of the long branch
-% represents the (potentially recombinant) origin of XB, and
-% the B.1 samples at the bottom left of the subgraph
-% may be misclassified by pangolin.
-
-% Further work is required to determine
-% why these additional samples were not gradually added
-% to the ARG by the daily sample matching process or
-% retrospective sample grouping.
-
-\scghdiscussion{780}
-
-\subsubsection{XC}
-\label{sec:dd_xc}
-The five XC samples descend from a well supported recombination node.
-The two parents that are both identical to actual samples
-(ERR4908740 on the left and DRR321274 on the right).
-There is one reversion (G21987A) on the left side of the breakpoint,
-but the site 21987 was masked during ARG inference
-because it was identified to be experiencing
-frequent mutation in a preliminary ARG.
-
-\subsubsection{XE/XH}
-\label{sec:dd_xexh}
-
-The XE samples descend from a well supported recombination node,
-with one mutation and one reversion (T22792C).
-This reversion was mapped onto the ARG
-as a part of postprocessing,
-and could be placed more parsimoniously by a single C22792T
-that groups together XH
-and the lineage leading to the BA.2 recombination nodes
-displayed on the lower left.
-Nevertheless, the exact arrangement of the mutations at position 22792
-should not affect the proposed breakpoint,
-which is very distant, at position 11283.
-This breakpoint coincides with a repeated homopolymeric sequence
-involving 5-base deletions,
-so that alignment errors could have a small effect
-on the breakpoint position.
-Note that the two lower recombination nodes are both well supported,
-the youngest being postulated as the origin of XAF (see below).
-% we leave investigation of this suggested series of
-% sequential recombination events to future work.
-
-\scghdiscussion{633}
-
-\subsubsection{XF}
-\label{sec:dd_xf}
-XF is a well-supported recombinant between Delta and Omicron BA.1,
-with the breakpoint on the far left of the genome.
-All XF samples are the unique descendants of
-a well supported recombination node,
-with both parent nodes (ERR6193241: AY.4 and ERR8042040: BA.1) and
-the single child node (ERR8076129) being samples.
-
-\subsubsection{XG}
-\label{sec:dd_xg}
-The three XG samples descend from a reconstructed internal node,
-which in turn descends from a well supported recombination node.
-The left parent is a sample (SRR20179863), and
-the right parent is a reconstructed internal node
-assigned to Omicron BA.2,
-which is supported by a sample (ERR9124067) that is identical to it.
-
-\subsubsection{XJ}
-\label{sec:dd_xj}
-All the XJ samples descend from a well supported recombination node.
-There are 14 samples (including e.g. ERR8683190)
-which sc2ts identified as descendants of the same recombination node
-but are labeled BA.2 by pangolin (displayed on the left), and
-which do not possess the XJ consensus mutations C22792T and C27945T.
-These BA.2 samples could be worthy of further investigation.
-
-\subsubsection{XL}
-\label{sec:dd_xl}
-All the XL samples descend from a well supported recombination node,
-with three child samples which are identical to it.
-The left parent is a sample (ERR7989424), and
-the right parent is a reconstructed internal node
-assigned to BA.2.5 with a large number of descendant samples.
-
-\subsubsection{XM/XAL}
-\label{sec:dd_xm}
-
-XM is a recombinant of BA.1 (left parent) and BA.2 (right parent).
-Sc2ts identifies multiple recombinant origins of XM.
-One of these origins leads to a dominant clade
-(node 1003220 at position 21,595),
-which also contains the only three XAL samples.
-
-Independent evidence for multiple XM origins
-comes from consideration of a 9-base deletion
-in the Spike gene at position 21633,
-which is characteristic of BA.2 but not BA.1
-\url{https://github.com/cov-lineages/pango-designation/issues/361}.
-Most of the XM-labeled samples in ARG have this deletion,
-but three (SRR19024311, ERR9447529, and SRR18775874) do not,
-suggesting that XM samples trace to at least two recombination events,
-one to the right of the deletion
-(which leads to the majority of the XM samples), and one to the right.
-In the subgraph, it can be seen that one of the samples, SRR18775874,
-has been (presumably wrongly) placed
-within the main XM deletion-containing clade.
-As sc2ts does not use deletions for inference,
-deletion remapping (section XXXX) has reverted the deletion in this case,
-creating many insertions on the branch above the sample (magenta).
-A better use of deletions in sc2ts would likely result
-in placing this sample within one of the two XM clades
-which do not have the deletion.
-
-\scghdiscussion{656}
-
-\subsubsection{XN/XAU}
-\label{sec:dd_xn}
-
-In the sc2ts ARG, the initial XN sample (ERR8986821)
-is well explained as a non-recombinant,
-copying from an inferred internal node
-assigned to BA.2 with only one mutation.
-This internal node is directly supported by the sample ERR9502469.
-All the other XN samples descend from ERR8986821.
-A small clade containing XAU is the sister to XN, and
-there are a series of samples (ERR9453882, ERR9502469, and ERR9823912)
-which support the gradual emergence of both groups
-from a BA.2 ancestral lineage via accumulation of single mutations.
-
-The evidence supporting the original designation of
-XN and XAU as recombinants may come
-from inaccurate or problematic data sources.
-In the case of XN, support for a recombinant BA.1+BA.2 origin
-is from three mutations that are claimed to be BA-1-specific:
-C241T, A2832G, and C3037T
-(\url{https://github.com/cov-lineages/pango-designation/issues/480}).
-However, the BA.1/BA.2 split
-(\url{https://github.com/cov-lineages/pango-designation/issues/361})
-does not list C241T and C3037T as BA.1-specific,
-meaning that these two mutations are not supportive of recombination.
-Moreover, although the remaining mutation
-claimed to support a recombinant origin, A2832G,
-is typical of BA.1 and not BA.2,
-it is seen in several BA.2 samples (e.g. ERR9453882),
-supporting the idea that it has independently arisen within BA.2.
-Hence, none of the sites used to designate XN as a recombinant reliably do so.
-
-In the case of XAU, a recombinant origin was proposed
-on the basis of the same three sites, plus C2470T
-(\url{https://github.com/cov-lineages/pango-designation/issues/894}).
-However, C2470T is not BA.1-specific
-(\url{https://github.com/cov-lineages/pango-designation/issues/361}).
-Moreover, some BA.2-designated samples such as ERR9823912 also have C2470T,
-so we do not believe this to be a recombinant-supporting mutation either.
-
-\scghdiscussiontwo{781}{XN}
-
-\scghdiscussiontwo{928}{XAU}
-
-\subsubsection{XP}
-\label{sec:dd_xp}
-
-This is a case where the Pango X lineage is designated
-on the basis of two loci on the far right of the genome:
-a single nucleotide mutation (A29510C) and
-a rare, large deletion at position 29734 in the S2M step loop
-(\url{https://github.com/cov-lineages/pango-designation/issues/481}),
-which is not common enough to map in sc2ts during postprocessing
-and hence not shown here.
-Therefore, unless this deletion were given more weight than a single locus,
-sc2ts would not identify XP as a recombinant,
-even if deletions were used during ARG inference.
-If this deletion were recurrent, the evidence from sc2ts suggests
-that XP might not be a recombinant,
-because it is only one mutation different from the sample ERR8628084.
-
-\scghdiscussion{775}
-
-\subsubsection{XQ/XR/XU/XAA/XAG/XAM}
-\label{sec:dd_xq}
-
-In the ARG, XR, XU, XAA, XAG, and XAM
-are nested under the XQ recombinant lineage.
-These lineages were designated around the same time,
-approximately within a five-month period (Mar. to Jul., 2022).
-The breakpoint intervals of these lineages,
-as suggested in the Pango designation issues,
-all occur at the 5' end of the genome,
-within a $\sim$5 kb segment ($\sim$4,300 to $\sim$9,400):
-% TODO: Add URLs to Pango issues?
-XQ (4321-5387), % https://github.com/cov-lineages/pango-designation/issues/468
-XAG (6515-8394), % https://github.com/cov-lineages/pango-designation/issues/709
-XU (6517-9345), % https://github.com/cov-lineages/pango-designation/issues/522
-XAM (8087-9193), % https://github.com/cov-lineages/pango-designation/issues/759
-and XAA (8392-9345). % https://github.com/cov-lineages/pango-designation/issues/664
-
-Upon closer inspection,
-the right-hand branch above the recombination node appears problematic,
-as it displays three reversion mutations,
-which revert mutations at positions 22792, 24023, and 25624
-in the ancestral lineage just above.
-These sites were excluded during ARG inference, and
-remapped onto the ARG by parsimony during ARG postprocessing.
-If these three sites were used during primary ARG inference,
-the right parent of the recombination node
-may have been chosen to be the ancestral BA.2 sample
-on the right-hand path with 18 child nodes.
-A more parsimonious reconstruction
-would then have the XQ clade as a sibling to
-a group of (monophyletic) XAA, XAM, XAG, and XR lineages.
-This would remove between 4 and 5 reversion mutations.
-
-\scghdiscussion{655}
-
-\subsubsection{XS}
-\label{sec:dd_xs}
-
-In the ARG, the XS samples appear to have originated from
-two sequential recombination events (separated by a single branch).
-However, the later event is likely an artifact,
-because the XS sample (SRR18343220) associated with
-this event failed detection-level QC
-(i.e., the minimum net number of supporting loci;
-see the lower copying pattern).
-In addition, this sample was sequenced using the Ion AmpliSeq V1 kit,
-which was also used to sequence eight other samples
-descending from the same event
-(clustered due to a shared sequencing error due to AmpliSeq).
-These eight samples would likely attach to the nodes
-descending from the first recombination event.
-We excluded this likely artifactual event from further analysis and
-consider XS to have emerged from a single recombinant origin.
-
-\scghdiscussion{782}
-
-\subsubsection{XW}
-\label{sec:dd_xw}
-
-All the XW samples descend from a well supported recombination node.
-Its left parent is a BA.1.1.15-labeled sample (ERR8069658), and
-its right parent is a BA.2-labeled sample (ERR9086669).
-
-\subsubsection{XY}
-\label{sec:dd_xy}
-
-All the XY samples descend from a well supported recombination node,
-through a shared reconstructed internal node.
-Its left parent is a reconstructed internal node assigned to BA.1.1,
-which is supported by a sample (ERR8627048) identical to it.
-Its right parent is a BA.2-labeled sample (ERR9399983).
-There are a pair of (orange) reversion mutations
-within the XY clade at position 23948:
-these were not used for inference
-but remapped using parsimony during postprocessing, and
-indicate that the topology within the XY clade
-could potentially be improved
-by alternative phylogenetic reconstruction methods.
-
-\subsubsection{Xx: XZ/XAC/XAD/XAE/XAP}
-\label{sec:dd_xx}
-
-This nested set of Pango X lineages descends
-from a single recombination node in the ARG,
-which we tentatively refer to as ``Xx''.
-Note that all were designated around the same time,
-with their samples collected
-roughly within a three-month period (Mar., 2022 to Jul., 2022).
-Several samples labeled as BA.2 by Pangolin also lie
-within this recombinant clade,
-including two that are identical
-to the proposed originating recombinant (ERR8146303 and ERR8163061).
-
-Exact relationships within the Xx clade are somewhat tentative.
-For example, there are three highly mutable sites (22792, 28877, and 28878)
-which were not used during inference,
-but were remapped onto the ARG during postprocessing.
-The inferred topology places these sites
-as recurrent (colored orange, grey, and faded yellow),
-but a more parsimonious arrangement would,
-for example, group the two XAD samples as siblings
-(removing the need to postulate mutations
-A1250G, A28877T, and G28878C to be recurrent).
-It is possible that a more parsimonious arrangement would also
-reduce the recurrent mutations and reversions seen at
-sites 22,792 (four orange mutations) and
-26,060 (two parallel brown mutations),
-but we do not show such a reconstruction,
-which is complicated by the presence of
-additional (non-displayed) samples descending from
-internal nodes in the plotted subgraph.
-
-For XAC, in particular, the Pango designation issue
-(\url{https://github.com/cov-lineages/pango-designation/issues/590}) suggests
-it to be a recombinant with two breakpoints:
-BA.2 on the left, BA.1 in the middle, and
-a switch back to a very small region of BA.2 on the right.
-This contrasts with the one-breakpoint BA.2+BA.1 origin inferred in the ARG.
-However, only two loci support the right hand BA.2 assignment:
-A29510C and a large deletion
-(deletions are treated as missing data during primary ARG inference).
-Sc2ts does not count this as enough evidence
-to insert an additional breakpoint.
-
-\scghdiscussion{631}
-
-\subsubsection{XAF}
-\label{sec:dd_xaf}
-
-Of the 21 XAF samples in the Viridian v04 dataset,
-only one XAF sample passed our QC filters and was added to the ARG.
-% many have greater than 500 missing Ns in their consensus sequences or
-% require a large number of mutations to incorporate into the ARG
-% (i.e. have a high HMM cost).
-However, we find a small recombinant BA1.1/BA.2 clade of 37 samples,
-associated with the BA.2-labeled sample ERR9755541,
-within which the single XAF sample is nested.
-The inferred parents and breakpoint at 10,447
-aligns with that proposed for XAF
-(\url{https://github.com/cov-lineages/pango-designation/issues/676}).
-Inspection of the omitted XAF samples with high HMM cost shows that
-they would attach as close sisters to the included XAF sample,
-albeit on branches with many new mutations.
-% Hence we believe the current XAF topology in the ARG is reasonable,
-% although somewhat tentative, given the number of included samples.
-
-The mutation-rich branch in the subgraph with
-a 5-base deletion (11283-11287) and insertion (11292-11296)
-may be due to a simple deletion alignment error.
-
-\scghdiscussion{638}
-
-\subsubsection{XAJ}
-\label{sec:dd_xaj}
-
-In the Pango designation issue
-(\url{https://github.com/cov-lineages/pango-designation/issues/826}),
-XAJ is suggested to be a recombinant between
-BA.2.12.1 (left parent) and BA.4 (right parent).
-However, XAJ is not inferred to be a recombinant lineage in the ARG.
-From the subgraph, which shows the 18 XAJ samples
-plus a small number of relevant BA.4 samples,
-it can be seen that 9 sites leading to XAJ indeed recur
-in the lineage to the closest BA.4 sample (SRR18848050),
-which strongly indicates that this is a recombinant
-that is missed by sc2ts.
-The reason that is is not a recombinant in the ARG is
-that 6 sites represent a 6bp deletion and so are not used for inference.
-The remaining three sites (T22917G, T23018G, and G23040A)
-are not enough to exceed our k=4 threshold  for recombinant designation.
-Including deletions in inference would certainly
-lead to XAJ being detected as a recombinant,
-and with a lower threshold of k=3
-the distribution of mutations on the ARG topology suggests
-that it is is even possible that XAJ is a 2-breakpoint recombinant
-with the region from 21765-23535
-coming from the BA.4 labeled sample SRR18848050.
-
-\scghdiscussion{634}
-
-\subsubsection{XAN/XAV}
-\label{sec:dd_xan}
-
-XAN and XAV are shown in this subgraph, as they share some mutations.
-In the Pango designation issues
-(\url{https://github.com/cov-lineages/pango-designation/issues/771};
-\url{https://github.com/cov-lineages/pango-designation/issues/911}),
-XAN and XAV are suggested to be a recombinant between
-BA.2 (left parent) and BA.5 (right parent).
-Both these Pango lineages were detected as non-recombinants by sc2ts,
-because the number of supporting loci fell below our threshold of detection.
-In the ARG, we found only one mutation (C9866T)
-supporting the suggested BA.2 parent for both XAN and XAV.
-
-\scghdiscussiontwo{783}{XAN}
-
-\scghdiscussiontwo{785}{XAV}
-
-\subsubsection{XAS}
-\label{sec:dd_xas}
-
-XAS was proposed as a recombination between
-BA.5 (left parent) and BA.2 (right parent).
-However, we find that this is only two mutations
-different from e.g. ERR9929846, which is a BA.4 sample.
-In fact, XAS only differs from the BA.4 root by a single mutation,
-C27945T, which was identified in the original Pango designation issue
-(\url{https://github.com/cov-lineages/pango-designation/issues/882})
-as an important ORF truncation mutation
-which is also present in a few BA.2.65 samples.
-It is possible that C27945T was acquired by recombination,
-but this is clearly below our threshold of detection.
-
-\scghdiscussion{784}
-
-\subsubsection{XAZ}
-\label{sec:dd_xaz}
-
-In the Pango designation issue
-(\url{https://github.com/cov-lineages/pango-designation/issues/797}),
-XAZ is suggested as a two-breakpoint recombinant between
-BA.2.5 and BA.5 (the middle parent),
-with three sites supporting the BA.2.5 parent
-on the left end (C2232T, C3317T, and T3358C).
-In the ARG, this is very close to being a recombinant of
-BA.2.5 (ERR9615610) on the left side of the breakpoint,
-as not only are the three mutations present in this sample,
-but there is another site which supports recombination 
-C1912T, which is not reported in the Pango designation issue).
-XAZ has C44T which is lacking in all BA.2.5 samples,
-leaving only three net supporting loci for a putative recombinant.
-However, position 44 is often clipped,
-so this is not strong evidence against a recombinant, and
-therefore XAZ is a marginal case of recombination.
-We found no evidence for the third segment
-inherited from a BA2.5 parent on the right end.
-In the ARG, all the XAZ samples form a clade.
-Here we display only the first 20 XAZ samples.
-
-\scghdiscussion{786}
-
-\subsubsection{XBB}
-\label{sec:dd_xbb}
-
-Most of the XBB samples descend from a well supported recombination node.
-In the Pango designation issue
-(\url{https://github.com/cov-lineages/pango-designation/issues/1058}),
-the Pango lineages of the suggested left and right parents are
-BJ.1 and BM.1.1.1, respectively.
-In the ARG, the Pango lineage of the right parent is identical,
-but that of the left parent is a related ancestral Pango lineage
-(BA.2.10 instead of BA.2.10.1.1 = BJ.1).
-This is likely due to the absence of BJ.1 in the ARG:
-only four BJ.1 samples exist in the Viridian v04 dataset,
-a single one passed sequence-level QC, and
-that was then rejected due to high HMM cost.
-Note that BJ.1 differs from BA.2.10 by over 20 mutations
-(\url{https://github.com/cov-lineages/pango-designation/issues/915}),
-explaining the large number of (gold) de novo mutations
-in the XBB copying pattern.
-Had BJ.1 samples been present in the ARG,
-the left branch of
-the XBB-associated recombination node
-may have attached to a BJ.1-labeled node, and
-the origin of XBB would required fewer mutations.
-
-\scghdiscussion{929}
-
-%% JK: commented this out as it's confusing the relationship between XBB.1 and
-%% XBB.x. Note that most methods pick out XBB.x as recombinant, so it's not that 
-%% remarkable.
-
-% \subsubsection{XBB.1}
-% \label{sec:dd_xbb.1}
-
-% XBB.1 is characterized by an additional mutation in Spike, G22317T (G252V) and
-% is not designated as a recombinant of XBB in the Pango designation issue
-% (https://github.com/cov-lineages/pango-designation/issues/1088).
-% The ARG agrees with this and places the characteristic mutation
-% above the root of a single non-recombinant XBB.1 clade.
-% We include this non-recombinant lineage
-% in order to highlight some well supported subsequent recombination nodes
-% within XBB.1, such as the (pink) one leading to two XBB.1-labeled samples
-% at the bottom left of the subgraph which we refer to as XBB.x.
-% XBB.x is well supported: its left parent is a reconstructed internal node,
-% which is directly supported by a BQ.1-labeled sample, and
-% its right parent is a BB.1.5.18-labeled sample (ERR10792363).
-% This highlights the automatic ability of sc2ts
-% to infer sequential recombination events
-% (i.e. recombinants-of-recombinants),
-% even those involving few samples.
-% A few other recombination events occur within XBB.1,
-% but they either fail QC checks, or
-% (as the other one shown here, to the bottom right)
-% lead to a single sample so may be
-% e.g. caused by accidental mixing of samples during sequencing.
-
-\subsubsection{XBD}
-\label{sec:dd_xbd}
-
-All the XBD samples descend from a well supported recombination node.
-Its left and right parents are reconstructed internal nodes,
-both of which are directly supported by samples
-(SRR21382561 and ERR9825609, respectively)
-that are identical to them.
-
-\subsubsection{XBE}
-\label{sec:dd_xbe}
-
-In the Pango designation issue
-(\url{https://github.com/cov-lineages/pango-designation/issues/1246}),
-XBE was suggested as a recombinant between
-BA.5.2 (left parent; most of the genome) and
-BE.4 (right parent; somewhere from 23,609 onwards). 
-However, sc2ts treats the 65 XBE samples
-as non-recombinant descendants of BA.5.2
-because the number of supporting sites
-fell below our threshold of detection.
-We have included the BE.4 lineage on the subgraph,
-revealing two mutations (C27800T, and G28681T)
-which occur in parallel above the root of the XBE clade
-and along the path ascending from the displayed BE.4 sample,
-suggesting that if we used k=2
-as the recombinant threshold,
-sc2ts would indeed detect XBE as a recombinant.
-
-\scghdiscussion{930}
-
-\subsubsection{XBF}
-\label{sec:dd_xbf}
-
-All the XBF samples descend from a well supported recombination node,
-with three mutations on the left of the breakpoint.
-Its left parent is a sample assigned to BA.5.2.1 (SRR20305553).
-Its right parent is a reconstructed internal node
-that is supported by a CJ.1-labeled sample (SRR21948068),
-which differs by three mutations.
-
-\subsubsection{XBG}
-\label{sec:dd_xbg}
-
-All the XBG samples descend from a well supported recombination node,
-with two mutations, one on each side of the breakpoint.
-Both its left and right parents are samples
-(SRR19920084 and ERR9852351, respectively).
-
-\subsubsection{XBH}
-\label{sec:dd_xbh}
-All the XBH samples descend from a well supported recombination node,
-despite two reversions (T222C and T12789C, respectively),
-which occur at excluded sites and
-were remapped by parsimony during postprocessing.
-If these two sites had been used during ARG inference,
-the branch attached to BA.2.1 would have been attached to
-the BA.2.5 shown on the top left instead,
-leading to a more parsimonious local topology
-with two mutations on the left branch. This would also remove the two
-blue lines on the left and side of the copying pattern.
-
-\subsubsection{XBQ/XBK/XBK.1}
-\label{sec:dd_xbq}
-
-XBK and XBQ are siblings, and neither appear to be recombinants in the ARG,
-as they are closely related to the non-recombinant node that originates CJ.1.3. 
-At the time of designating XBK
-a hypothesis was made that a missing common ancestor could exist 
-(\url{https://github.com/cov-lineages/pango-designation/issues/1440\#issuecomment-1355264973}),
-making XBK non-recombinant.
-Exactly such a missing ancestor
-is found as the upper internal node labeled CJ.1 in the ARG,
-constructed by sc2ts as part of its tree building steps.
-
-\scghdiscussiontwo{931}{XBQ}
-
-\scghdiscussiontwo{932}{XBK}
-
-\subsubsection{XBM}
-\label{sec:dd_xbm}
-
-All the XBM samples descend from a well supported recombination node.
-Its left parent is a reconstructed internal node
-that is directly supported by a BA.2.76-labeled sample identical to it.
-Its right parent is a BF.3-labeled sample.
-
-\subsubsection{XBR}
-\label{sec:dd_xbr}
-This one XBR sample descends from a well supported recombination node.
-Its left parent is a BN.3.1-labeled sample, and
-its right parent is a reconstructed internal node
-that is directly supported by a BQ.1.25-labeled sample identical to it.
-
-\subsection{Pango X lineages not added to the ARG}
-\label{sec:supplement_rejected_pango_x_lineages}
-
-Ten Pango X lineages which are present in the Viridian v04 dataset
-were not integrated into the ARG
-due to high HMM costs and/or insufficient sample support,
-but their samples were matched against the ARG during inference.
-% https://github.com/jeromekelleher/sc2ts-paper/issues/366
-Table~\ref{tab:extra_hmm_results} summarizes
-the HMM matching results of these Pango X lineages.
-All were identified as recombinants by sc2ts, except XBL.
-(For XBL, the evidence for recombination
-proposed in the Pango designation issue \cite{PangoIssueXBL}
-involves only three informative sites
-supporting the suggested rightmost parent, and
-therefore this likely fell below our threshold of detection.)
-% https://github.com/jeromekelleher/sc2ts-paper/issues/379
-Importantly, these results demonstrate that sc2ts can detect
-recombinants with multiple breakpoints (XAY, XBC, and XBT).
-% https://github.com/jeromekelleher/sc2ts-paper/discussions/776
-% https://github.com/jeromekelleher/sc2ts-paper/discussions/788
-XBC and XBT are relatively clean examples of multiple-breakpoint recombinants,
-with the Pango lineages of the parents and the breakpoint intervals
-concordant with those suggested in the Pango designation issues,
-but XAY is more complex, possibly involving up to six breakpoints as per sc2ts.
-
-\begin{table}
-\centering
-\caption{
-Sc2ts detection results for Pango X lineages not added to the ARG.
-The number of samples in Viridian v04 dataset
-assigned to these Pango X lineages,
-the number of samples which passed sequence-level quality control, and
-the number of recombination breakpoints
-in the Viterbi solutions of the samples
-are shown.
-}
-\label{tab:extra_hmm_results}
-\begin{tabular}{crrr}
-\toprule
-Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
-\midrule
-XBJ & 2 & 1 & 1\\
-XBP & 8 & 2 & 1\\
-XBS & 19 & 15 & 1\\
-XBW & 1 & 1 & 1\\
-XCA & 11 & 4 & 1\\
-XAK & 2 & 2 & 1\\   % Multiple breakpoint Pango X
-XAY & 15 & 11 & 6\\ % Multiple breakpoint Pango X
-XBC & 36 & 21 & 3\\ % Multiple breakpoint Pango X
-XBL & 2 & 2 & 0\\   % Multiple breakpoint Pango X
-XBT & 1 & 1 & 2\\   % Multiple breakpoint Pango X
-\bottomrule
-\end{tabular}
-\end{table}
-
-% 2 + 8 + 19+ 1 + 11+ 2 + 15+ 36+ 2 + 1
-% 97
-
-\subsection{Analysis of the Jackson et al. recombinants}
-\label{sec:jackson_recombs}
-
-Jackson et al.\cite{Jackson2021} conducted a detailed analysis of
-early recombinants involving B.1.1.7 (Alpha),
-including the first designated Pango X lineage, XA.
-The authors first performed a targeted search for candidate recombinants by
-scanning sampled sequences for sequence motifs associated with Alpha,
-followed by an analysis to reduce the list of plausible recombinants
-displaying signs of onward transmission.
-Here we compared the sc2ts detection results of these recombinants
-with the detection results reported by Jackson et al.
-
-Jackson et al. described
-four groups of recombinants (named A to D) and four singleton recombinants.
-Of the 16 samples analyzed by Jackson et al., 12 were included in the ARG;
-the remaining 4 were excluded
-because the sequences did not pass the QC filtering criteria or
-because of a high HMM cost.
-Groups A-D are summarized in Table~\ref{tab:jackson} and
-show excellent concordance
-in terms of the parental Pango lineages and breakpoint intervals.
-All four groups are directly associated with
-strongly supported recombination events in the ARG
-in terms of the number of averted mutations (A: 16, B: 6, C: 30, D: 18).
-In the ARG, the singleton CAMC-CBA018 has
-the parents B.1.177+B.1.1.7 and
-the breakpoint interval 16,177--20,133, and
-is strongly supported by 14 averted mutations.
-In the Jackson et al. results, this singleton is
-almost identical with the same parent lineages and
-a breakpoint interval of 11,395–-21,993.
-In the ARG, the singleton MILK-103C712 has
-the parents B.1.177.16+B.1.1.7 and
-a breakpoint interval of 24915--27972,
-again strongly supported by 11 averted mutations.
-In the Jackson et al. results,
-this lineage has the same parents,
-with the breakpoint interval 26,800–-27,974.
-In the ARG, the singleton QEUH-1067DEF has
-the parents B.1.1.7+B.1.177.9 and
-a breakpoint interval of 7,729--10,870,
-supported by 11 averted mutations.
-In the Jackson et al.\ results,
-this lineage has the same parents,
-with the breakpoint interval 6,953--10,872.
-
-See notebook
-\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_jackson\_recombinants.ipynb}
-{analysis\_jackson\_recombinants.ipynb}
-for details.
-
-\begin{table}
-\centering
-\caption{
-Comparison of recombination breakpoint
-intervals and parent lineages for Groups A-D
-reported by Jackson et al.\ with the corresponding
-recombination events in the sc2ts ARG.
-The second column gives the number of sequences in the group.
-See the text for details of the groups and the sequences included.
-limited to the samples considered by Jackson et al.
-The breakpoint coordinates in Table 1 of Jackson et al.\ have been altered as follows: 
-we subtract one to the left coordinates and add one to the right coordinates 
-to correspond to the \texttt{tskit} definition of inheritance on either side of a breakpoint,
-and add one to the right coordinate to make the intervals right-exclusive.
-}
-\label{tab:jackson}
-\begin{tabular}{lllr@{--}lr@{+}l}
-\toprule
-Group & Sequences & Method & \multicolumn{2}{c}{Breakpoint interval} & \multicolumn{2}{r}{Parent lineages} \\
-\midrule
-A (XA)       & 2   & Jackson        &  21,254&21,767 & B.1.177&B.1.1.7 \\
-             &     &\texttt{sc2ts} &  20,411&21,765 & B.1.177.18&B.1.1.7 \\
-\cmidrule{3-7}
-B            & 2   & Jackson        &  6,527&6,956 & B.1.36.28&B.1.1.7  \\
-             &     &\texttt{sc2ts} &   6,529&6,954 & B.1.36.28&B.1.1.7  \\
-\cmidrule{3-7}
-C            & 3   &Jackson         &  24,913&28,653 &  B.1.1.7&B.1.221 \\
-             &     &\texttt{sc2ts} &  25,997&27,972 &  B.1.1.7&B.1.221.1 \\
-\cmidrule{3-7}
-D            & 3   & Jackson        &  21,574&23,065 &  B.1.36.17&B.1.1.7 \\
-             &     &\texttt{sc2ts} &  22,445&23,063 &  B.1.36.39&B.1.1.7 \\
-\bottomrule
-\end{tabular}
-\end{table}
-
 \FloatBarrier
@@
 
+\newcommand{\cprheight}{1.5em}
+% Add some more space between rows. Note this it would be sligntly nicer if the 
+% label was aligned vertically in the center of the row, but this is good
+% enough for now.
+\renewcommand{\arraystretch}{2}
+
 \begin{figure}
 \centering
-\begin{tabular}{cc}
-\includegraphics[width=0.5\linewidth]{ripples_p4_sc2ts_events.pdf}&
-\includegraphics[width=0.5\linewidth]{ripples_p3_sc2ts_events.pdf}\\
-(A) & (B) \\
-\end{tabular}
-\caption{Intersection of the RIPPLES and sc2ts recombination events at
-different values of the RIPPLES parsimony parameter, p.}
-\label{fig:ripples_sc2ts_events}
+\begin{tabularx}{\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
+XC & \includegraphics[height=\cprheight]{static/XC.png}\\
+XBR & \includegraphics[height=\cprheight]{static/XBR.png}\\
+XA & \includegraphics[height=\cprheight]{static/XA.png}\\
+XS & \includegraphics[height=\cprheight]{static/XS.png}\\
+XL & \includegraphics[height=\cprheight]{static/XL.png}\\
+XBG &\includegraphics[height=\cprheight]{static/XBG.png}\\
+XQ & \includegraphics[height=\cprheight]{static/XQ++.png}\\
+XBD & \includegraphics[height=\cprheight]{static/XBD.png}\\
+XM & \includegraphics[height=\cprheight]{static/XM+XAL.png}\\
+XBH & \includegraphics[height=\cprheight]{static/XBH.png}\\
+XBB & \includegraphics[height=\cprheight]{static/XBB.png}\\
+XY & \includegraphics[height=\cprheight]{static/XY.png}\\
+XF & \includegraphics[height=\cprheight]{static/XF.png}\\
+XBF & \includegraphics[height=\cprheight]{static/XBF.png}\\
+XW & \includegraphics[height=\cprheight]{static/XW.png}\\
+XG & \includegraphics[height=\cprheight]{static/XG.png}\\
+\hline
+Xx & \includegraphics[height=\cprheight]{static/XZ++.png}\\
+XE/XH  & \includegraphics[height=\cprheight]{static/XE+XH.png}\\
+XBM & \includegraphics[height=\cprheight]{static/XBM.png}\\
+XJ & \includegraphics[height=\cprheight]{static/XJ.png}\\
+XAF & \includegraphics[height=\cprheight]{static/XAF.png}\\
+\end{tabularx}
+\caption{Copying patterns for the 21 recombination events associated with Pango 
+X lineages listed in Table~\ref{tab:pango_x_lineages} (in the same order). 
+The horizontal line separates Type I and Type II events.
+See also Document S3 for exact positions and nucleotide bases.}
+\label{fig:pango_x_copying_patterns}
 \end{figure}
+
+% Reset this in case we make any more tables beyond this
+\renewcommand{\arraystretch}{1}
 
@@
 
-\newcommand{\cprheight}{4.0em}
+\renewcommand{\cprheight}{4.0em}
 
@@
 
-\renewcommand{\cprheight}{1.5em}
-% Add some more space between rows. Note this it would be sligntly nicer if the 
-% label was aligned vertically in the center of the row, but this is good
-% enough for now.
-\renewcommand{\arraystretch}{2}
-
 \begin{figure}
 \centering
-\begin{tabularx}{\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
-XC & \includegraphics[height=\cprheight]{static/XC.png}\\
-XBR & \includegraphics[height=\cprheight]{static/XBR.png}\\
-XA & \includegraphics[height=\cprheight]{static/XA.png}\\
-XS & \includegraphics[height=\cprheight]{static/XS.png}\\
-XL & \includegraphics[height=\cprheight]{static/XL.png}\\
-XBG &\includegraphics[height=\cprheight]{static/XBG.png}\\
-XQ & \includegraphics[height=\cprheight]{static/XQ++.png}\\
-XBD & \includegraphics[height=\cprheight]{static/XBD.png}\\
-XM & \includegraphics[height=\cprheight]{static/XM+XAL.png}\\
-XBH & \includegraphics[height=\cprheight]{static/XBH.png}\\
-XBB & \includegraphics[height=\cprheight]{static/XBB.png}\\
-XY & \includegraphics[height=\cprheight]{static/XY.png}\\
-XF & \includegraphics[height=\cprheight]{static/XF.png}\\
-XBF & \includegraphics[height=\cprheight]{static/XBF.png}\\
-XW & \includegraphics[height=\cprheight]{static/XW.png}\\
-XG & \includegraphics[height=\cprheight]{static/XG.png}\\
-\hline
-Xx & \includegraphics[height=\cprheight]{static/XZ++.png}\\
-XE/XH  & \includegraphics[height=\cprheight]{static/XE+XH.png}\\
-XBM & \includegraphics[height=\cprheight]{static/XBM.png}\\
-XJ & \includegraphics[height=\cprheight]{static/XJ.png}\\
-XAF & \includegraphics[height=\cprheight]{static/XAF.png}\\
-\end{tabularx}
-\caption{Copying patterns for the 21 recombination events associated with Pango 
-X lineages listed in Table~\ref{tab:pango_x_lineages} (in the same order). 
-The horizontal line separates Type I and Type II events.
-See also Document S3 for exact positions and nucleotide bases.}
-\label{fig:pango_x_copying_patterns}
+\begin{tabular}{cc}
+\includegraphics[width=0.5\linewidth]{ripples_p4_sc2ts_events.pdf}&
+\includegraphics[width=0.5\linewidth]{ripples_p3_sc2ts_events.pdf}\\
+(A) & (B) \\
+\end{tabular}
+\caption{Intersection of the RIPPLES and sc2ts recombination events at
+different values of the RIPPLES parsimony parameter, p.}
+\label{fig:ripples_sc2ts_events}
 \end{figure}
-
-% Reset this in case we make any more tables beyond this
-\renewcommand{\arraystretch}{1}
 

### 2025-11-05 latex/main.tex
Updates from Overleaf

@@
 We hope that the ARG provided here will similarly
-enable a range recombination-aware analyses of SARS-CoV-2, and 
-catalyse a wave of methodological improvements fully integrating
+enable a range of recombination-aware analyses of SARS-CoV-2, and 
+catalyse a wave of methodological improvements which fully integrate
 recombination into pathogen genomics.
@@
 Muller2022bayesian}.
-By capturing recombination in real time at a global scale, our approach extends the established utility of comprehensive genealogies~\cite{Hinrichs2024,McBroome2024}, offers clear insight into recombination dynamics, and supports timely investigation of .
-The novelty of sc2ts and the inferred ARG motivates our  focus on methodological details. 
-Through these detailed analyses we validate the ARG itself,
-ensuring that it is a trustworthy basis for future research,
-and deepen our understanding of the inference method, and 
-how it may be adapted and improved (see Limitations of the Study).
-
-Study of the sc2ts ARG provides insights in the  contributions of this work are provided by our 
-insights into the evolution of SARS-CoV-2 obtained by studying
-the sc2ts ARG. Firstly, we have clarified the relationships among
-the recombinant Pango X-lineages, suggesting significant 
+By capturing recombination in real time at a global scale, our approach extends the established utility of comprehensive genealogies~\cite{Hinrichs2024,McBroome2024}, providing clear insight and enabling timely investigation of recombination-aware pathogen dynamics.
+The novelty of the inferred ARG motivates our detailed analyses of the methodology behind sc2ts, which we use to validate the ARG itself.
+This ensure that the ARG is a trustworthy basis for future research, deepens our understanding of the inference method, and suggests avenues for improvement (see Limitations of the Study).
+
+Study of the sc2ts ARG gives insight into the evolution of SARS-CoV-2, and this work therefore makes several additional contributions to the field. First, we clarify the relationships among
+recombinant Pango X-lineages, suggesting significant 
 clustering of named lineages under shared recombination events.
-Secondly, we have used the ARG to study breakpoint intervals 
-for recombination events, adding a novel temporal dimension
-and demonstrating a simple relationship with the 
+Second, we use the ARG to study breakpoint intervals 
+for recombination events, add a novel temporal dimension
+and show a simple relationship with the 
 size of the interval in base pairs derived from theory.
@@
 conservative parameter choices and real-time handling of 
-recombination events, we have provided a stringent lower
-bound on the amount of recombination that occured during
+recombination events, we  provide a stringent lower
+bound on the amount of recombination that occurred during
 the pandemic phase of SARS-CoV-2.
@@
 % SHZ: WIP
-We close by emphasizing several current limitations
-and potential directions to improve sc2ts,
-which we hope will also inspire future research
-to develop ARG inference methods for pathogens beyond SARS-CoV-2.
+We close by emphasizing several current limitations of sc2ts
+and suggest potential directions for improvement
+which we hope will inspire future development of ARG inference
+methods for pathogens beyong SARS-CoV2.
 

### 2025-11-05 latex/main.tex
First pass at first para

@@
 to an alternative genome rather than the true parent.
-Such genomic regions can be mistaken for evidence of recombination,
-because the LS HMM counts each base separately.
+Such genomic regions can be mistaken for evidence of recombination.
 We use a simple quality control (QC) measure
@@
 \subsection*{Comparison with UShER+RIPPLES}
+
+Things to say.
+\begin{itemize}
+\item Explain how it works. Post-hoc. Note methodological distinction 
+with sc2ts.
+\item Comparison at k/p=4 overall.
+\item Discussion of X lineages.
+\item Discussion of parameter choice and our lower bound on recomb.
+\end{itemize}
+
+The most comprehensive analysis on recombination in SARS-CoV-2 to 
+date has been performed using the combination of UShER and 
+RIPPLES~\cite{Turakhia2022}. RIPPLES works by taking a
+pre-existing UShER phylogeny and systematically breaking up
+sequences subtended by long branches into multiple segments 
+that are placed independently.
+A recombination event at a node is flagged if
+the overall parsimony of these multiple placements
+sufficiently improves over
+the original single-parent origin in the phylogeny.
+This approach has much in common with the LS HMM
+used in sc2ts, but with one important distinction.
+While sc2ts detects and incorporates recombination
+in ``real-time'' as sequences are added to the ARG in daily
+batches, RIPPLES operates on a complete phylogeny,
+attempting to detect recombination post-hoc among millions of nodes. 
+Like any phylogenetic reconstruction method
+that ignores recombination, UShER must force placement of recombinant
+sequences in places that minimize the discrepancy across the entire sequence,
+which may not be near any of the parent lineages. 
+These permanent misplacements distort the phylogeny~\cite{Schierup2000consequences}
+and subsequent tree reorganisations to maximise overall parsimony~\cite{DeMaio2023} 
+may break up the erroneous long branches used to detect 
+recombination.
+
+[Text from Sally]
+The main advantage of sc2ts over previous methods to infer recombination
+phylogenetically is that recombination events are explicitly handled as the ARG
+is built. Other methods (e.g., RIPPLES) first build a phylogeny, assuming no
+recombination, and then look for unusual features (e.g., unusually long
+branches). Phylogenetic methods that ignore recombination, however, distort the
+tree shape (REFS - Schierup and Hein 2000) and force placement of recombinant
+lineages in places that minimize the discrepancy across the entire sequence,
+rather than near either of the parent lineages.  As a consequence, many true
+recombination events can be missed because their branches are artificially
+inferred to be short on reconstructed trees that do not incorporate
+recombination. In addition, many of the recombinant sequences can be placed
+before their parents in time, because recombinants can appear to be more
+similar to ancestral sequences that are known with more uncertainty [OR because
+recombinants are placed anywhere on a tree that best fits].
+
+This ``parsimony improvement'' parameter ($p$)
+is equivalent to the sc2ts recombination penalty ($k$).
+
 RIPPLES~\cite{Turakhia2022} is a method for identifying
@@
 finds parsimonious placements for each segment independently.
-A recombination event at that node is then flagged if
-the overall parsimony of these multiple placements
-sufficiently improves over
-the original single-parent origin in the phylogeny.
-This ``parsimony improvement'' parameter ($p$)
-is equivalent to the sc2ts recombination penalty ($k$).
 
@@
 of the UShER+RIPPLES results.
+
+that from UShER+RIPPLES,
+largely reflecting differences in model formulation and parameter choice.
+When applied \textit{post hoc} to an UShER phylogeny of
+$\sim$1.6 million genomes sampled up to May, 2021,
+RIPPLES reported 589 unique recombination events with 43,104 descendant
+samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
+Using the same time interval, sc2ts \(k=4\) identifies
+14 high-confidence recombination events with
+172 descendant samples among 501,235 samples \(0.034\%\),
+a 40- to 80-fold reduction.
+Much of this difference is explained by
+the stricter parsimony penalty used here:
+increasing the penalty from p=3 (as used in the RIPPLES analysis) to
+k=4 greatly reduces false positives
+while retaining nearly all verified recombinant lineages,
+including those missed by RIPPLES at equivalent thresholds.
+These indirect comparisons suggest that
+our estimate represents a conservative lower bound.
+In future studies, dissecting the methodological differences
+between sc2ts and UShER+RIPPLES may
+inspire new ideas to develop or improve ARG building methods
+for SARS-CoV-2 (and potentially other pathogens).
 
@@
 which we hope will inspire future development of ARG inference
-methods for pathogens beyong SARS-CoV2.
-
-We suggest several ways to extend the HMM.
-First, reversion of deletions can indicate recombination,
-because bases that are lost are unlikely to be regained,
-except via recombination with a lineage that has those bases.
-We presently treated multi-base deletions as missing data, and
-added deletions to the ARG \textit{post hoc}.
-Instead, by explicitly modeling deletions (and more generally insertions) in the HMM,
+methods for pathogens beyond SARS-CoV2.
+
+There are several ways to extend the HMM.
+First, we currently treat multi-base deletions as missing data,
+adding indels to the ARG \textit{post hoc}.
+However, by explicitly modeling deletions as single HMM events, and
+accounting for fact that reversion of deletions is more likely to
+occur through recombination than back-mutation,
 we should be able to recover additional signals of recombination
 and reduce spurious recombinants.
-Second, the current HMM assumes a simplistic mutation model
-(where mutation to any other allele is equally probable) and
-a uniform mutation rate along the SARS-CoV-2 genome.
+Second, the current HMM assumes a simplistic model of equiprobable
+mutation between bases with a uniform mutation rate along the genome.
 However, genomic analyses have revealed
@@
 splitting of suspiciously long branches (e.g., Delta and Omicron BA.2).
-These methods could instead be run immediately
-after sample matching during ARG inference,
+These methods could instead be run during ARG inference,
+immediately after sample matching,
 reducing the number of ARG post-processing steps.
@@
 % in the viral context?
+% SHZ: Yes, agree with this.
 as a set of \emph{nodes}, representing a genome that existed at some specific time,
@@
 or in a vectorisable packed binary format for efficiency
-(used to attach sample identifiers and pango assignments in the final ARGs).
+(used to attach sample identifiers and Pango lineage assignments in the final ARGs).
 Tskit also provides a ``provenance'' table,
@@
 and most sites in this list were typical
-in terms of the numbers of mutations.
+in terms of the number of mutations.
 We therefore masked our 100 mutation-rich sites
@@
 It is a Hidden Markov Model (HMM) in which a focal genome
-is modelled as a
+is modeled as a
 sequence of nucleotides that are probabilistically emitted as
@@
 The HMM cost threshold approach outlined in the previous section is 
-effective at removing time-travelling samples, but it also filters out 
+effective at removing time-traveling samples, but it also filters out 
 important evolutionary events such as 
@@
 correlated error is non-trivial, and sc2ts has 5 parameters to fine
-tune the behaviour. Firstly, we require at least 10 samples in the
+tune the behavior. Firstly, we require at least 10 samples in the
 group to ensure there is sufficient support over a time window of 
@@
 Retrieving data from the base alignments and ARGs for 2.48 million samples 
-at the same time over all 27,507 sites took only 6 minutes,
-with a peak memory usage of about 5GiB of RAM
-on a standard desktop computer.
+at the same time over all 27,507 sites took a little under 5 minutes,
+with a peak memory usage of about 4.5GiB of RAM
+on a standard laptop.
 

### 2025-11-05 latex/main.tex
Second para

@@
 
-Things to say.
-\begin{itemize}
-\item Explain how it works. Post-hoc. Note methodological distinction 
-with sc2ts.
-\item Comparison at k/p=4 overall.
-\item Discussion of X lineages.
-\item Discussion of parameter choice and our lower bound on recomb.
-\end{itemize}
-
 The most comprehensive analysis on recombination in SARS-CoV-2 to 
@@
 and subsequent tree reorganisations to maximise overall parsimony~\cite{DeMaio2023} 
-may break up the erroneous long branches used to detect 
-recombination.
-
-[Text from Sally]
-The main advantage of sc2ts over previous methods to infer recombination
-phylogenetically is that recombination events are explicitly handled as the ARG
-is built. Other methods (e.g., RIPPLES) first build a phylogeny, assuming no
-recombination, and then look for unusual features (e.g., unusually long
-branches). Phylogenetic methods that ignore recombination, however, distort the
-tree shape (REFS - Schierup and Hein 2000) and force placement of recombinant
-lineages in places that minimize the discrepancy across the entire sequence,
-rather than near either of the parent lineages.  As a consequence, many true
-recombination events can be missed because their branches are artificially
-inferred to be short on reconstructed trees that do not incorporate
-recombination. In addition, many of the recombinant sequences can be placed
-before their parents in time, because recombinants can appear to be more
-similar to ancestral sequences that are known with more uncertainty [OR because
-recombinants are placed anywhere on a tree that best fits].
-
-This ``parsimony improvement'' parameter ($p$)
-is equivalent to the sc2ts recombination penalty ($k$).
-
-RIPPLES~\cite{Turakhia2022} is a method for identifying
-recombination events using a phylogeny inferred by UShER.
-This distinguishes RIPPLES from other recombination detection
-methods such as 3SEQ, which classify particular \emph{sequences}
-as recombinant or not and can only identify 
-potential onward transmission post-hoc. 
-The combination of UShER and RIPPLES
-is therefore the only method that is directly comparable to sc2ts. 
-RIPPLES works by first identifying long 
-branches in the input tree, by default those carrying 3 or more 
-mutations. For each candidate branch,
-RIPPLES systematically splits the sequence into 2 or 3 segments, and
-finds parsimonious placements for each segment independently.
+may break up the erroneous long branches used to detect recombination.
 
@@
 here, and do not perform QC filtering of either.
-Using the RIPPLES parameter $p=4$ (matching $k=4$ in the sc2ts ARG) resulted in 1174 reported recombination events.
+Using the RIPPLES parameter $p=4$ (matching $k=4$ used in the sc2ts ARG) 
+resulted in 1174 reported recombination events.
 We ran 3SEQ on each of the candidate trios (STAR Methods) and 601 
@@
 correspond to a recombination event in the sc2ts ARG can largely
-be explained by a combination of two factors.
-Firstly, 408 of these RIPPLES identified recombination events
-have one or more parents that are \emph{younger} than the
-recombinant sequence based on metadata. Thus, the parental
-sequences were simply not available to match against for 
-sc2ts (which enforces strict time ordering) when the 
-recombinant sequence was inserted. Secondly, of the 179
-remaining events with matching clades and in which the 
-sc2ts node has more than 4 mutations, of the total
-1,167 mutations, 452 of these occurred at sites that were 
-remapped during post-hoc parsimony. The remaining ``missing''
-events are likely attributable to the finer details of the 
-differences in tree building approaches between UShER and sc2ts.
-
+be explained by a combination the choice of sites 
+used for inference, and the strict time ordering imposed
+by sc2ts (Document~\ref{sec:detailed_ripples}).
 The 387 events detected by sc2ts and not by UShER+RIPPLES
@@
 to explain, which is beyond the scope of this work.
-However, the set does contain some well characterized events such as XBB.
-We also compared the recombination events
-output by RIPPLES with the Pango X lineages
-(Table~\ref{tab:ripples_x_lineages}).
-At p=4, we found that only 9 of the Pango X lineages were present and
-with unequivocal recombinant origin for 6 (contrast with 
-16 Type I events in Table~\ref{tab:pango_x_lineages}).
-We therefore reran at p=3, resulting in 4111 events overall,
-and increasing the intersection of events with sc2ts to 595 (70\%).
-At p=3, events involving XX of the 44 Pango X lineages
+However, the set does contain some well characterized events 
+associated with the Pango X lineages.
+Table~\ref{tab:ripples_x_lineages} shows the events detected 
+by UShER+RIPPLES associated with samples assigned to Pango X
+lineages. At the RIPPLES ``parsimony improvement'' parameter
+$p=4$ (equivalent to $k=4$ parameter used to build the sc2ts ARG)
+only 9 of the Pango X lineages are present and
+with unequivocal recombinant origin for 6 
+(contrast with 16 Type I events in Table~\ref{tab:pango_x_lineages}).
+We therefore reran RIPPLES with $p=3$, resulting in 4111 events overall,
+and increasing the intersection of events with sc2ts to 595 (70\%;
+Figure~\ref{fig:ripples_sc2ts_events}).
+At this more lenient threshold, 23 of the 44 Pango X lineages
 are present in the dataset, with unequivocal origins for 10 lineages, and
-7 lineages other lineages associated with two competing events.
-[More summary]
-[Note somewhere here that XB and XAJ are found to be recombinants 
-at p=4, showing that UShER is doing better here for some reason]
-
-In terms of \emph{detecting} recombination events, 
-sc2ts and the combination of UShER and RIPPLES have much in common, 
-and it is likely that very similar results could be obtained if run
-in the same way. That is, if we constructed an UShER tree by 
-incrementally adding the same daily batches as used for sc2ts
-and examined any newly-added long branches each day
-using RIPPLES, we would likely get very similar results to sc2ts
-in terms of recombinants detected.
-For a given matching operation, RIPPLES and the LS HMM 
-(under its current parameterisation) 
-are deterministically solving the same well-defined problem
-(albeit with radically different implementations) and they 
-must therefore produce similar answers. 
-Where sc2ts and UShER+RIPPLES diverge, however, is in terms 
-of how well such signals of recombination can be preserved
-over longer timescales. Continuing our thought experiment
-in which we run UShER and RIPPLES in daily batches, suppose
-that on a given day a branch with 4 mutations is identified 
-as being a recombinant. Unless this branch is specifically 
-``protected'' as representing a recombinant it may 
-be broken up by subsequent parsimony operations, thus 
-erasing the signal of recombination. This would appear
-to be the mostly likely explanation for the ``missing'' 
-X-lineages like XBB (detected by sc2ts at $k=4$ but 
-not by UShER+RIPPLES at $p=4$).
-
-See Document~\ref{sec:detailed_ripples} for more detailed analysis 
-of the UShER+RIPPLES results.
+7 lineages associated with two competing events 
+(Document~\ref{sec:detailed_ripples}).
+XB and XAJ are among the lineages detected by UShER+RIPPLES,
+strengthening evidence that these are false negatives 
+for sc2ts (Table~\ref{tab:pango_x_lineages}).
 
@@
 for SARS-CoV-2 (and potentially other pathogens).
+
+See Document~\ref{sec:detailed_ripples} for more detailed analysis 
+of the UShER+RIPPLES results.
 
@@
 
+Firstly, 408 of these RIPPLES identified recombination events
+have one or more parents that are \emph{younger} than the
+recombinant sequence based on metadata. Thus, the parental
+sequences were simply not available to match against for 
+sc2ts (which enforces strict time ordering) when the 
+recombinant sequence was inserted. Secondly, of the 179
+remaining events with matching clades and in which the 
+sc2ts node has more than 4 mutations, of the total
+1,167 mutations, 452 of these occurred at sites that were 
+remapped during post-hoc parsimony. The remaining ``missing''
+events are likely attributable to the finer details of the 
+differences in tree building approaches between UShER and sc2ts
+
 [TODO Discuss  the large effect recombinants inferred by both]
@@
 Clades are matched by finding the most recent common ancestor of
-the samples associated with each UShER/RIPPLES event in the sc2ts ARG.
+the samples associated with each UShER+RIPPLES event in the sc2ts ARG.
 The clade differences indicates how close this match is.

### 2025-11-05 latex/main.tex
Pass through RIPPLES

@@
 sequences in places that minimize the discrepancy across the entire sequence,
-which may not be near any of the parent lineages. 
+which may not be near any of the true parents. 
 These permanent misplacements distort the phylogeny~\cite{Schierup2000consequences}
 and subsequent tree reorganisations to maximise overall parsimony~\cite{DeMaio2023} 
-may break up the erroneous long branches used to detect recombination.
+may break up the artefactual long branches used to detect recombination.
 
@@
 lineages. At the RIPPLES ``parsimony improvement'' parameter
-$p=4$ (equivalent to $k=4$ parameter used to build the sc2ts ARG)
+$p=4$ (equivalent to $k=4$ used to build the sc2ts ARG)
 only 9 of the Pango X lineages are present and
@@
 
-that from UShER+RIPPLES,
-largely reflecting differences in model formulation and parameter choice.
-When applied \textit{post hoc} to an UShER phylogeny of
+The choice of the $p$ threshold has a large effect 
+on the number of recombination events detected by RIPPLES, and 
+previous studies have used the more lenient (default) value of 
+$3$\cite{Turakhia2022,Smith2023tracking}.
+For example, when applied to an UShER phylogeny of
 $\sim$1.6 million genomes sampled up to May, 2021,
@@
 samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
-Using the same time interval, sc2ts \(k=4\) identifies
+Over the same time interval, the sc2ts ARG ($k=4$) identifies
 14 high-confidence recombination events with
-172 descendant samples among 501,235 samples \(0.034\%\),
+172 descendant samples among 501,235 samples (0.034\%),
 a 40- to 80-fold reduction.
-Much of this difference is explained by
-the stricter parsimony penalty used here:
-increasing the penalty from p=3 (as used in the RIPPLES analysis) to
-k=4 greatly reduces false positives
-while retaining nearly all verified recombinant lineages,
-including those missed by RIPPLES at equivalent thresholds.
-These indirect comparisons suggest that
-our estimate represents a conservative lower bound.
-In future studies, dissecting the methodological differences
-between sc2ts and UShER+RIPPLES may
-inspire new ideas to develop or improve ARG building methods
-for SARS-CoV-2 (and potentially other pathogens).
-
-See Document~\ref{sec:detailed_ripples} for more detailed analysis 
-of the UShER+RIPPLES results.
+The sc2ts ARG identifies more of the well-characteristed X-lineages
+at a more stringent threshold, and therefore provides us with a 
+conservative lower bound
+on the prevalance of detectable recombination in SARS-CoV-2.
 

### 2025-11-05 latex/main.tex
Updates from Overleaf

--- 2025-11-05 latex/main.tex
@@
 \subsection*{Comparison with UShER+RIPPLES}
-
-Things to say.
-\begin{itemize}
-\item Explain how it works. Post-hoc. Note methodological distinction 
-with sc2ts.
-\item Comparison at k/p=4 overall.
-\item Discussion of X lineages.
-\item Discussion of parameter choice and our lower bound on recomb.
-\end{itemize}
-
-The most comprehensive analysis on recombination in SARS-CoV-2 to 
-date has been performed using the combination of UShER and 
-RIPPLES~\cite{Turakhia2022}. RIPPLES works by taking a
-pre-existing UShER phylogeny and systematically breaking up
-sequences subtended by long branches into multiple segments 
-that are placed independently.
-A recombination event at a node is flagged if
-the overall parsimony of these multiple placements
-sufficiently improves over
-the original single-parent origin in the phylogeny.
-This approach has much in common with the LS HMM
-used in sc2ts, but with one important distinction.
-While sc2ts detects and incorporates recombination
-in ``real-time'' as sequences are added to the ARG in daily
-batches, RIPPLES operates on a complete phylogeny,
-attempting to detect recombination post-hoc among millions of nodes. 
-Like any phylogenetic reconstruction method
-that ignores recombination, UShER must force placement of recombinant
-sequences in places that minimize the discrepancy across the entire sequence,
-which may not be near any of the parent lineages. 
-These permanent misplacements distort the phylogeny~\cite{Schierup2000consequences}
-and subsequent tree reorganisations to maximise overall parsimony~\cite{DeMaio2023} 
-may break up the erroneous long branches used to detect 
-recombination.
-
-[Text from Sally]
-The main advantage of sc2ts over previous methods to infer recombination
-phylogenetically is that recombination events are explicitly handled as the ARG
-is built. Other methods (e.g., RIPPLES) first build a phylogeny, assuming no
-recombination, and then look for unusual features (e.g., unusually long
-branches). Phylogenetic methods that ignore recombination, however, distort the
-tree shape (REFS - Schierup and Hein 2000) and force placement of recombinant
-lineages in places that minimize the discrepancy across the entire sequence,
-rather than near either of the parent lineages.  As a consequence, many true
-recombination events can be missed because their branches are artificially
-inferred to be short on reconstructed trees that do not incorporate
-recombination. In addition, many of the recombinant sequences can be placed
-before their parents in time, because recombinants can appear to be more
-similar to ancestral sequences that are known with more uncertainty [OR because
-recombinants are placed anywhere on a tree that best fits].
-
-This ``parsimony improvement'' parameter ($p$)
-is equivalent to the sc2ts recombination penalty ($k$).
-
 RIPPLES~\cite{Turakhia2022} is a method for identifying
@@
 finds parsimonious placements for each segment independently.
+A recombination event at that node is then flagged if
+the overall parsimony of these multiple placements
+sufficiently improves over
+the original single-parent origin in the phylogeny.
+This ``parsimony improvement'' parameter ($p$)
+is equivalent to the sc2ts recombination penalty ($k$).
 
@@
 of the UShER+RIPPLES results.
-
-that from UShER+RIPPLES,
-largely reflecting differences in model formulation and parameter choice.
-When applied \textit{post hoc} to an UShER phylogeny of
-$\sim$1.6 million genomes sampled up to May, 2021,
-RIPPLES reported 589 unique recombination events with 43,104 descendant
-samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
-Using the same time interval, sc2ts \(k=4\) identifies
-14 high-confidence recombination events with
-172 descendant samples among 501,235 samples \(0.034\%\),
-a 40- to 80-fold reduction.
-Much of this difference is explained by
-the stricter parsimony penalty used here:
-increasing the penalty from p=3 (as used in the RIPPLES analysis) to
-k=4 greatly reduces false positives
-while retaining nearly all verified recombinant lineages,
-including those missed by RIPPLES at equivalent thresholds.
-These indirect comparisons suggest that
-our estimate represents a conservative lower bound.
-In future studies, dissecting the methodological differences
-between sc2ts and UShER+RIPPLES may
-inspire new ideas to develop or improve ARG building methods
-for SARS-CoV-2 (and potentially other pathogens).
 

### 2025-11-05 latex/main.tex
Updates from Overleaf

@@
 important evolutionary events such as
-the major ``saltations'' that characterize many VOCs.
+``saltations'' that characterize major VOCs.
 These lineages are characterized by a high number of accumulated mutations,
@@
 To distinguish time travelers from true saltation lineages,
-we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added.
-If these samples form a sufficiently plausible group over successive days (under criteria
-described in the next paragraph) sc2ts adds the group to the ARG as a 
-``retrospective sample group'' using same mechanisms as standard
-daily sample clusters.
+we re-evaluate the evidence for each lineage with a high HMM cost
+as additional samples are added.
+If these samples form a sufficiently plausible group over successive days
+(under criteria described in the next paragraph)
+sc2ts adds the group to the ARG as a 
+``retrospective sample group'' using same mechanisms
+as standard daily sample clusters.
 
@@
 group are closely related. If we have at least 2 mutations 
-shared by all samples, there's at most 2 recurrent mutations
+shared by all samples, there are at most 2 recurrent mutations
 within the group and at most 5 mutations unique to each sample, 
@@
 Although this process is highly heuristic, it successfully 
-captured the majority of saltational lineages
+captured the majority of saltation lineages
 such as Alpha (Figure~\ref{fig:subgraph_alpha}), and the 

### 2025-11-05 latex/main.tex
Updates from Overleaf

@@
 a 40- to 80-fold reduction.
-The sc2ts ARG identifies more of the well-characteristed X-lineages
+The sc2ts ARG identifies more of the well-characterized X-lineages
 at a more stringent threshold, and therefore provides us with a 
 conservative lower bound
-on the prevalance of detectable recombination in SARS-CoV-2.
+on the frequency of detectable recombination events in SARS-CoV-2.
 

### 2025-11-05 latex/main.tex
Updates from Overleaf

@@
 insights into the relationships between previously identified
-recombinant lineages. [NOTE: mention overall estimate
-of recombination compared with UShER+RIPPLES?]
+recombinant lineages. 
 Together, the ARG, inference method and 
@@
 by running 3SEQ~\cite{Lam2018improved}
-on the putative recombinant sequence trios (STAR Methods)\sout{,
-and 484 of the 855 (57\%) were classified as recombinant.
-Validation status correlates strongly with our QC filter,
-with 95\% of the QC-passing recombinants
-and 29\% of the QC-failing recombinants
-being classified as recombinant.}
-.
+on the putative recombinant sequence trios (STAR Methods).
 Of the 855 recombination events identified by sc2ts, 3SEQ verified 57\% of them, rising to 95\% verification among the 354 recombinants that passed our QC filter. 

### 2025-11-05 latex/main.tex

@@
 
-% NOTE: currently 163 words
-
-Millions of SARS-CoV-2 genome sequences were collected in
-response to the COVID-19 pandemic, forming a dataset of unprecedented
-richness.
+% NOTE: currently 150 words
+
+Millions of SARS-CoV-2 genome sequences were collected 
+during the COVID-19 pandemic, forming a dataset of unprecedented richness.
 Estimated genealogies are fundamental to
-gleaning information from this ocean of data, 
+understanding this ocean of data, 
 and form the primary input to many downstream analyses.
@@
 the genealogy is a tree.
-Recombinant lineages have risen to global prevalence in SARS-CoV-2,
-and simple tree representations are therefore necessarily
+Recombinant lineages have risen to global prevalence 
+and simple tree representations are therefore 
 incomplete and potentially misleading.
 We present sc2ts, a method to infer reticulate genealogies 
-as an Ancestral Recombination Graph (ARG) for SARS-CoV-2
-that automatically detects and incorporates recombination.  
-We use sc2ts to infer an ARG of 2.48 million genomes,
-and show that it captures a broad range of 
-known evolutionary features and provides novel 
-insights into the relationships between previously identified
-recombinant lineages. 
-Together, the ARG, inference method and 
-suite of open-source analysis tools, 
-represent a powerful resource for future research.
+as an Ancestral Recombination Graph (ARG) in real time at pandemic scale.
+We infer an ARG for 2.48 million SARS-CoV-2 genomes,
+and show that this deeply validated resource clarifies
+the relationships between recombinant lineages,
+quantifies the rate of recombination over time
+and provides a conservative
+lower bound on detectable recombination.
+This ARG and rich supporting software ecosytem provide
+a strong platform for future research.
 

### 2025-11-05 latex/main.tex
Minor tweaks to results sections

@@
 enables pandemic-scale analysis on a laptop. 
-These resources
-will help us to learn from the vast troves of data collected during the
+These resources provide a strong platform for future research,
+helping us to learn from the vast troves of data collected during the
 pandemic, and to prepare for the next.
@@
 
-[FIXME Going in too fast here]
-Sc2ts reconstructs an ARG by incrementally adding
-daily batches of sequences (Figure~\ref{fig:methodology}).
-First, sc2ts infers likely ``copying paths'' connecting each sample
-in the batch to node(s) in the current ARG
+Sc2ts is a real-time inference method, updated incrementally 
+with daily batches of sequence data (Figure~\ref{fig:methodology}).
+For each batch, sc2ts first infers likely ``copying paths'' connecting each sample
+to node(s) in the current ARG
 using a simplified version of the Li and Stephens (LS) model~\cite{LiStephens2003}.
@@
 % the number of recombinants detected.
-(we use $k=4$ here; see STAR Methods and Discussion).
+(we use $k=4$ here; see STAR Methods).
 We used the highly efficient LS HMM implementation in tsinfer~\cite{Kelleher2019},
@@
 and their multi-base nature presents challenges
-to the Markovian structure of the LS HMM,
+to the LS HMM,
 we masked out all gap characters as missing data during primary inference.
 To approximate the effects of the most important deletions into the ARG,
-we remapped data using \textit{post hoc} parsimony at 75 sites
+we remapped data using post hoc parsimony at 75 sites
 involved in deletions with a frequency of
@@
 has been thoroughly characterized~\cite{Gutierrez2022} and
-may have misclassified by sc2ts due to limitations in
-handling deletions or matching against long branches (Document \ref{sec:dd_xb}).
+is likely misclassified by sc2ts because the required
+parent lineages were not present (Document \ref{sec:dd_xb}).
 
@@
 correspond to a recombination event in the sc2ts ARG can largely
-be explained by a combination the choice of sites 
+be explained by a combination of the choice of sites 
 used for inference, and the strict time ordering imposed

### 2025-11-05 latex/main.tex

--- 2025-11-05 latex/main.tex
@@
 on the putative recombinant sequences,
-summarized in Table~\ref{tab:recombinant_validation}.
+summarized in Table~\ref{tab:recombinant_validation}
+(see also Figures~\ref{fig:recombinant_evidence_cov_recomb},
+and \ref{fig:recombinant_evidence_rebar}).
 CovRecomb has a substantially lower validation rate
@@
 Rebar validation is intermediate, at 57\% among QC-passing recombinants.
-(See also Figures~\ref{fig:recombinant_evidence_cov_recomb},
-and \ref{fig:recombinant_evidence_rebar}.)
 % We don't talk about RIPPLES results yet as we haven't discussed it, a
@@
 among the most weakly supported, with 4 or 5 averted mutations.
-Figures~\ref{fig:recombinant_evidence}B and C shows the 
+Figures~\ref{fig:recombinant_evidence}B,C show the 
 distribution of recombination events as a function the 
@@
 Figures~\ref{fig:recombinant_evidence_cov_recomb} and
-~\ref{fig:recombinant_evidence_rebar} 
+\ref{fig:recombinant_evidence_rebar} 
 plot the same information as Figure~\ref{fig:recombinant_evidence} 

### 2025-11-05 latex/main.tex
Update intervals section

@@
 A basic assumption of methods to infer genealogies from viral 
-genetic sequence data is that recombination is negligible and 
+genetic data is that recombination is negligible and 
 the genealogy is a tree.
-Recombinant lineages have risen to global prevalence 
+However, recombinant lineages have risen to global prevalence 
 and simple tree representations are therefore 
@@
 % Accessed GISAID/EpiCov on 08 Oct., 2025
-At the peak in February 2022 about 80 thousand samples
+At the peak in February 2022 about 80,000 samples
 per day  were submitted to GISAID\cite{Shu2017},
@@
 The benefits of this simple model and the value of
-a comprehensive genealogy updated in real-time~\cite{McBroome2021,Kramer2023}
+a comprehensive genealogy updated in real time~\cite{McBroome2021,Kramer2023}
 were conclusively demonstrated, as the resource
@@
 We show that the dates agree reasonably well with
-NextStrain estimates (Figure~\ref{fig:pango_vs_nextstrain_node_times})
+Nextstrain estimates (Figure~\ref{fig:pango_vs_nextstrain_node_times})
 and overall accuracy is comparable to 
@@
 in the N-terminal domain of the S1 unit of Spike
-(which is a deletion hot spot~\cite{McCarthy2021}),
+(which is a deletion hotspot~\cite{McCarthy2021}),
 for example,
@@
 ``averted'' by recombination---is a measure of
-the evidence in favor of recombination (similar to the  
+the evidence in favour of recombination (similar to the  
 ``parsimony improvement'' score in RIPPLES~\cite{Turakhia2022}).
@@
 A,B,C show summaries of the 354 sc2ts recombination nodes passing QC filters
-colored by 3SEQ classification (recombinant: purple, non-recombinant: gray).
+colored by 3SEQ classification (recombinant: purple, non-recombinant: grey).
 (A) Upper: jittered scatterplot comparing number of mutations
@@
 Lower: recombination nodes classified by number of mutations averted
-by allowing recombination (i.e. $x-y$).
+by allowing recombination (i.e., $x-y$).
 (B) Distribution of time to the MRCA of the parents of each QC-passing recombination node.
@@
 
-The location of a recombination breakpoint is generally
-impossible to pinpoint exactly, and instead an interval
-containing the breakpoint can be inferred by
-examining the parental and recombinant sequences.
-Some methods~\cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}
-and approaches (e.g., Pango designation) infer this
-as the genomic distance between lineage-associated
-mutations that differentiate the two parental \emph{lineages}.
-Similarly, methods based on probabilistic representations of
-mutational profiles \cite{Varabyou2021rapid}
-can only be as precise as examination of the two
-parent lineages allows.
-However, genealogical methods (such as sc2ts and RIPPLES) are
-based on identifying exact parental \emph{sequences},
-so are in principle maximally precise.
-The breakpoint interval of a recombinant is bounded by
+The precise genomic location of a recombination breakpoint is 
+in general impossible to determine, but we can place 
+bounds by inspecting sequences of the recombinant and its parents.
+The breakpoint interval of a recombinant is defined by
 two positions: a left position where the recombinant
@@
 parent sequence but not the left. 
-How well the time and position of a recombination event
-can be inferred thus depends on the divergence between the
-parent sequences (measured as time to their MRCA).
+Genealogical methods such as sc2ts and UShER+RIPPLES
+use inferred parental sequences and are therefore 
+in principle maximally precise.
+Other methods are limited in precision by the various proxies 
+used to represent distinct parental \emph{lineages}~\cite{%
+Varabyou2021rapid,Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}.
+% How well the time and position of a recombination event
+% can be inferred thus depends on the divergence between the
+% parent sequences (measured as time to their MRCA).
 
 The distribution of breakpoint intervals of
-the QC-passing recombination events
+the QC-passing recombination events in the sc2ts ARG
 shows an overall enrichment of breakpoints
@@
 is considerably larger (up to $\sim$54\% of the genome).
-As expected, we found a negative relationship between
+As expected from theory (STAR Methods), we found a negative relationship between
 the length of a breakpoint interval and
 divergence between recombinant parents
-(Figure~\ref{fig:recombinant_intervals}D),
-which follows theoretical expectation (STAR Methods).
+(Figure~\ref{fig:recombinant_intervals}D).
 Breakpoint intervals tend to be the smallest
@@
 the age of the youngest parent and
-the age of the oldest child of the recombination node
-(either given as collection dates or estimated using tsdate).
+the age of the oldest child of the recombination node.
 We found that for $\sim$75\% of the QC-passing recombination events,
@@
 relative to the global case numbers (\cite{Dong2020}; left y-axis).
-Cases are color-coded by their Scorpio labels,
+Cases are colour-coded by their Scorpio labels,
 with the key showing those groupings
@@
 To compare the performance of sc2ts and UShER+RIPPLES 
-in terms of recombinant detection, we ran RIPPLES on the Viridian UShER phylogeny
+in terms of recombinant detection we ran RIPPLES on the Viridian UShER phylogeny,
 subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
-Note that we are interested in the direct output of sc2ts and UShER+RIPPLES
-here, and do not perform QC filtering of either.
-Using the RIPPLES parameter $p=4$ (matching $k=4$ used in the sc2ts ARG) 
-resulted in 1174 reported recombination events.
-We ran 3SEQ on each of the candidate trios (STAR Methods) and 601 
-were classified as recombinant, giving an overall validation rate of 51\%.
+Note that we are concerned with the direct output of sc2ts and UShER+RIPPLES, so QC filtering was not performed for either.
+Setting the ``parsimony improvement'' parameter to $p=4$, equivalent to $k=4$ in sc2ts, 
+RIPPLES reported 1174 recombination events.
+Running 3SEQ on each of the candidate trios (STAR Methods) identified 601 of these as recombinant, giving a concordance rate of 51\%.
 Aligning the events reported by both methods using the sets of descendant
@@
 by UShER+RIPPLES associated with samples assigned to Pango X
-lineages. At the RIPPLES ``parsimony improvement'' parameter
-$p=4$ (equivalent to $k=4$ used to build the sc2ts ARG)
+lineages. 
 only 9 of the Pango X lineages are present and
@@
 tskit software ecosystem, a rich and efficient platform for 
-recombination-aware analysis widely used in population genomics.
+recombination-aware analyses widely used in population genomics.
 This combination of a deeply validated ARG with a pre-existing
@@
 The novelty of the inferred ARG motivates our detailed analyses of the methodology behind sc2ts, which we use to validate the ARG itself.
-This ensure that the ARG is a trustworthy basis for future research, deepens our understanding of the inference method, and suggests avenues for improvement (see Limitations of the Study).
+This ensures that the ARG is a trustworthy basis for future research, deepens our understanding of the inference method, and suggests avenues for improvement (see Limitations of the Study).
 
@@
 We have developed and used filters
-to detect time traveling samples in the Viridian dataset and
+to detect time-traveling samples in the Viridian dataset and
 to exclude them from the ARG,
@@
 both of which have breakpoints near the 5' end of genome.
-We therefore compare the Usher topology against the first tree in the ARG (covering
+We therefore compare the UShER topology against the first tree in the ARG (covering
 positions 1-26858; the majority of the genome). To reduce plot density, we consider
@@
 One set of results is based on mutational profiles calculated from a global GISAID dataset,
-and the other set from a global NextStrain dataset.
+and the other set from a global Nextstrain dataset.
 The breakpoint intervals identified using RecombinHunt are defined
@@
 \subsubsection*{Comparison with UShER+RIPPLES}
-We began with the Viridian UShER tree downloaded from FigShare.
+We began with the Viridian UShER tree downloaded from Figshare.
 To simplify the comparison with sc2ts, we subset this 
@@
 (alongside recombination and single nucleotide mutations),
-which to-date has been understudied in great part due to
+which to date has been understudied in great part due to
 a lack of suitable data models, methods, and tools.
@@
 have one or more parents that are \emph{younger} than the
-recombinant sequence based on metadata. Thus, the parental
-sequences were simply not available to match against for 
-sc2ts (which enforces strict time ordering) when the 
-recombinant sequence was inserted. Secondly, of the 179
+recombinant sequence based on metadata. As sc2ts enforces strict time ordering,
+the parental sequences were simply not available for it to match against
+when the recombinant sequence was inserted. Secondly, of the 179
 remaining events with matching clades and in which the 

### 2025-11-05 latex/main.tex
Updates from Overleaf

@@
 
-The precise genomic location of a recombination breakpoint is 
-in general impossible to determine, but we can place 
-bounds by inspecting sequences of the recombinant and its parents.
-The breakpoint interval of a recombinant is defined by
+The location of a recombination breakpoint is generally
+impossible to pinpoint exactly, and instead an interval
+containing the breakpoint can be inferred by
+examining the parental and recombinant sequences.
+Some methods~\cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}
+and approaches (e.g., Pango designation) infer this
+as the genomic distance between lineage-associated
+mutations that differentiate the two parental \emph{lineages}.
+Similarly, methods based on probabilistic representations of
+mutational profiles \cite{Varabyou2021rapid}
+can only be as precise as examination of the two
+parent lineages allows.
+However, genealogical methods (such as sc2ts and RIPPLES) are
+based on identifying exact parental \emph{sequences},
+so are in principle maximally precise.
+The breakpoint interval of a recombinant is bounded by
 two positions: a left position where the recombinant
@@
 parent sequence but not the left. 
-Genealogical methods such as sc2ts and UShER+RIPPLES
-use inferred parental sequences and are therefore 
-in principle maximally precise.
-Other methods are limited in precision by the various proxies 
-used to represent distinct parental \emph{lineages}~\cite{%
-Varabyou2021rapid,Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}.
-% How well the time and position of a recombination event
-% can be inferred thus depends on the divergence between the
-% parent sequences (measured as time to their MRCA).
+How well the time and position of a recombination event
+can be inferred thus depends on the divergence between the
+parent sequences (measured as time to their MRCA).
 
 The distribution of breakpoint intervals of
-the QC-passing recombination events in the sc2ts ARG
+the QC-passing recombination events
 shows an overall enrichment of breakpoints
@@
 is considerably larger (up to $\sim$54\% of the genome).
-As expected from theory (STAR Methods), we found a negative relationship between
+As expected, we found a negative relationship between
 the length of a breakpoint interval and
 divergence between recombinant parents
-(Figure~\ref{fig:recombinant_intervals}D).
+(Figure~\ref{fig:recombinant_intervals}D),
+which follows theoretical expectation (STAR Methods).
 Breakpoint intervals tend to be the smallest
@@
 the age of the youngest parent and
-the age of the oldest child of the recombination node.
+the age of the oldest child of the recombination node
+(either given as collection dates or estimated using tsdate).
 We found that for $\sim$75\% of the QC-passing recombination events,

### 2025-11-05 latex/main.tex
Updates from Overleaf

@@
 by UShER+RIPPLES associated with samples assigned to Pango X
-lineages. 
+lineages, showing that at $p=4$
 only 9 of the Pango X lineages are present and

### 2025-11-05 latex/main.tex
Updates from Overleaf

@@
 lower bound on detectable recombination.
-This ARG and rich supporting software ecosytem provide
+This ARG and rich supporting software ecosystem provide
 a strong platform for future research.
@@
 we infer an ARG with 2.48 million whole genome sequences.
-We show that this ARG is highly congruent with the UShER phylogeny built 
+We show that the non-recombinant portions of this ARG are highly congruent with the UShER phylogeny built 
 from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020},
-and also accurately captures evolutionary features such as mutational 
+and that the ARG accurately captures evolutionary features such as mutational 
 spectra~\cite{Bloom2023} and deletions\cite{Li2023indels}.

### 2025-11-05 latex/main.tex
Pass through disco, including limitations of study
Closes #917
Closes #725

@@
 methods in human 
-genomics~\cite{Tagami2024tstrait,Mejia2025using} [NOTE add more refs 
-if we like this].
+genomics~\cite{Mejia2025using,Grundler2025geographic,
+Lehmann2025on,Terhorst2025accelerated}.
 We hope that the ARG provided here will similarly
@@
 recombination into pathogen genomics.
+We envisage the ARG as a long-term stable resource---building on the 
+high-quality, freely available reassemblies of
+the Viridian dataset~\cite{Hunt2024}---to facilitate the study 
+of the pandemic phase of SARS-CoV-2 in an open and reproducible manner.
 
@@
 Muller2022bayesian}.
-By capturing recombination in real time at a global scale, our approach extends the established utility of comprehensive genealogies~\cite{Hinrichs2024,McBroome2024}, providing clear insight and enabling timely investigation of recombination-aware pathogen dynamics.
-The novelty of the inferred ARG motivates our detailed analyses of the methodology behind sc2ts, which we use to validate the ARG itself.
-This ensures that the ARG is a trustworthy basis for future research, deepens our understanding of the inference method, and suggests avenues for improvement (see Limitations of the Study).
-
-Study of the sc2ts ARG gives insight into the evolution of SARS-CoV-2, and this work therefore makes several additional contributions to the field. First, we clarify the relationships among
+By capturing recombination in real time at a global scale, 
+our approach extends the established utility of comprehensive 
+genealogies~\cite{Hinrichs2024,McBroome2024}, 
+providing clear insight and enabling timely investigation of 
+recombination-aware pathogen dynamics.
+The novelty of the inferred ARG motivates our detailed 
+analyses of the methodology behind sc2ts, which we use to validate the ARG itself. 
+This ensures that the ARG is a trustworthy basis for future research, 
+deepens our understanding of the inference method, and 
+suggests avenues for improvement (see Limitations of the Study).
+
+Study of the sc2ts ARG gives insight into the evolution of SARS-CoV-2, 
+and this work therefore makes several additional contributions to the field. 
+First, we clarify the relationships among
 recombinant Pango X-lineages, suggesting significant 
@@
 
-% SHZ: WIP
-We close by emphasizing several current limitations of sc2ts
-and suggest potential directions for improvement
-which we hope will inspire future development of ARG inference
-methods for pathogens beyond SARS-CoV2.
-
-There are several ways to extend the HMM.
-First, we currently treat multi-base deletions as missing data,
-adding indels to the ARG \textit{post hoc}.
-However, by explicitly modeling deletions as single HMM events, and
-accounting for fact that reversion of deletions is more likely to
-occur through recombination than back-mutation,
-we should be able to recover additional signals of recombination
-and reduce spurious recombinants.
-Second, the current HMM assumes a simplistic model of equiprobable
-mutation between bases with a uniform mutation rate along the genome.
-However, genomic analyses have revealed
-transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024} and
-sequence context-dependent mutation rate variation
-across the SARS-CoV-2 genome~\cite{Haddox2025}.
-Using a more realistic mutation model and
-accounting for mutation rate heterogeneity
-could improve inference of local genetic relationships in the ARG.
-Third, geography can be accounted for in the HMM,
-favoring attachment of samples to geographically close lineages.
-This should also improve precise identification of recombinant parents,
-because parents sampled in the same region or geographically close regions
-are expected to be more plausible
-than parents sampled in geographically distant regions.
-
-% TODO: Add a short discussion of the node dates and
-% some references to the literature, closing issue 725.
-
-We have developed and used filters
-to detect time-traveling samples in the Viridian dataset and
-to exclude them from the ARG,
-yielding a set of samples with reliable collection dates.
-Future studies could leverage the ARG
-to flag wrongly dated samples or
-to infer the dates of samples
-with missing or incomplete collection dates in the metadata.
-
-It is challenging to accurately reconstruct global events
-when there are regions with poor sampling and
-jurisdictions with long delays
-before sequences become publicly
-available~\cite{Lancet2021,Kalia2021,Chen2022}.
-For example, revisiting the ARG topology periodically
-(e.g., rewinding the ARG a month and repeating
-with samples that became available later)
-might improve the accuracy of the genealogy.
-
-We have developed and used several methods
-to exclude spurious recombination events \textit{post hoc},
-including filtering by supporting loci per parent and
-splitting of suspiciously long branches (e.g., Delta and Omicron BA.2).
-These methods could instead be run during ARG inference,
-immediately after sample matching,
-reducing the number of ARG post-processing steps.
-
-Finally, an inherent limitation of detecting recombination from sequence data,
-which applies to all methods,
-is that it is impossible to detect events with few (or no) supporting bases,
-as these cannot be distinguished from mutation events.
-This limitation makes it particularly challenging
-to detect events near the termini of linear viral genomes.
+We have focused on applying sc2ts to the Viridian dataset rather 
+than the larger GISAID resource~\cite{Shu2017} for two reasons.
+Firstly, the Viridian dataset is of higher quality and major efforts
+have been made to remove sources of error. Applying sc2ts to 
+GISAID data is straightforward, but significant additional efforts 
+would be required to deal with the various data quality issues.
+Secondly, the Viridian data is free to download without restrictions,
+enabling us to share the resulting ARG in the same manner.
+
+% Need to be careful here to list stuff that sounds interesting and 
+% hard enough to not get asked to do it.
+The sc2ts method may be extended and improved in several directions.
+Firstly, the simplistic parameterisation of the LS HMM used 
+here could be extended to include more realistic
+models with mutation rate variation along the genome~\cite{Haddox2025}
+and transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024}.
+Secondly, the current handling of deletions by post-hoc parsimony
+could be improved either by a more nuanced data-masking strategy
+(in which, e.g., all but the right-most base of known deletions is 
+masked as missing data during HMM matching), or by the more 
+challenging approach of incorporating multi-base events into the HMM directly.
+Sc2ts is currently specialised for SARS-CoV-2 data, but the methods
+may be readily adapted to other pathogens.
+
+We have used the tsdate method~\cite{Wohns2022,Pope2025} to date the 
+sc2ts ARG, and initial results are promising. However, dating 
+of phylogenies in itself is a complex task~\cite{Kumar2016advances}
+and methods for incorporating information from metadata of uncertain 
+quality into large trees are in their 
+early stages~\cite{sanderson2021chronumental} and there is certainly
+substantial room for improvement.
+Much of the complexity of sc2ts lies in 
+dealing with ``time traveller'' samples, where sampling dates 
+provided in metadata are substantially incorrect. 
+The present ARG may provide 
+a way of improving this, by iteratively matching samples against
+it to estimate dates for all sequences, and adjusting the dates
+for time-travellers detected in this way. Sc2ts could then
+be run without the complex filters required to remove 
+time-travellers and unconditionally include all but the most divergent
+sequences. Other studies may also benefit from the 
+improved sample dates.
+
+[Final para is prelim. Acknowledge limitations, forward looking 
+for how we might deal with them]
+Finally, it is important to acknowledge that sc2ts performs best
+with dense sampling, and the dataset is heavily geographically biased.
+[cite ~\cite{Lancet2021,Kalia2021,Chen2022}?]
+Long branches, cause particular problems. Future improvements would 
+ideally detect when data is sparse and switch to more sophisticated
+models, as discussed in~\cite{Bisschop2025likelihoods}.
+
+% It is challenging to accurately reconstruct global events
+% when there are regions with poor sampling and
+% jurisdictions with long delays
+% before sequences become publicly
+% available~\cite{Lancet2021,Kalia2021,Chen2022}.
+% For example, revisiting the ARG topology periodically
+% (e.g., rewinding the ARG a month and repeating
+% with samples that became available later)
+% might improve the accuracy of the genealogy.
+
+% We have developed and used several methods
+% to exclude spurious recombination events \textit{post hoc},
+% including filtering by supporting loci per parent and
+% splitting of suspiciously long branches (e.g., Delta and Omicron BA.2).
+% These methods could instead be run during ARG inference,
+% immediately after sample matching,
+% reducing the number of ARG post-processing steps.
+
+% Finally, an inherent limitation of detecting recombination from sequence data,
+% which applies to all methods,
+% is that it is impossible to detect events with few (or no) supporting bases,
+% as these cannot be distinguished from mutation events.
+% This limitation makes it particularly challenging
+% to detect events near the termini of linear viral genomes.
 

### 2025-11-06 latex/main.tex
Updates from Overleaf

@@
 date has been performed using the combination of UShER and 
-RIPPLES~\cite{Turakhia2022}. RIPPLES works by taking a
+RIPPLES~\cite{Turakhia2022}. 
+% NOTE: not citing RIPPLES here as there is no real analysis
+% in the paper itself, it's just focused on the tool
+RIPPLES works by taking a
 pre-existing UShER phylogeny and systematically breaking up
@@
 Study of the sc2ts ARG gives insight into the evolution of SARS-CoV-2, 
-and this work therefore makes several additional contributions to the field. 
+which provides several further contributions to our understanding of the COVID-19 pandemic. 
 First, we clarify the relationships among
@@
 events over time being the product of global case numbers
-and lineage heterozygosity.
+and lineage diversity.
 Finally, and perhaps most fundamentally, through our 
@@
 would be required to deal with the various data quality issues.
-Secondly, the Viridian data is free to download without restrictions,
-enabling us to share the resulting ARG in the same manner.
+Secondly, the Viridian data is publicly available and redistributable,
+enabling us to share the resulting ARG without restrictions.
 
@@
 in terms of the number of mutations.
-We therefore masked our 100 mutation-rich sites
+We therefore masked our 100 most mutation-rich sites
 to remove sequencing errors and
@@
 
-\subsubsection*{Li and Stephens haplotype-copying model}
+\subsubsection*{Li and Stephens model}
 
@@
 Pango representative samples.
-We focus on the 1473 Pango lineages that have perfect ``origination events'' 
-(Document \ref{sec:supplement_pango_lineages}). We then omit
-omit X lineages, leaving 1329 lineages.
+We focus on 1286 Pango lineages comprising those which are defined by perfect ``origination events'' 
+(Document \ref{sec:supplement_pango_lineages}). but which are not acknowledged recombinant lineages (i.e. Pango X lineages and their descendants).
 To ensure that each sample is a tip in the simplified topologies,
@@
 We then further remove 4 lineages
-which are singleton descendants of a recombination event in the ARG,
-% NOTE: would have been simpler to say we just got rid of XBB and 
-% descendants.
-as well as 43 samples that are descendants of the recombinant XBB
-but which do not start with an ``X''.
+which are singleton descendants of a putative ARG recombination event,
 This results in a simplified ARG of 1282 samples,
@@
 we took the results of the event which had the highest number of epidemic recombinants.
+% Note from Shing: 
+% This is not clearly explained in the paper at all. I went to their code to see what it means, but their code is not easy to read. 
 For other Pango X lineages, we downloaded the full list of putative recombinants

### 2025-11-06 latex/main.tex
Updates from Overleaf

@@
 phylogenetic signal. 
-We therefore performed post-hoc parsinony mapping at 163 sites to include 
+We therefore performed post-hoc parsimony mapping at 163 sites to include 
 the effects of deletions and to ensure that all Pango-informative

### 2025-11-06 latex/main.tex
Comment out unnecessary Cell stuff

@@
 
-\section*{RESOURCE AVAILABILITY}
-
-%%%  The resource availability section is required
-%%%  for all research articles. This component
-%%%  has 3 subsections: "lead contact," "materials
-%%%  availability," and "data and code availability."
-%%%  All 3 subsections must be included, even if no
-%%%  unique materials were generated in the study.
-%%%  Do not edit or change the names of the subheadings.
-%%%  No other subheadings or text are allowed in the
-%%%  resource availability section.
-
-\subsection*{Lead contact}
-
-%%%  Authors are required to designate a lead contact,
-%%%  who will be responsible for communication with
-%%%  the journal before and after publication and is
-%%%  the arbiter of disputes, including concerns
-%%%  related to reagents or resource sharing. Only
-%%%  one author can be named the lead contact, and
-%%%  only the lead contact’s information may be
-%%%  provided in this section.
-
-Requests for further information and resources
-should be directed to and will be fulfilled
-by the lead contact, XX (XX@university.edu).
-
-\subsection*{Materials availability}
-
-%%%  This subsection must include a statement describing
-%%%  the availability of newly generated materials
-%%%  associated with the paper, including any conditions
-%%%  for access. If there are no newly generated materials
-%%%  associated with the paper, the statement should
-%%%  state this, e.g.: This study did not generate new
-%%%  materials.
-
-This study did not generate new materials.
+% \section*{RESOURCE AVAILABILITY}
+
+% %%%  The resource availability section is required
+% %%%  for all research articles. This component
+% %%%  has 3 subsections: "lead contact," "materials
+% %%%  availability," and "data and code availability."
+% %%%  All 3 subsections must be included, even if no
+% %%%  unique materials were generated in the study.
+% %%%  Do not edit or change the names of the subheadings.
+% %%%  No other subheadings or text are allowed in the
+% %%%  resource availability section.
+
+% \subsection*{Lead contact}
+
+% %%%  Authors are required to designate a lead contact,
+% %%%  who will be responsible for communication with
+% %%%  the journal before and after publication and is
+% %%%  the arbiter of disputes, including concerns
+% %%%  related to reagents or resource sharing. Only
+% %%%  one author can be named the lead contact, and
+% %%%  only the lead contact’s information may be
+% %%%  provided in this section.
+
+% Requests for further information and resources
+% should be directed to and will be fulfilled
+% by the lead contact, XX (XX@university.edu).
+
+% \subsection*{Materials availability}
+
+% %%%  This subsection must include a statement describing
+% %%%  the availability of newly generated materials
+% %%%  associated with the paper, including any conditions
+% %%%  for access. If there are no newly generated materials
+% %%%  associated with the paper, the statement should
+% %%%  state this, e.g.: This study did not generate new
+% %%%  materials.
+
+% This study did not generate new materials.
 
@@
 
-\section*{AUTHOR CONTRIBUTIONS}
-
-%%%  This component is required for most research papers.
-%%%  Mention each individual author with a statement
-%%%  outlining the contribution of each author to the work.
-
-Conceptualization, S.C.P. and S.Y.W.;
-methodology, A.B., S.C.P., and S.Y.W.;
-investigation, M.E., A.N.V., N.A.V., S.C.P., and S.Y.W.;
-writing-–original draft, S.C.P. and S.Y.W.;
-writing-–review \& editing, S.C.P. and S.Y.W.;
-funding acquisition, S.C.P. and S.Y.W.;
-resources, M.E.V and C.K.B.;
-supervision, A.B., N.L.W., and A.A.D.
-
-\section*{DECLARATION OF INTERESTS}
-
-%%%  This component is required for all articles, even
-%%%  if the authors have no competing interests; if
-%%%  this is the case, insert "The authors declare no
-%%%  competing interests." Please refer to the
-%%%  declaration of interests policy:
-%%%  https://www.cell.com/declaration-of-interests
-
-% TODO: Close https://github.com/jeromekelleher/sc2ts-paper/issues/922
-The authors declare no competing interests.
-
-\section*{DECLARATION OF GENERATIVE AI AND AI-ASSISTED TECHNOLOGIES}
-
-%%%  If generative AI and AI-assisted technologies
-%%%  were used in the writing process, this must
-%%%  be disclosed in the paper. This declaration
-%%%  does not apply to the use of basic tools for
-%%%  checking grammar, spelling, references, etc.
-%%%  If you have nothing to disclose, please do not
-%%%  include this component.
-
-During the preparation of this work, the author(s) used ChatGPT
-in order to obtain editorial input on the text.
-After using this tool or service, the author(s) reviewed
-and edited the content as needed and take(s) full responsibility for the content of the publication.
+% \section*{AUTHOR CONTRIBUTIONS}
+
+% %%%  This component is required for most research papers.
+% %%%  Mention each individual author with a statement
+% %%%  outlining the contribution of each author to the work.
+
+% Conceptualization, S.C.P. and S.Y.W.;
+% methodology, A.B., S.C.P., and S.Y.W.;
+% investigation, M.E., A.N.V., N.A.V., S.C.P., and S.Y.W.;
+% writing-–original draft, S.C.P. and S.Y.W.;
+% writing-–review \& editing, S.C.P. and S.Y.W.;
+% funding acquisition, S.C.P. and S.Y.W.;
+% resources, M.E.V and C.K.B.;
+% supervision, A.B., N.L.W., and A.A.D.
+
+% \section*{DECLARATION OF INTERESTS}
+
+% %%%  This component is required for all articles, even
+% %%%  if the authors have no competing interests; if
+% %%%  this is the case, insert "The authors declare no
+% %%%  competing interests." Please refer to the
+% %%%  declaration of interests policy:
+% %%%  https://www.cell.com/declaration-of-interests
+
+% % TODO: Close https://github.com/jeromekelleher/sc2ts-paper/issues/922
+% The authors declare no competing interests.
+
+% \section*{DECLARATION OF GENERATIVE AI AND AI-ASSISTED TECHNOLOGIES}
+
+% %%%  If generative AI and AI-assisted technologies
+% %%%  were used in the writing process, this must
+% %%%  be disclosed in the paper. This declaration
+% %%%  does not apply to the use of basic tools for
+% %%%  checking grammar, spelling, references, etc.
+% %%%  If you have nothing to disclose, please do not
+% %%%  include this component.
+
+% During the preparation of this work, the author(s) used ChatGPT
+% in order to obtain editorial input on the text.
+% After using this tool or service, the author(s) reviewed
+% and edited the content as needed and take(s) full responsibility for the content of the publication.
 

### 2025-11-06 latex/main.tex

@@
 
-A Viterbi algorithm can be used
-to find the most likely ``copying path'' of a focal sample
-through a reference panel of samples under the LS model.
-Specifically, sc2ts uses a Viterbi algorithm modified to operate on STS,
-which was originally developed to build genome-wide human genealogies and
-is capable of scaling to biobank-scale datasets \cite{Kelleher2019}.
-Using this Viterbi algorithm,
-sc2ts deterministically infers the copying path of a sample
-collected on a certain date through the samples
-collected on earlier dates in a given ARG.
-
-The SARS-CoV-2 sequences were determined
-from samples collected densely in time.
-SARS-CoV-2 phylogenies have low phylogenetic signals
-due to high sampling frequency
-(such that sampling and transmission outpace nucleotide substitution).
-Therefore, most samples can be explained by a small number of mutations.
-Leveraging this property of the dataset,
-we developed a likelihood-thresholding strategy
-to avoid needing to compute the likelihood of
-all possible copying paths through an ARG for every single sample.
-%%% TODO: This needs to be explained.
-
-[TODO explain Fig S1 and S2]
-
-Furthermore, to avoid performing redundant HMM matching on identical samples,
-new samples which were exactly matched to a (sample or non-sample) node in the ARG
-were not added immediately.
-Instead, the copying paths of these samples were recorded,
-and the samples were eventually added back to the ARG in a post-processing step.
+We use the Viterbi algorithm implemented in tsinfer~\cite{Kelleher2019}
+to efficiently find copying paths under this model. We updated this 
+implementation to include a ``likelihood threshold'', which allows 
+all possible paths through the HMM with likelihood less than a given
+value to be considered equally probable. This approach leads to major
+gains in likelihood compression, saving memory and reducing computation
+time. We then use this likelihood thresholding approach to iteratively
+process the samples for a given daily batch, first finding all exact 
+matches, then all with 1 mismatch, etc. By setting the appropriate
+likelihood threshold for the HMM, we are guaranteed that any HMM
+solutions found at this threshold are correct, and can then run the HMM again
+with a lower threshold for the remaining samples. Close matches
+are found very quickly using this approach, and it is only
+samples that are more distant from the current ARG that require
+significant processing time and memory. This is reflected in 
+the computaional resources reported in 
+Figures~\ref{fig:inference_resources},\ref{fig:samples_per_day}.
+
+% A Viterbi algorithm can be used
+% to find the most likely ``copying path'' of a focal sample
+% through a reference panel of samples under the LS model.
+% Specifically, sc2ts uses a Viterbi algorithm modified to operate on STS,
+% which was originally developed to build genome-wide human genealogies and
+% is capable of scaling to biobank-scale datasets \cite{Kelleher2019}.
+% Using this Viterbi algorithm,
+% sc2ts deterministically infers the copying path of a sample
+% collected on a certain date through the samples
+% collected on earlier dates in a given ARG.
+
+% The SARS-CoV-2 sequences were determined
+% from samples collected densely in time.
+% SARS-CoV-2 phylogenies have low phylogenetic signals
+% due to high sampling frequency
+% (such that sampling and transmission outpace nucleotide substitution).
+% Therefore, most samples can be explained by a small number of mutations.
+% Leveraging this property of the dataset,
+% we developed a likelihood-thresholding strategy
+% to avoid needing to compute the likelihood of
+% all possible copying paths through an ARG for every single sample.
+% %%% TODO: This needs to be explained.
+
+% [TODO explain Fig S1 and S2]
+
+% Furthermore, to avoid performing redundant HMM matching on identical samples,
+% new samples which were exactly matched to a (sample or non-sample) node in the ARG
+% were not added immediately.
+% Instead, the copying paths of these samples were recorded,
+% and the samples were eventually added back to the ARG in a post-processing step.
 

### 2025-11-06 latex/main.tex
Updates from Overleaf

@@
 (some events occurred at the same breakpoints).
-% NOTE: if we keep this we should redo on Yan's laptop and quote those numbers
 Using the tskit Python API,
-% TODO: Concrete numbers?
-the ARG can be loaded into memory in less than a second,
+% NOTE: numbers from Yan's laptop are:
+% ts = tszip.load(): user 484 ms, sys: 90.4 ms, total: 575 ms
+% sc2ts.node_data(ts, inheritance_stats=False): user 396 ms, sys: 42.8 ms, total: 439 ms
+the ARG can be loaded into memory in half a second,
 and the vectorized representation integrates tightly
@@
 including sample IDs and Pango assignments (stored as metadata in the ARG),
-can be derived as a Pandas data frame~\cite{Mckinney2010data} in 0.6 second.
+can be derived as a Pandas data frame~\cite{Mckinney2010data} in less than half a second.
 The ARG can therefore be easily accessed using familiar tools
@@
 (Document~\ref{sec:supplement_sc2ts_usher_imputation}).
-We compared the mutational spectra of major VOCs in the ARG
+We compared the mutational spectra of major Variants of Concern (VOCs) in the ARG
 against the UShER tree and a previous study~\cite{Bloom2023} (STAR Methods),
@@
 but are not incorporated into UShER trees and
-therefore must be accounted for separately.
+must be accounted for separately.
 Because deletions are difficult to distinguish
@@
 Figures~\ref{fig:recombinant_evidence}B,C show the 
-distribution of recombination events as a function the 
-divergence between the parents.
+distribution of recombination events as a function of the 
+divergence between their two parents.
 Figures~\ref{fig:recombinant_evidence_cov_recomb} and
@@
 % % might affect our interpretation of these findings?
-% [FIXME: This is direct output from ChatGPT.  Review!]
+% [TODO: This is direct output from ChatGPT.  Review!]
 % The dated ARG allows us to place recombination within its epidemiological
@@
 % % Maybe mention epistasis?
-% [FIXME: This is directish output from ChatGPT. Review!]
+% [TODO: This is directish output from ChatGPT. Review!]
 % The genomic distribution of breakpoints in the ARG is highly uneven, with a
@@
 
-[TODO fix the typesetting once all resources are compiled]
+[FIXME: fix the typesetting once all resources are compiled]
 
@@
 
-TODO add these to the list above with actual URLs
+FIXME: add these to the list above with actual URLs
 \begin{itemize}
@@
 Zarr format, resulting in a single 401 MiB file 
-(around 77X smaller than gzip compressed FASTA\cite{Czech2025}).
+(around 77$\times$ smaller than gzip compressed FASTA\cite{Czech2025}).
 
@@
 
-[TODO revise this paragraph to give some background info, and reference
+[FIXME: revise this paragraph to give some background info, and reference
 the Viridian work]
@@
 from Table S1 of Li et al.\cite{Li2023indels},
-we chose 75 sites to remap.
+we chose 75 sites for deletion-remapping (see Document \ref{sec:supplement_dels} for analysis of the major deletions identified from this process).
 In addition, we remapped data at 88 sites that were excluded
@@
 in \textit{Mathematica} 14.2 (Wolfram 2024) and
-available at GITHUB LINK.
+available at https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/TemporalTrends_RecEvent25August2025.nb.
 
@@
 % TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/684
-with XX mutations over the YYY sites shared between them.
+with [FIXME] XX mutations over the YYY sites shared between them.
 After site remapping, and including all mutations involving
@@
 As sc2ts does not use deletions for inference,
-deletion remapping (section XXXX) has reverted the deletion in this case,
+deletion remapping (STAR Methods) has reverted the deletion in this case,
 creating many insertions on the branch above the sample (magenta).
@@
 and 
-82 BA.2 samples share a recombination event [With what breakpoint and parents?].
+82 BA.2 samples share a recombination event [FIXME: With what breakpoint and parents?].
 We also ran RIPPLES with $p=4$, which reduced the overall number of
@@
 
-[TODO Discuss  the large effect recombinants inferred by both]
+[FIXME: Discuss  the large effect recombinants inferred by both]
 
@@
 rewire it post-hoc, owning to subsequent tree building) and the BA.2 event (doesn't
-pass QC.) - note this is discussed in more detail in document \ref{sec:spurious-ancestrally-important-recombinants}]
+pass QC.) - note this is discussed in more detail in document \ref{sec:spurious-ancestrally-important-recombinants} ]
 
@@
 
-[NOTE: this is an earlier draft of main text which explains the table in
+[FIXME: this is an earlier draft of main text which explains the table in
 detail]

### 2025-11-06 latex/main.tex

@@
 These statistical tests and graphics were conducted
-% TODO: The link!
 in \textit{Mathematica} 14.2 (Wolfram 2024) and
-available at https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/TemporalTrends_RecEvent25August2025.nb.
+available at
+\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/TemporalTrends\_RecEvent25August2025.nb}.
 

### 2025-11-06 latex/main.tex
Updates from Overleaf

@@
 
-To compare the performance of sc2ts and UShER+RIPPLES 
-in terms of recombinant detection we ran RIPPLES on the Viridian UShER phylogeny,
-subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
-Note that we are concerned with the direct output of sc2ts and UShER+RIPPLES, so QC filtering was not performed for either.
-Setting the ``parsimony improvement'' parameter to $p=4$, equivalent to $k=4$ in sc2ts, 
-RIPPLES reported 1174 recombination events.
-Running 3SEQ on each of the candidate trios (STAR Methods) identified 601 of these as recombinant, giving a concordance rate of 51\%.
+We compared the performance of sc2ts recombinant detection against RIPPLES run on the Viridian UShER phylogeny,
+subset to the 2,475,418 samples shared with the ARG (STAR Methods).
+To enable exact comparison, we omitted the QC filtering steps from both
+methods, and initially set the RIPPLES ``parsimony improvement'' parameter to $p=4$ (rather than the more lenient default of 3), to match the equivalent $k=4$ setting
+used in sc2ts (STAR Methods). Under this setting,
+RIPPLES reported 1174 recombination events of which 601 were designated as recombinant when running 3SEQ on the candidate trios (STAR Methods), giving a concordance rate of 51\%.
+This is notably lower than the equivalent concordance rate of 57\% among the 855 recombination events reported by sc2ts.
 Aligning the events reported by both methods using the sets of descendant
@@
 
-% \section*{RESOURCE AVAILABILITY}
+\section*{RESOURCE AVAILABILITY}
 
@@
 \begin{itemize}
-\item Snakemake workflow for aligning and converting the Viridian dataset to
-VCF Zarr:
-\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/viridian\_dataset/}
-\item Complete configuration for running sc2ts base ARG inference:
-\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}
-\item Snakemake workflow for ARG post-processing
-\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/Snakefile}
-\end{itemize}
-
-FIXME: add these to the list above with actual URLs
-\begin{itemize}
-    \item The ARG and the sequence alignments have been deposited at Zenodo and are publicly available.
-    \item All original code has been deposited in GitHub and is publicly available.
-    \item Jupyter notebooks to perform the analyses in this paper are available on GitHub.
+\item The sequences alignments in VCF Zarr format have been deposited at Zenodo and are publicly available at \url{https://zenodo.org/records/16314739}.
+\item The Snakemake workflow for aligning and converting the Viridian v05 dataset to VCF Zarr is publicly available at \url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/viridian\_dataset/}.
+\item The ARG in tskit format has been deposited at Zenodo and is publicly available at \url{https://zenodo.org/records/16313750}.
+\item The complete configuration file for running sc2ts to build the ARG above is publicly available at \url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}.
+\item The Snakemake workflow for ARG post-processing is publicly available at \url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/Snakefile}.
+\item The original source code of sc2ts is publicly available on GitHub at \url{https://github.com/jeromekelleher/sc2ts/}.
+\item The Jupyter notebooks and the Mathematica script to perform the analyses in this paper are publicly available on GitHub at \url{https://github.com/jeromekelleher/sc2ts-paper/}.
 \end{itemize}
@@
 
+[FIXME: add all acknowledgments]
+
 SHZ is supported by an NDPH Intermediate Research Fellowship
 (Oxford Population Health, University of Oxford).
+JK and YW acknowledge support from R01 HG012473 from the National Institutes of Health NHGRI, EPSRC (research grant EP/X024881/1) and the Robertson Foundation.
 
@@
 % complicating the story.
-For Delta, we used two seeds: one for B.1.617.1 and 1 samples assigned to
-B.1.617.2, [with more details and some references to back up our assertions.
+For Delta, we used two seeds: one for B.1.617.1 and 1 sample assigned to
+B.1.617.2, [FIXME more Delta details and some references to back up our assertions.
 There's some specific info in the Viridian paper on this, I'm sure. 
@@
 
-[FIXME: revise this paragraph to give some background info, and reference
+[FIXME: revise this paragraph to give some background info on Omicron, and reference
 the Viridian work]
@@
 
-At $p=3$, this resulted in a total of 4111 recombination events.
+Using the default ``parsimony improvement'' parameter of $p=3$, UShER+RIPPLES detected a total of 4111 recombination events.
 Most of these subtend $<10$ samples (2616 singletons),
-with 4 events subtending more than 100,000 samples.
+although four events subtended more than 100,000 samples.
 Analysis of the descending samples for each event shows that

### 2025-11-06 latex/main.tex

@@
 \section*{DISCUSSION}
-There are two major contributions of this work. The first is an
+There are two primary contributions of this work. The first is an
 Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
 Through multiple layers of validation against results from the 
-literature, we have demonstrated that this ARG successfully synthesises
+literature, we have demonstrated that this ARG synthesises
 the evolutionary features of genetic inheritance, point mutations,
@@
 recombination-aware analyses widely used in population genomics.
-This combination of a deeply validated ARG with a pre-existing
-and highly capable software platform forms a unique resource.
+This combination of a deeply validated pandemic-scale ARG with 
+a mature and highly capable software platform forms a unique resource.
 Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
-have helped spur the development of numerous ARG-aware 
+have helped spur the development of ARG-aware 
 methods in human 
@@
 enable a range of recombination-aware analyses of SARS-CoV-2, and 
-catalyse a wave of methodological improvements which fully integrate
+catalyse a wave of methodological developments integrating
 recombination into pathogen genomics.
@@
 of the pandemic phase of SARS-CoV-2 in an open and reproducible manner.
-
-The second major contribution is the method used to 
+% NOTE: still missing the "What" bit suggested by Jotun, where we 
+% list out some of the stuff that could be done. I'm not convinced
+% this works, though.
+
+The other primary contribution of this work is the method used to 
 infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
@@
 Muller2022bayesian}.
-By capturing recombination in real time at a global scale, 
-our approach extends the established utility of comprehensive 
-genealogies~\cite{Hinrichs2024,McBroome2024}, 
-providing clear insight and enabling timely investigation of 
-recombination-aware pathogen dynamics.
-The novelty of the inferred ARG motivates our detailed 
-analyses of the methodology behind sc2ts, which we use to validate the ARG itself. 
-This ensures that the ARG is a trustworthy basis for future research, 
-deepens our understanding of the inference method, and 
-suggests avenues for improvement (see Limitations of the Study).
-
-Study of the sc2ts ARG gives insight into the evolution of SARS-CoV-2, 
-which provides several further contributions to our understanding of the COVID-19 pandemic. 
+To ensure these innovations are well-founded, we have paid significant 
+attention to methodological detail, carefully validating the ARG
+(and consequently the method) against known results. 
+Sc2ts is currently specialised to take advantage of unique features 
+of the SARS-CoV-2 dataset, but the methods may be readily adapted
+to other pathogens. The LS HMM is the most important element of 
+sc2ts, efficiently finding parsimonious placements for samples 
+among all possible recombinant solutions. The current 
+parameterisation---emphasising interpretability---is simplistic,
+and could readily be extended to account
+for mutation rate variation along the genome~\cite{Haddox2025}
+or transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024}.
+Our analysis of deletions shows that valuable phylogenetic signal
+is lost when they are ignored, and an important avenue for 
+future work is to explore ways of incorporating them into the HMM.
+Another methodological innovation is the use of 
+the tsdate method~\cite{Wohns2022,Pope2025} to estimate internal node dates
+in the ARG. While results are encouraging, 
+dating of phylogenies is a complex task~\cite{Kumar2016advances}
+and methods for incorporating information from metadata of uncertain 
+quality into large genealogies are in their 
+early stages~\cite{sanderson2021chronumental}, and there is certainly
+substantial room for improvement.
+Sc2ts itself may benefit from these node dates, as 
+much of the method's complexity lies in dealing with 
+sample with incorrect sampling dates, and the ARG may provide 
+a means of pre-identifying such ``time travellers''.
+
+The secondary contributions of this work arise from 
+the detailed study of the sc2ts ARG, providing insights into the evolution of SARS-CoV-2
+and the COVID-19 pandemic. 
 First, we clarify the relationships among
@@
 clustering of named lineages under shared recombination events.
-Second, we use the ARG to study breakpoint intervals 
-for recombination events, add a novel temporal dimension
-and show a simple relationship with the 
+Second, we used the ARG to study breakpoint intervals 
+for recombination events, adding a novel temporal dimension
+and demonstrating a simple relationship with the 
 size of the interval in base pairs derived from theory.
@@
 the pandemic phase of SARS-CoV-2.
-Together, these  underscore the utility of the ARG itself, and suggest
-several important directions for future research.
-
-% Old discussions(s)
-
-% % P1. The ARG. Unique resource. Emphasise utility of UShER and how 
-% % this has more.
-% % SHZ: This makes it sound like we are just making analyses
-% % (which people can do already using existing methods and tools)
-% % easier to do
-% % rather than making a leap forward from anything else out there.
-% We present the first pandemic-scale ancestral recombination graph (ARG) for
-% SARS-CoV-2, reconstructed from 2.48 million genomes using \texttt{sc2ts}. This
-% ARG represents a step change in how the evolutionary history of this virus can
-% be represented and analysed. Like the UShER phylogeny, it provides a
-% comprehensive and continuously updatable record of global genomic diversity,
-% but it extends this paradigm by embedding recombination directly into the
-% genealogy. The ARG unifies point mutations, deletions, and recombination events
-% within a single, compact data structure, enabling analyses that
-% were previously fragmented across multiple methods and data models. By encoding
-% reticulate ancestry explicitly rather than assuming a single tree, the ARG
-% provides a complete and internally consistent representation of SARS-CoV-2
-% evolution. In this way, the
-% ARG offers the same accessibility and scalability that made the UShER tree
-% indispensable during the pandemic, while expanding its scope to encompass the
-% full spectrum of evolutionary processes shaping SARS-CoV-2.
-
-% % P2. This is a conservative lower-bound on recombination. 
-% % Discuss other estimates.
-% % 
-% % SHZ: Why should we care about getting this number/frequency right?
-% % Sure, popgen folks love estimating this, but to other people,
-% % existing estimates already tell us that
-% % detectable recombination in SARS-CoV-2 is (qualitatively) rare,
-% % even without the advances of sc2ts.
-% % Also, we are talking about the frequency of detectable recombination
-% % in the last sentence, right?
-% % The true frequencey of recombination is unknowable from this type of data.
-% % So, the best we are ever going to get is a better lower bound.
-% The recombination frequency estimated from the \texttt{sc2ts} ARG is
-% substantially lower than that implied by the combination of UShER+RIPPLES,
-% largely reflecting differences in model formulation and parameter choice.
-% RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
-% reported 589 unique recombination events with 43,104 descendant
-% samples---corresponding to roughly 2.7\% of all genomes being of recombinant
-% origin.  Using the same time interval, our inference at \(k=4\) identifies only
-% 14 high-confidence recombination events with 172 descendants among 501,235
-% genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
-% explained by the stricter parsimony penalty adopted here: increasing the
-% penalty from \(p=3\) (as used in the RIPPLES analysis) 
-% to \(k=4\) greatly reduces false
-% positives while retaining nearly all verified recombinant lineages, 
-% including those missed by RIPPLES at equivalent thresholds.  
-% We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
-% of recombination as
-% tree rearrangements proceed, thereby requiring the more lenient 
-% threshold in order to find known recombinats.
-% These comparisons suggest that our estimate represents a conservative 
-% lower bound. Further work will be required to determine
-% the true frequency of recombination in SARS-CoV-2 evolution.
-
-% % P3. Clarifies relationships between pango X lineages, simplifying
-% % the Omicron sublineage.
-% % 
-% % SHZ: I don't think knowing the relationships between Pango Xs is what's really important.
-% % Sure, phylogeneticists would get a kick out of it,
-% % but I don't think the majority of Cell readership would.
-% % I think what is important is knowing
-% % how many emerging recombinants carrying new combinations of mutations of concern
-% % we are actually dealing with.
-% % Clarifying the relationships between Pango Xs helps with this task.
-
-% The Pango X lineage designations represent the community’s best attempt to
-% catalogue recombinant SARS-CoV-2 lineages, but they arise from a largely
-% manual, volunteer-driven process that synthesizes multiple lines of evidence
-% and expert judgment rather than a single automated criterion.  As a result, the
-% system provides the closest available approximation to “truth data,” but
-% inevitably contains inconsistencies and over-splitting.  Within the
-% \texttt{sc2ts} ARG, 44 Pango X lineages are represented, of which 16 correspond
-% directly to recombination nodes (Type I events).  Examination of their
-% genealogical relationships shows that many of these 44 lineages share common
-% ancestral recombination events rather than arising independently.  In
-% particular, five lineages—XAA, XAG, XAM, XR, and XU—descend from the same
-% recombinant ancestor associated with XQ, while another five—XZ, XAC, XAD, XAE,
-% and XAP—share a second common ancestor that we refer to as “Xx.”  These
-% clusters account for roughly half of all X designations in the dataset,
-% indicating that the community process has at times produced multiple labels for
-% the descendants of a single recombinant lineage.  Conversely, several [TODO 
-% be more concrete]
-% designated X lineages show little or no evidence of recombination in the ARG,
-% suggesting that some may reflect convergent mutation or limited sampling rather
-% than hybrid origin.  Together, these results provide a data-driven synthesis in
-% which the 44 currently recognized recombinant lineages collapse into a smaller
-% number of well-supported ancestral events, simplifying the complex picture of
-% Omicron-era evolution and bringing the community nomenclature into closer
-% alignment with the underlying genealogy.
-
-% % P4. Results on global case numbers and frequency of recomb.
-% % SHZ: Should we mention how the reliability of global case number data
-% % and variation in sampling effort over time and across countries
-% % might affect our interpretation of these findings?
-% [TODO: This is direct output from ChatGPT.  Review!]
-% The dated ARG allows us to place recombination within its epidemiological
-% context, revealing clear temporal structure in when recombinant lineages arise.
-% The frequency of inferred recombination events follows global case numbers
-% through the pandemic, with peaks during the Delta and Omicron waves when both
-% infection prevalence and genetic diversity were high.  Quantitatively, the best
-% single predictor of weekly recombinant counts is the product of global cases
-% and lineage heterozygosity (\(nH\)), which explains nearly a quarter of the
-% variation in event frequency (\(R^{2}_{\mathrm{adj}} = 0.24\)).  This
-% relationship is consistent with the mechanistic expectation that recombination
-% requires both frequent co-infections (driven by high case numbers) and
-% genetically distinct co-infecting viruses (driven by lineage diversity).  In
-% other words, recombination becomes most detectable when epidemiological and
-% evolutionary opportunity coincide.  These results imply that the emergence of
-% novel recombinants is not uniform through time but concentrated into
-% predictable ``hot periods'' during large, antigenically diverse outbreaks, when
-% the ecological conditions for viral recombination are most favourable.
-
-% % P5. Confirm breakpoints around Spike. 
-% % SHZ: Maybe we can say it's a known (instead of robust)
-% % pattern of detectable recombination in CoVs in general?
-% % There are some citations for template switching hotspots in SC2.
-% % Also, it's not only about selection favoring new combinations of mutations in Spike.
-% % It's about combinations of mutations in, say, ORF1ab, getting selected against
-% % (or just low mutation rates).
-% % Maybe mention epistasis?
-% [TODO: This is directish output from ChatGPT. Review!]
-% The genomic distribution of breakpoints in the ARG is highly uneven, with a
-% pronounced enrichment near the $5'$ boundary of the Spike gene and across the
-% S1–S2 junction.  This pattern closely mirrors that observed in the
-% UShER+RIPPLES dataset \citep{Turakhia2022} and in earlier 
-% analyses \citep{Jackson2021} [more cites], but here it is recovered
-% independently over a much longer sampling period and under a more conservative
-% inference regime.  Under our more conservative estimate of the overall
-% prevalence of recombination, we find that this enrichment persists despite a substantial
-% reduction in the total number of inferred recombination events, indicating that
-% it reflects a genuine and stable feature of SARS-CoV-2 evolution rather than a
-% by-product of detection sensitivity.  Across all high-confidence events, the
-% median breakpoint interval length is $\sim 2.2$kb and declines with parental
-% divergence, consistent with expectations from sequence-based localization
-% accuracy.  These results show that even with a conservative lower bound on
-% recombination frequency, the concentration of breakpoints around Spike remains
-% a robust signature of coronavirus genome biology—likely reflecting a
-% combination of mechanistic biases in template switching and selective
-% constraints on Spike function.
-
-% % P6. Sc2ts method is a major advance. We validate with real data
-% % rather than simulations. Forward looking statement about simulations.
-% % 
-% % SHZ: Could at least some of this go into the Introduction?
-% In the broader context of pathogen genomics, \texttt{sc2ts} demonstrates that
-% it is now possible to reconstruct recombination-aware genealogies at the scale
-% of entire pandemics.  Although recombination is recognised as a major
-% evolutionary force in many pathogens—including RNA viruses such as influenza
-% and HIV, and numerous bacterial species—genealogical analyses have typically
-% relied on tree-based approximations that ignore or collapse recombination.
-% This simplification has limited the accuracy of downstream evolutionary and
-% epidemiological inference, even in well-studied systems.  \texttt{Sc2ts}
-% overcomes this barrier by directly inferring an ancestral recombination graph
-% from millions of viral genomes, producing the first coherent reticulate
-% genealogy for any pathogen dataset of comparable size.  The ability to
-% integrate recombination into genome-wide genealogies at this scale represents a
-% major conceptual step forward: it transforms recombination from a confounding
-% complication into an analyzable component of pathogen evolution.
-
-% % P7. The software ecosystem enables diverse besplike analysis.
-
-% [TODO: This is close, needs revision]
-% The practical value of the ARG is amplified by its integration into the broader
-% software ecosystems built around \texttt{tskit} and VCF~Zarr.  These mature,
-% open-source libraries provide highly efficient storage and analysis of tree
-% sequences and variant data, allowing millions of genomes to be explored
-% interactively within familiar environments such as Python and Jupyter
-% notebooks.  Analyses that previously required bespoke phylogenetic
-% pipelines—such as extracting local genealogies, summarising ancestral
-% relationships, mapping mutations to branches, or tracking recombinant ancestry
-% through time—can now be performed with concise, reproducible code on standard
-% hardware.  Because \texttt{tskit} and VCF~Zarr share a common data model, the
-% ARG integrates seamlessly with widely used tools in the PyData ecosystem,
-% enabling users to combine genealogical information with epidemiological,
-% geographic, or phenotypic metadata in the same workflow.  This interoperability
-% transforms what can be done with large viral genome collections, lowering the
-% barrier to sophisticated evolutionary analysis and making recombination-aware
-% genealogy inference accessible to a wide community of researchers.
-
-% % P8. Outlook
-
-% [TODO a bit muddled, refine]
-% Together, these results demonstrate that recombination in SARS-CoV-2 can be
-% systematically incorporated into genealogical inference at pandemic scale. The
-% \texttt{sc2ts} ARG captures the complexity of viral evolution in a form that is
-% both biologically realistic and computationally tractable, showing that
-% recombination can be modelled rather than ignored. Embedded within an
-% interoperable software ecosystem, it provides a comprehensive record of
-% SARS-CoV-2 ancestry and a practical framework for real-time,
-% recombination-aware genomic surveillance—turning large sequence archives into
-% dynamic evolutionary models that can be queried with concise, reproducible code. 
-% More broadly, this work marks a transition from
-% tree-based pandemic phylogenetics to recombination-aware genealogy inference.
-% While our focus is SARS-CoV-2, the approach offers a blueprint and a starting
-% point for other recombining pathogens: the core ideas are portable,
-% even though effective deployment will require pathogen-specific adjustments. 
-% We anticipate that this resource will clarify the recent history of
-% SARS-CoV-2 and catalyse methodological refinement and targeted biological
-% studies of recombination across systems where such extensions prove feasible.
-
-% [NOTE: This is Shing's WIP alternative]
-
-% %%% SHZ: WIP alternative.
-% We present the first pandemic-scale
-% ancestral recombination graph (ARG) for SARS-CoV-2,
-% reconstructed from $\sim$2.48 million genomes using \textit{sc2ts}.
-% We demonstrate how this ARG, for the first time,
-% enables systematic identification and characterization of
-% recombination events in fine-grained
-% detail--in terms of their parents, breakpoints, and timing of
-% emergence--providing richer and more nuanced
-% genealogical, genomic, and epidemiological context
-% than existing phylogenetic approaches.
-% Together, sc2ts and the ARG represent a new paradigm in
-% how we study the evolution of pathogens in the face of ongoing sampling.
-
-% Accounting for recombination is a key challenge in pathogen genomics, and
-% recombination has been largely ignored in analysis of viral genomes.
-% Existing phylogenetic approaches for pathogens detect recombination
-% \textit{post hoc} (e.g., UShER+RIPPLES),
-% treat it as a nuance (e.g., XX), or
-% discard its signals from downstream analyses (e.g., XX).
-% Sc2ts addresses this challenge
-% by using the ARG as the underlying data model,
-% explicitly representing recombination events as individual nodes and
-% relating them to specific parents in the ARG.
-% The ARG provides a coherent framework to determine and represent
-% the relationships between recombinants and their specific parents
-% (which are not possible using a tree-based data model, e.g., the MAT format of UShER).
-% The ARG reveals and clarifies the relationships between
-% individual samples of the same recombinant origin,
-% avoiding issues such as over-naming recombinants
-% (e.g., the samples assigned to
-% five differently labeled recombinant Pango lineages
-% descending from the “Xx” recombination event in the ARG).
-% Our results show how an ARG can considerably simplify
-% efforts to identify and monitor emerging recombinants,
-% facilitating communication of genomic surveillance results.
-
-% % SHZ: Very WIP
-% % Discuss the new preprint from Turakhia.
-% The timing of recombination events is of key interest in
-% pathogen genomic epidemiology.
-% because this information may yield insights into
-% how the emergence of a new recombinant might be
-% linked with changes in public health policies.
-
-% The overall frequency of detectable recombination events
-% estimated from the sc2ts ARG is substantially lower than
-% that from UShER+RIPPLES,
-% largely reflecting differences in model formulation and parameter choice.
-% When applied \textit{post hoc} to an UShER phylogeny of
-% $\sim$1.6 million genomes sampled up to May, 2021,
-% RIPPLES reported 589 unique recombination events with 43,104 descendant
-% samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
-% Using the same time interval, sc2ts \(k=4\) identifies
-% 14 high-confidence recombination events with
-% 172 descendant samples among 501,235 samples \(0.034\%\),
-% a 40- to 80-fold reduction.
-% Much of this difference is explained by
-% the stricter parsimony penalty used here:
-% increasing the penalty from p=3 (as used in the RIPPLES analysis) to
-% k=4 greatly reduces false positives
-% while retaining nearly all verified recombinant lineages,
-% including those missed by RIPPLES at equivalent thresholds.
-% These indirect comparisons suggest that
-% our estimate represents a conservative lower bound.
-% In future studies, dissecting the methodological differences
-% between sc2ts and UShER+RIPPLES may
-% inspire new ideas to develop or improve ARG building methods
-% for SARS-CoV-2 (and potentially other pathogens).
-
-% The ARG itself is a powerful resource
-% to support future studies of SARS-CoV-2.
-% Our results show that the ARG captures and unifies
-% important features of SARS-CoV-2 evolution,
-% which are shown in different studies
-% using various methods, tools, and approaches.
-% The ARG also compactly encodes genetic variation in
-% millions of SARS-CoV-2 genomes,
-% allowing for convenient and efficient reuse of
-% this valuable dataset for a range of
-% genetic and evolutionary analyses beyond those presented here,
-% for instance, to understand
-% how the variability in sequencing efforts
-% across time and location affects our ability
-% to reconstruct the origin of important recombinant lineages and
-% to estimate when they emerge.
-% Because building the ARG is computationally expensive, and
-% checking and improving various aspects of its quality is laborious,
-% we hope that making this resource openly available will
-% save significant time and effort for other researchers,
-% enabling immediate focus on analyzing the information
-% contained within the COVID-19 dataset.
-
-% % TODO: Briefly discussion software tools here or in a subsequent paragraph.
-
-% Recent developments in ARG methods have focused almost exclusively on
-% applications to human sequencing data.
-% The availability of human ARGs reconstructed from large datasets
-% has already had a tremendous impact on improving
-% our understanding of human genetic ancestry and past demography.
-% We anticipate similar advances to be made
-% by adapting existing inference methods for human ARGs
-% to use the sc2ts ARG for the genealogy-based study of
-% natural selection or the geographic spread of SARS-CoV-2 through time.
-
-% The unprecedented throughput and pace of genome sequencing are
-% creating ample opportunity to understand the impacts of recombination
-% on pathogen evolution and epidemiology.
-% We hope that this study will spur
-% development of ARG methods for pathogens,
-% ultimately making recombination-aware genome analysis
-% routine for all sorts of pathogens beyond SARS-CoV-2.
+Together these results underscore the utility of the ARG itself 
+and suggest important directions for future research.
 
@@
 
-We have focused on applying sc2ts to the Viridian dataset rather 
+We have focused on applying sc2ts to the Viridian dataset~\cite{Hunt2024} rather 
 than the larger GISAID resource~\cite{Shu2017} for two reasons.
@@
 GISAID data is straightforward, but significant additional efforts 
-would be required to deal with the various data quality issues.
+would be required to address data quality issues.
 Secondly, the Viridian data is publicly available and redistributable,
@@
 
-% Need to be careful here to list stuff that sounds interesting and 
-% hard enough to not get asked to do it.
-The sc2ts method may be extended and improved in several directions.
-Firstly, the simplistic parameterisation of the LS HMM used 
-here could be extended to include more realistic
-models with mutation rate variation along the genome~\cite{Haddox2025}
-and transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024}.
-Secondly, the current handling of deletions by post-hoc parsimony
-could be improved either by a more nuanced data-masking strategy
-(in which, e.g., all but the right-most base of known deletions is 
-masked as missing data during HMM matching), or by the more 
-challenging approach of incorporating multi-base events into the HMM directly.
-Sc2ts is currently specialised for SARS-CoV-2 data, but the methods
-may be readily adapted to other pathogens.
-
-We have used the tsdate method~\cite{Wohns2022,Pope2025} to date the 
-sc2ts ARG, and initial results are promising. However, dating 
-of phylogenies in itself is a complex task~\cite{Kumar2016advances}
-and methods for incorporating information from metadata of uncertain 
-quality into large trees are in their 
-early stages~\cite{sanderson2021chronumental} and there is certainly
-substantial room for improvement.
-Much of the complexity of sc2ts lies in 
-dealing with ``time traveller'' samples, where sampling dates 
-provided in metadata are substantially incorrect. 
-The present ARG may provide 
-a way of improving this, by iteratively matching samples against
-it to estimate dates for all sequences, and adjusting the dates
-for time-travellers detected in this way. Sc2ts could then
-be run without the complex filters required to remove 
-time-travellers and unconditionally include all but the most divergent
-sequences. Other studies may also benefit from the 
-improved sample dates.
-
-[Final para is prelim. Acknowledge limitations, forward looking 
-for how we might deal with them]
-Finally, it is important to acknowledge that sc2ts performs best
-with dense sampling, and the dataset is heavily geographically biased.
-[cite ~\cite{Lancet2021,Kalia2021,Chen2022}?]
-Long branches, cause particular problems. Future improvements would 
-ideally detect when data is sparse and switch to more sophisticated
-models, as discussed in~\cite{Bisschop2025likelihoods}.
-
-% It is challenging to accurately reconstruct global events
-% when there are regions with poor sampling and
-% jurisdictions with long delays
-% before sequences become publicly
-% available~\cite{Lancet2021,Kalia2021,Chen2022}.
-% For example, revisiting the ARG topology periodically
-% (e.g., rewinding the ARG a month and repeating
-% with samples that became available later)
-% might improve the accuracy of the genealogy.
-
-% We have developed and used several methods
-% to exclude spurious recombination events \textit{post hoc},
-% including filtering by supporting loci per parent and
-% splitting of suspiciously long branches (e.g., Delta and Omicron BA.2).
-% These methods could instead be run during ARG inference,
-% immediately after sample matching,
-% reducing the number of ARG post-processing steps.
-
-% Finally, an inherent limitation of detecting recombination from sequence data,
-% which applies to all methods,
-% is that it is impossible to detect events with few (or no) supporting bases,
-% as these cannot be distinguished from mutation events.
-% This limitation makes it particularly challenging
-% to detect events near the termini of linear viral genomes.
+It is important to acknowledge that sampling is nonuniform both
+geographically and over time~\cite{Lancet2021,Chen2022}. 
+Sc2ts performs best in the data-dense regime, when the underlying
+assumption that genetic diversity is gradually accumulating within
+the structure holds. Large changes (i.e, saltation events) present
+significant challenges and the results of the algorithm should 
+be regarded more critically in these situations.
+Future methodological improvements may combine the strengths 
+of sophisticated model-based inferences with the power 
+of non-parametric inference~\cite{Bisschop2025likelihoods}
+when such data sparsity is detected.
 
@@
 then running pangolin (version 4.3.1; pangolin-data version 1.29).
+Note that this version of pangolin uses UShER placement
+for classification~\cite{McBroome2024}.
 The resulting Pango and Scorpio designations were then stored as 

### 2025-11-06 latex/main.tex
Updates from Overleaf

@@
 %%%  components. Each component must be present.
-
-[FIXME: fix the typesetting once all resources are compiled]
-
 \begin{itemize}
+
 \item The sequences alignments in VCF Zarr format have been deposited at Zenodo and are publicly available at \url{https://zenodo.org/records/16314739}.
@@
 (Oxford Population Health, University of Oxford). 
-JK and YW acknowledge support from R01 HG012473 from the National Institutes of Health NHGRI, EPSRC (research grant EP/X024881/1) and the Robertson Foundation.
+% "ackknowledges support" here means "was awarded this grant 
+% and it supported this research"
+JK acknowledges support from R01 HG012473 from the National Institutes of Health NHGRI, EPSRC (research grant EP/X024881/1) and the Robertson Foundation.
 
@@
 RIPPLES events associated with Pango X lineages.
-Each row is a recombination event identified by RIPPLEs (at p = 3)
+Each row is a recombination event identified by RIPPLEs (at $p=3$)
 in which Pango X lineages appear,
@@
 where applicable, grouped by class.
-Events also present at p = 4 are indicated.
+Events also present at $p=4$ are indicated.
 Clades are matched by finding the most recent common ancestor of
 the samples associated with each UShER+RIPPLES event in the sc2ts ARG.
-The clade differences indicates how close this match is.
+The clade differences indicates the closeness of the match.
 }

### 2025-11-06 latex/main.tex
Spell check

@@
 of species~\cite{Felsenstein2004}, using small numbers of widely diverged
-genomes. In this context, recombination is a nuisance which must be 
-carefully accounted for~\cite{Rannala2025recombination},
-affecting the inference of species trees~\cite{Posada2002effect}
-and biasing downstream
+genomes. In this context, recombination is a nuisance 
+requiring careful consideration~\cite{Rannala2025recombination}
+as it affects the inference of species trees~\cite{Posada2002effect}
+and biases downstream
 analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
@@
 as of October 2025.
-Regularly reinferring trees de-novo from such rapidly
+Regularly re-inferring trees de-novo from such rapidly
 growing datasets is untenable.
@@
 broadening the range of potential applications~\cite{Bisschop2025likelihoods}.
-Another important point is that strong software support for ARGs now
+Another significant point is that strong software support for ARGs now
 exists, and the tskit library~\cite{Ralph2020} is the basis of a 
@@
 simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
-visualization~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit,Talbot2025argscape},
+visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit,Talbot2025argscape},
 inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
@@
 on genetic inheritance~\cite{Delaneau2019-wl,Browning2018-nk},
-typically parameterized by population-scaled mutation and
+typically parametrised by population-scaled mutation and
 recombination rates varying along the genome~\cite{Donnelly2010-coalescent}.
@@
 Illustrative subgraphs in the ARG.
-Boxes identify specific samples, colored by their Pango X classification;
+Boxes identify specific samples, coloured by their Pango X classification;
 smaller squares list extra samples descending from a node, hidden for brevity,
@@
 The descendants column shows the descendant samples broken down by Pango lineage.
-For Type III events, the focal pango, event (Type I or II) it is derived 
+For Type III events, the focal Pango, event (Type I or II) it is derived 
 from and the distance (path length and time in days) from that event are shown,
-as well as the number of mutations directly ancestral to the pango origin
-node. For Type IV events, the pango designation of the parent node and number
+as well as the number of mutations directly ancestral to the Pango origin
+node. For Type IV events, the Pango designation of the parent node and number
 of mutations are shown.
@@
 A,B,C show summaries of the 354 sc2ts recombination nodes passing QC filters
-colored by 3SEQ classification (recombinant: purple, non-recombinant: grey).
+coloured by 3SEQ classification (recombinant: purple, non-recombinant: grey).
 (A) Upper: jittered scatterplot comparing number of mutations
@@
 versus an alternative copying path disallowing recombination ($x$: top axis);
-recombination nodes in Table~\ref{tab:pango_x_lineages} are labeled starting with `X' and
+recombination nodes in Table~\ref{tab:pango_x_lineages} are labelled starting with `X' and
 those identified by Jackson et al.~\cite{Jackson2021} are prefixed with a `J',
@@
 in a recombinant (middle row) matches that of
-the left (upper row, colored green) 
-or right parent (lower row, colored blue),
+the left (upper row, coloured green) 
+or right parent (lower row, coloured blue),
 or where the recombination event requires
@@
 observed in the copying patterns suggests the existence of
-many genetically distinct, un-sampled ancestors
+many genetically distinct, unsampled ancestors
 and some caution should be exercised when interpreting them
@@
 (annotated with gene labels on top).
-Each line represents a breakpoint interval colored by its length.
+Each line represents a breakpoint interval coloured by its length.
 (B) Number of overlapping breakpoint intervals per site.
@@
 we note several sources of uncertainty.
-Global case numbers were underreported,
+Global case numbers were under reported,
 particularly during the Delta and Omicron waves~\cite{Wang2025}.
@@
 among all possible recombinant solutions. The current 
-parameterisation---emphasising interpretability---is simplistic,
+parametrisation---emphasising interpretability---is simplistic,
 and could readily be extended to account
@@
 
-It is important to acknowledge that sampling is nonuniform both
+It is important to acknowledge that sampling is non uniform both
 geographically and over time~\cite{Lancet2021,Chen2022}. 
@@
 (Oxford Population Health, University of Oxford). 
-% "ackknowledges support" here means "was awarded this grant 
+% "acknowledges support" here means "was awarded this grant 
 % and it supported this research"
@@
 % %%%  of the data; please do not use bold, italic,
-% %%%  colored text, or shading for this purpose. Use
+% %%%  coloured text, or shading for this purpose. Use
 % %%%  separate cells, not line breaks or spaces, for
 % %%%  all discrete data elements. Small embedded
-% %%%  graphics with color are OK.
+% %%%  graphics with colour are OK.
 
@@
 It is a Hidden Markov Model (HMM) in which a focal genome
-is modeled as a
+is modelled as a
 sequence of nucleotides that are probabilistically emitted as
@@
 significant processing time and memory. This is reflected in 
-the computaional resources reported in 
+the computational resources reported in 
 Figures~\ref{fig:inference_resources},\ref{fig:samples_per_day}.
@@
 without incurring too many mutations, particularly reversions.
-A value of $k$ too low could result in many artifactual recombination events,
+A value of $k$ too low could result in many artefactual recombination events,
 whereas a value of $k$ too high could miss many genuine recombination events.
@@
 reflecting their evolutionary relationships.
-We use a standard neighbor-joining algorithm\cite{Saitou1987}
+We use a standard neighbour-joining algorithm\cite{Saitou1987}
 to infer these within-cluster evolutionary relationships.
@@
 and then use biotite\cite{Kunzmann2018biotite,Kunzmann2023biotite}
-to compute the neighbor-joining tree.
+to compute the neighbour-joining tree.
 Then, mutations are mapped back onto this daily sample cluster tree using 
@@
 The HMM cost threshold approach outlined in the previous section is 
-effective at removing time-traveling samples, but it also filters out 
+effective at removing time-travelling samples, but it also filters out 
 important evolutionary events such as
@@
 pressures\cite{Otto2021origins,Markov2023}.
-To distinguish time travelers from true saltation lineages,
+To distinguish time travellers from true saltation lineages,
 we re-evaluate the evidence for each lineage with a high HMM cost
@@
 correlated error is non-trivial, and sc2ts has 5 parameters to fine
-tune the behavior. Firstly, we require at least 10 samples in the
+tune the behaviour. Firstly, we require at least 10 samples in the
 group to ensure there is sufficient support over a time window of 
@@
 % complicating the story.
-For Delta, we used two seeds: one for B.1.617.1 and 1 sample assigned to
-B.1.617.2, [FIXME more Delta details and some references to back up our assertions.
-There's some specific info in the Viridian paper on this, I'm sure. 
+For Delta, we used two seeds:
+one for B.1.617.1 (Kappa, which is sibling to Delta)
+and one sample assigned to B.1.617.2.
+[FIXME more Delta details and some references to back up our assertions.
+There's some specific info in the Viridian paper on this,
+I'm sure. 
 We should mention long branch splitting here too.]
-The resulting subgraph can be see in Figure~\ref{fig:subgraph_delta}.
+The resulting subgraph can be seen in Figure~\ref{fig:subgraph_delta}.
 
@@
 the Viridian work]
-For Omicron BA.1 and BA.2, we used SRR17041376 and SRR17461792, respectively.
+For Omicron BA.1 and BA.2,
+we used SRR17041376 and SRR17461792, respectively.
 We chose SRR17041376 (a sample dated 2021-11-12 from South Africa)
 as the BA.1 seed (https://github.com/jeromekelleher/sc2ts-paper/issues/264),
-as it could be well explained as as a non-recombinant given the ARG available on its collection date.
+as it could be well explained as as a non-recombinant
+given the ARG available on its collection date.
 For BA.2, we chose SRR17461792 (dated 2021-11-27 from South Africa),
@@
 Figure~\ref{fig:copying_pattern_examples} shows illustrative copying patterns
-(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}).
+(Q1 to Q4 correspond to the labelled quadrants in Figure~\ref{fig:recombinant_qc}).
 Q1 shows a well supported recombination event that passed QC (albeit marginally),
@@
 a combination of this Delta deletion and a known MNS (see below).
-Finally, Q4 shows an artifactual recombination event
+Finally, Q4 shows an artefactual recombination event
 associated with the A507T-T508C-G509A MNS,
@@
 using the mutational counts from an UShER phylogeny 
-analyzed by Bloom et al.\cite{Bloom2023},
+analysed by Bloom et al.\cite{Bloom2023},
 which was built using a Nextclade pipeline and
@@
 
-To analyze how well the ARG reflects the phylogenetic structure implicit in the
+To analyse how well the ARG reflects the phylogenetic structure implicit in the
 Pango lineage naming system, we considered the Pango assignments generated by
@@
 clade descending from a single originating node with that label. We identified
-putative origination nodes among those labeled with a given lineage as the node
+putative origination nodes among those labelled with a given lineage as the node
 with the maximum number of descendant samples and the earliest time (if there
@@
 
-Of the 2058 distinct pango lineages in the ARG, 1473 of these (comprising
+Of the 2058 distinct Pango lineages in the ARG, 1473 of these (comprising
 717798 samples) match perfectly, with unique origination events in the ARG
@@
 However, a characteristic three-base correlated mutation
-is instead associated with the parent node (labeled B.1.1) of the Alpha root,
+is instead associated with the parent node (labelled B.1.1) of the Alpha root,
 suggesting that this mutation was acquired prior to the emergence of Alpha.
@@
 In the ARG, all six of the shared Delta-Kappa mutations (red)
-occur above the B.1-labeled common ancestor
+occur above the B.1-labelled common ancestor
 of Kappa (left branch) and Delta (right branch).
 All nine of the Kappa-specific mutations (blue)
-occur above the B.1.617-labeled node leading to Kappa samples.
+occur above the B.1.617-labelled node leading to Kappa samples.
 Seven out of nine Delta-specific mutations (purple)
@@
 BA.1 and BA.2 differ from the non-Omicron B.1.1 ancestor
-(above the B.1.1.529-labeled MRCA of BA.1 and BA.2)
+(above the B.1.1.529-labelled MRCA of BA.1 and BA.2)
 by 27 mutations (including two deletions),
@@
 
-In cases where nested samples not assigned to the focal pango lineage are
-either from different pango X lineages 
-or given a generic BA.2 assignment. Of these 59 generic assignents
+In cases where nested samples not assigned to the focal Pango lineage are
+either from different Pango X lineages 
+or given a generic BA.2 assignment. Of these 59 generic assignments
 6 had conflicts in their UShER placements and could have been
@@
 four groups of recombinants (named A to D) and four singleton recombinants.
-Of the 16 samples analyzed by Jackson et al., 12 were included in the ARG;
+Of the 16 samples analysed by Jackson et al., 12 were included in the ARG;
 the remaining 4 were excluded
@@
 which sc2ts identified as descendants of the same recombination node
-but are labeled BA.2 by pangolin (displayed on the left), and
+but are labelled BA.2 by pangolin (displayed on the left), and
 which do not possess the XJ consensus mutations C22792T and C27945T.
@@
 descending from the first recombination event.
-We excluded this likely artifactual event from further analysis and
+We excluded this likely artefactual event from further analysis and
 consider XS to have emerged from a single recombinant origin.
@@
 All the XW samples descend from a well supported recombination node.
-Its left parent is a BA.1.1.15-labeled sample (ERR8069658), and
-its right parent is a BA.2-labeled sample (ERR9086669).
+Its left parent is a BA.1.1.15-labelled sample (ERR8069658), and
+its right parent is a BA.2-labelled sample (ERR9086669).
 
@@
 which is supported by a sample (ERR8627048) identical to it.
-Its right parent is a BA.2-labeled sample (ERR9399983).
+Its right parent is a BA.2-labelled sample (ERR9399983).
 There are a pair of (orange) reversion mutations
@@
 roughly within a three-month period (Mar., 2022 to Jul., 2022).
-Several samples labeled as BA.2 by Pangolin also lie
+Several samples labelled as BA.2 by Pangolin also lie
 within this recombinant clade,
@@
 The inferred topology places these sites
-as recurrent (colored orange, grey, and faded yellow),
+as recurrent (coloured orange, grey, and faded yellow),
 but a more parsimonious arrangement would,
@@
 However, we find a small recombinant BA1.1/BA.2 clade of 37 samples,
-associated with the BA.2-labeled sample ERR9755541,
+associated with the BA.2-labelled sample ERR9755541,
 within which the single XAF sample is nested.
@@
 with the region from 21765-23535
-coming from the BA.4 labeled sample SRR18848050.
+coming from the BA.4 labelled sample SRR18848050.
 
@@
 the XBB-associated recombination node
-may have attached to a BJ.1-labeled node, and
+may have attached to a BJ.1-labelled node, and
 the origin of XBB would required fewer mutations.
@@
 Its right parent is a reconstructed internal node
-that is supported by a CJ.1-labeled sample (SRR21948068),
+that is supported by a CJ.1-labelled sample (SRR21948068),
 which differs by three mutations.
@@
 Exactly such a missing ancestor
-is found as the upper internal node labeled CJ.1 in the ARG,
+is found as the upper internal node labelled CJ.1 in the ARG,
 constructed by sc2ts as part of its tree building steps.
@@
 Its left parent is a reconstructed internal node
-that is directly supported by a BA.2.76-labeled sample identical to it.
-Its right parent is a BF.3-labeled sample.
+that is directly supported by a BA.2.76-labelled sample identical to it.
+Its right parent is a BF.3-labelled sample.
 
@@
 XA, 
+XB, 
 XAC, 
@@
 XAJ,
-XB, 
 XBB, 
@@
 (%
-XAL,
 XAM,
@@
 XE,
+XM,
 XS,
@@
 and XAJ (Document~\ref{sec:dd_xaj}).
-See Document~\ref{sec:detailed_ripples} for more detailed analysis.
 
@@
 \caption{
-Daily statistics of samples analyzed
+Daily statistics of samples analysed
 during the ARG inference process.
@@
 Sites associated with multiple mutations in the subgraph
-have their mutations assigned a unique fill color
+have their mutations assigned a unique fill colour
 (orange, blue, green, red, etc);
@@
 outlined in green.
-These colors correspond to those used
+These colours correspond to those used
 in the BA.1/BA.2 Pango designation issue \cite{RambautOmicron2021}.
@@
 \includegraphics[width=\linewidth]{recombinant_evidence_cov_recomb}
-\caption{Properties of sc2ts recombination events, colored by CovRecomb 
+\caption{Properties of sc2ts recombination events, coloured by CovRecomb 
 classification. All other details as per Figure~\ref{fig:recombinant_evidence}.
@@
 \includegraphics[width=\linewidth]{recombinant_evidence_rebar}
-\caption{Properties of sc2ts recombination events, colored by rebar
+\caption{Properties of sc2ts recombination events, coloured by rebar
 classification. All other details as per Figure~\ref{fig:recombinant_evidence}.
@@
 on the left parent (x-axis) and the right parent (y-axis).
-The shaded regions highlight potentially artifactual recombination events,
+The shaded regions highlight potentially artefactual recombination events,
 which have fewer than 4 net supporting loci
 on one side or both sides of the suggested breakpoint.
-Colors indicate the primer scheme used for sequencing.
+colours indicate the primer scheme used for sequencing.
 The pie charts show breakdowns of the recombination events
@@
 Illustrative copying patterns
-drawn from the quadrants labeled in Figure~\ref{fig:recombinant_qc}. 
+drawn from the quadrants labelled in Figure~\ref{fig:recombinant_qc}. 
 Each copying pattern shows the 
@@
 ``copying path'' through the reference panel for a given focal sequence that
-maximises the likelihood under these parameters. Unseen states in the reference panel are shown as colored lines enclosed by
+maximises the likelihood under these parameters. Unseen states in the reference panel are shown as coloured lines enclosed by
 the grey box. The black arrow describes the true path through the data which leads to the emitted
@@
 shared by two siblings, and we create a new node to represent the ancestor
-on which this mutation occured. (B) Reversion pushing. Mutation A128C is
+on which this mutation occurred. (B) Reversion pushing. Mutation A128C is
 immediately reverted by C128A, and we create a new node to represent the 

### 2025-11-06 latex/main.tex
Updates from Overleaf

@@
 particularly challenging, however, as the initial emergences of 
-the saltation lineages were not densely sampled. In this case,
+these lineages were not densely sampled. In this case,
 there is little information for the retrospective sampling grouping mechanism
@@
 % complicating the story.
-For Delta, we used two seeds:
-one for B.1.617.1 (Kappa, which is sibling to Delta)
-and one sample assigned to B.1.617.2.
-[FIXME more Delta details and some references to back up our assertions.
-There's some specific info in the Viridian paper on this,
-I'm sure. 
-We should mention long branch splitting here too.]
-The resulting subgraph can be seen in Figure~\ref{fig:subgraph_delta}.
-
-[FIXME: revise this paragraph to give some background info on Omicron, and reference
-the Viridian work]
-For Omicron BA.1 and BA.2,
+For Delta
+(first detected in December, 2020 in India \cite{ECDCVOCUpdates}),
+% More samples were unconditionally added,
+% but the earliest are the seeds.
+we used two seeds:
+one sample (ERR5461562) assigned to B.1.617.1 (Kappa) and
+the other (ERR5676810) assigned to B.1.617.2 (Delta).
+Kappa is closely related to Delta, and
+therefore seeding it helps to reconstruct the origin of Delta.
+% See https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml
+Both of these seeds were sampled in the UK,
+the Kappa one in February, 2021 and the Delta one in March, 2021.
+% https://github.com/jeromekelleher/sc2ts-paper/issues/267
+% The reality is a bit more complicated,
+% but this is easy to explain.
+These are the earliest Kappa and Delta samples
+in the Viridian v04 dataset
+that were also used for Pango designation.
+% FIXME more Delta details and some references to back up our assertions
+% There's some specific info in the Viridian paper on this, I'm sure.
+% We should mention long branch splitting here too.]
+The resulting subgraph around Delta in the ARG
+can be seen in Figure~\ref{fig:subgraph_delta};
+for further explanation, see Document~\ref{sec:dd_delta}.
+
+% FIXME: revise this paragraph to give some background info on Omicron,
+% and reference the Viridian work]
+For Omicron BA.1 and BA.2
+(first detected in South Africa or Botswana
+in November, 2021 \cite{ECDCVOCUpdates}),
 we used SRR17041376 and SRR17461792, respectively.
-We chose SRR17041376 (a sample dated 2021-11-12 from South Africa)
-as the BA.1 seed (https://github.com/jeromekelleher/sc2ts-paper/issues/264),
-as it could be well explained as as a non-recombinant
+We chose SRR17041376 (dated November 12, 2021 from South Africa)
+as the BA.1 seed
+(\url{https://github.com/jeromekelleher/sc2ts-paper/issues/264}),
+as it could be well explained as a non-recombinant
 given the ARG available on its collection date.
-For BA.2, we chose SRR17461792 (dated 2021-11-27 from South Africa),
+For BA.2,
+we chose SRR17461792 (dated November 27, 2021 from South Africa),
 which is the earliest BA.2 sample present in the Viridian v04 dataset.
-It is dated later than SRR17041376 (the BA.1 seed),
-allowing time for the BA.1 lineage to accumulate enough sampled descendants
-and to get established in the ARG.
-See Figure~\ref{fig:subgraph_omicron} for an overview of the resulting 
-subgraph.
+This sample is dated later than SRR17041376 (the BA.1 seed),
+allowing time for the BA.1 lineage
+to accumulate enough sampled descendants and
+to get established in the ARG.
+The resulting subgraph around these Omicron lineages in the ARG
+can be seen in Figure~\ref{fig:subgraph_omicron}.
+For further explanation, see Document~\ref{sec:dd_omicron}.
 
@@
 XB, 
+XC, 
+XL, 
+XW,
 XAC, 
@@
 XBB, 
-XBM, 
-XC, 
-XL, 
-XW%
+XBM%
 )
@@
 (%
-XAM,
-XBD, 
-XBR, 
 XE,
@@
 XS,
-XY%
+XY,
+XAM,
+XBD, 
+XBR%
 )
@@
 one event subtending 1321 samples suggests that 
-XAF,
-XAL, 
 XE, 
@@
 XY, 
+XAF,
+XAL, 
 and 
-82 BA.2 samples share a recombination event [FIXME: With what breakpoint and parents?].
+82 BA.2 samples share a recombination event whose parents are BA.2 and BA.1 [FIXME: With what breakpoint and parents?].
 We also ran RIPPLES with $p=4$, which reduced the overall number of
@@
 a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
-8 have corresponding recombination events (4, at $p=4$).
+8 are found by RIPPLES (4, at $p=4$).
 Of the 5 Type II events
 (recombination nodes with more complex assocation with X lineages),
-two have close matches, with XBM matching exactly
+two have close matches in RIPPLES, with XBM matching exactly
 (including 2 descending BF.3 samples),
@@
 events are likely attributable to the finer details of the 
-differences in tree building approaches between UShER and sc2ts
-
-[FIXME: Discuss  the large effect recombinants inferred by both]
-
-Two further recombinants in the sc2ts ARG that are likely to be 
-spurious are important to discuss.
-[FIXME - discuss BA.5 (kept as a recombination event but only because we can't 
-rewire it post-hoc, owning to subsequent tree building) and the BA.2 event (doesn't
-pass QC.) - note this is discussed in more detail in document \ref{sec:spurious-ancestrally-important-recombinants} ]
+differences in tree building approaches between UShER and sc2ts.
+
+[
+FIXME: Discuss  the large effect recombinants inferred by both:
+sc2ts identifies two recombination events with \textgreater 10k descendants (specifically, 166,979 and 166,285 descendant samples), both of which we identify as spurious (see \ref{sec:spurious-ancestrally-important-recombinants}).
+RIPPLES at $p=4$ also identifies two recombination events with \textgreater 10k descendants (531,224 and 339,523 descendant samples); these do not correspond to the same events as in sc2ts.
+At $p=3$, RIPPLES identifies six recombination events with \textgreater 10k descendant samples, four of which have \textgreater 100k descendant samples.
+]
 
@@
 RIPPLES creates two independent recombinations.
-Similarly, XAE and ??? have exactly matching clades, but RIPPLES
+Similarly, XAE and [FIXME] ??? have exactly matching clades, but RIPPLES
 infers two independent recombination events while sc2ts groups them both

### 2025-11-06 latex/main.tex
Updates from Overleaf

@@
 (first detected in December, 2020 in India \cite{ECDCVOCUpdates}),
+% https://github.com/jeromekelleher/sc2ts-paper/issues/267
 % More samples were unconditionally added,
@@
 we used two seeds:
-one sample (ERR5461562) assigned to B.1.617.1 (Kappa) and
-the other (ERR5676810) assigned to B.1.617.2 (Delta).
-Kappa is closely related to Delta, and
+% https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml
+ERR5461562 assigned to B.1.617.1/Kappa
+(dated February 22, 2021 from the UK) and
+ERR5676810 assigned to B.1.617.2/Delta
+(dated March 23, 2021 from the UK).
+Kappa is closely related to Delta \cite{Stern2021}, and
 therefore seeding it helps to reconstruct the origin of Delta.
-% See https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml
-Both of these seeds were sampled in the UK,
-the Kappa one in February, 2021 and the Delta one in March, 2021.
-% https://github.com/jeromekelleher/sc2ts-paper/issues/267
-% The reality is a bit more complicated,
-% but this is easy to explain.
+% What we actually did is a bit more complicated,
+% as it was manual, but this is easier to explain.
 These are the earliest Kappa and Delta samples
 in the Viridian v04 dataset
-that were also used for Pango designation.
+% https://github.com/jeromekelleher/sc2ts-paper/issues/267
+% SHZ: I'd like to recheck this.
+that were used for the designation of these Pango lineages.
 % FIXME more Delta details and some references to back up our assertions
@@
 [
-FIXME: Discuss  the large effect recombinants inferred by both:
-sc2ts identifies two recombination events with \textgreater 10k descendants (specifically, 166,979 and 166,285 descendant samples), both of which we identify as spurious (see \ref{sec:spurious-ancestrally-important-recombinants}).
-RIPPLES at $p=4$ also identifies two recombination events with \textgreater 10k descendants (531,224 and 339,523 descendant samples); these do not correspond to the same events as in sc2ts.
+FIXME: Discuss the large effect recombinants inferred by both in more detail. E.g.:
+sc2ts identifies two recombination events with \textgreater 10k descendants (specifically, 166,979 and 166,285 descendant samples),
+both of which we identify as spurious (see \ref{sec:spurious-ancestrally-important-recombinants}).
+RIPPLES at $p=4$ also identifies two recombination events with \textgreater 10k descendants (531,224 and 339,523 descendant samples);
+these do not correspond to the same two spurious large-effect events in sc2ts.
 At $p=3$, RIPPLES identifies six recombination events with \textgreater 10k descendant samples, four of which have \textgreater 100k descendant samples.

### 2025-11-06 latex/main.tex
Update table Closes #850
Fixup RIPPLES text for pango X events

Final run and cleanup of the ripples nb

Supp and Methods should be done also now.

Pass through RIPPLES

@@
 Classical phylogenetics is largely concerned with inferring the evolutionary history 
-of species~\cite{Felsenstein2004}, using small numbers of widely diverged
+of multiple species~\cite{Felsenstein2004}, using small numbers of widely diverged
 genomes. In this context, recombination is a nuisance 
@@
 In contrast to the sophisticated model-based approaches
-used to perform species-level tree
+used to perform species tree
 inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
@@
 primarily due to computational scaling limitations,
-but also because of the complexity of modeling such datasets.
+but also because of the complexity of modelling such datasets.
 While several methods have been proposed to incorporate
@@
 recombination-aware inference is mostly limited to
-visualization, for example of small-scale phylogenetic
+visualisation, for example of small-scale phylogenetic
 networks~\cite{Huson2012,Vaughan2017,Solis-Lemus2017}
@@
 The global
-surveillance data collected during the COVID-19 pandemic presents fundamental challenges to existing approaches
+surveillance data collected during the COVID-19 pandemic
+presents fundamental challenges to existing approaches
 in two ways. Firstly, the sheer amount of data 
-overwhelmed even approximate distance-matrix based methods.
+overwhelmed even approximate distance matrix-based methods.
 % Accessed GISAID/EpiCov on 08 Oct., 2025
-At the peak in February 2022 about 80,000 samples
-per day  were submitted to GISAID\cite{Shu2017},
+At the peak in February, 2022 $\sim$80,000 samples per day
+were submitted to GISAID~\cite{Shu2017},
 accumulating to a total of over 17.5 million samples
 as of October 2025.
-Regularly re-inferring trees de-novo from such rapidly
+Regularly re-inferring trees de novo from such rapidly
 growing datasets is untenable.
@@
 Although several methods have been proposed to \emph{detect}
-recombinant samples~\cite{
-Varabyou2021rapid,% Bolotie
+recombinant samples~\cite{Varabyou2021rapid,%Bolotie
 Turakhia2022,%RIPPLES
-Zhou2023virusrecom,% CovRecomb
+Zhou2023virusrecom,
 Alfonsi2024,%RecombinHunt 
-Li2024CovRecomb%CovRecomb 
-},
+Li2024CovRecomb},
 this information is not incorporated into the phylogenies used 
@@
 We used the highly efficient LS HMM implementation in tsinfer~\cite{Kelleher2019},
-with some additional optimizations enabled by our simplified model,
+with some additional optimisations enabled by our simplified model,
 to find exact Viterbi solutions of these copying paths for each sample
@@
 the ARG can be loaded into memory in half a second,
-and the vectorized representation integrates tightly
+and the vectorised representation integrates tightly
 with powerful data science tools, such as NumPy~\cite{Harris2020array}.
@@
 (C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
-For more detailed visualizations of these subgraphs, see Document S2.
+For more detailed visualisations of these subgraphs, see Document S2.
 }
@@
 discovering and assessing evidence around these events.
-Table~\ref{tab:pango_x_lineages} summarizes
+Table~\ref{tab:pango_x_lineages} summarises
 these results for the 44 Pango X lineages present in the ARG
@@
 that they may be more parsimoniously explained as sublineages of XQ
-than de-novo recombinant lineages (Document~\ref{sec:dd_xq}). 
+than de novo recombinant lineages (Document~\ref{sec:dd_xq}). 
 The second cluster is based around a putative undocumented
@@
 XB, on the other hand,
-has been thoroughly characterized~\cite{Gutierrez2022} and
+has been thoroughly characterised~\cite{Gutierrez2022} and
 is likely misclassified by sc2ts because the required
@@
 
-The remaining 10 X lineages (97 samples) present in the dataset are not 
+The remaining 10 Pango X lineages (97 samples) present in the dataset are not
 included in the ARG because the HMM cost exceeds the inclusion threshold
@@
 associated with Table~\ref{tab:pango_x_lineages};
-and Document S2 for ARG visualizations of the Pango X lineages.
+and Document S2 for ARG visualisations of the Pango X lineages.
 
@@
 on the putative recombinant sequence trios (STAR Methods).
-Of the 855 recombination events identified by sc2ts, 3SEQ verified 57\% of them, rising to 95\% verification among the 354 recombinants that passed our QC filter. 
+Of the 855 recombination events identified by sc2ts,
+3SEQ verified 57\% of them,
+rising to 95\% verification among the 354 recombinants that passed our QC filter. 
 In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
@@
 This is expected, as CovRecomb is specifically intended
-to detect interlineage recombinants
+to detect inter-lineage recombinants
 and has previously been shown to be
@@
 To aid such interpretation and help identify QC issues,
-we developed a visualization tool
+we developed a visualisation tool
 inspired by SnpIt~\cite{OToole2024} and 
@@
 
-The precise genomic location of a recombination breakpoint is 
-in general impossible to determine, but we can place 
-bounds by inspecting sequences of the recombinant and its parents.
+The genomic location of a recombination breakpoint in general
+cannot be determined exactly at single base resolution,
+but we can place bounds
+by inspecting sequences of the recombinant and its parents.
 The breakpoint interval of a recombinant is defined by
@@
 RIPPLES~\cite{Turakhia2022}. 
-% NOTE: not citing RIPPLES here as there is no real analysis
+% NOTE: not citing RIVET here as there is no real analysis
 % in the paper itself, it's just focused on the tool
-RIPPLES works by taking a
-pre-existing UShER phylogeny and systematically breaking up
+RIPPLES works by taking an UShER phylogeny and systematically breaking up
 sequences subtended by long branches into multiple segments 
@@
 
-We compared the performance of sc2ts recombinant detection against RIPPLES run on the Viridian UShER phylogeny,
+We compared the performance of sc2ts recombinant detection 
+against RIPPLES run on the Viridian UShER phylogeny,
 subset to the 2,475,418 samples shared with the ARG (STAR Methods).
 To enable exact comparison, we omitted the QC filtering steps from both
-methods, and initially set the RIPPLES ``parsimony improvement'' parameter to $p=4$ (rather than the more lenient default of 3), to match the equivalent $k=4$ setting
-used in sc2ts (STAR Methods). Under this setting,
-RIPPLES reported 1174 recombination events of which 601 were designated as recombinant when running 3SEQ on the candidate trios (STAR Methods), giving a concordance rate of 51\%.
-This is notably lower than the equivalent concordance rate of 57\% among the 855 recombination events reported by sc2ts.
+methods, and initially set the RIPPLES ``parsimony improvement'' 
+parameter to $p=4$ (rather than the more lenient default of 3), 
+to match the equivalent $k=4$ setting used in sc2ts (STAR Methods). 
+Under this setting,
+RIPPLES reported 1174 recombination events of which 
+601 were validated by 3SEQ (51\%; STAR Methods).
+% JK: There's no point in making a meal of the 3SEQ validation rate,
+% This is notably lower than the equivalent 
+% concordance rate of 57\% among the 855 recombination events reported by sc2ts.
 Aligning the events reported by both methods using the sets of descendant
@@
 to explain, which is beyond the scope of this work.
-However, the set does contain some well characterized events 
+However, the set does contain some well characterised events 
 associated with the Pango X lineages.
@@
 Over the same time interval, the sc2ts ARG ($k=4$) identifies
-14 high-confidence recombination events with
+14 recombination events passing QC with
 172 descendant samples among 501,235 samples (0.034\%),
 a 40- to 80-fold reduction.
-The sc2ts ARG identifies more of the well-characterized X-lineages
+The sc2ts ARG identifies more of the well-characterised Pango X lineages
 at a more stringent threshold, and therefore provides us with a 
@@
 Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
-have helped spur the development of ARG-aware 
-methods in human 
+have helped spur the development of ARG-aware methods in human 
 genomics~\cite{Mejia2025using,Grundler2025geographic,
@@
 in the ARG. While results are encouraging, 
-dating of phylogenies is a complex task~\cite{Kumar2016advances}
-and methods for incorporating information from metadata of uncertain 
+dating of phylogenies is a complex task~\cite{Kumar2016advances},
+methods for incorporating information from metadata of uncertain 
 quality into large genealogies are in their 
@@
 The secondary contributions of this work arise from 
-the detailed study of the sc2ts ARG, providing insights into the evolution of SARS-CoV-2
+the detailed study of the sc2ts ARG,
+providing insights into the evolution of SARS-CoV-2
 and the COVID-19 pandemic. 
 First, we clarify the relationships among
-recombinant Pango X-lineages, suggesting significant 
+recombinant Pango X lineages, suggesting significant
 clustering of named lineages under shared recombination events.
@@
 the pandemic phase of SARS-CoV-2.
-Together these results underscore the utility of the ARG itself 
+Together, these results underscore the utility of the ARG itself 
 and suggest important directions for future research.
@@
 
-It is important to acknowledge that sampling is non uniform both
+It is important to acknowledge that sampling is non-uniform both
 geographically and over time~\cite{Lancet2021,Chen2022}. 
@@
 the structure holds. Large changes (i.e, saltation events) present
-significant challenges and the results of the algorithm should 
+significant challenges, and the results of the algorithm should 
 be regarded more critically in these situations.
@@
 We converted the Viridian v05 dataset\cite{Hunt2024}, consisting of 4,484,157 SARS-CoV-2
-consensus sequences (125GiB over 48 FASTA files) and 30 metadata fields (1.4
+consensus sequences (125 GiB over 48 FASTA files) and 30 metadata fields (1.4
 GiB TSV file) to VCF Zarr using a 
@@
 
-A classical tradeoff in bioinformatics is whether to store data 
+A classical trade-off in bioinformatics is whether to store data 
 in sample-major (FASTA) or variant-major (VCF) format, determining
@@
 Alignments read from the VCF Zarr file were first preprocessed
-to mask any non nucleotide characters (A,C,G or T) as missing data.
+to mask any non-nucleotide characters (A, C, G, or T) as missing data.
 % https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml
@@
 (Figure~\ref{fig:ls_diagram}).
+% TODO improve wording, lots of "model" here.
 The LS model is an HMM that models a focal genome
@@
 We first compute the Hamming distance between
-all pairs of samples using scipy~\cite{Virtanen2020scipy},
+all pairs of samples using SciPy~\cite{Virtanen2020scipy},
 and then use biotite\cite{Kunzmann2018biotite,Kunzmann2023biotite}
@@
 
-\subsubsection*{Inserting saltation lineages}
+\subsubsection*{Inserting saltational lineages}
 The HMM cost threshold approach outlined in the previous section is 
@@
 ``saltations'' that characterize major VOCs.
-These lineages are characterized by a high number of accumulated mutations,
-and are hypothesized to emerge from chronic infections
+These lineages are characterised by a high number of accumulated mutations,
+and are hypothesised to emerge from chronic infections
 during which the virus is exposed to individual-specific immune selective
@@
 pressures\cite{Otto2021origins,Markov2023}.
-To distinguish time travellers from true saltation lineages,
+To distinguish time travellers from true saltational lineages,
 we re-evaluate the evidence for each lineage with a high HMM cost
@@
 Distinguishing a truly emerging outbreak from different forms of
-correlated error is non-trivial, and sc2ts has 5 parameters to fine
-tune the behaviour. Firstly, we require at least 10 samples in the
+correlated error is non-trivial, and
+sc2ts has 5 parameters to fine-tune the behaviour.
+Firstly, we require at least 10 samples in the
 group to ensure there is sufficient support over a time window of 
-7 days. Then, we build a local tree of the samples and consider
+seven days. Then, we build a local tree of the samples and consider
 some tree metrics to evaluate whether the samples within the 
@@
 Although this process is highly heuristic, it successfully 
-captured the majority of saltation lineages
+captured the majority of saltational lineages
 such as Alpha (Figure~\ref{fig:subgraph_alpha}), and the 
@@
 there is little information for the retrospective sampling grouping mechanism
-to work with and it is very difficult to distinguish short-term time-travellers
+to work with and it is very difficult to distinguish short-term time travellers
 from true outbreak samples. For these cases, we chose specific
@@
 that were used for the designation of these Pango lineages.
-% FIXME more Delta details and some references to back up our assertions
-% There's some specific info in the Viridian paper on this, I'm sure.
-% We should mention long branch splitting here too.]
+%[should we mention long branch splitting here too?]
+% YW: I think we don't need to, actually. That's a subsequent thing, and also discusses in the "further explanation" link
 The resulting subgraph around Delta in the ARG
@@
 
-% FIXME: revise this paragraph to give some background info on Omicron,
-% and reference the Viridian work]
 For Omicron BA.1 and BA.2
@@
 we used SRR17041376 and SRR17461792, respectively.
-We chose SRR17041376 (dated November 12, 2021 from South Africa)
-as the BA.1 seed
-(\url{https://github.com/jeromekelleher/sc2ts-paper/issues/264}),
-as it could be well explained as a non-recombinant
-given the ARG available on its collection date.
+We chose SRR17041376 (dated November 6, 2021 from South Africa)
+for the BA.1 seed, as it is among the earliest African Viridian samples,
+and its initial placement in the ARG mirrors the initial placement of a subsequent high-quality COG-UK sample (ERR7443564).
 For BA.2,
@@
 which is the earliest BA.2 sample present in the Viridian v04 dataset.
-This sample is dated later than SRR17041376 (the BA.1 seed),
+Note that this sample is dated later than SRR17041376 (the BA.1 seed),
 allowing time for the BA.1 lineage
-to accumulate enough sampled descendants and
-to get established in the ARG.
+to become established and accumulate descendant samples in the ARG.
 The resulting subgraph around these Omicron lineages in the ARG
@@
 A similar result holds for the right interval ($I_{right}$),
-but with $P$ replaced with $1–P$.
+but with $P$ replaced with $1-P$.
 Summing the two intervals ($I_{left}+I_{right}$)
@@
 
-\subsubsection*{Analyzing global case counts}
+\subsubsection*{Analysing global case counts}
 
@@
 
-We selected a BA.1 seed sample that led to the fewest mutations.
+We selected a BA.1 seed sample that led to the fewest mutations (see \url{https://github.com/jeromekelleher/sc2ts-paper/issues/264}).
 This strategy resulted in five recurrent mutations
@@
 \label{sec:detailed_ripples}
-
-Using the default ``parsimony improvement'' parameter of $p=3$, UShER+RIPPLES detected a total of 4111 recombination events.
-Most of these subtend $<10$ samples (2616 singletons),
-although four events subtended more than 100,000 samples.
-Analysis of the descending samples for each event shows that
-10 of the 44 Pango X lineages present in the dataset
-% NOTE: listing them like this for sorting and manipulation purposes
-(%
-XA, 
-XB, 
-XC, 
-XL, 
-XW,
-XAC, 
-XAE, 
-XAJ,
-XBB, 
-XBM%
-)
-are associated unequivocally with a single recombination event.
-7 lineages 
-(%
-XE,
-XM,
-XS,
-XY,
-XAM,
-XBD, 
-XBR%
-)
-are associated with two events. Finally, 
-one event subtending 1321 samples suggests that 
-XE, 
-XH, 
-XJ, 
-XM, 
-XY, 
-XAF,
-XAL, 
-and 
-82 BA.2 samples share a recombination event whose parents are BA.2 and BA.1 [FIXME: With what breakpoint and parents?].
-We also ran RIPPLES with $p=4$, which reduced the overall number of
-events to 1174 (roughly 4-fold) but also approximately halved the number of
-events involving X lineages, missing well-characterized events,
-such as XBB (Table~\ref{tab:ripples_x_lineages}).
-
-To compare these results with the sc2ts ARG, we first matched clades
-by finding the most recent common ancestor (MRCA) of the samples
-descending from each RIPPLES event. Overall, clade agreement was
-high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG
-where descendants match exactly. Of these nodes, 68\% had no recombinant
-ancestry. Of the remainder, 595 could be matched with
-one of the 855 recombination events in the sc2ts ARG
-(Figure~\ref{fig:ripples_sc2ts_events}).
-Table~\ref{tab:ripples_x_lineages} shows a summary of
-all recombination events associated with Pango X lineages detected by RIPPLES,
-and their relationship with the sc2ts ARG.
-Of the 16 Class I events (Pango X lineages unequivocally associated with
-a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
-8 are found by RIPPLES (4, at $p=4$).
-Of the 5 Type II events
-(recombination nodes with more complex assocation with X lineages),
-two have close matches in RIPPLES, with XBM matching exactly
-(including 2 descending BF.3 samples),
-and XE/XH is similar (but without including the 2 XH samples in the clade).
-Two events in the RIPPLES set correspond to likely false negatives
-in sc2ts: XB (Document~\ref{sec:dd_xb})
-and XAJ (Document~\ref{sec:dd_xaj}).
-
-Firstly, 408 of these RIPPLES identified recombination events
-have one or more parents that are \emph{younger} than the
-recombinant sequence based on metadata. As sc2ts enforces strict time ordering,
-the parental sequences were simply not available for it to match against
-when the recombinant sequence was inserted. Secondly, of the 179
+We ran RIPPLES and aligned events with the sc2ts ARG as described in STAR
+Methods. We begin by describing the overall composition of events output 
+by RIPPLES at the two parameter values used.
+
+At $p=4$, RIPPLES returned a total of 1174 events, and 
+610 pass 3SEQ validation (51\%). 1052 events
+(89.6\%) were in close clade agreement with the sc2ts ARG,
+and 468 (40\%) corresponded to recombination events in the ARG
+(Figure~\ref{fig:ripples_sc2ts_events}A).
+We investigated the 701 RIPPLES events that could be aligned 
+with the ARG but were not close to a recombination node
+to determine why they were not found by sc2ts.
+For 408 of these RIPPLES identified recombination events,
+one or more parents were \emph{younger} than the
+recombinant sequence based on metadata,
+and therefore impossible under the strict time ordering 
+enforced by sc2ts.
+Secondly, of the 179
 remaining events with matching clades and in which the 
@@
 1,167 mutations, 452 of these occurred at sites that were 
-remapped during post-hoc parsimony. The remaining ``missing''
-events are likely attributable to the finer details of the 
+remapped during post-hoc parsimony (and thus the mutations
+were not observed by the LS HMM during primary inference). 
+The remaining ``missing'' events are likely attributable 
+to the finer details of the 
 differences in tree building approaches between UShER and sc2ts.
 
-[
-FIXME: Discuss the large effect recombinants inferred by both in more detail. E.g.:
-sc2ts identifies two recombination events with \textgreater 10k descendants (specifically, 166,979 and 166,285 descendant samples),
-both of which we identify as spurious (see \ref{sec:spurious-ancestrally-important-recombinants}).
-RIPPLES at $p=4$ also identifies two recombination events with \textgreater 10k descendants (531,224 and 339,523 descendant samples);
-these do not correspond to the same two spurious large-effect events in sc2ts.
-At $p=3$, RIPPLES identifies six recombination events with \textgreater 10k descendant samples, four of which have \textgreater 100k descendant samples.
-]
+At $p=3$, RIPPLES returned a total of 4111 events, and 
+2718 pass 3SEQ validation (66\%). 3755 events
+(91.3\%) were in close clade agreement with the sc2ts ARG,
+and 595 (14\%) corresponded to recombination events in the ARG
+(Figure~\ref{fig:ripples_sc2ts_events}B).
+We did not investigate the $p=3$ events any further as the additional
+events detected at the lower threshold are not expected to be detected
+by sc2ts at the parameter value used ($k=4$).
+
+\begin{table}
+\caption{
+Distribution of numbers of descendants for UShER+RIPPLES events
+and sc2ts. The first column shows the number of descendants (binned
+for larger values) and the others show the corresponding number of events.
+}
+\label{tab:ripples_sc2ts_event_descendants}
+\centering
+\begin{tabular}{lrrr}
+\toprule
+Count & RIPPLES p3 & RIPPLES p4 & sc2ts \\
+\midrule
+1 & 2616 & 757 & 525 \\
+2 & 575 & 178 & 124 \\
+3 & 232 & 65 & 47 \\
+4 & 124 & 38 & 32 \\
+5 & 74 & 14 & 17 \\
+5-100 & 437 & 108 & 96 \\
+100-1000 & 32 & 6 & 8 \\
+1000-10k & 15 & 6 & 4 \\
+10k-100k & 2 & 0 & 0 \\
+100k-1m & 4 & 2 & 2 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Next we considered the overall distribution of the numbers of descendants
+of the events returned by sc2ts and RIPPLES 
+(Table~\ref{tab:ripples_sc2ts_event_descendants}).
+Overall, the distributions were similar, with the majority having 
+very few descendants and being dominated by singletons.
+The events subtending $>$ 100,000 samples
+in all three cases are noteworthy (see
+Document~\ref{sec:spurious-ancestrally-important-recombinants} 
+for analysis of the two sc2ts events).
 
@@
 \toprule
-sc2ts event & in p4 & UShER desc & sc2ts class & clade diff & descendants \\
+sc2ts event & in $p=4$ & UShER desc & sc2ts class & clade diff & descendants \\
 \midrule
-\multicolumn{6}{l}{Equivalent Sc2ts and RIPPLEs recombination events}\\
+\multicolumn{6}{l}{Equivalent Sc2ts and RIPPLES recombination events}\\
+XC & $\checkmark$ & 5 & I & 0 & {XC:5} \\
 XBR & $\checkmark$ & 1 & I & 0 & {XBR:1} \\
 % XM3 & $\checkmark$ & 1 & I & 0 & {XM:1} \\
-XC & $\checkmark$ & 5 & I & 0 & {XC:5} \\
 XS & $\checkmark$ & 17 & I & 0 & {XS:17} \\
+XA & $\checkmark$ & 39 & I & 0 & {XA:39} \\
 XY &  & 23 & I & 0 & {XY:23} \\
 XW &  & 32 & I & 0 & {XW:32} \\
-XA & $\checkmark$ & 39 & I & 0 & {XA:39} \\
 XL &  & 64 & I & 0 & {XL:64} \\
-XBB &  & 6452 & I & 1 & {XBB:2361} \\
-XBM & $\checkmark$ & 12 & II & 0 & {XBM:10, BF.3:2} \\
+XBB &  & 6452 & I & 1 & {XBB:6452} \\
 \midrule
 \multicolumn{6}{l}{RIPPLEs recombination event close to sc2ts recombination event}\\
+XBM & $\checkmark$ & 12 & II & 0 & {XBM:10, BF.3:2} \\
+XM & $\checkmark$ & 8 & II & 0 & {XM:4, BA.2:4} \\
+Xx & $\checkmark$ & 18 & II & 0 & {XAC:18} \\
+Xx &  & 9 & II & 0 & {XAE:9} \\
+XQ &  & 1 & II & 0 & {XAM:1} \\
+XQ &  & 21 & II & 0 & {XAM:21} \\
 XE/XH &  & 1113 & II & 3 & {XE:1113} \\
-XQ &  & 1 & III & 0 & {XAM:1} \\
-XQ &  & 21 & III & 0 & {XAM:21} \\
-Xx &  & 9 & III & 0 & {XAE:9} \\
-Xx & $\checkmark$ & 18 & III & 0 & {XAC:18} \\
-XM & $\checkmark$ & 8 & III & 0 & {BA.2:4, XM:4} \\
 \midrule
 \multicolumn{6}{l}{RIPPLEs recombination event equivalant to sc2ts non-recombination event}\\
-XB & $\checkmark$ & 411 & IV & 0 & {B.1:219, XB:192} \\
+XB & $\checkmark$ & 411 & IV & 0 & {XB:192, B.1:219} \\
 \midrule
@@
 NA & $\checkmark$ & 19 & NA & 45190 & {XAJ:18, BA.4:1} \\
-NA &  & 20 & NA & 1061775 & {BA.1.15:1, XS:17, BA.1:1, BA.1.1:1} \\
 NA & $\checkmark$ & 35 & NA & 7147 & {XBD:30, BA.2.75:5} \\
-NA & $\checkmark$ & 36 & NA & 7146 & {XBD:30, XBR:1, BA.2.75:5} \\
-NA &  & 1321 & NA & 340501 & {XE:1113, XH:2, XJ:68, XY:23, BA.2:82, XM:29,
-XAL:3, XAF:1} \\
+NA & $\checkmark$ & 36 & NA & 7146 & {XBR:1, XBD:30, BA.2.75:5} \\
+NA &  & 20 & NA & 1061775 & {XS:17, BA.1:1, BA.1.1:1, BA.1.15:1} \\
+NA &  & 1321 & NA & 340501 & {XE:1113, XJ:68, BA.2:82, XM:29, XY:23,
+XAL:3, XAF:1, XH:2} \\
 \bottomrule
@@
 
-[FIXME: this is an earlier draft of main text which explains the table in
-detail]
-
-We ran fast-RIPPLES\cite{Smith2023tracking} on the Viridian UShER tree
-subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
-We first ran RIPPLES with parameter $p=4$, as being roughly comparable
-to the recombination penalty used in sc2ts ($k=4$; STAR Methods).
-This resulted in 1,174 recombination events reported, of which 460 could be mapped
-to one of the 855 recombination events in the sc2ts ARG (STAR Methods;
-Figure \ref{fig:ripples_sc2ts_events}A). Analysis of the descendant 
-samples showed that many Pango X lineages were not associated with
-recombination events (including XBB), and we therefore ran
-RIPPLES with the more lenient value of $p=3$ (the default).
-This resulted in a total of 4111 recombination events
-of which 595 have corresponding recombination events in the 
-sc2ts ARG (Figure \ref{fig:ripples_sc2ts_events}B).
-
+We next examined the RIPPLES events referencing Pango X lineages.
 Table~\ref{tab:ripples_x_lineages} shows a summary of all
-recombination events associated with Pango X lineages detected by RIPPLES,
-and their relationship with the sc2ts ARG.
-Of the 16 Class I events (Pango X lineages unequivocally associated with
+recombination events associated with Pango X lineages detected by RIPPLES
+at $p=3$ and their relationship with the sc2ts ARG.
+Of the 16 Class I events (X lineages unequivocally associated with
 a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
@@
 RIPPLES creates two independent recombinations.
-Similarly, XAE and [FIXME] ??? have exactly matching clades, but RIPPLES
+Similarly, XAC and XAE have exactly matching clades, but RIPPLES
 infers two independent recombination events while sc2ts groups them both
@@
 
+See notebook
+\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_ripples.ipynb}
+{analysis\_ripples.ipynb}
+for details.
+
 \subsection{Likely spurious recombinants with many descendants}
@@
 In the sc2ts ARG, there are two recombination events which have a large number of
-descendants but which are not classified as recombinants by Pango. The first of these
+descendants. The first of these
 corresponds to recombination event within BA.2 whose breakpoint is at position 27382
-on the far right of the genome (see supplementary document S3, node 1030562). Although
+on the far right of the genome.
+% JK not sure what this is referring to?
+% (see supplementary document S3, node 1030562). 
+Although
 supported by 5 sites, three of them are adjacent so that they only represent 3 loci,
-meaning that this event does not pass our recombinant quality control measure.
+meaning that this event does not pass our QC filter..
 
 The second of these involves a single breakpoint at position 26858, and is the origin
-of the BA.5 lineage in the ARG (see supplementary document S3, node 1189192). Although
-supported by 5 loci on the right, this recombinant also requires 3 de-novo mutations,
+of the BA.5 lineage in the ARG.
+% (see supplementary document S3, node 1189192). 
+Although supported by 5 loci on the right, this recombinant also requires 3 de-novo mutations,
 some of which are adjacent to the supporting loci. Furthermore, when the first BA.5
@@
 although subsequent tree building does not make this a parsimonious 
-way to rewire the final ARG.
+way to rewire the final ARG.  It is therefore likely to be spurious.
 

### 2025-11-06 latex/main.tex
Updates from Overleaf

@@
 RIPPLES~\cite{Turakhia2022}. 
-% NOTE: not citing RIVET here as there is no real analysis
+% NOTE: not citing RIPPLES here as there is no real analysis
 % in the paper itself, it's just focused on the tool
-RIPPLES works by taking an UShER phylogeny and systematically breaking up
+RIPPLES works by taking a
+pre-existing UShER phylogeny and systematically breaking up
 sequences subtended by long branches into multiple segments 
@@
 
-We compared the performance of sc2ts recombinant detection 
-against RIPPLES run on the Viridian UShER phylogeny,
+We compared the performance of sc2ts recombinant detection against RIPPLES run on the Viridian UShER phylogeny,
 subset to the 2,475,418 samples shared with the ARG (STAR Methods).
 To enable exact comparison, we omitted the QC filtering steps from both
-methods, and initially set the RIPPLES ``parsimony improvement'' 
-parameter to $p=4$ (rather than the more lenient default of 3), 
-to match the equivalent $k=4$ setting used in sc2ts (STAR Methods). 
-Under this setting,
-RIPPLES reported 1174 recombination events of which 
-601 were validated by 3SEQ (51\%; STAR Methods).
-% JK: There's no point in making a meal of the 3SEQ validation rate,
-% This is notably lower than the equivalent 
-% concordance rate of 57\% among the 855 recombination events reported by sc2ts.
+methods, and initially set the RIPPLES ``parsimony improvement'' parameter to $p=4$ (rather than the more lenient default of 3), to match the equivalent $k=4$ setting
+used in sc2ts (STAR Methods). Under this setting,
+RIPPLES reported 1174 recombination events of which 601 were designated as recombinant when running 3SEQ on the candidate trios (STAR Methods), giving a concordance rate of 51\%.
+This is notably lower than the equivalent concordance rate of 57\% among the 855 recombination events reported by sc2ts.
 Aligning the events reported by both methods using the sets of descendant
@@
 Over the same time interval, the sc2ts ARG ($k=4$) identifies
-14 recombination events passing QC with
+14 high-confidence recombination events with
 172 descendant samples among 501,235 samples (0.034\%),
@@
 \label{sec:detailed_ripples}
-We ran RIPPLES and aligned events with the sc2ts ARG as described in STAR
-Methods. We begin by describing the overall composition of events output 
-by RIPPLES at the two parameter values used.
-
-At $p=4$, RIPPLES returned a total of 1174 events, and 
-610 pass 3SEQ validation (51\%). 1052 events
-(89.6\%) were in close clade agreement with the sc2ts ARG,
-and 468 (40\%) corresponded to recombination events in the ARG
-(Figure~\ref{fig:ripples_sc2ts_events}A).
-We investigated the 701 RIPPLES events that could be aligned 
-with the ARG but were not close to a recombination node
-to determine why they were not found by sc2ts.
-For 408 of these RIPPLES identified recombination events,
-one or more parents were \emph{younger} than the
-recombinant sequence based on metadata,
-and therefore impossible under the strict time ordering 
-enforced by sc2ts.
-Secondly, of the 179
+
+Using the default ``parsimony improvement'' parameter of $p=3$, UShER+RIPPLES detected a total of 4111 recombination events.
+Most of these subtend $<10$ samples (2616 singletons),
+although four events subtended more than 100,000 samples.
+Analysis of the descending samples for each event shows that
+10 of the 44 Pango X lineages present in the dataset
+% NOTE: listing them like this for sorting and manipulation purposes
+(%
+XA, 
+XB, 
+XC, 
+XL, 
+XW,
+XAC, 
+XAE, 
+XAJ,
+XBB, 
+XBM%
+)
+are associated unequivocally with a single recombination event.
+7 lineages 
+(%
+XE,
+XM,
+XS,
+XY,
+XAM,
+XBD, 
+XBR%
+)
+are associated with two events. Finally, 
+one event subtending 1321 samples suggests that 
+XE, 
+XH, 
+XJ, 
+XM, 
+XY, 
+XAF,
+XAL, 
+and 
+82 BA.2 samples share a recombination event whose parents are BA.2 and BA.1 [FIXME: With what breakpoint and parents?].
+We also ran RIPPLES with $p=4$, which reduced the overall number of
+events to 1174 (roughly 4-fold) but also approximately halved the number of
+events involving X lineages, missing well-characterized events,
+such as XBB (Table~\ref{tab:ripples_x_lineages}).
+
+To compare these results with the sc2ts ARG, we first matched clades
+by finding the most recent common ancestor (MRCA) of the samples
+descending from each RIPPLES event. Overall, clade agreement was
+high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG
+where descendants match exactly. Of these nodes, 68\% had no recombinant
+ancestry. Of the remainder, 595 could be matched with
+one of the 855 recombination events in the sc2ts ARG
+(Figure~\ref{fig:ripples_sc2ts_events}).
+Table~\ref{tab:ripples_x_lineages} shows a summary of
+all recombination events associated with Pango X lineages detected by RIPPLES,
+and their relationship with the sc2ts ARG.
+Of the 16 Class I events (Pango X lineages unequivocally associated with
+a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
+8 are found by RIPPLES (4, at $p=4$).
+Of the 5 Type II events
+(recombination nodes with more complex assocation with X lineages),
+two have close matches in RIPPLES, with XBM matching exactly
+(including 2 descending BF.3 samples),
+and XE/XH is similar (but without including the 2 XH samples in the clade).
+Two events in the RIPPLES set correspond to likely false negatives
+in sc2ts: XB (Document~\ref{sec:dd_xb})
+and XAJ (Document~\ref{sec:dd_xaj}).
+
+Firstly, 408 of these RIPPLES identified recombination events
+have one or more parents that are \emph{younger} than the
+recombinant sequence based on metadata. As sc2ts enforces strict time ordering,
+the parental sequences were simply not available for it to match against
+when the recombinant sequence was inserted. Secondly, of the 179
 remaining events with matching clades and in which the 
@@
 1,167 mutations, 452 of these occurred at sites that were 
-remapped during post-hoc parsimony (and thus the mutations
-were not observed by the LS HMM during primary inference). 
-The remaining ``missing'' events are likely attributable 
-to the finer details of the 
+remapped during post-hoc parsimony. The remaining ``missing''
+events are likely attributable to the finer details of the 
 differences in tree building approaches between UShER and sc2ts.
 
-At $p=3$, RIPPLES returned a total of 4111 events, and 
-2718 pass 3SEQ validation (66\%). 3755 events
-(91.3\%) were in close clade agreement with the sc2ts ARG,
-and 595 (14\%) corresponded to recombination events in the ARG
-(Figure~\ref{fig:ripples_sc2ts_events}B).
-We did not investigate the $p=3$ events any further as the additional
-events detected at the lower threshold are not expected to be detected
-by sc2ts at the parameter value used ($k=4$).
-
-\begin{table}
-\caption{
-Distribution of numbers of descendants for UShER+RIPPLES events
-and sc2ts. The first column shows the number of descendants (binned
-for larger values) and the others show the corresponding number of events.
-}
-\label{tab:ripples_sc2ts_event_descendants}
-\centering
-\begin{tabular}{lrrr}
-\toprule
-Count & RIPPLES p3 & RIPPLES p4 & sc2ts \\
-\midrule
-1 & 2616 & 757 & 525 \\
-2 & 575 & 178 & 124 \\
-3 & 232 & 65 & 47 \\
-4 & 124 & 38 & 32 \\
-5 & 74 & 14 & 17 \\
-5-100 & 437 & 108 & 96 \\
-100-1000 & 32 & 6 & 8 \\
-1000-10k & 15 & 6 & 4 \\
-10k-100k & 2 & 0 & 0 \\
-100k-1m & 4 & 2 & 2 \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Next we considered the overall distribution of the numbers of descendants
-of the events returned by sc2ts and RIPPLES 
-(Table~\ref{tab:ripples_sc2ts_event_descendants}).
-Overall, the distributions were similar, with the majority having 
-very few descendants and being dominated by singletons.
-The events subtending $>$ 100,000 samples
-in all three cases are noteworthy (see
-Document~\ref{sec:spurious-ancestrally-important-recombinants} 
-for analysis of the two sc2ts events).
+[
+FIXME: Discuss the large effect recombinants inferred by both in more detail. E.g.:
+sc2ts identifies two recombination events with \textgreater 10k descendants (specifically, 166,979 and 166,285 descendant samples),
+both of which we identify as spurious (see \ref{sec:spurious-ancestrally-important-recombinants}).
+RIPPLES at $p=4$ also identifies two recombination events with \textgreater 10k descendants (531,224 and 339,523 descendant samples);
+these do not correspond to the same two spurious large-effect events in sc2ts.
+At $p=3$, RIPPLES identifies six recombination events with \textgreater 10k descendant samples, four of which have \textgreater 100k descendant samples.
+]
 
@@
 \toprule
-sc2ts event & in $p=4$ & UShER desc & sc2ts class & clade diff & descendants \\
+sc2ts event & in p4 & UShER desc & sc2ts class & clade diff & descendants \\
 \midrule
-\multicolumn{6}{l}{Equivalent Sc2ts and RIPPLES recombination events}\\
-XC & $\checkmark$ & 5 & I & 0 & {XC:5} \\
+\multicolumn{6}{l}{Equivalent Sc2ts and RIPPLEs recombination events}\\
 XBR & $\checkmark$ & 1 & I & 0 & {XBR:1} \\
 % XM3 & $\checkmark$ & 1 & I & 0 & {XM:1} \\
+XC & $\checkmark$ & 5 & I & 0 & {XC:5} \\
 XS & $\checkmark$ & 17 & I & 0 & {XS:17} \\
-XA & $\checkmark$ & 39 & I & 0 & {XA:39} \\
 XY &  & 23 & I & 0 & {XY:23} \\
 XW &  & 32 & I & 0 & {XW:32} \\
+XA & $\checkmark$ & 39 & I & 0 & {XA:39} \\
 XL &  & 64 & I & 0 & {XL:64} \\
-XBB &  & 6452 & I & 1 & {XBB:6452} \\
+XBB &  & 6452 & I & 1 & {XBB:2361} \\
+XBM & $\checkmark$ & 12 & II & 0 & {XBM:10, BF.3:2} \\
 \midrule
 \multicolumn{6}{l}{RIPPLEs recombination event close to sc2ts recombination event}\\
-XBM & $\checkmark$ & 12 & II & 0 & {XBM:10, BF.3:2} \\
-XM & $\checkmark$ & 8 & II & 0 & {XM:4, BA.2:4} \\
-Xx & $\checkmark$ & 18 & II & 0 & {XAC:18} \\
-Xx &  & 9 & II & 0 & {XAE:9} \\
-XQ &  & 1 & II & 0 & {XAM:1} \\
-XQ &  & 21 & II & 0 & {XAM:21} \\
 XE/XH &  & 1113 & II & 3 & {XE:1113} \\
+XQ &  & 1 & III & 0 & {XAM:1} \\
+XQ &  & 21 & III & 0 & {XAM:21} \\
+Xx &  & 9 & III & 0 & {XAE:9} \\
+Xx & $\checkmark$ & 18 & III & 0 & {XAC:18} \\
+XM & $\checkmark$ & 8 & III & 0 & {BA.2:4, XM:4} \\
 \midrule
 \multicolumn{6}{l}{RIPPLEs recombination event equivalant to sc2ts non-recombination event}\\
-XB & $\checkmark$ & 411 & IV & 0 & {XB:192, B.1:219} \\
+XB & $\checkmark$ & 411 & IV & 0 & {B.1:219, XB:192} \\
 \midrule
@@
 NA & $\checkmark$ & 19 & NA & 45190 & {XAJ:18, BA.4:1} \\
+NA &  & 20 & NA & 1061775 & {BA.1.15:1, XS:17, BA.1:1, BA.1.1:1} \\
 NA & $\checkmark$ & 35 & NA & 7147 & {XBD:30, BA.2.75:5} \\
-NA & $\checkmark$ & 36 & NA & 7146 & {XBR:1, XBD:30, BA.2.75:5} \\
-NA &  & 20 & NA & 1061775 & {XS:17, BA.1:1, BA.1.1:1, BA.1.15:1} \\
-NA &  & 1321 & NA & 340501 & {XE:1113, XJ:68, BA.2:82, XM:29, XY:23,
-XAL:3, XAF:1, XH:2} \\
+NA & $\checkmark$ & 36 & NA & 7146 & {XBD:30, XBR:1, BA.2.75:5} \\
+NA &  & 1321 & NA & 340501 & {XE:1113, XH:2, XJ:68, XY:23, BA.2:82, XM:29,
+XAL:3, XAF:1} \\
 \bottomrule
@@
 
-We next examined the RIPPLES events referencing Pango X lineages.
+[FIXME: this is an earlier draft of main text which explains the table in
+detail]
+
+We ran fast-RIPPLES\cite{Smith2023tracking} on the Viridian UShER tree
+subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
+We first ran RIPPLES with parameter $p=4$, as being roughly comparable
+to the recombination penalty used in sc2ts ($k=4$; STAR Methods).
+This resulted in 1,174 recombination events reported, of which 460 could be mapped
+to one of the 855 recombination events in the sc2ts ARG (STAR Methods;
+Figure \ref{fig:ripples_sc2ts_events}A). Analysis of the descendant 
+samples showed that many Pango X lineages were not associated with
+recombination events (including XBB), and we therefore ran
+RIPPLES with the more lenient value of $p=3$ (the default).
+This resulted in a total of 4111 recombination events
+of which 595 have corresponding recombination events in the 
+sc2ts ARG (Figure \ref{fig:ripples_sc2ts_events}B).
+
 Table~\ref{tab:ripples_x_lineages} shows a summary of all
-recombination events associated with Pango X lineages detected by RIPPLES
-at $p=3$ and their relationship with the sc2ts ARG.
-Of the 16 Class I events (X lineages unequivocally associated with
+recombination events associated with Pango X lineages detected by RIPPLES,
+and their relationship with the sc2ts ARG.
+Of the 16 Class I events (Pango X lineages unequivocally associated with
 a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
@@
 RIPPLES creates two independent recombinations.
-Similarly, XAC and XAE have exactly matching clades, but RIPPLES
+Similarly, XAE and [FIXME] ??? have exactly matching clades, but RIPPLES
 infers two independent recombination events while sc2ts groups them both
@@
 
-See notebook
-\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_ripples.ipynb}
-{analysis\_ripples.ipynb}
-for details.
-
 \subsection{Likely spurious recombinants with many descendants}
@@
 In the sc2ts ARG, there are two recombination events which have a large number of
-descendants. The first of these
+descendants but which are not classified as recombinants by Pango. The first of these
 corresponds to recombination event within BA.2 whose breakpoint is at position 27382
-on the far right of the genome.
-% JK not sure what this is referring to?
-% (see supplementary document S3, node 1030562). 
-Although
+on the far right of the genome (see supplementary document S3, node 1030562). Although
 supported by 5 sites, three of them are adjacent so that they only represent 3 loci,
-meaning that this event does not pass our QC filter..
+meaning that this event does not pass our recombinant quality control measure.
 
 The second of these involves a single breakpoint at position 26858, and is the origin
-of the BA.5 lineage in the ARG.
-% (see supplementary document S3, node 1189192). 
-Although supported by 5 loci on the right, this recombinant also requires 3 de-novo mutations,
+of the BA.5 lineage in the ARG (see supplementary document S3, node 1189192). Although
+supported by 5 loci on the right, this recombinant also requires 3 de-novo mutations,
 some of which are adjacent to the supporting loci. Furthermore, when the first BA.5
@@
 although subsequent tree building does not make this a parsimonious
-way to rewire the final ARG.  It is therefore likely to be spurious.
+way to rewire the final ARG.
 

### 2025-11-06 latex/main.tex
Fixup refs

--- 2025-11-06 latex/main.tex
@@
 lower bound on detectable recombination.
-This ARG and rich supporting software ecosystem provide
+This ARG and supporting tskit software ecosystem provide
 a strong platform for future research.
@@
 
-Ancestral recombination graphs, population genetics, recombination, tree sequence, SARS-CoV-2
+Ancestral recombination graphs, recombination, SARS-CoV-2, ARG, tskit
 
@@
 % all based on the assumption of a single panmictic population 
+% TODO could phrase this better, Sally didn't like it.
 A subtle but important point distinguishing recent work on ARGs
@@
 genomics~\cite{Mejia2025using,Grundler2025geographic,
-Lehmann2025on,Terhorst2025accelerated}.
+Lehmann2025args,Terhorst2025accelerated}.
 We hope that the ARG provided here will similarly

### 2025-11-07 latex/main.tex
Updates from Sally and Duncan

@@
 
-\author[1,6]{Shing~H.~Zhan}
-\author[1,5]{Yan~Wong}
-\author[2,5]{Anastasia~Ignatieva}
+\author[1,7,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
+\author[1,6,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
+\author[2,6,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
 \author[3]{Katherine~Eaton}
@@
 \author[1]{Benjamin~Jeffery}
-\author[1]{Duncan~S.~Palmer}
-\author[3]{Carmen~Lia~Murall}
-\author[4]{Sarah~P.~Otto}
-\author[1\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}
+\author[1,2,4,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
+\author[3,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
+\author[5,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
+\author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}
 
@@
 \affil[3]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
-\affil[4]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Canada}
-
-% \affil[1]{University A, London SW7 2AZ, UK}
-% \affil[2]{Department B, University B, Toronto, ON M5S 3H6, Canada}
-% \affil[3]{Division C, Department C, University C, Cambridge, MA, USA}
-
-\affil[5]{Joint second author}
-\affil[6]{Lead contact}
+\affil[4]{The Pioneer Centre for SMARTbiomed, Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, United Kingdom}
+\affil[5]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver BC Canada}
+
+\affil[6]{Joint second author}
+\affil[7]{Lead contact}
 
@@
 Estimated genealogies are fundamental to
-understanding this ocean of data, 
+understanding this ocean of data 
 and form the primary input to many downstream analyses.
@@
 the genealogy is a tree.
-However, recombinant lineages have risen to global prevalence 
+However, recombinant lineages have risen to global prevalence, 
 and simple tree representations are therefore 
@@
 as an Ancestral Recombination Graph (ARG) in real time at pandemic scale.
-We infer an ARG for 2.48 million SARS-CoV-2 genomes,
-and show that this deeply validated resource clarifies
-the relationships between recombinant lineages,
-quantifies the rate of recombination over time
-and provides a conservative
+We infer an ARG for 2.48 million SARS-CoV-2 genomes and provide a tskit software ecosystem to support further analyses and visualization.
+This rich and validated ARG resource clarifies
+the relationships among recombinant lineages,
+quantifies the rate of recombination over time,
+and provides a 
 lower bound on detectable recombination.
-This ARG and supporting tskit software ecosystem provide
-a strong platform for future research.
 
@@
 with the recombination penalty $k$
-defining the relative parsimoniousness of
+defining the relative parsimony of
 recombinant versus non-recombinant solutions to the HMM.
@@
 Inference took 21 days on a server with
-64 cores (2X AMD EPYC 7502) and 512 GB of RAM
+64 cores (2$\times$ AMD EPYC 7502) and 512 GB of RAM
 ($\sim$5 CPU years;
@@
 correctly identified many known
-high-frequency deletions (Table~\ref{tab:major_dels}) and
-recurrent deletions (Table~\ref{tab:rec_dels}).
+high-frequency (Table~\ref{tab:major_dels}) and
+recurrent (Table~\ref{tab:rec_dels}) deletions.
 For example,
@@
 can be attributed to the clarity of the rules under which new lineages 
-are designated and their explicitly phylogenetic basis.
+are designated and their explicit phylogenetic basis.
 The first recombinant lineage XA was designated in 2021
@@
 on the putative recombinant sequences,
-summarized in Table~\ref{tab:recombinant_validation}
+summarised in Table~\ref{tab:recombinant_validation}
 (see also Figures~\ref{fig:recombinant_evidence_cov_recomb},
@@
 which increases the chance of co-infection within an individual,
-allowing recombination.
+allowing recombination to occur.
 Second, we would expect recombination events to be detected
@@
 We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020}
-to relate the timing of recombination events with global case numbers.
-Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
+to relate the timing of recombination events with global case numbers. We binned case numbers ($n$) by week from January, 2020 to March, 2023 and
 sub-divided by the proportion in each Scorpio label inferred within the ARG
@@
 on the number of recombination events detected by RIPPLES, and 
-previous studies have used the more lenient (default) value of 
+previous studies have used the default value of 
 $3$\cite{Turakhia2022,Smith2023tracking}.
@@
 
-The secondary contributions of this work arise from 
+Further contributions of this work arise from 
 the detailed study of the sc2ts ARG,
@@
 First, we clarify the relationships among
-recombinant Pango X lineages, suggesting significant
-clustering of named lineages under shared recombination events.
+recombinant Pango X lineages, finding examples of
+clustering of named lineages under a single shared recombination event.
 Second, we used the ARG to study breakpoint intervals 
@@
 % and it supported this research"
+SO acknowledges support from the Natural Sciences and Engineering Research Council of Canada (RGPIN-2022-03726).
 JK acknowledges support from R01 HG012473 from the National Institutes of Health NHGRI, EPSRC (research grant EP/X024881/1) and the Robertson Foundation.
+DSP acknowledges funding administered by the Danish National Research Foundation in support of the Pioneer Centre for SMARTbiomed.
 
@@
 could be due to differences in the alignments used as input to UShER.
-See notebook
+See the notebook
 \href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/validate\_usher\_tree.ipynb}{validate\_usher\_tree.ipynb}
@@
 
-When we compare the intersection of the base sc2ts ARG
-(pre site-remapping), sc2ts is very slightly more parsimonious
-% TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/684
-with [FIXME] XX mutations over the YYY sites shared between them.
+% TODO implement this 
+% When we compare the intersection of the base sc2ts ARG
+% (pre site-remapping), sc2ts is very slightly more parsimonious
+% % TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/684
+% with XX mutations over the YYY sites shared between them.
 After site remapping, and including all mutations involving
@@
 associated with X lineages output by RIPPLES cannot be easily mapped 
-to the sc2ts ARG as one or more descendants is distantly placed.
+to the sc2ts ARG as one or more descendants are distantly placed.
 XAJ is an outlier for sc2ts, being the Pango X lineage identified as a 

### 2025-11-07 latex/main.tex
Abstract tweak

@@
 as an Ancestral Recombination Graph (ARG) in real time at pandemic scale.
-We infer an ARG for 2.48 million SARS-CoV-2 genomes and provide a tskit software ecosystem to support further analyses and visualization.
+We infer an ARG for 2.48 million SARS-CoV-2 genomes 
+which leverages the widely used tskit software ecosystem to support further analyses and 
+visualization.
 This rich and validated ARG resource clarifies
@@
 Classical phylogenetics is largely concerned with inferring the evolutionary history 
-of multiple species~\cite{Felsenstein2004}, using small numbers of widely diverged
+of multiple species~\cite{Felsenstein2004} using small numbers of widely diverged
 genomes. In this context, recombination is a nuisance 
@@
 We show that sc2ts automatically and accurately detects details about
-known recombinant lineages that were painstakingly collated in a community 
-process, and provides novel insights into the relationships between these
+known recombinant lineages that were painstakingly collated through a community effort, and provides novel insights into the relationships between these
 recombinant lineages. We compare the recombination events detected by 
@@
 For each batch, sc2ts first infers likely ``copying paths'' connecting each sample
-to node(s) in the current ARG
+to nodes in the current ARG
 using a simplified version of the Li and Stephens (LS) model~\cite{LiStephens2003}.
@@
 Overall, we also found close similarity between the ARG
-and UShER tree in terms of parsimoniousness 
+and UShER tree in terms of parsimony 
 (Document~\ref{sec:supplement_sc2ts_usher_parsimony})
@@
 There are two primary contributions of this work. The first is an
-Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
+ARG for 2.48 million SARS-CoV-2 genomes.
 Through multiple layers of validation against results from the 
@@
 DSP acknowledges funding administered by the Danish National Research Foundation in support of the Pioneer Centre for SMARTbiomed.
+For the purpose of Open Access, the authors have applied a CC BY public copyright licence to any Author Accepted Manuscript version arising from this submission.
 

### 2025-11-07 latex/main.tex
Visualization->visualisation

@@
 which leverages the widely used tskit software ecosystem to support further analyses and 
-visualization.
+visualisation.
 This rich and validated ARG resource clarifies
@@
 highlighting specific types of QC problems
-(and more details on the visualizations).
+(and more details on the visualisations).
 
@@
 they are likely to attach somewhere along the long branch
-in the subgraph visualization (with $> 30$ mutations).
+in the subgraph visualisation (with $> 30$ mutations).
 If samples from these parental lineages were available for attachment,

### 2025-11-07 latex/main.tex
Updates from Overleaf

@@
 % TODO could phrase this better, Sally didn't like it.
-A subtle but important point distinguishing recent work on ARGs
-from the classical literature is that the ARG 
-``data structure''\cite{Minichiello2006mapping,Wong2024} is no longer
-tightly coupled to the generative
-process\cite{Hudson1983,Griffiths1997ancestral}, substantially 
-broadening the range of potential applications~\cite{Bisschop2025likelihoods}.
-Another significant point is that strong software support for ARGs now
-exists, and the tskit library~\cite{Ralph2020} is the basis of a 
+As the range of applications for ARGs has broadened, the term is no
+longer synonymous with the coalescent with 
+recombination population genetic model \cite{Hudson1983,Griffiths1997ancestral}
+and an ARG is now more usually regarded as a \emph{data structure}
+for representing recombinant ancestry\cite{Minichiello2006mapping,Wong2024}.
+Strong software support for ARGs has facilitated these developments,
+and the tskit library~\cite{Ralph2020} is the basis of a 
 rapidly growing software ecosystem of

### 2025-11-07 latex/main.tex
Updates from Overleaf

@@
 \author[1]{Isobel~Guthrie}
-\author[1]{Benjamin~Jeffery}
+\author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
 \author[1,2,4,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
@@
 visualisation.
-This rich and validated ARG resource clarifies
+This rich and validated resource clarifies
 the relationships among recombinant lineages,
@@
 % all based on the assumption of a single panmictic population 
-% TODO could phrase this better, Sally didn't like it.
 As the range of applications for ARGs has broadened, the term is no
@@
 recombination population genetic model \cite{Hudson1983,Griffiths1997ancestral}
-and an ARG is now more usually regarded as a \emph{data structure}
+and an ARG is now more usually regarded as a data structure
 for representing recombinant ancestry\cite{Minichiello2006mapping,Wong2024}.
@@
 Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
-that incrementally adds samples over time in a similar manner 
+that sequentially adds samples over time in a similar manner 
 to UShER, but also automatically detects and incorporates recombination
@@
 we infer an ARG with 2.48 million whole genome sequences.
-We show that the non-recombinant portions of this ARG are highly congruent with the UShER phylogeny built 
+We show that the non-recombinant portions of this ARG are congruent with the UShER phylogeny built 
 from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020},
@@
 % concordance rate of 57\% among the 855 recombination events reported by sc2ts.
-Aligning the events reported by both methods using the sets of descendant
-samples (STAR Methods), 468 of the events are shared by both,
-with the intersection comprising 40\% of the RIPPLES events
-and 55\% of the sc2ts events (Figure~\ref{fig:ripples_sc2ts_events}). 
+We then compared
+the specific events reported by both methods by aligning them
+using sets of descendant samples (STAR Methods),
+giving 468 events shared between RIPPLES and sc2ts (Figure~\ref{fig:ripples_sc2ts_events}).
 The 706 recombination events output by RIPPLES that do not 
@@
 At this more lenient threshold, 23 of the 44 Pango X lineages
-are present in the dataset, with unequivocal origins for 10 lineages, and
+are identified as descendants of RIPPLES recombinant events,
+with unequivocal origins for 10 lineages, and
 7 lineages associated with two competing events 
@@
 would be required to address data quality issues.
+% NOTE Sally suggested that we say that we did this (run on GISAID) and cite 
+% the earlier preprint, but after discussion it was concluded that
+% this opens a major can of worms and we risk undermining the 
+% perceived novelty of this manuscript to guard against a very
+% minor technical point. If reviewers ask us to do things with
+% GISAID we can by all means explain we've already done it.
 Secondly, the Viridian data is publicly available and redistributable,
@@
 %%%  including grant numbers.
-
-[FIXME: add all acknowledgments]
 

### 2025-11-07 latex/main.tex
Updates from Overleaf

@@
 recombination into pathogen genomics.
+For instance, annotating the ARG with experimentally derived
+fitness estimates for mutations can help elucidate selection
+pressures on difference variants and understand their function;
+analysing the genomic positions of mutations occurring on long
+branches can illuminate the evolution of saltational variants.
 We envisage the ARG as a long-term stable resource---building on the 

### 2025-11-07 latex/main.tex
Updates from Overleaf

--- 2025-11-07 latex/main.tex
@@
 \author[3]{Katherine~Eaton}
-\author[1]{Isobel~Guthrie}
+\author[1,\orcidlink{0000-0002-3547-7226}]{Isobel~Guthrie}
 \author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
@@
 at 29,893 sites,
-and is stored in a 32 MB file.
+and is stored in a 32 MiB file.
 Of the nodes, 855 represent recombination events,
@@
 recombination into pathogen genomics.
+For instance, annotating the ARG with experimentally derived
+fitness estimates for mutations can help elucidate selection
+pressures on different recombinants;
+analysing the genomic positions of mutations occurring on long
+branches (which persist after accounting for recombination) can
+illuminate the evolution of saltational variants.
 We envisage the ARG as a long-term stable resource---building on the 
@@
 (Oxford Population Health, University of Oxford). 
+% JK: Commenting this out until exact wording is provided. I'm pretty
+% sure that "Wellcome trust" is wrong.
+% IG is supported by the Wellcome trust (DPhil in Genomic Medicine and Statistics).
+DSP acknowledges funding administered by the Danish National Research Foundation in support of the Pioneer Centre for SMARTbiomed. 
 % "acknowledges support" here means "was awarded this grant 
@@
 JK acknowledges support from R01 HG012473 from the National Institutes of Health NHGRI, EPSRC (research grant EP/X024881/1) and the Robertson Foundation.
-DSP acknowledges funding administered by the Danish National Research Foundation in support of the Pioneer Centre for SMARTbiomed.
 For the purpose of Open Access, the authors have applied a CC BY public copyright licence to any Author Accepted Manuscript version arising from this submission.
+
+Computation used the Oxford Biomedical Research Computing (BMRC) facility, a joint development between the Wellcome Centre for Human Genetics and the Big Data Institute supported by Health Data Research UK and the NIHR Oxford Biomedical Research Centre.
+The views expressed are those of the author(s) and not necessarily those of the NHS, the NIHR or the Department of Health.
 

### 2025-11-07 latex/main.tex
Updates from Overleaf

@@
 
-\author[1,7,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
-\author[1,6,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
-\author[2,6,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
-\author[3]{Katherine~Eaton}
+\author[1,2,8,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
+\author[1,7,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
+\author[3,7,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
+\author[4]{Katherine~Eaton}
 \author[1,\orcidlink{0000-0002-3547-7226}]{Isobel~Guthrie}
 \author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
-\author[1,2,4,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
-\author[3,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
-\author[5,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
+\author[1,3,5,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
+\author[4,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
+\author[6,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
 \author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}
@@
 Information and Discovery, University of Oxford, United Kingdom}
-\affil[2]{Department of Statistics, University of Oxford, United Kingdom}
-\affil[3]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
-\affil[4]{The Pioneer Centre for SMARTbiomed, Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, United Kingdom}
-\affil[5]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver BC Canada}
-
-\affil[6]{Joint second author}
-\affil[7]{Lead contact}
+\affil[2]{Infectious Disease Epidemiology Unit (IDEU), Nuffield Department of Population Health, University of Oxford, Oxford, United Kingdom}
+\affil[3]{Department of Statistics, University of Oxford, United Kingdom}
+\affil[4]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
+\affil[5]{The Pioneer Centre for SMARTbiomed, Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, United Kingdom}
+\affil[6]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver BC Canada}
+
+\affil[7]{Joint second author}
+\affil[8]{Lead contact}
 
@@
 of the pandemic phase of SARS-CoV-2 in an open and reproducible manner.
-% NOTE: still missing the "What" bit suggested by Jotun, where we 
-% list out some of the stuff that could be done. I'm not convinced
-% this works, though.
 
@@
 the tsdate method~\cite{Wohns2022,Pope2025} to estimate internal node dates
-in the ARG. While results are encouraging, 
-dating of phylogenies is a complex task~\cite{Kumar2016advances},
-methods for incorporating information from metadata of uncertain 
-quality into large genealogies are in their 
-early stages~\cite{sanderson2021chronumental}, and there is certainly
-substantial room for improvement.
-Sc2ts itself may benefit from these node dates, as 
-much of the method's complexity lies in dealing with 
-sample with incorrect sampling dates, and the ARG may provide 
-a means of pre-identifying such ``time travellers''.
+in the ARG, illustrating the reuse of methods initially developed 
+for population genetic applications. While results are encouraging, 
+dating of phylogenies is a complex task~\cite{Kumar2016advances}
+and there is certainly room for improvement.
 
@@
 the pandemic phase of SARS-CoV-2.
-Together, these results underscore the utility of the ARG itself 
-and suggest important directions for future research.
+These results underscore the utility of the ARG 
+and provide a template for future analyses in which 
+% JK: Going out on a limb here, please do change if 
+% we don't like it!!!
+recombination is fully integrated rather than treated
+as a nuisance.
 

### 2025-11-08 latex/main.tex
Sally's updates

@@
 as an Ancestral Recombination Graph (ARG) in real time at pandemic scale.
-We infer an ARG for 2.48 million SARS-CoV-2 genomes 
+We infer an ARG for 2.48 million SARS-CoV-2 genomes, 
 which leverages the widely used tskit software ecosystem to support further analyses and 
@@
 Classical phylogenetics is largely concerned with inferring the evolutionary history 
-of multiple species~\cite{Felsenstein2004} using small numbers of widely diverged
-genomes. In this context, recombination is a nuisance 
-requiring careful consideration~\cite{Rannala2025recombination}
-as it affects the inference of species trees~\cite{Posada2002effect}
-and biases downstream
-analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
-Modern pathogen datasets are not sparsely sampled from different species,
+of divergent species~\cite{Felsenstein2004}. Recombination among lineages, however, distorts the inference of species trees~\cite{Posada2002effect},
+ biases downstream
+analyses~\cite{Schierup2000consequences,Hedge2014bacterial}, and so requires careful consideration~\cite{Rannala2025recombination}.
+Modern pathogen datasets are not sparsely sampled from different species
+% Need to concrete. "large" means 100 to some people.
 but consist of tens to hundreds of thousands of genomes sampled from 
+% JK: deleted point about recomb here as TB, eg, doesn't
 a single species~\cite{
@@
 In contrast to the sophisticated model-based approaches
-used to perform species tree
-inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
-comprehensive phylogenies of these datasets are typically built with
+used to infer species trees~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
+comprehensive phylogenies of pathogen datasets are typically built with
 simpler distance methods % citation for this?
 or approximations~\cite{Price2009fasttree},
-primarily due to computational scaling limitations,
-but also because of the complexity of modelling such datasets.
+primarily due to computational scaling limitations.
 While several methods have been proposed to incorporate
@@
 Software support that explicitly \emph{uses}
-recombination-aware inference is mostly limited to
+recombination-aware inference is mostly restricted to
 visualisation, for example of small-scale phylogenetic
@@
 surveillance data collected during the COVID-19 pandemic
-presents fundamental challenges to existing approaches
-in two ways. Firstly, the sheer amount of data 
+presents fundamental challenges to existing approaches. 
+First, the sheer amount of data 
 overwhelmed even approximate distance matrix-based methods.
 % Accessed GISAID/EpiCov on 08 Oct., 2025
-At the peak in February, 2022 $\sim$80,000 samples per day
+At the peak in February 2022, $\sim$80,000 samples per day
 were submitted to GISAID~\cite{Shu2017},
@@
 of the pandemic response\cite{Hinrichs2024}.
-The second challenge posed is that recombination is an important force
-in SARS-CoV-2 evolution \cite{Jackson2021,VanInsberghe2021,Focosi2022,
+Recurrent recombination has posed an additional challenge to reconstructing the evolutionary history of
+SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,
 Gutierrez2022,Ignatieva2022,Tamura2023},
-with multiple deep recombination events now known to be present
-in the genealogy of the global population \cite{Roemer2023,Shiraz2023enhanced}.
+with multiple deep recombination events impacting both the genealogy and disease attributes of the virus \cite{Roemer2023,Shiraz2023enhanced}.
 Although several methods have been proposed to \emph{detect}
@@
 in innumerable studies~\cite{McBroome2021,Hadfield2018},
-and recombination continues to be treated as a 
-nuisance factor.
+and recombination continues to be treated in a post hoc manner, after phylogenetic reconstruction.
 
@@
 % all based on the assumption of a single panmictic population 
-As the range of applications for ARGs has broadened, the term is no
-longer synonymous with the coalescent with 
-recombination population genetic model \cite{Hudson1983,Griffiths1997ancestral}
-and an ARG is now more usually regarded as a data structure
-for representing recombinant ancestry\cite{Minichiello2006mapping,Wong2024}.
-Strong software support for ARGs has facilitated these developments,
-and the tskit library~\cite{Ralph2020} is the basis of a 
+As the range of applications for ARGs has broadened, the term has itself evolved from a specific model of coalescence with 
+recombination  \cite{Hudson1983,Griffiths1997ancestral}to a rich data structure representing recombinant ancestry\cite{Minichiello2006mapping,Wong2024}.
+The development of extensive software support has facilitated the broader application of ARGS. In particular, the collaboratively built tskit library~\cite{Ralph2020} underpins a 
 rapidly growing software ecosystem of
@@
 that sequentially adds samples over time in a similar manner 
-to UShER, but also automatically detects and incorporates recombination
-in real-time. Based on the Viridian dataset, a freely available  
-resource purged of various systematic errors and artifacts~\cite{Hunt2024},
+to UShER, while automatically detecting and incorporating recombination
+in real-time. Based on the freely-available Viridian dataset, a curated
+% NOTE: Viridian doesn't provide alignments, we do that with MAFFT
+collection of SARS-CoV-2 genomes purged of various systematic errors and artifacts~\cite{Hunt2024},
 we infer an ARG with 2.48 million whole genome sequences.
 We show that the non-recombinant portions of this ARG are congruent with the UShER phylogeny built 
-from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020},
-and that the ARG accurately captures evolutionary features such as mutational 
+from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020}.
+The ARG also accurately captures evolutionary features such as mutational 
 spectra~\cite{Bloom2023} and deletions\cite{Li2023indels}.
 We show that sc2ts automatically and accurately detects details about
-known recombinant lineages that were painstakingly collated through a community effort, and provides novel insights into the relationships between these
+known recombinant lineages that were painstakingly collated through a community effort and provides novel insights into the relationships between these
 recombinant lineages. We compare the recombination events detected by 
@@
 helping us to learn from the vast troves of data collected during the
-pandemic, and to prepare for the next.
+pandemic and to prepare for the next.
 

### 2025-11-08 latex/main.tex
Updates from Overleaf
Update for Zenodo release
Closes #961

@@
 
-\item The sequences alignments in VCF Zarr format have been deposited at Zenodo and are publicly available at \url{https://zenodo.org/records/16314739}.
+\item The ARG in tskit format has been deposited at Zenodo and is publicly
+available at \url{https://zenodo.org/records/17558489}. The UShER tree
+in tskit format is also available, as well as the intersection files.
+\item The sequences alignments in VCF Zarr format have been deposited at 
+Zenodo and are publicly available at \url{https://zenodo.org/records/16314739}.
 \item The Snakemake workflow for aligning and converting the Viridian v05 dataset to VCF Zarr is publicly available at \url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/viridian\_dataset/}.
-\item The ARG in tskit format has been deposited at Zenodo and is publicly available at \url{https://zenodo.org/records/16313750}.
-\item The complete configuration file for running sc2ts to build the ARG above is publicly available at \url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}.
-\item The Snakemake workflow for ARG post-processing is publicly available at \url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/Snakefile}.
-\item The original source code of sc2ts is publicly available on GitHub at \url{https://github.com/jeromekelleher/sc2ts/}.
+\item The complete configuration file for running sc2ts to build the ARG above is 
+publicly available at \url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}.
+\item The Snakemake workflow for ARG post-processing is publicly 
+available at \url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/Snakefile}.
+\item The original code of sc2ts is publicly available on 
+GitHub at \url{https://github.com/jeromekelleher/sc2ts/}.
 \item The Jupyter notebooks and the Mathematica script to perform the analyses in this paper are publicly available on GitHub at \url{https://github.com/jeromekelleher/sc2ts-paper/}.
@@
 %%%  including grant numbers.
+We would like to thank Jotun Hein for helpful comments on the manuscript.
 

### 2025-11-08 latex/main.tex
Updates from Overleaf

@@
 % Need to concrete. "large" means 100 to some people.
-but consist of tens to hundreds of thousands of genomes sampled from 
+% SO: The phrase tens to hundreds of thousands seems odd though given the millions available for SARS-CoV-2: maybe "thousands to millions" 
+% JK: How's that?
+and may contain millions of genomes sampled from 
 % JK: deleted point about recomb here as TB, eg, doesn't
@@
 using Pangolin~\cite{OToole2021assignment} (STAR Methods).
-All but 21,680 of the exported sample alignments (0.87\%) agreed
+Only a small minority of the exported sample alignments (21,680 ; 0.87\%) disagreed
 with the source consensus sequences in terms of Pango lineage assignment
@@
 versions 1.21 and 1.29 in the source Viridian metadata).
-A majority of these lineages (1473 of 2058)
+A majority of the Pango lineages (1473 of 2058)
+% SO: Not "these" as that seemed to refer to the conflicting nodes.
 have monophyletic origins in the ARG,
@@
 by a non-parametric method~\cite{Pope2025} (STAR Methods).
-We show that the dates agree reasonably well with
+We show that the dates agree well with
 Nextstrain estimates (Figure~\ref{fig:pango_vs_nextstrain_node_times})
@@
 \caption{
-Events in the ARG associated with Pango X lineages.
-Each row for Type I, III and IV events (see main text for definitions) 
-corresponds to
-the origin node for a given Pango lineage.
-For Type I and II recombination events,
+%SO: I found the legend hard to parse and revised it. OK?
+Events in the ARG associated with Pango X lineages. Each row corresponds to the origin node for a given Pango lineage (Types I, II, and IV, see text) or associated recombination event (Type II).
+For origin nodes that are recombinant (Type I) or near a recombination event (Type II),
 the ``averted'' column shows the number of additional mutations required
-without recombination (see main text).
+without recombination, as well as the position of informative sites to the left and right of the breakpoint.
 Parent lineages shown in boldface are nodes exactly matched by sampled sequences.
-The descendants column shows the descendant samples broken down by Pango lineage.
-For Type III events, the focal Pango, event (Type I or II) it is derived 
-from and the distance (path length and time in days) from that event are shown,
+The descendants column shows the number of descendants by Pango lineage.
+For Pango lineages (Type III) whose origin node was a descendant of a Type I or II recombination node, the distance (path length and time in days) from the recombination event is shown,
 as well as the number of mutations directly ancestral to the Pango origin
-node. For Type IV events, the Pango designation of the parent node and number
-of mutations are shown.
+node. Pango X lineages not associated with a recombination node (Type IV) are shown with the parent node and number of mutations.
 }
@@
 
-There are 5 partially concordant events (Type II) 
+There are five partially concordant events (Type II) 
 in Table~\ref{tab:pango_x_lineages}, which 
@@
 Other methods are limited in precision by the various proxies 
-used to represent distinct parental \emph{lineages}~\cite{%
+used to represent distinct parental lineages~\cite{%
 Varabyou2021rapid,Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}.
@@
 it may also be due to the higher levels of polymorphism in the Spike gene,
-making recombination events more likely to be detectable, and 
+making recombination events more likely to be detected and 
 allowing breakpoint intervals to be estimated more precisely.

### 2025-11-08 latex/main.tex
Updates from Overleaf

@@
 we note several sources of uncertainty.
-Global case numbers were under reported,
+Global case numbers were underreported,
 particularly during the Delta and Omicron waves~\cite{Wang2025}.
@@
 although similar conclusions were obtained
-when all 855 recombinants were considered (STAR Methods).
+when all 855 recombinants detected by sc2ts were considered (STAR Methods).
 

### 2025-11-08 latex/main.tex
Updates from Sally

@@
 \usepackage[normalem]{ulem}
+
+\newcommand{\red}[1]{\textcolor{red}{#1}}
 
@@
 that ignores recombination, UShER must force placement of recombinant
-sequences in places that minimize the discrepancy across the entire sequence,
+sequences in places that minimise the discrepancy across the entire sequence,
 which may not be near any of the true parents. 
-These permanent misplacements distort the phylogeny~\cite{Schierup2000consequences}
+These permanent misplacements distort the phylogeny~\cite{Schierup2000consequences},
 and subsequent tree reorganisations to maximise overall parsimony~\cite{DeMaio2023} 
-may break up the artefactual long branches used to detect recombination.
-
-We compared the performance of sc2ts recombinant detection 
+may break up the artefactually long branches used to detect recombination.
+
+We compared the performance of sc2ts to detect recombination 
 against RIPPLES run on the Viridian UShER phylogeny,
@@
 giving 468 events shared between RIPPLES and sc2ts (Figure~\ref{fig:ripples_sc2ts_events}).
-The 706 recombination events output by RIPPLES that do not 
+The 706 recombination events detected by RIPPLES that do not 
 correspond to a recombination event in the sc2ts ARG can largely
-be explained by a combination of the choice of sites 
-used for inference, and the strict time ordering imposed
-by sc2ts (Document~\ref{sec:detailed_ripples}).
+be explained by a combination 
+of the strict time ordering imposed by sc2ts
+and the differing sets of problematic sites masked
+by sc2ts and UShER
+(Document~\ref{sec:detailed_ripples}).
 The 387 events detected by sc2ts and not by UShER+RIPPLES
-would require a detailed dissection of the UShER tree
-to explain, which is beyond the scope of this work.
-However, the set does contain some well characterised events 
-associated with the Pango X lineages.
+ include some of the well characterised Pango X lineages.
 Table~\ref{tab:ripples_x_lineages} shows the events detected 
@@
 (contrast with 16 Type I events in Table~\ref{tab:pango_x_lineages}).
-We therefore reran RIPPLES with $p=3$, resulting in 4111 events overall,
-and increasing the intersection of events with sc2ts to 595 (70\%;
-Figure~\ref{fig:ripples_sc2ts_events}).
+We therefore reran RIPPLES with $p=3$, resulting in 4111 events overall
+and increasing the intersection of events with sc2ts to 595 
+(Figure~\ref{fig:ripples_sc2ts_events}).
 At this more lenient threshold, 23 of the 44 Pango X lineages
@@
 (Document~\ref{sec:detailed_ripples}).
-XB and XAJ are among the lineages detected by UShER+RIPPLES,
+Thus, a more lenient threshold ($p=3$) was needed for UShER+RIPPLES to detect a similar fraction of Pango X lineages as sct2ts with $k=4$, resulting in more false positives.
+That said, XB and XAJ are among the lineages detected by UShER+RIPPLES, even at $p=4$,
 strengthening evidence that these are false negatives 
@@
 RIPPLES reported 589 unique recombination events with 43,104 descendant
-samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
+samples---corresponding to $\sim$2.7\% of all samples of recombinant origin.
 Over the same time interval, the sc2ts ARG ($k=4$) identifies
@@
 We hope that the ARG provided here will similarly
-enable a range of recombination-aware analyses of SARS-CoV-2, and 
+enable a range of recombination-aware analyses of SARS-CoV-2 and 
 catalyse a wave of methodological developments integrating
@@
 pressures on different recombinants;
-analysing the genomic positions of mutations occurring on long
-branches (which persist after accounting for recombination) can
-illuminate the evolution of saltational variants.
+saltational variants can also be detected by analysing the rate and position of mutations along long
+branches (which persist after accounting for recombination).
 We envisage the ARG as a long-term stable resource---building on the 
@@
 among all possible recombinant solutions. The current 
-parametrisation---emphasising interpretability---is simplistic,
+parametrisation---emphasising interpretability---is simplistic
 and could readily be extended to account
@@
 is lost when they are ignored, and an important avenue for 
-future work is to explore ways of incorporating them into the HMM.
+future work is to explore ways of incorporating them directly into the HMM.
 Another methodological innovation is the use of 
@@
 for population genetic applications. While results are encouraging, 
-dating of phylogenies is a complex task~\cite{Kumar2016advances}
+dating of phylogenies is a complex task~\cite{Kumar2016advances},
 and there is certainly room for improvement.
@@
 Second, we used the ARG to study breakpoint intervals 
-for recombination events, adding a novel temporal dimension
-and demonstrating a simple relationship with the 
-size of the interval in base pairs derived from theory.
+for recombination events, showing that the relationship between the 
+size of the interval in base pairs and the divergence time of the parental lineages matches theoretical expectations.
 Thirdly, we demonstrate a clear relationship between
@@
 the pandemic phase of SARS-CoV-2.
-These results underscore the utility of the ARG 
-and provide a template for future analyses in which 
-% JK: Going out on a limb here, please do change if 
-% we don't like it!!!
-recombination is fully integrated rather than treated
-as a nuisance.
+These results underscore the utility of the ARG, and 
+the efficient and reproducible pipelines throughout this work 
+provide a template for future analyses in which recombination
+is accounted for by default.
 
@@
 based on information from COG-UK, GISAID, and ENA/SRA as our source.
-Samples which have missing or incomplete dates (e.g., 2020-01) were omitted,
-as were the $\sim$40,000 samples
-which had a collection date of December 31, 2020.
-As the sample count for this date is a major outlier
-for this period of the pandemic,
-it seems likely that this batch is highly enriched
+Samples that have missing or incomplete dates (e.g., 2020-01) were omitted. 
+Samples with a collection date of December 31, 2020
+were also omitted,
+as the number of sample on this date ($\sim$40,000) was a major outlier
+for this period of the pandemic and is likely enriched
 for incorrect collection dates.
 Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
-confirmed that some of the samples were not collected in 2020.
+confirmed that some of these samples were not collected in 2020.
 
@@
 % https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml
-Then, we filtered any samples that had missing data at more than 500 sites,
-excluding the 100 sites identified as problematic
-(this was set to 10,000 sites for the initial period up to 2020-03-01
-where sampling density is low).
-
+Then, excluding the 100 sites identified as problematic, we filtered any samples that had missing data at more than 500 sites (this was set to 10,000 sites for the initial period up to 2020-03-01
+when sampling density was low).
+% SO: Moved clause about 100 sites to the beginning so that it was clearer that 10,000 is in comparison to 500 not 100.
 \subsubsection*{Li and Stephens model}
@@
 
-We use the Viterbi algorithm implemented in tsinfer~\cite{Kelleher2019}
-to efficiently find copying paths under this model. We updated this 
+We use the efficient Viterbi algorithm
+(which finds the most likely sequence of hidden states in a HMM) 
+implemented in tsinfer~\cite{Kelleher2019}
+to find copying paths under this model. We updated this 
 implementation to include a ``likelihood threshold'', which allows 
 all possible paths through the HMM with likelihood less than a given
-value to be considered equally probable. This approach leads to major
+value to be considered equally improbable. This approach leads to major
 gains in likelihood compression, saving memory and reducing computation
@@
 inference by ``pushing in'' a new node which descends from the
-original parent, and carries all its mutations except those
+original parent and carries all its mutations except those
 causing the reversions on the newly added node.
-52,336 nodes were added to the ARG during primary inference by
-reversion pushing.
+Reversion pushing led to the addition of 52,336 nodes during primary inference.
 
@@
 substantially from their true collection date---pose significant problems
-for sc2ts, and are present at a significant frequency. For example, of the 
+for sc2ts and are present at a significant frequency. For example, of the 
 43 Scorpio~\cite{Colquhoun2023scorpio}
@@
 arise until late 2022. Inserting these time-travellers into the ARG 
-at the claimed sampling date causes major issues, and they must therefore be 
+at the claimed sampling date causes major artefacts, and they must therefore be 
 filtered out. We use the simple approach of filtering out samples that 
@@
 All the XL samples descend from a well supported recombination node,
-with three child samples which are identical to it.
+with three child samples, which are identical to it.
 The left parent is a sample (ERR7989424), and
@@
 assigned to these Pango X lineages,
-the number of samples which passed sequence-level quality control, and
+the number of samples that passed sequence-level quality control, and
 the number of recombination breakpoints

### 2025-11-09 latex/main.tex
Updates from Overleaf

@@
 
-\newcommand{\red}[1]{\textcolor{red}{#1}}
-
 \graphicspath{ {../figures/} }
@@
 \author[3,7,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
-\author[4]{Katherine~Eaton}
+\author[4,\orcidlink{0000-0001-6862-7756}]{Katherine~Eaton}
 \author[1,\orcidlink{0000-0002-3547-7226}]{Isobel~Guthrie}
@@
 Only a small minority of the exported sample alignments (21,680 ; 0.87\%) disagreed
+% NOTE Sally pointed out as potentially needing clarification.
 with the source consensus sequences in terms of Pango lineage assignment
@@
 the pandemic phase of SARS-CoV-2.
-These results underscore the utility of the ARG, and 
-the efficient and reproducible pipelines throughout this work 
-provide a template for future analyses in which recombination
-is accounted for by default.
+These results underscore the utility of the ARG 
+and provide a template for future analyses.
 
@@
 % SO: Moved clause about 100 sites to the beginning so that it was clearer that 10,000 is in comparison to 500 not 100.
+
 \subsubsection*{Li and Stephens model}

### 2025-11-09 latex/main.tex
Fix reference

@@
 efficient 
-algorithms~\cite{Kelleher2016,Ralph2020,Kelleher2020coalescent,Grundler2025geographic,Lehmann2025on}.
+algorithms~\cite{Kelleher2016,Ralph2020,Kelleher2020coalescent,Grundler2025geographic,Lehmann2025args}.
 The mature and feature-rich tskit library has interfaces in Python, C and Rust,

### 2025-11-09 latex/main.tex
duplicate PDF cross refs

@@
 \usepackage{authblk}
-\usepackage{hyperref}
+\usepackage[hypertexnames=false]{hyperref}
 \usepackage{booktabs}
@@
 RIPPLES~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
-detecting more of the well-characterized recombination events 
+detecting more of the well-characterised recombination events 
 curated by the Pango lineage designation community.

### 2025-11-11 src/run_lineage_imputation.py
Update unused imputation script

@@ -10,7 +10,7 @@ import sys
 import time
 from collections import defaultdict
 
-import sc2ts
+import sc2ts.debug as sd
 import tszip
 import sklearn.tree
 import numpy as np
@@ -723,7 +723,7 @@ if __name__ == "__main__":
     args = argparser.parse_args()
 
     ts = tszip.load(args.input_ts)
-    ti = sc2ts.info.TreeInfo(ts)
+    ti = sd.ArgInfo(ts)
 
 
     new_ts = lineage_imputation(

### 2025-11-18 latex/main.tex

@@
 
-% %%%  The resource availability section is required
-% %%%  for all research articles. This component
-% %%%  has 3 subsections: "lead contact," "materials
-% %%%  availability," and "data and code availability."
-% %%%  All 3 subsections must be included, even if no
-% %%%  unique materials were generated in the study.
-% %%%  Do not edit or change the names of the subheadings.
-% %%%  No other subheadings or text are allowed in the
-% %%%  resource availability section.
-
-% \subsection*{Lead contact}
-
-% %%%  Authors are required to designate a lead contact,
-% %%%  who will be responsible for communication with
-% %%%  the journal before and after publication and is
-% %%%  the arbiter of disputes, including concerns
-% %%%  related to reagents or resource sharing. Only
-% %%%  one author can be named the lead contact, and
-% %%%  only the lead contact’s information may be
-% %%%  provided in this section.
-
-% Requests for further information and resources
-% should be directed to and will be fulfilled
-% by the lead contact, XX (XX@university.edu).
-
-% \subsection*{Materials availability}
-
-% %%%  This subsection must include a statement describing
-% %%%  the availability of newly generated materials
-% %%%  associated with the paper, including any conditions
-% %%%  for access. If there are no newly generated materials
-% %%%  associated with the paper, the statement should
-% %%%  state this, e.g.: This study did not generate new
-% %%%  materials.
-
-% This study did not generate new materials.
+%%%  The resource availability section is required
+%%%  for all research articles. This component
+%%%  has 3 subsections: "lead contact," "materials
+%%%  availability," and "data and code availability."
+%%%  All 3 subsections must be included, even if no
+%%%  unique materials were generated in the study.
+%%%  Do not edit or change the names of the subheadings.
+%%%  No other subheadings or text are allowed in the
+%%%  resource availability section.
+
+\subsection*{Lead contact}
+
+%%%  Authors are required to designate a lead contact,
+%%%  who will be responsible for communication with
+%%%  the journal before and after publication and is
+%%%  the arbiter of disputes, including concerns
+%%%  related to reagents or resource sharing. Only
+%%%  one author can be named the lead contact, and
+%%%  only the lead contact’s information may be
+%%%  provided in this section.
+
+Requests for further information and resources
+should be directed to and will be fulfilled
+by the lead contact, Shing H. Zhan (shing.zhan@ndph.ox.ac.uk).
+
+\subsection*{Materials availability}
+
+%%%  This subsection must include a statement describing
+%%%  the availability of newly generated materials
+%%%  associated with the paper, including any conditions
+%%%  for access. If there are no newly generated materials
+%%%  associated with the paper, the statement should
+%%%  state this, e.g.: This study did not generate new
+%%%  materials.
+
+This study did not generate new materials.
 
@@
 
-SHZ is supported by an NDPH Intermediate Research Fellowship
-(Oxford Population Health, University of Oxford). 
+SHZ acknowledges support from
+Oxford Population Health, University of Oxford
+(NDPH Intermediate Research Fellowship)
+and
+% Apparently, it is now Johnson & Johnson Innovative Medicine.
+Janssen Pharmaceuticals
+(Oxford-Janssen Translational Genomics Fellowship).
 % JK: Commenting this out until exact wording is provided. I'm pretty
@@
 
-% \section*{AUTHOR CONTRIBUTIONS}
-
-% %%%  This component is required for most research papers.
-% %%%  Mention each individual author with a statement
-% %%%  outlining the contribution of each author to the work.
-
-% Conceptualization, S.C.P. and S.Y.W.;
-% methodology, A.B., S.C.P., and S.Y.W.;
-% investigation, M.E., A.N.V., N.A.V., S.C.P., and S.Y.W.;
-% writing-–original draft, S.C.P. and S.Y.W.;
-% writing-–review \& editing, S.C.P. and S.Y.W.;
-% funding acquisition, S.C.P. and S.Y.W.;
-% resources, M.E.V and C.K.B.;
-% supervision, A.B., N.L.W., and A.A.D.
-
-% \section*{DECLARATION OF INTERESTS}
-
-% %%%  This component is required for all articles, even
-% %%%  if the authors have no competing interests; if
-% %%%  this is the case, insert "The authors declare no
-% %%%  competing interests." Please refer to the
-% %%%  declaration of interests policy:
-% %%%  https://www.cell.com/declaration-of-interests
-
-% % TODO: Close https://github.com/jeromekelleher/sc2ts-paper/issues/922
-% The authors declare no competing interests.
-
-% \section*{DECLARATION OF GENERATIVE AI AND AI-ASSISTED TECHNOLOGIES}
-
-% %%%  If generative AI and AI-assisted technologies
-% %%%  were used in the writing process, this must
-% %%%  be disclosed in the paper. This declaration
-% %%%  does not apply to the use of basic tools for
-% %%%  checking grammar, spelling, references, etc.
-% %%%  If you have nothing to disclose, please do not
-% %%%  include this component.
-
-% During the preparation of this work, the author(s) used ChatGPT
-% in order to obtain editorial input on the text.
-% After using this tool or service, the author(s) reviewed
-% and edited the content as needed and take(s) full responsibility for the content of the publication.
+\section*{AUTHOR CONTRIBUTIONS}
+
+%%%  This component is required for most research papers.
+%%%  Mention each individual author with a statement
+%%%  outlining the contribution of each author to the work.
+
+% Credit system: https://credit.niso.org/
+Conceptualization, SHZ, SPO, JK;
+methodology, SHZ, YW, AI, KE, BJ, DSP, SPO, JK;
+investigation, SHZ, YW, AI, KE, IG, SPO, JK;
+writing--original draft, SHZ, JK;
+writing--review \& editing, everyone;
+funding acquisition, JK;
+resources, JK;
+supervision, JK.
+
+\section*{DECLARATION OF INTERESTS}
+
+%%%  This component is required for all articles, even
+%%%  if the authors have no competing interests; if
+%%%  this is the case, insert "The authors declare no
+%%%  competing interests." Please refer to the
+%%%  declaration of interests policy:
+%%%  https://www.cell.com/declaration-of-interests
+
+% TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/922
+The authors declare no competing interests.
+
+\section*{DECLARATION OF GENERATIVE AI AND AI-ASSISTED TECHNOLOGIES}
+
+%%%  If generative AI and AI-assisted technologies
+%%%  were used in the writing process, this must
+%%%  be disclosed in the paper. This declaration
+%%%  does not apply to the use of basic tools for
+%%%  checking grammar, spelling, references, etc.
+%%%  If you have nothing to disclose, please do not
+%%%  include this component.
+
+% https://github.com/jeromekelleher/sc2ts-paper/issues/923
+During the preparation of this work, the author(s) used ChatGPT
+in order to obtain editorial input on the text.
+After using this tool or service, the author(s) reviewed
+and edited the content as needed and
+take(s) full responsibility for the content of the publication.
 

### 2025-11-19 latex/main.tex
Tweaks for Acks

@@
 (Oxford-Janssen Translational Genomics Fellowship).
-% JK: Commenting this out until exact wording is provided. I'm pretty
-% sure that "Wellcome trust" is wrong.
-% IG is supported by the Wellcome trust (DPhil in Genomic Medicine and Statistics).
-DSP acknowledges funding administered by the Danish National Research Foundation in support of the Pioneer Centre for SMARTbiomed. 
+IG is supported by Wellcome (DPhil in Genomic Medicine and Statistics, 228319/Z/23/Z).
+DSP acknowledges funding administered by the Danish National Research Foundation
+in support of the Pioneer Centre for SMARTbiomed.
 % "acknowledges support" here means "was awarded this grant 
@@
 Conceptualization, SHZ, SPO, JK;
+% Methods dev, software dev, and approaches
 methodology, SHZ, YW, AI, KE, BJ, DSP, SPO, JK;
+% Data analysis and literature review/synthesis
 investigation, SHZ, YW, AI, KE, IG, SPO, JK;

### 2025-11-19 latex/main.tex
Tweak the ACKs

@@
 
-SHZ acknowledges support from
-Oxford Population Health, University of Oxford
-(NDPH Intermediate Research Fellowship)
-and
-% Apparently, it is now Johnson & Johnson Innovative Medicine.
-Janssen Pharmaceuticals
-(Oxford-Janssen Translational Genomics Fellowship).
+SHZ is supported by an NDPH Intermediate Research Fellowship (Oxford Population
+Health, University of Oxford).
 IG is supported by Wellcome (DPhil in Genomic Medicine and Statistics, 228319/Z/23/Z).
@@
 in support of the Pioneer Centre for SMARTbiomed.
-% "acknowledges support" here means "was awarded this grant 
-% and it supported this research"
 SO acknowledges support from the Natural Sciences and Engineering Research Council of Canada (RGPIN-2022-03726).
-JK acknowledges support from R01 HG012473 from the National Institutes of Health NHGRI, EPSRC (research grant EP/X024881/1) and the Robertson Foundation.
+JK acknowledges support from the
+Oxford-Janssen Translational Genomics Fellowship,
+research grant R01 HG012473 from the National Institutes of Health NHGRI, 
+research grant EP/X024881/1 from the Engineering and Physical Sciences Research Council, 
+and the Robertson Foundation. 
 For the purpose of Open Access, the authors have applied a CC BY public copyright licence to any Author Accepted Manuscript version arising from this submission.
 
-Computation used the Oxford Biomedical Research Computing (BMRC) facility, a joint development between the Wellcome Centre for Human Genetics and the Big Data Institute supported by Health Data Research UK and the NIHR Oxford Biomedical Research Centre.
-The views expressed are those of the author(s) and not necessarily those of the NHS, the NIHR or the Department of Health.
+Computation used the Oxford Biomedical Research Computing (BMRC) facility, a
+joint development between the Wellcome Centre for Human Genetics and the Big
+Data Institute supported by Health Data Research UK and the NIHR Oxford
+Biomedical Research Centre. The views expressed are those of the author(s) and
+not necessarily those of the NHS, the NIHR or the Department of Health.
 

### 2025-11-21 latex/main.tex
Updates from Shing

@@
 \item The original code of sc2ts is publicly available on 
-GitHub at \url{https://github.com/jeromekelleher/sc2ts/}.
+GitHub at \url{https://github.com/tskit-dev/sc2ts}.
 \item The Jupyter notebooks and the Mathematica script to perform the analyses in this paper are publicly available on GitHub at \url{https://github.com/jeromekelleher/sc2ts-paper/}.
@@
 Conceptualization, SHZ, SPO, JK;
-% Methods dev, software dev, and approaches
+% Methodology, software, resources, visualization
 methodology, SHZ, YW, AI, KE, BJ, DSP, SPO, JK;
-% Data analysis and literature review/synthesis
+% Invesigation, data curation, formal analysis, validation
 investigation, SHZ, YW, AI, KE, IG, SPO, JK;
@@
 writing--review \& editing, everyone;
-funding acquisition, JK;
 resources, JK;
-supervision, JK.
+supervision, JK;
+funding acquisition, JK.
 
@@
 
-% TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/922
+% https://github.com/jeromekelleher/sc2ts-paper/issues/922
 The authors declare no competing interests.
@@
 % (pre site-remapping), sc2ts is very slightly more parsimonious
-% % TODO: https://github.com/jeromekelleher/sc2ts-paper/issues/684
+% See https://github.com/jeromekelleher/sc2ts-paper/issues/684
 % with XX mutations over the YYY sites shared between them.

### 2025-11-23 latex/main.tex
Final changes pre submission

@@
 writing--original draft, SHZ, JK;
-writing--review \& editing, everyone;
+writing--review \& editing, SHZ, YW, AI, KE, IG, BJ, DSP, CLM, SPO, JK;
 resources, JK;

### 2025-12-20 latex/main.tex
Updates from Overleaf
Basic compiling version of document

@@
-% Notes on format:
-% The total word count of an article must be under 7,000 (including main figure
-% legends but excluding STAR Methods text, supplemental item legends, and
-% references section), and there should be no more than seven figures and/or
-% tables.
-
-\documentclass[12pt,letterpaper]{article}
-\usepackage[a4paper, total={7in, 10in}]{geometry}
-\renewcommand{\familydefault}{\sfdefault}
+% Use the basic LaTeX article class, 12pt text
+\documentclass[12pt]{article}
+
+% Science uses Times font. If you don't have this installed (most LaTeX installations will be
+% fine) or prefer the old Computer Modern fonts, comment out the following line
+% \usepackage{newtxtext,newtxmath}
+% Depending on your LaTeX fonts installation, you might get better results with one or both of these:
+%\usepackage{mathptmx}
+%\usepackage{txfonts}
+
+% Allow external graphics files
 \usepackage{graphicx}
-\usepackage{helvet}
-\usepackage{authblk}
-\usepackage[hypertexnames=false]{hyperref}
+
+% Use US letter sized paper with 1 inch margins
+\usepackage[letterpaper,margin=1in]{geometry}
+
+% Double line spacing, including in captions
+\linespread{1.5} % For some reason double spacing is 1.5, not 2.0!
+
+% One space after each sentence
+\frenchspacing
+
+% Abstract formatting and spacing - no heading
+\renewenvironment{abstract}
+	{\quotation}
+	{\endquotation}
+
+% No date in the title section
+\date{}
+
+% Reference section heading
+\renewcommand\refname{References and Notes}
+
+% Figure and Table labels in bold
+\makeatletter
+\renewcommand{\fnum@figure}{\textbf{Figure \thefigure}}
+\renewcommand{\fnum@table}{\textbf{Table \thetable}}
+\makeatother
+
+% Call the accompanying scicite.sty package.
+% This formats citation numbers in Science style.
+\usepackage{scicite}
+
+% Provides the \url command, and fixes a crash if URLs or DOIs contain underscores
+\usepackage{url}
+
+%%%%%%%%%%%% CUSTOM COMMANDS AND PACKAGES %%%%%%%%%%%%
+
 \usepackage{booktabs}
 \usepackage{tabularx}
-\usepackage{amsmath}
 \usepackage{amssymb}
-\usepackage{placeins}
-\usepackage{orcidlink}
-\usepackage[super,comma,sort&compress]
-   {natbib}\bibliographystyle{numbered}
-\usepackage[right]{lineno} \linenumbers
-\usepackage[normalem]{ulem}
-
-\graphicspath{ {../figures/} }
-
-\makeatletter
-\renewcommand{\maketitle}{\bgroup\setlength{\parindent}{0pt}
-\begin{flushleft}
-  \textbf{\@title}
-
-  \@author
-\end{flushleft}\egroup}
-\makeatother
-
-%%%  Insert title below; leave date empty.
-
-\title{A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2}
+\usepackage[hypertexnames=false]{hyperref}
+
+\graphicspath{{../figures}}
+
+% Authors can define simple custom commands e.g. as shortcuts to save on typing
+% Use \newcommand (not \def) to avoid overwriting existing commands.
+% Keep them as simple as possible and note the warning in the text below.
+% Example:
+\newcommand{\pcc}{\,cm$^{-3}$}	% per cm-cubed
+
+% Please DO NOT import additional external packages or .sty files.
+% Those are unlikely to work with our conversion software and will cause problems later.
+% Don't add any more \usepackage{} commands.
+
+%%%%%%%%%%%%%%%% TITLE AND AUTHORS %%%%%%%%%%%%%%%%
+
+% Title of the paper.
+% Keep it short and understandable by any reader of Science.
+% Avoid acronyms or jargon. Use sentence case.
+\def\scititle{
+A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2
+}
+% Store the title in a variable for reuse in the supplement (otherwise \maketitle deletes it)
+\title{\bfseries \boldmath \scititle}
+
+% Author and institution list.
+% Institution numbers etc. should be hard-coded, do *not* use the \footnote command.
+\author{
+	% You can write out first names or use initials - either way is acceptable, but be consistent
+	First~Author$^{1\ast\dagger}$,
+	A.~Scientist$^{2\dagger}$,
+	Someone~E.~Else$^{2}$\and
+	% Additional lines of authors should be inserted using the \and command (not \\)
+	% Institution list, in a slightly smaller font
+	\small$^{1}$Department, Institution, City \& Postal Code, Country.\and
+	\small$^{2}$Another Department, Different Institution, City \& Postal Code, Country.\and
+	% Identify at least one corresponding author, with contact email address
+	\small$^\ast$Corresponding author. Email: example@mail.com\and
+	% Joint contributions can be indicated like this
+	\small$^\dagger$These authors contributed equally to this work.
+}
+
 \date{}
 
-%%%  Author first and last names should be spelled
-%%%  out in their entirety (do not abbreviate "J.H.
-%%%  Watson" unless this is how the author's name
-%%%  always appears). Middle initials are OK. Do
-%%%  not include titles, positions, or degrees.
-
-%%%  Use numbered footnotes to indicate institutional
-%%%  affiliations. Authors may have multiple
-%%%  institutional affiliations, and affiliations
-%%%  may be shared among multiple authors.
-
-%%%  After the institutional affiliations, numbered
-%%%  footnotes may be used to indicate an author's
-%%%  present address, equal contribution status,
-%%%  and/or senior author status.
-
-%%%  The final numbered footnote should indicate
-%%%  which author is the lead contact (required).
-%%%  One author must be designated as the lead contact.
-%%%  There can be no more than one lead contact.
-
-%%%  Corresponding authors should be indicated with
-%%%  asterisks (*). Use 2 asterisks (**) for the
-%%%  second-listed corresponding author, 3 (***) for
-%%%  the third-listed, and so on. The lead contact
-%%%  must be a corresponding author. Additional
-%%%  authors may also serve as corresponding authors.
-
-% \author[1,\orcidlink{0000-0000-0000-0000}]{First Last}
-% \author[1,2]{First Middle Last}
-% \author[2,3]{First M.M. Last}
-% \author[3,*]{First Last, Jr.}
-% \author[3,4,5,**]{F. Middle Last}
-
-\author[1,2,8,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
-\author[1,7,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
-\author[3,7,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
-\author[4,\orcidlink{0000-0001-6862-7756}]{Katherine~Eaton}
-\author[1,\orcidlink{0000-0002-3547-7226}]{Isobel~Guthrie}
-\author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
-\author[1,3,5,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
-\author[4,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
-\author[6,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
-\author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}
-
-%%%  Institutional affiliations should contain the
-%%%  following information at minimum: department(s)/
-%%%  subunit(s), institution, city, state/region (if
-%%%  applicable), and country.
-
-\affil[1]{Big Data Institute, Li Ka Shing Centre for Health
-Information and Discovery, University of Oxford, United Kingdom}
-\affil[2]{Infectious Disease Epidemiology Unit (IDEU), Nuffield Department of Population Health, University of Oxford, Oxford, United Kingdom}
-\affil[3]{Department of Statistics, University of Oxford, United Kingdom}
-\affil[4]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
-\affil[5]{The Pioneer Centre for SMARTbiomed, Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, United Kingdom}
-\affil[6]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver BC Canada}
-
-\affil[7]{Joint second author}
-\affil[8]{Lead contact}
-
-%%%  List only one email address per corresponding author.
-
-\affil[*]{Correspondence: shing.zhan@ndph.ox.ac.uk}
-\affil[**]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
+% \author[1,2,8,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
+% \author[1,7,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
+% \author[3,7,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
+% \author[4,\orcidlink{0000-0001-6862-7756}]{Katherine~Eaton}
+% \author[1,\orcidlink{0000-0002-3547-7226}]{Isobel~Guthrie}
+% \author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
+% \author[1,3,5,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
+% \author[4,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
+% \author[6,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
+% \author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}
+
+% %%%  Institutional affiliations should contain the
+% %%%  following information at minimum: department(s)/
+% %%%  subunit(s), institution, city, state/region (if
+% %%%  applicable), and country.
+
+% \affil[1]{Big Data Institute, Li Ka Shing Centre for Health
+% Information and Discovery, University of Oxford, United Kingdom}
+% \affil[2]{Infectious Disease Epidemiology Unit (IDEU), Nuffield Department of Population Health, University of Oxford, Oxford, United Kingdom}
+% \affil[3]{Department of Statistics, University of Oxford, United Kingdom}
+% \affil[4]{National Microbiology Laboratory, Public Health Agency of Canada, Canada}
+% \affil[5]{The Pioneer Centre for SMARTbiomed, Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, United Kingdom}
+% \affil[6]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver BC Canada}
+
+% \affil[7]{Joint second author}
+% \affil[8]{Lead contact}
+
+% %%%  List only one email address per corresponding author.
+
+% \affil[*]{Correspondence: shing.zhan@ndph.ox.ac.uk}
+% \affil[**]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
 
@@
 simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019,Gopalan2025bridging}
-and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
+and inference~\cite{Rasmussen2014,Speidel2019,Kelleher2019,
 Wohns2022,Zhang2023,Gunnarsson2024,Deng2025robust} have led to a surge 
@@
 \begin{itemize}
-
 \item The ARG in tskit format has been deposited at Zenodo and is publicly
@@
 % Credit system: https://credit.niso.org/
-Conceptualization, SHZ, SPO, JK;
+Conceptualization, SHZ, CLM, SPO, JK;
 % Methodology, software, resources, visualization
@@
 
-% \section*{SUPPLEMENTAL INFORMATION INDEX}
-
-% %%%  Supplemental information must be uploaded as
-% %%%  separate files. In the main text, please list the
-% %%%  files to be included in a brief index. For details,
-% %%%  please review the supplemental information guidelines:
-
-% %%%  Journals with STAR Methods:
-% %%%  https://www.cell.com/STAR-supplemental-information
-
-% \begin{description}
-%   \item Figures S1-SX and their legends in a PDF
-%   \item Table S1. A descriptive title for an Excel file that was too large to appear in the PDF
-% \end{description}
-
-% \newpage
-
-% \section*{MAIN FIGURE TITLES AND LEGENDS}
-
-% %%%  At final submission, figure files MUST be
-% %%%  provided separately as high-resolution image
-% %%%  files. All of the panels for a figure should
-% %%%  be in the same file. Figures should have
-% %%%  clear labels/file names (Figure 1, Figure 2,
-% %%%  etc.).
-
-% %%%  Figure titles and legends should be placed
-% %%%  at the end of the main text. You do not
-% %%%  need to place the figures, nor their titles
-% %%%  and legends, within the main text. While
-% %%%  typesetting your article, our team will
-% %%%  place each figure in the best location
-% %%%  based on the final layout and on your
-% %%%  figure citations, e.g., (Figures 1A and 1B).
-
-% %%%  Please review the figure guidelines before
-% %%%  submitting your final materials:
-% %%%  https://www.cell.com/figureguidelines.
-
-% \noindent\includegraphics[width=0.85\linewidth]{Figure1.jpg}
-
-% \subsection*{Figure 1. A brief title that describes the entire figure without citing specific panels}
-
-% The figure legend can be all one paragraph and describe the images (A), graphs (B), and plots (C), etc., together.
-% \newline
-% (A) Or each panel or group of panels can be described separately, as shown here and below.
-% \newline
-% (B) Graph of X, Y, and Z.
-% \newline
-% (C and D) If panels are grouped like this, please explicitly describe each panel, e.g., ``Images showing SEM (C) and TEM (D).''
-% \newline
-% Please define all scale and error bars, and please review the Cell Press figure guidelines before submission: \href{https://www.cell.com/figureguidelines}{https://www.cell.com/figureguidelines}.
-Example figure created by Cassie Comeau, Cell Press.
-
-% \newpage
-
-% \section*{MAIN TABLES, INCLUDING TITLES AND LEGENDS}
-
-% %%%  Whenever possible, we encourage you to submit
-% %%%  your main-text tables as Microsoft Word documents,
-% %%%  using Word's table function. This will ensure
-% %%%  the best results during conversion. Tables
-% %%%  should be numbered Table 1, Table 2, etc. and
-% %%%  should not include subpanels (do not use Table 1A,
-% %%%  1B, etc.). Give each table a brief descriptive
-% %%%  title. Table legends are optional but encouraged.
-% %%%  Footnotes (superscript lowercase letters) should
-% %%%  be used where necessary to indicate some feature
-% %%%  of the data; please do not use bold, italic,
-% %%%  coloured text, or shading for this purpose. Use
-% %%%  separate cells, not line breaks or spaces, for
-% %%%  all discrete data elements. Small embedded
-% %%%  graphics with colour are OK.
-
-% %%%  Like figures, all tables must be cited within
-% %%%  the main text, and our typesetting team will
-% %%%  place the tables within the typeset paper at
-% %%%  the appropriate locations.
-
-% \subsection*{Table 1. A table with clear organization of data}
-
-% \begin{tabular}{|l | l | l | l|}
-%  \hline
-%  \textbf{Column 1} & \textbf{Column 2} & \textbf{Column 3} & \textbf{Column 4} \\ [1ex]
-%  \hline
-%  Row A\textsuperscript{a} & 6 & 87,837 & 787 \\ [1ex]
-%  \hline
-%  Row B & 7 & 78 & 5,415 \\ [1ex]
-%  \hline
-%  Row C & 545 & 778 & 7507 \\ [1ex]
-%  \hline
-%  Row D & 545 & 18,744 & 7,560 \\ [1ex]
-%  \hline
-%  Row E & 88 & 788 & 6,344 \\ [1ex]
-%  \hline
-% \end{tabular}
-
-% \bigskip
-
-% The table legend (optional) follows the table itself.
-% The legend should be used to provide additional info that relates to the table as a whole.
-
-% \textsuperscript{a}Footnotes can be used to provide additional info on specific content within the table,
-% such as this footnote to the first row (row A). Do not use footnotes in the table title.
-
-% \textsuperscript{b}More footnotes
-
-\newpage
-
-%%%  REFERENCES: As of 2023, all Cell Press journals
-%%%  use Numbered (AMA) style. We recommend placing
-%%%  your references in the included "references.bib"
-%%%  file.
-
-\bibliography{references}
+%%%%%%%%%%%%%%%% REFERENCES %%%%%%%%%%%%%%%
+
+\clearpage % Clear all remaining figures and tables then start a new page
+
+% The list of references goes after the main text and before the acknowledgements
+% When preparing an initial submission, we recommend you use BibTeX, like this:
+%
+\bibliography{references} % for a file named science_template.bib
+\bibliographystyle{sciencemag}
 
@@
 
-%%%  In your References, please include only articles
-%%%  that are published (online publication and
-%%%  preprint servers are OK). Unpublished data,
-%%%  submitted and/or accepted manuscripts, abstracts,
-%%%  and personal communications should be cited within
-%%%  the text only ("unpublished data," "data not
-%%%  shown," "Alice Smith, personal communication")
-%%%  and not included in the references list. Personal
-%%%  communication should be documented by a letter
-%%%  of permission. Whenever possible, please make
-%%%  sure your .bib file has the complete author lists
-%%%  for each item (at minimum, the first 11 authors
-%%%  listed).
-
-\newpage
-
-%%% All Cell Press \textbf{life and medical science} journals, and the multidisciplinary journal \textbf{iScience},
-%%% use the \textbf{\href{https://www.cell.com/star-authors-guide}{STAR Methods}} format for reporting materials and methods.
-
-%%% If you are publishing in a STAR Methods journal, please refer to the appropriate guide for authors:
-%%% Cell authors should download the guide https://www.cell.com/pb-assets/journals/research/cell/methods/Methods_Guide_Cell-1678470557763.pdf.
+%%%%%%%%%%%%%%%% ACKNOWLEDGEMENTS %%%%%%%%%%%%%%%
+
+\section*{Acknowledgments}
+Here you can thank helpful colleagues who did not meet the journal's authorship criteria, or
+provide other acknowledgements that don't fit the (compulsory) subheadings below.
+Formatting requirements for each of these sections differ between the \textit{Science}-family
+journals; consult the instructions to authors on the journal website for full details.
+\paragraph*{Funding:}
+List the grants, fellowships etc. that funded the research;
+use initials to specify which author(s) were supported by each source.
+Include grant numbers when appropriate or required by the funding agency.
+For example: F.~A. was funded by the Generous Science Agency grant~2372.
+\paragraph*{Author contributions:}
+List each author’s contributions to the paper.
+Use initials to abbreviate author names.
+\paragraph*{Competing interests:}
+Disclose any potential conflicts of interest for all authors, such as patent applications,
+additional affiliations, consultancies, financial relationships etc.
+See the journal editorial policies web page for types of competing interest that must be declared.
+If there are no competing interests, state:
+``There are no competing interests to declare.''
+\paragraph*{Data and materials availability:}
+Specify where the data, software, physical samples, simulation outputs or other materials
+underlying the paper are archived.
+They must be publicly accessible when the paper is published (without embargo) and enable
+readers to reproduce all the results in the paper.
+Contact the editor if you’re unsure what needs to be shared.
+
+Our preference is for digital material to be deposited in a suitable non-profit online data or
+software repository that issues the material with a DOI.
+Alternatively, an institutional repository, subject-based archive, commercial repository etc.
+is acceptable, as are (short) supplementary tables or a machine-readable supplementary data file.
+‘Available on request’ or personal web pages are not allowed.
+
+Cite the relevant DOI \cite{dataset}, URL \cite{example_url} or reference \cite{example2}
+in this statement.
+These \textit{do not} count towards the reference limit if they are only cited in the acknowledgements.
+Be specific and state a unique identifier -- such as an accession number, software version number
+or observation ID -- so readers can easily retrieve the exact material used.
+
+Declare any restrictions on sharing or re-use -- such as a Materials Transfer Agreement (MTA) or
+legal restrictions -- which must be approved by your editor.
+Unreasonable restrictions will preclude publication.
+Consult the journal's editorial policies web page for more details.
 
@@
 
-\FloatBarrier
+% \FloatBarrier
 

### 2025-12-20 latex/main.tex
Update abstract

@@
 \begin{document}
-
 \maketitle
 
-%%% The summary (abstract) should consist of a single paragraph of 150 words or
-%%% fewer.
-
-\section*{SUMMARY}
-
-% NOTE: currently 150 words
-
+% Abstract, in bold
+% There are strict length limits, and not all formats have abstracts.
+% Consult the journal instructions to authors for details.
+% Do not cite any references in the abstract.
+\begin{abstract}\bfseries
 Millions of SARS-CoV-2 genome sequences were collected 
@@
 lower bound on detectable recombination.
-
-\section*{KEYWORDS}
-
-%%%  Include up to 10 keywords, separated by commas.
-%%%  Keywords entered in EM are not carried over; only
-%%%  keywords included in the main text will be used
-%%%  in the final article metadata. Please note that
-%%%  for some journals, keywords are chosen by the editors.
-
-Ancestral recombination graphs, recombination, SARS-CoV-2, ARG, tskit
-
-\section*{INTRODUCTION}
+\end{abstract}
+
+% \section*{KEYWORDS}
+
+% Ancestral recombination graphs, recombination, SARS-CoV-2, ARG, tskit
+
 % SHZ: The first paragraph should explain
@@
 % can be treated.
+
+% The first paragraph of any Science paper does NOT have a heading
+% Nor is it indented
+\noindent
 Classical phylogenetics is largely concerned with inferring the evolutionary history 

### 2025-12-20 latex/main.tex
Partial updates to move Methods to supplement

@@
 
-\date{}
-
-% \author[1,2,8,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
-% \author[1,7,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
-% \author[3,7,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
-% \author[4,\orcidlink{0000-0001-6862-7756}]{Katherine~Eaton}
-% \author[1,\orcidlink{0000-0002-3547-7226}]{Isobel~Guthrie}
-% \author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
-% \author[1,3,5,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
-% \author[4,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
-% \author[6,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
-% \author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}
+% \author[1,2,8,*]{Shing~H.~Zhan}
+% \author[1,7]{Yan~Wong}
+% \author[3,7]{Anastasia~Ignatieva}
+% \author[4]{Katherine~Eaton}
+% \author[1]{Isobel~Guthrie}
+% \author[1]{Benjamin~Jeffery}
+% \author[1,3,5]{Duncan~S.~Palmer}
+% \author[4]{Carmen~Lia~Murall}
+% \author[6]{Sarah~P.~Otto}
+% \author[1,**]{Jerome~Kelleher}
 
@@
 % \affil[**]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}
-
 \begin{document}
+
 \maketitle
@@
 
-\section*{RESULTS}
-
-%%% The results may be divided with subheadings.
-%%% In our view, good subheadings convey information about the findings,
-%%% so we encourage you to be specific. For example, "Factor X requires factor Y to function in process Z"
-%%% is better than "Analysis of factors X and Y by approach Q."
-%%% We recommend that you use similar language in your figure titles for clarity and structural harmony.
-%%% Please do not use numbered subheadings. For some formats, ``results and discussion" may be combined under a single heading.
-
-%%% For initial submissions, figures and their titles and legends can be run-in with the text.
-%%% For final submissions, the image files must be uploaded as separate files, one complete file per figure.
-The figure titles and legends should be placed toward the end of the main text, after the declaration of interests.
-Our typesetters will place the figures and their titles and legends appropriately based on your in-text citations.
-
-%%% To cite/link to display items (figures and tables) and/or sections of the manuscript,
-%%% simply write, for example, ``(Figure 1)" or ``see discussion."
-%%% Our team will do the rest. Please note that we do not number sections or subsections.
+% Research Articles and Reviews split the text into sections using headings
 
@@
 
-\section*{STAR METHODS}
-
-%%%  The STAR Methods should appear in your main
-%%%  manuscript file after the figure legends, main
-%%%  table(s) and table legend(s).
-
-% \subsection*{Key resources table}
-
-% \textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}
-
-\subsection*{Method details}
-
-%%%  Please provide precise details of all the
-%%%  procedures in the paper (behavioral task,
-%%%  generation of reagents, biological assays,
-%%%  modeling, etc.) such that it is clear how,
-%%%  when, where, and why procedures were
-%%%  performed. We encourage authors to provide
-%%%  information related to the experimental
-%%%  design as suggested by NIH and ARRIVE
-%%%  guidelines (e.g., information about
-%%%  replicates, randomization, blinding, sample
-%%%  size estimation, and the criteria for
-%%%  inclusion and exclusion of any data or
-%%%  subjects).
+\begin{center}
+\section*{Supplementary Materials for\\ \scititle}
+
+% Author list for the supplement
+% Indicate the corresponding authors, but do NOT include institutions here
+% It would be nice if the template auto-generated this, but doing so is complicated...
+First~Author$^{\ast\dagger}$,
+A.~Scientist$^\dagger$,
+Someone~E.~Else\\ % we're not in a \author{} environment this time, so use \\ for a new line
+\small$^\ast$Corresponding author. Email: example@mail.com\\
+\small$^\dagger$These authors contributed equally to this work.
+\end{center}
+
+% Fill out the numbers for each type of supplementary material,
+% and delete any lines that aren't applicable.
+% These are just example numbers that don't match the rest of this template.
+\subsubsection*{This PDF file includes:}
+Materials and Methods\\
+Supplementary Text\\
+Figures S1 to S3\\
+Tables S1 to S4\\
+Captions for Movies S1 to S2\\
+Captions for Data S1 to S2
+
+\subsubsection*{Other Supplementary Materials for this manuscript:}
+Movies S1 to S2\\
+Data S1 to S2
+
+\newpage
+
+\subsection*{Materials and Methods}
 
